\LX{a} \HM{} \PH{à} \PS{conj.~} \SN \GE{and} \XV{O waɛ la a wa gaŋ.} \XE{S/he came and left.} \SN \GE{in order to} \XV{O waɛ la a na nyɛ te.} \XE{S/he has come in order to visit us.} 
\LX{a} \HM{} \PH{á} \PS{conj.~} \GE{but} \XV{A wola ka o ba wa?} \XE{But why didn't s/he come?.} \VA{ana}
\LX{a} \HM{} \PH{à} \PS{pron  (pl).~} \GE{they, them (nonhuman)} \XV{Vɛŋ ka a kpɛ a zage.} \XE{Let them go into the pen.} \XV{De a moɔ eŋ a.} \XE{Give them the grass.} 
\LX{a} \HM{} \PH{à} \PS{def article.~} \GE{the} \XV{A bie kpɛ la a die.} \XE{The child has gone into the room.} 
\LX{a} \HM{} \PH{á} \PS{v.~} \GE{to hate} \XV{Te a ba la.} \XE{We hate them.} \PD{aɛ, aara, aara, aareba, aaraa, aareba } 
\LX{aa} \HM{} \PH{áà} \PS{interj;.~} \GE{expression of surprise and regret} \XV{ A ŋmane ŋmarɛɛ la, aa.} \XE{The calabash is broken; what a pity!.} \XV{ Aa, ka N da baŋ ka lɛ N gaaɛ.} \XE{Oh, if I had known, I would have gone.} 
\LX{aa} \HM{} \PH{áá} \PS{interj.~} \GE{expression of regret} 
\LX{aa} \HM{} \PH{àà} \PS{interj.~} \GE{expression expressing "I told you so"} 
\LX{a-a} \HM{} \PH{á-à} \PS{interj.~} \SN \GE{expression of disapproval} \XV{A-a, ta de ona.} \XE{Oh no, don't go that way.} \SN \GE{ why; expression of doubt} \XV{A-a, fo naŋ ba gaa?} \XE{Well, you haven't gone yet?.} 
\LX{aae} \HM{} \PH{ááɪ́} \PS{interj.~} \GE{expression of pain} 
\LX{aa-ee} \HM{} \PH{áá-ɪ̀ɪ́} \PS{interj.~} \GE{expression of lack of confidence in someone's ability to perform a task} \XV{Aa-ee, fo koŋ daŋ ma ta.} \XE{I don't think you can get there before me.} 
\LX{aa-hii} \HM{} \PH{áá-hì᷉ì} \PS{interj.~} \GE{expression of total despair} 
\LX{aaŋ} \HM{} \PH{aaŋ} \PS{interj.~} \GE{well then} 
\LX{aane} \HM{} \PH{áánɪ̀} \PS{adj.~} \GE{infertile (woman )} \OP{aama} \TP{aanɛɛ} 
\LX{aba} \HM{} \PH{ábà} \PS{interj.~} \SN \GE{expression of despair over a repeated mishap} \XV{A bie la lee la, aba.} \XE{The child has fallen again, oh no!.} \SN \GE{expression of satisfaction or agreement} \XV{Aba, lɛ ka N boɔrɔ.} \XE{Good, that is what I want.} \SN \GE{expression of empathy} \XV{Aba, mɔ kyɛ so a bie ko᷉ɔ} \XE{Try and bathe the child.} 
\LX{abaa} \HM{} \PH{ábàà} \PS{interj.~} \SN \GE{exclamation of satisfaction or commendation} \XV{Abaa, fo waa bie.} \XE{Well done, you are a good child.} \SN \GE{exclamation of despair or being distraught} \XV{Abaa, vɛŋ a zɔɔre lɛ.} \XE{Oh, stop the fight.} 
\LX{abɛ} \HM{} \PH{àbɛ́} \PS{n.~} \SN \GE{palm tree} \SN \GE{palm fruit} 
\LX{abɛka᷉a} \HM{} \PH{àbɛ́ká᷉à} \PS{n.~} \GE{palm oil} \TP{abɛkaanɛɛ} 
\LX{abuobo} \HM{} \PH{ábúóbò} \PS{adv.~} \GE{which ones (nonhuman)} \XV{Abuobo ka fo boɔrɔ?} \XE{Which ones do you want?.} 
\LX{abɛbiri} \HM{} \PH{àbɛ̀bírì} \PS{n.~} \GE{palm fruit} \SG{abɛbiruu} \OP{abɛbie} \TP{abɛbiree} 
\LX{abɛda᷉a᷉} \HM{} \PH{àbɛ́dá᷉à᷉} \PS{n.~} \GE{palm wine} \SG{abɛdãão᷉} \TP{abɛdaanɛɛ} 
\LX{abɛsaare} \HM{} \PH{ábɛ́sáárɪ̀} \PS{n.~} \GE{palm-frond broom} \OP{abɛsaɛ} \TP{abɛsaarɛɛ} 
\LX{abɛteɛ} \HM{} \PH{àbɛ̀tɪ̀ɛ̀} \PS{n.~} \GE{palm tree} \OP{abɛteere} \TP{abɛteerɛɛ} 
\LX{Afereka} \HM{} \PH{áfɪ́rɪ́kà} \PS{n.~} \GE{Africa} \OP{Aferekare} \TP{Aferekarɛɛ} 
\LX{Aferekaneɛ} \HM{} \PH{áfíríkànɪ́ɛ́} \PS{n.~} \GE{African} \OP{Aferekanoba} 
\LX{afu} \HM{} \PH{àfú} \PS{n.~} \GE{preface (of a book)} \OP{afuri} \TP{afuree} 
\LX{age} \HM{} \PH{ágɪ́} \PS{v.~} \SN \GE{to dance} \SN \GE{to jump up and down} \PD{agɛɛ, agra, agra, agreba, agraa, agreba } \SE{} \GE{} 
\LX{age} \HM{} \PH{ágɪ̀} \PS{v.~} \SN \GE{to fly} \SN \GE{errect, ejaculate (penis)} \SN \GE{to jump over} \SN \GE{to be frightened; to miss a heartbeat} \PD{agɛɛ, agra, agra, agreba, agraa, agreba } \SE{pɔlaa age} \GE{to be frightened} 
\LX{agoo} \HM{} \PH{àgòò} \PS{v.~} \GE{to ask for permission to join someone (knock door)} \PD{-, -, -, -, -, - } 
\LX{agoo} \HM{} \PH{àgòò} \PS{n.~} \GE{a type of silky, colourful fabric} \OP{agoori} \TP{agooree} 
\LX{agye} \HM{} \PH{ádzé} \PS{n.~} \GE{bad luck} \OP{agyeri} \TP{agyeree} 
\LX{ahaŋ} \HM{} \PH{àháŋ} \PS{interj.~} \GE{an expression that denotes "well?"} \XV{Ahaŋ a polisi la waana.} \XE{All is well, as the police are coming.} 
\LX{ahâa} \HM{} \PH{àhá᷉á᷉} \PS{interj.~} \GE{that's right (expression of agreement)} \XV{Ahaa, o binzie la be.} \XE{That's right, that is its position.} 
\LX{ai} \HM{} \PH{àí} \PS{use word.~} \GE{no} 
\LX{aii} \HM{} \PH{áìì} \PS{interj.~} \GE{an expression of despair} \XV{Aii, a nɔgyɛle ŋmarɛɛ.} \XE{Oh no; the egg is broken!.} 
\LX{akaase} \HM{} \PH{àkáásɪ̀} \PS{interj.~} \GE{expression of boastful challenge} \XV{Akaase, neɛ kyebe na daŋ ma kɔnoɔre.} \XE{I dare anyone to finish his portion of the weeding before me.} 
\LX{akaasemanakei} \HM{} \PH{ákáásɪ̀mànà} \PS{interj.~} \GE{expression of disappointment at failure} \XV{A bie ba toɔ tuo daga, akaasemanakei.} \XE{The child has failed to carry the box, what a pity.} \VA{akaase}
\LX{akai} \HM{} \PH{àkàí} \PS{interj.~} \GE{expression of disgust and disagreement} \XV{ Akai, fo deme kyebe.} \XE{Too bad, you have no case.} 
\LX{aku} \HM{} \PH{ákú} \PS{interj.~} \GE{expression of despair, loss, empathy} \XV{A nannyigri la maaleŋ fa la n zɔmeŋɛ na, aku.} \XE{The thieves have robbed my friend again, how awful!.} \VA{akuroku}
\LX{akuroku} \HM{} \PH{àkúrókú} \PS{interj.~} \GE{emphatic form of aku} \VA{akuraku; aku}
\LX{akuu} \HM{} \PH{àkúú} \PS{interj.~} \GE{expression indicating one's inability to do a task} \XV{Akuu, ta ko ma, N koŋ toɔ̂ a tuori tuo.} \XE{Ei, don't kill me; I can't carry the load.} 
\LX{Ala} \HM{} \PH{álà} \PS{n.~} \GE{God} 
\LX{ala} \HM{} \PH{álà} \PS{interj.~} \GE{I swear} \XV{Ala, a dɔɔ ŋa naane so a woɔ.} \XE{I swear, the bag does not belong to this man.} 
\LX{alaakosebaroo} \HM{} \PH{àláákosɪ́bárʋ́ʊ́} \PS{interj.~} \GE{in the name of God} 
\LX{ale} \HM{} \PH{álɪ̀} \PS{v.~} \GE{to open (ring), to stretch out} \XV{Ka fooŋ ba ale a fenetere, o koŋ kpɛ a fo nubiri.} \XE{If you don't open up the ring you can't put it on your finger.} \PD{alɛɛ, alla, alla, aleba, allaa, - } 
\LX{alee} \HM{} \PH{àléé} \PS{v.~} \GE{expression ordering one to get out; away with you} \XV{A bibiiri naŋ di a bondirii baare la ka ba ba yeli ka alee, ka ba wa yi gaa sakuuri.} \XE{When the children had finished eating, their father ordered them to go to school at once.} \PD{-, -, -, -, -, - } 
\LX{aleɛma} \HM{} \PH{àlɪ́ɛ́mà} \PS{n.~} \GE{metaphor (literature)} 
\LX{aloo} \HM{} \PH{álòó} \PS{interj.~} \GE{hello} 
\LX{aloopelee} \HM{} \PH{álóópèléè} \PS{n.~} \GE{aeroplane} \OP{aloopelli} \TP{aloopellee} \VA{ayɔɔmapelee [áyɔÕÅ:mápeÕÄleÕÄ:]}
\LX{aloopeleeduoraa} \HM{} \PH{álóópèléédúóráà} \PS{n.~} \GE{airport} \OP{aloopeleeduori} 
\LX{alɛ} \HM{} \PH{álɛ̀} \PS{interj.~} \GE{expression of surprise mixed with fear} \XV{Alɛ, fentel diglii ka fo deɛnɛ ne?} \XE{Whoa, you're playing with electricity?.} 
\LX{ama} \HM{} \PH{àmá} \PS{conj.~} \GE{but, however} \XV{ A bɔlŋmeɛrebɛ mɔɛ la ama ba ba di.} \XE{The football players played well, but they did not win.} 
\LX{ama} \HM{} \PH{ámà} \PS{demonstrative pronoun.~} \GE{these (nonhuman)} \XV{A dogere ama daaroŋ tuo.} \XE{These pots are expensive.} 
\LX{ambaŋenaa} \HM{} \PH{ámbáŋɪ́náá} \PS{n.~} \GE{a type of small tree with large leaves and pink fruit} \OP{ambaŋene} \TP{ambaŋenɛɛ} 
\LX{Ameleka} \HM{} \PH{ámɪ́lɪ́kà} \PS{n.~} \GE{ America} \OP{amelekare} \TP{amelekarɛɛ} \VA{Hameleka}
\LX{aminekaŋa} \HM{} \PH{àmínékàŋá} \PS{conj.~} \SN \GE{perhaps, maybe} \XV{Aminekaŋa o koŋ wa pampana.} \XE{Perhaps s/he will not come now.} \SN \GE{sometimes} \VA{minekaŋa}
\LX{amma} \HM{} \PH{ámmá} \PS{function word.~} \GE{much more than (exaggeration)} \XV{A bie zoro to, amma soɔŋaa.} \XE{The boy runs really fast, much faster than the hare.} 
\LX{ammeseɛrɛ} \HM{} \PH{ámmɪ̀sɪ́ɛ́rɛ̀} \PS{n.~} \GE{thanks} \XV{Fo taa la ammeseɛrɛ ne fo naŋ toŋ a toma.} \XE{I am thankful to you for doing the work.} \TP{ammeseɛrɛɛ} 
\LX{ammine} \HM{} \PH{ámmìné} \PS{rel. interg. pron;pl.~} \GE{which people} 
\LX{ampe} \HM{} \PH{ámpé} \PS{n.~} \GE{a game traditionally played by women} \OP{amperi} \TP{amperee} 
\LX{aŋ} \HM{} \PH{ǎŋ} \PS{aux.v.~} \GE{to be alerted; used to caution or persuade} \XV{Ka fo wa gɛrɛ a libidaana na yiri, aŋ baŋ ne baare.} \XE{If you are going to that rich man’s house you'd better beware of the dogs.} 
\LX{aŋ} \HM{} \PH{áŋ} \PS{interrogative pronoun.~} \GE{who} \XV{Aŋ la ba nɔŋ bompaala?} \XE{Who does not like new things?.} 
\LX{ana} \HM{} \PH{áná} \PS{third pers. pron. pl.~} \GE{those (nonhuman)} \XV{Ana ka te na wuo.} \XE{We'll collect those.} 
\LX{aŋa} \HM{} \PH{̪áŋá} \PS{n.~} \GE{no one (retort; implies the matter is none of the asker's business)} \XV{Aŋ la wa a kyɛ? Aŋa la wa.} \XE{Who came here? No one. (Why do you care?).} 
\LX{aŋa} \HM{} \PH{àŋá} \PS{adv.~} \GE{like} \XV{O kono la aŋa bibile na.} \XE{S/he cries like a baby.} \VA{ŋa}
\LX{anaare} \HM{} \PH{ànáárɪ̀} \PS{adj.~} \GE{four} \OP{anaare} \TP{anaarɛɛ} \VA{naare}
\LX{anaŋ} \HM{} \PH{ánǎŋ} \PS{pron.pl.~} \GE{those ones} \XV{Te na iri la anaŋ naŋ sa᷉a te wuli a boŋkoɔrɔ.} \XE{We'll select the ones that are bad and show them to the seller.} 
\LX{anaŋ} \HM{} \PH{ánáŋ} \PS{pron.~} \GE{it (a problem, a situation)} \XV{Ka anaŋ waa lɛ, ta di a seemaa.} \XE{If that is the case, don't eat the food.} 
\LX{ananso} \HM{} \PH{ánánsʋ́} \PS{conj.~} \GE{because} \XV{O naŋ taa libie ananso ka o ba mɛ yiri.} \XE{S/he built a house because s/he has money.} \VA{shortened form of ana la so}
\LX{ananso} \HM{} \PH{ánánsʊ́} \PS{n.~} \GE{reason} \XV{A fo toma zagere gaabo ananso la wola?} \XE{What is your reason for refusing to go to work?.} 
\LX{andonɛɛ} \HM{} \PH{ándʋ́nɛ́ɛ́} \PS{n.~} \SN \GE{world} \XV{Andonɛɛ waa la yɛloŋ.} \XE{The world is large.} \SN \GE{people of the world} \XV{Andonɛɛ ba baŋ ka a wono ba mine.} \XE{People of the world do not kow that some people are suffering.} 
\LX{ane} \HM{} \PH{ànɪ̀} \PS{conj.~} \SN \GE{and} \XV{Dɛre ane Daŋyi maŋ daŋ la sakuuri gaabo.} \XE{Dɛre and Daŋyi always go to school early.} \SN \GE{although, despite} \XV{Ane ka zie e la kpeɛŋaa kyɛ noba mine naŋ ba sage dɔge seŋ lɛ puori tuubu.} \XE{Although times are hard, some people do not want to practise family planning.} \SN \GE{with} \XV{O pɔge daa la a dɔɔ ane o faŋa zaa ka o te le.} \XE{The woman pushed the man with all her strength and he fell over.} 
\LX{aneazaa} \HM{} \PH{ánɪ́ázàá} \PS{conj.~} \GE{despite, even though, in spite of} \XV{Aneazaa ka saa ba miire kyɛ kôɔ᷉ naŋ be la a baa poɔ.} \XE{Even though it has not been raining, there is still water in the pond.} 
\LX{anesɛ} \HM{} \PH{ánɪ́sɛ̀} \PS{v.~} \GE{welcome; (an expression of goodwill to a stranger who has just arrived)} \XV{Anesɛ, kpeɛrɛ waana.} \XE{Welcome, come in!.} \PD{-, -, -, -, -, - } 
\LX{aneɛ} \HM{} \PH{ánɪ́ɛ̀} \PS{n.~} \GE{effort, attempt} \XV{A tontonema da erɛ la aneɛ na baare a toma kyɛ ka saa ba sage.} \XE{The workers tried to finish the job but the rain would not allow them.} 
\LX{aŋgaara} \HM{} \PH{áŋgáárá} \PS{adj.~} \GE{having the blades burnt away, leaving the stalks (used for grass)} \XV{A kyeɛ zo kpɛ la a mɔ-aŋgaara poɔ.} \XE{The squirrel has escaped into the stalky bush.} \OP{aŋgaare} \TP{aŋgaarɛɛ} 
\LX{aŋgeli} \HM{} \PH{áŋgélí} \PS{n.~} \GE{a type of long grass used to weave sleeping mats} \OP{aŋgelli} \TP{aŋgelee} 
\LX{aŋgoɔle} \HM{} \PH{áŋgʋ́ɔ̀lɪ̀} \PS{v.~} \GE{to adorn, to decorate} \XV{Ba aŋgoɔle la a naa wiri.} \XE{They have decorated the chief's horse.} \PD{aŋgoɔlɛɛ, aŋgoɔlɔ, aŋgoɔlɔ, aŋgoɔleba, aŋgoɔlaa, - } 
\LX{aŋguoŋaa} \HM{} \PH{áŋgúóŋáà} \PS{n.~} \GE{a type of tree with round, pink fruit} \OP{aŋguoni/aŋguomo} \TP{aŋguonee} \VA{pl.var. aŋguôaa}
\LX{angyoɔso} \HM{} \PH{ándzʋ̀ɔ̀sʋ́} \PS{n.~} \GE{gratitude, thanks} \XV{Fo taa la angyoɔso ne a kyɔɔtaa fo naŋ ko ma.} \XE{Accept my thanks for the gift.} \OP{angyoɔsere} \TP{angyoɔserɛɛ} 
\LX{anii} \HM{} \PH{àníì} \PS{adj.~} \GE{eight} \OP{aniini} \TP{aniinee} 
\LX{aŋkaa} \HM{} \PH{àŋkàá} \PS{n.~} \GE{orange (fruit)} \SG{aŋkaaroo} \OP{aŋkaare/aŋkaɛ} \TP{aŋkarɛɛ} 
\LX{Aŋkara} \HM{} \PH{àŋkàrà} \PS{n.~} \GE{Accra} \OP{Aŋkarre} \TP{Aŋkarrɛɛ} 
\LX{aŋko} \HM{} \PH{áŋkò} \PS{n.~} \GE{river blindness} \OP{aŋkori} \TP{aŋkoree} 
\LX{aŋkorɔ} \HM{} \PH{áŋkʋ̀rɔ́} \PS{n.~} \GE{barrel, drum} \OP{aŋkorre} \TP{aŋkorɛɛ/aŋkorrɛɛ} \VA{aŋkora}
\LX{anlee} \HM{} \PH{ánléé} \PS{n.~} \SN \GE{a traditional women's dance} \SN \GE{music for such a dance} \OP{anleeri} \TP{anlerri/anleree} \VA{an-yee [án-yé:]}
\LX{annoo} \HM{} \PH{ánnʋ́ʊ́} \PS{interrogative pronoun.~} \GE{who is s/he?; which person?} \VA{shortened form of aŋ la o}
\LX{ansaŋ} \HM{} \PH{ánsáŋ} \PS{conj.~} \GE{before} \XV{Vɛŋ ka saa mi ansaŋ ka fo bore a kamaana.} \XE{Let it rain before you sow the maize.} 
\LX{ansiŋinee} \HM{} \PH{ánsíŋínéé} \PS{n.~} \SN \GE{a type of edible berry that turns violet when ripe} \SN \GE{the tree of this fruit} \SG{ansiŋinuu} \OP{ansiŋini} \VA{ansiinee}
\LX{ansoɔnee} \HM{} \PH{ánsʋ́ɔ́nɪ́́ɪ́} \PS{n.~} \GE{a woman who has recently given birth} \OP{ansoɔne/ansoɔmɔ} \TP{ansoɔnɛɛ} 
\LX{antere} \HM{} \PH{ántèré} \PS{v.~} \GE{good afternoon (Southern Dagaare dialect)} \PD{-, -, -, -, -, - } 
\LX{anuu} \HM{} \PH{ànúú} \PS{adj.~} \GE{five} \OP{anuuri} \TP{anuuree} 
\LX{anwoŋ} \HM{} \PH{ánwóŋ} \PS{n.~} \GE{one who is great or strong (physically or spiritually)} \OP{annwonni} \TP{anwonnee} 
\LX{apa} \HM{} \PH{ápà} \PS{interj.~} \GE{expression of unpleasant surprise} \XV{A soɔŋaa muri la a baa bare kyɛ zo, apa.} \XE{The rabbit dodged the dog and escaped, what a pity.} \VA{aparapa}
\LX{aparapa} \HM{} \PH{ápárápà} \PS{interj.~} \GE{emphatic expression of despair} \XV{Aparapa, a soɔŋaa ŋa meŋ la zoe?} \XE{Oh no, this hare too has escaped?.} \VA{apa}
\LX{ara} \HM{} \PH{àra} \PS{v.~} \GE{to gather up one's courage} \XV{A pɔge ara kyɛ daa la a dɔɔlee eŋ tambɔgeŋ.} \XE{The woman gathered up her courage and pushed the man into a ditch.} \PD{-, -, -, -, -, - } 
\LX{arambiri} \HM{} \PH{árámbírí} \PS{n.~} \GE{kidney} \XV{A dɔɔta yeli ka a baala arambiri kaŋa sa᷈a᷈ɛ la.} \XE{The doctor said one of the patient's kidneys had failed.} \SG{arambiruu} \OP{arembie} \TP{arambiiree} \VA{arambiri [árámbírí]; pl arambie [árambíè]}
\LX{are} \HM{} \PH{árɪ́} \PS{v. pl.~} \GE{to cut, harvest (maize)} 
\LX{are} \HM{} \PH{árɪ́} \PS{v.~} \GE{to be related through the mother's line} \XV{N are la Kusielee.} \XE{I am maternally related to the Kusielee clan.} \PD{arɛɛ, ara, ara, areba, araa, - } 
\LX{are} \HM{} \PH{àrɪ̀} \PS{v.~} \SN \GE{to stand} \SN \GE{to wait} \XV{N baŋ ka o na are la ka N gaa te pɔge o.} \XE{I know s/he will wait so that I go to meet him/her.} \SN \GE{to stop} \XV{A wɔɔkye kyɛŋɛɛ la tasoga tasoga leɛ are.} \XE{The watch worked for a while and then stopped.} \PD{arɛɛ, ara, ara, areba, araa, - } 
\LX{are} \HM{} \PH{árɪ̀} \PS{v.~} \GE{to tear off from the main body (branch or limb of an animal)} \XV{A maŋgo wɔŋ zue la ka a ulee kaŋa are le.} \XE{The mango tree has borne so many fruits that one of the branches tore off.} \PD{arɛɛ, ara, ara, areba, araa, areba } 
\LX{areba} \HM{} \PH{árɪ́bá} \PS{n.~} \GE{maternal uncle} \OP{aremine} \TP{arrɛɛ} 
\LX{arekpoŋ} \HM{} \PH{árɪ́kpóŋ} \PS{n.~} \GE{eldest niece} \OP{arekponni} \TP{arekponnee} 
\LX{arelee} \HM{} \PH{árɪ́lèé} \PS{n.~} \GE{nephew or niece} \OP{arebilii} \TP{arebilee/arebillee} 
\LX{aremateke} \HM{} \PH{àrɪ́mátɪ́kɪ̀} \PS{n.~} \GE{arithmetic} \OP{arematekere} \VA{kuntaŋ (Akan)}
\LX{arezie} \HM{} \PH{[àrɪ̀zíé]} \PS{n.~} \SN \GE{standing place, spot} \SN \GE{post, position, rank} \XV{O arezie kpoŋ.} \XE{S/he holds a high position.} \SN \GE{position, stand (for or against an idea)} \XV{O ba wuli o arezie a nɔkpeɛne poɔ.} \XE{She has not indicated her position with regard to the argument..} \OP{areziiri} \TP{areziiree} 
\LX{aroozie} \HM{} \PH{árʋ́ʊ́zíé} \PS{n.~} \SN \GE{standing place; position} \SN \GE{adverb of place (on, under)} \OP{arooziiri} \TP{arooziiree} 
\LX{aseŋ} \HM{} \PH{àsɪ́ŋ} \PS{adv.~} \GE{for instance, like} 
\LX{asibiti} \HM{} \PH{ásíbítì} \PS{n.~} \GE{hospital, clinic} \OP{asibitiri} \TP{asibitiree} \VA{asiiti, [ásí:tì], asimeti, [ásímétì], asebere.}
\LX{ata} \HM{} \PH{átà} \PS{adj.~} \GE{three} \OP{atare} \TP{atarɛɛ} \SE{ata ata} \GE{in threes} 
\LX{awae} \HM{} \PH{àwáɪ̀} \PS{adj.~} \GE{nine} \OP{awarre} \TP{awarɛɛ/awarrɛɛ} \VA{wae}\SE{awae awae} \GE{nine each} 
\LX{awola} \HM{} \PH{àwʊ́là} \PS{adv.~} \SN \GE{how much (price)} \XV{Awola ka fo da a kparoo?} \XE{How much did the shirt cost you?.} \SN \GE{how many (nonhuman)} \XV{Teere awola ka yɛ sɛle?} \XE{How many trees did you plant?.} 
\LX{awommu} \HM{} \PH{àwómmú} \PS{n.~} \GE{hardship, suffering} \XV{Ka kɔŋeŋ le awommu maŋ pɔge la noba yaga.} \XE{When there is a famine, many people experience a lot of suffering.} \TP{awonnee} \VA{awonaa [awona:]}
\LX{awulo} \HM{} \PH{áwúló} \PS{n.~} \GE{continuous, loud, sharp cry of pain} \XV{A pɔge na bie naŋ kpi la lɔɔrɔ awulo.} \XE{The woman who lost her son is crying hysterically.} \TP{awulee} 
\LX{ayi} \HM{} \PH{áyì} \PS{adj.~} \GE{two} \OP{ayiri} \TP{ayiree} \SE{ayi ayi} \GE{in twos} 
\LX{ayililili} \HM{} \PH{ayililili} \PS{interjection.~} \GE{joyous, shrill shouting in commendation (usually by women)} \XV{A pɔgeba ŋmɛ la kulinni, ayililili, eŋ a bawaseɛrebɛ.} \XE{The women made shrill noises of commendation for the dancers.} \VA{ayiyiyiyi}
\LX{Ayɔɔ} \HM{} \PH{áyɔ̀ɔ̀} \PS{n.~} \GE{name traditionallly given to a female child born on a market day} \OP{Ayɔɔmine} \TP{Ayɔɔrɛɛ} 
\LX{ayoɔbo} \HM{} \PH{àyʊ̀ɔ̀b} \PS{adj.~} \GE{six} \OP{ayoɔre} \TP{ayoɔrɛɛ} 
\LX{ayɔpo᷉i᷉} \HM{} \PH{àyʊ̀ɔ̀pó᷉ì᷉} \PS{adj.~} \GE{seven} \OP{ayoɔponni} \TP{ayoɔponnee} 
\LX{Ayoree} \HM{} \PH{áyóréé} \PS{n.~} \GE{name of a clan} \OP{Ayoruu} 
\LX{Ayuo} \HM{} \PH{áyúó} \PS{n.~} \GE{name traditionally given to a female child born on a funeral day} \OP{Ayumine} \TP{Ayuuree} 
\LX{Azaasoba} \HM{} \PH{àzàásʊ́bá} \PS{n.~} \GE{almighty} 
\LX{âa} \HM{} \PH{à᷉á᷉} \PS{adv.~} \SN \GE{yes (answer to a call)} \XV{N boɔle la a bie ka o sage ka "a᷉a."} \XE{I called the child and he answered, "Yes.".} \SN \GE{what? (exclamation of disbelief)} \XV{Ka Naa boɔle fo ka fo zagere, a᷉a.} \XE{You refuse to answere the chief's call, what?.} 
\LX{âaa} \HM{} \PH{à᷉á᷉́á᷉́} \PS{part.~} \GE{answer in agreement} \XV{Fo na gaa la? âaa.} \XE{Will you go? Yes, I will.} 
\LX{a᷉̀aa} \HM{} \PH{à᷉á᷉à᷉} \PS{interj.~} \GE{is that so? (expression of uncertain surprise)} \XV{Bayuoː A pɔgebil-pɛɛre na taa la poɔ. Ayuoː A᷉aa?} \XE{Bayuo: That teenage girl is pregnant. Ayuo: Is that true?.} 
\LX{âo} \HM{} \PH{a᷉ʋ᷉̀} \PS{interj.~} \GE{of course (expression of certainty)} \XV{A. Fo na kuli la be? B. âo.} \XE{A.Will you go home? B. Of course I willl.} 
\LX{ba} \HM{} \PH{bá} \PS{adv.~} \GE{not (negative particle)} \XV{N ba nyɛ o zenɛ.} \XE{I have not seen him today.} 
\LX{ba} \HM{} \PH{bà} \PS{pron.~} \SN \GE{they (human)} \XV{Ba na wa la.} \XE{They will come.} \SN \GE{their (human)} \XV{Te de la ba bondirii biŋ.} \XE{We have reserved their food.} \SN \GE{them (human)} \XV{Te de la a bondirii ko ba.} \XE{We have given them the food.} 
\LX{ba} \HM{} \PH{bà} \PS{v.~} \GE{to fix a pointed object firmly in the ground} \XV{Ba ba la daare gɔlle a dabɔgɔ zaa.} \XE{They have put up planks around the whole garden.} \PD{baɛ, baara, baara, bareba, baaraa, - } 
\LX{ba} \HM{} \PH{bà} \PS{v.~} \GE{to take a long drink from a container} \XV{A saana de la a ko᷉ɔ ba lɛɛ ka a ŋmane wa te kyɛre.} \XE{The stranger drank deeply until the calabash was empty.} \PD{baɛ, baara, baara, bareba, baaraa, - } 
\LX{ba} \HM{} \PH{bá} \PS{v.~} \GE{to gallop} \PD{baɛ, baara, baara, bareba, baaraa, - } 
\LX{ba} \HM{} \PH{bâ} \PS{n.~} \SN \GE{father} \SN \GE{friend (between males)} \SN \GE{Mr} \OP{bamine} \TP{barɛɛ} 
\LX{baa} \HM{} \PH{báà} \PS{conj.~} \GE{or; (normally used in question tag)} \XV{Te na gaa la, baa wolɔ?} \XE{We'll go, won't we?.} \VA{bee}
\LX{baa} \HM{} \PH{báá} \PS{n.~} \SN \GE{dog} \SN \GE{a small black ground ant-like creature which burrows into the ground with the tip of its abdomen} \SN \GE{a type of children's game} \OP{baare} \TP{baarɛɛ} 
\LX{baa} \HM{} \PH{bàá} \PS{n.~} \GE{body of water} \OP{bare} \TP{barɛɛ} 
\LX{baa} \HM{} \PH{bàà} \PS{v.~} \SN \GE{to grow up, to flourish} \XV{Sêɛ saŋa, moɔ maŋ baa la wogri lɛ.} \XE{During the rainy season, grass normally grows tall.} \SN \GE{to mature} \XV{A pɔgelee naŋ ba baa azuiŋ o ba ta serɛ kuluu.} \XE{The girl has not yet matured enough, so she is not marriageable.} \SN \GE{to survive, to live} \XV{A baala naŋ ba kpi zenɛ o na baa la.} \XE{Since the patient has not died today, s/he will survive.} \PD{-, -, -, -, -, - } 
\LX{baabaa} \HM{} \PH{bààbáá} \PS{n.~} \GE{father (familiar term of address used by toddlers)} \OP{baabaamine} \TP{baabaarɛɛ} 
\LX{Baabol} \HM{} \PH{báábʋ̀l} \PS{n.~} \GE{Bible} \OP{baabola} \TP{baabolɛɛ/baabollɛɛ} 
\LX{baaebaae} \HM{} \PH{bááɪ́bááɪ́} \PS{interj.~} \GE{goodbye} 
\LX{baage} \HM{} \PH{báágɪ̀} \PS{n.~} \GE{bag} \OP{baagere} \TP{bagerɛɛ} \VA{baagye}
\LX{baala} \HM{} \PH{bààlá} \PS{n.~} \GE{patient, sick person} \OP{baalba} \TP{baalbɛɛ} 
\LX{baalaa} \HM{} \PH{bààlàà} \PS{adj.~} \SN \GE{slim, slender, thin} \XV{N da nyɛ la pɔgebil baalaa kaŋa a seɛre zie.} \XE{I saw a slender girl at the dance.} \SN \GE{narrow} \XV{A sori waa la baalaa lɛ.} \XE{The road is very narrow.} \OP{baale} \TP{baalɛɛ} 
\LX{baale} \HM{} \PH{báálɪ̀} \PS{v.~} \GE{to ache slightly (disconfort of the stomach)} \XV{Ka maaŋ iri ba gaa bangyeraa n poɔ maŋ baala la.} \XE{If I don’t go to toilet for the day I normally have a slight stomach ache.} \PD{baalɛɛ, baala, baala, baaleba, baalaa, - } 
\LX{baale} \HM{} \PH{báálɪ̀} \PS{v.~} \GE{to separate (grains from unwanted material)} \XV{A pɔge baale iri la a seŋkãabie yi a pɛgere poɔ.} \XE{The woman separated the groundnuts from the shells.} \PD{baalɛɛ, baala, baala, baaleba, baalaa, - } 
\LX{baalebayiri} \HM{} \PH{bààlbáyírì} \PS{n.~} \GE{hospital} \OP{baalebayie} \TP{baalebayiree} 
\LX{baalimbo} \HM{} \PH{bààlímbò} \PS{n.~} \GE{the act of supporting someone by holding them by the midsection} \XV{Ba kpɛ la a baala baalimbo gaa ne a asibiti.} \XE{They supported the patient by the midsection and took him to the hospital.} \OP{baalimbori} \TP{baalimboree} \VA{boolimbo}
\LX{baalombiri} \HM{} \PH{bààlʋ́mbírì} \PS{n.~} \GE{germ, bacterium} \OP{baalombie} \TP{baalombiiree} 
\LX{baalomboɔlaa} \HM{} \PH{báálʋ̀mbʋ̀ɔ̀làá} \PS{n.~} \GE{feast organised for a friend} \XV{Te daare gaa la baalomboɔlaa Loraa poɔ.} \XE{We attended a friendship feast in Lawra.} \XV{baalomboɔle} \XE{.} 
\LX{baaloŋ} \HM{} \PH{báálʋ́ŋ} \PS{n.~} \GE{friendship} \XV{Ba nyɔge la baaloŋ saŋa zaa.} \XE{They have been friends for a long time.} \OP{baalondeme} 
\LX{baaloŋ} \HM{} \PH{bààlʊ́ŋ} \PS{n.~} \GE{illness, sickness} \XV{A baala baaloŋ seɛ o la zenɛ.} \XE{The patient's illness has improved today.} 
\LX{baaloŋ} \HM{} \PH{bààlʊ̀ŋ} \PS{adj.~} \SN \GE{calm} \XV{Zenɛ, a mane ko᷉ɔ᷉ e la baaloŋ.} \XE{Today, the sea is calm.} \SN \GE{gentle, gently} \XV{A doɔ baaloŋ na wa la kyɛ.} \XE{That gentle man came here.} \XV{A doɔ wa la baaloŋ.} \XE{The man is gentle.} \XV{De a bileɛ baaloŋ.} \XE{Take the baby gently.} \SN \GE{soft (objects, sounds)} \XV{A kapuro ŋa e la baaloŋ yelmeŋɛ.} \XE{This pillow is really soft.} \SN \GE{smooth (powdery substances)} \XV{Bɛnzɔŋ maŋ waa la baaloŋ.} \XE{Bean flour is usually smooth.} 
\LX{baalonlɔnnaa} \HM{} \PH{bààlʋ̀nlɔ́nnáá} \PS{n.~} \GE{contagious disease} \OP{baalonlɔnnɔ} \TP{baalonlɔnnɛɛ} 
\LX{baalonnooraa} \HM{} \PH{bààlʋ́nnʋ̀ʋ̀ráà} \PS{n.~} \GE{recurring illness} \XV{baalonnoore} \XE{} 
\LX{baalugiri} \HM{} \PH{bààlùgìrí} \PS{n.~} \GE{armpit} \SG{baalugiruu} \OP{baalugo} \TP{baalugiree} \VA{bagelugri}
\LX{baaluu} \HM{} \PH{bààlúú} \PS{n.~} \SN \GE{gallon measure} \SN \GE{balloon} \SG{baaluuruu} \OP{baaluuri} \TP{baaluuree} 
\LX{baa-ne-ŋmaaŋa} \HM{} \PH{bàá-nɪ́-ŋmááŋà} \PS{n.~} \GE{Ursa Major (constellation); also known as Big Dipper, Plough, Great Bear} 
\LX{baapɛmpɛle} \HM{} \PH{bàápɛ̀mpɛ́lɛ̀} \PS{n. pl.~} \GE{lily pads} 
\LX{baaraa} \HM{} \PH{bááráá} \PS{n.~} \GE{end, last} \XV{A baaraa zaa te na nyɛ la a nannyige.} \XE{At the end of it all, we will find the thief.} 
\LX{baaraakyi} \HM{} \PH{báárááʧì} \PS{n.~} \GE{bridge} \OP{baaraakyiri} \TP{baaraakyiree} 
\LX{baare} \HM{} \PH{báárɪ̀} \PS{v.~} \SN \GE{to finish, to end, to complete} \XV{A tontonema baare la a toma.} \XE{The workers have finished the work.} \SN \GE{to be used up} \XV{A baaloŋ vɛŋɛɛ la ka a dɔɔ bɛroŋ na baare la o zaa.} \XE{Sickness has caused that man to lose weight.} \PD{baarɛɛ, baara, baara, baareba, baaraa, - } 
\LX{baasaabiri} \HM{} \PH{bààsààbírí} \PS{n.~} \SN \GE{a seed from a type of tree} \SN \GE{a type of large bead made from this seed} \SG{baasaabiruu} \OP{baasaabie} \TP{baasaabiiree} 
\LX{Baasaale} \HM{} \PH{bàásààlɪ̀} \PS{n.~} \GE{Baasare (ethnic group)} \SG{baasaaloo} 
\LX{baasakuuri} \HM{} \PH{báásàkúúrì} \PS{n.~} \GE{bicycle} \OP{baasakue} \TP{baasakuuree} 
\LX{baataŋaa} \HM{} \PH{báátáŋáá} \PS{n.~} \GE{a type of tree that resembles the shea nut tree} \OP{baataŋema} \TP{baataŋemɛɛ/baataanɛɛ} 
\LX{babammo} \HM{} \PH{bábámmʊ́} \PS{n.~} \GE{ignorance} 
\LX{babaŋena} \HM{} \PH{bábáŋɪ́ná} \PS{n.~} \GE{one who is innocent} \OP{babaŋenema} \TP{babaŋenemɛɛ} 
\LX{babaŋyeli} \HM{} \PH{bábàŋyélí} \PS{n.~} \GE{innocence} \VA{dakumo}
\LX{babere} \HM{} \PH{bàbɪ̀rɪ́} \PS{n.~} \GE{the breast of a dog} \OP{baberɛ} \TP{baberɛɛ} 
\LX{babeɛ} \HM{} \PH{bábɪ́ɛ́} \PS{n.~} \GE{a friend that cannot be trusted} \OP{babeere} \TP{babeerɛɛ} 
\LX{babe᷉e} \HM{} \PH{bàbɪ́᷉ɪ́᷉} \PS{n.~} \GE{broth from dog meat} \TP{babeenɛɛ} 
\LX{babeɛraa} \HM{} \PH{bàbɪ̀ɛ́ráà} \PS{n.~} \GE{bridge} \OP{babeɛre} \TP{babeɛrɛɛ} \VA{baaraakyi, takoriko.}
\LX{babiŋ} \HM{} \PH{bábíŋ} \PS{n.~} \GE{a type of tree} \OP{babinni} \TP{babinnee} 
\LX{babini} \HM{} \PH{bàbíní} \PS{n.~} \GE{dog faeces} \OP{babinee} \TP{babinnee} 
\LX{babiri} \HM{} \PH{bábírì} \PS{n.~} \GE{bosom friend} \OP{babie} \TP{babiiree} \VA{batuo}
\LX{babogi} \HM{} \PH{bàbógí} \PS{n.~} \GE{the deepest part of a body of water} \OP{babogri} \TP{babogree} 
\LX{babullaa} \HM{} \PH{bàbùlláà} \PS{n.~} \GE{spring, pond where water is drawn in small amounts (usually during dry season)} \OP{babulli} 
\LX{badaa} \HM{} \PH{bàdáá} \PS{n.~} \GE{male dog} \OP{badaare} \TP{badaarɛɛ} 
\LX{badaa-de-nyoɔre} \HM{} \PH{bàdáádíɲɛʊ́ɔ́rɪ́} \PS{n.~} \GE{the last day of a festival} \TP{badaadenyoɔrɛɛ} \VA{baraadenyoɔre}
\LX{badalane} \HM{} \PH{bàdàlánɪ́} \PS{n.~} \GE{uncastrated male dog} \OP{badalama} \TP{badalamɛɛ} 
\LX{badalee} \HM{} \PH{bàdàléé} \PS{n.~} \GE{male puppy} \OP{badabilii} \TP{badabilee/badabillee} 
\LX{badavaraa} \HM{} \PH{bàdàváràà} \PS{n.~} \GE{castrated dog} \OP{badavarre} \TP{badavarɛɛ/badavarrɛɛ} 
\LX{badie} \HM{} \PH{bàdìé} \PS{n.~} \GE{the deepest part of a body of water} \OP{baderi} \TP{baderee} 
\LX{badoge} \HM{} \PH{bàdʊ̀gɪ̀} \PS{n.~} \GE{a plant with a sweet smell, used to season sauce} \OP{badogre} \TP{badogrɛɛ} 
\LX{badɔgerɔ} \HM{} \PH{bàdɔ̀grɔ́} \PS{n.~} \GE{female dog that has recently given birth} \OP{badɔgere} \TP{badɔgerɛɛ} \VA{badɔgera}
\LX{badɛre} \HM{} \PH{bádɛ́rɪ́} \PS{n.~} \SN \GE{spider} \SN \GE{cunning person (fig.)} \OP{badɛrre} \TP{badɛrɛɛ} 
\LX{bafoɔlaa} \HM{} \PH{bàfʊ́ɔ́láá} \PS{n.~} \SN \GE{valley} \SN \GE{creek} \OP{bafoɔle} \TP{baroɔlɛɛ} 
\LX{baga} \HM{} \PH{bágà} \PS{n.~} \GE{a triangular garment worn by men as underwear} \OP{bagere} \TP{bagerɛɛ} \VA{lere, para}
\LX{baga} \HM{} \PH{bàgà} \PS{n.~} \GE{not caring, unconcerned} \XV{Ka onaŋ ba wa a pɔgebeɛloŋ kyɛ n baga kyebe.} \XE{I don't care if she doesn't attend the wedding.} 
\LX{bagaŋ} \HM{} \PH{bágáŋ} \PS{n.~} \GE{a mad dog} \OP{baganne} \TP{baganɛɛ/bagannɛɛ} \VA{bagao}
\LX{baganaa} \HM{} \PH{bàgànàà} \PS{n.~} \GE{dog leash} \OP{baganne} \TP{baganɛɛ/bagannɛɛ} 
\LX{bagane} \HM{} \PH{bàgánɪ̀} \PS{n.~} \GE{dog skin, dog hide} \OP{bagama} \TP{bagannɛɛ} 
\LX{bagao} \HM{} \PH{bàgáʋ̀} \PS{n.~} \GE{mad dog} \OP{bagagre} \TP{bagagrɛɛ} \VA{bagaŋ}
\LX{bagbɛre} \HM{} \PH{bàgbɛ́rɪ̀} \PS{n.~} \GE{leg of a dog} \OP{bagbɛɛ} \TP{bagbɛɛrɛɛ} 
\LX{bagbɛre} \HM{} \PH{bàgbɛ́rɪ̀} \PS{n.~} \GE{valley} \OP{bagbɛɛ} \TP{bagbɛɛrɛɛ} 
\LX{bagboŋ} \HM{} \PH{bágbòŋ} \PS{n.~} \GE{fool} \OP{bagbonni} \TP{bagbonnee} 
\LX{bagbori} \HM{} \PH{bàgbòrí} \PS{n.~} \GE{trough for a dog} \OP{bagboe} \TP{bagboree} 
\LX{bageba} \HM{} \PH{bágɪ́bá} \PS{n.~} \GE{a type of sacred cult which initiates boys and girls into adulthood} \TP{bagebɛɛ} \VA{bɔgebɔ}
\LX{bageboge} \HM{} \PH{bàgɪ̀bʋ̀gɪ̀} \PS{n.~} \GE{the act of divining} \XV{Kiritabiiri kyiiruŋ la bageboge.} \XE{It is a taboo for Christians to participate  in divining.} \VA{bɔgeboge}
\LX{bageboge-dagoli} \HM{} \PH{bàgɪ̀bʊ̀g-dàgólí} \PS{n.~} \GE{divining stick} \OP{bageboge-dagolo} \TP{bageboge-dagolee} \VA{bɔgeboge-dagoli}
\LX{bageboge-karaa} \HM{} \PH{bàgɪ̀bʊ̀g-kàràá} \PS{n.~} \GE{chief diviner} \OP{bagebogekarre} \TP{bagebogekarɛɛ} \VA{bɔgeboge-karaa}
\LX{bageboge-woɔ} \HM{} \PH{bàgɪ̀bʊ̀gɪ̀-wʊ́ɔ́} \PS{n.~} \GE{divining bag} \OP{bageboge-woore} \TP{bageboge-woorɛɛ} \VA{bɔgeboge-woɔ}
\LX{bagere} \HM{} \PH{bàgrɪ́} \PS{n.~} \GE{sacrifice for gods} \OP{baga} \TP{bagɛɛ} \VA{bɔgere}
\LX{bagukyɔfeɛle} \HM{} \PH{bágùʧɔ̀nfɪ́ɛ́lɪ́} \PS{n.~} \GE{very early morning (usually cold)} 
\LX{baguo} \HM{} \PH{bágúó} \PS{n.~} \GE{morning} \OP{baguuri} \TP{baguuree} \VA{beguo}
\LX{bagupi} \HM{} \PH{bàgùpí} \PS{n.~} \GE{early morning} \OP{baguipiri} \TP{bagupiree} \VA{bebipi}
\LX{bagupipi} \HM{} \PH{bágùpípí} \PS{n.~} \GE{very early morning} \OP{bagupipiri} \TP{bagupipiree} \VA{bagupi}
\LX{bagyeɛnɛ} \HM{} \PH{bàdzɪ̀ɛ̀nɛ́} \PS{n.~} \GE{a type of children's dance} \TP{bagyeɛnɛɛ} 
\LX{bagyoɔ} \HM{} \PH{bàdzʊ́ɔ́} \PS{n.~} \SN \GE{squirrel-like rodent (believed to be very sexually active)} \SN \GE{sex maniac (male)} \VA{bagyoa, bazoɔ}
\LX{bagyɔroo} \HM{} \PH{bàdzɔ̀rʊ́ʊ́} \PS{n.~} \GE{a thick, fleshy, poisonous creeping plant} 
\LX{ba-irime} \HM{} \PH{bá-írímé} \PS{n.~} \GE{a type of wild mouse found along river valleys which has tender skin} \OP{ba-irri} \TP{ba-iree} 
\LX{bakana} \HM{} \PH{bàkáná} \PS{n.~} \GE{broth from dog meat, made with flour} \TP{bakanɛɛ} 
\LX{bakɔ} \HM{} \PH{bàkɔ̀} \PS{n.~} \GE{a type of yam (early maturing)} \OP{bakɔɛ} \TP{bakɔrɛɛ} 
\LX{bakore} \HM{} \PH{bákʊ́rɪ́} \PS{n.~} \GE{trusted friend} \OP{bakorɔ/bakoremine} \TP{bakorɛɛ/bakorrɛɛ} 
\LX{bakpoŋ} \HM{} \PH{bàkpóŋ} \PS{n.~} \SN \GE{large stream, river} \SN \GE{(capitalized) the name of a tributary of the Black Volta River in the Upper West Region} \OP{bakponni} \TP{bakponnee} 
\LX{balaa} \HM{} \PH{báláà} \PS{n.~} \GE{waistband (pair of shorts)} \OP{balle} \TP{balɛɛ} 
\LX{balambaaraa} \HM{} \PH{bálámbááráá} \PS{n.~} \GE{peg used for tethering animals} \OP{balambaare} \TP{balambaarɛɛ} \VA{balambiri}
\LX{balambiri} \HM{} \PH{bálámbírí} \PS{n.~} \GE{peg used for tethering animals} \OP{balambie} \TP{balambiiree} \VA{balambaaraa}
\LX{balantamm} \HM{} \PH{bàlàntàmm} \PS{adv.~} \GE{large and heavy} \XV{Wɔɔ waa la kpoŋi balantamm.} \XE{The elephant is enormously large.} 
\LX{balante} \HM{} \PH{bálántɪ́} \PS{n.~} \GE{belt} \SG{balanteerɛɛ} \OP{balantere} \TP{balantere} \VA{bɛlɛnte}
\LX{bale} \HM{} \PH{bàlɪ̀} \PS{v.~} \SN \GE{to be tired} \XV{O toŋ la a toma te zuo lɛnso ka o bale.} \XE{He has overworked himself; that's why he is tired.} \SN \GE{to be fed up} \XV{Bebiri zaa ka o maŋ wa sɔre libie, maa eŋ bale la o yɛlɛ.} \XE{He comes to beg for money every day; I'm fed up with him.} \SN \GE{to be weak} \XV{Ka neɛŋ kore o maŋ bale la.} \XE{One becomes weak when one is old.} \SN \GE{to be soft} \XV{A gane koe la, yɛ mere ko᷉ɔ ka o bale.} \XE{The animal skin is dry and hard, so sprinkle water on it to make it soft.} \PD{balɛɛ, balla, balla, baleba, balaa, baleba } 
\LX{bale} \HM{} \PH{bálɪ́} \PS{n.~} \GE{tribe, ethnic group} \XV{Dagaaba bale kaŋa yuori la Kowɛrɛɛ.} \XE{Kowɛrɛɛ is one of the tribes among the Dagaaba.} \OP{balmine} \TP{balɛɛ} 
\LX{bale} \HM{} \PH{bálɪ́} \PS{v.~} \SN \GE{avoid, evade} \XV{O daare to ma lɛnso ka o maŋ wa nyɛ ma bale.} \XE{She insulted me the last time, and that's why she avoids me when she sees me.} \SN \GE{divert} \XV{Ba naŋ da beɛrɛ a baa ba bale la a sori.} \XE{When they were constucting the bridge, they diverted the road.} \PD{balɛɛ, bala, bala, baleba, balaa, baleba } 
\LX{bale} \HM{} \PH{bálɪ́} \PS{adj.~} \GE{of a kind} \XV{A toma ŋa bale ba kye kyɛ.} \XE{We do not have his kind of work here.} 
\LX{balee} \HM{} \PH{bàléé} \PS{n.~} \GE{brook, stream} \OP{babilii} \TP{babillee} 
\LX{balee} \HM{} \PH{bàléé} \PS{n.~} \GE{puppy} \OP{babilii} \TP{babillee} 
\LX{balee} \HM{} \PH{bàlɪ̀ɪ̀} \PS{adj.~} \GE{of a dubious and dangerous nature} \XV{Pɔge koŋ sage toŋ tombalee ŋa tɔ.} \XE{A woman will not do this sort of shady thing.} 
\LX{baleŋkpɔgre} \HM{} \PH{bàlɪ́ŋkpɔ̀grì} \PS{n.~} \GE{tick} \OP{baleŋkpɔgɔ} \TP{baleŋkpɔgrɛɛ} 
\LX{balɔre} \HM{} \PH{bálɔ́rɪ́} \PS{n.~} \SN \GE{ugly person or thing} \SN \GE{one without strength} \OP{balɔɛ} \TP{balɔrɛɛ} 
\LX{balɔrroŋ} \HM{} \PH{bálɔ̀rrʊ́ŋ} \PS{n.~} \GE{not being strong} 
\LX{balɔrroŋ} \HM{} \PH{bálɔ̀rrʊ́ŋ} \PS{n.~} \GE{act of being ugly} 
\LX{bam} \HM{} \PH{bǎm} \PS{n.~} \GE{a type of large yam} \OP{bama} \TP{bamɛɛ} 
\LX{bama} \HM{} \PH{bámà} \PS{demonstrative pronoun.~} \GE{these (human)} \XV{A noba bama la maŋ toŋ a toma wie᷉u.} \XE{These are the people who normally do the work quickly.} \XV{Bibiiri bama la ka ba iri.} \XE{These are the children that have been chosen.} 
\LX{bambaala} \HM{} \PH{bàmbààlá} \PS{n.~} \GE{unassuming person; poor person} \OP{bambaaleba} \TP{bambaalebɛɛ} 
\LX{bambaalaa} \HM{} \PH{bámbááláá} \PS{n.~} \GE{small fish, fingerling} \OP{bambaale} \TP{bambaalɛɛ} 
\LX{bambalaaraa} \HM{} \PH{bámbálááráá} \PS{n.~} \GE{swallow (bird)} \OP{bambalaare} \TP{bambalaarɛɛ} 
\LX{bambaŋ} \HM{} \PH{bámbáŋ} \PS{adv.~} \GE{hot, steamy} \XV{A zee, re ba taa diibu, a toloŋ bambaŋ.} \XE{The soup is too hot; it can't be eaten.} \XV{Tentoloŋ saŋa a ŋmenaa maŋ toloŋ bambaŋ.} \XE{During the hot season, the sun is always extremely hot.} 
\LX{bambiŋi} \HM{} \PH{bàmbìŋ} \PS{n.~} \GE{an arrangement made in advance by two parties} 
\LX{bambugo} \HM{} \PH{bámbúgó} \PS{n.~} \GE{confused person; fool} \OP{bambugri} \TP{ bambugree} 
\LX{bambuŋi} \HM{} \PH{bàmbúŋ} \PS{n.~} \GE{salp (fish of the class Thaliacea)} \OP{babunni} \TP{bambunnee} 
\LX{bambuŋi} \HM{} \PH{bàmbúŋ} \PS{n.~} \GE{locally made gunpowder} \OP{babunni} \TP{bambunnee} 
\LX{bamm} \HM{} \PH{bàmm} \PS{adv.~} \GE{big and heavy} \XV{A nannyige zu mɔrɛɛ bamm lɛ.} \XE{The thief's head is badly swollen.} 
\LX{bammo} \HM{} \PH{bàmmʊ́} \PS{n.~} \GE{intelligence, wit} \XV{Kɔkɔlaa taa la bammo gaŋ baa.} \XE{The cat is more intelligent than the dog.} 
\LX{bamoɔ} \HM{} \PH{bàmʊ́ɔ́} \PS{n.~} \GE{a type of tall, strong grass found around bodies of water, whose dry stalks are used for weaving} \SG{bamoɔnoo} \OP{bamoore} \TP{bamoorɛɛ} 
\LX{baŋa} \HM{} \PH{báŋá} \PS{n.~} \GE{bangle} \OP{banne} \TP{bannɛɛ} \VA{bɔŋɔ}
\LX{baŋa} \HM{} \PH{bàŋá} \PS{n.~} \GE{lizard} \OP{banne} \TP{bannɛɛ} \VA{shortened form of bandaa}
\LX{bandaa} \HM{} \PH{bándáá} \PS{n.~} \GE{lizard (male)} \OP{bandaare} \TP{bandarɛɛ/bandaarɛɛ} \VA{baŋa, bɔndaa}
\LX{bandaakpogilo} \HM{} \PH{bàndáákpóglò} \PS{n.~} \GE{the act of hitting the opponent's armpits with one's knees to weaken the opponent (a wrestling move)} \XV{Bayuo to la Dɛre ka Dɛre nyɔge o lɔɔ tɔ bandaakpogilo.} \XE{Bayuo insulted Dɛre, and Dɛre threw Bayuo down and used his knees to hit his armpits.} \TP{bandaakpogilee} 
\LX{bandakarekpele} \HM{} \PH{bàndàkárkpɪ́lɪ́} \PS{n.~} \GE{Agama agama; also known as red-headed rock agama} \OP{bandakarekpelɛ} \TP{bandakarekpelɛɛ} \VA{karekpele}
\LX{bandakarema} \HM{} \PH{bàndàkàrɪ̀mà} \PS{n.~} \GE{large male lizard} \TP{bandakaremɛɛ} 
\LX{bandanyene} \HM{} \PH{bàndàɲɛɪ́nɪ́} \PS{n.~} \GE{tooth of a lizard} \OP{bandanyemɛ} \TP{bandanyenɛɛ} 
\LX{bandapoleɛne} \HM{} \PH{bàndàpʊ́lɪ́ɛ́nɪ́} \PS{n.~} \GE{female lizard} \OP{bandapoleɛmɛ} \TP{bandapoleɛmɛɛ/bandapoleɛnɛɛ} \VA{poleɛne}
\LX{bandazeɛraa} \HM{} \PH{bàndàzɪ̀ɛ́ráá} \PS{n.~} \GE{kite; small bird of prey (family Accipitridae)} \OP{bandazeɛre} \TP{bandazeɛrɛɛ} 
\LX{bandazoore} \HM{} \PH{bàndàzʊ́ʊ́rɪ́} \PS{n.~} \GE{tail of a lizard} \OP{bandazoɛ} \TP{bandazoorɛɛ} 
\LX{baŋe} \HM{} \PH{bàŋ} \PS{v.~} \SN \GE{to know (something or someone)} \XV{N baŋe la a pɔge ŋa serɛ, o yuori la Dabuo.} \XE{I know this woman's husband; he is called Dabuo.} \SN \GE{to be aware of} \XV{N baŋe ne la o waabo.} \XE{I know he's coming.} \PD{baŋɛɛ, banna, banna, banema, bannaa, banema } \VA{bɔŋe}
\LX{baŋena} \HM{} \PH{bàŋɪ́ná} \PS{n.~} \SN \GE{area between the shoulder and neck} \SN \GE{trickster} \OP{baŋema/banema} \TP{baŋemɛɛ/banemɛɛ} \VA{bɔŋenɔ}
\LX{baŋenaa} \HM{} \PH{báŋɪ́náá} \PS{n.~} \GE{a type of tree that has medicinal properties} \OP{baŋene} \TP{baŋenɛɛ} \VA{bɔŋenaa}
\LX{baŋene} \HM{} \PH{bàŋɪ̀nɪ́} \PS{n.~} \GE{cleverness} \XV{Nannyige maŋ taa la baŋene, ka o nyɔgebo maŋ e tuo.} \XE{A thief usually has such cleverness that it is difficult to catch him.} \VA{bɔŋene}
\LX{banfane} \HM{} \PH{bànfánɪ́} \PS{n.~} \GE{a bangle that is believed to have medicinal value and may be worn to cure a disease} \OP{banfama} \TP{banfanɛɛ} 
\LX{baŋgori} \HM{} \PH{báŋgórí} \PS{n.~} \GE{high-heeled shoes} \SG{baŋgoruu} \OP{baŋgorri} \TP{baŋgoree/baŋgorree.} 
\LX{bangye} \HM{} \PH{bándzɪ́} \PS{n.~} \GE{bank} \OP{bangyere} \TP{bangyerɛɛ} 
\LX{bangyeraa} \HM{} \PH{bándzɪ́ráá} \PS{n.~} \SN \GE{latrine; washroom} \XV{Pampana ka fooŋ mɛ yiri a seŋ ka bagyeraa bebe.} \XE{Modern houses are supposed to have built-in washrooms.} \SN \GE{feces} \XV{A bie nyɛ la bini meele a sɛŋ.} \XE{The child has soiled the mat with feces.} \OP{bangyerre} \TP{bangyerɛɛ/bangyerrɛɛ} 
\LX{ba-nimie-anaare} \HM{} \PH{bà-nímíè-ànàárɪ̀} \PS{n.~} \GE{a dog, usually with big spots above the eyes that look like a second pair of eyes, believed to have supernatural powers} \OP{ba-nimie-anaare} \TP{ba-nimie-anaarɛɛ} 
\LX{baŋkaroŋ} \HM{} \PH{bàŋkàrʊ́ŋ} \PS{n.~} \GE{a type of brown bird with long, thin legs} \OP{baŋkarre} \TP{baŋkarɛɛ/baŋkarrɛɛ} 
\LX{baŋkoroŋ} \HM{} \PH{bàŋkʊ́rʊ́ŋ} \PS{n.~} \GE{foreknowledge} 
\LX{baŋŋmeɛle} \HM{} \PH{bàŋŋmɪ́ɛ́lɪ́} \PS{n.~} \GE{twisted bangle, often with a spiritual or medicinal significance} \OP{ baŋŋmeɛlɛ} \TP{baŋŋmeɛlɛɛ} 
\LX{bantakperee} \HM{} \PH{bàntákpérèé} \PS{n.~} \GE{a type of spotted lizard (believed to be poisonous)} \OP{bantakperri} \TP{bantakperree} 
\LX{bantoɔ} \HM{} \PH{bántʊ́ɔ́} \PS{n.~} \GE{rectal bulb syringe} \OP{bantoore} \TP{bantoorɛɛ} \VA{bɔntoɔ}
\LX{banweebaa} \HM{} \PH{bànwɪ́ɪ́báá} \PS{n.~} \GE{carnivorous bird (bigger than a kite but strong and fast)} \OP{banweere} \TP{banweerɛɛ} 
\LX{banyaa} \HM{} \PH{bàɲáà} \PS{n.~} \GE{chest of a dog} \OP{banyaɛ/banyaane} \TP{banyaanɛɛ} 
\LX{banyagere} \HM{} \PH{bàɲágrɪ́} \PS{n.~} \GE{intestine of a dog} \OP{banyaga} \TP{banyagɛɛ/banyagerɛɛ} 
\LX{banyaŋaa} \HM{} \PH{bàɲáŋáá} \PS{n.~} \GE{female dog} \OP{banyaŋene /banyanne} \TP{banyaŋenɛɛ/banyannɛɛ} 
\LX{Banyene} \HM{} \PH{bàɲɪ̀nɪ̀} \PS{n.~} \GE{homeland of the Banyenɛɛ clan} 
\LX{Banyenɛɛ} \HM{} \PH{bàɲɪ́nɛ́ɛ́} \PS{n.~} \GE{name of a clan} \SG{banyenoo} 
\LX{banyɛtɔgle} \HM{} \PH{báɲɛtɔ́glɪ́} \PS{n.~} \GE{imitation} \XV{Bibiiri taa la banyɛtɔgle, azuiŋ neŋkponni seŋ ka ba taa ba menne velaa.} \XE{Children like to immitate, so adults should behave well.} 
\LX{banzeɛ} \HM{} \PH{bànzɪ̀ɛ̀} \PS{n.~} \GE{copper bangle} \OP{banzeere} \TP{banzeerɛɛ} 
\LX{banzeɛraa} \HM{} \PH{bǎnzɪ́ɛ́ráá} \PS{n.~} \GE{bird that eats lizards} \OP{banzeɛre} \TP{banzeɛrɛɛ} 
\LX{banzɔlɔ} \HM{} \PH{bànzɔ̀lɔ̀} \PS{n.~} \GE{chain} \OP{banzɔlle} \TP{banzɔllɛɛ} \VA{bɔnzɔlɔ}
\LX{ba-ɔge} \HM{} \PH{bà-ɔ̀gɪ́} \PS{n.~} \GE{a dog's nest, usually for a female that has recently given birth} \OP{ba-ɔgre} \TP{ba-ɔgɛɛ/ba-ɔgrɛɛ} 
\LX{bapare} \HM{} \PH{bápárɪ́} \PS{n.~} \GE{close friend} \OP{baparemine} \TP{baparɛɛ} \VA{batuo}
\LX{bapare} \HM{} \PH{bàpárɪ́} \PS{n.~} \GE{end of a valley} \OP{baparɛɛ} 
\LX{bapiiraa} \HM{} \PH{bàpííráá} \PS{n.~} \GE{a puppy earmarked for someone} \OP{bapiiri} \TP{bapiiree} 
\LX{bapimbaare} \HM{} \PH{bápímbáárɪ̀} \PS{n.~} \GE{a small part of the stomach of a ruminant that has many compartments} \OP{bapimbaare} \TP{bapimbaarɛɛ} 
\LX{bara} \HM{} \PH{bárá} \PS{n.~} \GE{boastful threat, often considered inappropriate or rude} \XV{Dakoraa naŋ baŋ ka dɔɔ zaa kye a yiri poɔ la so ka o wa fare a bara ama taaba.} \XE{Knowing that all the men had left the house, Dakoraa went to make boastful threats.} \TP{barɛɛ} 
\LX{baraa} \HM{} \PH{báráá} \PS{n.~} \GE{not making any effort; helpless} 
\LX{bare} \HM{} \PH{bàrɪ̀} \PS{v.~} \SN \GE{to leave, to release, to let go} \XV{Ba bare la sakubiiri.} \XE{They have let the schoolchildren go for vacation.} \SN \GE{to discontinue, to abstain, to cease, to stop} \XV{A baala bare la a dɔɔta gaabo.} \XE{The patient has stayed away from the hospital.} \SN \GE{to divorce (initiated by husband)} \XV{A dɔɔ gaa bare la o pɔge.} \XE{The man has divorced his wife.} \SN \GE{not having power} \XV{A dɔɔ ŋa eŋ barɛɛ la te zuo, o koŋ to᷉ɔ a dɛre do.} \XE{This man is so useless that he cannot climb the ladder.} \PD{barɛɛ, bara, bara, bareba, baraa, bareba } 
\LX{bare} \HM{} \PH{bàrɪ̀} \PS{v.~} \SN \GE{to cultivate seeds} \XV{A pɔge bare la kommie a dabɔge poɔ.} \XE{The woman has cultivated tomato seeds in the garden.} \SN \GE{to offer firstfruits to the gods} \XV{Bɔgebogereba maŋ bare la bompaala kyɛ baŋ di.} \XE{Diviners normally offer firstfruits to their gods before eating.} \PD{barɛɛ, bara, bara, bareba, baraa, bareba } 
\LX{bare gyɛle} \HM{} \PH{} \PS{v.~} \GE{to be careless; to not pay attention} \PD{barɛɛ gyɛle, bara gyɛle, gyɛle bara, -, -, - } 
\LX{bare yɛŋe} \HM{} \PH{bàrɪ̀ yɛ́ŋ} \PS{v.~} \GE{to feel free, to relax} \PD{-, -, -, -, -, - } 
\LX{barekanneɛbaa} \HM{} \PH{bàrɪ̀kánnɪ́ɛ́báá} \PS{n.~} \GE{local spirits (alcohol)} \OP{barekanneɛbaare} \TP{barekanneɛbaarɛɛ} 
\LX{barre} \HM{} \PH{bárrɪ́} \PS{v.~} \GE{to haggle} \PD{barrɛɛ, barra, barra, bareba, barraa, - } 
\LX{ba-sakɔre} \HM{} \PH{bà-sakɔrɪ́} \PS{} \GE{lazy and useless dog} 
\LX{basasale} \HM{} \PH{bàsàsàlɪ́} \PS{n.~} \GE{slippery waterway} \TP{basasalɛɛ} 
\LX{ba-soɔmee} \HM{} \PH{bà-sʊ́ɔ́mɪ́ɪ́} \PS{n.~} \GE{weak and sickly dog} \VA{ba-sakɔre}
\LX{bataabo} \HM{} \PH{bátáábʊ́} \PS{n.~} \GE{the state of not having anything} 
\LX{batara} \HM{} \PH{bátárá} \PS{n.~} \GE{poor or needy person} \OP{batareba} \TP{batarebɛɛ} 
\LX{batuo} \HM{} \PH{bátúó} \PS{n.~} \GE{bosom friend} \OP{batuuri} \TP{batuuree} \VA{bakore bapare, bameŋɛ, babiri}
\LX{bavilime} \HM{} \PH{bàvílímé} \PS{n.~} \GE{lake} \OP{bavilimeri} \TP{bavilimee} 
\LX{bawaa} \HM{} \PH{báwáá} \PS{n.~} \GE{a type of dance} \OP{bawaare} \TP{bawaarɛɛ} 
\LX{Bayaayiree} \HM{} \PH{báyààyíréé} \PS{n.~} \GE{people of Bayaayiri} \SG{bayaayiruu} 
\LX{Bayaayiri} \HM{} \PH{báyààyírí} \PS{n.~} \SN \GE{a settlement in the Upper West Region} \SN \GE{ancestors' dwelling; next world} 
\LX{Bayuo} \HM{} \PH{báyúó} \PS{n.~} \GE{name traditionally given to a male child born on a funeral day} \TP{bayumine} 
\LX{bazele} \HM{} \PH{bàzɪ̀lɪ́} \PS{n.~} \GE{tongue of a dog} \OP{bazelɛ} \TP{bazelɛɛ} 
\LX{baze᷉e} \HM{} \PH{bàzɪ́᷉ɪ̀᷉} \PS{n.~} \GE{blood of a dog} \TP{bazeenɛɛ} 
\LX{bazoɔ} \HM{} \PH{bàzʊ́ɔ̀} \PS{n.~} \GE{a type of wild fox} \TP{bazoorɛɛ} \VA{bazoa, bagyoɔ}
\LX{bazoraa} \HM{} \PH{bàzóráà} \PS{n.~} \GE{fast dog} \OP{bazorri} \TP{bazorree} 
\LX{bazugiri} \HM{} \PH{bázúgrí} \PS{n.~} \GE{a type of wild mouse that has thick fur} \OP{bazugo} \TP{bazugee/bazugiree} 
\LX{ba᷉abo᷉o} \HM{} \PH{bà᷉à᷉bʊ́᷉ʊ́᷉} \PS{n.~} \GE{bustard} \OP{ba᷉aboomɔ/ba᷉aboone} \TP{ba᷉aboomɛɛ/ba᷉aboonɛɛ} 
\LX{be} \HM{} \PH{bě} \PS{v.~} \GE{to exist, to be there} \XV{A baala be la asibiti poɔ.} \XE{The patient is in the hospital.} \PD{bee, bere, bere, beribe, beraa, beribe } \SE{be nimiri} \GE{to be awake, to be alert} 
\LX{be} \HM{} \PH{bě} \PS{loc.~} \GE{there (place)} \XV{N ba nyɛ fo be.} \XE{I did not see you there.} 
\LX{be} \HM{} \PH{bɪ̀} \PS{v.~} \SN \GE{to become boiled or cooked} \XV{A nɛne ba be, o varema.} \XE{The meat is not well boiled, it is hard.} \SN \GE{to be smooth (flour)} \XV{A mansinni ɛre la a bɛnzɔŋ ka a be.} \XE{The machine has ground the bean flour very smooth.} \SN \GE{to mature, to ripen (fruit, millet)} \XV{Ka taŋaaŋ wa be o maŋ pɔre le la.} \XE{A ripe shea fruit normally falls off the tree.} \SN \GE{to ferment (alcohol)} \XV{A da᷉a naŋ ba be azuiŋ a ba bugiro.} \XE{The pito is not yet fermented, so it is not alcholic.} \SN \GE{to fortify (spiritually)} \XV{Pɔlebilii yaga maŋ boɔrɔ ka ba tegere ba menne a be.} \XE{Today, most young men want to fortify themselves through spritual powers.} \PD{beɛ,  berɛ, beerɛ, berebɛ, beraa, berebɛ } 
\LX{be} \HM{} \PH{bɪ́} \PS{conj.~} \GE{or} \XV{Fo na gaa la be?} \XE{Will you go or not?.} \VA{shortened form of bee}
\LX{bɛ} \HM{} \PH{bɛ̀} \PS{v.~} \GE{to be sick, to be ill} \XV{Ka saaŋ poɔ a bie o na bɛ la.} \XE{If the child is out in the downpour he will get sick.} \PD{bɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ } 
\LX{bɛ} \HM{} \PH{bɛ̀} \PS{v.~} \GE{to seal, to close a hole with a muddy substance} \XV{Ka fooŋ ba bɛ a peɛ voɔre ka saaŋ wa miire a die na kyire la.} \XE{If you do not patch the hole in the roof, the room will leak when there is rain.} \PD{bɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ } 
\LX{bɛ} \HM{} \PH{bɛ́} \PS{interrogative.~} \GE{where} \XV{A libie bɛ?} \XE{Where is the money?.} \VA{bɛɛ}
\LX{bebe} \HM{} \PH{bèbé} \PS{adj.~} \GE{present, available} 
\LX{bɛbɛgere} \HM{} \PH{bɛ̀bɛ̌grɪ̀} \PS{n.~} \GE{mud} \OP{bɛbɛgɛ} \TP{bɛbɛgerɛɛ} 
\LX{bebelaa} \HM{} \PH{bɪ́bɪ́láá} \PS{n.~} \GE{termite hill} \OP{bebelle} \TP{bebellɛɛ} \VA{belaa}
\LX{bebelebiri} \HM{} \PH{bɪ́bɪ̀lɪ̀bírì} \PS{n.~} \GE{white ant} \SG{bebelebiruu} \OP{bebelebie} \TP{bebelebiiree} 
\LX{beberaa} \HM{} \PH{bébéráà} \PS{n.~} \GE{a persimist} \OP{beberri} \VA{beberee}
\LX{beberee} \HM{} \PH{bébéréè} \PS{n.~} \GE{one who is malicious} \OP{beberri} \TP{beberree} \VA{bebberaa}
\LX{beberiyɛlɛ} \HM{} \PH{bébèrìyɛ́lɛ́} \PS{n.~} \GE{malicious talk} \SG{beberiyeli} 
\LX{beberuŋ} \HM{} \PH{bébérúŋ} \PS{n.~} \GE{the act of causing trouble} 
\LX{bebibaaraa} \HM{} \PH{bèbíbááráá} \PS{n.~} \GE{last day} \OP{bebibaare} \TP{bebibaarɛɛ} 
\LX{bebibalaa} \HM{} \PH{bèbíbàlàà} \PS{n.~} \GE{particular day} 
\LX{bebidɛŋe} \HM{} \PH{bèbídɛ́ŋ} \PS{n.~} \GE{first day} \OP{bebidɛnne} \TP{bebidɛnnɛɛ} \VA{bebidɛndɛŋ}
\LX{bebifaa} \HM{} \PH{bèbífáá} \PS{n.~} \GE{sabbath; sacred day} \OP{bebifaare} \TP{bebifaarɛɛ} 
\LX{bebigbuli} \HM{} \PH{bèbígbùlì} \PS{n.~} \GE{whole day} \OP{bebigbulo} \TP{bebigbulee} 
\LX{bebikore} \HM{} \PH{bèbíkʊ́rɪ́} \PS{n.~} \GE{a former day, a day in the past} \OP{bebikorɔ/bebikora} \TP{bebikorɛɛ} 
\LX{bebipaalaa} \HM{} \PH{bèbípááláá} \PS{n.~} \GE{new day; different day} \OP{bebipaala} \TP{bebipaalɛɛ} 
\LX{bebiri} \HM{} \PH{bèbírì} \PS{n.~} \GE{date, day} \OP{bebie} \TP{bebiiri/bebiiree} 
\LX{bebisaana} \HM{} \PH{bèbísááná} \PS{n.~} \GE{strange day} \OP{bebisaama} 
\LX{bebisoŋ} \HM{} \PH{bèbísʊ́ŋ} \PS{n.~} \GE{good day, holy day} \OP{bebisonne} \TP{bebisonnɛɛ} 
\LX{bebitɔɔre} \HM{} \PH{bèbítɔ̀ɔ̀rɪ́} \PS{n.~} \GE{distant date} \OP{bebitɔbɔ} \TP{bebitɔɔrɛɛ} 
\LX{bebivuo} \HM{} \PH{bèbívúó} \PS{n.~} \GE{free day} \OP{bebivori} \TP{bebivoree} 
\LX{bebiyeni} \HM{} \PH{bèbíyéní} \PS{n.~} \GE{one day, a single day} \OP{bebiyaga} \VA{bebiyenaa}
\LX{bee} \HM{} \PH{bèé} \PS{v.~} \GE{to exist; to be present} \XV{O bee be.} \XE{S/he is there.} \PD{-, -, -, -, -, - } \VA{be; shortened form of be la}
\LX{bee} \HM{} \PH{bɪ́ɪ́} \PS{conj.~} \GE{or} \XV{O gaaɛ la bee o ba gaa?} \XE{Has s/he gone or not?.} \VA{beɛ, baa}
\LX{beɛ} \HM{} \PH{bɪ̀ɛ̀} \PS{v.~} \SN \GE{to move out of the way (allowing another to pass)} \SN \GE{to divert} \SN \GE{to pass someone by, to overtake} \PD{beɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, - } 
\LX{beɛ} \HM{} \PH{bɪ̀ɛ̀} \PS{v.~} \GE{to slice (meat)} \XV{A nɛŋŋmaara beɛ la a nɛne eŋ a doge poɔ.} \XE{The butcher sliced the meat and put it into the pot.} \PD{beɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, - } \VA{beɛle}
\LX{beɛ} \HM{} \PH{bíɛ́} \PS{adj.~} \GE{funny; strange} \OP{beere} \TP{beerɛɛ} \VA{beɛbeɛ}
\LX{bɛɛ} \HM{} \PH{bɛ́ɛ̀} \PS{adv.~} \GE{where (emphatic)} \XV{A libie bɛɛ?} \XE{Where the hell is the money?.} \VA{bɛ}
\LX{bɛɛ} \HM{} \PH{bɛ̀ɛ́ɛ̀} \PS{n.~} \GE{sheep} 
\LX{beɛbeɛ} \HM{} \PH{bɪ́ɛ́bɪ́ɛ́} \PS{adj.~} \GE{strange} \VA{beɛ}
\LX{beɛlaa} \HM{} \PH{bɪ́ɛ́láá} \PS{adj.~} \GE{small, little (quantity, time)} \OP{beɛlɛ} 
\LX{beele} \HM{} \PH{bɪ́ɪ́lɪ̀} \PS{v.~} \GE{to envy (with evil intent)} \PD{beelɛɛ, beelɛ, beelɛ, beelebɛ, beelaa, beelebɛ } 
\LX{beele} \HM{} \PH{bɪ́ɪ́lɪ̀} \PS{v.~} \GE{to move slightly away from a spot} \PD{beelɛɛ, beelɛ, beelɛ, beelebɛ, beelaa, beelebɛ } 
\LX{beɛle} \HM{} \PH{bɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{to accompany} \PD{beɛlɛɛ, beɛlɛ, beɛlɛ, beɛlebɛ, beɛlaa, beɛlebɛ } \VA{beɛre}
\LX{beɛle} \HM{} \PH{bɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{to cut in strips  (meat, hide)} \PD{beɛlɛɛ, beɛlɛ, beɛlɛ, beɛlebɛ, beɛlaa, beɛlebɛ } \VA{beɛre}
\LX{beɛlɛbeɛlɛ} \HM{} \PH{bɪ́ɛ́lɛ́bɪ́ɛ́lɛ́} \PS{adv.~} \GE{slowly; bit by bit} 
\LX{beene} \HM{} \PH{bɪ́ɪ́nɪ́} \PS{n.~} \SN \GE{strong person} \SN \GE{one with supernatural power} \OP{beemɛ} \TP{beemɛɛ} 
\LX{beene} \HM{} \PH{bɪ́ɪ́nɪ́} \PS{adj.~} \SN \GE{well cooked, boiled} \SN \GE{ripe (fruit)} \OP{beemɛ} \TP{beenɛɛ} 
\LX{beene} \HM{} \PH{bɪ̀ɪ̀nɪ̀} \PS{v.~} \SN \GE{to move majestically} \SN \GE{to roam round leisurely} \PD{beenɛɛ, beenɛ, beenɛ, beenemɛ, beenaa, - } 
\LX{bɛɛne} \HM{} \PH{bɛ́ɛ́nɪ́} \PS{n.~} \GE{hemp seeds} 
\LX{beere} \HM{} \PH{bɪ̀ɪ̀rɪ̀} \PS{v.~} \SN \GE{to kill, to finish off, to beat mercilessly} \SN \GE{to boil pito in the final stage of production} \PD{beerɛɛ, beerɛ, beerɛ, beerebɛ, beeraa, - } 
\LX{beɛre} \HM{} \PH{bɪ́ɛ́rɪ̀} \PS{n.~} \GE{older brother} \OP{beɛremine} \TP{beɛrɛɛ} 
\LX{beɛre} \HM{} \PH{bɪ́ɛ́rɪ́} \PS{v.~} \GE{to follow up} \PD{beɛrɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ } 
\LX{beɛre} \HM{} \PH{bɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to cut into strips} \PD{beɛrɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebeɛraa } \VA{beɛle}
\LX{bege} \HM{} \PH{bɪ̀gɪ̀} \PS{v.~} \GE{to stop up, to close a passage} \PD{begɛɛ, begerɛ, begeraa, begerebɛ, begeraa, begerebɛ } 
\LX{bege} \HM{} \PH{bɪ̀gɪ́} \PS{n.~} \GE{law} \OP{begɛ} \TP{begɛɛ} \VA{merɛ (Akan)}
\LX{bɛge} \HM{} \PH{bɛ̀gɪ̀} \PS{v.~} \SN \GE{to embank} \SN \GE{to block with a dam} \PD{bɛgɛɛ, bɛgerɛ, bɛgerɛ, bɛgerebɛ, bɛgeraa, bɛgerebɛ } 
\LX{begebiŋi-duoraa} \HM{} \PH{bɪ̀gɪ̀bíŋ-duoraa} \PS{n.~} \GE{parliament house} \OP{begebiŋi-duori} \TP{begebiŋi-duoree} 
\LX{begebinne} \HM{} \PH{bɪ̀gɪ̀bínné} \PS{n.~} \SN \GE{lawmaker} \SN \GE{member of parliament} \OP{begebinime} \TP{begebinimee} 
\LX{bɛgelaa} \HM{} \PH{bɛ̀gɪ̀làá} \PS{n.~} \GE{a shell of a particular type of seed, used to scrape inside of cooking utensils} \OP{bɛgele} \TP{bɛgelɛɛ} 
\LX{begere} \HM{} \PH{bɪ̀gɪ̀rɪ́} \PS{n.~} \GE{wing} \OP{begɛ} \TP{begerɛɛ} 
\LX{belaa} \HM{} \PH{bɪ́láá} \PS{n.~} \GE{anthill} \OP{belle} \TP{bellɛɛ} \VA{shortened form of bebelaa}
\LX{belaa} \HM{} \PH{bìláà} \PS{adv.~} \GE{interog. what} \XV{Belaa ka fo erɛ?} \XE{What are you doing?.} 
\LX{belaa} \HM{} \PH{bɪ́láá} \PS{adv.~} \SN \GE{small (quantity), few} \XV{A laa poɔ bondirii waa la belaa lɛ.} \XE{The amount of food is very small.} \XV{Zenɛ a tontonema waa la belaa.} \XE{Today the workers are few.} \SN \GE{short while (time)} \XV{Vɛŋ ka belaa ka fo gaa.} \XE{Wait for a short while before you go.} 
\LX{bele} \HM{} \PH{bɪ́lɪ́} \PS{v.~} \GE{to do extraneous work} \PD{belɛɛ, belɛ, belɛ, belbɛ, belaa, - } 
\LX{bele} \HM{} \PH{bɪ́lɪ́} \PS{n.~} \GE{yeast} \OP{belɛ} \TP{belɛɛ} 
\LX{belefele} \HM{} \PH{bɪ̀lɪ̀fɪ̀lɪ̀} \PS{adv.~} \GE{big and fat (person)} 
\LX{bɛlɛmbɛgere} \HM{} \PH{bɛ̀lɛ̀mbɛ́grɪ́} \PS{n.~} \GE{nightengale} \OP{bɛlɛmbɛgɛ} \TP{bɛlɛmbɛgerɛɛ} \VA{bɛlembɛgre}
\LX{bɛlɛnte} \HM{} \PH{bɛ́lɛ́ntɪ̀} \PS{n.~} \GE{belt} \OP{bɛlɛntere} \TP{bɛlɛnteerɛɛ} 
\LX{bɛlle} \HM{} \PH{bɛ̀llɪ̀} \PS{v.~} \GE{to deceive} \PD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa, - } 
\LX{bɛlle} \HM{} \PH{bɛ̀llɪ̀} \PS{v.~} \GE{to coax (a baby)} \PD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa, - } 
\LX{bɛlle} \HM{} \PH{bɛ́llɪ̀} \PS{v.~} \GE{to look at closely, to inspect} \PD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa, - } 
\LX{bɛllebɛllɛ} \HM{} \PH{bɛ̀lbɛllɛ} \PS{n.~} \GE{bean flour shaped into slabs and steamed} \TP{bɛllebɛllɛɛ} \VA{tumpanne}
\LX{bɛmbɛŋ} \HM{} \PH{bɛ́mbɛ́ŋ} \PS{adv.~} \GE{extremely black} \XV{A peroo e la sɔgelaa bɛmbɛŋ.} \XE{The sheep is extremely black.} 
\LX{bɛmbɛŋ} \HM{} \PH{bɛ̀mbɛ̌ŋ} \PS{v.~} \GE{to move in a shaky manner} \PD{bɛmbɛŋɛɛ, bɛmbɛnɛ, bɛmbɛnɛ, bɛmbɛnemɛ, bɛmbɛnaa, - } 
\LX{bɛmbiri} \HM{} \PH{bɛ́mbírì} \PS{n.~} \GE{bean seed} \OP{bɛmbie} \TP{bɛmbiiree} 
\LX{bɛmbugo} \HM{} \PH{bɛ́mbùgó} \PS{n.~} \GE{barn for storing beans} \OP{bɛmbugri} \TP{bɛmbugree} 
\LX{bɛmbuli} \HM{} \PH{bɛ́mbúlí} \PS{n.~} \GE{bean sprout} \OP{bɛmbulo} \TP{bɛmbullee} 
\LX{bɛmpaala} \HM{} \PH{bɛ́mpáálà} \PS{n.~} \GE{new, fresh beans} \TP{bɛmpaalɛɛ} 
\LX{bɛmpɛɛre} \HM{} \PH{bɛ́mpɛ̀ɛ̀rɪ́} \PS{n.~} \GE{unripened bean seed} \OP{bɛmpɛbɛ} \TP{bɛmpɛbɛɛ} 
\LX{bɛmpɛge} \HM{} \PH{bɛ́mpɛ́gɪ́} \PS{n.~} \GE{bean husk} \OP{bɛmpɛgre/bɛmpɛgɛ} \TP{bɛmpɛgrɛɛ} 
\LX{bɛmpelaa} \HM{} \PH{bɛ́mpɪ́láá} \PS{n.~} \GE{white beans} \OP{bɛmpeɛle} \TP{bɛmpeɛlɛɛ} 
\LX{bɛndaa} \HM{} \PH{bɛ́ndàà} \PS{n.~} \GE{bean stalk} \OP{bɛndaare} \TP{bɛndarɛɛ/bɛndaarɛɛ} 
\LX{bɛndaa} \HM{} \PH{bɛ́ndáá} \PS{n.~} \GE{a cooked bean dish (usually not seasoned)} \OP{bɛndaare} \TP{bɛndarɛɛ/bɛndaarɛɛ} 
\LX{bɛŋe} \HM{} \PH{bɛ̂ŋ} \PS{v.~} \SN \GE{to sift, to separate grains from unwanted material} \SN \GE{to skim (usually by tapping water with the back of the hand or a container to push away debris)} \SN \GE{to move unsteadily, to sway} \PD{bɛŋɛɛ, bɛnnɛ, bɛnnɛ, bɛnemɛ, bɛnnaa, bɛnemɛ } \VA{pɛŋe}
\LX{bɛŋe} \HM{} \PH{bɛ̀ŋ} \PS{v.~} \GE{to pretend to invite someone to share something (but not whole-heartedly)} \PD{bɛŋɛɛ, bɛnnɛ, bɛnnɛ, bɛnnaa, bɛnemɛ, - } 
\LX{bɛŋɛ} \HM{} \PH{bɛ́ŋɛ̀} \PS{n.~} \GE{beans} \SG{bɛmmo} \TP{bɛnnɛɛ} 
\LX{benee} \HM{} \PH{bènéé} \PS{adv.~} \GE{that place} 
\LX{bɛŋgbɛre} \HM{} \PH{bɛ̀ŋgbɛ̀rɪ́} \PS{n.~} \GE{a variety of beans} \OP{bɛŋgbɛrre} \TP{bɛŋgbɛrɛɛ} 
\LX{bɛŋkorɔ} \HM{} \PH{bɛ́ŋkʊ́rɔ́} \PS{n.~} \GE{old beans} \TP{bɛŋkorɛɛ} \VA{bɛŋkora}
\LX{bɛŋkuoŋaa} \HM{} \PH{bɛ́ŋkúóŋáá} \PS{n.~} \GE{dry bean} \OP{bɛŋkuoni} \TP{bɛŋkuonee} 
\LX{benne} \HM{} \PH{bɪ̀nnɪ̀} \PS{v.~} \GE{to intrude} \PD{bennɛɛ, bennɛ, bennɛ, benemɛ, bennaa, benemɛ } 
\LX{bennɛ} \HM{} \PH{bɪ́nnɛ́} \PS{n.~} \GE{intruder (especially to meals)} \OP{benemɛ} \TP{benemɛɛ} 
\LX{benni} \HM{} \PH{benni} \PS{v.~} \GE{to doubt} \PD{bennee, benne, benne, benime, bennaa, - } 
\LX{bentere} \HM{} \PH{bɪ̀ntɪ́rɪ̀} \PS{n.~} \GE{a large gourd drum} \OP{benteɛ} \TP{benterɛɛ} \VA{bontere}
\LX{bɛnwɔne} \HM{} \PH{bɛ́nwɔ́nɪ́} \PS{n.~} \GE{the fruit of a bean plant} \OP{bɛnwɔmɔ} \TP{bɛnwɔnnɛɛ} 
\LX{bɛnzɔŋe} \HM{} \PH{bɛ́nzɔ́ŋ} \PS{n.~} \GE{bean flour} \OP{bɛnzɔnne} \TP{bɛnzɔnnɛɛ} 
\LX{bɛnzɔɔre} \HM{} \PH{bɛ́nzɔ̀ɔ̀rɪ́} \PS{n.~} \GE{tender (seedless) fruit of a bean plant} \OP{bɛnzɔbɔ} \TP{bɛnzɔrɛɛ} 
\LX{beo} \HM{} \PH{bɪ̀ʊ̀} \PS{adv.~} \GE{big and stagnant} 
\LX{bɛraa} \HM{} \PH{bɛ́ráá} \PS{n.~} \SN \GE{trap, snare} \SN \GE{bait (fishhook)} \SN \GE{poison} \OP{bɛrre} \TP{bɛrɛɛ} 
\LX{bɛraa} \HM{} \PH{bɛ̀ràá} \PS{n.~} \GE{razor} \OP{bɛrre} \TP{bɛrrɛɛ} 
\LX{bere} \HM{} \PH{bɪ́rɪ́} \PS{n.~} \SN \GE{a type of small plant whose leaves and flowers are used as vegetables} \SN \GE{the leaves of this plant} \TP{berrɛɛ} \VA{berenɛmɛ}
\LX{bere} \HM{} \PH{bɪ̀rɪ́} \PS{n.~} \SN \GE{breast} \OP{berɛ} \TP{berɛɛ} 
\LX{bere} \HM{} \PH{bɪ̀rɪ̀} \PS{v.~} \GE{protect by supernatural means} \PD{berɛɛ, berɛ, berɛ, berebɛ, beraa, beremɛ } 
\LX{bere} \HM{} \PH{bɪ̀rɪ̀} \PS{adj.~} \GE{powerful (supernaturally)} 
\LX{berɛ} \HM{} \PH{bìrɛ́} \PS{v.~} \SN \GE{to wish} \XV{O berɛ ŋa zie da koŋ nyaa.} \XE{S/he wished day would never come.} \XV{O berɛ ona la da na gaa a tigiri.} \XE{S/he wished s/he would be the one to attend the feast.} \SN \GE{Before he realized it, a stranger arrived} \PD{-, -, -, -, -, - } 
\LX{bɛre} \HM{} \PH{bɛ̀rɪ̀} \PS{v.~} \SN \GE{to set (a trap)} \SN \GE{to poison} \XV{Dabe᷉ɛ deme maŋ boɔrɔ ka ba bɛre ba dendɔmɔ ko.} \XE{Cowards always try to kill their enemies by poisoning them.} \PD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ } 
\LX{bɛre} \HM{} \PH{bɛ̀rɪ̀} \PS{v.~} \GE{to ponder, to wonder} \XV{O da di la a bondirii tasoga leɛ zeŋe bɛre.} \XE{After eating for some time, she sat back and thought.} \PD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ } 
\LX{bɛre} \HM{} \PH{bɛ́rɪ̀} \PS{v.~} \SN \GE{to exclude, to separate from the herd} \XV{A naakyeenɛ bɛre la naadɔgerɔ na bare.} \XE{The cattle herder has separated the cow that had recently calved from the herd.} \SN \GE{to detach or cut off (a large piece)} \XV{A naŋkpaana bɛre la a naabo gbɛre koɔre.} \XE{The butcher has cut off and sold the thigh of the cow.} \PD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ } 
\LX{berebaaraa} \HM{} \PH{bɪ̀rɪ̀bááráá} \PS{n.~} \GE{last born (of humans and other mammals)} \OP{berebaare} 
\LX{berekaroo} \HM{} \PH{bɪ̀rìkàrʊ́ʊ́} \PS{n.~} \SN \GE{a type of disease of the breast} \SN \GE{a type of charm believed to inflict this disease} \OP{berekarre} \TP{berekarɛɛ} 
\LX{berekpulo} \HM{} \PH{bɪ̀rɪ̀kpúló} \PS{n.~} \GE{hard, firm breasts (of a young woman)} \TP{berkpulee} 
\LX{berekuuri} \HM{} \PH{bìrɪ̀kúúrí} \PS{n.~} \GE{thick centre of the breast} \SG{berekuuree} \OP{berekubo} 
\LX{beremɛ} \HM{} \PH{bɪ́rɪ́mɛ́} \PS{n.~} \GE{one who possesses supernatural powers} \TP{beremɛɛ} 
\LX{berenɛmɛ} \HM{} \PH{bɪ̀rɪ̀nɛ́mɛ́} \PS{n.~} \GE{the flower of the bere plant} \SG{berenɛnoo} \TP{berenɛmɛɛ} \VA{bere}
\LX{berenoɔ} \HM{} \PH{bɪ̀rɪ̀nʊ́ɔ̀} \PS{n.~} \GE{a bird used in rituals to induce breast milk in a nursing mother} \OP{berenoore} \TP{berenoorɛɛ} 
\LX{berenoɔre} \HM{} \PH{bɪ̀rɪ̀nʊ́ɔ́rɪ̀} \PS{n.~} \GE{nipple (of a breast)} \OP{berenɔɛ} \TP{berenɔrɛɛ} 
\LX{bɛren-uri} \HM{} \PH{bɛ̀rɪ̀n-ùrí} \PS{n.~} \GE{pollen from the kenaf plant (a fibre-producing plant similar to hibiscus cannabinus) which causes itching} \TP{bɛren-uree} \VA{bɛren-uri}
\LX{berese} \HM{} \PH{bɪ̀rɪ̀sɪ̊̀} \PS{adv.~} \GE{extremely large and fat} 
\LX{berewoɔ} \HM{} \PH{bɪ̀rɪ̀wʊ́ɔ́} \PS{n.~} \GE{slang term for female breast} \OP{berewoore} \TP{berewoorɛɛ} 
\LX{bereyɛre} \HM{} \PH{bɪ̀rɪ̀yɛ́rɪ́} \PS{n.~} \GE{a type of wild plant that is used as a vegetable} \TP{bereyɛrɛɛ} 
\LX{beroŋ} \HM{} \PH{bɪ̀rʊ́ŋ} \PS{n.~} \GE{milk} 
\LX{bɛroŋ} \HM{} \PH{bɛ́rʊ́ŋ} \PS{adj.~} \GE{fat, thick} \OP{bɛrɛ} \TP{bɛrɛɛ/bɛrrɛɛ} 
\LX{berondɔndɔre} \HM{} \PH{bɪ̀rʊ́ndɔ̀ndɔ̀rɪ́} \PS{n.~} \GE{curdled milk} \TP{berondɔndɔrɛɛ} 
\LX{beroŋka᷈a} \HM{} \PH{bɪ́rʊ́ŋká᷈à᷈} \PS{n.~} \GE{fat from fresh milk; butter} \SG{beroŋkaamo} \TP{beroŋkaanɛɛ} 
\LX{berroŋ} \HM{} \PH{bɪ̀rrʊ́ŋ} \PS{n.~} \GE{the state of being supernaturally powerful} 
\LX{be᷉e} \HM{} \PH{bɪ́᷉ɪ́᷉} \PS{n.~} \SN \GE{concentrated extraction from something (tasty, interesting)} \SN \GE{broth from meat} \TP{beenɛɛ} 
\LX{be᷉ɛ} \HM{} \PH{bɪ́᷉ɛ̀᷉} \PS{n.~} \GE{beer} \SG{beɛnoo} \TP{beɛnɛɛ} 
\LX{be᷉ɛo} \HM{} \PH{bɪ̀᷉ɛ̀᷉ʊ́᷉} \PS{n.~} \GE{multitude, crowd} \OP{beɛnɛɛ} 
\LX{bibile} \HM{} \PH{bìbílé} \PS{n.~} \GE{small child} \OP{bibilii} 
\LX{bibile-deɛnaa} \HM{} \PH{bìbílédɪ́ɛ́náá} \PS{n.~} \GE{toy} \OP{bibiledeɛne} \TP{bibiledeɛnɛɛ} 
\LX{bibilpɔɔ} \HM{} \PH{bìbìlpɔ̀ɔ́} \PS{n.~} \GE{young child} \OP{bibilpɔɔre} \TP{bibilpɔɔrɛɛ} 
\LX{bibore} \HM{} \PH{bíbʊ̀rɪ̀} \PS{n.~} \GE{the act of playing chess} 
\LX{bidɛndɛŋe} \HM{} \PH{bìdɛ̀ndɛ́ŋ} \PS{n.~} \GE{firstborn (human)} \OP{bidɛndɛnne} \TP{bidɛndɛnnɛɛ} \VA{bidɛŋ, bidɛnne, bidɛnnɛɛ}
\LX{bidɛŋe} \HM{} \PH{bìdɛ́ŋ} \PS{n.~} \GE{firstborn (human)} \OP{bidɛnne} \TP{bidɛnnɛɛ} \VA{shortened form of bidɛndɛŋ}
\LX{bidɔŋe} \HM{} \PH{bìdɔ̂ŋ} \PS{n.~} \GE{sibling (clan level)} \OP{bidɔnne} \TP{bidɔnnɛɛ} 
\LX{bidɔɔ} \HM{} \PH{bìdɔ́ɔ́} \PS{n.~} \GE{son} \OP{bidɔbɔ} \TP{bidɔɔrɛɛ} 
\LX{bie} \HM{} \PH{bíé} \PS{n.~} \GE{child} \OP{biiri} \TP{biiree} 
\LX{bie} \HM{} \PH{bíé} \PS{n.~} \GE{day, date} \XV{Fo sori na yiibu bie la dabuo?} \XE{What date have you set for your journey?.} \OP{biiri} \TP{biiree} 
\LX{bie} \HM{} \PH{bíè} \PS{n.~} \GE{a type of game played using marbles} \OP{biree/biiree} 
\LX{bie} \HM{} \PH{bíé} \PS{adv.~} \GE{tomorrow} \VA{bieo}
\LX{bie} \HM{} \PH{bìè} \PS{v.~} \GE{to come to pass} \XV{A pɔlbile na de la pɔge, a bebiri la naŋ ba bie.} \XE{The young man will marry, it just has not happened yet.} \PD{-, biere, biere, bieribe, bieraa, bieribe } 
\LX{biebie} \HM{} \PH{bíébíé} \PS{adv.~} \GE{daily; every day} 
\LX{bieo} \HM{} \PH{bíèó} \PS{adv.~} \GE{tomorrow} \XV{Bieo, ba na wa la.} \XE{Tomorrow, they will come.} \VA{bie}
\LX{bieri} \HM{} \PH{bíérí} \PS{n.~} \GE{stinginess, unwillingness to share or open up} 
\LX{bifaa} \HM{} \PH{bìfáá} \PS{n.~} \GE{bad child} \OP{bifaare} \TP{bifaarɛɛ} 
\LX{bifaa} \HM{} \PH{bífáá} \PS{n.~} \GE{bad pebble} \OP{bifaare} \TP{bifaarɛɛ} 
\LX{bife᷉e} \HM{} \PH{bífɪ́᷉ɪ́᷉} \PS{n.~} \GE{small quantity of pebbles} \OP{bifeene} \TP{bifeenɛɛ} 
\LX{bifo᷉ɔo} \HM{} \PH{bífʊ́᷉ɔ́᷉ʊ́᷉} \PS{n.~} \GE{light pebble} \OP{bifoɔne} \TP{bifoɔnɛɛ} 
\LX{bigaalaa} \HM{} \PH{bìgáálàá} \PS{n.~} \GE{any bedding for a child} \OP{bigaale} 
\LX{bigaale} \HM{} \PH{bìgààlɪ̀} \PS{n.~} \GE{postnatal care} \OP{bigaale ̪[bìgáálɪ́]} 
\LX{bigane} \HM{} \PH{bìgánɪ̀} \PS{n.~} \GE{leather spread for a baby cot} \OP{bigama} \TP{bigamɛɛ} 
\LX{bigeŋ} \HM{} \PH{bìgèŋ} \PS{adv.~} \GE{all over (area)} 
\LX{bigeŋ} \HM{} \PH{bìgèŋ} \PS{adv.~} \GE{big heavy sound (fall of a big wall, gunshot)} 
\LX{bigileŋ} \HM{} \PH{̪bìgìlèŋ} \PS{adj.~} \GE{large and heavy} \XV{A kuuri waa la bigileŋ lɛ.} \XE{The stone is large and heavy.} 
\LX{bigiri} \HM{} \PH{bígírí} \PS{n.~} \SN \GE{fibre} \SN \GE{kenaf; a plant that produces fibre} \OP{bige} \TP{bigee} 
\LX{bigiri} \HM{} \PH{bígírí} \PS{adj.~} \GE{fibre-like} \XV{A maŋgo deɛ waa la bigiri.} \XE{The mango is fibrous.} 
\LX{bigiri} \HM{} \PH{bìgìrì} \PS{v.~} \GE{to describe} \PD{bigiree, bigire, bigire, bigiribe, bigiraa, bigiribe } 
\LX{biiree} \HM{} \PH{bììréé} \PS{n.~} \GE{sand} \OP{biiri} \TP{biiree} 
\LX{biiridalempoɔ} \HM{} \PH{bììrìdàlɪ̀mpʊ̀ɔ̀} \PS{n.~} \GE{desert} \OP{biiridalempoore} \TP{biiridalemporɛɛ} \VA{biirikpoŋ}
\LX{biirikpoŋi} \HM{} \PH{bíírìkpôŋ} \PS{n.~} \GE{desert} \OP{biirikponni} \TP{biirikponnee} \VA{biiridalempoɔ}
\LX{bikaa} \HM{} \PH{bìkáà} \PS{n.~} \GE{the act of caring for children} 
\LX{bikaale} \HM{} \PH{bíkààlɪ̀} \PS{n.~} \GE{the act of gathering pebbles} 
\LX{bikaara} \HM{} \PH{bìkáárá} \PS{n.~} \GE{caretaker of a child} \OP{bikaareba} \TP{bikarebɛɛ} 
\LX{biko} \HM{} \PH{bìkʊ́} \PS{n.~} \GE{the act of killing children} 
\LX{bikoŋkonaa} \HM{} \PH{bìkóŋkónáá} \PS{n.~} \GE{a child that cries a lot} \OP{bikoŋkonni} \TP{bikoŋkonnee} 
\LX{bikoɔraa} \HM{} \PH{bìkʊ́ɔ́ráá} \PS{n.~} \GE{industrious person} \OP{bikoɔre} 
\LX{bikore} \HM{} \PH{bìkʊ́rɪ́} \PS{n.~} \GE{old chap, child} \OP{bikorɔ/bikoreba} \TP{bikorɛɛ} 
\LX{bikpeebɛ} \HM{} \PH{bìkpɪ́ɪ́bɛ́} \PS{n.~} \GE{orphan} \OP{bikpeere/bikpeemine} \TP{bikpeerɛɛ} 
\LX{biku᷉u} \HM{1} \PH{bìkú᷉ù᷉} \PS{n.~} \GE{baby cot} \OP{bikuuni} \TP{bikuunee} 
\LX{biku᷉u} \HM{2} \PH{bìkú᷉ú᷉} \PS{n.~} \GE{dead child} \OP{bikuuni} \TP{bikuunee} 
\LX{bikyelaa} \HM{} \PH{bìʧɪ́láà} \PS{n.~} \GE{smart child} \OP{bikelle} \TP{bikyelɛɛ} 
\LX{bikyɛnaa} \HM{} \PH{bìʧɛ́náá} \PS{n.~} \GE{toddler} \OP{bikyɛnne/bikyɛnemɛ} \TP{bikyɛnemɛɛ} 
\LX{bildɛŋe} \HM{} \PH{bìldɛ́ŋ} \PS{n.~} \GE{firstborn (nonhuman)} \OP{bildɛnne} \TP{bildɛnnɛɛ} 
\LX{bile} \HM{} \PH{bile} \PS{adj.~} \SN \GE{small} \XV{A kparoo waa la bile; o koŋ kpɛ a bie.} \XE{The shirt is small; it won't fit the child.} \SN \GE{younger} \OP{bilii} 
\LX{bile} \HM{} \PH{bílé} \PS{n.~} \GE{offspring (nonhuman)} \OP{bilii} \TP{bilee} 
\LX{bileɛ} \HM{} \PH{bìlɪ́ɛ́} \PS{n.~} \GE{young baby} \OP{bileɛre} \TP{bileɛrɛɛ} \VA{bilɛɛ}
\LX{bilɛɛ} \HM{} \PH{bìlìɛ̀} \PS{n.~} \GE{baby (newly born)} \OP{bilɛɛre} \TP{bilɛɛrɛɛ} \VA{bileɛ}
\LX{bilembileŋ} \HM{} \PH{bílémbíléŋ} \PS{adj.~} \GE{round and smooth} 
\LX{bilembileŋ} \HM{} \PH{bìlèmbíléŋ} \PS{n.~} \GE{caterpillar track, continous track} \OP{bilembilenni} \TP{bilembilennee} 
\LX{bilembileŋ} \HM{} \PH{bìlèmbíléŋ} \PS{adj.~} \GE{large and heavy, with weight distributed unevenly} 
\LX{bilentimm} \HM{} \PH{bìlèntìmm} \PS{adj.~} \GE{bloated and large} 
\LX{bilibala} \HM{} \PH{bìlìbálá} \PS{adv.~} \GE{grabbing on to a huge object by throwing both arms around it} \XV{O le nyɔge la a bombɛroŋ na lɛ bilibala.} \XE{S/he firmly grabbed the fat one.} 
\LX{billi} \HM{} \PH{bìllì} \PS{v.~} \GE{to roll (a heavy object along the ground)} \PD{billee, bille, bille, bilibe, billaa, - } 
\LX{billi} \HM{} \PH{bíllí} \PS{v.~} \GE{to roll between the palms; to mould} \PD{billee, bille, bille, bilibe, billaa, - } 
\LX{bimaraa} \HM{} \PH{bìmáráá} \PS{n.~} \GE{baby sling} \OP{bimarre} \VA{maraa}
\LX{bimbiŋ} \HM{} \PH{bímbíŋ} \PS{adv.~} \GE{extremely dark (rain clouds)} \XV{A saa koe la bimbiŋ lɛ.} \XE{The rain clouds have gotten awfully thick.} 
\LX{bimbiŋe} \HM{} \PH{bìmbìŋé} \PS{n.~} \GE{pen, poultry shed} \OP{bimbinni} \TP{bimbinnee} \VA{nɔzage}
\LX{bimeŋɛ} \HM{} \PH{bìmɪ́ŋɛ́} \PS{n.~} \GE{good child} \OP{bimenne} \TP{bimennɛɛ} 
\LX{bimm} \HM{} \PH{bìmm} \PS{adv.~} \GE{large and solid} \XV{A bugo e la kpoŋ bimm lɛ.} \XE{The granary is big and solid.} 
\LX{bimm} \HM{} \PH{} \PS{adj.~} \GE{solid} \XV{A taabo daa waa la bimm lɛ.} \XE{The timber log is solid.} 
\LX{bimpeɛle} \HM{} \PH{bìmpɪ́ɛ́lɪ́} \PS{n.~} \SN \GE{white faeces} \SN \GE{fear (used with nyɛ...)} \XV{Ba toŋ yeli ka n pɔge ma beɛrɛ la ka o deɛ yɔ nyeɛrɛ bimpeɛle.} \XE{A message was sent to inform my wife that her mother was sick and she is in utter distress and fear.} \TP{bimpeɛlɛɛ} 
\LX{bimporɔ} \HM{} \PH{bìmpʊ́rɔ́} \PS{n.~} \GE{flatulence} \OP{bimporre} \TP{bimporɛɛ/bimporrɛɛ} \VA{bimpora}
\LX{bin-eɛraa} \HM{} \PH{bìn-ɪ́ɛ́ráá} \PS{n.~} \GE{cloth used for cleaning faeces (of a baby)} \OP{bin-eɛre} \TP{bin-eɛrɛɛ} 
\LX{bini} \HM{} \PH{bíní} \PS{n.~} \GE{excrement} \OP{n/a} \TP{binee} 
\LX{biŋi} \HM{} \PH{bìŋ} \PS{v.~} \SN \GE{to put down, to place, to set out} \SN \GE{to store (up)} \PD{biŋee, binne, binne, binime, binnaa, binime } 
\LX{biŋibiŋi} \HM{} \PH{bìŋìbíŋí} \PS{v.~} \GE{to shift unsteadily, teeter} \XV{A tanzuŋ waa la bile, ka a doge biŋibiŋine.} \XE{The pot holder is small and so the pot is moving unsteadily.} \PD{biŋibiŋee, biŋibiŋine, biŋibiŋine, biŋibiŋinime, biŋibiŋinaa, biŋibiŋinime } 
\LX{biŋkɔnnyeɛrɛ} \HM{} \PH{bìŋkɔ́nɲɪ́ɛ́rɛ́} \PS{n.~} \GE{one suffering from diarrhoea} \OP{biŋkɔnnyeɛrebɛ} \TP{biŋkɔnnyeɛrebɛɛ} 
\LX{biŋko᷉ɔ} \HM{} \PH{bǐŋkʊ́᷉ɔ́᷉} \PS{n.~} \GE{watery faeces, diarrhoea} \TP{biŋkoɔnɛɛ} 
\LX{binkyoo} \HM{} \PH{bìnʧóó} \PS{n.~} \GE{open place full of defecation} \OP{binkyoori} \TP{binkyooree} 
\LX{binnyeɛbo} \HM{} \PH{bìnɲɛɪ́ɛ́bʊ́} \PS{n.~} \GE{act of misbehavior} 
\LX{binɔnaa} \HM{} \PH{bìnɔ̀náá} \PS{n.~} \GE{beloved child; favorite} \OP{binɔnne} 
\LX{binɔŋe} \HM{} \PH{bìnɔ̀ŋ} \PS{n.~} \GE{the act of loving children; the act of pampering} \OP{binɔnema} 
\LX{bintanyeɛrɛ} \HM{} \PH{bìntàɲɪ́ɛ́rɛ́} \PS{n.~} \GE{a type of diarrhoea} \OP{bintanyeɛre} \TP{bintanyeɛrɛɛ} 
\LX{binze᷉e} \HM{} \PH{bìnzɪ́᷉ɪ́} \PS{n.~} \GE{dysentery} \TP{binzeenɛɛ} 
\LX{bipeɛ} \HM{} \PH{bìpɪ́ɛ́} \PS{n.~} \GE{baby cot} \OP{bipɛre} \TP{bipɛrɛɛ} \VA{biku᷉u}
\LX{bipɔlbile} \HM{} \PH{bìpɔ́lbílé} \PS{n.~} \GE{a lad; young man} 
\LX{bipɔlee} \HM{} \PH{bìpɔ́lɪ́ɪ́} \PS{n.~} \GE{young man, lad} \OP{bipɔlɔ} \TP{bipɔlɛɛ/bipɔllɛɛ} \VA{bipɔlee, bipɔlbile}
\LX{bipɔllee} \HM{} \PH{bìpɔ́lléé} \PS{n.~} \GE{young man} \OP{bipɔlɔ} \TP{bipɔlɛɛ} \VA{bipɔlee, bipɔlbile}
\LX{bipoɔnaa} \HM{} \PH{bìpʊ́ɔ́náá} \PS{n.~} \GE{pampered child} \OP{bipoɔne} \TP{bipoɔnɛɛ} 
\LX{bipuulee} \HM{} \PH{bìpúúléé} \PS{n.~} \GE{young woman} \OP{bipuuli} 
\LX{biraa} \HM{} \PH{bíráá} \PS{n.~} \GE{one who stammers} \OP{biribe} \TP{biribee} 
\LX{biri} \HM{} \PH{bírí} \PS{v.~} \GE{cause debris to drop (down from a height)} \XV{Gyaŋgbonni biri la tɛne a dasɛrɛɛ poɔ eŋe a ko᷉ɔ poɔ.} \XE{The mice broke debris from the ceiling into the water.} \PD{biree, bire, bire, biribe, biraa, biribe } 
\LX{biri} \HM{} \PH{bírì} \PS{v.~} \GE{to stutter, to stammer} \PD{biree, bire, bire, biribe, biraa, - } 
\LX{biri} \HM{} \PH{bìrì} \PS{v.~} \GE{to pass by narrowly, to slip past} \XV{O deɛ biri la a taŋa kyɛ le.} \XE{S/he just passed the hill a little and fell.} \PD{biree, bire, bire, biribe, biraa, biribe } 
\LX{biri} \HM{1} \PH{bírì} \PS{n.~} \GE{chaff from malt} \OP{biree} 
\LX{biri} \HM{2} \PH{bírì} \PS{n.~} \GE{seed, nut} \OP{bie} \TP{biiree} 
\LX{biri} \HM{3} \PH{bírì} \PS{n.~} \GE{something of worth} \XV{Te zenɛ toma ŋa eŋ yi la biri.} \XE{Our work today has been fruitful.} 
\LX{biribiri} \HM{} \PH{bìrìbìrì} \PS{adv.~} \GE{rushing in large numbers e.g. in a stampede} 
\LX{biribiri} \HM{} \PH{bíríbírí} \PS{adv.~} \SN \GE{completely, totally} \XV{O zagerɛɛ biribiri ka o koŋ gaa.} \XE{He has completely refused to go.} \SN \GE{completely black} 
\LX{biribiri} \HM{} \PH{bìrìbírí} \PS{v.~} \GE{to slip, to not fit in (joint)} \XV{A sori na taa la kusiwara te zuo ka fo maŋ kyɛnɛ ka fo gbɛɛ bibire.} \XE{That road has too much gravel, so your feet slip around as you walk on it.} \PD{biribiree, biribire, biribire, biribiribe, biribiraa, - } 
\LX{birikyi} \HM{} \PH{bìríʧì} \PS{n.~} \SN \GE{brake} \SN \GE{brick} \OP{birikyiri} \TP{birikyiree} \VA{birisi}
\LX{birikyilaa} \HM{} \PH{bìríʧìláá} \PS{n.~} \GE{mould (for bricks)} \OP{birikyilaa} \TP{birikyilaarɛɛ} 
\LX{birikyisaabo} \HM{} \PH{bìríʧìsáábʊ́} \PS{n.~} \GE{brake pad} \TP{birikyisaarɛɛ} 
\LX{birisi} \HM{} \PH{bìrísì} \PS{n.~} \GE{brick} \SG{birisiruu} \OP{birisiri} \TP{birisiree} \VA{birikyi}
\LX{biriyokpekperaa} \HM{} \PH{bìríyǒkpɪ́kpìráà} \PS{n.~} \GE{wild duck-like water bird} \OP{biriyokpekperre} \TP{biriyokpekperɛɛ} 
\LX{biriyonaameɛo} \HM{} \PH{bíríyònààmɪ̀ɛ́ʊ} \PS{n.~} \GE{very large duck} \OP{biriyonaameɛne} \TP{biriyonaameɛnɛɛ} 
\LX{biriyuo} \HM{} \PH{bìríyúó} \PS{n.~} \GE{duck} \SG{biriyoruu} \OP{biriyori/biriyoe} \TP{biriyoree} \VA{biriyori}
\LX{bisaa} \HM{} \PH{bìsàá} \PS{n.~} \GE{father (human)} \OP{bisaamine} \TP{bisaarɛɛ} 
\LX{bisoŋ} \HM{} \PH{bìsʊ́ŋ} \PS{n.~} \GE{good child, well-behaved child} \OP{bisonne} \TP{bisonnɛɛ} 
\LX{bitɔgelɔ} \HM{} \PH{bìtɔ́glɔ̀} \PS{n.~} \GE{midwife} \OP{bitɔgelebɔ/bitɔgeleba} \TP{bitɔgelebɛɛ} \VA{bitɔgela}
\LX{bitoroo} \HM{} \PH{bìtòróò} \PS{n.~} \GE{virgin} \OP{bitorri} \TP{bitoree} 
\LX{bituuro} \HM{} \PH{bìtúúró} \PS{n.~} \GE{baby adopter, nanny} \OP{bituuribo} \TP{bituuribee} 
\LX{biweɛmɛ} \HM{} \PH{bìwɪ́ɛ́mɛ́} \PS{n.~} \GE{strayed or wayward child} \OP{biweɛne} \TP{biweɛnɛɛ} 
\LX{biweɛraa} \HM{} \PH{bìwɪ́ɛ́ráá} \PS{n.~} \GE{spoilt child} \OP{biweɛre} 
\LX{biwolaa} \HM{} \PH{bìwʊ́láá} \PS{n.~} \SN \GE{problem child} \SN \GE{beloved child} \OP{biwolle} \TP{biwolɛɛ} 
\LX{biyaalaa} \HM{} \PH{bìyááláá} \PS{n.~} \GE{babysitter} \OP{biyaalba} \TP{biyaalbɛɛ} 
\LX{biyaale} \HM{} \PH{bìyáálɪ́} \PS{n.~} \GE{the act of babysitting} 
\LX{biyaaleyieluŋ} \HM{} \PH{bìyáálɪ̀yíélúŋ} \PS{n.~} \GE{lullaby} \OP{biyaaleyiele} \VA{biyagleyieluŋ}
\LX{biyoɔraa} \HM{} \PH{bìyʊ́ɔ́ráá} \PS{n.~} \GE{well travelled person; one who roams a lot} \OP{biyoɔre} \TP{biyoɔrɛɛ} 
\LX{biyopore} \HM{} \PH{bìyópʊ̀rɪ̀} \PS{} \GE{naming ceremony (baby)} 
\LX{biyuori} \HM{} \PH{bìyúórí} \PS{n.~} \GE{popular person} \OP{biyoe} \TP{biyoree} 
\LX{bizɛgɛ} \HM{} \PH{bìzɛ̀gɛ́} \PS{n.~} \GE{young baby} \OP{bizɛgre} \TP{bizɛgɛɛ} 
\LX{bizɔnzɔɔ} \HM{} \PH{bìzɔ̀nzɔ́ɔ́} \PS{n.~} \GE{young baby} \OP{bizɔnzɔɔbɔ} \TP{bizɔnzɔɔrɛɛ} 
\LX{bo} \HM{} \PH{bʋ́} \PS{v.~} \GE{to gather momentum (rain seen coming from afar in full scale)} \PD{boɛ, boorɔ, boorɔ, booreba, booraa, booreba } 
\LX{bo} \HM{} \PH{bʊ̀} \PS{v.~} \SN \GE{to confer, to deliberate, to plan together} \XV{Ba na bo la sɛre kyɛ gaa a kɔɔto.} \XE{They will confer before they go to court.} \SN \GE{to suppose} \XV{Fo bo ka boŋ ka a saana na taa wa ne?} \XE{What do you suppose the stranger will bring along?.} \PD{boɛ, boorɔ, boorɔ, booreba, booraa, booreba } 
\LX{bɔ} \HM{} \PH{bɔ} \PS{v.~} \SN \GE{to search; to look for} \XV{N bie bɔrɛɛ la soŋ ma ka N bɔ.} \XE{My child is lost; help me look for her.} \SN \GE{to court, to woo} \XV{Dagaaba maŋ bɔ la pɔge kyɛ de.} \XE{The Dagaaba usally court a woman before marriage.} \SN \GE{to engage in sexual intercourse, to make love} \XV{Ka fooŋ bɔ neɛ pɔge ba na ŋmɛ fo la sane yaga.} \XE{If you make love to a married woman, you will be heavily fined.} \PD{bɔɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ } 
\LX{bɔ} \HM{} \PH{bɔ̀} \PS{v.~} \GE{to clean a very dirty place or thing} \PD{bɔɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ } 
\LX{bobɛllɛ} \HM{} \PH{bʊ̀bɛ́llɛ́} \PS{n.~} \GE{a type of carnivorous bird} \OP{bobɛlle} 
\LX{bɔbɔe} \HM{} \PH{bɔ́bɔ́ɪ̀} \PS{n.~} \GE{duck} \OP{bɔbɔɛ} \TP{bɔbɔrɛɛ} 
\LX{bobɔge} \HM{} \PH{bʊ̀bɔ̀gɪ́} \PS{n.~} \GE{huge carnivorous bird} \OP{bobɔgre} \TP{bobɔgrɛɛ} 
\LX{bɔbɔlaa} \HM{} \PH{bɔ́bɔ́láá} \PS{n.~} \GE{a wild animal like a dog} \OP{bɔbɔlle} \TP{bɔbɔlɛɛ} 
\LX{boboo} \HM{} \PH{bʊ̀bʊ́ʊ́} \PS{adj.~} \GE{extremely voluminous} 
\LX{boboruu} \HM{} \PH{bóbórúú} \PS{n.~} \GE{a type of termite} \OP{boborri/boboree} \TP{boborree} 
\LX{bodaa} \HM{} \PH{bʊ̀dáá} \PS{n.~} \GE{male goat} \OP{bodaare} \TP{bodaarɛɛ/bodaɛ} 
\LX{bodaa-ansiinee} \HM{} \PH{bʊ̀dáá-ánsíínéé} \PS{n.~} \GE{a type of tree with edible fruit} \OP{bodaa-ansiini} \VA{boraa-ansiinee}
\LX{bodaanonnoonaa} \HM{} \PH{bʊ̀dáánʊ́nnʊ́ʊ́náá} \PS{n.~} \GE{a type of tree with edible fruit} \OP{bodaanonnoone} \TP{bodaanonnoonɛɛ} \VA{bononnoonaa}
\LX{bodaasensɔɔre} \HM{} \PH{bʊ̀dáásɪ̌nsɔ́ɔ́rɪ́} \PS{n.~} \GE{wild tigernut (not edible)} \OP{bodaasensɔbɔ} \TP{bodaasensɔɔrɛɛ} 
\LX{bodalane} \HM{} \PH{bʊ́dàlánɪ́} \PS{n.~} \GE{male goat (uncastrated)} \OP{bodalama} \TP{bodalamɛɛ} 
\LX{bodalee} \HM{} \PH{bʊ̀dàléé} \PS{n.~} \GE{ young male goat} \OP{bodabilii .} \TP{bodabilee} 
\LX{bodateɛne} \HM{} \PH{bʊ̀dàtɪ́ɛ́nɪ́} \PS{n.~} \GE{male goat with long hair on the chin; bearded male goat} 
\LX{bodavaraa} \HM{} \PH{bʊ̀dàváràà} \PS{n.~} \GE{castrated male goat} \OP{bodavarre} \TP{bodavarɛɛ/bodavarrɛɛ} 
\LX{bɔe} \HM{} \PH{bɔ̀ɪ́} \PS{n.~} \GE{servant} \OP{bɔemine} 
\LX{bɔge} \HM{} \PH{bɔ́gɪ́} \PS{v.~} \GE{to grind into smooth flour} \PD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \VA{bage}
\LX{bɔge} \HM{} \PH{bɔ́gɪ́} \PS{v.~} \SN \GE{to hit with many arrows; to sting or bite many times (insects)} \SN \GE{to apply poisonous lotion to arrows} \PD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \VA{bage}
\LX{bɔge} \HM{} \PH{bɔ́gɪ́} \PS{v.~} \GE{to put a fence around something} \PD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \VA{bage}
\LX{bɔge} \HM{} \PH{bɔ́gɪ́} \PS{n.~} \SN \GE{shoulder} \SN \GE{fore limb of an animal} \OP{bɔgere} \TP{bɔgerɛɛ} \VA{bɔga/baga}
\LX{bɔgebie} \HM{} \PH{bɔ̀gìbíé} \PS{n.~} \GE{misfortune} \OP{bɔgebiiri} \TP{bɔgebiiree} \VA{bɔɔbie/bagebie}
\LX{bɔgebiesoba} \HM{} \PH{bɔ̀gɪ̀bíésʋ́bá} \PS{n.~} \GE{unfortunate person; unlucky person} \OP{ bɔgebiedeme} \VA{bagebiesoba}
\LX{bɔgebogerɔ} \HM{} \PH{bɔ̀gɪ̀bʊ́grɔ́} \PS{n.~} \GE{sorcerer, soothsayer} \OP{bɔgebogerebɔ} \TP{bɔgebogerebɛɛ} \VA{bagebogera}
\LX{bɔgedie} \HM{} \PH{bɔ̀gɪ̀dìè} \PS{n.~} \GE{room for divining} \OP{bɔgederi} \TP{bɔgederee} 
\LX{bɔgele} \HM{} \PH{bɔ̀glɪ̀} \PS{v.~} \SN \GE{to stalk (hunting)} \XV{Naŋkpaana zaa maŋ bɔgele la wɛdoŋɔ a pɔge o ŋmɛ.} \XE{A hunter will normally stalk a wild animal to get close enough to shoot.} \SN \GE{to feel, to move one's hand over} \XV{A zɔŋɔ bɔgele la a gado zu te nyɛ a o kparoo.} \XE{The blind man felt along the bed and found his shirt.} \PD{bɔgelɛɛ, bɔgelɔ, bɔgelɔ, bɔgeleba, bɔgelaa, - } 
\LX{bɔgemaale} \HM{} \PH{bɔ̀gɪ̀mààlɪ̀} \PS{n.~} \GE{the act of making sacrifice} \OP{bɔgemaaleba} \TP{bɔgemaalbɛɛ} \VA{bagemaale}
\LX{bɔgeŋmampegere} \HM{} \PH{bɔ̀gɪ̀ŋmǎmpɪ́gɪ́rɪ́} \PS{n.~} \GE{shoulder} \VA{bageŋmampere}
\LX{bɔgeŋmampere} \HM{} \PH{bɔ̀gɪ̀ŋmàmpɪ́rɪ́} \PS{n.~} \GE{shoulder blade} \OP{bɔgeŋmamperre} \TP{bɔgeŋmamperɛɛ} \VA{bɔgeŋmampegere}
\LX{bɔgeŋmane} \HM{} \PH{bɔ̀gɪ̀ŋmánɪ́} \PS{n.~} \GE{forearm} \OP{bɔgeŋmama} \TP{bɔgeŋmamɛɛ} \VA{bageŋmane}
\LX{bogere} \HM{} \PH{bʊ́grɪ̀} \PS{v.~} \GE{to break into little bits or pulp} \PD{bogerɛɛ, bogerɔ, bogerɔ, bogerebɔ, bogeraa, bogerebɔ } 
\LX{bɔgere} \HM{} \PH{bɔ̌grɪ́} \PS{n.~} \GE{part, portion of a whole} \TP{bɔgerɛɛ} \VA{bɔgre}
\LX{bɔgere} \HM{} \PH{bɔ̀gɪ̀rɪ́} \PS{n.~} \SN \GE{fetish, object believed to embody a spirit} \SN \GE{sacrifice, offering to a divinity} \OP{bɔgɔ/bɔga} \TP{bɔgerɛɛ/bɔgɛɛ} \VA{bagere}
\LX{bɔgesaa} \HM{} \PH{bɔ̀gsàà} \PS{n.~} \GE{counsellor for a person being initiated into the bagre cult} \OP{bɔgesaamine} \TP{bɔgesaarɛɛ} \VA{bagesaa}
\LX{bɔgesoŋ} \HM{} \PH{bɔ̀gɪ̀sʋ̂ŋ} \PS{n.~} \GE{luck} \OP{bɔgesonne} \TP{bɔgesonnɛɛ} \VA{bagesoŋ}
\LX{bɔgetaŋkpolo} \HM{} \PH{bɔ̀gɪ̀táŋkpóló} \PS{n.~} \GE{cult members, those who have gone through the initiation process for a cult} \OP{bɔgetaŋkpolli} \TP{bɔgetaŋkpolee} \VA{bagetaŋkpolo}
\LX{bogi} \HM{} \PH{bògí} \PS{n.~} \SN \GE{hole} \SN \GE{grave, tomb} \SN \GE{pit} \SN \GE{ditch} \SN \GE{limit} \OP{bogiri} \TP{bogree} 
\LX{bogigbɛre-katakye} \HM{} \PH{bògìgbɛre-kátákyè} \PS{n.~} \GE{subway, underground train} \OP{bogigbɛre-katakyeri} \TP{bogigbɛre-katakyeree} 
\LX{bogikantoo} \HM{} \PH{bògíkántóó} \PS{n.~} \GE{a type of rodent (not edible)} \OP{bogikantori} \TP{bogikantoree} 
\LX{bogisulee} \HM{} \PH{bògsùlèè} \PS{n.~} \GE{round, deep tomb (like a well)} \OP{bogisulli} \TP{bogisullee} 
\LX{bogiti} \HM{} \PH{bógítì} \PS{n.~} \GE{bucket} \OP{bogitiri} \TP{bogitiree} 
\LX{bogi-uuno} \HM{} \PH{bògì-ù᷉ù᷉nó} \PS{n.~} \GE{one who buries the dead} \OP{bogi-uunimo} \TP{bogi-uunimee} 
\LX{bogi-u᷉} \HM{} \PH{bògì-ù᷉ù᷉} \PS{n.~} \GE{the act of burying a human corpse} 
\LX{bogi-u᷉u-faare} \HM{} \PH{bògì-ù᷉ù᷉-fáárɪ́} \PS{n.~} \GE{xylophone music signaling the completion of the burial of the dead} \TP{bogi-u᷉u-faarɛɛ} 
\LX{bogi-u᷉u-naazoɔ} \HM{} \PH{bògì-ù᷉ù᷉-náázʊ́ɔ́} \PS{n.~} \GE{chief undertaker (one who buries the dead)} \OP{bogi-u᷉u-naazoore} \TP{bogi-u᷉u-naazoorɛɛ} 
\LX{bogi-u᷉u-saabo} \HM{} \PH{bògì-ù᷉ù᷉-sáábʊ́} \PS{n.~} \GE{ritual TZ for those who bury the dead} \TP{bogi-u᷉u-saarɛɛ} 
\LX{bogi-u᷉u-sawɔlɔ} \HM{} \PH{bògì-ù᷉ù᷉-sáwɔ́lɔ́} \PS{n.~} \GE{ritual dish for those who bury the dead} \TP{bogi-u᷉u-sawɔlɛɛ} 
\LX{bogi-u᷉u-zanna} \HM{} \PH{bògì-ù᷉ù᷉-zànná} \PS{n.~} \GE{amateur undertaker} \OP{bogi-u᷉u-zanema} \TP{bogi-u᷉u-zanemɛɛ} 
\LX{bogi-u᷉u᷉-libie} \HM{} \PH{bògì-ù᷉ù᷉-líbíè} \PS{n.~} \GE{fee for those who bury the dead} \TP{bogi-u᷉u᷉-libiiree} 
\LX{bogivuo} \HM{} \PH{bògìvùò} \PS{n.~} \GE{uncovered tomb} \OP{bogivori} \TP{bogivoree} 
\LX{bogiwɛraa} \HM{} \PH{bògwɛ́ráà} \PS{n.~} \GE{oblong tomb} \OP{bogiwɛrre} 
\LX{bogizeɛ} \HM{} \PH{bògzɪ̀ɛ̀} \PS{n.~} \GE{grave, deep pit} \OP{bogizeere} \TP{bogizeerɛɛ} 
\LX{bɔgɔ} \HM{} \PH{bɔ́gɔ̀} \PS{n.~} \GE{track, path (usually left by passing through grassy bush)} \OP{bɔgere} \TP{bɔgerɛɛ} \VA{bɔga}
\LX{bɔgɔ} \HM{} \PH{bɔ́gɔ́} \PS{n.~} \GE{shoulder} \OP{bɔgere} \TP{bɔgerɛɛ} \VA{bɔge/bɔga/bage}
\LX{bogɔbogɔ} \HM{} \PH{bʊ́gɔ́bʊ́gɔ́} \PS{adv.~} \GE{ground into pulp} \VA{bogaboga}
\LX{boguolo} \HM{} \PH{bʊ̀gúóló} \PS{n.~} \GE{goats given to one to raise in return for some} \TP{boguolee} 
\LX{bog-uuno} \HM{} \PH{bòg-úúnó} \PS{n.~} \GE{one who buries the dead; undertaker} \OP{bog-uunimo} \TP{bog-uunimee} 
\LX{bokolɔ} \HM{} \PH{bʊ̀kʊ̀lɔ̀} \PS{n.~} \GE{tasteless powdered stuff, tobacco} \OP{bokolle} \TP{bokolɛɛ} \VA{bokola}
\LX{bokoŋ} \HM{} \PH{bʊ̀kóŋ} \PS{n.~} \GE{lean or thin goat} \OP{bokonni} \TP{bokonnee} 
\LX{bokontawɛ} \HM{} \PH{bʊ̀kòntáwɛ́} \PS{n.~} \GE{a daily marketplace} \OP{bokontawɛre} \TP{bokontawɛrɛɛ} 
\LX{bokore} \HM{} \PH{bʊ̀kʊ́rɪ́} \PS{n.~} \GE{old goat} \OP{bokora} \TP{bokorɛɛ} 
\LX{bokpɛ} \HM{} \PH{bʊ̀kpɛ̀} \PS{v.~} \GE{to sympathise with} \PD{bokpɛɛ, boorɔkpeɛrɛ, bokpeɛrɛ, bokpeɛrebɛ, bokpeɛraa, - } 
\LX{bɔlaa} \HM{} \PH{bɔ̀làà} \PS{adv.~} \GE{ long, round and smooth} 
\LX{bolaakaatoo} \HM{} \PH{bʊ̀lààkààtʊ̀ʊ̀} \PS{adv.~} \GE{talking continuously, noisy} 
\LX{bolaare} \HM{} \PH{bʊ̀láárɪ́} \PS{v.~} \GE{to burst strongly} \XV{A lɔɔre kɔbɔ da bolaarɛɛ la.} \XE{The lorry tyre blasted with a loud noise.} \PD{bolaarɛɛ, bolaara, bolaara, bolaareba, bolaaraa, bolaareba } \VA{belaare}
\LX{bole} \HM{} \PH{bʊ́lɪ́} \PS{v.~} \GE{to smear all over} \PD{bolɛɛ, bola, bola, bolba, bolaa, - } 
\LX{bɔle} \HM{} \PH{bɔ́lɪ́} \PS{n.~} \GE{ball} \OP{bɔlɔ} \TP{bɔlɛɛ/bɔllɛɛ} 
\LX{bɔle} \HM{} \PH{bɔ̀lɪ̀} \PS{n.~} \GE{field rat} \SG{bɔloo} \OP{bɔlɔ} \TP{bɔllɛɛ} 
\LX{bolebole} \HM{} \PH{bʊ̀lbʊ́lɪ́} \PS{v.~} \GE{to speak unintelligibly, to mumble} \PD{bolebolɛɛ, bolebolɔ, bolebolɔ, bolebolebɔ, bolebolaa, bolebolebɔ } 
\LX{bolee} \HM{} \PH{bʊ̀léé} \PS{n.~} \GE{kid (young goat)} \OP{bobili} \TP{bobillee} 
\LX{bɔlembɔ} \HM{} \PH{bɔ̀lɪ̀mbɔ́} \PS{n.~} \GE{bottle} \SG{bɔlembɔroo} \OP{ bɔlembɔre} \TP{bɔlembɔrɛɛ} \VA{pilintoɔ}
\LX{bolenaa} \HM{} \PH{bʊ̀lénáá} \PS{n.~} \SN \GE{tethered goat} \SN \GE{one restricted from moving} \OP{bolenni} \TP{bolennee} 
\LX{bo-leŋi} \HM{} \PH{bʊ̀-léŋ} \PS{n.~} \GE{the act of tethering goats} 
\LX{boli} \HM{} \PH{bólí} \PS{v.~} \GE{to be foolish} \PD{bolee, bolo, bolo, bolibo, bolaa, bolibo } 
\LX{boli} \HM{} \PH{'bòlí} \PS{adj.~} \GE{pure (honey)} 
\LX{bɔll} \HM{} \PH{bɔ̀ll} \PS{adj.~} \GE{long and oblong} 
\LX{bolle} \HM{} \PH{bʊ̀llɪ} \PS{v.~} \GE{to murmur; to talk inaudibly} \PD{bollɛɛ, bolla, bolla, boleba, bollaa, boleba } 
\LX{bɔlle} \HM{} \PH{bɔ̀llɪ̀} \PS{v.~} \SN \GE{to move all over an object (insects)} \SN \GE{to smear liquid on a surface in an untidy way} \PD{bɔllɛɛ, bɔllɔ, bɔllɔ, bɔlbɔ, bɔllaa, - } 
\LX{bolɔ} \HM{} \PH{bʋ̀lɔ́} \PS{n.~} \GE{a type of vegetable} \VA{bola}
\LX{bolɔbolɔ} \HM{} \PH{bʋ́lɔ́bʋ́lɔ́} \PS{adv.~} \GE{lukewarm  (water)} \VA{bolabola}
\LX{bɔlɔmbɔlɔŋ} \HM{} \PH{bɔ́lɔ́mbɔ́lɔ́ŋ} \PS{adj.~} \GE{smooth and shiny} 
\LX{boluu} \HM{} \PH{bòlúú} \PS{pron. interj.~} \GE{what is it?} 
\LX{bombaala} \HM{} \PH{bʋ́mbààlá} \PS{n.~} \GE{a sick thing} \OP{bombaaleba} \TP{bombaalɛɛ} 
\LX{bombaalaa} \HM{} \PH{bʋ́mbààlàà} \PS{n.~} \GE{a slim or narrow thing} \OP{bombaale} 
\LX{bombaaraa} \HM{} \PH{bʋ́mbààráà} \PS{n.~} \GE{something that can grow} \OP{bombaare} 
\LX{bombaaraa} \HM{} \PH{bʋ́mbááráá} \PS{n.~} \GE{the last one} \OP{bombaare} 
\LX{bombalaa} \HM{} \PH{bʋ́mbáláá} \PS{n.~} \GE{something which is avoided} \OP{bomballe} 
\LX{bombalaa} \HM{} \PH{bʊ́mbàlàà} \PS{n.~} \GE{kind of thing (referring to something huge or strange)} \XV{Yeŋ ka fo ta pɔge a bombalaa ŋa?} \XE{Where did you get this kind of thing?.} 
\LX{bombaltakyelmɛ} \HM{} \PH{bòmbàltáʧɪ́mɛ́} \PS{n.~} \GE{things of all sorts} \TP{bombaltakyelmɛɛ} 
\LX{bombaŋe} \HM{} \PH{bʋ́mbàŋ} \PS{n.~} \SN \GE{the act of knowing} \SN \GE{knowledge} 
\LX{bombannaa} \HM{} \PH{bʊ́mbànnáà} \PS{n.~} \GE{something which is known} \OP{bombanne} \VA{bombaŋenaaːbombaŋene}
\LX{bombare} \HM{} \PH{bʋ̌mbàrɪ̀} \PS{n.~} \SN \GE{the act of releasing; the act of setting animals free} \SN \GE{the act of offering firstfruits to gods} \OP{bombarre} 
\LX{bombeɛ} \HM{} \PH{bʋ́mbɪ́ɛ́} \PS{n.~} \GE{something strange or mysterious} \OP{bombeere} \TP{bombeerɛɛ} 
\LX{bombeene} \HM{} \PH{bʋ́mbɪ́ɪ́nɪ́} \PS{n.~} \GE{something that is cooked or ripe} \OP{bombeemɛ} \TP{bombeenɛɛ} 
\LX{bombeɛraa} \HM{} \PH{bʋ́mbɪ̀ɛ́ráà} \PS{n.~} \SN \GE{sliced strips (of leather)} \SN \GE{that which is patched or to be patched (as in car tyre)} \OP{bombeɛre} \TP{bombɛrɛɛ} 
\LX{bombeɛre} \HM{} \PH{bʋ́mbɪ̀ɛ̀rɪ́} \PS{n.~} \GE{boils} \OP{bombɛbɛ} \TP{bombɛrɛɛ} 
\LX{bombiiri} \HM{} \PH{bʊ́mbíírí} \PS{n.~} \GE{very young ones; unhatched embryos in eggs} \TP{bombiiree} 
\LX{bombile} \HM{} \PH{bòmbílé} \PS{n.~} \GE{ young donkey; foal} \OP{bombilii} \TP{bombillee} 
\LX{bombile} \HM{} \PH{bʊ́mbílé} \PS{n.~} \GE{small thing} \OP{bombilii} 
\LX{bombiri} \HM{} \PH{bʊ́mbírì} \PS{n.~} \GE{seed; cereal} \SG{bombiruu} \OP{bombie} \TP{bombiiree} 
\LX{bombo} \HM{} \PH{bǒmbò} \PS{n.~} \SN \GE{a small plant with a jellylike tuber} \SN \GE{the tuber or flour made from the tuber of this plant} \SN \GE{fried dish made from the flour of this plant} \SG{bomboruu (for plant or tuber)} \OP{bombori} \TP{bomboree} 
\LX{bombɔ} \HM{} \PH{bʋ́mbɔ́} \PS{n.~} \GE{quest for wealth} 
\LX{bombo sawɔlɔ} \HM{} \PH{bòmbò sáwɔ́lɔ́} \PS{n.~} \GE{fried duff from bombo (flour)} \TP{bombo sawɔlɛɛ} \VA{bombo sawala}
\LX{bombɔge} \HM{} \PH{bʋ́mbɔ́gɪ́} \PS{n.~} \GE{the act of grinding into fine pulp} \OP{bombɔgre} \TP{bombɔgrɛɛ} 
\LX{bombɔgɔ} \HM{} \PH{} \PS{n.~} \GE{the forearm or leg of an animal} \OP{bombɔgerɔ} \TP{bombɔgerɛɛ} \VA{bombɔge}
\LX{bombolaa} \HM{} \PH{bʋ́mbʋ́láá} \PS{n.~} \GE{a thing that is not of the same species; a strange thing} \OP{bombolle} \TP{bombollɛɛ} 
\LX{bombone} \HM{} \PH{bʋ́mbʋ́nɪ̀} \PS{n.~} \GE{a useless thing} \OP{bomboma} \TP{bombonnɛɛ} 
\LX{bomboŋɔ} \HM{} \PH{bʋ́mbʋ̀ŋɔ́} \PS{n.~} \GE{something that cannot be classified} \OP{bombonne} \TP{bombonnɛɛ} 
\LX{bombonoone} \HM{} \PH{bǒmbònʋ̀ʋ̀nɪ̀} \PS{n.~} \GE{a type of tree} \OP{bombonooma} \TP{ bombonoonɛɛ.} 
\LX{bomboɔraa} \HM{} \PH{bʋ́mbʋ́ɔ́ráá} \PS{n.~} \GE{something that is sought ; something that is wanted or liked} \OP{bomboɔre} \TP{bomboɔrɛɛ.} 
\LX{bomboɔre} \HM{} \PH{bʊ́mbʊ́ɔ́rɪ́} \PS{n.~} \SN \GE{things one wishes to have} \SN \GE{wealth} 
\LX{bombooree} \HM{} \PH{bʋ́mbʋ́ʋ́rɪ́} \PS{n.~} \SN \GE{something that is kept for posterity} \SN \GE{seed for sowing} \OP{bomboora} \TP{bomboorɛɛ} 
\LX{bomboorɔ} \HM{} \PH{bʊ́mbʊ́ʊ́rɔ́} \PS{n.~} \GE{seed for planting} \SG{bombooroo} \TP{bomboorɛɛ} \VA{bomboora}
\LX{bomboraa} \HM{} \PH{bʋ́mbʋ̀ráà} \PS{n.~} \GE{something that is sown; a crop} \OP{bomborre} \TP{ bomborrɛɛ.} 
\LX{bombore} \HM{} \PH{bʋ́mbʋ̀rɪ̀} \PS{n.~} \GE{the act of sowing} 
\LX{bɔmbɔrezeɛ} \HM{} \PH{bɔ̀mbɔ̀rzɪ̀ɛ́} \PS{n.~} \GE{a type of snake} \OP{bɔmbɔrezeere} \TP{bɔmbɔrezeerɛɛ} 
\LX{bɔmbɔroŋ} \HM{} \PH{bɔ̌mbɔ̀rʋ́ŋ} \PS{n.~} \GE{a type of snake} \OP{bɔmbɔrre} \TP{bɔmbɔrɛɛ} 
\LX{bombugilaa} \HM{} \PH{bʋ́mbúglàá} \PS{n.~} \GE{something that is nursed} \OP{bombugili} 
\LX{bombulaa} \HM{} \PH{bʋmbulaa} \PS{n.~} \GE{something that can sprout from a seed} \OP{bombulli} 
\LX{bombuli} \HM{} \PH{bʊ́mbúlí} \PS{n.~} \GE{shoot of a new plant} \OP{bombulo} \TP{bombullee} 
\LX{bombuuraa} \HM{} \PH{bʋ́mbúúràá} \PS{n.~} \SN \GE{a measure} \SN \GE{something that is respected} \OP{bombuuri} 
\LX{bommaalaa} \HM{} \PH{bʋ́mmààláà} \PS{n.~} \GE{something that is made} \OP{bommaale} 
\LX{bommaale} \HM{} \PH{bʋ́mmààlɪ̀} \PS{n.~} \GE{the act of making} \OP{bommaaleba} 
\LX{bommaane} \HM{} \PH{bʋ́mmàànɪ́} \PS{n.~} \SN \GE{an egg that is immature} \SN \GE{something that is tiny} \OP{bommaama} 
\LX{bommaaroŋ} \HM{} \PH{bʋ́mmáárʋ́ŋ} \PS{n.~} \GE{a wet, cold, or fresh thing} \OP{bommaara} 
\LX{bommannaa} \HM{} \PH{bʋ́mmánnáà} \PS{n.~} \GE{a measure; something that is measured} 
\LX{bompaala} \HM{} \PH{bʋ́mpáálà} \PS{n.~} \SN \GE{new or fresh things} \SN \GE{firstfruits of crops} \OP{bompaala} \TP{bompalɛɛ} 
\LX{bompaare} \HM{} \PH{bʋ́mpáárɪ̀} \PS{n.~} \GE{the vulva of any animal} \OP{bompɛɛrɛɛ} 
\LX{bompaare} \HM{} \PH{bǒmpáárɪ̀} \PS{n.~} \GE{the vulva of a donkey} \OP{bompaɛ} \TP{ bompɛɛrɛɛ} 
\LX{bompaare} \HM{} \PH{bǒmpáárɪ̀} \PS{n.~} \GE{foot rot} \OP{bompaɛ} \TP{ bompɛɛrɛɛ} 
\LX{bompare} \HM{} \PH{bʋ́mpárɪ̀} \PS{n.~} \SN \GE{residue} \SN \GE{bottom} \OP{bomparɛɛ} 
\LX{bompeelaa} \HM{} \PH{bʋ́mpɪ́ɪ́láá} \PS{n.~} \GE{a very young animal} \OP{bompeele} \TP{bompeelɛɛ} 
\LX{bompeelaa} \HM{} \PH{bʋ́mpɪ́ɪ́láá} \PS{n.~} \GE{a piece torn from a piece of cloth} \OP{bompeele} \TP{bompeelɛɛ} 
\LX{bompɛɛre} \HM{} \PH{bʋ́mpɛ̀ɛ̀rɪ́} \PS{n.~} \GE{an immature thing} \OP{bompɛbɛ} 
\LX{bompege} \HM{} \PH{bʋ́mpɪ̀gɪ̀} \PS{n.~} \GE{the act of decorating, improving or adding worth to something} 
\LX{bompegɛ} \HM{} \PH{bʋ́mpɪ́gɛ́} \PS{n.~} \GE{a piece of a broken container (calabash)} \OP{bompegerre} \TP{ bompegerɛɛ.} 
\LX{bompɛge} \HM{} \PH{bʋ́mpɛ́gɪ̀} \PS{n.~} \GE{the act of washing the shell or bark of something} 
\LX{bompegelaa} \HM{} \PH{bʋ́mpɪ́gláà} \PS{n.~} \GE{something that is moving and must be intercepted} \OP{bompegele} \TP{bompegelɛɛ} 
\LX{bompɛgelaa} \HM{} \PH{bʋ́mpɛ́gláà} \PS{n.~} \GE{something that is carried by holding it to one's side} \OP{bompɛgele} \TP{bompɛgelɛɛ} 
\LX{bompegeraa} \HM{} \PH{bʋ́mpɪ́gráà} \PS{n.~} \GE{something that is adorned} \OP{bompegere} \TP{bompegerɛɛ} 
\LX{bompɛgeraa} \HM{} \PH{bʊ́pɛ́gráá} \PS{n.~} \SN \GE{something that has been washed} \SN \GE{something for washing} \OP{bompɛgere} \TP{bompɛgerɛɛ} 
\LX{bompelaa} \HM{} \PH{bʋ́mpɪ̀làá} \PS{n.~} \GE{a white thing} \OP{bompeɛle} \TP{bompeɛlɛɛ} 
\LX{bompɛlaa} \HM{} \PH{bʋ́mpɛ́láá} \PS{n.~} \GE{something that can be peeled} \OP{bompɛlle} \TP{ bompɛllɛɛ} 
\LX{bompɛle} \HM{} \PH{bʋ́mpɛ́lɪ́} \PS{n.~} \GE{the act of peeling} 
\LX{bompɛle} \HM{} \PH{bʊḿpɛ̀lì} \PS{n.~} \GE{peel} \OP{bompɛlɛ} \TP{bompɛllɛɛ} 
\LX{bompɛnnaa} \HM{} \PH{bʊ́mpɛ̀nnáá} \PS{n.~} \GE{something that is carved} \OP{bompɛnne} \TP{bompɛnnɛɛ} 
\LX{bompɛnne} \HM{} \PH{bʋ́mpɛ̀nnɪ̀} \PS{n.~} \GE{the act of carving} 
\LX{bompere} \HM{} \PH{bʋ́mpɪ̀rɪ̀} \PS{n.~} \GE{the act of dehusking (removing the dry outer cover of a grain)} 
\LX{bompere} \HM{} \PH{bʊ́pɪ́rɪ́} \PS{n.~} \GE{the act of scratching (as by fowl)} 
\LX{bompi} \HM{} \PH{bʋ́mpí} \PS{n.~} \SN \GE{the process of becoming distended or bloated} \SN \GE{forerunning for something} 
\LX{bompigiraa} \HM{} \PH{bʋ́mpǐgráà} \PS{n.~} \GE{something that is to be watched or ought to be watched} \OP{bompigiri} 
\LX{bompiiraa} \HM{} \PH{bʋ́mpììráà} \PS{n.~} \SN \GE{a lost thing that is found; an unexpected find} \SN \GE{an easy find} \OP{bompiiri} 
\LX{bompiiraa} \HM{} \PH{bʊ́mpííráá} \PS{n.~} \GE{a distended thing} \OP{bompiiri} 
\LX{bompiiraa} \HM{} \PH{bʊ́mpííráá} \PS{n.~} \GE{something that has been earmarked for someone} \OP{bompiiri} 
\LX{bompiire} \HM{} \PH{bʋ́mpííré} \PS{n.~} \GE{one who finds a lost item} \OP{bompiiribe} \TP{bompiiribee} 
\LX{bompɔ} \HM{} \PH{bʋ́mpɔ́} \PS{n.~} \GE{the act of swearing or taking an oath on something} 
\LX{bompɔge} \HM{} \PH{bʋ́mpɔ́gɪ́} \PS{n.~} \GE{a female thing} \OP{bompɔgeba} \TP{bompɔgrɛɛ} 
\LX{bompɔge} \HM{} \PH{bʋ́mpɔ̀gɪ̀} \PS{n.~} \GE{the act of confining animals in enclosures} 
\LX{bompɔgeraa} \HM{} \PH{bʋ́mpɔ̀gráà} \PS{n.~} \GE{something that is confined in an enclosure} \OP{bompɔgere} 
\LX{bompɔle} \HM{} \PH{bʋ́mpɔ́lɪ̀} \PS{n.~} \GE{a young thing} \OP{bompɔlɔ} \TP{ bompɔllɛɛ} 
\LX{bompollaa} \HM{} \PH{bʊ́mpʊ́lláá} \PS{n.~} \GE{something on which an oath is taken} \OP{bompolle} 
\LX{bompoɔ} \HM{} \PH{bʋ́mpʋ́ɔ́} \PS{n.~} \GE{a pregnant thing (non human)} \OP{bompoore} \TP{ bompoorɛɛ.} 
\LX{bompoɔraa} \HM{} \PH{bʋ́mpʋ́ɔ́ráá} \PS{n.~} \GE{something by which one swears} \OP{bompoɔre} \TP{ bompoɔrɛɛ} 
\LX{bomporaa} \HM{} \PH{bʋ́mpʋ́ráá} \PS{n.~} \GE{something that is poured} \OP{bomporre} \TP{ bomporrɛɛ} 
\LX{bompore} \HM{} \PH{bʋ́mpʋ̀rɪ̀} \PS{n.~} \GE{plentiful things} \OP{bomporre} 
\LX{bompore} \HM{} \PH{bʋ́mpʋ́rɪ́} \PS{n.~} \GE{something that is small and short; a little thing} \OP{bompora} \TP{bomporɛɛ} 
\LX{bompore} \HM{} \PH{bʋ́mpʋ́rɪ̀} \PS{n.~} \GE{the act of pouring} 
\LX{bompɔre} \HM{} \PH{bʋ́mpɔ́rɪ́} \PS{n.~} \GE{the act of plucking something; (fruits or groundnuts)} 
\LX{bompɔrɔ} \HM{} \PH{bʊ́mpɔ́rɔ̀} \PS{n.~} \GE{harvester} \OP{bompɔreba} \TP{bompɔrebɛɛ} 
\LX{bɔmpɔrɔ} \HM{} \PH{bɔ̀mpɔ́rɔ́} \PS{n.~} \GE{mumps} \XV{ .} \XE{} \OP{bɔmpɔrre} \TP{ bɔmpɔrɛɛ.} 
\LX{bompugo} \HM{} \PH{bʋ́mpúgó} \PS{n.~} \GE{an ash-coloured thing} \OP{bompugri} \TP{ bompugree} 
\LX{bompulonpuloŋ} \HM{} \PH{bʋ́mpùlònpúlóŋ} \PS{n.~} \GE{various things mixed up} 
\LX{bompuuraa} \HM{} \PH{bʋ́mpùùráà} \PS{n.~} \GE{something that has flowered} \OP{bompuuri} \TP{bompuuree.} 
\LX{bompuuraa} \HM{} \PH{bʋ́mpùùráà} \PS{n.~} \GE{something that is or can be praised} \OP{bompuuri} \TP{bompuuree.} 
\LX{bompuuri} \HM{} \PH{bʋ́mpúúrí} \PS{n.~} \GE{flowers of things} \TP{bompuuree} 
\LX{boŋ} \HM{} \PH{bóŋ} \PS{interrogative  pronoun.~} \GE{what?} \XV{Boŋ ka fo na e zenɛ?} \XE{What will you do today?.} 
\LX{bonaare} \HM{} \PH{bʋ̀náárɪ̀} \PS{adj.~} \GE{four times} 
\LX{bon-agraa} \HM{} \PH{bʋ́ŋágráà} \PS{n.~} \SN \GE{winged creature} \SN \GE{ insect} \OP{boŋ-agre} 
\LX{bonda} \HM{} \PH{bʋ́ndà} \PS{n.~} \GE{the act of buying} 
\LX{bondaa} \HM{} \PH{bòndáá} \PS{n.~} \GE{male donkey} \OP{bondaare} \TP{bondarɛɛ} 
\LX{bondaa} \HM{} \PH{bʋ́ndáá} \PS{n.~} \GE{a male thing} \OP{bondaare} \TP{bondarɛɛ} 
\LX{bondaa} \HM{} \PH{bʋ́ndáà} \PS{n.~} \GE{the act of pushing} \OP{bonda} 
\LX{bɔndaa} \HM{} \PH{bɔ̀ndáá} \PS{n.~} \GE{lizard (variation of ‘bandaa’(individual) and ‘baŋa’(dialectal))} \OP{bɔndaare} \TP{bɔndarɛɛ} \VA{bandaa}
\LX{bondaana} \HM{} \PH{bʋ́ndááná} \PS{n.~} \GE{rich person; one who has wealth} \XV{.} \XE{} \OP{bondaanema/bondaama} \TP{ bondaamɛɛ/bondaanɛɛ} 
\LX{bondaaraa} \HM{} \PH{bʋ́ndààráà} \PS{n.~} \GE{something that is bought; something that is meant for sale} \OP{bondaare} 
\LX{bondaga} \HM{} \PH{bòndàgà} \PS{n.~} \GE{carefree} \OP{bondagre} \TP{ bondagrɛɛ.} 
\LX{bonde} \HM{} \PH{bʋ́ndɪ́} \PS{n.~} \SN \GE{the act of taking} \SN \GE{the act of harvesting} 
\LX{bondegere} \HM{} \PH{bʋ́ndɪ̀grɪ} \PS{n.~} \GE{a dwarflike thing} \OP{bondegɛ} 
\LX{bondɛgere} \HM{} \PH{bʋ́ndɛ̌grɪ̀} \PS{n.~} \GE{a dirty thing} \OP{bondɛgɛ} 
\LX{bondemannewullaa} \HM{} \PH{bʊ́ndɪ́mánnɪ́wúlláá} \PS{n.~} \GE{example} \OP{bondemannewulli} 
\LX{bondi} \HM{} \PH{bʋ́ndì} \PS{n.~} \GE{the act of eating} 
\LX{bondikaama} \HM{} \PH{bʊ̀ndìkáámá} \PS{n.~} \GE{delicious and assorted food; luncheon} \TP{bondikaanɛɛ} 
\LX{bondikaraa} \HM{} \PH{bʊ́ndìkáráá} \PS{n.~} \GE{raw food; food that is not well cooked} \OP{bondikare} \TP{bondikarɛɛ} 
\LX{bondi-kpaa} \HM{} \PH{bondi-kpaa} \PS{n.~} \GE{leftover food} \OP{bondi-kpaare} 
\LX{bondimaala} \HM{} \PH{bʊ́ndímáálá} \PS{n.~} \GE{cook, one who prepares food} 
\LX{bondiraa} \HM{} \PH{bʊ́ndìráà} \PS{n.~} \SN \GE{food; something that is edible} \SN \GE{a sharp thing (blade)} \OP{bondirri} \TP{bondirree} 
\LX{bondire} \HM{} \PH{bʊ́ndíré} \PS{n.~} \SN \GE{one who eats} \SN \GE{rightful owner} \OP{bondiribe} \TP{bondiribee} 
\LX{bondirii} \HM{} \PH{bʊ́ndìríì} \PS{n.~} \GE{food, things to eat} \TP{bondirree} 
\LX{bondi-sa᷉a᷉} \HM{} \PH{bʊ́ndì-sà᷉à} \PS{n.~} \GE{the act of wasting food} 
\LX{bondizie} \HM{} \PH{bʊ́ndìzíé} \PS{n.~} \GE{restaurant} \OP{bondiziiri} \TP{bondiziiree} 
\LX{bondogelaa} \HM{} \PH{bʋ́ndʋ́gláá} \PS{n.~} \GE{something that is used for cooking} \OP{bondogele} 
\LX{bondɔgelaa} \HM{} \PH{bʋ́ndɔ́gláá} \PS{n.~} \GE{an extra thing} \OP{bondɔgele} 
\LX{bondogeraa} \HM{} \PH{bʋ́ndʋ́gráá} \PS{n.~} \GE{a boiled thing} \OP{bondogere} 
\LX{bondɔgerɔ} \HM{} \PH{bʊ́ndɔ̀gerɔ̀} \PS{n.~} \GE{an animal that has recently littered} \OP{bondɔgere} \TP{bondɔgerɛɛ} 
\LX{bondonaa} \HM{} \PH{bʊ́ndʊ̀náà} \PS{n.~} \SN \GE{something that bites} \SN \GE{insect, bug} \OP{bondonne} \TP{bondonnɛɛ} \VA{kyibe}
\LX{bondori} \HM{} \PH{bʋ́ndórí} \PS{n.~} \GE{ heap} \OP{bondoree} \TP{bondorree} 
\LX{bone} \HM{} \PH{bʊ́nɪ̀} \PS{n.~} \GE{thing} \OP{boma} \TP{bonnɛɛ} 
\LX{bone} \HM{} \PH{bʋ́nɪ́} \PS{adj.~} \GE{worthless or useless} \OP{boma} 
\LX{boŋemaŋ} \HM{} \PH{bʋ́ŋɪ̀máŋ} \PS{adj.~} \GE{hazy} 
\LX{bɔŋenaa} \HM{} \PH{bɔ́ŋɪ́náá} \PS{n.~} \GE{type of tree, medicinal} \OP{bɔŋene} \TP{bɔŋenɛɛ} 
\LX{bon-ennaa} \HM{} \PH{bʊ́n-ɪ́nnáà} \PS{n.~} \GE{container} \OP{bon-enne} \TP{bon-ennɛɛ} 
\LX{bɔŋenɔ} \HM{} \PH{bɔ̀ŋɪ́nɔ́} \PS{n.~} \GE{witty person} \OP{bɔŋenema} \TP{bɔŋenemɛɛ} 
\LX{bonfa} \HM{} \PH{́bʋ́nfá} \PS{n.~} \GE{the act of robbing} 
\LX{bonfaa} \HM{} \PH{bʋ́nfáá} \PS{n.~} \GE{a bad or evil thing} \OP{bonfaare} \TP{ bonfaarɛɛ} 
\LX{bonfaa} \HM{} \PH{bʊ́nfáà} \PS{n.~} \GE{robbery} 
\LX{bonfaara} \HM{} \PH{bʋ́nfáárá} \PS{n.~} \GE{robber, thief} \OP{bonfareba} \TP{ bonfabarɛɛ.} 
\LX{bonfaaraa} \HM{} \PH{bʋ́nfááráá} \PS{n.~} \GE{something that is robbed} \OP{bonfaare} \TP{ bonfaarɛɛ} 
\LX{bonfane} \HM{} \PH{bʋ́nfànɪ́} \PS{n.~} \GE{a rough thing} \OP{bonfama} \TP{ bonfannɛɛ.} 
\LX{bonfarafara} \HM{} \PH{bʋ́nfàràfàrà} \PS{n.~} \GE{a rough, scaly thing} 
\LX{bonfeeloo} \HM{} \PH{bʋ́fɪ́ɪ́lʋ́ʋ́} \PS{n.~} \GE{an opening that is small and tight} \OP{bonfeelɛ} \TP{ bonfeelɛɛ.} 
\LX{bonfegɛ} \HM{} \PH{bʋ́nfɪ̀gɛ́} \PS{n.~} \GE{narrow space} \XV{see [bʊ́nfɔ́glʊ́ʊ]́} \XE{} 
\LX{bonfe᷉ê} \HM{} \PH{bʋ́nfɪ́᷉ɪ́᷉} \PS{n.~} \GE{a small quantity} \OP{bonfeene} \TP{bonfeenɛɛ} 
\LX{bonfogeloo} \HM{} \PH{bʋ́nfʋ́glʋ́ʋ́} \PS{n.~} \GE{a narrow thing} \OP{bofogela} \TP{ bonfogelɛɛ} 
\LX{bonfogiluu} \HM{} \PH{bʋ́nfóglúú} \PS{n.~} \GE{a seedless pod} \OP{bonfogilo} \TP{ bonfogilee} 
\LX{bonfole} \HM{} \PH{bʋ́nfʋ̀lɪ́} \PS{n.~} \GE{an immature thing} \OP{bonfola} \TP{ bonfolɛɛ.} 
\LX{bonfolo} \HM{} \PH{bʋ́nfòló} \PS{n.~} \GE{powerful spiritual beings} \OP{bonfolo} \TP{ bonfolee.} 
\LX{bonfɔraa} \HM{} \PH{bʊ́nfɔ́ráá} \PS{n.~} \GE{a narrow thing} \OP{bonfɔrre} \TP{bonfɔrɛɛ} 
\LX{bonfo᷉ɔo} \HM{} \PH{bʋ́nfʋ́᷉ɔ́̀᷉́ʋ́᷉} \PS{n.~} \GE{a light thing} \OP{bonfoɔne} \TP{bonfoɔnɛɛ} 
\LX{bonfuraa} \HM{} \PH{bʋ́nfùráà} \PS{n.~} \GE{something that is used to plug an opening} \OP{bonfurri} \TP{ bonfurree.} 
\LX{bonfuuraa} \HM{} \PH{bʋ́nfùùráá} \PS{n.~} \SN \GE{something that is cool} \SN \GE{something that threatens} \OP{bonfuuri} \TP{bonfuuree.} 
\LX{boŋganaa} \HM{} \PH{bʋŋganaa} \PS{n.~} \GE{bed, something that is slept on} \OP{boŋganne} \TP{ boŋganɛɛ.} 
\LX{boŋgbuli} \HM{} \PH{bʊ́ŋgbùlí} \PS{n.~} \SN \GE{a whole thing} \SN \GE{tumour} \OP{boŋgbulo} \TP{boŋgbulee} 
\LX{boŋgoraa} \HM{} \PH{bʋ́ŋgʋ̀ráà} \PS{n.~} \GE{something that is abundant} \OP{boŋgorre} \TP{boŋgorɛɛ} 
\LX{boŋgori} \HM{} \PH{bʋ́ŋgòrí} \PS{n.~} \GE{half-ripe fruit} \OP{boŋgoe} \TP{ boŋgoree.} 
\LX{boŋgori} \HM{} \PH{bóŋgórí} \PS{n. pl.~} \GE{shoes} 
\LX{boŋgu} \HM{} \PH{bʋ́ŋgù} \PS{n.~} \GE{the act of watching over something (crops from animals)} 
\LX{boŋguolaa} \HM{} \PH{bʋ́ŋgúóláá} \PS{n.~} \SN \GE{something that is raised or reared} \SN \GE{an animal given to someone to be raised} \OP{boŋguoli} \TP{boŋguolee} 
\LX{boŋguoli} \HM{} \PH{bʊ́ŋgúólí} \PS{n.~} \GE{domestic animals and birds (poultry and animal husbandry)} \OP{boŋguolo} 
\LX{boŋguoli} \HM{} \PH{bʊ̌ŋgúólì} \PS{n.~} \GE{the rearing of animals} 
\LX{boŋguolo} \HM{} \PH{bʊ́ŋgúólò} \PS{n.~} \SN \GE{} \VA{doŋguolo}
\LX{boŋguraa} \HM{} \PH{bʋ́ŋgùráà} \PS{n.~} \GE{something jumped upon} \OP{boŋgurri} 
\LX{boŋguuro} \HM{} \PH{bʊ́ŋgúúro} \PS{n.~} \GE{one who watches over things} \OP{boŋguuribo} \TP{boŋguuribee} 
\LX{bonii} \HM{} \PH{bʋ̀níì} \PS{adj.~} \GE{eight times} 
\LX{boŋkaa} \HM{} \PH{bʋ́ŋkáá} \PS{n.~} \GE{invocation of spirits} \OP{boŋkaareba} \TP{boŋkaarebɛɛ} \VA{boŋkaale}
\LX{boŋkaa} \HM{} \PH{bʋ́ŋkàà} \PS{n.~} \SN \GE{the act of taking care of something} \SN \GE{the act of staring at something} 
\LX{boŋkaara} \HM{} \PH{bʊ́ŋkáára} \PS{n.~} \GE{one who watches over something} \OP{boŋkaareba} \TP{boŋkaarebɛɛ} 
\LX{boŋkaŋa} \HM{} \PH{bʊ́ŋkáŋ} \PS{n.~} \GE{something} \OP{bommine} \VA{boŋkaŋa}
\LX{boŋkapala} \HM{} \PH{boŋkapala} \PS{n.~} \GE{a type of plant} \TP{boŋkapalɛɛ} 
\LX{boŋkaraa} \HM{} \PH{bʋ́ŋkáráá} \PS{n.~} \GE{a raw, uncooked, or unripe thing} \OP{boŋkare} \TP{ boŋkarɛɛ.} 
\LX{boŋkâa} \HM{} \PH{bʋ́ŋká᷉à᷉} \PS{n.~} \SN \GE{oil or fat from an animal} \SN \GE{something oily or full of fat} \SG{boŋkaanoo} \TP{boŋkaanɛɛ} 
\LX{boŋkoŋi} \HM{} \PH{bʋ́ŋkǒŋ} \PS{n.~} \GE{a lean thing} \OP{boŋkonni} \TP{boŋkonnee} 
\LX{boŋkoɔraa} \HM{} \PH{bʋ́ŋkʊ́ɔ́ráá} \PS{n.~} \SN \GE{crop; something that is cultivated} \SN \GE{something used for ploughing or farming} \OP{boŋkoɔre, boŋkoɔrɛɛ} 
\LX{boŋkoɔraa} \HM{} \PH{bʋ́ŋkʋ́ɔ́ráà} \PS{n.~} \GE{something meant for sale (goods)} \OP{boŋkoɔre} \TP{boŋkoɔrɛɛ} 
\LX{boŋkorabinzie} \HM{} \PH{bʊ́ŋkʊ́rábìnzíé} \PS{n.~} \GE{museum} \OP{boŋkorabinziiri} 
\LX{boŋkore} \HM{} \PH{bʋ́ŋkʊ́rɪ́} \PS{n.~} \GE{old thing} \OP{boŋkorɔ} \TP{boŋkorɛɛ} 
\LX{boŋkôɔ} \HM{} \PH{bʋ́ŋkʋ́᷉ɔ́᷉} \PS{n.~} \GE{a watery thing} \TP{ boŋkoɔnɛɛ} 
\LX{boŋkpɔgɔ} \HM{} \PH{bʊ́ŋkpɔ́gɔ̀} \PS{n.~} \SN \GE{lumps} \SN \GE{a type of disease whose symptoms are swellings and boils} \TP{boŋkpɔgerɛɛ} \VA{boŋkpɔga/boŋkpaga}
\LX{boŋkuoŋaa} \HM{} \PH{bʋ́ŋkúóŋáá} \PS{n.~} \GE{a dry thing} \OP{boŋkuoni} \TP{ boŋkuonee} 
\LX{boŋku᷉u} \HM{} \PH{bʊ́ŋkú᷉ú᷉} \PS{n.~} \GE{a dead thing} \OP{bokuuni} \TP{ boŋkuunee} 
\LX{bonkyaaraa} \HM{} \PH{bʊ́nʧááráá} \PS{n.~} \GE{gift} \OP{bonkyaare} \TP{bonkyaarɛɛ} 
\LX{bonkyeɛraa} \HM{} \PH{bʋ́nʧɪ́ɛ́ráá} \PS{n.~} \SN \GE{something that is contributed} \SN \GE{something that is cut} \SN \GE{something used for cutting} \OP{bonkyeɛre} \TP{ bonkyeɛrɛɛ} 
\LX{bonkyɛlee} \HM{} \PH{bʋ́nʧɛ́lɪ́ɪ́} \PS{n.~} \GE{a half thing} \OP{bonkyɛlle} 
\LX{bonkyêɛ᷉} \HM{} \PH{bʋ́nʧɪ́᷉ɛ́᷉} \PS{n.~} \GE{a torn thing} \OP{bonkyeɛne/bonkyeɛnɛ} \TP{ bonkyeɛnɛɛ} 
\LX{bonlaane} \HM{} \PH{bʋ́nláánɪ́} \PS{n.~} \GE{a worthless thing} \OP{bonlaama} \TP{ bonlaanɛɛ.} 
\LX{bonlane} \HM{} \PH{bʋ́nlánɪ́} \PS{n.~} \GE{uncastrated animal} \OP{bonlama} \TP{ bonlamɛɛ} 
\LX{bonlee} \HM{} \PH{bʊ́nléé} \PS{n.~} \GE{a small thing e.g. object/non human} \VA{bʊ́mbílé}
\LX{bonlee} \HM{} \PH{bònléé} \PS{n.~} \GE{colt, young donkey} \OP{bombilii} 
\LX{bonnaakale} \HM{} \PH{bǒnnààkálɪ́} \PS{n.~} \GE{the act of rolling over and over. (donkey having a sun bath)} \TP{bonnakalɛɛ} 
\LX{bonnaŋ} \HM{} \PH{bʋ́nnáŋ} \PS{n.~} \GE{scarcity} \OP{bonnanne} \TP{ bonnannnɛɛ.} 
\LX{bonne} \HM{} \PH{bʋ̀nnɪ̀, bʊ́nnɪ́} \PS{v.~} \GE{to doubt; to speculate, to guess} \PD{bonnɛɛ, bonnɔ, bonnɔ, bonema, bonnaa, bonema } 
\LX{bonneɛ} \HM{} \PH{bʋ́nnɪ́ɛ́} \PS{n.~} \SN \GE{the act of trampling on things} \SN \GE{the act of ironing clothes} \OP{bonneɛne} \TP{bonneɛnɛɛ} 
\LX{bonni} \HM{} \PH{bónní} \PS{v.~} \SN \GE{to burn (pain)} \SN \GE{to burn witth fire} \PD{bonnee, bonno, bonno, bonnaa, bonimo, - } 
\LX{bonnimiri} \HM{} \PH{bʋ́nnímírì} \PS{n.~} \GE{something that is very valuable} \OP{bonnimie} \TP{ bonimiiree.} 
\LX{boŋŋmaa} \HM{} \PH{bʋ́ŋŋmáá} \PS{n.~} \GE{a piece of something} \OP{boŋŋmaara} \TP{ boŋŋmaarɛɛ} 
\LX{boŋŋmaa} \HM{} \PH{bʊ́ŋŋmàà} \PS{n.~} \GE{the practice of cutting things} 
\LX{boŋŋmaa} \HM{} \PH{bʊ́ŋŋmàà} \PS{n.~} \GE{confining of domestic animals that have been ranging freely (usually at the beginning of the farming season)} 
\LX{boŋŋmare} \HM{} \PH{bʋ́ŋŋmárɪ́} \PS{n.~} \GE{the act of breaking things} \OP{boŋŋmarre} \TP{ boŋŋmarɛɛ.} 
\LX{boŋŋmare} \HM{} \PH{bʊ́ŋŋmàrɪ́} \PS{n.~} \GE{something that is spoilt (food)} \TP{boŋŋmarɛɛ} 
\LX{boŋŋmeɛraa} \HM{} \PH{bʊ́ŋŋmɪ́ɛ́ráà} \PS{n.~} \GE{musical instrument} \OP{boŋŋmeɛre} 
\LX{Boŋŋo} \HM{} \PH{bóŋŋò} \PS{n.~} \GE{a traditional festival of Jirapa  traditional area} 
\LX{bonnoɔ} \HM{} \PH{bʋ́nnʋ̀ɔ̀} \PS{n.~} \GE{a sweet thing} \OP{bonnoore} \TP{ bonnoorɛɛ} 
\LX{bonnyaa} \HM{} \PH{bʋ́nɲáà} \PS{n.~} \SN \GE{chest of;} \SN \GE{most important part of} \OP{bonnyaane} \TP{bonnyaanɛɛ} 
\LX{bonnyaanaa} \HM{} \PH{bʋ́nɲáànáà} \PS{n.~} \GE{a clever or wise animal} \OP{bonnyaane} 
\LX{bonnyaara} \HM{} \PH{bʋ́nɲàárɪ̀} \PS{n.~} \GE{coarseness (flour)} \TP{bonnyaarɛɛ} 
\LX{bonnyagere} \HM{} \PH{bʊ́nɲágá} \PS{n.~} \GE{intestines of nonhuman creatures} \OP{bonnyaga} \TP{bonnyagerɛɛ} 
\LX{bonnyaŋaa} \HM{} \PH{bʋ́nɲáŋáá} \PS{n.~} \GE{female} \OP{bonnyaŋene/bonnyaane} \TP{bonnyaanɛɛ} 
\LX{bonnyaŋaa} \HM{} \PH{bònɲàŋáà} \PS{n.~} \GE{female donkey} \OP{boŋnyaŋene} \TP{ boŋnyaŋenɛɛ.} 
\LX{bonnyuu} \HM{} \PH{bʋ́nɲúú} \PS{n.~} \SN \GE{something that smells} \SN \GE{smelling medicinal lotion} \OP{bonnyuuri} \TP{bonyuuree} 
\LX{bonnyuu} \HM{} \PH{bʊ́nɲúú} \PS{n.~} \GE{the neck of something} \OP{bonnyuuri} \TP{bonyuuree} 
\LX{bonnyuuraa} \HM{} \PH{bʋ́nɲúúráá} \PS{n.~} \SN \GE{something that is drunk; a drink} \SN \GE{something that is used for drinking} \OP{bonnyuuri} 
\LX{boŋo} \HM{} \PH{̀bòŋó} \PS{n.~} \GE{donkey} \OP{bonni} \TP{ bonnee.} 
\LX{boŋɔ} \HM{} \PH{bʊ̀ŋɔ́} \PS{adv.~} \GE{done anyhow} \VA{boŋa}
\LX{bɔŋɔ} \HM{} \PH{bɔ̀ŋɔ́} \PS{n.~} \GE{lizard} \OP{bɔnne} \TP{bɔnnɛɛ} \VA{baŋa/bɔŋa}
\LX{bɔŋɔ} \HM{} \PH{bɔ́ŋɔ́} \PS{n.~} \GE{bangle} \OP{bɔnne} \TP{bɔnnɛɛ} \VA{baŋa}
\LX{bonɔbonɔ} \HM{} \PH{bʊ̀nɔ̀bʊ̀nɔ̀} \PS{adv.~} \GE{oily-looking} \XV{A zeɛre zu deɛ e la bonɔbonɔ lɛ.} \XE{The surface of the soup looks oily.} \VA{bonabona}
\LX{bononnoonaa} \HM{} \PH{bʊ̀nʊ́nnʊ́ʊ́náá} \PS{n.~} \GE{a type of tree with juicy fruits} \VA{bodaanonnoonaa}
\LX{bonoone} \HM{} \PH{bʋ̀nʋ̀ʋ̀nɪ̀} \PS{n.~} \GE{a type of tree with fleshy edible fruits; medicinal} \OP{bonooma} \TP{ bonoonɛɛ} 
\LX{bonsaalaa} \HM{} \PH{bʋ́nsààlàà} \PS{n.~} \GE{something that is smooth} \OP{bonsaale} \TP{ bonsaalɛɛ.} 
\LX{bonsaaloŋ} \HM{} \PH{bʊ́nsààlʊ́ŋ} \PS{n.~} \GE{a slippery thing} \OP{bonsaale} \TP{bonsaalɛɛ} 
\LX{bonsaana} \HM{} \PH{bʋ́nsáánà} \PS{n.~} \GE{a new or strange thing} \OP{bonsaama} \TP{ bonsaanɛɛ} 
\LX{bonsaana} \HM{} \PH{bʋ́nsááná} \PS{n.~} \GE{one who destroys} \OP{bonsaanema} 
\LX{bonsaanaa} \HM{} \PH{bʋ́nsáánáá} \PS{n.~} \GE{a spoilt thing} \OP{bonsaane} \TP{ bonsaanɛɛ} 
\LX{bonsammo} \HM{} \PH{bʊ́nsàmmʊ} \PS{n.~} \GE{unshelled nut} \OP{bonsama} \TP{bonsamɛɛ} 
\LX{bonsaraa} \HM{} \PH{bʋ́nsáràà} \PS{n.~} \GE{a young female} \OP{bonsarre} \TP{ bonsarɛɛ} 
\LX{bonsâa} \HM{} \PH{bʋ́nsà̂à᷉} \PS{n.~} \GE{the act of being wasteful, destructive} 
\LX{bonseenaa} \HM{} \PH{bʋ́nsɪ́ɪ́náà} \PS{n.~} \GE{something that is touched} \OP{boseene} \TP{ bonseenɛɛ.} 
\LX{bonseɛraa} \HM{1} \PH{bʊ́nsɪ́ɛ́ráá} \PS{n.~} \SN \GE{something that is worn on the waist} \SN \GE{something that is sewn} \OP{bonseɛre} \TP{ bonseɛrɛɛ} 
\LX{bonseɛraa} \HM{1} \PH{bʊ́nsɪ́ɛ́ráá} \PS{n.~} \GE{that which is better than} \OP{bonseɛre} \TP{ bonseɛrɛɛ} 
\LX{bonseɛraa} \HM{2} \PH{bʊ́nsɪ́ɛ́ráá} \PS{n.~} \GE{something that is roasted} \OP{bonseɛre} \TP{bonseɛrɛɛ} 
\LX{bonsɛgeraa} \HM{} \PH{bʊ́nsɛ́gráá} \PS{n.~} \GE{something that is written} \OP{bonsɛgere} \TP{ bonsɛgerɛɛ} 
\LX{bonsɛgeraa} \HM{} \PH{bʊ́nsɛ̀gráà} \PS{n.~} \GE{something that is used to make another thing sturdy} \OP{bonsɛgere} 
\LX{bonsennaa} \HM{} \PH{bʊ̀nsɪ́nnáà} \PS{n.~} \GE{something that is the right size} \OP{bonsenne} \TP{bonsennɛɛ} 
\LX{bonserɛ} \HM{} \PH{bʋ́nsɪ̀rɛ́} \PS{n.~} \GE{a valuable thing} \OP{bonserre} \TP{ bonserɛɛ/bonserrɛɛ} 
\LX{bonsigiraa} \HM{} \PH{bʋ́nsìgráà} \PS{n.~} \SN \GE{something whose price is reduced} \SN \GE{something that is lowered (height)} \OP{bonsigiri} \TP{bonsigiree} 
\LX{bonso} \HM{} \PH{bʋ́nsʋ́} \PS{n.~} \GE{the act of owning} 
\LX{bonso} \HM{} \PH{bónsó} \PS{adv.~} \GE{why?} \XV{Bonso ka fo ba gaa a daa?} \XE{Why have you not gone to the market?.} 
\LX{bonso} \HM{} \PH{bónsó} \PS{conj.~} \GE{because} \XV{Ba ba gaa a sori bonso a lɔɔre sa᷉aɛ la.} \XE{They could not go on the trip because the car broke down.} 
\LX{bonsɔgelaa} \HM{} \PH{bʋ́nsɔ́gláà} \PS{n.~} \SN \GE{a black thing} \SN \GE{a hidden thing} \OP{bonsɔgelɔ} \TP{ bonsɔgelɛɛ} 
\LX{bonsɔgele} \HM{} \PH{bʊ́nsɔ́glɪ́} \PS{n.~} \GE{the act of hiding things} 
\LX{bonsoŋ} \HM{} \PH{bʋ́nsʋ́ŋ} \PS{n.~} \GE{a good or blessed thing} \OP{bonsonne} \TP{bonsonnɛɛ} 
\LX{bonsooraa} \HM{} \PH{bʊ́nsʊ́ʊ́ráá} \PS{n.~} \GE{a thing that is owned; a belonging} \OP{bonsoore} \TP{bonsoorɛɛ} 
\LX{bonsooraa} \HM{} \PH{bʊ́nsʊ́ʊ́ráà} \PS{n.~} \GE{something that is asked for or wanted} 
\LX{bonsɔraa} \HM{} \PH{bʋ́nsɔ̀ráà} \PS{n.~} \GE{something that is begged for} \OP{bonsɔrre} \TP{bonsɔrɛɛ} 
\LX{bonsɔraa} \HM{} \PH{bʊ́sɔ́ráá} \PS{n.~} \GE{something that has been counted} \XV{A pɔge iri la a bonsɔraa kaŋa.} \XE{The woman took one of the already counted ones.} \OP{bonsɔrre} 
\LX{bonsu} \HM{} \PH{bʋ́nsù} \PS{n.~} \GE{the act of wearing (clothes)} 
\LX{bonsu} \HM{} \PH{bʋ́nsù} \PS{n.~} \GE{the act of pushing things into an opening} 
\LX{bonsugo} \HM{} \PH{bʋ́nsúgò} \PS{n.~} \GE{things in great numbers;} \OP{bonsugri} \TP{ bonsugree} 
\LX{bonsuuraa} \HM{} \PH{bʋ́nsúů̀ráà} \PS{n.~} \GE{garment} \OP{bonsuuri} \TP{ bonsuuree} 
\LX{bontaa} \HM{} \PH{bʋ́ntáà} \PS{n.~} \GE{the act of owning} 
\LX{bontage} \HM{} \PH{bʋ́ntàgɪ́} \PS{n.~} \GE{a thick thing} \OP{bontagre} \TP{ bontagrɛɛ} 
\LX{bontage} \HM{} \PH{bʊ́ntágɪ̀} \PS{n.~} \GE{the act of pulling things} 
\LX{bontale} \HM{} \PH{bʋ́ntálɪ̀} \PS{n.~} \GE{a lobe of  (kola)} \OP{bontala} \TP{bontalɛɛ} 
\LX{bontaŋa} \HM{} \PH{bòtáŋá} \PS{n.~} \GE{camel} \OP{bontanne} \TP{bontannɛɛ} 
\LX{bontara} \HM{} \PH{bʋ́ntárá} \PS{n.~} \GE{rich man; one who owns many things} \OP{bontareba} \TP{bontarebɛɛ.} 
\LX{bonteɛ} \HM{} \PH{bʋ́ntɪ̀ɛ̀} \PS{n.~} \SN \GE{the act of shooting} \SN \GE{the act of supporting things} 
\LX{bonteɛraa} \HM{} \PH{bʋ́ntɪ̀ɛ̀ráà} \PS{n.~} \GE{something that supports something else} \OP{bonteɛre} \TP{bonteɛrɛɛ} 
\LX{bonteɛraa} \HM{} \PH{bʋ́ntɪ̀ɛ̀ráà} \PS{n.~} \GE{a boil} \OP{bonteɛre} \TP{bonteɛrɛɛ} 
\LX{bonteɛraa} \HM{} \PH{bʋ́ntɪ̀ɛ̀ráà} \PS{n.~} \GE{something that is shot at} \OP{bonteɛre} \TP{bonteɛrɛɛ} 
\LX{bonteɛrɛ} \HM{} \PH{bʊ́ntɪ́ɛ́rɛ́} \PS{n.~} \GE{shooter} \OP{bonteɛrebɛ} \TP{bonteɛrebɛɛ} 
\LX{bontɛge} \HM{} \PH{bʊ́tɛ́gɪ̀} \PS{n.~} \GE{the act of exchanging} 
\LX{bontɛgeraa} \HM{} \PH{bʊ́ntɛ̀gráà} \PS{n.~} \GE{something that is traded for} \OP{bontɛgere} \TP{bontɛgerɛɛ} 
\LX{bontɛgerɛ} \HM{} \PH{bʊ́ntɛ́grɛ́} \PS{n.~} \GE{one who exchanges} \OP{bontɛgerebɛ} 
\LX{bontegeroŋ} \HM{} \PH{bʋ́ntɪ́grʋ́ŋ} \PS{n.~} \GE{a heavy thing} \OP{ bontegerɛ} \TP{bontegerɛɛ} 
\LX{bonteraa} \HM{} \PH{bʋ́ntɪ́ráà} \PS{n.~} \GE{something that is given out or offered} \OP{bonterre} 
\LX{bontere} \HM{} \PH{bʋ́ntɪ́rɪ́} \PS{n.~} \GE{the act of giving out things} 
\LX{bontere} \HM{} \PH{bʊ̀ntɪ́rɪ} \PS{n.~} \GE{a big drum made of animal skin and a gourd} \VA{bentere}
\LX{bonto} \HM{} \PH{bʋ́ntʋ́} \PS{n.~} \GE{act of insaulting of things} 
\LX{bonto} \HM{} \PH{bòntó} \PS{n.~} \GE{type of cigar made be the Dagaabe} \OP{bontori} \TP{bontoree} 
\LX{bontobo} \HM{} \PH{bʋ́ntòbó} \PS{n.~} \SN \GE{a type of plant; medicinal} \SN \GE{the immediate tender leaves of a plant (bean leaves as vegetable for soup)} \SN \GE{thing with prominent ears} 
\LX{bontoɔlaa} \HM{} \PH{bʋ́ntʋ́ɔ́láá} \PS{n.~} \GE{that which is sent by a messenger} \OP{bontoɔle/bontoɔlɔ} \TP{bontoɔlɛɛ} 
\LX{bontɔraa} \HM{} \PH{bʊ́ntɔ̀ráà} \PS{n.~} \GE{something that is shared} \OP{bontɔrre} 
\LX{bontɔre} \HM{} \PH{bʋ́ntɔ̀rɪ́} \PS{n.~} \GE{a worthy thing} \OP{bontɔɛ} \TP{bontɔrɛɛ} 
\LX{bontɔre} \HM{} \PH{bʊ́ntɔ̀rɪ̀} \PS{n.~} \GE{the act of giving out or sharing to as many as possible} 
\LX{bontuo} \HM{} \PH{bʋ́ntúó} \PS{n.~} \SN \GE{a bitter thing} \SN \GE{a poisonous or dangerous thing} \OP{bontuuri} \TP{bonturee.} 
\LX{bonuu} \HM{} \PH{bʋ̀núú} \PS{adj.~} \GE{five times} \OP{bonuuri} \TP{bonuuree} 
\LX{bonvaane} \HM{} \PH{bʋ́nvàànɪ́} \PS{n.~} \GE{a shrivelled thing} \OP{bonvaama} \TP{ bonvaanɛɛ.} 
\LX{bonvaraa} \HM{} \PH{bʋ́nváràà} \PS{n.~} \GE{castrated animal} \OP{bonvarre} \TP{ bonvarrɛɛ.} 
\LX{bonvelaa} \HM{} \PH{bʋ́nvɪ̀làá} \PS{n.~} \GE{a good thing} \OP{bonveɛle} \TP{ bonveɛlɛɛ} 
\LX{bonveraa} \HM{} \PH{bʋ́nvɪ̀ráà} \PS{n.~} \SN \GE{an implement used in throwing stones (catapult)} \SN \GE{something that has been shot at by means of a catapult} \OP{bonverre} 
\LX{bonveraa} \HM{} \PH{bʋ́nvɪ̀ráà} \PS{n.~} \GE{a shift of a bone} \OP{bonverre} 
\LX{bonvere} \HM{} \PH{bʋ́nvɪ̀rɪ̀} \PS{n.~} \GE{the act of spraining (body part)} 
\LX{bonvere} \HM{} \PH{bʋ́nvɪ̀rɪ̀} \PS{n.~} \GE{the act of shooting with a catapult} 
\LX{bonvoɔloŋ} \HM{} \PH{bʋ́nvʋ̀ɔ̀lʋ̌ŋ} \PS{n.~} \GE{a slimy thing} \OP{bonvoɔle} \TP{ bonvoɔlɛɛ.} 
\LX{bonvooraa} \HM{} \PH{bʋ́nvʋ́ʋ́ráá} \PS{n.~} \GE{something that breathes} \OP{bonvoore} \TP{ bonvoorɛɛ.} 
\LX{bonvore} \HM{} \PH{bʋ́nvʋ́rɪ̀} \PS{n.~} \GE{a living thing} \OP{bonvoɛ/bonvoba} \TP{bonvorɛɛ} 
\LX{bonvuo} \HM{} \PH{bʋ́nvùó} \PS{n.~} \GE{an open thing; an uncovered enclosure or container} \OP{bonvori} \TP{ bonvoree} 
\LX{bonvuuraa} \HM{} \PH{bʋ́nvúúráá} \PS{n.~} \SN \GE{a thing that crawls (e.g. creature without legs)} \SN \GE{a thing that is dragged along the ground} \OP{bonvuuri} \TP{ bonvuuree.} 
\LX{bonweɛ} \HM{} \PH{bʋ́nwɪ́ɛ́} \PS{n.~} \GE{a common thing} \OP{bonweere} \TP{bonweerɛɛ.} 
\LX{bon-yaga} \HM{} \PH{bʋ́ŋ-yágà} \PS{n.~} \GE{many things} \TP{bon-yagrɛɛ} \VA{boŋyaga}
\LX{bonyaŋaa} \HM{} \PH{bʊ̀nɲáŋáá} \PS{n.~} \GE{female mother goat} \OP{bonyaŋene} \TP{bonyaŋenɛɛ} 
\LX{bonyanlori} \HM{} \PH{bʊ̀ɲánlórí} \PS{n.~} \GE{very large female goat} \TP{bonyanloree} 
\LX{bonyɛ} \HM{} \PH{bʋ̀ɲɛ́} \PS{n.~} \GE{wicked ideas} \XV{A dɔɔ suuri da iri ne la o pɔge kyɛ o da ba de bonyɛ tuuro ne o.} \XE{The man was annoyed with his wife but he did not have any wicked ideas towards her.} 
\LX{bonyɛ} \HM{} \PH{bʋ̀ɲɛ́} \PS{adj.~} \GE{having wicked ideas} 
\LX{bon-yɛɛ} \HM{} \PH{bʋn-yɛɛ} \PS{n.~} \SN \GE{unimportant things} \SN \GE{wild crops} \SG{bon-yɛroo} \TP{bon-yɛrɛɛ} \VA{boŋyɛɛ}
\LX{bon-yeɛle} \HM{} \PH{bʋ́n-yɪ́ɛ́lɪ́} \PS{n.~} \GE{winnowed thing} 
\LX{bon-yɛraa} \HM{} \PH{bʋ́n-yɛ̀ráà} \PS{n.~} \SN \GE{something that is hung around the neck} \SN \GE{garment} \OP{bon-yɛrre} \TP{bon-yɛrrɛɛ} 
\LX{bon-yɔmɔ} \HM{} \PH{bʋ́n-yɔ̀mɔ́} \PS{n.~} \GE{plentiful and free thing(s)} \TP{ bon-yɔmɛɛ.} 
\LX{bon-yoɔraa} \HM{} \PH{bʋ́n-yʋ́ɔ́ráá} \PS{n.~} \SN \GE{something that roams or wanders about} \SN \GE{rheumatism} \OP{bon-yoɔre} \TP{ bon-yoɔrɛɛ} 
\LX{bon-yuo} \HM{} \PH{bʋ́n-yúó} \PS{n.~} \GE{different one} \XV{A bie yage la  a kparoo na bare kyɛ de bonyuo.} \XE{The boy took of that shirt and took a different one.} \OP{bon-yuobo} \VA{boŋyuo}
\LX{bon-yuori} \HM{} \PH{bún-yúórì} \PS{n.~} \GE{a popular or prominent thing} \OP{bon-yoe} \TP{ bon-yoree.} 
\LX{bonzaa} \HM{} \PH{bʊ̂nzàá} \PS{n.~} \SN \GE{anything, everything} \XV{Fo na nyɛ la bonzaa fo naŋ boɔrɔ.} \XE{You will get anything that you want.} \SN \GE{everyone} \XV{Bonzaa naŋ wa kyɛ taa la puoraa.} \XE{Thanks to everyone who has come here.} 
\LX{bonzagelaa} \HM{} \PH{bʊ́nzágláá} \PS{n.~} \GE{an empty thing} \OP{bonzagele} 
\LX{bonzeɛ} \HM{} \PH{bʊ́nzɪ̀ɛ̀} \PS{n.~} \GE{a red thing} \OP{bonzeere} \TP{bonzeerɛɛ} 
\LX{bonzoolɔ} \HM{} \PH{bʊ́nzʊ́ʊ́lɔ́} \PS{n.~} \GE{a thing with hairy ends (corncob)} \TP{bonzoolɛɛ} \VA{bonzoola}
\LX{boo} \HM{} \PH{bóò} \PS{pron.~} \GE{} \VA{buori}
\LX{boo} \HM{} \PH{bʊ̀ʊ̀} \PS{adj.~} \GE{voluminous (dress)} 
\LX{boɔ} \HM{} \PH{bʊ́ɔ́} \PS{n.~} \GE{goat} \OP{boore} \TP{boorɛɛ} 
\LX{bɔɔ} \HM{} \PH{bɔ̀ɔ̀} \PS{v.~} \GE{to dress very well} \PD{bɔɔɛ, bɔɔrɔ, bɔɔrɔ, bɔɔrebɔ, bɔɔraa, bɔɔrebɔ } 
\LX{boɔbo} \HM{} \PH{bʋ́ɔ́bʋ̀} \PS{n.~} \SN \GE{something that is preferred} \SN \GE{the act of liking or searching for something} 
\LX{bɔɔderoo} \HM{} \PH{bɔ̀ɔ̀dɪ́rʊ́ʊ́} \PS{n.~} \GE{plantain} \SG{bɔɔderoo} \OP{bɔɔdeɛ} \TP{bɔɔdeerɛɛ} 
\LX{bɔɔdoa} \HM{} \PH{bɔ̀ɔ̀dʊ́áá} \PS{n.~} \GE{towel} \OP{bɔɔdoore} \TP{bɔɔdoorɛɛ} 
\LX{boɔlaa} \HM{} \PH{bʋ̀ɔ̀làá} \PS{n.~} \SN \GE{call} \SN \GE{invitation} 
\LX{boɔle} \HM{} \PH{bʊ̀ɔ̀lɪ̀} \PS{v.~} \SN \GE{to call} \SN \GE{to invite} \PD{boɔlɛɛ, boɔlɔ, boɔlɔ, boɔleba, boɔlaa, boɔleba } 
\LX{boolimbo} \HM{} \PH{bòòlímbò} \PS{n.~} \GE{taking one by the shoulders and chest (patient) (typically used with kpɛ̧)} \OP{boolimbori} \TP{boolimboree} 
\LX{boɔraa} \HM{} \PH{bʋ̀ɔ̀ráà} \PS{n.~} \SN \GE{sack} \SN \GE{one hundred pounds} \OP{boɔre} \TP{boɔrɛɛ} 
\LX{boore} \HM{} \PH{bʋ̀ʋ̀rɪ̀} \PS{v.~} \GE{to attack suddenly and fiercely, to rush on} \PD{boorɛɛ, boorɔ, boorɔ, boorebɔ, booraa, boorebɔ } 
\LX{boore} \HM{} \PH{bʊ̌ʊ́rɪ} \PS{n.~} \GE{of same stalk, species, descendant} \OP{boora} \TP{boorɛɛ} \VA{booree}
\LX{boore} \HM{} \PH{bʊ̀ʊ̀rɪ́} \PS{n.~} \GE{aggrement, plan} \TP{boorɛɛ} 
\LX{boɔre} \HM{} \PH{bʋ̀ɔ̀rɪ́} \PS{v.~} \GE{to smear with thick paste} \PD{boɔrɛɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ } 
\LX{bɔɔre} \HM{} \PH{bɔ̀ɔ̀rɪ́} \PS{n.~} \GE{sacrifice, deity} \OP{bɔbɔ} \VA{bɔgre}
\LX{boorɔ} \HM{} \PH{bʊ́ʊ́rɔ́} \PS{n.~} \SN \GE{off same type, family} \SN \GE{seed for planting} \SG{booroo} \VA{boora}
\LX{boɔrɔ} \HM{} \PH{bʊ́ɔ́rɔ̀} \PS{v.~} \SN \GE{to like, to love; to desire, to want} \SN \GE{to need} \PD{bɔɛ, boɔrɔ, boɔrɔ, boɔreba, boɔraa, - } 
\LX{boɔso} \HM{} \PH{bʋ́ɔ́sʋ́} \PS{n.~} \GE{bus} \OP{boɔsere} \TP{ boɔserɛɛ} 
\LX{bopeɛ} \HM{} \PH{bʋ̀pɪ́ɛ́} \PS{n.~} \GE{traditional storey building} \OP{bopeere} \TP{bopeerɛɛ} \VA{peɛzudie}
\LX{boraa} \HM{} \PH{bʊ́ráá} \PS{n.~} \GE{a stick used in sowing grains} \OP{borre} \TP{borrɛɛ} \VA{boredaa}
\LX{bore} \HM{} \PH{bʊ̀rɪ̀} \PS{v.~} \GE{to break part of a lump off in big chunks} \PD{borɛɛ, bora, bora, boreba, boraa, - } 
\LX{bore} \HM{} \PH{bʊ̀rɪ̀} \PS{v.~} \GE{to sow, to plant} \PD{borɛɛ, bora, bora, boreba, boraa, boreba } 
\LX{bɔre} \HM{} \PH{bɔ́rɪ̀} \PS{v.~} \GE{to break off a large chunk (from a bigger mass)} \PD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔreba, bɔraa, bɔreba } 
\LX{bɔre} \HM{} \PH{bɔ̀rɪ̀} \PS{v.~} \GE{to blossom, to flourish (plants)} \PD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔrebɔ, bɔraa, - } 
\LX{bɔre} \HM{} \PH{bɔ̀rɪ̀} \PS{v.~} \GE{to get lost, to go astray; to make a mistake} \PD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔrebɔ, bɔraa, - } 
\LX{bɔre} \HM{} \PH{bɔ́rɪ́} \PS{n.~} \GE{puff adder; a thick large poisonous snake} \OP{bɔrre} \TP{bɔrɛɛ} \VA{bɔsere}
\LX{boredaa} \HM{} \PH{bʋ̀rɪ̀dàá} \PS{n.~} \GE{a stick for sowing} \OP{boredaare} \TP{ boredarɛɛ/boredaarɛɛ.} 
\LX{bɔreferoo} \HM{} \PH{bɔ̀rɪ́fɪ́rʊ́ʊ́} \PS{n.~} \GE{pawpaw} \OP{bɔreferre} \TP{bɔreferrɛɛ} 
\LX{Bɔrefɔ} \HM{} \PH{bɔ̀rɪ̀fɔ́} \PS{n.~} \GE{English} \OP{bɔrefɔre} \TP{ bɔrefɔrɛɛ} \VA{bɔrɔ/naasaaleŋ}
\LX{boreŋmaakpantoli} \HM{} \PH{bʊ̀rɪ̀ŋmàáakpántólí} \PS{n.~} \GE{a type of large, beautiful bird with a sweet voice} \OP{boreŋmaakpantolo} \TP{boreŋmaakpantolee} 
\LX{boreŋmaao} \HM{} \PH{bʋ̀rɪ̊ŋmààʋ́} \PS{n.~} \GE{a large bird with a strong sweet voice} \OP{boreŋmaane} \TP{ boreŋmaanɛɛ.} 
\LX{boreŋmanlee} \HM{} \PH{bʊ̀rɪ̀ŋmánléé} \PS{n.~} \GE{calabash for sowing} \OP{boreŋmambili} \TP{boreŋmambilee} 
\LX{borenyɛ} \HM{} \PH{bʋ́rɪ́ɲɛ̀} \PS{n.~} \GE{Christmas} \OP{borenyɛre} \TP{ borenyɛrɛɛ} 
\LX{boribara} \HM{} \PH{bòrìbàrà} \PS{adv.~} \GE{done poorly} 
\LX{boribori} \HM{} \PH{bòrìbórí} \PS{v.~} \GE{to carry out poorly (cleaning, talk)} \PD{boriboree, boriboro, boriboro, boriboribo, boriboraa, - } 
\LX{borɔbɛ} \HM{} \PH{bʊ̀rɔ̀bɛ́} \PS{n.~} \GE{pineapple fruit or plant} \OP{borɔbɛre} \TP{borɔbɛrɛɛ} 
\LX{boroboro} \HM{} \PH{bòròbòrò} \PS{n.~} \GE{bread} \OP{boroborri} \TP{boroboree} 
\LX{borɔborɔ} \HM{} \PH{bʋ̀rɔ̀bʋ̀rɔ̀} \PS{adj.~} \GE{multicoloured} 
\LX{boroborozɔŋ} \HM{} \PH{bòròbòròzɔ̂ŋ} \PS{n.~} \GE{bread flour} \TP{boroborozɔnnɛɛ} 
\LX{borɔso} \HM{} \PH{bʋ̀rɔ́sʋ̀} \PS{n.~} \GE{ brush} \OP{borɔsore} \TP{ borɔsorɛɛ} 
\LX{boroto} \HM{} \PH{bòròtò} \PS{adj.~} \GE{ big and round} 
\LX{boroto} \HM{} \PH{bʊ̀rʊ̀tʊ̊} \PS{adv.~} \GE{fall flat} \XV{A bie zoro gɛrɛ te ŋmɛ gbɛre a gaa te boroto o naŋ le.} \XE{As the boy ran along he stumbled on a stone and crashed to the ground.} 
\LX{bosaraa} \HM{} \PH{bʋ̀sáràà} \PS{n.~} \GE{young female goat} \OP{bosarre} \TP{bosarrɛɛ} 
\LX{bosere} \HM{} \PH{bʋ̀sɪ́rɪ̀} \PS{n.~} \GE{a necklace made of huge beads} \OP{boseɛ/boserre} \TP{boserɛɛ} 
\LX{bovammo} \HM{} \PH{bʋ̀vàmmʋ́} \PS{n.~} \GE{a type of bow, usually not made from bamboo} \OP{bovama} \TP{ bovannɛɛ.} 
\LX{bovaraa} \HM{} \PH{bʋ̀váràá} \PS{n.~} \GE{castrated goat} \OP{bovarre} \TP{ bovarrɛɛ} 
\LX{bowola} \HM{} \PH{bʊ́wʊ́lá} \PS{adv.~} \GE{how many times} 
\LX{boyaa} \HM{} \PH{[bóyáá} \PS{interj.~} \GE{expression of surprise} \VA{boyaaŋ}
\LX{boyaaŋ} \HM{} \PH{bòyááŋ} \PS{interj.~} \GE{expression of surprise} 
\LX{boyoɔbo} \HM{} \PH{bʋ̀yʋ̀ɔ́bʋ́} \PS{adj.~} \GE{six times} 
\LX{boyɔpôi} \HM{} \PH{bʋ̀yɔ̀pó̂í} \PS{adj.~} \GE{seven times} 
\LX{bozeɛ} \HM{} \PH{bʊ̀zɪ́ɛ́} \PS{n.~} \GE{dog} \OP{bozeere} \TP{bozeerɛɛ} 
\LX{bɔ᷉ee} \HM{} \PH{bɔ̀᷉ɪ̀᷉̂ɪ̀᷉} \PS{adv.~} \GE{calmly, slowly} \XV{O yeli la ayɛlɛ bɔ᷉ee lɛ.} \XE{She spoke gently.} 
\LX{bo᷉o} \HM{} \PH{bʋ̀᷉ʊ̀᷉} \PS{adv.~} \GE{oily-looking (top of sauce)} 
\LX{bo᷉ɔ} \HM{} \PH{bʊ̀᷉ɔ̀᷉} \PS{v.~} \GE{to scoop up a large amount of something} \XV{ O bo᷉ɔ la a ka᷉a poli o eŋɛ zaa.} \XE{She scooped up a large amount of lotion and rubbed it all over her body.} \PD{bo᷉ɔɛ, boɔnɔ, boɔnɔ, boɔnema, boɔnaa, - } 
\LX{bo᷉ɔ} \HM{} \PH{bʊ̀᷉ɔ̀᷉} \PS{v.~} \GE{snub} \XV{A dɔɔ suuri iree la ka o maŋ deɛ bo᷉ɔ neɛ te lɔɔ.} \XE{The man is so annoyed that he snubs anybody he comes across.} \PD{bo᷉ɔɛ, boɔnɔ, boɔnɔ, boɔnema, boɔnaa, - } 
\LX{bu} \HM{} \PH{bù} \PS{v.~} \SN \GE{to get wet} \XV{Saa poɔ la a tontonema ka ba bonsuuri zaa bu.} \XE{The rain beat down on the workers and soaked all their clothes.} \SN \GE{to mix dry powder with liquid or oil (mortar)} \XV{A demeɛrɛ bu la tɛne na mɛ a dankyini.} \XE{The mason has prepared mortar to build the wall with.} \PD{bue, buuro, buuro, buuribo, buuraa, buuribo } 
\LX{bu} \HM{} \PH{bú} \PS{v.~} \GE{to measure (liquid), to calculate} \XV{Ba bu la ka᷉a᷉ bɔlembɔre ata eŋ a doge poɔ.} \XE{The have measured out three bottles of oil into the pot.} \PD{bue, buuro, buuro, buuribo, buuraa, buuribo } 
\LX{bu} \HM{} \PH{bú} \PS{v.~} \SN \GE{to ponder, to reflect} \XV{N bu la a yɛlɛ n zu poɔ lɛɛ kyɛ N naŋ ba baŋ a pare.} \XE{I've pondered over the issue for some time now, but I can't seem to come to terms with it.} \SN \GE{to respect , to value} \XV{Ka fooŋ ba bu ma N meŋ koŋ bu fo.} \XE{If you don't respect me, I wont respect you either.} \PD{bue, buuro, buuro, buuribo, buuraa, buuribo } 
\LX{bufuli} \HM{} \PH{bùfúlí} \PS{n.~} \GE{a blood-sucking maggot-like worm, usually found under floors} \OP{bufulo} \TP{bufulee} 
\LX{bugagu} \HM{} \PH{bùgàgú} \PS{n.~} \GE{a type of tree, medicinal} \OP{bugaguri} \TP{bugaguree} 
\LX{bugi} \HM{} \PH{bùgì} \PS{v.~} \SN \GE{to be confused, to be perplexed} \SN \GE{to be drunk} \XV{O ya᷉a᷉ nyu la da᷉a᷉ bugi.} \XE{As usual, he has been drinking and is drunk.} \PD{bugee., bugiro, bugiro, bugiribo, bugiraa, - } 
\LX{bugi} \HM{} \PH{búgí} \PS{v.~} \GE{to pound into powder} \XV{Ba bugi a kyooma na tɔ ka᷉a᷉.} \XE{The shea nuts have been pounded into powder in order to prepare shea butter.} \PD{bugee, bugro, bugro, bugribo, bugraa, - } 
\LX{bugi} \HM{} \PH{búgí} \PS{v.~} \GE{to make fresh mounds (farming)} \XV{A koɔraa bugi la o weɛ yuoni ŋa azuiŋ o bɛŋɛ wɔŋɛɛ la soŋ.} \XE{The farmer made fresh mounds; consequently, he had a good yield of beans.} \PD{bugee, bugro, bugro, bugribo, bugraa, - } 
\LX{bugikpara} \HM{} \PH{bùgìkpárá} \PS{n.~} \GE{a moulded granary with two apartments} \OP{bugkparre} \TP{bugkparɛɛ} 
\LX{bugili} \HM{} \PH{búgílì} \PS{v.~} \GE{to tend (seed)} \PD{bugilee, bugilo, bugilo, buglibo, bugilaa, - } \VA{buuli/buoli}
\LX{bugiliŋe} \HM{} \PH{bùglíŋé} \PS{n.~} \GE{lid for a barn or granary} \OP{bugilinni} \TP{bugilinnee} 
\LX{bugimoɔ} \HM{} \PH{bùgìmʊ́ɔ́} \PS{n.~} \GE{a type of grass, the straw of which is mixed with clay to build a granary} \SG{bugimoɔnoo} \TP{bugimoɔnɛɛ} 
\LX{bugipelaa} \HM{} \PH{bùgìpɪ̀làá} \PS{n.~} \GE{a small granary} \OP{bugipeɛle} \TP{ bugipeɛlɛɛ} 
\LX{bugita} \HM{} \PH{búgítà} \PS{n.~} \GE{kettle} \OP{bugitare} \TP{ bugitarɛɛ} 
\LX{bugo} \HM{} \PH{bùgó} \PS{n.~} \GE{granary} \OP{bugri} \TP{bugree} 
\LX{bukulo} \HM{1} \PH{búkúló} \PS{adv.~} \GE{perish completely} \XV{ A vu᷉u᷉ di la  yiri zaa bukulo.} \XE{Fire has burnt the house completely.} 
\LX{bulee} \HM{} \PH{búléé} \PS{n.~} \GE{well} \OP{bulli} \TP{ bullee} 
\LX{buli} \HM{} \PH{búlí} \PS{v.~} \SN \GE{to germinate, to sprout, to grow out from} \XV{Ba bore la kyi kyɛ a naŋ ba buli sɛre.} \XE{They have sown millet, but it has not germinated yet.} \SN \GE{to spring up, to ooze (water)} \XV{Te tu la a bulee ka o sigi kyɛ o naŋ ba buli ko᷉ɔ᷉.} \XE{We have dug the well quite deep, but water has not started to flow.} \PD{bulee, bulo, bulo, bulibo, bulaa, bulibo } 
\LX{buli} \HM{} \PH{búlí} \PS{n.~} \GE{sprout} \OP{bulo} 
\LX{buliba} \HM{} \PH{búlíbà} \PS{v.~} \GE{to run fast} \PD{bulibaɛ, bulobaara, bulibaara, bulibaareba, bulibaaraa, - } 
\LX{bulinimiri} \HM{} \PH{bùlnímírì} \PS{n.~} \GE{the point in a well from which water flows out} \OP{bulinimie} \TP{bulinimiiree} 
\LX{bulli} \HM{} \PH{bùllì} \PS{v.~} \GE{to collect bits of water flowing from a well or hole (e.g. during draught)} \PD{bullee, bullo, bullo, bulibo, bullaa, bulibo } 
\LX{bulo} \HM{} \PH{bùlò} \PS{n.~} \GE{soup from seed or nuts (beans, groundnuts)} \TP{bulee/bullee} 
\LX{bulugiri} \HM{} \PH{bùlùgrí} \PS{n.~} \GE{a container made from a gourd} \OP{bulugo} \TP{bulugiree} \VA{buluuri}
\LX{buluuri} \HM{} \PH{bùlùùrí} \PS{n.~} \GE{a fat goat} \TP{buluuree.} \VA{bulugiri}
\LX{bumbunturi} \HM{} \PH{bùmbùntùrì} \PS{n.~} \GE{a type of tree with edible fruits} \OP{bumbunturri} \TP{bumbunturee} 
\LX{buŋi} \HM{} \PH{búŋ} \PS{v.~} \GE{to get mouldy (seed)} \PD{buŋee, buŋino, buŋino, buŋinimo, buŋinaa, buŋinimo } 
\LX{buo} \HM{} \PH{búò} \PS{pron.~} \GE{which one} \VA{buori}
\LX{buoli} \HM{} \PH{búólì} \PS{v.~} \GE{to nurse (a seedling)} \PD{buolee, buolo, buolo, buolibo, buolaa, buolibo } \VA{buuli/bugili}
\LX{buori} \HM{} \PH{bùòrì} \PS{v.~} \GE{to hurt} \PD{buoree, buoro, buoro, buoribo, buoraa, buoribo } 
\LX{buori} \HM{} \PH{bùórì} \PS{pron.~} \GE{which} \VA{buo/booː abuo/abobo}
\LX{buri} \HM{} \PH{bùrì} \PS{v.~} \GE{to soften by moistening} \XV{O buri la a gane a na to᷉ɔ᷉sɛ a woɔ.} \XE{He softened the leather so he could sew the bag.} \PD{buree, buro, buro, buribo, buraa, buribo } 
\LX{buri} \HM{} \PH{bùrì} \PS{v.~} \GE{to gain weight} \XV{Beri ayi ŋa a dɔɔ ŋmeene na leɛ buri la o zaa.} \XE{Lately, that lean man has been putting on weght.} \PD{buree, buro, buro, buribo, buraa, buribo } 
\LX{buri} \HM{} \PH{búrí} \PS{v.~} \GE{to immerse a container in liquid or powder to fill it} \PD{buree, buro, buro, buribo, buraa, buro } 
\LX{buriburi} \HM{} \PH{bùrìbúrí} \PS{v.~} \GE{to be confused} \PD{buriburee, buriburo, buriburo, buriburibo, buriburaa, buriburibo } 
\LX{buriburi} \HM{} \PH{búríbúrí} \PS{n.~} \GE{state of confusion} 
\LX{burimo} \HM{} \PH{bùrìmó} \PS{n.~} \GE{a type of wild animal the size of a zebra} \OP{burri} \TP{burree} 
\LX{buro} \HM{} \PH{bùró} \PS{adj.~} \GE{in dry powdered form (flour)} \XV{O nɔŋ la zɔmburo mogebo.} \XE{He likes eating unsoaked flour.} \OP{burri} \TP{burree} 
\LX{buroŋ-buroŋ} \HM{} \PH{búróŋ-búróŋ} \PS{adv.~} \GE{done haphazardly (without order)} \XV{A yɛlɛ deɛ waa la buromburoŋ lɛ.} \XE{The case is just in a confused state.} \VA{/bùròmbùròŋ/}
\LX{buroni} \HM{} \PH{bùròní} \PS{n.~} \GE{white man} \OP{buroniri} \TP{buroniree} 
\LX{burri} \HM{} \PH{búrrì} \PS{v.~} \GE{to fool, to outwit} \XV{imperf. .perf. . agent:} \XE{} \PD{burree, burro, burro, buribo, burraa, burro } 
\LX{buseli} \HM{} \PH{búsélí} \PS{n.~} \GE{very heavy object} \OP{busele} \TP{ buselee} 
\LX{buulaa} \HM{} \PH{bùùláá} \PS{n.~} \GE{playground} \OP{buuli} \TP{buulee} 
\LX{buuli} \HM{} \PH{bùùlí} \PS{n.~} \GE{an insect that makes honey in the ground} \OP{buulo} \TP{buulee} \VA{wirikuoluŋ}
\LX{buuli} \HM{} \PH{búúlí} \PS{n.~} \GE{bells worn on the wrist to play xylophone} \OP{buulo} \TP{buulee} 
\LX{buuli} \HM{} \PH{búúlì} \PS{v.~} \GE{nurse (seed)} \XV{O buuli la kommie a dabɔge poɔ.} \XE{She has nursed tomoto seed in the garden.} \PD{buulee, buulo, buulo, buulibo, buulaa, buulibo } \VA{buoli/bugili}
\LX{buuli} \HM{} \PH{bùùlì} \PS{v.~} \GE{to boil (first stage in making porridge paste; saabo)} \PD{buulee, buulo, buulo, buulibo, buulaa, buulo } 
\LX{buuli} \HM{} \PH{bùùlì} \PS{v.~} \GE{to murmur, to grumble} \XV{Ka o poɔŋ sa᷉a᷉ o maŋ buuli la yɛlɛ yaga.} \XE{When she is sad, she grumbles a lot.} \PD{buulee, buulo, buulo, buulibo, buulaa, buulo } 
\LX{buulo} \HM{} \PH{búúló} \PS{n.~} \GE{something that complements or adorns another thing} \XV{Nɔtoro e la dɔɔ noɔre buulo.} \XE{A mustache complements a man's mouth.} 
\LX{buuluŋ} \HM{} \PH{búúlúŋ} \PS{n.~} \GE{porridge} \TP{buulunee} 
\LX{buuluŋ} \HM{} \PH{bùúlúŋ} \PS{n.~} \GE{fleshy portion (meat)} \OP{buulo} \TP{buulee} 
\LX{buulunwala} \HM{} \PH{bùùlùnwálá} \PS{n.~} \GE{light porridge} \TP{buulunwalɛɛ} 
\LX{buulunwoɔlɔ} \HM{} \PH{bùùlùnwʊ́ɔ́lɔ́} \PS{n.~} \GE{porridge used in the preparation of TZ} \TP{buulunwoɔlɛɛ} \VA{buulunwoɔla}
\LX{da} \HM{} \PH{dà} \PS{part.~} \GE{past tense marker} \XV{Ba da wa la kyɛ.} \XE{They came here.} 
\LX{da} \HM{} \PH{dà} \PS{n.~} \GE{time in the past; olden days} 
\LX{da} \HM{} \PH{dá} \PS{v.~} \SN \GE{to push (pl)} \SN \GE{to blow (wind)} \SN \GE{to wander (in the bush)} \XV{A naŋkpaana da la moɔ yaga kyɛ ba nyɛ nɛne.} \XE{The hunter wandered all day but did not get any game.} \PD{daɛ, daara, daara, daareba, daaraa, daareba } \SE{da lɔɔ} \GE{to push or blow down} 
\LX{da} \HM{} \PH{dà} \PS{v.~} \GE{to buy} \PD{daɛ, daara, daara, dareba, daaraa, daareba } 
\LX{daa} \HM{} \PH{dàá} \PS{n.~} \GE{stick, piece of wood} \OP{daare} \TP{darɛɛ} 
\LX{daa} \HM{} \PH{dàá} \PS{n.~} \GE{market; place for buying and selling} \XV{Ba maŋ ti la sampare daa poɔ.} \XE{Sheds are often put up at the marketplace.} \SN \GE{day or occasion for buying and selling} \XV{Wa daa ane Kaane daa la maŋ laŋ kpɛ.} \XE{The Wa market day and Kaane market day fall on the same day.} \OP{daare} \TP{darɛɛ} 
\LX{daa} \HM{} \PH{dàá} \PS{n.~} \GE{week (six-day)} \XV{O naŋ wa kyɛ, o daare ayi la.} \XE{It is two weeks since he came here.} \OP{daare} \TP{darɛɛ} 
\LX{daa} \HM{} \PH{dàá} \PS{n.~} \GE{cost, price} \XV{Se᷉ɛ᷉ saŋa bondirii daa maŋ do la.} \XE{The cost of food rises during the middle of the farming season.} \OP{daare} \TP{darɛɛ} \VA{daaroŋ}
\LX{daa} \HM{} \PH{dàá} \PS{n.~} \GE{gun} \OP{daare} \TP{darɛɛ} 
\LX{daa} \HM{} \PH{dáá} \PS{adv.~} \GE{last time} \XV{O daa wa la kyɛ?} \XE{Did s/he come here last time?.} \VA{daare}
\LX{daa} \HM{} \PH{dáà} \PS{v.~} \GE{to push} \PD{daaɛ, daara, daara, dareba, daaraa, dareba } 
\LX{daa} \HM{} \PH{dáá} \PS{adj.~} \GE{male (nonhuman)} \OP{daare/daɛ} \TP{darɛɛ/daarɛɛ} 
\LX{daadaa} \HM{} \PH{dáádáá} \PS{adv.~} \GE{straight ahead} 
\LX{daa-daa} \HM{} \PH{dàá-dàá} \PS{adv.~} \GE{weekly; from time to time} \XV{Daa-daa ka a dɔɔta maŋ wa.} \XE{The doctor comes every week.} 
\LX{daale} \HM{} \PH{dààlɪ̀} \PS{v.~} \GE{to float; to be suspended (on liquid)} \XV{Ka fooŋ kyiri ka᷉a᷉ eŋ ko᷉ɔ᷉ poɔ a maŋ iri daale la.} \XE{If oil is poured into water it will float on the top.} \PD{daalɛɛ, daala, daala, daaleba, daalaa, daaleba } 
\LX{daama} \HM{} \PH{dáámá} \PS{n.~} \GE{state of ease; freedom} \XV{Ka a toma ŋaŋ wa baare N na nyɛ la daama fẽẽ.} \XE{When this work is over I shall be free a bit.} \TP{daamɛɛ} 
\LX{Daana} \HM{} \PH{dááná} \PS{n.~} \GE{God} 
\LX{daana} \HM{} \PH{dááná} \PS{n.~} \GE{owner, lord, master} \OP{daanema} \TP{daanemɛɛ} \VA{daŋena}
\LX{daane} \HM{} \PH{dàànɪ̀} \PS{v.~} \SN \GE{to change, to make different} \XV{O naŋ gaa o sori na te  waana la ka o daane (la) o zaa.} \XE{Returning from his trip he had changed completely.} \SN \GE{to steer, to redirect, to turn away} \XV{O poɔ sa᷉aɛ la ka o daane o niŋe bare ziyuo.} \XE{She was sad and turned her face away.} \PD{daanɛɛ, daana, daana, daanema, daanaa, daanema } 
\LX{daanyu} \HM{} \PH{dá᷉á᷉̂ɲú} \PS{n.~} \GE{the act of drinking alcohol} 
\LX{daanyuuro} \HM{} \PH{dá᷉á᷉́̂ɲúúrò} \PS{n.~} \SN \GE{an alcoholic, a habitual drinker of alcohol} \SN \GE{a person who is drinking alcohol} \OP{daanyuuribo} \TP{daanyuribee} 
\LX{daare} \HM{} \PH{dààrɪ̀} \PS{v.~} \GE{to get out of control. (bad behaviour)} \XV{Ka bibileŋ ba nyɛ wuluu velaa o bile saŋa, o maŋ daare la.} \XE{If a child does not get good guidance during his early life, he gets out of hand.} \PD{daarɛɛ, daara, daara, daareba, daaraa, daareba } 
\LX{daare} \HM{} \PH{dààrɪ́} \PS{n.~} \GE{firewood (collective)} \TP{daarɛɛ} 
\LX{daare} \HM{} \PH{dáárɪ́} \PS{adv.~} \SN \GE{day before yesterday;} \SN \GE{last time} \VA{daa}
\LX{daaroŋ} \HM{} \PH{dààrʋ́ŋ} \PS{n.~} \GE{cost of} \XV{Uoni saŋa kommie daaroŋ maŋ e la kpeɛŋaa.} \XE{During the dry season the cost of tomatoes goes high.} 
\LX{daayi} \HM{} \PH{dááyí} \PS{v.~} \GE{to stand out} \PD{daa yiee, daara yire, daayire, daayiribe, daayiraa, - } 
\LX{dabaala} \HM{} \PH{dàbáálà} \PS{n.~} \GE{effort, struggle} \TP{dabaalɛɛ} \VA{dabagla}
\LX{dabaaraa} \HM{} \PH{dàbááráá} \PS{n.~} \GE{a three-forked stick on which pots or calabashes are placed} \OP{dabaare} \TP{dabaarɛɛ.} 
\LX{dabagelakala} \HM{} \PH{dàbáglàkálà} \PS{interj.~} \GE{an expression of surprise at something great} \XV{A pɔge to᷉ɔ lɔɔ la a dɔɔ, dabagelakala.} \XE{The woman was able to throw the man down; that's great!.} 
\LX{dabaŋkalɛ} \HM{} \PH{dàbàŋkálɛ́} \PS{n.~} \GE{looking back with dissatisfaction; had I known, regret, remorse} \XV{Selaa de zeɛ la a nɔlee kaŋa, ka N da baŋ ka lɛ N pɔge a.} \XE{A hawk has snatched one of the chickens; if I had known, I would have put them in a pen.} 
\LX{dabaŋkoroŋ} \HM{} \PH{dàbàŋkʊ́rʊ́ŋ} \PS{n.~} \GE{prior knowledge} \XV{A seŋ ka karemamine maŋ de karembiiri dabaŋkoroŋ wuli ba.} \XE{Teachers should base their teaching on pupils' prior knowledge.} 
\LX{dabeɛmɛ} \HM{} \PH{dàbɪ̀ɛ́mɛ̀} \PS{n.~} \GE{coward} \XV{Nyɛ, a dɔɔŋ leɛ zoro, dabeɛmɛ la.} \XE{Look, the man is running away, he is a coward.} \TP{dabeɛmɛɛ} 
\LX{dabegere} \HM{} \PH{dábɪ́gɪ́rɪ́} \PS{n.~} \SN \GE{piece of, portion} \SN \GE{paragraph} \OP{dabegɛ} \TP{dabegerɛɛ} 
\LX{dabegere} \HM{} \PH{dábɪ́gɪ́rɪ́} \PS{n.~} \SN \GE{a type of white ant} \SN \GE{the nest of such ants} \OP{dabegɛ} \TP{dabegerɛɛ} 
\LX{dabe᷈e} \HM{} \PH{dàbɪ́᷉ɪ́᷉} \PS{n.~} \SN \GE{juice} \SN \GE{semen} \OP{dabeene} \TP{dabeemɛɛ/dabeenɛɛ} 
\LX{dabe᷈ɛ} \HM{} \PH{dàbɪ́᷈ɛ́} \PS{n.~} \SN \GE{fear} \SN \GE{danger, fearsome} \XV{A vu᷉uzele taa la dabe᷉ɛ.} \XE{The burning flame is looking dangerous.} \OP{dabeɛmɛ} \TP{dabeɛnɛɛ} 
\LX{dabilikpeo} \HM{} \PH{dàbìlkpɪ́ʋ́} \PS{n.~} \GE{a type of tree with medicinal properties and edible fruits} 
\LX{dabilkaŋkanyogelɔ} \HM{} \PH{dàbìlkáŋkáɲʊ́glɔ́} \PS{n.~} \GE{small twigs used as firewood, kindling} \TP{dabilkaŋkanyoglɛɛ} \VA{dabilkaŋkanyogla}
\LX{dabiltuo} \HM{} \PH{dàbìltúó} \PS{n.~} \GE{a bitter herb used to treat abdominal problems} \OP{dabiltuuri} \TP{dabiltuuree.} 
\LX{dabodabo} \HM{} \PH{dábʋ́dábʋ́} \PS{n.~} \GE{duck} \OP{dabodabore} \TP{dabodaborɛɛ.} 
\LX{dabɔge} \HM{} \PH{dàbɔ̀gɪ́} \PS{n.~} \GE{a type of large bird (carnivorous)} \OP{dabɔgere} \TP{dabɔgerɛɛ} 
\LX{dabogi} \HM{} \PH{dàbògí} \PS{n.~} \GE{hole in a tree} \OP{dabogri} \TP{dabogree.} 
\LX{dabɔgɔ} \HM{} \PH{dàbɔ́gɔ̀} \PS{n.~} \GE{fenced enclosure; garden} \OP{dabɔgre} \TP{dabɔgrɛɛ} \VA{dabɔga/dabɔge}
\LX{dabɔle} \HM{} \PH{dàbɔ́lɪ́} \PS{n.~} \GE{round smooth stick, club} \OP{dabɔlɔ} \TP{dabɔlɛɛ} 
\LX{dabɔlmakasogi} \HM{} \PH{dábɔ́lmákásògí} \PS{n.~} \GE{a type of tall straw hat} \OP{dabɔlmakasogiri} \TP{dabɔlmakasogiree} \VA{dabɔlkasogi/dabɔlkosogi}
\LX{dabolore} \HM{} \PH{dàbʋ́lʋ̀rɪ̀} \PS{n.~} \GE{a type of tree with medicinal properties and edible fruits} \TP{dabolorɛɛ} 
\LX{daboorabɔge} \HM{} \PH{dàbʊ̀ʊ̀rábɔ́gɪ́} \PS{n.~} \GE{praying mantis} \OP{daboorabɔgere} \TP{daboorabɔgerɛɛ} 
\LX{daboze᷈e} \HM{} \PH{dàbòzɪ́᷉ɪ́᷉̂} \PS{n.~} \GE{a type of small brown bird} \OP{dobazeene} \TP{dobazeenɛɛ} 
\LX{Dabuo} \HM{} \PH{dàbúó} \PS{n.~} \GE{name traditionally given to one born after the death of the father} \OP{Dabumine} \TP{Dabuuree} 
\LX{dabuo} \HM{} \PH{dàbúó} \PS{n.~} \GE{abandoned house} \OP{dabori} \TP{daboree} 
\LX{̀dabuori} \HM{} \PH{dàbúóri} \PS{interrogative pronoun.~} \GE{when?} \XV{Dabuori ka fo tu fo sori na?} \XE{When will you make your trip?.} 
\LX{dadagere} \HM{} \PH{dàdàgɪ́rɪ́} \PS{n.~} \GE{a kind of fish} \OP{dadaga} \TP{dadagerɛɛ} 
\LX{daduŋi} \HM{} \PH{dàdúŋí} \PS{n.~} \GE{wood weevil} \OP{dadunni} \TP{dadunnee} 
\LX{daga} \HM{} \PH{dágà} \PS{n.~} \GE{box} \OP{dagere} \TP{dagerɛɛ} 
\LX{Dagaare} \HM{} \PH{dàgàárɪ̀} \PS{n.~} \GE{Dagaare; a language of the Moore-Gurma family} \TP{dagaarɛɛ} \VA{Dagara}
\LX{dagaareŋ} \HM{} \PH{dàgààrɪ̀ŋ} \PS{adj.~} \GE{done according to the practices of the Dagaaba} \XV{Ziyuo vɛŋɛɛ la ka Dagaareŋ deɛ bɔrɔ.} \XE{Civilisation is killing most of the traditional practices of the Dagaaba.} 
\LX{dagalaa} \HM{} \PH{dàgàláà} \PS{n.~} \GE{a large, long plank used to support weak structures} \OP{dagalle} \TP{dagallɛɛ} 
\LX{dagaŋgaanaa} \HM{} \PH{dàgàŋgàànáà} \PS{n.~} \GE{large vine of a creeping or climbing plant} \OP{dagaŋgaane} \TP{dagaŋgaanɛɛ} \VA{gaŋgaanaa}
\LX{dagaŋgaŋ} \HM{} \PH{dàgàŋgáŋ} \PS{n.~} \GE{wit} \XV{A bie taa la dagaŋgaŋ, fo na yeli ka so᷉ɔ.} \XE{The child is so full of wit that it might be attributed to witchraft.} \OP{dagaŋganne} \TP{dagaŋgannɛɛ} 
\LX{Dagao} \HM{} \PH{dàgáʋ̀} \PS{n.~} \SN \GE{native speaker of Dagaare} \SN \GE{the homeland of the speakers of Dagaare} \OP{Dagaaba} \TP{Dagaabɛɛ} \VA{Dagara}
\LX{Dagara (ND)} \HM{} \PH{dàgàrà} \PS{n.~} \GE{the Dagaare language} \VA{Dagao}
\LX{dagaraa} \HM{} \PH{dàgàráà} \PS{n.~} \SN \GE{cross; symbol of Christianity} \SN \GE{crossbeam} \OP{dagarre} \TP{dagarrɛɛ} 
\LX{Dagbɛmmo} \HM{} \PH{dàgbɛ́mmʊ́} \PS{n.~} \GE{a native speaker of the Dagbaane language} \OP{Dagbɛnne} \TP{Dagbɛnnɛɛ} 
\LX{Dagbɛnne} \HM{} \PH{Dàgbɛ́nnɪ́} \PS{n.~} \GE{Dagbaane language} \TP{dagbɛnnɛɛ} 
\LX{Dagbɛnnoo} \HM{} \PH{dàgbɛ́nnʋ́ʋ́} \PS{n.~} \GE{a native speaker of the Dagbaane language} \OP{Dagbɛnne} \TP{Daggbɛnnɛɛ.} \VA{Dagbɛmmo}
\LX{dagboli} \HM{} \PH{dàgbólí} \PS{n.~} \GE{large piece of wood (firewood)} \OP{dagbolo} \TP{dagbolee} 
\LX{dagbollo} \HM{} \PH{dàgbólló} \PS{n.~} \GE{the fifth day from now} \OP{dagbolli} \TP{dagbollee} 
\LX{dagbuli} \HM{} \PH{dàgbùlí} \PS{n.~} \GE{the act of buying as a whole} \OP{dagbulo} \TP{dagbullee} 
\LX{dagbulo} \HM{} \PH{dàgbùló} \PS{n.~} \GE{appellation of a kinship group} 
\LX{dageliŋe} \HM{} \PH{dágelíŋé} \PS{n.~} \GE{lid of a box} \OP{dagelinni} \TP{daglinnee.} 
\LX{dagoli} \HM{} \PH{dagoli} \PS{n.~} \GE{cudgel; a throwing stick (usually hook-shaped at one end)} \OP{dagolo} \TP{dagolee/dagollee} 
\LX{daguo} \HM{} \PH{dàgùó} \PS{n.~} \GE{club with a knobbed end; a short, large-headed piece of wood used for breaking mud-built walls and lumps of earth} \OP{dagori} \TP{dagoree/dagorree.} \VA{guo}
\LX{dagyu} \HM{} \PH{dàdzú} \PS{n.~} \GE{a type of large, hairy spider; believed to be poisonous} \OP{dagyuri} \TP{dagyuree.} 
\LX{daha} \HM{} \PH{dáhá} \PS{n.~} \GE{a type of game similar to chess} \OP{dahare} \VA{dara}
\LX{dakala} \HM{} \PH{dàkálá} \PS{n.~} \GE{dried sperm} \TP{dakallɛɛ} 
\LX{dakogi} \HM{} \PH{dàkógí} \PS{n.~} \GE{chair, stool} \OP{dakogiri} \TP{dakogiree} 
\LX{dakogidɛllaa} \HM{} \PH{dàkògìdɛ́lláá} \PS{n.~} \GE{chair with a back} \OP{dakogidɛlle} 
\LX{dakogilee} \HM{} \PH{dàkògìléé} \PS{n.~} \GE{small stool} \OP{dakogibilii} \TP{dakogibilee} 
\LX{dakoɔre} \HM{} \PH{dàkʋ̀ɔ̀rɪ́} \PS{n.~} \GE{bachelor} \OP{dakɔbɔ} \TP{dakoɔrɛɛ.} 
\LX{dakɔɔre} \HM{} \PH{dàkɔ́ɔ́rɪ́} \PS{n.~} \GE{armlet made from enamel} \OP{dakɔbɔ} \TP{dakɔɔrɛɛ} 
\LX{dakɔɔre} \HM{} \PH{dàkɔ́ɔ́rɪ́} \PS{n.~} \GE{cob (maize)} \OP{dakɔbɔ} \TP{dakɔɔrɛɛ} 
\LX{dakoɔroŋ} \HM{} \PH{dàkʊ́ɔ́rʊ́ŋ} \PS{n.~} \GE{celibacy} 
\LX{Dakoraa} \HM{} \PH{dàkʋ́ráá} \PS{n.~} \GE{name traditionally given to a reincarnated male child} \OP{dakormine} \TP{dakorɛɛ/dakorrɛɛ.} 
\LX{dakoraa} \HM{} \PH{dàkòràá} \PS{n.~} \GE{walking stick with a curved top} \OP{dakorri} \TP{dakoree/dakorree} 
\LX{dakoroŋ} \HM{} \PH{dàkʊ̀rʊ̀ŋ} \PS{n.~} \GE{former times; olden days} \VA{dasaŋa}
\LX{Dakpaa} \HM{} \PH{dàkpàá} \PS{n.~} \SN \GE{name of a village} \SN \GE{a kinship group} 
\LX{dakpaa} \HM{} \PH{dàkpáà} \PS{n.~} \GE{a type of wasp} \OP{dakpaare} \TP{dakpaarɛɛ} 
\LX{Dakpaaloo} \HM{} \PH{dàkpáálʊ́ʊ́} \PS{n.~} \GE{member of the Dakpaa kinship group} \OP{Dakpaalɛɛ} 
\LX{dakpalori} \HM{1} \PH{dàkpàlórí} \PS{n.~} \GE{a type of small bug that sucks blood (usually found in crevices of walls)} \SG{dakpaloruu} \OP{dakpaloree} \TP{dakpalorree} 
\LX{dakpele} \HM{} \PH{dàkpɪ́lɪ́} \PS{n.~} \GE{hard wood} \OP{dakpelɛ} \TP{dakpelɛɛ} 
\LX{dakpiri} \HM{} \PH{dàkpíri} \PS{n.~} \GE{stump} \SG{dakpirruu} \OP{dakpirri} \TP{dakpiree/dakpirree} \VA{dakpaa}\VA{daŋmaparekpiri}
\LX{Dakpɔge} \HM{} \PH{dàkpɔ́gɪ́} \PS{n.~} \GE{a kinship group} \OP{dakpɔgɔ} 
\LX{dakuoŋaa} \HM{} \PH{dàkúóŋáá} \PS{n.~} \GE{dry or dead wood} \OP{dakuoni/dakuomo} \TP{dakuonee} 
\LX{dakyage} \HM{} \PH{dàʧàgɪ́} \PS{n.~} \GE{lazy chair (for men), deckchair or sun lounger} \OP{dakyagre} \TP{dakyagrɛɛ} 
\LX{dakyara} \HM{} \PH{dàʧárà} \PS{n.~} \GE{forked piece of wood} \OP{dakyɛrre} \TP{dakyɛrrɛɛ} 
\LX{dakyɛ} \HM{} \PH{dàʧɛ́} \PS{n.~} \GE{carelessness} \XV{N nɛɛse dakyɛ la so ka a bie kpi.} \XE{The child died because of the nurse's carelessness.} 
\LX{dakyeɛ} \HM{} \PH{dàʧɪ́ɛ} \PS{n.~} \GE{brother- or sister-in-law} \OP{dakyeere/dakyemine} \TP{dakyeerɛɛ} 
\LX{Dakyeeloo} \HM{} \PH{dàʧɪ́ɪ́lʋ́ʋ́} \PS{n.~} \GE{member of some type of kinship (Dakyeɛ) group} \OP{dakyeelɛɛ} 
\LX{dakyeɛraa} \HM{} \PH{dàʧɪ̀ɛ̀ráà} \PS{n.~} \GE{tree meant to be felled} \OP{dakyeɛre} 
\LX{dakyeɛraa} \HM{} \PH{dàʧɪ́ɛ́ráá} \PS{n.~} \GE{something used for cutting wood} \OP{dakyeɛre} 
\LX{dakyeɛrɛko} \HM{} \PH{dàʧɪ́ɛ́rɛ́kò} \PS{n.~} \GE{woodpecker} \OP{dakyeɛrɛkori} \TP{dakyeɛrɛkoree} \VA{tekoliŋkuoraa}
\LX{dakyeerekpaglaga} \HM{} \PH{dàʧɪ̀ɪ̀rɪ̀kpàglágá} \PS{n.~} \GE{a type of tree} \OP{dakyeerekpaglagre} \TP{dakyeerekpaglagrɛɛ} 
\LX{dakye-nɔguu} \HM{} \PH{dàʧɪ̀-nɔ́gùu]} \PS{n.~} \GE{parrot} \OP{dakye-nɔguuri} \TP{dakye-nɔguuree} 
\LX{dakyeraa} \HM{} \PH{dàʧɪ̀ráa} \PS{n.~} \SN \GE{canary} \SN \GE{parrot} \SN \GE{weaver (bird) or weaverbird} \OP{dakyerre} \TP{dakyerrɛɛ} \VA{dakyeɛ, dakyere, dakyerɛɛ}
\LX{dala} \HM{} \PH{dàlà} \PS{n.~} \GE{conceited talk; arrogance, self-importance} \XV{A pɔgelee zagere la a toma kyɛ yele o dala.} \XE{The girl refused to do the work and put on airs.} 
\LX{dalaa} \HM{} \PH{dàlàá} \PS{n.~} \GE{small drum, made with hide and the rim of a pot} \OP{dalle} \TP{dallɛɛ} \VA{dalaraa}
\LX{dalaraa (ND)} \HM{} \PH{dàláràà} \PS{n.~} \GE{} \OP{dalarre} \TP{dalarɛɛ} \VA{dalaa}
\LX{dalee} \HM{} \PH{dàléé} \PS{n.~} \GE{ dry twig} \OP{dabili} \TP{dabilee} \VA{delee}
\LX{daleerɛ} \HM{} \PH{dàlɪ́ɪ́rɛ́} \PS{n.~} \GE{total knowledge of} \XV{A nannyige daleerɛ zaa ka a polisri baŋ.} \XE{The police have gotten to know all the thief's tricks.} 
\LX{dalege} \HM{} \PH{dàlɪ́gɪ́} \PS{n.~} \GE{stampede} \OP{dalegre} \TP{dalegrɛɛ} 
\LX{dalempoɔ} \HM{} \PH{dàlɪ́mpʋ́ɔ́} \PS{n.~} \GE{an area without vegetation cover} \OP{dalempoore} \TP{dalempoorɛɛ} \VA{dampoɔ}
\LX{dalendeɛloo} \HM{} \PH{dàlɪ́ndɪ́ɛ́lʊ̀ʊ́} \PS{n.~} \GE{leech; a blood-sucking aquatic worm} \OP{dalendeɛlɛ} \TP{dalendeɛlɛɛ} 
\LX{daliiri} \HM{} \PH{dàlíírì} \PS{n.~} \GE{substance} \XV{O yɛlɛ na o naŋ yeli ba taa daliiri.} \XE{There is no substance in his/her speech.(is useless).} \TP{daliiree} 
\LX{dalogeraa} \HM{} \PH{dàlʋ́gráá} \PS{n.~} \GE{wood bug} \OP{dalogere} \TP{dalogerɛɛ} 
\LX{dalogi} \HM{} \PH{dàlógì} \PS{n.~} \GE{an area where woman pile up firewood for use in the kitchen} \OP{dalogri} \TP{dalogree} 
\LX{daloraa} \HM{} \PH{dàlòràá} \PS{n.~} \GE{neat stack of firewood} \OP{dalorri} \TP{daloree} 
\LX{dalugiri} \HM{} \PH{dàlúgrí} \PS{n.~} \GE{Y-shaped plank that supports roof beams} \OP{dalugo} \TP{dalugiree} 
\LX{daluri} \HM{} \PH{dàlúri} \PS{n.~} \GE{a type of calabash} \OP{dalurri} \TP{daluree} 
\LX{damadama} \HM{} \PH{dàmàdàmà} \PS{adj.~} \GE{big, large, extensive} 
\LX{damagara} \HM{} \PH{dàmàgàrà} \PS{n.~} \GE{act of showing wit; adeptness} \OP{damagarre} \TP{damagarɛɛ.} 
\LX{damannaa} \HM{} \PH{dàmánnáá} \PS{n.~} \SN \GE{measuring stick} \SN \GE{example, comparison} \SN \GE{simile (lit.)} \OP{damanne} \TP{damaannɛɛ} 
\LX{dambo} \HM{} \PH{dàmbó} \PS{n.~} \GE{(zɛ---) of soup not spiced} \OP{dambori} \TP{ damboree} 
\LX{damboli} \HM{} \PH{dàmbólì} \PS{n.~} \GE{stupid or foolish person} \OP{dambolo} \TP{dambolee} 
\LX{damɔlɔ} \HM{} \PH{dàmɔ́l} \PS{n.~} \GE{cheap price (be inexpensive)} 
\LX{dampoɔ} \HM{} \PH{dàmpʊ́ɔ́} \PS{n.~} \GE{short form of  dalempoɔ} \OP{dampoore} \TP{dampoorɛɛ} 
\LX{dampuori} \HM{} \PH{dàmpúórí} \PS{n.~} \GE{the back of a house} \OP{dampoe} \TP{damporee} 
\LX{damui} \HM{} \PH{dàmùí} \PS{n.~} \GE{a type of tree with medicinal properties} 
\LX{daŋ} \HM{} \PH{dàŋ} \PS{adj.~} \GE{before} \XV{O  daŋ kpɛ la aloopelee.} \XE{S/he has travelled by plane before.} 
\LX{daŋaa} \HM{} \PH{dàŋàá} \PS{n.~} \GE{hearth; fireplace for cooking} \OP{danne} \TP{dannɛɛ} 
\LX{daŋaa} \HM{} \PH{dàŋàà} \PS{adj.~} \GE{big and heavy} 
\LX{dandamiine} \HM{} \PH{dàndàmííné} \PS{n.~} \GE{a type of red ant (found in trees)} \OP{dandamiime} \TP{dandamiimee/dandamiinee} \VA{gyangyamiime}
\LX{dandayieni} \HM{1} \PH{dándáyíéní} \PS{n.~} \GE{a type of small toad} \OP{dandayieme} \TP{dandayiemee/dandayienee} \VA{dayieni}
\LX{dandɔɔ} \HM{} \PH{dàndɔ́ɔ́} \PS{n.~} \GE{owner (male)} \OP{dandɔbɔ} \TP{dandɔbɛɛ} 
\LX{danduli} \HM{} \PH{dàndúlí} \PS{n.~} \GE{caterpillar, worm} \OP{dandulo} \TP{dandulee} \VA{dunduli}
\LX{daŋe} \HM{} \PH{dáŋ} \PS{v.~} \GE{to be ahead of; to beat, to be the first to do something} \XV{N na daŋ fo la ta.} \XE{I will get there before you.} \PD{daŋɛɛ, danna, danna, danema, dannaa, danema } 
\LX{daŋe} \HM{} \PH{dàŋ} \PS{v.~} \SN \GE{be early} \XV{Zenɛ, o daŋe la toma waabo.} \XE{Today s/he came to work early.} \SN \GE{come earlier} \XV{Zenɛ, o daŋe la toma waabo gaŋe bebiri zaa.} \XE{Today s/he came to work earlier than all other days.} \PD{daŋɛɛ, danna, danna, danema, dannaa, danema } 
\LX{daneenɛ} \HM{} \PH{dànɪ́ɪ́nɛ́} \PS{n.~} \GE{the fifth day from today} \OP{daneene} \TP{daneenɛɛ} \VA{daneŋenɛ}
\LX{danegere} \HM{} \PH{dànɪ́grɪ́} \PS{n.~} \GE{a type of honeycomb where bees store their food} \OP{danegɛ} \TP{danegerɛɛ} \VA{sedanegre}
\LX{daŋena} \HM{} \PH{dáŋɪ́ná} \PS{n.~} \GE{stack of pots on top of one another} \OP{daŋene} \TP{daŋenɛɛ} 
\LX{daneŋenɛ} \HM{} \PH{dànɪ́ŋɪ́nɛ́} \PS{n.~} \GE{the sixth day from today} \OP{daneŋene} \TP{daneŋenɛɛ} 
\LX{danfaraŋ} \HM{} \PH{dànfàràŋ} \PS{adj.~} \GE{numb} 
\LX{daŋgbala} \HM{} \PH{dáŋgbála} \PS{n.~} \GE{bare and open place} \OP{daŋgballe} \TP{daŋgbalɛɛ} 
\LX{daŋgbale} \HM{} \PH{dàŋgbàlɪ́} \PS{n.~} \GE{walking stick} \OP{daŋgbala} \TP{daŋgbalɛɛ/daŋgballɛɛ} 
\LX{daŋgoɔne} \HM{} \PH{dàŋgʊ́ɔ́nɪ́} \PS{n.~} \GE{cliff} \OP{daŋgoɔmɔ} \TP{daŋgoɔnɛɛ} 
\LX{daŋkasaŋ} \HM{} \PH{dáŋkásáŋ} \PS{adj.~} \GE{bloated-looking} 
\LX{daŋkuuri} \HM{} \PH{dàŋkúúrí} \PS{n.~} \GE{stone used to support a cooking pot in the hearth} \OP{daŋkubo/daŋkurri} \TP{daŋkuree/daŋkuuree} 
\LX{dankyimmiri} \HM{} \PH{dànʧǐmmírì} \PS{n.~} \GE{layer of a wall; horizontal layer of plaster} \OP{dankyimmie} \TP{dankyimmiree} 
\LX{dankyini} \HM{} \PH{dànʧíní} \PS{n.~} \GE{wall} \OP{dankyime} \TP{dankyimee/dankyinnee.} \VA{daŋkyini}
\LX{dankyiŋkogilaa} \HM{} \PH{dànʧìnkógláá} \PS{n.~} \GE{inner corner of a wall e.g. a room} \OP{dankyiŋkogli/dankyiŋkoglo} \TP{dankyiŋkoglee} 
\LX{dankyin-ŋmaa} \HM{} \PH{dǎnʧǐn-ŋmáá} \PS{n.~} \GE{broken wall (reduced in height)} \OP{dankyin-ŋmaara} \TP{dankyin-ŋmaarɛɛ} 
\LX{dankyinsɛgeraa} \HM{} \PH{dǎnʧǐnsɛ́gráá} \PS{n.~} \GE{chalkboard (writing wall)} \OP{dankyinsɛgre} \TP{dankyinsɛgrɛɛ} 
\LX{dankyinsɔglaa} \HM{} \PH{dǎnʧǐnsɔ́gláá} \PS{n.~} \GE{blackboard (black wall)} \OP{dankyinsɔgle} 
\LX{dankyinsɔɔraa} \HM{} \PH{dǎnʧǐnsɔ́ɔ́ráá} \PS{n.~} \GE{blackboard (blackened wall)} \OP{dankyinsɔɔre} \TP{dankyinsɔɔrɛɛ} 
\LX{dankyo} \HM{} \PH{dánʧó} \PS{n.~} \GE{single-barreled gun} \OP{dankyori} \TP{dankyoree} 
\LX{daŋmaa} \HM{} \PH{dàŋmáá} \PS{n.~} \GE{log} \OP{daŋmaara} \TP{daŋmaarɛɛ} 
\LX{daŋmeɛlɛ} \HM{} \PH{dàŋmɪ́ɛ́lɛ́} \PS{n.~} \GE{the sixth day from today} \OP{daŋmeɛlɛ} \TP{ daŋmeɛlɛɛ} 
\LX{daŋmeɛraa} \HM{} \PH{dàŋmɪ́ɛ́ráà} \PS{n.~} \GE{stick used to beat someone} \OP{daŋmeɛre} 
\LX{daŋmeɛraa} \HM{} \PH{dàŋmɪ́ɛ́ráà} \PS{n.~} \GE{solution, answer to a problem} \OP{daŋmeɛre} 
\LX{daŋŋa} \HM{} \PH{dáŋŋá} \PS{n.~} \GE{mischief} 
\LX{danne} \HM{} \PH{dánnɪ̀} \PS{v.~} \GE{to praise} \PD{dannɛɛ, danna, danna, danema, dannaa, danema } \VA{dɛnne}
\LX{dannoŋ} \HM{} \PH{dànnʊ́ŋ} \PS{n.~} \GE{praise} 
\LX{danooraa} \HM{} \PH{dànʊ́ʊ́ráá} \PS{n.~} \GE{a type of wood bug} \OP{danoore} \TP{danoorɛɛ} 
\LX{danse} \HM{} \PH{dánsɪ̀} \PS{n.~} \GE{a paid dance session} \OP{dansere} \TP{danserɛɛ} 
\LX{danseɛ} \HM{} \PH{dánsɪ́ɛ́} \PS{n.~} \GE{witness} \OP{danseere} \TP{danseerɛɛ} 
\LX{danseɛsoba} \HM{} \PH{dánsɪ́ɛ́sʊ́bá} \PS{n.~} \GE{witness} \OP{danseɛdeme} 
\LX{Danteɛ} \HM{} \PH{dàntɪ́ɛ́} \PS{n.~} \GE{a town in the Upper West Region} 
\LX{Danteɛloo} \HM{} \PH{dàntɪ́ɛ́lʋ́ʋ́} \PS{n.~} \GE{a member of the Danteɛ clan} \OP{Danteɛlɛɛ} 
\LX{dantiŋ} \HM{} \PH{dàtíŋ} \PS{n.~} \GE{a type of tree} \OP{dantinni} \TP{dantinnee} 
\LX{danwe᷉ɛ} \HM{} \PH{dànwé᷉ɛ̀᷉} \PS{adj.~} \GE{first, ahead} \XV{Bayuo la a zoobu danwe᷉ɛ soba.} \XE{Bayuo is the first in the race.} 
\LX{danyagela} \HM{} \PH{dàɲáglá} \PS{n.~} \GE{the sixth day coming} \OP{danyaegle} \TP{danyagelɛɛ} 
\LX{danyagere} \HM{} \PH{dàɲágrɪ} \PS{n.~} \GE{root} \OP{danyaga} \TP{danyagerɛɛ} 
\LX{danyaŋe} \HM{} \PH{dàɲáŋ} \PS{adj.~} \GE{quickly, fast} 
\LX{Daŋyi} \HM{} \PH{dàŋyí} \PS{n.~} \GE{name traditionally given to the firstborn of a set of twins} \VA{Zeɛmɛ}
\LX{daoo} \HM{} \PH{dàʊ̀ʊ̀} \PS{adv.~} \GE{large and firm-looking} 
\LX{da-ɔɔraa} \HM{} \PH{dà-ɔ́ɔ́ráá} \PS{n.~} \GE{chewing stick (tooth cleaner)} \OP{da-ɔɔre} 
\LX{dapaaloŋ} \HM{} \PH{dàpáálʊ́ŋ} \PS{n.~} \SN \GE{vanity} \SN \GE{splendour, glory} 
\LX{dapare} \HM{} \PH{dàpàrɪ́} \PS{n.~} \GE{dwelling place of the dead; heaven, eternity} 
\LX{daparedaa} \HM{} \PH{dàpárɪ́dàá} \PS{n.~} \GE{death, next world} 
\LX{daparekpeɛŋaa} \HM{} \PH{dàpàrɪ̀kpɪ́ɛ́ŋáá} \PS{n.~} \GE{strong spirit} \OP{daparekpeɛne} \TP{daparekpeɛnɛɛ} 
\LX{daparenuulee} \HM{} \PH{dàpàrɪ̀núúléé} \PS{n.~} \GE{cockalo} \OP{daparenuuli} 
\LX{daparoo} \HM{} \PH{dàpárʊ́ʊ́} \PS{n.~} \GE{soul or spirit of the dead} \SG{daparrɛɛ} \TP{daparɛɛ} 
\LX{dapeɛlkammo} \HM{} \PH{dàpɪ́ɛ́lkámmʊ́} \PS{n.~} \GE{a type of bean} \TP{dapeɛlkamɛɛ} 
\LX{dapɛɛre} \HM{} \PH{dàpɛ̀ɛ̀rɪ́} \PS{n.~} \GE{immature tree or wood} \OP{dapɛbɛ} \TP{dapɛbɛɛ} 
\LX{dapɛge} \HM{} \PH{dàpɛ́gɪ́} \PS{n.~} \GE{bark (of tree)} \OP{dapɛgre/dapɛgɛ} \TP{dapɛgrɛɛ} 
\LX{dapɛle} \HM{} \PH{dàpɛ́lɪ́} \PS{n.~} \GE{splinter of wood} \OP{dapɛlɛ} \TP{dapɛlɛɛ/dapɛllɛɛ} 
\LX{dapɛlemɛ} \HM{} \PH{dàpɛ́lmɛ́} \PS{n.~} \GE{splinters of wood} \TP{dapɛlmɛɛ} 
\LX{dapɛlɛŋe} \HM{} \PH{dàpɛ̀lɛ̀ŋ} \PS{n.~} \GE{a type of tiny wild bat} \OP{dapɛlɛnne} \TP{dapɛlɛnnɛɛ} 
\LX{dapɛloo} \HM{} \PH{dàpɛ́lʋ́ʋ́} \PS{n.~} \GE{splinter of wood} 
\LX{daperaa} \HM{} \PH{dàpɪ́ràa} \PS{n.~} \GE{small dove or pigeon (grey and red)} \OP{daperre} \TP{daperɛɛ/daperrɛɛ} 
\LX{daperekalonni} \HM{} \PH{dàpɪ́rɪ́kálónnì} \PS{n.~} \GE{a type of trap for birds (usually made of bamboo and twine)} \TP{daperekalonnee} 
\LX{daperenyaa} \HM{} \PH{dàpɪ̀rɪ̀ɲáà} \PS{n.~} \GE{the breast of a type of dove} 
\LX{dara} \HM{} \PH{dárá} \PS{n.~} \GE{a game played by moving small pins into marked-out holes} \OP{darre} \TP{ darɛɛ/darrɛɛ} 
\LX{daransaŋ} \HM{} \PH{dàránsáŋ} \PS{n.~} \GE{heavy, cast-aluminium cooking pot} \OP{daransanne} \TP{daransannɛɛ} 
\LX{dare} \HM{} \PH{dárɪ́} \PS{v.~} \GE{to be hysterical} \PD{darɛɛ, dara, dara, dareba, daraa, dareba } 
\LX{dare} \HM{} \PH{dàrɪ̀} \PS{v.~} \SN \GE{to harden} \SN \GE{to get acustomed to; to be used to something unpleasant} \PD{darɛɛ, dara, dara, dareba, dara, dareba } 
\LX{dasaŋa} \HM{} \PH{dàsáŋà} \PS{n.~} \GE{olden times} \TP{dasaŋɛɛ} \VA{dakoroŋ}
\LX{dasaraa} \HM{} \PH{dàsáràà} \PS{n.~} \GE{plank} \OP{dasarre} \TP{dasarrɛɛ} 
\LX{Dasa᷉a᷉} \HM{} \PH{dàsá᷉á᷉} \PS{n.~} \GE{name traditionally given to a child born on a rainy day} \OP{dasaamine} \TP{dasaanɛɛ} 
\LX{daseelɛ} \HM{} \PH{dàsɪ̀ɪ̀lɛ̀} \PS{n.~} \GE{suicide} \TP{daseelɛɛ} 
\LX{dasɛgeraa} \HM{} \PH{dàsɛ́gráà} \PS{n.~} \GE{wood for filling in gaps (especially in mud roofing)} \OP{dasɛgere} \TP{daseɛgerɛɛ} 
\LX{dasɛrɛɛ} \HM{} \PH{dàsɛ́rɛ́ɛ́} \PS{n.~} \GE{small sticks that are used to close a gap in a thatch earth-made roof of a building} \SG{dasɛroo} \TP{dasɛrrɛɛ} \VA{dasɛgeraa}
\LX{dasɛroo} \HM{} \PH{dàsɛ́rʋ́ʋ́} \PS{n.~} \GE{small stick that is used to close a gap in a thatch earth-made roof of a building} \OP{dasɛrɛɛ} \TP{dasɛrrɛɛ} 
\LX{dasugo} \HM{} \PH{dàsúgò} \PS{n.~} \GE{many kinds of trees} \OP{dasugiri} \TP{dasugree} 
\LX{dasuoluŋ} \HM{} \PH{dàsùólúŋ} \PS{n.~} \GE{shadow} \OP{dasuolo} \TP{dasuolee} \VA{salinsuuri}
\LX{dasɛgeraa} \HM{} \PH{dàsɛ́gɪ́ráá} \PS{n.~} \GE{piece of wood used to cover gaps between bigger plunks of wood in an earth-constructed roof} \TP{dasɛgere, dasɛgerɛɛ} 
\LX{dataa} \HM{} \PH{dàtáá} \PS{n.~} \SN \GE{rival (between males)} \SN \GE{antagonist (lit.)} \OP{dataare} \TP{ dataarɛɛ} 
\LX{datara} \HM{} \PH{dàtárá} \PS{n.~} \GE{third day from now} \OP{datarre} \TP{datarrɛɛ} 
\LX{datuo} \HM{1} \PH{dàtúó} \PS{n.~} \GE{a type of tree with medicinal properties and edible fruits} \OP{datuuri} \TP{datuuree} 
\LX{datuo} \HM{2} \PH{dàtúó} \PS{n.~} \GE{being expensive} \OP{datuuri} \TP{datuuree} 
\LX{dave} \HM{} \PH{dàvɪ́} \PS{n.~} \GE{spoke} \OP{davere} \TP{daverɛɛ} 
\LX{daveene} \HM{} \PH{dàvɪ̀ɪ̀nɪ̀} \PS{n.~} \GE{a type of small bird with very thin legs} \OP{daveemɛ} \TP{daveenɛɛ} \VA{veene}
\LX{daveraŋkantoll} \HM{} \PH{dàvɪ̀ràŋkántóll} \PS{n.~} \GE{last warning, last chance} 
\LX{davoro} \HM{} \PH{dàvòró} \PS{n.~} \GE{courtyard} \OP{davorri} \TP{davorree} 
\LX{davuo} \HM{} \PH{dàvùó} \PS{n.~} \GE{a hollowed out log} \OP{davori} \TP{davoree} 
\LX{dawɛge} \HM{} \PH{dàwɛ́gɪ́} \PS{n.~} \GE{a big plunk of wood used in the construction of earth-made roof} \OP{dawɛgre} \TP{dawɛgrɛɛ} \VA{wɛge}
\LX{dawɛrɛ} \HM{} \PH{dàwɛ̀rɪ̀} \PS{n.~} \GE{person who saws or splits wood} \OP{dawɛrebɛ} \TP{dawɛrebɛɛ} 
\LX{daworo} \HM{} \PH{dáwóró} \PS{n.~} \GE{a type of bell used in music and to call for attention} \OP{dawoe} \TP{ daworee.} 
\LX{dayeni} \HM{} \PH{dàyénì} \PS{adj.~} \GE{once} \OP{dayeme} \TP{dayennee} 
\LX{dayere} \HM{} \PH{dàyéré} \PS{n.~} \GE{the day after tomorrow} \TP{dayeree} 
\LX{dayieni} \HM{} \PH{dáyíéní} \PS{n.~} \GE{a type of small toad} \OP{dayieme} \TP{dayienee} \VA{dandayieni}
\LX{dayuu} \HM{} \PH{dàyúú} \PS{n.~} \GE{giant rat} \OP{dayuuri dayuuree} 
\LX{dazɛllaa} \HM{} \PH{dàzɛ́lláá} \PS{n.~} \GE{swing} \OP{dazɛlle} \TP{dazɛllɛɛ} 
\LX{dazugiri} \HM{} \PH{dàzùgrí} \PS{n.~} \GE{} \OP{dazugo} \TP{dazugiree} \VA{shortened form of kɔndazugri}
\LX{dazugivu᷉u} \HM{} \PH{dàzùgvù᷉ù} \PS{n.~} \GE{intense fire; hell; inferno} \TP{dazugvuunee} 
\LX{da᷉a} \HM{} \PH{dá᷉à᷉} \PS{n.~} \SN \GE{brewed drink, alcoholic beverage (pito)} \SN \GE{container of drink} \SG{daamo} \OP{daanɛɛ} 
\LX{dâabeene} \HM{} \PH{dá᷉á᷉bɪ́ɪ́nɪ́} \PS{n.~} \GE{fermented pito} \OP{dâabeemɛ} 
\LX{dâabeerɛ} \HM{} \PH{dá᷉á᷉bɪ̀ɪ̀rɛ́} \PS{n.~} \GE{second stage in brewing pito} \TP{dâabeerɛɛ} 
\LX{dâabele} \HM{} \PH{dá᷉ábɪ́lɪ́} \PS{n.~} \GE{pito yeast} \OP{dâabelɛ} \TP{dâabelɛɛ} 
\LX{dâabinnaa} \HM{} \PH{dá᷉á᷉bìnnáà} \PS{n.~} \GE{pito reserved for a regular customer} \OP{dâabinni} \TP{dâabinnee} 
\LX{da᷉abiri} \HM{} \PH{dá᷉á᷉bírí} \PS{n.~} \GE{residue from pito (local beer)} \TP{da᷉abiree} 
\LX{dâabiri} \HM{} \PH{dá᷉á᷉bírì} \PS{n.~} \GE{chaff of pito malt} \TP{dâabiree/dâabirree.} 
\LX{dâada} \HM{} \PH{dá᷉á᷉dà} \PS{n.~} \GE{the act of buying alcohol} 
\LX{dâadaara} \HM{} \PH{dá᷉á᷉dáárá} \PS{n.~} \GE{one who buys alcohol} \OP{dâadareba} \TP{dâadarebɛɛ} 
\LX{dâadagoli} \HM{} \PH{dá᷉á᷉dàgólí} \PS{n.~} \GE{one who loves drinking alcholic beverages} \OP{dâadagolo} \TP{dâadagolee} 
\LX{dâadie} \HM{} \PH{dá᷉á᷉dìè} \PS{n.~} \GE{pito shop} \OP{dâaderi} \TP{dâaderee} 
\LX{dâadoge} \HM{} \PH{dá᷉á᷉dʋ́gɪ́} \PS{n.~} \GE{pito pot or pot of pito} \OP{dâadoggre} \TP{dâadogrɛɛ} 
\LX{dâadogerɔ} \HM{} \PH{dá᷉á᷉dʊ́grɔ́} \PS{n.~} \GE{brewer, one who brews alcohol} \OP{dâadogerebɔ} \TP{dâadogerebɛɛ} 
\LX{da᷉afaare} \HM{} \PH{dá᷉áfáárɪ́} \PS{n.~} \GE{bad alcoholic beverage} \TP{da᷉a᷉faarɛɛ} 
\LX{da᷈afaraa} \HM{} \PH{dá᷈áfáráá} \PS{n.~} \GE{well-fermented pito} \OP{dââfarre} \TP{dââfarrɛɛ} 
\LX{da᷉agyemɛ} \HM{} \PH{dá᷉ágyɪ́mɛ́} \PS{n.~} \GE{pito soda} 
\LX{dâagyemɛ} \HM{} \PH{dá᷉á᷉̂gyɪ̀mɛ̀} \PS{n.~} \GE{soda from pito brewing} \TP{dâagyemɛɛ/dâagyennɛɛ} 
\LX{dâakaraa} \HM{} \PH{dá᷉á᷈káráá} \PS{n.~} \GE{unfermented pito} \OP{dâakare} \TP{ dâakarɛɛ.} 
\LX{dâakɔgere} \HM{} \PH{dá᷉á᷉̂kɔ́grɪ́} \PS{n.~} \GE{the first stage of pito brewing} \TP{dâakɔgerɛɛ} 
\LX{dâakonni} \HM{} \PH{dá᷉á᷉̂kònnì} \PS{n.~} \GE{pito that is fermenting} 
\LX{da᷉akoɔre} \HM{} \PH{dá᷉á᷉kʊ́ɔ́rɪ́} \PS{n.~} \GE{residue of pito from the last brewing stage} \TP{da᷉akoɔrɛɛ} 
\LX{da᷉akuoli} \HM{} \PH{dá᷉á᷉kúólì} \PS{n.~} \GE{drunkenness, condition of being drunk} 
\LX{da᷈akuolo} \HM{} \PH{dá᷈á᷈kúóló} \PS{n.~} \GE{drunk} \OP{da᷈akuolibo} \TP{da᷈akuolibee} 
\LX{da᷉aŋmane} \HM{} \PH{dá᷉á᷉ŋmánɪ́} \PS{n.~} \SN \GE{calabash used for drinking pito} \SN \GE{a calabash of pito} \OP{da᷉aŋmama} \TP{da᷉aŋmamɛɛ} 
\LX{da᷉apare} \HM{} \PH{dá᷉á᷉párɪ́} \PS{n.~} \GE{residue of pito from the pot} \TP{da᷉aparɛɛ} 
\LX{da᷈apelaa} \HM{} \PH{dá᷈á᷈pɪ̀làá} \PS{n.~} \GE{palm wine} \OP{da᷈apeɛle} \TP{da᷈apeɛlɛɛ} 
\LX{dâapɔge} \HM{} \PH{dá᷉á᷉̂́pɔ́gɪ́} \PS{n.~} \GE{ pito brewer or seller} \OP{dâapɔgeba} \TP{dâapɔgebɛɛ/dâapɔgrɛɛ} 
\LX{dâasenyoɔle} \HM{} \PH{dá᷉á᷉sɪ́ɲʋ́ɔ́lɪ́} \PS{n.~} \SN \GE{ sugar ant} \SN \GE{one who likes drinking alcohol} \OP{dâasenyoɔlɔ/da᷉asenyoɔla} \TP{dâasenyoɔlɛɛ} 
\LX{dâa-so} \HM{} \PH{dá᷉á᷉-sò} \PS{n.~} \GE{ a ritual that involves bathing a corpse with pito before burial} 
\LX{da᷉asunni} \HM{} \PH{dá᷉á᷉súnní} \PS{n.~} \GE{sticky substance from pito} \SG{da᷉asunnuu} \TP{da᷉asunnee} 
\LX{da᷉asuoro} \HM{} \PH{dá᷉á᷉súóró} \PS{n.~} \GE{a corpse that is to be bathed with pito before burial} \OP{da᷉asuoribo} \TP{da᷉asuoribee} 
\LX{da᷉atambalaaraa} \HM{} \PH{dá᷉á᷉támbálááráá} \PS{n.~} \GE{sweet pito} \TP{da᷉atambalaarɛɛ} 
\LX{da᷈ateɛ} \HM{} \PH{dá᷈á᷈tɪ́ɛ́} \PS{n.~} \GE{grapevine} \OP{da᷈ateere} \TP{dâateerɛɛ} 
\LX{dâatuo} \HM{} \PH{dá᷉á᷉túó} \PS{n.~} \GE{pito meant for a ritual} \OP{dâatuuri} \TP{dâatuuree} 
\LX{da᷉azeɛ} \HM{} \PH{dá᷉á᷉zìɛ́} \PS{n.~} \GE{locally brewed beer (pito)} \TP{da᷉azerɛɛ/da᷉azeerɛɛ} 
\LX{dâazu} \HM{} \PH{dá᷉á᷉zú} \PS{n.~} \GE{the first pito to be poured from the pot} \OP{dâazuri} \TP{dâazuree} 
\LX{da᷉azukpala} \HM{} \PH{dá᷉á᷉zúkpálá} \PS{n.~} \GE{pito without yeast froth} \TP{da᷉azukpalɛɛ} \VA{da᷉a᷉zukpoli}
\LX{de} \HM{} \PH{dɪ̌} \PS{v.~} \SN \GE{to take} \SN \GE{to accept} \SN \GE{to believe} \PD{deɛɛ, derɛ, derɛ, derebɛ, deraa, derebɛ } 
\LX{dɛ} \HM{} \PH{dɛ̀} \PS{v.~} \GE{pl. to cut to make even (fingernails)} \PD{dɛɛ, deɛrɛ, deɛrɛ, deɛrɛ, deɛraa, deɛrebɛ } 
\LX{de pɔge} \HM{} \PH{dɪ́ pɔ́gɪ́} \PS{v.~} \GE{to marry a woman} \PD{-, -, -, -, -, - } 
\LX{dɛbiri} \HM{} \PH{dɛ́bírì} \PS{n.~} \GE{stair} \OP{dɛbie} \TP{dɛbiiree} \VA{dabiri}
\LX{dɛbogi} \HM{} \PH{dɛ́bògí} \PS{n.~} \GE{step on a ladder; stairs} \OP{dɛbogiri} \TP{dɛbogiree} 
\LX{dedee} \HM{} \PH{dèdéé} \PS{adv.~} \GE{see dedeekɔɔ} 
\LX{dedeekɔɔ} \HM{} \PH{dèdéékɔ́ɔ́} \PS{n.~} \GE{act of standing steadily e.g. a child beginning to walk} \VA{dedeekporo, zazaakɔɔ}
\LX{dɛdɛrɛɛ} \HM{} \PH{dɛ́dɛ́rɛ́ɛ́} \PS{adv.~} \GE{shallow} 
\LX{deɛ} \HM{} \PH{dɪ́ɛ̀} \PS{adv.~} \GE{just} \XV{N deɛ ba baŋ a pare.} \XE{I just don't understand it.} 
\LX{deɛ} \HM{} \PH{dɪ́ɛ̀} \PS{v.~} \GE{to remain somewhere} \XV{O gaa la a tigiri te deɛ.} \XE{S/he went to the feast and never returned.} \PD{deɛɛ, deɛrɛ, deɛrɛ, deɛrebɛ, deɛraa, deɛrebɛ } 
\LX{deedee} \HM{} \PH{déédéé} \PS{adj.~} \GE{straight, upright} 
\LX{deɛlaa} \HM{} \PH{dɪ̀ɛ̀láá} \PS{n.~} \GE{nightjar} \OP{deɛle/deɛlɛ/deɛlɛɛ.} 
\LX{deɛle} \HM{} \PH{dɪ́ɛ́lɪ́} \PS{v.~} \SN \GE{to glide away from the target, to deviate} \SN \GE{to deflect, to dodge} \PD{deɛlɛɛ, deɛlɛ, deɛlɛ, deɛlebɛ, deɛlaa, deɛlebɛ } 
\LX{deɛle} \HM{} \PH{dɪ̀ɛ̀lɪ̀} \PS{v.~} \SN \GE{to spread out (to dry)} \SN \GE{to expose completely} \SN \GE{to float on liquid} \PD{deɛlɛɛ, deɛlɛ, deɛlɛ, deɛlebɛ, deɛlɛ, deɛlebɛ } 
\LX{deɛlɛɛpo᷈o᷈} \HM{} \PH{dɪ́ɛ́lɛ́ɛ́pʊ́᷈ʊ̀᷈} \PS{n.~} \GE{drizzle} \OP{deɛlɛɛpoone} \TP{deɛlɛɛpoonɛɛ} 
\LX{deɛlɛɛpʊ᷉ʊ} \HM{} \PH{dɪ́ɛ́lɛ́ɛ́pʊ́᷉ʊ́} \PS{n.~} \GE{drizzle} 
\LX{deɛmɛ} \HM{} \PH{dɪ̀ɛ̀mɛ́} \PS{n.~} \SN \GE{in-law (father or mother)} \SN \GE{son-in-law} \OP{deɛmine} \TP{deɛmɛɛ} 
\LX{deɛmɛkonlɔŋ} \HM{} \PH{dɪ́ɛ́mɛ́kʊ́nlɔ́ŋ} \PS{n.~} \GE{the first layer of TZ in a bowl} \OP{deɛmɛkonlɔnne} \TP{deɛmɛkonlɔnnɛɛ} 
\LX{deɛmpɔge} \HM{} \PH{dɪ̀᷉ɛ̀pɔ́gɪ́} \PS{n.~} \GE{mother-in-law} \OP{deɛmpɔgeba/deɛmpɔgemine} \TP{deɛmpɔgrɛɛ} 
\LX{deɛnaa} \HM{} \PH{dɪ́ɛ́náá} \PS{n.~} \GE{plaything, toy} \OP{deɛne} \TP{deɛnɛɛ} 
\LX{deɛndɔɔ} \HM{} \PH{dè᷉ɛdɔ́ɔ́} \PS{n.~} \GE{father-in-law} \OP{deɛndɔbɔ/deɛndɔɔmine} \TP{deɛ᷉dɔɔrɛɛ} 
\LX{deɛne} \HM{} \PH{dɪ̀ɛ̀nɪ́} \PS{n.~} \SN \GE{play} \XV{Bibiiri nɔŋe la deɛne yaga.} \XE{Children love play.} \XV{A bibiiri ba nɔŋe a deɛne ŋa.} \XE{The children do not like this play.} \SN \GE{joke} \XV{Bonso ka fo de a yelnimize ŋa eŋe deɛne?} \XE{Why do you take this serious matter for a joke?.} \SN \GE{drama} \XV{Zenɛ te na kaa la karembiiri deɛne na naŋ kyaare ‘bibiiri emmaaroŋ yɛlɛ’.} \XE{Today we shall watch a drama piece by pupils on the title ‘children’s health’.} \TP{deɛnɛɛ} \VA{de᷉ɛdeɛnɛ}
\LX{deɛraa} \HM{} \PH{dɪ́ɛ́ráá} \PS{n.~} \GE{an instrument for scraping and smoothing the edge of objects} \OP{deɛre} \TP{deɛrɛɛ} 
\LX{deɛre} \HM{} \PH{dɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to trim, to cut to make even (mouth of calabash)} \PD{deɛrɛɛ, deɛrɛ, deɛrɛ, deɛrebɛ, deɛraa, deɛrebɛ } 
\LX{dɛge} \HM{} \PH{dɛ́gɪ́} \PS{v.~} \GE{to be dirty} \PD{dɛgɛɛ, dɛgrɛ, dɛgrɛ, -, -, - } 
\LX{degedaga} \HM{} \PH{dɪ̀gɪ̀dàgà} \PS{adv.~} \GE{thick, heavy, and immovable} 
\LX{dɛgelɛkɛlɛ} \HM{} \PH{dɛ̀glɛ̀kɛ̀lɛ̀} \PS{adj.~} \GE{fatty and plentiful e.g of a big dinner (food) on a rich couple’s wedding day} 
\LX{degere} \HM{} \PH{dɪ̀grɪ́} \PS{n.~} \GE{dwarf, one stunted in growth} \OP{degɛ} \TP{degerɛɛ} 
\LX{dɛgere} \HM{} \PH{dɛ́grɪ̀} \PS{n.~} \GE{dirt} \OP{dɛgɛ} \TP{dɛgerɛɛ} 
\LX{dɛgere iribu} \HM{} \PH{dɛ̌grɪ́ íríbú} \PS{n.~} \SN \GE{circumcision (male)} \SN \GE{excision (female)} 
\LX{degili} \HM{} \PH{dèglì} \PS{v.~} \GE{not to be of substance} \PD{degilee, degile, degile, degilibe, degilaa, degilibe } 
\LX{degili} \HM{} \PH{dèglì} \PS{v.~} \GE{to marry one's sibling's husband} \PD{degilee, degile, degile, degilibe, degilaa, degilibe } 
\LX{deli} \HM{} \PH{délí} \PS{v.~} \GE{to have one's eyes dazzled; to look without seeing} \PD{delee, dele, dele, delibe, delaa, delibe } 
\LX{dɛlle} \HM{} \PH{dɛ́llɪ́} \PS{v.~} \GE{to lean against, to rely upon for support} \PD{dɛllɛɛ, dɛllɛ, dɛllɛ, dɛlebɛ, dɛllaa, dɛlebɛ } 
\LX{dɛlle} \HM{} \PH{dɛ́llɪ́} \PS{v.~} \GE{to stage a corpse} \PD{dɛllɛɛ, dɛllɛ, dɛllɛ, dɛlebɛ, dɛllaa, dɛlebɛ } 
\LX{dɛlle} \HM{} \PH{dɛ̀llɪ̀} \PS{v.~} \GE{to remove the top layer of liquid, to decant, to skim} \PD{dɛllɛɛ, dɛllɛ, dɛllɛ, dɛlebɛ, dɛllaa, dɛlebɛ } 
\LX{deme} \HM{} \PH{dèmé} \PS{pron.~} \GE{own pl. (yours/mine)} \XV{Fo deme la gaŋ n deme.} \XE{Yours are more than mine.} 
\LX{dɛmɛdɛmɛ} \HM{} \PH{dɛ̀mɛ̀dɛ́mɛ́} \PS{n.~} \GE{making an effort} 
\LX{dɛmɛdɛmɛ} \HM{} \PH{dɛ̀mɛ̀dɛ̀mɛ̀} \PS{adv.~} \SN \GE{full to capacity with liquid e.g. a swollen foot} \SN \GE{big, fat and tall or long e.g a huge python} 
\LX{demee} \HM{} \PH{déméé} \PS{n.~} \GE{unidentified people} \XV{Yeŋ demee la wa?} \XE{Which people have come?.} 
\LX{demee} \HM{} \PH{déméé} \PS{n.~} \GE{owners} \XV{A noore taa la demee.} \XE{The fowls have owners.} 
\LX{demm} \HM{} \PH{dèmm} \PS{adv.~} \GE{large and upright} 
\LX{dɛmmo} \HM{} \PH{dɛ̀mmʋ́} \PS{n.~} \GE{conversation, talk, a pending problem} \OP{dɛmɛ} \TP{ dɛmɛɛ} 
\LX{dendâee} \HM{} \PH{dɪ̀ndà᷉ɪ̀᷉ɪ̀᷈} \PS{adv.~} \GE{appearing suddenly and frighteningly} 
\LX{dendege} \HM{} \PH{dèndégé} \PS{n.~} \GE{a woman given in marriage to her sister's husband} \OP{dendegri} \TP{dedegree} 
\LX{dɛndɛlɛŋ} \HM{} \PH{dɛ̀ndɛ̀lɛ̀ŋ} \PS{adv.~} \GE{long and wavy} 
\LX{dendeŋe} \HM{} \PH{d̀endèŋé} \PS{adj.~} \GE{upright} \OP{dendenni} \TP{dendennee} 
\LX{dɛndɛŋe} \HM{} \PH{dɛ̀ndɛ́ŋ} \PS{adj.~} \GE{first, earlier} \VA{dɛŋ}
\LX{dendeŋi} \HM{} \PH{děndéŋ} \PS{adv.~} \GE{upright, straight} 
\LX{dendeŋŋe} \HM{} \PH{déndéŋŋé} \PS{n.~} \SN \GE{dagger} \SN \GE{the longest toe of a bird} \OP{dendenni} \TP{dendennee} 
\LX{dendɔgaŋe} \HM{} \PH{dɪ̀ndɔ́gǎŋ} \PS{n.~} \GE{bastard} \OP{dendɔganne} \TP{dedɔgannɛɛ} \VA{dendɔgaŋa}
\LX{dendɔlee} \HM{} \PH{dɪ̀ndɔ́léé} \PS{n.~} \GE{bastard} \OP{dendɔbilii} \TP{dendɔbillee} 
\LX{dendɔmɔ} \HM{} \PH{dɪ́ndɔ́mɔ́} \PS{n.~} \GE{enemy} \OP{dendɔnne} \TP{dendɔnnɛɛ} \VA{dendɔma, dɔndɔmɔ}
\LX{dendɔre} \HM{} \PH{dɪ̀ndɔ́rɪ̀} \PS{n.~} \SN \GE{doorway} \SN \GE{frontage of the main house} \OP{dendɔɛ} \TP{dendɔrɛɛ} 
\LX{dɛŋe} \HM{} \PH{dɛ́ŋ} \PS{adj.~} \GE{first} 
\LX{dɛŋɛɛ} \HM{} \PH{dɛ̀ŋɛ̀ɛ̀} \PS{adj.~} \GE{plentiful} 
\LX{dɛŋeneŋ} \HM{} \PH{dɛ̀ŋɪ́nɪ́ŋ} \PS{adj.~} \GE{first instance} 
\LX{deŋgeluu} \HM{} \PH{dèŋgélúú} \PS{n.~} \GE{sweet potato} \OP{deŋgele} \TP{ deŋgellee} 
\LX{deŋgereŋ} \HM{} \PH{dèŋgèrèŋ} \PS{adv.~} \GE{upright, stiff, unbending and strong} 
\LX{deni} \HM{} \PH{dèní} \PS{pron.~} \GE{that which belongs to} \XV{A dagere poɔ n deni la bile.} \XE{Among the boxes mine(the one that belongs to me) is the smallest.} 
\LX{dɛnne} \HM{} \PH{dɛ̀nnɪ̀} \PS{v.~} \SN \GE{to straighten} \XV{Da maaroŋ ka ba maŋ dɛnne, ba ba dɛnnɛ dakuoŋaa.} \XE{You can straighten a fresh stick, but not a dry one.} \SN \GE{to roam} \XV{A teŋɛ zaa ka o dɛnne kyɛ naŋ ba bale.} \XE{She has roamed the whole town but is not tired yet.} \SN \GE{to be full to capacity (with liquid)} \XV{Saa mie la ka a baa dɛnne o zaa.} \XE{It has rained and the stream is full to capacity.} \PD{dɛnnɛɛ, dɛnnɛ, dɛnnɛ, dɛnemɛ, dɛnnaa, dɛnemɛ } 
\LX{dɛnne} \HM{} \PH{dɛ́nnɪ́} \PS{v.~} \GE{praise} \PD{dɛnnɛɛ, dɛnnɛ, dɛnnɛ, dɛnemɛ, dɛnnaa, - } \VA{danne}
\LX{denni} \HM{} \PH{dénnì} \PS{v.~} \GE{to carry on one's head without the support of one's hands} \PD{dennee, denne, denne, denime, dennaa, denimee } \VA{delli}
\LX{deŋo} \HM{} \PH{dɪ̀ŋʊ́} \PS{n.~} \GE{preference, wish (not caring about others)} 
\LX{deraa} \HM{} \PH{dɪ̀ràà} \PS{adj.~} \GE{strong and hard, unbendable} 
\LX{Dɛre} \HM{} \PH{dɛ́rɪ́} \PS{proper noun.~} \GE{name traditionally given to a male child born after the death of an infant boy of the same mother} 
\LX{dɛre} \HM{} \PH{dɛ́rɪ́} \PS{n.~} \GE{house mouse} \OP{dɛrɛɛ} 
\LX{dɛre} \HM{} \PH{dɛ́rɪ̀} \PS{n.~} \GE{ladder} \OP{dɛɛ} \TP{dɛrɛɛ} 
\LX{dɛre} \HM{} \PH{dɛ́rɪ́} \PS{v.~} \GE{to boil pito at the second stage of brewing} \PD{dɛrɛɛ, dɛrɛ, dɛrɛ, dɛrebɛ, dɛraa, - } 
\LX{deree} \HM{} \PH{déréé} \PS{n.~} \GE{negligence, carelessness} 
\LX{dɛremaakonokono} \HM{} \PH{dɛ́rɪ́máákónókónó} \PS{n.~} \GE{a traditional women's dance} 
\LX{Dɛrepɔge} \HM{} \PH{dɛ̀rɪ̀pɔ́gɪ́} \PS{ n.~} \GE{name traditionally given to a female child born after the death of an infant girl of the same mother} 
\LX{deri} \HM{} \PH{dèrì} \PS{v.~} \SN \GE{to harden, to dry (wet or muddy place or substance)} \XV{Pampana ŋa ka ba salle a dankyini o naŋ ba deri.} \XE{The wall has just been plastered and is not hardened yet.} \SN \GE{to go into a state of convulsion, to collapse in hysteria} \XV{Ka kɔmmaaroŋ waare pɔgenyaŋaa naŋ zoro ɔɔre o na deri la.} \XE{If cold water is accidently sprayed with water she will collapse in cold.} \PD{deree, dere, dere, deribe, deraa, deribe } 
\LX{detaareŋ} \HM{} \PH{dɪ̀táárɪ̌ŋ} \PS{adv.~} \GE{3 years ago} 
\LX{deyɛŋ} \HM{} \PH{dɪ̀yɛ̌ŋ} \PS{adv.~} \GE{last year} 
\LX{dêe} \HM{} \PH{dɪ̀̂ɪ̀̂} \PS{adv.~} \GE{unlimited} 
\LX{de᷉ɛ} \HM{} \PH{dɪ̀᷉ɛ̀᷉} \PS{v.~} \GE{to play (child)} \PD{de᷉ɛ᷉ɛ᷉, deɛnɛ, deɛnɛ, deɛnemɛ, deɛnaa, - } 
\LX{dêɛ} \HM{} \PH{dɪ̂̀᷉ɛ̀᷈} \PS{v.~} \GE{to play, to joke} \PD{dêɛɛ, deɛnɛ, deɛnɛ, deɛnemɛ, deɛnaa, deɛnemɛ } 
\LX{dɛ᷉ɛ} \HM{} \PH{dɛ̀᷉ɛ̀᷉} \PS{adv.~} \GE{open and endless e.g. grassland, desert, ocean} 
\LX{de᷉ɛbaŋ} \HM{} \PH{dɪ̀᷉ɛ̀bàŋ} \PS{n.~} \GE{drama, play} 
\LX{de᷉ɛdeɛnɛ} \HM{} \PH{dɪ̀᷉ɛ̀dɪ́ɛ́nɛ́} \PS{n.~} \SN \GE{one who plays e.g. of a child} \SN \GE{a character in a story, drama} \OP{de᷉ɛdeɛmɛ} \TP{de᷉ɛdeɛmɛɛ} \VA{deɛne}
\LX{de᷉ɛde᷉ɛkaraa} \HM{} \PH{dɪ̀᷉ɛ̀dɪ̀᷉ɛ̀kàràà} \PS{n.~} \SN \GE{lead player} \SN \GE{protagonist (lit.)} \OP{de᷉ɛde᷉ɛkarre} \TP{de᷉ɛde᷉ɛkarɛɛ} 
\LX{dêɛkakalle} \HM{} \PH{dɪ̂̀ɛ̀kákàllɪ́} \PS{n.~} \GE{rough, dangerous play e.g. throwing hard objects at each other} \TP{  dêɛkakalɛɛ/dêɛkakallɛɛ.} 
\LX{de᷉ɛkoɔbo} \HM{} \PH{dɪ̀᷉ɛ̀kʊ́ɔ́bʊ́} \PS{n.~} \GE{farmwork done for one's in-law} \TP{de᷉ɛkɔrɛɛ} 
\LX{de᷉ɛkpa᷉ao} \HM{} \PH{dɪ̀ɛ̀kpá᷉áʊ́} \PS{n.~} \GE{guineafowl meant for one's in-law} \OP{de᷉ɛkpinni} \TP{de᷉ɛkpinnee} 
\LX{de᷉ɛkuori} \HM{} \PH{dɪ́᷉ɛ́᷉kúórí} \PS{n.~} \GE{funeral of an in-law} \OP{de᷉ɛkoe} \TP{de᷉ɛkoree} 
\LX{de᷉ɛmakoma} \HM{} \PH{dɪ̀᷉ɛ̀màkʊ́má} \PS{n.~} \GE{grandmother-in-law} \OP{de᷉ɛmakommine} 
\LX{dêɛne} \HM{} \PH{dɪ̂̀ɛ̀nɪ́} \PS{n.~} \GE{play, joke} \TP{deɛnɛɛ} 
\LX{de᷉ɛsa᷉akoma} \HM{} \PH{dɪ̀᷉ɛ̀᷉sà᷉à᷉kʊ́má} \PS{n.~} \GE{grandfather-in-law} \OP{de᷉ɛsa᷉akommine} 
\LX{de᷉ɛyiri} \HM{} \PH{dɪ̀᷉ɛ̀᷉yírí} \PS{n.~} \GE{house of an in-law} \OP{de᷉ɛyie} \TP{de᷉ɛyiree} 
\LX{di} \HM{} \PH{dì} \PS{v.~} \SN \GE{to eat, to consume} \XV{Bibiiri seŋ ka ba maŋ di bondirii kyɛ gaa sakuuri.} \XE{} \SN \GE{to burn,  to blaze} \XV{Vu᷉u᷉ di la a tuu zaa.} \XE{The whole shrub has been burnt down by fire.} \SN \GE{to erode, to make sore} \XV{A bie noɔre die la ka ba sage na di bonzaa.} \XE{The child has a sore mouth and refuses to eat anything.} \SN \GE{to overcome, to win} \SN \GE{to assume a throne} \SN \GE{to judge} \XV{Ba di la ba zɔɔre sɛrɛɛ na baare.} \XE{They have finished judging that case of fighting.} \PD{diee, dire, dire, diribe, diraa, diribe } \SE{di danseɛ} \GE{to give witness} \SE{di dapaaloŋ} \GE{to show elegance} \SE{di dɛmmo} \GE{to converse} \SE{di dɔɔloŋ} \GE{to show bravery} \SE{di dɔgɛɛ} \GE{to suffer} \SE{di kannyiri} \GE{to be patient} \SE{di naŋ} \GE{to be scarce} \SE{di naaloŋ} \GE{to gain wealth or authority} \SE{di neɛ} \GE{(of a person) to be worthless} \SE{di noɔ} \GE{to enjoy} \SE{di sane} \GE{to owe} \SE{di sɛrɛɛ} \GE{to judge} \SE{di suuri} \GE{to forgive} \SE{di vi} \GE{to be ashamed} \SE{di yuori} \GE{to have an unworthy name} \SE{kôɔ naŋ di neɛv} \GE{to drown} 
\LX{di daa} \HM{} \PH{dì dàá} \PS{v.~} \GE{to haggle, to negotiate a price} \PD{-, -, -, -, -, - } 
\LX{di dɛgere} \HM{} \PH{dì dɛ́gɪ́rɪ́} \PS{v.~} \GE{to suffer} \PD{die dɛgere, dire dɛgere, -, -, -, - } 
\LX{di dɛmɛ} \HM{} \PH{dì dɛ́mɛ́} \PS{v.~} \GE{to chat, to converse} \PD{-, -, -, -, -, - } 
\LX{di dɔgɛɛ} \HM{} \PH{di ̀dɔ́gɛ́ɛ́} \PS{v.~} \GE{to suffer} \PD{-, -, -, -, -, - } 
\LX{di faŋa} \HM{} \PH{dì fáŋá} \PS{v.~} \GE{wild power} \PD{-, -, -, -, -, - } 
\LX{di kannyiri} \HM{} \PH{dì kànɲìrì} \PS{v.~} \GE{to be patient} \PD{-, -, -, -, -, - } 
\LX{di naaloŋ} \HM{} \PH{dì náálʊ́ŋ} \PS{v.~} \GE{to ascend to the throne} \PD{-, -, -, -, -, - } 
\LX{di naŋe} \HM{} \PH{dì náŋ} \PS{v.~} \GE{to be scarce} \PD{-, -, -, -, -, - } 
\LX{di neɛ} \HM{} \PH{dì nɪ́ɛ́} \PS{v.~} \GE{to be hopeless} \PD{-, -, -, -, -, - } 
\LX{di sane} \HM{} \PH{dì sánɪ̀} \PS{v.~} \GE{to owe, to be indebted to} \PD{-, -, -, -, -, - } 
\LX{di seebo} \HM{} \PH{dì sɪ́ɪ́bʊ́} \PS{v.~} \GE{to harvest, to collect (honey from a hive)} \PD{-, -, -, -, -, - } 
\LX{di sɛrɛɛ} \HM{} \PH{dì sɛ̀rɛ́ɛ́} \PS{v.~} \GE{to judge} \PD{-, -, -, -, -, - } 
\LX{di suuri} \HM{} \PH{dì súúrí} \PS{v.~} \GE{to forgive} \PD{-, -, -, -, -, - } 
\LX{di vi} \HM{} \PH{dì ví} \PS{v.~} \GE{to be ashamed} \PD{-, -, -, -, -, - } 
\LX{dibogi} \HM{} \PH{dìbògí} \PS{n.~} \GE{where the family line begins and where it ends} \OP{dibogri} \TP{dibogree} 
\LX{die} \HM{} \PH{dìè} \PS{n.~} \SN \GE{room} \SN \GE{compound, house} \OP{deri} \TP{deree} 
\LX{diebie} \HM{} \PH{dìébíè} \PS{n.~} \SN \GE{cat} \SN \GE{son of the house} \OP{diebiiri} \TP{diebiiree} \VA{yiribie}
\LX{diebugo} \HM{} \PH{dìèbùgó} \PS{n.~} \GE{one who knows nothing outside home (untravelled)} \OP{diebugri} \TP{diebugree} \VA{diene}
\LX{dieli} \HM{} \PH{dìèlì} \PS{v.~} \GE{to float} \PD{dielee, diele, diele, dielibe, dielaa, dielibe } 
\LX{dieneɛ} \HM{} \PH{dìènɪ́ɛ́} \PS{n.~} \GE{a close relative} \OP{dienoba} \VA{diebugo}
\LX{difudeɛ} \HM{} \PH{dìfùdɪ́ɛ́} \PS{n.~} \GE{cheap} \TP{difudeerɛɛ} 
\LX{diganaa} \HM{} \PH{dìgánáà} \PS{} \GE{bedroom} 
\LX{digaŋene} \HM{} \PH{dìgàŋɪ̀nɪ́} \PS{n.~} \GE{gluttony} 
\LX{digaŋgaŋ} \HM{} \PH{dìgàŋgáŋ} \PS{n.~} \GE{craftiness} 
\LX{digbogbo} \HM{} \PH{dìgbògbò} \PS{adj.~} \GE{gluttonous, eating too much} 
\LX{digi} \HM{} \PH{digi} \PS{v.~} \SN \GE{to chase, to run after} \SN \GE{to drive away} \PD{digee, digire, digre, digribe, digiraa, digiribe } 
\LX{digoo} \HM{} \PH{dìgòò} \PS{n.~} \GE{a disease affecting bean plants} 
\LX{dii} \HM{} \PH{díì} \PS{v.~} \SN \GE{to press down e.g. with a heavier object} \XV{A teɛ da le dii la a lɔɔre ka noba perre.} \XE{The tree fell on the car and injured some people.} \SN \GE{to preserve (bean seed in ashes)} \XV{A yidaampɔge dii la a bɛŋɛ ka a na sa᷉a᷉.} \XE{The head of the women preserved the bean seed in ash to prevent it from getting destroyed by insects.} \SN \GE{to pile up} \XV{A boɔre la dii taa do teɛ a dasɛrɛɛ.} \XE{The bags are packed on top of each other up to the ceiling.} \PD{diiee, diire, dire, diiribe/diiribee, diiraa, - } 
\LX{dii} \HM{} \PH{díí} \PS{v.~} \SN \GE{to store grain in ashes} \XV{A yidaampɔge dii la a bɛŋɛ ne tampɛloŋ ka a na ta sa᷉a᷉.} \XE{The head of the women preserved the bean seed in ash to prevent it from getting destroyed by insects.} \SN \GE{plural of dii (to pile objects)} \PD{diiee, diire, dire, diiribe/diiribee, diiraa, diiribe } 
\LX{diibu} \HM{} \PH{dííbú} \PS{n.~} \GE{food} \TP{diiree} 
\LX{dikpaare} \HM{} \PH{dìkpáárɪ́} \PS{n.~} \GE{back of a room} \OP{dikpaɛ} \TP{dikpaarɛɛ} 
\LX{dikpeɛraa} \HM{} \PH{dìkpɪ́ɛ́ráá} \PS{n.~} \GE{living room} \OP{dikpeɛre} 
\LX{dilɔɔraa} \HM{} \PH{dìlɔ́ɔ́ràá} \PS{n.~} \GE{a room thatched with wood and earth} \OP{dilɔɔre} \TP{ dilɔɔrɛ} 
\LX{dindigili} \HM{} \PH{díndíglí} \PS{n.~} \GE{electric fish} \OP{dindigile} \TP{dindigilee} \VA{dindigiri}
\LX{dindigiri} \HM{} \PH{díndígirí} \PS{n.~} \GE{catfish, a type of fish that gives you an electric shock when touched} \OP{dindige/dindigire} \TP{dindigiree} \VA{dindigli}
\LX{dipanaa} \HM{} \PH{dìpànáá} \PS{n.~} \GE{plastered room} \OP{dipanne} \TP{dipannɛɛ} 
\LX{dire} \HM{} \PH{dìré} \PS{v.~} \GE{to be sharp} \PD{diee, dire, dire, diribe, diraa, dire } 
\LX{diruŋ} \HM{} \PH{dìrúŋ} \PS{n.~} \SN \GE{food, plenty of food} \SN \GE{feast} 
\LX{disogɔ} \HM{} \PH{dìsʊ́gɔ́} \PS{n.~} \GE{middle room} \OP{disogere} \TP{disogerɛɛ} \VA{disoga}
\LX{diteenɛ} \HM{} \PH{dìtɪ́ɪ́nɛ́} \PS{n.~} \GE{mediator, spokesman in a marriage agreement} \OP{diteenemɛ} \TP{diteenɛɛ} 
\LX{divuo} \HM{} \PH{dìvùó} \PS{n.~} \GE{abandoned room; room without a roof} \OP{divori} \TP{ divoree} 
\LX{do} \HM{} \PH{dò} \PS{v.~} \GE{to roll along the ground (circular object)} \PD{doe, duoro, duoro, duori, duoraa, duoro } 
\LX{do} \HM{} \PH{dó} \PS{v.~} \SN \GE{to climb up, to ascend} \SN \GE{to soar} \SN \GE{to mate, to crossbreed (nonhuman)} \SN \GE{to climb out of the ground into a new nest (trapped termites)} \PD{doe, duoro, duoro, duoribo, duoraa, duoribo } \SE{do saa} \GE{to be high} 
\LX{dɔ} \HM{} \PH{dɔ̀} \PS{v.~} \GE{to spill over e.g. boiling water from yam being cooked over fire} \PD{dɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\LX{dɔ} \HM{} \PH{dɔ̀} \PS{v.~} \GE{to pay the last bride price} \PD{dɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\LX{dɔ} \HM{} \PH{dɔ́} \PS{v.~} \GE{to weed} \PD{dɔɛ, doɔrɔ, doɔrɔ, doɔreba, doɔraa, doɔrebɔ } 
\LX{doba} \HM{} \PH{dʊ́bà} \PS{n.~} \GE{island} \OP{dobare} \TP{dobarɛɛ} 
\LX{dobaa} \HM{} \PH{dòbáá} \PS{n.~} \GE{pig (domesticated)} \OP{dobaare} \TP{dobaarɛɛ} 
\LX{dobadaa} \HM{} \PH{dòbàdáá} \PS{n.~} \GE{male pig} \OP{dobadaare} \TP{dobadaarɛɛ} \VA{dodaa}
\LX{dobagbɛre} \HM{} \PH{dòbàgbɛ́rɪ̀} \PS{n.~} \GE{the mane of a pig} \OP{dogbɛɛ} \TP{dogbɛɛrɛɛ} 
\LX{dobanɛne} \HM{} \PH{dòbànɛ́nɪ̀} \PS{n.~} \GE{pork} \OP{dobanɛmɛ} \TP{dobanɛmɛɛ/dobanɛnnɛɛ} 
\LX{dobapɔle} \HM{} \PH{dòbàpɔ́lɪ̀} \PS{n.~} \GE{young  pig} \OP{dobapɔlɔ} \TP{ dopɔllɛɛ} \VA{dopɔle}
\LX{dobasugo} \HM{} \PH{dòbàsúgò} \PS{n.~} \GE{herd of swine} \OP{dobasugri} \TP{dobasugree} 
\LX{dodaa} \HM{} \PH{dòdáá} \PS{n.~} \GE{wild male pig} \OP{dodaare} \TP{dodaarɛɛ} 
\LX{dɔdɔgeroŋ} \HM{} \PH{dɔ́dɔ́grʊ̂ŋ} \PS{n.~} \GE{wretchedness, bad treatment} 
\LX{dodoo} \HM{} \PH{dʋ̀dʋ́ʋ́} \PS{adj.~} \GE{large and bloated} 
\LX{dodoo} \HM{} \PH{dʊ̀dʊ́ʊ́} \PS{n.~} \GE{an illness that causes bloating of the whole body} \OP{dodoore} \TP{doddoorɛɛ} 
\LX{dɔɛ} \HM{} \PH{dɔ́ɛ́} \PS{n.~} \GE{last bride price} 
\LX{doge} \HM{} \PH{dʋ́gɪ̀} \PS{n.~} \GE{large pot} \OP{dogre} \TP{dogrɛɛ.} 
\LX{doge} \HM{} \PH{dʋ́gɪ́} \PS{v.~} \SN \GE{to boil, to cook} \SN \GE{to brew} \SN \GE{to become very hot} \PD{dogɛɛ, dogerɔ, dogerɔ, dogereba, dogeraa, dogerɔ } 
\LX{dɔge} \HM{} \PH{dɔ̀gɪ̀} \PS{v.~} \SN \GE{to bear (child), to give birth} \SN \GE{to flower (grass pollen)} \SN \GE{take something from the top of another} \XV{A bie dɔge la a nɛne a saabo zu.} \XE{The child has taken the meat which was on top of the food.} \PD{dɔgɛɛ, dɔgrɔ, dɔgrɔ, dɔgreba, dɔgraa, dɔgreba } 
\LX{dogebile} \HM{} \PH{dʋ́gɪ́bílé} \PS{n.~} \GE{small pot} \OP{dogebilii} \TP{dogebillee} 
\LX{dɔgɛɛ} \HM{} \PH{dɔ́gɛ́ɛ́} \PS{n.~} \GE{suffering} 
\LX{dogekoma} \HM{} \PH{dʋ́gkʋ́má} \PS{n.~} \GE{a type of tree with medicinal properties} 
\LX{dogele} \HM{} \PH{dʋ́glɪ̀} \PS{v.~} \GE{to put over a fire} \PD{dogelɛɛ, dogerɔ, dogelɔ, dogelebɔ, dogelɔ, dogelebɔ } 
\LX{dɔgele} \HM{} \PH{dɔ̀glɪ̀} \PS{v.~} \SN \GE{to place on top of something} \XV{De a nɛne dɔgele dankyini zu ka baare ta wa de o.} \XE{Place the meat on the wall so that dogs do not eat it.} \SN \GE{to exceed, to be extra} \XV{A bie yɔɔ la a libie ka a dɔgele ka a seŋ ka fo leɛ de a dɔgelɔ ko o.} \XE{The child has overpaid and so the extra money should be returned to him.} \PD{dɔglɛɛ, dɔglɔ, dɔglɔ, dɔgleba, dɔglaa, - } 
\LX{dɔgele} \HM{} \PH{dɔ̀glɪ̀} \PS{v.~} \GE{to mix beans with flour} \PD{dɔglɛɛ, dɔglɔ, dɔglɔ, dɔgleba, dɔglaa, - } 
\LX{dogelee} \HM{} \PH{dʊ́gléé} \PS{n.~} \GE{small pot} \OP{dugili} 
\LX{dɔgere} \HM{} \PH{dɔ́grɪ̀} \PS{v.~} \GE{to punish, to make to suffer} \PD{dɔgerɛɛ, dɔgerɔ, dɔgerɔ, dɔgerebɔ, dɔgeraa, dɔgerebɔ } 
\LX{dɔgerɔ} \HM{} \PH{dɔ́grɔ} \PS{n.~} \GE{relative (blood)} \OP{dɔgerebɔ} \TP{dɔgerebɛɛ} 
\LX{dɔgerɔlenne} \HM{} \PH{dɔ̀grɔ́lénné} \PS{n.~} \GE{ear of maize with multicoloured kernels} \TP{dɔgerɔlennee} 
\LX{dɔgeroŋ} \HM{} \PH{dɔ̀grʊ̂ŋ} \PS{n.~} \GE{descent} 
\LX{dɔgeroŋ} \HM{} \PH{dɔ́grʊ̂ŋ} \PS{n.~} \GE{penalty, punishment} 
\LX{dogi} \HM{} \PH{dògì} \PS{v.~} \GE{to remove an object from the top of something} \XV{O dogi la a kuuri na naŋ pare a dankyini zu.} \XE{S/he has removed the stone from the top of the wall.} \PD{dogee, dogiro, dogiro, dogribo, dogiraa, dogiribo } 
\LX{dogo} \HM{} \PH{dʊ̀gʊ́} \PS{n.~} \GE{a type of dance} \TP{dogrɛɛ} 
\LX{dogodogo} \HM{} \PH{dógódògò} \PS{n.~} \GE{power of charm in magic} \OP{dogodogri} \TP{dogodogree} 
\LX{dɔkele} \HM{} \PH{dɔ̀kélé} \PS{n.~} \GE{kernels from dawdaw fruit} \TP{dɔkelee} \VA{tokele}
\LX{dɔkone} \HM{} \PH{dɔ̀kʊ́nɪ́} \PS{n.~} \GE{kenkey (thick corn paste shaped into balls and boiled in leaves)} \OP{dɔkona} \TP{dɔkonnɛɛ} \VA{dakone}
\LX{dole} \HM{} \PH{dʋ̀lɪ̀} \PS{v.~} \GE{to be unaware} \PD{dolɛɛ, dollɔ, dollɔ, dolebɔ, dollaa, dolebɔ } 
\LX{dole} \HM{} \PH{dʊ́lɪ́} \PS{v.~} \GE{to raise (one's head) to face upwards (when moving through tall grass)} \PD{dolɛɛ, dollɔ, dollɔ, dolebɔ, dolaa, dolebɔ } 
\LX{dɔle} \HM{} \PH{dɔ́lɪ́} \PS{adv.~} \GE{pop out smoothly e.g. from something slippery} 
\LX{doledole} \HM{} \PH{dʊ̀lɪ̀dʊ́lɪ́} \PS{v.~} \GE{to swing one's head around} \PD{doledolɛɛ, doledolɔ, doledolɔ, doldolebɔ, doledolaa, doledolebɔ } 
\LX{dolee} \HM{} \PH{dolee} \PS{n.~} \GE{small wild pig} \OP{dobili} 
\LX{doli} \HM{} \PH{dólí} \PS{n.~} \GE{a type of tree whose bark is used as a disinfectant for washing pito pots} \OP{dolo} \TP{dolee/dollee} 
\LX{dɔlle} \HM{} \PH{dɔ́llɪ̀} \PS{v.~} \GE{to stretch, to lengthen} \PD{dɔllɛɛ, dɔllɔ, dɔllɔ, dɔlebɔ, dɔllaa, dɔlebɔ } 
\LX{dɔlɔ} \HM{} \PH{dɔ́lɔ́́} \PS{n.~} \GE{way of doing something} \OP{dɔlɔdɔlɔ} 
\LX{dɔlɔdɔlɔ} \HM{} \PH{dɔ̀lɔ̀dɔ̀lɔ̀} \PS{adv.~} \GE{slimy and slippery} 
\LX{doloŋ} \HM{} \PH{dʊ̀lʊ̀ŋ} \PS{n.~} \GE{right side} \OP{dolle} \TP{dollɛɛ} 
\LX{domadoma} \HM{} \PH{dʊ̀màdʊ̀mà} \PS{adv.~} \GE{bloated and soft} 
\LX{dombaala} \HM{} \PH{dʊ̀mbààlá} \PS{n.~} \SN \GE{sick animal} \SN \GE{domesticated animal} \OP{dombaale} \TP{dombaalɛɛ.} 
\LX{dombɛroŋ} \HM{} \PH{dʊ̀mbɛ́rʊ̂ŋ} \PS{n.~} \SN \GE{large animal} \SN \GE{famous animal} \OP{dombɛrɛ} \TP{ dombɛrɛɛ/dombɛrrɛɛ.} 
\LX{dombɛroŋ} \HM{} \PH{dʊ̀mbɛ́rʊ́ŋ} \PS{n.~} \GE{large animal;} \OP{dombɛrɛ} \TP{dombɛrɛɛ/dombɛrrɛɛ} 
\LX{domm} \HM{} \PH{dòmm} \PS{adv.~} \GE{tall and big} 
\LX{domɔ} \HM{} \PH{dʊ̀mɔ́} \PS{n.~} \GE{totem; the sacred bird, animal, or insect of a clan} \OP{donne} \TP{domɛɛ} \VA{domɔ}
\LX{dompane} \HM{} \PH{dʋ̀mpánɪ̀} \PS{n.~} \GE{a type of snake that is short and poisonous} \OP{dompama} \TP{ dompannɛɛ.} 
\LX{dompɛge} \HM{} \PH{dʋ̀mpɛÕÅgɪ́} \PS{n.~} \GE{shell of an animal} \XV{ pl.} \XE{dompɛgere.} 
\LX{dompelaa} \HM{} \PH{dʊ̀mpɪ̀làá} \PS{n.~} \SN \GE{white animal} \SN \GE{an ordinary animal (believed not to have any spiritual powers)} \OP{dompeɛle} \TP{dompeɛlɛɛ} 
\LX{dompoll} \HM{} \PH{dʊ̀mpʊ̀ll} \PS{n.~} \GE{a material thing believed to be owed to a child from birth} \OP{dompolɔ} \TP{dompolɛɛ/dompollɛɛ} 
\LX{dompoɔ} \HM{} \PH{dʋ̀mpʋ́ɔ́} \PS{n.~} \GE{pregnant animal} \XV{pl.} \XE{} \OP{dompoore} \TP{ dompoorɛɛ.} 
\LX{doŋa} \HM{} \PH{dʊ́ŋá} \PS{n.~} \GE{animal} \OP{donne} \TP{donnɛɛ} \VA{doŋɔ}
\LX{doŋa} \HM{} \PH{dʊ́ŋá} \PS{n.~} \GE{mosquito} \OP{donne} \TP{donnɛɛ} \VA{doŋɔ}
\LX{dondaa} \HM{} \PH{dʋ̀ndáá} \PS{n.~} \GE{male animal} \XV{ pl.} \XE{} \OP{dondaare} \TP{ dondaarɛɛ.} 
\LX{dondigiraa} \HM{} \PH{dʊ̀ndígráá} \PS{n.~} \GE{something that drives mosquitoes away} \OP{dondigiri} \TP{dondigiree} 
\LX{dondigire} \HM{} \PH{dʊ̀ndígré} \PS{n.~} \GE{an insect-repellent plant used in keeping mosquitoes at bay} \OP{dondigiri} \TP{ dondigiree} 
\LX{dondoli} \HM{} \PH{dóndólí} \PS{n.~} \GE{dry land} \OP{dondolo} \TP{dondolee} 
\LX{dondolo} \HM{} \PH{dòndóló} \PS{n.~} \GE{planks used as seats (usually in front of a house)} 
\LX{dɔndɔŋ-karaa} \HM{} \PH{dɔ́ndɔ̀ŋ-kàràá} \PS{n.~} \GE{chief enemy} \OP{dɔndɔŋ-karre} \TP{dɔndɔŋ-karɛɛ} 
\LX{dɔndɔre} \HM{} \PH{dɔ̀ndɔ̀rɪ́} \PS{n.~} \GE{curd; coagulated milk} \OP{dɔndɔrre} \TP{dɔndɔrɛɛ} 
\LX{dɔndɔre} \HM{} \PH{dɔ̀ndɔ̀rɪ́} \PS{n.~} \GE{shade caused by clouds coming between the earth and the sun} \OP{dɔndɔrre} \TP{dɔndɔrɛɛ} 
\LX{doŋe} \HM{} \PH{dʊ̀ŋ} \PS{v.~} \GE{to bite} \SN \GE{to sting} \PD{doŋɛɛ, donɔ, donɔ, donemɔ, donaa, donemɔ } 
\LX{doŋe} \HM{} \PH{dʋ́ŋ} \PS{v.~} \GE{to be at close range} \XV{A saa da doŋe te la ka te zo fa᷉a᷉ kyɛ ka poɔ te.} \XE{The rain was so close we could not escape and therefore were caught in the downpour.} \PD{doŋɛɛ, donnɔ, donnɔ, donemɔ, donnaa, donemɔ } 
\LX{dɔŋe} \HM{} \PH{dɔ̌ŋ} \PS{n.~} \GE{enmity, hatred} \OP{dɔnne} \TP{dɔnnɛɛ} 
\LX{dɔŋe} \HM{} \PH{dɔ̀ŋ} \PS{v.~} \GE{to be lean, to be thin} \PD{dɔŋɛɛ, dɔnnɔ, dɔnnɔ, dɔnemɔ, dɔnnaa, dɔnemɔ } 
\LX{donɛɛ} \HM{} \PH{dʊnɛɛ} \PS{n.~} \GE{humankind} \OP{n/a} \TP{n/a} \VA{andonɛɛ}
\LX{doŋenokono} \HM{} \PH{dʊ́ŋɪ́nʊ́kʊ́nʊ́} \PS{n.~} \SN \GE{a type of tree whose leaves are edible} \SN \GE{the leaves of this tree} \TP{doŋenokonɛɛ} 
\LX{donfaa} \HM{} \PH{dʋnfaa} \PS{n.~} \GE{dangerous animal} \XV{ pl.} \XE{} \OP{donfaare} \TP{ donfaarɛɛ.} 
\LX{donforoŋ} \HM{} \PH{dʊ̀nfʊ̀rʊ̀ŋ} \PS{adv.~} \GE{large/big and puffed} \XV{A pɔge e la bɛroŋ donforoŋ lɛ.} \XE{The woman is big and puffed.} 
\LX{doŋguolaa} \HM{} \PH{dʊ̀ŋgúóláà} \PS{n.~} \GE{an animal that has been raised, a domestic animal} \OP{doŋguoli} 
\LX{doŋguolo} \HM{} \PH{dʊ̀ŋgúólò} \PS{n.~} \GE{animals that are shared between an original owner and a caretaker who raises them} \VA{boŋguolo}
\LX{doŋi} \HM{} \PH{dǒŋ} \PS{n.~} \GE{a Muslim religious festival} \OP{donni} \TP{donnee} 
\LX{doŋkaraa} \HM{} \PH{dʋŋkaraa} \PS{n.~} \GE{a strong-smelling plant used for chasing away mosquitoes} 
\LX{doŋkommiruu} \HM{} \PH{dʊ̀ŋkómmírúú} \PS{n.~} \GE{a type of tiny mosquito (bites only during day)} \SG{doŋkommiruu} \OP{doŋkommie} \TP{doŋkommiree/doŋkommiiree} 
\LX{doŋkɔɔseɛ} \HM{} \PH{dʋ̀ŋkɔ̀ɔ̀sɪ́ɛ́} \PS{n.~} \GE{wasp; a stinging flying insect that resembles a bee but longer} \OP{doŋkɔɔseere} \TP{doŋkɔɔseerɛɛ} 
\LX{doŋkuuree} \HM{} \PH{dʋkuuree} \PS{n.~} \GE{a type of short, very poisonous snake} \XV{ pl.} \XE{doŋkuuri/doŋkuuree.} 
\LX{donkyeenɛ} \HM{} \PH{dùnʧɪ́ɪ́nɛ́} \PS{n.~} \GE{shepherd, herdsman} \OP{donkyeenemɛ} \TP{donkyeemɛɛ} 
\LX{donnyaŋaa} \HM{} \PH{dʋ́nɲáŋáá} \PS{n.~} \GE{adult female animal} \OP{donyaŋene} \TP{donyaŋenɛɛ} 
\LX{donnyaŋaa} \HM{} \PH{dòɲáŋáá} \PS{n.~} \GE{wild female boar} \OP{donnyaŋene} \TP{donnyaŋenɛɛ} 
\LX{donsaana} \HM{} \PH{dʋ́nsáánà} \PS{n.~} \GE{strange animal} \XV{ pl.} \XE{} \OP{donsaama} \TP{ donsaanɛɛ.} 
\LX{donseɛ} \HM{} \PH{donsɪɛ} \PS{n.~} \GE{the waist of an animal} \XV{ pl.} \XE{} \OP{donseere} \TP{ donseerɛɛ.} 
\LX{donsɔgelaa} \HM{} \PH{dʊ̀nsɔ́gláá} \PS{n.~} \SN \GE{black animal} \SN \GE{an animal believed to have spiritual powers} \OP{donsɔgelɔ} 
\LX{donsugo} \HM{} \PH{dʊnsugo} \PS{n.~} \GE{large herds of animals} \OP{donsugri} \TP{donsugree} 
\LX{doɔ} \HM{} \PH{dʋ́ɔ̀} \PS{v.~} \GE{to raise one's head, to look up} \XV{O doɔ la o zu dayeni kyɛ leɛ gaŋ} \XE{S/he raised his/her head once and laid back down.} \PD{doɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\LX{doɔ} \HM{} \PH{dʋ́ɔ̀} \PS{v.~} \GE{to cause to become worse (behaviour)} \XV{A dɔɔ ŋmɛ la a bie doɔ bare.} \XE{The man worsened the situation by beating the child.} \PD{doɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\LX{doɔ} \HM{} \PH{dʋ̀ɔ̀} \PS{v.~} \GE{to ooze; to overflow} \XV{A baa paalɛɛ la doɔrɔ.} \XE{The river is overflowing its banks.} \PD{doɔɛ, doɔrɔ, -, -, -, - } 
\LX{doɔ} \HM{} \PH{dʋ̀ɔ̀} \PS{v.~} \GE{to fetch water with a large container} \PD{doɔɛ, doɔrɔ, -, -, -, - } 
\LX{doɔ} \HM{} \PH{dʋ̀ɔ́} \PS{n.~} \SN \GE{dawadawa tree} \SN \GE{fruit of the dawadawa tree (edible)} \OP{dɔre} \TP{dɔrɛɛ} 
\LX{dɔɔ} \HM{} \PH{dɔ̀ɔ̀} \PS{v.~} \SN \GE{to squat} \SN \GE{to bow (as in greeting)} \SN \GE{to excrete} \SN \GE{to perch} \PD{dɔɔɛ, dɔɔrɔ, dɔɔrɔ, dɔɔrebɔ, dɔɔraa, dɔɔrebɔ } 
\LX{dɔɔ} \HM{} \PH{dɔ́ɔ́} \PS{n.~} \SN \GE{man} \SN \GE{friend of the same sex} \OP{dɔbɔ/dɔba} \TP{dɔɔrɛɛ} 
\LX{dɔɔbile} \HM{} \PH{dɔ̀ɔ̀bílé} \PS{n.~} \GE{boy} \OP{dɔɔbilii} 
\LX{doɔbo} \HM{} \PH{dʋ́ɔ́bʋ̀} \PS{n.~} \GE{the act of weeding} 
\LX{Dɔɔda᷈a᷈} \HM{} \PH{} \PS{n.~} \GE{a male name} \OP{dɔɔda᷈a᷈nɪ᷈} \TP{dɔɔda᷈a᷈nɛ᷈ɛ᷈} 
\LX{doogɔ} \HM{} \PH{dʊ́ʊ́gɔ́} \PS{n.~} \GE{plenty of enjoyment (food and drink)} \TP{doogɛɛ} \VA{dooga}
\LX{dɔɔkɔɔteɛlɔɔ} \HM{} \PH{dɔ̀ɔ̀kɔ̀ɔ̀tɪ́ɛ́lɔ́ɔ́} \PS{n.~} \GE{a squatting kick game} 
\LX{doɔkpangyogele} \HM{} \PH{dʋ̀ɔ́kpá́ndzʋ̀glɪ̀} \PS{n.~} \GE{a large catapillar found on the leaves of the dawadawa tree} \OP{doɔkpangyoglɔ} \TP{ doɔkpangyoglɛɛ} \VA{doɔkpankyogele}
\LX{dɔɔ-la-pɔge} \HM{} \PH{dɔ̀ɔ̀-la-́pɔ́gɪ́} \PS{n.~} \GE{a man with feminine features} \TP{dɔɔ-la-pɔgrɛɛ} 
\LX{doɔle} \HM{} \PH{dʋ̀ɔ́lɪ́} \PS{n.~} \GE{xylophone mallet} \OP{doɔlɔ} \TP{doɔlɛɛ} 
\LX{doɔle} \HM{} \PH{dʋ̀ɔ̀lɪ̀} \PS{v.~} \GE{to walk carefully and majestically} \PD{doɔlɛɛ, doɔlɔ, doɔlɔ, doɔlebɔ, doɔlaa, doɔlebɔ } 
\LX{dɔɔlee} \HM{} \PH{dɔ̀ɔ̀léé} \PS{n.~} \GE{boy} \OP{dɔɔbilii} \TP{dɔɔbilee} \VA{dɔɔbile}
\LX{doɔloŋ} \HM{} \PH{dʊ̀ɔ́lʊ́ŋ} \PS{n.~} \GE{mudfish (lives in the mud during dry season)} \OP{doɔlɔ} \TP{doɔlɛɛ} 
\LX{dɔɔloŋ} \HM{} \PH{dɔ̀ɔ̀lʊ́ŋ} \PS{n.~} \SN \GE{bravery, dexterity} \SN \GE{penis (figurative)} 
\LX{dɔɔne} \HM{} \PH{dɔ̀ɔ̀nɪ́} \PS{n.~} \SN \GE{noise} \XV{N woŋ la dɔɔne a die poɔ, neɛkaŋa be la be.} \XE{I have heard some noise in the room, there must be someone in there.} \SN \GE{movement} \XV{Te nyɛ la a baala nɔgbɛmɛ dɔɔne kyɛ te ba woŋi a yɛlɛ o naŋ yeli.} \XE{We saw some movement of the patient’s lips but we did not hear what s/he said.} \SN \GE{disturbance, worry} \XV{A pɔge yeli ka o ba boɔrɔ dɔɔne ka anaŋ so ka o yi.} \XE{The woman said she did not want trouble and that is why she divorced.} 
\LX{dɔɔnyaŋaa} \HM{} \PH{dɔ̀ɔ̀ɲáŋáá} \PS{n.~} \GE{old man} \OP{ dɔɔnyaŋene} \TP{dɔɔnyaŋenɛɛ} 
\LX{doɔnyeraa} \HM{} \PH{dʊ̀ɔ̀ɲɪ́ráá} \PS{n.~} \GE{a type of small ant} \OP{doɔnyere} \TP{doɔnyerɛɛ} 
\LX{dɔɔ-ɔre} \HM{} \PH{dɔ̀ɔ̀-ɔ́rɪ́} \PS{n.~} \GE{infertile man} \TP{dɔɔ-ɔrɛɛ} \VA{dɔɔ-are}
\LX{doore} \HM{} \PH{dʊ̀ʋ̀rɪ̀} \PS{v.~} \GE{to continuously heave chunks of load or matter} \XV{Ba doore la ko᷉ɔ̂ lɛɛ te bale ba zaa.} \XE{They heaved water until they became tired.} \PD{doorɛɛ, doorɔ, doorɔ, doorebɔ, dooraa, doorebɔ } 
\LX{doɔre} \HM{} \PH{dʊ̀ɔ̀rɪ́} \PS{adj.~} \GE{yellow} \TP{doɔrɛɛ} 
\LX{dɔɔta} \HM{} \PH{dɔ́ɔ́tà} \PS{n.~} \SN \GE{doctor} \SN \GE{hospital, clinic} \OP{dɔɔtare} \TP{dɔɔtarɛɛ} 
\LX{dɔɔtayiri} \HM{} \PH{dɔ́ɔ́táyírí} \PS{n.~} \GE{hospital} \OP{dɔɔtayie} \TP{dɔɔtayiree} 
\LX{dopɔge} \HM{} \PH{dòpɔ́gɪ́} \PS{n.~} \GE{female bush pig} \OP{dopɔgeba} \TP{dopɔgerɛɛ} 
\LX{dopɔle} \HM{} \PH{dòpɔ́lɪ̀} \PS{n.~} \GE{young bush pig} \OP{dopɔlɔ} \TP{dopɔlɛɛ/dopɔllɛɛ} 
\LX{dore} \HM{} \PH{dʊ̀rɪ̀} \PS{n.~} \GE{thud; sound of an impact e.g. a heavy log falling} 
\LX{dɔre} \HM{} \PH{dɔ̀rɪ̀} \PS{v.~} \GE{to advise, to caution} \PD{dɔrɛɛ, dɔrɔ, dɔrɔ, dɔrebɔ, dɔraa, dɔrebɔ } 
\LX{dɔre} \HM{} \PH{dɔ́rɪ́} \PS{n.~} \GE{fruits of doɔ} \TP{dɔrɛɛ} 
\LX{doreba} \HM{} \PH{dʊ́rɪ́bà} \PS{n.~} \GE{driver} \OP{dorebare} \TP{dorebarɛɛ} 
\LX{doree} \HM{} \PH{dòrèè} \PS{adj.~} \GE{large and solid} 
\LX{dɔrekabakoma} \HM{} \PH{dɔ̀rɪ̀kábákʊ́má} \PS{n.~} \GE{one who incites hatred} \OP{dɔrekabakomamine} 
\LX{dori} \HM{} \PH{dòrì} \PS{v.~} \GE{remove bits of the top} \PD{doree, doro, doro, doribo, doraa, doribo } 
\LX{dori} \HM{} \PH{dórí} \PS{n.~} \GE{heap} \OP{doe} \TP{doree/dorree} 
\LX{dori} \HM{} \PH{dórí} \PS{n.~} \GE{joint (stalk of grass)} \OP{doe} \TP{doree/dorree} 
\LX{dɔsaŋkpare} \HM{} \PH{dɔ́sáŋkparɪ́} \PS{n.~} \GE{peels of dawadawa fruits} \TP{dɔsaŋkparɛɛ} \VA{dɔseŋkpare}
\LX{dɔsempeɛlɛ} \HM{} \PH{dɔ̀sɪ́mpɪ́ɛ́lɛ́} \PS{n.~} \GE{immature dawadawa fruits} \SG{dɔsempeɛloo} \OP{dɔsempeɛle} \TP{dɔsempeɛlɛɛ} 
\LX{dosoraa} \HM{} \PH{dòsáràà} \PS{n.~} \GE{young female bush pig} \OP{dosarre} \TP{ donsarɛɛ/donsarrɛɛ} 
\LX{dɔzɔŋe} \HM{} \PH{dɔ̀zɔ̂ŋ} \PS{n.~} \GE{daw-daw flour; pulp from daw-daw fruit} \OP{dɔzɔnne} \TP{dɔzɔnnɛɛ} \VA{dɔzɔŋa}
\LX{dɔzunni} \HM{} \PH{dɔ̀zúnnì} \PS{n.~} \GE{daw-daw seeds} \SG{dɔzunnuu} \OP{dɔzunni} \TP{dɔzunnee} 
\LX{dɔ᷉ɔ} \HM{} \PH{dɔ̀᷈ɔ̀} \PS{v.~} \SN \GE{to disturb, to move (rustle)} \SN \GE{to stir, to mix} \SN \GE{to worry} \PD{dɔ᷉ɔɛ, dɔɔnɔ, dɔɔnɔ, dɔɔnema, dɔɔnaa, dɔɔnema } 
\LX{dɔ᷉ɔ} \HM{} \PH{dɔ̀᷈ɔ̀} \PS{v.~} \GE{to sprain e.g. a leg} \PD{dɔ᷉ɔɛ, dɔɔnɔ, dɔɔnɔ, dɔɔnema, dɔɔnaa, dɔɔnema } 
\LX{du} \HM{} \PH{dú} \PS{v.~} \SN \GE{to rub against} \SN \GE{to iron (clothes)} \PD{due, duuro, duuro, duuribo, duuraa, duuribo } 
\LX{dudugiri} \HM{} \PH{dùdùgrí} \PS{n.~} \GE{a box made of grass} \OP{dudugo} \TP{dudugiree} 
\LX{dugi} \HM{} \PH{dúgí} \PS{n.~} \GE{swim} \VA{dúú}
\LX{dugi} \HM{} \PH{dùgì} \PS{v.~} \GE{to be eaten up to powder e.g. wood infected with bugs} \PD{dugee, dugiro, -, -, -, - } \VA{dúú}
\LX{dukuree} \HM{} \PH{dúkúréé} \PS{n.~} \GE{large sores caused by scabies} \TP{dukurree} \VA{dikuree}
\LX{duli} \HM{} \PH{dúlí} \PS{v.~} \SN \GE{to work strenuously on a difficult task} \SN \GE{to eat a large amount (fig.)} \PD{dulee, dulo, dulo, dulibo, dulaa, dulibo } 
\LX{duluŋ} \HM{} \PH{dùlûŋ} \PS{n.~} \GE{a type of large bird} \OP{ dulli} \TP{dullee} 
\LX{dunduli} \HM{} \PH{dùndùlí} \PS{n.~} \SN \GE{caterpillar} \SN \GE{worm} \SN \GE{maggot} \OP{dundulo} \TP{dundulee/dundullee} \VA{kpaŋkyori, kpaŋkyobo}
\LX{dundulivarekpɛnyoɔre} \HM{} \PH{dúndùlìváríkpɛ́ɲʊ́ɔ́rɪ́} \PS{n.~} \GE{small green caterpillar that springs from place to place} \OP{dundulivarekpɛnyɔɛ} 
\LX{dundulkyoo} \HM{} \PH{dùndùlʧóò} \PS{n.~} \GE{cocoon} \OP{dundulkyoori} \TP{dundulkyooree} 
\LX{dundulvarkpɛnyoɔre} \HM{} \PH{dùndùlvárkpɛ̀ɲʊ́ɔ́rɪ́} \PS{n.~} \GE{a type of small green caterpillar that moves by loops} \OP{dundulvarkpɛnyɔɛ} 
\LX{dunduŋkolo} \HM{} \PH{dùndǔŋkòlò} \PS{n.~} \GE{fried yam wedges} \SG{dunduŋkoluu} \TP{dunduŋkolee} 
\LX{duni} \HM{} \PH{dúni} \PS{n.~} \GE{knee} \OP{dumo} \TP{dunee/dunnee} 
\LX{duŋi} \HM{} \PH{dúŋ} \PS{n.~} \GE{type of weevil (found in wood )} \OP{dunni} \TP{dunee/dunnee} 
\LX{dunselee} \HM{} \PH{dùnséléé} \PS{n.~} \GE{kneecap} \OP{ dunsebilii} \TP{dunsebilee} 
\LX{duo} \HM{} \PH{dùó} \PS{n.~} \GE{wild pig, boar} \OP{dori} \TP{doree/dorree} 
\LX{duo} \HM{} \PH{dùò} \PS{v.~} \SN \GE{to be seated all over the place in great numbers (patients at consulting room)} \SN \GE{to prepare malt for brewing local beer;} \PD{duoe, duoro, duoro, duoribo, duoraa, duoribo } 
\LX{duogo} \HM{} \PH{dúógó} \PS{n.~} \GE{giant} \OP{duogori} \TP{duogoree} 
\LX{Duoŋ} \HM{} \PH{dúóŋ} \PS{n.~} \GE{town in the Upper West Region noted for treatment of bone fractures} 
\LX{duoraa} \HM{} \PH{dùóráa} \PS{n.~} \SN \GE{open space for a gathering;} \SN \GE{resting place for sheherds and their animals} \OP{duori} \TP{duoree} 
\LX{duori} \HM{} \PH{dúórì} \PS{v.~} \SN \GE{to urinate} \SN \GE{to ejaculate} \SN \GE{to set high up (child on shoulders)} \SN \GE{to raise yam mounds} \SN \GE{to set a bow ready for use} \PD{duoree, duoro, duoro, duoribo, duoraa, duoribo } 
\LX{duori} \HM{} \PH{dùòrì} \PS{v.~} \GE{to set out in heaps e.g. tomatoes for sale} \PD{duoree, duoro, duoro, duoribo, duoraa, duoribo } 
\LX{duoro} \HM{} \PH{dùòró} \PS{n.~} \GE{news} \TP{duoree} 
\LX{duor-tere} \HM{} \PH{dùòr-tɪ́rɪ́} \PS{n.~} \GE{annoucement, broadcast, information} \OP{duor-terre} \TP{duorterrɛɛ} 
\LX{duor-terɛ} \HM{} \PH{dùòr-tɪ́rɛ́} \PS{n.~} \GE{one who gives information, messenger, newscaster} \OP{duor-terebɛ} \TP{duor-terebɛɛ} 
\LX{duoruŋ} \HM{} \PH{dúórǔŋ} \PS{n.~} \GE{urine} 
\LX{duoruŋkɔnsɔgerɔ} \HM{} \PH{dúórǔŋkɔ́nsɔ́grɔ́} \PS{n.~} \GE{the condition of having difficulty passing urine} \OP{duoruŋkɔnsɔgre} \TP{duoruŋkɔnsɔgrɛɛ} 
\LX{duorunlaa} \HM{} \PH{dúórùnláá} \PS{n.~} \GE{chamber-pot} \OP{duorunlaare} \TP{duorunlaarɛɛ} 
\LX{duorunwoɔ} \HM{} \PH{dúórùnwʊ̀ɔ̀} \PS{n.~} \GE{bladder} \OP{duorunwoore} \TP{duorunwoorɛɛ} 
\LX{duorunze᷉e᷉} \HM{} \PH{dúórùnzɪ́᷉ɪ́᷉} \PS{n.~} \GE{bilharziasis} \TP{duorunzeenɛɛ} 
\LX{duulaa} \HM{} \PH{dùùláà} \PS{n.~} \SN \GE{row} \SN \GE{line (poetry)} \OP{duuli} 
\LX{dõo} \HM{} \PH{dʋ̂̀ʋ} \PS{adv.~} \GE{hazy, misty} 
\LX{e} \HM{} \PH{ɪ̀} \PS{v.~} \SN \GE{to act, to do} \XV{Te na e o la velaa.} \XE{We shall do him good.} \SN \GE{to let} \XV{E ka te gaa pampana.} \XE{Let us go now.} \SN \GE{to fetch} \XV{A dɔɔ e la a kyi ko a pɔge.} \XE{The man has fetched  millet for the woman.} \SN \GE{to be} \XV{A bie e la kpeɛŋaa.} \XE{The child is strong.} \PD{eɛɛ, erɛ, erɛ, erebɛ, eraa, erebɛ } 
\LX{ɛbaa} \HM{} \PH{ɛ́bàà} \PS{n.~} \GE{crocodile} \OP{ɛrre} \TP{ɛrrɛɛ} 
\LX{eɛ} \HM{} \PH{ìɛ} \PS{v.~} \SN \GE{to pinch} \SN \GE{to hint} \XV{Ba eɛ la a yɛlɛ fe᷉e᷉ ka N woŋ.} \XE{They have given me a hint of the situation.} \PD{eɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ } 
\LX{eɛ eŋe} \HM{} \PH{ɪ̀ɛ̀ ɪ̀ŋ} \PS{v.~} \GE{to flee, to run away} \XV{A bie zu pɔre la maŋgo ka a daana waana ka o eɛ eŋ.} \XE{The boy stole a mango and when he saw the owner coming he took to his heels.} \PD{eɛ eŋɛɛ, eɛrɛ ennɛ, eɛ ennɛ, eɛ enemɛ, eɛ ennaa, - } 
\LX{eele} \HM{} \PH{ɪ́ɪ́lɪ́} \PS{n.~} \SN \GE{horn} \SN \GE{antenna} \OP{eelɛ} \TP{eelɛɛ} 
\LX{eɛne} \HM{} \PH{ɪ̀ɛ̀nɪ̀} \PS{n.~} \GE{hippopotamus} \OP{eɛmɛ} \TP{eɛnɛɛ} 
\LX{eɛne} \HM{} \PH{ɪ́ɛ́nɪ́} \PS{n.~} \GE{claw} \SG{eɛnoo} \OP{eɛmɛ} \TP{eɛnɛɛ} 
\LX{eɛne} \HM{} \PH{ɪ̀ɛ̌nɪ́} \PS{n.~} \GE{a type of deer} \OP{eɛmɛ} \TP{eɛmɛɛr} 
\LX{eɛne} \HM{} \PH{ɪ̀ɛ́nɪ́} \PS{adj.~} \GE{tiny, e.g. fingernail} \OP{eɛmɛ} \TP{eɛmɛɛ} 
\LX{eɛraa} \HM{} \PH{ɪ́ɛ̀ráá} \PS{n.~} \GE{hook, e.g. fishhook} \OP{eɛre} \TP{eɛrɛɛ} 
\LX{eɛre} \HM{} \PH{ɪ́ɛ́rɪ̀} \PS{v.~} \GE{to wipe away e.g. dirt, tears} \PD{eɛrɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ } 
\LX{eɛre} \HM{} \PH{ɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to claw at} \PD{eɛrɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ } 
\LX{eɛre-miri} \HM{} \PH{ɪ̀ɛ̀rɪ̀-mírì} \PS{n.~} \GE{fishing line} \OP{eɛre-mie} \TP{eɛre-miiree} 
\LX{ɛɛ᷉} \HM{} \PH{ɛɛ᷉} \PS{use word.~} \GE{} \VA{o᷉o}
\LX{e-la} \HM{} \PH{ɪ̀-là} \PS{v.~} \GE{to tease} \PD{e-laɛ, erɛ laara, e-laara, e-laareba, e-laaraa, e-lareba } 
\LX{e-laara} \HM{} \PH{ɪ̀-láárá} \PS{n.~} \GE{one who teases} \OP{e-lareba} \TP{e-larebɛɛ} 
\LX{ɛl-ɛloo} \HM{} \PH{ɛ̀l-ɛ̀lʊ́ʊ́} \PS{interjection.~} \GE{it serves you well} 
\LX{ɛlemɛ} \HM{} \PH{ɛ̀lɪ̀mɛ́} \PS{n.~} \GE{shellfish} \OP{ɛlemɛre} \TP{ɛlemɛɛ} 
\LX{embaare} \HM{} \PH{imbaari} \PS{n.~} \GE{laziness} \TP{embarɛɛ} \VA{entuo}
\LX{emii} \HM{} \PH{ɪ́míì} \PS{n.~} \GE{act of desperation; impatience} \TP{emmiinee} 
\LX{emmaa} \HM{} \PH{ɪ̀máà} \PS{n.~} \GE{the condition of being in a good state} 
\LX{emmaaroŋ} \HM{} \PH{ɪ̀mmáárʊ́ŋ} \PS{n.~} \GE{health} \OP{emmaara} \TP{emmaarɛɛ} 
\LX{emmii} \HM{} \PH{ɪ̀mmìì} \PS{n.~} \GE{a body that is in a painful condition} \OP{emmiime} \TP{emmiinee} 
\LX{emmiiruŋ} \HM{} \PH{ɪ̀mmììrûŋ} \PS{n.~} \SN \GE{a body in pain} \SN \GE{stunted body} \TP{emmiiree} 
\LX{emmiiruŋ} \HM{} \PH{ɪ̀mmììrûŋ} \PS{adj.~} \GE{stunted} 
\LX{Emɔ} \HM{} \PH{ɪ́mɔ̀} \PS{n.~} \SN \GE{the name of a clan in Southern Dagao} \SN \GE{appellation of this group} \SG{emoɔloo} \OP{emoɔlɔ} \TP{emoɔlɛɛ} 
\LX{empoɔne} \HM{} \PH{ɪ́mpʊ́ɔ́nɪ́} \PS{n.~} \GE{ocean} \OP{mpoɔnɛɛ} \VA{po}
\LX{eŋ} \HM{} \PH{ɪ́ŋ} \PS{part.~} \GE{as} \XV{Maa eŋ koŋ gaa.} \XE{As for me I wont go.} 
\LX{eŋ noɔre} \HM{} \PH{ɪ̀ŋ nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to promise} \PD{-, -, -, -, -, - } 
\LX{eŋ poɔ} \HM{} \PH{ɪ̀ŋ pʊ́ɔ́} \PS{v.~} \GE{to empregnate} \PD{-, -, -, -, -, - } 
\LX{eŋ ze᷉ɛ᷉} \HM{} \PH{ɪ̀ŋ zɪ᷉̀ɛ᷉} \PS{v.~} \GE{to punish severely} \PD{eŋɛɛ ze᷉ɛ᷉, ennɛ ze᷉ɛ᷉, -, -, -, - } 
\LX{endaa} \HM{} \PH{ɪ̀ndàá} \PS{n.~} \GE{frame (of a person's body)} \OP{endaare} \TP{endarɛɛ/endaarɛɛ} 
\LX{endaa waaloŋ} \HM{} \PH{ɪ̌ndàá-wàálʊ̂ŋ} \PS{n.~} \GE{skin colour, appearance of the body} 
\LX{eŋe} \HM{} \PH{ɪ̀ŋ} \PS{v.~} \SN \GE{to put in} \XV{o ɔŋ la kôɔ eŋe a doge poɔ.} \XE{S/he has put water into the pot.} \SN \GE{to put on (shoes, ring, spectacles)} \SN \GE{to respect} \XV{A seŋe ka bibile maŋ eŋe neŋkpoŋ.} \XE{The young should respect the elderly.} \PD{eŋɛɛ, ennɛ, ennɛ, enemɛ, ennaa, - } 
\LX{eŋɛ} \HM{} \PH{ɪ́ŋɛ́} \PS{pron.~} \GE{what, who} \XV{Boŋ eŋɛ ka fo wa?} \XE{What have you come for?.} 
\LX{eŋɛ} \HM{} \PH{ɪ́ŋɛ́} \PS{n.~} \GE{body (surface of)} \OP{enne} \TP{ennɛɛ} 
\LX{eŋe daa} \HM{} \PH{ɪ̀ŋ dàà} \PS{v.~} \GE{to set a date} \PD{-, -, -, -, -, - } 
\LX{eŋe daa} \HM{} \PH{ɪ̀ŋ dàà} \PS{v.~} \GE{to beat} \PD{-, -, -, -, -, - } 
\LX{eŋene} \HM{1} \PH{ɪ́ŋɪ́nɪ́} \PS{n.~} \GE{main hole leading to a rodent's living room in its burrow} \XV{A teŋɛ deme zaa taa la eŋene kora a naa.} \XE{All the people in town have respect for the chief.} \OP{enne} \TP{eŋenɛɛ} 
\LX{eŋene} \HM{2} \PH{ɪ́ŋɪ́nɪ́} \PS{n.~} \GE{respect, regard for, honour} \XV{N teɛre ka o deɛ la N eŋene.} \XE{I think he has just targeted me.} \TP{ennɛɛ} \VA{enne}
\LX{enfuomaala} \HM{} \PH{ɪ̀nfùòmáálá} \PS{n.~} \GE{one who makes pictures} \OP{enfuomaaleba} \TP{enfuomaalebɛɛ} 
\LX{enfuomaale} \HM{} \PH{ɪ̀nfùòmàálɪ̀} \PS{n.~} \GE{the act of making a picture (draw)} 
\LX{enfuoni} \HM{} \PH{ɪ̀nfúóní} \PS{n.~} \GE{image, photograph, picture} \OP{enfuomo} \TP{enfuonee} 
\LX{enfuori} \HM{} \PH{ɪ̀nfùòrì} \PS{n.~} \GE{the act of bewitching; the act of casting a spell} 
\LX{eŋgane} \HM{} \PH{ɪ̀ŋgánɪ̀} \PS{n.~} \GE{skin (human body)} \SG{eŋganoo} \OP{eŋgama} \TP{eŋgannɛɛ} 
\LX{Eŋgelɛ} \HM{} \PH{ɪ́ŋgɪ̀lɛ̀} \PS{n.~} \GE{England} \TP{eŋgelɛɛ} 
\LX{Eŋgilisi} \HM{} \PH{ɪ́ŋgìlìsì} \PS{n.~} \GE{English} \OP{ɪŋlisiri} \TP{ɪŋgilisiree} 
\LX{eŋkɔbɔ} \HM{} \PH{ɪ̀ŋkɔ́bɔ́} \PS{n.~} \GE{skeleton} \VA{eŋkɔba}
\LX{eŋkpaŋkpoli} \HM{} \PH{ɪ́ŋkpáŋkpòlì} \PS{n.~} \GE{naked body} \VA{eŋkpoŋkpoli, eŋkpoli}
\LX{eŋkpoli} \HM{} \PH{iŋkpoli} \PS{n.~} \GE{naked body} \OP{eŋkpolo} \TP{eŋkpolee} \VA{eŋkpoŋkpoli, eŋkpaŋkpoli}
\LX{eŋkpoŋkpoli} \HM{} \PH{ɪ̀ŋkpóŋkpólì} \PS{n.~} \GE{naked} \OP{eŋkpoŋkpolo} \TP{eŋkpoŋkpolee} \VA{eŋkpoli, eŋkpaŋkpoli}
\LX{enne} \HM{} \PH{inni} \PS{n.~} \GE{honour} \TP{ennɛɛ} \VA{eŋene}
\LX{ɛnne} \HM{} \PH{ɛ̀nnɪ̀} \PS{v.~} \SN \GE{to suckle, to nurse} \SN \GE{to suck e.g. breast or finger} \PD{ɛnnɛɛ, ɛnnɛ, ɛnnɛ, ɛnemɛ, ɛnnaa, ɛnemɛ } 
\LX{ennoɔ} \HM{} \PH{ɪ̀nnʊ̀ɔ́} \PS{n.~} \GE{happiness} \OP{ennoore} \TP{ennoorɛɛ} 
\LX{ensɔgelaa} \HM{} \PH{ɪ́nsɔ́gláá} \PS{n.~} \GE{black skin} \OP{ensɔglɔ} 
\LX{entaa} \HM{} \PH{ɪ̀ntáá} \PS{n.~} \GE{mutual respect} 
\LX{entage} \HM{} \PH{ɪ̌ntágɪ́} \PS{n.~} \GE{muscular body} \OP{entagre} \TP{entagrɛɛ} 
\LX{entole} \HM{} \PH{ɪ́ntʊ́lɪ̀} \PS{n.~} \GE{fever, rise in body temperature} \OP{entolɔ/entola} \TP{entolɛɛ} 
\LX{entoloŋ} \HM{} \PH{ɪ́ntʊ́lʊ̂ŋ} \PS{n.~} \GE{fever} \OP{entola} \TP{entollɛɛ} 
\LX{entukpalaa} \HM{} \PH{ɪ̂ntùkpàláà} \PS{n.~} \GE{extreme laziness} \OP{entukpalle} \TP{entukpalɛɛ} 
\LX{entuo} \HM{} \PH{ɪ̀ntúó} \PS{n.~} \GE{laziness} \OP{entuuri} \TP{entuuree} \VA{embaare}
\LX{eŋyoɔre} \HM{} \PH{ɪ̀ŋyʊ̀ɔ̀rɪ̀} \PS{adj.~} \GE{that which does not produce excitement} \VA{enyoɔre}
\LX{enzeɛ} \HM{} \PH{ǐnzɪ̀ɛ́} \PS{n.~} \SN \GE{light skin} \SN \GE{leprosy} \OP{enzeere} \TP{enzeerɛɛ} 
\LX{ɛre} \HM{} \PH{ɛ́rɪ́} \PS{v.~} \GE{to crawl; to slide along on the buttocks} \PD{ɛrɛɛ, ɛrɛ, ɛrɛ, ɛrebɛ, ɛraa, ɛrebɛ } 
\LX{ɛre} \HM{} \PH{ɛ́rɪ̀} \PS{v.~} \GE{to grind} \PD{ɛrɛɛ, ɛrɛ, ɛrɛ, ɛrebɛ, ɛraa, ɛrebɛ } 
\LX{ɛre} \HM{} \PH{ɛ̀rɪ̀} \PS{v.~} \GE{to fill a hole or space tightly to disallow access} \PD{ɛrɛɛ, ɛrɛ, ɛrɛ, ɛrebɛ, ɛraa, ɛre-ɛrɛ } 
\LX{ɛre-eŋ} \HM{} \PH{ɛ̀rɪ̀-ɪ̀ŋ} \PS{n.~} \SN \GE{exaggeration} \SN \GE{metaphor (lit.)} 
\LX{ɛregbɛrɛ} \HM{} \PH{ɛ̀rɪ̀gbɛ̀rɛ̀} \PS{adv.~} \GE{fleshy and fatty (meat)} 
\LX{eroŋ} \HM{} \PH{ɪ̀rʊ̂ŋ} \PS{n.~} \GE{conduct} \OP{erre} \TP{errɛɛ} 
\LX{eroŋ} \HM{} \PH{ɪ̀rʊ̂ŋ} \PS{n.~} \GE{verb} \OP{erre} \TP{errɛɛ} 
\LX{etɔ} \HM{} \PH{ɪ́tɔ̀} \PS{n.~} \GE{the name of a clan in Southern Dagao} \OP{etoɔlɔ} \TP{etoɔlɛɛ} 
\LX{Etore} \HM{} \PH{ítʊ́rɪ̀} \PS{n.~} \GE{Easter} \OP{itorɛɛ} 
\LX{ewa} \HM{} \PH{ɪ́wà} \PS{n.~} \SN \GE{a kinship group} \SN \GE{appellation of this group} \SG{ewaaloo} \OP{ewaala} \TP{ewaalɛɛ} 
\LX{ewɛ} \HM{} \PH{ɪ́wɛ̀} \PS{n.~} \GE{the name of a clan in Southern Dagao} \OP{eweɛlɛ} \TP{eweɛlɛɛ} 
\LX{e᷉ɛ} \HM{} \PH{ɪ́᷉ɛ́᷉} \PS{n.~} \GE{} \SG{eɛnoo} \OP{e᷉ɛ} \TP{enɛɛ} \VA{eɛne}
\LX{e᷈ɛ} \HM{} \PH{ɪ́᷉ɛ̀᷉} \PS{v.~} \SN \GE{to mark the foundation of a building} \SN \GE{to cut a very small piece} \PD{eɛɛ, eɛnɛ, eɛnɛ, eɛnemɛ, eɛnaa, eɛnemɛ } 
\LX{e᷉ɛo} \HM{} \PH{ɪ̀᷉ɛ̀᷉ʊ́᷉} \PS{n.~} \GE{type of deer} \OP{eɛne} \TP{eɛnɛɛ} 
\LX{eɛne} \HM{} \PH{ɪ̀ɛ̀ɪ̀} \PS{v.~} \GE{to claw e.g a cat climbing a wall} \PD{-, -, -, -, -, - } 
\LX{fa} \HM{} \PH{fà} \PS{v.~} \SN \GE{to be bad} \XV{Ka fooŋ ba kyɛlle a bie ŋa na fa la, bonso ka o ba maŋ kpɛ die wi᷉eu?} \XE{If you are not careful, this child will go bad, why doesn't s/he come early?.} \SN \GE{to become poor} \XV{A bie baaloŋ vɛŋɛɛ la ka o tuuli e te᷉e lɛɛ deɛ te fa o zaa.} \XE{The child's sickness has caused her to search so much for a cure that she has nothing left.} \PD{faɛ, faara, faaraa, fareba, -, fareba } 
\LX{fa} \HM{} \PH{fá} \PS{v.~} \SN \GE{to rob, to cheat} \XV{N dɔɔ, fo baŋ ka ka maaŋ yɔɔ libie zaa poɔ a anaŋ ka fo fa ma la eŋ?} \XE{My friend, do you know that if I paid anything more, you would have robbed me?.} \SN \GE{pl. of 'faa' (snatch from someone)} \XV{A seŋ ka ba pɔge a nannyige eŋ deŋɛ lɛ, N ba nyɛ teŋɛ na deme o naŋ ba fa.} \XE{It will be good if this thief is imprisoned for life, because there is no town he has not robbed.} \PD{faɛ, faara, faara, fareba, faaraa, fareba } 
\LX{faa} \HM{} \PH{fáá} \PS{adj.~} \SN \GE{bad} \XV{Saa naŋ piili miibu ŋaa, beri ayi ŋa zaa ka te teŋɛ sori ya᷉a na waa faa lɛ.} \XE{With the setting in of the rains, the road leading to our village will soon be very bad.} \SN \GE{ugly} \XV{A pɔge veɛlɛ la kyɛ o serɛ faa.} \XE{The woman is beautiful but her husband is ugly.} \OP{faare} \TP{faarɛɛ} 
\LX{faa} \HM{} \PH{fáà} \PS{v.~} \SN \GE{to take by force, to snatch} \XV{A bie ŋa yɛlɛ kpeɛŋaa, ka onaŋ sɔre o tɔ bone ka o mɔŋ o maŋ faa la.} \XE{This boy is ruthless, for if he begs from his colleague and is refused to give he takes the thing by force.} \SN \GE{to save} \XV{Kɔŋ da kpɛ la a bikpeebɛ ka o kono kyɛ ka pɔge-nimbaaloŋ kaŋ la da faa o ne bondirii.} \XE{The orphan was hungry and began crying, but a benevolent woman came to his aid with food.} \SN \GE{to get free, to get out of one's grip or hold} \XV{N nyɔge la a kpa᷉ao kyɛ ka o faa.} \XE{I caught the guinea fowl but it slipped free and escaped.} \PD{faaɛ, faara, faara, fareba, faaraa, faareba } 
\LX{faa le} \HM{} \PH{fáà lè} \PS{v.~} \GE{to drop, to fall} \PD{faa lee, faara lere, faa lere, faa leribe, faa leraa, faa leribe } 
\LX{Faaloŋ} \HM{} \PH{fààlʊ̂ŋ} \PS{n.~} \GE{the name of a village located near Lawra, a District headquarters in the UWR of Ghana} 
\LX{faaloŋ} \HM{} \PH{fààlʊ̂ŋ} \PS{n.~} \GE{the condition of being bad or wicked} 
\LX{faara} \HM{} \PH{fáárà} \PS{n.~} \GE{priest} \OP{faaramine} \TP{faarɛɛ} 
\LX{faara} \HM{} \PH{fáárá} \PS{n.~} \GE{saviour} \OP{faareba} 
\LX{faara} \HM{} \PH{fáárá} \PS{n.~} \GE{one who robs} \OP{faareba} 
\LX{faare} \HM{} \PH{fáárɪ̀} \PS{v.~} \GE{to intercede, to mediate (in a fight)} \PD{faarɛɛ, faara, faara, faareba, faaraa, faareba } 
\LX{faare} \HM{} \PH{fààrɪ̀} \PS{v.~} \GE{to stroke, to massage} \XV{Kɔlaa maŋ nɔŋ ka ba faara o tegitegi lɛ.} \XE{A cat likes to be stroded all the time.} \PD{faarɛɛ, faara, faara, faareba, faaraa, faareba } 
\LX{faare} \HM{} \PH{fààrɪ̀} \PS{n.~} \GE{a boys' game of snatching others' arrows} \TP{faarɛɛ} 
\LX{faaregyɔ} \HM{} \PH{fààrɪ́ɲɔ́} \PS{v.~} \GE{to not utter a word; to be quiet} \PD{-, -, -, -, -, - } 
\LX{fabo} \HM{} \PH{fàbʊ̀} \PS{adv.~} \GE{not responding to anything; speechless} 
\LX{fabo} \HM{} \PH{fàbʊ̀} \PS{adv.~} \GE{tasteless} 
\LX{faee} \HM{} \PH{fàɪ̀ɪ̀} \PS{adv.~} \GE{empty-looking} 
\LX{fafara} \HM{} \PH{fàfàrà} \PS{adv.~} \GE{suddenly, without permission (of entering)} 
\LX{fage} \HM{} \PH{fàgɪ̀} \PS{adv.~} \GE{plenty (and jumbled up), e.g. of different things} \XV{Naŋkoɔre boma da pale a daa lɛ fage.} \XE{There were many assorted goods in the market.} 
\LX{fage} \HM{} \PH{fágɪ́} \PS{v.~} \GE{to peck, to eat (of birds)} \PD{fagɛɛ, fagera, fagra, fagreba, fageraa, fagreba } 
\LX{fagere} \HM{} \PH{fàgrɪ̀} \PS{adv.~} \GE{sharp (of a bite)} 
\LX{fagoo} \HM{} \PH{fàgʊ̀ʊ̀} \PS{adv.~} \GE{full of rubbish, dirty} 
\LX{falala} \HM{} \PH{fàlàlà} \PS{adv.~} \GE{long, smooth, and straight} \VA{falalla}
\LX{falala} \HM{} \PH{fálálá} \PS{adj.~} \GE{smooth and straight} \VA{falalla}
\LX{falalla} \HM{} \PH{fálállá} \PS{adv.~} \GE{} \VA{falala}
\LX{falanfuuri} \HM{} \PH{fálánfúúrí} \PS{n.~} \GE{lungs} \OP{falanfugo} \TP{falanfugree} 
\LX{fale} \HM{} \PH{fálɪ̀} \PS{v.~} \GE{to slap} \PD{falɛɛ, falla, falla, faleba, falaa, faleba } 
\LX{fale pareŋ} \HM{} \PH{fálɪ̀ párɪ̀ŋ} \PS{v.~} \GE{to spank (a child)} \PD{-, -, -, -, -, - } 
\LX{fambataabo} \HM{} \PH{fǎmbátáábʊ́} \PS{n.~} \GE{weakness, powerlessness} 
\LX{faŋa} \HM{} \PH{fàŋá} \PS{n.~} \GE{strength; be strong (physically)} \OP{fanne} \TP{fannɛɛ} 
\LX{faŋa} \HM{} \PH{fáŋá} \PS{adj.~} \GE{young; used for calves} \TP{fanne} 
\LX{fananna} \HM{} \PH{fánánná} \PS{adv.~} \GE{very clean} 
\LX{fane} \HM{} \PH{fànɪ̀} \PS{v.~} \GE{to become rough (skin)} \PD{fanɛɛ, fana, fana, fanema, fanaa, fanema } 
\LX{fane} \HM{} \PH{fànɪ́} \PS{adj.~} \GE{rough (skin)} \OP{fama} 
\LX{fanfane} \HM{} \PH{fànfánɪ̀} \PS{n.~} \GE{soap} \OP{fanfama} \TP{fanfanɛɛ/fanfannɛɛ} 
\LX{fanne} \HM{} \PH{fànnɪ̀} \PS{v.~} \GE{to sell one's wares at a low price"?} \PD{fannɛɛ, fanna, fanna, fanema, fannaa, fanema } 
\LX{fanne} \HM{} \PH{fánnɪ̀} \PS{v.~} \GE{to think} \PD{fannɛɛ, fanna, fanna, fanema, fannaa, - } 
\LX{Farafara} \HM{} \PH{fàràfàrà} \PS{n.~} \SN \GE{a person of the Gurunɛ ethnicity} \SN \GE{the language spoken by this group} \SG{farafaroo} \OP{farafarɛɛ} 
\LX{farata} \HM{} \PH{fàràtà} \PS{} \GE{tall and straight} 
\LX{fare} \HM{} \PH{fárɪ́} \PS{v.~} \SN \GE{to foam (yeast in alcohol; discharge from eyes)} \SN \GE{to boil (porridge)} \PD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\LX{fare} \HM{} \PH{fárɪ́} \PS{v.~} \GE{to build mounds around crops (farming corn)} \PD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\LX{fare} \HM{} \PH{fárɪ́} \PS{v.~} \GE{to pluck leaves in large quantities} \PD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\LX{fare} \HM{} \PH{fárɪ̀} \PS{v.~} \GE{to make bare; (to shave hair off completely)} \PD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\LX{fare} \HM{} \PH{fárɪ́} \PS{adv.~} \GE{completely} \XV{A bibiiri zaa zo kulee la fare.} \XE{All the children have gone without any left behind.} 
\LX{fare bini} \HM{} \PH{fárɪ́ bíní} \PS{v.~} \SN \GE{shit plenty} \SN \GE{say digusting things} \PD{-, -, -, -, -, - } 
\LX{fare yɛlɛ} \HM{} \PH{fárɪ́ yɛ́lɛ́} \PS{v.~} \SN \GE{to talk a great deal} \SN \GE{to complain} \PD{-, -, -, -, -, - } 
\LX{fasane} \HM{} \PH{fàsánɪ̀} \PS{n.~} \GE{(with ̀ŋmɛ') thought} \XV{A dɔɔ da ŋmɛ la fasane, a wa te yeli ka o ba la boɔrɔ a o pɔge.} \XE{The man thought for a while and finally said he did not want his wife.} \OP{fasama} \TP{fasamɛɛ/fasannɛɛ} 
\LX{fa᷉a} \HM{} \PH{fà᷉̀à̀᷉} \PS{adv.~} \GE{slowly (of a fall)} \XV{O kyeelɛɛ la gɛrɛ fa᷉a na te le.} \XE{S/he is falling slowly.} 
\LX{fa᷉a} \HM{} \PH{fà᷉à᷉} \PS{v.~} \SN \GE{to be unable} \XV{O gaa la o sori na fa᷉a.} \XE{S/he has failed to go on her trip.} \SN \GE{to have no choice} \XV{A pɔmpeŋ sa᷉aɛ la ka noba fa᷉a kyɛ sigi baa te ɔŋ ko᷉ɔ.} \XE{The water tap is spoilt, so people have no choice but to go the stream to fetch water.} \PD{faa᷉ɛ, faana, faana, faanema, faanaa, faa᷉faana } 
\LX{fa᷉afa᷉a} \HM{} \PH{fa᷉́a᷉́fa᷉́a᷉́} \PS{n.~} \GE{whistle used in times of war or danger} \OP{fa᷉afaane} \TP{fa᷉afaanɛɛ} 
\LX{febo} \HM{} \PH{fɪ́bʊ́} \PS{adv.~} \GE{piercingly; the manner a sharp object enters another suddenly and firmly (a thorn in the body)} 
\LX{fee} \HM{} \PH{féé} \PS{adv.~} \GE{completely (of drinking)} 
\LX{feɛ} \HM{} \PH{fɪ̀ɛ̀} \PS{v.~} \SN \GE{to whip} \SN \GE{to unshell e.g. rice or corn} \PD{feɛɛ, feɛrɛ, feɛrɛ, feɛrebɛ, feɛraa, feɛrebɛ } 
\LX{fɛɛ} \HM{} \PH{fɛ́ɛ́} \PS{adv.~} \SN \GE{happening in some manner e.g. completely} \XV{A baala nyu la a tẽẽ a de goɔoŋ e fɛɛ.} \XE{The patient took his medication and slept soundly.} \SN \GE{becoming airless instantly (tyre)} \XV{Pimbiri kyɔge la a kuriwiri kɔbɔ ka o e fɛɛ, o naŋ yi.} \XE{The tyre of the bicycle was pierced by a nail and deflated at once.} 
\LX{fɛɛ} \HM{} \PH{fɛ̀ɛ̀} \PS{adv.~} \GE{deflating completely of quiet sound of pent up air or gas escaping} 
\LX{feelaa} \HM{} \PH{fɪ́ɪ́láà} \PS{adj.~} \GE{young and tender, at the very infant age e.g. chicken, kid} \XV{A nɔnyaŋaa neɛ la nɔbil feelaa kaŋa ko.} \XE{The mother hen has stepped on one of the very young chicks and killed it.} \OP{feelɛ} 
\LX{feele} \HM{} \PH{fɪ̀ɪ̀lɪ̀} \PS{v.~} \GE{to make tight} \PD{feelɛɛ, feelɛ, feelɛ, feelbɛ, feelaa, feelbɛ } 
\LX{feele} \HM{} \PH{fɪ́ɪ́lɪ̀} \PS{v.~} \GE{to examine} \PD{feelɛɛ, feelɛ, feelɛ, feelbɛ, feelaa, feelbɛ } 
\LX{feele} \HM{} \PH{fɪ́ɪ́lɪ́} \PS{n.~} \GE{tsetse fly} \OP{feelɛ} \TP{feelɛɛ} 
\LX{feɛle} \HM{} \PH{fɪ̀ɛ̀lɪ̀} \PS{v.~} \SN \GE{to pass closely by another e.g. a blow missing the target narrowly} \SN \GE{to pass gas silently} \PD{feɛlɛɛ, feɛlɛ, feɛlɛ, feɛlbɛ, feɛlaa, feɛlbɛ } 
\LX{feɛle} \HM{} \PH{fɪ́ɛ́lɪ́} \PS{v.~} \GE{to sneak out} \PD{feɛlɛɛ, feɛlɛ, feɛlɛ, feɛlbɛ, feɛlaa, feɛlbɛ } 
\LX{feelee} \HM{} \PH{fɪ́ɪ́lɪ́ɪ́} \PS{n.~} \GE{chance, space to operate} \XV{Feelee kye a die poɔ, fo koŋ nyɛ zie biŋi fo sɛŋɛ.} \XE{There is no space in the room; you can't find space for your bed.} \OP{feelɛ} \TP{feelɛɛ} 
\LX{feeloo} \HM{} \PH{fɪ́ɪ́lʊ́ʊ́} \PS{adj.~} \GE{narrow, tight} \OP{feelɛ} 
\LX{feere} \HM{} \PH{fɪ̀ɪ̀rɪ̀} \PS{v.~} \SN \GE{to blow one's nose} \SN \GE{to hold, press, or tie tightly} \PD{feerɛɛ, feerɛ, feerɛ, feerebɛ, feeraa, - } 
\LX{feerɔɔ} \HM{} \PH{fɪ̀ɪ̀rɔ̀ɔ̀} \PS{adv.~} \GE{suddenly disappearing into an opening e.g. into a hole or room} \XV{Bayuo to la o beɛre kyɛ zo gaa e die feerɔɔ, o naŋ kpɛ.} \XE{Bayuo insulted his elder brother and ran into the house at once.} 
\LX{fefeɛ} \HM{} \PH{fɪ́fɪ̀ɛ́} \PS{n.~} \GE{crevice} \OP{fefeere} \TP{fefeerɛɛ} 
\LX{feferɛ} \HM{} \PH{fɪ́fɪ́rɛ́} \PS{adj.~} \GE{least} \XV{Ka fooŋ ba maale fo kyi ŋa biŋi,  fo yeli ka fo erɛ feferɛ ka noore wa la eŋɛ.} \XE{If you don't keep your millet well the least move away and fowls swoop on it.} \VA{fifire}
\LX{fege} \HM{} \PH{fɪ̀gɪ̀} \PS{v.~} \GE{to cut out a tiny opening} \XV{A daa kpeɛŋaa, a pɛnnepɛnɛ ba toɔ na fege o zie zaa.} \XE{The wood is so hard that the carver can hardly make any headway to carve out even a small hole.} \PD{fegɛɛ, fegrɛ, -, fegrebɛ, fegraa, fegrebɛ } 
\LX{fɛge} \HM{} \PH{fɛ̀gɪ̀} \PS{v.~} \SN \GE{to indent} \SN \GE{to become one-eyed} \PD{fɛgɛɛ, fɛgrɛ, fɛgrɛ, fɛgrebɛ, fɛgraa, fɛgfɛgrɛ } 
\LX{fɛge} \HM{} \PH{fɛ́gɪ́} \PS{n.~} \GE{one-eyed person} \OP{fɛgre} \TP{fɛgrɛɛ} 
\LX{fɛge} \HM{} \PH{fɛ́gɪ́} \PS{adv.~} \GE{fall lightly, such that you don't hear anything e.g. dry leaf} 
\LX{fegɛfegɛ} \HM{} \PH{fɪ̀gɛ̀fɪ̀gɛ̀} \PS{adv.~} \GE{flashing intermittently e.g. lightening} \VA{felɛfelɛ}
\LX{fegele} \HM{} \PH{fɪ̀gɪ̀lɪ̀} \PS{v.~} \GE{to be tight} \PD{-, feglɛ, -, -, -, - } \VA{fogele}
\LX{fɛgelɛgɛ} \HM{} \PH{fɛgelɛgɛ} \PS{adv.~} \GE{out of energy} \VA{fɛgetɛgɛ}
\LX{fegere} \HM{} \PH{fɪ́gɪ́rɪ̀} \PS{v.~} \GE{sob} \PD{fegrɛɛ, fegrɛ, fegrɛ, fegrebɛ, fegraa, fegrebɛ } 
\LX{fɛgetɛgɛ} \HM{} \PH{fɛ́gtɛ́gɛ́} \PS{adv.~} \GE{completely out of air or energy} 
\LX{fele} \HM{} \PH{fɪ̀lɪ́} \PS{n.~} \GE{sore, wound} \OP{felɛ} \TP{felɛɛ} 
\LX{felɛfelɛ} \HM{} \PH{fɪ̀lɛ̀fɪ̀lɛ̀} \PS{adv.~} \GE{flash brightly} 
\LX{fɛlefɛle} \HM{} \PH{fɛ́lfɛ́l} \PS{adj.~} \GE{tasteless} 
\LX{fɛlɛlɛ} \HM{} \PH{fɛ́lɛ́lɛ́} \PS{adv.~} \GE{utterly, completely} \XV{A dɔɔ ku᷉u maa la maa eŋ noɔre fɛlɛlɛ, ŋmenaa ŋa ka maa ne o puori taa daa poɔ} \XE{The man's death has utterly surprised me, because we greeted each other this afternoon in the marketplace.} 
\LX{felko᷉ɔ} \HM{} \PH{fɪ̀lkʊ̀᷉ɔ́᷉} \PS{n.~} \GE{pus} \TP{felkoɔnɛɛ} 
\LX{felmaara} \HM{} \PH{filmaara} \PS{n.~} \GE{sores over the body thought to be incurable e.g leprosy} \VA{felmaaroŋ}
\LX{felmaaroŋ} \HM{} \PH{fɪ̀lmáárʊ́ŋ} \PS{n.~} \GE{sore caused by leprosy or yaws} \OP{felmaara} \TP{felmaarɛɛ} \VA{felmaara}
\LX{felɔɔ} \HM{} \PH{fɪ̀lɔ̀ɔ̀} \PS{adv.~} \GE{of a sparkle} 
\LX{felpɛmpɛle} \HM{} \PH{fɪ̀lpɛ̌mpɛ́lɪ́} \PS{n.~} \GE{scar} \SG{felpɛmpɛloo} \OP{felpɛmpɛlɛ} \TP{felpɛmpɛlɛɛ/felpɛmpɛllɛɛ} 
\LX{felzeɛ} \HM{} \PH{fɪ̀lzɪ̀ɛ̀} \PS{n.~} \GE{sore without pus} \OP{felzeere} \TP{felzeerɛɛ} 
\LX{felɛfelɛ} \HM{} \PH{fɪ́lɛ́fɪ́lɛ́} \PS{adv.~} \GE{glittery and fashing with light} 
\LX{femm} \HM{} \PH{fɪ́mm} \PS{adv.~} \GE{getting away fast and far e.g. bullet} 
\LX{fɛmm} \HM{} \PH{fɛ̀mm} \PS{adv.~} \GE{lying flat e.g. vegetation or hair} 
\LX{fenlege} \HM{} \PH{fɪ́nlɪ́gɪ́} \PS{adv.~} \GE{laboured e.g. breathing} 
\LX{fenne} \HM{} \PH{fɪ̀nnɪ̀} \PS{v.~} \GE{to begin to ripen (first few fruits)} \PD{ fennɛɛ, fennɛ, fennɛ, fenemɛ, fennaa, - } 
\LX{fenne} \HM{} \PH{fɪ́nnɪ́} \PS{v.~} \GE{to cut a very small piece} \PD{-, -, -, -, -, - } 
\LX{fentelɛ} \HM{} \PH{fɪ̀ntɪ́lɛ̀} \PS{n.~} \GE{lamp} \OP{fentelle} \TP{fentelɛɛ/fentellɛɛ} 
\LX{fentellaa} \HM{} \PH{fɪ́ntɪ́lláá} \PS{n.~} \GE{local lamp} \OP{fentellaare} \TP{fentellaarɛɛ} 
\LX{fentere} \HM{} \PH{fɪ́ntɪ́rɪ́} \PS{n.~} \GE{ring e.g. that put on the finger} \OP{fenteɛ} \TP{fenterɛɛ/fenterrɛɛ} 
\LX{fentere} \HM{} \PH{fɪ́ntɪ́rɪ́} \PS{n.~} \GE{the act of standing at attention} \OP{fenteɛ} \TP{fenterɛɛ/fenterrɛɛ} 
\LX{fɛraa} \HM{} \PH{fɛ́ráá} \PS{adj.~} \GE{narrow, tight} \OP{fɛrre} 
\LX{fere} \HM{} \PH{fɪ̀rɪ̀} \PS{v.~} \SN \GE{to force, to pester} \SN \GE{to be morally obliged to do something} \XV{A fere ka te sonnɔ nembaalba.} \XE{We are obliged to care for the helpless.} \PD{ferɛɛ, ferɛ, ferɛ, ferebɛ, feraa, - } \VA{fɛre}
\LX{fere} \HM{} \PH{fèrè} \PS{n.~} \GE{regret; bitter experience} \XV{Ka fooŋ wa deɛnɛ ne a badɔgrɔ bile fo na wa yeli ka fere.} \XE{If you play around that newborn puppy you will come to regret it.} \TP{feree/ferree} 
\LX{fɛre} \HM{} \PH{fɛ̀rɪ̀} \PS{v.~} \GE{to be sqeezed up in between} \PD{fɛrɛɛ, fɛrɛ, fɛrɛ, fɛrbɛ, fɛraa, fɛrbɛ } 
\LX{fɛre} \HM{} \PH{fɛ́rɪ́} \PS{adv.~} \GE{sound of drop of something small and light} 
\LX{feree} \HM{} \PH{féréé} \PS{n.~} \GE{the fruit of the w??kamp?re plant} \VA{wɔŋkampɔre}
\LX{fereko} \HM{} \PH{fɪ́rɪ́kʊ́} \PS{n.~} \GE{a type of bawaa dance performed by moving to and fro to the sides (left and right)} \OP{ferekore} \TP{ferekorɛɛ} \VA{feroo}
\LX{fɛre-taa-kyɛ-moorɔ} \HM{} \PH{fɛ́rɪ̀tááʧɛ́mʊ́ʊ́rɔ́} \PS{n.~} \GE{one who has something but lies to others that one does not have it} \VA{fɛre-bin-kyɛ-moorɔ/fɛre-taa-kyɛ-moora}
\LX{feroo} \HM{} \PH{fɪ́rʊ́ʊ́} \PS{n.~} \GE{force} 
\LX{feroo} \HM{} \PH{fir??} \PS{n.~} \GE{} \VA{fereko}
\LX{feroo toma} \HM{} \PH{fɪ́rʊ́ʊ́ tʊ́má} \PS{n.~} \GE{duty, obligation} 
\LX{fe᷉e} \HM{} \PH{fɪ᷉́ɪ᷉́} \PS{adv.~} \SN \GE{for a short time; in a short time} \SN \GE{few} \SN \GE{small, little} 
\LX{fe᷉ɛ} \HM{} \PH{fɪ́᷉ɛ́᷉} \PS{v.~} \GE{to be grown to some extent; to not be small} \XV{A da ba seŋ ka o la taa poɔ pampana, a bie na o naŋ taa naŋ ba fe᷉ɛ.} \XE{She should not be pregnant now because the baby she has is still too young.} \PD{fe᷉ɛɛ, feɛnɛ, feɛnɛ, feɛnemɛ, feɛnaa, - } 
\LX{fieli} \HM{} \PH{fíélí} \PS{adj.~} \GE{slim and flat (nape)} 
\LX{fieli} \HM{} \PH{fíélì} \PS{v.~} \GE{to fail} \PD{fielee, fiele, fiele, fielbe, fielaa, fielbe } 
\LX{fieluu} \HM{} \PH{fíélúú} \PS{n.~} \GE{the act of failing} 
\LX{fifire} \HM{} \PH{fífírè} \PS{adv.~} \GE{scarcely out of side than something happens} \XV{A dɔɔ yeli ka o erɛ fifire ka a kɔlaa de a nɛne.} \XE{No sooner had the man left than the cat snatched the meat.} \VA{feferɛ}
\LX{figi} \HM{} \PH{fígí} \PS{v.~} \GE{to throw out one's hand to signal to stop} \PD{figee, figire, figire, figiribe, figiraa, figiribe } 
\LX{figiraa} \HM{} \PH{fígràá} \PS{n.~} \GE{fan} \OP{figiri} \TP{figiree} 
\LX{figiri} \HM{} \PH{fígrí} \PS{v.~} \SN \GE{to fan} \SN \GE{to flap one's wings} \PD{figee, figire, figire, figiribe, figiraa, figiribe } 
\LX{fii} \HM{} \PH{fìì} \PS{adv.~} \GE{manner of movement e.g. of a heavy object in the air} 
\LX{filoo} \HM{} \PH{fílóó} \PS{adv.~} \GE{at glance} \XV{N nyɛ la a bie filoo kyɛ ka o kpɛ a lɔɔre.} \XE{I saw the child at a glance before he entered the lorry.} 
\LX{finni} \HM{} \PH{fìnnì} \PS{v.~} \GE{to cheat} \PD{finnee, finne, finne, finime, finnaa, finime } \VA{fuuli}
\LX{firimeŋ} \HM{} \PH{fìrìmèŋ} \PS{adv.~} \GE{appearing neatly cut e.g. growth} 
\LX{fo} \HM{} \PH{fʊ̀} \PS{pron.~} \SN \GE{you (second-person singular)} \XV{Fo na gaa la bee fo koŋ gaa?} \XE{Will you go or not?.} \SN \GE{your (second-person singular possessive)} \XV{Fo yuori la boŋ?} \XE{What is your name?.} 
\LX{fɔbo} \HM{} \PH{fɔ̀b} \PS{adv.~} \GE{very wet} 
\LX{foge} \HM{} \PH{fʊ̀gɪ̀} \PS{v.~} \SN \GE{to unshell e.g. paddy rice} \SN \GE{to poke, to push quickly} \SN \GE{to foam (soap)} \PD{fogɛɛ, fogera, fogera, fogereba, fogeraa, fogereba } \VA{feɛ}
\LX{fogele} \HM{} \PH{fʊ́glɪ́} \PS{v.~} \GE{to be tight e.g. passengers in a crowded bus} \XV{A bibiiri fogele la taa a lɔɔre poɔ.} \XE{The children are squeezed tightly in the lorry.} \PD{fogelɛɛ, fogelɔ, fogelɔ, fogelebɔ, fogelaa, fogelebɔ } 
\LX{fɔgele} \HM{} \PH{fɔ̀glɪ̀} \PS{v.~} \SN \GE{to press together} \SN \GE{to strangle} \PD{fɔgelɛɛ, fɔgelɔ, fɔgelɔ, fɔgelebɔ, fɔgelaa, fɔgeleba } 
\LX{fogeloo} \HM{} \PH{fʊ́glʊ́ʊ́} \PS{adj.~} \GE{narrow} \XV{A die waa la fogeloo lɛ.} \XE{The room is very small.} \OP{fogelɔ} 
\LX{fogi} \HM{} \PH{fógí} \PS{v.~} \GE{to pound e.g. clay for modeling} \PD{fogee, fogiro, fogiro, fogiribo, fogiraa, fogiribo } 
\LX{fogi} \HM{} \PH{fógì} \PS{v.~} \GE{to pull out from a tight hold} \XV{Dasaŋa na ba maŋ fogiri la nannyige nubi-eɛmɛ.} \XE{In the olden days, thieves had their fingernails pulled out.} \PD{fogee, fogiro, fogiro, fogiribo, fogiraa, fogiribo } 
\LX{fogili} \HM{} \PH{fòglì} \PS{v.~} \GE{to be seedless e.g. bean pod} \XV{Ka saaŋ ba mi a bɛŋɛ na fogili la.} \XE{If it does not rain the beans will not form seed.} \PD{fogilee, fogilo, fogilo, fogilibo, fogilaa, fogilibo } 
\LX{fogiluu} \HM{} \PH{fóglúú} \PS{adj.~} \GE{seedless} \OP{fogilo} 
\LX{fole} \HM{} \PH{fʊ̀lɪ́} \PS{adj.~} \GE{immature, unripe} \OP{folɔ} \TP{folɛɛ/follɛɛ} 
\LX{foli} \HM{} \PH{fòlì} \PS{adj.~} \GE{not hidden} \OP{folo} \TP{folee} 
\LX{folinfoori} \HM{} \PH{fóli̊nfǒórí} \PS{n.~} \GE{a type of broadleaf grass used as pig feed} \OP{folinfobo} \TP{folinfooree} \VA{folonfoori}
\LX{fɔlle} \HM{} \PH{fɔ̀llɪ̀} \PS{v.~} \GE{to smear with} \PD{fɔllɛɛ, fɔllɔ, fɔllɔ, fɔlebɔ, fɔllaa, fɔlebɔ } 
\LX{Folo} \HM{} \PH{fóló} \PS{n.~} \GE{Fulani (an African ethnic group)} \OP{folibo} \TP{folee/follee} 
\LX{fomeŋɛ} \HM{} \PH{fʊ́mɪ́ŋɛ́} \PS{pron.~} \GE{yourself} 
\LX{fomfom} \HM{} \PH{fómfóm} \PS{n.~} \GE{type of small cake-like food prepared from flour} 
\LX{fomfom} \HM{} \PH{fómfóm} \PS{adl.~} \GE{deepset (eyes)} 
\LX{foŋ} \HM{} \PH{fʊ́ŋ} \PS{pron.~} \GE{then you (used in conditional forms)} \XV{Ka saaŋ wa mi  foŋ bore a kamaana.} \XE{If it rains, you should sow the maize.} 
\LX{fonaano} \HM{} \PH{fʊ́náánʊ́} \PS{n.~} \GE{sleeveless singlet} \OP{fonaanore} \TP{fonaanorɛɛ/fonaanorrɛɛ} \VA{fenaano}
\LX{fɔŋe} \HM{} \PH{fɔ́ŋ} \PS{v.~} \GE{to become lean, to shrink} \PD{fɔŋɛɛ, fɔŋenɔ/fɔnnɔ, fɔŋenɔ/fɔnnɔ, fɔŋemɔ/fɔnemɔ, fɔŋenaa/fɔnnaa, fɔŋenemɔ/fɔnemɔ } 
\LX{fonne} \HM{} \PH{fʊ̀nnɪ̀} \PS{v.~} \SN \GE{to sniff (dog on a trail)} \SN \GE{to snuff} \PD{fonnɛɛ, fonnɔ, fonnɔ, fonnemɔ, fonnaa, fonnemɔ } 
\LX{fɔnne} \HM{} \PH{fɔ̀nnɪ̀} \PS{v.~} \GE{to grab, to hold} \PD{fɔnnɛɛ, fɔnnɔ, fɔnnɔ, fɔnemɔ, fɔnnaa, fɔnemɔ } 
\LX{fɔŋɔ} \HM{} \PH{fɔ́ŋɔ̀} \PS{n.~} \GE{suburb, section} \OP{fɔnne} \TP{fɔnnɛɛ} \VA{fɔŋa}
\LX{fonɔfonɔ} \HM{} \PH{fʊ́nɔ́fʊ́nɔ́} \PS{adj.~} \GE{soft, silky, and  light (cloth)} \VA{fonafona}
\LX{foo} \HM{} \PH{fòò} \PS{adv.~} \GE{looking open and endless e.g. a deep hole} 
\LX{foɔ} \HM{} \PH{fʊ́ɔ́} \PS{v.~} \GE{to fill holes in a wall with mortar} \PD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\LX{foɔ} \HM{} \PH{fʊ́ɔ́} \PS{v.~} \GE{to pile up a heap of guinea corn heads} \PD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\LX{foɔ} \HM{} \PH{fʊ́ɔ́} \PS{v.~} \GE{to place branches or thorns around an object for protection against animals} \PD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\LX{foɔ} \HM{} \PH{fʊ́ɔ́} \PS{v.~} \GE{to whip} \PD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\LX{foɔlaa} \HM{} \PH{fʊ́ɔ́láá} \PS{n.~} \GE{passageway} \OP{foɔle} \TP{foɔlɛɛ} 
\LX{foɔle} \HM{} \PH{fʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to draw one's appendages against the body; to stand at attention} \PD{foɔlɛɛ, foɔlɔ, foɔlɔ, foɔlbɔ, foɔlaa, - } 
\LX{foɔle} \HM{} \PH{fʊ́ɔ́lɪ̀} \PS{v.~} \GE{to sneak, to move stealthily} \PD{foɔlɛɛ, foɔlɔ, foɔlɔ, foɔlbɔ, foɔlaa, - } 
\LX{fooŋ} \HM{} \PH{fʊ́ʊ́ŋ} \PS{pron.~} \GE{you (emphatic)} \XV{Fo yɔɔpɔge balɛɛ la fooŋ na gaa te ɛre a zɔŋe?} \XE{Your sister is tired, you will have to go and grind the flour.} 
\LX{foone} \HM{} \PH{fʊ̀ʊ̀nɪ̀} \PS{v.~} \SN \GE{to shrink e.g. a fruit} \SN \GE{to deflate e.g. a tyre} \PD{foonɛɛ, foona, foona, foonema, foonaa, foonema } 
\LX{foone} \HM{} \PH{fʊ́ʊ́nɪ́} \PS{adj.~} \GE{shrinked} \OP{fooma} \TP{foonɛɛ/foomɛɛ} 
\LX{foore} \HM{} \PH{fʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to hiss e.g. a snake} \PD{foorɛɛ, foora, foora, fooreba, fooraa, fooreba } 
\LX{foore} \HM{} \PH{fʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to become bloated} \PD{foorɛɛ, foora, foora, fooreba, fooraa, fooreba } 
\LX{foore} \HM{} \PH{fʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to rage} \PD{foorɛɛ, foora, foora, fooreba, fooraa, fooreba } 
\LX{foɔre} \HM{} \PH{fʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to smear with thick liquid} \PD{foɔrɛɛ, foɔrɔ, foɔrɔ, foɔrba, foɔraa, foɔrba } 
\LX{fɔɔre} \HM{} \PH{fɔ̀ɔ̀rɪ̀} \PS{n.~} \GE{greed} \TP{fɔɔrɛɛ} 
\LX{foori} \HM{} \PH{fòòrí} \PS{adj.~} \SN \GE{empty} \SN \GE{empty-handed (typically used with nu)} \XV{O bɛlle ma la ne nufoori.} \XE{S/he deceived me with an empty hand.} 
\LX{fɔre} \HM{} \PH{fɔ̀rɪ̀} \PS{v.~} \GE{to hold, to squeeze tightly (by the throat)} \PD{fɔrɛɛ, fɔrɔ, fɔrɔ, fɔrbɔ, fɔraa, fɔrbɔ } 
\LX{fori} \HM{} \PH{fórì} \PS{v.~} \GE{to loosen, to slip off, to put off a dress around the waist} \PD{foree, foro, foro, forbo, foraa, forbo } 
\LX{fori} \HM{} \PH{fórí} \PS{use word.~} \GE{since} \XV{Fo naŋ fori yi, gaa kyɛ wa.} \XE{Since you have already prepared, go and come.} 
\LX{fori} \HM{} \PH{fòrì} \PS{v.~} \SN \GE{to force out from its cover or shell. (boiled bambara beans)} \SN \GE{to treat hide by removing the fur. (bleeching of the skin)} \SN \GE{to develop to an appreciable level; to mature} \XV{A pɔgelee ŋa naŋ ba fori kyɛ o pɔloo zuoe la.} \XE{This girl is not yet worth a woman but she is quite proud.} \PD{foree, foro, foro, forbo, foraa, forbo } 
\LX{fɔrɔ} \HM{} \PH{fɔ̀rɔ̀} \PS{n.~} \GE{yaws} \TP{fɔrrɛɛ} 
\LX{forɔforɔ} \HM{} \PH{fʊ́rɔ̀fʊ́rɔ̀} \PS{adj.~} \GE{rough} \VA{forafora}
\LX{forre} \HM{} \PH{fʊ́rrɪ́} \PS{v.~} \SN \GE{to scrape leaves off a branch} \SN \GE{to run very fast (through vegetation)} \PD{forrɛɛ, forra, forra, foreba, forraa, foreba } 
\LX{foto} \HM{} \PH{fótó} \PS{n.~} \GE{photograph} \OP{fotori} \TP{fotoree/fotorree} 
\LX{fo᷉o} \HM{} \PH{fʊ̀᷉ʊ̀᷉} \PS{adj.~} \GE{dry and light e.g. grass} 
\LX{fo᷉o} \HM{} \PH{fʊ̌᷉ʊ᷉́} \PS{pron.~} \GE{you} \XV{Fo᷉o naane ka ba boɔrɔ.} \XE{They are not looking for you.} \VA{fooŋ}
\LX{fo᷉ɔ} \HM{} \PH{fʊ́᷉ɔ́᷉} \PS{v.~} \GE{to make lighter} \XV{E ka a tuobu fo᷉ɔ fe᷉e ka lɛ naane N koŋ to᷉ɔ a tuo.} \XE{Make the load a little light otherwise I cannot carry it.} \PD{fo᷉ɔɛ, foɔnɔ, foɔnɔ, foɔnemɔ, foɔnaa, foɔnemɔ } 
\LX{fo᷉ɔo} \HM{} \PH{fʊ́᷉ɔ́᷉ʊ̀᷉} \PS{adj.~} \GE{light (weight)} \OP{foɔne} \TP{foɔnɛɛ} 
\LX{fu} \HM{} \PH{fú} \PS{n.~} \GE{rumour, hint} \XV{N woŋee fu kyɛ N ba baŋ ka serɛ la be.} \XE{I have heard a rumour but I am not suer whether it is true.} \TP{furee} 
\LX{fu} \HM{} \PH{fú} \PS{v.~} \GE{to blow (wind)} \PD{fue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\LX{fu} \HM{} \PH{fù} \PS{v.~} \SN \GE{to parboil, to steam (meat)} \SN \GE{to slough} \SN \GE{to make lather} \PD{fue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\LX{fugi} \HM{} \PH{fùgì} \PS{v.~} \GE{to threaten} \PD{fugee, fugiro, fugiro, fugiribo, fugiraa, fugiribo } 
\LX{fugi} \HM{} \PH{fùgì} \PS{v.~} \GE{to bluff} \PD{fugee, fugiro, fugiro, fugiribo, fugiraa, fugiribo } 
\LX{fugoŋ} \HM{} \PH{fúgóŋ} \PS{adv.~} \GE{falling suddenly into e.g. a hole} \XV{O faa le a bogiŋ lɛ fogoŋ.} \XE{S/he fell into the hole at once.} 
\LX{fulee} \HM{} \PH{fúléé} \PS{adv.~} \GE{eat gulpingly e.g of a hungry person} 
\LX{fuli} \HM{} \PH{fùlì} \PS{adv.~} \GE{exit suddenly} 
\LX{fulo} \HM{} \PH{fúló} \PS{adv.~} \GE{entering neatly, without friction} \XV{Naazoɔ deɛ e la n nyɔbogiriŋ lɛ fulo.} \XE{A fly shot into my nose at once.} 
\LX{fulofulo} \HM{} \PH{fúlófúló} \PS{adj.~} \GE{roundish e.g of aperture like a hole} 
\LX{fulonfugiri} \HM{} \PH{fúlónfúgrí} \PS{n.~} \GE{lung} \SG{fulonfuuruu} \OP{fulonfugo} \TP{fulonfugiree} 
\LX{fululu} \HM{} \PH{fúlúlú} \PS{adv.~} \GE{appearing round and deep e.g. a hole} 
\LX{fumini} \HM{} \PH{fúmìnì} \PS{n.~} \GE{needle} \OP{fumime} \TP{fuminee/fuminnee} 
\LX{fuoli} \HM{} \PH{fúólì} \PS{v.~} \SN \GE{to whistle} \SN \GE{to squeak (of a wheel)} \PD{fuolee, fuolo, fuolo, fuolibo, fuolaa, fuolibo } 
\LX{fuomii} \HM{} \PH{fùòmíì} \PS{n.~} \SN \GE{gluttony, voracious eating, wanting cheap things} \XV{A dɔɔ ŋa eŋ daŋ ba iri bonzaa, fuomii yoŋ ka o nɔŋ.} \XE{This man has never contributed anything, he only likes taking things for free.} \SN \GE{freeloading; wanting cheap things} \VA{fuomiŋ}
\LX{fuomiŋ} \HM{} \PH{fuomiŋ} \PS{n.~} \GE{} \VA{fuomii}
\LX{fuori} \HM{} \PH{fùòrì} \PS{v.~} \GE{to bewitch, to take the soul from the body} \PD{fuoree, fuoro, fuoro, fuoribo, fuoraa, fuoribo } 
\LX{fuori} \HM{} \PH{fúórì} \PS{v.~} \GE{to sip} \PD{fuoree, fuoro, fuoro, fuoribo, fuoraa, fuoribo } 
\LX{furi} \HM{} \PH{fùrì} \PS{v.~} \GE{to block an opening with an object, to seal} \PD{furee, furo, furo, furibo, furaa, furibo } 
\LX{furi} \HM{} \PH{fùrì} \PS{v.~} \GE{to open one's eyes widely e.g. when surprised} \PD{furee, furo, furo, furibo, furaa, furibo } 
\LX{furi} \HM{} \PH{fùrì} \PS{v.~} \GE{to stoop} \PD{furee, furo, furo, furibo, furaa, furibo } 
\LX{furi} \HM{} \PH{fúrí} \PS{v.~} \GE{to cave in} \PD{furee, furo, furo, furibo, furaa, furibo } 
\LX{furi} \HM{} \PH{fúrí} \PS{n.~} \GE{Islamic marriage rite} \OP{furri} \TP{furee/furree} 
\LX{furi kaa} \HM{} \PH{fùrì kàà} \PS{v.~} \GE{to stare, to glare} \PD{furi kaaɛ, furi kaaraa, furi kaaraa, furi kaareba, furi kaara, furi kaareba } 
\LX{fuu} \HM{} \PH{fúù} \PS{n.~} \GE{the cast-off skin of a snake} \OP{fuuri} \TP{fuuree} 
\LX{fuu} \HM{} \PH{fùù} \PS{v.~} \GE{to frighten, to threaten} \PD{fuue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\LX{fuu} \HM{} \PH{fúù} \PS{v.~} \GE{to speed} \PD{fuue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\LX{fuu} \HM{} \PH{fúú} \PS{n.~} \GE{a game where bare arrows are used} \OP{fuuri} \TP{fuuree} 
\LX{fuuli} \HM{} \PH{fúúlí} \PS{v.~} \GE{to cheat; to make a fool of} \PD{fuulee, fuulo, fuulo, fuulibo, fuulaa, fuulibo } \VA{finni}
\LX{fuuli} \HM{} \PH{fùùlì} \PS{v.~} \GE{to constrict, to make rounded (opening)} \PD{fuulee, fuulo, fuulo, fuulibo, fuulibo, fuulfuulo } 
\LX{fuuri} \HM{} \PH{fúúrí} \PS{n.~} \SN \GE{foam, bubbles, lather, froth} \SN \GE{rest , cool off} \OP{fue} \TP{fuuree} 
\LX{fuuri} \HM{} \PH{fùùrì} \PS{v.~} \GE{to cool off} \PD{fuuree, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\LX{fuuri} \HM{} \PH{fùúrí} \PS{n.~} \GE{treachery} 
\LX{fuurofuuro} \HM{} \PH{fúúrófúúró} \PS{n.~} \GE{bits and pieces (of information)} 
\LX{ga} \HM{} \PH{gà} \PS{v.~} \GE{to eat TZ without sorce/soup} \PD{gaɛ, gaara, gaara, garba, gaaraa, gaareba } 
\LX{ga} \HM{} \PH{gá} \PS{v.~} \GE{to run wildly or in a frenzy e.g. cow stung by tsetse fly} \PD{gagaɛ, gaara, gaara, gareba, gaaraa, gareba } 
\LX{gaa} \HM{} \PH{gáà} \PS{n.~} \GE{ebony (tree or fruit)} \OP{gaare} \TP{garɛɛ/gaarɛɛ} 
\LX{gaa} \HM{} \PH{gàà} \PS{v.~} \GE{to leave} \PD{gaaɛ, gɛrɛ, gɛrɛ, gɛrebɛ, gɛraa, gɛrebɛ } 
\LX{gaa} \HM{} \PH{gáà} \PS{v.~} \GE{to scoop pulp e.g. soil} \PD{gaaɛ, gaara, gaara, gaareba, gaaraa, gaareba } 
\LX{gaabu} \HM{} \PH{gáábú} \PS{n.~} \GE{onion} \OP{gaaburi} \TP{gaaburee/gaaburree} \VA{zammo}
\LX{gaafara} \HM{} \PH{gááfárà} \PS{n.~} \GE{excuse, sorry} \SG{gaafaroo} \OP{gaafarre} \TP{gaafarɛɛ} 
\LX{gaale} \HM{} \PH{gáálɪ̀} \PS{v.~} \SN \GE{to place flat on the ground, to put to bed} \SN \GE{to attend prenatal or postnatal care} \PD{gaalɛɛ, gaala, gaala, gaalba, gaalaa, gaaleba } 
\LX{Gaana} \HM{} \PH{gáånà} \PS{n.~} \GE{Ghana} \OP{gaanarɪ} \TP{gaanɛɛ} 
\LX{gaane} \HM{} \PH{gàànɪ̀} \PS{v.~} \SN \GE{to stride; to take strides of equal length with energy and confidence} \SN \GE{to wear by passing a strip between the buttocks (triangular underwear)} \XV{Neŋkponni mine taa la parre a maŋ de gaane yelmaale daare.} \XE{Some elderly men have triangular underwear which they wear on festive occasions.} \PD{gaanɛɛ, gaana, gaana, gaanema, gaanema, gaangaana } 
\LX{gaanoo} \HM{} \PH{} \PS{n.~} \GE{stride} \VA{gaane}
\LX{gabaa} \HM{} \PH{gàbàà} \PS{adv.~} \GE{open and extensive} 
\LX{gabadayaa (hsa)} \HM{} \PH{gàbádáyáá} \PS{adj.~} \GE{straightaway} \VA{gaba}
\LX{gabadii} \HM{} \PH{gàbàdìì} \PS{n.~} \GE{a type of woolen cloth used to make suits} 
\LX{gabaŋgabaŋ} \HM{} \PH{gàbàŋgàbàŋ} \PS{adv.~} \GE{anyhow} 
\LX{gadaa} \HM{} \PH{gàdàà} \PS{n.~} \SN \GE{ebony stick} \SN \GE{sign of extradition} \XV{A tendaana ba la nɔgraa gadaa.} \XE{The landlord has ordered the sojourner to leave.} 
\LX{gado} \HM{} \PH{gádó} \PS{n.~} \GE{bed} \OP{gadori} \TP{gadoree/gadorree} \VA{godo}
\LX{gagaraa} \HM{} \PH{gágáráà} \PS{n.~} \SN \GE{traitor} \SN \GE{One who denies the truth and treads on falsehood, inconsistent} \OP{gagarre} \TP{gagarɛɛ} \VA{minaafigi, minaafigri, minaafigree}
\LX{gakogili} \HM{} \PH{gákóglíí} \PS{n.~} \GE{dried berry of the ebony tree} \SG{gakogiluu} \OP{gakogilo} \TP{gakogilee} 
\LX{galaa} \HM{} \PH{gàláà} \PS{n.~} \GE{fool} \OP{galle} \TP{gallɛɛ} 
\LX{galbie} \HM{} \PH{gálbíé} \PS{n.~} \GE{a great person} \OP{galbiiri} \TP{galbiiree} 
\LX{galbie} \HM{} \PH{gálbíé} \PS{n.~} \GE{a child born after more than 9 months of pregnancy} \OP{galbiiri} \TP{galbiiree} 
\LX{gale} \HM{} \PH{gálɪ́} \PS{v.~} \GE{to speak in anger; to reproach} \PD{galɛɛ, gala, gala, galeba, galaa, galeba } 
\LX{gale} \HM{} \PH{gàlɪ̀} \PS{v.~} \GE{to miss the regular time} \XV{O daŋe ba gale toma waabo bebiyeni gba.} \XE{S/he never missed going to work even for a day.} \PD{galɛɛ, gala, gala, galba, galaa, galba } \VA{galle}
\LX{galegale} \HM{} \PH{gálgál} \PS{adv.~} \GE{of quick fast steps (run)} 
\LX{galegale} \HM{} \PH{gálɪ́gálɪ́} \PS{n.~} \GE{dog} 
\LX{galema} \HM{} \PH{gálɪmá́} \PS{adj.~} \GE{strange} \OP{galemalema} \VA{galee}
\LX{galeŋgaa} \HM{} \PH{gàlɪ́ŋgáà} \PS{n.~} \GE{crow} \OP{galeŋgaare} \TP{galeŋgaarɛɛ} 
\LX{galeŋgoɔnaa} \HM{} \PH{gàlɪ̀ŋgʊ́ɔ́náá} \PS{n.~} \GE{heron} \OP{galeŋgoɔne} \TP{galeŋgoɔnɛɛ} \VA{gɔleŋgoɔnaa}
\LX{galiŋkoori} \HM{} \PH{gàlíŋkóórì} \PS{n.~} \GE{millipede} \SG{galiŋkooruu} \OP{galiŋkoro} \TP{galiŋkoree} 
\LX{galle} \HM{} \PH{gàllɪ̀} \PS{v.~} \GE{to pass, to miss} \XV{Ka pɔgeŋ wa galle o pɛgebo kaapɔge poɔ la ka o taa.} \XE{If a woman misses her period, then she may be pregnant.} \PD{gallɛɛ, galla, galla, galeba, gallaa, - } 
\LX{galle} \HM{} \PH{gàllɪ̀} \PS{v.~} \GE{manoevre with skill and expertise} \XV{Te karema maŋ galle la a kɔkɔre ka noɔ kpɛ fo.} \XE{Our teacher manipulates the language such that you enjoy his talk.} \PD{gallɛɛ, galla, galla, galeba, gallaa, - } 
\LX{Galyuoni} \HM{} \PH{gàlyúóní} \PS{n.~} \GE{name traditionally given to a child born after 9 months of pregnancy} \VA{Yuoni}
\LX{gambanna} \HM{} \PH{gámbánná} \PS{n.~} \GE{one who is literate} \OP{gambanema} \TP{gambannɛɛ} 
\LX{gammiri} \HM{} \PH{gámmírì} \PS{n.~} \GE{woven twine string worn by women around the waist} \SG{gammiruu} \OP{gammie} \TP{gammiiree} 
\LX{gammisɔgelaa} \HM{} \PH{gámmísɔ́gláà} \PS{n.~} \GE{woven leather waist string} \OP{gammisɔgelɔ} 
\LX{gammizoolɔ} \HM{} \PH{gámmízóólɔ̀} \PS{n.~} \GE{waist strings for adolescent girl} \OP{gammizoolɛɛ} \VA{gammizoola}
\LX{gampɛle} \HM{} \PH{gámpɛ́lɪ̀} \PS{n.~} \SN \GE{sheet of paper} \SN \GE{page of a book} \OP{ gampɛlɛ} 
\LX{gamporrizie} \HM{} \PH{gámpórrìzíé} \PS{n.~} \GE{post office} \OP{gamporriziiri} \TP{gamporriziiree} 
\LX{gaŋa} \HM{} \PH{gàŋá} \PS{n.~} \GE{An enclosure for livestock e.g. cattle} \OP{ganne} \TP{gannɛɛ} \VA{gaŋe}
\LX{ganaa} \HM{} \PH{gànàá} \PS{n.~} \SN \GE{strip of leather} \SN \GE{twine for chief mourners of a deceased person} \OP{ganne} \TP{gannɛɛ} 
\LX{ganaaganaa} \HM{} \PH{gánáágánáá} \PS{adv.~} \GE{small and stretched like a string, thin} 
\LX{ganaaganaa} \HM{} \PH{gànààgànàà} \PS{adv.~} \GE{large and stretched} \XV{A dɔbɔ ti la taa ganaaganaa, ka ba faara faana.} \XE{The men hooked each other tightly and fiercely such that people were finding it difficult to separate them.} 
\LX{gandaa} \HM{} \PH{gàndáá} \PS{n.~} \SN \GE{hero, great person, protagonist} \SN \GE{hero in literature} \OP{gandaare} \TP{gandaarɛɛ} 
\LX{gandaa} \HM{} \PH{gàndáá} \PS{n.~} \GE{placenta} \OP{gandaare} \TP{gandaarɛɛ} 
\LX{gane} \HM{} \PH{gánɪ̀} \PS{n.~} \SN \GE{book} \SN \GE{hide, skin} \OP{gama} \TP{gannɛɛ/gamɛɛ} 
\LX{gaŋe} \HM{} \PH{gáŋ} \PS{adj.~} \GE{mad or insane (dogs)} \XV{Bagaŋe zaameŋ doŋe la n baa.} \XE{A mad dog bit my dog yesterday.} 
\LX{gaŋe} \HM{} \PH{gâŋ} \PS{v.~} \SN \GE{to lie down} \SN \GE{to sleep} \PD{gaŋɛɛ, gana, gana, ganema, ganaa, ganema } 
\LX{gaŋe} \HM{} \PH{gàŋɪ̀} \PS{v.~} \SN \GE{to step over} \XV{A baa age gaŋ la a bogi.} \XE{The dog jumped over the hole.} \SN \GE{to overtake,} \XV{Ba lɔɔre zo wa gaŋ la te deni na bare.} \XE{Their lorry overtook ours.} \SN \GE{to be more than, to exceed} \XV{A kparoo waa la kpoŋ gaŋ a bie.} \XE{The shirt is bigger than the child.} \PD{gaŋɛɛ, gaŋena/gana, gana, gaŋenema/ganema, gaŋenaa/ganaa, gaŋenema/ganema } 
\LX{gaŋene} \HM{1} \PH{gàŋɪ̀nɪ́} \PS{n.~} \GE{next one} \XV{N daga maŋ biŋ la a gaŋene die na poɔ.} \XE{My box is kept in the room after the first one.} 
\LX{gaŋene} \HM{2} \PH{gàŋɪ̀nɪ́} \PS{n.~} \GE{that which is too much} 
\LX{gaŋgaa} \HM{} \PH{gàŋgàá} \PS{n.~} \GE{drum (general)} \OP{gaŋgarɪ} \TP{gaŋgarɛɛ} 
\LX{gaŋgaanaa} \HM{} \PH{gǎŋgàànáà} \PS{n.~} \SN \GE{reed} \SN \GE{vine} \OP{gaŋgaane} \TP{gaŋgaanɛɛ} 
\LX{gaŋgadaa} \HM{} \PH{gàŋgàdáá} \PS{n.~} \GE{small drum with a quieter sound} \OP{gaŋgadaare} \TP{gaŋgadaarɛɛ} 
\LX{gaŋgadalee} \HM{} \PH{gàŋgàdàléé} \PS{n.~} \GE{drumstick} \OP{gaŋgadabilii} \VA{gaŋgadabili}
\LX{gaŋgakpoŋ} \HM{} \PH{gàŋgàkpóŋ} \PS{n.~} \GE{bass drum} \OP{gaŋgakponni} \TP{gaŋgakponnee} 
\LX{gaŋgalee} \HM{} \PH{gàŋgàléé} \PS{n.~} \GE{small drum; alto drum} \OP{gaŋgabilii} \TP{gaŋgabilee} 
\LX{gaŋganyaŋaa} \HM{} \PH{gàŋgàɲáŋáá} \PS{n.~} \GE{drum with a deep, loud sound} \OP{gaŋganyaŋene} \TP{gaŋganyaŋenɛɛ} 
\LX{gangyinni} \HM{} \PH{gàndzínní} \PS{n.~} \GE{the act of doing things in excess e.g. eating} \TP{gangyinnee} \VA{gangyinnuŋ}
\LX{gangyinnuŋ} \HM{} \PH{gàndzínnúŋ} \PS{n.~} \GE{the act of doing things in excess} \VA{gangyinnuu, gangyinni}
\LX{gaŋkanna} \HM{} \PH{gáŋkánná} \PS{n.~} \GE{reader, letter or book reader} \OP{gaŋkanema} \TP{gaŋkanemɛɛ} 
\LX{gaŋkoŋi} \HM{} \PH{gàŋkǒŋ} \PS{n.~} \GE{funeral string/strip given without any accompanying funeral donation} \OP{gaŋkonni} \TP{gaŋkonnee} 
\LX{gaŋkoŋi} \HM{} \PH{gáŋkóŋ} \PS{n.~} \GE{dry and crumpled animal hide} \OP{gaŋkonni} \TP{gaŋkonnee} 
\LX{gaŋkuoŋaa} \HM{} \PH{gàŋkúóŋáá} \PS{n.~} \GE{dried strip of leather} \OP{gaŋkuoni} \TP{gaŋkuonee} 
\LX{ganlaa} \HM{} \PH{gánláá} \PS{n.~} \GE{a type of dance} \OP{ganlaare} \TP{ganlaarɛɛ} 
\LX{ganlɔgɔ} \HM{} \PH{gànlɔ̀gɔ̀} \PS{n.~} \GE{small leather bag worn on the shoulder} \OP{ganlɔgre} \TP{ganlɔgrɛɛ} 
\LX{gaŋŋmaara} \HM{} \PH{gáŋŋmáárá} \PS{n.~} \GE{writer} \OP{gaŋŋmaareba} \TP{gaŋŋmaarebɛɛ} \VA{gansɛgrɛ}
\LX{gansɛgerɛ} \HM{} \PH{gánsɛ́grɛ̀} \PS{n.~} \GE{writer, letter writer} \OP{gansegerebɛ} \TP{ganɛgerebɛɛ} \VA{gaŋŋmaara}
\LX{gansɔraa} \HM{} \PH{gánsɔ́ráá} \PS{n.~} \GE{a book for reading} \OP{gansɔrre} \TP{gansɔrrɛɛ} 
\LX{gantoɔle} \HM{} \PH{gántʊ́ɔ́lɪ̀} \PS{n.~} \GE{mail} \SG{gantoɔlaa} \TP{gantoɔlɛɛ} 
\LX{gara} \HM{} \PH{gàrá} \PS{n.~} \GE{tapeworm} \OP{garre} \TP{garɛɛ/garrɛɛ} 
\LX{gara} \HM{} \PH{gárá} \PS{n.~} \GE{hobble (rope for holding two legs of an animal)} \OP{garre} \TP{garɛɛ/garrɛɛ} \VA{gbara}
\LX{garaa} \HM{} \PH{gáráá} \PS{adj.~} \GE{small and pointed} 
\LX{garaa} \HM{} \PH{gàráà} \PS{n.~} \GE{ridge} \OP{garre} \TP{garrɛɛ} 
\LX{gare} \HM{} \PH{gárɪ́} \PS{v.~} \SN \GE{to stumble, to trip over} \SN \GE{to tie or hold legs together} \PD{garɛɛ, gara, gara, gareba, garaa, gareba } 
\LX{gare} \HM{} \PH{gàrɪ̀} \PS{v.~} \GE{to place across, to close (road)} \XV{A dɔɔ de la daa gare a sori.} \XE{The man has placed a plank across the road.} \PD{garɛɛ, gara, gara, gareba, garaa, gareba } 
\LX{gare} \HM{} \PH{gárɪ́} \PS{v. adj.~} \GE{go bad, sour (food)} 
\LX{garegyerɛ} \HM{} \PH{gárɪ́yɪ̀rɛ} \PS{n.~} \GE{a shrub with milky sap, poisonous} \SG{garegyeroo} \TP{gargyerɛɛ} 
\LX{Garegyerɛbaa} \HM{} \PH{gàrɪ́dzìrɛ́bàá} \PS{n.~} \GE{One of the main district headquarters of the Dagaare speaking area whose dialeɔt is often referred to as the ‘central dialect’} \VA{Gyerɛbaa}
\LX{garepegele} \HM{} \PH{gàrɪ̀pɪ̀glɪ̀} \PS{v.~} \GE{to crisscross} \XV{A teɛ le la garepegele a sori.} \XE{The tree fell and crossed/bocked the road.} \PD{garepegelɛɛ, garapegelɛ, garapegelɛ, garepegelebɛ, garepegelaa, garepeglebɛ } 
\LX{garewaa} \HM{} \PH{gárɪ́wáá} \PS{n.~} \GE{four-gallon tin} \OP{garewaare} \TP{garewaarɛɛ} 
\LX{garoŋ} \HM{} \PH{gárʊ́ŋ} \PS{n.~} \GE{sour-tasting substance} 
\LX{garre} \HM{} \PH{gárrɪ́} \PS{v.~} \GE{to deny} \PD{garrɛɛ, garra, garra, gareba, garaa, gareba } 
\LX{gasalaa} \HM{} \PH{gàsàlàà} \PS{adj.~} \GE{gaping} 
\LX{ga᷉a} \HM{} \PH{gà᷉à᷉} \PS{adv.~} \GE{outstretched and long} \XV{A sɔrsɔrɔ te᷉ɛ la o nu ga᷉a ka nembaaloŋ erɛ a noba ka ba korɔ o libie.} \XE{The beggar stretched out his hand ernestly and the onlookers were moved and began offering him money.} 
\LX{gba} \HM{} \PH{gbá} \PS{v.~} \SN \GE{to gather substantive debris} \SN \GE{to beat someone completely (gambling)} \PD{gbaɛ, gbaara, gbaara, gbarba, gbaaraa, gbarba } 
\LX{gba} \HM{} \PH{gbà} \PS{v.~} \GE{to close tightly, to lock} \PD{gbaɛ, gbaara, gbaara, gbareba, gbaaraa, gbareba } 
\LX{gbaa} \HM{} \PH{gbáà} \PS{n.~} \GE{viper} \OP{gbaare} \TP{gbaarɛɛ} 
\LX{gbaalaa} \HM{} \PH{gbááláá} \PS{n.~} \GE{large, sturdy broom for sweeping pavement} \OP{gbaale} \TP{gbaalɛɛ} 
\LX{gbaale} \HM{} \PH{gbáálɪ́} \PS{n.~} \GE{large earthenware basin} \OP{gbaala} \TP{gbaalɛɛ} 
\LX{gbaale} \HM{} \PH{gbààlɪ̀} \PS{v.~} \GE{to gather extensively} \XV{A koɔraa gbaale la o weɛ kɛkɛrɛɛ baare.} \XE{The farmer has finished gathering the stalks on his farm.} \PD{gbaalɛɛ, gbaala, gbaala, gbaaleba, gbaalaa, gbaaleba } 
\LX{gbaale} \HM{} \PH{gbáálɪ̀} \PS{v.~} \GE{to follow in large numbers (typically used with tu)} \XV{Polsiri nyɔge la a nannyige ka bibiiri gbaale tu.} \XE{The police arrested the thief and children crowded behind them.} \PD{gbaalɛɛ, gbaala, gbaala, gbaalba, gbaalaa, gbaalba } 
\LX{gbaama} \HM{} \PH{gbààmà} \PS{n.~} \GE{weak appendages} \TP{gbaamɛɛ} 
\LX{gbaane} \HM{} \PH{gbàànɪ́} \PS{adj.~} \GE{lean} \OP{gbaama} 
\LX{gbaga} \HM{} \PH{gbágá} \PS{n.~} \GE{animal cage} \OP{gbagre} \TP{gbagrɛɛ} \VA{gbɔɡɔ}
\LX{gbage} \HM{} \PH{gbàgɪ̀} \PS{v.~} \GE{to face} \XV{A peroo yie la a de o niŋe gbage sapare kyɛ deɛ zoro.} \XE{When the sheep left the pen it just went running toward the east.} \PD{gbagɛɛ, gbagra, gbagra, gbagreba, gbagraa, gbagreba } \VA{gbɔge}
\LX{gbagelaŋ} \HM{} \PH{gbàglàŋ} \PS{adj.~} \GE{large and strong e.g. cob of corn} 
\LX{gbagele} \HM{} \PH{gbáglɪ́} \PS{n.~} \GE{bone with meat} \OP{gbagla} \TP{gbaglɛɛ} 
\LX{gbala} \HM{} \PH{gbàlà} \PS{n.~} \GE{a type of women's hairstyle} \XV{A pɔgepaalaa poŋi la gbala.} \XE{The bride has shaved gbala on her head.} \OP{gballe} \TP{gbalɛɛ} 
\LX{gbalaa} \HM{} \PH{gbàláà} \PS{n.~} \SN \GE{pole for plucking} \SN \GE{a type of tree that has aerial roots, milky sap, and (usually) a large amount of shade} \OP{gballe} \TP{gballɛɛ} \VA{laŋgbaraa, kaŋkaŋgbalaa}
\LX{gbale} \HM{} \PH{gbàlɪ́} \PS{adj.~} \GE{It is not for nothing that the child is crying} \OP{gbala} \TP{gbalɛɛ} 
\LX{gbale} \HM{} \PH{gbàlɪ́} \PS{adj.~} \GE{not wearing anything arround the waist} \XV{Maria taa la se-gbale.} \XE{Mary does not have any beads around her waist.} \OP{gbala} \TP{gbalɛɛ} 
\LX{gbale} \HM{} \PH{gbàlɪ́} \PS{adj.~} \GE{blunt, not sharp} \XV{A sɔ-gbale na koŋ to᷉ɔ ŋmaa a nɛne.} \XE{That blunt knife can't cut the meat.} \OP{gbala} \TP{gbalɛɛ} 
\LX{gbale} \HM{} \PH{gbàlɪ̀} \PS{v.~} \GE{to make blunt} \PD{gbalɛɛ, gballa, gballa, gbaleba, gballaa, gbaleba } 
\LX{gbambelɛ} \HM{} \PH{gbàmbɪ̀lɛ̀} \PS{n.~} \GE{albino} \OP{gbambelle} \TP{gbambelɛɛ} 
\LX{gbambiri} \HM{} \PH{gbàmbírì} \PS{n.~} \GE{an object with two distinct sides (cowrie) thrown in gambling to decide the winner} \SG{gbambiruu} \OP{gbambie} \TP{gbambiiree} 
\LX{gbambiri} \HM{} \PH{gbàmbìrì} \PS{n.~} \GE{a type of musical instrument, made of metal tin, a wooden board and spokes. Spoke-like metals are fixed on the board and fixed into the tin. It is played like the guitar} \TP{gbambiree} 
\LX{gbammo} \HM{} \PH{gbàmmʊ́} \PS{n.~} \SN \GE{comb (of a rooster)} \SN \GE{a variety of maize that usually has big cobs and pure white kernels} \OP{gbama/gbanne} \TP{gbamɛɛ} \VA{gyampegre}
\LX{gbandige} \HM{} \PH{gbǎndígé} \PS{n.~} \GE{area of jurisdiction} \OP{gbandigri} 
\LX{gbandiruŋ} \HM{} \PH{gbǎndírúŋ} \PS{n.~} \GE{inheritance} \TP{gbandiree} 
\LX{gbaŋe} \HM{} \PH{gbáŋ} \PS{n.~} \GE{gambling} \OP{gbanne} \TP{gbannɛɛ} \VA{gbaŋa}
\LX{gbaŋe} \HM{} \PH{gbǎŋ} \PS{n.~} \GE{albino} \OP{gbanne} \TP{gbannɛɛ} \VA{gbanzeɛ}
\LX{gbaŋe} \HM{} \PH{gbáŋ} \PS{v.~} \GE{to shoot an arrow as far as one can} \PD{gbaŋɛɛ, gbana, gbana, gbanema, gbanaa, gbanema } 
\LX{gbanfolɔ} \HM{} \PH{gbànfʊ̀là} \PS{n.~} \GE{a dish made from mashed fresh corn} \TP{gbanfolɛɛ/gbanfollɛɛ} \VA{gbanfola}
\LX{gbaŋgbaa} \HM{} \PH{gbàŋgbáà} \PS{n.~} \GE{slave} \OP{gbaŋgbaare} \TP{gbaŋgbaarɛɛ} 
\LX{gbaŋgbaaloŋ} \HM{} \PH{gbàŋgbààlʊ́ŋ} \PS{n.~} \GE{slavery} 
\LX{gbaŋgbakyeraa} \HM{} \PH{gbáŋgbákyɪ́ráà} \PS{n.~} \GE{stone axe (chopper?)} \OP{gbaŋgbakyerre} \TP{gbaŋgbakyerɛɛ/gbaŋgbakyerrɛɛ} 
\LX{gbaŋgbalaa} \HM{} \PH{gbáŋgbáláà} \PS{n.~} \SN \GE{a clearing} \SN \GE{spot for threshing grain} \OP{gbaŋgballe} \TP{gbaŋgbalɛɛ} \VA{gbeŋgbelaa}
\LX{gbaŋgbaŋ} \HM{} \PH{gbáŋgbáŋ} \PS{n.~} \GE{midday} \OP{gbaŋgbanne} \TP{gbaŋgbannɛɛ} 
\LX{gbaŋgbaŋ} \HM{} \PH{gbàŋgbàŋ} \PS{adv.~} \GE{wide, open, and endless} 
\LX{gbaŋgbele} \HM{} \PH{gbàŋgbɪ́lɪ́} \PS{adv.~} \GE{raw and hard (fruit)} 
\LX{gbaŋkeroo} \HM{} \PH{gbàŋkéróó} \PS{n.~} \GE{a type of wild plant with edible fruit} \OP{gbaŋkerri} \TP{gbaŋkeree/gbaŋkerree} 
\LX{gbanna} \HM{} \PH{gbánná} \PS{n.~} \GE{small farm near the house} \OP{gbanne} \TP{gbannɛɛ} \VA{gbannaa}
\LX{gbannaa} \HM{} \PH{gbánnáá} \PS{n.~} \GE{crop field} \OP{gbanne} \TP{gbannɛɛ} \VA{gbanna}
\LX{gbannaa} \HM{} \PH{gbànnà} \PS{n.~} \GE{see gbannaa (crop field)} \OP{gbanne} \TP{gbannɛɛ} \VA{gbanna}
\LX{gbantieloo} \HM{} \PH{gbàntíélóó} \PS{n.~} \GE{the act of shooting an arrow as high as one can. (used with "saluoni')} \XV{O gbaŋ la saluoni gbantieloo.} \XE{He has shot an arrow into the sky.} 
\LX{gbantori} \HM{} \PH{gbàntòrí} \PS{n.~} \GE{ulcer} \OP{gbantoe} \TP{gbantoree} \VA{gbontori}
\LX{gbanzeɛ} \HM{} \PH{gbànzɪ̀ɛ́} \PS{n.~} \GE{very light-skinned person} \OP{gbanzeere} \TP{gbanzeerɛɛ} \VA{gbaŋ}
\LX{gbara} \HM{} \PH{gbàrá} \PS{n.~} \GE{a deaf or mute person} \OP{gbarre} \TP{gbarɛɛ} 
\LX{gbare} \HM{} \PH{gbárɪ́} \PS{v.~} \GE{to struggle in agony e.g. dying} \XV{Lɔɔre ŋmɛ la naabo ka o le a sori zu deɛ gbara.} \XE{The lorry hit the cow and it fell on the road struggling.} \PD{gbarɛɛ, gbara, gbara, gbarba, gabaraa, gabrba } 
\LX{gbare} \HM{} \PH{gbárɪ́} \PS{v.~} \GE{to tie legs together firmly} \XV{Ba naneŋ nyɔge la a naaweɛmɛ na, a nyɔge gbare.} \XE{When they eventually caught that wild cow, they tied it down by the legs.} \PD{gbarɛɛ, gbara, gbara, gbarba, gabaraa, gabrba } 
\LX{gbare} \HM{} \PH{gbàrɪ̀} \PS{v.~} \GE{to shut off from hearing} \PD{gbarɛɛ, gbara, gbara, gbareba, gbaraa, gbareba } 
\LX{gbarema} \HM{} \PH{gbàrɪ̀má} \PS{n.~} \GE{small child} \OP{gbarre} \TP{gbaremɛɛ} 
\LX{gbateele} \HM{} \PH{gbàtɪ́ɪ́lɪ́} \PS{n.~} \GE{a kind of grass which is heated and smashed to explode like knockout} \SG{gbateeloo} \OP{gbateelɛ} \TP{gbateelɛɛ} 
\LX{gbatɛre} \HM{} \PH{gbàtɛ́rɪ́} \PS{n.~} \GE{small, shallow pond (usually in the bush)} \OP{gbatɛrre} \TP{gbatɛrɛɛ} 
\LX{gbe} \HM{} \PH{gbè} \PS{v.~} \SN \GE{to grind coarsely} \SN \GE{to knaw at something hard; (unripe fruit)} \PD{gbee, gbiere, gbiere, gbieribe, gbieraa, gbegbiere } 
\LX{gbɛbaŋa} \HM{} \PH{gbɛ́báŋ} \PS{n.~} \GE{ankle bangle} \OP{gbɛbanne} \TP{gbɛbannɛɛ} \VA{gbɛbɔŋɔ/gbɛbaŋe}
\LX{gbɛbi-eɛne} \HM{} \PH{gbɛ́bì-ɪ́ɛ́nɪ́} \PS{n.~} \SN \GE{toenail} \SN \GE{claw} \OP{gbɛbi-eɛmɛ} \TP{gbɛbi-eɛnɛɛ} 
\LX{gbɛbinyoɔre} \HM{} \PH{gbɛ́bìɲʊ́ɔ́rɪ́} \PS{n.~} \GE{tip of the toe} \OP{gbɛbinyɔɛ} \TP{gbɛbinyɔrɛɛ} 
\LX{gbɛbiri} \HM{} \PH{gbɛ́bìrí} \PS{n.~} \GE{toe} \OP{gbɛbie} \TP{gbɛbiree/gbɛbiiree} 
\LX{gbɛbogi} \HM{} \PH{gbɛ́bògí} \PS{n.~} \SN \GE{footprint} \SN \GE{spot} \OP{gbɛbogri} \TP{gbɛbogree} 
\LX{gbɛɛ yaga} \HM{} \PH{gbɛ́ɛ́ yágá} \PS{adj.~} \GE{often} \VA{buyibuyi}
\LX{Gbeɛdomɔ} \HM{} \PH{gbɪ̀ɛ̀dʊmɔ́} \PS{n.~} \GE{land area of the clan members (Gbeɛdomɛɛ)} \VA{Gbeɛdoma}
\LX{Gbeɛdomoo} \HM{} \PH{gbɪ̀ɛ̀dʊ́mʊ́ʊ́} \PS{n.~} \GE{a member of the gbeɛdoma clan} \OP{Gbeɛdomɛɛ} 
\LX{gbeɛle} \HM{} \PH{gbɪ̀ɛ̀lɪ̀} \PS{n.~} \GE{a type of fetish manifested in a stone; a stone believed to have supernatural power and treated as a god. The clan god of the Gbeɛdomɛɛ} \OP{gbeɛlɛ} \TP{gbeɛlɛɛ} 
\LX{gbeɛle} \HM{} \PH{gbɪ́ɛ́lɪ́} \PS{v.~} \GE{to eye with suspicion, to look angrily at} \PD{gbeɛle, gbeɛlɛ, gbeɛlɛ, gbeɛlebɛ, gbeɛlaa, gbeɛlebɛ } 
\LX{gbeɛle kuuri} \HM{} \PH{gbɪ̀ɛ́lɪ̀ kúúrí} \PS{n.~} \SN \GE{round, smooth stone} \SN \GE{such a stone believed to be a symbol of a fetish} \OP{gbeɛle kubo} \TP{gbeɛle kuuree} \VA{gbeɛle}
\LX{gbeɛre} \HM{} \PH{gbɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to be completely out of energy; to be spent} \PD{gbeɛre, gbeɛrɛ, gbeɛrɛ, gbeɛrebɛ, gbeɛraa, gbeɛrebɛ } 
\LX{gbeɛre} \HM{} \PH{gbɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to be felled by a rainstorm (of grass)} \PD{gbeɛre, gbeɛrɛ, gbeɛrɛ, gbeɛrebɛ, gbeɛraa, gbeɛrebɛ } 
\LX{gbeɛre} \HM{} \PH{gbɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to cut bits and pieces of meat} \PD{gbeɛre, gbeɛrɛ, gbeɛrɛ, gbeɛrebɛ, gbeɛraa, gbeɛrebɛ } 
\LX{gbeɛrɛ} \HM{} \PH{gbɪ́ɛ́rɛ́} \PS{n.~} \GE{bits and pieces of meat} \TP{gbeɛrɛɛ} 
\LX{gbɛfaa} \HM{} \PH{gbɛ́fáá} \PS{n.~} \GE{ill luck; unluckiness} \OP{gbɛfaare} \TP{gbɛfaarɛɛ} 
\LX{gbɛgbɛ} \HM{} \PH{gbɛ̀gbɛ́} \PS{n.~} \GE{wide, shallow basket} \OP{gbɛgbɛre} \TP{gbɛgbɛrɛɛ} 
\LX{gbɛgere} \HM{} \PH{gbɛ̀grɪ̀} \PS{v.~} \GE{to be miserly, to economize} \PD{gbɛgerɛɛ, gbɛgerɛ, gbɛgerɛ, gbɛgerebɛ, gbɛgeraa, gbɛgerebɛ } 
\LX{gbɛgere} \HM{} \PH{gbɛ́grɪ́} \PS{v.~} \GE{to be loose, to be shaky (object in a hole)} \XV{A daa bogi ba sigi yaga lɛnso ka o gbɛgrɛ.} \XE{The hole for the pole is not deep enough; that is why it is shaky.} \PD{gbɛgrɛɛ, gbɛgrɛ, gbɛgrɛ, gbɛgrebɛ, gbɛgraa, gbɛgrebɛ } 
\LX{gbegi} \HM{} \PH{gbégí} \PS{v.~} \GE{to pile up e.g. firewood} \PD{gbegee, gbegre, gbegire, gbegiribe, gbegiraa, gbegire } 
\LX{gbegili} \HM{} \PH{gbèglì} \PS{adj.~} \GE{hard and rigid} 
\LX{gbegili} \HM{} \PH{gbèglì} \PS{v.~} \GE{to become dry and hard} \PD{gbegilee, gbegile, gbegile, gbeglibe, gbegilaa, gbegilibe } 
\LX{gbɛkpara} \HM{} \PH{gbɛ́kpárá} \PS{n.~} \GE{double legged} \OP{gbɛkparre} \TP{gbɛkparɛɛ} 
\LX{gbɛkpeɛŋaa} \HM{} \PH{gbɛ́kpɪ́ɛ́ŋáá} \PS{n.~} \GE{fleet footed} \OP{gbɛkpeɛne} 
\LX{gbɛkpoŋi} \HM{} \PH{gbɛ́kpóŋ} \PS{n.~} \GE{thigh} \OP{gbɛkponni} \TP{gbɛkponnee} 
\LX{gbelaa} \HM{} \PH{gbèláá} \PS{n.~} \GE{the bone of the forhead} \OP{gbelaare} \TP{gbelaarɛɛ} 
\LX{gbelaa} \HM{} \PH{gbɪ̀làà} \PS{adv.~} \GE{protruding, straight and hard} 
\LX{gbele} \HM{} \PH{gbɪ́lɪ́} \PS{n.~} \GE{a type of small fox with bushy fur} \OP{gbelɛ} \TP{gbelɛɛ/gbellɛɛ} 
\LX{gbele} \HM{} \PH{gbɪ́lɪ́} \PS{v.~} \GE{to work very hard} \PD{gbelɛɛ, gbelɛ, gbelɛ, gbelbɛ, gbelaa, - } 
\LX{gbɛle} \HM{} \PH{gbɛ́lɪ̀} \PS{n.~} \GE{a long, thin strip of material e.g. rubber} \OP{gbɛlɛ} \TP{gbɛlɛɛ/gbɛllɛɛ} 
\LX{gbɛle} \HM{} \PH{gbɛ́lɪ́} \PS{v.~} \SN \GE{to wriggle one's trunk (like in dancing)} \SN \GE{way of begging where one wriggles one's wrist} \PD{gbɛlɛɛ, gbɛlɛ, gbɛlɛ, gbɛlbɛ, gbɛlaa, gbɛlbɛ } 
\LX{gbɛlee} \HM{} \PH{gbɛ́léé} \PS{n.~} \GE{foot} \OP{gbɛbilii} \TP{gbɛbilee} 
\LX{gbɛlɛɛ} \HM{} \PH{gbɛ̀lɛ̀ɛ̀} \PS{adj.~} \GE{loose and sagging e.g. a rope} 
\LX{gbeleŋgbeleŋ} \HM{} \PH{gbèlèŋgbéléŋ} \PS{n.~} \GE{flat metal musical instrument; a cymbal} \OP{gbeleŋgbelli} \TP{gbeleŋgbelee} \VA{lato/nato}
\LX{gbɛlgbɛle} \HM{} \PH{gbɛ̀lgbɛ̀l} \PS{adv.~} \GE{long and flowing} 
\LX{gbelime} \HM{} \PH{gbèlìmé} \PS{n.~} \GE{bell} \OP{gbelli} \TP{gbellee} 
\LX{gbelime} \HM{} \PH{gbèlìmé} \PS{n.~} \GE{hour} \OP{gbelli} \TP{gbellee} 
\LX{gbelli} \HM{} \PH{gbéllì} \PS{v.~} \GE{to turn the head away abruptly} \PD{gbellee, gbelle, gbelle, gbelibe, gbellaa, gbelibe } 
\LX{gbɛlɔlɔraa} \HM{} \PH{gbɛ́lɔ̀lɔ̀ráà} \PS{n.~} \GE{calf (human body)} \OP{gbɛlɔlɔrre} \TP{gbɛlɔlɔrɛɛ} 
\LX{gbɛloŋkoraa} \HM{} \PH{gbɛ́lòŋkòràá} \PS{n.~} \GE{the region at the back of the knee joint} \OP{gbɛloŋkorri} \TP{gbɛloŋkoree} 
\LX{gbɛneɛraa} \HM{} \PH{gbɛ́níɛ́ráá} \PS{n.~} \GE{footprint} \OP{gbɛneɛre} \TP{gbɛneɛrɛɛ} 
\LX{gbeŋgbegri} \HM{} \PH{gbèŋgbégrí} \PS{n.~} \SN \GE{river bank} \SN \GE{riverbed (dry)} \OP{gbeŋgbege} \TP{gbeŋgbegree} 
\LX{gbeŋgbelaa} \HM{} \PH{gbɪ́ŋgbɪ́láá} \PS{n.~} \GE{ground where millet is threshed and winnowed out} \OP{gbeŋgbelle} \TP{gbeŋgbelɛɛ/gbeŋgbellɛɛ} \VA{gbaŋgbalaa}
\LX{gbeŋgbeŋ} \HM{} \PH{gbéŋgbéŋ} \PS{adv.~} \GE{firmly fixed} 
\LX{gbeŋini} \HM{} \PH{gbéŋìní} \PS{n.~} \GE{lion} \OP{gbeŋime} \TP{gbennee} 
\LX{gbeŋkeroo} \HM{} \PH{gbèŋkéróó} \PS{n.~} \GE{a shrub with brown hard fruits that are edible. Also the fruit of such a tree} \OP{gbeŋkerri} \TP{gbeŋkeree} \VA{naakyeemɛgbeŋkeroo}
\LX{gbɛnne} \HM{} \PH{gbɛ̀nnɪ̀} \PS{v.~} \GE{to pluck (leaves)} \PD{gbɛnnɛɛ, gbɛnnɛ, gbɛnnɛ, gbɛnemɛ, gbɛnnaa, gbɛnemɛ } 
\LX{gbɛnne} \HM{} \PH{gbɛ́nnɪ́} \PS{v.~} \GE{to chip the rim or edge of something} \PD{gbɛnnɛɛ, gbɛnnɛ, gbɛnnɛ, gbɛnemɛ, gbɛnnaa, gbɛnemɛ } 
\LX{gbɛre} \HM{} \PH{gbɛ́rɪ̀} \PS{n.~} \SN \GE{leg} \SN \GE{foot} \OP{gbɛɛ} \TP{gbɛɛrɛɛ} 
\LX{gbɛre} \HM{} \PH{gbɛ́rɪ́} \PS{n.~} \GE{person with a physical disability that prevents them from walking} \OP{gbɛrre} \TP{gbɛrɛɛ} \VA{gbɛɛ (1pl)}
\LX{gbɛre} \HM{} \PH{gbɛ̀rɪ́} \PS{n.~} \GE{mane} \OP{gbɛrre} \TP{gbɛrɛɛ} 
\LX{gbɛre} \HM{} \PH{gbɛ̀rɪ̀} \PS{v.~} \GE{to be unable to walk due to a physical disability} \PD{gbɛrɛɛ, gbɛrɛ, gbɛrɛ, gbɛrebɛ, gbɛraa, gbɛrebɛ } 
\LX{gbɛrɛ} \HM{} \PH{gbɛ̀rɛ̀} \PS{n.~} \GE{sponge} \SG{gbɛrrɛɛ} \TP{gbɛrɛɛ} 
\LX{gbɛregbɛre} \HM{} \PH{gbɛ̀rɪ̀gbɛ́rɪ́} \PS{v.~} \GE{crawl, walk with serious difficulty} \PD{gbɛregbɛrɛɛ, gbɛregbɛrɛ, gbɛregbɛrɛ, gbɛregbɛrbɛ, gbɛregbɛraa, gbɛregbɛrbɛ } 
\LX{gberi} \HM{} \PH{gbèrì} \PS{adv.~} \GE{of a twine snapping sound} 
\LX{gberi} \HM{} \PH{gbérì} \PS{adv.~} \GE{stopping suddenly} 
\LX{gbɛroo} \HM{} \PH{gbɛ̀rʊ̀ʊ̀} \PS{n.~} \GE{weak and dejected; paralysis} 
\LX{gbɛsampane} \HM{} \PH{gbɛ́sámpánɪ́} \PS{n.~} \GE{top of foot} \OP{gbɛsampama} \TP{gbɛsampanɛɛ/gbɛsampannɛɛ} 
\LX{gbɛsugili} \HM{} \PH{gbɛ́súglí} \PS{n.~} \GE{ankle} \OP{gbɛsugilo} \TP{gbɛsugilee} 
\LX{gbɛtanlori} \HM{} \PH{gbɛ́tánlórí} \PS{n.~} \GE{half-bent leg on the other} \OP{gbɛtanloe} \TP{gbɛtanloree} 
\LX{gbɛtɛre} \HM{} \PH{gbɛ̀tɛ́rɪ́} \PS{n.~} \GE{small pond, usually in the bush} \OP{gbɛtɛrre} \TP{gbɛtɛrɛɛ/gbɛtɛrrɛɛ} \VA{gbatare}
\LX{gbɛtoloŋ} \HM{} \PH{gbɛ́tʊ́lʊ́ŋ} \PS{n.~} \SN \GE{the condition of being fast-footed; quickness} \SN \GE{intelligence} \OP{gbɛtola.} 
\LX{gbɛtooraa} \HM{} \PH{gbɛ́tʊ́ʊ́ráá} \PS{n.~} \GE{sock} \OP{gbɛtoore} \TP{gbɛtoorɛɛ} 
\LX{gbɛvelaa} \HM{} \PH{gbɛ́vɪ́làá} \PS{n.~} \GE{good luck, good fortune} \OP{gbɛveɛle} 
\LX{gbɛyeni} \HM{} \PH{gbɛ́yénì} \PS{n.~} \GE{one-legged thing} 
\LX{gbe᷉ɛ} \HM{} \PH{gbɪ́᷉ɛ́} \PS{v.~} \GE{to cut a bit tiny} \PD{gbeɛ᷉ɛ, gbeɛnɛ, gbeɛnɛ, gbeɛnemɛ, gbeɛnaa, gbeɛnemɛ } 
\LX{gbɛ᷉ɛ} \HM{} \PH{gbɛ̀᷉ɛ̀᷉} \PS{adv.~} \GE{dull, sleepy} 
\LX{gbeɛle;} \HM{} \PH{gbɪ́ɛ́lɪ̀} \PS{v.~} \GE{to look at sternly} \PD{-, -, -, -, -, - } 
\LX{gbi} \HM{} \PH{gbí} \PS{v.~} \GE{to kneel} \PD{gbie, gbiire, gbiire, gbiribe, gbiiraa, gbiribe } 
\LX{gbi} \HM{} \PH{gbí} \PS{v.~} \GE{to break down a mud or earth-made roof} \PD{gbie, gbiire, gbiire, gbiribe, gbiiraa, gbiribe } 
\LX{gbi} \HM{} \PH{gbí} \PS{v.~} \GE{to harvest yams to make room for seed yams to grow} \PD{gbie, gbiire, gbiire, gbiribe, gbiiraa, gbiribe } 
\LX{gbie} \HM{} \PH{gbìé} \PS{n.~} \GE{forehead} \SG{gberuu} \OP{gberi} \TP{gberee} 
\LX{gbiele} \HM{} \PH{gbíélé} \PS{n.~} \SN \GE{hands and arms} \XV{N ba taa gbiele derɛ biiri.} \XE{My arms are too week to curdle children.} 
\LX{gbieli} \HM{} \PH{gbíélí} \PS{n.~} \GE{elbow} \OP{gbiele} \TP{gbielee} \VA{shortened form of kpaŋkpannyugbieli, kpaŋkpaŋgbieli}
\LX{gbieni} \HM{} \PH{gbìènì} \PS{v.~} \GE{to crawl in pain, to crawl painfully} \PD{gbienee, gbiene, gbiene, gbienime, gbienaa, gbienime } 
\LX{gbigiri} \HM{} \PH{gbígrì} \PS{v.~} \GE{to ruminate, to chew cud} \PD{gbigiree, gbigire, gbigire, gbigiribe, gbigiraa, gbigiribe } 
\LX{gbigiri} \HM{} \PH{gbígrì} \PS{v.~} \GE{to shake to test firmness e.g. a stem} \PD{gbigiree, gbigire, gbigire, gbigiribe, gbigiraa, gbigiribe } 
\LX{gbili} \HM{} \PH{gbìlí} \PS{adj.~} \GE{silent, quiet} \OP{gbile} \TP{gbilee} 
\LX{gbiŋ-gbiŋ} \HM{} \PH{gbîŋ-gbîŋ} \PS{adv.~} \GE{manner of running that is uninterrupted} 
\LX{gbini} \HM{} \PH{gbíní} \PS{n.~} \GE{bundle} \OP{gbime} 
\LX{gbiŋi} \HM{} \PH{gbíŋ} \PS{v.~} \GE{to cause to tremble or shake} \XV{Bibiiri deɛnɛ la a peɛ lɛnso ka o gbine.} \XE{Children are playing on the roof; that is why the roof is shaking.} \PD{gbiŋee, gbine, gbine, gbinime, gbinaa, gbinime } 
\LX{gbiŋi} \HM{} \PH{gbíŋ} \PS{v.~} \GE{to hypnotise; to put into utter shock such that the person or thing cannot react} \XV{A dɔɔta kyɔge o la pimbiri ka o deɛ gbiŋi.} \XE{The doctor gave him an injection and he was put out.} \PD{gbiŋee, gbine, gbine, gbinime, gbinaa, gbinime } 
\LX{gbireŋgbireŋ} \HM{} \PH{gbíréŋgbíréŋ} \PS{adv.~} \GE{solid and firm} 
\LX{gbiri} \HM{} \PH{gbírì} \PS{v.~} \SN \GE{to sleep} \SN \GE{to coagulate, to clot} \PD{gbiree, gbire, gbire, gbiribe, gbiraa, gbiribe } 
\LX{gbo} \HM{} \PH{gbô} \PS{n.~} \GE{heart} \SG{gboruu} \OP{gbori} \TP{gboree} 
\LX{gbo} \HM{} \PH{gbò} \PS{adv.~} \GE{a loud sound, usually made when two solid things collide, e.g two bulls in a fight when their heads crash} 
\LX{gbogbo} \HM{} \PH{gbògbò} \PS{adj.~} \GE{gluttonous} 
\LX{gbɔge} \HM{} \PH{gbɔ̀gɪ́} \PS{adj.~} \GE{not full e.g. sack of grain} \OP{gbɔgere} 
\LX{gbɔɡe} \HM{} \PH{gbɔ̀ɡɪ̀} \PS{v.~} \SN \GE{to be close to, to be next to} \XV{O zeŋ gbɔge la a baala.} \XE{He is seated close to the patient.} \SN \GE{to drink as a couple from one calabash at the same time (traditional kiss)} \PD{gbɔɡɛɛ, gbɔɡerɔ, gbɔɡerɔ, gbɔɡerebɔ, gbɔɡeraa, - } 
\LX{gbɔgele} \HM{} \PH{gbɔ̀glɪ̀} \PS{v.~} \GE{to hold too many things at a time} \PD{gbɔgelɛɛ, gbɔgelɔ, gbɔgelɔ, gbɔgelebɔ, gbɔgelaa, - } 
\LX{gbɔgele} \HM{} \PH{gbɔ̀glɪ̀} \PS{v.~} \GE{to assemble poorly e.g. broken-down machine} \PD{gbɔgelɛɛ, gbɔgelɔ, gbɔgelɔ, gbɔgelebɔ, gbɔgelaa, - } 
\LX{gbogili} \HM{} \PH{gbóglì} \PS{v.~} \SN \GE{to pack many objects on top of one another} \SN \GE{to stand up against each other (quarrel)} \PD{gbogilee, gbogilo, gbogilo, gbogilibo, gbogilaa, - } 
\LX{gbogiri} \HM{} \PH{gbògrì} \PS{v.~} \SN \GE{to gnaw flesh from a bone} \SN \GE{to lick a container clean e.g. a palatable soup pot} \PD{gbogree, gbogiro, gbogiro, gbogiribo, gbogiraa, - } 
\LX{gbɔgɔ} \HM{} \PH{gbɔ̀gɔ̀} \PS{n.~} \GE{a type of reptile similar to the monitor lizard} \OP{gbɔgre} \TP{gbɔgrɛɛ} 
\LX{gbɔgɔ} \HM{} \PH{gbɔ́gɔ́} \PS{n.~} \GE{strong open cage for carrying animals} \OP{gbɔgre} \TP{gbɔgrɛɛ} \VA{gbaga}
\LX{gbole} \HM{} \PH{gbʊ̀lɪ̀} \PS{v.~} \GE{to look up towards the sky} \XV{A naŋkpaana gbole kaa la a nuulee naŋ dɔɔ a teɛ zu.} \XE{The hunter looked up to see the bird in the tree.} \PD{gbolɛɛ, gbollɔ, gbollɔ, gboleba, gbollaa, gboleba } 
\LX{gboli} \HM{} \PH{gbòlí} \PS{adj.~} \GE{bare-headed (carrying a load on the head without a pad); without attachment} \OP{gbolo} \TP{gbolee} 
\LX{gboli} \HM{} \PH{gbólí} \PS{n.~} \GE{leather string worn around the waist} \SG{gbolee/gbollee} \OP{gbolo} \TP{gbollli} 
\LX{gboli} \HM{} \PH{gbólí} \PS{v.~} \GE{to search ernestly; to rummage} \PD{gbolee, gbolo, gbolo, gbolibo, gbolaa, gbolibo } 
\LX{gboligboli} \HM{} \PH{gbólgból} \PS{adj.~} \GE{sound of a loud echoing and persistent call} 
\LX{gbolli} \HM{} \PH{gbòllì} \PS{v.~} \GE{to harass; to pursue something or someone eagerly} \PD{gbollee, gbollo, gbollo, gbolbo, gbollaa, gbolibo } 
\LX{gbomo} \HM{} \PH{gbómò} \PS{n.~} \GE{lumps of matter used by a rodent to block its hole against intruders} \TP{gbonnee} 
\LX{gboŋe} \HM{} \PH{gbʊ́ŋ} \PS{v.~} \GE{to put pressure on (demanding action)} \PD{gboŋɛɛ, gbonnɔ, gbonnɔ, gbonema, gbonnaa, gbonema } 
\LX{gboŋgboli} \HM{} \PH{gbóŋgbólí} \PS{n.~} \GE{big rodent} \OP{gboŋgbolo} \TP{gboŋgbolee} 
\LX{gboŋgboŋ} \HM{} \PH{gbóŋgbóŋ} \PS{adj.~} \GE{all of, completely} 
\LX{gboŋgboŋ} \HM{} \PH{gbòŋgbôŋ} \PS{n.~} \GE{bucket; metal container} \OP{gboŋgbonni} \TP{gboŋgbonee} 
\LX{gboŋgbopelaa} \HM{} \PH{gbòŋgbòpɪ̀làá} \PS{n.~} \GE{jackal} \OP{gboŋgbopeɛle} \TP{gboŋgbopeɛlɛɛ} 
\LX{gboŋgbori} \HM{} \PH{gbòŋgbòrí} \PS{n.~} \GE{hyena} \OP{gboŋgboe} \TP{gboŋgboree} 
\LX{gboŋi} \HM{1} \PH{gbòŋ} \PS{v.~} \GE{to press against between two things.(finger between two objects)} \PD{gboŋee, gbonno, gbonno, gbonimo, gbonnaa, gbonimo } 
\LX{gboŋi} \HM{2} \PH{gbòŋ} \PS{v.~} \SN \GE{to pile up; to pack together} \SN \GE{to block an entrance (of a rodent's hole)} \PD{gboŋee, gbonno, gbonno, gbonimo, gbonnaa, gbonimo } 
\LX{gbontori} \HM{} \PH{gbòntòrí} \PS{n.~} \GE{ulcer} \OP{gbontoe} \TP{gbontoree} 
\LX{gbore} \HM{} \PH{gbʊ̀rɪ̀} \PS{v.~} \SN \GE{to hold water in the mouth} \SN \GE{to conceal a hole (of a rodent)} \PD{gborɛɛ, gborɔ, gborɔ, gboreba, gboraa, gboreba } 
\LX{gboregbore} \HM{} \PH{gbʊ̀rgbʊ̀rɪ́} \PS{v.~} \GE{to gargle} \PD{gborgborɛɛ, gborgborɔ, gborgborɔ, gborgboreba, gboregboraa, gborgboreba } 
\LX{gbori} \HM{} \PH{gbórí} \PS{v.~} \GE{to move suddenly and strongly, to struggle} \XV{Vũũ nyege la a bie ka o gbori  na tage yi a daa a doge lɔɔ.} \XE{The fire burnt the child and as he struggled to move away he pushed the pot over.} \PD{gboree, gboro, gboro, gboribo, gboraa, gboribo } \VA{gbori}
\LX{gbori} \HM{} \PH{gbòrí} \PS{n.~} \GE{ship, boat, canoe} \OP{gboe} \TP{gboree/gborree} 
\LX{gbori} \HM{} \PH{gbòrí} \PS{n.~} \GE{door flap} \OP{gboe} \TP{gboree/gborree} 
\LX{gbori} \HM{} \PH{gbòrí} \PS{n.~} \GE{trough} \OP{gboe} \TP{gboree/gborree} 
\LX{gbori} \HM{} \PH{gbòrí} \PS{n.~} \GE{board on which meat is dressed and cut} \OP{gboe} \TP{gboree/gborree} 
\LX{gboribiri} \HM{} \PH{gbòrbírì} \PS{n.~} \GE{key} \OP{gboribie} \TP{gboribiree/gboribiiree} 
\LX{gborigbori} \HM{} \PH{gbòrgbórí} \PS{v.~} \GE{to struggle, especially on the ground} \PD{gborgboree, gborgboro, gborgboro, gborgborbo, gborgboraa, - } 
\LX{gborikuŋkuni} \HM{} \PH{gbòrìkúŋkúní} \PS{n.~} \GE{padlock} \OP{saafemamine} \VA{saafema}
\LX{gborivuuri} \HM{} \PH{gbòrvúúrì} \PS{n.~} \GE{paddle} \OP{gborivue} \TP{gborivuuree} 
\LX{gborogboro} \HM{} \PH{gbòrògbòrò} \PS{adj.~} \GE{very large e.g. grains} 
\LX{gboroto} \HM{} \PH{gbórótó} \PS{adv.~} \GE{the manner of meeting face-to-face suddenly} 
\LX{gbotole} \HM{} \PH{gbʊ̀tʊ́lɪ́} \PS{n.~} \GE{bush baby (animal)} \OP{gbotola} \TP{gbotolɛɛ/gbotollɛɛ} 
\LX{gbugbulee} \HM{} \PH{gbúgbúlèè} \PS{n.~} \GE{tadpole} \OP{gbugbulli} \TP{gbugbullee} 
\LX{gbuli} \HM{} \PH{gbùlí} \PS{adj.~} \GE{whole} \OP{gbulo} \TP{gbulo gbulo} 
\LX{gbulli} \HM{} \PH{gbùllì} \PS{v.~} \SN \GE{to make into a ball} \SN \GE{to come together in large numbers} \XV{A pɔge mɔŋe la a fanfane baare a nyɔge gbulli gbulli sidi sidi.} \XE{When the woman finished making the soap she made it into balls of one cedi each.} \PD{-,  gbullo,  gbullo, gbulibo, gbullaa, gbulibo } 
\LX{gbutuli} \HM{} \PH{gbùtùlí} \PS{n.~} \GE{uninformed person; fool} \OP{gbutulo} \TP{gbutulee} 
\LX{gbuuli} \HM{} \PH{gbúúlí} \PS{v.~} \GE{see guuli (make a knot)} \PD{gbuulee, gbuulo, gbuulo, gbuulibo, gbuulaa, - } 
\LX{gɛbɛɛ} \HM{} \PH{gɛ̀bɛ́ɛ́} \PS{n.~} \GE{fool, lunatic} 
\LX{gegili} \HM{} \PH{gèglì} \PS{v.~} \GE{be half-closed} \PD{gegilee, gegile, gegile, gegilibe, gegilaa, gegilibe } \VA{ogili}
\LX{geli} \HM{} \PH{} \PS{v.~} \GE{to lift over something heavy lying flat on the ground} \XV{Ba geli la a kuri gbori tiili a dankyini poɔ.} \XE{They have lifted over the iron gate unto the wall.} \PD{gelee, gelle, gelle, gelibe, -, gelibe } 
\LX{geli} \HM{} \PH{} \PS{v.~} \GE{to burn badly (food)} \PD{gelee, gelle, gelle, gelibe, -, gelibe } 
\LX{geli} \HM{} \PH{} \PS{v.~} \GE{to be mad, to behave abnormally or foolishly} \PD{gelee, gelle, gelle, gelibe, -, gelibe } 
\LX{gelime} \HM{} \PH{gélímè} \PS{adj/n.~} \GE{insanity} 
\LX{gemee} \HM{} \PH{géméé} \PS{adj.~} \GE{mad, insane} 
\LX{genegene} \HM{} \PH{gènègènè} \PS{adv.~} \SN \GE{tinted (colour)} \XV{A boɔ e la zeɛ genegene.} \XE{The goat is tinted brown.} \SN \GE{not appearing clearly} \XV{Zie naŋ da wa nyaana genegene la ka o wa.} \XE{S/he came when dawn had not clearly set in.} 
\LX{geŋgeŋ} \HM{} \PH{géŋgéŋ} \PS{adj.~} \GE{firmly fixed} 
\LX{geŋi} \HM{} \PH{gêŋ} \PS{v.~} \GE{to stretch out to reach e.g. to look over a wall} \PD{geŋee, geŋine/genne, geŋine/genne, geŋinime/genime, geŋinaa/gennaa, genime } 
\LX{geŋi} \HM{} \PH{gêŋ} \PS{v.~} \GE{to be mad, to be angry} \PD{geŋee, geŋine/genne, geŋine/genne, geŋinime/genime, geŋinaa/gennaa, genime } 
\LX{geregere} \HM{} \PH{gèrègèrè} \PS{adv.~} \GE{moderately hard e.g. half-cooked yam} 
\LX{gieri} \HM{} \PH{gíérì} \PS{v.~} \GE{to belch} \PD{gieree, giere, giere, gieribe, gieraa, gieribe } 
\LX{go} \HM{} \PH{gó} \PS{v.~} \GE{to stoop, to bend down} \PD{goe, guoro, guoro, guoribo, goraa, guoribo } 
\LX{go} \HM{} \PH{gò} \PS{v.~} \GE{watch over} \PD{goe, goro, goro, goribo, goraa, - } \VA{gu}
\LX{gobaa} \HM{} \PH{gòbàà} \PS{n.~} \GE{shame, disgrace} 
\LX{gɔbaa} \HM{} \PH{gɔ̀báà} \PS{n.~} \SN \GE{skillful person; one full of dexterity} \SN \GE{artist (music, visual arts)} \OP{gɔɔre} \TP{gɔɔrɛɛ} 
\LX{gɔbenɛ} \HM{} \PH{gɔ́bɪ̀nɛ̀} \PS{n.~} \GE{governor, government} \OP{gɔbenne} \TP{gɔbenɛɛ} 
\LX{gɔbenɛnte} \HM{} \PH{gɔbɪnɛntɪ} \PS{n.~} \GE{government} \OP{gɔbɪnɛntɪrɪ} \TP{gɔbɪnɛntɪrɛɛ} 
\LX{gɔfuuraa} \HM{} \PH{gɔ́fúúráá} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{gɔfuuri} \TP{gɔfuuree} 
\LX{gɔge} \HM{} \PH{gɔ́gɪ́} \PS{v.~} \GE{to nod} \PD{gɔgɛɛ, gɔgerɔ, gɔgerɔ, gɔgerebɔ, gɔgraa, gɔgerebɔ } 
\LX{gogo} \HM{} \PH{gógó} \PS{n.~} \SN \GE{clock} \SN \GE{hour} \OP{gogori} \TP{gogoree} 
\LX{gɔgɔ} \HM{} \PH{gɔ́gɔ̀} \PS{n.~} \SN \GE{a type of tree whose bark is used in tanning} \SN \GE{to type of black bird with a long tail} \OP{gɔgɔre} \TP{gɔgɔrɛɛ} 
\LX{gɔgɔ} \HM{} \PH{gɔ̀gɔ́} \PS{v.~} \GE{to bend halfway; to stoop} \PD{gɔgɔɛ, gɔgɔɔrɔ, gɔgɔɔrɔ, gɔgɔrebɔ, gɔgɔraa, gɔgɔrebɔ } 
\LX{gogobiri} \HM{} \PH{gògòbírí} \PS{n.~} \GE{zebu (bull with a humped back)} \OP{gogobie} \TP{gogobiiree} 
\LX{gogoo} \HM{} \PH{gògóó} \PS{adj.~} \GE{standing out with long appendages e.g. pointing} 
\LX{gogoɔ} \HM{} \PH{gʊ̀gʊ̀ɔ́} \PS{n.~} \GE{sawfish} \OP{gogoɔre} \TP{gogoɔrɛɛ} 
\LX{golebɛlɛ} \HM{} \PH{gʊ̀lɪ̀bɛ̀lɛ̀} \PS{adj.~} \GE{lethargic and unsteady (very sick person)} 
\LX{goligoli} \HM{} \PH{gòlgólí} \PS{v.~} \GE{to scribble} \PD{goligolee, goligolo, goligolo, golgolibo, golgolaa, golgolibo } 
\LX{gollaa} \HM{} \PH{gólláà} \PS{n.~} \GE{a game in which players hop on one leg} \OP{golli} \TP{gollee} 
\LX{gɔllaa} \HM{} \PH{gɔ́lláá} \PS{n.~} \GE{a type of small, fatty field mouse} \OP{gɔlle} \TP{gɔllɛɛ} \VA{mɔɔtere}
\LX{golle} \HM{} \PH{gʊ̀llɪ̀} \PS{v.~} \GE{to mix up any which way (clothes)} \PD{gollɛɛ, gollɔ/golla, gollɔ/golla, golba, gollaa, - } 
\LX{gɔlle} \HM{} \PH{gɔ̀llɪ̀} \PS{v.~} \SN \GE{to go round, to detour} \SN \GE{to surround, to encircle} \PD{gɔllɛɛ, gɔllɔ, gɔllɔ, gɔlebɔ, gɔllaa, gɔlebɔ } 
\LX{golli} \HM{} \PH{góllì} \PS{v.~} \GE{to hop on one leg} \PD{gollee, gollo, gollo, golibo, gollaa, golibo } 
\LX{golɔgolɔ} \HM{} \PH{gʊ̊lɔ́gʊ́lɔ́} \PS{adj.~} \GE{soft and wavy (hair)} 
\LX{goloŋgoloŋ} \HM{} \PH{gòlòŋgólóŋ} \PS{adv.~} \GE{long and twisted} 
\LX{golɔnsomm} \HM{} \PH{gʊ̀lɔ̀nsʊ̀mm} \PS{adv.~} \GE{roundish and big} 
\LX{golɔnsomm} \HM{} \PH{gʊ́lɔ́nsʊ́mm} \PS{adv.~} \GE{roundish and small} 
\LX{gɔlɔnteebo} \HM{} \PH{gɔ́lɔ́ntɪ̀ɪ̀bʊ́} \PS{n.~} \GE{chameleon} \OP{gɔlɔnteere} \TP{gɔlɔnteerɛɛ} \VA{gɔmɔ/gɔmɔteebo}
\LX{gɔmberaa} \HM{} \PH{gɔ́mbíráá} \PS{n.~} \GE{state of sleeplessness} \OP{gɔmberre} \TP{gɔmberɛɛ} 
\LX{gombɔle} \HM{} \PH{gòmbɔ́lɪ́} \PS{n.~} \GE{kapok fruit} \OP{gombɔlɔ} \TP{gombɔlɛɛ} 
\LX{gɔmbulo} \HM{} \PH{gɔ̀mbùlò} \PS{n.~} \GE{unique sound in poetry, sound that is pleasing to the ear} 
\LX{gɔmɔ} \HM{} \PH{gɔ́mɔ̀} \PS{n.~} \GE{noise, disturbance} \OP{gɔnne} \TP{gɔnnɛɛ} \VA{gɔnne}
\LX{gomɔgomɔ} \HM{} \PH{gʊ̀mɔ̀gʊ̀mɔ̀} \PS{adv.~} \GE{not straight forward} \VA{gomagoma}
\LX{gɔnda} \HM{} \PH{gɔ̌ndà} \PS{n.~} \GE{pawpaw, papaya} \SG{gɔndaroo} \OP{gɔndare} \TP{gɔndarɛɛ} 
\LX{gondoŋ} \HM{} \PH{gòndóŋ} \PS{n.~} \GE{skirt} \OP{gondonni} \TP{gondonnee} 
\LX{gone} \HM{} \PH{gʊ́nɪ́} \PS{n.~} \GE{structural flaw} \OP{gonne} \TP{gonɛɛ/gonnɛɛ} 
\LX{gɔŋe} \HM{} \PH{gɔ́ŋ} \PS{adj.~} \GE{crooked, bent} 
\LX{gɔŋe} \HM{} \PH{gɔ̀ŋ} \PS{v.~} \GE{to make crooked, to incline toward} \PD{gɔŋɛɛ, gɔnnɔ, gɔnnɔ, gɔnema, gɔnnaa, gɔnema } 
\LX{gɔŋe} \HM{} \PH{gɔ́ŋ} \PS{v.~} \GE{to make noise, to quarrel} \PD{gɔŋɛɛ, gɔnnɔ, gɔnɔ, gɔnemɔ, gɔnnaa, gɔnemɔ } 
\LX{gɔŋ-eŋ} \HM{} \PH{gɔ̀ŋ-ɪ̀ŋ} \PS{v.~} \GE{to redirect or state to suit one's purpose} \XV{A yɛlɛ ba waa lɛ a pɔge gɔŋ-eŋɛɛ la.} \XE{That is not the true state of the matter; the woman has tried to add something to make it look like that.} \PD{gɔŋ-eŋɛɛ, gɔnnɔ-ennɛ, gɔŋ-ennɛ, gɔŋ-ennaa, gɔŋ-enemɛ, - } 
\LX{gɔŋgbire} \HM{} \PH{gɔ́ŋgbírè} \PS{n.~} \GE{one who is asleep} \OP{gɔŋgbiribe} \TP{gɔŋgbiribee} 
\LX{gɔŋgɔlɔŋ} \HM{} \PH{gɔ̀ŋgɔ̀lɔ̀ŋ} \PS{adj.~} \GE{long and winding} 
\LX{gɔŋgɔŋe} \HM{} \PH{gɔ̀ŋgɔ́ŋ} \PS{v.~} \GE{to walk wearily and unsteadily; to trudge} \PD{gɔŋgɔŋɛɛ, gɔŋgɔnɔ, gɔŋgɔnɔ, gɔŋgɔnema, gɔŋgɔnaa, gɔŋgɔnema } 
\LX{goŋmane} \HM{} \PH{góŋmánɪ́} \PS{n.~} \GE{one hundred kola nuts} \OP{goŋmama} \TP{goŋmannɛɛ} 
\LX{gonnaabo} \HM{} \PH{gónnáábʊ́} \PS{n.~} \GE{plant louse} \OP{gonnii} \TP{gonniinee} 
\LX{gɔnne} \HM{} \PH{gɔ́nnɪ̀} \PS{n.~} \GE{noise} \OP{gɔmɔ} \TP{gɔnnɛɛ} 
\LX{gonni} \HM{} \PH{gònní} \PS{n.~} \GE{cotton} \SG{gonnuu} \OP{n/a} \TP{gonnee} 
\LX{goŋo} \HM{} \PH{gòŋó} \PS{n.~} \GE{cotton tree} \OP{gonni} \TP{gonnee} 
\LX{gɔŋɔ} \HM{} \PH{gɔ́ŋɔ́} \PS{n.~} \GE{type of guinea corn} \TP{gɔnnɛɛ} 
\LX{gonseɛ} \HM{} \PH{gònsɪ̀ɛ́} \PS{n.~} \GE{fall; season of intermitent wind} 
\LX{gɔntɛre} \HM{} \PH{gɔ̀ntɛ́rɪ́} \PS{v.~} \GE{to talk incessantly} \PD{gɔntɛrɛɛ, gɔnɔtɛrɛ, gɔnɔtɛrɛ, gɔntɛrebɛ, -, - } 
\LX{gɔntɛre} \HM{} \PH{gɔ̀ntɛ́rɪ́} \PS{v.~} \GE{to burst out, especially when angry} \PD{gɔntɛrɛɛ, gɔnɔtɛrɛ, gɔnɔtɛrɛ, gɔntɛrebɛ, -, - } 
\LX{gɔnzuuro} \HM{} \PH{gɔ́nzúúrò} \PS{n.~} \GE{sleepiness} \XV{O taa la gɔnzuuro.} \XE{He/she is feeling sleepy.} \TP{gɔnzuuree} 
\LX{goɔ} \HM{} \PH{gʊ́ɔ̀} \PS{n.~} \GE{thorn} \OP{goore} \TP{goorɛɛ} 
\LX{goɔ} \HM{} \PH{gʊ̀ɔ̀} \PS{v.~} \GE{to bend at the trunk} \XV{Pɔgeba maŋ goɔ la kaabo kyɛ ŋma.} \XE{Women usually hold down guinea corn to cut the head.} \PD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ } 
\LX{goɔ} \HM{} \PH{gʊ̀ɔ̀} \PS{v.~} \GE{to prevent from progressing; to cause to return or refrain from something} \PD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ } 
\LX{goɔ} \HM{} \PH{gʊ̀ɔ̀} \PS{v.~} \GE{to bite (snake)} \PD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ } 
\LX{gɔɔ} \HM{} \PH{gɔ̀ɔ́} \PS{n.~} \GE{left hand, left side} \OP{gɔɔe} \TP{gɔɔrɛɛ} 
\LX{gɔɔ} \HM{} \PH{gɔ̀ɔ́} \PS{adj.~} \GE{left-handed} \OP{gɔɔe} \TP{gɔɔrɛɛ} 
\LX{gɔɔ} \HM{} \PH{gɔ̀ɔ̀} \PS{v.~} \GE{to go round and round} \PD{gɔɔɛ, gɔɔrɔ, gɔɔrɔ, gɔɔrebɔ, gɔɔraa, gɔɔrebɔ } \VA{goɔre}
\LX{goole} \HM{} \PH{gʊ̀ʊ̀lɪ̀} \PS{v.~} \GE{to mix, to stir (thick paste)} \PD{goolɛɛ, goolɔ, goolɔ, gooleba, goolaa, gooleba } 
\LX{goone} \HM{} \PH{gʊ̀ʊ̀nì} \PS{v.~} \GE{to walk majestically with an up-and-down body movement} \XV{A daanedann da maŋe la a naa niŋŋ kyɛ goone iri.} \XE{The praise singer would walk majestically with up-and-down-movement of the body in front of the King.} \PD{goonɛɛ, goonɔ, goonɔ, goonema, goonaa, goonema } 
\LX{goɔraa} \HM{} \PH{gʊ̀ɔ̀ràá} \PS{n.~} \GE{a type of small tree with medicinal properties} \OP{goɔre} \TP{goɔrɛɛ} 
\LX{goɔre} \HM{} \PH{gʊ́ɔ́rɪ́} \PS{v.~} \GE{to walk or run half stooping} \PD{goɔrɛɛ, goɔrɔ, goɔrɔ, goɔreba, goɔraa, goɔreba } 
\LX{goɔre} \HM{} \PH{gʊ̀ɔ̀rɪ̀} \PS{v.~} \SN \GE{to go round and round} \SN \GE{to splash mashed ashes round the outer wall of a house (ritual to rid the house of activities of witchraft)} \PD{goɔrɛɛ, goɔrɔ, goɔrɔ, goɔreba, goɔraa, - } 
\LX{gɔɔrombiri} \HM{} \PH{gɔ́ɔ́rʊ́mbírí} \PS{n.~} \GE{literary device (lit.)} \SG{gɔɔrombiruu} \OP{gɔɔrombie} \TP{gɔɔrombiree} 
\LX{gɔpelaa} \HM{} \PH{gɔ̀pɪ̀làá} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{gɔpeɛle} \TP{gɔpeɛlɛɛ} 
\LX{gora} \HM{} \PH{gʊ́rà} \PS{n.~} \SN \GE{hunchback} \SN \GE{disease of the back} \OP{gorre} \TP{gorrɛɛ} \VA{gorɔ}
\LX{goraa} \HM{} \PH{gòràà} \PS{adj.~} \GE{curved} 
\LX{goraa} \HM{} \PH{gòràà} \PS{adj.~} \GE{ashamed} 
\LX{gore} \HM{} \PH{gʊ̀rɪ̀} \PS{v.~} \SN \GE{to be abundant} \SN \GE{to be overburdened with weight} \PD{gorɛɛ, gora, gora, goreba, goraa, - } 
\LX{gore} \HM{} \PH{gʊ́rɪ́} \PS{v.~} \GE{to pluck feathers or fur} \PD{gorɛɛ, gora, gora, gorba, goraa, - } 
\LX{gɔre} \HM{} \PH{gɔ̀rɪ̀} \PS{v.~} \SN \GE{to bend} \SN \GE{to mould a hoe into a curved shape} \PD{gɔrɛɛ, gɔrɔ, gɔrɔ, gɔrebɔ, gɔraa, gɔrebɔ } 
\LX{gɔre} \HM{} \PH{gɔ́rɪ́} \PS{n.~} \GE{scythe} \OP{gɔrre} \TP{gɔrɛɛ} 
\LX{gorgore} \HM{} \PH{gʊ̀rgʊ́rɪ́} \PS{v.~} \GE{to ruffle} \PD{gorgorɛɛ, gorgora, gorgora, gorgorba, gorgoraa, gorgorba } 
\LX{gori} \HM{} \PH{górí} \PS{n.~} \GE{dowry, bride price} \OP{gorri} \TP{goree} 
\LX{gori} \HM{} \PH{gòrì} \PS{v.~} \GE{to bend an object (into a sharp curve)} \PD{goree, goro, goro, gorbo, goraa, gorbo } 
\LX{gori} \HM{} \PH{górí} \PS{n.~} \GE{big pile of; load} \OP{goe} \TP{goree} 
\LX{gori-gori-gori} \HM{} \PH{gór-gór-gór} \PS{adv.~} \GE{motionless} 
\LX{gɔsɔgelaa} \HM{} \PH{gɔ́sɔ́glàá} \PS{n.~} \GE{a type of thorn tree} \OP{gɔsɔgelɔ} \TP{gɔsɔgelɛɛ} 
\LX{gɔteɛ} \HM{} \PH{gɔ̂tɪ̀ɛ́} \PS{n.~} \GE{a tree with thorns} \OP{gɔteere} \TP{gɔteerɛɛ} 
\LX{gozambuuli} \HM{} \PH{gòzàmbúúlí} \PS{n.~} \GE{the fruit of a type of thorn tree} \OP{gozambuulo} \TP{gozambuulee} \VA{gɔzambuuli}
\LX{gozaŋaa} \HM{} \PH{gózàŋàà} \PS{n.~} \GE{a type of thorn tree with yellow fruit} \OP{gozanne} \TP{gozannɛɛ} \VA{gɔzaŋaa}
\LX{gɔzeɛ} \HM{} \PH{gɔ̀zɪ̀ɛ́} \PS{n.~} \GE{a type of tree with medicinal properties} 
\LX{go᷉ɔ} \HM{} \PH{gʊ́᷉ɔ́} \PS{n.~} \SN \GE{kiln (place for baking clay items to make them hard)} \SN \GE{arena, grouping point (ko-go᷉ɔ)} \OP{goɔne} \TP{goɔnɛɛ} 
\LX{go᷉ɔo} \HM{} \PH{gʊ́᷉ɔ̀ʊ́} \PS{n.~} \GE{sleep} \OP{goɔnɛɛ} 
\LX{gu} \HM{} \PH{gù} \PS{v.~} \SN \GE{to watch out for} \SN \GE{to look after} \SN \GE{to protect, to defend} \SN \GE{to wait for (someone),} \SN \GE{to waylay (someone)} \PD{gue, guuro, guuro, guribo, guuraa, guribo } \VA{go}\SE{gu zɔɔ} \GE{to ambush} 
\LX{gu zɔɔ} \HM{} \PH{gù zɔ̀ɔ́} \PS{v.~} \GE{to ambush} \PD{gue  zɔɔ, guuro zɔɔ, zɔɔ guuro, -, -, - } 
\LX{gufu} \HM{} \PH{gúfù} \PS{n.~} \GE{interpreter} \OP{gufuri} \TP{gufuree/gufurree} 
\LX{Gumo} \HM{} \PH{gúmó} \PS{n.~} \GE{a town in the Sisaalaland noted for soothsaying and divining} \TP{gumee} 
\LX{gunni} \HM{} \PH{gúnní} \PS{v.~} \GE{to stalk (hunter tracking game)} \PD{gunnee, gunno, gunno, gunimo, gunnaa, - } 
\LX{gunni} \HM{} \PH{gùnnì} \PS{v.~} \GE{to pick tiny leaves from plants} \PD{gunnee, gunno, gunno, gunimo, gunnaa, - } 
\LX{guo} \HM{} \PH{gùò} \PS{n.~} \GE{heavy club with knobbed end} \OP{gori} \TP{goree} \VA{daguo}
\LX{guoli} \HM{} \PH{gùòlì} \PS{v.~} \GE{to coil (rope)} \PD{guolee, guolo, guolo, guolibo, guolaa, guolibo } 
\LX{guoli} \HM{} \PH{gúólì} \PS{v.~} \SN \GE{to bring up (child)} \SN \GE{to raise (animals)} \PD{guolee, guolo, guolo, guolibo, guolaa, guolibo } 
\LX{guoli} \HM{} \PH{gúólí} \PS{n.~} \GE{pot rest; pad on which pots are placed} \OP{guolo} \TP{guolee} 
\LX{guori} \HM{} \PH{gúórí} \PS{n.~} \GE{kola nuts} \SG{guoruu} \OP{goe} \TP{goree} 
\LX{guri} \HM{} \PH{gùri} \PS{v.~} \SN \GE{to throw one's body onto another (wrestling, catching a fowl)} \SN \GE{to fold up (leg)} \PD{guree, guro, guro, guribo, guraa, - } 
\LX{guri pɔge} \HM{} \PH{gùrì pɔ́gɪ́} \PS{v.~} \GE{to rape} \PD{-, -, -, -, -, - } 
\LX{gurri} \HM{} \PH{gúrrí} \PS{n.~} \GE{mushroom} \SG{gurruu} \TP{gurree} 
\LX{guruŋ} \HM{} \PH{gùrúŋ} \PS{n.~} \GE{mushroom} \SG{gurruu} \OP{gurri} \TP{gurree} 
\LX{guu} \HM{} \PH{gúú} \PS{v.~} \GE{to be unable, to fail} \PD{guue, guuro, guuro, guuribo, guuraa, guuribo } 
\LX{guu} \HM{} \PH{gùú} \PS{adj.~} \GE{blunt, not sharp} \OP{guuri} \TP{guuree} 
\LX{guu} \HM{} \PH{gùù} \PS{v.~} \GE{to be blunt} \PD{guue, guuro, guuro, guurbo, guuraa, guurbo } 
\LX{guuli} \HM{} \PH{gúúlì} \PS{v.~} \GE{to delay someone} \PD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \VA{gbuuli}
\LX{guuli} \HM{} \PH{gúúlì} \PS{v.~} \GE{to be not yet healed of a sprain} \PD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \VA{gbuuli}
\LX{guuli} \HM{} \PH{gúúlì} \PS{v.~} \GE{to delay someone} \PD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \VA{gbuuli}
\LX{guuli} \HM{} \PH{gúúlí} \PS{n.~} \GE{knot} \OP{guulo} \TP{guulee} 
\LX{gyaaŋkoli} \HM{} \PH{dzàáŋkólí} \PS{n.~} \GE{the sacrum and coccyx} \OP{gyaaŋkolo} \TP{gyaaŋkolee} \VA{gyikoli}
\LX{gyaare} \HM{} \PH{dzààrɪ̀} \PS{v.~} \GE{to rake, to sweep roughly} \PD{gyaarɛɛ, gyaara, gyaara, gyaareba, gyaaraa, gyaareba } 
\LX{gyaaroo} \HM{} \PH{dzáárʊ́ʊ́} \PS{n.~} \GE{a type of wild mouse} \OP{gyaara} \TP{gyaarɛɛ} 
\LX{Gyabuni} \HM{} \PH{dzàbùnì} \PS{n.~} \GE{a type of idol with magical powers among the Dagaaba} \OP{gyabunni/gyabumo} \TP{gyabunee} 
\LX{gyabuno} \HM{} \PH{dzàbùnò} \PS{n.~} \GE{fetish} \OP{gyabunni/gyabumo} \TP{gyabunnee/gyabumee} \VA{gyabuni}
\LX{gyaga} \HM{} \PH{dzàgà} \PS{adj.~} \GE{useless} 
\LX{gyage} \HM{} \PH{dzàgɪ̀} \PS{v.~} \SN \GE{to face a huge task} \SN \GE{jack} \PD{gyagɛɛ, gyagera, gyagera, gyagereba, gyageraa, gyagereba } 
\LX{gyagele} \HM{} \PH{dzàglɪ̀} \PS{v.~} \GE{to hang on to, to support lightly} \PD{gyaglɛɛ, gyagela, gyagela, gyageleba, gyagelaa, gyageleba } 
\LX{gyammo} \HM{} \PH{dzàmmʊ́} \PS{n.~} \GE{clitoris} \OP{gyama} \TP{gyamɛɛ/gyannɛɛ} \VA{gyàmbírì,gyambie,gyambiiree}
\LX{gyampegere} \HM{} \PH{dzàmpɪ́grɪ́} \PS{n.~} \GE{comb (of a rooster)} \OP{gyampegɛ} \TP{gyampegɛɛ/gyampegerɛɛ} \VA{gbammo}
\LX{gyaŋe} \HM{} \PH{dzáŋ} \PS{n.~} \GE{see gyɛŋ} 
\LX{gyaŋgbaa} \HM{} \PH{dzáŋgbàá} \PS{n.~} \GE{something that is acidic; usually harmful} \OP{gyaŋgbaare} \TP{gyaŋgbaarɛɛ} 
\LX{gyaŋgbeɛ} \HM{} \PH{dzáŋgbɪ́ɛ́} \PS{n.~} \GE{small child} \OP{gyaŋgbeere} \TP{gyaŋgbeerɛɛ} 
\LX{gyaŋgbɔgɔ} \HM{} \PH{dzǎŋgbɔ̀gɔ̀} \PS{n.~} \GE{a type of monitor} \OP{gyaŋgbɔgre} \TP{gyaŋgbɔgrɛɛ} 
\LX{gyaŋgboŋi} \HM{} \PH{dzàŋgbòŋ} \PS{n.~} \GE{house mouse} \OP{gyaŋgbonni} \TP{gyaŋgbonnee} \VA{gyoŋgboŋi}
\LX{gyaŋgboro} \HM{} \PH{dzàŋgbòrò} \PS{n.~} \GE{fried dough made from the leaves of a small plant} \TP{gyaŋgboree} 
\LX{gyaŋgboro} \HM{} \PH{dzáŋgbóró} \PS{n.~} \GE{a type of small plant with edible leaves} \SG{gyaŋgboruu} \TP{gyaŋgboree} 
\LX{gyangyagera} \HM{} \PH{dzàndzágrá} \PS{n.~} \GE{one who behaves abnormally; insane person} \SG{gyangyagerebɛɛ} \TP{gyangyagereba} 
\LX{gyangyagera} \HM{} \PH{dzàndzágrá} \PS{n.~} \GE{fried corn or bean flour mixed with vegetables} \TP{gyangyagrɛɛ} 
\LX{gyarema} \HM{} \PH{dzàrɪ̀má} \PS{n.~} \GE{a type of yam farm where sticks with branches are used to support the vine of the yam} \TP{gyaremɛɛ} 
\LX{gyarima} \HM{} \PH{dzárɪ́má} \PS{n.~} \GE{a type of tree with medicinal properties} \TP{gyaremɛɛ} 
\LX{gya᷉ayo} \HM{} \PH{dza᷉ayo} \PS{n.~} \GE{thanks} \OP{gya᷉ayori} \TP{gya᷉ayoree} 
\LX{gyɛ} \HM{} \PH{dzɛ́} \PS{v.~} \SN \GE{to reject, to refuse to accept (in protest or annoyance)} \SN \GE{to abandon} \PD{gyɛɛ, gyeɛrɛ, gyeɛrɛ, gyeɛrebɛ, gyeɛraa, gyeɛrebɛ } 
\LX{gyee} \HM{} \PH{dzɪ̀ɪ̀} \PS{v.~} \GE{short form of 'gyeere' to move away from; to change environment} \PD{gyeeɛ, gyeerɛ, gyeerɛ, gyeerebɛ, -, - } 
\LX{Gyɛɛmaan} \HM{} \PH{dzɛ́ɛ́máán} \PS{n.~} \GE{Germany} \TP{gyɛɛmanɛɛ} 
\LX{gyeɛne} \HM{} \PH{dzɪ̀ɛ̀nɪ̀} \PS{v.~} \GE{to lurk; wishing to do something} \XV{A noɔ gyeɛnɛ la a kyi.} \XE{The fowl is lurking around the millet.} \PD{gyeɛnɛɛ, gyeɛnɛ, gyeɛnɛ, gyeɛnemɛ, gyeɛnaa, gyeɛnemɛ } 
\LX{gyeɛne} \HM{} \PH{dzɪ́ɛ́nɪ́} \PS{n.~} \GE{see gyɛngyeɛne (protrusion)} \OP{gyeɛmɛ} \TP{gyeɛmɛɛ} 
\LX{gyeɛraa} \HM{} \PH{dzɪ̀ɛ̀ràà} \PS{n.~} \GE{plume, long feather (of a rooster)} \OP{gyeɛre} \TP{gyeɛrɛɛ} 
\LX{gyeere} \HM{} \PH{dzɪ̀ɪ̀rɪ̀} \PS{v.~} \GE{to circle around from front to back of an object} \PD{gyeerɛɛ, gyeerɛ, gyeerɛ, gyeerebɛ, gyeeraa, gyeerebɛ } 
\LX{gyefa} \HM{} \PH{dzɪ́fà} \PS{n.~} \GE{pocket} \OP{gyefare} \TP{gyefarɛɛ} 
\LX{gyefa} \HM{} \PH{dzɪ́fà} \PS{n.~} \GE{unclean thing} \OP{gyefare} \TP{gyefarɛɛ} 
\LX{gyege} \HM{} \PH{dzɪ̀gɪ̀} \PS{v.~} \GE{to be fully grown, to be strong, to be powerful} \PD{gyegɛɛ, gyegerɛ, gyegerɛ, gyegerebɛ, gyegeraa, gyegerebɛ } 
\LX{gyegele} \HM{} \PH{dzɪ́glɪ́} \PS{n.~} \GE{a spiritually powerful person} \OP{gyegelle} \TP{gyegelɛɛ} 
\LX{gyegere} \HM{} \PH{dzɪ̀grɪ̀} \PS{v.~} \GE{to display one's supernatural power, to challenge} \PD{gyegerɛɛ, gyegerɛ, gyegerɛ, gyegerebɛ, gyegeraa, gyegerebɛ } 
\LX{gyegiri} \HM{} \PH{dzégrí} \PS{v.~} \GE{to mix by placing things in between others} \PD{gyegiree, gyegire, gyegire, gyegiribe, gyegiraa, gyegiribe } 
\LX{gyegiri} \HM{} \PH{dzégrí} \PS{v.~} \GE{to counter someone's speech; to challenge} \PD{gyegiree, gyegire, gyegire, gyegiribe, gyegiraa, gyegiribe } 
\LX{gyelaa} \HM{} \PH{dzɪ̀làá} \PS{n.~} \GE{carnivorous animal} \OP{gyelaare} \TP{gyelaarɛɛ} 
\LX{Gyɛle} \HM{} \PH{dzɛ̀lɨ̀} \PS{n.~} \SN \GE{a stream in Kaane} \SN \GE{a native named after this stream} \OP{gyɛlɛ} \TP{gyɛlɛɛ} 
\LX{gyɛle} \HM{} \PH{dzɛ́lɪ́} \PS{v.~} \SN \GE{to rummage; to look through searching} \SN \GE{to analyse} \PD{gyɛlɛɛ, gyɛlɛ, gyɛlɛ, gyɛlebɛ, gyɛlaa, gyɛlebɛ } 
\LX{gyɛle} \HM{} \PH{dzɛ́lɪ̀} \PS{n.~} \GE{egg} \SG{gyɛlee} \OP{gyɛlɛ} \TP{gyɛllɛɛ} 
\LX{gyɛle} \HM{} \PH{dzɛ̀lɪ́} \PS{n.~} \GE{thought (used with bare---) not giving thought; absent minded} \XV{A porisi bare la gyɛle ka a nannyige zo.} \XE{The police did not pay attention, so the thief escaped.} 
\LX{gyɛlemɛ} \HM{} \PH{dzɛ̀lɪ̀mɛ́} \PS{n.~} \GE{smallpox} \TP{gyɛlmɛɛ} 
\LX{gyɛleŋmɛne} \HM{} \PH{dzɛ́lɪ́ŋmɛ́nɪ́} \PS{n.~} \SN \GE{snail} \SN \GE{shellfish} \OP{gyɛleŋmɛmɛ} \TP{gyɛleŋmɛmɛɛ} 
\LX{gyell} \HM{} \PH{dzɪ̀ll} \PS{adj.~} \GE{protuding e.g. arrow stuck in an animal} 
\LX{gyɛlpɛge} \HM{} \PH{dzɛ́lpɛ́gɪ̀} \PS{n.~} \GE{eggshell} \OP{gyɛlpɛgre/gyɛlpɛgɛ} \TP{gyɛlpɛgrɛɛ} 
\LX{gyɛlzizeɛ} \HM{} \PH{dzɛ̀lzìzɪ̀ɛ́} \PS{n.~} \GE{yolk} \OP{gyɛlzeere} \TP{gyɛlzeerɛɛ} 
\LX{gyemɛ} \HM{} \PH{dzɪ̀mɛ̀} \PS{n.~} \SN \GE{convulsion, disease of children} \SN \GE{short form of "da᷉a᷉gyemɛ"} \TP{gyemɛɛ} 
\LX{gyɛmɛgyɛmɛ} \HM{} \PH{dzɛ́mɛ́dzɛ́mɛ́} \PS{adj.~} \GE{slowly and unnoticed, jokingly} \XV{A kɔlaa e la gyɛmɛgyɛmɛlɛ kyɛ ɔɔ a nɛne zaa.} \XE{The cat ate all the meat without anyone noticing it.} 
\LX{gyena} \HM{} \PH{dzɪ́nà} \PS{n.~} \GE{fairy} \OP{gyenne} \TP{gyennɛɛ} 
\LX{gyɛndaa} \HM{} \PH{dzɛ̀ndàá} \PS{n.~} \GE{spindle} \OP{gyɛndaare} \TP{gyɛndarɛɛ} 
\LX{gyɛŋe} \HM{} \PH{dzɛ̌ŋ} \PS{adj.~} \GE{extensive; deep} \XV{A naŋkpaana kpɛ la mɔ-gyɛŋe.} \XE{The hunter has gone into the deep bush.} \OP{gyɛnne} \TP{gyɛnnɛɛ} \VA{gyaŋ}
\LX{gyɛŋe} \HM{} \PH{dzɛ́ŋ} \PS{v.~} \SN \GE{to move unsteadily} \SN \GE{to move aside} \PD{gyɛŋɛɛ, gyɛŋenɛ/gyɛnnɛ, gyɛŋenɛ/gyɛnnɛ, gyɛŋenaa/gyɛnnaa, gyɛŋenemɛ/gyɛnemɛ, - } 
\LX{gyɛŋe} \HM{} \PH{dzɛ̀ŋ} \PS{n.~} \GE{thought, mind} \OP{gyɛnne} \TP{gyɛnnɛɛ} \VA{gyaŋe}
\LX{gyɛnee} \HM{} \PH{dzɛ̀nɪ́ɪ́} \PS{n.~} \GE{groundnut, peanut (Southern Dagaare dialect)} \TP{gyɛnnɛɛ} 
\LX{gyɛŋgɛlɛ} \HM{} \PH{dzɛ́ŋgɛ́lɛ́} \PS{n.~} \GE{measles} \TP{gyɛŋgɛlɛɛ} 
\LX{gyɛngyeɛne} \HM{} \PH{dzɛ̀ndzɪ́ɛ́nɪ́} \PS{n.~} \GE{protrusion, part sticking out} \OP{gyɛngyeɛmɛ} \TP{gyɛngyeɛmɛɛ} 
\LX{gyɛngyɛŋ} \HM{} \PH{dzɛ́ndzɛ́ŋ} \PS{n.~} \GE{circle arrangement for advancement. (war)} 
\LX{gyɛŋgyɛŋ} \HM{} \PH{dzɛ̌ŋdzɛ̌ŋ} \PS{n.~} \GE{mind, thought} 
\LX{gyɛngyɛre} \HM{} \PH{dzɛ̀ngyɛ̀rɪ́} \PS{n.~} \GE{a climbing plant} \OP{gyɛngyɛrre} \TP{gyɛngyɛrɛɛ} 
\LX{gyenne} \HM{} \PH{dzɪ̀nnɪ̀} \PS{v.~} \GE{to test} \PD{gyennɛɛ, gyennɛ, gyennɛ, gyenemɛ, gyennaa, gyenemɛ } 
\LX{gyenne} \HM{} \PH{dzɪ̀nnɪ̀} \PS{v.~} \GE{to waylay, to plan evil against} \PD{gyennɛɛ, gyennɛ, gyennɛ, gyenemɛ, gyennaa, gyenemɛ } 
\LX{gyɛnne} \HM{} \PH{dzɛ̀nnɪ́} \PS{v.~} \GE{please, beg} \XV{Gyɛnne, ɔŋe kõɔ ko ma ka N nyu.} \XE{Please, give me some water to drink.or I beg you, give me some water to drink.} \PD{-, -, -, -, -, - } 
\LX{gyennoo} \HM{} \PH{dzɪ́nnʊ́ʊ́} \PS{n.~} \GE{test, examination} \TP{gyennɛɛ} 
\LX{gyɛnterɛ} \HM{} \PH{dzɛ́ntɪ́rɛ̀} \PS{n.~} \SN \GE{prostitute} \SN \GE{not sticking to one thing, jumping from one thing to another} \OP{gyɛnterre} \TP{gyɛnterrɛɛ} 
\LX{gyeraa} \HM{} \PH{dzɪ̀ráà} \PS{n.~} \GE{woven grass bracelet worn on the wrist} \OP{ gyerre} \TP{gyerrɛɛ} 
\LX{gyeraa} \HM{} \PH{dzɪ́ráá} \PS{adj.~} \GE{short (distance)} 
\LX{gyere} \HM{} \PH{dzɪ́rɪ́} \PS{v.~} \SN \GE{to cross between (two objects)} \SN \GE{to stumble} \PD{gyerɛɛ, gyerɛ, gyerɛ, gyerebɛ, gyeraa, gyerebɛ } 
\LX{gyere} \HM{} \PH{dzɪ́rɪ̀} \PS{v.~} \GE{to subside (of rain)} \PD{gyerɛɛ, gyerɛ, -, -, -, - } \VA{yoɔ}
\LX{yoɔ} \HM{} \PH{yʊ̀ɔ̀} \PS{v.~} \GE{to stop raining} \PD{-, -, -, -, -, - } \VA{gyere}
\LX{gyere} \HM{} \PH{dzɪ̀rɪ̀} \PS{v.~} \GE{to nock; to place across e.g. arrow on the bow} \PD{gyerɛɛ, gyerɛ, gyerɛ, gyerebɛ, gyeraa, gyerebɛ } 
\LX{gyere} \HM{} \PH{dzɪ̀rɪ̀} \PS{v.~} \GE{to pierce through to the other side} \PD{gyerɛɛ, gyerɛ, gyerɛ, gyerebɛ, gyeraa, gyerebɛ } 
\LX{gyere} \HM{} \PH{dzɪ́rɪ́} \PS{n.~} \GE{portion unattended to (part of farm)} \OP{gyerre} \TP{gyerɛɛ/gyerrɛɛ} 
\LX{gyɛre} \HM{} \PH{dzɛ́rɪ́} \PS{v.~} \SN \GE{to be mixed up (different types)} \SN \GE{to be confused, to be disturbed (mad)} \SN \GE{to not be in good health} \PD{gyɛrɛɛ, gyɛrɛ, gyɛrɛ, gyɛrbɛ, gyɛraa, gyɛrbɛ } 
\LX{gyɛre} \HM{} \PH{dzɛ̀rɪ̀} \PS{v.~} \GE{to stand firm, to not fall e.g. after a stumble} \PD{gyɛrɛɛ, gyɛrɛ, gyɛrɛ, gyɛrebɛ, gyɛraa, gyɛrebɛ } 
\LX{Gyerɛbaa} \HM{} \PH{dzɪ̀rɛ̀bàá} \PS{n.~} \GE{Jirapa (the main town in Central Dagao)} 
\LX{gyerɛgyerɛ} \HM{} \PH{dzɪ́rɛ́dzɪ̀rɛ̀} \PS{n.~} \GE{a type of skin disease} 
\LX{gyeremɛ} \HM{} \PH{dzɪ́rɪ́mɛ́} \PS{n.~} \SN \GE{respect, honour, homage} \SN \GE{fifty fifty (gambling)} \TP{gyeremɛɛ} 
\LX{gyeremiri} \HM{} \PH{dzɪ́rɪ́mírí} \PS{n.~} \GE{rope made from a type of grass (kantɛrɛɛ)} \OP{gyeremie} \TP{gyeremiiree/gyeremiree} 
\LX{gyeremuri} \HM{} \PH{dzɪ̀rìmúrí} \PS{n.~} \GE{a type of river grass stalk} \OP{gyeremue} \TP{gyeremuree} 
\LX{gyerre} \HM{} \PH{dzɪ̀rrɪ̀} \PS{v.~} \GE{used with ...taa; to crisscross; lying across one another} \XV{A bibiiri le gyerre la taa.} \XE{The children fell crisscrossing each other.} \XV{A teere le gyerre la taa.} \XE{The trees fell across one another.} \PD{gyerrɛɛ, gyerrɛ, gyerrɛ, gyerbɛ, gyerraa, - } 
\LX{gyieli} \HM{} \PH{dzìèlì} \PS{v.~} \SN \GE{to note, to examine} \SN \GE{to count} \PD{gyielee, gyiele, gyiele, gyielbe, gyielaa, - } 
\LX{gyieri} \HM{} \PH{dzìèrì} \PS{v.~} \GE{to grind into coarse powder} \PD{gyieree, gyiere, gyiere, gyieribe, gyieraa, - } 
\LX{gyiili} \HM{} \PH{dzííĺí} \PS{n.~} \SN \GE{tendon} \SN \GE{artery} \SN \GE{nerve} \SG{gyiiluu} \OP{gyiile} \TP{gyiilee} 
\LX{gyikoli} \HM{} \PH{dzìkólí} \PS{n.~} \GE{part of body} \OP{gyikolo} \TP{gyikolee/gyikollee} \VA{gyaaŋkoli}
\LX{gyildaa} \HM{} \PH{dzìldáá} \PS{n.~} \GE{less soudning one of the pair of xylophones} \OP{gyidaare} \TP{gyildaarɛɛ} 
\LX{gyildoɔle} \HM{} \PH{dzìldʊ́ɔ́lɪ́} \PS{n.~} \GE{xylophone mallet} \SG{gyildoɔloo} \OP{gyildoɔlɔ} \TP{gyildoɔlɛɛ} 
\LX{gyili} \HM{} \PH{dzìlí} \PS{n.~} \GE{xylophone} \OP{gyile} \TP{gyilee/gyillee} 
\LX{gyili} \HM{} \PH{dzílí} \PS{v.~} \GE{to surround, to go behind} \XV{Teere are gyili la a bulee.} \XE{The well is surrounded by trees.} \PD{gyilee, gyile, gyile, gyilibe, gyilaa, gyilibe } 
\LX{gyilkoɔre} \HM{} \PH{dzìlkʊ́ɔ́rɪ̀} \PS{n.~} \GE{gourd used in making the xylophone} \OP{gyilkɔɛ} \TP{gyilkɔrɛɛ} 
\LX{gyilnyaa} \HM{} \PH{dzìlɲáà} \PS{n.~} \GE{the base chord area of the xylophone} \OP{gyilnyaane} \TP{gyilnyaanɛɛ} 
\LX{gyilnyaŋaa} \HM{} \PH{dzìlɲáŋáá} \PS{n.~} \GE{the more sounding between a pair of xylophones} \OP{gyilnyanne} \TP{gyilnyannɛɛ} 
\LX{gyiri} \HM{} \PH{dzírí} \PS{n.~} \GE{cordlike talisman worn on the waist, believed to have supernatural power} \OP{gyirri} \TP{gyiree} 
\LX{gyiri} \HM{} \PH{dzírí} \PS{v.~} \GE{to bring about a change that makes life difficult} \XV{A zie gyiri ka donɛɛ kpeɛbo e kpeɛŋaa.} \XE{Society has declined and life has become difficult.} \PD{gyiree, gyire, gyire, gyirbe, gyiraa, - } 
\LX{gyirimuri} \HM{} \PH{dzìrìmúrí} \PS{n.~} \GE{sulphur} \OP{gyirimue} \TP{gyirimuree} 
\LX{gyoboraa} \HM{} \PH{dzʊ̀bʊ̀ràà} \PS{n.~} \GE{carnivore, e.g. a leopard} \OP{gyoborre.} \TP{gyoborɛɛ/gyoborrɛɛ} 
\LX{gyoge} \HM{} \PH{dzʊ̀gɪ̀} \PS{v.~} \SN \GE{to pull mouth or lips in front to show anger. (typically used with noɔre)} \SN \GE{to weed badly} \PD{gyogrɛɛ, gyogra, gyogra, gyogreba, gyograa, - } \VA{gyogere}
\LX{gyogere} \HM{} \PH{dzʊ̀grɪ̀} \PS{v.~} \GE{to clear growth badly (barbing or hoeing)} \PD{gyogerɛɛ, gyogera, gyogera, gyogereba, gyogeraa, gyogereba } \VA{gyoge}
\LX{gyogili} \HM{} \PH{dzòglì} \PS{v.~} \GE{to hold on to lightly and gently} \PD{gyogilee, gyogilo, gyogilo, gyogilibo, gyogilaa, gyogilibo } 
\LX{gyogo} \HM{} \PH{dzógó} \PS{n.~} \GE{behaviour, character} \OP{gyogiri} \TP{gyogiree} 
\LX{gyoll} \HM{} \PH{dzʊ̀ll} \PS{adv.~} \GE{appearing alone} 
\LX{gyolom} \HM{} \PH{dzólóm} \PS{adv.~} \GE{quietly} \VA{gyoloŋ, gyolomm}
\LX{gyolomm} \HM{} \PH{dzolomm} \PS{adv.~} \GE{quietly and slowly (on tiptoe)} \VA{gyolom}
\LX{gyoŋene} \HM{} \PH{dzʊ̀ŋɪ̀nɪ̀} \PS{v.~} \GE{to make unlively, to make dull} \PD{gyoŋenɛɛ, gyoŋena, gyoŋena, gyoŋenema, gyoŋenaa, gyoŋenema } \VA{gyoone}
\LX{gyoŋgboŋi} \HM{} \PH{dzòŋgbǒŋ} \PS{n.~} \GE{house mouse} \OP{gyoŋgbonni} \TP{gyoŋgbonee} \VA{gyaŋgboŋi}
\LX{gyoŋgoroŋ} \HM{} \PH{dzóŋgóróŋ} \PS{adj.~} \GE{a little higher or taller than} 
\LX{gyongyolingyo} \HM{} \PH{dzǒngyólǐndzó} \PS{adv.~} \GE{squat squarely, with knees fully bent} 
\LX{gyongyoŋ} \HM{} \PH{dzòngyóŋ} \PS{v.~} \GE{to move on tiptoe} \PD{gyongyoŋee, gyongyono, gyongyono, gyongyonimo, gyongyonaa, - } \VA{gyɔngyɔŋ}
\LX{gyɔŋgyɔŋ} \HM{} \PH{dzɔ̀ŋdzɔ́ŋ} \PS{n.~} \GE{to move haphazardly on one's toes} \VA{gyongyoŋ}
\LX{gyoone} \HM{} \PH{dzʊ̀ʊ̀nɪ̀} \PS{v.~} \GE{to become inactive} \PD{gyoonɛɛ, gyoona, gyoona, gyoonema, gyoonaa, - } \VA{gyoŋene}
\LX{gyore} \HM{} \PH{dzʊ́rɪ́} \PS{v.~} \GE{to work at or make badly, to spoil e.g. haircut} \PD{gyorɛɛ, gyora, gyora, gyoreba, gyoraa, gyoreba } 
\LX{gyɔreba} \HM{} \PH{dzɔ́rɪ́bá} \PS{n.~} \GE{long, loose dress worn by women; tunic} \OP{gyɔrebare} \TP{gyɔrebarɛɛ} 
\LX{gyɔ᷉ɔ} \HM{} \PH{dzɔ̀᷉ɔ̀} \PS{adv.~} \GE{tall and lanky} \VA{gyɔgyɔ᷉ɔ}
\LX{gyuuni} \HM{} \PH{dzùùnì} \PS{n.~} \GE{vulture} \OP{gyuumo} \TP{gyuunee} 
\LX{ha} \HM{} \PH{há} \PS{v.~} \GE{to hire} \PD{haɛ, haara, haara, hareba, haaraa, - } 
\LX{haa} \HM{} \PH{háà} \PS{v.~} \GE{to open by stretching out two ends e.g. lips} \PD{haaɛ, haara, haara, haareba, haaraa, haareba } 
\LX{haa (cd)} \HM{} \PH{hàá} \PS{adv.~} \GE{all} \VA{zaa}
\LX{haale} \HM{} \PH{háalì́} \PS{adv.~} \GE{even, extremely} \VA{haali}
\LX{hạạli} \HM{} \PH{hạ́ạ́lì} \PS{adv.~} \GE{ greatly, very much} 
\LX{hạạli} \HM{} \PH{hạ́ạ́lì} \PS{conj.~} \GE{even if} \XV{Haali ka a bieŋ wa kyɛ te koŋ o a gane.} \XE{Even if the child comes we will not give him/her the book.} 
\LX{haanaa} \HM{} \PH{háánáá} \PS{n.~} \GE{blackberry} \OP{haane} \TP{haanɛɛ} 
\LX{haare} \HM{} \PH{háárɪ̀} \PS{v.~} \GE{to shout at to frighten or drive away} \PD{haarɛɛ, haara, haara, haareba, haaraa, haareba } 
\LX{hage} \HM{} \PH{hhágɪ́} \PS{v.~} \SN \GE{tight} \XV{A dɔɔ hage la o pɔge nyuu taa lɛ.} \XE{The man held tight to his wife’s neck.} \SN \GE{tighten} \XV{Ka fooŋ wa nyɔge foŋ hage taa lɛ.} \XE{When you get hold of it tighten your grip.} \PD{hagɛɛ, hagera, hagera, hagereba, hageraa, hagereba } 
\LX{ha-haa} \HM{} \PH{hà-háá} \PS{adv.~} \GE{wide open} 
\LX{hai} \HM{} \PH{hháí} \PS{interjection.~} \GE{so?} 
\LX{hale} \HM{} \PH{hálɪ̀} \PS{v.~} \GE{to smell strongly and unpleasantly, to reek} \PD{halɛɛ, hala, halla, haleba, halaa, haleba } 
\LX{hale} \HM{} \PH{hálɪ̀} \PS{v.~} \GE{to speak in anger} \PD{halɛɛ, hala, halla, haleba, halaa, haleba } 
\LX{halhale} \HM{} \PH{hàlhálɪ́} \PS{v.~} \GE{to strive with, to struggle} \PD{halhalɛɛ, halhala, halhala, halhaleba, halhalaa, - } 
\LX{haligba} \HM{} \PH{hálìgbà} \PS{conjunction.~} \GE{even} 
\LX{hama} \HM{} \PH{hámà} \PS{n.~} \GE{hammer} \OP{hamare} \TP{hamarɛɛ} 
\LX{hama} \HM{} \PH{hàmá} \PS{adv.~} \GE{plenty} 
\LX{hamama} \HM{} \PH{hàmàmà} \PS{adj.~} \GE{resounding (singing)} 
\LX{hamamma} \HM{} \PH{hàmàmmà} \PS{adj.~} \GE{loudly and continuously} 
\LX{hamm} \HM{} \PH{hàmm} \PS{adv.~} \GE{large} 
\LX{haŋe} \HM{} \PH{hháŋ} \PS{v.~} \SN \GE{to suffer} \SN \GE{to perspire, sweat (nd)} \PD{haŋɛɛ, hana, hana, hanema, hanaa, hanema } 
\LX{hankyere} \HM{} \PH{hánʧɪ́rɪ́} \PS{n.~} \GE{handkerchief} \OP{hankyeɛ} \TP{hankyerɛɛ} 
\LX{hap (nd)} \HM{} \PH{hháp} \PS{v.~} \GE{to clutch tightly} \PD{habɛɛ, hapra, hapra, hapreba, hapraa, hapreba } 
\LX{harahara} \HM{} \PH{hháráhhárá} \PS{adv.~} \GE{rough} 
\LX{hare} \HM{} \PH{hàrɪ̀} \PS{v.~} \GE{of the stomach (overwhelmed by hunger)} \PD{hanrɛɛ,, hara, hara, hareba, haraa, hareba } 
\LX{hɛɛ} \HM{} \PH{hhɛ́ɛ̀} \PS{interjection.~} \GE{caution, be careful} 
\LX{hɛɛ} \HM{} \PH{hɛ́ɛ́} \PS{adv.~} \GE{overtaken completely e.g. sleep} 
\LX{hɛge} \HM{} \PH{hɛ̀gɪ̀} \PS{v.~} \GE{to increase e.g. fire} \PD{hɛgɛɛ, hɛgerɛ, hɛgerɛ, hɛgerebɛ, hɛgeraa, hɛgerebɛ } 
\LX{hɛgɛkɛ} \HM{} \PH{hɛ̀gɛ̀kɛ̀} \PS{adv.~} \GE{large and red e.g. bonfire} 
\LX{hɛgele} \HM{} \PH{hhɛ́glɪ́} \PS{v.~} \GE{to cramp} \PD{hɛgelɛɛ, hɛgelɛ, hɛgelɛ, hɛgelebɛ, hɛgelaa, hɛgelebɛ } 
\LX{hɛgelɛgɛ} \HM{} \PH{hhɛ́gɪ́lɛ́gɛ́} \PS{adv.~} \GE{tired and helpless} \VA{sɛgelɛgɛ}
\LX{hɛgere} \HM{} \PH{hhɛ́grɪ̀} \PS{v.~} \GE{to pant e.g. dog} \PD{hɛgerɛɛ, hɛgerɛ, hɛgerɛ, hɛgerebɛ, hɛgeraa, hɛgerebɛ } 
\LX{helli} \HM{} \PH{hhéllí} \PS{v.~} \GE{sharp shout e.g. during hearty laughter; the sharp cry of a dog when hit} \PD{hellee, helle, helle, helibe, hellaa, - } 
\LX{hɛmɛmmɛ} \HM{} \PH{hɛ̀mɛ̀mmɛ̀} \PS{adj.~} \GE{very sweet} 
\LX{heŋi} \HM{} \PH{hhéŋ} \PS{v.~} \GE{to draw in one's stomach} \PD{heŋee, heŋine/henne, heŋine/henne, henime, hennaa, henime } 
\LX{hɛrɛ} \HM{} \PH{hɛ̀rɛ̀} \PS{adv.~} \GE{manner of taking a big sip of fluid} 
\LX{heremm} \HM{} \PH{hérèmm} \PS{adj.~} \GE{very tight} 
\LX{hieri} \HM{} \PH{hhíérì} \PS{v.~} \GE{to yawn} \PD{hieree, hiere, hiere, hieribe, hieraa, hieribe } 
\LX{hii} \HM{} \PH{hì᷉ì} \PS{interj.~} \GE{expression of sorrow (sobbing)} 
\LX{hiini} \HM{} \PH{hììnì} \PS{v.~} \GE{to cry earnestly, to wail, to sob, to moan} \PD{hiinee, hiine, hiine, hiinime, hiinaa, hiinime } 
\LX{hiri} \HM{} \PH{hìrì} \PS{v.~} \GE{to get to the climax} \PD{hiree, hire, -, -, -, - } 
\LX{hiri} \HM{} \PH{hìrì} \PS{v.~} \GE{to become hot} \PD{hiree, hire, -, -, -, - } 
\LX{hiri} \HM{} \PH{hìrì} \PS{v.~} \GE{to become serious} \PD{hiree, hire, -, -, -, - } 
\LX{hogoko} \HM{} \PH{hògòkò} \PS{adv.~} \GE{intensely red} 
\LX{holi} \HM{} \PH{hólí} \PS{n.~} \GE{triangular blade} \OP{holo} \TP{hollee} 
\LX{hoo} \HM{} \PH{hòò} \PS{adv.~} \GE{open and endless} 
\LX{hoo} \HM{} \PH{hóó} \PS{interj.~} \GE{hooting} 
\LX{hoone} \HM{} \PH{hʊ̀ʊ̀nɪ̀} \PS{v.~} \GE{to growl} \PD{hoonɛɛ, hoonɔ, hoonɔ, honema, hoonaa, honema } 
\LX{hɔɔpɛmpɛlɛŋ} \HM{} \PH{hɔ̀ɔ̀pɛ́mpɛ́lɛ́ŋ} \PS{n.~} \GE{spleen} \OP{hɔɔpɛmpɛlle} \TP{hɔɔpɛmpɛlɛɛ} \VA{zagepɛmpɛlɛŋ}
\LX{ho᷉o} \HM{} \PH{hʊ̀᷉ʊ̀} \PS{interj.~} \GE{expression of dispair, depression} 
\LX{hunihuni} \HM{} \PH{hùnìhúní} \PS{v.~} \GE{to speak undertone} \PD{hunihunee, hunihuno, hunihuno, hunihunimo, -, - } \VA{vunnii}
\LX{hunihuni} \HM{} \PH{hùnìhúní} \PS{v.~} \GE{to gossip} \PD{hunihunee, hunihuno, hunihuno, hunihunimo, -, - } \VA{vunnii}
\LX{hunni} \HM{} \PH{hhúnnì} \PS{v.~} \GE{whisper into the ear, gossip} \PD{hunnee, hunno, hunno, hunimo, hunnaaa, - } 
\LX{huuni} \HM{} \PH{hhúúní} \PS{v.~} \SN \GE{to hum (dispirited)} \SN \GE{of a dog's expression of a familiar person; welcome} \PD{huunee, huuno, huuno, huunimo, huunaa, - } 
\LX{iini} \HM{} \PH{ììnì} \PS{v.~} \GE{to get charged up} \XV{Ba to la a dɔɔ ka o iini gbie kyɛ sɔrɔ yɛlɛ.} \XE{They insulted the man and he got charged up and started grumbling.} \PD{-, -, -, -, -, - } 
\LX{ingyi} \HM{} \PH{índzì} \PS{n.~} \GE{engine} \OP{ingyime} \TP{ingyinnee} 
\LX{inni} \HM{} \PH{ìnnì} \PS{v.~} \GE{to forget} \PD{innee, inne, inne, inime, innaa, - } \VA{yinni}
\LX{iree} \HM{} \PH{ìréè} \PS{n.~} \GE{duiker (small antelope)} \OP{irri} \TP{irree} 
\LX{iri} \HM{} \PH{ìrì} \PS{v.~} \GE{to rise, to stand up} \PD{iree, ire, ire, iribe, iraa, iribe } 
\LX{iri} \HM{} \PH{írì} \PS{v.~} \SN \GE{to choose, to pick} \SN \GE{to take out (from a container)} \SN \GE{to subtract, to take away} \SN \GE{to create} \PD{iree, ire, ire, iribe, iraa, iribe } 
\LX{iri noɔre} \HM{} \PH{írí nʊ́ɔ́rɪ̀} \PS{v.~} \SN \GE{to answer, to respond (to a question)} \SN \GE{to explain} \SN \GE{to fulfil a promise} \PD{iree noɔre, ire noɔre, -, -, -, - } 
\LX{iri pɔge} \HM{} \PH{írì pɔ́gɪ̀} \PS{v.~} \GE{to elope with a woman} \PD{-, -, -, -, -, - } 
\LX{iri suuri} \HM{} \PH{ìrì súúrí} \PS{v.~} \GE{to annoy} \PD{iree suuri, ire suuri, -, -, -, - } 
\LX{iri yuori} \HM{} \PH{írí yúórì} \PS{v.~} \GE{to praise} \PD{iree yuori, ire yuori, -, -, -, - } 
\LX{iribare} \HM{} \PH{i̊́rì bàrɪ̀} \PS{n.~} \GE{subtraction} 
\LX{iriŋ} \HM{} \PH{íríŋ} \PS{n.~} \GE{kind, race, species} \TP{irree} 
\LX{iriŋ-iriŋ} \HM{} \PH{írìŋ-írìŋ} \PS{adv.~} \GE{of all kinds, types} \VA{iruŋ-iruŋ}
\LX{irizeɛ} \HM{} \PH{ìrìzìɛ̀} \PS{n.~} \GE{a type of small but fast and clever antelope} \OP{irizeere} \TP{irizeerɛɛ} 
\LX{ka} \HM{} \PH{ká} \PS{rel. pron.~} \GE{that} \XV{O yeli ka ba gɛrɛ.} \XE{S/he said s/he would not go.} 
\LX{ka} \HM{} \PH{kà} \PS{conj.~} \SN \GE{ if} \XV{Ka saaŋ mi ba na bore la.} \XE{If it rains, they will sow their crops.} \SN \GE{and} \XV{O maale la diibu de wa ko te.} \XE{S/he cooked and brought it to us.} 
\LX{ka} \HM{} \PH{ká} \PS{v.~} \GE{to bail out (water)} \PD{kaɛ, kaara, kaara, kaareba, kaaraa, - } 
\LX{kaa} \HM{} \PH{káà} \PS{v.~} \GE{to curse} \PD{kaaɛ, kaara, kaara, kaareba, kaaraa, - } \VA{kaale}
\LX{kaa} \HM{} \PH{káà} \PS{v.~} \GE{to invoke} \PD{kaaɛ, kaara, kaara, kaareba, kaaraa, - } \VA{kaale}
\LX{kaa} \HM{} \PH{kàà} \PS{v.~} \SN \GE{to look, to watch} \SN \GE{to take care of} \PD{kaaɛ, kaara, kaara, kaareba, kaaraa, - } 
\LX{kaa ti᷉i} \HM{} \PH{kàà tí᷉í} \PS{v.~} \GE{look closely} \PD{kaaɛ ti᷉i, kaara ti᷉i, -, -, -, - } 
\LX{kaabanyɛ} \HM{} \PH{káábányɛ́} \PS{n.~} \GE{direspect} 
\LX{kaabo} \HM{} \PH{káábʊ̀} \PS{n.~} \GE{millet plant} \OP{kyi} \TP{kyiree} 
\LX{kaale} \HM{} \PH{káálɪ́} \PS{v.~} \SN \GE{to invoke} \SN \GE{to perform a libation} \PD{kaalɛɛ, kaala, kaala, kaaleba, kaalaa, kaaleba } \VA{kaa}
\LX{kaale} \HM{} \PH{kààlɪ̀} \PS{v.~} \GE{to gather e.g. soil, grains} \PD{kaalɛɛ, kaala, kaala, kaaleba, kaalaa, kaaleba } 
\LX{kaalɔɔ} \HM{} \PH{káálɔ̀ɔ́} \PS{adv.~} \GE{little thing eaten instantly} 
\LX{kaama} \HM{} \PH{káámá} \PS{n.~} \GE{lively state} \TP{kaamɛɛ} 
\LX{kaamaako} \HM{} \PH{káámáákò} \PS{n.~} \GE{persistence} 
\LX{kaapɔge} \HM{} \PH{kàápɔ́gɪ̀} \PS{adv.~} \GE{maybe, perhaps} \XV{Saa koe la kaapɔge o na mi la.} \XE{Rain clouds have gathered, it may rain.} 
\LX{kaaraa} \HM{} \PH{kááráá} \PS{n.~} \GE{lumpy paste made from bean flour} \OP{kaarɛɛ} 
\LX{kaare} \HM{} \PH{káárɪ̀} \PS{n.~} \GE{car} \OP{kaɛ/kɛɛ} \TP{kaarɛɛ} 
\LX{kaare} \HM{} \PH{kààrɪ̀} \PS{v.~} \GE{to drag e.g. feet on the ground} \PD{kaarɛɛ, kaara, kaara, kaareba, kaaraa, kaareba } 
\LX{kaare} \HM{1} \PH{káárɪ́} \PS{v.~} \GE{to clear one's throat; to cough up (phlegm)} \PD{kaarɛɛ, kaara, kaara, kaareba, kaaraa, kaareba } 
\LX{kaare} \HM{2} \PH{káárɪ́} \PS{v.~} \SN \GE{to move away} \SN \GE{to move away immediately, not to return (typically used with gaba)} \XV{Ba iri la kparoo ko a bie ka o deɛ kaare gaba.} \XE{A shirt was given to the boy and he left at once.} \PD{kaarɛɛ, kaara, kaara, kaareba, kaaraa, kaareba } 
\LX{kaarebiri} \HM{} \PH{kàárɪ̀bírí} \PS{n.~} \GE{sputum} \SG{kaarebiruu} \OP{kaarebie} \TP{kaarebiree/kaarebiiree} 
\LX{kaaterɛ} \HM{} \PH{káàtɪ̀rɛ́} \PS{n.~} \GE{cursed one} \OP{kaaterre} \TP{kaaterɛɛ} 
\LX{kabiri} \HM{} \PH{kábírì} \PS{n.~} \GE{millet grain} \SG{kabiruu} \OP{kabie} \TP{kabiiree} 
\LX{kabooroo} \HM{} \PH{kábʊ́ʊ́rʊ́ʊ́} \PS{n.~} \GE{seed millet} \OP{kaboora} \TP{kaboorɛɛ} 
\LX{kabɔrɔ} \HM{} \PH{kábɔ́rɔ̀} \PS{n.~} \GE{fresh, mature guinea corn} \TP{kabɔrɛɛ} 
\LX{kabulo} \HM{} \PH{kábúló} \PS{n.~} \GE{sown sprouts of millet} \TP{kabulee} 
\LX{kadaa} \HM{} \PH{[kádàá]} \PS{n.~} \SN \GE{millet stalk} \SN \GE{long stick for threshing millet} \OP{kadaare} \TP{kadarɛɛ/kadaarɛɛ} 
\LX{kadaara} \HM{} \PH{kádààrà} \PS{n.~} \GE{millet or guinea corn not that is properly matured} \TP{kadaarɛɛ} 
\LX{kadori} \HM{} \PH{kádórí} \PS{n.~} \GE{heap of millet} \OP{kadorri} \TP{kadoree} 
\LX{kafini} \HM{} \PH{kàfìní} \PS{n.~} \GE{lance} \OP{kafime} \TP{kafinnee} 
\LX{kagbin} \HM{} \PH{kágbín} \PS{n.~} \GE{ears of corn or millet tied together} \OP{ kagbime} \TP{kagbinnee} 
\LX{kagbulo} \HM{} \PH{kágbùlò} \PS{n.~} \GE{unbroken millet seeds} \TP{kagbulee} 
\LX{kage} \HM{} \PH{kàgɪ̀} \PS{v.~} \GE{to hold firmly on to} \PD{kagɛɛ, kagera, kagera, kagereba, kageraa, kagereba } 
\LX{kage} \HM{} \PH{kágɪ́} \PS{adv.~} \GE{stay intently} 
\LX{kagelaŋ} \HM{} \PH{kàglàŋ} \PS{adv.~} \GE{large and long} 
\LX{kagere} \HM{} \PH{kágrɪ̀} \PS{v.~} \GE{to beckon, to touch to call the attention of} \PD{kagerɛɛ, kagera, kagera, kagereba, kageraa, kagereba } 
\LX{kagere} \HM{} \PH{kágrɪ̀} \PS{v.~} \GE{to bite or cut lightly} \PD{kagerɛɛ, kagera, kagera, kagereba, kageraa, kagereba } 
\LX{kagesaga} \HM{} \PH{kágɪ́ságá} \PS{adv.~} \GE{small and dry} 
\LX{kagesaga} \HM{} \PH{kàgɪ̀sàgà} \PS{adv.~} \GE{large and dry} 
\LX{kaka} \HM{} \PH{kàkà} \PS{n.~} \GE{hippopotamus} \OP{kakare} \TP{kakarɛɛ} 
\LX{kakaduro} \HM{} \PH{kákádúró} \PS{n.~} \GE{ginger} \SG{kakaduruu} \OP{kakadurri} \TP{kakaduree} 
\LX{kakalaa} \HM{} \PH{kákáláá} \PS{n.~} \GE{a type of tree; has extremely hard wood} \OP{ kakalle} \TP{kakalɛɛ} 
\LX{kakpennɛ} \HM{} \PH{kákpɪ́nnɛ̀} \PS{n.~} \GE{untidy millet} \TP{kakpennɛɛ} 
\LX{Kakube} \HM{} \PH{kákúbé} \PS{n.~} \GE{a farming festival of the Nandom people} \TP{kakubee} 
\LX{kakubo} \HM{} \PH{kákúbò} \PS{n.~} \GE{millet mixed with stones} \TP{kakubee} 
\LX{kakumo} \HM{} \PH{kǎkúmò} \PS{n.~} \GE{donkey} \OP{kakunni} \TP{kakunnee} 
\LX{kakyeerɛ} \HM{} \PH{káʧɪ́ɪ́rɛ́} \PS{n.~} \GE{second-grade millet} \TP{kakyeerɛɛ} 
\LX{kakyeɛrɛ} \HM{} \PH{káʧɪ́ɛ́rɛ́} \PS{n.~} \GE{one who harvests millet} \OP{kakyeɛrebɛ} \TP{kakyeɛrebɛɛ} 
\LX{kakyo} \HM{} \PH{káʧó} \PS{n.~} \GE{basket strainer} \OP{ kakyori} \TP{kakyoree} 
\LX{Kala} \HM{} \PH{kàlà} \PS{n.~} \GE{a type of fetish} 
\LX{kala} \HM{} \PH{kàlà} \PS{n.~} \GE{dog} \OP{kalle} \TP{kalɛɛ/kallɛɛ} 
\LX{kalakala} \HM{} \PH{kàlàkàlà} \PS{adv.~} \GE{slightly bitter} 
\LX{kalbiri} \HM{} \PH{kalbiri} \PS{n.~} \GE{dawdaw spice (seed)} \SG{kalbiruu} \OP{kalbie} \TP{kalbire/kalbiiree} 
\LX{kale} \HM{} \PH{kálɪ́} \PS{n.~} \GE{suffering, woe as a result of another pesrson's action} \TP{kalɛɛ} 
\LX{kale} \HM{} \PH{kálɪ̀} \PS{n.~} \GE{spice made from dawdaw seed} \SG{kaloo} \TP{kallɛɛ} 
\LX{kale} \HM{} \PH{kálɪ̀} \PS{v.~} \GE{to scoop water in small amounts e.g. washing hands} \PD{kalɛɛ, kalla, kalla, kaleba, kallaa, kaleba } 
\LX{kale} \HM{} \PH{kálɪ̀} \PS{v.~} \GE{to make a short comment on another person’s speech} \PD{kalɛɛ, kalla, kalla, kaleba, kallaa, kaleba } 
\LX{kalekale} \HM{} \PH{kákál} \PS{adv.~} \GE{foul-smelling} 
\LX{kalekale} \HM{} \PH{kàlkàlɪ̀} \PS{v.~} \GE{to move in a jerky manner} \PD{kalekalɛɛ, kalekala, kalkala, kalekaleba, kalekalaa, kalekaleba } 
\LX{kaleŋkaa} \HM{} \PH{kálɪ̀ŋkáá} \PS{n.~} \GE{gall} \OP{kaleŋkaare} \TP{kalen.kaarɛɛ} \VA{yamm}
\LX{kaleŋkaloɔraa} \HM{} \PH{kálɪ́ŋkálʊ́ɔ́ráá} \PS{n.~} \GE{a round smooth stone used in grinding} \OP{kaleŋkaloɔre} \TP{kaleŋkaloɔrɛɛ} \VA{kaleŋkoɔraa}
\LX{kaleŋkawerɛ} \HM{} \PH{kálɪ́ŋkàwɪ̀rɛ̀} \PS{n.~} \GE{a type of wasp} \SG{kaleŋkaweroo} \TP{kaleŋkawerɛɛ} 
\LX{kaleŋkoɔraa} \HM{} \PH{kálɪ́ŋkʊ́ɔ̀ráá} \PS{n.~} \GE{round stone} \OP{kaleŋkoɔre} \TP{kaleŋkoɔrɛɛ} \VA{kaleŋkaloɔraa}
\LX{kaleŋmeɛ} \HM{} \PH{kálŋmɪ́ɛ̀} \PS{n.~} \GE{a type of bird} \OP{kaleŋmeɛne} \TP{kaleŋmeɛnɛɛ} \VA{kareŋmeɛ}
\LX{kalenyaa} \HM{} \PH{kàlɲàá} \PS{n.~} \GE{a type of spiky grass that causes skin iritation} \OP{kalenyaane} \TP{kalenyaanɛɛ} 
\LX{kalere} \HM{} \PH{kàlɪ́rɪ́} \PS{n.~} \GE{oath in which one gives away one's food if one fails to twist a finger on another while eating. (children)} \OP{kalerre} \TP{kalerɛɛ} 
\LX{kalinfuoro} \HM{} \PH{kálínfúóró} \PS{n.~} \GE{bit of the sign} \TP{kalinfuoree} 
\LX{kalpigiraa} \HM{} \PH{kálpígráà} \PS{n.~} \GE{ball of dawdaw spice} \OP{kalpigiri.} 
\LX{kalsoro} \HM{} \PH{kálesóró} \PS{n.~} \GE{water used to wash dawdaw seed for spice} \TP{kalesoree} 
\LX{kama} \HM{} \PH{kámá} \PS{adj.~} \GE{close eyes shortly} \XV{O poɔ zaa᷉ bɛ o la tensoga ka o daŋ gaabo kyɛ o nimri zaa᷉ ba kyɛ kama.} \XE{Last night, s/he had an upset stomch and went to bed early, but s/he did not get a wink of sleep.} 
\LX{kama} \HM{} \PH{kàmà} \PS{adj.~} \GE{cut sharply} 
\LX{kamaabiri} \HM{} \PH{kámáábírì} \PS{n.~} \GE{kernel (of corn, maize)} \SG{kamaabiruu} \OP{kamaabie} \TP{kamaabiiree} 
\LX{kamaadaa} \HM{} \PH{kámáádàá} \PS{n.~} \GE{maize stalk} \OP{kamaadaare} \TP{kamaadaarɛɛ} 
\LX{kamaadakɔɔre} \HM{} \PH{kámáádàkɔ́ɔ́rɪ́} \PS{n.~} \GE{corncob} \OP{kamaadakɔbɔ} \TP{kamaadakɔɔrɛɛ} 
\LX{kamaalɔnzeere} \HM{} \PH{kámáálɔ̌nzɪ́ɪ́rɪ́} \PS{n.~} \GE{corn silk} \OP{kamaalɔnzeerɛɛ} 
\LX{kamaane} \HM{} \PH{kàmáánɪ̀} \PS{n.~} \GE{maize} \SG{kamaanoo} \OP{kamaana} \TP{kamaanɛɛ} 
\LX{kamaapɛge} \HM{} \PH{kámáápɛ́gɪ́} \PS{n.~} \GE{corn husk} \OP{kamaapɛgre} \TP{kamaapɛgrɛɛ} 
\LX{kamakama} \HM{} \PH{kámákámá} \PS{adj.~} \GE{brittle and light} 
\LX{kamalaɛ} \HM{} \PH{kámàláɪ̀} \PS{n.~} \GE{carbide lamp; headlamp used by hunters at night} \OP{kamalaare} \TP{kamalaarɛɛ} 
\LX{kamasɛle} \HM{} \PH{kàmàsɛ́lɪ̀} \PS{n.~} \GE{a type of thorny, poisonous tree} \TP{kamasɛlɛɛ} 
\LX{kamɔnɔ} \HM{} \PH{kámɔ́nɔ́} \PS{n.~} \GE{driver} \OP{kamɔnema} \TP{kamɔnnɛɛ} 
\LX{kampala} \HM{} \PH{kàmpàlà} \PS{n.~} \GE{a type of smart hunting dog} \OP{kampalle} \TP{kampalɛɛ/kampallɛɛ} 
\LX{kampele} \HM{} \PH{kámpɪ̀l} \PS{n.~} \GE{city, town} \OP{kampelɛ} \TP{kampelɛɛ} 
\LX{kampili} \HM{} \PH{kámpílí} \PS{n.~} \GE{thatched hut} \OP{kampile} \TP{kampilee} \VA{kampiluu}
\LX{kampoŋi} \HM{} \PH{kámpʊ́ŋ} \PS{n.~} \GE{camp} \OP{kamponne} \TP{kamponɛɛ} 
\LX{kana} \HM{} \PH{káná} \PS{n.~} \GE{thick paste made from flour and broth} \TP{kanɛɛ/kannɛɛ} 
\LX{kanaakanaa} \HM{} \PH{kànààkànàà} \PS{adj.~} \GE{stretched to the limit; tight} 
\LX{kanaŋe} \HM{} \PH{kánáŋ} \PS{n.~} \GE{a type of red insect with a split tail} \OP{kananne} \TP{kanannɛɛ} \VA{kanaŋa}
\LX{kananna} \HM{} \PH{kánánná} \PS{adj.~} \GE{not coming forth} \XV{A saana wa la kyɛ deɛ e kananna.} \XE{The stranger came here and made a permanent stay.} 
\LX{kanatɛgɛ} \HM{} \PH{kánátɛ́gɛ́} \PS{n.~} \GE{reward, advantage} \OP{kanatɛgre} \TP{kanatɛgrɛɛ} 
\LX{kaŋazaa} \HM{} \PH{káŋázáá} \PS{adj.~} \GE{anyone, each one} 
\LX{kandazugidaa} \HM{} \PH{kàndàzùgdáá} \PS{n.~} \GE{wood of the kandazugiri tree} \OP{kandazugdaare} \TP{kandazugdarɛɛ} 
\LX{kandazugiri} \HM{} \PH{kàndàzùgdáá} \PS{n.~} \SN \GE{a type of small tree with multiple branches and yellow fruits that are edible} \SN \GE{the fruits of this tree} \OP{kandazugo} \TP{kandazugiree} \VA{kɔndazugiri}
\LX{kane} \HM{} \PH{kànɪ̀} \PS{v.~} \GE{to be overmature e.g. okro fruits} \PD{kanɛɛ, kanna, kanna, kanema, kannaa, kanema } 
\LX{kanea} \HM{} \PH{kànìá} \PS{n.~} \GE{hurricane lamp} \OP{kaneɛne} \TP{kaneɛnɛɛ} 
\LX{kaneɛbiri} \HM{} \PH{kànɪ́ɛ́bírí} \PS{n.~} \GE{lampshade} \OP{kaneɛbie} \TP{kaneɛbiree} 
\LX{kaŋemaŋ} \HM{} \PH{káŋmáŋ} \PS{adv.~} \GE{light (weight)} 
\LX{kanfaraŋ} \HM{} \PH{káfáráŋ} \PS{adv.~} \GE{light (weight)} 
\LX{kaŋkaaraamoorɛɛ} \HM{} \PH{káŋkááráámʊ́ʊ́rɛ́ɛ́} \PS{n.~} \GE{expression signifying the end of a folktale} 
\LX{kaŋkalaŋ} \HM{} \PH{kàŋkàlàŋ} \PS{adv.~} \GE{appearing suddenly} \XV{Te zeŋe dire seemaa wa berɛ ka saana la kaŋkalaŋ.} \XE{As we sat down eating a stranger suddenly appeared.} 
\LX{kaŋkalare} \HM{} \PH{káŋkálàrɪ̀} \PS{n.~} \GE{persistence in doing something despite knowing it is beyond one's ability} \TP{kaŋkalarɛɛ} 
\LX{kaŋkale} \HM{} \PH{kǎŋkál} \PS{n.~} \GE{a type of medicinal stew food made from flour and broth} \TP{kaŋkalɛɛ} 
\LX{kaŋkambulo} \HM{} \PH{kaŋkambulo} \PS{n.~} \GE{chronic ulcer} 
\LX{kaŋkaŋ} \HM{} \PH{káŋkáŋ} \PS{adv.~} \GE{bad-smelling} 
\LX{kaŋkaŋaa} \HM{} \PH{kǎŋkáŋàá} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{kaŋkanne} \TP{kaŋkannɛɛ} 
\LX{kaŋkandoraa} \HM{} \PH{káŋkándòráá} \PS{n.~} \GE{a type of bird with tan plumage and a thick beak} \OP{kaŋkandori} \TP{kaŋkandoree} \VA{kaŋkanduo}
\LX{kaŋkanduo} \HM{} \PH{káŋkǎndúó} \PS{n.~} \GE{a type of small bird with a thick beak that eats fruits of trees} \OP{kaŋkandori, kaŋkandorri, kaŋkandoree} \VA{kaŋkandobaa/kaŋkandoraa}
\LX{kaŋkaŋgbalaa} \HM{} \PH{kàŋkàŋgbàláà} \PS{n.~} \GE{a type of shady tree} \OP{kaŋkaŋgballe} \TP{kaŋkaŋgbalɛɛ} \VA{gbalaa}
\LX{kaŋkankyekyeɛ} \HM{} \PH{kákánʧɪ́ʧɪ́ɛ́} \PS{n.~} \GE{a type of tree with milky sap} \OP{kaŋkankyekyeere} \TP{kaŋkankyekyeerɛɛ} 
\LX{kaŋkaŋkyekyeɛ} \HM{} \PH{káŋkáŋʧɪ̀ʧɪ̀ɛ́} \PS{n.~} \GE{a type of fig tree} \OP{kaŋkaŋkyekyeere} \TP{kaŋkaŋkyekyeerɛɛ} 
\LX{kaŋkansaalaa} \HM{} \PH{káŋkánsààláà} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{kaŋkansaale} \TP{kaŋkansaalɛɛ} 
\LX{kaŋkansɔglaa} \HM{} \PH{káŋkánɔ́gláá} \PS{n.~} \GE{a type of fig tree with aerial roots} \OP{kaŋkansɔglɔ} \TP{kaŋkansɔglɛɛ} 
\LX{kaŋkanzeɛ} \HM{} \PH{ káŋkánzɪ̀ɛ́} \PS{n.~} \GE{a type of fig tree with medicinal properties and edible fruits} \OP{ kaŋkanzeere} \TP{ kaŋkanzeerɛɛ} 
\LX{kankyeemɛ} \HM{} \PH{kánʧɪ́ɪ́mɛ́} \PS{n.~} \GE{one hired to cook for another} \TP{kankyeemɛɛ} 
\LX{kankyɛle} \HM{} \PH{kánʧɛ́lɪ́} \PS{n.~} \GE{a type of large tree with large, shiny leaves} \OP{kankyɛlɛ} \TP{kankyɛlɛɛ} 
\LX{kankyɛlpelaa} \HM{} \PH{kánʧɛ̀lpɪ̀làá} \PS{n.~} \GE{a type of tree} \OP{kankyɛlpeɛle} \TP{kankyɛlpeɛlɛɛ} 
\LX{kankyiri} \HM{} \PH{kànʧírí} \PS{n.~} \GE{a type of small plant with edible leaves} \TP{kankyiree} 
\LX{kankyirikpoŋ} \HM{} \PH{kánʧíríkpóŋ} \PS{n.~} \GE{variety of (see kankyiri)} \OP{kankyirikponni} \TP{kankyirikponnee} 
\LX{kaŋmɛ} \HM{} \PH{káŋmɛ̀} \PS{n.~} \GE{act of threshing millet} 
\LX{kaŋmɛdaa} \HM{} \PH{káŋmɛ̀dàá} \PS{n.~} \GE{stick for threshing millet} \OP{kaŋmɛdaare} \TP{kaŋmɛdarɛɛ/kaŋmɛdaarɛɛ} 
\LX{kaŋmeɛrɛ} \HM{} \PH{káŋmɪ́ɛ́rɛ́} \PS{n.~} \GE{one engaged in the threshing of millet} \OP{kaŋmeɛrebɛ} \TP{kaŋmeɛrebɛɛ} 
\LX{kanne} \HM{} \PH{kànnɪ̀} \PS{v.~} \SN \GE{to teach} \SN \GE{to read} \PD{kannɛɛ, kanna, kanna, kanema, kannaa, kanema } 
\LX{kanne} \HM{} \PH{kánnɪ̀} \PS{v.~} \GE{to doubt} \PD{kannɛɛ, kanna, kanna, kanema, kannaa, kanema } 
\LX{kannyiri} \HM{} \PH{kànɲìrì} \PS{n.~} \GE{be patient} \TP{kannyiree} 
\LX{kansɛ} \HM{} \PH{kánsɛ́} \PS{n.~} \GE{cakes} \OP{kansɛre} \TP{kansɛrɛɛ} \VA{sɛnsɛre, sensɛ}
\LX{kansele} \HM{} \PH{kánsɪ̀l} \PS{n.~} \GE{council} \OP{kanselɛ} \TP{kanselɛɛ} 
\LX{kansɔge} \HM{} \PH{kánsɔ́gɪ́} \PS{n.~} \GE{foreign (not locally made) hoe} \OP{kansɔgre} \TP{kansɔgrɛɛ} 
\LX{kantakye} \HM{} \PH{kántáʧè} \PS{n.~} \GE{train} \OP{kantakyeri} \TP{kantakyeree} 
\LX{kantɛroo} \HM{} \PH{kántɛ́rʊ́ʊ́} \PS{n.~} \GE{a type of strong grass that grows on riverbeds} \OP{kantɛrɛɛ} \TP{kantɛroo} 
\LX{kanwaa} \HM{} \PH{kánwàà} \PS{adv.~} \GE{supposed} \XV{O kanwaa libie na baarɛɛ.} \XE{His supposed money is finished.} 
\LX{kanwaa} \HM{} \PH{kánwáá} \PS{n.~} \GE{hard pebble of acid} \SG{kanwaarɛɛ} \OP{kanwaare} 
\LX{kanwaa-ŋmen-yelmanna} \HM{} \PH{kánwàá-ŋmɪ́n-yèlmǎnnà} \PS{n.~} \GE{false prophet} \OP{kanwaa-ŋmen-yelmanema} \TP{kanwaa-ŋmen-yelmanemɛɛ} 
\LX{kapaalaa} \HM{} \PH{kápááláá} \PS{n.~} \GE{raised shed for guinea corn or millet} \OP{kapaale} \TP{kapaalɛɛ} 
\LX{kapala} \HM{} \PH{kàpálà} \PS{n.~} \GE{mashed yam, plantain or cassava} \SG{kapaloo} \TP{kapalɛɛ/kapallɛɛ} \VA{kapalo (Saanchi) -mashed yams, pl. kapala}
\LX{kapenta} \HM{} \PH{kápɪ́ntà} \PS{n.~} \GE{carpenter} \OP{kapentarɪ} 
\LX{kapile} \HM{} \PH{kàpílè} \PS{adj.~} \GE{crafty} 
\LX{kapoŋi} \HM{} \PH{kápòŋ} \PS{n.~} \GE{millet flower} \OP{kaponni} \TP{kaponnee} 
\LX{kapuro} \HM{} \PH{kàpùrò} \PS{n.~} \GE{pillow} \OP{kapurri} \TP{kapuree/kapurree} 
\LX{kara} \HM{} \PH{kárà} \PS{v.~} \GE{to prepare, to get ready} \XV{O kara la a saama karoo.} \XE{S/he is preparing for the visitors.} \PD{karɛɛ, kara, kara, kareba, karaa, kareba } 
\LX{kara} \HM{} \PH{kárá} \PS{n.~} \GE{a type of grass used for thatching} 
\LX{karaa} \HM{} \PH{káráá} \PS{adj.~} \SN \GE{unripe} \SN \GE{not cooked through} \SN \GE{not hot enough (of iron in fire)} \OP{kare} \TP{karrɛɛ} 
\LX{karaa} \HM{} \PH{káráá} \PS{adj.~} \GE{uncompromising} \OP{kare} \TP{karrɛɛ} 
\LX{karaa} \HM{} \PH{kàráà} \PS{n.~} \GE{cobra} \OP{karre} \TP{karrɛɛ} 
\LX{karaa} \HM{} \PH{kàráà} \PS{n.~} \GE{clerk} \OP{karre} \TP{karrɛɛ} 
\LX{karaa} \HM{} \PH{káráá} \PS{n.~} \GE{file} \OP{karre} \TP{karɛɛ/karrɛɛ} 
\LX{karaa} \HM{} \PH{kàràà} \PS{adj.~} \GE{large} \OP{karre} 
\LX{karambanne} \HM{} \PH{kárámbánnɪ́} \PS{n.~} \GE{wit} \TP{karambannɛɛ} 
\LX{karansunni} \HM{} \PH{kàrànsúnnì} \PS{n.~} \GE{kerosene} \TP{karansunnee} 
\LX{karasa} \HM{} \PH{kàràsà} \PS{adj.~} \GE{bad-smelling} 
\LX{karawaara} \HM{} \PH{káráwáárá} \PS{n.~} \GE{green grass that has not been burnt completyely} \SG{karawaaroo} \TP{karawaarɛɛ} 
\LX{kare} \HM{} \PH{kàrɪ̀} \PS{v.~} \SN \GE{to drive away} \SN \GE{to chase, to run after} \SN \GE{to sharpen or polish a weapon or tool (arrow or knife)} \SN \GE{to grow tall; to cause to rise tall or high} \SN \GE{to get overgrown (okro)} \PD{karɛɛ, kara, kara, kareba, karaa, kareba } 
\LX{kare} \HM{} \PH{kárɪ́} \PS{adv.~} \GE{completely (of people going away)} \XV{A bibiiri zaaŋ zo, kare.} \XE{All the children have run away, leaving none behind.} 
\LX{kare} \HM{} \PH{kárɪ́} \PS{v.~} \GE{to break (small objects)} \PD{karɛɛ, kara, kara, kareba, karaa, kareba } 
\LX{karekare} \HM{} \PH{kàrɪ̀kárɪ́} \PS{v.~} \GE{to attempt to make a move} \XV{Meŋ iri karekare ka te nyɛ ka o na sage la me.} \XE{Also make an effort and let's see whether he will agree.} \PD{karekarɛɛ, karekara, karekara, karekareba, karekaraa, - } 
\LX{karekpe} \HM{} \PH{kàrɪ́kpé} \PS{n.~} \GE{spitting cobra} \OP{karekperi} \TP{karekperee} 
\LX{karekpele} \HM{} \PH{kárɪ́kpɪ́lɪ́} \PS{n.~} \GE{male lizard; a small reptile of the squamata family that is red-headed} \OP{karekpelɛ} \TP{karekpelɛɛ} \VA{shortened form of bandakarekpele}
\LX{karekye} \HM{} \PH{kárɪ́ʧɪ̀} \PS{n.~} \GE{clerk} \OP{karekyere} \TP{karekyerɛɛ} 
\LX{karema} \HM{} \PH{kárɪ́má} \PS{n.~} \GE{teacher, one who deals with studies} \OP{karemamine} \TP{karemɛɛ} 
\LX{karemogɔ} \HM{} \PH{kárɪ̀mʊ́gá} \PS{n.~} \GE{fortune-teller, diviner} \OP{karemogre} \TP{karemogrɛɛ} 
\LX{kareŋmeɛ} \HM{} \PH{kárɪ́ŋmɪ́ɛ̀} \PS{n.~} \GE{a type of bird with long, thin legs} \OP{kareŋmeɛne} \TP{kareŋmeɛnɛɛ} \VA{kalŋmeɛ}
\LX{karenteɛ} \HM{} \PH{kàrɪ́ntɪ́ɛ́} \PS{n.~} \GE{machete, cutlass} \OP{karenteere} \TP{karenteerɛɛ} 
\LX{karingyoŋ} \HM{} \PH{káríndzóŋ} \PS{n.~} \GE{a type of musical instrument} \OP{karingyonni} \TP{karingyonnee} \VA{karangyoŋ/karengyoŋ}
\LX{kasɛɛte} \HM{} \PH{kàsɛ́ɛ́tɪ́} \PS{n.~} \GE{cassette tape} \OP{kasɛɛtere} \TP{kasɛɛterɛɛ} 
\LX{kasɛlemɛ} \HM{} \PH{kásɛ́lɪ́mɛ́} \PS{n.~} \GE{millet seedling} \VA{kasɛlle}
\LX{kasɛlle} \HM{} \PH{kásɛ́llɪ́} \PS{n.~} \GE{millet seedling meant for planting} \VA{kasɛlemɛ}
\LX{kasogi} \HM{} \PH{kásògí} \PS{n.~} \GE{a small cage used as a nest for young chicks} \OP{kasogri} \TP{kasogree} \VA{kosogi}
\LX{katakye} \HM{} \PH{kátáʧè} \PS{n.~} \GE{train} \OP{katakyeri} \TP{katakyeree} 
\LX{kataŋa} \HM{} \PH{kátáŋà} \PS{n.~} \GE{a large granary for storage of foodstuff} \VA{katɛgɛ}
\LX{kataweɛ} \HM{} \PH{kàtàwɪ́ɛ́} \PS{n.~} \GE{umbrella} \OP{kataweere} \TP{kataweerɛɛ} 
\LX{katɛgɛ} \HM{} \PH{kàtɛ́gɛ̀} \PS{n.~} \GE{large granary} \OP{katɛgre} \TP{katɛgrɛɛ} \VA{kataŋa}
\LX{katoŋe} \HM{} \PH{kàtʊ̂ŋ} \PS{n.~} \GE{civet} \OP{katonne} \TP{katonnɛɛ} 
\LX{ka-uri} \HM{} \PH{ká-úrí} \PS{n.~} \GE{dusty chaff from millet (irritates the skin)} \TP{ka-uree} 
\LX{kayeɛle} \HM{} \PH{káyɪ́ɛ́lɪ̀} \PS{n.~} \GE{the act of winnowing millet} 
\LX{kayeɛlɛ} \HM{} \PH{káyɪ́ɛ́lɛ́} \PS{n.~} \GE{one engaged in winnowing millet} \OP{kayeɛlbɛ} \TP{kayeɛlbɛɛ} 
\LX{kayɛre} \HM{} \PH{káyɛ́rɪ́} \PS{n.~} \GE{half-wild millet} \OP{kayɛɛ} \TP{kayɛɛrɛɛ} 
\LX{kazama} \HM{} \PH{kázàmá} \PS{n.~} \GE{poorly matured millet} \TP{kazamɛɛ} 
\LX{kazeɛ} \HM{} \PH{kázɪ̀ɛ́} \PS{n.~} \GE{guinea corn} \SG{kazeroo (plant)} \TP{kazeerɛɛ} 
\LX{kazu} \HM{} \PH{kázú} \PS{n.~} \GE{ear of millet} \OP{kazuri} \TP{kazuree} 
\LX{ka᷈a} \HM{} \PH{ká᷉á} \PS{n.~} \SN \GE{oil; butter} \SN \GE{fat} \SG{kaanoo} \TP{kaanɛɛ} 
\LX{ka᷈a} \HM{} \PH{ká᷉á} \PS{n.~} \GE{nectar e.g. in honey} \SG{kaanoo} \TP{kaanɛɛ} 
\LX{ka᷉aboli} \HM{} \PH{ká᷉ábòlí} \PS{n.~} \GE{pure liquid oil} \OP{ka᷉abolo} \TP{ka᷉abolee} 
\LX{ka᷈aburi} \HM{} \PH{ká᷉ábúrí} \PS{n.~} \GE{burnt residue from shea butter} \OP{ka᷈aburri} \TP{ka᷈aburee} 
\LX{ka᷉aburituo} \HM{} \PH{ká᷉ábùrìtúó} \PS{n.~} \GE{type of residue from the last stage of shea butter preparation} \OP{ka᷉aburituuri} \TP{ka᷉aburituuree} 
\LX{ka᷉aze᷉ɛ} \HM{} \PH{ká᷉ázɪ́᷉ɛ́} \PS{n.~} \GE{debris from final preparation of shea butter oil} \TP{ka᷉aze᷉ɛnɛɛ} 
\LX{keɛre} \HM{} \PH{kɪ́ɛ́rɪ̀} \PS{v.~} \GE{to make a cluckling noise to call fowl or drive animals away} \PD{keɛrɛɛ, keɛrɛ, keɛrɛ, keɛrebɛ, keɛraa, keɛrebɛ } 
\LX{kegireŋ} \HM{} \PH{kégréŋ} \PS{adj.~} \GE{small and hanging} \XV{A bie yɛre la wo-kegireŋ kaŋa.} \XE{The child carried some small animal skin bag.} 
\LX{kegireŋ} \HM{} \PH{kègrèŋ} \PS{adj.~} \GE{large and hanging} 
\LX{kegisege} \HM{} \PH{kégségé} \PS{adv.~} \GE{small but compact} \XV{A noɔɛ la a e kegisegee lɛ.} \XE{the child has grown and is plump.} 
\LX{kegisege} \HM{} \PH{kègsègè} \PS{adv.~} \GE{large and compact} \XV{Wɛnaabo maŋ waa la kegisegee lɛ.} \XE{The buffalo is usually big and plump.} 
\LX{kei} \HM{} \PH{kéì} \PS{interjection.~} \GE{stop it} 
\LX{kei} \HM{} \PH{kèí} \PS{interjection.~} \GE{expression of disapproval} 
\LX{kɛkɛre} \HM{} \PH{kɛ̀kɛ́rɪ̀} \PS{n.~} \SN \GE{stalk (of maize, millet)} \SN \GE{small bundle of stalks used for gathering threshed millet} \SG{kɛkɛroo} \OP{kɛkɛrɛɛ} \TP{kɛkɛrrɛɛ} 
\LX{kele} \HM{} \PH{kélé} \PS{n.~} \GE{very hard seeds of some fruits; e.g baobab (adansonia digitata) gourd} \VA{tokele}
\LX{kɛlɛkɛlɛ} \HM{} \PH{kɛ́lɛ́kɛ́lɛ́} \PS{adj.~} \GE{lightweight (thickness)} 
\LX{kelewele} \HM{} \PH{kéléwèlé} \PS{n.~} \GE{fried plantain chips} \TP{kelewellee} 
\LX{kelifelli} \HM{} \PH{kéféll} \PS{adj.~} \GE{light (weight)} 
\LX{keloo} \HM{} \PH{kélòó} \PS{adv.~} \GE{eat or chew instantly} 
\LX{keloo} \HM{} \PH{kèlòò} \PS{adv.~} \GE{glance at once} 
\LX{kenekene} \HM{} \PH{kɪ́nkɪ̀n} \PS{adv.~} \GE{full to capacity liquid} 
\LX{keŋkeleŋ} \HM{} \PH{kéŋkéléŋ} \PS{adj.~} \GE{light and dangling} \XV{A dɔɔ de la ŋmampege keŋkeleŋ kaŋa yɛre a baa.} \XE{The man has hung some light broken calabash on the neck of the dog.} 
\LX{keŋkeleŋ} \HM{} \PH{kèŋkèlèŋ} \PS{adj.~} \GE{large and dangling} \XV{Ka a bieŋ yɛre a wo-keŋkeleŋ ŋa o koŋ tõɔ kyɛŋe ne.} \XE{If the boy hands this large skin bag on his shoulder he cannot walk with it.} 
\LX{keŋkeŋ} \HM{} \PH{kéŋkéŋ} \PS{adj.~} \GE{very full (stomach, balloon)} 
\LX{kere} \HM{} \PH{kèré} \PS{n.~} \GE{dog} \OP{kerri} \TP{keree} 
\LX{kerekere} \HM{} \PH{kérékéré} \PS{adv.~} \GE{lightweight (thickness)} 
\LX{kerekere} \HM{} \PH{kɪ́rɪ́kɪ́rɪ́} \PS{n.~} \GE{wheel from the bucket} \OP{kerekerre} \TP{kerekerɛɛ} 
\LX{kerekere} \HM{} \PH{kɪ̀rɪ̀kɪ́rɪ́} \PS{v.~} \GE{to struggle} \PD{kerekerɛe,, kerekerɛ, kerekerɛ, kerekerebɛ, kerekeraa, - } 
\LX{kerekerekeree} \HM{} \PH{kérékérékèrèé} \PS{n.~} \GE{nursery rhyme} \OP{kerekerri.} 
\LX{kereŋkereŋ} \HM{} \PH{kéréŋkéréŋ} \PS{adv.~} \GE{full to capacity (tyre, balloon, or stomach)} 
\LX{kerese} \HM{} \PH{kérésé} \PS{adj.~} \GE{small and distended} \XV{A bibaala puori waa la kerese lɛ.} \XE{The sick child’s stomach is distended.} 
\LX{kerese} \HM{} \PH{kèrèsè} \PS{adj.~} \GE{big and distended (stomach)} 
\LX{keri} \HM{} \PH{kèrì} \PS{v.~} \GE{come round clumsiness} \XV{Zenɛ ka a boɔ dɔge a bile naŋ ba keri} \XE{The goat just literred today the kid has not come round yet.} \PD{keree, kere, kere, keribe, keraa, keribe } 
\LX{keri} \HM{} \PH{kérí} \PS{v.~} \SN \GE{to measure, to weigh} \SN \GE{to face a difficult task} \XV{O keri la toma ko ma ka N puro ne wale.} \XE{She has burdened me with a heavy job, and I am sweating.} \PD{keree, kere, kere, keribe, keraa, keribe } 
\LX{keri} \HM{} \PH{kèrì} \PS{adv.~} \GE{onomatopoeic; the way a type of sound is produced when a small object hits a bigger (stiff) object} 
\LX{kieluu} \HM{} \PH{kíélúú} \PS{adj.~} \GE{burnt-smelling} 
\LX{kieluu} \HM{} \PH{kíélúú} \PS{adj.~} \GE{high-pitched and distorted (singing)} 
\LX{kieree} \HM{} \PH{kíéréé} \PS{n.~} \GE{brim of hat e.g. cowboy hat, bowler hat} \OP{kieri} 
\LX{ko} \HM{} \PH{kʊ́} \PS{v.~} \GE{to kill} \PD{koɛ, koora, koora, kooreba, kooraa, kooreba } 
\LX{ko} \HM{} \PH{kʊ́} \PS{v.~} \GE{to put out (fire)} \PD{koɛ, koora, koora, kooreba, kooraa, kooreba } 
\LX{ko} \HM{} \PH{kʊ́} \PS{v.~} \GE{to fold (cloth)} \PD{koɛ, koora, koora, kooreba, kooraa, kooreba } 
\LX{ko} \HM{} \PH{kò} \PS{v.~} \GE{to dry} \PD{ koe,  koro, koro, koribo, koraa, koribo } 
\LX{ko} \HM{} \PH{kó} \PS{v.~} \GE{to gather e.g rain clouds} \PD{ koe,  koro, -, -, -, - } 
\LX{ko} \HM{} \PH{kʊ̀} \PS{v.~} \GE{to give} \PD{koɛ, kora, kora, koreba, koraa, koreba } 
\LX{kɔ} \HM{} \PH{kɔ̌} \PS{v.~} \SN \GE{to hoe} \SN \GE{to cultivate, to farm} \PD{kɔɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\LX{ko suuri} \HM{} \PH{kʊ́ súúrí} \PS{v.~} \GE{to arouse someone's anger} \PD{koɛ suuri, koora suuri, -, -, -, - } 
\LX{kobeɛre} \HM{} \PH{kòbɪ́ɛ́rɪ̀} \PS{n.~} \GE{first weekday of post-funeral rites} \OP{kobeɛre /kóbɪ́ɛ́rɪ́/} \TP{kobeɛrɛɛ} 
\LX{Kɔbenɛ} \HM{} \PH{kɔ́bɪ́nɛ́} \PS{n.~} \GE{an annual festival celebrated by the people of Lawra} \TP{kɔbenɛɛ} 
\LX{kobo} \HM{} \PH{ko̊bò} \PS{n.~} \GE{penny} \OP{kogri} \TP{kogree} 
\LX{koboŋkoboŋ} \HM{} \PH{kòbòŋkóbóŋ} \PS{adj.~} \GE{rugged, uneven e.g. road} 
\LX{koboɔ} \HM{} \PH{kòbʊ́ɔ́} \PS{n.~} \GE{goat for funeral rites} \OP{koboore} \TP{koboorɛɛ} 
\LX{kobora} \HM{} \PH{kʊ̀bʊ̀rà} \PS{n.~} \GE{leader, greatest one among a group} \OP{koborre} \TP{koborɛɛ} 
\LX{kodaa} \HM{} \PH{kòdàá} \PS{n.~} \GE{funeral week} \OP{kodare} \TP{kodarɛɛ} 
\LX{koda᷉a} \HM{} \PH{kòdá᷉á} \PS{n.~} \GE{funeral drink} \SG{koda᷉ao} \TP{kodaanɛɛ} 
\LX{kodie} \HM{} \PH{kòdìè} \PS{n.~} \GE{first weekend of post-funeral rites celebration} \OP{koderi} \TP{koderee} 
\LX{kodɔgerɔ} \HM{} \PH{kʊ̊dɔ̀grɔ̀} \PS{n.~} \GE{foetus (that causes the death of a parent during labour)} \OP{kodɔgere} \TP{koɖgerɛɛ} 
\LX{kodɔɔ} \HM{} \PH{kòdɔ́ɔ́} \PS{n.~} \GE{widower} \OP{ko-dɔbɔ} \TP{ko-dɔɔrɛɛ} 
\LX{kodu} \HM{} \PH{kòdú} \PS{n.~} \GE{banana} \SG{koduruu} \OP{koduri} \TP{koduree} 
\LX{kofi} \HM{} \PH{kófì} \PS{n.~} \GE{coffee} \OP{kofiri} \TP{kofiree} 
\LX{koganaa} \HM{} \PH{kògànàá} \PS{n.~} \GE{strip of leather, cloth, or twine for a bereaved person} \OP{koganne} \TP{kogannɛɛ} 
\LX{kɔge} \HM{} \PH{kɔ̀gɪ̀} \PS{v.~} \SN \GE{to be near, to be close to} \SN \GE{to put by the side to support (pillow)} \SN \GE{to be lean} \PD{kɔgɛɛ, kɔgrɔ, kɔgrɔ, kɔgreba, kɔgraa, - } 
\LX{kɔge} \HM{} \PH{kɔ́gɪ́} \PS{v.~} \GE{to boil ground malt at the the first stage in the preparation of pito} \PD{kɔgɛɛ, kɔgera, kɔgera, kɔgereba, kɔgeraa, kɔgereba } 
\LX{kɔgele} \HM{} \PH{kɔ̀glɪ̀} \PS{v.~} \SN \GE{to keep company} \SN \GE{to give aid} \PD{kɔgelɛɛ, kɔgela, kɔgela, kɔgeleba, kɔgelaa, kɔgeleba } 
\LX{kɔgelɔ} \HM{} \PH{kɔ̀glɔ̀} \PS{n.~} \GE{a type of fetish or god} \TP{kɔgelɛɛ} 
\LX{kɔgeraa} \HM{} \PH{kɔ́gráá} \PS{n.~} \GE{raised lump on the floor of a room used as a seat} \OP{kɔgere} \TP{kɔgerɛɛ} 
\LX{kogi} \HM{} \PH{kógí} \PS{n.~} \SN \GE{seat, stool, chair} \SN \GE{throne} \OP{kogree} 
\LX{kogi} \HM{} \PH{kògí} \PS{adl.~} \GE{thick and hard} 
\LX{kogilii} \HM{} \PH{kógilíí} \PS{adj.~} \GE{dry and wrinkled; used on humans, animals and objects like containers} \OP{kogilo} \VA{kogiluu}
\LX{kogilimala} \HM{} \PH{kógílímálá} \PS{adj.~} \GE{very lean and worn out} 
\LX{kogiloŋ} \HM{} \PH{kóglóŋ} \PS{adj.~} \GE{small and curved} 
\LX{kogiluu} \HM{} \PH{kógílúú} \PS{adj.~} \GE{lean e.g a person, objects like containers, animal and humans} \OP{kogilo} 
\LX{Kogiri} \HM{} \PH{kógrì} \PS{n.~} \GE{name of a village in Central Dagao} \TP{kogiree} 
\LX{kogiri} \HM{} \PH{kógrí} \PS{adj.~} \GE{near the side of} \XV{A teɛ peɛle la mɔkogiri.} \XE{The tree is near the edge of the bush.} 
\LX{kogiri} \HM{} \PH{kógrì} \PS{v.~} \GE{to wink} \PD{kogree, kogiro, kogiro, kogiribo, kogiraa, kogiribo } 
\LX{kogiro} \HM{} \PH{kógró} \PS{n.~} \GE{basket seat} \OP{kogiri} \TP{kogiree} 
\LX{kogiroŋ} \HM{} \PH{kògròŋ} \PS{adj.~} \GE{oblong in shape} 
\LX{kogluu} \HM{} \PH{} \PS{n.~} \GE{} \VA{kogilii}
\LX{kogo} \HM{} \PH{kógò} \PS{n.~} \GE{mahogany} \OP{kogri} \TP{kogree} \VA{kogi}
\LX{kogo᷉ɔ} \HM{} \PH{kógʊ́᷉ɔ́} \PS{n.~} \GE{funeral gound (arena of dirge singers)} \OP{kogoɔne} \TP{kogoɔnɛɛ} 
\LX{kɔkɔbaŋ} \HM{} \PH{kɔ̀kɔ́báŋ} \PS{n.~} \GE{neck bangle} \OP{kɔkɔbanne} \TP{kɔkɔbannɛɛ} \VA{kɔkɔbɔŋ}
\LX{kɔkɔkpɔgere} \HM{} \PH{kɔ̀kɔ́kpɔ́grɪ̀} \PS{n.~} \GE{goitre} \OP{kɔkɔkpɔgɔ} \TP{kɔkɔkpɔgerɛɛ} 
\LX{kokolaa} \HM{} \PH{kókóláà} \PS{n.~} \GE{shuttle (weaving)} \OP{kokolli} \TP{kokolee/kokollee} 
\LX{kokolaa} \HM{} \PH{kókóláà} \PS{n.~} \GE{a type of river god or fetish} \OP{kokolli} \TP{kokolee/kokollee} 
\LX{kɔkɔlaa} \HM{} \PH{kɔ̀kɔ̀láà} \PS{n.~} \GE{cat} \OP{kɔkɔlle} \TP{kɔkɔllɛɛ} \VA{kɔlaa}
\LX{kɔkɔlɛre} \HM{} \PH{kɔ̀kɔ́lɛ̀rɪ́} \PS{n.~} \GE{necklace} \SG{kɔkɔlɛroo} \OP{kɔkɔlɛɛ} 
\LX{kɔkɔliŋe} \HM{} \PH{kɔ̀kɔ́líŋé} \PS{n.~} \GE{velum} \OP{kɔkɔlinni} \TP{kɔkɔlinnee} 
\LX{kɔkɔllee} \HM{} \PH{kɔ̀kɔ̀llée} \PS{n.~} \GE{kitten} \OP{kɔkɔlbilii} \TP{kɔkɔlbillee} 
\LX{kokolo} \HM{} \PH{kòkòlò} \PS{n.~} \GE{a type of tree whose ashes are used as salt} \OP{kokolli} \TP{kokollee} 
\LX{kɔkɔlɔŋgoɔne} \HM{} \PH{kɔ̀kɔ́lɔ̀ŋgʊ́ɔ́nɪ́} \PS{n.~} \GE{the trachea} \OP{kɔkɔlɔŋgoɔmɔ} \TP{kɔkɔlɔŋgoɔnɛɛ} 
\LX{kɔkɔlweɛmɛ} \HM{} \PH{kɔ̀kɔ̀lwɪ́ɛ̀mɛ́} \PS{n.~} \GE{wild cat} \OP{kɔkɔlweɛne} \TP{kɔkɔlweɛnɛɛ} \VA{kɔlweɛmɛ}
\LX{kɔkɔlwoɔ} \HM{} \PH{kɔ̀kɔ̌lwʊ́ɔ́} \PS{n.~} \GE{bag made from cat hide} \OP{kɔkɔlwoore} \TP{kɔkɔlwoorɛɛ} 
\LX{kɔkɔmaraa} \HM{} \PH{kɔ̀kɔ́máráá} \PS{n.~} \GE{tight-fitting necklace} \OP{kɔkɔmarre} \TP{kɔkɔmarrɛɛ} 
\LX{kokomo} \HM{} \PH{kòkòmò} \PS{adj.~} \GE{the nature of the sound of when one takes a bite at something hard. e.g. a bone} 
\LX{kokoori} \HM{} \PH{kòkòòrí} \PS{n.~} \GE{pan flute} \OP{kokoe} \TP{kokoree/kokorree} 
\LX{kɔkɔraa} \HM{} \PH{kɔ́kɔ́ràà} \PS{n.~} \GE{one who does dirty things (obscene language)} \OP{kɔkɔrre} 
\LX{kɔkɔre} \HM{} \PH{kɔ́kɔ́rɪ̀} \PS{n.~} \SN \GE{throat} \SN \GE{voice} \SN \GE{language} \OP{kɔkɔrɛɛ/kɔkɔɛ} \TP{kɔkɔrrɛɛ} 
\LX{kɔkɔreyɛlɛ} \HM{} \PH{kɔ́kɔ̀rɪ̀yɛ́lɛ́} \PS{n.~} \GE{obscene things} 
\LX{kɔkɔroŋ} \HM{} \PH{kɔ́kɔ́rʊ́ŋ} \PS{n.~} \GE{the act of being uncouth} 
\LX{kɔkɔsɔgelaa} \HM{} \PH{kɔ̀kɔ́sɔ̀glàà} \PS{n.~} \GE{breastbone} \OP{kɔkɔsɔgele} \TP{kɔkɔsɔgelɛɛ} 
\LX{kɔkɔvɔlaa} \HM{} \PH{kɔ́kɔ́vɔ́làà} \PS{n.~} \GE{esophagus, gullet} \OP{kɔkɔvɔlle} \TP{kɔkɔvɔllɛɛ} 
\LX{kɔkɔyoraa} \HM{} \PH{kɔ̀kɔ́yòráá} \PS{n.~} \GE{larynx, voice box, Adam's apple} \SG{kɔkɔyoruu} \OP{kɔkɔyorri} \TP{kɔkɔyoree} 
\LX{koko᷉ɔ} \HM{} \PH{kòkʊ́᷉ɔ́} \PS{n.~} \GE{food or drink given to mourners at a funeral} \TP{kokoɔnɛɛ} 
\LX{kokyi} \HM{} \PH{kòʧí} \PS{n.~} \GE{millet for funeral rites} \TP{kokyiree} 
\LX{kolaa} \HM{} \PH{kʊ́láá} \PS{n.~} \SN \GE{river, source of water people must fetch} \SN \GE{river god} \OP{kolle} \TP{kollɛɛ} 
\LX{kɔlaa} \HM{} \PH{kɔ̀láà} \PS{n.~} \GE{cat} \OP{kɔlle} \TP{kɔllɛɛ} \VA{kɔkɔlaa}
\LX{kolaa nyɔge} \HM{} \PH{kʊ́láá ɲɔ́gɪ̀} \PS{intransitive v.~} \GE{to drown} 
\LX{kolaanoɔre} \HM{} \PH{kʊ́láánʊ́ɔ́rɪ́} \PS{n.~} \GE{south of Ghana, Kumasi} 
\LX{kolakola} \HM{} \PH{kʊ́låkʊ́lå} \PS{adv.~} \GE{soft e.g. cloth, leather} \VA{kolɔkolɔ}
\LX{kolaŋene} \HM{} \PH{kòláŋɪ́nɪ́} \PS{n.~} \GE{dirge} \TP{laŋenɛɛ} 
\LX{kɔlbaa} \HM{} \PH{kɔ́lbáá} \PS{n.~} \GE{bottle} \OP{kɔlbaare} \TP{kɔlbaarɛɛ} 
\LX{kolboll} \HM{} \PH{kʊ̀bʊ̀ll} \PS{adv.~} \GE{rising suddenly and in great quantity e.g. dust, smoke} 
\LX{kɔle} \HM{} \PH{kɔ́lɪ́} \PS{v.~} \GE{to be loose within something; to be too small for} \XV{A dɔɔ kuree kɔlɔ la a o bidɔɔ seɛ.} \XE{The man's pair of shorts is too large for his son.} \PD{kɔlɛɛ, kɔlɔ, kɔlɔ, kɔlebɔ, kɔlaa, kɔlebɔ } 
\LX{kolee} \HM{} \PH{kóléé} \PS{n.~} \GE{tin ; cup} \OP{kobilii} \TP{kobilee/kollee} 
\LX{kɔlema} \HM{} \PH{kɔ̀lɪ̀má} \PS{n.~} \GE{well} \OP{kɔlle} \TP{kɔlɛɛ} \VA{kolaa (kʊlaa), kolli}
\LX{kolemɛleŋ} \HM{} \PH{kʊ́lmɛ́lɪ́ŋ} \PS{adj.~} \GE{match or correspond very well e.g. rice and beans} 
\LX{kɔleŋkɔ} \HM{} \PH{kɔ́lɪ́ŋkɔ́} \PS{n.~} \GE{a type of gourd with a long neck} \OP{kɔleŋkɔre} \TP{kɔleŋkɔrɛɛ} 
\LX{kolezare} \HM{} \PH{kòlɪ̀zárɪ́} \PS{n.~} \GE{funeral donation} \OP{kolezaɛ} \TP{kolezɛrɛɛ} 
\LX{kolimasogilo} \HM{} \PH{kòlìmásògìló} \PS{n.~} \GE{a legendary bird believed to be as big as the ostrich and wicked} \OP{kolimasogili} \TP{kolimasoglee} 
\LX{kolkol} \HM{} \PH{kʊ̀lkʊ̀l} \PS{adv.~} \GE{in great numbers} 
\LX{kolkole} \HM{} \PH{kʊ̀lkʊ́l} \PS{v.~} \GE{to engage in affectionate play; usually between lovers} \PD{kolkolɛɛ, kolkolɔ/kolkola, kolkolɔ/kolkola, kolkoleba, kolkollaa, - } 
\LX{kolkoli} \HM{} \PH{kòlkól} \PS{v.~} \GE{to be all over another e.g. a dog welcoming an acquaintance} \PD{kololee, kolkolo, kolkolo, kolkolaa, kolkolibo, kolkolo } 
\LX{kɔl-kɔl-kɔl} \HM{} \PH{kɔ̀l-kɔ̀l-kɔ̀l} \PS{adj.~} \GE{in a long line e.g a column of school children} 
\LX{koll} \HM{} \PH{kʊ̀ll} \PS{adv.~} \GE{rising into the air e.g. smoke, dust} \VA{kull}
\LX{kolle} \HM{} \PH{kʊ̀llɪ̀} \PS{v.~} \GE{to gather or cut vegetation in great quantities} \PD{kollɛɛ, kolla, kolla, koleba, kollaa, koleba } 
\LX{kɔlle} \HM{} \PH{kɔ̀llɪ̀} \PS{v.~} \GE{to cluck} \PD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlebɔ, kɔllaa, kɔlebɔ } 
\LX{kɔlle} \HM{} \PH{kɔ̀llɪ̀} \PS{v.~} \GE{to roam} \PD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlebɔ, kɔllaa, kɔlebɔ } 
\LX{kɔlle} \HM{} \PH{kɔ́llɪ́} \PS{v.~} \GE{to gallop gracefully} \PD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlbɔ, kɔllaa, - } 
\LX{kɔlle} \HM{} \PH{kɔ́llɪ́} \PS{v.~} \GE{to smooth e.g. potter's pot} \PD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlbɔ, kɔllaa, - } 
\LX{kolnoɔ} \HM{} \PH{kʊ́lnʊ́ɔ̀} \PS{n.~} \GE{moorhen} \OP{kolnoore} \TP{kolnoorɛɛ} 
\LX{kolɔ} \HM{} \PH{kʊ́là} \PS{v.~} \GE{to be unperturbed; to exhibit coolness. (used with negative marker)} \XV{A bie kaŋa ba kolɔ kyɛ kaŋa nyegnyege.} \XE{One of the children is calm but one is the boisterous type.} \PD{-, -, -, -, -, - } \VA{kola}
\LX{kolokolo} \HM{} \PH{kòlókòló} \PS{n.~} \GE{turkey} \OP{kolokolli} \TP{kolokollee} 
\LX{kolokolo} \HM{} \PH{kólókóló} \PS{adv.~} \GE{light} \VA{koloŋkoloŋ}
\LX{kɔlɔkɔlɔ} \HM{1} \PH{kɔ̀lɔ̀kɔ̀lɔ̀} \PS{adv.~} \GE{to begin to brighten e.g day breaking} 
\LX{kɔlɔkɔlɔ} \HM{2} \PH{kɔ̀lɔ̀kɔ̀lɔ̀} \PS{adv.~} \GE{not firm (object smaller than its fitting)} 
\LX{koloŋkoloŋ} \HM{} \PH{kólóŋkólóŋ} \PS{adv.~} \GE{watery; diluted e.g porridge} \VA{kolokolo}
\LX{koltaŋaa} \HM{} \PH{kʊ̀ltáŋáà} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{koltaŋema} \TP{koltaŋemɛɛ} 
\LX{kolyaaraa} \HM{} \PH{kôlyááráá} \PS{n.~} \GE{a type of bird of the suborder alcedines that feeds on fish} \OP{kolyaare} \TP{kolyaarɛɛ} \VA{kɔn-yaara}
\LX{komaale} \HM{} \PH{kòmààlɪ̀} \PS{n.~} \GE{funeral rites} 
\LX{komaaroŋ} \HM{} \PH{kòmáárʊ́ŋ} \PS{n.~} \GE{a funeral where the corpse is not yet buried} \OP{komaara} \TP{komaarɛɛ} 
\LX{kɔmaaroŋ} \HM{} \PH{kɔ̀mààrʊ́ŋ} \PS{n.~} \GE{freshly weeded or hoed-up place} \OP{kɔmaara} \TP{kɔmaarɛɛ} 
\LX{Komaase} \HM{} \PH{kʊ̀máásɪ́} \PS{n.~} \GE{main city in the Ashanti Region of Ghana} \OP{Komaasere} \TP{Komaaserɛɛ} 
\LX{komakoma} \HM{} \PH{kʊ̀màkʊ̀mà} \PS{adj.~} \GE{mild e.g. drink, alcohol} 
\LX{kɔmbanna} \HM{} \PH{kɔ̀mbánná} \PS{n.~} \GE{swimmer} \OP{kɔmbanema} \TP{kɔmbanemɛɛ} 
\LX{kɔmbiri} \HM{} \PH{kɔ́mbírí} \PS{n.~} \GE{drop of water} \OP{kɔmbie} \TP{kɔbiiree} 
\LX{kɔmbɔ} \HM{} \PH{kɔ̀mbɔ́} \PS{n.~} \GE{dialog, conversation} \TP{kɔmbɔrɛɛ} 
\LX{kɔmbɔge} \HM{} \PH{kɔ̀mbɔ̀gɪ̀} \PS{n.~} \GE{water-way} \VA{kɔmbɔgeraa}
\LX{kɔmbɔgeraa} \HM{} \PH{kɔ̀mbɔ́gráá} \PS{n.~} \GE{waterway} \OP{kɔmbɔgre} \TP{kɔmbɔgrɛɛ} \VA{kɔmbɔge}
\LX{kɔmbɔŋenɔ} \HM{} \PH{kɔ́mbɔ́ŋɪ́nɔ́} \PS{n.~} \GE{swimmer} \OP{kɔmbɔŋenemɔ} \TP{kɔmbɔŋenemɛɛ} \VA{kɔmbɔŋena}
\LX{kɔmbɔnne} \HM{} \PH{kɔ̀mbɔnnɪ́̀} \PS{n.~} \GE{Akan language} \TP{kɔmbɔnnɛɛ} 
\LX{kɔmbɔŋɔ} \HM{} \PH{kɔ̀mbɔ̀ŋɔ́} \PS{n.~} \GE{Akan person (one who can swim)} \OP{kɔmbɔnne} \TP{kɔmbɔnnɛɛ} 
\LX{komboroŋ} \HM{} \PH{kʊ̀mbʊ̀rʊ̀ŋ} \PS{adj.~} \GE{wrinkled} 
\LX{kɔmbuli} \HM{} \PH{kɔ̀mbúlí} \PS{n.~} \GE{period of water shortage e.g. during drought} \TP{kɔmbulee} 
\LX{komiri} \HM{} \PH{kòmírí} \PS{n.~} \GE{a string or strip of cloth tied round the wrist or waist of the bereaved during the funeral rites} \VA{koganaa}
\LX{kɔmmaaroŋ} \HM{} \PH{kɔ̌mmáárʊ́ŋ} \PS{n.~} \GE{cold water} 
\LX{kɔmmaaroŋ} \HM{} \PH{kɔ̌mmáárʊ́ŋ} \PS{n.~} \GE{trouble free; peace} 
\LX{kɔmmiiiruŋ} \HM{} \PH{kɔ̀mmíírûŋ} \PS{n.~} \GE{sour water} \TP{kɔmmiiree} 
\LX{kɔmmiri} \HM{} \PH{kɔ̀mmírì} \PS{n.~} \GE{current (running water)} \OP{kɔmmie} \TP{kɔmmiiree} \VA{kombiri}
\LX{kommiruu} \HM{} \PH{kómmírúú} \PS{n.~} \SN \GE{tomato} \SN \GE{eggplant} \SG{kommiruu} \OP{kommie} \TP{kommiiree} 
\LX{kɔmɔkɔmɔ} \HM{} \PH{kɔ̀mɔ̀kɔ̀mɔ̀} \PS{adv.~} \GE{not hard} 
\LX{komoɔreŋ} \HM{} \PH{kómʊ́ɔ́rɪ́ŋ} \PS{n.~} \GE{a type of funeral dirge} 
\LX{kɔmpaala} \HM{} \PH{kɔ́mpáálá} \PS{n.~} \GE{new water e.g bottled water for sale} \TP{kɔmpaalɛɛ} 
\LX{kɔmpare} \HM{} \PH{kɔ́mpárɪ́} \PS{n.~} \SN \GE{the under part of water, residue} \SN \GE{the southern part of Ghana} \OP{kɔmparɛɛ} 
\LX{kɔmpiiri} \HM{} \PH{kɔ́mpíírí} \PS{n.~} \GE{floodwaters} \OP{kɔmpie} \TP{kɔmpiree/kɔmpiiree} 
\LX{kɔmpɔge} \HM{} \PH{kɔ̀mpɔ̀gɪ̀} \PS{n.~} \GE{abundant waters, flood} \OP{kɔmpɔgre} \TP{kɔmpɔgrɛɛ} \VA{kɔmpu᷉o}
\LX{kɔmpu᷉o} \HM{} \PH{kɔmpu᷉o} \PS{adv.~} \GE{period of excessive water e.g during the peak of heavy rains} \VA{kɔmpɔge}
\LX{koŋ} \HM{} \PH{kʊ̀ŋ} \PS{negative marker.~} \GE{not} \XV{N koŋ gaa be zenɛ.} \XE{I will not go there today.} 
\LX{kona} \HM{} \PH{kʊ́ná} \PS{adv.~} \GE{in large quantity, immeasurable} \XV{A kyi ba taa kona, a koŋ kpɛ a bugo.} \XE{There is a great deal of millet; the granary cannot contain it.} 
\LX{konaabo} \HM{} \PH{kònáábʊ́} \PS{n.~} \GE{cow meant for funeral rites} \OP{konii} \TP{koniinee} 
\LX{kɔndaare} \HM{} \PH{kɔ̀ndáárɪ̀} \PS{n.~} \GE{waves} \TP{kɔndaarɛɛ} 
\LX{kɔndaare} \HM{} \PH{kɔ̀ndàárɪ̀} \PS{n.~} \GE{water for sale} \TP{kɔndaarɛɛ} 
\LX{kɔndazugiri} \HM{} \PH{kɔ̀ndàzùgrí} \PS{n.~} \SN \GE{a type of wild dwarf tree, short and stout with complex branches} \SN \GE{the yellow fruit of this plant (edible)} \SG{kɔndazugiruu} \OP{kɔndazugo} \TP{kɔndazugiree} \VA{dazugri}
\LX{kɔndazugovu᷈u᷈} \HM{} \PH{kɔ̀ndàzúgóvù᷈ú᷈} \PS{n.~} \SN \GE{furnace} \SN \GE{hell} \TP{kɔndazugovuunee} 
\LX{kɔndoge} \HM{} \PH{kɔ̌ndʊ́gɪ̀} \PS{n.~} \GE{water pot} \OP{kɔndogre} \TP{kɔndogrɛɛ} 
\LX{kɔndugzie} \HM{} \PH{kɔ̌ndúgzíé} \PS{n.~} \GE{swimming pool} \OP{kɔndugziiri} \TP{kɔndugziiree} 
\LX{Kɔnduoraa} \HM{} \PH{kɔ̌ndúóràá} \PS{n.~} \GE{a river in the Upper West Region of Ghana} \OP{kɔnduori} \TP{kɔnduoree} 
\LX{kɔŋe} \HM{} \PH{kɔ̀ŋ} \PS{v.~} \GE{to contract leprosy} \PD{kɔŋɛɛ, kɔnnɔ, kɔnnɔ, kɔnema, kɔnnaa, kɔnema } 
\LX{kɔŋe} \HM{} \PH{kɔ̀ŋ} \PS{v.~} \GE{to be unavailable} \XV{Ka saaŋ ba miire te maŋ kɔŋe la ko᷉ɔ.} \XE{If it does not rain, we run out of water.} \PD{kɔŋɛɛ, kɔnnɔ, kɔnnɔ, kɔnema, kɔnnaa, kɔnema } 
\LX{kɔŋe} \HM{} \PH{kɔ̂ŋ} \PS{n.~} \GE{hunger} \TP{kɔnnɛɛ} \VA{koŋ (saanchi)}
\LX{konɛɛ konɛɛ} \HM{} \PH{kʊ́nɛ́ɛ́kʊ́nɛ́ɛ́} \PS{adv.~} \GE{on and on in the same direction} \XV{A pɔgelee baŋe la seɛre, o deɛ taaɛ konɛɛ, konɛɛ o naŋ erɛ.} \XE{The girl knows how to dance, she is just dancing gracefully along.} 
\LX{kɔŋgaŋene} \HM{} \PH{kɔ̀ŋgáŋɪ́nɪ́} \PS{adj.~} \GE{a place beyond the waters} 
\LX{koŋi} \HM{} \PH{kǒŋ} \PS{adj.~} \GE{mangled} \OP{konni} 
\LX{koŋi} \HM{} \PH{kòŋ} \PS{v.~} \SN \GE{to cry, to lament} \SN \GE{to crow (rooster)} \PD{koŋee, kono, kono, konimo, konaa, konimo } 
\LX{koŋi} \HM{} \PH{kòŋ} \PS{v.~} \GE{to ferment} \PD{koŋee, kono, kono, konimo, konaa, konimo } 
\LX{kɔŋkaare} \HM{} \PH{kɔ̌ŋkáárɪ́} \PS{n.~} \SN \GE{water that has been bailed out} \SN \GE{water that must be bailed out} 
\LX{kɔŋkare} \HM{} \PH{kɔ̂ŋkárɪ́} \PS{n.~} \GE{plain drinking water} \TP{kɔŋkarɛɛ} 
\LX{koŋkogiri} \HM{} \PH{kóŋkógrí} \PS{n.~} \GE{boundary (of a field)} \OP{koŋkobo} \TP{koŋkogiree} 
\LX{koŋkola} \HM{} \PH{kʊ́ŋkʊ́lá} \PS{n.~} \GE{farming field} \OP{koŋkolle} \TP{koŋkolɛɛ/koŋkollɛɛ} \VA{koŋkolɔ}
\LX{koŋkombiri} \HM{} \PH{kǒŋkǒmbírì} \PS{n.~} \GE{funeral dirge text} \SG{koŋkombiruu} \OP{koŋkombie} \TP{koŋkombiiree} 
\LX{koŋkommoola} \HM{} \PH{kòŋkǒmmʊ́ʊ́lá} \PS{n.~} \GE{incessant crying and for no apparent reason e.g a child} \VA{koŋkommoolɔ}
\LX{konkon} \HM{} \PH{kʊ́nkʊ́n} \PS{adv.~} \GE{completely, flat (of falling)} 
\LX{kɔŋkɔŋ} \HM{} \PH{kɔ́ŋkɔ́ŋ} \PS{n.~} \GE{palm tree} \OP{kɔŋkɔnne} \TP{kɔŋkɔnnɛɛ} 
\LX{kon-kon-kon} \HM{} \PH{kʊ́n-kʊ́n-kʊ́n} \PS{adv.~} \GE{to fall helplessly} 
\LX{kɔŋkɔŋɔ} \HM{} \PH{kɔ̀ŋkɔ̀ŋɔ́} \PS{n.~} \GE{leper} \OP{kɔŋkɔnne} \TP{kɔŋkɔnnɛɛ} 
\LX{kɔŋkpaare} \HM{} \PH{kɔ̀ŋkpáárí} \PS{n.~} \GE{backstroke (swimming)} \OP{kɔŋkpaɛ} 
\LX{kɔnkyɔkyɔgere} \HM{} \PH{kɔ̀nʧɔ́ʧɔ̀grɪ́} \PS{n.~} \GE{swamp} \TP{kɔnkyɔkyɔgerɛɛ} 
\LX{kɔnkyɔrɛɛ} \HM{} \PH{kɔ̌nkyɔ̀rɛ́ɛ́} \PS{n.~} \GE{a type of bath where soap and sponge are not used} \OP{kɔnkyɔrre} \TP{kɔnkyɔrrɛɛ} 
\LX{kɔnkyuuri} \HM{} \PH{kɔ̀nʧúúrì} \PS{n.~} \GE{waterfall} \TP{kɔnkyuuree} 
\LX{kɔnnaŋ} \HM{} \PH{kɔ́nnáŋ} \PS{n.~} \GE{water scarcity} \TP{kɔnnannɛɛ} 
\LX{konne} \HM{} \PH{kʊ́nnɪ̀} \PS{v.~} \GE{to make a short audible exhalation through the nose in to stop inhaling bad smell} \PD{konnɛɛ, konna, konna, konema, konnaa, konema } 
\LX{kɔnne} \HM{} \PH{kɔ̀nnɪ́} \PS{n.~} \GE{leprosy} \OP{kɔnne} \TP{kɔnnɛɛ} 
\LX{konni} \HM{} \PH{kònní} \PS{n.~} \GE{cry} \TP{konnee} 
\LX{konni} \HM{} \PH{kónnì} \PS{v.~} \GE{to catch a falling object in midair (to prevent from falling)} \PD{konnee, konno, konno, konimo, konnaa, konimo } 
\LX{kɔnnimiri} \HM{} \PH{kɔ̂nnímírì} \PS{n.~} \SN \GE{spring} \SN \GE{water hole} \SG{kɔnnimiiruu} \OP{kɔnnimie} \TP{kɔnnimiiree} \VA{bulnimiri}
\LX{koŋŋmɛbaa} \HM{} \PH{kʊ́ŋŋmɛ́báá} \PS{n.~} \GE{back of the head of the dog} 
\LX{konno} \HM{} \PH{kónnò} \PS{n.~} \GE{mouse nest} \VA{konzagera}
\LX{kɔnnoŋ} \HM{} \PH{kɔ́nnʊ́ŋ} \PS{n.~} \GE{the condition of having leprosy} \VA{kɔnnoo}
\LX{kɔnnoɔre} \HM{} \PH{kɔ̀nnʊ́ɔ́rɪ́} \PS{n.~} \SN \GE{beach, coastline} \SN \GE{edge} \OP{kɔnnɔɛ} \TP{kɔnnɔrɛɛ} 
\LX{kɔnnyuuri} \HM{} \PH{kɔ̀nɲúúrì} \PS{n.~} \GE{thirst} \SG{kɔnyuuree} 
\LX{kɔnnyuuri} \HM{} \PH{kɔ́nɲúúrí} \PS{n.~} \GE{potable water, water meant for drinking} 
\LX{konobara} \HM{} \PH{kónóbàrá} \PS{n.~} \GE{lamentation} 
\LX{kɔnɔŋmaara} \HM{} \PH{kɔ́nɔ́ŋmáárá} \PS{n.~} \GE{leader of working farmers in the farm} \OP{kɔnɔŋmaareba} \TP{kɔnɔŋmaarebɛɛ} 
\LX{konono} \HM{} \PH{kʊ́nʊ́nʊ́} \PS{adv.~} \SN \GE{smooth (flour)} \SN \GE{soft (body)} \SN \GE{unasuming (behaviour)} 
\LX{konoɔ} \HM{} \PH{kónʊ́ɔ́} \PS{n.~} \GE{final funeral of the Sisala (an ethnic group in the UWR)} \OP{konoore} \TP{konoorɛɛ} 
\LX{koŋ-ɔɔ-moɔ} \HM{} \PH{kʊ́ŋ-ɔ́ɔ́-mʊ́ɔ́} \PS{n.~} \GE{dog} 
\LX{kon-ɔɔnɛnesoba} \HM{} \PH{kʊ́nɔ́ɔ́nɛ́nɪ̀sʊ́bá} \PS{n.~} \GE{vegetarian} \OP{konɔɔnɛnedeme} \TP{konɔɔnɛnedemee} \VA{kon-ɔɔnɛne}
\LX{kɔnoɔre} \HM{} \PH{kɔ́nʊ́ɔ́rɪ̀} \PS{n.~} \GE{a farmer's portion during farming} \OP{kɔnɔɛ} \TP{kɔnɔrɛɛ} 
\LX{kɔnɔtɛne} \HM{} \PH{kɔ́nɔ́tɛ́nɪ́} \PS{n.~} \GE{earth or soil used to mark out the boundaries of portions of land for weeding} \TP{kɔnɔtɛnɛɛ} 
\LX{kɔnɔtuuro} \HM{} \PH{kɔ́nɔ́túúró} \PS{n.~} \GE{deputy leader of working farmers in the farm} \OP{kɔnɔtuuribo} \TP{kɔnɔtuuribee} 
\LX{kɔnsaaloŋ} \HM{} \PH{kɔ́nsáálʊ́ŋ} \PS{n.~} \GE{plain fluid from a swelling (may be mixed with blood)} \TP{kɔnsaalɛɛ} 
\LX{kɔnsaane} \HM{} \PH{kɔ̌nsáánɪ́} \PS{n.~} \GE{water with marshed poridge} \TP{kɔnsaanɛɛ} 
\LX{kɔnsa᷈a} \HM{} \PH{kɔ̀nsà᷈à} \PS{n.~} \GE{the act of polluting water} 
\LX{kɔnso} \HM{} \PH{kɔ̀nsò} \PS{n.~} \GE{the act of bathing} \TP{kɔnsoree} 
\LX{kɔnsobɔɔdoaa} \HM{} \PH{kɔ̀nsòbɔ̀ɔ̀dʊ́áá} \PS{n.~} \GE{bath towel} \OP{kɔnsobɔɔdoore} \TP{kɔnsobɔɔdorɛɛ} 
\LX{kɔnsɔgerɔ} \HM{} \PH{kɔ̀nsɔ́grɔ́} \PS{n.~} \GE{dirty water} \TP{kɔnsɔgerɛɛ} \VA{kɔnsɔgera}
\LX{kɔnsogɔ} \HM{} \PH{kɔ́nsʊ́gɔ́} \PS{n.~} \SN \GE{the middle of water e.g. a stream} \SN \GE{the act of keeping afloat in water with body upright} \OP{kɔnsogere} \TP{kɔnsogerɛɛ} \VA{kɔnsoga}
\LX{kɔnsolaa} \HM{} \PH{kɔ̀nsòláá} \PS{n.~} \GE{large jar used for bathing} \OP{kɔnsolaare} \TP{kɔnsolaarɛɛ} 
\LX{kɔnsoŋ} \HM{} \PH{kɔ̀nsʊ́ŋ} \PS{n.~} \GE{holy water} \OP{kɔnsonne} \TP{kɔnsonnɛɛ} 
\LX{kɔnsuori} \HM{} \PH{kɔ̀nsúórì} \PS{n.~} \GE{bathwater} \TP{kɔnsuoree} 
\LX{kɔnsuoro} \HM{} \PH{kɔ̀nsúóró} \PS{n.~} \GE{one who is bathing} \OP{kɔnsuoribo} \TP{kɔnsuoribee} 
\LX{kɔntere} \HM{} \PH{kɔ́ntɪ́r} \PS{n.~} \GE{cheek} \OP{kɔnteɛ} \TP{kɔnteerɛɛ} 
\LX{kɔntɛtɛgere} \HM{} \PH{kɔ̀ntɛ́tɛ̀grɪ́} \PS{n.~} \GE{puddles} \OP{kɔntɛtɛgerɛɛ} 
\LX{kɔntɔmbile} \HM{} \PH{kɔ́ntɔ́mbílé} \PS{n.~} \GE{fairy} \OP{kɔntɔmbilii} \TP{kɔntɔmbilee} \VA{kɔntɔmɔ, kɔntɔnzuzeɛ}
\LX{kɔntɔmbɔgɔ} \HM{} \PH{kɔ̀ntɔ̀mbɔ́gɔ́} \PS{n.~} \GE{the act of divining through the power of fairy spirits} \OP{kɔntɔmbɔgɔrɔ} \TP{kɔntɔmbɔgrɛɛ} 
\LX{kɔntɔmɔ} \HM{} \PH{kɔ̀ntɔ́mɔ̀} \PS{n.~} \GE{fairy, dwarf (spiritual being)} \OP{kɔtɔnne} \TP{kɔtɔnnɛɛ} \VA{kɔntɔmbile}
\LX{kɔntɔŋgbelime} \HM{} \PH{kɔ̀ntɔ̀ŋgbèlìmé} \PS{n.~} \GE{diviner's bell} \OP{kɔntɔŋgbelli} \TP{kɔntɔŋgbellee} 
\LX{kɔntɔn-ire} \HM{} \PH{kɔ̀ntɔ̀n-íré} \PS{n.~} \GE{one possessed by a spirit} \OP{kɔntɔn-iribe} \TP{kɔntɔn-iribee} \VA{kɔntɔŋ-ire}
\LX{kɔntɔnkyi} \HM{} \PH{kɔ̀ntɔ̀nʧì} \PS{n.~} \GE{a type of guinea corn with short thick bunch of seed} \TP{kɔntɔnkyiree} 
\LX{kɔntɔnnaa} \HM{} \PH{kɔ̀ntɔ̀nnáá} \PS{n.~} \GE{two-shilling coin} \OP{kɔntɔnnanne} \TP{kɔntɔnnannɛɛ} 
\LX{kɔntɔn-woɔ} \HM{} \PH{kɔ̀ntɔ̀n-wʊ́ɔ́} \PS{n.~} \GE{soothsayer's bag} \OP{kɔntɔn-woore} \TP{kɔntɔn-woorɛɛ} 
\LX{kɔnvɔlɔ} \HM{} \PH{kɔ̀nvɔ̀lɔ́} \PS{n.~} \GE{liquid residue of shea butter} \OP{kɔnvɔlle} \TP{kɔnvɔlɛɛ/kɔnvɔllɛɛ} \VA{kɔvɔlɔ}
\LX{kɔn-yaaraa} \HM{} \PH{kɔ̀n-yáárá} \PS{n.~} \GE{a type of bird that feeds on fish} \OP{kɔn-yaare} \TP{kɔn-yaarɛɛ} \VA{kolyaaraa}
\LX{kɔn-yuori} \HM{} \PH{kɔn-yuori} \PS{n.~} \GE{pot for water} \OP{kɔn-yoe} \TP{kɔn-yoree} 
\LX{konzagera} \HM{} \PH{kónzàgrà} \PS{n.~} \GE{mouse nest} \TP{konzagerɛɛ} \VA{konno, konzage}
\LX{kɔnzuluŋ} \HM{} \PH{kɔ̀nzùlúŋ} \PS{n.~} \GE{deep water} \OP{kɔnzululli} \TP{kɔnzulullee} 
\LX{kɔnzunni} \HM{} \PH{kɔ́nzúnní} \PS{n.~} \GE{debris of water} \TP{kɔnzunnee} 
\LX{koɔ} \HM{} \PH{kʊ̀ɔ̀} \PS{v.~} \GE{to have an upset stomach, intestinal illness} \PD{koɔɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\LX{kɔɔ} \HM{} \PH{kɔ̀ɔ̀} \PS{v.~} \GE{to break} \PD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\LX{kɔɔ} \HM{} \PH{kɔ̀ɔ̀} \PS{v.~} \GE{to dance} \PD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\LX{kɔɔ} \HM{} \PH{kɔ̀ɔ̀} \PS{v.~} \GE{to become charred} \PD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\LX{kɔɔ} \HM{} \PH{kɔ̀ɔ̀} \PS{v.~} \GE{to stampede, sudden flight or dispersion} \PD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\LX{kɔɔ} \HM{} \PH{kɔ̀ɔ́} \PS{adv.~} \GE{one hundred} 
\LX{kɔɔ} \HM{} \PH{kɔ́ɔ́} \PS{adj.~} \GE{appear broken} 
\LX{koɔbo} \HM{} \PH{kʊ́ɔ́bʊ̀} \PS{n.~} \GE{the act of farming} \TP{kɔrɛɛ} 
\LX{kookeree} \HM{} \PH{kòòkéréé} \PS{n.~} \GE{one's best attempt; effort} 
\LX{kɔɔkyaporo} \HM{} \PH{kɔ̀ɔ̀ʧàpòrò} \PS{n.~} \GE{bone marrow} \SG{kɔɔkyaporuu} \OP{kɔɔkyaporri} \TP{kɔɔkyaporree} 
\LX{koɔle} \HM{} \PH{kʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to gather; to bring together} \PD{koɔlɛɛ, koɔlɔ, koɔlɔ, koɔleba, koɔlaa, koɔlɔ } 
\LX{koɔle} \HM{} \PH{kʊ́ɔ́lɪ̀} \PS{v.~} \GE{to get sick} \PD{koɔlɛɛ, koɔlɔ, koɔlɔ, koɔleba, koɔlaa, koɔleba } 
\LX{kɔɔle} \HM{} \PH{kɔ́ɔ́lɪ́} \PS{adj.~} \GE{burnt, charred} \XV{A vu᷉u toloŋ, ka fooŋ deɛnɛ kamaa-kɔɔle ka fo na sɛ.} \XE{The fire is quite hot, and if you are not careful the maize will get burnt.} \OP{kɔɔlɔ} 
\LX{kɔɔlmare} \HM{} \PH{kɔ̀ɔ̀lmàrɪ̀} \PS{v.~} \GE{to hurt verbally e.g. insult, chastise} \PD{kɔɔlmarɛɛ, kɔɔlɔmara, kɔɔlɔmara, kɔɔlemareba, kɔɔlɔmaraa, - } 
\LX{koɔlɔ} \HM{} \PH{kʊ̀ɔ̀lɔ̀} \PS{n.~} \SN \GE{farmland} \SN \GE{environment} \OP{koɔle} \TP{koɔlɛɛ} \VA{koɔla}
\LX{kɔɔloŋ} \HM{} \PH{kɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{hair, fur} \OP{kɔɔlɔ} \TP{kɔɔlɛɛ} 
\LX{koone} \HM{} \PH{kʊ̀ʊ̀nɪ̀} \PS{v.~} \GE{to groan (in pain)} \PD{koonɛɛ, koonɔ/koona, koonɔ/koona, koonema, koonaa, - } 
\LX{kooraa} \HM{} \PH{kʊ́ʊ́ràà} \PS{n.~} \GE{umbilical hernia} \OP{koore} \TP{koorɛɛ} \VA{kpograa, kpogre, kpogrɛɛ}
\LX{koɔraa} \HM{} \PH{kʊ́ɔ́ràà} \PS{n.~} \GE{farmer} \OP{koɔrebɔ} \TP{n/a} 
\LX{koore} \HM{} \PH{kʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to snore} \PD{koorɛɛ, koorɔ/koora, koorɔ/koora, kooreba, kooraa, kooreba } \VA{korŋzaa}
\LX{koore} \HM{} \PH{kʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to collect or move a load at once} \PD{koorɛɛ, koorɔ/koora, koorɔ/koora, kooreba, kooraa, kooreba } \VA{korŋzaa}
\LX{koɔre} \HM{} \PH{kʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to sell} \PD{koɔrɛɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\LX{koɔre} \HM{} \PH{kʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to scrape, to scratch} \PD{koɔrɛɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\LX{koɔre} \HM{} \PH{kʊ́ɔ́rɪ́} \PS{v.~} \GE{to tap} \PD{koɔrɛɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\LX{koɔre} \HM{} \PH{kʊ̀ɔ́rɪ̀} \PS{n.~} \GE{group of stars; galaxy} \SG{koɔroo} \OP{koɔrɛɛ} 
\LX{koɔre} \HM{1} \PH{kʊ́ɔ́rɪ̀} \PS{n.~} \GE{gourd} \OP{kɔɛ} \TP{kɔrɛɛ} 
\LX{koɔre} \HM{2} \PH{kʊ́ɔ́rɪ̀} \PS{n.~} \GE{residue of final beer} \TP{koɔrɛɛ} 
\LX{kɔɔre} \HM{} \PH{kɔ́ɔ́rɪ́} \PS{n.~} \SN \GE{bone} \SN \GE{armlet made from elephant tusk} \OP{kɔbɔ} \TP{kɔɔrɛɛ} 
\LX{kɔɔre} \HM{} \PH{kɔ̀ɔ̀rɪ̀} \PS{v.~} \SN \GE{to delay} \XV{Ta kɔɔre ma, N boɔrɛɛ wĩẽũ.} \XE{Don’t delay me I am in a haste.} \SN \GE{keep long} \XV{Zenɛ a bie kɔɔre la a sakuuri poɔ.} \XE{The child has kept long in school today.} \SN \GE{to be late} \XV{N karembie kɔɔre la sakuuri gaabo ka ba ko o dɔgeroŋ.} \XE{The pupil was late to school and they gave him some punishment.} \PD{kɔɔrɛɛ, kɔɔrɔ, kɔɔrɔ, kɔɔreba, kɔɔraa, kɔɔreba } 
\LX{kɔɔto} \HM{} \PH{kɔ́ɔ́tʊ̀} \PS{n.~} \GE{court} \OP{kɔɔtere} \TP{kɔɔterɛɛ} 
\LX{kɔɔyɛ} \HM{} \PH{kɔ́ɔ́yɛ̀} \PS{n.~} \GE{forest} \OP{kɔɔyɛre} \TP{kɔɔyɛrɛɛ} 
\LX{kɔɔzii} \HM{} \PH{kɔ̀ɔ̀zìì} \PS{n.~} \GE{a type of tiny bird that moves in large groups} \OP{kɔɔziiri} \TP{kɔɔziiree} 
\LX{kopeɛ} \HM{} \PH{kòpɪ́ɛ́} \PS{n.~} \GE{ground immediately surrounding the stage of the corpse} \OP{kopeere} \TP{kopeerɛɛ} 
\LX{kɔpo} \HM{} \PH{kɔ́pʊ̀} \PS{n.~} \GE{cup} \OP{kɔpore} \TP{kɔporɛɛ} 
\LX{ko-pɔge} \HM{} \PH{kò-pɔ́gɪ́} \PS{n.~} \GE{widow} \OP{ko-pɔgeba} \TP{ko-pɔgrɛɛ} 
\LX{ko-puoraa} \HM{} \PH{kò-pùòràá} \PS{n.~} \GE{belated condolence (usually from someone who misses a funeral)} 
\LX{kɔpuori} \HM{} \PH{kɔ́pùòrí} \PS{n.~} \GE{the act of weeding again} \OP{kopoe} \TP{koporee} 
\LX{koraa} \HM{} \PH{kòràá} \PS{n.~} \GE{village; bush settlement} \OP{korri} \TP{korree} 
\LX{kɔraa} \HM{} \PH{kɔ̀ràà} \PS{n.~} \GE{hook} \OP{kɔrre} \TP{kɔrɛɛ} \VA{takɔraa}
\LX{kore} \HM{} \PH{kʊ̀rɪ̀} \PS{v.~} \SN \GE{to be old (not young)} \SN \GE{to be old (not new)} \PD{korɛɛ, kora, kora, koreba, koraa, koreba } 
\LX{kore} \HM{} \PH{kʊ́rɪ́} \PS{adj.~} \GE{old} \OP{kora} \TP{korɛɛ} 
\LX{kɔre} \HM{} \PH{kɔ̀rɪ́} \PS{n.~} \SN \GE{pocket} \SN \GE{bag} \OP{kɔɛ} \TP{kɔrɛɛ} 
\LX{kɔre} \HM{} \PH{kɔ̀rɪ̀} \PS{v.~} \GE{to scoop out; to empty} \XV{A yuori ko᷉ɔ zaa ka ba kɔre.} \XE{They have taken all the water from the pot.} \PD{kɔrɛɛ, kɔrɔ, kɔrɔ, kɔreba, kɔraa, - } 
\LX{kɔre} \HM{} \PH{kɔ̀rɪ̀} \PS{adv.~} \GE{how it sounds when a stone hits an object} 
\LX{kɔre} \HM{} \PH{kɔ́rɪ̀} \PS{v.~} \GE{to cough} \PD{kɔrɛɛ, kɔrɔ, kɔrɔ, kɔreba, kɔraa, kɔreba } 
\LX{kɔrɛɛ} \HM{} \PH{kɔ́rɛ́ɛ́} \PS{adv.~} \GE{moving at once nearer another e.g. child to mother} 
\LX{korekore} \HM{} \PH{kʊ̀rɪ̀kʊ́rɪ́} \PS{v.~} \GE{to break into bits and pieces (grass)} \PD{korkorɛɛ, korekora, korekora, korkoreba, korkoraa, - } 
\LX{korekore} \HM{} \PH{kʊ̀rɪ̀kʊ̀rɪ̀} \PS{adv.~} \GE{many, in great numbers} 
\LX{kɔrema} \HM{} \PH{kɔ̀rɪ́má} \PS{n.~} \GE{partridge} \OP{kɔrre} \TP{kɔrrɛɛ} 
\LX{koremaŋ} \HM{} \PH{kʊ̀rɪ̀màŋ} \PS{adj.~} \GE{unkempt} \VA{koremɔŋ}
\LX{korepere} \HM{} \PH{kʊ́rɪ́pɪ́rɪ́} \PS{n.~} \GE{a type of monkey with thick fur} \OP{korepeɛ} \TP{koreperɛɛ} 
\LX{kɔretɔ} \HM{} \PH{kɔ́rɪ́tɔ́} \PS{n.~} \GE{sickness of animals (fur falls off)} \OP{kɔretɔre} \TP{kɔretɔrɛɛ} 
\LX{korewayiri} \HM{} \PH{kʊ̀rɪ̀wàyírí} \PS{n.~} \GE{one who has has spent all their youth outside their original homeland} 
\LX{kori} \HM{} \PH{kòrì} \PS{v.~} \GE{to embrace, to hug} \PD{koree, koro, koro, koro, koraa, koro } 
\LX{kori} \HM{} \PH{kórí} \PS{n.~} \GE{laziness} \TP{koree} 
\LX{kori} \HM{} \PH{kórí} \PS{n.~} \GE{difficulty, suffering} \TP{koree} 
\LX{kori} \HM{} \PH{kórí} \PS{adv.~} \GE{the manner of a sound from hitting a hard surface} 
\LX{korimo} \HM{} \PH{kòrímò} \PS{n.~} \GE{a type of lizard believed to be poisonous} \OP{korri} \TP{korree} 
\LX{koringyoŋ} \HM{} \PH{kóríndzóŋ} \PS{n.~} \GE{violin} \OP{koringyonni} \TP{koringyonnee} 
\LX{koringyoŋ} \HM{} \PH{koríngyóŋ} \PS{n.~} \GE{violin, (harp)} \OP{koringyonni} \TP{koringyonnee} 
\LX{koro} \HM{} \PH{kòró} \PS{n.~} \GE{pot for storing TZ (thick paste made from flour)} \OP{korri} \TP{koree/korree} 
\LX{korodie} \HM{} \PH{kòródìé} \PS{n.~} \GE{kitchen} \OP{koroderi} \TP{koroderee/korodrree} 
\LX{korombo} \HM{} \PH{kórómbò} \PS{adv.~} \GE{to empty totally, to finish with completely} 
\LX{koroŋ} \HM{} \PH{kʊ́rʊ́ŋ} \PS{adj.~} \GE{a long time ago} 
\LX{kɔroŋ} \HM{} \PH{kɔ́rʊ́ŋ} \PS{n.~} \GE{cough} 
\LX{kɔroŋkpoŋ} \HM{} \PH{kɔ́rʊ́ŋkpóŋ} \PS{n.~} \GE{mycobacterium tuberculosis} \OP{kɔroŋkponni} \TP{kɔroŋkponnee} 
\LX{koroso} \HM{} \PH{kórósó} \PS{adv.~} \GE{lean and dry} \XV{A balee ba dire lɛnso ka o waa a koroso lɛ.} \XE{The puppy does not eat well and that is why it is so lean and bony.} 
\LX{kɔrre} \HM{} \PH{kɔ́rrɪ̀} \PS{v.~} \GE{to slaughter, to cut the throat of} \PD{kɔrrɛɛ, kɔrrɔ, kɔrrɔ, kɔreba, kɔrraa, kɔreba } 
\LX{kɔsɔrre} \HM{} \PH{kɔ́sɔ́rrɪ́} \PS{n.~} \GE{request for farm help} \TP{kɔsɔrɛɛ} 
\LX{kɔtɔ} \HM{} \PH{kɔ́tɔ́} \PS{n.~} \GE{crab} \OP{kɔtɔre} \TP{kɔtɔrɛɛ} 
\LX{kotoloŋ} \HM{} \PH{kòtʊ́lʊ́ŋ} \PS{n.~} \GE{the early part of mourning at the funeral} \OP{kotolɔ/kotola} 
\LX{kotontona} \HM{} \PH{kòtʊ́ntʊ́ná} \PS{n.~} \GE{messenger who announces funeral rites} \OP{kotontonema} \TP{kotontomɛɛ} 
\LX{kotuo} \HM{} \PH{kòtúó} \PS{n.~} \GE{a funeral where mourning is in a frenzy} \OP{kotuuri} \TP{kotuuree} 
\LX{kotuosobɔ} \HM{} \PH{kòtúósʊ́bá} \PS{n.~} \GE{bereaved person (close relative)} \OP{kotuodeme} \VA{kotuosoba}
\LX{kɔvɔlɔ} \HM{} \PH{kɔ̀vɔ̀lɔ́} \PS{n.~} \GE{} \TP{kɔvɔlɛɛ} \VA{kɔnvɔlɔ}
\LX{kovuoro} \HM{} \PH{kòvúóró} \PS{n.~} \GE{condolences by well wishers who could not attend the funeral rites} \TP{kovuoree} 
\LX{Koware} \HM{} \PH{kʊ́wárɪ} \PS{n.~} \GE{homeland of the Kowarɛe clan} 
\LX{Kowaroo} \HM{} \PH{kʊ́wárʊ́ʊ̊} \PS{n.~} \GE{a member of the Koware clan} \OP{Kowarɛɛ} 
\LX{koyeŋi} \HM{} \PH{kòyěŋ} \PS{n.~} \GE{wake keeping (funeral)} \OP{koyenni} \TP{koyenee} 
\LX{koyiri} \HM{} \PH{kóyírí} \PS{n.~} \GE{house of a bereaved family} \OP{koyie} \TP{koyiree} 
\LX{ko᷉ɔ} \HM{} \PH{kʊ́ɔ́} \PS{n.~} \SN \GE{water} \SN \GE{pus} \OP{koɔnɛɛ} \TP{koɔnoo} 
\LX{ko᷉ɔ} \HM{} \PH{kʊ́᷅ɔ́} \PS{adj.~} \GE{water-like, liquid} 
\LX{kpa} \HM{} \PH{kpá} \PS{v.~} \SN \GE{to boil} \SN \GE{to dismantle (usually at the joints)} \SN \GE{to lament; to mourn} \PD{kpaɛ, kpaara, kpaara, kpaareba, kpaaraa, - } 
\LX{kpa} \HM{} \PH{kpà} \PS{v.~} \GE{to lock} \PD{kpaɛ, kpaara, kpaara, kpareba, kpaara, - } 
\LX{kpaa} \HM{} \PH{kpàà} \PS{v.~} \SN \GE{to be unsuccessful} \XV{A karema goɔ la a bie kpaa.} \XE{The teacher was unable to stop the child.} \SN \GE{to leave some of} \XV{Ba di la a bondirii kyɛ ka a kpaa.} \XE{They ate the food and some was left over.} \PD{kpaaɛ, kpaara, kpaara, kpaareba, kpaaraa, - } 
\LX{kpaa} \HM{} \PH{kpáà} \PS{v.~} \GE{to remove by pulling out} \PD{kpaaɛ, kpaara, kpaara, kpaareba, kpaaraa, - } 
\LX{kpaa} \HM{} \PH{kpàá} \PS{adj.~} \GE{leftover} \XV{A baa ŋa ba nɔŋ bondi-kpaa.} \XE{This dog does not like leftover food.} \OP{kpaare} \TP{kpaarɛɛ} 
\LX{kpaa} \HM{} \PH{kpáá} \PS{interj.~} \GE{surprised} 
\LX{kpaagyiri} \HM{} \PH{kpáádzrí} \PS{n.~} \GE{strong whip (made of tendon and leather)} \TP{kpaagyiree} \VA{gyiri}
\LX{kpaakpaa} \HM{} \PH{kpàákpàá} \PS{n.~} \GE{palm kernel} \OP{kpaakpaare} \TP{kpaakpaarɛɛ} 
\LX{kpaakpaabiri} \HM{} \PH{kpàákpàábírì} \PS{n.~} \GE{palm kernel seed} \SG{kpaakpaabiruu} \OP{kpaakpaabie} \TP{kpaakpaabiree/kpaakpaabiiree} \VA{kpaakpaa}
\LX{kpaakyinni} \HM{} \PH{kpááʧìnnì} \PS{v.~} \GE{to leap from a height to land} \PD{kpaakyinnee, kpaarakyinne, kpaakyinne, kpaakyinime, kpaakyinnaa, - } 
\LX{kpaale} \HM{} \PH{kpáálɪ̀} \PS{v.~} \GE{to remind} \PD{kpaalɛɛ, kpaala, kpaala, kpaaleba, kpaalaa, - } 
\LX{kpaalzeɛ} \HM{} \PH{kpààlzɪ̀ɛ́} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{kpaalzeere} \TP{kpaalzeerɛɛ} 
\LX{kpaamoɔ} \HM{} \PH{kpáámʊ́ɔ́} \PS{n.~} \GE{a type of grass} \SG{kpaamoɔnoo} \TP{kpaamoɔnɛɛ} 
\LX{kpaane} \HM{} \PH{kpàànɪ̀} \PS{v.~} \SN \GE{to become tight} \SN \GE{to go hunting} \PD{kpaanɛɛ, kpaana, kpaana, kpaanema, kpaanaa, - } 
\LX{kpaanyo} \HM{} \PH{kpàányʊ̀} \PS{n.~} \GE{nothing} \VA{kpaanyɔ}
\LX{kpaaraa} \HM{} \PH{kpááràá} \PS{n.~} \GE{whip} \OP{kpaare} \TP{kpaarɛɛ} \VA{kpeeraa, kpeere}
\LX{kpaare} \HM{} \PH{kpáárɪ́} \PS{n.~} \GE{nape of neck} \OP{kpaɛ} \TP{kpaarɛɛ} 
\LX{kpaare} \HM{} \PH{kpáárɪ̀} \PS{v.~} \SN \GE{to tap (on a door)} \SN \GE{to roof by knocking nails to hold roof} \SN \GE{to harvest termites (at the end of the process one knocks on the pot containing termites)} \SN \GE{to fly at top speed} \PD{-, -, -, -, -, - } 
\LX{kpaare} \HM{} \PH{kpààrɪ̀} \PS{v.~} \SN \GE{to come together} \SN \GE{to assemble, to meet together} \PD{kpaarɛɛ, kpaara, kpaara, kpaareba, kpaaraa, kpaareba } 
\LX{kpaasere} \HM{} \PH{kpáásɪ́rɪ́} \PS{v.~} \GE{to end up at a point, not returning} \XV{Ba toŋ la a bie ka o deɛ te kpaasere.} \XE{The child they sent has ended up not returning.} \PD{kpaaserɛɛ, kpaaserɛ, kpaaserɛ, kpaaserebɛ, kpaaseraa, - } 
\LX{kpaba} \HM{} \PH{kpàbà} \PS{adv.~} \GE{sweating profusely} 
\LX{kpabogi} \HM{} \PH{kpàbògí} \PS{n.~} \GE{point between the neck and the head} \OP{kpabogri} \TP{kpabogree} 
\LX{kpaga} \HM{} \PH{kpágà} \PS{adv.~} \GE{refuse to do anything} 
\LX{kpage} \HM{} \PH{kpàgɪ́} \PS{adj.~} \GE{thick, strong} 
\LX{kpagelaga} \HM{} \PH{kpàglàgà} \PS{adv.~} \SN \GE{thick} \SN \GE{motionless} 
\LX{kpagele} \HM{} \PH{kpàglɪ̀} \PS{v.~} \GE{to harden} \PD{kpagelɛɛ, kpagela, kpagela, kpageleba, kpagelaa, - } 
\LX{kpageloo} \HM{} \PH{kpáglʊ́ʊ́} \PS{adj.~} \SN \GE{hard, firm} \SN \GE{difficult} \OP{kpagela} 
\LX{kpagyɛle} \HM{} \PH{kpádzɛ́lɪ́} \PS{n.~} \GE{the pointed part of the back of the head} \OP{kpagyɛlɛ} \TP{kpagyɛlɛɛ} 
\LX{kpagyiili} \HM{} \PH{kpàdzíílí} \PS{n.~} \GE{main vein connecting the back of the head and the neck} \OP{kpagyiile} \TP{kpagyiilee} 
\LX{kpakyaga} \HM{} \PH{kpáʧàgá} \PS{adj.~} \GE{between} \OP{kpakyagre} \TP{kpakyagrɛɛ} 
\LX{kpala} \HM{} \PH{kpálà} \PS{n.~} \GE{a type of powder for the eyelashes} \OP{kpalɛɛ} 
\LX{kpala} \HM{} \PH{kpàlá} \PS{n.~} \SN \GE{nakedness, nudity} \SN \GE{the act of engaging in a fight (wrestling)} \TP{kpalɛɛ} 
\LX{kpalaa} \HM{} \PH{kpàlàá} \PS{n.~} \GE{sling} \OP{kpalle} \TP{kpallɛɛ} 
\LX{kpalaŋ} \HM{} \PH{kpàlàŋ} \PS{adj.~} \GE{quickly, at once} 
\LX{kpalaŋkpalaŋ} \HM{} \PH{kpáláŋkpáláŋ} \PS{adv.~} \GE{in a uneventful manner} \XV{A dɔɔ entuo na toŋ la a toma kpalaŋkpalaŋ lɛ.} \XE{That lazy man did the work haphazardly.} 
\LX{kpale} \HM{} \PH{kpálɪ́} \PS{v.~} \GE{to become bare (bald)} \XV{N sa᷉akoma zu zaaŋ kpale.} \XE{My granfather is completely bald.} \PD{kpalɛɛ, kpala, kpala, kpaleba, kpalaa, kpaleba } 
\LX{kpale} \HM{} \PH{kpàlɪ̀} \PS{adj.~} \SN \GE{bare; empty} \SN \GE{for nothing} \XV{O wa la kyɛ kpale lɛ, o ba nyɛ bonzaa.} \XE{She came here for nothing; she did not get anything.} \OP{kpala} \TP{kpalɛɛ} 
\LX{kpale} \HM{} \PH{kpàlɪ̀} \PS{v.~} \GE{to make firm (loose soil)} \PD{kpalɛɛ, kpalla, kpalla,  kpalba, kpallaa, - } 
\LX{kpaleŋkpaa} \HM{} \PH{kpàlɪ̀ŋkpàá} \PS{n.~} \GE{louse} \OP{kpaleŋkpaare} \TP{kpaleŋkpaarɛɛ} 
\LX{kpalle} \HM{} \PH{kpállɪ̀} \PS{v.~} \SN \GE{to remind} \SN \GE{to caution} \SN \GE{to remould the blade of an overused implement (axe)} \PD{kpallɛɛ, kpalla, kpalla, kpalba, kpallaa, - } 
\LX{kpambeɛ} \HM{} \PH{kpǎmbɪ́ɛ̀} \PS{n.~} \GE{spokesman} \OP{kpambeere} \TP{kpambeerɛɛ} 
\LX{kpambiri} \HM{} \PH{kpàmbírí} \PS{n.~} \GE{wooden bell} \OP{kpambie} \TP{kpambiiree} 
\LX{kpammo} \HM{} \PH{kpàmmó} \PS{adv.~} \GE{thick and hard} 
\LX{kpaŋaa} \HM{} \PH{kpàŋàà} \PS{adj.~} \GE{large and hollow} 
\LX{kpane} \HM{} \PH{kpànɪ̀} \PS{n.~} \SN \GE{lead} \SN \GE{spear, lance} \OP{ kpama} 
\LX{kpaŋe} \HM{} \PH{kpáŋ} \PS{v.~} \GE{to become rigid and hard (of a surface)} \PD{kpaŋɛɛ, kpana, kpana, kpanema, kpanaa, - } 
\LX{kpaŋe} \HM{} \PH{kpàŋɪ́} \PS{adj.~} \GE{rigid and hard (surface, object)} \OP{kpama} 
\LX{kpaŋenyaŋa} \HM{} \PH{kpàŋɪ̀nyàŋà} \PS{adv.~} \GE{large and dirty (container)} 
\LX{kpaŋkpane} \HM{} \PH{kpáŋkpánɪ́} \PS{n.~} \GE{arm} \OP{kpaŋkpama} \TP{kpaŋkpannɛɛ} 
\LX{kpaŋkpaŋkpe᷉ɛ} \HM{} \PH{kpǎŋkpǎŋkpɪ́᷈ɛ́᷉} \PS{n.~} \GE{force} \OP{kpaŋkpaŋkpeɛne} \TP{kpaŋkpaŋkpeɛnɛɛ} 
\LX{kpaŋkpannyugbieli} \HM{} \PH{kpàŋkpànnyúgbíélí} \PS{n.~} \GE{elbow} \OP{kpaŋkpannyugbiele} \TP{kpaŋkpannyugbielee} \VA{gbieli}
\LX{kpaŋkpazegle} \HM{} \PH{kpáŋkpázɪ́glí} \PS{n.~} \GE{hard, pimple-like swelling} \OP{kpaŋkpazeglɛ} \TP{kpaŋkpazeglɛɛ} \VA{zaŋkpegiri}
\LX{kpaŋkpele} \HM{} \PH{kpáŋkpɪ́lɪ́} \PS{adj.~} \GE{extremely hard} 
\LX{kpaŋkpire} \HM{} \PH{kpǎŋkpírè} \PS{n.~} \GE{race, running competition ( typically used with ŋmɛ)} \XV{Ba ŋmɛ la kpaŋkpire te le dɔgele taa.} \XE{They raced neck and neck and fell over each other.} \OP{kpaŋkpirri} \TP{kpaŋkpiree/kpaŋkpirree} 
\LX{kpaŋkpoli} \HM{} \PH{kpáŋkpòlì} \PS{n.~} \GE{chicken without feathers} \OP{kpaŋkpolo} \TP{kpaŋkpolee} \VA{kpoŋkpoli}
\LX{kpanne} \HM{} \PH{kpànnɪ̀} \PS{v.~} \SN \GE{to hypnotize} \SN \GE{to be overwhelmed by cold} \PD{kpannɛɛ, kpanna, kpanna, kpanema, kpannaa, - } 
\LX{kpantoli} \HM{} \PH{kpántólí} \PS{n.~} \SN \GE{patch of hair as a hairstyle} \SN \GE{crest (of a bird)} \OP{kpantolo} \TP{kpantolee} \VA{kpontoli; kpontolo' kpontolee}
\LX{kparata} \HM{} \PH{kpárátá} \PS{adv.~} \GE{lie prostrate; flat with face upwards} 
\LX{kpardiiraa} \HM{} \PH{kpǎrdííráá} \PS{n.~} \GE{smock with multiple folds} \OP{kpardiiri} \TP{kpardiiree} 
\LX{Kpare} \HM{} \PH{kpárɪ́} \PS{n.~} \GE{name of a town in the Upper West Region} 
\LX{kpare} \HM{} \PH{kpárɪ́} \PS{v.~} \SN \GE{to tie together (to make a bow)} \SN \GE{to suffer, to stuggle on the ground} \PD{kparɛɛ, kpara, kpara, kpareba, kparaa, - } 
\LX{kpare} \HM{} \PH{kpàrɪ̀} \PS{v.~} \SN \GE{to pair up} \SN \GE{closed to} \SN \GE{to ride one bike or animal together (two or more people)} \PD{kparɛɛ, kpara, kpara, kpareba, kparaa, - } 
\LX{kparoo} \HM{} \PH{kpàrʊ́ʊ̀} \PS{n.~} \GE{top (clothing)} \OP{kparre} \TP{kparrɛɛ} 
\LX{kparpi᷉i} \HM{} \PH{kpàrpí᷉ì᷉} \PS{n.~} \GE{needle} \OP{kparpiime} \TP{kparpiimee/kparpiinee} 
\LX{kparsirikyi} \HM{} \PH{kpǎrsíríʧì} \PS{n.~} \GE{expensive beautiful gown} \OP{kparsirikyiri} \TP{kparsirikyiree} 
\LX{kparyoluŋ} \HM{} \PH{kpǎryòlǔŋ} \PS{n.~} \GE{top wear without a supporting underwear} \OP{kparyolli} \TP{kparyollee} 
\LX{kpateɛrɛ} \HM{} \PH{kpàtɪ́ɛ́rɛ́} \PS{n.~} \GE{supporter, backer} \OP{kpateɛrebɛ} \TP{kpateɛrebɛɛ} 
\LX{kpatere} \HM{} \PH{kpàtɪ́rɪ́} \PS{n.~} \SN \GE{gorilla} \SN \GE{chimpanzee} \OP{kpateɛ} \TP{kpateerɛɛ} 
\LX{kpawerɛ} \HM{} \PH{kpàwɪ́rɛ́} \PS{n.~} \GE{back case; not concerned} 
\LX{kpawoɔ} \HM{} \PH{kpàwʊ́ɔ} \PS{n.~} \GE{hat (especially one made from cloth) (northern Dagaare dialect)} \OP{kpawoore} \TP{kpawoorɛɛ} 
\LX{kpa᷉a} \HM{} \PH{kpà᷉à᷉} \PS{v.~} \GE{to warn} \PD{kpaa᷉ɛ, kpaana, kpaana, kpaanema, kpaanaa, kpaanema } 
\LX{kpa᷉a} \HM{} \PH{kpá᷉à᷉} \PS{v.~} \GE{cook very palatable food, fry (saucetew)} \PD{kpa᷉aɛ, kpaana, kpaana, kpaanema, kpaanaa, - } 
\LX{kpa᷉a toori} \HM{} \PH{kpà᷉à᷉ tòòrí} \PS{v.~} \SN \GE{to insist} \SN \GE{to harden one's heart} \PD{-, -, -, -, -, - } 
\LX{kpa᷉aloɔraa} \HM{} \PH{kpá᷉álʊ́ɔ́ráá} \PS{n.~} \GE{a type of small brown bird} \OP{kpa᷉aloɔre} \TP{kpa᷉aloɔrɛɛ} 
\LX{kpa᷈ao} \HM{} \PH{kpá᷉áʊ́} \PS{n.~} \GE{guinea fowl} \OP{kpinni} \TP{kpinnee} 
\LX{kpe} \HM{} \PH{kpɪ́} \PS{v.~} \SN \GE{to shell a bit of (maize seed from cob)} \SN \GE{to wink (to give someone a sign)} \PD{kpeɛ, kpeerɛ, kpeerɛ, kpeerebɛ, kpeeraa, - } 
\LX{kpɛ} \HM{} \PH{kpɛ̀} \PS{v.~} \SN \GE{to enter} \SN \GE{to dwell in, to inhabit} \SN \GE{to gain fertility (farmland)} \SN \GE{to fit (dress)} \SN \GE{to be well attended} \SN \GE{to feel hurt} \SN \GE{to come around (of the a season)} \PD{kpɛɛ, kpeɛrɛ, kpeɛrɛ, kpeɛrebɛ, kpeɛraa, - } 
\LX{kpɛ bire} \HM{} \PH{kpɛ̀ bìrè} \PS{v.~} \GE{to struggle, to wrestle, to contend} \PD{-, -, -, -, -, - } 
\LX{kpɛ nyuo} \HM{} \PH{kpɛ̀ ɲúó} \PS{v.~} \GE{to seek protection from} \PD{-, -, -, -, -, - } 
\LX{kpɛde} \HM{} \PH{kpɛ́dɪ́} \PS{n.~} \GE{ownership (wealth)} 
\LX{kpɛɛ} \HM{} \PH{kpɛ́ɛ̀} \PS{n.~} \GE{malt} \TP{kpɛɛrɛɛ} 
\LX{kpeebɛ} \HM{} \PH{kpɪ́ɪ́bɛ̀} \PS{adj.~} \GE{orphaned} \OP{kpeere} \TP{kpeerɛɛ} 
\LX{kpeebo} \HM{} \PH{kpɪ̀ɪ̀bʊ́} \PS{n.~} \GE{orphan} \OP{kpeere} \TP{kpeerɛɛ} 
\LX{kpeele} \HM{} \PH{kpɪ́ɪ́lɪ̀} \PS{v.~} \GE{to bend slightly to the side} \PD{kpeelɛɛ, kpeelɛ, kpeelɛ, kpeelbɛ, kpeelaa, - } 
\LX{kpeele} \HM{} \PH{kpɪ̀ɪ́lɪ̀} \PS{adj.~} \GE{small, tiny, immature} \OP{kpeelɛ} 
\LX{kpeeloŋ} \HM{} \PH{kpɪ̀ɪ̀lʊ́ŋ} \PS{n.~} \GE{the state of being an orphan} 
\LX{kpeeloŋ} \HM{} \PH{kpɪ́ɪ́lʊ́ŋ} \PS{n.~} \GE{hip joint} \OP{kpeelɛ} \TP{kpeelɛɛ} 
\LX{kpeemɛ} \HM{} \PH{kpɪ́ɪ́mɛ́} \PS{n.~} \GE{departed soul, the dead} \OP{kpenne} \TP{kpeemɛɛ/kpennɛɛ} 
\LX{kpeɛmɛ} \HM{} \PH{kpɪ́ɛ́mɛ́} \PS{n.~} \GE{older sister} \OP{kpeɛmine} \TP{kpeɛmɛɛ} \VA{kpe᷉ɛ᷉ɛ}
\LX{kpeɛŋaa} \HM{} \PH{kpɪ́ɛ́ŋáá} \PS{adj.~} \SN \GE{hard, strong, difficult} \SN \GE{well, healthy} \OP{kpeɛne} 
\LX{kpeeraa} \HM{} \PH{kpɪ́ɪ́ráà} \PS{n.~} \GE{whip} \OP{kpeere} \TP{kpeerɛɛ} \VA{kpaaraa}
\LX{kpeɛre} \HM{} \PH{kpɪ̀ɛ̀rɪ̀} \PS{n.~} \GE{popular entrance of a rodent's hole} \TP{kpeɛrɛɛ} 
\LX{kpeɛrentɔ} \HM{} \PH{kpɪ́ɛ́rɪ́ntɔ̀} \PS{n.~} \GE{neighbour} 
\LX{kpegeke} \HM{} \PH{kpegeke} \PS{adv.~} \GE{direct to the point (usually unleashing pain)} 
\LX{kpɛgelaa} \HM{} \PH{kpɛ́gláá} \PS{n.~} \SN \GE{supporting object (for a boiling pot on a hearth.)} \SN \GE{attachment} \OP{kpɛgele} \TP{kpɛgelɛɛ} 
\LX{kpɛgele} \HM{} \PH{kpɛ̀glɪ̀} \PS{v.~} \SN \GE{to place by the side (provide support for a cooking pot on the hearth)} \SN \GE{to adjoin} \PD{kpɛgelɛɛ, kpɛgelɛ, kpɛgelɛ, kpɛgelebɛ, kpɛgelaa, - } 
\LX{kpɛgelɛŋ} \HM{} \PH{kpɛ̀glɛ̀ŋ} \PS{adv.~} \GE{loose and flabby (diseased arm)} 
\LX{kpegere} \HM{} \PH{kpɪ́grɪ̀} \PS{v.~} \GE{to be curious (asking questions to find information)} \PD{kpegerɛɛ, kpegerɛ, kpegerɛ, kpegerebɛ, kpegeraa, kpegerebɛ } 
\LX{kpɛgere} \HM{} \PH{kpɛ́grì} \PS{v.~} \GE{to disturb or worry someone} \PD{kpɛgerɛɛ, kpɛgerɛ, kpɛgerɛ, kpɛgerebɛ, kpɛgeraa, - } 
\LX{kpegileŋ} \HM{} \PH{kpégléŋ} \PS{adv.~} \GE{small but compact} 
\LX{kpegiri} \HM{} \PH{kpégrí} \PS{n.~} \GE{gizzard} \OP{kpege} \TP{kpegiree} 
\LX{kpegiri} \HM{} \PH{kpégrì} \PS{v.~} \SN \GE{to shiver (from cold)} \SN \GE{to move one's torso quickly and vigorously (usually in dancing)} \PD{kpegiree, kpegire, kpegire, kpegiribe, kpegiraa, - } 
\LX{kpegiruu} \HM{} \PH{kpégrúú} \PS{n.~} \GE{a type of dance} 
\LX{kpekpe} \HM{} \PH{kpɪ́kpɪ́} \PS{n.~} \GE{pair of scissors} \OP{kpekpere} \TP{kpekperɛɛ} 
\LX{kpele} \HM{} \PH{kpɪ́lɪ́} \PS{v.~} \SN \GE{to develop a bad eye (black of eye looking whitish)} \SN \GE{to cause to be bare (growth)} \PD{kpelɛɛ, kpelɛ, kpelɛ, kpelbɛ, kpelaa, - } 
\LX{kpele} \HM{} \PH{kpɪ̀lɪ̀} \PS{adj.~} \GE{hard, difficult} \OP{kpelɛ} \TP{kpelɛɛ} 
\LX{kpele} \HM{} \PH{kpélé} \PS{adv.~} \GE{break completely} \VA{kpelekpele}
\LX{kpelɛ} \HM{} \PH{kpɪ́lɛ́} \PS{n.~} \GE{nothing (typically used with negative forms 'ba', koŋ)} \XV{Ka fooŋ gaa a bondaana ŋa zie kyɛ fo koŋ nyɛ kpelɛ zaa.} \XE{Even if you go to this this rich you will not get anything.} 
\LX{kpɛle} \HM{} \PH{kpɛ́lɪ́} \PS{v.~} \SN \GE{to faint} \SN \GE{to crave} \XV{O nyɛ la pegebil velaa kaŋa a deɛ kpɛlɛ o yɛlɛ.} \XE{He is just crazy about a beautiful girl he saw.} \PD{kpɛlɛɛ, kpɛlɛ, kpɛlɛ, kpɛlebɛ, kpɛlaa, kpɛlebɛ } 
\LX{kpɛle} \HM{} \PH{kpɛ́lɪ̀} \PS{v.~} \GE{to pluck} \PD{kpɛlɛɛ, kpɛlɛ, kpɛlɛ, kpɛlebɛ, kpɛlaa, kpɛlebɛ } 
\LX{kpɛlemɛ} \HM{} \PH{kpɛ́lɪ́mɛ̀} \PS{adj.~} \GE{snatch, do or act at once (hungry dog snatching a bone)} 
\LX{kpɛlɛŋ} \HM{} \PH{kpɛ́lɛ́ŋ} \PS{adv.~} \GE{everything} 
\LX{kpɛlɛŋ kpɛlɛŋ} \HM{} \PH{kpɛ́lɛ́ŋkpɛ̀lɛ̀ŋ} \PS{adv.~} \GE{finish completely} 
\LX{kpeleŋkpeleŋ} \HM{} \PH{kpèlèŋkpéléŋ} \PS{adj.~} \GE{uncompromising state (quarel)} 
\LX{kpɛlɛŋkpɛlɛŋ} \HM{} \PH{kpɛ́lɛ́ŋkpɛ́lɛ́ŋ} \PS{adv.~} \GE{finished copletely} \VA{[kpɛ́lɛ́ŋkpɛ̀lɛ̀ŋ]}
\LX{kpɛlewɛlɛ} \HM{} \PH{kpɛ́lɪ́wɛ́lɛ́} \PS{adv.~} \GE{not firm or strong} 
\LX{kpɛlewɛlɛ} \HM{} \PH{kpɛ́lwɛ́lɛ́} \PS{adv.~} \GE{weak and loose (pair of scissors)} 
\LX{kpeli} \HM{} \PH{kpélí} \PS{n.~} \GE{throw of dice in the game bie} \OP{kpele} \TP{kpelee} \VA{kpoli}
\LX{kpell} \HM{} \PH{kpéll} \PS{adv.~} \GE{breaking of a hard objecct} 
\LX{kpell} \HM{} \PH{kpɪ́ll} \PS{adv.~} \GE{spread out openly and in great quantity} 
\LX{kpelle} \HM{} \PH{kpɪ́llɪ̀} \PS{n.~} \GE{grass strings worn on the waist} \TP{kpellɛɛ} 
\LX{kpɛlle} \HM{} \PH{kpɛ́llɪ́} \PS{n.~} \GE{loud and sharp cry} \TP{kpɛllɛɛ} 
\LX{kpeme} \HM{} \PH{kpèmé} \PS{n.~} \GE{grass used for making hat} \OP{kpemee} \TP{kpennee} 
\LX{kpeme} \HM{} \PH{kpémé} \PS{adv.~} \GE{exerting something hefty (high price)} 
\LX{kpememe} \HM{} \PH{kpɪ́mɪ́mɪ́} \PS{adv.~} \GE{very sweet} 
\LX{kpɛnezu} \HM{} \PH{kpɛ́nɪ́zú} \PS{n.~} \GE{a small black blood-sucking fly; causes river blindness} \OP{kpɛnezuri} \TP{kpɛnezurree} 
\LX{kpeŋkpegiri} \HM{} \PH{kpéŋkpégrí} \PS{n.~} \GE{highest point (xylophone sound)} \OP{kpeŋkpege} \TP{kpeŋkpegiree} 
\LX{kpɛŋkpɛlemɛ} \HM{} \PH{kpɛ̌ŋkpɛ́lɪ́mɛ́} \PS{n.~} \GE{bits and pieces} \XV{A bɔgebogera die kpɛŋkpɛlemɛ ba di ba.} \XE{There are very many pieces of items in the diviner's room.} \TP{kpɛŋkpɛlemɛɛ} 
\LX{kpɛŋkpɛlɛŋ} \HM{} \PH{kpɛ́ŋkpɛ́lɛ́ŋ} \PS{adv.~} \GE{of a hanging bit} 
\LX{kpeŋkperɛŋ} \HM{} \PH{kpɪ́ŋkpɪ́rɛ́ŋ} \PS{adv.~} \GE{smallish but strong (short person)} 
\LX{kpɛŋkpɛrɛŋ} \HM{} \PH{kpɛ́ŋkpɛ́rɛ́ŋ} \PS{adj.~} \GE{spilled all over} 
\LX{kpenne} \HM{} \PH{kpɪ́nnɪ̀} \PS{v.~} \GE{to break off bits of the tip (of a fire log)} \PD{kpennɛɛ, kpennɛ, kpennɛ, kpenemɛ, kpennaa, kpenemɛ } 
\LX{kpenne} \HM{} \PH{kpɪ́nnɪ̀} \PS{n.~} \GE{ancestor} \OP{kpemɛ} \TP{kpennɛɛ} 
\LX{kperaa} \HM{} \PH{kpɪ́ráà} \PS{n.~} \GE{disease of the hip} \OP{kperre} 
\LX{kperaa} \HM{} \PH{kpɪ̀ràà} \PS{adv.~} \GE{exposed, seen clearly} 
\LX{kpere} \HM{} \PH{kpɪ̀rɪ̀} \PS{v.~} \GE{to be stunted in growth or development} \PD{kperɛɛ, kperɛ, kperɛ, kperebɛ, kperaa, - } 
\LX{kpere} \HM{} \PH{kpɪ́rɪ̀} \PS{v.~} \SN \GE{to break off an attached bit from the main body} \SN \GE{to sneeze} \PD{kperɛɛ, kperɛ, kperɛ, kperebɛ, kperaa, - } 
\LX{kpere} \HM{} \PH{kpɪ́rɪ́} \PS{v.~} \GE{to litter} \PD{kperɛɛ, kperɛ, kperɛ, kperebɛ, kperaa, - } 
\LX{kperɛ} \HM{} \PH{kpɪ́rɛ̀} \PS{adj.~} \SN \GE{small} \SN \GE{worn out (broom)} 
\LX{kpɛre} \HM{} \PH{kpɛ́rɪ́} \PS{v.~} \SN \GE{to spill} \SN \GE{to cook a small quantity of} \PD{kpɛrɛɛ, kpɛrɛ, kpɛrɛ, kpɛrebɛ, kpɛraa, - } 
\LX{kperefene} \HM{} \PH{kpɪ́rɪ́fɪ́nɪ́} \PS{adj.~} \GE{muscular and stocky} 
\LX{kperekpere} \HM{} \PH{kpɪ̀rɪ̀kpɪ́rɪ́} \PS{v.~} \GE{to struggle} \PD{kperekperɛɛ, kperekperɛ, kperekperɛ, kperekperebɛ, kperekperaa, - } 
\LX{kperekpere} \HM{} \PH{kpɪ̀rɪ̀kpɪ̀rɪ̀} \PS{adv.~} \GE{rough, forceful} \XV{O bie ŋa waa la kperekpere zuo, o ŋmare la a kyaanaa.} \XE{This child is too rough; he will break the mirror.} 
\LX{kperɛkperɛ} \HM{} \PH{kpɪ́rɛ̀kpírɛ̀} \PS{adj.~} \GE{little bits} 
\LX{kpɛrɛkpɛrɛ} \HM{} \PH{kpɛ́rɛ́kpɛ́rɛ́} \PS{adv.~} \GE{bits and pieces} 
\LX{kperemɛŋ} \HM{} \PH{kpɪ́rɪ̀mɛ́ŋ} \PS{adv.~} \GE{stunted, short, and small} 
\LX{kperemene} \HM{} \PH{kpɪ̀rìmɪ̀nɪ̀} \PS{adj.~} \GE{all over (fruits on a tree or scratches on the body)} \XV{A maŋgo wɔŋe,ɛ la kperemenee lɛ.} \XE{The mango tree has borne a lot of fruit.} \VA{kpurimunu}
\LX{kpɛrɛndɛŋ} \HM{} \PH{kpɛ́rɛ́ndɛ́ŋ} \PS{adv.~} \GE{best of (deed)} \XV{O maale la a bondirii de wa biŋi kpɛndɛŋ.} \XE{She cooked the food and served it neatly.} 
\LX{kpɛrɛŋkpɛrɛŋ} \HM{} \PH{kpɛ́rɛ́ŋkpɛ́rɛ́ŋ} \PS{adv.~} \GE{smeared all over, in a mess} \XV{A bie le la mɛmɛrɛɛ poɔ a e o zaa kpɛrɛŋkpɛrɛŋ lɛ} \XE{The boy fell in the mud and is soiled all over.} 
\LX{kperese} \HM{} \PH{kpírɪ́sɪ́} \PS{adv.~} \GE{very sweet} 
\LX{kpɛrewɛrɛ} \HM{} \PH{kpɛ́rɪ́wɛ́rɛ́} \PS{adv.~} \GE{small quantity} \XV{Zenɛ N koŋ to᷉ɔ nɛne da, n libie deɛ waa la kpɛrewɛrɛɛ lɛ.} \XE{Today I cannot buy meat; my money is very small.} \VA{kpɛrewɛrɛɛ}
\LX{kperi} \HM{} \PH{kpérí} \PS{v.~} \GE{to chop (meat, tree branches)} \PD{kperee, kpere, kpere, kperibee, kperaa, - } 
\LX{kperime} \HM{} \PH{kpérímé} \PS{n.~} \GE{type of food (full of chopped meat)} \TP{kperimee} 
\LX{kperiwere} \HM{} \PH{kpéríwéré} \PS{adv.~} \GE{few, not much} 
\LX{kpɛsenneŋenɛ} \HM{} \PH{kpɛ́sɪ́nnɪ́ŋɪ́nɛ́} \PS{n.~} \GE{pulp from pito malt} 
\LX{kpɛtaa} \HM{} \PH{kpɛ́táá} \PS{n.~} \SN \GE{be suited to; fit well} \SN \GE{a union in which the group members work for each other in turns} \OP{kpɛtaare} \TP{kpɛtaarɛɛ} 
\LX{kpɛwe᷉ɛ} \HM{} \PH{kpɛ́wɪ́᷉ɛ́} \PS{n.~} \GE{water from prepared malt} \OP{kpɛweɛne} \TP{kpɛweɛnɛɛ} 
\LX{kpɛzie} \HM{} \PH{kpɛ̀zíé} \PS{n.~} \GE{abode, living place} \OP{kpɛziiri} \TP{kpɛziiree} 
\LX{kpe᷉ɛ} \HM{} \PH{kpɪ́᷉ɛ́} \PS{n.~} \GE{strength, power, force} \OP{kpeɛne} \TP{kpeɛnɛɛ} \VA{kpeɛ᷉o}
\LX{kpe᷉ɛ} \HM{} \PH{kpɪ̀᷉ɛ̀} \PS{v.~} \GE{to be older than} \PD{kpeɛ᷉ɛ, kpeɛnɛ, kpeɛnɛ, kpeɛnemɛ, kpeɛnaa, - } 
\LX{kpe᷉ekpe᷉e} \HM{} \PH{kpɪ́᷉ɪ́kpɪ́᷉ɪ́} \PS{adv.~} \GE{not bright (fire)} 
\LX{kpe᷉ɛo} \HM{} \PH{kpɪ́᷉ɛ́ʊ̀} \PS{n.~} \GE{strength} \OP{kpeɛne} \VA{kpe᷉ɛ}
\LX{kpi} \HM{} \PH{kpî} \PS{v.~} \SN \GE{to die, to be dead} \SN \GE{to wither (plant)} \PD{kpie, kpiire, kpiire, kpiiribe, kpiiraa, kpiiribe } 
\LX{kpi mare} \HM{} \PH{kpì màrɪ̀} \PS{v.~} \GE{to cling to, to persist} \PD{kpi marɛɛ, kpiire mara, kpimara, kpimareba, kpimaraa, - } 
\LX{kpibifi} \HM{} \PH{kpíbífí} \PS{adj.~} \GE{small, tiny} 
\LX{kpieni} \HM{} \PH{kpíéní} \PS{adj.~} \GE{snatch (hawk picking a chicken)} 
\LX{kpiiri} \HM{} \PH{kpììrì} \PS{v.~} \GE{to sew the end of; to hem (dress)} \PD{kpiiree, kpiire, kpiire, kpiiribe, kpiiraa, kpiiribe } 
\LX{kpimare} \HM{} \PH{kpìmàrɪ̀} \PS{v.~} \GE{to be glued to (work)} \PD{-, -, -, -, -, - } 
\LX{kpinni} \HM{} \PH{kpìnnì} \PS{v.~} \GE{to quench, to extinguish} \PD{kpinnee, kpinne, kpinne, kpinime, kpinnaa, kpinime } 
\LX{kpinni} \HM{} \PH{kpínnì} \PS{n.~} \GE{remind} \VA{pinni}
\LX{kpirikpiri} \HM{} \PH{kpíríkpírí} \PS{adv.~} \GE{very black} \VA{malmal, bɛmbɛŋ}
\LX{kpiroo} \HM{} \PH{kpíróó} \PS{adv.~} \GE{protruding in a small manner} 
\LX{kpo} \HM{} \PH{kpo} \PS{v.~} \SN \GE{to break open an egg} \SN \GE{to remove the bark of a tree} \PD{kpoe, kpuoro, kpuoro, kpuoribo, kpuoraa, - } 
\LX{kpo} \HM{} \PH{kpô} \PS{n.~} \SN \GE{funeral donation} \SN \GE{meeting place} \OP{kpori} \TP{kporee} 
\LX{kpɔge} \HM{} \PH{kpɔ́gɪ́} \PS{v.~} \GE{to develop a lump; to swell up} \PD{kpɔgɛɛ, kpɔgrɔ, kpɔgrɔ, kpɔgreba, kpɔgraa, - } 
\LX{kpɔge} \HM{} \PH{kpɔ́gɪ̀} \PS{v.~} \GE{to carry, to pick up wholly (mud)} \PD{kpɔgɛɛ, kpɔgera, kpɔgera, kpɔgereba, kpɔgeraa, - } 
\LX{kpɔge} \HM{} \PH{kpɔ́gɪ́} \PS{v.~} \GE{carry (pl)  carrying muddy stuff severally} \PD{kpɔgɛɛ, -, -, -, -, - } 
\LX{kpogeraa} \HM{} \PH{kpʊ́gráá} \PS{n.~} \GE{haenia} \OP{kpogere} \TP{kpogerɛɛ} 
\LX{kpɔgeraa} \HM{} \PH{kpɔ́gràá} \PS{n.~} \GE{a type of tree with hard but edible fruit} \OP{kpɔgere} \TP{kpɔgerɛɛ} \VA{kpɔgetere}
\LX{kpɔgere} \HM{} \PH{kpɔ́grɪ́} \PS{n.~} \GE{lump} \OP{kpɔgɔ} \TP{kpɔgerɛɛ} 
\LX{kpɔgetere} \HM{} \PH{kpɔ́gɪ́tɪ́rɪ́} \PS{n.~} \GE{} \VA{kpɔgeraa}
\LX{kpogi} \HM{} \PH{kpògì} \PS{v.~} \GE{to bully} \PD{kpogee, kpogro, kpogro, kpogribo, kpograa, - } 
\LX{kpogili} \HM{} \PH{kpòglì} \PS{n.~} \GE{helmet} \OP{kpogilo} \TP{kpogilee} 
\LX{kpogili} \HM{} \PH{kpòglì} \PS{v.~} \GE{to mature, to grow big} \PD{kpogilee, kpogilo, kpogilo, kpogilibo, kpogilaa, - } 
\LX{kpogili} \HM{} \PH{kpóglí} \PS{v.~} \GE{to wait without getting attention; to sit aimlessly} \PD{kpogilee, kpogilo, kpogilo, kpogilibo, kpogilaa, - } 
\LX{kpogilo} \HM{} \PH{kpóglò} \PS{n.~} \GE{boiled balls of bean and corn flour} \SG{kpogiluu} \TP{kpogilee} 
\LX{kpogilokolo} \HM{} \PH{kpòglòkòlò} \PS{adv.~} \GE{appearance (many big items)} 
\LX{kpogiri} \HM{} \PH{kpógrí} \PS{v.~} \GE{to tap the inner side of a mortar (creating a beat)} \PD{kpogiree, kpogiro, kpogiro, kpogiribo, kpogiraa, - } 
\LX{kpogo} \HM{} \PH{kpógó} \PS{adv.~} \GE{refusing utterly to do something; (run an erand)} 
\LX{kpokpo} \HM{} \PH{kpókpó} \PS{adj.~} \GE{permanent} 
\LX{kpokpori} \HM{} \PH{kpòkpórí} \PS{n.~} \GE{mound} \OP{kpokporee} \TP{kpokporuu} 
\LX{kpolaa} \HM{} \PH{kpʊ̀làà} \PS{adj.~} \GE{large and blunt} 
\LX{kpolaa} \HM{} \PH{kpòlàà} \PS{adv.~} \GE{not hidden (truth)} 
\LX{kpolee} \HM{} \PH{kpóléé} \PS{adj.~} \GE{small and round} \OP{kpolee kpolee} 
\LX{kpɔlɛɛ} \HM{} \PH{kpɔ̀lɛ̀ɛ̀} \PS{adv.~} \GE{blump} 
\LX{kpɔlɛɛkpɔlɛɛ} \HM{} \PH{kpɔ́lɛ́ɛ́kpɔ́lɛ́ɛ́} \PS{adv.~} \GE{appearing with bulging parts} 
\LX{kpoli} \HM{} \PH{kpólí} \PS{v.~} \SN \GE{to get to extreme; (red hot metal)} \SN \GE{to go bad, to spoil (rotten eggs)} \PD{kpolee, kpolo, kpolo, kpolibo, kpolaa, - } 
\LX{kpoli} \HM{} \PH{kpòlì} \PS{adj.~} \GE{bare, naked} \OP{kpolo} \TP{kpolee/kpollee} 
\LX{kpoli} \HM{} \PH{kpólí} \PS{n.~} \GE{a player letting go a dice of the opponent in the game of "bie" as a sign of total control of the game} \OP{kpoli} \TP{kpolee} \VA{kpeli}
\LX{kpolikpala} \HM{} \PH{kpòlìkpàlå} \PS{adv.~} \GE{appearance of something big with appendages} 
\LX{kpoliŋkpoori} \HM{} \PH{kpólíŋkpòòrí} \PS{n.~} \GE{a type of small ants} \OP{kpoliŋkpogo} \TP{kpoliŋkpooree} 
\LX{kpolo} \HM{} \PH{kpòló} \PS{n.~} \GE{termites (trapped for chicken feed.)} \SG{kpoluu} \TP{kpolee} 
\LX{kpolɔ} \HM{} \PH{kpʊ̀lɔ̀} \PS{n.~} \GE{inner feeling; anger} \VA{kpola}
\LX{kpoloŋkpoloŋ} \HM{} \PH{kpólóŋkpólóŋ} \PS{adj.~} \GE{bare and clean (shaven head)} 
\LX{kpolore} \HM{} \PH{kpʊ́lʊ́rɪ́} \PS{n.~} \GE{small, gourdlike flute} \OP{kpoloɛ} \TP{kpolorɛɛ} 
\LX{kpommu} \HM{} \PH{kpòmmú} \PS{n.~} \GE{cassava; plant} \OP{kpoŋŋo} \TP{kponnee} 
\LX{kpoŋi} \HM{} \PH{kpôŋ} \PS{adj.~} \GE{big, large, great} \OP{kponni} \TP{kponnee} 
\LX{kpoŋi} \HM{} \PH{kpôŋ} \PS{v.~} \GE{to knock on the head slightly} \PD{kpoŋee, kponno, kponno, kponimo, kponnaa, - } 
\LX{kpɔŋkɔrɔŋ} \HM{} \PH{kpɔ́ŋkɔ́rɔ́ŋ} \PS{adv.~} \GE{small and dented} 
\LX{kpɔŋkɔrɔŋ} \HM{} \PH{kpɔ̀ŋkɔ̀rɔ̀ŋ} \PS{adv.~} \GE{large and dented} 
\LX{kpoŋkpogiri} \HM{} \PH{kpóŋkpógrì} \PS{n.~} \GE{wickedness; deliberate unfriendly behaviour} 
\LX{kpoŋkpoli} \HM{} \PH{kpókpòlì} \PS{n.~} \GE{} \OP{kpoŋkpolo} \TP{kpoŋkpolee} \VA{kpaŋkpoli}
\LX{kpoŋkpolimo} \HM{} \PH{kpǒŋkpólímó} \PS{n.~} \GE{coop for fowl} \OP{kpoŋkpolli} \TP{kpoŋkpolee} 
\LX{kpoŋkpolo} \HM{} \PH{kpóŋkpóló} \PS{n.~} \GE{second-rate xylophone notes (usually laid out on the ground)} \SG{kpoŋkpoluu} \TP{kpoŋkpolee} 
\LX{kpoŋkponti} \HM{} \PH{kpòŋkpòntí} \PS{n.~} \SN \GE{dried cassava tubers} \SN \GE{food made from cassava flour} \OP{kpoŋkpontiri} \TP{kpoŋkpontiree} 
\LX{kponni} \HM{} \PH{kpónnì} \PS{v.~} \SN \GE{to hit a rigid object to clean of dirt (dust)} \SN \GE{to empty powdery stuff completely from a container} \PD{kponnee, kponno, kponno, kponimo, kponnaa, - } 
\LX{kpoŋŋo} \HM{} \PH{kpǒŋŋó} \PS{n.~} \SN \GE{cassava plant} \SN \GE{dried cassava tubers} \OP{kponnee} 
\LX{kpontoli} \HM{} \PH{kpóntólí} \PS{n.~} \GE{} \VA{kpantoli}
\LX{kpoo} \HM{} \PH{kpóó} \PS{adj.~} \GE{expression of surprise and dismay} \VA{kpaa}
\LX{kpore} \HM{} \PH{kpʊ̀rɪ̀} \PS{v.~} \SN \GE{to crack one's knuckles} \SN \GE{to seal completely (hole)} \PD{kporɛɛ, kpora, kporɔ, kporeba, kporaa, - } 
\LX{kpore} \HM{1} \PH{kpʊ́rɪ́} \PS{n.~} \GE{snuffbox (small container for snuff)} \OP{ kporɛɛ} 
\LX{kpore} \HM{2} \PH{kpʊ́rɪ́} \PS{n.~} \GE{lump (on wood)} \OP{kporɛɛ} \TP{kporrɛɛ} 
\LX{kpɔre} \HM{} \PH{kpɔ́rɪ̀} \PS{adj.~} \GE{weak, broken (pot, calabash)} \OP{kpɔrɔ} \TP{kpɔrɛɛ} 
\LX{kpori} \HM{} \PH{kpórì} \PS{v.~} \GE{to empty} \XV{O kpori la o woɔ libie zaa da ne yiri.} \XE{He emptied his purse to buy the house.} \PD{kporee, kporo, kporo, kporibo, kporaa, - } 
\LX{kpoteɛ} \HM{} \PH{kpʊ́tɪ́ɛ́} \PS{n.~} \GE{lumps, protrusions} 
\LX{kpugiloŋ} \HM{} \PH{kpúglóŋ} \PS{adj.~} \GE{small, solid, and round} 
\LX{kpugiloŋ} \HM{} \PH{kpùglòŋ} \PS{adj.~} \GE{large, solid, and round} 
\LX{kpulee} \HM{} \PH{kpúléé} \PS{adv.~} \GE{small and rounded on top} 
\LX{kpulee} \HM{} \PH{kpùlèè} \PS{adv.~} \GE{large and rounded on top} 
\LX{kpuleekpulee} \HM{} \PH{kpúléékpúléé} \PS{n.~} \GE{groundnut cake} \VA{kpulikpuli}
\LX{kpuli} \HM{} \PH{kpúlí} \PS{adj.~} \SN \GE{round-headed} \SN \GE{hornless (ram)} \OP{kpullo} \TP{kpulee} 
\LX{kpulikpuli} \HM{} \PH{kpùlìkpúlí} \PS{v.~} \GE{to walk with difficulty; to stutter (leper)} \PD{kpulikpulee, kpulikpulo, kpulikpulo, kpulikpulbo, kpulikpulaa, - } 
\LX{kpulli} \HM{} \PH{kpùllì} \PS{v.~} \GE{to conclude; to bring to the end} \PD{kpullee, kpullo, kpullo, kpulbo, kpullaa, - } 
\LX{kpuŋkpuli} \HM{} \PH{kpúŋkpúlí} \PS{n.~} \GE{round headed truncheon} \OP{kpuŋkpulo} \TP{kpuŋkpulee} 
\LX{kpurimunu} \HM{} \PH{kpùrìmùnù} \PS{adv.~} \SN \GE{in great numbers (fruit)} \SN \GE{of the fall of fruits} \VA{kperemene}
\LX{kubogi} \HM{} \PH{kúbògí} \PS{n.~} \GE{stone cave} \OP{kubogiri} \TP{kubogiree} 
\LX{kugbe} \HM{} \PH{kùgbé} \PS{n.~} \GE{coconut} \OP{kugberi} \TP{kugberee} 
\LX{kugbeɛ} \HM{} \PH{kùgbɪ̀ɛ́} \PS{n.~} \GE{worn-out hoe blade} \OP{kugbeere} \TP{kugbeerɛɛ} 
\LX{kukpoŋi} \HM{} \PH{kùkpôŋ} \PS{n.~} \GE{large hoe} \OP{kukponni} \TP{kukpomo/kukponnee} 
\LX{kukurbompaala} \HM{} \PH{kùkúrbʊ́mpáálà} \PS{n.~} \GE{firstfruits} \SG{kukurbompaalɛɛ} \VA{bompaala}
\LX{kukuri} \HM{} \PH{kùkúrì} \PS{n.~} \GE{hoe handle} \OP{kukue} \TP{kukuree} 
\LX{kukuribone} \HM{} \PH{kùkúrìbʊ̀n} \PS{n.~} \GE{farm produce} \OP{kukuriboma} 
\LX{kukurkpaare} \HM{} \PH{kúkúkpáárɪ́} \PS{n.~} \GE{nape of the neck of a hoe handle} \OP{kukukpaɛ} \TP{kukukpaarɛɛ} 
\LX{kuli} \HM{} \PH{kúlí} \PS{v.~} \GE{to go home} \PD{kulee, kulo, kulo, kulbo, kulaa, - } 
\LX{kulikuli} \HM{} \PH{kúlkúl} \PS{adv.~} \GE{very bad-smelling} 
\LX{kulinni} \HM{} \PH{kúlínní} \PS{n.~} \GE{shrill noise of praise} \TP{kulinnee} 
\LX{kull} \HM{} \PH{kùll} \PS{adv.~} \GE{rising in a cloud} \VA{koll}
\LX{kulli} \HM{} \PH{kùllì} \PS{v.~} \SN \GE{to mind} \SN \GE{to not regard, to ignore (typically used with ba)} \XV{A baa da vuolo yaga lɛ kyɛ a nannyige deɛ da ba kullo a.} \XE{The dog was barking intensely, but the thief just ignored it.} \PD{kullee, kullo, kullo, kulibo, kullaa, - } \VA{kpulli/kpullo}
\LX{kultaa} \HM{} \PH{kúltáá} \PS{n.~} \GE{marriage} \OP{kultaarɛɛ} 
\LX{kultaa wɛltaa} \HM{} \PH{kúltáá wɛ́ltáá} \PS{n.~} \GE{divorce} 
\LX{kuŋkuŋ} \HM{} \PH{kúŋkúŋ} \PS{adv.~} \GE{dark and threatening e.g. black storm clouds} 
\LX{kuŋkunee} \HM{} \PH{kúŋkúnéè} \PS{n.~} \GE{hillock, raised ground} \OP{kuŋkunni} \TP{kuŋkunnee} 
\LX{kuŋkuŋlere} \HM{} \PH{kúŋkúnléré} \PS{n.~} \GE{one who has epilepsy} \OP{kuŋkunlerbe} \TP{kuŋkunlerbee} 
\LX{kuŋkunni} \HM{} \PH{kùŋkúnnì} \PS{n.~} \GE{epilepsy} \TP{kuŋkunnee} 
\LX{kunoɔre} \HM{} \PH{kùnʊ́ɔ̀rɪ̀} \PS{n.~} \GE{edge of hoe blade} \OP{kunɔɛ} \TP{kunɔrɛɛ} 
\LX{kuntuŋi} \HM{} \PH{kúntúŋ} \PS{n.~} \GE{blanket} \OP{kuntunni} \TP{kuntunnee} 
\LX{kuntuŋi} \HM{} \PH{kùntùŋ} \PS{n.~} \GE{wolf} \OP{kuntunni} \TP{kuntunnee} 
\LX{kuntuŋi} \HM{} \PH{kùntùŋ} \PS{n.~} \GE{blanket} \OP{kuntunni} \TP{kuntunnee} 
\LX{kunyoɔre} \HM{} \PH{kùɲʊ́ɔ́rɪ́} \PS{n.~} \GE{hoe, plough tip} \OP{kunyɔɛ} \TP{kùnyɔrɛɛ} 
\LX{kuo} \HM{} \PH{kúó} \PS{n.~} \GE{a type of a wild mouse} \OP{kuuri} \TP{kuuree} 
\LX{kuoli} \HM{} \PH{kùòlì} \PS{v.~} \GE{to get drunk} \PD{kuolee, kuolo, kuolo, kuolbo, kuolaa, - } 
\LX{kuoli} \HM{} \PH{kùòlì} \PS{v.~} \GE{to apply fertilizer to crops} \PD{kuolee, kuolo, kuolo, kuolbo, kuolaa, - } 
\LX{kuoluŋ} \HM{} \PH{kùólǔŋ} \PS{n.~} \GE{manure} \TP{kuoklunnee} 
\LX{kuomo} \HM{} \PH{kúómó} \PS{n.~} \GE{a type of water plant} \OP{kuomee} 
\LX{kuoŋaa} \HM{} \PH{kúóŋáá} \PS{adj.~} \GE{dry} \OP{kuoni} \TP{kuonee} 
\LX{kuori} \HM{} \PH{kùórí} \PS{n.~} \GE{funeral} \OP{koe} \TP{koree} 
\LX{kuori} \HM{} \PH{kúórì} \PS{v.~} \GE{to tap with an object} \PD{kuoree, kuoro, kuoro, kuoribo, kuoraa, - } 
\LX{kuree} \HM{} \PH{kùrèè} \PS{n.~} \GE{shorts} \OP{kurri} \TP{kurree} 
\LX{kuri} \HM{} \PH{kúrí} \PS{n.~} \SN \GE{metal} \SN \GE{iron} \SG{kuruu} \OP{kurri} \TP{kuree} 
\LX{kuri} \HM{} \PH{kùrì} \PS{v.~} \GE{to forge, to shape metal} \PD{kuree, kuro, kuro, kuribo, kuraa, kuribo } 
\LX{kuri} \HM{} \PH{kùrì} \PS{v.~} \GE{to plant calabash seeds} \PD{kuree, kuro, kuro, kuribo, kuraa, kuribo } 
\LX{kuri} \HM{} \PH{kúrí} \PS{v.~} \GE{to scrape} \PD{kuree, kuro, kuro, kuribo, kuraa, kuribo } 
\LX{kuri} \HM{} \PH{kùrì} \PS{adv.~} \GE{onomatopoeia (thud of a heavy solid object falling)} 
\LX{kuri zu mare} \HM{} \PH{} \PS{v.~} \GE{to persevere} \PD{-, -, -, -, -, - } 
\LX{kuribini} \HM{} \PH{kùrìbíní} \PS{n.~} \GE{slag (ore after iron is extracted)} \SG{kuribinuu} \TP{kuribinee/kurbinnee} 
\LX{kuridaa} \HM{} \PH{kùrìdàá} \PS{n.~} \GE{iron bar} \OP{kuridaare} \TP{kuriaarɛɛ} 
\LX{kurigbɛɛgbori} \HM{} \PH{kúrígbɛ́ɛ́gbòrí} \PS{n.~} \GE{train} \OP{kurigbɛɛgboe} \TP{kurigbɛɛgboree} 
\LX{kurigbɛɛsori} \HM{} \PH{kúrígbɛ́ɛ́sórí} \PS{n.~} \GE{rail line} \OP{kurigbɛɛsoe} \TP{kurigbɛɛsoree} 
\LX{kurigbɛwogri} \HM{} \PH{kúrígbɛ́wógrì} \PS{n.~} \GE{trousers} \TP{kurigbɛwogree} 
\LX{kurikuni} \HM{} \PH{kùrìkúní} \PS{n.~} \SN \GE{tortoise} \SN \GE{ball of seeds in the calabash fruit} \SN \GE{padlock} \OP{kurikumo} \TP{kurikunee} \VA{kuri/kuŋkuni}
\LX{kurikuri} \HM{} \PH{kùrìkúrí} \PS{v.~} \GE{to search} \PD{kurikuree, kurikuro, kurikuro, kurikuribo, kurikuraa, - } 
\LX{kurikuro} \HM{} \PH{kùrìkúró} \PS{n.~} \GE{blacksmith} \OP{kurikuribo} \TP{kurikuribee} \VA{kurkuro, kurkuribo}
\LX{kurimoŋ} \HM{} \PH{kúrímóŋ} \PS{adv.~} \GE{stunted e.g. malnourished child} 
\LX{kuriwiri} \HM{} \PH{kúríwìrí} \PS{n.~} \GE{bicycle} \OP{kuriwie} \TP{kurwiree} 
\LX{kuriyoluŋ} \HM{} \PH{kúríyólúŋ} \PS{n.~} \GE{the act of wearing shorts or trousers without underwear} \OP{kuriyolli} \TP{kuriyolee} 
\LX{kurizeɛ} \HM{} \PH{kùrìzɪ̀ɛ́} \PS{n.~} \SN \GE{brass} \SN \GE{copper} \OP{kurizeere} \TP{kurizeerɛɛ} 
\LX{kurkumpɛge} \HM{} \PH{kùrkûmpɛ́gɪ̀} \PS{n.~} \GE{shell (of turtle)} \OP{kurkumpɛgre} \TP{kurkumpɛgrɛɛ} 
\LX{kurusu} \HM{} \PH{kúrúsú} \PS{adv.~} \GE{small and stunted} 
\LX{kusaalaa} \HM{} \PH{kúsààlàà} \PS{n.~} \GE{smooth stone used by a potter to smooth pots} \OP{kusaale} \TP{kusaalɛɛ} 
\LX{kusaalaa} \HM{} \PH{kúsàálàá} \PS{n.~} \GE{sharpening stone} \OP{kusaale} \TP{kusaalɛɛ} 
\LX{kuseɛbaa} \HM{} \PH{kúsɪ̀ɛ̀bàá} \PS{n.~} \GE{very large hoe} \OP{kuseɛbaare} \TP{kuseɛbaarɛɛ} 
\LX{kusibiri} \HM{} \PH{kúsíbírì} \PS{n.~} \GE{a very small stone} \OP{kusibie} \TP{kusibiree/kusibiiree} 
\LX{Kusieli} \HM{} \PH{kúsíélí} \PS{n.~} \GE{homeland of the Kusielee clan} 
\LX{Kusieluu} \HM{} \PH{kúsíélúů} \PS{n.~} \GE{a member of the Kusielee clan} \OP{Kusielee} 
\LX{kusiri} \HM{} \PH{kúsírì} \PS{n.~} \GE{stone} \OP{kusie} \TP{kusiruu} 
\LX{kusiwara} \HM{} \PH{kúsíwàrà} \PS{n.~} \GE{gravel} \TP{kusiwarɛɛ} 
\LX{kutaayi} \HM{} \PH{kútááyí} \PS{n.~} \GE{file} \OP{kutaayiri} \TP{kutaayiree} 
\LX{kutobo} \HM{} \PH{kùtòbó} \PS{n.~} \GE{hoe (traditional Dagaaba hoe)} \TP{kutooree} 
\LX{kutoori} \HM{} \PH{kùtóórí} \PS{n.~} \GE{plough breast} \OP{kutooree} 
\LX{kuunileɛiruu} \HM{} \PH{kúúnílɪ́ɛ́-írúú} \PS{n.~} \GE{resurrection} 
\LX{kuuri} \HM{} \PH{kùùrí} \PS{n.~} \GE{hoe} \OP{kue} \TP{kuree} 
\LX{kuuri} \HM{} \PH{kùùrì} \PS{v.~} \GE{to dredge; to remove from the bottom of a container of liquid} \PD{kuuree, kuuro, kuuro, kuuribo, kuuraa, - } 
\LX{kuuri} \HM{} \PH{kúúrì} \PS{v.~} \GE{to quickly dodge an object in the air} \PD{kuuree, kuuro, kuuro, kuuribo, kuuraa, - } 
\LX{kuuri} \HM{} \PH{kúúrì} \PS{n.~} \SN \GE{stone} \SN \GE{fetish stone} \OP{kubo} \TP{kuuree} \VA{kusiri, kusibe, kusirree, kusiruu}
\LX{ku᷉u} \HM{1} \PH{kú᷉ú} \PS{n.~} \SN \GE{death} \SN \GE{corpse} \OP{kuuni} \TP{kuunee} 
\LX{ku᷉u} \HM{2} \PH{kú᷉ù} \PS{n.~} \GE{} \OP{kuuni} \TP{kuunee} \VA{shortened form of biku᷉u}
\LX{ku᷉ubogi} \HM{} \PH{kú᷉úbògí} \PS{n.~} \GE{grave} \OP{ku᷉ubogri} \TP{ku᷉ubogree} 
\LX{ku᷉uyee} \HM{} \PH{kú᷉ú᷉yéè} \PS{n.~} \GE{a type of small owl believed to foresee evil happenings} \OP{ku᷉uyeeri} \TP{ku᷉uyeeree} 
\LX{kya} \HM{} \PH{ʧá} \PS{v.~} \SN \GE{to make a gift} \SN \GE{to be divided, to be separated} \SN \GE{to split; to gnaw roughly} \PD{kyaɛ, kyaara, kyaara, kyareba, kyaaraa, - } 
\LX{kyaakyee} \HM{} \PH{ʧááʧèè} \PS{interjection.~} \GE{expression announcing arrival} 
\LX{kyaaleŋ} \HM{} \PH{ʧáálɪ́ŋ} \PS{adv.~} \GE{no use going back; just do it to put an end to} 
\LX{Kyaana} \HM{} \PH{ʧáánà} \PS{n.~} \GE{China} \TP{Kyaannɛɛ} 
\LX{kyaanaa} \HM{} \PH{ʧáánáá} \PS{n.~} \GE{a mirror, that which can be seen through} \OP{ kyaane} \TP{kyaanɛɛ} 
\LX{kyaane} \HM{} \PH{ʧàànɪ̀} \PS{v.~} \GE{to give light, to make clear} \XV{A lɔɔre vu᷉u᷉ kyaane la a sori zaa.} \XE{The car's lights lit up the whole road.} \PD{kyaanɛɛ, kyaana, kyaana, kyaanema, kyaanaa, - } 
\LX{kyaaraa} \HM{} \PH{ʧááráá} \PS{n.~} \GE{comb} \OP{kyaare} \TP{kyaarɛɛ} 
\LX{kyaaraa} \HM{} \PH{ʧààráà} \PS{n.~} \GE{hall, large room} \OP{kyaare} \TP{kyaarɛɛ} 
\LX{kyaare} \HM{} \PH{ʧààrɪ̀} \PS{v.~} \SN \GE{to face} \XV{Ba zeŋ kyaare la a dendɔre.} \XE{They sat facing the door.} \SN \GE{to have diarrhoea} \XV{Ka fooŋ di sɛremaane yaga fo na baŋ kyaare la.} \XE{If you eat too much pepper, you may get diarrhoea.} \SN \GE{to shade} \XV{A ŋmenaa tuo de kataweɛ kyaare a bie.} \XE{The sun is very hot; please use an umbrella over the child.} \SN \GE{to comb} \XV{A seŋ ka te maŋ kyaare bibiiri zuri baaloŋ.} \XE{We should learn to comb children's hair gently.} \SN \GE{to give in place of; to pay} \XV{O de la bɛŋɛ kookoolaa wa kyaare ne zeɛ deni.} \XE{She gave a bowl of beans in place of a bowl of millet.} \PD{-, kyaara, kyaara, kyaareba, -, - } 
\LX{kyaare} \HM{} \PH{ʧáárɪ́} \PS{v.~} \SN \GE{to exclaim, to shout loudly (in joy or victory)} \SN \GE{extremely boiled (oil)} \PD{kyaarɛɛ, kyaara, kyaara, kyaareba, kyaaraa, - } 
\LX{kyaare} \HM{} \PH{ʧààrɪ̀} \PS{adj.~} \GE{really} \XV{O kyaare boɔrɔ la a pɔge.} \XE{He really loves the woman.} 
\LX{kyabaa} \HM{} \PH{ʧábáá} \PS{n.~} \GE{pincers, something that clamps} \OP{kyaebre} \TP{kyaberɛɛ} 
\LX{kyabo} \HM{} \PH{ʧábʊ́} \PS{adv.~} \GE{} \VA{kyape᷉}
\LX{kyafo᷉o᷉} \HM{} \PH{ʧàfʊ̀᷉ʊ̀᷉} \PS{adj.~} \GE{like foam; ruffled (husks in water)} 
\LX{kyagaaloŋ} \HM{} \PH{ʧágààlʊ́ŋ} \PS{n.~} \GE{carelessness} \XV{O kyagaaloŋ la so ka a zeɛre kɔɔ.} \XE{It is because of your carelessness that the soup got burnt.} \VA{kyagaleŋ, kyagaloŋ}
\LX{kyagalaa} \HM{} \PH{ʧágálåà} \PS{n.~} \GE{careless person} 
\LX{kyagaleŋ} \HM{} \PH{ʧágálɪ́ŋ} \PS{n.~} \GE{} \VA{kyagaaloŋ}
\LX{kyagaloŋ} \HM{} \PH{ʧágàlʊ́ŋ} \PS{n.~} \GE{} \VA{kyagaaloŋ}
\LX{kyage} \HM{} \PH{ʧágɪ́} \PS{v.~} \GE{to check} \XV{Kyage a libie noɔre ka te nyɛ.} \XE{Check the amount of money; let's see.} \PD{kyagɛɛ, kyagera, kyagera, kyagereba, kyageraa, - } 
\LX{kyage} \HM{} \PH{ʧàgɪ́} \PS{adj.~} \GE{not straight, crooked, twisted} \XV{A o lɔɔre gbɛŋmɛ werekao na poɔ la ka o seɛ deɛ kyage.} \XE{His waist got twisted as result of the injury he sustained from the lorry accident.} \OP{kyagere} \TP{kyagerɛɛ} 
\LX{kyage} \HM{} \PH{ʧàgɪ̀} \PS{v.~} \SN \GE{supported in between; (Y-shaped wood)} \SN \GE{grab firmly with fingers (around the neck)} \PD{-, -, -, -, -, - } 
\LX{kyagele} \HM{} \PH{ʧàgɪ̀lɪ̀} \PS{v.~} \SN \GE{to hold gently} \XV{O kyagele la a bie nubaala poɔ gaa ne asibiti.} \XE{He held the child's wounded hand gently as he took him to the hospital.} \SN \GE{to be well} \XV{O yeli ka a yiri kyagelɛɛ la.} \XE{She said the house was peaceful.} \SN \GE{to settle temporarily} \XV{Vɛŋ ka N kyagele kyɛ ka fo gaa kyɛ wa.} \XE{Let me rest here while you go and come back.} \PD{kyagelɛɛ, kyagela, kyagela, kyageleba, kyagelaa, - } 
\LX{kyagere} \HM{} \PH{ʧágɪ́rɪ̀} \PS{v.~} \GE{to hobble, to limp} \PD{kyagerɛɛ, kyagera, kyagera, kyagereba, kyageraa, kyagereba } 
\LX{kyakoli} \HM{} \PH{ʧákòlí} \PS{n.~} \GE{spoon} \OP{kyakolo/kyakolli} \TP{kyakolee/kyakollee} \VA{kyɛkoli}
\LX{kyakol-kyaaraa} \HM{} \PH{ʧákòl-kyááráá} \PS{n.~} \GE{fork} \OP{kyakol-kyaare} \TP{kyakol-kyaarɛɛ} 
\LX{kyakya} \HM{} \PH{ʧàʧá} \PS{v.~} \GE{to talk incessantly, to chirp} \PD{kyakyaɛ, kyakyaara, kyakyaara, kyakyareba, kyakyaaraa, - } 
\LX{kyåkyá} \HM{} \PH{ʧáʧá} \PS{n.~} \GE{betting, gambling, competition} \XV{Loto e la e la kyakya naŋ bɛllɛ faara noba libie.} \XE{The Lotto is a form of gambling that extracts money from people.} \OP{kyakyare} \TP{kyakyarɛɛ} 
\LX{kyakya᷉a} \HM{} \PH{ʧàʧá᷉á} \PS{adv.~} \GE{perforated, full of holes such that you can see through} \XV{A pelee waa la kyakya᷉a lɛ o koŋ to᷉ɔ eŋ a kyi.} \XE{The basket has sizeable spaces in between its straws and so cannot hold the millet.} 
\LX{kyalakyala} \HM{} \PH{ʧàlàʧàlà} \PS{adv.~} \SN \GE{not thick; having some gaps (vegetation)} \XV{A tuu poɔ waa la kyalakyala, a kpa᷉a᷉o᷉ koŋ baŋ sɔgle be.} \XE{The thicket is not thick, so the guinea fowl cannot hide there.} \SN \GE{dark but having some amount of visibility (dawn)} \XV{Zie waa la kyalakyala o na nyɛrɛ la a sori.} \XE{It is dark but light enough for him to find his way.} 
\LX{kyamakyama} \HM{} \PH{ʧámáʧámá} \PS{adv.~} \GE{brittle} \XV{Nɔgyɛlepɛgere man waa la kyamakyama lɛ.} \XE{Eggshells are normally brittle.} 
\LX{kyankyampelaa} \HM{} \PH{ʧánʧámpɪ̀làá} \PS{n.~} \GE{cattle egret} \OP{kyankyampeɛle} \TP{kyankyampeɛlɛɛ} 
\LX{kyankyamponaa} \HM{} \PH{ʧánʧámpʊ́nàá} \PS{n.~} \GE{a type of dark blue insect that emits light at night} \OP{kyankyamponne} \TP{kyankyamponnɛɛ} 
\LX{kyaperɛpe᷉} \HM{} \PH{ʧàpɪ́rɛ́pɪ́᷉} \PS{adv.~} \GE{gripping firmly and suddenly} \VA{kyape᷉}
\LX{Kyapore} \HM{} \PH{ʧápʊ́rɪ̀} \PS{n.~} \GE{a village in the Upper West Region of Ghana} \OP{kyaporre} \TP{kyaporrɛɛ} 
\LX{kyaporo} \HM{} \PH{ʧàpórò} \PS{n.~} \GE{brain mass} \OP{kyaporri} \TP{kyaporree} \VA{taporo}
\LX{kyara} \HM{} \PH{ʧárá} \PS{adj.~} \GE{fork shape (intersection)} \OP{kyarre} \TP{kyarɛɛ/kyarrɛɛ} 
\LX{kyara} \HM{} \PH{ʧárà} \PS{n.~} \GE{woman (nickname)} \OP{kyarre} \TP{kyarrɛɛ} 
\LX{kyaraakyaraa} \HM{} \PH{ʧárááʧáráá} \PS{adj.~} \GE{forked looking} 
\LX{kyarakyara} \HM{} \PH{ʧàràʧàra} \PS{n.~} \GE{liquid mixed roughage} 
\LX{kyarewara} \HM{} \PH{ʧárɪ́wárá} \PS{adv.~} \GE{few (items), little (money)} 
\LX{kya᷉a} \HM{} \PH{ʧà᷉à} \PS{adj.~} \SN \GE{lit  up} \XV{Zie waa la kya᷉a dabe᷉ɛ kyebe.} \XE{It is day; there is no need to fear.} \SN \GE{full of tiny holes so that one can see through (light cloth)} 
\LX{kya᷈e} \HM{} \PH{ʧà᷉ɪ̀} \PS{adj.~} \GE{new, shining} 
\LX{kye} \HM{} \PH{ʧɪ̀} \PS{v.~} \SN \GE{to sprout} \SN \GE{to make a small cut on (tattooing)} \SN \GE{to spilt open a gourd to make calabashes} \SN \GE{to mark out the mouth of a hole to be dug (grave)} \PD{kyeɛɛ, kyeerɛ, kyeerɛ, kyeerebɛ, kyeeraa, kyeerebɛ } 
\LX{kye} \HM{} \PH{ʧɪ́} \PS{v.~} \GE{to be smart; to awaken} \PD{kyeɛɛ, kyeerɛ, kyeerɛ, kyeerebɛ, kyeeraa, kyeerebɛ } 
\LX{kye} \HM{} \PH{ʧé} \PS{particle.~} \GE{not} \XV{Zenɛ o kye be.} \XE{Today he is not there.} 
\LX{kyɛ} \HM{} \PH{ʧɛ̀} \PS{v.~} \SN \GE{to fell (tree)} \SN \GE{to strike (matches)} \SN \GE{to peck (of fowl)} \SN \GE{to make small cuts on the skin (herbal treatment)} \SN \GE{to start or lead a song} \PD{kyɛɛ, kyeɛrɛ, kyeɛrɛ, kyeɛrebɛ, kyeɛraa, kyeɛrebɛ } 
\LX{kyɛ} \HM{} \PH{ʧɛ́} \PS{conj.~} \SN \GE{but} \XV{Wagre gɛrɛ la kyɛ o naŋ ba wa.} \XE{Time is running out, but s/he has not yet come.} \SN \GE{and} \XV{O zoro la bangyeraa kyɛ tiire.} \XE{S/he has diarrhoea and is vomiting.} 
\LX{kyɛ} \HM{} \PH{ʧɛ̂} \PS{adj.~} \GE{here} \XV{Sazɔŋ ba miire Ghana kyɛ.} \XE{Snow does not fall here in Ghana.} 
\LX{kyea} \HM{} \PH{ʧɪ́á} \PS{interjection.~} \GE{no (expressing disagreement)} 
\LX{kyɛbaa} \HM{} \PH{ʧɛ́báá} \PS{n.~} \GE{trap, pincers} \OP{kyɛrre} \TP{kyɛrɛɛ} \VA{kyabaa}
\LX{kyebɛ} \HM{} \PH{ʧɪ́bɛ́} \PS{adj.~} \GE{fast and cunning} \XV{A bie taa la kyebɛ te gaŋ ona ka lɔɔre ŋmɛ.} \XE{The child is just too restless, he has been hit by a car.} \VA{kyebɛkyebɛ}
\LX{kyɛbɛɛ} \HM{} \PH{ʧɛ́bɛ́ɛ́} \PS{n.~} \GE{fool} 
\LX{kyɛbɛɛ} \HM{} \PH{ʧɛ̀bɛ̀ɛ̀} \PS{adv.~} \GE{loose and hanging} \XV{A wono la a baala nyɛ o naŋ e o noɔre kyɛbɛɛ lɛ.} \XE{The patient is suffering; see how his mouth is hanging open.} 
\LX{kyeɛ} \HM{} \PH{ʧɪ́ɛ̀} \PS{n.~} \GE{ground squirrel} \OP{kyeere} \TP{kyeerɛɛ} 
\LX{kyeɛ} \HM{} \PH{ʧɪ̀ɛ́} \PS{adj.~} \SN \GE{faded} \XV{A kpaaroo ŋa da waa la zeɛ kyɛ o yi la kyeɛ a tɔgele bompelaa.} \XE{This shirt was red but it has faded and now looks white.} \SN \GE{outmoded} \XV{Fo yieli la yieluŋ ŋa lɛɛ ka o te yi kyeɛ.} \XE{You have sung this song so much that it has lost its sweetness.} 
\LX{kyeele} \HM{} \PH{ʧɪ́ɪ́lɪ́} \PS{v.~} \GE{to incline} \PD{kyeelɛɛ, kyeelɛ,, kyeelɛ,, kyeelbɛ, kyeelaa, - } 
\LX{kyeele} \HM{} \PH{ʧɪ̀ɪ̀lɪ̀} \PS{v.~} \GE{to scratch with a pointed object} \PD{kyeelɛɛ, kyeelɛ,, kyeelɛ,, kyeelbɛ, kyeelaa, - } 
\LX{kyeɛle} \HM{} \PH{ʧɪ̀ɛ́lɪ̀} \PS{v.~} \GE{to spread across in some order (lines)} \PD{kyeɛlɛɛ, kyeɛlɛ, kyeɛlɛ, kyeɛlbɛ, kyeɛlaa, - } 
\LX{kyeɛlɛ} \HM{} \PH{ʧɪ́ɛ́lɛ́} \PS{n.~} \GE{faulty arrangement in the game of marbles; usually a wrong move where four or more marbles of same ??? are in a continuous row} \TP{kyeɛlɛɛ} 
\LX{kyeene} \HM{} \PH{ʧɪ̀ɪ̀nɪ̀} \PS{v.~} \GE{to wipe (sweat from forehead with finger); to lick out a food container (usually with finger)} \PD{kyeenɛɛ, kyeenɛ, kyeenɛ, kyeenemɛ, kyeenaa, - } 
\LX{kyeene} \HM{} \PH{ʧɪ́ɪ́nɪ́} \PS{n.~} \GE{metal ankle gingle for dancing} \OP{kyeemɛ} \TP{kyeemɛɛ/kyeenɛɛ} 
\LX{kyeenɛ} \HM{} \PH{ʧɪ́ɪ́nɛ́} \PS{v.~} \GE{one who shepherds} \PD{kye᷉e᷉ɛ᷉, kyeenɛ, kyeenɛ, kyeenemɛ, kyeenaa, - } 
\LX{kyeɛraa} \HM{} \PH{ʧɪ́ɛ́ráá} \PS{n.~} \GE{lighter, something used to create a spark to produce fire} \OP{kyeɛre} \TP{kyeɛrɛɛ} 
\LX{kyeɛraa} \HM{} \PH{ʧɪ̀ɛ̀ràà} \PS{n.~} \GE{contribution} \XV{N teɛre ka ba zaaŋ yɔɔ ba kyeɛraa.} \XE{I think everyone has paid his contribution.} \OP{kyeɛre} 
\LX{kyeere} \HM{} \PH{ʧɪ́ɪ́rɪ́} \PS{v.~} \GE{to cry loudly (from pain)} \PD{kyeerɛɛ, kyeerɛ, kyeerɛ, kyeerebe, kyeeraa, kyeerebɛ } \VA{kyeere}
\LX{kyeɛre} \HM{} \PH{ʧɪɛrɪ} \PS{v.~} \SN \GE{to contribute} \SN \GE{to harvest yams} \PD{-, kyeɛrɛ, -, -, -, - } 
\LX{kyɛgaleŋ} \HM{} \PH{ʧɛ́gálɪ̀ŋ} \PS{} \GE{careless} \VA{kyagaloŋ/kyagaleŋ}
\LX{kyegi} \HM{} \PH{ʧègì} \PS{v.~} \SN \GE{to pay, to make a ritual offering (for herbal treatment)} \SN \GE{to pay the dowry or brideprice for} \PD{kyegee, kyegiri, kyeire, kyegiribe, kyegiraa, - } 
\LX{kyegiri} \HM{} \PH{ʧégírí} \PS{n.~} \GE{payment; ritual payment (pride price)} 
\LX{kyɛkuuri} \HM{} \PH{ʧɛ́kúúrí} \PS{v.~} \GE{to duck} \PD{kyɛkuuree, kyeɛrɛkuuro, kyeɛrekuuro, kyeɛrekuuribo, kyeɛrekuuraa, - } 
\LX{kyekye} \HM{} \PH{ʧèʧé} \PS{n.~} \GE{sewing machine} \OP{kyekyeri} \TP{kyekyeree} 
\LX{kyɛkyɛre} \HM{} \PH{ʧɛ̀ʧɛ̀rɪ́} \PS{adj.~} \GE{striped design (black and white)} \TP{kyɛkyɛrɛɛ} 
\LX{kyelaa} \HM{} \PH{ʧɪ̀làà} \PS{adv.~} \GE{pointed and large} 
\LX{kyelaa} \HM{} \PH{ʧɪ́láá} \PS{adj.~} \GE{pointed and small} 
\LX{kyele} \HM{} \PH{ʧɪ́lɪ́} \PS{v.~} \GE{to awaken; to come around} \XV{Pampana ŋa ka o iri go᷉ɔ᷉o᷉ poɔ vɛŋ ka niŋe kyele.} \XE{He has just woken up from sleep; allow him to come around.} \PD{kyelɛɛ, kyelɛ, kyelɛ, kyelbɛ, kyelaa, - } 
\LX{kyɛle} \HM{} \PH{ʧɛ́lɪ̀} \PS{v.~} \GE{to stretch out the hand earnestly to receive} \PD{kyɛlɛɛ, kyɛllɛ, kyɛllɛ, kyɛlbɛ, kyɛllaa, - } 
\LX{kyɛlee} \HM{} \PH{ʧɛ́lɪ́ɪ́} \PS{adj.~} \SN \GE{half of} \SN \GE{part of} \OP{kyɛlle} \VA{kyɛlɛɛ}
\LX{kyɛlɛɛ} \HM{} \PH{ʧɛ́lɛ̀ɛ́} \PS{adj.~} \GE{} \OP{kyɛlle} \VA{kyɛlee}
\LX{kyɛlɛɛkyɛlɛɛ} \HM{} \PH{ʧɛ́lɛ́ɛ́ʧɛ̀lɛ̀ɛ̀} \PS{adj.~} \GE{done half and half} 
\LX{kyɛleere} \HM{} \PH{ʧɛ̀lɪ́ɪ́rɪ́l} \PS{n.~} \GE{pointed iron bar for boring holes} \TP{kyɛleerɛɛ} 
\LX{kyɛlemɛŋ} \HM{} \PH{ʧɛ́lɪ́mɛ́ŋ} \PS{adv.~} \GE{narrow (strip of cloth)} \XV{A naŋkpaana kyeele la pɛnne kyɛlemɛŋ lɛ leŋ ne a o gbɛre naŋ ŋmaa.} \XE{The hunter tore a narrow piece of cloth and used it to tie his wounded leg.} 
\LX{kyelenkyooraa} \HM{} \PH{ʧɪ́lɪ́nʧʊ́ʊ̊ráà} \PS{n.~} \GE{mole} \OP{kyelenkyoore} \TP{kyelenkyoorɛɛ} \VA{kyenkyooraa/kyore}
\LX{kyeli} \HM{} \PH{ʧélì} \PS{v.~} \GE{to gaze at with mind completely drawn, dumbfounded} \XV{O nyɛ la a pɔge velaa na a deɛ kyeli eŋ.} \XE{He saw that beautiful woman and just fixed both his eyes and mind on her.} \PD{kyelee, kyelle, kyelle, kyelbe, kyellaa, - } 
\LX{kyell} \HM{} \PH{ʧɪ́ll} \PS{adv.~} \GE{protruding with a pointed end} 
\LX{kyelle} \HM{} \PH{ʧɪ̀llɪ̀} \PS{v.~} \GE{to rejoice, to hope} \PD{kyellɛɛ, kyellɛ, kyellɛ, kyellebɛ, kyellaa, kyellebɛ } 
\LX{kyɛlle} \HM{} \PH{ʧɛ́llɪ̀} \PS{v.~} \SN \GE{to listen} \XV{Ka a karemaŋ wa yele a yɛlɛ yɛ kyɛlle velaa.} \XE{When the teacher is talking, listen attentively.} \SN \GE{to notice, to be aware} \XV{N ba kyɛlle ka daa la zenɛ.} \XE{I'm not aware that today is the market day.} \SN \GE{I'm in a hurry; I can't wait for you} \SN \GE{to stay away; to be irretrievable} \XV{A bie gaa nyɛ deɛne a deɛ kyɛlle a be.} \XE{The child went and saw there was a play going on, so s/he never showed up again.} \PD{kyɛllɛɛ, kyɛllɛ, kyɛllɛ, kyɛlebɛ, kyɛllaa, kyɛlebɛ } 
\LX{kyɛloo} \HM{} \PH{ʧɛ̀lʊ́ʊ́} \PS{n.~} \GE{striped non-venomous snake} \OP{kyɛlle} \TP{kyɛlɛɛ} 
\LX{kyempereŋ} \HM{} \PH{ʧɪ̀mpɪ̀rɪ̀ŋ} \PS{adv.~} \GE{utter darkness} 
\LX{kyempereŋ} \HM{} \PH{ʧɪ́mpɪ́rɪ́ŋ} \PS{adv.~} \GE{small but sturdy} 
\LX{kyɛŋ} \HM{} \PH{ʧɛ̂ŋ} \PS{v.~} \GE{to walk} \PD{kyɛŋɛɛ, kyɛnɛ, kyɛnɛ, kyɛnemɛ, kyɛnaa, - } 
\LX{kyɛnaa} \HM{} \PH{ʧɛ́nàá} \PS{adv.~} \GE{this place} \VA{shortened form of kyɛ ŋa}
\LX{kyɛnɛ} \HM{} \PH{ʧɛ̀nɛ́} \PS{adv.~} \GE{still} \XV{O naŋ kyɛnɛ ba iri.} \XE{S/he is still not up.} 
\LX{kyɛnɛ} \HM{} \PH{ʧɛ̀nɛ̀} \PS{n.~} \GE{friend (used among females)} \OP{kyɛmine} \TP{kyɛnnɛɛ} 
\LX{kyenekɔnɔ} \HM{} \PH{ʧɪ̀nɪ̀kɔ̀nɔ} \PS{adj.~} \GE{tall and twisted} 
\LX{kyeŋi} \HM{} \PH{êŋ} \PS{v.~} \GE{to stand up against in argument} \XV{Karemamine mine ba nɔŋ ka karembibiiri kyenne ba.} \XE{Some teachers dislike students arguing with them.} \PD{kyennee, kyenne, kyenne, kyenime, kyennaa, - } 
\LX{kyeŋkɔleŋ} \HM{} \PH{ʧɪ́ŋkɔ̀lɪ́ŋ} \PS{n.~} \GE{a type of small bird that goes around in groups of two or three} \OP{kyeŋkɔlle} \TP{kyeŋkɔllɛɛ} 
\LX{kyenkyamm} \HM{} \PH{ʧɪ́nʧámm} \PS{adv.~} \GE{stupified, lost as to what move to make} 
\LX{kyɛnkyɛmpelaa} \HM{} \PH{ʧɛnʧɛ́mpɪ̀làá} \PS{n.~} \GE{egret} \OP{ kyɛnkyɛmpeɛle} \TP{kyɛnkyɛmpeɛlɛɛ} \VA{kyankyampelaa}
\LX{kyɛnkyɛmpelaanagbare} \HM{} \PH{ʧɛ́nʧɛ́mpɪ́láánàgbárɪ́} \PS{n.~} \GE{a type of plant with medicinal properties} \OP{kyɛnkyɛmpelaanagbarre} \TP{kyɛnkyɛmpelaanagbarɛɛ} 
\LX{kyɛnkyɛŋ} \HM{} \PH{ʧɛ̀nʧɛ̀ŋ} \PS{adj.~} \GE{often} \XV{Saa ba la miire kyɛnkyɛŋ.} \XE{It does not rain often, lately.} 
\LX{kyenkyeremɛ} \HM{} \PH{ʧɪ̌nʧɪ́rɪ́mɛ́} \PS{adj.~} \GE{get lost, disappear} \XV{Ba yɔ lɔnne la a kuuri lɛɛ ka o deɛ te kpɛ kyenkyeremɛ.} \XE{The hoe was passed from one person to another until it disappeared.} 
\LX{kyɛnkyɛrɛŋ} \HM{} \PH{ʧɛ́nʧɛ́rɛ́ŋ} \PS{n.~} \GE{confused state; crossroads} \XV{N boɔrɔ ka N gaa tomɔ kyɛ ka n bie yoŋ la die poɔ, N deɛ kpɛ la kyɛnkyɛrɛŋ poɔ.} \XE{I want to go to work but my child is alone in the house; I'm completely confused.} 
\LX{kyɛnkyɛweerɛ} \HM{} \PH{ʧɛ́nʧɛ́wɪ́ɪ́rɛ́} \PS{n.~} \GE{belt of pearls (cowries and beads)} \SG{kyɛnkyɛweeroo} \TP{kyɛnkyɛweerɛɛ} \VA{(kyɛweerɛ)}
\LX{kyenkyoɔlaa} \HM{} \PH{ʧɪ́nʧʊ̀ɔ̀láà} \PS{n.~} \GE{morning star} \OP{kyenkyoɔle} \TP{kyenkyoɔlɛɛ} 
\LX{kyɛnne} \HM{} \PH{ʧɛ́nnɪ́} \PS{n.~} \GE{inheritance} \TP{kyɛnnɛɛ} 
\LX{kyɛnne} \HM{} \PH{ʧɛ́nnɪ̀} \PS{n.~} \SN \GE{the act of walking} \SN \GE{journey} \XV{Fo kyɛnne be soŋ?} \XE{How was your journey?.} \TP{kyɛnnɛɛ} 
\LX{kyɛnne} \HM{} \PH{ʧɛ̀nnɪ̀} \PS{n.~} \GE{friendship between females} \TP{kyɛnnɛɛ} 
\LX{kyɛnse} \HM{} \PH{ʧɛ́nsɪ̀} \PS{n.~} \GE{iron sheet} \OP{kyɛnsere} \TP{kyɛnserɛɛ} \VA{kyanse}
\LX{kyɛparepeɛ} \HM{} \PH{ʧɛ́párɪ́pɪ́ɛ́} \PS{n.~} \GE{a type of brown, spotted nocturnal bird} \OP{kyɛparepeere} \TP{kyɛparepeerɛɛ} 
\LX{kyeraa} \HM{} \PH{ʧɪ̀ràà} \PS{n.~} \SN \GE{a type of tree that bears pink fruits with white edible flesh} \SN \GE{the edible part of this tree's fruit} \OP{kyerre} \TP{kyerɛɛ/kyerrɛɛ} 
\LX{kyeraa} \HM{} \PH{ʧɪ̀ráà} \PS{n.~} \GE{chisel} \OP{kyerre} \TP{kyerɛɛ/kyerrɛɛ} 
\LX{kyere} \HM{} \PH{ʧɪ́rɪ̀} \PS{v.~} \SN \GE{to sneeze} \SN \GE{to spit forcefully} \SN \GE{to roll a small stone in the game baa} \PD{kyerɛɛ, kyerɛ, kyerɛ, kyerebɛ, kyeraa, kyerebɛ } 
\LX{kyere} \HM{} \PH{ʧɪ́rɪ́} \PS{v.~} \SN \GE{to hasten} \SN \GE{of the body; shake; irrestive} \PD{kyerɛɛ, kyerɛ, kyerɛ, kyerebɛ, kyeraa, kyerebɛ } 
\LX{kyere} \HM{} \PH{ʧɪ̀rɪ̀} \PS{v.~} \GE{to lance (to cut through the skin to let out pus)} \PD{kyerɛɛ, kyerɛ, kyerɛ, kyerebɛ, kyeraa, kyerebɛ } 
\LX{kyɛre} \HM{} \PH{ʧɛ̀rɪ̀} \PS{v.~} \SN \GE{to pay a dowry or bride price} \SN \GE{to untie, to set free} \SN \GE{to come across (road)} \PD{kyɛrɛɛ, kyɛrɛ, kyɛrɛ, kyɛrebɛ, kyɛraa, kyerebɛ } 
\LX{kyɛre} \HM{} \PH{ʧɛ́rɪ̀} \PS{v.~} \SN \GE{to left over, to remain} \SN \GE{to come after (in birth order)} \XV{Maa la kyɛre N yɔɔpɔge.} \XE{My sister comes after me.} \PD{kyɛrɛɛ, kyɛrɛ, kyɛrɛ, kyɛrebɛ, kyɛraa, - } 
\LX{kyɛre} \HM{} \PH{ʧɛ̀rɪ́} \PS{n.~} \GE{small body of water on a farm} \OP{kyɛrre} \TP{kyɛrɛɛ/kyɛrrɛɛ} 
\LX{kyɛrɛ} \HM{} \PH{ʧɛ̀rɛ̀} \PS{n.~} \GE{a type of guinea corn} \TP{kyɛrɛɛ} 
\LX{kyerebaa} \HM{} \PH{ʧɪ̀rɪ̀báá} \PS{n.~} \GE{a type of brown insect usually found on the tree and fruit} \OP{kyerbaare} \TP{kyerbaarɛɛ} 
\LX{kyɛrebo} \HM{} \PH{ʧɛ́rɪ́bʊ́} \PS{n.~} \GE{bride price} 
\LX{kyerekyere} \HM{} \PH{ʧɪ˥rɪ́ʧɪ́rɪ́} \PS{adj.~} \GE{utter silence or quiet (empty house)} 
\LX{kyerelaane} \HM{} \PH{ʧɪ́rɪ́láánɪ́} \PS{n.~} \GE{ripe fruit of the archie tree} \OP{kyerelaama} \TP{kyerelaamɛɛ} 
\LX{kyeremɛ} \HM{} \PH{ʧɪ́rɪ́mɛ́} \PS{n.~} \GE{facial incision(s), tattoo(s)} \OP{kyeremɛɛ} \VA{yaakɔɔre}
\LX{kyɛremɛ} \HM{} \PH{ʧɛ̀rɪ̀mɛ́} \PS{n.~} \GE{stick for stirring soup} \TP{kyɛremɛɛ} 
\LX{kyerenaabo} \HM{} \PH{ʧɪ̀rɪ̀náábʊ́} \PS{n.~} \GE{a type of red insect} \OP{kyerenii} \TP{kyereniinee} 
\LX{kyerɔɔ} \HM{} \PH{ʧɪ̀rɔ̀ɔ̀} \PS{adv.~} \GE{disappearing instantly} 
\LX{kyɛrre} \HM{} \PH{ʧɛ̀rrɪ̀} \PS{v.~} \GE{to scoop liquid to the last bit (water)} \PD{kyɛrrɛɛ, kyɛrrɛ, kyɛrrɛ, kyɛrebɛ, kyɛrraa, - } 
\LX{kyerri} \HM{} \PH{ʧérrì} \PS{v.~} \GE{to scamper} \PD{kyerree, kyerre, kyerre, kyeribe, kyerraa, - } 
\LX{kyesel} \HM{} \PH{ʧɪ́sɪ̀l} \PS{n.~} \GE{chisel} \OP{kyeselɛ} \TP{kyeselɛɛ} 
\LX{kye᷉e} \HM{} \PH{ʧɪ̀ɪ̀} \PS{v.~} \GE{to follow around, to herd} \XV{A pɔge daare kye᷉ la serɛ a tigiri zie lɛɛ ka zie te soɔ.} \XE{The woman followed her husband all day at the feast.} \PD{kye᷉eɛ, kye᷉enɛ, kyeenɛ, kyeenemɛ, kyeenaa, kyeenemɛ } 
\LX{kye᷉e} \HM{} \PH{ʧɪ́ɪ̀} \PS{v.~} \GE{to fry} \XV{N nɔŋ la zoma ba naŋ maŋ kye᷉e.} \XE{I like fish that is fried.} \PD{kye᷉eɛ, kye᷉enɛ, kyeenɛ, kyeenemɛ, kyeenaa, kyeenemɛ } 
\LX{kye᷉ɛ} \HM{} \PH{ʧɪ̀᷉ɛ̀} \PS{v.~} \SN \GE{to tear} \XV{Ka fooŋ tage a kparoo o na kye᷉ɛ la.} \XE{If you pull the shirt, it will tear.} \SN \GE{to dash, to run quickly} \XV{A yan-yaa da zeɛ la kparoo kaŋa deɛ leɛ kyeɛ ne.} \XE{The lunatic snatched one of the shirts and dashed away with it.} \PD{kye᷉ɛɛ, kyeɛnɛ, kyeɛnɛ, kyeɛnemɛ, kyeɛnaa, kyeɛnemɛ } 
\LX{kye᷉ɛ} \HM{} \PH{ʧɪ́ɛ́} \PS{adj.~} \GE{torn (shirt)} \XV{Kpare kye᷉ɛ la ka a sɔresɔrɔ su.} \XE{The beggar wore a torn shirt.} 
\LX{kyi} \HM{} \PH{ʧî} \PS{n.~} \GE{millet} \SG{kyiruu (bag of) kaboo (for plant)} \TP{kyiree} 
\LX{kyi} \HM{} \PH{ʧí} \PS{v.~} \SN \GE{to direct toward, to pinpoint} \XV{Maa la ka o kyi kyɛ yele a yɛlɛ.} \XE{S/he directed the talk toward me.} \SN \GE{to hate} \XV{O deɛ kyi ma, o yeli ka o koŋ wuli ma a pɔge yiri zie.} \XE{She just hates me; she refuses to show me her house.} \PD{kyiee, kyire, kyire, kyiribe, kyiraa, kyiribe } 
\LX{kyibe} \HM{} \PH{ʧìbé} \PS{n.~} \GE{lice} \SG{kyibuu} \TP{kyibee} 
\LX{kyibuu} \HM{} \PH{ʧìbúú} \PS{n.~} \GE{flea} \OP{kyibe} \TP{kyibee} 
\LX{kyieree} \HM{} \PH{ʧìèréè} \PS{v.~} \GE{a large black bird with red tail feathers} \PD{-, -, -, -, -, - } 
\LX{kyieri} \HM{} \PH{ʧíèri} \PS{v.~} \SN \GE{to argue with} \SN \GE{to shout with joy} \PD{kyieree, kyiere, kyiere, kyieribe, kyieraa, kyieribe } 
\LX{kyieruŋ} \HM{} \PH{ʧíérúŋ} \PS{n.~} \GE{the ability to remember or recall information} 
\LX{kyigi} \HM{} \PH{ʧìgì} \PS{v.~} \GE{to cover entirely with a cuplike object} \PD{kyigee, kyigre, kyigre, kyigribe, kyigraa, - } 
\LX{kyiiri} \HM{} \PH{ʧíírì} \PS{v.~} \SN \GE{to forbid} \SN \GE{to deny} \PD{kyiiree, kyiire, kyiire, kyiiribe, kyiiraa, kyiiribe } 
\LX{kyiiruŋ} \HM{} \PH{ʧíírúŋ} \PS{n.~} \GE{taboo, prohibition} \OP{kyiire} \TP{kyiiree} 
\LX{kyileŋko} \HM{} \PH{ʧíléŋkó} \PS{adj.~} \GE{expert} \OP{kyileŋkori} \TP{kyileŋkoree} 
\LX{kyilli} \HM{} \PH{ʧíllì} \PS{v.~} \GE{to alter, to change} \PD{kyillee, kyille, kyille, kyilibe, kyillaa, kyilibe } 
\LX{kyinni} \HM{} \PH{ʧìnnì} \PS{v.~} \GE{to land suddenly and stongly (from a leap)} \PD{kyinnee, kyinne, kyinne, kyinime, kyinnaa, kyinnaa } 
\LX{kyiri} \HM{} \PH{ʧírì} \PS{v.~} \GE{to pour} \SN \GE{to leak} \PD{kyiree, kyire, kyire, kyiribe, kyiraa, kyirre } \SE{kyiri zee} \GE{to bleed} 
\LX{kyo} \HM{} \PH{ʧó} \PS{v.~} \GE{to aim at} \PD{kyoe, kyuoro, kyuoro, kyuoribo, kyuoraa, - } 
\LX{kyo} \HM{} \PH{ʧʊ̀} \PS{v.~} \SN \GE{to suck} \SN \GE{shooting contest (gambling with arrows)} \PD{kyoɛ, kyoora, kyoora, kyooreba, kyooraa, - } 
\LX{kyɔ} \HM{} \PH{ʧɔ́} \PS{v.~} \SN \GE{to hide behind an object} \SN \GE{to start a tune} \SN \GE{to bear dawdaw fruits} \SN \GE{to tune up to play xylophone text} \PD{kyɔɛ, kyoɔrɔ, kyoɔrɔ, kyoɔreba, kyoɔraa, - } 
\LX{kyɔ} \HM{} \PH{ʧɔ́} \PS{adj.~} \GE{at first; in the beginning} \XV{o naŋ kyɔ wa o ba soore te.} \XE{When s/he first came, s/he did not ask us.} 
\LX{kyɔ-de} \HM{} \PH{ʧɔ́-dɪ́} \PS{v.~} \GE{to take by surprise; to attack suddenly} \PD{kyɔ-deɛ, kyoɔrɔ-derɛ, kyɔ-derɛ, kyɔ-derebɛ, kyɔ-deraa, - } 
\LX{kyɔge} \HM{} \PH{ʧɔ́gɪ́} \PS{v.~} \GE{to sprain} \PD{kyɔgɛɛ, kyɔgrɔ, kyɔgrɔ, kyɔgreba, kyɔgraa, kyɔgreba } 
\LX{kyɔge} \HM{} \PH{ʧɔ̀gɪ̀} \PS{v.~} \SN \GE{to chook, to stab} \SN \GE{to pierce, to inject} \PD{kyɔgɛɛ, kyɔgrɔ, kyɔgrɔ, kyɔgreba, kyɔgraa, kyɔgreba } 
\LX{kyɔge a mɛrɛɛ poɔ} \HM{} \PH{ʧɔ̀gɪ̀ à mɛ̀rɛ́ɛ pʊ́ɔ́} \PS{v.~} \GE{straight to the point; hitting the nail on the head} \PD{-, -, -, -, -, - } 
\LX{kyɔgelɔkɔlɔ} \HM{} \PH{ʧɔ́gɪ́lɔ́kɔ́lɔ́} \PS{adj.~} \GE{slim, (appendages)} 
\LX{kyogi} \HM{} \PH{ʧógí} \PS{v.~} \SN \GE{to continue, to resume} \SN \GE{to be in line with} \PD{kyogee, kyogro, kyogro, kyogribo, kyogaa, kyogribo } 
\LX{kyogi} \HM{} \PH{ʧògì} \PS{v.~} \GE{to pay a bride price} \PD{kyogee, kyogro, kyogro, kyogribo, kyograa, - } 
\LX{kyogili} \HM{} \PH{ʧóglì} \PS{v.~} \SN \GE{to stack delicately} \SN \GE{pick ears} \PD{kyogilee, kyogilo, kyoglo, kyogilibo, kyogilaa, - } 
\LX{kyogilokolo} \HM{} \PH{ʧóglókóló} \PS{adj.~} \GE{tall and high up into the sky} \VA{kyogilokolokolo}
\LX{kyogilokolo} \HM{} \PH{ʧóglókóló} \PS{n.~} \SN \GE{tower} \SN \GE{storey building} \VA{kyogilokolokolo}
\LX{kyogiri} \HM{} \PH{ʧógrì} \PS{v.~} \GE{to trot (of a horse)} \PD{kyogiree, kyogiro, kyogiro, kyogiribo, kyogiraa, - } 
\LX{kyɔkyɔɔ} \HM{} \PH{ʧɔʧɔɔ} \PS{adv.~} \GE{appearing very long (dresσ} 
\LX{kyɔlaa} \HM{} \PH{ʧɔ̀làà} \PS{adj.~} \GE{long and pointed} 
\LX{kyɔle} \HM{} \PH{ʧɔ́lɪ́} \PS{v.~} \GE{to roam, to go around} \PD{kyɔlɛɛ, kyɔlɔ, kyɔlɔ, kyɔlbɔ, kyɔlaa, - } 
\LX{kyɔlemɔ} \HM{} \PH{ʧɔ́lɪ́mɔ́} \PS{n.~} \GE{spear; lance} \OP{kyɔlle} \TP{kyɔllɛɛ} \VA{kpane}
\LX{kyolenkyooraa} \HM{} \PH{ʧʊ́lɪ́nʧʊ́ʊ́ráá} \PS{n.~} \GE{mole; small black mouse with a strong smell} \OP{kyolenkyoore} \TP{kyolenkyoorɛɛ} \VA{kyore}
\LX{kyoli} \HM{} \PH{ʧólí} \PS{v.~} \SN \GE{to pierce through} \SN \GE{to pass by} \SN \GE{to forget} \SN \GE{to overdo; to do wrong} \PD{kyolee, kyolo, kyolo, kyolbo, kyolaa, - } 
\LX{kyɔlkyɔl} \HM{} \PH{ʧɔ̀lʧɔ̀l} \PS{adv.~} \GE{long (snake)} 
\LX{kyololo} \HM{} \PH{ʧólóló} \PS{adv.~} \GE{of light beam; gentle lighting} 
\LX{kyɔmpoore} \HM{} \PH{ʧɔ̀mpʊ̀ʊ̀rɪ̀} \PS{n.~} \GE{a challenge between two people in which one who wakes up first sprays the counterpart with water} \TP{kyɔpoorɛɛ} 
\LX{kyɔmpuuri} \HM{} \PH{ʧɔ̀mpúúrí} \PS{n.~} \GE{betting in which the earlier to wake up sprays water on his counterpart} \TP{kyɔmpuuree} 
\LX{kyɔŋ} \HM{} \PH{ʧɔ̀ŋ} \PS{adv.~} \GE{first thing done} \XV{Ko᷉ɔ nyegrɛ la see ka te kyɔŋ iri gaa a baa.} \XE{Water is scarce so we have to get up early and go to the stream.} 
\LX{kyone} \HM{} \PH{ʧʊ́nɪ́} \PS{adj.~} \GE{bulging} 
\LX{kyɔŋe} \HM{} \PH{ʧɔ́ŋ} \PS{v.~} \GE{to file along in line; to follow one another} \PD{kyɔŋɛɛ, kyɔnɔ, kyɔnɔ, kyɔnemɔ, kyɔnaa, - } \VA{tɔŋ}
\LX{kyɔŋe} \HM{} \PH{ʧɔ̂ŋ} \PS{v.~} \GE{fetch or light from a fire} \PD{kyɔŋɛɛ, kyɔnnɔ, kyɔnnɔ, kyɔnema, kyɔnnaa, - } \VA{tɔŋ}
\LX{kyonɛɛkyonɛɛ} \HM{} \PH{ʧʊ́nɛ́ɛ́ʧʊ́nɛ́ɛ́} \PS{adj.~} \GE{lighting in sparks} 
\LX{kyɔnfeɛle} \HM{} \PH{ʧɔ̀nfɪ́ɛ́lɪ́} \PS{n.~} \GE{very early (morning)} \OP{kyɔnfeɛlɛ} \TP{kyɔnfeɛlɛɛ} 
\LX{kyoŋi} \HM{} \PH{ʧôŋ} \PS{v.~} \SN \GE{to take over (piece of work)} \SN \GE{to catch in midair} \PD{kyoŋee, kyonno, kyonno, kyonimo, kyonnaa, - } 
\LX{kyɔnkyɔlemo} \HM{} \PH{ʧɔ̀nʧɔ́lɪ́mɔ́} \PS{n.~} \GE{hanging bits of an object} 
\LX{kyonkyolo} \HM{} \PH{ʧònʧóló} \PS{n.~} \GE{loose skirt} \TP{kyonkyolee} 
\LX{kyɔnkyɔlɔŋ} \HM{} \PH{ʧɔ̀nʧɔ̀lɔ̀ŋ} \PS{adv.~} \GE{long looking (rope)} 
\LX{kyonkyoŋ} \HM{} \PH{ʧonʧoŋ} \PS{v.~} \GE{to jump around and over a spot} \PD{kyonkyoŋee, kyonkyono, kyonkyono, kyonkyonimo, kyonkyonaa, - } \VA{kyoŋkyoŋ}
\LX{kyonni} \HM{} \PH{ʧónnì} \PS{v.~} \GE{to pound fried shea nuts into a thick paste to make shea butter} \PD{kyonnee, kyonno, kyonno, kyonimo, kyonnaa, - } 
\LX{kyonni} \HM{} \PH{ʧónní} \PS{v.~} \SN \GE{to drip} \SN \GE{to trim} \PD{kyonnee, kyonno, kyonno, kyonimo, kyonnaa, kyonimo } 
\LX{kyoo} \HM{} \PH{ʧóò} \PS{n.~} \GE{nest (in tree)} \OP{kyoori} \TP{kyooree} \VA{kyogi, kyogri, kyogree}
\LX{kyɔɔ} \HM{} \PH{ʧɔ̀ɔ̀} \PS{v.~} \GE{to hold gingerly, hanging from the tips of fingers, teeth, claws, or beak} \PD{kyɔɔɛ, kyɔɔrɔ, kyɔɔrɔ, kyɔɔrebɔ, kyɔɔraa, - } 
\LX{kyɔɔ} \HM{} \PH{ʧɔ̀ɔ̀} \PS{adj.~} \GE{long (dress)} 
\LX{kyoone} \HM{} \PH{ʧʊ́ʊ́nɪ́} \PS{n.~} \GE{shea nut} \OP{kyoomɔ} \TP{kyoonɛɛ} 
\LX{kyoɔne} \HM{} \PH{ʧʊ̀ɔ̀nɪ̀} \PS{adj.~} \GE{stripped} 
\LX{kyoɔnɔkyoɔnɔ} \HM{} \PH{ʧʊ̀ɔ̀nɔ̀ʧʊ́ɔ́nɔ̀} \PS{adj.~} \GE{stripped in many colours} 
\LX{kyɔɔɔ} \HM{} \PH{ʧɔ̀ɔ̀ɔ̀} \PS{adv.~} \GE{long and draping (of a hanging object)} 
\LX{kyooraa} \HM{} \PH{ʧʊ́ʊ́ráá} \PS{v.~} \GE{hernia} \PD{-, -, -, -, -, - } 
\LX{kyoore} \HM{} \PH{ʧʊ́ʊ́rɪ́} \PS{v.~} \SN \GE{a clicking sound used to call a dog} \SN \GE{expression of strong dislike} \PD{kyoorɛɛ, kyoora, kyoora, kyooreba, kyooraa, - } 
\LX{kyoore} \HM{} \PH{ʧʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to suck to the last bit} \PD{kyoorɛɛ, kyoora, kyoora, kyooreba, kyooraa, - } 
\LX{kyoɔre} \HM{} \PH{ʧʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to filter, to strain} \PD{kyoɔrɛɛ, kyoɔrɔ, -, -, kyoɔraa, kyoɔre } 
\LX{kyoɔre} \HM{} \PH{ʧʊ̀ɔ̀rɪ̀} \PS{v-adj.~} \GE{stand loosely, helplessly (in a confused state)} \XV{o kyoɔre are la a teɛ pare kyɛ kaara a zɔɔre.} \XE{S/he stood under the tree, perplexed, watching the fight.} 
\LX{kyɔɔtaa} \HM{} \PH{ʧɔ̀ɔ̀tàá} \PS{n.~} \GE{offer, gift} \OP{kyɔɔtaare} \TP{kyɔɔtaarɛɛ} 
\LX{kyore} \HM{} \PH{ʧʊ́rɪ́} \PS{v.~} \SN \GE{to be worn} \SN \GE{to be totally destroyed} \PD{kyorɛɛ, kyorɔ, kyorɔ, kyoreba, kyoraa, - } 
\LX{kyore} \HM{} \PH{ʧʊ́rɪ́} \PS{n.~} \GE{} \OP{kyorre} \TP{kyorɛɛ} \VA{shortened form of kyolenkyooraa}
\LX{kyɔre} \HM{} \PH{ʧɔ̀rɪ̀} \PS{v.~} \SN \GE{to be saturated with water} \SN \GE{to pound in order to shell (nuts)} \SN \GE{to loosen soil} \SN \GE{to do things anyhow (speaking)} \PD{kyɔrɛɛ, kyɔrɔ, kyɔrɔ, kyɔreba, kyɔraa, kyɔreba } \SE{saa kyore} \GE{period of intense rain; downpour} 
\LX{kyɔre} \HM{1} \PH{ʧɔ́rɪ́} \PS{v.~} \SN \GE{to slice; to chop into bits} \SN \GE{to sprinkle water on} \PD{kyɔrɛɛ, kyɔrɔ, kyɔrɔ, kyɔreba, kyɔraa, kyɔreba } 
\LX{kyɔre} \HM{2} \PH{ʧɔ́rɪ́} \PS{v.~} \GE{to do things anyhow (speaking)} \PD{-, -, -, -, -, - } 
\LX{kyori} \HM{} \PH{ʧórí} \PS{v.~} \GE{to cut or chop into pieces} \PD{kyoree, kyoro, kyoro, kyoribo, kyoraa, - } 
\LX{kyoriforo} \HM{} \PH{ʧórífóró} \PS{adv.~} \GE{squating position (open legs)} 
\LX{kyoroso} \HM{} \PH{ʧʊ́rʊ́sʊ́} \PS{adj.~} \SN \GE{of worth} \SN \GE{appearance like a twisted mouth} 
\LX{kyo᷉} \HM{} \PH{ʧó᷉} \PS{adv.~} \GE{silent, not uttering a word} 
\LX{kyo᷉o} \HM{} \PH{ʧʊ̀᷉ʊ̀᷉} \PS{adv.~} \GE{of the splash of water, when one jumps into it} 
\LX{kyo᷉ɔ} \HM{} \PH{ʧʊ́ɔ̀} \PS{v.~} \GE{to leap} \PD{kyo᷉ɔɛ, kyoɔnɔ, kyoɔnɔ, kyoɔnema, kyoɔnaa, - } 
\LX{kyo᷉ɔ} \HM{} \PH{ʧʊ̀ɔ̀} \PS{v.~} \GE{to spring, to jump} \PD{kyo᷉ɔɛ, kyoɔnɔ, kyoɔnɔ, kyoɔnemɔ, kyoɔnaa, - } 
\LX{kyo᷉ɔ} \HM{} \PH{ʧʊ̀ɔ̀} \PS{v.~} \GE{to leap, to sprint} \PD{kyo᷉ɔɛ, kyoɔnɔ, kyoɔnɔ, kyoɔnema, kyoɔnaa, - } 
\LX{kyugiloŋ} \HM{} \PH{ʧùglòŋ} \PS{adj.~} \GE{heavy and hung out (head)} 
\LX{kyugiloŋ} \HM{} \PH{ʧúglóŋ} \PS{adv.~} \GE{small and hung down (head)} 
\LX{kyuli} \HM{} \PH{ʧùlì} \PS{v.~} \GE{to bend one's head down} \PD{kyulee, kyullo, kyullo, kyulibo, kyullaa, - } 
\LX{kyulokyulo} \HM{} \PH{ʧúlóʧúló} \PS{adj.~} \GE{confined within a small area (giving assistance)} 
\LX{kyulom} \HM{} \PH{ʧúlóm} \PS{adv.~} \GE{disappearing at once into water or a hole} \VA{kyuloŋ}
\LX{kyuloŋ} \HM{} \PH{ʧúlóŋ} \PS{adv.~} \GE{} \VA{kyulom}
\LX{kyumm} \HM{} \PH{ʧúmm} \PS{adj.~} \GE{going straight down} 
\LX{kyunkyuŋ} \HM{} \PH{ʧúnʧúŋ} \PS{adj.~} \GE{very deep} 
\LX{kyuoli} \HM{} \PH{ʧùòlì} \PS{v.~} \GE{to dazzle the eyes (of light)} \PD{kyuolee, kyuolo, kyuolo, kyuolbo, kyuolaa, - } 
\LX{kyuori} \HM{} \PH{ʧúórì} \PS{v.~} \GE{empty (stomach)} \PD{kyuoree, kyuoro, kyuoro, kyuoribo, kyuoraa, - } 
\LX{kyuri} \HM{} \PH{ʧúrí} \PS{v.~} \SN \GE{to fall through} \SN \GE{to abort (pregnancy)} \PD{kyuree, kyuro, kyuro, kyuribo, kyuraa, - } 
\LX{kyuu} \HM{} \PH{ʧúù} \PS{n.~} \SN \GE{moon} \SN \GE{month} \SG{kyuuruu} \OP{kyuuri} \TP{kyuuree} 
\LX{kyuu} \HM{} \PH{ʧùú} \PS{n.~} \GE{poultry lice} \OP{kyuuri} \TP{kyuuree} 
\LX{kyuu} \HM{} \PH{ʧúú} \PS{adv.~} \GE{sleepy} 
\LX{kyuukyuu} \HM{} \PH{ʧúùʧúù} \PS{adv.~} \GE{monthly} 
\LX{kyuupaalaa} \HM{} \PH{ʧúúpáálàà} \PS{n.~} \SN \GE{new month} \SN \GE{crescent moon, new moon} 
\LX{kyuupaaleŋ} \HM{} \PH{ʧúúpààlɪ̀ŋ} \PS{n.~} \GE{full moon} 
\LX{kyuupelaa} \HM{} \PH{ʧúúpɪ̀làá} \PS{n.~} \GE{moonlight} \OP{kyuupeɛle} \TP{kyuupeɛlɛɛ} 
\LX{kyuuri} \HM{} \PH{ʧùùrì} \PS{v.~} \GE{to pour out from a height onto (of a waterfall)} \PD{kyuuree, kyuuro, kyuuro, kyuuribo, kyuuraa, - } 
\LX{la} \HM{} \PH{lá} \PS{particle.~} \SN \GE{emphatic particle be; is, are} \XV{Ona la na gaa.} \XE{S/he is the one who will go.} \SN \GE{helping verb} \XV{O na gaa la.} \XE{S/he will go.} 
\LX{la} \HM{} \PH{là} \PS{v.~} \GE{to laugh} \PD{laɛ, laara, laara, laareba, laaraa, laareba } 
\LX{la} \HM{} \PH{là} \PS{adv.~} \GE{again} \XV{O la gaaɛ la.} \XE{S/he has gone again.} 
\LX{la} \HM{} \PH{lá} \PS{v.~} \GE{to open or break an object made of lobes. (kola nut, fruit)} \PD{laɛ, laara, laara, lareba, laaraa, - } 
\LX{laa} \HM{} \PH{láá} \PS{n.~} \GE{dish, bowl} \OP{laare} \TP{laarɛɛ} 
\LX{laadimee} \HM{} \PH{làadíméé́} \PS{n.~} \GE{great suffering} \TP{laadinnee} 
\LX{laafeɛ} \HM{} \PH{làafɪ́ɛ́̀} \PS{n.~} \GE{health} \OP{laafeere} \TP{laafeerɛɛ} 
\LX{laalee} \HM{} \PH{lààléè} \PS{n.~} \GE{small eartheware dish for serving soup} \OP{laabilii} \TP{laabilee} 
\LX{laalɛŋ} \HM{} \PH{lààlɛ́ŋ} \PS{n.~} \GE{medium-size bowl} 
\LX{laalere} \HM{} \PH{lààlɪ́rɪ̀} \PS{n.~} \GE{flute made from stalk} \OP{laaleɛ} \TP{laalerɛɛ/laalerrɛɛ} 
\LX{laane} \HM{} \PH{láánɪ́} \PS{adj.~} \GE{useless} \OP{laama} \TP{laamɛɛ} 
\LX{laane} \HM{} \PH{láánɪ́} \PS{adj.~} \GE{ripe archie-fruit (showing a split)} \OP{laama} \TP{laamɛɛ} 
\LX{laane} \HM{} \PH{làànɪ̀} \PS{v.~} \GE{to grope, to move one's hand over to feel} \PD{laanɛɛ, laana, laana, laaneba, laanaa, - } 
\LX{laare} \HM{} \PH{láárɪ̀} \PS{v.~} \SN \GE{to wave e.g. greeting} \SN \GE{to move or spread out e.g. people in a circle} \SN \GE{to brush aside} \PD{laarɛɛ, laara, laara, laareba, laaraa, laareba } 
\LX{laare} \HM{} \PH{láárɪ̀} \PS{v.~} \GE{to hover above e.g. bird} \PD{laarɛɛ, laara, laara, laareba, laaraa, laareba } 
\LX{laare} \HM{} \PH{lààrɪ̀} \PS{v.~} \SN \GE{to spread all over} \XV{Saa mie la ka kõɔ laare a bombulo zaa.} \XE{The rain has caused a near flooding of all the crops.} \SN \GE{to move a finger gently over} \XV{A dɔɔta de la tẽẽ kãã mine laare ne a bie nimie.} \XE{The doctor applied some creamy medicine to the child’s eyes.} \PD{laarɛɛ, laara, laara, laareba, laaraa, - } 
\LX{laare} \HM{} \PH{lààrɪ̀} \PS{v.~} \GE{to assemble; to bring together} \PD{laarɛɛ, laara, laara, laareba, laaraa, - } 
\LX{laatereke} \HM{} \PH{láátɪ́rɪ́kɪ̀} \PS{n.~} \GE{electricity} \OP{laaterekere} \TP{laaterekerɛɛ} 
\LX{labiri} \HM{} \PH{lábírì} \PS{n.~} \GE{axe blade} \OP{labie} \TP{labiree} 
\LX{ladiraa} \HM{} \PH{ládìráà} \PS{n.~} \GE{sharp axe} \OP{ladirri} \TP{ladiree} 
\LX{laga} \HM{} \PH{lágà} \PS{n.~} \GE{spoon} \OP{lagre} \TP{lagrɛɛ} 
\LX{laga} \HM{} \PH{làgá} \PS{n.~} \GE{groin} \OP{lagre} \TP{lagɛɛ} 
\LX{lagara} \HM{} \PH{lágàrà} \PS{n.~} \GE{a type of creeping grass} \SG{lagaroo} \OP{lagarre} \TP{lagarɛɛ} 
\LX{lage} \HM{} \PH{lágɪ̀} \PS{v.~} \GE{to strip off hard material e.g. bark} \PD{lagɛɛ, lagra, lagra, lagreba, lagraa, lagreba } \VA{geli}
\LX{lage} \HM{} \PH{lágɪ̀} \PS{v.~} \GE{to remove malted grain e.g guinea corn from the place it was left to sprout} \PD{lagɛɛ, lagra, lagra, lagreba, lagraa, lagreba } \VA{geli}
\LX{lage} \HM{} \PH{làgɪ̀} \PS{v.~} \GE{to surround} \PD{lagɛɛ, lagra, lagra, lagreba, lagraa, lagreba } 
\LX{lagere} \HM{} \PH{lágrɪ́} \PS{v.~} \GE{to meet, to move across} \XV{A dataare na lagere la taa a naayiri poɔ'} \XE{The opposing parties will meet at the chief's palace.} \PD{lagerɛɛ, lagera, lagera, lagereba, lagraa, - } 
\LX{laguu} \HM{} \PH{lágùú} \PS{n.~} \GE{blunt axe} \OP{laguuri} \TP{laguuree} 
\LX{lakpoŋi} \HM{} \PH{lákpôŋ} \PS{n.~} \GE{large axe} \OP{lakponni} \TP{lakponnee} 
\LX{lalee} \HM{} \PH{lál̊èé} \PS{n.~} \GE{small axe} \OP{labilii} \TP{labilee} 
\LX{laloɔrɔ} \HM{} \PH{lálʊ́ɔ́rɔ́} \PS{n.~} \GE{playmate} \OP{laloɔreba} \TP{laloɔrɛɛ} 
\LX{lambiri} \HM{} \PH{lǎmbírì} \PS{n.~} \GE{testicle} \SG{lambiruu} \OP{lambie} \TP{lambiiree} 
\LX{lambori} \HM{} \PH{làmbórí} \PS{n.~} \SN \GE{side} \SN \PS{postposition.~} \GE{side (of something)} \OP{lamboe/lambobo} \TP{lamboree} \VA{lambobo (1pl)}
\LX{lamboriŋ} \HM{} \PH{làmbórîŋ} \PS{loc.~} \GE{beside} 
\LX{lambosentaa-naare} \HM{} \PH{làmbósɪ́ntáá-náárɪ̀} \PS{n.~} \GE{square} 
\LX{lammo} \HM{} \PH{làmmʊ́} \PS{n.~} \GE{gathering} 
\LX{lampare} \HM{} \PH{làmpàrɪ̀} \PS{n.~} \GE{groin} \OP{lampare} \TP{lamparɛɛ} 
\LX{lampɛ} \HM{} \PH{lâmpɛ́} \PS{n.~} \GE{cradle cover} \OP{lampɛre} \TP{lampɛrɛɛ} 
\LX{lampɛgere} \HM{} \PH{lámpɛ́gɪ́rɪ́} \PS{n.~} \GE{cheek} \OP{lampɛgɛ} \TP{lampɛgɛɛ/lampɛgerɛɛ} 
\LX{lampoŋ} \HM{} \PH{lǎmpóŋ} \PS{n.~} \GE{levy, tax} \OP{lamponni} \TP{lamponnee} 
\LX{lamponderɛ} \HM{} \PH{lǎmpóndɪ́rɛ́} \PS{n.~} \GE{tax collector} \OP{lamponderebɛ} \TP{lamponderebɛɛ} 
\LX{laŋa} \HM{} \PH{láŋá} \PS{n.~} \GE{a tribe (Sisaale)} \OP{laŋema} \TP{laŋemɛɛ} 
\LX{lane} \HM{} \PH{lánɪ́} \PS{n.~} \GE{scrotum} \OP{lama} \TP{lannɛɛ} \VA{lambiri, lambie}
\LX{Laŋe} \HM{} \PH{làŋɪ́} \PS{n.~} \GE{area inhabited by the Sisaala} 
\LX{laŋe} \HM{} \PH{làŋɪ̀} \PS{v.~} \SN \GE{to gather, to get together} \SN \GE{to add} \SN \GE{to be jointly owned} \PD{laŋɛɛ, lanna, -, -, -, - } 
\LX{laŋe} \HM{} \PH{láŋɪ́} \PS{v.~} \GE{to slosh} \PD{laŋɛɛ, laŋena/lanna, lanna, laŋenema/lanema, laŋenaa, - } 
\LX{laŋgbaraa} \HM{} \PH{láŋgbáráà} \PS{n.~} \GE{a long pole with a hook for plucking things high up} \OP{laŋgbarre} \TP{laŋgbarɛɛ} \VA{laŋgbalaa, gbalaa}
\LX{laŋkoroo} \HM{} \PH{láŋkóróò} \PS{n.~} \GE{corner} \OP{laŋkorri} \TP{laŋkorree} 
\LX{laŋkpaa} \HM{} \PH{lâŋkpáà} \PS{n.~} \GE{groin gland} \OP{laŋkpaare} \TP{laŋkpaarɛɛ} \VA{loŋkoroo}
\LX{laŋkpeɛrɛ} \HM{} \PH{làŋkpɪ́ɛ́rɛ́} \PS{n.~} \GE{neighbour} \OP{laŋkpeɛrebɛ} \TP{laŋkpeɛrebɛɛ} 
\LX{lanleebu} \HM{} \PH{lànléébú} \PS{n.~} \GE{unison} 
\LX{lanleebu} \HM{} \PH{lànléébú} \PS{n.~} \GE{rhythm} 
\LX{lanna} \HM{} \PH{lànná} \PS{n.~} \GE{tree squirrel} \OP{lanne} \TP{lannɛɛ} \VA{lɔnnɔ}
\LX{lanne} \HM{} \PH{lánnɪ̀} \PS{v.~} \GE{to lean against} \PD{lannɛɛ, lanna, lanna, lanema, lannaa, lanema } 
\LX{larata} \HM{} \PH{lárátá} \PS{adj.~} \GE{white, very clean e.g. cloudless full moon} 
\LX{lare} \HM{} \PH{lárì} \PS{n.~} \GE{axe} \OP{laɛ} \TP{larɛɛ} 
\LX{lare} \HM{} \PH{lárɪ́} \PS{v.~} \GE{to become very hot} \PD{larɛɛ, lara, lara, lareba, laraa, - } 
\LX{lare} \HM{} \PH{làrì} \PS{v.~} \GE{to lock} \PD{larɛɛ, lara, lara, lareba, laraa, - } 
\LX{lare} \HM{} \PH{làrì} \PS{v.~} \GE{to stack one container in another} \PD{larɛɛ, lara, lara, lareba, laraa, - } 
\LX{le} \HM{} \PH{lè} \PS{v.~} \SN \GE{to fall (down)} \SN \GE{to be guilty} \SN \GE{to settle e.g. sediment in liquid} \SN \GE{to reduce in pain e.g. stomach ache} \PD{lee, lere, lere, leribe, leraa, leribe } 
\LX{le} \HM{} \PH{lé} \PS{v.~} \GE{to be bent on} \XV{A porisi le ka o gaa ne a doreba kɔɔto.} \XE{The police officer is bent on prosecuting the driver in court.} \PD{-, -, -, -, -, - } 
\LX{lɛ} \HM{} \PH{lɛ̌} \PS{adv.~} \SN \GE{so, usually} \XV{Lɛ ka a waa.} \XE{So it is.} \SN \GE{as it is, at that point} \XV{Meŋ bare a zoma lɛ.} \XE{Also, stop there with the insults.} \SN \GE{what} \XV{N na te woŋi la lɛ o naŋ na yeli.} \XE{I will go and listen to what s/he will say.} 
\LX{le ko᷉ɔ} \HM{} \PH{lè kʊ̀᷉ɔ́} \PS{adj.~} \GE{small} 
\LX{le ko᷉ɔ} \HM{} \PH{lè kʊ̀᷉ɔ́} \PS{adj.~} \GE{watery} 
\LX{le lɔɔ} \HM{} \PH{lè lɔ́g} \PS{adj.~} \GE{guilty of the law} 
\LX{le nimiri} \HM{} \PH{lè nímírì} \PS{adj.~} \GE{pleasant} 
\LX{le tobo} \HM{} \PH{lè tòbó} \PS{adj.~} \GE{beautiful} 
\LX{lɛbiri} \HM{} \PH{lɛ̀bírì} \PS{n.~} \GE{bead} \SG{lɛbiruu} \OP{lɛbie} \TP{lɛbiiree} 
\LX{leɛ} \HM{} \PH{lɪ̀ɛ̀} \PS{v.~} \SN \GE{to return, to go back} \SN \GE{to transfer to a different container} \SN \GE{to change to something different e.g. behaviour} \PD{leɛɛ, leɛrɛ, leɛrɛ, leɛrebɛ, leɛraa, leɛrebɛ } \SE{leɛ gaa} \GE{} \SE{leɛ puori} \GE{} 
\LX{leɛ} \HM{} \PH{lɪ̀ɛ̀} \PS{v.~} \SN \GE{to prepare self in readiness e.g. going on a journey} \SN \GE{to prepare an offering for a guest to take away} \PD{leɛɛ, leɛrɛ, leɛrɛ, leɛrebɛ, leɛraa, leɛrebɛ } \SE{leɛ gaa} \GE{} \SE{leɛ puori} \GE{} 
\LX{leɛ} \HM{} \PH{lɪ́ɛ̀} \PS{v.~} \GE{to replace, to take the place of} \PD{leɛɛ, leɛrɛ, leɛrɛ, leɛrebɛ, leɛraa, - } 
\LX{lɛɛ} \HM{} \PH{lɛ̀ɛ̂} \PS{use word.~} \GE{is that so?} \XV{a. A dɔɔ ŋmaa na gaa la dansareka kyɛ yi. b. Lɛɛ?} \XE{a. That short man is an ex-convict. b. Is that so?.} 
\LX{lɛɛ} \HM{} \PH{lɛ́ɛ́} \PS{adv.~} \SN \GE{for a while (wait)} \XV{o are la be lɛɛ wa te bale.} \XE{S/he waited until s/he became tired.} \SN \GE{continuous and for quite sometime} \XV{Ba yeli ka a sori e la ŋmaa kyɛ ka te kyɛŋe o lɛɛ ka ŋmenaa te kpɛ.} \XE{We were told the journey was not long but walked and walked until dusk.} \VA{lɛɛŋ}
\LX{leɛ-eebo} \HM{} \PH{lìɛ̀-ɪ́ɪ́bʊ́} \PS{v.~} \GE{to redo; to go over} \PD{-, -, -, -, -, - } 
\LX{leɛgɔɔtɔbaa} \HM{} \PH{lɪ́ɛ́gɔ̀ɔ́tɔ̀báá} \PS{n.~} \GE{a local dish made from fried corn and vegetable} \TP{leɛgɔɔtɔbaarɛɛ} 
\LX{leɛne} \HM{} \PH{lɪ́ɛ́nɪ̀} \PS{n.~} \GE{a type of tree with sweet and sour fruits and medicinal properties} \OP{leɛmɛ} \TP{leɛmɛɛ} 
\LX{leɛne} \HM{} \PH{lɪ́ɛ́nɪ̀} \PS{n.~} \GE{large umbrella} \OP{leɛmɛ} \TP{leɛmɛɛ} 
\LX{leere} \HM{} \PH{lɪ̀ɪ̀rɪ̀} \PS{v.~} \SN \GE{to manouvre, to weave through} \XV{O de la a bɔle leere o taaba zaa kyɛ eŋ gool.} \XE{He dribbled the ball through the other players and scored a goal.} \SN \GE{to go through and select from many objects leaving some behind} \XV{O dɔ la a moɔ kyɛ leere a bombulo bare.} \XE{He removed the weeds, leaving the crops.} \SN \GE{to exclude} \XV{Ba poŋi la a bibiiri diibu kyɛ te leere a entuo soba bare.} \XE{They shared the food among the children, excluding the lazy one.} \SN \GE{to get to the bottom of (hole), knooks (cheeks)} \XV{O naŋ ɔɔ a seŋka᷉a baare la ka o leere o yagekɔbɔ.} \XE{It is after she had finished eating the groudnuts that she brushed her cheeks using her tongue.} \PD{leerɛɛ, leerɛ, leerebɛ, leerebɛ, leeraa, - } 
\LX{leɛre} \HM{} \PH{lɪ̀ɛ̀rɪ̀} \PS{v.~} \SN \GE{to change (appointment, time, place)} \XV{Ba leɛre la a kpaaroŋ biŋ bieo.} \XE{They have rescheduled the meeting for tomorrow.} \SN \GE{to return, to give back} \XV{o leɛre la a o gane ko o.} \XE{She has returned his book to him.} \SN \GE{to repudiate, to turn round} \XV{A nannyige boɔrɔ ka o leɛre a yɛlɛ tuo a o tutɔ.} \XE{The thief is trying to shift the theft onto his friend.} \PD{leɛ rɛɛ, leɛrɛ, leɛre terɛ, leɛre terebɛ, leɛ teraa, - } 
\LX{leɛre} \HM{} \PH{lɪ́ɛ́rɪ̀} \PS{v.~} \SN \GE{to exchange} \XV{O de la bɛŋɛ leɛre ne kyi.} \XE{S/he substituted beans for millet.} \SN \GE{change position} \XV{A dɔɔ leɛre la a boore kpɛ ne maaroŋ} \XE{The man moved the goats from the sun into the shade.} \SN \GE{substitute, take the place of} \XV{A bɔlŋmeɛrɛ kaŋa perɛɛ la ka bonyuo kpɛ leɛre o zu.} \XE{As one of the football players was injured they brought in another player to substitute him.} \PD{-, -, -, -, -, - } 
\LX{leɛre} \HM{} \PH{lɪ́ɛ́rɪ̀} \PS{v.~} \SN \GE{to wave one's hands or other objects in the air or in the face of another} \XV{A zɔɔzɔɔreba da leɛrɛ la nuuri ennɛ taa.} \XE{The fighters were throwing their hands in each other's face.} \XV{A lɔɔre naŋ da wa ire n ma da leɛre la o nu ko ma ka n poɔ zaa sãã.} \XE{As the lorry took off my mother waved me but I was very sad.} \SN \GE{wave hand vigorously in the face of another, usually in anger} \XV{Dɔbɔ mine bayi da toora la taa kyɛ leɛrɛ nuuri ennɛ taa.} \XE{Two men were waving their hands angrily in each other's face.} \PD{-, -, -, -, -, - } 
\LX{lege} \HM{} \PH{lɪ̀gɪ̀} \PS{v.~} \GE{to apply to a spot} \XV{A te᷉e-erɛ de la te᷉e lege a baala tobo ne o gbɛɛ.} \XE{The herbalist applied some ointment to the patient's ears and legs.} \PD{legɛɛ, legerɛ, legerɛ, legerebɛ, legeraa, - } 
\LX{lege} \HM{} \PH{lɪ̀gɪ̀} \PS{v.~} \GE{to fortify oneself} \XV{Ba lege la a naŋkpaana ŋa mɔyɔ te᷉e.} \XE{This hunter has been fortified against the attack of dangerous animals.} \PD{legɛɛ, legerɛ, legerɛ, legerebɛ, legeraa, - } 
\LX{lege} \HM{} \PH{lɪ́gɪ́} \PS{v.~} \SN \GE{to wag (dog's tail)} \XV{A baa da zo tuori la o daana a legerɛ o zoore.} \XE{The dog ran to meet its master and wagged its tail.} \SN \GE{to struggle} \PD{legɛɛ, legerɛ, legerɛ, legerɛ, legeraa, - } 
\LX{legelege} \HM{} \PH{lɪ̀gɪ̀lɪ́gɪ́} \PS{v.~} \GE{to struggle under a tight grip} \XV{A dɔɔ da feere la a nannyige kɔkɔre ka o te legelegerɛ.} \XE{The man held the thief's throat and he struggled to breathe.} \PD{legelegɛɛ, legelegɛ, legelegɛ, legelegerebɛ, legelegeraa, - } 
\LX{legere} \HM{} \PH{lɪ́gɪ́rɪ̀} \PS{v.~} \GE{to tickle} \PD{legrɛɛ, legrɛ, legrɛ, legrebɛ, legraa, legrebɛ } 
\LX{legere} \HM{} \PH{lɪ́gɪ́rɪ̀} \PS{v.~} \GE{to stalk} \PD{legrɛɛ, legrɛ, legrɛ, legrebɛ, legraa, legrebɛ } 
\LX{legere} \HM{} \PH{lɪ́gɪ́rɪ̀} \PS{v.~} \GE{to eat a great deal} \PD{legrɛɛ, legrɛ, legrɛ, legrebɛ, legraa, legrebɛ } 
\LX{legere} \HM{} \PH{lìgɪ̀rɪ̀} \PS{v.~} \GE{to move noiselessly to avoid being noticed} \PD{legerɛɛ, legerɛ, legerɛ, legerebɛ, legeraa, - } 
\LX{lɛgerɛŋ} \HM{} \PH{lɛ́gɪ̀rɛ́ŋ} \PS{adv.~} \GE{long, loose, and hanging} 
\LX{legiri} \HM{} \PH{légirí} \PS{v.~} \GE{to fill in what is missing e.g. sowing of grains} \XV{Kɔrre da pere la kyi yaga, ka a koɔreba la te legiri.} \XE{Partridges have scatched most of the millet that was sown and the farmers had to resow.} \PD{legiree, legire, legire, legiribe, legiraa, - } 
\LX{legiri} \HM{} \PH{légirí} \PS{v.~} \GE{to argue} \XV{Dagaaba ba nɔŋ bibile naŋ legire neŋkponni.} \XE{The Dagaaba do not take kindly to children who argue with elderly people.} \PD{legiree, legire, legire, legiribe, legiraa, - } 
\LX{lekere} \HM{} \PH{lɪ́kɪ́rɪ́} \PS{n.~} \GE{boastfulness; assuming (typically used with yɛlɛ)} \XV{A bikpoŋ na erɛ la lekere yɛlɛ yelmeŋɛ.} \XE{That big boy behaves in a boastful manner.} 
\LX{lelaare} \HM{} \PH{lèlààrɪ̀} \PS{v.~} \GE{to take complete control;} \PD{lelaarɛɛ, lerelaara, lelaara, lelaareba, lelaaraa, - } 
\LX{lɛmɛlɛmɛ} \HM{} \PH{lɛ́mɛ́lɛ́mɛ́} \PS{adv.~} \GE{slightly sweet} 
\LX{lɛmpɛrɛŋ} \HM{} \PH{lɛ́mpɛ́rɛ́ŋ} \PS{adv.~} \GE{short and even at the top} 
\LX{lɛnɛɛ} \HM{} \PH{lɛ̀nɛ́ɛ̂} \PS{conj.~} \SN \GE{what} \XV{N ba baŋe lɛnɛɛ N naŋ na e pampana.} \XE{I don’t know what to do next.} \SN \GE{how} \XV{Fo baŋe la lɛnɛɛ ba naŋ maŋ sɛ kparoo?} \XE{Do you know how a shirt is sewn?.} 
\LX{leŋi} \HM{} \PH{lèŋ} \PS{v.~} \SN \GE{to fasten, to bind e.g. load} \SN \GE{to tether} \PD{leŋee, lene, lene, lenime, lenaa, - } 
\LX{leŋi miri} \HM{} \PH{léŋí mírì} \PS{v.~} \GE{to commit suicide} \PD{leŋee miri, lene miri, mi-lene, -, -, - } 
\LX{leŋi noɔre} \HM{} \PH{léŋ nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to fast; to refrain from eating} \PD{leŋee noɔre, lene noɔre, nɔ-lene, -, -, - } 
\LX{leŋi woɔ} \HM{} \PH{léŋ wʊ́ɔ́} \PS{v.~} \GE{to pack bags} \PD{leŋee woɔ, lene woɔ, wo-lene, -, -, - } 
\LX{lɛnlɛŋko} \HM{} \PH{lɛ́nlɛ́ŋkò} \PS{use word.~} \GE{expression to soothe a child; usual tune in a cradle song} 
\LX{lenlenni} \HM{} \PH{lènlènní} \PS{n.~} \GE{a type of skin disease} \TP{lenlennee} 
\LX{lɛnne} \HM{} \PH{lɛ̀nnɪ̀} \PS{v.~} \SN \GE{to lick, to taste} \SN \GE{to offer a sample of food for sale} \PD{lɛnnɛɛ, lɛnnɛ, lɛnnɛ, lɛnemɛ, lɛnnaa, lɛnemɛ } 
\LX{lɛŋŋɛlɛŋŋɛ} \HM{} \PH{lɛ́ŋŋɛ́lɛ́ŋŋɛ́} \PS{n.~} \GE{sickle} \OP{lɛŋŋɛlɛnne} \VA{sɔgɔraa}
\LX{lenni} \HM{} \PH{lénní} \PS{v.~} \GE{to be spotted} \PD{lennee, lenne, lenne, lenime, lennaa, - } 
\LX{lenni} \HM{} \PH{lénní} \PS{v.~} \GE{to be interpersed} \PD{lennee, lenne, lenne, lenime, lennaa, - } 
\LX{lenni} \HM{} \PH{lénní} \PS{v.~} \SN \GE{to augment the main drummer's beat (of support xylophonist)} \SN \GE{play a musical instrument in support of the lead player} \PD{lennee, lenne, lenne, lenime, lennaa, - } 
\LX{lɛntɛre} \HM{} \PH{lɛ́ntɛ́rɪ́} \PS{n.~} \GE{thinking too much of oneself} \TP{lɛntɛrɛɛ} 
\LX{leraa} \HM{} \PH{lɪ́ráá} \PS{n.~} \GE{safety pin} \OP{lerre} \TP{lerɛɛ/lerrɛɛ} 
\LX{lere} \HM{} \PH{lérè} \PS{n.~} \GE{a type of loincloth for men} \OP{lerri} \TP{leree/lerree} 
\LX{lere} \HM{} \PH{lɪ́rɪ́} \PS{v.~} \SN \GE{to interweave} \SN \GE{crisscross e.g sitting with legs crossed} \PD{lerɛɛ, lerɛ, lerɛ, lerebɛ, leraa, - } 
\LX{lɛre} \HM{} \PH{lɛ̀rɪ́} \PS{n.~} \GE{bead} \OP{lɛɛ} \TP{lɛrɛɛ} 
\LX{lɛre} \HM{} \PH{lɛ́rɪ́} \PS{v.~} \GE{to behave badly to get attention e.g. of a child} \PD{lɛrɛɛ, lɛrɛ, lɛrɛ, lɛrebɛ, lɛraa, - } \VA{lɛroo}
\LX{lɛre} \HM{} \PH{lɛ́rɪ̀} \PS{v.~} \GE{to replace, to put in the place of} \XV{A bɔlŋmeɛrɛ kaŋa perɛɛ la ka boŋyuo kpɛ lɛre o zu.} \XE{One of the football players got injured and another replaced him.} \PD{lɛrɛɛ, lɛrɛ, lɛrɛ, lɛrebɛ, lɛraa, - } 
\LX{leree} \HM{} \PH{lèrèè} \PS{adj.~} \GE{naked, bare bodied} 
\LX{leremɛ} \HM{} \PH{lɪ́rɪ́mɛ́} \PS{adj.~} \GE{interwoven} 
\LX{lerese} \HM{} \PH{lɪ́rɪ́sɪ́} \PS{adv.~} \GE{small but intact (person)} 
\LX{lezaɛ anaare} \HM{} \PH{lɪ̀záɛ̀ ánáárɪ̀} \PS{adv.~} \GE{eighty} 
\LX{lezaɛ ata} \HM{} \PH{lɪ̀záɛ̀ átà} \PS{adv.~} \GE{sixty} 
\LX{lezaɛ ata ne pie} \HM{} \PH{lɪ̀záɛ̀ átà nɪ́ píé} \PS{adv.~} \GE{seventy} 
\LX{lezaɛ ayi} \HM{} \PH{lɪ̀záɛ̀ áyì} \PS{adv.~} \GE{forty} 
\LX{lezaɛ ayi ne pie} \HM{} \PH{lɪ̀záɛ̀ áyì nɪ́ píé} \PS{adv.~} \GE{fifty} 
\LX{lezaɛ ayi ŋmaa anuu} \HM{} \PH{lɪ̀záɛ́ áyì ŋmàà ànúú} \PS{adv.~} \GE{thirty-five} 
\LX{lezare} \HM{} \PH{lɪ̀zárɪ̀} \PS{adv.~} \GE{twenty} 
\LX{lezare ne ayi} \HM{} \PH{lɪ̀zárɪ̀ nɪ́ píé nɪ́ yéní} \PS{adv.~} \GE{thirty-one} 
\LX{lezare ne yeni} \HM{} \PH{lɪ̀zárɪ̀ nɪ́ yénì} \PS{adv.~} \GE{twenty-one} 
\LX{libibinzie} \HM{} \PH{lìbíbìnzíé} \PS{n.~} \GE{bank} \OP{libibinziiri} \TP{libibinziiree} 
\LX{libibɔ} \HM{} \PH{líbíbɔ́} \PS{n.~} \GE{quest for money} 
\LX{libibugo} \HM{} \PH{libibùgó} \PS{n.~} \GE{safe for money} \OP{libibugiri} \TP{libibugiree} 
\LX{libidaana} \HM{} \PH{líbídááná} \PS{n.~} \GE{rich person} \OP{libidaanema} \TP{libidaanemɛɛ} 
\LX{libidi} \HM{} \PH{líbídì} \PS{n.~} \GE{the act of spending money} 
\LX{libima} \HM{} \PH{líbímá} \PS{n.~} \GE{capital} 
\LX{libipelaa} \HM{} \PH{lìbípɪ̀làá} \PS{n.~} \GE{cowry} \OP{libipeɛle} \TP{libipeɛlɛɛ} 
\LX{libiri} \HM{} \PH{lìbírì} \PS{n.~} \GE{money} \SG{libiruu} \OP{libie} \TP{libiiree} 
\LX{libizɔŋe} \HM{} \PH{líbízɔ́ŋ} \PS{n.~} \SN \GE{gold dust} \SN \GE{plentiful money} \TP{libizɔnnɛɛ} 
\LX{libu} \HM{} \PH{líbú} \PS{n.~} \GE{a funeral rite for an adult woman} \TP{libiree} \VA{ligi}
\LX{liemu} \HM{} \PH{lìèmú} \PS{n.~} \GE{orange (fruit)} \OP{liemuri} \TP{liemuree} 
\LX{lige} \HM{} \PH{lìgè} \PS{adj.~} \GE{dark} \TP{ligiree} 
\LX{ligi} \HM{} \PH{lígí} \PS{v.~} \GE{to get dark} \PD{ligee, ligre, ligre, ligraa, -, - } 
\LX{ligi} \HM{} \PH{lígí} \PS{v.~} \GE{to shroud} \PD{ligee, ligre, ligre, ligraa, -, - } 
\LX{ligi} \HM{} \PH{lìgì} \PS{v.~} \GE{to withhold information, to leave someone in the dark} \PD{ligee, ligire, ligire, ligiraa, -, - } 
\LX{ligi} \HM{} \PH{lìgì} \PS{v.~} \GE{to perform a burial rite for a woman} \PD{ligee, ligire, ligire, ligiraa, -, - } 
\LX{ligi} \HM{} \PH{lígí} \PS{n.~} \GE{a burial rite for women} \VA{libu}
\LX{ligiki} \HM{} \PH{lìgìkì} \PS{adv.~} \GE{deep and dark} 
\LX{ligiligi} \HM{} \PH{ligiligi} \PS{adj.~} \GE{silent} 
\LX{ligiruŋ} \HM{} \PH{lígirúŋ} \PS{n.~} \GE{darkness} \OP{ligire} \TP{ligiree} 
\LX{liiri} \HM{} \PH{líírí} \PS{v.~} \GE{to patch, to mend} \PD{liiree, liire, liire, liiraa, -, - } 
\LX{liiri} \HM{} \PH{lììrì} \PS{v.~} \GE{to get out of sight} \XV{Pampana ŋa ka ba deɛ kyɛŋ liiri la a taŋa puori} \XE{They just went behind the hill and out of sight a short while ago.} \PD{liiree, liire, liire, liiribe, liiraa, - } 
\LX{limbo} \HM{} \PH{límbò} \PS{n.~} \SN \GE{a transit place to heaven} \SN \GE{a state of confusion} \OP{limbori} \TP{limboree} 
\LX{limelime} \HM{} \PH{lìmèlìmè} \PS{adv.~} \GE{between light and darkness} \XV{Zimaa limelime ka noore maŋ gaŋe.} \XE{Fowls sleep when darkness is setting in.} 
\LX{limm} \HM{} \PH{lìmm} \PS{adv.~} \GE{dark, not clear} \XV{Ka zieŋ wa soɔrɔ vaare poɔ maŋ waa la limm lɛ.} \XE{When it is dusk, it is dark in the thick vegetation.} 
\LX{limpiriŋ} \HM{} \PH{lìmpìrìŋ} \PS{adj.~} \GE{oval shape and dark, not discernable} 
\LX{liŋe} \HM{} \PH{líŋé} \PS{n.~} \GE{lid, cover} \OP{linni} \TP{linnee} 
\LX{liŋgbiraa} \HM{} \PH{lìŋgbíráá} \PS{n.~} \GE{a type of tree with medicinal properties and edible fruits} \OP{liŋgbirri} \TP{liŋgbiree} 
\LX{linni} \HM{} \PH{lìnnì} \PS{v.~} \GE{to sink, to immerse e.g. as in water} \PD{linnee, linne, linne, linime, linnaa, - } 
\LX{linni} \HM{} \PH{lìnnì} \PS{v.~} \GE{to glean} \PD{linnee, linne, linne, linime, linnaa, - } 
\LX{linni} \HM{} \PH{línnì} \PS{v.~} \GE{to cover up someone e.g. from being embarrassed} \XV{A bie na la zu a libie kyɛ ka o ma linni o.} \XE{That boy stole the money but his mother covered him up.} \PD{linnee, linne, linne, linime, linnaa, - } 
\LX{liroo} \HM{} \PH{líróó} \PS{adv.~} \GE{small and protruding} 
\LX{Lobaa} \HM{} \PH{lòbàà} \PS{n.~} \GE{a person from the Lobi ethnic group} \OP{luori} \TP{lobree} 
\LX{lɔbɔ} \HM{} \PH{lɔ́bɔ́} \PS{n.~} \SN \GE{riddle} \SN \GE{A charm believed to be needle-like and has super natural power that can inflict sickness on someone} \VA{lɔba, lɔɔbiri}
\LX{lɔɛ duoraa} \HM{} \PH{lɔ́ɛ̀ dúóráá} \PS{n.~} \GE{parking lot} \OP{lɔɛ duori} \TP{lɔɛ duoree} 
\LX{loge} \HM{} \PH{lʊ̀gɪ̀} \PS{v.~} \GE{to penetrate soft ground} \PD{logɛɛ, logra, logra, logreba, lograa, - } 
\LX{lɔge} \HM{} \PH{lɔ́gì} \PS{n.~} \GE{law} \TP{lɔgerɛɛ} 
\LX{lɔge} \HM{} \PH{lɔ̀gɪ́} \PS{n.~} \GE{a type of wild mouse} \OP{lɔgere} \TP{lɔgerɛɛ} \VA{lɔɔ}
\LX{logɛɛ} \HM{} \PH{lʊ̊gɛɛ} \PS{adj.~} \GE{powdery} 
\LX{logere} \HM{} \PH{lʊ́gɪ́rɪ̀} \PS{v.~} \GE{to cause to move up and down e.g. muscle under the skin} \PD{logerɛɛ, logera, logera, logereba, logeraa, - } 
\LX{logerɔŋ} \HM{} \PH{lʊ̀gɪ̀rɔ̀ŋ} \PS{adv.~} \GE{round and fat} \XV{Nyɛ bufuli naŋ nyu ze᷉e pa᷉a waa logerɔŋ lɛ.} \XE{See how round and fat the ground maggot looks after sucking blood.} 
\LX{lɔgetɔgɔ} \HM{} \PH{lɔ̀gɪ̀tɔ̀gɔ̀} \PS{adv.~} \GE{fat, flabby, and heavy} 
\LX{logi} \HM{} \PH{lògí} \PS{n.~} \SN \GE{poison} \SN \GE{venom e.g. of snake} \OP{logiri} \TP{logiree} 
\LX{logi} \HM{} \PH{lògí} \PS{n.~} \GE{quiver; sheath for arrows} \OP{logiri} \TP{logiree} 
\LX{logi} \HM{} \PH{lógì} \PS{n.~} \GE{rest stop, encampment} \XV{Moɔre da maŋ ŋmɛ logi Kaleo poɔ kyɛ gaa kɔmpare.} \XE{The Mossi used to camp at Kaleo on their way to southern Ghana.} \OP{logiri} \TP{logiree} 
\LX{logi} \HM{} \PH{lógì} \PS{n.~} \GE{farm for crops that produce nuts} \OP{logiri} \TP{logiree} 
\LX{logi} \HM{} \PH{lògì} \PS{v.~} \GE{to make a small hole} \XV{A bie logi la bogi a dankyini seɛ de a libiri u᷉u᷉.} \XE{The child  made a hole at the base of the wall and buried the money there.} \PD{logee, logro, logro, logribo, lograa, - } \VA{loguu}
\LX{logi} \HM{} \PH{lógí} \PS{adj.~} \GE{deep} \VA{logoko}
\LX{logiri} \HM{} \PH{lògirì} \PS{v.~} \GE{to remove forcibly, to take out from a hidden or tight position e.g. crevice, hole} \XV{A kɔrre logiri la a bɛŋɛ ba naŋ bore.} \XE{The partridges have scratched to remove the beans that were sown.} \PD{logiree, logiro, logiro, logribo, logiraa, - } 
\LX{logiri noɔre} \HM{} \PH{lógírí nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to pester, to bother (for information)} \XV{O logiri la o serɛ noɔre ka o na yeli a zɔɔre pare ko o.} \XE{She pestered her husband to tell her the cause of the fight.} \PD{logiree, logiro, logiro, logribo, logiraa, - } 
\LX{logitogo} \HM{} \PH{lògìtògò} \PS{adv.~} \GE{deep and open (large)} \XV{Ba yage la a baa lɛ logitogo.} \XE{They have dug the pond very big.} 
\LX{logitogo} \HM{} \PH{lógítógó} \PS{adv.~} \GE{shallow and small} \XV{A kparoo zomɔ deɛ waa la lɛ logitogo.} \XE{The pocket of the shirt is very shallow.} 
\LX{lɔgɔ} \HM{} \PH{lɔ̀gɔ̀} \PS{n.~} \GE{a type of small bag in which personal documents are kept and usually carried about} \OP{lɔgre} \TP{lɔgrɛɛ} 
\LX{lɔgɔ} \HM{} \PH{lɔ́gɔ́} \PS{n.~} \GE{state of helplessness} \XV{A bie bebiri zaa baaloŋ vɛŋɛɛ ka o dɔgerebɔ lɔgɔ zaa yi.} \XE{The child's frequent illness has made his parents helpless.} \OP{lɔgere} \TP{lɔgerɛɛ} 
\LX{logoko} \HM{} \PH{lògòkò} \PS{adv.~} \GE{deep set and large} \XV{A dɔɔ nimie logoko.} \XE{The man's eyes are deep set.} 
\LX{logoko} \HM{} \PH{lógókó} \PS{adv.~} \GE{deep set and small} \XV{A bie nimie logoko.} \XE{The child's eyes are deep set.} 
\LX{lohoo} \HM{} \PH{lòhòó} \PS{n.~} \GE{a type of tree with edible fruits} \OP{lohoori} \TP{lohooree} 
\LX{lɔlɔraa} \HM{} \PH{lɔ̀lɔ̀ráà} \PS{n.~} \GE{calf (part of leg)} \OP{lɔlɔrre} \TP{lɔlɔrɛɛ} 
\LX{lɔlɔrezeɛ} \HM{} \PH{lɔ̀lɔ̀rɪ̀zɪ̀ɛ́} \PS{n.~} \GE{tree, is medicinal} \OP{lɔlɔrezeere} \TP{lɔlɔrezeerɛɛ} 
\LX{lɔmɛɛ} \HM{} \PH{lɔ̀mɛ̀ɛ̀} \PS{adv.~} \GE{fat and soft} 
\LX{lomm} \HM{} \PH{lʊ̀mm} \PS{adv.~} \GE{round and big} 
\LX{lomm} \HM{} \PH{lòmm} \PS{adv.~} \GE{balloon-like} 
\LX{lomolomo} \HM{} \PH{lómólómó} \PS{adv.~} \GE{jellylike; soft} \XV{Duoronwoɔ maŋ waa la lomolomo lɛ.} \XE{The urinary bladder is soft and jellylike.} 
\LX{lomɔlomɔ} \HM{} \PH{lʊ̀mɔ̀lʊ̀mɔ̀} \PS{adv.~} \GE{soft e.g. flesh, fruit} 
\LX{lɔŋe} \HM{} \PH{lɔ̂ŋ} \PS{v.~} \GE{to spread, to move from one person or area to another e.g. disease, fire} \PD{lɔŋɛɛ, lɔnnɔ, lɔnnɔ, lɔnema, lɔnnaa, lɔnema } \VA{lɔnne}
\LX{lɔŋe} \HM{} \PH{lɔ́ŋ} \PS{v.~} \GE{to sag, to be slack e.g. drying line} \PD{lɔŋɛɛ, lɔnnɔ, lɔnnɔ, lɔnemɔ, lɔnnaa, - } 
\LX{loŋgoroŋ} \HM{} \PH{lʊ̀ŋgʊ̀rʊ̀ŋ} \PS{adj.~} \GE{cloudy} 
\LX{loŋkoroo} \HM{} \PH{lóŋkóróò} \PS{n.~} \GE{hidden corner} \XV{A waabo zo kpɛ la a die loŋkoroo kaŋa poɔ.} \XE{The sake ran into one of the dark corners of the room.} \OP{loŋkorri} \TP{loŋkoree} 
\LX{lonlommu} \HM{} \PH{lǒnlǒmú} \PS{n.~} \GE{a type of fish that is not edible} \OP{lonlonni} \TP{lonlonnee} 
\LX{lɔnne} \HM{} \PH{lɔ́nnɪ̀} \PS{v.~} \GE{to move from one point to another e.g. on a tree} \XV{A ŋmaaŋa lɔnne la a teɛ ulli zaa kyɛ baŋ sigi.} \XE{The monkey climbed through all the branches of the tree before coming down.} \PD{lɔŋɛɛ, lɔnnɔ, lɔnnɔ, lɔnema, lɔnnaa, lɔnema } \VA{loɔre}
\LX{lonni} \HM{} \PH{lònnì} \PS{v.~} \GE{to hurry (Southern Dagaare dialect)} \PD{lonnee, lonno, lonno, lonimo, lonaa, - } 
\LX{loŋŋo} \HM{} \PH{lǒŋŋò} \PS{n.~} \GE{a type of X-shaped small drum. Usually held under the armpit and played with a stick} \OP{lonni} \TP{lonnee} 
\LX{lɔnnɔ} \HM{} \PH{lɔ̀nnɔ́} \PS{n.~} \GE{squirrel} \OP{lɔnne} \TP{lɔnnɛɛ} \VA{lanna}
\LX{lɔŋɔ} \HM{} \PH{lɔ́ŋɔ́} \PS{n.~} \GE{frog} \OP{lɔnne} \TP{lɔnnɛɛ} \VA{lɔŋa}
\LX{loɔ} \HM{} \PH{lʊ́ɔ́} \PS{n.~} \GE{the immediate front, under of one's self} \OP{loɔre} \TP{loɔrɛɛ} 
\LX{lɔɔ} \HM{} \PH{lɔ̀ɔ́} \PS{n.~} \GE{mouse} \OP{lɔɔre} \TP{lɔɔrɛɛ} \VA{lɔge}
\LX{lɔɔ} \HM{} \PH{lɔ́ɔ́} \PS{v.~} \SN \GE{to drop, to fall (from one's hands)} \SN \GE{to throw down (wrestling)} \SN \GE{to subdue, to overpower (argument)} \SN \GE{to be greater than, to be more than} \XV{A pɔgeba veɛlɛ la kyɛ a bonsɔgelaa naŋ lɔɔ ba zaa.} \XE{The women are beautiful, but the black one beats them all.} \SN \GE{to put drops into (eyes)} \SN \GE{to dash a buyer for an item bought; (tomatoes)} \PD{lɔɔɛ, lɔɔrɔ, lɔɔrɔ, lɔreba, lɔɔraa, lɔreba } \SE{lɔɔ baare} \GE{to throw away, to get rid of, to discard} 
\LX{lɔɔbiri} \HM{} \PH{lɔ́ɔ́bírí} \PS{n.~} \GE{a charm believed to be mobile and have destructive powers} \OP{lɔɔbie} \TP{lɔɔbiiree} 
\LX{lɔɔlage} \HM{} \PH{lɔ́ɔ́làgɪ̀} \PS{v.~} \GE{to surround} \PD{lɔɔlagɛɛ, lɔɔrɔlagra, lɔɔlagra, lɔɔlagreba, lɔɔlagraa, - } 
\LX{loɔraa} \HM{} \PH{lʊ́ɔ̀ráá} \PS{n.~} \GE{leopard} \OP{loɔre} \TP{loɔrɛɛ} \VA{wɛnyuo, nyuo}
\LX{loɔre} \HM{} \PH{lʊ́ɔ́rɪ́} \PS{v.~} \GE{to move from one to the other e.g. branches of a tree} \XV{A lɔnnɔ lɔnne la a ulee te ta o manzelɛ.} \XE{The squirrel moved along the branch up to the tip.} \XV{A bie dɔgereba taa la ba bibaal loɔre ne a asibitiri yaga kyɛ baŋ nyɛ sammo.} \XE{The sick child’s parents took him through several hospitals before he was cured.} \PD{loɔrɛɛ, loɔrɔ, loɔrɔ, loɔreba, loɔraa, - } 
\LX{loɔre} \HM{1} \PH{lʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to joke} \PD{loɔrɛɛ, loɔrɔ, loɔrɔ, loɔreba, loɔraa, - } 
\LX{loɔre} \HM{2} \PH{lʊ̀ɔ̀rì} \PS{v.~} \GE{to bend into a half-circle} \PD{loɔrɛɛ, loɔrɔ, loɔrɔ, loɔreba, loɔraa, - } 
\LX{lɔɔre} \HM{1} \PH{lɔ́ɔ́rɪ̀} \PS{n.~} \GE{lorry} \OP{lɔɛ} \TP{lɔrrɛɛ} 
\LX{lɔɔre} \HM{2} \PH{lɔ́ɔ́rɪ̀} \PS{n.~} \GE{items given to a buyer for free e.g pepper} \XV{lɔbɔ} \XE{} 
\LX{lɔɔ-wullo} \HM{} \PH{lɔ́ɔ-́wúlló} \PS{n.~} \GE{scribe} \OP{lɔɔ-wulibo} 
\LX{lɔre} \HM{} \PH{lɔ̀rɪ̀} \PS{v.~} \GE{to soak, to immerse} \XV{A pɔge lɔre la  muri a na wuo peɛ.} \XE{The woman has soaked straw to weave a basket.} \PD{lɔrɛɛ, lɔrɔ, lɔrɔ, lɔreba, lɔraa, lɔreba } 
\LX{lɔre} \HM{} \PH{lɔ̀rɪ̀} \PS{v.~} \GE{to be ugly} \XV{A pɔge bonsuuri la so ka o lɔre.} \XE{The woman's clothing has made her ugly.} \PD{lɔrɛɛ, lɔrɔ, lɔrɔ, lɔreba, lɔraa, lɔreba } 
\LX{lɔre} \HM{} \PH{lɔ̀rɪ̀} \PS{adj.~} \GE{weak, useless} \XV{Ka a bieŋ ba lɔre o da na tuo la a daga ŋa eŋ.} \XE{If the child had not been weak, he could, at least, have carried this box.} 
\LX{lɔre} \HM{} \PH{lɔ́rɪ̀} \PS{v.~} \GE{to be lucky} \XV{N zaameŋ ba yeli ka fo na lɔre la zenɛ, ana naane ka fo ba toɔle fo libie.} \XE{Didn't I say you would be lucky today? See, your father has sent you money.} \PD{lɔrɛɛ, lɔrɔ, lɔrɔ, lɔreba, lɔraa, lɔreba } \VA{nɔre}
\LX{lɔre} \HM{} \PH{lɔ̀rɪ̀} \PS{adv.~} \GE{onomatopoeiac, sounding like a hard object hitting a soft object} 
\LX{loregboro} \HM{} \PH{lʊ̀rɪ̀gbʊ̀rʊ̀} \PS{adv.~} \GE{crumbled} 
\LX{lɔrehazie} \HM{} \PH{lɔ́rɪ́házìé} \PS{n.~} \GE{car rental company} \OP{lɔrehaziiri} \TP{lɔrehaʊziiree} 
\LX{lori} \HM{} \PH{lòrí} \PS{n.~} \GE{pidgeon} \SG{loruu} \OP{loe} \TP{loree} 
\LX{lori} \HM{} \PH{lórì} \PS{v.~} \SN \GE{to untie} \SN \GE{to put on weight} \XV{Se᷉ɛ᷉ saŋa don-lenni maŋ ŋmeene la kyɛ ka ba wa bare a uoni ka lee, lori a zaa.} \XE{During the farming season, tethered animals are lean, but they bounce back in weight when ranging free during the dry season.} \PD{-, loro, loro, loribo, -, - } 
\LX{lori} \HM{} \PH{lórí} \PS{n.~} \GE{meteorite} \OP{loe} \TP{loree} 
\LX{lori} \HM{} \PH{lòrì} \PS{adv.~} \GE{very full (from eating)} \XV{A diibu noma la ka a bie di tege te lori o zaa.} \XE{The food was palatable and the child ate and became very full.} 
\LX{lori noɔre} \HM{} \PH{lórì nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to break a fast} \PD{-, -, -, -, -, - } 
\LX{lori pɛnne} \HM{} \PH{lórì pɛ́nnɪ̀} \PS{v.~} \GE{to tell lies} \PD{-, -, -, -, -, - } 
\LX{loroto} \HM{} \PH{lórótó} \PS{adv.~} \GE{jellylike e.g. blister on the palm} 
\LX{loroto} \HM{} \PH{lórótó} \PS{adv.~} \GE{on continuous terrain (smooth-moving vehicle)} 
\LX{lo᷉ilo᷉i} \HM{} \PH{ló᷉íló᷉í} \PS{adv.~} \GE{bright red} 
\LX{lugi} \HM{} \PH{lùgì} \PS{v.~} \GE{to induce} \PD{lugee, lugiro, lugiro, lugiribo, lugiraa, - } 
\LX{lugi} \HM{} \PH{lùgì} \PS{v.~} \GE{to swell a little e.g. armpit, groin} \PD{lugee, lugiro, lugiro, lugiribo, lugiraa, - } 
\LX{lugiri} \HM{} \PH{lúgirí} \PS{n.~} \GE{a forked beam that supports the main beam in a swish-thatched roof} \OP{lugo} \TP{lugiree} 
\LX{lugiri} \HM{} \PH{lúgirí} \PS{v.~} \GE{to loosen (a nut on a bolt), to ease} \PD{lugiree, lugiro, lugiro, lugiribo, lugiribo, lugiraa } 
\LX{luŋguri} \HM{} \PH{lùŋgúrì} \PS{n.~} \GE{lemon} \SG{luŋguruu} \OP{luŋgue} \TP{luŋgurree} 
\LX{luori} \HM{} \PH{lùòrì} \PS{v.~} \SN \GE{to become bowed e.g. legs} \XV{A bie baaloŋ eɛ la kao gbɛe luori.} \XE{The child's sickness has caused his legs to become bowed.} \SN \GE{to extend the size of a vessel} \XV{A pɔgebeɛlebɛ da luori la poore na di dooga.} \XE{The bridemaids had prepared their stomachs to feast.} \PD{luoree, luoro, luoro, luoribo, luoraa, - } 
\LX{luou} \HM{} \PH{lùòú} \PS{n.~} \GE{west} \TP{luoree} \VA{saniŋe}
\LX{luri} \HM{} \PH{lùrì} \PS{v.~} \GE{to place underneath to hide e.g. armpit} \PD{luree, luro, luro, luribo, luraa, - } 
\LX{luri} \HM{} \PH{lúrì} \PS{v.~} \GE{to be filled with e.g. smoke} \XV{A vu᷉u᷉ ba dire ka nyoore luri a korodie zaa.} \XE{The fire is not burning cleanly, so the kitchen is choked with smoke.} \PD{luree, luro, luro, luribo, luraa, - } 
\LX{luri} \HM{} \PH{lúrì} \PS{v.~} \GE{to play a gourd drum vigorously} \XV{A kɔŋŋmeɛrɛ luri la a koɔre eŋ a gyile poɔ ne toori.} \XE{The gourd drummer played the drum alongside the xylophone with artistry.} \PD{luree, luro, luro, luribo, luraa, - } 
\LX{luroluro} \HM{} \PH{lúrólúró} \PS{adj.~} \GE{in a circular shape (usually within a small enclosure)} \XV{A nɔzage waa la luroluro lɛ.} \XE{The hen coop is roudish within.} 
\LX{luroluro} \HM{} \PH{lùròlùrò} \PS{adv.~} \GE{in a circular shape (usually within a big enclosure)} \XV{A taŋa bogi e la luroluro.} \XE{The inner part of the cave is large and round.} 
\LX{lurri} \HM{} \PH{lúrrì} \PS{v.~} \GE{to smooth out a mound by covering all protruding matter with soil} \XV{Ba maŋ lurri la vubiri ka o naa ta ona ka waa to᷉o᷉ nyɛ.} \XE{A yam mound is usually enlarged by adding more soil to enable the yam bear a good tuber.} \PD{lurree, lurro, luribo, lurraa, -, - } 
\LX{luuraa} \HM{} \PH{lùùráà} \PS{n.~} \GE{denial} \OP{luuri} \TP{luuree} 
\LX{luuri} \HM{} \PH{lùùrì} \PS{v.~} \GE{to duck, to bend to pass under a hanging object e.g. beam} \PD{-, luuro, luuro, luuribo, luuraa, lure } 
\LX{luuri} \HM{} \PH{lùùrì} \PS{v.~} \SN \GE{to renounce one's word or promise, to deny} \XV{O daare yeli ka o baŋe la a nannyige kyɛ ka zenɛ ka leɛ luuri o noɔre.} \XE{S/he said s/he knew the thief, but today s/he denies that s/he said so.} \SN \GE{to return something in exchange for something else} \XV{A dɔɔ luuri la n libie ko ma kyɛ de o noɔ.} \XE{The man has returned my money and taken back his fowl.} \PD{-, luuro, luuro, luuribo, luuraa, lure } 
\LX{ma} \HM{} \PH{mǎ} \PS{n.~} \GE{mother, female that has reproduced} \OP{mamine} \TP{marɛɛ} 
\LX{ma} \HM{} \PH{mà} \PS{v. pl.~} \GE{to stick to} \XV{Goore ma la a kparoo.} \XE{Thorns are stuck to the shirt.} 
\LX{ma} \HM{} \PH{má} \PS{particle.~} \GE{suppose, hope; it is used only in question form} \XV{Ma fo na gaa la?} \XE{I suppose/hope you will go?.} 
\LX{ma} \HM{} \PH{má} \PS{pronoun.~} \GE{me (first person singular)} \XV{O de la a gane ko ma.} \XE{S/he has given the book to me.} 
\LX{maa} \HM{} \PH{máâ} \PS{pronoun.~} \SN \GE{I (first person singular personal emphatic)} \XV{O yeli ka maa la sa᷉a.} \XE{S/he said it was I who was wrong.} \SN \GE{me (first person singular personal emphatic)} \XV{Maa la so a gane.} \XE{The book belongs to me.} 
\LX{maa} \HM{} \PH{máà} \PS{v.~} \SN \GE{to be wet} \XV{Saa mie la ka a teŋɛ zaa maa.} \XE{It has rained, making the ground wet.} \SN \GE{to be cold (objects)} \XV{A baala nyu la te᷉e᷉ ka o eŋɛ leɛ maa a zaa.} \XE{The patient has taken medicine and his temperature has gone down.} \SN \GE{to make peaceful} \XV{Zenɛ a zɔɔzɔɔreba na ba wa lɛnso ka a zie maa a ŋaa.} \XE{Today, those quarrelsome men did not come and that is why the place is peaceful.} \PD{maaɛ, maana, maana, maanema, maanaa, maanema } 
\LX{maa} \HM{} \PH{màà̀} \PS{adv.~} \SN \GE{move streamingly (used only with liquid substances)} \XV{O ma naŋ da kpi o deɛ kono la nentaŋ maaa.} \XE{When her mother died, she was shedding tears uncontrolably.} \SN \GE{slimy} 
\LX{maa gbɛre} \HM{} \PH{máá gbɛ́rɪ̀} \PS{v.~} \GE{to delay, to be late} \XV{Ba naŋ boɔle o o ba maa gbɛre kyɛ wa.} \XE{When he was called, he did not waste time but responded promptly.} \PD{-, -, -, -, -, - } 
\LX{maa noɔre} \HM{} \PH{máá nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to be surprised} \PD{-, -, -, -, -, - } 
\LX{maa suuri} \HM{} \PH{máà súúrí} \PS{v.~} \GE{to appease, to pacify} \PD{-, -, -, -, -, - } 
\LX{maalam} \HM{} \PH{máálàm} \PS{n.~} \GE{Islamic cleric} \OP{maalammine} \TP{maalonnɛɛ} 
\LX{maale} \HM{} \PH{mààlɪ̀} \PS{v.~} \SN \GE{to make, to construct, to draw} \XV{A kɔnarata maale la baraakyi ŋmaa a baa.} \XE{The contractor has built a bridge across the river.} \SN \GE{to cook} \XV{Boŋ bondirii ka a bondimaala na maale zenɛ?} \XE{What food will the cook cook today?.} \PD{maalɛɛ, maala, maala, maaleba, maalaa, - } 
\LX{maale} \HM{} \PH{mààlɪ̀} \PS{v.~} \SN \GE{to bless, to make good, to clean, to make beautiful} \XV{Ba maale la a die poɔ veɛlaa.} \XE{They have cleaned the room well.} \SN \GE{to pacify, to offer sacrifice} \XV{Dagaaba mine maŋ maale la bagere.} \XE{Some Dagaaba sacrifice to their gods.} \PD{maalɛɛ, maala, maala, maaleba, maalaa, - } 
\LX{maale} \HM{} \PH{máálɪ̀} \PS{v.~} \GE{to cause to calm down, to reduce the intensity of} \XV{A yidaandɔɔ maale la a yɛlɛ ka a zɔɔre baare.} \XE{The family head calmed tempers and ended the quarrel.} \PD{-, -, -, -, -, - } 
\LX{maalekondi} \HM{} \PH{mààlɪ̀kʊ́ndí} \PS{n.~} \GE{a type of ritual food that is eaten only by specific people i.e. those who are members of the ritual} 
\LX{maaleŋ} \HM{} \PH{mààlɪ̀ŋ} \PS{adv.~} \GE{just too much} \XV{Saa la maaleŋ miire tu.} \XE{The rain just pours too much.} 
\LX{maaleŋ} \HM{} \PH{mààlɪ̀ŋ} \PS{adv.~} \GE{again} \XV{Saa la maaleŋ miire tu.} \XE{The rain just pours too much.} 
\LX{mạạli} \HM{} \PH{mạ́ạ́lì} \PS{v.~} \GE{get burnt e.g part of the body} \PD{-, -, -, -, -, - } 
\LX{maane} \HM{} \PH{màànɪ̀} \PS{adj.~} \SN \GE{flattened} \XV{A kuree feele o la ka o pare deɛ waa maane lɛ} \XE{The undersized pair of knickers has made his buttocks flat.} \SN \GE{premature (egg)} \OP{maama} \TP{maanɛɛ} 
\LX{maanee} \HM{} \PH{máánɪ́ɪ́} \PS{n.~} \GE{okra} \SG{maanoo} \OP{maana} \TP{maanɛɛ} \VA{saalommaane}
\LX{maaŋmeɛ} \HM{} \PH{máàŋmɪ́ɛ̀} \PS{n.~} \GE{a kind of tall, brown, lean bird} \OP{maaŋmeere} 
\LX{maare} \HM{} \PH{máárɪ̀} \PS{v.~} \GE{to hold, to support with open palms facing upwards e.g. baby} \PD{maarɛɛ, maara, maara, maareba, maaraa, - } 
\LX{maaroŋ} \HM{} \PH{máárʊ́ŋ} \PS{adj.~} \GE{slow} \XV{O taa la maaroŋ zuo lɛnso ka ba wa pɔge kpaaroŋ.} \XE{He is very slow and that is why he did not come to the meeting.} 
\LX{maaroŋ} \HM{} \PH{máárʊ́ŋ} \PS{adj.~} \SN \GE{cold, cool} \XV{Firigyi ko᷉ɔ᷉ maŋ e la maaroŋ.} \XE{Water stored in the fridge is always cold.} \SN \GE{wet} \XV{Ka saaŋ mi a zie zaa maŋ e la maaroŋ lɛ.} \XE{When it rains, the whole place becomes wet.} 
\LX{maaroŋ} \HM{} \PH{máárʊ́ŋ} \PS{n.~} \SN \GE{cold, coolness} \XV{Yuoni ŋa a maaroŋ zuoe la.} \XE{This year, the cold is excessive.} \SN \GE{shade} \XV{Tentoloŋ saŋa dendɔre teɛ maaroŋ maŋ tere la zemmo.} \XE{During the hot season, the shade of trees is always inviting.} 
\LX{maaroŋ} \HM{} \PH{máárʊ́ŋ} \PS{n.~} \GE{peace} \XV{Ka zieŋ na maale ferebo ka maaroŋ be be.} \XE{If development is to take place there must be peace.} 
\LX{maaroŋ} \HM{} \PH{máárʊ́ŋ} \PS{n.~} \GE{fever} \XV{Maaroŋ kpɛ la dɔɔ ka o gaa asibiti.} \XE{The man has a fever and has gone to the hospital.} 
\LX{maasa} \HM{} \PH{máásà} \PS{n.~} \GE{a type of cake made from millet flour} 
\LX{maayee} \HM{} \PH{mááyéé} \PS{adv.~} \GE{empty} \XV{Nannyigiri da ŋmaa la a woɔ pare iri a libie, ka a dɔɔ da wa tuno a woɔ ka maayee.} \XE{Thieves had cut the bottom of the bag and removed all the man's money, and when he later put his hand into the bag, it was empty.} 
\LX{mabie} \HM{} \PH{mábíé} \PS{n.~} \GE{sibling, relative} \OP{mabiiri} \TP{mabiiree} 
\LX{mabiekontapɔgema} \HM{} \PH{mábíékʊ́ntápɔ́gɪ́má} \PS{n.~} \GE{a type of small, short snake that is highly venomous} \TP{mabiekontapɔgemɛɛ} 
\LX{mabile} \HM{} \PH{màbílé} \PS{n.~} \GE{maternal aunt, mother's sister} \OP{mabilii, mabilimine} \TP{mabilee} 
\LX{mabo} \HM{} \PH{màbʊ̀} \PS{adv.~} \GE{marshy and messy} 
\LX{maee} \HM{} \PH{màɪ̀ɪ̀} \PS{adj.~} \GE{red} \XV{A pɔgelee niŋ waa la maee lɛ.} \XE{The girl's face is looking red.} 
\LX{maemae} \HM{} \PH{máɪ́máɪ́} \PS{adj.~} \GE{bright red} \XV{A fele e la zeɛ maemae lɛ.} \XE{The saw is bright red.} 
\LX{magaka} \HM{} \PH{màgakà̀} \PS{adv.~} \GE{deep red} 
\LX{mage} \HM{} \PH{mágɪ́} \PS{v.~} \SN \GE{to write} \XV{Karekyere maŋ mage la gane velaa.} \XE{Literate people write well.} \SN \GE{to make beautiful patterns; usually on some clothes} \XV{A teelɛ mage la a kparoo niŋe ne puori zaa.} \XE{The tailor has designed both the front and back of the shirt.} \PD{magɛɛ, magra, magra, magreba, magraa, - } 
\LX{magedalee} \HM{} \PH{mágɪ́dàléé} \PS{n.~} \GE{writing stick, pen} \OP{magedabilii} \TP{magedabillee} 
\LX{magelakala} \HM{} \PH{màgɪ̀làkàlà} \PS{adj.~} \GE{skinny} \XV{A dɔɔ wogi na bɛɛ la a te waa magelakala lɛ.} \XE{That tall man has been sick for a long time hence his being skinny.} 
\LX{magelaŋ} \HM{} \PH{màgɪ̀làŋ} \PS{adj.~} \GE{large and flat} \VA{maŋgalaŋ}
\LX{magere} \HM{} \PH{mágɪ́rɪ̀} \PS{n.~} \GE{sign, mark} \TP{magrɛɛ} 
\LX{magetaga} \HM{} \PH{mágɪ̀tàgà} \PS{adv.~} \GE{thick and marshy} \VA{magtaga}
\LX{makoma} \HM{} \PH{màkʊ́má} \PS{n.~} \GE{grandmother} \OP{makommine} \TP{makomɛɛ/makonnɛɛ} 
\LX{makoŋkommuulitoori} \HM{} \PH{màkʊ́ŋkʊ́mmúúlítòòrí} \PS{n.~} \GE{great-grandmother} \OP{makoŋkommuulitobo} \TP{makoŋkommuulitooree} 
\LX{makontulo} \HM{} \PH{màkʊ́ntúló} \PS{n.~} \GE{great-grandmother} 
\LX{makpe᷉ɛ᷉} \HM{} \PH{màkpɪ́᷈ɛ́᷉} \PS{n.~} \GE{stepmother; older wife of one’s biological father or father’s elder brother’s wife} \OP{makpeɛmine} \TP{makpeɛnɛɛ/makpeɛmɛɛ} 
\LX{makpoŋ} \HM{} \PH{màkpóŋ} \PS{n.~} \GE{} \VA{makpe᷉ɛ᷉}
\LX{malaa} \HM{} \PH{màláà} \PS{n.~} \GE{a type of bird with deep blue and red eyes} \OP{ malle} \TP{mallɛɛ} 
\LX{malammalam} \HM{} \PH{màlàmmálám} \PS{adj.~} \GE{soiled e.g. mud} 
\LX{malantamm} \HM{} \PH{màlàntàmm} \PS{adj.~} \GE{bundled up} 
\LX{male} \HM{} \PH{màlɪ̀} \PS{v.~} \GE{to claim or pretend to} \XV{O male ka o so la ko᷉ɔ᷉} \XE{S/he claims to have taken his/her bath.} \XV{O male ka o taa la libie kyɛ ka o bie kaŋa zaa ba gaa sakuuri.} \XE{S/he claims to be rich but none of his /her children has been to school.} \PD{malɛɛ, malla, malla, maleba, malaa, maleba } 
\LX{male} \HM{} \PH{màlɪ̀} \PS{v.~} \GE{to suffer repercussions, to punish} \XV{Fomeŋɛ ka na male.} \XE{You will only end up punishing yourself.} \XV{Ka fo saaŋ yeli ka fo gaa sakuuri ka fo zagere fomeŋɛ ka male.} \XE{If your father sends you to school and you refuse you only punish yourself.} \PD{malɛɛ, malla, malla, maleba, malaa, maleba } 
\LX{male} \HM{} \PH{màlì} \PS{adv.~} \GE{anomatopoeic word expressing the way the soud of a cut} 
\LX{malefa} \HM{} \PH{málɪ́fà} \PS{n.~} \GE{gun} \OP{malefare} \TP{malefarɛɛ} 
\LX{malefabiri} \HM{} \PH{málɪ́fàbírì} \PS{n.~} \GE{bullet} \SG{malefabiruu} \OP{malefabie} \TP{malefabiiree} 
\LX{malefala} \HM{} \PH{màlɪ̀fàlà} \PS{adv.~} \GE{tasteless} 
\LX{malefate᷉e} \HM{} \PH{málɪ́fàtɪ̀᷉ɪ́᷉} \PS{n.~} \GE{gunpowder} \OP{malefateenɛɛ} 
\LX{malemale} \HM{} \PH{málɪ́málɪ́} \PS{adv.~} \GE{very black} \VA{kpirikpiri}
\LX{malemale} \HM{} \PH{màlɪ̀málɪ́} \PS{v.~} \GE{to struggle to do} \PD{malemalɛɛ, malemala, malemala, malemaleba, malemalaa, - } 
\LX{mana} \HM{} \PH{máná} \PS{n.~} \GE{rubber, elastic} \SG{manoo} \OP{manne} \TP{mannɛɛ} 
\LX{mana (Hausa)} \HM{} \PH{mànà} \PS{particle.~} \GE{of course} \XV{kɔlaa la kpɛ a die ona la ɔɔ a nɛne mana.} \XE{A cat went into the room; of course it is the one that has eaten the meat.} 
\LX{manamana} \HM{} \PH{mànàmànà} \PS{adj.~} \GE{sticky} 
\LX{mane} \HM{} \PH{mànɪ́} \PS{n.~} \GE{large body of water} \OP{mama} \TP{mannɛɛ} 
\LX{maŋgalaŋ} \HM{} \PH{maŋgalaŋ} \PS{adj.~} \GE{flattened} \VA{magelaŋ}
\LX{maŋgaŋene} \HM{} \PH{màŋgáŋɪ́nɪ́} \PS{n.~} \GE{the other side of the sea; overseas regions} \TP{maŋgaŋenɛɛ} 
\LX{maŋgbɛre} \HM{} \PH{máŋgbɛ́rɪ́} \PS{n.~} \GE{tributary (stream)} \OP{maŋgbɛɛ} \TP{maŋgbɛɛrɛɛ} 
\LX{maŋkeni} \HM{} \PH{mâŋkèní} \PS{n.~} \GE{cocoyam} \SG{maŋkenuu} \OP{maŋkeme} \TP{maŋkenee/maŋkennee} 
\LX{mankyɛlɛ} \HM{} \PH{mánʧɛ̀lɛ̀} \PS{n.~} \GE{diarrhoea} 
\LX{mankyere} \HM{} \PH{mánʧɪ̀rɪ̀} \PS{n.~} \GE{match} \OP{ mankyeɛ} \TP{mankyerɛɛ} 
\LX{manne} \HM{} \PH{mánnɪ̀} \PS{v.~} \SN \GE{to measure} \XV{Kookoolaa ka o manne ne a kyi.} \XE{They used a measuring bowl to measure the millet.} \XV{Ba de la kɛkɛre manne ne a dendɔre wogiruŋ.} \XE{They used a stalk to measure the height of the door.} \SN \GE{to compare} \XV{Sakubie koŋ tõɔ de omeŋɛ te manne ne karema.} \XE{A pupil cannot compare herself to a teacher.} \PD{mannɛɛ, manna, manna, manema, mannaa, - } 
\LX{manne} \HM{} \PH{mánnɪ̀} \PS{v.~} \GE{to narrate, to explain} \XV{O manne la a zɔɔre pare.} \XE{S/he explained the cause of the fight.} \XV{N naŋ da gaa n zɔmeŋɛ dɔgoo bebiri tigiri N da kuli te manne la a zaa ko n dɔgereba.} \XE{After attending my best friend’s birthday party, I went home and narrated everything to my parents.} \PD{mannɛɛ, manna, manna, manema, mannaa, - } 
\LX{mannoɔre} \HM{} \PH{mànnʊ́ɔ́r} \PS{n.~} \GE{beach} \OP{mannɔɛ} \TP{mannɔrɛɛ} 
\LX{mansugiri} \HM{} \PH{mànsùgirì} \PS{n.~} \GE{a type of wild mushroom with medicinal properties} \SG{mansugruu} \OP{mansugo} \TP{mansugree} 
\LX{mansugo} \HM{} \PH{mànsùgò} \PS{n.~} \SN \GE{a type of stomach ailment; hernia} \SN \GE{type of medicine for hernia} \OP{mansugiri} \TP{mansugiree} 
\LX{mantare} \HM{} \PH{màntàrɪ̀} \PS{n.~} \SN \GE{Southern Dagaare dialect} \SN \GE{the area where the Southern Dagaare dialect is spoken} \OP{mantarɛɛ} 
\LX{mantaroo} \HM{} \PH{màntàrʊ́ʊ́} \PS{n.~} \GE{speaker of Southern Dagaare dialect} \OP{mantarɛɛ} 
\LX{manteɛ} \HM{} \PH{mǎntɪ̀ɛ́} \PS{n.~} \GE{rubber tree} \OP{manteere} \TP{manteerɛɛ} 
\LX{manzelɛ} \HM{} \PH{mànzɪ́lɛ́} \PS{n.~} \GE{tip of (branch of tree)} \OP{manzelle} \TP{manzelɛɛ} 
\LX{mapo} \HM{} \PH{mápʊ̀} \PS{n.~} \GE{map} \OP{mapore} \TP{maporɛɛ} 
\LX{mara} \HM{} \PH{màrà} \PS{n.~} \GE{courage, persistence, steadfastness} \TP{marɛɛ} 
\LX{mara} \HM{} \PH{màrà} \PS{adv.~} \GE{onomatopoeia, sounding like the sound coming from one stepping into muddy matter} 
\LX{maraa} \HM{} \PH{máráá} \PS{n.~} \GE{necklace} \OP{marre} \TP{marrɛɛ} 
\LX{maraa} \HM{} \PH{máráá} \PS{n.~} \GE{glue} \OP{marre} \TP{marrɛɛ} 
\LX{maraa} \HM{} \PH{máráá} \PS{n.~} \GE{cloth used to carry a baby on the back} \OP{marre} \TP{marrɛɛ} 
\LX{maraa} \HM{} \PH{màràà} \PS{adj.~} \GE{not responding to anything, quiet, mute} \XV{A pɔge suuri naŋ iri na fo maŋ yeli la yɛlɛ ko o ka e maraa.} \XE{The woman is annoyed so when you talk to her she stays mute.} 
\LX{maraawa} \HM{} \PH{màrááwà} \PS{n.~} \GE{a type of tree animal} \OP{maraaware} \TP{maraawarɛɛ} 
\LX{maramara} \HM{} \PH{màràmàrà} \PS{adj.~} \GE{marshy} \XV{Saa mie la ko᷉ɔ laare a tɛne ka a waa maramara lɛ.} \XE{The soil has been beaten by rain and has become marshy.} 
\LX{marata} \HM{} \PH{màràtà} \PS{adv.~} \GE{refusing to respond to anything} 
\LX{mare} \HM{} \PH{màrɪ̀} \PS{v.~} \GE{to stick to, to attach to} \XV{Dunduli mare la a vaa poɔ.} \XE{There is a caterpillar hanging on the leaf.} \PD{marɛɛ, mara, mara, mareba, maraa, - } 
\LX{mare} \HM{} \PH{màrɪ̀} \PS{v.~} \GE{to carry on one's back e.g. a baby} \XV{A pɔge mare la a bie ka na gbiri.} \XE{The woman has carried the baby on her back in order to make it sleep.} \PD{marɛɛ, mara, mara, mareba, maraa, - } 
\LX{mare} \HM{} \PH{màrɪ̀} \PS{v.~} \GE{to light, to build a fire} \XV{Te maŋ mare vu᷉u᷉ maale seemaa.} \XE{We build a fire before we cook food.} \PD{marɛɛ, mara, mara, mareba, maraa, - } 
\LX{mare} \HM{} \PH{márɪ́} \PS{v.~} \GE{to detatch (a limb from the body)} \PD{marɛɛ, mara, mara, mareba, maraa, - } 
\LX{marekpɛ} \HM{} \PH{màrɪ̀kpɛ̀} \PS{v.~} \GE{to tease} \XV{Koŋkombie mine maŋ marekpɛ la a kotuo deme.} \XE{Some lyrics of funeral dirges tease the bereaved family.} \PD{-, -, -, -, -, - } 
\LX{mɛ} \HM{} \PH{mɛ̀} \PS{v.~} \SN \GE{to build e.g. a house} \SN \GE{to shape (pottery)} \PD{mɛɛ, meɛrɛ, meɛrɛ, meɛrebɛ, -, - } 
\LX{meɛ} \HM{} \PH{mɪ́ɛ̀} \PS{v.~} \GE{to incriminate, to make an accomplice} \XV{A bibarewono na zu la o ba libie kyɛ de meɛ ne o yɔɔlee.} \XE{That naughty boy stole his father's money and incriminated his younger brother.} \PD{meɛɛ, meɛnɛ, meɛnɛ, meɛnemɛ, meɛnaa, - } 
\LX{mɛɛ} \HM{} \PH{mɛ̀ɛ̀} \PS{adv.~} \GE{oozing (oily liquid)} \XV{Ba kyere la a gbɛre ka ko᷉o᷉ zoro mɛɛ.} \XE{They cut the swelling on his leg and pus was oozing.} 
\LX{meɛbo} \HM{} \PH{mɪ́ɛ́bʊ́} \PS{n.~} \SN \GE{structure} \XV{A pɔge meɛbo wuli ka o taa la seɛ yelmeŋɛ.} \XE{The woman’s structure shows she has a really large waist.} \SN \GE{character; a distinguishing feature} \XV{A pɔlebile na meɛbo ba veɛlɛ.} \XE{That young man's character is not good.} 
\LX{meele} \HM{} \PH{mɪ̀ɪ̀lɪ̀} \PS{v.~} \SN \GE{to wipe off (dirt, water)} \SN \GE{to soil, to make dirty} \PD{meelɛɛ, meelɛ, meelɛ, meelebɛ, -, - } 
\LX{meele} \HM{} \PH{mɪ́ɪ́lɪ́} \PS{v.~} \GE{to try to incriminate someone; to challenge} \XV{O sɔgele gaa toŋ la a toma o yoŋ, o deɛ boɔrɔ ka o meele ma.} \XE{S/he did the work alone behind me; I think this is aimed tarnishing my image.} \PD{meelɛɛ, meelɛ, meelɛ, meelebɛ, meelaa, - } 
\LX{meelɛ} \HM{} \PH{mɪ̀ɪ́lɛ̀} \PS{n.~} \SN \GE{a type of grass that is thin but strong} \SN \GE{decorative ribbons worn across the chest} \SN \GE{string scale; a locally made scale made from a string and a flat piece of wood or container} \TP{meelɛɛ} 
\LX{meele eŋ} \HM{} \PH{mɪ̀ɪ̀lɪ̀ ɪ̀ŋ} \PS{v.~} \GE{to turn something to replace something else} \XV{Waabo doŋ la a nɔnyaŋaa ko ka ba nyɔge nɔdaa meele eŋ a nɔbilii.} \XE{The mother fowl was killed by a snake and so a cock was tamed to replace it.} \PD{-, -, -, -, -, - } 
\LX{meel-eŋ} \HM{} \PH{mɪ̀ɪ̀l-ɪ̀ŋ} \PS{n.~} \GE{metaphor (literature)} \XV{A bie zoobu wuli ka soɔŋaa la.} \XE{The child's speed makes him a rabbit.} 
\LX{meɛloŋ} \HM{} \PH{mɪ̀ɛ̀lʊ́ŋ} \PS{n.~} \GE{dew; drops of water on vegetation} \OP{meɛlɛ} \TP{meɛlɛɛ} 
\LX{meeloo} \HM{} \PH{mɪ́ɪ́lʊ́ʊ́} \PS{adj.~} \GE{thin} \OP{meelɛ} 
\LX{meemee} \HM{} \PH{mɪ̀ɪ̀mɪ́ɪ́} \PS{v.~} \GE{to swing, to turn round (buttocks in a chair)} \XV{A bie meemee la pare ka o koŋ gaa.} \XE{The child swung his bottom, indicating that he would not go.} \PD{meemeeɛ, meemeenɛ, meemeenɛ, meemeenemɛ, meemeenaa, - } 
\LX{meɛroŋ} \HM{} \PH{mɪ̀ɛ́rʊ̂ŋ} \PS{n.~} \SN \GE{structure, makeup} \SN \GE{character, behaviour} 
\LX{mɛfaa} \HM{} \PH{mɛ́fáá} \PS{n.~} \GE{bad character} \OP{mɛfaare} \TP{mɛfaarɛɛ} 
\LX{mɛgelɛŋ} \HM{} \PH{mɛ́gɪ́lɛ́ŋ} \PS{adv.~} \GE{flat} \XV{Lɔɔre do la a bogiti ka o e ŋmɛgelɛŋ lɛ.} \XE{A car ran over the bucket and made it flat.} 
\LX{megemege} \HM{} \PH{mìgɪ̀mɪ́gɪ́} \PS{v.~} \GE{to wriggle e.g. fish in muddy water} \PD{megemegɛɛ, megemegrɛ, megemegrɛ, megemegrebɛ, megemegraa, - } 
\LX{mɛgeyɛgɛ} \HM{} \PH{mɛ́gɪ́yɛ́gɛ́} \PS{adj.~} \GE{loose and flabby (of a fat body)} 
\LX{melaa} \HM{1} \PH{mɪ́láá} \PS{n.~} \GE{fence} \OP{melle} \TP{mellɛɛ} 
\LX{melaa} \HM{2} \PH{mɪ́láá} \PS{n.~} \GE{talebearer} \OP{melebɛ} \TP{melebɛɛ} 
\LX{mele} \HM{} \PH{mɪ́lɪ́} \PS{v.~} \SN \GE{to fence in} \XV{Ba maŋ mele la dabɔgɔ ka donne na ta sa᷉a᷉ o poɔ bombulo.} \XE{Gardens are fenced to prevent animals from destroying the plants.} \SN \GE{to entangle} \XV{Ba ba leŋ a boɔ lɛnso ka o mele kpi.} \XE{The goat was not tethered well and so it got entangled and died.} \SN \GE{to spread all over an object (ants)} \XV{Zuluŋ mele la boku᷉u᷉.} \XE{Soldier ants are all over the carcas.} \SN \GE{to tattle, to gossip} \XV{Bibiiri nɔŋ la mele taa.} \XE{Children like tattling on each other.} \PD{melɛɛ, melɛ, melɛ, melbɛ, melaa, - } 
\LX{mɛle} \HM{} \PH{mɛ́lɪ̀} \PS{n.~} \GE{medal} \OP{mɛlɛ} \TP{mɛllɛɛ} 
\LX{mɛlefɛlɛ} \HM{} \PH{mɛ̀lɪ̀fɛ̀lɛ̀} \PS{adj.~} \GE{tasteless} \XV{N nɔŋ la zɛvaare zeɛre kyɛ ama deɛ waa la mɛlefɛlɛ lɛ.} \XE{I like vegetable soup but this one is tasteless.} 
\LX{melekala} \HM{} \PH{mɪ̀lɪ̀kàlà} \PS{adj.~} \GE{large and twisted} 
\LX{melekala} \HM{} \PH{mɪ́lɪ́kálá} \PS{adj.~} \GE{small and twisted} 
\LX{melele} \HM{} \PH{mɪ́lɪ́lɪ́} \PS{adv.~} \GE{thin, very slim} 
\LX{melemele} \HM{} \PH{mɪ̀lɪ̀mɪ́lɪ́} \PS{v.~} \GE{to move with unsteady legs} \PD{melemelɛɛ, melemelɛ, melemelɛ, melemelbɛ, melemelaa, - } 
\LX{mɛlemɛle} \HM{} \PH{mɛ̀lɪ́mɛ̀lɪ̀} \PS{adj.~} \GE{blowing severely with cold} \XV{Saseɛ saŋa baguo ɔɔre maŋ erɛ la mɛlemɛle lɛ, Dagao poɔ.} \XE{During hamattan, early mornings are normally very cold in the Dagaabaland.} 
\LX{mɛlɛmɛlɛ} \HM{} \PH{mɛ́lɛ́mɛ́lɛ́} \PS{adj.~} \GE{tasty; slightly sweet} 
\LX{mɛlɛmmɛlɛŋ} \HM{} \PH{mɛ́lɛ́mmɛ́lɛ́ŋ} \PS{adj.~} \GE{sweet or beautiful (kind words, deeds)} \XV{O yeli la a yɛlɛ mɛlɛmmɛlɛŋ lɛ, ka fo na teɛre ka yelimeŋɛ la.} \XE{She narrated the case so beautifully that you would think it was true.} 
\LX{mɛlɛntɛmm} \HM{} \PH{mɛ́lɛ́ntɛ́mm} \PS{adj.~} \SN \GE{exactly; precisely} \XV{A bie manne la a yɛlɛ na o naŋ nyɛ mɛlɛntɛmm wa gaa.} \XE{The boy narrated the story in a precise manner.} \SN \GE{meticulous} 
\LX{meletaa} \HM{} \PH{mɪ́lɪ́táá} \PS{n.~} \GE{the practice of talebearing} \XV{Ka noba laŋkpeɛboŋ na noma a seŋ ka ba bare meletaa.} \XE{If people want to live in harmony, they must put an end to talebearing.} 
\LX{melle} \HM{} \PH{mìllɪ̀} \PS{v.~} \GE{to twist} \XV{Soɔ da kyebe kyɛ a dɔɔ da de la o nu melle ne a miri ŋmaa.} \XE{There was no knife, but the man twisted the rope until it snapped.} \PD{mellɛɛ, mellɛ, mellɛ, melbɛ, mellaa, - } 
\LX{meloo} \HM{} \PH{mélóó} \PS{adv.~} \GE{eating instantly and hurriedly; especially with tasty food} \XV{O nɔnoɔ zuoe la, a be᷉e᷉ ka o de e meloo.} \XE{S/he has a taste for delicious things, so s/he licked the broth at once.} 
\LX{mɛmeɛrɛ} \HM{} \PH{mɛ̀mɪ́ɛ́rɛ́} \PS{n.~} \SN \GE{potter} \SN \GE{builder} \OP{mɛmeɛrebɛ} \TP{mɛmeɛrebɛɛ} 
\LX{mɛmɛrɛɛ} \HM{} \PH{mɛ̀mɛ́rɛ́ɛ́} \PS{n.~} \SN \GE{mud} \SN \GE{marsh} \TP{mɛmɛrrɛɛ} 
\LX{meŋ} \HM{} \PH{mɪ̌ŋ} \PS{adv.~} \GE{also, too} \XV{Fo meŋ na gaa la be?} \XE{Will you go too?.} 
\LX{menɛ} \HM{} \PH{mɪ́nɛ́} \PS{n.~} \GE{a type of thin grass} \TP{menɛɛ} 
\LX{meŋɛ} \HM{} \PH{mɪ́ŋɛ́} \PS{n.~} \GE{self} \XV{Maa meŋɛ la na gaa a kuori.} \XE{I myself will attend the funeral.} \OP{menne} \VA{meŋa}
\LX{meŋɛ} \HM{} \PH{mɪ́ŋɛ́} \PS{adj.~} \GE{well behaved} \XV{O teɛre ka o waa la meŋɛ zaa kyɛ maa ba teɛre.} \XE{S/he claims to be good, but I doubt it.} 
\LX{menekana} \HM{} \PH{mɪ̀nɪ̀kànà} \PS{adj.~} \GE{crooked, twisted} 
\LX{menne} \HM{} \PH{mɪ́nnɪ́} \PS{v.~} \GE{to throw off something heavy e.g. wrestling} \XV{A dɔɔkpeɛŋaa na are la dendeŋ kyɛ menne a o tɔ na bare ka o te le.} \XE{The strong man heaved and threw his opponent away.} \PD{mennɛɛ, mennɛ, mennɛ, menemɛ, mennaa, - } 
\LX{menne} \HM{} \PH{mɪ́nnɪ́} \PS{v.~} \GE{to swallow food in large chunks} \XV{Kɔŋ da kpɛ o la te zuo ka o maŋ deɛ ŋmaa a saabo menne bare.} \XE{The man was so hungry that he swallowed the food in chunks.} \PD{mennɛɛ, mennɛ, mennɛ, menemɛ, mennaa, - } 
\LX{mennoŋ} \HM{} \PH{mɪ́nnʊ́ŋ} \PS{n.~} \GE{righteousness} \XV{O mennoŋ zuiŋ noba zaa nɔŋ la o yɛlɛ.} \XE{All people like him because of his righteousness.} 
\LX{menselɛ} \HM{} \PH{mɪ́nsɪ́lɛ́} \PS{n.~} \GE{well done (praise)} \XV{Gandaa menselɛ ka fooŋ da ba peɛle a bie kpie.} \XE{Well done; if you had not been around, the child would have died.} \OP{menselle} \TP{mensellɛɛ} \VA{nenselɛ}
\LX{mere} \HM{} \PH{mɪ́rɪ́} \PS{v.~} \GE{to shake in pain or distress e.g. hand in contact with fire} \XV{Vu᷉u nyege o la ka o mere nu.} \XE{Fire burned her hand and she shook it.} \PD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\LX{mere} \HM{} \PH{mɪ́rɪ́} \PS{v.~} \GE{to broadcast e.g rice seed during sowing} \XV{Ba kyɔre la baa mere a mui.} \XE{They have harrowed the river valley and broadcast the rice.} \PD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\LX{mere} \HM{} \PH{mɪ́rɪ́} \PS{v.~} \GE{to have a stillbirth (nonhumans)} \XV{A naapoɔ na da bɛɛ la lɛnso ka o mere bare.} \XE{That pregnant cow was sick and that is why it had a stillbirth.} \PD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\LX{mere} \HM{} \PH{mɪ̀rɪ̀} \PS{v.~} \GE{to stick the flap of an object into an opening} \XV{Pɔgeba maŋ mere la ba wagyɛ eŋ kyɛ zeŋ.} \XE{Women usually push the flaps of their cloth in between their legs as they sit.} \PD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\LX{mɛre} \HM{} \PH{mɛ́rɪ́} \PS{v.~} \GE{to cook (struggling with)} \XV{A baala mɛre la seemaa fe᷉e᷉ meŋ di.} \XE{The patient has cooked some small food to eat.} \PD{mɛrɛɛ, mɛrɛ, mɛrɛ, mɛrebɛ, mɛraa, - } 
\LX{mere bare} \HM{} \PH{mɪ́rɪ́ bàrɪ̀} \PS{v.~} \SN \GE{to have a stillbirth (of animals)} \SN \GE{to desert unweaned young (of animals)} \XV{A peroo dɔgɛɛ la kyɛ mere bare ka a bilii kpi.} \XE{The sheep littered but deserted the lambs and they died.} \PD{mere barɛɛ, merɛbara, merebara, merebareba, merebaraa, - } 
\LX{mere pare} \HM{} \PH{mɪ̀rɪ̀ párɪ̀} \PS{v.~} \SN \GE{to lie helpless} \SN \GE{to die} \PD{merɛɛ pare, merɛ pare, pare merɛ, pare merebɛ, -, - } 
\LX{mɛrɛɛ} \HM{} \PH{mɛ̀rɛ́ɛ̀} \PS{n.~} \GE{pus} \OP{mɛrre} \TP{mɛrrɛɛ} 
\LX{merete} \HM{} \PH{mèrètè} \PS{adj.~} \GE{unresponsive, mute} \XV{Ba yeli la a bie baaloŋ ko o saa ka o e merete.} \XE{When they reported the child's mother, he refused to do anything.} 
\LX{meri} \HM{} \PH{mérì} \PS{v.~} \GE{to twist one's lips (to show indifference); to refuse to care} \XV{A bie lee la ka leɛ kaa kyɛ meri o noɔre.} \XE{When the child fell, the brother looked at him but refused to bother.} \PD{meree, mere, mere, meribe, meraa, - } 
\LX{meri noɔre} \HM{} \PH{mérì nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to twist one's lips (to show indifference)} \PD{-, -, -, -, -, - } 
\LX{meri noɔre} \HM{} \PH{mérì nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to die} \PD{-, -, -, -, -, - } 
\LX{meruu} \HM{} \PH{mèrúú} \PS{adj.~} \GE{very small} 
\LX{mi} \HM{} \PH{mí} \PS{v.~} \GE{to rain} \PD{mie, miire, -, -, -, - } 
\LX{mi} \HM{} \PH{mí} \PS{v.~} \GE{to spin (thread)} \XV{A zɔŋɔ mi la miri ka o gbɛkpoŋ te zogere.} \XE{The blind person spun rope until her thigh became rough.} \PD{mie, miire, -, -, -, - } 
\LX{mii} \HM{} \PH{míì} \PS{v.~} \SN \GE{to shake, to shiver} \XV{Entole kpɛ la a bie lɛnso ka o maŋ mii a lɛ.} \XE{The child has a fever and that is why he is shivering like that.} \SN \GE{to become frightened, to tremble} \XV{A sogya tanne la a pɔge ka o mii.} \XE{The soldier shouted at the woman and she became fightened.} \SN \GE{to push, to shake e.g. the branch of a tree to knock down fruits} \XV{A pɔge do mii la a taŋaa ko o bibiiri ka ba di.} \XE{The woman climbed and shook the shea nut tree to knock down the fruits for her children.} \PD{miie, miine, miine, miinime, miinaa, - } \VA{miiu}
\LX{mii} \HM{} \PH{mìì} \PS{v.~} \SN \GE{to be sour (taste)} \XV{O za᷉a᷉ tuŋ la nu a zeɛre poɔ lɛnso ka a mii.} \XE{S/he put her hand into the soup and that is why it has become sour.} \SN \GE{to be sore} \XV{A bie za᷉a᷉ tɔ o nubiri, zenɛ o zaaŋ mii.} \XE{The child bruised his finger yesterday and today it is quite sore.} \SN \GE{to be angry, to be annoyed} \XV{Ba to o la lɛnso ka o mii.} \XE{They insulted him and that is why he is angry.} \SN \GE{to learn from past unpleasant experience} \XV{A tontona naŋ zagere toma gaabo na ba ba yɔɔ o, azuiŋ o miie la; zenɛ o ba nagera toma.} \XE{Since the worker refused to go to work, he was not paid, and as a result, he did not skip work today.} \PD{miie, miine, miine, miinime, miinaa, miinime } 
\LX{miii} \HM{} \PH{mììì} \PS{adv.~} \GE{swarming all over e.g. ants} \XV{A nyere deɛ da te paalɛɛ a korodie zaa, miloo lɛ.} \XE{The ants had invaded and covered the whole kitchen.} 
\LX{miire} \HM{1} \PH{mììré} \PS{v.~} \GE{to desist, to not dare} \XV{Ma fo to ma la, aŋ miire n yiri waabo.} \XE{Since you have insulted me, you dare not come to my house.} \PD{-, -, -, -, -, - } 
\LX{miire} \HM{2} \PH{mììré} \PS{v.~} \GE{to be careful; usually to warn or caution} \PD{miiree, miire, miire, miiribe, miiraa, - } 
\LX{miiri} \HM{} \PH{mììrì} \PS{v.~} \GE{to not attempt} \PD{miiree, miire, miire, miiribe, miiraa, - } 
\LX{miiri} \HM{} \PH{mììrì} \PS{v.~} \GE{to turn taste sour} \PD{miiree, miire, miire, miiribe, miiraa, - } 
\LX{miiruŋ} \HM{} \PH{mììrùŋ} \PS{n.~} \SN \GE{sour sap} \SN \GE{type of plant with sour sap} 
\LX{miiruŋ} \HM{} \PH{mììrùŋ} \PS{adj.~} \GE{sour} 
\LX{mikalaa} \HM{} \PH{mǐkàláà} \PS{n.~} \GE{thick, strong rope} \OP{mikalle} \TP{mikallɛɛ} 
\LX{milemmilem} \HM{} \PH{mílémílém} \PS{adv.~} \GE{intact, untouched} 
\LX{miliki} \HM{} \PH{mílíkì} \PS{n.~} \GE{milk} \SG{milikiruu} \OP{milikiri} \TP{milikiree} 
\LX{milikolo} \HM{} \PH{mìlìkòlò} \PS{adj.~} \GE{twisted} 
\LX{milli} \HM{} \PH{mìllì} \PS{v.~} \SN \GE{to get acquainted with; to be used to} \XV{Bie naŋ ba milli neɛ koŋ sage ka a soba de o.} \XE{A child that is not acquainted with someone will refuse to be cuddled by that person.} \SN \GE{all over} \PD{millee, mille, mille, milibe, millaa, - } \VA{milli pɔge}
\LX{milli pɔge} \HM{} \PH{mìllì-pɔ̀gɪ̀} \PS{v.~} \GE{to cover all over e.g. flies over a rotten carcass} \XV{Naazoore milli pɔge la a nɛmpoɔnaa zaa.} \XE{The rotten meat is completely covered with flies.} \PD{-, -, -, -, -, - } 
\LX{miloo} \HM{} \PH{mìlòò} \PS{adj.~} \GE{oily and dirty looking} 
\LX{miloo} \HM{} \PH{mìlòò} \PS{adj.~} \GE{existing in great numbers at a particular spot e.g. ants} 
\LX{mine} \HM{} \PH{mìné} \PS{adj.~} \SN \GE{some} \XV{A boore mine zoe la.} \XE{Some of the goats have escaped.} \SN \GE{pl. morpheme ( human type only)} \XV{Mamine la maŋ ko  biiri bere ka ba ɛnne.} \XE{Mothers provide their breasts to babies to suck.} 
\LX{minimini} \HM{} \PH{mìnìmìnì} \PS{adv.~} \GE{streaming continuously e.g. tears} \XV{A bie poɔ beɛrɛ la ka o kono minimini lɛ.} \XE{The child has a stomach ache and he is sheding tears uncontrollably.} 
\LX{mipelaa} \HM{} \PH{mípɪ́láá} \PS{n.~} \GE{a type of rope used in rituals} \OP{mipeɛle} 
\LX{miri} \HM{} \PH{mírì} \PS{n.~} \GE{rope, twine, string} \OP{mie} \TP{miiree} 
\LX{misiri} \HM{} \PH{mìsírì} \PS{n.~} \GE{mosque} \OP{misirri} \TP{misirree} \VA{musiri}
\LX{miti} \HM{} \PH{mítì} \PS{n.~} \GE{meeting} \XV{mitiri} \XE{mitiree.} \VA{English}
\LX{mizeɛ} \HM{} \PH{mízɪ̀ɛ́} \PS{n.~} \GE{type of very strong red twine used in making bird trap} \OP{mizeere} \TP{mizeerɛɛ} 
\LX{mmeŋɛ} \HM{} \PH{mmɪ́ŋɛ́} \PS{pronoun.~} \GE{myself} \XV{Ka maaŋ beɛrɛ kyɛ ba gaa asibiti mmeŋɛ la ka N erɛ.} \XE{If I'm sick and refuse to go the hospital, I have only myself to blame.} \VA{mmeŋa}
\LX{mɔ} \HM{} \PH{mɔ́} \PS{v.~} \SN \GE{to wrestle} \XV{A pɔgebɔ da zɔɔ la taa, a mɔ te le.} \XE{The women fought, wrestled, and fell.} \SN \GE{to try hard} \XV{Ka fooŋ ba mɔ fo koŋ to᷉ɔ᷉ a teɛ do.} \XE{If you don't try hard, you can't climb the tree.} \PD{mɔɛ, moɔrɔ, moɔrɔ, moɔreba, moɔraa, - } 
\LX{mɔ-aŋgaara} \HM{} \PH{mɔ̀-áŋgáárá} \PS{n.~} \GE{half-burnt grass} \OP{mɔ-aŋgaare} \TP{mɔ-aŋgaarɛɛ} 
\LX{mɔdoŋɔ} \HM{} \PH{mɔ̀dʊ́ŋɔ́} \PS{n.~} \GE{wild animal} \OP{mɔdonne} \TP{mɔdonnɛɛ} \VA{mɔdoŋa, wɛdoŋɔ}
\LX{mɔgaabo} \HM{} \PH{mɔ̀gáábʊ̀} \PS{n.~} \GE{game hunting expedition from a pitched tent over a length of time} 
\LX{moge} \HM{} \PH{mʊ́gɪ̀} \PS{v.~} \SN \GE{to suck e.g. sweets} \SN \GE{to eat powdery food} \SN \GE{to eat up e.g. wood, of white ants} \PD{mogɛɛ, mogrɔ/mogra, mogerɔ/mogra, mogreba, mograa, - } 
\LX{mɔge} \HM{} \PH{mɔ́gɪ́} \PS{v.~} \GE{to get dirty} \XV{Ka fooŋ kpɛ nɔzage fo nɔɔteɛ maŋ mɔge la nɔbini.} \XE{If you go into a hen coop, your shoes will get dirty with fowl droppings.} \PD{mɔgɛɛ, mɔgrɔ, mɔgrɔ, mɔgreba, -, - } 
\LX{mɔgele} \HM{} \PH{mɔ̀gɪ̀lɪ̀} \PS{v.~} \SN \GE{to beseech, to supplicate} \SN \GE{to persuade} \SN \GE{to console, to comfort e.g. one in pain or distress} \SN \GE{to beg for forgiveness} \PD{mɔgelɛɛ, mɔgelɔ, mɔgelɔ, mɔgeleba, mɔgelaa, - } 
\LX{mɔgeraa} \HM{} \PH{mɔ̀gɪ̀ráá} \PS{n.~} \GE{a type of giant white ants} \OP{mɔgere} \TP{mɔgerɛɛ} 
\LX{mogere} \HM{} \PH{mʊ̀gɪ̀rɪ̀} \PS{v.~} \GE{to suck at, to lick e.g. fingers} \PD{mogerɛɛ, mogera, mogera, mogereba, mogeraa, mogereba } \VA{moore}
\LX{mogere noɔre} \HM{} \PH{mʊ̀gɪ̀rɪ̀ nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to kiss (on the mouth)} \PD{-, -, -, -, -, - } 
\LX{mogi} \HM{} \PH{mógí} \PS{v.~} \GE{to eat delicious food voraciously} \PD{mogee, mogiro, mogiro, mogiribo, mogiraa, - } \VA{yogi}
\LX{moimoi} \HM{} \PH{móímóí} \PS{adj.~} \GE{very red} 
\LX{mɔkangyi} \HM{} \PH{mɔ́kándzí} \PS{n.~} \SN \GE{a type of tree with medicial properties and edible fruits} \SN \GE{pawpaw} \OP{mɔkangyiri} \TP{mɔkangyiree} 
\LX{mokanzie} \HM{} \PH{mókánzíé} \PS{n.~} \GE{pawpaw} \OP{mokanziri} \TP{mokanziree} 
\LX{mɔkpeɛrɛ} \HM{} \PH{mɔ̀kpɪ́ɛ́rɛ́} \PS{n.~} \SN \GE{bush dweller} \SN \GE{one who lives in a faraway country} \OP{mɔkpeɛrebɛ} \TP{mɔkpeɛrebɛɛ} 
\LX{mole} \HM{} \PH{mʊ̀lɪ̀} \PS{n.~} \GE{become weak; used for metal that has become weak e.g. iron} 
\LX{mole} \HM{} \PH{mʊ̀lɪ̀} \PS{n.~} \GE{get wet} 
\LX{mɔll} \HM{} \PH{mɔ̀ll} \PS{adv.~} \GE{long and smooth} 
\LX{mollaa} \HM{} \PH{mʊ́lláà} \PS{n.~} \GE{a type of game in which a loop of string is hidden in soil to be found by other players using twigs} \OP{molle} \TP{mollɛɛ} 
\LX{molle} \HM{} \PH{mʊ́llɪ̀} \PS{v.~} \SN \GE{to deceive, to hide from} \SN \GE{to play the game mollaa} \PD{mollɛɛ, mollɔ/molla, mollɔ/molla, molbɔ/molba, mollaa, - } 
\LX{mɔlɔ} \HM{} \PH{mɔ̀lɔ́} \PS{adj.~} \SN \GE{cheap} \SN \GE{easy} \VA{mɔlɛɛ}
\LX{molɔmolɔ} \HM{} \PH{mʊ́lɔ́mʊ́lɔ́} \PS{adj.~} \GE{not stiff; flexible} \XV{A tammo waa la molɔmolɔ lɛ.} \XE{The bow is very flexible.} 
\LX{mɔlɔmɔlɔ} \HM{} \PH{mɔ́lɔ́mɔ́lɔ́} \PS{adj.~} \GE{seemingly easy, like a joke} 
\LX{mɔloo} \HM{} \PH{mɔ̀lʊ́ʊ́} \PS{n.~} \GE{a type of large antelope} \OP{mɔlle} \TP{mɔlɛɛ} 
\LX{mɔmaara} \HM{} \PH{mɔ̀màárà} \PS{n.~} \GE{fresh grass} \TP{mɔmaarɛɛ} 
\LX{momm} \HM{} \PH{mómm} \PS{adv.~} \GE{wholly and completely} 
\LX{momone} \HM{} \PH{mʊ̀mùnɪ́} \PS{n.~} \GE{dried salted fish} \TP{momonɛɛ} 
\LX{mɔnaabo} \HM{} \PH{mɔ̀náábʊ́} \PS{n.~} \GE{bush cow} \OP{mɔnii} \TP{mɔniinee} \VA{wɛnaabo}
\LX{monaŋe} \HM{} \PH{mʊ̀náŋ} \PS{n.~} \GE{hippopotamus} \OP{monanne} \TP{monannɛɛ} 
\LX{mone} \HM{} \PH{mʊ̀nɪ́} \PS{n.~} \GE{serious condition of piles} \OP{momɔ} \TP{monɛɛ/monnɛɛ} 
\LX{mone} \HM{} \PH{mʊ̀nɪ̀} \PS{v.~} \GE{to become serious} \PD{monɛɛ, mona, mona, monema, monaa, - } 
\LX{mɔŋe} \HM{} \PH{mɔ̀ŋ} \PS{v.~} \SN \GE{to stir (food)} \SN \GE{to steer, to drive (car)} \SN \GE{to paddle (boat)} \PD{mɔŋɛɛ, mɔnɔ, mɔnɔ, mɔnemɔ, mɔnaa, mɔnemɔ } 
\LX{mɔŋe} \HM{} \PH{mɔ́ŋ} \PS{v.~} \GE{to refuse to give} \XV{A entuo daana sɔre la a bondaana libiri ka o mɔŋ o.} \XE{The lazy man begged for money from the rich man, but he refused him.} \PD{mɔŋɛɛ, mɔnɔ, mɔnnɔ, mɔnemɔ, mɔnaa, mɔnemɔ } 
\LX{monɛɛ} \HM{} \PH{mʊ̀nɛ̀ɛ̀} \PS{adj.~} \GE{red} 
\LX{mɔneɛ} \HM{} \PH{mɔ̀nɪ́ɛ́} \PS{n.~} \GE{bushman, uncivilised person} \OP{mɔnoba} 
\LX{monɛɛmonɛɛ} \HM{} \PH{mʊ́nɛ́ɛ́mʊ́nɛ́ɛ́} \PS{adj.~} \GE{red looking; bloodshot (eyes)} 
\LX{moŋgbɛnɛɛ} \HM{} \PH{mʊ́ŋgbɛ́nɛ́ɛ́} \PS{n.~} \GE{infected millet seeds} \SG{moŋgbɛnnoo} 
\LX{moŋgolɔ} \HM{} \PH{mʊ̀ŋgʊ̀lɔ̀} \PS{n.~} \GE{a type of fall in which the person rolls over} \XV{A da᷉a bugro lee la willi moŋgolɔ.} \XE{The drunk man fell and rolled over.} \VA{moŋgola}
\LX{monne} \HM{} \PH{mʊ́nnɪ̀} \PS{v.~} \SN \GE{to lift and carry a heavy load without assisstance} \SN \GE{to dip to fill} \SN \GE{to gather up (smile to laugh)} \PD{monnɛɛ, monna, monna, monema, monnaa, - } 
\LX{montole} \HM{} \PH{mʊ̀ntʊ̀lɪ̀} \PS{n.~} \GE{a type of small round beans} \OP{montola} \TP{montollɛɛ} 
\LX{mɔnyuu} \HM{} \PH{mɔ̀ɲúú} \PS{n.~} \GE{a type of grass with a strong smell and has medicinal use} \OP{mɔnyuuri} \TP{mɔnyuuree} 
\LX{moɔ} \HM{} \PH{mʊ̀ɔ̀} \PS{v.~} \SN \GE{to make red} \SN \GE{to be ripe} \SN \GE{to be serious} \XV{Ka fooŋ boɔrɔ ka fo paase gyennoo fo maŋ moɔ nimiri ne zannoo.} \XE{If you want to pass a test, you need to be serious about your studies.} \SN \GE{to fetch fire (red coals)} \PD{moɔɛ, moɔnɔ, moɔnɔ, moɔnema, moɔnaa, moɔnema } 
\LX{moɔ} \HM{} \PH{mʊ̀ɔ̀} \PS{adv.~} \GE{really} \XV{yeŋ moɔ ka a bie yi?} \XE{Really, where does the child come from?.} \VA{mɔɔ}
\LX{moɔ} \HM{} \PH{mʊ́ɔ́} \PS{n.~} \SN \GE{grass, weeds} \SN \GE{bush, wilderness} \SG{moɔnoo} \TP{moɔnɛɛ} 
\LX{moole} \HM{} \PH{mʊ̀ʊ̀lɪ̀} \PS{v.~} \SN \GE{to be stunted, to be weak} \SN \GE{to cook in small quantity} \SN \GE{to cook poorly e.g. corn paste} \PD{moolɛɛ, moolɔ, moolɔ, mooleba, -, - } 
\LX{moɔle} \HM{} \PH{mʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to stretch (hair)} \PD{moɔlɛɛ, moɔlɔ, moɔlɔ, moɔleba, moɔlaa, moɔleba } 
\LX{moɔle} \HM{} \PH{mʊ́ɔ́lɪ́} \PS{v.~} \GE{to announce} \XV{Ba moɔle ka neɛzaa ta wa gɛrɛ tomɔ bie.} \XE{It has been announced that no one should go to work tomorrow.} \PD{moɔlɛɛ, moɔlɔ, moɔlɔ, moɔleba, moɔlaa, moɔleba } 
\LX{moɔle} \HM{} \PH{mʊ́ɔ́lɪ́} \PS{v.~} \GE{to vanish, to sneak} \XV{Kɔkɔlaa wa la a yiri ka gyaŋgbonni zaa moɔle.} \XE{A cat has come to the house and all the mice have vanished.} \PD{moɔlɛɛ, moɔlɔ, moɔlɔ, moɔleba, moɔlaa, moɔleba } 
\LX{moɔlɔ} \HM{} \PH{mʊ́ɔ́lɔ̀} \PS{n.~} \GE{established behaviour, character; (past)} \XV{A bie de la o saa moɔlɔ.} \XE{The child has taken after his father.} \TP{moɔlɛɛ} \VA{moɔla}
\LX{mooloo} \HM{} \PH{mʊ́ʊ́lʊ́ʊ́} \PS{adj.~} \GE{weak, stunted} 
\LX{moɔraa} \HM{} \PH{mʊ́ɔ́ráà} \PS{adv.~} \GE{like, similar manner} \VA{moɔreŋ}
\LX{moore} \HM{} \PH{mʊ́ʊ́rɪ̀} \PS{v.~} \SN \GE{to empathise; to sympathise with} \XV{Koŋkonimo maŋ kono moorɔ korɔ la kotuo deme.} \XE{Funeral dirge singers usually empathise with the bereaved family.} \SN \GE{to express regret for not being able to provide something} \XV{A da᷉a᷉pɔge moore la a da᷉a᷉.} \XE{The pito seller regrets that the pito is gone.} \PD{moorɛɛ, moorɔ, moorɔ, mooreba, mooraa, - } 
\LX{moore} \HM{} \PH{mʊ́ʊ́rɪ̀} \PS{v.~} \SN \GE{to dissolve, to diminish} \XV{A zeɛre nyɛnnoŋ ba moore.} \XE{The salt in the soup is not dissolved.} \SN \GE{to rain lightly} \XV{A saa ba mi soŋ mooroo ka o moore.} \XE{It did not rain hard; it just drizzled.} \SN \GE{to intercede, to suggest} \XV{N daare moore la o yiri kuluu ka o nyare.} \XE{When I suggested his going home, he flared up.} \PD{moorɛɛ, moorɔ, moorɔ, mooreba, mooraa, - } 
\LX{moore} \HM{} \PH{mʊ̀ʊ̀rɪ} \PS{v.~} \GE{to suck e.g. finger} \PD{moorɛɛ, moora, moora, mooreba, mooraa, - } \VA{mogre}
\LX{moɔre} \HM{} \PH{mʊ́ɔ̀rɪ̀} \PS{v.~} \SN \GE{to smear with thick liquid e.g. mud} \SN \GE{to camouflage} \XV{O ku᷉u deɛ moɔre ma la.} \XE{He camouflaged us before his death.} \PD{moɔrɛɛ, moɔrɔ, moɔrɔ, moɔreba, moɔraa, - } 
\LX{moore tere} \HM{} \PH{mʊ́ʊ́rɪ́ tɪ̀rɪ̀} \PS{v.~} \GE{to sympathise with} \PD{-, -, -, -, -, - } 
\LX{moɔreŋ} \HM{} \PH{mʊ́ɔ́rɪ́ŋ} \PS{adj.~} \GE{similar to how the Mosi people do it} \VA{komoɔreŋ}
\LX{moɔreŋ} \HM{} \PH{mʊ́ɔ́rɪ́ŋ} \PS{pp.~} \GE{like, according to} \XV{ŋmaaŋa moɔreŋ, dabe᷉ɛ sobaŋ taa yiri.} \XE{According to the monkey, he who runs away lives to fight another day.} \VA{moɔraa}
\LX{mooreterɛ} \HM{} \PH{mʊ́ʊ́rtɪ́rɛ́} \PS{n.~} \GE{consoler} \OP{moorterebɛ} \TP{mooreterebɛɛ} 
\LX{moɔrɔ} \HM{} \PH{mʊ́ɔ́rɔ́} \PS{n.~} \GE{illness of sexual misconduct; adultery} \TP{moɔrɛɛ} 
\LX{moorteroo} \HM{} \PH{mʊ́ʊ́rtɪ́rʊ́ʊ́} \PS{n.~} \GE{the act of consoling} 
\LX{mɔpelaa} \HM{} \PH{mɔ̀pɪ́láà} \PS{n.~} \GE{a stretch of bush made mainly of dry tall grass} \OP{mɔpeɛle} \TP{mɔpeɛlɛɛ} 
\LX{more} \HM{} \PH{mʊ̀rɪ̀} \PS{v.~} \GE{to disappear, to vanish} \PD{morɛɛ, morɔ, morɔ, morba, moraa, morba } 
\LX{more} \HM{} \PH{mʊ̀rɪ̀} \PS{v.~} \GE{to pop out e.g. seeds from a fruit in the mouth} \PD{morɛɛ, morɔ, morɔ, morba, moraa, morba } 
\LX{more} \HM{} \PH{mʊ́rɪ́} \PS{v.~} \SN \GE{to smile} \SN \GE{to initiate (typically used with laare)} \PD{morɛɛ, morɔ, morɔ, moreba, moraa, moreba } 
\LX{mɔre} \HM{} \PH{mɔ́rɪ́} \PS{n.~} \GE{a type of strong grass used for arrow shafts} \OP{mɔɛ} \TP{mɔrɛɛ} 
\LX{mɔre} \HM{} \PH{mɔ́rɪ́} \PS{v.~} \GE{to swell} \PD{mɔrɛɛ, mɔrɔ, mɔrɔ, mɔreba, mɔraa, mɔreba } 
\LX{moremore} \HM{} \PH{mʊ̀rɪ̀mʊ́rɪ́} \PS{v.~} \GE{to vanish at once} \PD{-, -, -, -, -, - } 
\LX{morɔ} \HM{} \PH{mʊ́rɔ́} \PS{n.~} \GE{herb tablet for external use} \OP{morre} \TP{morɛɛ} \VA{mora}
\LX{moromoro} \HM{} \PH{mórómóró} \PS{adj.~} \GE{soft, not solid} 
\LX{morɔmorɔ} \HM{} \PH{mʊ́rɔ́mʊ́rɔ́} \PS{adv.~} \SN \GE{showing slightly (smile)} \XV{O nyɛ la wagyɛ paalaa a monne laare morɔmorɔ lɛ.} \XE{She has got a new coth and this has made her beam with smiles.} \SN \GE{get lost mysteriously} \XV{A libie deɛ bɔrɛɛ la morɔmorɔ lɛ.} \XE{The money just got lost mysteriously.} 
\LX{mɔrɔmɔrɔ} \HM{} \PH{mɔ̀rɔ̀mɔ̀rɔ̀} \PS{adv.~} \GE{marshy, soft, and loose, so that it can't be held together} 
\LX{moroo} \HM{} \PH{mʊ́rʊ́ʊ́} \PS{n.~} \GE{local herbal medicine in tablet form used in massaging only} \OP{morre} \TP{morɛɛ/morrɛɛ} \VA{morɔ}
\LX{mɔrɔtɔ} \HM{} \PH{mɔrɔtɔ} \PS{adj.~} \GE{long and flat} 
\LX{mɔsɔglaa} \HM{} \PH{mɔ̀sɔ́glàá} \PS{n.~} \GE{a type of grass whose presence indicates fertile land} \OP{mɔsɔglɔ} 
\LX{mɔsogɔ} \HM{} \PH{mɔ̀sʊ́gɔ́} \PS{n.~} \GE{bush, the middle of the wilderness} \OP{mɔsogre} \TP{mɔsogrɛɛ} 
\LX{mɔtɔ} \HM{} \PH{mɔ́tɔ́} \PS{adv.~} \GE{very full} \XV{A boɔraa e la mɔtɔ o naŋ paale.} \XE{The sack is completely full.} 
\LX{mɔvɔ} \HM{} \PH{mɔ̀vɔ̀} \PS{n.~} \GE{part of the rainy season when there is thick growth of grass} \TP{mɔvɔrɛɛ} 
\LX{mɔwegi} \HM{} \PH{mɔ̀wégí} \PS{n.~} \GE{wild bush} \OP{mɔwegiri} \TP{mɔwegiree} 
\LX{mui} \HM{} \PH{mùí} \PS{n.~} \GE{rice} \TP{muunee} 
\LX{muli} \HM{} \PH{múlí} \PS{v.~} \GE{to remove hair or feathers from skin (usually by applying hot water)} \PD{mulee, mulo, mulo, mulibo, mulaa, - } 
\LX{mulli} \HM{} \PH{mùllì} \PS{v.~} \GE{to dive, to go underwater} \PD{mullee, mullo, mullo, mullibo, mullaa, mullibo } 
\LX{muni} \HM{} \PH{múní} \PS{n.~} \SN \GE{meaning} \XV{A dɔɔ ba baŋe a yelbiri ŋa muni.} \XE{the man does not know the meaning of this word.} \SN \GE{reason} \XV{A dɔɔ ba wuli a o polisiri ŋmeɛbo muni} \XE{The man does not know the reason why he was beaten by the police.} 
\LX{muni} \HM{} \PH{múní} \PS{adv.~} \SN \GE{bottom} \XV{A bie nu koŋ ta a doge muni.} \XE{The child’s hand cannot get to the bottom of the pot.} \SN \GE{under} \XV{A saama zeŋe la a maŋgo teɛ muni.} \XE{The strangers are sitting under the mango tree.} 
\LX{munii} \HM{} \PH{múníí} \PS{adj.~} \GE{tiny, small} \XV{A bɛmmo ŋa biri maŋ waa la munii lɛ.} \XE{This variety of bean has a tiny seed.} \OP{muno} \TP{munnee} 
\LX{muŋkuri} \HM{} \PH{mùŋkúrí} \PS{n.~} \GE{skirt} \OP{muŋkurri} \TP{muŋkuree} 
\LX{munkyara} \HM{} \PH{mùnʧárá} \PS{n.~} \GE{a type of smock with an opening at the back} \OP{munkyarre} \TP{munkyarɛɛ} 
\LX{mupɛge} \HM{} \PH{múpɛ́gɪ́} \PS{n.~} \GE{grass husk} \OP{mupɛgre} \TP{mupɛgrɛɛ} 
\LX{mupelaa} \HM{} \PH{mùpílàá} \PS{n.~} \GE{a type of straw from tall grass used in lighting fires} \OP{mupeɛle} \TP{mupeɛlɛɛ} 
\LX{muri} \HM{} \PH{múrí} \PS{n.~} \SN \GE{straw from a type of tall grass used in making mats} \SN \GE{a bundle of such straw} \OP{mie} \TP{miiree} 
\LX{muri} \HM{} \PH{múrí} \PS{n.~} \GE{snake} \OP{mie} \TP{miiree} 
\LX{muri} \HM{} \PH{mùrì} \PS{v.~} \SN \GE{to dodge, to escape} \SN \GE{to slide out e.g. rope around one's neck} \SN \GE{to deny an earlier statement} \PD{muree, muro, muro, muribu, muraa, muribo } 
\LX{muro} \HM{} \PH{mùró} \PS{n.~} \GE{the posture of lying facedown} \XV{Ba mɔɛ la te le vɔgele muro.} \XE{They wrestled and fell facedown.} 
\LX{muu} \HM{} \PH{múú} \PS{n.~} \GE{deaf and mute person} \OP{muuni} \TP{muunee} 
\LX{muu} \HM{} \PH{múú} \PS{adj.~} \GE{deaf and mute} 
\LX{muuli} \HM{} \PH{múúlí} \PS{v.~} \GE{to peep into or through} \PD{muulee, muulo, muulo, muulibo, muulaa, - } 
\LX{muuli} \HM{} \PH{mùùlì} \PS{v.~} \GE{to make a short visit (typically used with nyɛ)} \XV{N baŋ ka fo pagra la kyɛ fo naŋ muuli a baala kyɛ gɛrɛ.} \XE{I know you are in a hurry, but just pop in to visit the patient and then go.} \PD{muulee, muulo, muulo, muulibo, muulaa, - } 
\LX{muu-u} \HM{} \PH{mùú-ú̀} \PS{n.~} \GE{moo (sound made by cows)} \OP{muuni} \TP{muunee} 
\LX{muuu} \HM{} \PH{múúù} \PS{n.~} \SN \GE{onomatopoeia; the characteristic sound made by a cow} \SN \GE{cow} 
\LX{N} \HM{} \PH{ɪ̀n} \PS{pronoun.~} \GE{I (first person singular nominative)} \XV{N na gaa la daa.} \XE{I will go to the market.} \SN \GE{my (first person singular possessive)} \XV{N libie baarɛɛ la.} \XE{My money is gone.} 
\LX{na} \HM{} \PH{ná} \PS{pronoun.~} \GE{that (third person demonstrative)} \XV{A bi ŋmeene na beɛrɛ la.} \XE{That lean child is sick.} 
\LX{na} \HM{} \PH{nà} \PS{particle.~} \GE{will, shall (future tense marker)} \XV{A baala na gaa la asibiti.} \XE{The patient will go to the hospital.} 
\LX{ŋa} \HM{} \PH{ŋá} \PS{pronoun.~} \GE{this (third person demonstrative)} \XV{A dɔɔ ŋa zaameŋ wa la kyɛ.} \XE{This man came here yesterday.} 
\LX{ŋa} \HM{} \PH{ŋá} \PS{adv.~} \GE{similar to, like} \XV{A dɔɔ waa ŋa sogya na.} \XE{The man is like a soldier.} 
\LX{Naa} \HM{} \PH{nàá} \PS{n.~} \GE{name traditionally given to a second male twin} \OP{Namine} 
\LX{naa} \HM{} \PH{nàá} \PS{n.~} \SN \GE{chief, king} \SN \GE{grandparent} \SN \GE{rich person} \OP{namine} \TP{narɛɛ} 
\LX{naa} \HM{} \PH{nàá} \PS{modal verb.~} \SN \GE{will/shall (emphatic 'na')} \XV{Te naa wa la.} \XE{We shall come.} \SN \GE{should have} \XV{Fo da naa taa la a bie gaa ne asibiti.} \XE{You should have taken the child to the hospital.} \SN \GE{would have} \XV{A boɔso ba wa wiẽu ka lɛ naane fo da naa wa waana ko o gaaɛ.} \XE{The bus came late, otherwise by the time you came it would have left.} 
\LX{naa} \HM{} \PH{náá} \PS{adv.~} \GE{not} \XV{Onɔ naa ka N yele.} \XE{I'm not referring to that one.} \VA{shortened form of naane}
\LX{naabaa} \HM{} \PH{nààbáá} \PS{n.~} \GE{a type of small black bird regarded a totem of a clan (Zagɛɛ)} \OP{naabaare} \TP{naabaarɛɛ} 
\LX{naabaaraa gane} \HM{} \PH{náábááráá gánɪ̀} \PS{n.~} \GE{bona fide property; status not changeable; that which naturally belongs to a specific set} \XV{Ka Dagabieŋ baa o areyiriŋ kyɛ o naŋ waa la naabaaraa gane.} \XE{Among the Dagaaba, even if a child is brought up in his mother's home, he still remains the kin of the father.} 
\LX{naabere} \HM{} \PH{náábɪ̀rɪ́} \PS{n.~} \GE{udder} \OP{naaberɛ} \TP{naaberɛɛ} 
\LX{naabileŋmene} \HM{} \PH{náábíléŋmɪ́nɪ̀} \PS{n.~} \GE{God, supreme being} \OP{naabileŋmemɛ} 
\LX{naabilfaŋ} \HM{} \PH{náábìlfáŋ} \PS{n.~} \GE{a large calf} \OP{naabilfanne} \TP{naabilfannɛɛ} 
\LX{naabil-paaŋaa} \HM{} \PH{náábìl-pááŋáá} \PS{n.~} \GE{calf, baby cow} \OP{naabil-paane} \TP{naabil-paanɛɛ} 
\LX{naabini} \HM{} \PH{náábíní} \PS{n.~} \GE{cow dung} \SG{naabinuu} \TP{naabinee} 
\LX{naabo} \HM{} \PH{náábʊ́} \PS{n.~} \GE{cow} \OP{nii} \TP{niinee} 
\LX{naada} \HM{} \PH{náádà} \PS{n.~} \GE{the act of buying cattle} 
\LX{naadaara} \HM{} \PH{náádáárá} \PS{n.~} \GE{one who buys cattle} \OP{naadareba} \TP{naadarebɛɛ} 
\LX{naadaaraa} \HM{} \PH{náádàáráà} \PS{n.~} \GE{a cow that has been bought} \OP{naadaare} 
\LX{naadalane} \HM{} \PH{náádalánɪ́̀} \PS{n.~} \GE{uncastrated bull} 
\LX{naadalee} \HM{} \PH{náádàléé} \PS{n.~} \GE{young bull, calf} \OP{naadabilii} \TP{naadabilee} 
\LX{naadaŋaa} \HM{} \PH{náádáŋáá} \PS{n.~} \GE{kraal for cattle} \OP{naadaŋene} \TP{naadaŋenɛɛ} 
\LX{naadavaraa} \HM{} \PH{náádàváràà} \PS{n.~} \GE{castrated bull} 
\LX{naadɛre} \HM{} \PH{náádɛ́rɪ́} \PS{n.~} \GE{bullock, bull} \OP{naadɛɛ} \TP{naadɛrɛɛ} 
\LX{naadɔgrɔ} \HM{} \PH{náádɔ̀grɔ́} \PS{n.~} \GE{a cow that has recently given birth} 
\LX{naaduoraa} \HM{} \PH{nåådúóráá} \PS{n.~} \GE{resting ground for cattle} \OP{naaduori} \TP{naaduoree} 
\LX{naafaa} \HM{} \PH{nààfáà} \PS{n.~} \GE{that which is advantageous} \XV{Ka anaŋ waa lɛ ka lɛ naafaa la kyaare te.} \XE{If that is what it has come to, then luck is on our side.} 
\LX{naaga} \HM{} \PH{nààgà} \PS{n.~} \GE{a type of guinea corn} \OP{naagare} \TP{naagarɛɛ} 
\LX{naagaaraa} \HM{} \PH{náágááráá} \PS{n.~} \GE{feral cow} \OP{naagaare} \TP{naagaarɛɛ} 
\LX{naagale} \HM{} \PH{náágálɪ́} \PS{n.~} \GE{a type of dance which is meant to invite other dancers to the stage} \OP{naagalle} \TP{naagalɛɛ} 
\LX{naagaŋa} \HM{} \PH{náágàŋá} \PS{n.~} \GE{kraal for cattle} \OP{naaganne} \TP{naagannɛɛ} 
\LX{naagane} \HM{} \PH{náágánɪ̀} \PS{n.~} \GE{cowhide} \OP{naagama} \TP{naagannɛɛ} 
\LX{naakoɔraa} \HM{} \PH{náákʊ́ɔ́ráá} \PS{n.~} \GE{working bullock} \OP{naakoɔre} \TP{naakoɔrɛɛ} 
\LX{naakoɔrɔ} \HM{} \PH{náákʊ́ɔ́rɔ́} \PS{n.~} \GE{cattle seller} \OP{naakoɔrebɔ} \TP{naakoɔrebɛɛ} 
\LX{naakyeemɛgbeŋkeroo} \HM{} \PH{nááʧɪ́ɪ́mɛ́gběŋkéróó} \PS{n.~} \GE{a type of wild small tree with edible fruit} \OP{naakyeemɛgbeŋkerri} \TP{naakyeemɛgbeŋkeree} \VA{gbeŋkeroo}
\LX{naakyeenɛ} \HM{} \PH{nááʧɪ́ɪ́nɛ́} \PS{n.~} \GE{cattle herdsman} \OP{naakyeemɛ} \TP{naakyeemɛɛ} 
\LX{naalee} \HM{} \PH{nááléé} \PS{n.~} \GE{calf, heifer; a young cow or bull} \OP{naabilii} \TP{naabillee} 
\LX{naamaaroŋ} \HM{} \PH{nàámáárʊ́ŋ} \PS{n.~} \GE{large umbrella} \OP{naamaara} \TP{naamaarɛɛ} 
\LX{naamanyeɛnɛ} \HM{} \PH{náámáɲɪ́ɛ́nɛ́} \PS{adj.~} \GE{common, unimportant} 
\LX{naameɛ} \HM{} \PH{nààmɪ̀ɛ́} \PS{n.~} \GE{water turtle} \OP{naameene} \TP{naameenɛɛ} \VA{naameɛo}
\LX{naameɛo} \HM{} \PH{nààmɪ̀ɛ̀ʊ̀} \PS{n.~} \GE{turtle} \OP{naameɛne} \TP{naameɛnɛɛ} \VA{naameɛ}
\LX{naamoɔ} \HM{} \PH{náámʊ́ɔ́} \PS{n.~} \GE{a type of grass} \SG{naamoɔnoo} \OP{naamoore} \TP{naamoɔnɛɛ} 
\LX{naamu} \HM{} \PH{nàámù} \PS{n.~} \GE{head cloth} \OP{naamuri} \TP{naamuree} 
\LX{naane} \HM{} \PH{nàànɪ̀} \PS{v.~} \GE{to prepare, to get ready, to gather momentum to do something} \PD{naanɛɛ, naana, naana, naanema, naanaa, - } 
\LX{naane} \HM{} \PH{náánɪ́} \PS{aux. v.~} \GE{not} \XV{Ona naane de a libie.} \XE{S/he is not the one who took the money.} 
\LX{naanɛne} \HM{} \PH{náánɛ́nɪ̀} \PS{n.~} \GE{beef} \OP{naanɛmɛ} \TP{naanɛnnɛɛ} 
\LX{Naaŋmene} \HM{} \PH{nààŋmɪ́nɪ̀} \PS{n.~} \GE{God} \XV{naaŋmemɛ} \XE{} \OP{naaŋmemɛ} \TP{naaŋmenɛɛ} 
\LX{naanyaŋaa} \HM{} \PH{nááɲáŋàá} \PS{n.~} \GE{cow that has had a calf} \OP{naanyaŋene} \TP{naanyaŋenɛɛ} 
\LX{naapɔge} \HM{} \PH{nààpɔ́gɪ́} \PS{n.~} \GE{daughter-in-law} \OP{naapɔgeba/naapɔgemine} \TP{naapɔgrɛɛ} 
\LX{naapɔge} \HM{} \PH{náápɔ́gɪ́} \PS{n.~} \GE{female cow} \OP{naapɔgebɔ} \TP{naapɔgebɛɛ} 
\LX{naapɔle} \HM{} \PH{náápɔ́lɪ̀} \PS{n.~} \GE{young cow} \OP{naapɔlɔ} 
\LX{naara} \HM{} \PH{náárà} \PS{n.~} \GE{a variety of sorghum} \SG{naaroo} \OP{naare} \TP{naarɛɛ} 
\LX{naare} \HM{} \PH{náárɪ̀} \PS{adj.~} \GE{four} \OP{naare} \TP{naarɛɛ} 
\LX{naare} \HM{} \PH{nààrɪ̀} \PS{v.~} \SN \GE{strife to snatch from the ground} \SN \GE{to attack, usually violently} \PD{naarɛɛ, naara, naara, naareba, naaraa, - } 
\LX{naare} \HM{} \PH{náárɪ̀} \PS{v.~} \GE{to delay, to be late} \XV{A dɔɔta naŋ woŋ a bie baaloŋ o ba naare fe᷉e zaa.} \XE{When the doctor learnt of the child's sickness, he did not delay at all.} \PD{naarɛɛ, naara, naara, naareba, naaraa, - } 
\LX{naasaa} \HM{} \PH{náásàà} \PS{n.~} \GE{dragonfly} \OP{naasaare} \TP{naasaarɛɛ} 
\LX{naasaalaa} \HM{} \PH{nààsááláá} \PS{n.~} \GE{white person, European} \OP{nasamine} \TP{naasaalɛɛ} 
\LX{naasaal-kuuri} \HM{} \PH{nààsààl-kúúrí} \PS{n.~} \GE{plough} \OP{naasaal-kue} \TP{naasaal-kuree} 
\LX{naasaalsensɛroo} \HM{} \PH{nààsàálsɪ́nsɛ́rʊ́ʊ́} \PS{n.~} \GE{biscuits} \SG{naasaalsensɛroo} \OP{naasaalsensɛ} \TP{naasaalsensɛrɛɛ} 
\LX{naasaalsori} \HM{} \PH{nààsààlsórí} \PS{n.~} \GE{road} \OP{naasaalsoe} \TP{naasaalsoree} 
\LX{naasaal-taŋaa} \HM{} \PH{nààsààl-tàŋáà} \PS{n.~} \GE{mango} \OP{naasaal-taŋema} \TP{naasaal-taŋemɛɛ} \VA{maŋgo}
\LX{naasaal-teŋɛ} \HM{} \PH{nààsààl-tɪ́ŋɛ́} \PS{n.~} \GE{Europe} \OP{naasaal-tenne} \TP{naasaal-tennɛɛ} \VA{nasa-teŋɛ}
\LX{naasaraa} \HM{} \PH{náásáràà} \PS{n.~} \GE{adult female cow that has not littered yet} \OP{naasarre} \TP{naasarrɛɛ} \VA{naasarelee}
\LX{naasarelee} \HM{} \PH{náásárɪ́léé} \PS{n.~} \GE{adult female cow that has not littered yet} \OP{naasarebilii} \TP{naasarebillee} \VA{'naasaraa}
\LX{naaseere} \HM{} \PH{nààsɪ̀ɪ̀rɪ̀} \PS{n.~} \GE{a diseaese of the gums and teeth that produces sores} \OP{naaseerɛɛ} 
\LX{naaweɛmɛ} \HM{} \PH{nàáwɪ́ɛ́mɛ́} \PS{n.~} \GE{strayed cow} \OP{naaweɛne} \TP{naaweɛnɛɛ} 
\LX{naaweɛzoore} \HM{} \PH{nááwɪ́ɛ́zʊ́ʊ́rɪ́} \PS{n.~} \GE{a type of grass} \OP{naaweɛzoɛ} \TP{naaweɛzoorɛɛ} 
\LX{naayiri} \HM{} \PH{nààyírì} \PS{n.~} \SN \GE{palace, chief's house} \SN \GE{place of refuge} \OP{naayie} \TP{naayiree} 
\LX{naayoɔ} \HM{} \PH{nààyʊ́ɔ} \PS{n.~} \GE{young one, child} \OP{naayoore} \TP{naayoorɛɛ} 
\LX{naayoɔre} \HM{} \PH{nààyʊ̀ɔ̀rɪ̀} \PS{n.~} \SN \GE{penis of a cow} \SN \GE{a type of whip made of leather} \OP{naayɔɛ} \TP{naayɔrɛɛ} 
\LX{naazage} \HM{} \PH{náázàgɪ́} \PS{n.~} \GE{cattle pen} \OP{naazagre} \TP{naazagrɛɛ} 
\LX{naazeɛraa} \HM{} \PH{náázɪ̀ɛ̀ráà} \PS{n.~} \GE{hostile cow (usually one that has recently given birth)} \OP{naazeɛre} 
\LX{naazoŋmaa} \HM{} \PH{náázʊ̀ŋmáá} \PS{n.~} \GE{a cow that has lost part or all its tail} \OP{naazoŋmaara} \TP{naazoŋmaarɛɛ} 
\LX{naazoɔ} \HM{} \PH{náázʊ́ɔ̀} \PS{n.~} \GE{fly, housefly} \OP{naazoore} \TP{naazoorrɛɛ} \VA{zoɔ}
\LX{naazo-sɔglaa} \HM{} \PH{náázʊ̀-sɔ́gláá} \PS{n.~} \GE{black fly} \OP{naazo-sɔglɔ} \TP{naazo-sɔglɛɛ} 
\LX{naazozuri} \HM{} \PH{náázʊ̀zúrí} \PS{n.~} \GE{type of beads that look like the heads of flies} \TP{naazozuree} 
\LX{nabɛllebie} \HM{} \PH{nàbɛ́llɪ́bíé} \PS{n.~} \GE{camouflage} \OP{nabɛllebiiri} \TP{nabɛllebiiree} 
\LX{nabie} \HM{} \PH{nàbíé} \PS{n.~} \GE{royal person, prince} \OP{nabiiri} \TP{nabiiree} 
\LX{nabiiluŋ} \HM{} \PH{nàbììlúŋ} \PS{n.~} \GE{the condition of being a prince} 
\LX{naboma} \HM{} \PH{nàbʊ́má} \PS{n.~} \GE{riches, treasure} \OP{nabomɛɛ} 
\LX{nadire} \HM{} \PH{nàdíré} \PS{n.~} \GE{one who is qualified to be chief, chief-elect} \OP{ nadiribe} \TP{nadiribee} 
\LX{naferemɛ} \HM{} \PH{náfɪ́rɪ́mɛ́} \PS{n.~} \SN \GE{heel} \SN \GE{hoof} \SN \GE{hind toe (of a bird)} \OP{naferre} \TP{naferɛɛ/naferemɛɛ} 
\LX{nagbare} \HM{} \PH{nàgbárí} \PS{n.~} \GE{shin} \OP{nagbarɛɛ} \TP{nagharrɛɛ} 
\LX{nagbime} \HM{} \PH{nàgbímé} \PS{n.~} \GE{heavy footsteps} \TP{nagbimee} 
\LX{nage} \HM{} \PH{nágɪ́} \PS{v.~} \GE{to beat, to hit hard (especially on the back)} \PD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\LX{nage} \HM{} \PH{nágɪ́} \PS{v.~} \GE{to produce a lot of} \PD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\LX{nage} \HM{} \PH{nàgɪ̀} \PS{v.~} \GE{to bend backward} \PD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\LX{nage} \HM{} \PH{nàgɪ̀} \PS{v.~} \GE{to miss work} \XV{Daa zaa o maŋ nage la toma dayeni.} \XE{Every week she misses work once.} \PD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\LX{nage} \HM{} \PH{nàgɪ́} \PS{adj.~} \GE{crooked, bent backwards} \OP{nagre} 
\LX{nagenyuu} \HM{} \PH{nàgìɲúú} \PS{n.~} \GE{meningitis} 
\LX{nakogi} \HM{} \PH{nàkógí} \PS{n.~} \GE{throne} \OP{nakogri} \TP{nakogree} 
\LX{nakpi} \HM{} \PH{nàkpí} \PS{adj.~} \GE{mortal} 
\LX{nakyige} \HM{} \PH{nàʧígé} \PS{n.~} \GE{that which is great} \OP{nakyigri} \TP{nakyigree} 
\LX{nakyige} \HM{} \PH{nàʧígé} \PS{n.~} \GE{four-legged animal} \OP{nakyigri} \TP{nakyigree} 
\LX{nakyige} \HM{} \PH{nàʧígé} \PS{n.~} \GE{long, heavy stride} \OP{nakyigri} \TP{nakyigree} 
\LX{nama} \HM{} \PH{nàmá} \PS{n.~} \GE{carnivorous animal e.g. hyena} \TP{namɛɛ} 
\LX{namba} \HM{} \PH{námbà} \PS{n.~} \GE{number} \OP{nambare} \TP{nambarɛɛ} 
\LX{nambaŋka} \HM{} \PH{nàmbàŋkà} \PS{n.~} \GE{state of being probable} 
\LX{nambiyaalaa} \HM{} \PH{námbìyáálàà} \PS{n.~} \GE{centipede} \OP{nambiyaale} \TP{nambiyaalɛɛ} 
\LX{nambɔgelɔ} \HM{} \PH{nǎmbɔ́glɔ́} \PS{n.~} \GE{hunter} \OP{nambɔgele} \TP{nambɔgelɛɛ} 
\LX{nambonna} \HM{} \PH{nàmbʊ̀nná} \PS{n.~} \GE{doubtfulness} 
\LX{namelemɛ} \HM{} \PH{nàmɪ́lɪ́mɛ́} \PS{n.~} \GE{knock-knee} \TP{namelemɛɛ} 
\LX{namonɔ} \HM{} \PH{nàmʊ̀nɔ̀} \PS{n.~} \GE{body decoration e.g. tattoo} \OP{namonne} \TP{namonɛɛ/namonnɛɛ} \VA{namona}
\LX{naŋ} \HM{1} \PH{nǎŋ} \PS{adv.~} \SN \GE{since} \XV{O naŋ gaa  a kɔɔrɛɛ la.} \XE{It has been a long time since she went.} \SN \GE{yet} \XV{O naŋ ba gaa.} \XE{She has not gone yet.} \XV{O suuri naŋ naŋ ba sigi na te koŋ baŋ maale a zɔɔre.} \XE{Since his anger has not cooled off yet, we cannot arbitrate.} 
\LX{naŋ} \HM{2} \PH{nǎŋ} \PS{adv.~} \SN \GE{still} \XV{O naŋ zeŋ la be.} \XE{She is still sitting there.} \SN \GE{kindly, please} \XV{Naŋ ko ma ko᷉ɔ ka N nyu.} \XE{Kindly give me some water to drink.} 
\LX{nana} \HM{} \PH{nàná} \PS{adv.~} \GE{slowly and painfully e.g. a very sick person} 
\LX{naŋa} \HM{} \PH{náŋá} \PS{n.~} \GE{scorpion} \OP{nanne} \TP{nannɛɛ} 
\LX{nandaa} \HM{} \PH{nàndàá} \PS{n.~} \GE{limb} \OP{nandaare} \TP{nandarɛɛ} 
\LX{nandaana} \HM{} \PH{nándááná} \PS{n.~} \GE{poor person, needy person} \OP{ nandaanema} 
\LX{naŋe} \HM{} \PH{náŋɪ̀} \PS{n.~} \GE{poverty; scarcity} \XV{O toma lee la ka naŋ kpɛ o.} \XE{His business has collapsed and poverty has overtaken him.} \TP{nannɛɛ} 
\LX{naneŋ} \HM{} \PH{nànɪ́ŋ} \PS{adj.~} \GE{at last} \XV{Saa naneŋ mie la.} \XE{At last it has rained.} 
\LX{naŋene} \HM{} \PH{náŋɪ̀nɪ̀} \PS{n.~} \GE{hippopotamus} \OP{naŋema} \TP{naŋenɛɛ} 
\LX{naŋgbagele} \HM{} \PH{nǎŋgbáglɪ́} \PS{n.~} \GE{bony meat; bone left over after the flesh has been removed} \OP{naŋgbagela} \TP{naŋgbagelɛɛ} 
\LX{naŋgbeɛnɛ} \HM{} \PH{nǎŋgbɪ́ɛ́rɛ́} \PS{n.~} \SN \GE{bits of meat cut from from different parts of the animal} \SN \GE{profit} \XV{Boŋ naŋgbeɛrɛ ka fo nyɛ a toma ŋa poɔ?} \XE{What profit have you made from this work?.} \OP{ naŋgbeɛmɛ} \TP{naŋgbeɛrɛɛ} \VA{naŋgbeɛnɛ}
\LX{naŋkoɔredie} \HM{} \PH{nàŋkʊ̀ɔ̀rɪ̀dìé} \PS{n.~} \GE{store} \OP{naŋkoɔrederi} \TP{nankoɔrederee} 
\LX{naŋkoɔrɔ} \HM{} \PH{nàŋkʊ́ɔ́rɔ́} \PS{n.~} \GE{trader} \OP{naŋkoɔreba} \TP{naŋkoɔrebɛɛ} \VA{naŋkoɔreba}
\LX{naŋkoɔse} \HM{} \PH{nàŋkʊ̀ɔ̀sɪ́} \PS{n.~} \GE{butcher} \OP{naŋkoɔsere} \TP{naŋkoɔserɛɛ} 
\LX{naŋkpaana} \HM{} \PH{nǎŋkpááná} \PS{n.~} \GE{hunter} \OP{naŋkpaama/naŋkpaane} \TP{naŋkpaanɛɛ} 
\LX{naŋmane} \HM{} \PH{nàŋmánɪ́} \PS{n.~} \GE{pigeon} \OP{naŋmama} \TP{naŋmannɛɛ} 
\LX{nanna-nanna} \HM{} \PH{nánná-nánná} \PS{adv.~} \GE{seldom} \XV{O da ba nagera yiri kyɛ o ma naŋ kpi o maŋ nanna-nanna la wa kuli wa.} \XE{She used to come home regularly but since her mother's death she seldom comes home.} 
\LX{nannyige} \HM{} \PH{nànnyìgé} \PS{n.~} \GE{thief} \OP{nannyigiri} \TP{nannyigiree} 
\LX{nansoba} \HM{} \PH{nánsʊ̀bá} \PS{n.~} \GE{poor person, needy person} 
\LX{nantɛŋ} \HM{} \PH{nàntɛ̀ŋɪ̀} \PS{n.~} \GE{very large gun} \OP{nantɛnne} \TP{nantɛnnɛɛ} \VA{nantɛŋ}
\LX{nanyuuri} \HM{} \PH{nàɲúúrí} \PS{n.~} \GE{potato; sweet potato} \SG{nanyuuruu} \OP{nanyie} \TP{nanyiiree} 
\LX{Napɔge} \HM{} \PH{nàpɔ́gɪ́} \PS{n.~} \GE{name traditionally given to a younger female twin} \OP{napɔgeba} \TP{napɔgebɛɛ} 
\LX{napoloŋ} \HM{} \PH{nàpʊ̀lʊ́ŋ} \PS{n.~} \GE{calf (part of the leg)} \OP{napolle} \TP{napollɛɛ} 
\LX{naraa} \HM{} \PH{nàràà} \PS{adv.~} \GE{large} 
\LX{nare} \HM{} \PH{nárɪ́} \PS{v.~} \GE{to prepare; to get ready} \PD{narɛɛ, nara, nara, nareba, naraa, - } 
\LX{nare} \HM{} \PH{nàrɪ̀} \PS{v.~} \GE{to increase, to get out of control} \XV{A vu᷉u narɛɛ la ka ba kpinne faana.} \XE{The fire has gotten out of control and they are struggling to put it out.} \PD{narɛɛ, nara, nara, nareba, naraa, - } 
\LX{nasa-kuuri} \HM{} \PH{nàsà-kúúrí} \PS{n.~} \GE{plough} \OP{nasa-kue} 
\LX{natɛbaa} \HM{} \PH{nátɛ́báá} \PS{n.~} \GE{large boil or abscess} \OP{natɛrre} \TP{natɛrɛɛ} 
\LX{nato} \HM{} \PH{nàtó} \PS{n.~} \GE{a pair of metal high sounding instrument, used to call people's attention and as a musical instrument} \OP{natori} \TP{natoree} \VA{lato; latori}
\LX{nawale} \HM{} \PH{náwálɪ́} \PS{n.~} \GE{ulcer} \OP{nawalle} \TP{nawalɛɛ} \VA{nayagere}
\LX{nayagere} \HM{} \PH{nàyàgɪ̀rɪ́} \PS{n.~} \GE{ulcer} \OP{nayaga} \TP{nayagerɛɛ} \VA{nawale}
\LX{ne} \HM{} \PH{nɪ̂} \PS{conj.~} \SN \GE{and} \XV{Ko᷉ɔ ne vu᷉u᷉ koŋ to᷉ɔ᷉ laŋ kpɛ.} \XE{Water and fire cannot coexist.} \SN \GE{with} \XV{Te maŋ kyɛ la daa ne lare.} \XE{We cut a tree with an axe.} \VA{shortened form of ane}
\LX{nɛbiri} \HM{} \PH{nɛ́bìrí} \PS{n.~} \GE{smaller stone used in grinding on the main grindstone} \OP{nɛbie} \TP{nɛbiiree} 
\LX{nɛbo} \HM{} \PH{nɛ́bʊ́} \PS{v.~} \SN \GE{to dress well} \SN \GE{to tuck in one's shirt} \PD{nɛbɛɛ, nɛbrɛ, nɛbraa, nɛbrebɛ, nɛbraa, - } 
\LX{nɛ-bɔgeraa} \HM{} \PH{nɛ́-bɔ́gɪ́ràá} \PS{n.~} \GE{smooth grinding stone} \OP{nɛ-bɔgere} \TP{nɛ-bɔgerɛɛ} 
\LX{neɛ} \HM{} \PH{nɪ́ɛ́} \PS{v.~} \SN \GE{to trample, to stamp on} \SN \GE{to cheat} \PD{neɛɛ, neɛrɛ, neɛrɛ, neɛrebɛ, neɛraa, neɛrebɛ } 
\LX{neɛ} \HM{} \PH{nɪ́ɛ̀} \PS{v.~} \SN \GE{to step on} \XV{N ba baŋ kyɛ neɛ o gbɛre kyɛ ka o suuri iri.} \XE{I stepped on her foot unintentionally, but she was angry.} \SN \GE{to grind} \XV{A te᷉e᷉-erɛ neɛ la a te᷉e᷉ eŋ ko᷉ɔ᷉ poɔ ko a baala.} \XE{The herbal doctor ground the medicine and mixed it with water for the patient.} \PD{neɛɛ, neɛrɛ, neɛrɛ, neɛrebɛ, neɛraa, neɛrebɛ } 
\LX{neɛ} \HM{} \PH{nɪ̀ɛ̀} \PS{n.~} \GE{person} \OP{noba} \TP{nerrɛɛ} 
\LX{neɛne} \HM{} \PH{nɪ̀ɛ̀nɪ̀} \PS{v.~} \SN \GE{to push through or under} \XV{Kɛkɛpele neɛne la a bie nubi-e᷉ɛ.} \XE{A piece of stalk has lodged itself under the child's fingernail.} \SN \GE{to refer to indirectly, to gossip} \XV{A pɔge ŋa maŋ neɛnɛ la o yontaa.} \XE{This woman always gossips about her rival.} \SN \GE{to tease} \PD{neɛnɛɛ, neɛnɛ, neɛnɛ, neɛnemɛ, neɛnaa, - } 
\LX{neere} \HM{} \PH{nɪ̀ɪ̀rɪ̀} \PS{v.~} \GE{to dodge by squeezing through} \XV{A nannyige neere tu la a lɔɔre takoroo zo.} \XE{The thief squeezed himself through the car window and escaped.} \PD{neerɛɛ, neerɛ, neerɛ, neerebɛ, neeraa, - } 
\LX{neɛre} \HM{} \PH{nɪ́ɛ́rɪ̀} \PS{n.~} \GE{grindstone} \OP{nɛɛ} \TP{nɛrɛɛ} 
\LX{neɛre} \HM{} \PH{nɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to hide under cover} \XV{Kɔkɔlaa digi la a gyaŋgboŋo ka o zo kpɛ neɛre moɔ poɔ.} \XE{The cat chased the mouse and it ran into hiding under the grass.} \PD{-, -, -, -, -, - } 
\LX{neɛrekuuri} \HM{} \PH{nɪ́ɛ́rɪ́kúúrí} \PS{n.~} \GE{small stone used in grinding on the main grindstone} \OP{neɛrekubo} \TP{neɛrekuuree} \VA{nɛbiri}
\LX{neɛre-saare} \HM{} \PH{nɪ́ɛ́rsáárɪ́} \PS{n.~} \GE{small, soft broom used to gather flour} \OP{neɛrsaɛ} \TP{neɛrsaarɛɛ} 
\LX{neɛzaa} \HM{} \PH{nɪ́ɛ̀zàá} \PS{n.~} \GE{each person, everyone} \XV{Ba yeli ka neɛzaa gaa a kpaaroŋ.} \XE{It has been said that every person should attend the meeting.} 
\LX{nɛgaa} \HM{} \PH{nɛ́gáá} \PS{n.~} \GE{a type of cedar-like tree with very hard wood} \OP{nɛgere} \TP{nɛgerɛɛ} 
\LX{negebaga} \HM{} \PH{nɪ̀gɪ̀bàgà} \PS{adj.~} \GE{heavily built; usually referring to people} 
\LX{nɛgetɛgɛɛ} \HM{} \PH{nɛ̀gɪ̀tɛ́gɛ́ɛ́} \PS{adj.~} \GE{curved inward} \XV{A a die peɛ deɛ waa la nɛgetɛgɛɛ lɛ.} \XE{The roof of the house is looking sunken in.} \VA{nɛgtɛgɛ}
\LX{nɛgetɛgɛɛ} \HM{} \PH{nɛ̀gɪ́tɛ́gɛ́ɛ́} \PS{n.~} \GE{depression, place that is curved inward} \XV{Saa mie la ka ko᷉ɔ are a peɛ nɛgetɛgɛɛ poɔ.} \XE{It has rained and water has collected in a depression in the roof.} \OP{nɛgtɛgere} \TP{nɛgtɛgerɛɛ} 
\LX{negiri} \HM{} \PH{négirí} \PS{v.~} \GE{to bend the knees slightly as a sign of respect; usually by girls while the boys salute} \PD{negiree, negire, negire, negiribe, negiraa, - } 
\LX{nɛguo} \HM{} \PH{nɛ́gùó} \PS{n.~} \GE{raised ground on which grindstones are placed} \OP{nɛgori} \TP{nɛgoree} 
\LX{nɛ-gyieraa} \HM{} \PH{nɛ́-dzèráà} \PS{n.~} \GE{rough grindstone} \OP{nɛ-gyieri} 
\LX{nɛku-bile} \HM{} \PH{nɛ́kú-bílé} \PS{n.~} \GE{smaller version of a neɛrekuuri} \OP{nɛku-bilii} \TP{nɛku--bilee} 
\LX{nɛku-kpoŋ} \HM{} \PH{nɛ̀kú-kpôŋ} \PS{n.~} \GE{larger version of neɛrekuuri} \OP{nɛku-kponni} \TP{nɛku-kponnee} 
\LX{nɛle} \HM{} \PH{nɛ́lɪ́} \PS{v.~} \GE{to fill completely} \XV{A bogi da nɛle ne la ɔnne.} \XE{The burrow was full of mice.} \PD{nɛlɛɛ, nɛlɛ, nɛlaa, nɛlebɛ, nɛlaa, - } 
\LX{nelebala} \HM{} \PH{nɪ̀lɪ̀bàlà} \PS{adj.~} \GE{large and bent backward} 
\LX{nɛll} \HM{} \PH{nɛ̀ll} \PS{adv.~} \GE{long and unwieldy} 
\LX{nembaala} \HM{} \PH{nɪ́mbààlá} \PS{n.~} \GE{destitute, weak} \OP{nembaalba} \TP{nembaalbɛɛ} 
\LX{nembaaloŋ} \HM{} \PH{nɪ́mbààlʊ́ŋ} \PS{n.~} \GE{pity} 
\LX{nembaal-zoro} \HM{} \PH{nɪ́mbààl-zóró} \PS{n.~} \GE{merciful person} \OP{nembaal-zoribo} \TP{nembaal-zoribee} 
\LX{nembeɛ} \HM{} \PH{nɪ́mbɪ̀ɛ́} \PS{n.~} \GE{a mysterious person} \OP{nembeere} \TP{nembeerɛɛ} 
\LX{nembeene} \HM{} \PH{nɪ́mbɪ́ɪ́nɪ́} \PS{n.~} \GE{someone with supernatural powers} \OP{nembeemɛ} \TP{nembeemɛɛ} 
\LX{nɛmbeene} \HM{} \PH{nɛ́mbɪ́ɪ́nɪ́} \PS{n.~} \GE{cooked meat} \OP{nɛmbeemɛ} \TP{nɛmbeenɛɛ} 
\LX{nembeleŋ} \HM{} \PH{némbéléŋ} \PS{adj.~} \GE{bent slightly backwards} 
\LX{nɛmbɛlɛŋ} \HM{} \PH{nɛ̀mbɛ̀lɛ̀ŋ} \PS{adv.~} \GE{long and crooked} 
\LX{nembe᷉ɛo} \HM{} \PH{nɪ́mbɪ̀᷉ɛ̀᷉ʊ́᷉} \PS{n.~} \GE{crowd of people; many people} \OP{nebeɛne/nembeɛmɛ} \TP{nembeɛnɛɛ/nembeɛmɛɛ} 
\LX{nembieri} \HM{} \PH{nɪ́mbíérí} \PS{n.~} \GE{malicious person} \OP{nembebe} 
\LX{nɛmbiri} \HM{} \PH{nɛ́mbírì} \PS{n.~} \GE{boneless meat} \SG{nɛmbiruu} \OP{nɛmbie} \TP{nɛmbiiree} 
\LX{nemboŋ} \HM{} \PH{nímbóŋ} \PS{n.~} \GE{what sort of person, who} \XV{Nemboŋ la na to᷉ɔ᷉ di naaloŋ Dagaaba laŋkpeɛbo poɔ?} \XE{Who qualifies to be a chief among the Dagaaba?.} 
\LX{nembone} \HM{} \PH{nɪ́mbʊ́nɪ́} \PS{n.~} \GE{worthless person} \OP{nemboma} \TP{nembonnɛɛ} 
\LX{nembɔnnaa} \HM{} \PH{nɪ́mbɔ̀nnáà} \PS{n.~} \GE{acquaintance} \OP{nembɔnne} \TP{nembɔnnɛɛ} 
\LX{nembon-yɛlɛ} \HM{} \PH{nɪ́bʊ́n-yɛ́lɛ́} \PS{n.~} \GE{useless talk} 
\LX{nemboore} \HM{} \PH{nɪ́mbʊ́ʊ́rɪ́} \PS{n.~} \GE{type of person} \XV{Maa ba baŋ nemboore na zaa naŋ la o.} \XE{I don't know what type of person she is.} \OP{nemboorɔ} 
\LX{nembuo} \HM{} \PH{nɪ́mbúò} \PS{n.~} \GE{which person; who} \OP{nembuobo} 
\LX{nemburee} \HM{} \PH{nɪ́mbùréé} \PS{n.~} \GE{unintelligent person, confused person} \OP{nemburri} \TP{nemburree} 
\LX{nɛmmaaroŋ} \HM{} \PH{nɛ́mmáárʊ́ŋ} \PS{n.~} \GE{fresh meat} \OP{nɛmmaara} \TP{nɛmmaarɛɛ} 
\LX{nemmeŋɛ} \HM{} \PH{nɪ́mmɪ́ŋɛ́} \PS{n.~} \GE{good person} \OP{nemenne} \TP{nemmennɛɛ} 
\LX{nempaalaa} \HM{} \PH{nímpááláá} \PS{n.~} \GE{new person} \OP{nempaalba} \TP{nempaalɛɛ} 
\LX{nempelaa} \HM{} \PH{nɪ́mpɪ̀làá} \PS{n.~} \SN \GE{caucasian; white person} \SN \GE{simple person, straightforward person} \OP{nempeɛle} \TP{nempeɛlɛɛ} 
\LX{nempɔgeraa} \HM{} \PH{nɪ́mpɔ̀gɪ̀ráà} \PS{n.~} \GE{prisoner} \OP{nempɔgere} \TP{nempɔgerɛɛ} 
\LX{nempɔgere-die} \HM{} \PH{nɪ́mpɔ̀gɪ̀rdìé} \PS{n.~} \GE{prison} \OP{nempɔgere-deri} \TP{nempɔgere-deree} 
\LX{nɛmpoɔnaa} \HM{} \PH{nɛ́mpʊ́ɔ́náá} \PS{n.~} \GE{rotten meat} \OP{nɛmpoɔne} \TP{nɛmpoɔnɛɛ} 
\LX{nendaare} \HM{} \PH{nɪ̀ndààrɪ̀} \PS{adv.~} \GE{one day; in the future} \XV{O zagere la sakuuri gaabo nendaare kaŋa o na koŋ la dabaŋkalɛ.} \XE{S/he has refused to go to school; one day she will regret it.} 
\LX{nendɛgere} \HM{} \PH{nɪ́ndɛ́grí} \PS{n.~} \SN \GE{dirty person} \SN \GE{unscruplous person} \OP{nendɛgɛ} 
\LX{nɛne} \HM{} \PH{nɛ́nɪ̀} \PS{n.~} \SN \GE{meat} \SN \GE{flesh} \OP{ nɛmɛ} 
\LX{nɛnɛ} \HM{} \PH{nɛ́nɛ́} \PS{adv.~} \GE{precisely} \XV{A sokyɛnɛ manne la o sori yɛlɛ nɛnɛ lɛɛŋ te baare.} \XE{The traveler narrated precisely what she/he encountered on the journey.} 
\LX{nɛŋɛɛ} \HM{} \PH{nɛ̀ŋɛ̀ɛ̀} \PS{adj.~} \GE{saggy} 
\LX{nɛŋɛɛ} \HM{} \PH{nɛ́ŋɛ́ɛ́} \PS{adj.~} \GE{saggy} 
\LX{nɛnefɛne} \HM{} \PH{nɛ̀nɪ̀fɛ́nɪ́} \PS{v.~} \GE{to try to get attention} \XV{O naŋ waa biyeni na ana la so ka o nɛnefɛnɛ.} \XE{She is an only child; that is why she wants so much attention.} \PD{nɛnefɛnɛɛ, nɛnefɛnɛ, nɛnefɛnɛ, nɛnefɛnemɛ, nɛnefɛnaa, - } 
\LX{nenfaa} \HM{} \PH{nɪ̀nfáà} \PS{n.~} \GE{bad person, wicked person} \OP{nenfaare} \TP{nenfaarɛɛ} 
\LX{nenfole} \HM{} \PH{nɪ́nfʊ́lɪ́} \PS{n.~} \GE{albino} \OP{nenfola} \TP{nenfolɛɛ} 
\LX{neŋ-iraa} \HM{} \PH{nɪ́ŋ-íráà} \PS{n.~} \GE{chosen person} \OP{neŋ-irri} \TP{neŋ-irree} 
\LX{neŋkaa} \HM{} \PH{nɪ́ŋkàà} \PS{n.~} \SN \GE{the act of looking at people} \SN \GE{the act of caring for people} \OP{neŋkaareba} \TP{neŋkaarebɛɛ} 
\LX{neŋkaara} \HM{} \PH{nɪ́ŋkáárá} \PS{n.~} \SN \GE{one who looks at people} \SN \GE{one who takes care of people, caregiver} \OP{neŋkaareba} \TP{neŋkaarebɛɛ} 
\LX{neŋkaaraa} \HM{} \PH{nɪ̌ŋkààráà} \PS{n.~} \GE{one who is cared for} \OP{neŋkaare} 
\LX{neŋkanoo} \HM{} \PH{nɪ́ŋkànʊ́ʊ̀} \PS{n.~} \GE{mature person, old person} \OP{neŋkana/neŋkama} \TP{neŋkanɛɛ} 
\LX{neŋkaraa} \HM{} \PH{nɪ́ŋkáráá} \PS{n.~} \GE{an unruly person} \OP{neŋkare} \TP{neŋkarɛɛ} 
\LX{neŋkaraa} \HM{} \PH{nɪ́kàràà} \PS{n.~} \GE{large person} 
\LX{nɛŋkaraa} \HM{} \PH{nɛ́ŋkáráá} \PS{n.~} \GE{raw meat} \OP{nɛŋkare} \TP{nɛŋkarɛɛ} 
\LX{nɛŋka᷉a} \HM{} \PH{nɛ́ŋká᷉à᷉} \PS{n.~} \GE{fatty meat} \SG{nɛŋkaao} \OP{nɛŋkaanɛɛ} 
\LX{neŋkɔge} \HM{} \PH{nɪ́ŋkɔ́gɪ́} \PS{n.~} \GE{lean person} \OP{neŋkɔgre} \TP{neŋkɔgrɛɛ} 
\LX{neŋkoorɔ} \HM{} \PH{nɪ́ŋkʊ́ʊ́rɔ̀} \PS{n.~} \GE{murderer} \OP{ neŋkoorebɔ} \VA{neŋkoora}
\LX{nɛŋkoorɔ} \HM{} \PH{nɛ́ŋkʊ́ʊ́rɔ̀} \PS{n.~} \GE{butcher} \OP{nɛŋkooreba} 
\LX{neŋkore} \HM{} \PH{nɪ́ŋkʊ́rɪ́} \PS{n.~} \GE{old person} \OP{neŋkora} 
\LX{neŋkpeɛŋaa} \HM{} \PH{nɪ́ŋkpɪ́ɛ́ŋáá} \PS{n.~} \SN \GE{healthy person} \SN \GE{strong person} \OP{ neŋkpeɛne} 
\LX{neŋkpoŋi} \HM{} \PH{nɪ́ŋkpôŋ} \PS{n.~} \SN \GE{adult, elder} \SN \GE{leader} \OP{neŋkponni} \TP{neŋkponnee} 
\LX{neŋkpɔre} \HM{} \PH{nɪ́ŋkpɔ́rɪ́} \PS{n.~} \GE{weak and sickly person} \OP{neŋkpɔɛ} \TP{neŋkpɔrɛɛ} 
\LX{nɛŋkuoŋaa} \HM{} \PH{nɛ́ŋkúóŋáá} \PS{n.~} \GE{smoked meat} \OP{nɛŋkuoni} \TP{nɛŋkuonee} \VA{nɛŋku᷉oaa}
\LX{nenlaane} \HM{} \PH{nínlåånɪ́} \PS{n.~} \GE{worthless person} \OP{nenlaama} \TP{nelaanɛɛ} 
\LX{nenlaara} \HM{} \PH{nɪ́nláárá} \PS{n.~} \GE{person who laughs at others; scornful person} \OP{nenlaareba} \TP{nenlaarebɛɛ} 
\LX{nenlaaraa} \HM{} \PH{nɪ́nlààràà} \PS{n.~} \GE{person who is scorned, laughingstock} \OP{nenlaare} \VA{neŋ-e-laaraa}
\LX{nɛŋŋmaara} \HM{} \PH{nɛ́ŋŋmáárá} \PS{n.~} \GE{butcher} \OP{nɛŋŋmaareba} \VA{naŋkoɔse, naŋkoɔsere, naŋkoɔserɛɛ}
\LX{nen-nyɔgera} \HM{} \PH{nɪ́n-ɲɔ́gɪ́rà} \PS{n.~} \GE{conqueror} \OP{ nen-nyɔgereba} \TP{nen-nyɔgerebɛɛ} \VA{nen-nyɔgerɔ}
\LX{nen-nyɔgeraa} \HM{} \PH{nɪ́n-ɲɔ́gɪ́ráá} \PS{n.~} \GE{captive} \OP{ nen-nyɔgere} \TP{nen-nyɔgerɛɛ} 
\LX{nensaala} \HM{} \PH{nɪ̌nsààlá} \PS{n.~} \GE{human being} \OP{nensaaleba} \TP{nensaalebɛɛ} 
\LX{nensaalabie} \HM{} \PH{nɪ́nsààlàbíé} \PS{n.~} \GE{son of man} \OP{nensaalabiiri} \TP{nensaalabiiree} 
\LX{nensaaloŋ} \HM{} \PH{nɪ́nsààlʊ́ŋ} \PS{n.~} \SN \GE{humanity} \SN \GE{something that is given a human face} \XV{Ka a karemazuzeɛŋ da ba de nensaaloŋ kpɛ ne a bie dɔgeroŋ poɔ a ŋa a bie yi a sakuuriŋ.} \XE{But for the headmaster giving a human face to the boy’s punishment the boy would have been out of school.} \SN \GE{personification (literature)} 
\LX{nensaal-sɔgelaa} \HM{} \PH{nɪ́nsáál-sɔ́gláá} \PS{n.~} \GE{the typical African person} \OP{nensaal-sɔglɔ} 
\LX{nenseɛ} \HM{} \PH{nɪ́nsɪ́ɛ̀} \PS{n.~} \GE{ghost, evil spirit, apparition} \OP{nenseere} \TP{nenseerɛɛ} 
\LX{nenseɛbaa} \HM{} \PH{nɪ̀nsɪ̀ɛ̀báá} \PS{n.~} \SN \GE{a fairly grown person in age} \XV{A pɔgelee serɛ e la nenseɛbaa.} \XE{The young woman's husband is middle aged.} \SN \GE{a fairly big person (size)} \OP{nenseɛbaare} \TP{nenseɛbaarɛɛ} 
\LX{nenselɛ} \HM{} \PH{nɪ̀nsɪ̊lɛ} \PS{interj.~} \GE{great, well done} \VA{menselɛ}
\LX{nensenyɛrɛ} \HM{} \PH{nɪ́nsɪ́ɲɛ́rɛ̀} \PS{n.~} \GE{one who sees spirits} \OP{nensenyɛrebɛ} \TP{nensenyɛrebɛɛ} 
\LX{nensɔglaa} \HM{} \PH{nɪ́nsɔ́gláá} \PS{n.~} \GE{black person} \OP{nensɔglɔ} 
\LX{nensoŋ} \HM{} \PH{nɪ́nsʊ́ŋ} \PS{n.~} \GE{good person, holy person,} \OP{nensonne} \TP{nensonnɛɛ} 
\LX{nentaare} \HM{} \PH{nɪ́ntáárɪ̀} \PS{n.~} \GE{saliva} \TP{nentaarɛɛ} 
\LX{nentaboli} \HM{} \PH{nɪ́ntábòlí} \PS{n.~} \GE{large glob of saliva} \OP{nentabolo} \TP{nentabolee} 
\LX{nentage} \HM{} \PH{nɪ́ntágɪ́} \PS{n.~} \SN \GE{a thickset person} \SN \GE{person of substance; wealthy} \OP{nentagere} \TP{nentagerɛɛ} 
\LX{nentage} \HM{} \PH{nɪ́ntàgɪ̀} \PS{n.~} \SN \GE{act of tugging e.g in a tussle} \SN \GE{the act of taking people to court} 
\LX{nentakyeɛrɛ} \HM{} \PH{nɪ́ntáʧɪɛrɛ} \PS{n.~} \SN \GE{dried saliva around the mouth} \SN \GE{dried tears on the cheek} \OP{nentakyeɛre} \TP{nentakyeɛrɛɛ} 
\LX{nentaŋ} \HM{} \PH{nɪ́ntáŋ} \PS{n.~} \GE{tears} \OP{nentanne} \TP{nentannɛɛ} \VA{ŋmentɔŋ, ŋmentɔnnɛɛ}
\LX{nentegeroŋ} \HM{} \PH{nɪ́ntɪ́grʊ́ŋ} \PS{n.~} \GE{important person; person of worth} \OP{nentegerɛ} 
\LX{nenweɛ} \HM{} \PH{nɪ́nwɪ́ɛ́} \PS{n.~} \SN \GE{a person who is not related to another} \SN \GE{unwanted person} \OP{neweere} \TP{nenweerɛɛ} 
\LX{nɛnzeɛ} \HM{} \PH{nɛ́nzɪ̀ɛ́} \PS{n.~} \GE{lean meat} \SG{nɛnzeroo} \OP{nɛnzeere} \TP{nɛnzeerɛɛ} 
\LX{nɛre} \HM{} \PH{nɛ̀rɪ̀} \PS{v.~} \GE{to be buried underneath e.g. in the soil} \PD{nɛrɛɛ, nɛrɛ, nɛrɛ, nɛrebɛ, nɛraa, - } 
\LX{nerroŋ} \HM{} \PH{nɪ́rrʊ́ŋ} \PS{n.~} \GE{consanguinity, blood ties, blood relationship} \XV{A ba seŋ ka naasaalaa toma kaara nerroŋ.} \XE{Government employment should not be based on blood ties.} 
\LX{nibiri} \HM{} \PH{nìbirì̀} \PS{n.~} \GE{emotion} \TP{nibiree} 
\LX{nii} \HM{} \PH{níì} \PS{adj.~} \GE{eight} 
\LX{nimibaaloŋ} \HM{} \PH{nímíbààlʊ̂ŋ} \PS{n.~} \GE{kindness} 
\LX{nimibaare} \HM{} \PH{nímíbáárɪ́} \PS{n.~} \GE{bloodshot} \TP{nimibaarɛɛ} \VA{nimibeɛmɛ}
\LX{nimibiri} \HM{} \PH{nímíbírì} \PS{n.~} \GE{pupil (of eye)} \SG{nimibiruu} \OP{nimibie} \TP{nimibiiree} 
\LX{nimibuli} \HM{} \PH{nímíbúlí} \PS{n.~} \GE{eye with a cataract} \OP{nimibulo} \TP{nimibulee} 
\LX{nimifaa} \HM{} \PH{nímífáá} \PS{n.~} \GE{evil eye} \OP{nimifaare} \TP{nimifaarɛɛ} 
\LX{nimifaa} \HM{} \PH{nímífáá} \PS{n.~} \GE{wrong side (of fabric)} \OP{nimifaare} \TP{nimifaarɛɛ} 
\LX{nimifuri} \HM{} \PH{nímífùrì} \PS{v.~} \SN \GE{to be wide eyed} \SN \GE{to show wit} \PD{-, -, -, -, -, - } 
\LX{nimigane} \HM{} \PH{nímígánɪ̀} \PS{n.~} \GE{eyelid} \OP{nimigama} \TP{nimiganɛɛ} 
\LX{nimigbaŋkɔɔloŋ} \HM{} \PH{nímígbàŋkɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{eyebrow} 
\LX{nimigbɛlɛɛ} \HM{} \PH{nímígbɛ̀lɛ̀ɛ̀} \PS{n.~} \GE{the state of being considerate} \VA{nimimaaroŋ}
\LX{nimigboŋkogiri} \HM{} \PH{nímígbòŋkógrí} \PS{n.~} \GE{brow, the part of the face where the eyebrow is located} \OP{nimigboŋkogiro} \TP{nimigboŋkogiree} \VA{(1p) nimigboŋkobo}
\LX{nimikaare} \HM{} \PH{nímíkáárɪ̀} \PS{n.~} \GE{eyeballs} \OP{nimikaare} 
\LX{nimiko} \HM{} \PH{nímíkò} \PS{n.~} \GE{act of doing wicked things} 
\LX{nimikɔɔloŋ} \HM{} \PH{nímíkɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{eyelash} 
\LX{nimiko᷉ɔ} \HM{} \PH{nímíkʊ́᷉ɔ́᷉} \PS{n.~} \SN \GE{watery eyes} \SN \GE{tears} \TP{nimikoɔnɛɛ} 
\LX{nimikuoŋaa} \HM{} \PH{nímíkúóŋáá} \PS{n.~} \GE{wickedness} \OP{nimikuoni} \TP{nimikuonee} \VA{nimitiiruŋ}
\LX{nimikyaane} \HM{} \PH{nímíʧáánɪ́} \PS{n.~} \GE{eye glasses, spectacles} \OP{nimikyaane} \TP{nimikyaanaa} 
\LX{nimilaa} \HM{} \PH{nímíláá} \PS{n.~} \GE{eye socket} \OP{nimilaare} \TP{nimilaarɛɛ} 
\LX{nimimaaroŋ} \HM{} \PH{nímímáárʊ́ŋ} \PS{n.~} \GE{kindness, having pity on others} \OP{nimimaara} \TP{nimimaarɛɛ} 
\LX{niminyɛrɛ} \HM{} \PH{nímínyɛ́rɛ̀} \PS{n.~} \GE{visionary, seer} \OP{ niminyɛrebɛ} \TP{niminyɛrebɛɛ} 
\LX{nimipeɛnɛ} \HM{} \PH{nímípɪ́ɛ́nɛ́} \PS{n.~} \GE{one who claims to be wise} \OP{nimipeɛnemɛ} \TP{nimipeɛnemɛɛ} 
\LX{nimipe᷉ɛ} \HM{} \PH{nímípɪ̀᷉ɛ̀᷉} \PS{n.~} \GE{the act of claiming to be wise} 
\LX{nimipe᷉ɛ᷉} \HM{} \PH{nímípɪ́᷉ɛ́᷉} \PS{n.~} \GE{the act of showing oneself to be wiser than others} 
\LX{nimipɔge} \HM{} \PH{nímípɔ̀gɪ̀} \PS{n.~} \GE{the act of blindfolding or confusing others} 
\LX{nimipoŋi} \HM{} \PH{nímípôŋ} \PS{n.~} \GE{puss-like fluid on the surface of the eyeball} \OP{nimiponni} \TP{nimiponnee} \VA{nimipuuri}
\LX{nimipuuri} \HM{} \PH{nímípúúrí} \PS{n.~} \GE{puss-like fluid on the surface of the eyeball} \OP{nimipuuree} \VA{nimipoŋi}
\LX{nimiri} \HM{} \PH{nímírì} \PS{n.~} \SN \GE{eye} \SN \GE{stinginess; avarice} \XV{N ba taa la libie kyɛ nimiri koŋ sage ka o da tiivi paalaa.} \XE{My father has the money but his stinginess would not allow him to buy a new TV.} \SN \GE{covetousness} \XV{A bie ŋa taa la nimiri, o maŋ nyɛ la o tɔ zaa bone boɔrɔ o.} \XE{This child is covetous; he wants anything that another child has.} \SG{nimiruu} \OP{nimie} \TP{nimiiree} 
\LX{nimiri} \HM{} \PH{nímírì} \PS{n.~} \GE{success} \XV{O toma taa la nimiri pampana.} \XE{That job of his is becoming fruitful.} \SG{nimiruu} \OP{nimie} \TP{nimiiree} 
\LX{nimiri-soba} \HM{} \PH{nímírì-sʊ̀bá} \PS{n.~} \SN \GE{miser} \SN \GE{one who covets} \OP{nimiri-deme} \TP{nimi-demee} 
\LX{nimisogɔ} \HM{} \PH{nímísʊ́gɔ́} \PS{n.~} \GE{face} \XV{A pɔge kaa la a o bibaala nimisogɔ kyɛ viiri zu.} \XE{The woman looked into her sick child's face and shook her head.} \OP{nimisogere} \TP{nimisogerɛɛ} 
\LX{nimisogɔ} \HM{} \PH{nímísʊ́gɔ́} \PS{adv.~} \GE{in front of, ahead of} \XV{A gane biŋ la n nimisogɔ.} \XE{The book is placed in front of me.} 
\LX{nimisoɔ} \HM{} \PH{nímísʊ́ɔ̀} \PS{n.~} \GE{dizziness, confusion} \OP{nimisoɔre} \TP{nimisoɔrɛɛ} 
\LX{nimitiiruŋ} \HM{} \PH{nímítììrúŋ} \PS{n.~} \GE{wickedness} \VA{nimikuoŋaa}
\LX{nimitoloŋ} \HM{} \PH{nímítʊ́lʊ́ŋ} \PS{n.~} \GE{haste} 
\LX{nimitɔɔre} \HM{} \PH{nímítɔ́ɔ́rɪ́} \PS{prep.~} \SN \GE{ahead, in front of} \XV{Ka fooŋ mɔnɔ lɔɔre a seŋ ka fo kaara nimitɔɔre.} \XE{If you drive you must look ahead.} \SN \GE{forward (distance, direction)} \XV{O yeli ka N gaa nimitɔɔre fe᷉e᷉ kyɛ leɛ tu nudoloŋ.} \XE{She directed that I should go ahead a little and turn right.} 
\LX{nimiyuo} \HM{} \PH{nímíyúò} \PS{n.~} \GE{an action that creates the impression that one is more intelligent than another} \XV{Te naŋ gaa a nɛne pommu Bayuo da boɔrɔ ka o wuli nimiyuo kyɛ ka te ba sage.} \XE{When we went to share the meat Bayuo wanted to cheat but we did not give him the chance.} 
\LX{nimiyuoro} \HM{} \PH{nímíyúóró} \PS{n.~} \GE{well-informed person, one who is abreast of the times} \OP{nimiyuoribo} \TP{nimiyuoribee} 
\LX{nimizeɛ} \HM{} \PH{nímízɪ̀ɛ́} \PS{n.~} \GE{seriousness} \OP{nimizeere} \TP{nimizeerɛɛ} 
\LX{niŋe} \HM{} \PH{níŋè} \PS{n.~} \GE{face} \XV{A bi-entuo na ba maŋ boɔrɔ na pɛge o niŋe baguo.} \XE{That lazy child does not like washing his face in the morning.} \OP{ninni} \TP{ninnee} \SE{niŋe sógɔ́ --in front, before, niŋesogre, niŋesogrɛɛ} \GE{in front, before} 
\LX{niŋe} \HM{} \PH{níŋé} \PS{adv.~} \GE{in front of, ahead of} \XV{O de la a kogi wa biŋ a saana niŋeŋ.} \XE{He placed the seat in front of the visitor.} 
\LX{nirii} \HM{} \PH{nìríì} \PS{n.~} \GE{a type of creeping plant with edible seeds} \TP{nirree} 
\LX{ŋma} \HM{} \PH{ŋmá} \PS{v.~} \GE{to conquer, to plunder} \PD{ŋmaɛ, ŋmaara, ŋmaara, ŋmaareba, ŋmaaraa, - } 
\LX{ŋma zirii} \HM{} \PH{ŋmàà zìríì} \PS{v.~} \GE{to lie} \XV{O ŋmaa la zirii ka o baare la a toma.} \XE{He lied that he had finished the work.} \PD{-, -, -, -, -, - } 
\LX{ŋmaa} \HM{} \PH{ŋmàà} \PS{v.~} \GE{to cut} \XV{Nɛne la ka a pɔge da ŋmaara ka a soɔ ŋmaa o nubiri.} \XE{The woman was cutting meat when the knife cut her finger.} \PD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\LX{ŋmaa} \HM{} \PH{ŋmàà} \PS{v.~} \GE{to write} \XV{Ayuo ŋmaa la sɛnsɛŋe gane gaa ko o sɛne.} \XE{Ayuo has written a love letter to her boyfriend.} \PD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\LX{ŋmaa} \HM{} \PH{ŋmàà} \PS{v.~} \SN \GE{to cross (river)} \XV{A mane paalɛɛ la ka ba maŋ de gbori ŋmaa ne.} \XE{The river is flooded and they are using a boat to cross it.} \SN \GE{to go across to meet} \XV{Polisiri ŋmaa la a sori nyɔgere a nannyigiri.} \XE{The police crossed the road and arrested the thieves.} \PD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\LX{ŋmaa} \HM{} \PH{ŋmàà} \PS{v.~} \GE{to prevent} \PD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\LX{ŋmaa} \HM{} \PH{ŋmáá} \PS{adj.~} \GE{short} \XV{O serɛ la a dɔɔ ŋmaa na.} \XE{That short man is her husband.} \OP{ŋmaara} 
\LX{ŋmaa bare} \HM{} \PH{ŋmàà bàrɪ̀} \PS{v.~} \GE{to exclude; to put aside} \PD{ŋmaa barɛɛ, ŋmaara bara, ŋmaabara, ŋmaabareba, -, - } 
\LX{ŋmaa seɛ} \HM{} \PH{ŋmàá sɪ́ɛ́} \PS{v.~} \GE{to conceive, to be pregnant} \PD{ŋmaaɛ seɛ, ŋmaara seɛ, se-ŋmaara, se-ŋmareba, -, - } 
\LX{ŋmaa vooroŋ} \HM{} \PH{ŋmàà vʊ́ʊ́rʊ́ŋ} \PS{v.~} \GE{to stop breathing, to die} \PD{-, -, -, -, -, - } 
\LX{ŋmaabini} \HM{} \PH{ŋmàábíní} \PS{n.~} \GE{a type of plant} \OP{ŋmaabinee} 
\LX{ŋmaadaa} \HM{} \PH{ŋmààdáá} \PS{n.~} \GE{mature big male monkey} \OP{ŋmaadaare} \TP{ŋmaadarɛɛ} 
\LX{ŋmaadasaa} \HM{} \PH{ŋmààdàsàà} \PS{n.~} \GE{a huge monkey} \OP{ŋmaadasaare} \TP{ŋmaadasaarɛɛ} 
\LX{ŋmaakalaa} \HM{} \PH{ŋmààkàlàá} \PS{n.~} \GE{small but smart type of monkey} \OP{ŋmaakalle} \TP{ŋmaakalɛɛ} 
\LX{ŋmaakyeraa} \HM{} \PH{ŋmààʧɪ̀ràá} \PS{n.~} \SN \GE{a type of tree with medicinal properties} \SN \GE{a type of small monkey} \OP{ŋmaakyerre} \TP{ŋmaakyerɛɛ} 
\LX{ŋmaaŋa} \HM{} \PH{ŋmááŋà} \PS{n.~} \GE{monkey} \OP{ŋmaane} \TP{ŋmaanɛɛ} 
\LX{ŋmaapare} \HM{} \PH{ŋmààpárɪ́} \PS{n.~} \GE{name of a popular market in Teza, a village in the Upper West region of Ghana} \TP{ŋmaaparɛɛ} 
\LX{ŋmaaraa} \HM{} \PH{ŋmààráà} \PS{n.~} \GE{an evil spirit with powers of witchcraft, usually in the form of an animal that haunts people} \OP{ŋmaare} \TP{ŋmaarɛɛ} \VA{vuuluŋ}
\LX{ŋmaasugo} \HM{} \PH{ŋmààsúgó} \PS{n.~} \GE{band of monkeys} \OP{ŋmaasugri} \TP{ŋmaasugree} 
\LX{ŋmaataŋaa} \HM{} \PH{ŋmààtàŋáá} \PS{n.~} \GE{a type of tree whose fruits are eaten by birds and animals} \OP{ŋmaataŋema} \TP{ŋmaatannɛɛ} 
\LX{ŋmaateɛ} \HM{} \PH{ŋmààtɪ́ɛ́} \PS{n.~} \GE{a type of tree} \OP{ŋmaateere} \TP{ŋmaateerɛɛ} 
\LX{ŋmaazeɛ} \HM{} \PH{ŋmààzɪ̀ɛ̀} \PS{n.~} \GE{a type of monkey} \OP{ŋmaazeere} \TP{ŋmaazeerɛɛ} 
\LX{ŋmalaŋmeɛ} \HM{} \PH{ŋmáláŋmɪ́ɛ́} \PS{n.~} \GE{a type of  bird with thin long legs and a long beak} \OP{ŋmalaŋmeere} \TP{ŋmalaŋmeerɛɛ} \VA{ŋmalanweɛ}
\LX{ŋmamberɛ} \HM{} \PH{ŋmámbɪ́rɛ́} \PS{n.~} \GE{a type of small weed with white flowers and medicinal properties} \SG{ŋamberoo} \OP{ŋmambere/ŋmamberoo} \TP{ŋamberɛɛ} 
\LX{ŋmammo} \HM{} \PH{ŋmámmʊ́} \PS{n.~} \GE{a type of cake made from millet flour} \OP{ŋmama} \TP{ŋmamɛɛ} 
\LX{ŋmampapaa} \HM{} \PH{ŋmámpàpáà} \PS{n.~} \GE{a type of pigeon with red eyes} \OP{ŋmampapaare} \TP{ŋmampapaarɛɛ} 
\LX{ŋmampelaa} \HM{} \PH{ŋmámpɪ̀láá} \PS{n.~} \GE{wild grey pigeon} \OP{ŋmampeɛle} \TP{ŋmampeɛlɛɛ} 
\LX{ŋmampelaa} \HM{} \PH{ŋmàmpɪ̀làá} \PS{n.~} \GE{neat calabash used for serving water or poridge} \OP{ŋmampeɛle} \TP{ŋmampeɛlɛɛ} 
\LX{ŋmamporɔ} \HM{} \PH{ŋmàmpʊ́rɔ́} \PS{n.~} \GE{a red round insect} \OP{ŋmamporre} \TP{ŋmamporɛɛ} \VA{ŋmampora}
\LX{ŋmamporezupelaa} \HM{} \PH{ŋmǎmpʊ̀rzúpɪ́làá} \PS{n.~} \GE{a character in Dagaaba folktales which has supernatural power and which it uses to test the ego of other characters} \OP{ŋmamporzupeɛle} \VA{ŋmamporzuzeɛ}
\LX{ŋmamporzuzeɛ} \HM{} \PH{ŋmàmpʊ́rzúzɪ̀ɛ́} \PS{n.~} \GE{} \VA{ŋmamporzupelaa}
\LX{ŋmandɛre} \HM{} \PH{ŋmándɛ́rè} \PS{n.~} \GE{spider} \OP{ŋmandɛɛ} \TP{ŋmandɛrɛɛ} 
\LX{ŋmane} \HM{} \PH{ŋmánɪ́} \PS{n.~} \SN \GE{calabash, bottle gourd} \SN \GE{calabash plant} \OP{ŋmama} \TP{ŋmannɛɛ} 
\LX{ŋmanee} \HM{} \PH{ŋmánɪ́ɪ̀} \PS{n.~} \GE{dove} \OP{ŋmama} \TP{ŋmannɛɛ} 
\LX{ŋmaŋgbɛre} \HM{} \PH{ŋmàŋgbɛ́rɪ́} \PS{n.~} \GE{vine of the calabash plant} \OP{ŋmaŋgbɛɛ} \TP{ŋmaŋgbɛɛrɛɛ} 
\LX{ŋmaŋkuŋkummu} \HM{} \PH{ŋmàŋkùŋkùmmú} \PS{n.~} \GE{calabash seed in a ball form} \OP{ŋmaŋkuŋkumo} \TP{ŋmaŋkuŋkunni} 
\LX{ŋmantummu} \HM{} \PH{ŋmǎntúmmú} \PS{n.~} \GE{unpolished calabash} \OP{ŋmantumo} \TP{ŋmantumee} 
\LX{ŋmantuo} \HM{} \PH{ŋmàntúó} \PS{n.~} \GE{bitter calabash} \OP{ŋmantuuri} \TP{ŋmantuuree} 
\LX{ŋmanzeɛ} \HM{1} \PH{ŋmǎnzɪ̀ɛ́} \PS{n.~} \GE{a type of dove with pink plumage} \SG{ŋmanzeroo} \OP{ŋmanzeere} \TP{ŋmanzeerɛɛ} 
\LX{ŋmanzeɛ} \HM{2} \PH{ŋmǎnzɪ̀ɛ́} \PS{n.~} \GE{calabash used in the preparation of oil} \OP{ŋmanzeere} \TP{ŋmanzeerɛɛ} 
\LX{ŋmaraa} \HM{} \PH{ŋmáràà} \PS{n.~} \SN \GE{moon} \SN \GE{month} \OP{ŋmarre} \TP{ŋmarrɛɛ} 
\LX{ŋmare} \HM{} \PH{ŋmárɪ̀} \PS{v.~} \SN \GE{to break e.g. pot} \SN \GE{to crush} \PD{ŋmarɛɛ, ŋmara, ŋmara, ŋmareba, ŋmaraa, ŋmareba } 
\LX{ŋmare} \HM{} \PH{ŋmárɪ́} \PS{v.~} \GE{to change money into smaller units} \PD{ŋmarɛɛ, ŋmara, ŋmara, ŋmareba, ŋmaraa, ŋmareba } 
\LX{ŋmarebiri} \HM{} \PH{ŋmárɪ́bírì} \PS{n.~} \GE{star} \SG{ŋmarebiruu} \OP{ŋmarebie} \TP{ŋmarebiiree} 
\LX{ŋmaroo} \HM{} \PH{ŋmárʊ̀ʊ̀} \PS{n.~} \GE{change (money)} \SG{ŋmarrɛɛ} \TP{ŋmarre} 
\LX{ŋmɛ} \HM{} \PH{ŋmɛ̀} \PS{v.~} \SN \GE{to beat, to hit, to strike} \SN \GE{to clap, to applaud} \PD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \SE{ŋmɛ kpɛlle} \GE{shout; scream} \SE{ŋmɛ gbelime} \GE{ring (bell)} 
\LX{ŋmɛ} \HM{} \PH{ŋmɛ̀} \PS{v.~} \GE{to thresh} \XV{Dagaaba maŋ ŋmɛ kyi kyɛ baŋ yeɛle.} \XE{Among the Dagaaba, millet is threshed before being winnowed out.} \PD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \SE{ŋmɛ kpɛlle} \GE{shout; scream} \SE{ŋmɛ gbelime} \GE{ring (bell)} 
\LX{ŋmɛ} \HM{} \PH{ŋmɛ̀} \PS{v.~} \GE{to bump into, to knock against} \XV{A lɔɛ ama la za᷉a᷉ ŋmɛ taa na.} \XE{These are the two cars that collided yesterday.} \PD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \SE{ŋmɛ kpɛlle} \GE{shout; scream} \SE{ŋmɛ gbelime} \GE{ring (bell)} 
\LX{ŋmɛ} \HM{} \PH{ŋmɛ̀} \PS{v.~} \GE{to fire, to shoot (gun)} \XV{Ba ŋmɛ la malefa ka a nuuli iri agere.} \XE{When they fired the gun, the birds flew away.} \PD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \SE{ŋmɛ kpɛlle} \GE{shout; scream} \SE{ŋmɛ gbelime} \GE{ring (bell)} 
\LX{ŋmɛ} \HM{} \PH{ŋmɛ̀} \PS{v.~} \GE{to begin to hatch} \PD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \SE{ŋmɛ kpɛlle} \GE{shout; scream} \SE{ŋmɛ gbelime} \GE{ring (bell)} 
\LX{ŋmɛ} \HM{} \PH{ŋmɛ̀} \PS{v.~} \GE{to apply by dusting (powder)} \PD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \SE{ŋmɛ kpɛlle} \GE{shout; scream} \SE{ŋmɛ gbelime} \GE{ring (bell)} 
\LX{ŋmɛ ɔge} \HM{} \PH{ŋmɛ̀ ɔ̀gɪ́} \PS{v.~} \SN \GE{to burrow to make a nest} \XV{Soɔŋaa maŋ ŋmɛ la ɔge kpɛ gaŋ be.} \XE{The rabbit digs a warren where it sleeps.} \SN \GE{to settle down in one place} \XV{A pɔlebile nyɛ la n yɔɔpuulee a deɛ ŋmɛ te yiriŋ.} \XE{The young man saw my sister and has settled down in our house.} \PD{-, -, -, -, -, - } 
\LX{ŋmɛ sane} \HM{} \PH{ŋmɛ̀ sánɪ̀} \PS{v.~} \SN \GE{to make a net loss} \XV{Ka fooŋ da kommie sidi kɔɔ kyɛ te koɔre sidi lezaɛ anaare fo ŋmɛ la sane sidi lezare.} \XE{If you buy tomatoes for one hundred cedis and sell them for eighty cedis, you have made a loss of twenty cedis.} \SN \GE{to spend money on something that is not of much worth to oneself} \XV{Ka kuoriŋ ko fo yiriŋ fo maŋ ŋmɛ la sane.} \XE{When there is a funeral in your house you normally spend a lot.} \PD{-, -, -, -, -, - } 
\LX{ŋmee} \HM{} \PH{ŋmɪ́ɪ̀} \PS{v.~} \GE{to crush grains out of fresh heads of millet (usually after roasting)} \PD{ŋmeɛɛ, ŋmeɛnɛ, ŋmeɛnɛ, ŋmeɛnemɛ, ŋmenaa, - } \VA{ŋmeɛ}
\LX{ŋmeɛ} \HM{} \PH{ŋmɪ́ɛ} \PS{v.~} \GE{to crush} \PD{ŋmeɛɛ, ŋmeɛnɛ, ŋmeɛnɛ, ŋmeɛnemɛ, ŋmeɛnaa, - } \VA{ŋmee}
\LX{ŋmeɛ} \HM{} \PH{ŋmɪ́ɛ} \PS{adj.~} \GE{scorched, withered e.g. green vegetation turning brown} 
\LX{ŋmeɛle} \HM{} \PH{ŋmɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{to twist; to wring e.g. hand} \PD{ŋmeɛlɛɛ, ŋmeɛlɛ, ŋmeɛlɛ, ŋmeɛlebɛ, ŋmeɛlaa, ŋmeɛlebɛ } \VA{nyɛ}
\LX{ŋmeɛle} \HM{} \PH{ŋmɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{to curl out, to ooze (smoke)} \XV{Nyoore ŋmeɛlɛ la a sege poɔ, ba mare la vu᷉u᷉.} \XE{Smoke is oozing from the hut, an indication that they have made fire.} \PD{ŋmeɛlɛɛ, ŋmeɛlɛ, ŋmeɛlɛ, ŋmeɛlebɛ, ŋmeɛlaa, ŋmeɛlebɛ } \VA{nyɛ}
\LX{ŋmeɛle} \HM{} \PH{ŋmɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{to pass gas, to break wind, to fart} \XV{A bie poɔ pie la ka o deɛ maŋ ŋmeɛle bimporɔ bare.} \XE{The child has a bloated stomach and he has terrible gas.} \PD{ŋmeɛlɛɛ, ŋmeɛlɛ, ŋmeɛlɛ, ŋmeɛlebɛ, ŋmeɛlaa, ŋmeɛlebɛ } \VA{nyɛ}
\LX{ŋmeɛle} \HM{} \PH{ŋmɪ́ɛ́lɪ́} \PS{adj.~} \GE{twisted} \XV{A nɔ pare ŋmeɛle na koŋ la nyɛ.} \XE{That fowl with the twisted tail will not lay eggs again.} 
\LX{ŋmeene} \HM{} \PH{ŋmɪ̀ɪ̀nɪ̀} \PS{v.~} \GE{to become thin or small} \PD{ŋmeenɛɛ, ŋmeenɛ, ŋmeenɛ, ŋmeenemɛ, ŋmeenaa, - } 
\LX{ŋmeene} \HM{} \PH{ŋmɪ́ɪ́nɪ́} \PS{adj.~} \GE{lean} \OP{ŋmeemɛ} 
\LX{ŋmeeo} \HM{} \PH{ŋmɪ́ɪ́ʊ́} \PS{n.~} \GE{wicked and secret act} 
\LX{ŋmɛgele} \HM{} \PH{ŋmɛ̀gɪ̀lɪ̀} \PS{v.~} \SN \GE{to flatten} \XV{Kuuri ka o de tɔ ne a koŋkoŋ te ŋmɛgele o bare.} \XE{She hit the tin with a stone until she flattened it out.} \SN \GE{to suffer from hunger} \XV{Zaameŋ zaa ba ba maale diibu ka a bibiiri deɛ ŋmɛgelɛ.} \XE{They have cooked food yesterday and the children are suffering from hunger.} \PD{ŋmɛglɛɛ, ŋmɛglɛ, ŋmɛglɛ, ŋmɛglebɛ, ŋmɛglaa, ŋmɛglebɛ } 
\LX{ŋmɛgelɛŋ} \HM{} \PH{ŋmɛ́gɪ́lɛ́ŋ} \PS{adj.~} \GE{flat e.g. run-over container} 
\LX{ŋmɛgeloo} \HM{} \PH{ŋmɛ́gɪ́lʊ́ʊ́} \PS{adj.~} \GE{flattened out e.g. tin run over by a car} \OP{ŋmɛgelɛ} 
\LX{ŋmegere} \HM{} \PH{ŋmɪ́gɪrɪ́} \PS{v.~} \GE{to shrug one's shoulders} \XV{Ba toŋ la a bie ka o ŋmegere o bɔgeŋmampegɛ kyɛ zagere.} \XE{When they sent the child, he shrugged his shoulders and refused to go.} \PD{ŋmegrɛɛ, ŋmegrɛ, ŋmegrɛ, ŋmegrebɛ, ŋmegraa, - } 
\LX{kɔŋkɔre} \HM{} \PH{kɔ̀ŋkɔ̀rɪ́} \PS{n.~} \GE{forest} \VA{tetige}
\LX{koɔle sɔgelaa} \HM{} \PH{kʋ́ɔ́lɪ́sɔ̀gɪ́làá} \PS{n.~} \GE{fertile land} 
\LX{ŋmegere} \HM{} \PH{ŋmɪ́gɪrɪ́} \PS{v.~} \GE{to swallow with difficulty} \XV{A bɔbɔe ŋmegere la a pɔntere eŋ.} \XE{The duck was able to swallow the toad.} \PD{ŋmegrɛɛ, ŋmegrɛ, ŋmegrɛ, ŋmegrebɛ, ŋmegraa, - } 
\LX{ŋmɛgere} \HM{} \PH{ŋmɛ̀gɪ̀rɪ̀} \PS{v.~} \GE{to press, to squeeze out} \XV{Ba ŋmɛgere la sunni yi a teɛ ŋa poɔ.} \XE{They have been able to squeeze some gum out of this tree.} \PD{ŋmɛgrɛɛ, ŋmɛgrɛ, ŋmɛgrɛ, ŋmɛgrebɛ, ŋmɛgraa, - } 
\LX{ŋmɛgere} \HM{} \PH{ŋmɛ̀gɪ̀rɪ̀} \PS{v.~} \GE{to weep inwardly, to suffer inwardly} \XV{Ba ŋmɛ la a bie kyɛ dabe᷉ɛ᷉ ba sage ka o koŋ ka o fa᷉a᷉ kyɛ ŋmɛgerɛ.} \XE{The child was beaten and, being afraid to cry, he wept inwardly.} \PD{ŋmɛgrɛɛ, ŋmɛgrɛ, ŋmɛgrɛ, ŋmɛgrebɛ, ŋmɛgraa, - } 
\LX{ŋmɛleŋmɛle} \HM{} \PH{ŋmɛ̀lɪ̀ŋmɛ́lɪ́} \PS{v.~} \GE{to writhe in pain} \XV{O ŋmɛ la a waabo ka o ŋmɛleŋmɛle.} \XE{He hit the snake and it writhed.} \PD{-, -, -, -, -, - } 
\LX{ŋmɛmɛɛ} \HM{} \PH{ŋmɛ̀mɛ̀ɛ̀} \PS{adj.~} \GE{flat and twisted e.g. nose} 
\LX{ŋmɛmm} \HM{} \PH{ŋmɛ́mm} \PS{adv.~} \GE{stand against something quietly (hide)} 
\LX{ŋmenaa} \HM{} \PH{ŋmɪ́nàá} \PS{n.~} \SN \GE{sun} \SN \GE{noon} \SN \GE{daytime} 
\LX{ŋmenaa-diibu} \HM{} \PH{ŋmɪ́náá-dííbú} \PS{n.~} \GE{lunch} 
\LX{ŋmendaare} \HM{} \PH{ŋmɪ́ndáárɪ́} \PS{n.~} \GE{Sabbath} \OP{ ŋmendɛɛ} 
\LX{ŋmendie} \HM{} \PH{ŋmɪ̀ndìè} \PS{n.~} \GE{church} \OP{ŋmenderi} \TP{ŋmenderee} 
\LX{ŋmene} \HM{} \PH{ŋmɪ́nɪ̀} \PS{n.~} \GE{god, spirit} \OP{ŋmemɛ} \TP{ŋmennɛɛ} 
\LX{ŋmenɛŋmenɛ} \HM{} \PH{ŋmɪ́nɛ́ŋmɪ́nɛ́} \PS{adj.~} \GE{firmly and beautifully woven} 
\LX{ŋmɛŋgɛlɛŋ} \HM{} \PH{ŋmɛ́ŋgɛ́lɛ́ŋ} \PS{adj.~} \GE{flattened out of shape} 
\LX{ŋmɛnne} \HM{} \PH{ŋmɛ́nnɪ̀} \PS{v.~} \SN \GE{to squeeze with the fingers e.g. breastfeeding} \SN \GE{to press to feel (swollen leg)} \PD{ŋmɛnnɛɛ, ŋmɛnnɛ, ŋmɛnnɛ, ŋmɛnemɛ, ŋmɛnnaa, ŋmɛnemɛ } 
\LX{ŋmentantuo} \HM{} \PH{ŋmɪ́ntántúó} \PS{n.~} \GE{hot or bright afternoon} \OP{ŋmentantuuri} \TP{ŋmentantuuree} 
\LX{ŋmentɔmpɔle} \HM{} \PH{ŋmɪ́ntɔ̀mpɔ́lɪ̀} \PS{n.~} \GE{late morning} \OP{ŋmentɔmpɔlɔ} 
\LX{ŋmentɔŋ} \HM{} \PH{ŋmɪ́ntɔ̀ŋ} \PS{n.~} \GE{sunshine, daylight} \OP{ŋmentɔnne} \TP{ŋmentɔnnɛɛ} 
\LX{ŋmentɔŋgɔŋ} \HM{} \PH{ŋmɪ́ntɔ̀ŋgɔ́ŋ} \PS{n.~} \GE{early afternoon (between 1 and 2 pm)} \OP{ŋmentɔŋgɔnne} \TP{ŋmentɔŋgɔnnɛɛ} 
\LX{ŋmentɔŋkpeɛraa} \HM{} \PH{ŋmɪ́ntɔ̀ŋkpɪ́ɛ́ráá} \PS{n.~} \GE{sunset} \OP{ŋmentɔŋkpeɛre} \TP{ŋmentɔŋkpeɛrɛɛ} 
\LX{ŋmen-yɛlɛ} \HM{} \PH{ŋmɪ́n-yɛ́lɛ́} \PS{n.~} \GE{work for God} 
\LX{ŋmen-yelyelli} \HM{} \PH{ŋmɪ́n-yèlyèllì} \PS{v.~} \GE{to divine, to prophesy} \PD{-, -, -, -, -, - } 
\LX{ŋmenzone} \HM{} \PH{ŋmɪ́nzʊ́nɪ́} \PS{n.~} \GE{blasphemy} \OP{ ŋmenzomɔ} \VA{ŋmenzoma}
\LX{ŋmeregɛ} \HM{} \PH{ŋmírɪ́gɛ́} \PS{adv.~} \GE{totally} \XV{Ba toŋ la a bi entuo na ka o zagere ŋmeregɛ.} \XE{They sent the lazy boy and he refused totally.} 
\LX{ŋmeɛlaa} \HM{} \PH{ŋmɪ́ɛ́láá} \PS{n.~} \GE{skipping rope} \OP{ŋmeɛle} \TP{ŋmeɛlɛɛ} 
\LX{ŋmobofo} \HM{} \PH{ŋmʊ̀bʊ̀fʊ̀} \PS{adj.~} \GE{big and ruffled} 
\LX{ŋmolle} \HM{} \PH{ŋmʊ̀llɪ̀} \PS{v.~} \GE{to stuff into the mouth (food)} \PD{ŋmollɛɛ, ŋmolla, ŋmolla, ŋmoleba, ŋmollaa, - } 
\LX{ŋmolɔŋŋmooraa} \HM{} \PH{ŋmolɔŋŋmooraa} \PS{n.~} \GE{a type of insect that feeds on yam leaves} \OP{ŋmolɔŋŋmoore} \TP{ŋmolɔŋŋmoorɛɛ} \VA{ŋmolaŋŋmooraa}
\LX{ŋmonne} \HM{} \PH{ŋmʊ́nnɪ́} \PS{v.~} \SN \GE{to make spotted} \SN \GE{to beautify} \PD{ŋmonnɛɛ, ŋmonnɔ, ŋmonnɔ, ŋmoneba, ŋmonnaa, - } 
\LX{ŋmonɔŋmonɔ} \HM{} \PH{ŋmʊ́náŋmʊ́ná} \PS{adj.~} \SN \GE{spotted} \SN \GE{beautiful} \VA{ŋmonaŋmona}
\LX{ŋmoo} \HM{} \PH{ŋmʊ́ʊ̀} \PS{v.~} \GE{to become scorched (green vegetation)} \PD{ŋmooɛ, ŋmoona, -, ŋmoonema, ŋmoonaa, - } 
\LX{ŋmoone} \HM{} \PH{ŋmʊ̀ʊ̀ne} \PS{v.~} \SN \GE{to mutter, to grumble} \SN \GE{to growl e.g. dog} \PD{ŋmoonɛɛ, ŋmoona, ŋmoona, ŋmoonema, ŋmoonaa, ŋmoonema } 
\LX{ŋmore} \HM{} \PH{ŋmʊ́rɪ́} \PS{v.~} \GE{to crush} \XV{Ka aloopeleeŋ le o maŋ ŋmore la o zaa.} \XE{An aircraft that crashes is always crushed completely.} \PD{ŋmorɛɛ, ŋmora, ŋmora, ŋmoreba, ŋmoraa, ŋmoreba } 
\LX{ŋmore} \HM{} \PH{ŋmʊ́rɪ́} \PS{v.~} \GE{to rush; to move suddenly and fast} \XV{A bie lere la ka o ma ŋmore te nyɔge.} \XE{The child was going to fall, but the mother quickly took hold of him.} \PD{ŋmorɛɛ, ŋmora, ŋmora, ŋmoreba, ŋmoraa, ŋmoreba } 
\LX{no} \HM{} \PH{nʊ̀} \PS{v.~} \GE{to grind a object by rubbing it against a surface} \PD{noɛ, noona, noona, noonema, noonaa, - } 
\LX{nɔbɛraa} \HM{} \PH{nɔ́bɛ́ráá} \PS{n.~} \GE{irritableness} \OP{nɔbɛrre} \TP{nɔbɛrɛɛ} 
\LX{nɔbinnaa} \HM{} \PH{nɔ́bìnnáá} \PS{n.~} \GE{agreement} \OP{nɔbinni} 
\LX{nɔbirime} \HM{} \PH{nɔ́bìrìmé} \PS{n.~} \GE{fowl with stunted feathers} \OP{nɔbirimeri} \TP{nɔbirimee} 
\LX{nɔbɔŋɔ} \HM{} \PH{nɔ́bɔ́ŋɔ́} \PS{n.~} \GE{bit e.g. of a horse} \OP{nɔbɔnne} \TP{nɔbɔnnɛɛ} \VA{nɔbaŋa}
\LX{nɔboɔroŋ} \HM{} \PH{nɔ́bʊ́ɔ́rʊ́ŋ} \PS{n.~} \GE{troubling another person; irritation} 
\LX{nɔboyi} \HM{} \PH{nɔ́bʊ́yì} \PS{n.~} \GE{double standards} 
\LX{nɔdaa} \HM{} \PH{nɔ́dáá} \PS{n.~} \GE{rooster} \OP{nɔdaare} \TP{nɔdaarɛɛ} \VA{(1p) nɔraa}
\LX{nɔdaakommiruu} \HM{} \PH{nɔ́dáákómmírúú} \PS{n.~} \GE{a type of caterpillar, similar in appearance to a maggot, with a red head; typically found in the ground} \OP{nɔdaakommie} \TP{nɔdaakommiree} 
\LX{nɔdakpoŋ} \HM{} \PH{nɔ̀dàkpóŋ} \PS{n.~} \GE{capon} \OP{nɔdakponni} \TP{nɔdakpomo/nɔdakponnee} 
\LX{nɔdalee} \HM{} \PH{nɔ̀dàléé} \PS{n.~} \GE{cockerel} \OP{nɔdabilii} \TP{nɔdabillee} 
\LX{nɔdo} \HM{} \PH{nɔ́dó} \PS{n.~} \GE{food} 
\LX{nɔdɔɔne} \HM{} \PH{nɔ́dɔ̀ɔ̀nɪ́} \PS{n.~} \GE{that which is too little to satisfy (food)} \TP{nɔdɔɔnɛɛ} 
\LX{nɔ-ennaa} \HM{} \PH{nɔ́-ɪ̀nnáà} \PS{n.~} \GE{vow, promise} \OP{nɔ-enne} \TP{nɔ-ennɛɛ} 
\LX{nɔfaa} \HM{} \PH{nɔ́fáá} \PS{n.~} \GE{rudeness; foul mouth} \OP{nɔfaare} \TP{nɔfaarɛɛ} 
\LX{nɔfolee} \HM{} \PH{nɔ́fóléé} \PS{n.~} \GE{inability to keep a secret} 
\LX{nɔfo᷉ɔo} \HM{} \PH{nɔ́fʊ᷉ɔ́᷉ʊ́᷉} \PS{n.~} \GE{quick at talking} \OP{nɔfoɔne} \TP{nɔfoɔnɛɛ} 
\LX{nɔgbɛmbebe} \HM{} \PH{nɔ́gbɛ́mbébé} \PS{n.~} \GE{curse, foul language} 
\LX{nɔgbɛne} \HM{} \PH{nɔ́gbɛ́nɪ̀} \PS{n.~} \GE{lip} \SG{nɔgbɛnnoo} \OP{ nɔgbɛmɛ} \TP{nɔgbɛnnɛɛ} 
\LX{nɔgeraa} \HM{} \PH{nɔ́gráá} \PS{n.~} \GE{sojourner} \OP{nɔgere} \TP{nɔgerɛɛ} 
\LX{nɔgɔɔ} \HM{} \PH{nɔ́gɔ̀ɔ́} \PS{n.~} \GE{one-winged fowl} \OP{nɔgɔɔre} \TP{nɔgɔɔrɛɛ} 
\LX{nɔgoolɔ} \HM{} \PH{nɔ̀gʊ̀ʊ̀lɔ̀} \PS{n.~} \GE{moustache} \VA{nɔgoola}
\LX{nɔgo᷉o} \HM{} \PH{nɔ́gʊ̀᷉ʊ́᷉} \PS{n.~} \GE{moustache} \OP{nɔgoone} \TP{nɔgoonɛɛ} 
\LX{nɔhaaroŋ} \HM{} \PH{nɔ́háárʊ́ŋ} \PS{n.~} \GE{troublesomeness, pestering behaviour} \VA{nɔyaaroŋ}
\LX{nɔkɔnnoɔ} \HM{} \PH{[nɔ́kɔ̀nnʊ́ɔ́]} \PS{n.~} \GE{sweet talk} \OP{nɔkɔnnoore} \TP{nɔkɔnnoorɛɛ} 
\LX{nɔkɔɔloŋ} \HM{} \PH{nɔ́kɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{feathers of fowl} \TP{nɔkɔɔlɛɛ} 
\LX{nɔkɔrema} \HM{} \PH{nɔ́kɔ́rɪ́má} \PS{n.~} \GE{a type of fowl with feathers like those of a partridge} \OP{nɔkɔrre} \TP{nɔkɔrrɛɛ} 
\LX{nɔkpage} \HM{} \PH{nɔ́kpágɪ́} \PS{n.~} \GE{gluttonism} \VA{nɔkpogi}
\LX{nɔkpankyoɔne} \HM{} \PH{nɔ́kpánʧʊ́ɔ́nɪ́} \PS{n.~} \GE{a type of fowl with ash-coloured feathers} \OP{nɔkpankyoɔmɔ} \TP{nɔkpankyoɔnɛɛ} 
\LX{nɔkpeɛne} \HM{} \PH{nɔ́kpɪ́ɛ́nɪ́} \PS{n.~} \GE{argument} \TP{nɔkpeɛnɛɛ} 
\LX{nɔkpogi} \HM{} \PH{nɔ́kpógí} \PS{n.~} \GE{gluttonism} \OP{nɔkpogiri} \TP{nɔkpogiree} \VA{nɔkpage}
\LX{nɔlaŋ} \HM{} \PH{nɔ́làŋ} \PS{n.~} \GE{unity; agreement} 
\LX{nɔlee} \HM{} \PH{nɔ̀léé} \PS{n.~} \GE{chick} \OP{nɔbilii} \TP{nɔbillee} 
\LX{nɔlibie} \HM{} \PH{nɔ́líbíé} \PS{n.~} \GE{a type of fowl with spotted feathers} 
\LX{nɔll} \HM{} \PH{nɔ̀ll} \PS{adj.~} \GE{oval and large} 
\LX{nɔll} \HM{} \PH{nɔ́ll} \PS{adj.~} \GE{oval and small} 
\LX{nɔluuri} \HM{} \PH{nɔ́lùùrì} \PS{n.~} \GE{the act of renouncing an oath} 
\LX{nɔmaa} \HM{} \PH{nɔ́máà} \PS{n.~} \GE{surprise} \OP{nɔmaama} \TP{nɔmaanɛɛ} 
\LX{nɔmelime} \HM{} \PH{nɔ́mélímé} \PS{n.~} \GE{sweets} \OP{nɔmelimee} 
\LX{nɔmelime} \HM{} \PH{nɔ́mélímé} \PS{adj.~} \GE{craving for sweet things} 
\LX{nɔmm} \HM{} \PH{nɔ̀mm} \PS{adv.~} \GE{at rest; without movement} 
\LX{nɔmmo} \HM{} \PH{nɔ́mmʊ́} \PS{n.~} \GE{love; strong affection or likeness for} 
\LX{nomɔ} \HM{} \PH{nʊ́má} \PS{adj.~} \SN \GE{sweet, tasty} \XV{A tigiri bondirii da nomɔ la yaga.} \XE{The food at the party was really tasty.} \SN \GE{interesting} \XV{A senselloŋ da nomɔ la ka a bibiiri la paale zie zaa.} \XE{The story was interesting and so the children laughed off their heads.} \VA{noma}
\LX{nɔmoɔnaa} \HM{} \PH{nɔ́mʊ́ɔ́náá} \PS{n.~} \GE{lips decorated with red lipstick} \OP{nɔmoɔne/nɔmoɔmɔ} \TP{nɔmoɔnɛɛ} 
\LX{nɔmuri} \HM{} \PH{nɔ́múrí} \PS{n.~} \GE{a small, smooth stick fixed in a hole made in the lip} \OP{nɔmue} \TP{nɔmuuree} 
\LX{nɔmuu} \HM{} \PH{nɔ́mùú} \PS{n.~} \GE{dumbness} \OP{nɔmuuni} \TP{nɔmuunee} 
\LX{nɔŋe} \HM{} \PH{nɔ́ŋ} \PS{v.~} \SN \GE{to massage} \SN \GE{to make soft by applying substances and wringing out (leather)} \SN \GE{to knead} \PD{nɔŋɛɛ, nɔnɔ, nɔnɔ, nɔnema, nɔnaa, - } 
\LX{nɔŋe} \HM{} \PH{nɔ̀ŋ} \PS{v.~} \SN \GE{to like, to love} \SN \GE{to prefer} \PD{nɔŋɛɛ, nɔnɔ, nɔnɔ, nɔnema, nɔnaa, nɔnema } 
\LX{nɔnɛne} \HM{} \PH{nɔ́nɛ́nɪ̀} \PS{n.~} \GE{chicken (meat)} \SG{nɔnɛnoo} \OP{nɔnɛmɛ} \TP{nɔnɛnnɛɛ} 
\LX{nɔnne} \HM{} \PH{nɔ̀nnɪ́} \PS{n.~} \GE{love, affection} \TP{nɔnnɛɛ} 
\LX{nɔnne boore} \HM{} \PH{nɔ̀nnɪ́ bʊ̀ʊ̀rɪ́} \PS{n.~} \GE{the act of taking sides} 
\LX{nɔnɔ} \HM{} \PH{nɔ́nɔ́} \PS{n.~} \GE{lover, one who likes} \OP{nɔnema} \TP{nɔnemɛɛ} \VA{nɔna}
\LX{nɔnɔmaa} \HM{} \PH{nɔ̀nɔ̀màà} \PS{n.~} \GE{a type of tree with medicinal properties and edible fruits} \OP{nɔnɔmaɛ} 
\LX{nɔnoɔ} \HM{} \PH{nɔ́nʊ̀ɔ́} \PS{n.~} \GE{convincing talk; sweet mouth} \OP{nɔnoore} \TP{nɔnoorɛɛ} 
\LX{nɔnyaŋaa} \HM{} \PH{nɔ́ɲáŋáá} \PS{n.~} \GE{adult female hen (one that has hatched chicks)} \OP{nɔnyaŋene} \TP{nɔnyaŋenɛɛ} 
\LX{nɔnyeɛraa} \HM{} \PH{ńɔɲɪ́ɛ́ráà} \PS{n.~} \GE{laying fowl} \OP{nɔnyeɛre} \TP{nɔnyeɛrɛɛ} 
\LX{nɔnyemɛ} \HM{} \PH{nɔ́ɲɪ́mɛ́} \PS{n.~} \GE{disturbing sharp utterance e.g. barking puppy} \TP{nɔnyemɛɛ} 
\LX{noɔ} \HM{} \PH{nʊ̀ɔ́} \PS{n.~} \GE{happiness, joy} \XV{Pɔgebilii yaga maŋ teɛre ka noɔ yoŋ la be serekuli poɔ.} \XE{Most young women think that marriage carries with it only joy.} \OP{noore} \TP{noorɛɛ} 
\LX{noɔ} \HM{} \PH{nʊ́ɔ̀} \PS{n.~} \GE{fowl} \OP{noore} \TP{noorɛɛ} 
\LX{noɔ} \HM{} \PH{nʊ̀ɔ̀} \PS{v.~} \GE{to grow fat or large} \PD{noɔɛɛ, noɔnɔ, noɔnɔ, noɔnema, noonaa, - } 
\LX{noɔ} \HM{} \PH{nʊ̀ɔ́} \PS{adj.~} \GE{sweet} \XV{N lɛnne la a tii ka a waa noɔ lɛ.} \XE{I tasted the tea and it was very sweet.} \OP{noɔne} \TP{noɔnɛɛ} 
\LX{nɔɔ} \HM{} \PH{nɔ́ɔ́} \PS{adv.~} \GE{very one, exactly} \XV{O ne o zu nɔɔ la da wa a kyɛ.} \XE{He is the very one who came here.} 
\LX{nɔɔfeele} \HM{} \PH{nɔ̀ɔ̀fɪ́ɪ́lɪ́} \PS{n.~} \GE{a hard, sharp nail on the back of the leg of a full-grown male fowl} \OP{nɔɔfeele,} \TP{nɔɔfeelɛɛ} 
\LX{nɔɔganaa} \HM{} \PH{nɔ̀ɔ̀gànàá} \PS{n.~} \GE{big vein running from the heel to the calf} \OP{nɔɔganne} \TP{nɔɔganɛɛ} 
\LX{nɔɔkyoone} \HM{} \PH{nɔ̀ɔ̀kyʊ́ʊ́nɪ́} \PS{n.~} \GE{ankle} \OP{nɔɔkyooma} \TP{nɔɔkyoonɛɛ} 
\LX{noɔ-noɔre} \HM{} \PH{nʊ́ɔ́-nʊ́ɔ́rɪ̀} \PS{n.~} \GE{the habit of always eating} 
\LX{nɔɔpelaa} \HM{} \PH{nɔ́ɔ́pɪ́láá} \PS{n.~} \GE{abscess on the sole of the foot} 
\LX{noore} \HM{} \PH{nʊ̀ʊ̀rɪ̀} \PS{v.~} \SN \GE{to slither (snake)} \SN \GE{to move aside; usually backwards and away from} \PD{noorɛɛ, noora, noora, nooreba, nooraa, nooreba } 
\LX{noore} \HM{} \PH{nʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to search, to look for} \PD{noorɛɛ, noora, noora, nooreba, nooraa, nooreba } 
\LX{noɔre} \HM{} \PH{nʊ́ɔ́rè} \PS{n.~} \SN \GE{mouth} \SN \GE{beak, bill} \OP{nɔɛ} \TP{nɔrɛɛ} 
\LX{noɔre} \HM{} \PH{nʊ́ɔ́rè} \PS{n.~} \GE{edge} \OP{nɔɛ} \TP{nɔrɛɛ} 
\LX{noɔre} \HM{} \PH{nʊ́ɔ́rè} \PS{n.~} \GE{exact quantity} \OP{nɔɛ} \TP{nɔrɛɛ} 
\LX{noɔre} \HM{} \PH{nʊ́ɔ́rè} \PS{n.~} \GE{statement} \OP{nɔɛ} \TP{nɔrɛɛ} 
\LX{noɔre} \HM{} \PH{nʊ́ɔ́rè} \PS{n.~} \GE{the habit of talking a lot} \OP{nɔɛ} \TP{nɔrɛɛ} 
\LX{noɔre-emmo} \HM{} \PH{nʊ́ɔ́rɪ́-ɪ́mmʊ́} \PS{n.~} \GE{promise} 
\LX{nɔɔsaŋkpaare} \HM{} \PH{nɔ̀ɔ̀sàŋkpáàrɪ́} \PS{n.~} \SN \GE{heel} \SN \GE{hind toe of a bird} \OP{nɔɔsaŋkpaɛ} \TP{nɔɔsaŋkpaarɛɛ} 
\LX{nɔɔtemiri} \HM{} \PH{nɔ̀ɔ̀tɪ́mírì} \PS{n.~} \GE{shoelaces} \OP{nɔɔtemie} \TP{nɔɔtemiiree} 
\LX{nɔɔtere} \HM{} \PH{nɔ̀ɔ̀tɪ́rɪ̀} \PS{n.~} \GE{footwear, sandal, shoe} \SG{nɔɔteroo} \OP{nɔɔteɛ} \TP{nɔɔteerɛɛ} 
\LX{nɔɔtere} \HM{} \PH{nɔ́ɔ́tɪ́rɪ́} \PS{n.~} \GE{boil, abscess} \SG{nɔɔteroo} \OP{nɔɔteɛ} \TP{nɔɔteerɛɛ} 
\LX{nɔpollaa} \HM{} \PH{nɔ́pʊ́lláà} \PS{n.~} \GE{vow, promise} \OP{nɔpolle} \TP{nɔpollɛɛ} 
\LX{nɔpɔllee} \HM{} \PH{nɔ́pɔ́lléé} \PS{n.~} \GE{young fowl} \OP{nɔpɔlbilii} \TP{nɔpɔlbillee} \VA{nɔpɔle; nɔpɔlɔ}
\LX{nore} \HM{} \PH{nʊ̀rɪ̀} \PS{v.~} \SN \GE{to bend inward, usually with force} \SN \GE{to hide something} \SN \GE{to appear without warning e.g. forcing a seed out of a fruit} \PD{norɛɛ, norɔ, norɔ, noreba, noraa, - } 
\LX{nɔre} \HM{} \PH{nɔ̀rɪ̀} \PS{v.~} \GE{to be lucky} \XV{Zenɛ karembiiri banaŋ naŋ ba wa sakuuri wiẽu nɔrɛɛ la ne a karema potuo na naŋ ba wa sakuuri, o da na eŋe  ba zẽɛ.} \XE{Today, those pupils who came to school late were lucky that wicked teacher did not come to school since he would have punished them severely.} \PD{nɔrɛɛ, nɔrɔ, nɔrɔ, nɔreba, nɔraa, - } 
\LX{nɔsare} \HM{} \PH{nɔ́sárɪ́} \PS{n.~} \GE{a hen trough} \VA{nɔsɛre}
\LX{nɔsɛre} \HM{} \PH{nɔ́sɛ́rɪ́} \PS{n.~} \GE{drinking water container for fowl} \OP{nɔsaɛ} \TP{nɔsɛrɛɛ} \VA{nɔsare}
\LX{nɔtɛɛtɛɛ} \HM{} \PH{nɔ́tɛ́ɛ́tɛ̀ɛ̀} \PS{} \GE{incompatible statements; usually made by the same person} 
\LX{nɔtɛnzuŋi} \HM{} \PH{nɔ́tɛ́nzúŋ} \PS{n.~} \GE{pursed lips; protruding lips to show anger, annoyance} \OP{nɔtɛnzunni} \TP{nɔtɛnzunee/nɔtɛnzunee} 
\LX{nɔtore} \HM{} \PH{nɔ́tʊ́rɪ̀} \PS{n.~} \GE{the act of pressing the lips together (sign of annoyance)} 
\LX{nɔtoro} \HM{} \PH{nɔ̀tòrò} \PS{n.~} \GE{moustache} \OP{nɔtorri} \TP{nɔtoree} 
\LX{nɔtuo} \HM{} \PH{nɔ́túó} \PS{n.~} \GE{abusiveness; rudeness of language} 
\LX{nɔtuo} \HM{} \PH{nɔ́túó} \PS{n.~} \GE{hungering} 
\LX{nɔ-uunaa} \HM{} \PH{nɔ́-úúnáá} \PS{n.~} \GE{brooding fowl} \OP{nɔ-uuni} \TP{nɔ-uunee} 
\LX{nɔvelaa} \HM{} \PH{nɔ́vɪ́láá} \PS{n.~} \GE{encouraging talk} 
\LX{nɔwogi} \HM{} \PH{nɔ́wógì} \PS{n.~} \GE{dog} \OP{nɔwogri} \TP{nɔwogree} 
\LX{nɔwɔɔ} \HM{} \PH{nɔ́wɔ̀ɔ́} \PS{n.~} \GE{abrassiveness} \OP{nɔwɔɔre} \TP{nɔwɔɔrɛɛ} \VA{nɔtuo}
\LX{nɔyaaroŋ} \HM{} \PH{nɔ́yáárʊ́ŋ} \PS{n.~} \GE{trouble causing, pestering} \VA{nɔyaaroŋ}
\LX{nɔyoɔre} \HM{} \PH{nɔ́yʊ̀ɔ̀rɪ̀} \PS{n.~} \GE{pestering e.g. a little child} 
\LX{nɔzage} \HM{} \PH{nɔ́zàgɪ́} \PS{n.~} \GE{pen for hens} \OP{nɔzagere} \TP{nɔzagerɛɛ} 
\LX{nɔzagere} \HM{} \PH{nɔ́zàgɪ̀rɪ̀} \PS{n.~} \GE{disobedience} 
\LX{nɔzeɛ} \HM{} \PH{nɔ́zɪ́ɛ́} \PS{n.~} \GE{red mouth} 
\LX{nɔzɔbɔ} \HM{} \PH{nɔ́zɔ̀bɔ́} \PS{n.~} \GE{the act of pestering someone to annoy that person} \VA{nɔzɔba}
\LX{nɔzoolɔ} \HM{} \PH{nɔ́zʊ́ʊ́lɔ́} \PS{n.~} \GE{fringe} \OP{nɔzoolɛɛ} \VA{nɔzoola}
\LX{nsɛ} \HM{} \PH{nsɛ̀} \PS{n.~} \SN \GE{Mr} \SN \GE{great one} 
\LX{nu} \HM{} \PH{nû} \PS{n.~} \GE{hand, arm} \SG{nuuruu} \OP{nuuri} \TP{nuuree} 
\LX{nubaŋa} \HM{} \PH{núbáŋá} \PS{n.~} \GE{bangle for the wrist} \OP{nubanne} \TP{ nubannɛɛ.} \VA{nu bɔŋɔ}
\LX{nubidaa} \HM{} \PH{núbìdáà} \PS{n.~} \GE{thumb} \OP{nubidaare} \TP{nubidaarɛɛ} 
\LX{nubidori} \HM{} \PH{núbìdórí} \PS{n.~} \GE{finger joint; knuckle} \OP{nubidorri} \TP{nubidoree} 
\LX{nubi-eɛne} \HM{} \PH{núbì-ɪ́ɛ́nɪ́} \PS{n.~} \GE{fingernail} \OP{nubi-eɛmɛ} \TP{nubi-eɛnɛɛ} \VA{nubi-e᷉ɛ/nubi-e᷉mɛ}
\LX{nubilee} \HM{} \PH{núbìléé} \PS{n.~} \GE{smallest finger, pinky} \OP{nubibilii} \TP{nubilbillee} 
\LX{nubiri} \HM{} \PH{núbìrì} \PS{n.~} \GE{finger} \OP{nubie/nubiiri} \TP{nubiiree} 
\LX{nudɛgere} \HM{} \PH{núdɛ́gɪ́rɪ́} \PS{n.~} \GE{dirty hand} \OP{nudɛgɛ} \TP{nudɛgrɛɛ} 
\LX{nudoloŋ} \HM{} \PH{núdʊ̀lʊ̂ŋ} \PS{n.~} \SN \GE{right hand} \SN \GE{to the right (direction)} \OP{nudolle} \TP{nudollɛɛ} \VA{doloŋ}
\LX{nufoori} \HM{} \PH{núfóórì} \PS{n.~} \GE{closed empty hand} \XV{A bie ũũ la libiri o nu kaŋa poɔ kyɛ yeli ka N iri kaŋa ka N ŋmɛ a nugɔɔ, ka o yuoro ka nufoori la.} \XE{The girl hid a coin in one of her closed hands and asked me to choose one, and when I chose her left hand, it was empty.} \OP{nufoe} \TP{nuforee} 
\LX{nufoɔ᷉o} \HM{} \PH{núfʊ́᷉ɔ́᷉ʊ́᷉} \PS{n.~} \GE{ability to act fast with the hand} \OP{nufoɔne} \TP{nufoɔnɛɛ} 
\LX{nugɔɔ} \HM{} \PH{núgɔ̀ɔ̀} \PS{n.~} \GE{left hand} \OP{nugɔɔre} \TP{nugɔɔrɛɛ} \VA{gɔɔ}
\LX{nugɔɔ} \HM{} \PH{núgɔ̀ɔ̀} \PS{postposition.~} \GE{towards the left direction} \VA{gɔɔ}
\LX{nugunno} \HM{} \PH{núgúnnó} \PS{n.~} \GE{very poor, helpless} \TP{nugunnee} 
\LX{nukpara} \HM{} \PH{núkpárá} \PS{n.~} \GE{both hands put together} \OP{nukparre} \TP{nukparrɛɛ} 
\LX{nukuni} \HM{} \PH{núkúní} \PS{n.~} \GE{fist} \OP{nukumo} \TP{nukunee} 
\LX{nukuoŋaa} \HM{} \PH{núkúóŋáá} \PS{n.~} \GE{dry hand} \OP{nukuoni} \TP{nukuonee} 
\LX{nulaama} \HM{} \PH{núláámá} \PS{n.~} \GE{idle hands} 
\LX{nulaayɛlɛ} \HM{} \PH{núlàáyɛ́lɛ} \PS{n.~} \GE{work of idle hands} 
\LX{nuperaa} \HM{} \PH{núpɪ́ràà} \PS{n.~} \GE{finger snap} \OP{nuperre} \TP{nuperrɛɛ} 
\LX{nupoɔ} \HM{} \PH{núpʊ́ɔ́} \PS{n.~} \GE{palm (of hand)} \OP{nupoore} \TP{nupoorɛɛ} 
\LX{nupuori} \HM{} \PH{núpúórì} \PS{n.~} \GE{back of the hand} \OP{nupoe} \TP{nuporee} 
\LX{nusampane} \HM{} \PH{núsámpánɪ́} \PS{n.~} \GE{palm (of hand)} \OP{nusampama} \TP{nusampanɛɛ} 
\LX{nusugili} \HM{} \PH{núsúgílí} \PS{n.~} \GE{wrist} \OP{nusugilo} \TP{nusugilee} 
\LX{nutaa} \HM{} \PH{nútàà} \PS{n.~} \GE{wealthiness} 
\LX{nutale} \HM{} \PH{nútálɪ̀} \PS{n.~} \GE{palm (of hand)} \OP{nutala} \TP{nutalɛɛ} 
\LX{nutara} \HM{} \PH{nútárá} \PS{n.~} \GE{wealthy person} \OP{nutareba} \TP{nutarebɛɛ} 
\LX{nute᷉ɛ} \HM{} \PH{nútɪ́᷉ɛ́᷉} \PS{n.~} \GE{the act of giving, benevolence} 
\LX{nuu} \HM{} \PH{núú} \PS{adj.~} \GE{five} \OP{nuuri} \TP{nuuree} 
\LX{nuulee} \HM{} \PH{núúléé} \PS{n.~} \GE{bird} \OP{nuuli} 
\LX{nuuleemako᷉ɔko᷉ɔ} \HM{} \PH{núúléémákʊ́᷉ɔ́kʊ́᷉ɔ́} \PS{n.~} \GE{a type of plant with medicinal properties} 
\LX{nuuleemako᷉ɔ᷉} \HM{} \PH{núúléémákʊ̀᷉ɔ̀᷉} \PS{n.~} \GE{a type of small creeping plant} \OP{nuuleemakoɔnɛɛ} \VA{nuuleemako᷉ɔ᷉ko᷉ɔ᷉}
\LX{nuulikɔɔloŋ} \HM{} \PH{nùúlìkɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{feather} \OP{nuulikɔɔlo} 
\LX{nuulikyoo} \HM{} \PH{nùúlìʧóò} \PS{n.~} \GE{bird nest} \OP{nuulikyoori} \TP{nuulikyooree} 
\LX{nuuliparezoore} \HM{} \PH{núúlìpárɪ̀zʊ́ʊ́rɪ́n} \PS{n.~} \GE{pintailed bird} \OP{nuuliparezoɛ} \TP{nuuliparezoorɛɛ} 
\LX{nuulitige} \HM{} \PH{núùlitígé} \PS{n.~} \GE{flock (of birds) -- singular or plural agreement} \TP{nuulitigree} 
\LX{nuwogi} \HM{} \PH{núwógí} \PS{n.~} \SN \GE{light-fingeredness} \SN \GE{thief} \OP{nuwogiri} \TP{nuwogiree} 
\LX{nuyoɔre} \HM{} \PH{núyʊ̀ɔ̀rɪ̀} \PS{n.~} \GE{something that causes boredom, inactivity} 
\LX{nu-yuo} \HM{} \PH{nú-yùò} \PS{n.~} \GE{generosity} \OP{nu-yuori} \TP{nu-yuoree} 
\LX{nuzagele-pie} \HM{} \PH{núzáglɪ̀-píé} \PS{n.~} \GE{empty handedness} 
\LX{nuzɛgere} \HM{} \PH{núzɛ́grɪ́} \PS{n.~} \GE{handshake} 
\LX{nyaa} \HM{} \PH{ɲáà} \PS{n.~} \GE{chest} \OP{nyaɛ} \TP{nyaane/nyaanɛɛ} 
\LX{nyaa} \HM{} \PH{ɲàà} \PS{v.~} \SN \GE{to break e.g. light of day} \SN \GE{to recover oneself; to come round} \SN \GE{to awaken} \PD{nyaaɛ, nyaana, nyaana, nyaanema, nyaanaa, - } 
\LX{nyaa} \HM{} \PH{ɲàà} \PS{adj.~} \GE{wise, crafty} 
\LX{nyaaba} \HM{} \PH{ɲáábà} \PS{n.~} \GE{disgrace, shame} \TP{nyaabɛɛ} \VA{nyaama}
\LX{nyaalbini} \HM{} \PH{ɲààlbíní} \PS{n.~} \GE{a disease that causes itchy rashes} \TP{nyaalbinee} 
\LX{nyaalbiri} \HM{} \PH{ɲààlbírí} \PS{n.~} \GE{guinea worm (the worm)} \OP{nyaalbie} \TP{nyaalbiiree} 
\LX{nyaale} \HM{} \PH{ɲáálɪ̀} \PS{v.~} \GE{to move fire closely over a surface or move an object closely over fire} \PD{nyaalɛɛ, nyaala, nyaala, nyaaleba, nyaalaa, - } 
\LX{nyaalenyieli} \HM{} \PH{ɲààlɪ̀nyíélì} \PS{v.~} \GE{to walk with buttocks thrust backward and legs apart (because of sores on the buttocks or groin)} \PD{nyaalenyielee, nyaalenyiele, nyaalenyiele, nyaalenyielibe, nyaalenyielaa, - } 
\LX{nyaalombiri} \HM{} \PH{ɲáàlʊ̂mbírì} \PS{n.~} \GE{guinea worm (the worm)} \OP{nyaalombie} \TP{nyaalombiiree} 
\LX{nyaalompoɔ} \HM{} \PH{ɲáálʊ́mpʊ́ɔ́} \PS{n.~} \GE{undeveloped or dormant guinea-worm disease} \OP{nyaalompoore} \TP{nyaalompoore} 
\LX{nyaaloŋ} \HM{} \PH{ɲààlʊ́ŋ} \PS{n.~} \GE{guinea worm disease} \OP{nyiile} \TP{nyiilee} 
\LX{nyaanaa} \HM{} \PH{ɲàànáà} \PS{n.~} \GE{craft, cunning} \OP{nyaane} 
\LX{nyaane} \HM{} \PH{ɲààné} \PS{n.~} \SN \GE{wit} \SN \GE{wisdom} \TP{nyaanɛɛ} 
\LX{nyaanema} \HM{} \PH{ɲàànɪ́mà} \PS{n.~} \GE{cat} \OP{nyaanemamine} \TP{nyaanemɛɛ} 
\LX{nyaanoo} \HM{} \PH{ɲáánɔ́ɔ́} \PS{adv.~} \GE{at once} \XV{A wuo la a boma nyaanɔɔ ka saa na poɔ a.} \XE{S/he collected the things at once to prevent them being soaked by the rain.} 
\LX{nyaanyieli} \HM{} \PH{ɲàànyíélì} \PS{v.~} \GE{to walk with legs apart (usually because of sores on the buttocks)} \PD{nyaanyielee, nyaalanyiele, nyaanyiele, nyaanyielibe, nyaanyielaa, - } 
\LX{nyaanyoo} \HM{} \PH{ɲáányʊ́ʊ́} \PS{adv.~} \SN \GE{dim} \SN \GE{light (weight)} 
\LX{nyaaraa} \HM{} \PH{ɲááráá} \PS{n.~} \SN \GE{rake, fork;} \SN \GE{a type of small plant that grows on waste land} \SN \GE{bean soup, soup without okra} \OP{nyaare} \TP{nyaarɛɛ} \VA{gbaalaa}
\LX{nyaaraa} \HM{} \PH{ɲááráá} \PS{n.~} \GE{broom made from this plant} \OP{nyaare} \TP{nyarɛɛ} 
\LX{nyaare} \HM{} \PH{ɲààrɪ̀} \PS{v.~} \GE{to grind roughly e.g. in order to peel the husks off seeds} \PD{nyaarɛɛ, nyaara, nyaara, nyaareba, nyaaraa, - } 
\LX{nyaatafale} \HM{} \PH{ɲáátáfálɪ́} \PS{n.~} \GE{emptiness} 
\LX{nyabipeɛ} \HM{} \PH{ɲábìpɪ́ɛ́} \PS{n.~} \GE{ribcage} \OP{nyabipɛre} \TP{nyabipɛrɛɛ} 
\LX{nyabiri} \HM{} \PH{ɲábìrí} \PS{n.~} \SN \GE{rib} \SN \GE{a disease that causes a cramp in the chest} \OP{nyabie} \TP{nyabiree/nyabiiree} 
\LX{nyage} \HM{} \PH{ɲágɪ́} \PS{v.~} \GE{to become very angry, to become wild} \PD{nyagɛɛ, nyagra, nyagra, nyagreba, nyagraa, - } 
\LX{nyagele} \HM{} \PH{ɲàgɪ̀lɪ̀} \PS{v.~} \GE{to be entangled in spiky material (thorns)} \PD{nyaglɛɛ, nyagla, nyagla, nyagleba, nyaglaa, - } 
\LX{nyagele} \HM{} \PH{ɲàgɪ̀lɪ̀} \PS{v.~} \GE{to draw arms in a bid to fight (typically used with kpaŋkpama)} \PD{nyaglɛɛ, nyagla, nyagla, nyagleba, nyaglaa, - } 
\LX{nyagere} \HM{} \PH{ɲágɪ́rɪ́} \PS{n.~} \SN \GE{bowel} \SN \GE{tree root} \OP{nyaga} \TP{nyagrɛɛ} 
\LX{nyagere} \HM{} \PH{ɲàgɪ̀rɪ̀} \PS{v.~} \SN \GE{bite suddenly and sharply (dog bite)} \XV{A baa digi te nyɔge la a nannyige gbɛre e nygere o naŋ doŋ.} \XE{The dog ran after the thief and bit tore at his leg instantly.} \SN \GE{get stimulated} \XV{Saa ba miire azuiŋ n eŋɛ ba nyagera ne weɛ gaabo.} \XE{It does not rain and so I am not encouraged to go to the farm.} \PD{nyagrɛɛ, nyagra, nyagra, nyagreba, nyagraa, - } 
\LX{nyageroŋ} \HM{} \PH{ɲágrʊ́ŋ} \PS{adv.~} \SN \GE{quickly} \XV{Kyɛŋe nyageroŋ ka te gaa wi᷉eu.} \XE{Walk fast so that we can get there quickly.} \SN \GE{intelligently} \XV{A pɔge da e la nyãgeroŋ ne o naŋ tage a dɔɔ yi a soriŋ, ka lɛ naane a lɔɔre da na ŋmɛ o la.} \XE{The woman was smart by pulling the man out of the way otherwise the lorry would have hit him.} 
\LX{nyagesaga} \HM{} \PH{ɲágɪ́ságá} \PS{adj.~} \GE{jumbled, dry, and bare e.g. metal scrap, bones} \XV{A kuridaare e la nyagesaga lɛ a davorikpoŋi poɔ.} \XE{The scrap of metal was jumbled up in the yard.} 
\LX{nyakpuli} \HM{} \PH{ɲákpúlí} \PS{n.~} \GE{sternum} \OP{nyakpulo} \TP{nyakpulee} 
\LX{nyale} \HM{} \PH{ɲàlɪ́} \PS{adv.~} \GE{suddenly, usually used to describe the sudden manner some animals excrete} \XV{Ba ŋmɛ la a bagao ka o nyɛ bini nyale.} \XE{They hit the mad dog and it defecated suddenly.} 
\LX{nyalle} \HM{} \PH{ɲàllɪ̀} \PS{v.~} \GE{to eat with very little sauce} \PD{nyallɛɛ, nyalla, nyalla, nyaleba, nyallaa, - } 
\LX{nyam} \HM{} \PH{ɲàm} \PS{v.~} \GE{to take heart, to be patient} \PD{nyamɛɛ, nyamena, nyamena, nyamena, nyamenaa, - } 
\LX{nyama} \HM{} \PH{ɲàmà} \PS{adv.~} \GE{gripping suddenly} \XV{A kɔkɔlaa vare guri la a gyaŋgboŋ lɛ nyama.} \XE{The cat jumped on the mouse, gripping it firmly.} 
\LX{nyamaa} \HM{} \PH{ɲàmàà} \PS{adv.~} \GE{mixed up, congested} 
\LX{nyamama} \HM{} \PH{ɲámámá} \PS{adv.~} \GE{rising in intensity (pain)} \XV{Ka gyaŋgboŋiŋ wa peɛle fo kyɛ doŋ a be maŋ buori nyamama lɛ.} \XE{When a mouse gnaws at your foot, the pain rises slowly.} 
\LX{nyambaraŋ} \HM{} \PH{ɲàmbàràŋ} \PS{adj.~} \GE{itchy} \XV{Ka neɛŋ puri wale eŋe ka-uri poɔ, o eŋɛ maŋ waa la nyambaraŋ lɛ.} \XE{If you sweat on top of millet chaff your body is normally itchy.} 
\LX{nyaŋ} \HM{} \PH{ɲáŋ} \PS{v.~} \GE{to age, to become old} \PD{nyaŋɛɛ, nyaŋena, nyaŋena, nyaŋenema, nyaŋenaa, - } 
\LX{nyaŋe} \HM{} \PH{ɲáŋ} \PS{adj.~} \GE{hasty} \XV{Kyɛŋ gbɛnyaŋe ka te gaa, zie soɔrɔ la.} \XE{Walk fast; let's go, for the night is falling on us.} 
\LX{nyaŋema} \HM{} \PH{ɲáŋímà} \PS{n.~} \GE{accident} \TP{nyaŋemɛɛ} 
\LX{nyaŋgalaŋ} \HM{} \PH{ɲàŋgàlàŋ} \PS{adv.~} \GE{thick and stiff} 
\LX{nyaraa-nyaraa} \HM{} \PH{ɲárááɲáráá} \PS{adj.~} \GE{with small pointed appendages e.g. crab} 
\LX{nyaranyara} \HM{} \PH{ɲàràɲàrà} \PS{adj.~} \GE{rough, rugged e.g. stony ground} 
\LX{nyarata} \HM{} \PH{ɲárátá} \PS{adv.~} \GE{swiftly e.g. movement of lizard} \VA{nyaratatata}
\LX{nyaremaŋ} \HM{} \PH{ɲàrɪ̀màŋ} \PS{adj.~} \GE{hot e.g. food} 
\LX{nyarewara} \HM{} \PH{ɲàrɪ̀wàrà} \PS{adj.~} \GE{grainy, stony e.g. unplastered floor} 
\LX{nyaroo} \HM{} \PH{ɲárʊ́ʊ́} \PS{adj.~} \GE{small and lean} 
\LX{nyawele} \HM{} \PH{ɲåwɪ́lɪ́} \PS{n.~} \GE{one that is the greatest} \OP{nyawelɛ} \TP{nyawelɛɛ} 
\LX{nyɛ} \HM{} \PH{ɲɛ́} \PS{v.~} \SN \GE{to see} \SN \GE{to find} \SN \GE{to get, to obtain} \PD{nyɛɛ, nyɛrɛ, nyɛrɛ, nyɛrebɛ, nyɛraa, - } 
\LX{nyɛ} \HM{} \PH{ɲɛ̀} \PS{v.~} \SN \GE{to defecate} \SN \GE{pass substance through the anus (lay eggs)} \PD{nyɛɛ, nyɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\LX{nyɛ} \HM{} \PH{ɲɛ̀} \PS{v.~} \GE{to produce tubers or nuts e.g. yam} \PD{nyɛɛ, nyɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\LX{nyɛ} \HM{} \PH{ɲɛ̀} \PS{v.~} \SN \GE{to drain a pond of its water; usually to enable people catch the fish in the pond} \SN \GE{to drill or scoop matter out to make a hole} \PD{nyɛɛ, nyɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\LX{nyɛ eebo} \HM{} \PH{ɲɛ́ ɪ́ɪ́bʊ́} \PS{v.~} \GE{succeed} \PD{-, -, -, -, -, - } 
\LX{nyɛ kparɛɛ} \HM{} \PH{ɲɛ́ kpárɛ́ɛ́} \PS{v.~} \GE{to suffer} \PD{nyɛɛ kparɛɛ, nyɛrɛ kparɛɛ, -, -, -, - } 
\LX{nyɛ pare} \HM{} \PH{ɲɛ́ párɪ́} \PS{v.~} \GE{to lie down helplessly, to crumple} \PD{nyɛɛ pare, nyeɛrɛ pare, -, -, -, - } 
\LX{nyɛ zɛ} \HM{} \PH{ɲɛ̀ zɛ̀} \PS{v.~} \GE{to suffer} \PD{-, -, -, -, -, - } 
\LX{nyɛ ferɛ} \HM{} \PH{ɲɛ̀ fɪ́rɛ̀} \PS{v.~} \GE{to suffer} \PD{-, -, -, -, -, - } 
\LX{nyɛ yuori} \HM{} \PH{ɲɛ́ yúórí} \PS{v.~} \GE{get fame, (be) powerful} \PD{-, -, -, -, -, - } 
\LX{nyɛ zɛ} \HM{} \PH{ɲɛ̀ zɛ̀} \PS{v.~} \GE{to suffer greatly} \PD{nyɛ zɛɛ, nyeɛrɛ zeɛrɛ, nyɛ zeɛrɛ, nyɛ zeɛraa, -, - } 
\LX{nyɛdanaao} \HM{} \PH{ɲɛ̀dànáábʊ́} \PS{n.~} \GE{a type of tiny brown bird that moves in large groups that lay eggs together} \OP{nyɛdanii} 
\LX{nyeɛ} \HM{} \PH{ɲɪ̀ɛ́} \PS{n.~} \GE{snot} \TP{nyeɛnɛɛ} 
\LX{nyeɛne} \HM{} \PH{ɲɪ̀ɛ́nɪ́} \PS{n.~} \SN \GE{bulb, tuber} \SN \GE{descendant of mother's line} \OP{nyeɛmɛ} \TP{nyeɛnɛɛ} 
\LX{nyeɛre} \HM{} \PH{ɲɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to test, to try out, to pilot} \PD{nyeɛrɛɛ, nyeɛrɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\LX{nyeɛre} \HM{} \PH{ɲɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to pay for the cost of herbs after treatment} \PD{nyeɛrɛɛ, nyeɛrɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\LX{nyɛgɛkɛ} \HM{} \PH{ɲɛ́gɛ́kɛ́} \PS{adj.~} \GE{warmish} 
\LX{nyegenyege} \HM{} \PH{ɲɪ̀gɪ̀ɲɪ̀gɪ̀} \PS{n.~} \GE{restlessness} \XV{O taa la nyegenyege zuo, o na ŋmare la ŋmane.} \XE{His restlessness is great; he will break the calabsh.} 
\LX{nyegenyege} \HM{} \PH{ɲɪ̀gɪ̀ɲɪ̀gɪ̀} \PS{adj.~} \GE{restless, restive} \XV{O waa la nyegenyege lɛnso noba yaga zoro o la.} \XE{He is very restive, so many people fear him.} 
\LX{nyegɛnyegɛ} \HM{} \PH{ɲɪ́gɛ́ɲɪ́gɛ́} \PS{adj.~} \GE{very sharp and shiny e.g. blade} 
\LX{nyegere} \HM{} \PH{ɲɪ̀grɪ̀} \PS{v.~} \GE{to be scarce} \XV{Yuoni ŋa boŋkoɔre yaga ba maale azuiŋ bondirii na nyegere la.} \XE{This year's crop yield was poor and so food will be scarce.} \PD{-, nyegrɛ, -, -, -, - } 
\LX{nyegere} \HM{} \PH{ɲɪ̀grɪ̀} \PS{v.~} \GE{to massage with a hot, dry object} \XV{Daa daare kyɔge la a koɔraa gbɛre ka ba voɔ kyɛ nyegere.} \XE{A thorn pierced and lodged itself in the farmer's foot, and after removing it they used hot water to massage it.} \PD{-, nyegrɛ, -, -, -, - } 
\LX{nyegere} \HM{} \PH{ɲɪ́grɪ́} \PS{v.~} \SN \GE{to flash e.g. lightening} \SN \GE{give way momentarily (limb as a result of sharp twist)} \XV{A dɔɔ gbɛbaala na nyegerɛɛ la ka o le.} \XE{The man's wounded leg made a snapping noise and he fell.} \PD{nyegrɛɛ, nyegrɛ, nyegrɛ, nyegrebɛ, nyegraa, - } 
\LX{nyelaa} \HM{} \PH{ɲɪ̀ràà} \PS{adj.~} \GE{not hidden, plainly visible} \XV{A goɔ kɔɔ eŋ la a gbɛre poɔ kyɛ o sa᷉a la kperaa lɛ} \XE{The thorn has pierced the foot, but it can be seen clearly.} 
\LX{nyɛle} \HM{} \PH{ɲɛ́lɪ̀} \PS{v.~} \GE{to melt} \XV{ŋmenaa ŋmɛ la a ka᷉a ka a nyɛle.} \XE{The oil melted as a result of the sun beating down on it.} \PD{nyɛllɛɛ, nyɛllɛ, nyɛllɛ, nyɛleba, nyɛllaa, nyɛleba } 
\LX{nyelele} \HM{} \PH{ɲɪ́lɪ́lɪ́} \PS{adv.~} \GE{slightly} \XV{A vũũ da sããɛ̃ nyelele lɛ gaŋe zaameŋ.} \XE{The light appeared slightly less bright than it was yesterday.} 
\LX{nyelennyeleŋ} \HM{} \PH{ɲélénɲéléŋ} \PS{adj.~} \GE{bare and shiny e.g. razor shaved head} \XV{A bie zu pu᷉oe la ka ba fare ka o e nyelennyeleŋ lɛ.} \XE{The child has a cut on his head and they have shaven it clean.} 
\LX{nyɛlle} \HM{} \PH{ɲɛ́llɪ́} \PS{v.~} \GE{to shine, to sparkle} \XV{A pɔge du la a talaa ka o te nyɛllɛ.} \XE{The womann scrubbed the pan until it shone.} \PD{nyɛllɛɛ, nyɛllɛ, nyɛllɛ, nyɛlebɛ, nyɛllaa, - } 
\LX{nyɛlle} \HM{} \PH{ɲɛ̀llɪ̀} \PS{v.~} \GE{to covet, to aim at getting something} \XV{N kuriwiri korɛɛ la N nyɛllɛ la bopaalaa.} \XE{My bicycle is old, so I'm aiming at getting a new one.} \PD{nyɛllɛɛ, nyɛllɛ, nyɛllɛ, nyɛlebɛ, nyɛllaa, - } 
\LX{nyɛlmaanyu} \HM{} \PH{ɲɛ́lmáányú} \PS{n.~} \GE{a type of small plant with a strong smell and medicinal properties} \OP{nyɛlmaanyuri/nyɛlmaanyuuri} \TP{nyɛlmaanyuree/nyɛlmaanyuuree} 
\LX{nyelɔɔ} \HM{} \PH{ɲɪ́lɔ́ɔ́} \PS{adj.~} \GE{tiny} 
\LX{nyɛlugiri} \HM{} \PH{ɲɛ̀lúgrí} \PS{n.~} \GE{support stick for a yam plant} \OP{nyɛlugo} \TP{nyɛlugee/nyɛlugree} \VA{nyalugri}
\LX{nyembini} \HM{} \PH{ɲɪ́bɪ́nɪ́} \PS{n.~} \GE{matter (dirt) in between the teeth} \TP{nyembinee} 
\LX{nyembulo} \HM{} \PH{ɲɪ́mbúló} \PS{n.~} \GE{adult} \OP{ nyembulibo} 
\LX{nyempara} \HM{} \PH{ɲɪ́mpàrá} \PS{n.~} \GE{unchiselled teeth} 
\LX{nyempɛge dalee} \HM{} \PH{ɲɪ́mpɛ́gɪ̀ dàléé} \PS{n.~} \GE{tooth stick, toothbrush} \OP{nyempɛge dabilii} \TP{nyempɛge dabilee} 
\LX{nyene} \HM{} \PH{ɲɪ́nɪ̀} \PS{n.~} \SN \GE{tooth, fang (snake)} \SN \GE{toothache} \OP{nyemɛ} \TP{nyennɛɛ} 
\LX{nyeŋgoŋe} \HM{} \PH{ɲíŋgʊ́ŋ} \PS{n.~} \GE{the state of grimacing (typically used with "vere")} \XV{A dɔɔ da tage la a miri te vere nyeŋgoŋ.} \XE{The man tugged at the rope until he grimaced.} \OP{nyeŋgonne} \TP{nyeŋgonnɛɛ} 
\LX{nyeŋkaane} \HM{} \PH{ɲɪ́ŋkáánɪ́} \PS{n.~} \GE{molar} \SG{nyeŋkaanoo} \OP{nyeŋkaama} \TP{nyeŋkaanɛɛ} 
\LX{nyɛnnoŋ} \HM{} \PH{ɲɛ̀nnʊ́ŋ} \PS{n.~} \GE{salt} \VA{nyɛnnoo, yaaroŋ}
\LX{Nyɛntɔge} \HM{} \PH{ɲɛ́ntɔ́gɪ̀} \PS{n.~} \GE{a meteorite revered to be a fetish in the Sisaalaland} \OP{nyɛntɔgre} \TP{nyɛntɔgrɛɛ} 
\LX{nyeraa} \HM{} \PH{ɲìráá} \PS{n.~} \GE{ant} \OP{nyere} \TP{nyerɛɛ} 
\LX{nyeraa} \HM{} \PH{ɲɪ̀ràà} \PS{adj.~} \GE{not hidden; clearly visible} 
\LX{nyɛraa} \HM{} \PH{ɲɛ́ráá} \PS{n.~} \GE{seer} \OP{nyɛrebɛ} \TP{nyɛrebɛɛ} 
\LX{nyere} \HM{} \PH{ɲɪ̀rɪ̀} \PS{v.~} \GE{to bare one's teeth; to laugh (scornfully)} \PD{nyerɛɛ, nyerɛ, nyerɛ, nyerebɛ, nyeraa, - } 
\LX{nyɛre} \HM{} \PH{ɲɛ́rɪ̀} \PS{v.~} \GE{to miss (a target)} \PD{nyɛrɛɛ, nyɛrɛ, nyɛrɛ, nyɛrebɛ, nyɛraa, - } \VA{nyare}
\LX{nyɛre} \HM{} \PH{ɲɛ̀rɪ̀} \PS{v.~} \GE{to be fierce, to get very angry} \PD{nyɛrɛɛ, nyɛrɛ, nyɛrɛ, nyɛrebɛ, nyɛraa, - } \VA{nyare}
\LX{nyerebaa} \HM{} \PH{ɲɪ̀rɪ̀bàá} \PS{n.~} \GE{a type of ant with a painful sting} \OP{nyerebaare} \TP{nyerebaarɛɛ} \VA{nyeresɔgelaa}
\LX{nyeregaŋguu} \HM{} \PH{ɲɪ̀rɪ̀gàŋgúú} \PS{n.~} \GE{a type of ant} \OP{nyeregaŋguuri} \TP{nyeregaŋguuree} 
\LX{nyeregbere} \HM{} \PH{ɲɪ̀rɪ̀gbɪ̀rɪ̀} \PS{adj.~} \GE{mixed} 
\LX{nyerekɔɔloŋ} \HM{} \PH{ɲɪ̀rɪ̀kɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{a type of ant that flies} \OP{nyerekɔɔlɔ} \TP{nyerekɔɔlɛɛ} 
\LX{nyerenyaga} \HM{} \PH{ɲɪ̀rɪ̀nyàgà} \PS{adv.~} \GE{uncovered, exposed completely} 
\LX{nyeresaare} \HM{} \PH{ɲɪ̀rɪ̀sáárɪ́} \PS{n.~} \GE{a type of small slim dark brown ant; usually in a swarm and on trees} \SG{nyeresaaroo} \OP{nyeresaare} \TP{nyeresaarɛɛ} \VA{nyeresalensaare}
\LX{nyeresalensaare} \HM{} \PH{ɲɪ̀rɪ̀sálɪ́nsáárɪ̀} \PS{n.~} \GE{a type of tiny ant that lives together in large numbers} \SG{nyeresalensaaroo} \TP{nyeresalensaarɛɛ} \VA{nyeresaare}
\LX{nyeresɔgelaa} \HM{} \PH{ɲɪ́rɪ́sɔ́gláá} \PS{n.~} \GE{a type of small black ant, usually found in the house; it has a painful sting} \OP{nyeresɔgelɔ} \VA{nyerebaa}
\LX{nyɛtaa} \HM{} \PH{ɲɛ́táá} \PS{v.~} \GE{confer} \PD{-, -, -, -, -, - } 
\LX{nyieli} \HM{} \PH{ɲíélì} \PS{v.~} \GE{to tempt; to show something to someone but refuse to give it to him/her} \PD{nyíéliee, nyíéle, nyíélie, nyíélaa, -, - } 
\LX{nyigiri} \HM{} \PH{ɲígírì} \PS{v.~} \GE{to close both eyes tightly and simultaneously towards someone (insulting gesture)} \PD{nyigree, nyigre, nyigre, nyigiribe, nyigraa, nyigiribe } 
\LX{nyigiri} \HM{} \PH{ɲìgìrì} \PS{v.~} \GE{to steal} \PD{nyigiree, nyigire, nyigire, nyigiribe, nyigiraa, nyigiribe } 
\LX{nyiile} \HM{} \PH{ɲììlé} \PS{n.~} \GE{guinea worm disease} \OP{nyiili} \TP{nyiilee} 
\LX{nyiili} \HM{} \PH{ɲììlì} \PS{v.~} \GE{to touch gently} \PD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\LX{nyiili} \HM{} \PH{ɲììlì} \PS{v.~} \GE{to incense} \PD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\LX{nyiili} \HM{} \PH{ɲììlì} \PS{v.~} \GE{to shudder} \PD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\LX{nyiili} \HM{} \PH{ɲììlì} \PS{v.~} \GE{to swell (breast)} \PD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\LX{nyiiri} \HM{} \PH{ɲììrì} \PS{v.~} \GE{an intransitive verb; used with ‘nyemɛ’ to bare teeth} \XV{Basɔgelaa kaŋa nyiiri la nyemɛ eŋe a nannyige.} \XE{A black dog bared its teeth at the thief.} \PD{nyiiree, nyiire, nyiire, -, nyiiraa, - } 
\LX{nyimenyime} \HM{} \PH{ɲíményímé} \PS{adj.~} \GE{tasteful} 
\LX{nyɔbaare} \HM{} \PH{ɲɔ́báárɪ́} \PS{n.~} \GE{disease of the nose} \TP{nyɔbaarɛɛ} 
\LX{nyobogdaa} \HM{} \PH{ɲòbògìdáá} \PS{n.~} \GE{bridge of nose} \OP{nyobogdaare} \TP{nyobogdaarɛɛ} 
\LX{nyɔbogiri} \HM{} \PH{ɲɔ̀bògirí} \PS{n.~} \SN \GE{nostril} \SN \GE{nose} \OP{nyɔbogo} \TP{nyɔbogiree} \VA{nyobogi}
\LX{nyɔfeerɛ} \HM{} \PH{ɲɔ̀fɪ́ɪ́rɛ́} \PS{n.~} \GE{cold (illness)} \TP{nyɔfeerɛɛ} 
\LX{nyɔfuli} \HM{} \PH{ɲɔ̀fùlí} \PS{n.~} \GE{dried snot} \SG{nyɔfuluu} \OP{nyɔfulo} \TP{nyɔfulee} 
\LX{nyoge} \HM{} \PH{ɲʊ̀gɪ̀} \PS{v.~} \GE{to show off} \PD{nyogɛɛ, nyogrɔ/nyogra, nyogrɔ/nyogra, nyogreba, nyograa, - } 
\LX{nyoge} \HM{} \PH{ɲʊ̀gɪ̀} \PS{v.~} \GE{to become sour or bad e.g. sauce} \PD{nyogɛɛ, nyogrɔ/nyogra, nyogrɔ/nyogra, nyogreba, nyograa, - } 
\LX{nyoge} \HM{} \PH{ɲʊ́gɪ́} \PS{v.~} \GE{used with ‘tɔ’ (tɔ-nyoge) punish severely (physically)} \PD{-, -, -, -, -, - } 
\LX{nyoge} \HM{} \PH{ɲʊ́gɪ́} \PS{n.~} \GE{concoction believed to be powerful} 
\LX{nyoge} \HM{} \PH{ɲʊ́gɪ́} \PS{n.~} \GE{filth} 
\LX{nyɔge} \HM{} \PH{ɲɔ́gɪ̀} \PS{v.~} \SN \GE{to hold} \XV{A bibiiri nyɔge la taa nuuri kyerre te gɛrɛ.} \XE{The children held each others' hands and scampered away.} \SN \GE{to catch, to arrest} \XV{A polisi nyɔge la a nanyige baguo ŋa.} \XE{The police arrested the thief this morning.} \PD{nyɔgɛɛ, nyɔgrɔ, nyɔgrɔ, nyɔgreba, nyɔgraa, nyɔgreba } 
\LX{nyɔge} \HM{} \PH{ɲɔ́gɪ̀} \PS{v.~} \GE{to choke} \XV{A baa vɔlɔ la a nɛne yaga lɛ ka a te nyɔge o.} \XE{The dog swallowed the meat so fast that the meat choked it.} \PD{nyɔgɛɛ, nyɔgrɔ, nyɔgrɔ, nyɔgreba, nyɔgraa, nyɔgreba } 
\LX{nyɔge} \HM{} \PH{ɲɔ́gɪ̀} \PS{v.~} \GE{to infect} \XV{A bie kɔroŋ na nyɔge la o yɔɔ.} \XE{The child's cold has infected his sister.} \PD{nyɔgɛɛ, nyɔgrɔ, nyɔgrɔ, nyɔgreba, nyɔgraa, nyɔgreba } 
\LX{nyɔge kukuri} \HM{} \PH{ɲɔ́gɪ́ kúkúrì} \PS{v.~} \GE{to farm} \PD{nyɔgɛɛ kukuri, nyɔgrɔ kukuri, -, -, -, - } 
\LX{nyɔge noɔre} \HM{} \PH{ɲɔ́gɪ́ nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to be surprised} \PD{-, -, -, -, -, - } 
\LX{nyoge-kaa-kpɛre} \HM{} \PH{ɲʊ́gɪ́-káá-kpɛ́rɪ́} \PS{n.~} \GE{over show off that ends badly} 
\LX{nyogele} \HM{} \PH{ɲʊ̀gɪ̀lɪ̀} \PS{v.~} \GE{to dress well} \PD{-, -, -, -, -, - } 
\LX{nyogele} \HM{} \PH{ɲʊ̀gɪ̀lɪ̀} \PS{v.~} \GE{to be congested} \PD{nyoglɛɛ, nyoglɔ, nyoglɔ, nyoglebɔ, nyoglaa, - } 
\LX{nyogere} \HM{} \PH{ɲʊ́gɪ́rɪ̀} \PS{n.~} \GE{pride} 
\LX{nyɔgetaa} \HM{} \PH{ɲɔ́gɪ́táá} \PS{n.~} \GE{unity} 
\LX{nyɔgezie} \HM{} \PH{ɲɔ́gɪ̀zíé} \PS{n.~} \GE{handle, part of an object to hold} \OP{nyɔgeziiri} \TP{nyɔgeziiree} 
\LX{nyoll} \HM{} \PH{ɲʊ̀ll} \PS{adj.~} \GE{heavy} 
\LX{nyoll} \HM{} \PH{ɲʊ́ll} \PS{adj.~} \GE{small and light} 
\LX{nyoŋgoloŋ} \HM{} \PH{ɲʊ̀ŋgʊ̀lʊ̀ŋ} \PS{adj.~} \GE{mixed up e.g. heap of things} 
\LX{nyoŋgoloŋ} \HM{} \PH{ɲʊ̀ŋgʊ̀lʊ̀ŋ} \PS{adj.~} \GE{thick and stiff e.g. paste} 
\LX{nyonne} \HM{} \PH{ɲʊ́nnɪ́} \PS{v.~} \GE{to dry up, to evaporate} \PD{nyonnɛɛ, nyonna, nyonna, nyonnema, nyonnaa, nyonnema } 
\LX{nyonne} \HM{} \PH{ɲʊ̀nnɪ̀} \PS{v.~} \SN \GE{to sniff} \SN \GE{to search; to delve into} \SN \GE{to test; to examine} \PD{nyonnɛɛ, nyonna, nyonna, nyonnema, nyonnaa, nyonnema } 
\LX{nyonono} \HM{} \PH{ɲʊ́nʊ́nʊ́} \PS{n.~} \GE{mercy, pity} 
\LX{nyonsomm} \HM{} \PH{ɲʊ̀nsʊ̀mm} \PS{adj.~} \GE{biggish} 
\LX{nyonyonne} \HM{} \PH{ɲʊ́nyʊ́nnɪ̀} \PS{n.~} \GE{nausea} 
\LX{nyonyonne} \HM{} \PH{ɲʊ́nyʊ́nnɪ̀} \PS{n.~} \GE{dirt} 
\LX{nyoɔ} \HM{} \PH{ɲʊ́ɔ̀} \PS{v.~} \GE{to fire, to bake to harden (clay)} \PD{nyoɛ, nyoɔnɔ, nyoɔnɔ, nyoɔnema, nyoɔnaa, nyoɔnema } 
\LX{nyɔɔ} \HM{} \PH{ɲɔ̀ɔ̀} \PS{v.~} \GE{to have sexual intercourse (human)} \PD{nyɔɔɛ, nyɔɔnɔ/nyɔɔrɔ, nyɔɔnɔ/nyɔɔrɔ, nyɔɔnema, nyɔɔnaa/nyɔɔraa, nyɔɔnema } 
\LX{nyoɔle} \HM{} \PH{ɲʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to begin rising} \PD{nyoɔlɛɛ, nyoɔlɔ, nyoɔlɔ, nyoɔlba, nyoɔlaa, - } 
\LX{nyoɔle} \HM{} \PH{ɲʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to close partially e.g. of eyes in strong light} \PD{nyoɔlɛɛ, nyoɔlɔ, nyoɔlɔ, nyoɔlba, nyoɔlaa, - } 
\LX{nyoɔnaa} \HM{} \PH{ɲʊ́ɔ́náá} \PS{n.~} \GE{oven} \OP{nyoɔne} \TP{nyoɔnɛɛ} 
\LX{nyoore} \HM{} \PH{ɲʊ̀ʊ́rɪ̀} \PS{n.~} \GE{smoke} \TP{nyoorɛɛ} \VA{zoore}
\LX{nyoore} \HM{} \PH{ɲʊ̀ʊ́rɪ̀} \PS{n.~} \GE{pride} \TP{nyoorɛɛ} \VA{zoore}
\LX{nyoore} \HM{} \PH{ɲʊ̀ùrɪ̀} \PS{v.~} \SN \GE{to make hard} \XV{Zenɛ ka ba mɛ a dankyini ŋa bieo ka o na nyoore.} \XE{This wall was only built today, so by tomorrow it will get hard.} \SN \GE{to dry up} \XV{Uoni saŋa bare yaga maŋ nyoore la.} \XE{Many ponds get dried up during the dry season.} \PD{nyoorɛɛ, nyoora, nyoora, nyooreba, nyooraa, - } 
\LX{nyoore} \HM{} \PH{ɲʊ̀ùrɪ̀} \PS{v.~} \GE{to test the capability of e.g. a tool or a person} \PD{nyoorɛɛ, nyoora, nyoora, nyooreba, nyooraa, - } 
\LX{nyoɔre} \HM{} \PH{ɲʊ́ɔ́rɪ́} \PS{n.~} \GE{nose} \OP{nyɔɛ} \TP{nyɔrɛɛ} 
\LX{nyoɔre} \HM{} \PH{ɲʊ́ɔ́rɪ́} \PS{n.~} \GE{life} \OP{nyɔɛ} \TP{nyɔrɛɛ} 
\LX{nyoɔre} \HM{} \PH{ɲʊ́ɔ́rɪ́} \PS{n.~} \GE{tip of e.g. stick} \OP{nyɔɛ} \TP{nyɔrɛɛ} 
\LX{nyɔpolema} \HM{} \PH{ɲɔ̀pʊ́lɪ́má} \PS{n.~} \GE{summit, highest point} \OP{nyɔpolle} \TP{nyɔpollɛɛ} 
\LX{nyore} \HM{} \PH{ɲʊ̀rɪ̀} \PS{v.~} \GE{to shrivel, to wrinkle (cloth)} \PD{nyorɛɛ, nyora, nyora, nyoreba, nyoraa, nyoreba } 
\LX{nyore} \HM{} \PH{ɲʊ̀rɪ̀} \PS{v.~} \GE{to take courage} \PD{nyorɛɛ, nyora, nyora, nyoreba, nyoraa, nyoreba } 
\LX{nyɔre} \HM{} \PH{ɲɔ̀rɪ̀} \PS{v.~} \GE{to despise, to hate} \PD{nyɔrɛe,, nyɔrɔ, nyɔrɔ, nyɔreba/nyɔrebɔ, nyɔraa, - } \VA{nyɔroo}
\LX{nyoremaŋ} \HM{} \PH{ɲʊ̀rɪ̀màŋ} \PS{adj.~} \GE{bushy, unkempt (hair)} 
\LX{nyɔtage} \HM{} \PH{ɲɔ̀tàgɪ̀} \PS{n.~} \GE{sniveling} 
\LX{nyɔvore} \HM{} \PH{ɲɔ̀vʊ́rɪ̀} \PS{n.~} \GE{life} \OP{nyɔvoɛ} \TP{nyɔvorɛɛ} 
\LX{nyu} \HM{} \PH{ɲù} \PS{v.~} \GE{to smell (scent)} \XV{A zɛkpaane da nyu la noɔ lɛ.} \XE{The simmering sauce smelt nice.} \PD{-, -, nyue, nyuuro, nyuuro, nyuuraa } 
\LX{nyu} \HM{} \PH{ɲú} \PS{v.~} \SN \GE{to drink} \SN \GE{to suck (of a mosquito)} \PD{nyue, nyuuro, nyuuro, nyuuribo, nyuuraa, nyuuribo } 
\LX{nyu} \HM{} \PH{ɲú} \PS{v.~} \GE{to smoke (pipe)} \PD{nyue, nyuuro, nyuuro, nyuuribo, nyuuraa, nyuuribo } 
\LX{nyubuli} \HM{} \PH{ɲúbúlí} \PS{n.~} \GE{yam sprout} \OP{nyubulo} \TP{nyubulee} 
\LX{nyudɛre} \HM{} \PH{ɲúdɛ́rɪ́} \PS{n.~} \GE{large yam tuber} \OP{nyudɛɛ} \TP{nyudɛrɛɛ/nyudɛrrɛɛ} \VA{wadɛre}
\LX{nyumoɔ} \HM{} \PH{ɲúmʊ́ɔ́} \PS{n.~} \GE{field prepared to make yam mounds} \TP{nyumoɔnɛɛ} \VA{vutɛne}
\LX{nyunni} \HM{} \PH{ɲùnnì} \PS{v.~} \SN \GE{to smell} \XV{A zɔŋɔ nyunni la a kparoo, a yeli ka o deni ba e o.} \XE{The blind man smelt the shirt and said it was not his.} \SN \GE{to sniff e.g. dog} \PD{nyunnee, nyunno, nyunno, nyunimo, nyunnaa, nyunimo } 
\LX{nyuo} \HM{} \PH{ɲùó} \PS{n.~} \SN \GE{navel} \SN \GE{see wɛnyuo} \OP{nyuuri} \TP{nyuuree} 
\LX{nyusege} \HM{} \PH{ɲúsɪ́gɪ́} \PS{n.~} \GE{yam barn} \OP{nyusegre} \TP{nyusegrɛɛ} \VA{wasege}
\LX{nyuu} \HM{} \PH{ɲúú} \PS{n.~} \SN \GE{neck} \SN \GE{smell, odour} \OP{nyuuri} \TP{nyuuree} 
\LX{nyuu} \HM{} \PH{ɲùú} \PS{n.~} \GE{a sensation, good or bad, detected by inhaling} 
\LX{nyuu} \HM{} \PH{ɲúú} \PS{adj.~} \GE{stinky} \OP{nyuuri} \TP{nyuuree} 
\LX{nyuuli} \HM{} \PH{ɲúúlì} \PS{v.~} \GE{to empower supernaturally} \PD{nyuulee, nyuulo, nyuulo, nyuulibo, nyuulaa, - } 
\LX{nyu-uo} \HM{} \PH{ɲú-úó} \PS{n.~} \GE{yam barn} \OP{nyu-uori} \TP{nyu-uoree} \VA{nyu-sege}
\LX{nyuuri} \HM{} \PH{ɲúúrí} \PS{n.~} \GE{envy,  jealousy} \OP{n/a} \TP{nyuuree} 
\LX{nyuuri} \HM{} \PH{ɲúúrì} \PS{n.~} \GE{yam} \OP{nyie} \TP{nyiree/nyiiree} \VA{waa}
\LX{nyuweɛle} \HM{} \PH{ɲùwɪ́ɛ́lɪ́} \PS{n.~} \GE{yam seed} \SG{nyuweɛloo} \OP{nyuweɛlɛ} \TP{nyuweɛlɛɛ} \VA{waweɛlɛ}
\LX{nyuzɛle} \HM{} \PH{ɲùzɛ̀lɪ́} \PS{n.~} \GE{umbilical cord} \SG{nyuzɛloo} \OP{nyuzɛlɛ} \TP{nyuzɛlɛɛ/nyuzɛllɛɛ} 
\LX{o} \HM{} \PH{ʊ̀} \PS{pronoun.~} \GE{s/he, it (third person singular)} 
\LX{ɔge} \HM{} \PH{ɔ́gɪ̀} \PS{n.~} \GE{ground den, lair, hole} \OP{ɔgre} \TP{ɔgrɛɛ} 
\LX{ɔgere} \HM{} \PH{ɔ́gɪ́rɪ́} \PS{v.~} \GE{to struggle to vomit; to retch} \PD{ɔgrɛɛ, ɔgrɔ, ɔgrɔ, ɔgreba, ɔgraa, - } 
\LX{ɔgere} \HM{} \PH{ɔ́gɪ́rɪ̀} \PS{v.~} \GE{to get choked e.g during eating when a child is unable to swallow a muscle of food and strains to bring the food out} \PD{-, -, -, -, -, - } 
\LX{ogi} \HM{} \PH{ògì} \PS{v.~} \GE{to bring out of the mouth} \PD{ogee, ogro, ogro, ogribo, ograa, - } 
\LX{ogi} \HM{} \PH{ógì} \PS{v.~} \SN \GE{to segregate; to discriminate} \SN \GE{to dislocate (joint)} \PD{ogee, ogro, ogro, ogribo, ograa, - } 
\LX{ogili} \HM{} \PH{ògìlì} \PS{adj.~} \GE{not properly fitted or seated (lid, cover)} 
\LX{oii} \HM{} \PH{óìì} \PS{interjection.~} \GE{expression of despair} 
\LX{ɔlaa} \HM{} \PH{ɔ́làà} \PS{n.~} \GE{large black fish} \OP{ɔlle} \TP{ɔllɛɛ} 
\LX{oli-oli} \HM{} \PH{òlì-ólí} \PS{v.~} \GE{to be restless, to be unsettled} \PD{oli-olee, oli-olo, oli-olo, oli-olibo, oli-olaa, oli-olibo } 
\LX{ɔlɔ-ɔlɔ} \HM{} \PH{ɔ̀lɔ̀-ɔ̀lɔ̀} \PS{adj.~} \GE{slimy and dirty} 
\LX{ɔlɔ-ɔlɔ} \HM{} \PH{ɔ̀lɔ̀-ɔ̀lɔ̀} \PS{adj.~} \GE{purple} 
\LX{ɔmm} \HM{} \PH{ɔ̀mm} \PS{adv.~} \GE{idling} 
\LX{onaŋ} \HM{} \PH{ʊ́náŋ} \PS{pronoun.~} \GE{that one (nonhuman, third person demonstrative)} 
\LX{ɔŋe} \HM{} \PH{ɔ̀ŋ} \PS{v.~} \SN \GE{to draw something liquid e.g. water, to fetch something powdery e.g. flour, soil} \SN \GE{to harvest vegetables} \SN \GE{to harvest dawdaw fruits or leaves} \PD{ɔŋɛɛ, ɔnnɔ, ɔnnɔ, ɔnema, ɔnnaa, ɔnnema } 
\LX{ɔŋe} \HM{} \PH{ɔ̀ŋ} \PS{v.~} \GE{to smell (bad odour)} \PD{ɔŋɛɛ, ɔnnɔ, ɔnnɔ, ɔnema, ɔnnaa, ɔnnema } 
\LX{ɔŋ-eŋ} \HM{} \PH{ɔ̀ŋ-ɪ̀ŋ} \PS{v.~} \GE{to flee} \PD{ɔŋ-eŋɛɛ, ɔnnɔ-ennɛ, ɔŋ-ennɛ, ɔŋ-enemɛ, ɔŋ-ennaa, ɔŋ-enemɛ } 
\LX{oŋgoloŋ} \HM{} \PH{óŋgólóŋ} \PS{adj.~} \GE{circular, spherical in shape and small} 
\LX{ɔnne} \HM{} \PH{ɔ̀nnɪ̀} \PS{v.~} \GE{to smoke, to put under extreme heat (meat)} \PD{ɔnnɛɛ, ɔnnɔ, ɔnnɔ, ɔnema, ɔnnaa, - } 
\LX{onɔ} \HM{} \PH{ʊ́nɔ́} \PS{pronoun.~} \GE{s/he, it (third person, demonstrative)} \XV{Onɔ la na yi.} \XE{S/he is the one that will go out.} \VA{ona/one}
\LX{ɔŋɔ} \HM{} \PH{ɔ́ŋɔ́} \PS{n.~} \GE{wild mouse} \OP{ɔnne} \TP{ɔnnɛɛ} \VA{ɔŋa}
\LX{ɔŋ-ɔɔbo} \HM{} \PH{ɔ̀ŋ-ɔ́ɔ́bʊ́} \PS{n.~} \GE{valuable, useful} 
\LX{ɔnsɔɔre} \HM{} \PH{ɔ̀nsɔ́ɔ́rɪ́} \PS{n.~} \GE{} \OP{ɔnsɔbɔ} \VA{ɔŋ-sɔɔre}
\LX{ɔŋ-sɔɔre} \HM{} \PH{ɔ̌ŋ-sɔ̊́ɔ́rɪ́} \PS{n.~} \GE{a type of wild mouse that lives along river banks; has a dark brown colour} \OP{ɔŋ-sɔbɔ} \TP{ɔŋ-sɔɔrɛɛ} \VA{ɔnsɔɔre}
\LX{oo} \HM{} \PH{òó} \PS{use word.~} \GE{yes (acceptance)} 
\LX{ɔɔ} \HM{} \PH{ɔ̀ɔ̀} \PS{v.~} \SN \GE{to chew} \SN \GE{to ache; to throb with pain} \PD{ɔɔɛ, ɔɔrɔ, ɔɔrɔ, ɔɔreba, ɔɔraa, ɔɔreba } 
\LX{ɔɔ-doŋɔ} \HM{} \PH{ɔ̀ɔ̀-dʊ́ŋó} \PS{n.~} \SN \GE{intestinal worm} \SN \GE{roundworm} \OP{ɔɔ-donne} \TP{ɔɔ-donnɛɛ} 
\LX{ooi} \HM{} \PH{óóì} \PS{use word.~} \GE{yes (response to call)} 
\LX{ɔɔkpoŋi} \HM{} \PH{ɔ́ɔ́kpóŋ} \PS{n.~} \GE{winter} \OP{ɔɔkponni} \TP{ɔɔkponnee} 
\LX{ɔɔkyi} \HM{} \PH{ɔ́ɔ́ʧí} \PS{n.~} \GE{goose bumps i.e. small bumps on skin caused by cold} \OP{ɔɔkyiree} 
\LX{oɔle} \HM{} \PH{ʊ́ɔ́lɪ́} \PS{v.~} \GE{to happen gradually e.g. stomach ache} \PD{oɔlɛɛ, oɔlɔ, oɔlɔ, oɔlba, oɔlaa, - } 
\LX{ɔ-ɔlle} \HM{} \PH{ɔ́-ɔ́llɪ̀} \PS{n.~} \GE{moss (usually green matter found in unkempt wet places e.g bath house)} \TP{ɔ-ɔllɛɛ} 
\LX{oone} \HM{} \PH{ʊ̀ʊ̀nɪ̀} \PS{v.~} \SN \GE{to clear one's throat} \SN \GE{to push (giving birth)} \PD{oonɛɛ, oona, oona, oonema, oonaa, - } 
\LX{oɔne} \HM{} \PH{ʊ̀ɔ̀nɪ̀} \PS{v.~} \GE{to grip firmly with claws} \PD{oɔnɛɛ, oɔnɔ, oɔnɔ, oɔnema, oɔnaa, - } 
\LX{ɔɔnyuko᷉ɔ} \HM{} \PH{ɔ́ɔ́nyúkʊ́᷉ɔ́} \PS{n.~} \GE{a type of tree with edible fruits} \OP{ɔɔnyuko᷉ɔnɛɛ} \TP{ɔɔnyukoɔnoo} 
\LX{ɔɔnyuko᷉ɔ} \HM{} \PH{ɔɔnyúkʊ̀᷉ɔ́᷉} \PS{n.~} \GE{a type of small tree whose leaves are edible and have medicinal properties} \TP{ɔɔnyukoɔnɛɛ} 
\LX{ɔɔraa} \HM{} \PH{ɔ́ɔ́ráá} \PS{n.~} \GE{chewer} \OP{ɔɔreba} \TP{ɔɔrebɛɛ} 
\LX{ɔɔraa-piiri} \HM{} \PH{ɔ́ɔ́ráá-pìírí} \PS{n.~} \GE{safe abode} \OP{ɔɔraa-pie} 
\LX{oore} \HM{} \PH{ʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to dash up against} \PD{oorɛɛ, oora, oora, ooreba, ooraa, - } 
\LX{ɔɔre} \HM{} \PH{ɔ́ɔ́rɪ̀} \PS{n.~} \GE{cold} \OP{ɔɔrɛɛ} 
\LX{ɔɔtoorebare} \HM{} \PH{ɔ̀ɔ̀tʊ́ʊ́rɪ́bàrɪ̀} \PS{n.~} \GE{October (month)} 
\LX{ɔraa} \HM{} \PH{ɔ́ràà} \PS{n.~} \GE{a type of climbing plant with edible fruit} \OP{ɔrrɪ} \TP{ɔrrɛɛ} 
\LX{ɔraa} \HM{} \PH{ɔ́ràà} \PS{n.~} \GE{portable cage (for carrying chicks)} \OP{ɔrrɪ} \TP{ɔrrɛɛ} 
\LX{ore} \HM{} \PH{ʊ́rɪ́} \PS{v.~} \SN \GE{to wash very dirty things e.g clothes} \SN \GE{to make one suffer severely e.g. beat} \PD{-, -, -, -, -, - } 
\LX{ore} \HM{} \PH{ʊ́rɪ́} \PS{adj.~} \GE{filled to the brim} 
\LX{ɔre} \HM{} \PH{ɔ̀rɪ̀} \PS{v.~} \GE{to prepare bean leaf broth for soup} \PD{ɔrɛɛ, ɔrɔ, ɔrɔ, ɔreba, ɔraa, ɔreba } 
\LX{ɔre} \HM{} \PH{ɔ̀rɪ̀} \PS{v.~} \GE{to throw hard against e.g. ball against a player} \PD{ɔrɛɛ, ɔrɔ, ɔrɔ, ɔreba, ɔraa, ɔreba } 
\LX{ɔre} \HM{} \PH{ɔ̀rɪ́} \PS{adj.~} \GE{impotent} \OP{ɔrɔ} \TP{ɔrɛɛ} 
\LX{ɔre} \HM{} \PH{ɔ́rɪ́} \PS{v.~} \SN \GE{to begin to rot} \SN \GE{to stink} \SN \GE{to be useless (person)} \PD{ɔrɛɛ, ɔrɔ, ɔrɔ, ɔreba, ɔraa, ɔreba } 
\LX{oree} \HM{} \PH{óre̊e̊} \PS{adj.~} \GE{round and open looking e.g. a hole} 
\LX{ɔregbɔrɔ} \HM{} \PH{ɔ̀rɪ̀gbɔ̀rɔ̀} \PS{adv.~} \GE{messy looking} 
\LX{ɔresɔglaa} \HM{} \PH{ɔ́rɪ́sɔ́gláà} \PS{n.~} \GE{a type of climbing plant with edible fruit} \OP{ɔresɔglɔ} 
\LX{orɔba} \HM{} \PH{ʊ̂rɔ́bà} \PS{n.~} \GE{rubber} \OP{orɔbare} \TP{orɔbarɛɛ} 
\LX{ɔrɔsɔ} \HM{} \PH{ɔ̀rɔ̀sɔ̀} \PS{adv.~} \GE{paying dearly for being obstinate} \XV{O yeli ka o ba wono yɛlɛ, o taraa be ɔrɔsɔ.} \XE{He says he is hardheaded; he is up to suffer for it.} 
\LX{ó᷉̊ó} \HM{} \PH{ʊ́᷉-ʊ̀᷉} \PS{interjection.~} \GE{expression of surprise} 
\LX{o᷉o} \HM{} \PH{ʊ́᷉ʊ́᷉} \PS{use word.~} \GE{yes (agreement)} \VA{ɛ᷉ɛ}
\LX{o᷉ɔ} \HM{} \PH{ʊ́᷉ɔ́᷉} \PS{n.~} \GE{a pot with small holes all over; colander} \OP{oɔne} \TP{oɔnɛɛ} 
\LX{o᷉-o᷉} \HM{} \PH{ʊ́-ʊ̂} \PS{use word.~} \GE{why so?} 
\LX{o᷉-o᷉} \HM{} \PH{ʊ́-ʊ̀} \PS{use word.~} \GE{oh no} 
\LX{o᷉-o᷉o} \HM{} \PH{ʊ́᷉-ʊ́᷉ʊ́᷉} \PS{interjection.~} \GE{no (disagreement)} 
\LX{pa} \HM{} \PH{pá} \PS{v.~} \GE{to set up the beginning of something e.g. weaving} \PD{paɛ, paara, paara, pareba, paaraa, - } 
\LX{pa} \HM{} \PH{pá} \PS{interjection.~} \GE{an expression of surprise} \VA{paa}
\LX{pa} \HM{} \PH{pà} \PS{v.~} \GE{to be full} \PD{paɛ, paara, paara, pareba, paaraa, - } 
\LX{paa} \HM{} \PH{páá} \PS{interj.~} \GE{expression of fear and surprise} \VA{pa}
\LX{paalaa} \HM{} \PH{pááláá} \PS{adj.~} \GE{new} \OP{paale} \TP{paalɛɛ} 
\LX{paalaa} \HM{} \PH{pááláá} \PS{n.~} \SN \GE{raised dais, platform} \SN \GE{stage for laying a corpse in state} \OP{paale} \TP{paalɛɛ} 
\LX{paalaa} \HM{} \PH{pààláà} \PS{n.~} \GE{space in between four mounds} \OP{paale} \TP{paalɛɛ} 
\LX{paale} \HM{} \PH{pààlɪ̀} \PS{v.~} \SN \GE{to fill} \SN \GE{to add} \PD{paalɛɛ, paala, paala, paaleba, -, - } 
\LX{paaloŋ} \HM{} \PH{páálʊ́ŋ} \PS{n.~} \GE{area of jurisdiction (country, district, kingdom)} 
\LX{paaloŋ} \HM{} \PH{páálʊ́ŋ} \PS{adj.~} \GE{first, fresh instance} 
\LX{paalonwuluugane} \HM{} \PH{páálʊ́nwúlúúgánɪ̀} \PS{n.~} \GE{map} \OP{paalonwuluugama} 
\LX{paanaa} \HM{} \PH{pàànáá} \PS{n.~} \GE{ground cricket} \OP{paane} \TP{paanɛɛ} \VA{paaŋaa - mole cricket (Durant)}
\LX{paane} \HM{} \PH{pàànɪ̀} \PS{v.~} \GE{to blame} \PD{paanɛɛ, paana, paana, paanema, paanaa, - } 
\LX{paantakyeraa} \HM{} \PH{pààntáʧɪ́ráá} \PS{n.~} \GE{wild flying cricket} \OP{paantakyerre} 
\LX{paapaa} \HM{} \PH{páápàà} \PS{n.~} \GE{pope} \OP{paapaamine/paapaare} \TP{paapaarɛɛ} 
\LX{paare} \HM{} \PH{páárɪ̀} \PS{n.~} \GE{vagina, vulva} \OP{paɛ} \TP{pɛɛrɛɛ} 
\LX{paare} \HM{} \PH{pààrɪ̀} \PS{v.~} \GE{to scrape e.g. roasted yam} \PD{paarɛɛ, paara, paara, paareba, paaraa, - } 
\LX{paare} \HM{} \PH{pààrɪ̀} \PS{v.~} \GE{to weed e.g. grass} \PD{paarɛɛ, paara, paara, paareba, paaraa, - } 
\LX{paare} \HM{} \PH{páárɪ̀} \PS{v.~} \GE{to pass (a test)} \PD{paarɛɛ, paara, paara, paareba, paaraa, - } 
\LX{paasepɔɔto} \HM{} \PH{páásɪ́pɔ́ɔ́tʊ̀} \PS{n.~} \GE{passport} \OP{paasepɔɔtere} \TP{paasepɔɔterɛɛ} 
\LX{paate} \HM{} \PH{páátɪ̀} \PS{interjection.~} \GE{expression of surprise astonishment e.g. one found in a challenging situation} 
\LX{paawa} \HM{} \PH{pááwà} \PS{n.~} \GE{power, electric light} \OP{paaware} \TP{paawarɛɛ} 
\LX{paawerɛ} \HM{} \PH{pááwɪ́rɛ̀} \PS{n.~} \GE{powder} \OP{paawerre} \TP{paawerɛɛ/paawerrɛɛ} 
\LX{pabo} \HM{} \PH{pábʊ́} \PS{adj.~} \GE{completely} 
\LX{padaa} \HM{} \PH{pàdàá} \PS{n.~} \GE{wood used to make a stage for a corpse} \OP{padaare} \TP{padaarɛɛ} 
\LX{page} \HM{} \PH{págɪ́} \PS{v.~} \GE{to hasten, to hurry} \PD{pagɛɛ, pagera, pagereba, pageraa, pagereba, pagepagera } 
\LX{pagere} \HM{} \PH{págrɪ̀} \PS{v.~} \GE{to compete, to challenge} \PD{pagrɛɛ, pagra, pagra, pagreba, pagraa, - } 
\LX{pakɔɔloŋ} \HM{} \PH{pákɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{female pubic hair} 
\LX{pakyoŋ} \HM{} \PH{pàʧʋ́ŋ} \PS{n.~} \GE{a large black ant with a painful sting} \OP{pakyonne} \TP{ pakyonnɛɛ.} 
\LX{palampiiri} \HM{} \PH{pálámpììrí} \PS{n.~} \SN \GE{butterfly} \SN \GE{moth} \OP{palampie} \TP{palampiiree} \VA{palampigri/palimpiiri}
\LX{palle} \HM{} \PH{pàllɪ̀} \PS{v.~} \GE{to plait, to braid} \PD{pallɛɛ, palla, palla, paleba, pallaa, - } 
\LX{pamm} \HM{} \PH{pámm} \PS{adj.~} \GE{straight onto the target} 
\LX{pampana} \HM{} \PH{pámpánà} \PS{adv.~} \GE{now} 
\LX{pampuntɛgɛ} \HM{} \PH{pámpùntɛ̀gɛ́} \PS{n.~} \GE{the act of being inward looking; concerned with only immediate relation} \OP{pampuntɛgre} 
\LX{panana} \HM{} \PH{pànànà} \PS{adv.~} \GE{lit up e.g. car headlight in the dark} 
\LX{panapana} \HM{} \PH{pànàpànà} \PS{adj.~} \GE{colourful, beautiful} 
\LX{pane} \HM{} \PH{pánɪ́} \PS{n.~} \GE{doorway cover made of straw} \OP{pama} \TP{panɛɛ} 
\LX{paŋe} \HM{} \PH{pàŋ} \PS{v.~} \SN \GE{to beat down wet rough surface to make it even (local way to plaster) e.g floor} \SN \GE{fig. walk without any footwear on, barefooted} \PD{paŋɛɛ, pana, pana, panema, panaa, - } 
\LX{paŋe} \HM{} \PH{páŋ} \PS{v.~} \SN \GE{to beat mercilessily} \SN \GE{to gather plenty of leaves} \SN \GE{to sew (akn)} \PD{paŋɛɛ, pana, pana, panema, panaa, - } 
\LX{papeɛmɛ} \HM{} \PH{pápɪ́ɛ́mɛ́} \PS{n.~} \GE{waist string with cowries} \OP{papeɛne} \TP{papeɛnɛɛ} 
\LX{papoɔ} \HM{} \PH{pàpʊ́ɔ́} \PS{n.~} \GE{space in between mounds} \OP{papoore} \TP{papoorɛɛ} 
\LX{papoɔ} \HM{} \PH{pápʊ́ɔ́} \PS{n.~} \GE{inner vagina} 
\LX{para} \HM{} \PH{párá} \PS{n.~} \GE{triangular loincloth} \OP{parre} \TP{parɛɛ} 
\LX{para} \HM{} \PH{pàrá} \PS{v.~} \GE{to walk quickly} \PD{parɛɛ, para, para, pareba, paraa, - } 
\LX{parapara} \HM{} \PH{pàràpàrà} \PS{adv.~} \GE{hurried manner} \XV{A tontonema da nyɔge la a toma paraparapara, ka ba baare a.} \XE{The worker started the work hurriedly and soon finished.} 
\LX{parata} \HM{} \PH{párátá} \PS{adj.~} \GE{white and spotless} \VA{perete}
\LX{pare} \HM{} \PH{párɪ̀} \PS{n.~} \GE{buttocks, bottom} \OP{parɛɛ} \VA{tɛgɛ}
\LX{pare} \HM{} \PH{párɪ̀} \PS{n.~} \GE{meaning} \XV{A yɛlɛ ŋa pare la wola?} \XE{What is the meaning of this?.} \XV{A yɛlɛ pare da ba kyaane.} \XE{The meaning of what was said was not clear.} \OP{parɛɛ} \VA{tɛgɛ}
\LX{pare} \HM{} \PH{párɪ̀} \PS{v.~} \GE{to understand} \XV{A bie da ba woŋi a yɛlɛ pare.} \XE{The child did not understand what was said.} \PD{-, -, -, -, -, - } 
\LX{pare} \HM{} \PH{pàrɪ́} \PS{n.~} \GE{raised tomb} \OP{parɛɛ} 
\LX{pare} \HM{} \PH{párɪ̀} \PS{postposition.~} \GE{under} \XV{A tontonema balɛɛ la gaa zeŋe teɛ pare ka ba pɛnne.} \XE{The workers were tired and sat under the tree to rest.} 
\LX{pare} \HM{} \PH{pàrè} \PS{v.~} \SN \GE{to place on top} \SN \GE{to overtake, to pass} \PD{parɛɛ, para, para, pareba, paraa, pareba } 
\LX{pare} \HM{} \PH{pàrè} \PS{adj.~} \GE{more than, over} 
\LX{pare-araa} \HM{} \PH{párɪ̀-áráá} \PS{n.~} \GE{the act of having time} \XV{N ba taa pare-araa na gaa be zenɛ, n toma yaga.} \XE{I have no time to go there today, because I have a lot of work.} 
\LX{parebaaraa} \HM{} \PH{párbàáráà} \PS{n.~} \GE{last one} \OP{parbaare} 
\LX{parebiri} \HM{} \PH{párɪ̀bírì} \PS{n.~} \GE{anus} \OP{parebie} \TP{parebiiree} \VA{parebogi}
\LX{parebogi} \HM{} \PH{párbógì} \PS{n.~} \GE{anus} \OP{parebogiri} \TP{parebogiree} \VA{parebiri}
\LX{parɛɛ} \HM{} \PH{párɛ́ɛ́} \PS{n.~} \GE{species; kind} \OP{parɛɛ parɛɛ} 
\LX{parefe᷉ɛ} \HM{} \PH{párɪ́fɪ́᷉ɛ́} \PS{adv.~} \GE{in vain} \XV{N viiri la o yiri te bale kyɛ ka parefẽɛ.} \XE{I have gone to his house several times but all without success.} 
\LX{parefuri} \HM{} \PH{párfùrì} \PS{n.~} \GE{act of dragging one's feet e.g. cowardice} 
\LX{parekolee} \HM{} \PH{pàrèkòlèé} \PS{n.~} \GE{piglet} \OP{parekobilii} \VA{dobalee, dobabilii}
\LX{parekonɛne} \HM{} \PH{pàrɪ́kònɛ́nɪ̀} \PS{n.~} \GE{pork} \OP{parekonɛmɛ} \TP{porkonɛnnɛɛ} 
\LX{parekyara} \HM{} \PH{párʧárá} \PS{n.~} \GE{The part in between the legs e.g. when in astride position} \OP{parekyarre} \TP{parekyarɛɛ} \VA{parekyɛrɛ}
\LX{parekyɛrɛ} \HM{} \PH{párʧɛ́rɛ̀} \PS{n.~} \GE{} \OP{ parekyɛrre} \TP{parekyɛrɛɛ} \VA{parekyara}
\LX{parema} \HM{} \PH{párɪ́má} \PS{n.~} \GE{extra} 
\LX{parenyɔge} \HM{} \PH{párɲɔ́gɪ́} \PS{n.~} \GE{courage} 
\LX{parepe᷉ɛ} \HM{} \PH{párɪ̀pɪ̀ɛ̀} \PS{n.~} \GE{act of struggling it out with another} \XV{Ba yɛlɛ na e la parepe᷉ɛ yɛlɛ.} \XE{That problem of theirs is a difficult one.} 
\LX{parepe᷉ɛ᷉} \HM{} \PH{párɪ̀pɪ̀᷉ɛ̀᷉} \PS{n.~} \GE{struggle} 
\LX{parepi᷉i} \HM{} \PH{párɪ́pí᷉ì} \PS{n.~} \GE{the end of the anus} \OP{parepiime} 
\LX{parepoɔroŋ} \HM{} \PH{párɪ́pʊ́ɔ́rʊ́ŋ} \PS{n.~} \GE{a type of wild mouse that reproduces prolifically} \OP{parepoɔrɔ} \TP{parepoɔrɛɛ} 
\LX{parepu᷉o᷉} \HM{} \PH{párɪ̀pù᷉ò᷉} \PS{n.~} \GE{fear} 
\LX{paresɛgere} \HM{} \PH{párɪ̀sɛ́grɪ̀} \PS{n.~} \GE{buttock} \OP{paresɛgɛ} \TP{paresɛgerɛɛ} 
\LX{paresunsugiri} \HM{} \PH{párɪ̀súnsúgrí} \PS{n.~} \GE{the knob-like appendage attached to the area above the anus of birds} \OP{paresunsugo} \TP{paresunsugree} 
\LX{paretaa} \HM{} \PH{pàrɪ̀táà} \PS{n.~} \GE{game} \TP{paretaarɛɛ} 
\LX{paretɛleŋ} \HM{} \PH{párɪ́tɛ́lɪ́ŋ} \PS{adj.~} \GE{at once} 
\LX{parikonyaŋaa} \HM{} \PH{pàríkóɲáŋáá} \PS{n.~} \GE{sow (pig)} \OP{parikonyaŋene} 
\LX{pataase} \HM{} \PH{pàtáásɪ́} \PS{n.~} \GE{type of local spirit (alcoholic drink)} \OP{pataasere} \TP{pataaserɛɛ} 
\LX{pazɔŋɔ} \HM{} \PH{pàzɔ̌ŋ} \PS{n.~} \GE{blood-sucking fly} \OP{pazɔnne} \TP{pazɔnnɛɛ} \VA{pazɔŋa}
\LX{pa᷉a} \HM{} \PH{pá᷉à᷉} \PS{conjunction.~} \GE{then, and now} \XV{O waɛ la pa᷉a wuli o a toma.} \XE{He has come, you should then show him the work.} 
\LX{pa᷉a} \HM{} \PH{pá᷉à} \PS{adj.~} \GE{what else} \XV{Libie kyebe N pa᷉a la e wolɔ?} \XE{There is no money, what else can I do?.} 
\LX{pa᷉a} \HM{} \PH{pà᷉à᷉} \PS{v.~} \GE{to warn, to caution} \PD{pa᷉aɛ, paana, paana, paanema, paanaa, - } 
\LX{pa᷉apere} \HM{} \PH{pa᷉a᷉pɪ́rɪ́} \PS{n.~} \GE{paperlike spider egg sac} \OP{pa᷉ape᷉ɛ} \TP{pa᷉apeerɛɛ} 
\LX{pa᷉atakyeraa} \HM{} \PH{pá᷉á᷉táʧɪ́ráá} \PS{n.~} \GE{a type of cricket} \OP{pa᷉atakyerre} \TP{pa᷉atakyerɛɛ} 
\LX{pe} \HM{} \PH{pɪ́} \PS{v.~} \GE{to be fed up with} \PD{peɛ, peerɛ, peerɛ, peerebɛ, peeraa, peerebɛ } 
\LX{pɛ} \HM{} \PH{pɛ̀} \PS{v.~} \GE{to squeeze out e.g. milk from a cow} \PD{-, -, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{pɛ} \HM{} \PH{pɛ̀} \PS{v.~} \GE{to open one's eyes widely} \PD{-, -, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{pɛ} \HM{} \PH{pɛ̀} \PS{v.~} \GE{to spiritually deprive another of prowess} \PD{-, -, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{pɛ} \HM{} \PH{pɛ́} \PS{v.~} \GE{to harvest nuts from the ground} \PD{pɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{pɛ} \HM{} \PH{pɛ́} \PS{adj.~} \GE{same, equal} \VA{pɛpɛ}
\LX{pedaa} \HM{} \PH{pɪ́dáá} \PS{n.~} \GE{ram} \OP{padaare} \TP{pedaɛ/padaarɛɛ} 
\LX{peda-eelɛ} \HM{} \PH{pɪ́dá-ɪ́ɪ́lɛ́} \PS{n.~} \GE{male sheep with horns} 
\LX{pedavaraa} \HM{} \PH{pídàváràà} \PS{n.~} \GE{castrated ram} \OP{pedavarre} \TP{pedavarɛɛ} 
\LX{pee} \HM{} \PH{pèè} \PS{adv.~} \GE{open and large} 
\LX{peɛ} \HM{} \PH{pɪ́ɛ́} \PS{n.~} \GE{roof} \OP{peere} \TP{peerɛɛ} 
\LX{peɛ} \HM{} \PH{pɪ̀ɛ́} \PS{n.~} \GE{basket} \OP{pɛre} \TP{pɛrɛɛ} 
\LX{peɛ} \HM{} \PH{pɪ̀ɛ̀} \PS{v.~} \SN \GE{to drink (taking in large chunks)} \SN \GE{to take a wife or house maid from her parents} \PD{peɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{peelaa} \HM{} \PH{pɪ́ɪ́láà} \PS{adj.~} \GE{infant} \OP{peele} 
\LX{peele} \HM{} \PH{pɪ́ɪ́lɪ̀} \PS{v.~} \GE{to tear into shreds e.g. rag} \PD{peelɛɛ, peelɛ, peelɛ, peelbɛ, peelaa, peelbɛ } 
\LX{peɛle} \HM{} \PH{pɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{} \PD{peɛlɛɛ, peɛlɛ, peɛlɛ, peɛlebɛ, peɛlaa, peɛlebɛ } \SE{peɛle eelee} \GE{to blow a horn} 
\LX{peɛle} \HM{} \PH{pɪ́ɛ́lɪ̀} \PS{v.~} \GE{to be near; to be close to} \PD{peɛlɛɛ, peɛlɛ, peɛlɛ, peɛlebɛ, peɛlaa, peɛlebɛ } 
\LX{peɛle-peɛle} \HM{} \PH{pɪ́ɛ́lɪ́-pɪ́ɛ́lɪ́} \PS{adj.~} \GE{having white spots} 
\LX{peɛloŋ} \HM{} \PH{pɪ́ɛ́lʊ́ŋ} \PS{n.~} \SN \GE{whiteness} \SN \GE{food} 
\LX{peɛmɛ} \HM{} \PH{pɪ́ɛ́mɛ́} \PS{n.~} \SN \GE{decorations} \SN \GE{beautiful clothing} \OP{peɛne} \TP{peɛnɛɛ/peɛmɛɛ} 
\LX{peɛmo} \HM{} \PH{pɪ́ɛ́mʊ́} \PS{n.~} \SN \GE{courtship} \SN \GE{sexual intercourse} 
\LX{peere} \HM{} \PH{pɪ̀ɪ̀rɪ̀} \PS{v.~} \SN \GE{to sweep, to dust, to brush off (shirt)} \SN \GE{to cleanse; to reinstate someone's soul believed to have deserted her/him by performing certain rituals to cleanse the person of an evil spirit, which is believed to have overshadowed the person’s actual soul} \PD{peerɛɛ, peerɛ, peerebɛ, peeraa, -, - } 
\LX{peɛre} \HM{} \PH{pɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to investigate} \PD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{peɛre} \HM{} \PH{pɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to not be fussy about choices (typically used with negative marker ba)} \PD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{peɛre} \HM{1} \PH{pɪ́ɛ́rɪ́} \PS{v.~} \GE{to line up; to place side by side} \PD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{peɛre} \HM{2} \PH{pɪ́ɛ́rɪ́} \PS{v.~} \GE{to tap (to call attention)} \PD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\LX{peɛrɛ} \HM{} \PH{pɪ́ɛ́rɛ̀} \PS{n.~} \GE{a type of small, black potato} \SG{peɛroo} \OP{peɛrɛ} \TP{peɛrɛɛ} 
\LX{pɛɛre} \HM{} \PH{pɛ́ɛ́rɪ́} \PS{adj.~} \GE{immature} \OP{pɛbɛ} 
\LX{pɛɛtɛɛre} \HM{} \PH{pɛ́ɛ́tɛ́ɛ́rɪ́} \PS{n.~} \GE{small, flat fish; tilapia} \SG{pɛɛtɛroo} \OP{pɛɛtɛbɛ} \TP{pɛɛtɛbɛɛ} \VA{pɛɛtere}
\LX{peeteree} \HM{} \PH{pɪ́ɪ́tɪ́rɪ́ɪ́} \PS{adj.~} \GE{everything about something, the whole truth} \XV{A polisiri da soore la noba yaga a na baŋe a yɛlɛ peeteree zaa.} \XE{The police questioned many people in a bid to get to the bottom of the matter.} \TP{peeterɛɛ} 
\LX{pege} \HM{} \PH{pɪ̀gɪ̀} \PS{v.~} \SN \GE{to decorate (dress)} \SN \GE{to make tasty, delicious} \PD{pegɛɛ, pegrɛ, pegrɛ, pegrebɛ, pegraa, - } 
\LX{pegɛ} \HM{} \PH{pɪ́gɛ́} \PS{adj.~} \GE{not whole; e.g. broken piece of calabash} 
\LX{pɛge} \HM{} \PH{pɛ́gɪ̀} \PS{v.~} \GE{to wash e.g. clothes, utensils} \PD{-, pɛgrɛ, pɛgrɛ, pɛgrebɛ, -, - } 
\LX{pɛge} \HM{} \PH{pɛ́gɪ́} \PS{n.~} \GE{hard external covering of seeds, nuts, animals, trees} \OP{pɛgre/pɛgɛ} \TP{pɛgrɛɛ} 
\LX{pɛge yi} \HM{} \PH{pɛ́gɪ̀ yí} \PS{adj.~} \GE{clean, bright} 
\LX{pɛgɛkɛ} \HM{} \PH{pɛ̀gɛ̀kɛ̀} \PS{adj.~} \GE{deep and wide} 
\LX{pegelaa} \HM{} \PH{pìgláà} \PS{n.~} \GE{row (of farm mounds)} \OP{pegle} \TP{peglɛɛ} 
\LX{pegele} \HM{} \PH{pɪ̀glɪ̀} \PS{v.~} \GE{cut across, in order to meet at a point} \PD{peglɛɛ, peglɛ, peglɛ, peglebɛ, peglaa, - } 
\LX{pegele} \HM{} \PH{pɪ̀glɪ̀} \PS{adj.~} \GE{not staight on e.g. feet} 
\LX{pɛgele} \HM{} \PH{pɛ̀glɪ̀} \PS{v.~} \GE{to carry on one's side, shoulder, in arms} \PD{pɛglɛɛ, pɛglɛ, pɛglɛ, pɛglebɛ, pɛglaa, - } 
\LX{pegerɛ} \HM{} \PH{pɪ́grɛ̀} \PS{n.~} \GE{wings} \SG{pegroo} 
\LX{pegi} \HM{} \PH{pègì} \PS{v.~} \GE{to flap one's wings} \PD{pegee, pegre, pegre, pegribe, pegraa, - } 
\LX{pɛglɛŋ} \HM{} \PH{pɛ́glɛ́ŋ} \PS{n.~} \GE{small and flat} \OP{pɛglɛŋ pɛglɛŋ} \VA{pɛŋgɛlɛŋ}
\LX{pɛkogo} \HM{} \PH{pɛ̀kógó} \PS{n.~} \GE{sticks put under a basket as a protective base} \OP{pɛkogri} \TP{pɛkogree} 
\LX{pekɔɔloŋ} \HM{} \PH{pɪ́kɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{wool} 
\LX{pɛkpoŋ} \HM{} \PH{pɛ̀kp̂oŋ} \PS{n.~} \GE{large basket} \OP{pɛkponni} \TP{pɛkponnee} 
\LX{pekpoŋi} \HM{} \PH{pɪ̀kpôŋ} \PS{n.~} \GE{large roof} \OP{pekponni} \TP{pekponnee} 
\LX{pekyeenɛ} \HM{} \PH{ṕʧɪ́ɪ́nɛ́} \PS{n.~} \GE{shepherd} \OP{pekyeemɛ} \TP{pekyeenɛɛ} 
\LX{pelaa} \HM{} \PH{pɪ́làá} \PS{adj.~} \GE{white} \OP{peɛle} 
\LX{pɛlaa} \HM{} \PH{pɛ́làà} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{pɛlle} \TP{pɛllɛɛ} 
\LX{pelaaya (Eng)} \HM{} \PH{pɪ́lááyà} \PS{n.~} \GE{pliers} \OP{pelaayare} \TP{pelaayarɛɛ} 
\LX{pele} \HM{} \PH{pɪ́lɪ̀} \PS{v.~} \SN \GE{to split a piece from a stalk} \SN \GE{to leave and fail to return} \PD{pelɛɛ, pellɛ, pellɛ, pelebɛ, pelaa, - } 
\LX{pele} \HM{} \PH{pɪ́lɪ́} \PS{v.~} \SN \GE{to tear, to break off (splinter)} \SN \GE{to discontinue (visits)} \SN \GE{to continue without going back to former place} \PD{pelɛɛ, pellɛ, pellɛ, pelbɛ, pellaa, - } 
\LX{pɛle} \HM{} \PH{pɛ̀lɪ̀} \PS{v.~} \GE{to be white} \PD{-, -, -, -, -, - } 
\LX{pɛle} \HM{} \PH{pɛ̀lɪ̀} \PS{v.~} \GE{to open up; be happy} \XV{Zenɛ o pɛle la poɔ manne a yɛlɛ ko te.} \XE{Today s/he open up by explaining the issues to us.} \PD{-, -, -, -, -, - } 
\LX{pɛle} \HM{} \PH{pɛ́lɪ́} \PS{v.~} \SN \GE{to peel (yam)} \SN \GE{to clear grass by weeding with a hoe} \PD{-, -, -, -, -, - } 
\LX{pelee} \HM{} \PH{pɪ́léè} \PS{n.~} \GE{lamb} \OP{pebilii} 
\LX{pelee} \HM{} \PH{pèléè} \PS{n.~} \GE{small basket} \OP{ pebilii} \VA{pɛlee}
\LX{peleemaze᷈e᷈} \HM{} \PH{pɪ́léèmàzɪ́᷈ɪ̀᷈} \PS{n.~} \GE{a type of with pink coloured edible fruits} 
\LX{pelempeleŋ} \HM{} \PH{pélémpéléŋ} \PS{adv.~} \GE{bare, empty} 
\LX{pelɛmpelɛŋ} \HM{} \PH{pɪ̀lɛ̀mpɪ́lɛ́ŋ} \PS{adv.~} \GE{all over the place, disorganised} 
\LX{pelewelee} \HM{} \PH{pɪ̀liɪ̀wɪ̀lɪ̀ɪ̀} \PS{adj.~} \GE{plenty of things found everywhere e.g. assorted items} \XV{A weɛ tomɔ deɛ waa la pelewelee lɛ.} \XE{There is plenty of work on the farm.} \VA{pelewele}
\LX{peli} \HM{} \PH{pélì} \PS{n.~} \GE{spirit, heart} \OP{pele} \TP{pelee} 
\LX{peli} \HM{} \PH{pél} \PS{n.~} \GE{fright} 
\LX{pell} \HM{} \PH{pɪ́ll} \PS{adj.~} \GE{full to capacity} \VA{pelpel}
\LX{pell} \HM{} \PH{péll} \PS{adj.~} \GE{rise into the air e.g. flame} 
\LX{pelɔɔ} \HM{} \PH{pɪ̀lɔ̀ɔ̀} \PS{adv.~} \GE{suddenly and quickly e.g fish splashing in water} 
\LX{pelpel} \HM{} \PH{pɪ́lpɪ́l} \PS{adv.~} \GE{full to the brim} 
\LX{pelpele} \HM{} \PH{pìlpɪ́l} \PS{v.~} \GE{to litter with broken pieces e.g. of straw} \PD{pelpelɛɛ, pelpelɛ, pelpelɛ, pelpelebɛ, pelpelaa, - } 
\LX{pempeleŋ} \HM{} \PH{pémpéléŋ} \PS{adj.~} \GE{small e.g. piece of cloth} 
\LX{pɛmpɛlɛŋ} \HM{} \PH{pɛ́mpɛ́lɛ́ŋ} \PS{n.~} \GE{flat metal (cover of a tin)} \OP{pɛmpɛlle} \TP{pɛmpɛlɛɛ} 
\LX{pɛmpɛnɛɛ} \HM{} \PH{pɛ́mpɛ́nɛ́ɛ́} \PS{n.~} \GE{wild gum; rubber} \OP{pɛmpɛnne} \TP{pɛmpɛnnɛɛ} 
\LX{pɛmpɛre} \HM{} \PH{pɛ̀mpɛ̀rɪ̀} \PS{n.~} \GE{oblique position (bending very low)} 
\LX{pɛmpɛre} \HM{} \PH{pɛ̀mpɛ̀rɪ̀} \PS{v.~} \GE{to bend very low and in an oblique position} \PD{pɛmpɛrɛɛ, pɛmpɛrɛ, pɛmpɛrɛ, pɛmpɛrebɛ, pɛmpɛraa, - } 
\LX{pemperi} \HM{} \PH{pèmpèrì} \PS{v.~} \GE{to bend or place something in an oblique position} \PD{pemperee, pempere, pempere, pemperibe, pemperaa, - } \VA{pɛmpɛre}
\LX{pɛnaa} \HM{} \PH{pɛ̀náà} \PS{n.~} \GE{musical instrument made from seasoned straw} \OP{pɛnne} \TP{pɛnnɛɛ} 
\LX{pɛŋe} \HM{} \PH{pɛ̀ŋ} \PS{v.~} \SN \GE{to groan} \SN \GE{to grunt (from effort)} \PD{pɛŋɛɛ, pɛnɛ, pɛnɛ, pɛnemɛ, pɛnaa, pɛnemɛ } 
\LX{pɛŋe} \HM{} \PH{pέŋì} \PS{} \GE{to skim (usually by tapping water with the back of the hand or a container to push away debris)} 
\LX{pɛŋe} \HM{} \PH{pɛ̂ŋ} \PS{v.~} \SN \GE{to borrow} \SN \GE{to lend} \PD{pɛŋɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, pɛnemɛ } 
\LX{pɛŋe} \HM{} \PH{pɛ̂ŋ} \PS{v.~} \GE{to tap on water to clear the top of unwanted material} \PD{pɛŋɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, pɛnemɛ } \VA{bɛŋ}
\LX{penɛne} \HM{} \PH{pɪ́nɛ́nɪ̀} \PS{n.~} \GE{mutton} \OP{penɛmɛ} \TP{penɛnnɛɛ} 
\LX{pɛŋkɔɔloŋ} \HM{} \PH{pɛ̌ŋkɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{pubic hair} 
\LX{penne} \HM{} \PH{pɪ́nnɪ̀} \PS{v.~} \GE{to make delicious} \PD{pennɛɛ, pennɛ, pennɛ, penemɛ, pennaa, - } 
\LX{pɛnne} \HM{} \PH{pɛ̀nnɪ̀} \PS{v.~} \GE{to carve} \PD{pɛnnɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, - } 
\LX{pɛnne} \HM{} \PH{pɛ̀nnɪ̀} \PS{v.~} \GE{to rest} \PD{pɛnnɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, - } 
\LX{pɛnne} \HM{} \PH{pɛ̀nnɪ́} \PS{n.~} \GE{lower part of abdomen} \OP{pɛmɛ} \TP{pɛnnɛɛ} 
\LX{pɛnne} \HM{} \PH{pɛ́nnɪ̀} \PS{n.~} \GE{cloth} \OP{pɛmɛ} \TP{pɛnnɛɛ} 
\LX{pɛnnoŋ} \HM{} \PH{pɛ̀nnʊ̂ŋ} \PS{n.~} \GE{} \VA{pɛnnoo}
\LX{pɛnnoo} \HM{} \PH{pɛ̀nnʊ̀ʊ́} \PS{n.~} \GE{air, breeze} \VA{pɛnnoŋ}
\LX{pɛnnu} \HM{} \PH{pɛ́nnù} \PS{n.~} \GE{hand measure (about 9 inches) of woven strip of cloth} \OP{pɛnnuuri} \TP{pɛnnuuree} 
\LX{penti} \HM{} \PH{péntì} \PS{n.~} \GE{paint} \OP{pentiri} \TP{pentiree} 
\LX{penyaŋaa} \HM{} \PH{pɪ́ɲáŋàá} \PS{n.~} \GE{ewe, adult female sheep} \OP{penyaŋene/penyanne} \TP{penyaŋenɛɛ/penyannɛɛ} 
\LX{pɛnzage} \HM{} \PH{pɛ́nzágɪ́} \PS{n.~} \GE{rag} \OP{pɛnzagra} \TP{pɛnzagrɛɛ} 
\LX{pɛpɛlaa} \HM{} \PH{pɛ́pɛ́láà} \PS{n.~} \GE{a type of small, carnivorous bird} \OP{pɛpɛlle} 
\LX{peraa} \HM{} \PH{pɪ́ráá} \PS{n.~} \SN \GE{deliberateness} \SN \GE{obstinacy} 
\LX{peraa} \HM{} \PH{pɪ́ráá} \PS{n.~} \GE{hyperbole (literature)} 
\LX{peraa} \HM{} \PH{pɪ́ràà} \PS{n.~} \GE{castanet; a pair of metal rings worn on the fingers and clicked together rhythmically} \OP{perre} \TP{perrɛɛ} 
\LX{pere} \HM{} \PH{pɪ́rɪ̀} \PS{v.~} \GE{push away; usually a small object with the foot} \PD{-, -, -, -, -, - } 
\LX{pere} \HM{} \PH{pɪ́rɪ́} \PS{n.~} \GE{a loop used in making the rim of a basket} \OP{perɛɛ} 
\LX{pere} \HM{} \PH{pɪ̀rɪ̀} \PS{v.~} \SN \GE{remove from; shell (groundnuts), unwrap (parcel),} \SN \GE{get wounded} \XV{O daare zɔ᷉ɔ᷉la kuriwiri te le ka o gbɛre pere.} \XE{He rode a bicycle, crashed and got his leg wounded.} \PD{perɛɛ, perɛ, perɛ, perebɛ, peraa, - } 
\LX{pere} \HM{} \PH{pɪ́rɪ̀} \PS{v.~} \GE{push away (small object), ignore} \XV{Ka fooŋ ba boɔrɔ a kuuri ŋa kyɛ ta pere o bare.} \XE{Even if you don't like this stone don't push it away.} \PD{perɛɛ, perɛ, perɛ, perebɛ, peraa, - } 
\LX{pere} \HM{} \PH{pɪ́rɪ̀} \PS{n.~} \GE{} \OP{peere} \VA{peroo}
\LX{pɛre} \HM{} \PH{pɛ́rɪ́} \PS{v.~} \GE{to stand abreast, to be equal e.g. height} \PD{pɛrɛɛ, pɛrɛ, pɛrɛ, pɛrebɛ, pɛraa, - } 
\LX{pɛre} \HM{} \PH{pɛ̀rɪ̀} \PS{v.~} \GE{to bend in an oblique manner} \PD{pɛrɛɛ, pɛrɛ, pɛrɛ, pɛrebɛ, pɛrɛ, - } 
\LX{perebare} \HM{} \PH{pɪ́rɪ́bàrɪ̀} \PS{n.~} \GE{outer toe of a fowl} \OP{perebarre} \TP{perebarɛɛ} 
\LX{perekara} \HM{} \PH{pɪ̀rɪ̀kàrà} \PS{adj.~} \GE{astride, firm and immovable} \XV{A sogya yaare la o gbɛɛ perekara ka tɔloo zie kyebe.} \XE{The soldier stood firmly with his feet astride such that no one can pass by him.} 
\LX{perekatɔɔre} \HM{} \PH{pɪ́rɪ́kátɔ́ɔ́rɪ̀} \PS{n.~} \GE{cemetery} \OP{perekatɔɛ} \TP{perekatɔɔrɛɛ} 
\LX{perepere} \HM{} \PH{pìrɪ̀pɪ́rɪ́} \PS{v.~} \GE{to struggle e.g. insect on a hot object} \PD{pereperɛɛ, pereperɛ, pereperɛ, pereperebɛ, pereperaa, - } 
\LX{perɛperɛ} \HM{} \PH{pɪ́rɛ́pɪ́rɛ́} \PS{adj.~} \GE{not stuck together e.g. well-cooked rice} 
\LX{pɛrɛpɛrɛ} \HM{} \PH{pɛ̀rɛ̀pɛ́rɛ́} \PS{n.~} \GE{food made of mashed yam} 
\LX{perete} \HM{} \PH{pɪ́rɪ́tɪ́} \PS{adv.~} \GE{spotlessly white} \VA{parata}
\LX{periperi} \HM{} \PH{pèrìpérí} \PS{v.~} \GE{to make an effort to overcome, to try hard} \XV{O ba taa a eebo kyɛ o meŋ aŋ peripere la.} \XE{She has not got the means, but she is also still trying hard.} \PD{periperee, peripere, peripere, periperibe, periperaa, - } 
\LX{peroo} \HM{} \PH{pɪ́rʊ́ʊ́} \PS{n.~} \GE{sheep} \OP{peere} \TP{peerɛɛ} \VA{pere}
\LX{pesaraa} \HM{} \PH{pɪ́sáráà} \PS{n.~} \GE{young adult female sheep} \SG{pesarrɛɛ} \OP{pesarre} 
\LX{pɛteɛ} \HM{} \PH{pɛ̀tɪ́ɛ́} \PS{n.~} \GE{medium-sized basket} \OP{pɛteere} \TP{pɛteerɛɛ} \VA{pɛtelee}
\LX{petooŋmaa} \HM{} \PH{pɪ́tòòŋmáá} \PS{n.~} \GE{one who will not listen to anything} \OP{petooŋmaara} \TP{petooŋmaarɛɛ} 
\LX{pɛyagelaa} \HM{} \PH{pɛ̀yàgláà} \PS{n.~} \GE{a basket, usually hung, in which calabashes are stored} \OP{pɛyagle} \TP{pɛyaglɛɛ} 
\LX{pezage} \HM{} \PH{pɪ́zàgɪ́} \PS{n.~} \GE{sheep pen} \OP{pezagre} \TP{pezagrɛɛ} 
\LX{pe᷉} \HM{} \PH{pɪ̀᷉} \PS{adv.~} \GE{onomatopoeic word (weighty object smashing a smaller object)} 
\LX{pe᷉e} \HM{} \PH{pɪ́᷉ɪ́᷉} \PS{adj.~} \GE{unrelenting, steadfast, not giving way} 
\LX{pe᷉ɛ} \HM{} \PH{pɪ́᷉ɛ́᷉} \PS{adv.~} \GE{completely e.g. finished} \XV{A ko᷉ɔ zaaŋ baare pe᷉ɛ.} \XE{All the water is finished completely.} 
\LX{pe᷉ɛ} \HM{} \PH{pɪ̀᷉ɛ̀᷉} \PS{v.~} \GE{pull apart in a bid to open e.g pulling the eyelid(s) apart} \PD{pe᷉ɛɛ, peɛnɛ, peɛnɛ, peɛnemɛ, peɛnaa, - } 
\LX{pe᷉ɛ} \HM{} \PH{pɪ̀᷉ɛ̀᷉} \PS{v.~} \SN \GE{to decorate well} \SN \GE{to dress gorgeously} \PD{pe᷉ɛɛ, peɛnɛ, peɛnɛ, peɛnemɛ, peɛnaa, - } 
\LX{pe᷉ɛ} \HM{} \PH{pɪ̀᷉ɛ̀᷉} \PS{v.~} \GE{to court} \PD{pe᷉ɛɛ, peɛnɛ, peɛnɛ, peɛnemɛ, peɛnaa, - } 
\LX{pɛ᷉ɛ} \HM{} \PH{pɛ̀᷉ɛ̀᷉} \PS{adv.~} \GE{fatty; oily} 
\LX{pe᷉ɛo} \HM{} \PH{pɪ́᷉ɛ́᷉ʊ́᷉} \PS{n.~} \GE{a type of wild cat (dark spotted)} \OP{peɛne} \TP{peɛnɛɛ} 
\LX{pe᷉epe᷉e} \HM{} \PH{pɪ́᷉ɪ́᷉pɪ́᷉ɪ́᷉} \PS{adv.~} \GE{very sharp (taste, sting)} 
\LX{pi} \HM{} \PH{pí} \PS{v.~} \GE{to become bloated, to swell up} \PD{pie, piire, piire, piiribe, piiraa, - } 
\LX{pi} \HM{} \PH{pí} \PS{v.~} \GE{to reserve, to claim in advance} \PD{pie, piire, piire, piiribe, piiraa, - } 
\LX{pi} \HM{} \PH{pì} \PS{v.~} \GE{to be in fear; to clutch self tightly in fear} \PD{pie, piire, piire, piiribe, piiraa, - } 
\LX{pibɔŋɔ} \HM{} \PH{pìbɔ̀ŋɔ́} \PS{n.~} \GE{a type of lizard believed to be poisonous} \OP{pibɔnne} \TP{pibɔnnɛɛ} \VA{pibaŋa}
\LX{pie} \HM{} \PH{píé} \PS{adj.~} \GE{ten} \OP{piiri} \TP{piiree} 
\LX{pie ane anaare} \HM{} \PH{pí é ánɪ́ ánáárɪ̀} \PS{adj.~} \GE{fourteen} \OP{pie ne anaare} 
\LX{pie ane anii} \HM{} \PH{pí é ánɪ́ áníì} \PS{adj.~} \GE{eighteen} \OP{pie ne aniini} 
\LX{pie ane anuu} \HM{} \PH{pí é áne anúú} \PS{adj.~} \GE{fifteen} \OP{pie ne anuuri} \TP{pie ne anuuree} 
\LX{pie ane ata} \HM{} \PH{pí é ánɪ́ átà} \PS{adj.~} \GE{thirteen} \OP{pie ne atare} 
\LX{pie ane ayi} \HM{} \PH{pí é ánɪ́ áyì} \PS{adj.~} \GE{twelve} \OP{pie ne ayiri} 
\LX{pie ane ayoɔbo} \HM{} \PH{píé  ánɪ́ àyʊ̀ɔ̀bʊ́} \PS{adj.~} \GE{sixteen} \OP{pie ne ayoɔre} 
\LX{pie ane ayɔpo᷉i} \HM{} \PH{píé ánɪ́ àyɔ̀po᷉́ì} \PS{adj.~} \GE{seventeen} \OP{pie ne ayɔponni} 
\LX{pie ane yeni} \HM{} \PH{píé ánɪ́ yénì} \PS{adj.~} \GE{eleven} \OP{pie ne yeme} 
\LX{pie ne awae} \HM{} \PH{pie nɪ awaɪ} \PS{adj.~} \GE{nineteen} 
\LX{pieri} \HM{} \PH{pìèrì} \PS{v.~} \SN \GE{to cause the skin to peel e.g as a result of scrapes on rough surfaces} \SN \GE{to cause to be bare (bush)} \PD{pieree, piere, piere, pieribe, pieraa, pieribe } 
\LX{pieto} \HM{} \PH{píétó} \PS{n.~} \GE{loincloth} \OP{pietori} \TP{pietoree} 
\LX{pigepige} \HM{} \PH{pígépígé} \PS{adj.~} \GE{moving around craftily} 
\LX{pigi} \HM{} \PH{pìgì} \PS{v.~} \SN \GE{to obstruct (view)} \SN \GE{to shield, to prevent} \SN \GE{to trap in an enclosure e.g. house, hole} \PD{pigee, pigre, pigre, pigribe, pigraa, pigribe } 
\LX{pigi} \HM{} \PH{pígí} \PS{v.~} \GE{to make a ball e.g. from mud} \PD{pigee, pigre, pigre, pigribe, pigraa, - } 
\LX{pigiri} \HM{} \PH{pìgirì} \PS{v.~} \GE{to look out for, to watch over} \PD{pigiree, pigire, pigire, pigiribe, pigiraa, - } 
\LX{piili} \HM{} \PH{píílì} \PS{v.~} \GE{to begin; to start} \PD{piilee, piile, piile, piilibe, piilaa, piilibe } 
\LX{piiri} \HM{} \PH{píírì} \PS{v.~} \GE{to massage with warm water e.g. on bruises} \PD{piiree, piire, piire, piiribe, piiraa, - } 
\LX{piiri} \HM{} \PH{píírí} \PS{n.~} \GE{rock} \OP{pie} \TP{piiree} \VA{pie (Southern Dagaare dialect)}
\LX{piiri} \HM{} \PH{pììrì} \PS{v.~} \GE{to recover a lost object by chance} \PD{-, piire, piiribe, piiraa, -, - } 
\LX{piirimpi᷉i᷉} \HM{} \PH{pììrímpí᷉ì᷉} \PS{n.~} \GE{a type of lizard} \OP{piirimpiime} 
\LX{pikuuri} \HM{} \PH{pìkúúrì} \PS{n.~} \GE{large piece of rock} \OP{pikubo} \TP{pikuuree} 
\LX{pileesekɔɔ} \HM{} \PH{píléésɪ́kɔ́ɔ́} \PS{n.~} \GE{a type of bird} \OP{pileesekɔɔre} 
\LX{pilempileŋ} \HM{} \PH{pìlèmpíléŋ} \PS{adv.~} \GE{gusty} 
\LX{pilepile} \HM{} \PH{pílépílé} \PS{adv.~} \GE{in a fast manner} 
\LX{pili} \HM{} \PH{pílí} \PS{n.~} \GE{a type of mat made of tiny straw and thread used to sleep on} \OP{pile} \TP{pilee} 
\LX{pilintoɔ} \HM{} \PH{pìlìntʊ́ɔ́} \PS{n.~} \GE{bottle} \OP{pilintoore} \TP{pilintoorɛɛ} 
\LX{pilipili} \HM{} \PH{pìlìpílí} \PS{v.~} \GE{to cover up or confuse by clever actions} \PD{pilipilee, pilipile, pilipile, pilipilibe, pilipilaa, - } 
\LX{pilli} \HM{} \PH{pìllì} \PS{v.~} \GE{to roll up (mat)} \PD{pillee, pille, pille, pilibe, pillaa, pilibe } 
\LX{pilli} \HM{} \PH{pìllì} \PS{v.~} \SN \GE{to cover (with a cloth)} \SN \GE{cover the top of a building} \PD{pillee, pille, pille, pilibe, pillaa, pilibe } 
\LX{pilli} \HM{} \PH{pìllì} \PS{v.~} \GE{to roll the tip of a feather in the ear to scratch it} \PD{pillee, pille, pille, pilibe, pillaa, pilibe } 
\LX{pilli pɔge} \HM{} \PH{pilli pɔgɪ} \PS{v.~} \GE{to overwhelm with a swarm (bee attack)} \PD{pilili pɔgɛɛ, pille pɔgrɔ, pilli-pɔgrɔ, -, -, - } 
\LX{pimbiri} \HM{} \PH{pímbírì} \PS{n.~} \SN \GE{nail} \SN \GE{arrowhead} \SN \GE{syringe} \SG{pimbiruu} \OP{pimbie} \TP{pimbiiree} 
\LX{pimbitaaraa} \HM{} \PH{pìmbítááráà} \PS{n.~} \GE{screw,  bolt} \OP{pimbitaare} \TP{pimbitaarɛɛ} 
\LX{pimpiŋ} \HM{} \PH{pímpíŋ} \PS{adj.~} \GE{far away} 
\LX{pimpiŋ} \HM{} \PH{pìmpíŋ} \PS{v.~} \GE{to attempt to confuse, to cover a deed} \PD{pimpiŋee, pimpine, pimpine, pimpinime, pimpinaa, - } 
\LX{piŋgaraa} \HM{} \PH{píŋgáráà} \PS{n.~} \GE{pickaxe} \OP{piŋgarre} \TP{piŋgarrɛɛ} 
\LX{piŋi} \HM{} \PH{píŋ} \PS{v.~} \SN \GE{to shake to remove e.g. dust from cloth} \SN \GE{to beat mercilessly (figurative)} \PD{piŋee, pine, pine, pinime, pinaa, pinime } 
\LX{piŋi} \HM{} \PH{píŋ} \PS{v.~} \GE{to hide or take cover behind an object} \PD{piŋee, pine, pine, pinime, pinaa, pinime } 
\LX{pinni} \HM{} \PH{pínnì} \PS{v.~} \SN \GE{to tap on to empty e.g. pocket} \SN \GE{to shake e.g. to remove unwanted bodies (bugs from a mat)} \SN \GE{to perform the final threshing to remove bits of grain} \PD{pinnee, pinne, pinne, pinime, pinnaa, - } 
\LX{pinni} \HM{} \PH{pínnì} \PS{v.~} \GE{to remind} \PD{pinnee, pinne, pinne, pinime, pinnaa, - } 
\LX{pipiiri} \HM{} \PH{pìpììrí} \PS{n.~} \GE{a type of tiny black flying insect that moves in groups} 
\LX{pipiree} \HM{} \PH{pìpíréé} \PS{n.~} \SN \GE{chaff} \SN \GE{residue of the matter left after chewing and extracting the liquid from some nuts like coconut} \OP{pipirri} \TP{pipirree} 
\LX{piraa} \HM{} \PH{píráá} \PS{n.~} \GE{button} \OP{pirri} 
\LX{pire} \HM{} \PH{pìré} \PS{n.~} \GE{new land cleared for farming} \OP{pirri} \TP{piree/pirree} 
\LX{pirempireŋ} \HM{} \PH{pírémpíréŋ} \PS{adv.~} \GE{full (tyre, stomach)} 
\LX{pirepire} \HM{} \PH{pírépíré} \PS{n.~} \GE{the state of complete emptiness; the state of having nothing} \XV{N da boɔrɔ ka N gaa sori kyɛ pirepire ba mare ma.} \XE{I had wanted to go on a trip, but I had nothing on me.} 
\LX{piri} \HM{} \PH{pírí} \PS{v.~} \SN \GE{to compete to win e.g. vacant seat} \SN \GE{to speak at the same time (two or more people)} \PD{piree, pire, pire, piribe, piraa, - } 
\LX{piri} \HM{} \PH{pírì} \PS{v.~} \GE{to lance, to make a small cut to release pus} \PD{piree, pire, pire, piribe, piraa, - } 
\LX{piri} \HM{} \PH{pìrì} \PS{v.~} \SN \GE{to button up e.g. shirt} \SN \GE{to strap on e.g. wristwatch} \SN \GE{to put a noose on or off e.g. neck or leg of an animal} \PD{piree, pire, pire, piribe, piraa, - } 
\LX{piriyoro} \HM{} \PH{píríyòró} \PS{adv.~} \GE{completely scattered in different directions e.g. members of a search party} 
\LX{pi᷈i} \HM{} \PH{pí᷈ì᷈} \PS{n.~} \GE{arrow} \SG{pi᷈iu} \OP{piime} \TP{piinee} 
\LX{pi᷈i biri} \HM{} \PH{pí᷈ì᷈ bírì} \PS{n.~} \GE{arrowhead} \OP{pi᷈i bie} \VA{pimbiri}
\LX{pi᷉ikaraa} \HM{} \PH{pí᷉í᷉kàràà} \PS{n.~} \GE{sharpened arrow} \OP{pi᷉ikarre} \TP{pi᷉ikarrɛɛ} 
\LX{pi᷉i-mɔre} \HM{} \PH{pí᷉í᷉-mɔ́rɪ́} \PS{n.~} \GE{arrow shaft} \OP{pi᷉i-mɔɛ} \TP{pi᷉i-mɔrɛɛ} 
\LX{pi᷉i-sabɔle} \HM{} \PH{pí᷉í᷉sàbɔ́l} \PS{n.~} \GE{a type of square-shaped arrow without spikes} \OP{pi᷉i-sabɔlɔ} \TP{pi᷉i-sabɔlɛɛ} 
\LX{pi᷉i-sabɔle} \HM{} \PH{pí᷉í᷉-sàbɔ́lɪ́} \PS{n.~} \GE{finless arrow} \OP{pi᷉i-sabɔlɔ} \TP{pi᷉i-sabɔlɛɛ} 
\LX{pi᷉iseɛbaa} \HM{} \PH{pí᷉í᷉sɪ́ɛ́báá} \PS{n.~} \GE{large, long, flat needle for weaving} \OP{pi᷉iseɛre} \VA{pi᷉iseɛraa}
\LX{pi᷉i-takɔraa} \HM{} \PH{pi̊́᷉í-tàkɔ̀ràá} \PS{n.~} \GE{a type of arrow with spikes on the sides} \OP{pi᷉i-takɔrre} \TP{pi᷉i-takɔrɛɛ} 
\LX{pi᷉iteɛraa} \HM{} \PH{pí᷉ìtɪ̀ɛ̀ráà} \PS{n.~} \GE{an arrow that is shot out; arrow for shooting} \OP{piiteɛre} 
\LX{pi᷉izɛle} \HM{} \PH{pí᷉í᷉zɛ́lɪ́} \PS{n.~} \GE{poisoned arrow} \OP{pi᷉izɛlɛ} 
\LX{po} \HM{} \PH{pʊ̀} \PS{v.~} \GE{to pull a grass stalk off the base} \PD{poɛ, poorɔ/poora, poorɔ/poora, pooreba, pooraa, - } 
\LX{po} \HM{} \PH{pò} \PS{v.~} \GE{to dig for termites (to feed young chicks)} \PD{poe, puoro, puoro, puoribo, puoraa, - } 
\LX{po} \HM{} \PH{pò} \PS{v.~} \GE{to ooze, to flow (water)} \PD{poe, puoro, puoro, puoribo, puoraa, - } 
\LX{pɔ} \HM{} \PH{pɔ́} \PS{v.~} \GE{to swear; to make a vow} \PD{pɔɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\LX{pɔ saa} \HM{} \PH{pɔ́ sáá} \PS{v.~} \GE{to jump from a very high point e.g. lizard from the tip of a tree branch} \PD{-, -, -, -, -, - } 
\LX{pobeɛmɛ} \HM{} \PH{pʊ̀bɪ́ɛ́mɛ́} \PS{n.~} \GE{stomachache, upset stomach} \TP{pobeɛmɛɛ} 
\LX{pobieri} \HM{} \PH{pʊ̀bìèrí} \PS{n.~} \GE{wickedness} \TP{pobieree} \VA{powolaa}
\LX{poboɔbo} \HM{} \PH{pʊ̀bʊ́ɔ́bʊ̀} \PS{n.~} \GE{will, desire} 
\LX{pobu} \HM{} \PH{pòbú} \PS{adv.~} \GE{filled with (tears in the eyes)} 
\LX{podɛgere} \HM{} \PH{pʊ̀dɛ́gɪ́rɪ̀} \PS{n.~} \GE{evil thought} 
\LX{pode᷉o} \HM{} \PH{pʊ̀dɪ̀᷉ʊ́᷉} \PS{n.~} \GE{preference} 
\LX{pɔ-eŋe} \HM{} \PH{pɔ́-ɪ̀ŋ} \PS{v.~} \GE{to curse; putting a malicious spell on someone} \PD{pɔ eŋɛɛ, poɔrɔ ennɛ, pɔ-ennɛ, pɔ-enemɛ, pɔ-ennaa, - } 
\LX{pofaa} \HM{} \PH{pʊ̀fáá} \PS{n.~} \GE{state of being evil-minded} \OP{pofaare} \TP{pofaarɛɛ} 
\LX{pogaale} \HM{} \PH{pʊ̀gáálɪ̀} \PS{n.~} \GE{prenatal care} \OP{pogaaɭɛɛ} 
\LX{pɔge} \HM{} \PH{pɔ̀gɪ̀} \PS{v.~} \SN \GE{to close, to shut} \SN \GE{to cover} \SN \GE{to imprison} \PD{pɔgɛɛ, pɔgrɔ, pɔgrɔ, pɔgreba, pɔgraa, pɔgreba } 
\LX{pɔge} \HM{} \PH{pɔ́gɪ́} \PS{n.~} \SN \GE{female} \SN \GE{woman} \SN \GE{wife} \OP{pɔgebɔ} \TP{pɔgebɛɛ} 
\LX{pɔge} \HM{} \PH{pɔ́gɪ̀} \PS{v.~} \SN \GE{to meet the presence of} \XV{Daŋ iribu wa pɔge ma kyɛ ka N gaa.} \XE{Get up early to come and meet me before I go.} \SN \GE{to hit a target} \XV{A naŋkpaana ŋmɛ la walaa kyɛ o ba pɔge o.} \XE{The hunter shot an antelope but did not hit the target.} \SN \GE{to be possible} \XV{A baa kye a dieŋ; o naa pɔge la yeŋe poɔ.} \XE{The dog is not in the room; it may be outside.} \PD{pɔgɛɛ, pɔgrɔ, pɔgrɔ, pɔgreba, pɔgraa, pɔgreba } 
\LX{pɔgebaa} \HM{} \PH{pɔ̀gɪ̀báá} \PS{n.~} \SN \GE{a woman who has a sharp tongue} \SN \GE{prostitute} \OP{pɔgebaare} \TP{pɔgebaarɛɛ} 
\LX{pɔgebeɛle} \HM{} \PH{pɔ̀gɪ̀bɪ̀ɛ̀lɪ̀} \PS{n.~} \GE{the act of taking a bride to the bridegroom} 
\LX{pɔgebeɛlɛ} \HM{} \PH{pɔ̀gɪ̀bɪ́ɛ́lɛ} \PS{n.~} \GE{bridesmaid} \OP{pɔgebeɛlbɛ} \TP{pɔgebeɛloo} 
\LX{pɔgebeɛloŋ} \HM{} \PH{pɔ̀gɪ̀bɪ̀ɛ̀lʊ̌ŋ} \PS{n.~} \GE{marriage feast (taking the bride to the bridegroom)} 
\LX{pɔgebeɛre} \HM{} \PH{pɔ̀gɪ̀bɪ́ɛ́rɪ́} \PS{n.~} \GE{a stage in courtship} 
\LX{pɔge-bɔ} \HM{} \PH{pɔ̀gɪ̀bɔ́} \PS{n.~} \GE{courtship} 
\LX{Pɔgeda᷈a᷈} \HM{} \PH{pɔ̀gɪ̀dá᷉à} \PS{n.~} \GE{name traditionally given to a female child born on day of locally brewed beer} \OP{Pɔgedaamine} \TP{Pɔgedaanɛɛ} 
\LX{pɔge-de} \HM{} \PH{pɔ̀gɪ́-dɪ́} \PS{n.~} \GE{marriage} 
\LX{pɔgederaa} \HM{} \PH{pɔ̀gɪ̀dɪ́ráá} \PS{n.~} \GE{a woman that is qualified to be married} \OP{pɔgederre} \TP{pɔgederɛɛ} 
\LX{pɔgederɛ} \HM{} \PH{pɔ̀gɪ̀dɪ́rɛ́} \PS{n.~} \SN \GE{bridegroom} \SN \GE{a man who is eligible to marry} 
\LX{pɔgederoŋ} \HM{} \PH{pɔ̀gɪ̀dɪ́rʋ́ŋ} \PS{n.~} \GE{marriage (action of groom)} 
\LX{pɔgedɔgerɔ} \HM{} \PH{pɔ̀gɪ̀dɔ̀gɪ̀rɔ́} \PS{n.~} \GE{a woman who has just given birth to a child} \OP{pɔgedɔgereba} \TP{pɔgedɔgerebɛɛ} 
\LX{pɔgefaa-ŋmenaa} \HM{} \PH{pɔ́gɪ́fáá-ŋɪ́náá} \PS{n.~} \GE{early evening (4 pm)} 
\LX{pɔgegbeɛloo} \HM{} \PH{pɔ̀gɪ̀bɪ́ɛ́lʊ́ʊ́} \PS{n.~} \GE{one who leads the bride to the bridegroom's house} \OP{pɔgegbeɛlebɛ} \TP{pɔgegbeɛlebɛɛ} 
\LX{pɔgegori} \HM{} \PH{pɔ̀gígórí} \PS{n.~} \GE{bride price} \OP{pɔgegoe} \TP{pɔgegoree} 
\LX{pɔgeguri} \HM{} \PH{pɔ̀gɪ̀gùrì} \PS{n.~} \GE{rape} 
\LX{pɔge-iraa} \HM{} \PH{pɔ̀gɪ́-íráá} \PS{n.~} \GE{a woman who has eloped with her fiance} \OP{pɔge-irri} 
\LX{pɔgekɔŋ} \HM{} \PH{pɔ̀gɪ̀kɔ̂ŋ} \PS{n.~} \GE{unavailability of mariagable women} 
\LX{pɔgekoɔre} \HM{} \PH{pɔ̀gɪ̀kʊ́ɔ́rɪ́} \PS{n.~} \SN \GE{unmarried woman} \SN \GE{widower} \OP{pɔgekɔbɔ} \TP{pɔgekɔbɛɛ} 
\LX{pɔgekuree} \HM{} \PH{pɔ̀gɪ̀kúréé} \PS{n.~} \GE{skirt} \OP{ pɔgekurri} \TP{pɔgekurree} 
\LX{pɔge-kyaaraa} \HM{} \PH{pɔ̀gɪ̀-ʧàáráá} \PS{n.~} \GE{woman offered for marriage} \OP{pɔge-kyaare} \TP{pɔge-kyaarɛɛ} 
\LX{pɔgekyeɛ} \HM{} \PH{pɔ̀gɪ̀ʧɪ́ɛ́} \PS{n.~} \GE{sister-in-law} \OP{pɔgekyeere\\pɔgekyemine} \TP{pɔgekyeerɛɛ} 
\LX{pɔgekyɛre} \HM{} \PH{pɔ̀gɪ̀ʧɛ̀rɪ̀} \PS{n.~} \GE{payment of bride price} 
\LX{pɔge-la-dɔɔ} \HM{} \PH{pɔ̀gɪ̀-lá-dɔ́ɔ́} \PS{n.~} \GE{woman with manly features} \OP{pɔge-la-dɔbɔ} 
\LX{pɔgele} \HM{} \PH{pɔ̀gɪ̀lɪ̀} \PS{v.~} \GE{to tame} \PD{pɔgelɛɛ, pɔgelɔ,, pɔgelɔ,, pɔgeleba, pɔgelaa, - } 
\LX{pɔgelee} \HM{} \PH{pɔ̀gɪ̀léé} \PS{n.~} \SN \GE{girl} \SN \GE{junior wife (second of two wives)} \OP{pɔgebilii} \TP{pɔgebillee} 
\LX{pɔgeloŋ} \HM{} \PH{pɔ̀gɪ̀lʊ́ŋ} \PS{n.~} \GE{womanliness} 
\LX{pɔge-nɔgeraa} \HM{} \PH{pɔ̀gɪ̀-nɔ̀gɪ̀ràá} \PS{n.~} \GE{woman (a derogatory term implying that the woman is a sojourner and therefore has no power)} \OP{pɔge-nɔgere} \TP{pɔge-nɔgerɛɛ} 
\LX{pɔgenyaŋaa} \HM{} \PH{pɔ̀gɪ̀ɲáŋáá} \PS{n.~} \GE{old woman} \OP{pɔgnyaŋene/pɔgenyanne} \TP{pɔgenyannɛɛ} 
\LX{pɔgenyɔgeraa} \HM{} \PH{pɔgeɲɔgeraa} \PS{n.~} \GE{a woman who has eloped with a man} \OP{pɔgenyɔgerre} 
\LX{pɔgenyuo} \HM{} \PH{pɔ̀gɪ̀ɲúó} \PS{n.~} \GE{woman with a big navel (name)} \OP{pɔgenyuuri/pɔgenyumine} 
\LX{pɔge-pɛgebo} \HM{} \PH{pɔ̀gɪ̀-pɛ́gɪ́bʊ́} \PS{n.~} \GE{menstruation} 
\LX{pɔgepe᷉ɛ᷉} \HM{} \PH{pɔ̀gɪ̀pɪ̀᷉ɛ̀᷉} \PS{n.~} \GE{courtship} 
\LX{pɔgepiiraa} \HM{} \PH{pɔ̀gɪ̀pííráá} \PS{n.~} \SN \GE{fiancee, betrothed woman} \SN \GE{engaged woman for marriage} \OP{pɔgepiiri} \TP{pɔgepiiree} 
\LX{pɔgepoɔ} \HM{} \PH{pɔ̀gìpʊ́ɔ́} \PS{n.~} \GE{pregnant woman} \OP{pɔgepoore} \TP{pɔgepoorɛɛ} 
\LX{Pɔgesaa} \HM{} \PH{pɔ̀gɪ̀sáá} \PS{n.~} \GE{name traditionally given to a female child born on a rainy day} \OP{pɔgesamine} 
\LX{pɔgesaraa} \HM{} \PH{pɔ̀gɪ̀sáráà} \PS{n.~} \GE{young woman} \OP{pɔgesarre} \TP{pɔgesarrɛɛ} 
\LX{pɔgesarebie} \HM{} \PH{pɔ̀gɪ̀sárɪ́bíé} \PS{n.~} \GE{a woman's firstborn} \OP{pɔgesarebiiri} \TP{pɔgesarebiiree} 
\LX{pɔgesarelee} \HM{} \PH{pɔ̀gɪ̀sárɪ́léé} \PS{n.~} \GE{young woman} \OP{pɔgesarebilii} 
\LX{pɔgeseɛ} \HM{} \PH{pɔ̀gɪ̀sɪ́ɛ́} \PS{} \GE{woman who is menstruating} 
\LX{pɔgeseɛ} \HM{} \PH{pɔ̀gɪ̀sɪ́ɛ́} \PS{n.~} \GE{a type of yam} \OP{pɔgeseere} \TP{pɔgeseerɛɛ} 
\LX{pɔgeserekulo} \HM{} \PH{pɔ̀gɪ̀sɪ̀rɪ̀kúlò} \PS{n.~} \GE{marriageable woman} \OP{ pɔgeserekulibo} \TP{pɔgeserekulibee} 
\LX{pɔgeseɛ} \HM{} \PH{pɔ̀gɪ̀sɪ́ɛ́} \PS{n.~} \SN \GE{a woman’s cloth measuring about 72 x 45 inches worn on the waist} \SN \GE{any such size of cloth} \SG{seerɛɛ} \OP{pɔgeseere} \TP{pɔge} 
\LX{pɔgesoga} \HM{} \PH{pɔ̀gɪ̀sʊ́gá} \PS{n.~} \GE{middle wife (one in between the first and last of three or more wives)} \OP{pɔgesogre} \TP{pɔgesogrɛɛ} 
\LX{pɔgesoɔbɔ} \HM{} \PH{pɔ̀gɪ̀sʊ́ɔ́bɔ́} \PS{n.~} \GE{witch} \OP{pɔgesoɔre} \TP{pɔgesoɔrɛɛ} 
\LX{Pɔgesugilo} \HM{} \PH{pɔ̀gɪ̀súgiló} \PS{n.~} \GE{female name traditionally given to the second born after twins} 
\LX{pɔgesugo} \HM{} \PH{pɔ̀gɪ̀súgó} \PS{n.~} \GE{a large group of women} \OP{pɔgesugri} \TP{pɔgesugree} 
\LX{pɔge-tu} \HM{} \PH{pɔ̀gɪ̀-tù} \PS{n.~} \GE{taking home of a woman who failed to divorce} 
\LX{pɔge-tu} \HM{} \PH{pɔ̀gɪ̀-tú} \PS{n.~} \GE{final stage in courtship} 
\LX{pɔgetuuro} \HM{} \PH{pɔ̀gɪ̀túúró} \PS{n.~} \GE{one who takes a woman from her family to her husband's house} \OP{pɔgetuuribo} \TP{pɔgetuurbee} 
\LX{Pɔgeweɛ} \HM{} \PH{pɔ̀gɪ̀wɪ̀ɛ́} \PS{n.~} \GE{name traditionally given to a female child born in the farm/bush (outside of the normal living environment)} 
\LX{pɔgeyaa} \HM{} \PH{pɔ̀gɪ̀yáá} \PS{n.~} \GE{daughter} \OP{pɔgeyaare} \TP{pɔgeyaarɛɛ} 
\LX{pɔgeyagede} \HM{} \PH{pɔ̀gɪ̀yágɪ́dɪ́} \PS{n.~} \GE{polygamy} 
\LX{pɔgeyi-gane} \HM{} \PH{pɔ̀gɪ̀yí-gánɪ̀} \PS{n.~} \GE{writ of divorce} \OP{pɔgeyi-gama} \TP{pɔgeyi-gannɛɛ} 
\LX{pɔgeyiraa} \HM{} \PH{pɔ̀gɪ̀yíráá} \PS{n.~} \GE{divorced woman} \OP{pɔgeyirri} \TP{pɔgeyirree} 
\LX{Pɔgezeɛ} \HM{} \PH{pɔ̀gɪ̀zɪ̀ɛ́} \PS{n.~} \SN \GE{light-skinned woman} \SN \GE{female name} \OP{pɔgezeere/pɔgezeɛmine} 
\LX{pogitogo} \HM{} \PH{pógítógó} \PS{adv.~} \GE{very short and small} \VA{pogitogoo}
\LX{pokpaane} \HM{} \PH{pʊ̀kpàànɪ̀} \PS{n.~} \GE{fasting} \TP{pokpaanɛɛ} \VA{pokpegri}
\LX{pokpegiri} \HM{} \PH{pʊ̀kpégrí} \PS{n.~} \GE{closed-heartedness, without an open mind, wickedness} \OP{pokpege} \TP{pokpegree} \VA{pokpaane}
\LX{pokpɛlɛnkyɛ} \HM{} \PH{pʊ̀kpɛ́lɛ́nʧɛ́} \PS{n.~} \GE{the act of being potentially dangerous} 
\LX{pokpɔgere} \HM{} \PH{pʊ̀kpɔ́grí} \PS{n.~} \GE{tumor in the stomach} \OP{pokpɔgɔ} \TP{pokpɔgrɛɛ} 
\LX{pokuoŋaa} \HM{} \PH{pʊ̀kůóŋáá} \PS{n.~} \GE{sharp stomachache without diarrhoea} \OP{pokuoni} \TP{pokuonee} 
\LX{pokyiiri} \HM{} \PH{pʊ́ʧíírí} \PS{n.~} \GE{confession} 
\LX{pɔlaa} \HM{} \PH{pɔ̀làà} \PS{n.~} \GE{heartbeat, pulse} \OP{pɔlle} \TP{pɔllɛɛ} 
\LX{pɔlaare} \HM{} \PH{pɔ́láárɪ̀} \PS{v.~} \GE{to swear vehemently} \PD{pɔlaarɛɛ, poɔrɔlaara, pɔlaara, pɔlaareba, pɔlaaraa, - } 
\LX{pɔlbile} \HM{} \PH{pɔ́lbílé} \PS{n.~} \GE{young man} \XV{pɔlbilii} \XE{} \OP{pɔlbilee} \VA{pɔle, pɔlee}
\LX{pɔle} \HM{} \PH{pɔ́lɪ̀} \PS{n.~} \GE{young boy} \OP{pɔlbilii} \VA{pɔlee, pɔlbile}
\LX{pɔle} \HM{} \PH{pɔ̀lɪ̀} \PS{v.~} \SN \GE{to grow to maturity} \SN \GE{to be full of life} \SN \GE{to become beautiful} \PD{pɔlɛɛ, pɔllɔ, pɔllɔ, pɔleba, -, - } 
\LX{polee} \HM{} \PH{póléé} \PS{n.~} \GE{small farm} 
\LX{polee} \HM{} \PH{pʊ̀léé} \PS{n.~} \GE{early pregnancy} 
\LX{pɔlee} \HM{} \PH{pɔ́lɪ́ɪ́} \PS{n.~} \GE{young man} \VA{pɔle, pɔlbile}
\LX{poleɛne} \HM{} \PH{pʊ́lɪ́ɛ́nɪ́} \PS{n.~} \GE{female lizard} \OP{poleɛmɛ} \TP{poleɛmɛɛ} 
\LX{polepole} \HM{} \PH{pʊ́lpʊ́lɪ́} \PS{adj.~} \GE{very tall and pointed summit. (church tower)} 
\LX{pɔlese} \HM{} \PH{pɔ́lɪ́sɪ̀} \PS{v.~} \GE{to polish} \PD{pɔleseɛ, pɔleseerɛ, pɔleseerɛ, pɔleseerebɛ, pɔleseeraa, - } 
\LX{poli} \HM{} \PH{pólí} \PS{v.~} \GE{smear plenty of oil producing a shining effect} \PD{polee, polo, polo, polibo, polaa, polibo } 
\LX{poli} \HM{} \PH{pòlì} \PS{adv.~} \GE{flaring up abruptly (fire)} 
\LX{polige} \HM{} \PH{pʊ̀lígè} \PS{n.~} \GE{the state of not being open-minded} 
\LX{pɔlkakyeerɛ} \HM{} \PH{pɔ̌lkáʧɪ́ɪ́rɛ́} \PS{n.~} \GE{the act of being overly proud of one’s self} 
\LX{poll} \HM{} \PH{pʊ́ll} \PS{adj.~} \GE{small and sharp} \OP{poll poll} 
\LX{polle} \HM{} \PH{pʊ̌llɪ̀} \PS{v.~} \GE{to take an oath; to make a promise} \GE{an oath} \PD{pollɛɛ, polla, polla, poleba, pollaa, poleba } \SE{polle noɔre, polle teebo} \GE{to vow} 
\LX{polle} \HM{} \PH{pʊ̌llɪ̀} \PS{v.~} \GE{to sharpen (a point)} \GE{an oath} \PD{pollɛɛ, polla, polla, poleba, pollaa, poleba } \SE{polle noɔre, polle teebo} \GE{to vow} 
\LX{pɔllɔ} \HM{} \PH{pɔ̀llɔ̀} \PS{n.~} \GE{pathway used by wild rodents} \OP{pɔlle} \TP{pɔllɛɛ} \VA{pɔle}
\LX{pololo} \HM{} \PH{pʊ́lʊ́lʊ́} \PS{adj.~} \GE{pointed and sharp} 
\LX{pɔloo} \HM{} \PH{pɔ́lʊ́ʊ́} \PS{n.~} \GE{liveliness} \XV{Yɛ yieluŋ na taa la pɔloo.} \XE{That song of yours is lively.} 
\LX{pɔloo} \HM{} \PH{pɔ́lʊ́ʊ́} \PS{n.~} \SN \GE{pride} \SN \GE{vanity} \XV{A pɔgelee na veɛlɛ la kyɛ o erɛ la pɔloo.} \XE{That young lady is beautiful but she is proud.} 
\LX{pɔlteɛ} \HM{} \PH{pɔ̀ltìɛ̀} \PS{n.~} \GE{frightening} 
\LX{pɔmɛɛ} \HM{} \PH{pɔ̀mɛ̀ɛ̀} \PS{adv.~} \GE{fat and round} 
\LX{pomm} \HM{} \PH{pòmm} \PS{adv.~} \GE{bloated like a balloon} 
\LX{pompoluoree} \HM{} \PH{pómpólúóréé} \PS{n.~} \GE{a type of wild tree with edible fruit} \OP{pompoluori} \TP{ pompoluoree} \VA{polimpoluoree}
\LX{pompoŋɔ} \HM{} \PH{pʊ́mpʊ́ŋɔ̀} \PS{n.~} \GE{a type of poisonous green-trunked tree with white nectar} \OP{pomponne} \TP{pomponnɛɛ} \VA{pompoŋa}
\LX{pɔmpɔre} \HM{} \PH{pɔ̌mpɔ̀rɪ́} \PS{adj.~} \GE{very young and immature e.g. puppy} 
\LX{ponaŋa} \HM{} \PH{pʊ́náŋ} \PS{n.~} \GE{severe stomachache} \OP{ponanne} \TP{ponannɛɛ} \VA{ponaŋe}
\LX{poŋi} \HM{} \PH{pôŋ} \PS{n.~} \GE{pollen} \OP{ponni} \TP{ponnee} 
\LX{poŋi} \HM{} \PH{pòŋ} \PS{v.~} \SN \GE{to shave (hair)} \SN \GE{to remove corn from the cob} \PD{poŋee, pono, pono, ponimo, ponaa, ponimo } 
\LX{poŋi} \HM{} \PH{póŋ} \PS{v.~} \GE{to share, to divide} \PD{poŋee, pono, pono, ponimo, ponaa, - } 
\LX{ponne} \HM{} \PH{pʊ́nnɪ́} \PS{v.~} \SN \GE{to drop in bits e.g. rain} \SN \GE{to test tightness e.g. string of bow, guitar} \PD{ponnɛɛ, ponnɔ, ponnɔ, ponema, ponnaa, - } 
\LX{ponne} \HM{} \PH{pʊ̀nnɪ̀} \PS{v.~} \GE{to covet; to make advances to mate with e.g. animals} \PD{-, -, -, -, -, - } 
\LX{ponne} \HM{} \PH{pʊ̀nnɪ̀} \PS{adj.~} \GE{not up to standard} 
\LX{pɔnne} \HM{} \PH{pɔ̀nnɪ̀} \PS{v.~} \SN \GE{to hold onto someone's chest or belly, but not firmly} \XV{A saa naŋ iri la ka a bie zɛle o yɔɔlee pɔnne a zɛlzɛlɛ zoro ne.} \XE{As the rain started, the child picked his brother and half dragged him away.} \SN \GE{to be calm} \XV{Saseɛ daare erɛ la kyɛ saa mie la ka zie pɔnne.} \XE{It used to be windy but with the coming of rain, the weather is calm.} \PD{pɔnnɛɛ, pɔnnɔ, pɔnnɔ, pɔnema, pɔnnaa, - } 
\LX{pɔntenakaare} \HM{} \PH{pɔ̀ntɪ́nàkáárɪ́} \PS{n.~} \GE{one that is here and there} \TP{pɔntenakaarɛɛ} 
\LX{pɔntere} \HM{} \PH{pɔ̀ntɪ́rɪ̀} \PS{n.~} \GE{toad} \OP{pɔnteɛ} \TP{pɔnteerɛɛ} 
\LX{poo} \HM{} \PH{pʊ̀ʊ̀} \PS{adv.~} \GE{onomatopoeia; like noise made when one wades into water} 
\LX{po-o} \HM{} \PH{pʊ́-ʊ́} \PS{n.~} \GE{ocean} \TP{poorɛɛ} \VA{mpoɔne, mpoɔnɛɛ (Akan)}
\LX{poɔ} \HM{} \PH{pʊ́ɔ́} \PS{n.~} \GE{stomach, abdomen} \OP{poore} \TP{poorɛɛ} 
\LX{poɔ} \HM{} \PH{pʊ́ɔ́} \PS{n.~} \GE{part (of a group)} \XV{A pɔgelee ba poɔ a zɔɔreŋ.} \XE{The girl is not part of the fight.} \OP{poore} \TP{poorɛɛ} 
\LX{poɔ} \HM{} \PH{pʊ́ɔ́} \PS{n.~} \GE{pregnancy} \OP{poore} \TP{poorɛɛ} 
\LX{poɔ} \HM{} \PH{pʊ́ɔ́} \PS{v.~} \GE{to beat down on by rain} \PD{poɔɛɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\LX{poɔ} \HM{} \PH{pʊ́ɔ̀} \PS{v.~} \GE{to add to; to include} \PD{poɔɛɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\LX{poɔ} \HM{} \PH{pʊ́ɔ́} \PS{prep.~} \SN \GE{in, inside} \XV{Ka diibuŋ be fo noɔre poɔ fo koɳ to᷉ɔ᷉ yeli yɛlɛ velaa.} \XE{If there is food in your mouth, you cannot talk well.} \SN \GE{among, within} \XV{O poɔ la a bibiiri na naŋ daŋe waaboŋ.} \XE{S/he is among the children who came early.} 
\LX{pɔɔ} \HM{} \PH{pɔ̀ɔ́} \PS{v.~} \GE{to be too small, to be too young} \PD{pɔɔɛ, pɔɔrɔ, pɔɔrɔ, pɔɔrebɔ, pɔɔraa, - } 
\LX{poɔle} \HM{} \PH{pʊ́ɔ́lɪ́} \PS{v.~} \SN \GE{to cause wetness of e.g. floor} \SN \GE{to soak herbs (preparing medicine)} \PD{poɔlɛɛ, poɔlɔ, poɔlɔ, poɔlbɔ, poɔlaa, - } 
\LX{poɔle} \HM{} \PH{pʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to make the end of something pointed} \PD{poɔlɛɛ, poɔlɔ, poɔlɔ, poɔlbɔ, poɔlaa, - } 
\LX{poɔle} \HM{} \PH{pʊ́ɔ́lɪ́} \PS{adj.~} \GE{pointed end} \XV{A bugo pare waa la poɔle lɛ.} \XE{The bottom of the granary is pointed.} 
\LX{poomɔ} \HM{} \PH{pʊ́ʊ́mɔ̀} \PS{n.~} \GE{bean weevil} \TP{poomɛɛ} \VA{pooma}
\LX{poɔnaa} \HM{} \PH{pʊ́ɔ́náá} \PS{n.~} \GE{hide and seek (children's game)} \OP{poɔne} \TP{poɔnɛɛ} 
\LX{poone} \HM{} \PH{pʊ̀ʊ̀nɪ̀} \PS{v.~} \GE{to persist e.g. talk, argue} \XV{O maŋ boɔle la ba woŋi yeli kyɛ meŋ aŋ poonɔ pare.} \XE{When he calls, there comes no answer, but he persists in calling earnestly.} \PD{poonɛɛ, poonɔ, poonɔ, poonema, poonaa, - } 
\LX{poore} \HM{} \PH{pʊ̌ʊ̌rɪ́} \PS{n.~} \GE{challenge} \TP{poorɛɛ} 
\LX{poore} \HM{} \PH{pʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to challenge, to compete} \PD{poorɛɛ, poora, poora, pooreba, pooraa, - } 
\LX{poore} \HM{} \PH{pʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to spray e.g. water onto someone} \PD{poorɛɛ, poora, poora, pooreba, pooraa, - } 
\LX{poore} \HM{} \PH{pʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to harvest grass stalks} \PD{poorɛɛ, poora, poora, pooreba, pooraa, - } 
\LX{poɔre} \HM{} \PH{pʊ̀ɔ̀rɪ̀} \PS{v.~} \SN \GE{to rub to clear e.g. hand over face just from sleep} \SN \GE{to clutch with both hands and feet e.g in climbing a tree or wall} \PD{poɔrɛɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\LX{pɔɔre} \HM{} \PH{pɔ́ɔ́rɪ́} \PS{n.~} \GE{a short and poisonous snake} \VA{mabiekontapɔgema}
\LX{poɔroŋ} \HM{} \PH{pʊ́ɔ́rʊ́ŋ} \PS{n.~} \GE{rot, rotten thing} \OP{poɔrɔ} 
\LX{popaale} \HM{} \PH{pʊ́páálɪ́} \PS{n.~} \GE{disease that causes swollen feet and a distended stomach} 
\LX{popeɛloŋ} \HM{} \PH{pʊ́pɪ́ɛ́lʊ́ŋ} \PS{n.~} \GE{happiness} 
\LX{popɛge} \HM{} \PH{pòpɛ́gɪ́} \PS{n.~} \GE{outer shell} \OP{popɛgere/popɛgɛ} \TP{popɛgerɛɛ} 
\LX{popelaa} \HM{} \PH{pʊ́pɪ́láá} \PS{n.~} \GE{stomachache} 
\LX{popelaa} \HM{} \PH{pʊ́pɪ́láá} \PS{adv.~} \SN \GE{prostrate} \SN \GE{open minded} 
\LX{popɛle} \HM{} \PH{pʊ̀pɛ̀lɪ̀} \PS{n.~} \SN \GE{happiness, joy (taa popɛle)} \SN \GE{choice, wish} 
\LX{popo} \HM{} \PH{pòpò} \PS{n.~} \GE{motorcycle} \OP{popori} \TP{poporee} 
\LX{pɔpɔrɛɛ} \HM{} \PH{pɔ́pɔ́rɛ́ɛ́} \PS{n.~} \GE{} \SG{pɔpɔroo} \VA{shortened form of sepɔpɔrɛɛ}
\LX{pɔraa} \HM{} \PH{pɔ́ràà} \PS{n.~} \SN \GE{an ordeal to find a wrongdoer} \SN \GE{the condition of hanging in the balance} \OP{pɔrre} \TP{pɔrrɛɛ} 
\LX{pore} \HM{} \PH{pʊ́rɪ́} \PS{adj.~} \GE{tiny, small} \OP{porre} \TP{porrɛɛ} 
\LX{pore} \HM{} \PH{pʊ̀rɪ̀} \PS{v.~} \GE{to name, to mention} \PD{porɛɛ, porɔ, porɔ, poreba, poraa, - } 
\LX{pore} \HM{} \PH{pʊ́rɪ̀} \PS{v.~} \GE{to pour out} \PD{porɛɛ, porɔ, porɔ, poreba, poraa, - } 
\LX{pore} \HM{} \PH{pʊ́rɪ̀} \PS{v.~} \GE{to buy herbal medicine} \PD{porɛɛ, porɔ, porɔ, poreba, poraa, - } 
\LX{pore} \HM{} \PH{pʋ́rɪ́} \PS{adj.~} \GE{small in size} 
\LX{pore} \HM{} \PH{pʊ̀rɪ̀} \PS{adj.~} \GE{abundant in quantity} 
\LX{pɔre} \HM{} \PH{pɔ́rɪ̀} \PS{v.~} \SN \GE{to pluck (fruit)} \SN \GE{to die (baby)} \PD{pɔrɛɛ, pɔrɔ, pɔrɔ, pɔreba, pɔraa, - } 
\LX{pɔre} \HM{} \PH{pɔ́rɪ́} \PS{v.~} \GE{to carry out a test to find a wrongdoer} \PD{pɔrɛɛ, pɔrɔ, pɔrɔ, pɔreba, pɔraa, - } 
\LX{poree} \HM{} \PH{póréé} \PS{n.~} \GE{pair of pegs on which pots and calabashes are placed} \OP{porri} \TP{porree} 
\LX{pori} \HM{} \PH{pòrì} \PS{v.~} \GE{to cause to blister, to scald (of fire)} \PD{poree, poro, poro, poribo, poraa, - } 
\LX{pori} \HM{} \PH{pòrì} \PS{adj.~} \GE{sound of a big falling object from a height (a big fruit)} 
\LX{pori} \HM{} \PH{pórí} \PS{adj.~} \GE{sound of a small object falling from a height (small fruit)} 
\LX{pori} \HM{} \PH{pórí} \PS{v.~} \GE{to struggle e.g. dying fowl} \PD{poree, poro, poro, poribo, poraa, - } 
\LX{poriko} \HM{} \PH{póríkó} \PS{n.~} \GE{pig} \OP{porikori} \TP{porikoree} \VA{pariko}
\LX{porisi} \HM{} \PH{pórísì} \PS{n.~} \GE{police} \OP{porisiri} \TP{porisiree} 
\LX{poritiisan} \HM{} \PH{pòrítɪ́ɪ́sa} \PS{n.~} \GE{police station} \OP{poritiisanne} \TP{poritiisanɛɛ/poritiisannɛɛ} 
\LX{porɔ} \HM{} \PH{pʊ́rɔ́} \PS{n.~} \GE{fail to respond smoothly or click (archer's arrow)} \XV{A tammo ŋmɛ la porɔ ka a tanteɛrɛ teɛ a soɔŋaa nyɛre.} \XE{The bow failed to release the arrow and so the archer failed to hit the hare.} \OP{porre} \TP{porɛɛ} 
\LX{poroso} \HM{} \PH{pòròsò} \PS{adv.~} \GE{large and round e.g. balloon} 
\LX{poroworo} \HM{} \PH{pʊ̀rɪ̀wʊ̀rʊ̀} \PS{adj.~} \GE{countless and all over the place e.g. piglets of a sow} 
\LX{porre} \HM{} \PH{pʊ́rrɪ̀} \PS{v.~} \GE{to force air through closed lips (rude way of disagreeing; lie)} \PD{porrɛɛ, porra, porra, poreba, porraa, - } 
\LX{porre} \HM{} \PH{pʊ́rrɪ̀} \PS{v.~} \GE{to break into tiny bits} \PD{porrɛɛ, porra, porra, poreba, porraa, - } 
\LX{porri} \HM{} \PH{pórrì} \PS{v.~} \GE{to report} \PD{porree, porro, porro, poribo, porraa, - } 
\LX{posaanaa} \HM{} \PH{posaanaa} \PS{n.~} \GE{sorrow, unhappiness} \OP{posaane/posaama} \TP{posaanɛɛ} 
\LX{posa᷉a} \HM{} \PH{pʊ̀sàà} \PS{n.~} \GE{sadness} 
\LX{posogedaa} \HM{} \PH{pòsʊ̀gdàá} \PS{n.~} \GE{backbone} \OP{posogedaare} \TP{posogedaarɛɛ} 
\LX{posoge-kpoŋi} \HM{} \PH{pósʊ́g-kpôŋ} \PS{n.~} \GE{spinal tuberculosis} 
\LX{posogɔ} \HM{} \PH{pósʊ́gɔ́} \PS{n.~} \GE{mid-back} \OP{posogere} \TP{posogerɛɛ} \VA{posoga}
\LX{poteɛrɛ} \HM{} \PH{pʊ́tɪ́ɛ́rɛ́} \PS{n.~} \GE{thoughts, wishes} 
\LX{poteɛroŋ} \HM{} \PH{pʊ̀tɪ́ɛ́rʊ́ŋ} \PS{n.~} \GE{thought} \OP{poteɛrɛ} 
\LX{po-tu} \HM{} \PH{pò-tù} \PS{n.~} \GE{the act of following} 
\LX{potulimo} \HM{} \PH{pótúlímó} \PS{adj.~} \GE{backwards} \VA{potulo}
\LX{potulo} \HM{} \PH{pótúló} \PS{adj.~} \GE{backwards} \VA{potulimo}
\LX{potuo} \HM{} \PH{pòtúó} \PS{n.~} \GE{wickedness, malice} \OP{potuuri} \TP{potuuree} 
\LX{potuoluŋ} \HM{} \PH{po̊túólúŋ} \PS{n.~} \GE{wickedness} 
\LX{potuosobɔ} \HM{} \PH{pòtúósʊ́bá} \PS{n.~} \GE{wicked person} \OP{ potuodeme} \VA{potuosoba}
\LX{potuuro} \HM{} \PH{pòtúúró} \PS{n.~} \GE{follower, disciple} \OP{ potuuribo} 
\LX{pou} \HM{} \PH{póú} \PS{adv.~} \GE{onomatopoeia; sound of blast (balloon)} 
\LX{povelaa} \HM{} \PH{pʊ̀vɪ̀làá} \PS{adj.~} \GE{sincere, pure} \OP{poveɛle} 
\LX{powɛle} \HM{} \PH{pʊ́wɛ́lɪ́} \PS{n.~} \GE{confession} 
\LX{powolaa} \HM{} \PH{pʊ́wʊláá́} \PS{n.~} \GE{wickedness} \OP{powolle} \VA{pobieri}
\LX{po-yiraa} \HM{} \PH{pʊ̀-yíráá} \PS{n.~} \GE{miscarriage} \OP{po-yirri} \VA{po-yi}
\LX{pozagela} \HM{} \PH{pʊ́zágɪ́là} \PS{n.~} \GE{innocence} 
\LX{pozagelaa} \HM{} \PH{pʊ́zágɪ́láá} \PS{n.~} \GE{empty stomach} \OP{pozagele} 
\LX{pozagelaa} \HM{} \PH{pʊ́zágɪ́láá} \PS{n.~} \GE{the state of being free of malice toward another} \OP{pozagele} 
\LX{pozu} \HM{} \PH{pòzú} \PS{n.~} \GE{gossip} \OP{pozuri} \TP{pozuree} 
\LX{po᷉ɔ} \HM{} \PH{pʊ̀᷉ɔ̀᷉} \PS{v.~} \SN \GE{to interrupt} \XV{A bie ba tu o noɔre baare kyɛ ka a karema po᷉ɔ kpɛ, yeli ka zirii la.} \XE{The child had not finished telling his story when the teacher interrupted and said it was a lie.} \XV{A pɔgebilii deɛnɛ la velaa zaa ka dɔɔbilii mine wa põɔ kpɛ ka a deɛne ŋmaa.} \XE{The girls were happily playing when some boys interrupted them and the play ended.} \SN \GE{to perform a task earlier than expected} \XV{A tontonema da põɔ iree la gaa o toma zie a na te de ba sanyɔɔ.} \XE{The workers left before dawn to the work place in order to receive their salaries.} \PD{po᷉ɔɛ, poɔnɔ, poɔnɔ, poɔnema, poɔnaa, - } 
\LX{po᷉ɔ} \HM{} \PH{pʊ́᷉ɔ́᷉} \PS{v.~} \GE{to rot} \PD{po᷉ɔ᷉ɛ᷉, poɔnɔ, poɔnɔ, poɔnema, poɔnaa, - } 
\LX{po᷉ɔ} \HM{} \PH{pʊ́᷉ɔ́᷉} \PS{v.~} \GE{to play a game of hide and seek} \PD{po᷉ɔ᷉ɛ᷉, poɔnɔ, poɔnɔ, poɔnema, poɔnaa, - } 
\LX{pu} \HM{} \PH{pù} \PS{v.~} \GE{to flower} \PD{pue, puuro, puuro, puuribo, puuraa, - } \VA{puuri}
\LX{pu} \HM{} \PH{pù} \PS{v.~} \GE{to pluck very tender fruits e.g. okra} \PD{pue, puuro, puuro, puuribo, puuraa, - } \VA{puuri}
\LX{pu} \HM{} \PH{pú} \PS{v.~} \GE{direct to} \XV{O naŋ yeli a yɛlɛ o ba pore neɛ zaa yuori kyɛ te baŋe la neɛ na o naŋ pu.} \XE{S/he did not mention any person’s name in the talk but we know who it was directed to.} \PD{-, -, -, -, -, - } 
\LX{pugi} \HM{} \PH{pùgì} \PS{v.~} \GE{praise, adore} \PD{-, -, -, -, -, - } 
\LX{pugi} \HM{} \PH{púgí} \PS{v.~} \SN \GE{to disturb e.g. dust, water} \SN \GE{to knead e.g. flour or shea butter dough} \PD{pugee, pugiro, pugiro, pugiribo, pugiraa, - } 
\LX{pugi} \HM{} \PH{pùgì} \PS{v.~} \GE{to praise} \PD{pugee, pugiro, pugiro, pugiribo, pugiraa, - } 
\LX{pugo} \HM{} \PH{púgó} \PS{adj.~} \GE{gray} \OP{pugiri} \TP{pugiree} 
\LX{puli} \HM{} \PH{púlí} \PS{adv.~} \GE{under, below} 
\LX{puli} \HM{} \PH{pùlí} \PS{n.~} \GE{an emergency exit of rodents that live in holes} \OP{pulo} \TP{pulee} 
\LX{puli} \HM{} \PH{púlí} \PS{n.~} \GE{hind toe of a bird} \OP{pulli} \TP{pulee} 
\LX{pulli} \HM{} \PH{pùllì} \PS{v.~} \GE{to mix} \XV{Ba de la bɛŋɛ pulli ne a mui.} \XE{They mixed the rice with beans.} \PD{pulee, pullo, pullo, pulibo, pullaa, pulibo } 
\LX{pulli} \HM{} \PH{púllì} \PS{v.~} \SN \GE{to repeat} \SN \GE{to fold over} \PD{pullee, pullo, pullo, pulibo, pullaa, pulibo } 
\LX{pulluu} \HM{} \PH{púllúú} \PS{n.~} \SN \GE{repetition} \SN \GE{refrain (poem, song)} 
\LX{pulompuloŋ} \HM{} \PH{pùlòmpúlóŋ} \PS{adj.~} \GE{mixed up e.g. different items} 
\LX{puloŋ} \HM{} \PH{púlóŋ} \PS{n.~} \GE{sound of a small object falling into water} 
\LX{pulopulo} \HM{} \PH{púlópúló} \PS{adj.~} \GE{less serious than} \XV{Ka fooŋ la yi tensogɔ saŋa, a feɛbo ŋa fo ba naŋ feɛ fo zenɛ na waa la pulopulo.} \XE{If you go out in the night again, the punishment you have received today will be less serious than what you will get.} 
\LX{pulopulo} \HM{} \PH{púlópúló} \PS{adj.~} \GE{incomparable} \XV{Tentoŋ saŋa ŋmenaa maŋ tasoga puri la ka vũũ meŋ waa pulopulo.} \XE{During the hot season sometimes the heat of the sun is incomparable to fire.} 
\LX{pulopulo} \HM{} \PH{pùlópùló} \PS{adj.~} \GE{made of different varieties} \XV{A tigiri daare o da di la bondi pulopulo te zuo lɛnso ka o poɔ da beɛrɛ.} \XE{On the feast day s/he ate too many assorted dishes and that was why s/he had a stomach ache.} 
\LX{puluŋ} \HM{} \PH{pùlúŋ} \PS{n.~} \GE{a type of grass with sharp tips that can pierce soft or light objects e.g. feet} 
\LX{pumpuŋkɔrɔ} \HM{} \PH{púmpúŋkɔ̀rɔ̀} \PS{n.~} \GE{a type of poisonous plant} \OP{pumpuŋkɔrɛɛ} 
\LX{pumpuroŋ} \HM{} \PH{púmpúróŋ} \PS{adv.~} \GE{small, round, and hairy e.g. tail end of a squirrel} 
\LX{pumpuroŋ} \HM{} \PH{pùmpùròŋ} \PS{adv.~} \GE{large, round, and furry} 
\LX{puo} \HM{} \PH{púò} \PS{n.~} \SN \GE{small farm near the house} \SN \GE{farmland} \OP{poe} \TP{poree} 
\LX{puoli} \HM{} \PH{púólí} \PS{v.~} \GE{to saturate with water} \PD{puolee, puolo, puolo, puolibo, puolaa, - } 
\LX{puoraa} \HM{} \PH{pùòràá} \PS{n.~} \SN \GE{greetings} \SN \GE{gratitude} \OP{puoro} 
\LX{puoraa karetaa} \HM{} \PH{pùòràá káretàá} \PS{n.~} \GE{postcard} \OP{puoraa karetaare} \TP{puoraa karetaarɛɛ} 
\LX{puori} \HM{} \PH{pùòrí} \PS{n.~} \GE{back (body)} \OP{poe} \TP{poree} 
\LX{puori} \HM{} \PH{púórì} \PS{n.~} \GE{stomach; the organ in humans and animals in which food is stored} \OP{poe} \TP{puoree} 
\LX{puori} \HM{} \PH{pùòrì} \PS{v.~} \SN \GE{to pray} \SN \GE{to greet} \SN \GE{to thank} \PD{puoree, puoro, puoro, puoribo, puoraa, puoribo } 
\LX{puori} \HM{} \PH{pùòrí} \PS{adv.~} \SN \GE{behind} \SN \GE{backward (direction)} 
\LX{puoriŋ} \HM{} \PH{púòríŋ} \PS{conj.~} \GE{after} 
\LX{puoruu} \HM{} \PH{púórúú} \PS{n.~} \GE{prayer} \OP{puoro} 
\LX{puree} \HM{} \PH{pùrèé} \PS{n.~} \GE{paternal aunt, father's sister} \OP{purri/purimine} \TP{purree} 
\LX{puri} \HM{} \PH{pùrì} \PS{v.~} \SN \GE{to burst, to break} \SN \GE{to appear suddenly} \SN \GE{to rise (sun)} \PD{puree, puro, puro, puribo, puraa, puribo } \SE{puri ŋmenaa} \GE{sunrise} 
\LX{puri wale} \HM{} \PH{pùrì wálɪ́} \PS{v.~} \SN \GE{to sweat, to perspire} \SN \GE{to struggle, to be in a difficult situation} \PD{-, -, -, -, -, - } 
\LX{puro} \HM{} \PH{púrò} \PS{n.~} \GE{a type of tree that has sour sap which is used in preparing porridge} \OP{purri} \TP{purree} 
\LX{puro} \HM{} \PH{púrò} \PS{n.~} \GE{a type of small tree squirrel} \OP{purri} \TP{purree} 
\LX{puu} \HM{} \PH{pùù} \PS{v.~} \GE{to praise, to congratulate} \PD{puue, puuro, puuro, puuribo, puuraa, - } 
\LX{puulee} \HM{} \PH{púúléé} \PS{n.~} \GE{lady (typically used with bi-)} \OP{puuli} 
\LX{puuli} \HM{} \PH{púúlì} \PS{n.~} \GE{a funeral rhythm; a type of musical performance at a funeral rites session} \OP{puulo} \TP{puulee} 
\LX{puuri} \HM{} \PH{púúrí} \PS{n.~} \GE{flowers} \SG{puuruu} \TP{puuree} \VA{tepuuri}
\LX{puuri} \HM{} \PH{pùùrì} \PS{v.~} \GE{to spray from the mouth (usually chewed material as medicine)} \PD{puuree, puuro, puuro, puuribo, puuraa, - } \VA{pu}
\LX{puuri} \HM{} \PH{pùùrì} \PS{v.~} \GE{to harvest okra prematurely} \PD{puuree, puuro, puuro, puuribo, puuraa, - } \VA{pu}
\LX{pu᷉o} \HM{} \PH{pù᷉ò᷉} \PS{v.~} \GE{to make a hole through an object} \PD{pu᷉oe, puono, puono, puono, puonimo, puonaa } 
\LX{pɛrɛpɛrɛ} \HM{} \PH{pɛ́rɛ́pɛ́rɛ́} \PS{adv.~} \GE{completely} \XV{A nimibaaloŋ vɛŋɛɛ la ka a dɔɔ zɔŋe pɛrɛpɛrɛ.} \XE{The eye disease has caused him to be blind completely.} 
\LX{sa} \HM{} \PH{sá} \PS{v.~} \GE{to administer an enema} \PD{saɛ, saara, saara, sareba, saaraa, - } 
\LX{saa} \HM{} \PH{sàá} \PS{n.~} \GE{father} \OP{saamine} \TP{saarɛɛ} \VA{ba}
\LX{saa} \HM{} \PH{sáá} \PS{n.~} \GE{rain} \TP{saarɛɛ} 
\LX{saa} \HM{} \PH{sáá} \PS{adv.~} \GE{even if; short form of saareŋ} \XV{Ka fooŋ saa be gaa kyɛ fo koŋ nyɛ o.} \XE{Even if you go you wont see him/her.} \VA{shortened form of saareŋ}
\LX{saabo} \HM{} \PH{sáábʊ́} \PS{n.~} \GE{flour paste (staple food of the Dagaare-speaking people)} \TP{saarɛɛ} \VA{TZ}
\LX{saabuulaa} \HM{} \PH{sààbùùláà} \PS{n.~} \GE{porridge set for preparing saabo} \OP{saabuuli} 
\LX{saadaare} \HM{} \PH{sáádáárɪ́} \PS{adv.~} \GE{rainy, wet day} 
\LX{saadoge} \HM{} \PH{sààdʊ́gɪ́} \PS{n.~} \GE{large cooking pot} \OP{saadogere} \TP{saadogerɛɛ} 
\LX{saafebiri} \HM{} \PH{sààfɪ̀bírì} \PS{n.~} \GE{key} \OP{saafebie} \TP{saafebiruu} 
\LX{saafee} \HM{} \PH{sáàfɪ́ɪ́} \PS{n.~} \GE{lock} \OP{saafeere} \TP{saafeerɛɛ} \VA{safoa, safoɔ}
\LX{sa-ageraa} \HM{} \PH{sà-ágɪ́ráà} \PS{n.~} \GE{rainclouds that gather and suddenly give way to clear weather} \OP{sa-agere} \TP{sa-agerɛɛ} 
\LX{saakaŋkaŋaa} \HM{} \PH{sáákǎŋkàŋáá} \PS{n.~} \GE{a type of tree} \OP{saakaŋkanne} \TP{saakaŋkannɛɛ} \VA{kaŋkaŋaa}
\LX{saakaraa} \HM{} \PH{sààkáráá} \PS{n.~} \GE{flour paste not fully cooked, still part raw} \OP{saakarare} \TP{saakarɛɛ} 
\LX{saakɔntuo} \HM{} \PH{sáákɔ̌ntúó} \PS{n.~} \GE{Water from the first rain of the onset of the rainy season} \OP{saakɔntuuri} \TP{saakɔntuuree} 
\LX{saako᷉ɔ᷉} \HM{} \PH{sààkʊ́᷉ɔ́᷉} \PS{n.~} \GE{water in mashed poridge} \TP{saakoɔnɛɛ} 
\LX{saala} \HM{} \PH{sáálá} \PS{n.~} \GE{charcoal} \SG{saaloo} \TP{saalɛɛ} 
\LX{saalaa} \HM{} \PH{sààlàà} \PS{adj.~} \SN \GE{smooth} \SN \GE{slippery} \OP{saale} 
\LX{saalaa} \HM{} \PH{sááláá} \PS{n.~} \GE{object used for sharpening implements} \OP{saale} \TP{saaɛɛ} 
\LX{saalaa} \HM{} \PH{sààlàà} \PS{n.~} \GE{a type of tree with hard wood} \OP{saale} 
\LX{saalaa} \HM{} \PH{sààláá} \PS{n.~} \GE{bowl of T.Z} \OP{saalaare} \TP{saalaarɛɛ} 
\LX{saale} \HM{} \PH{sààlɪ̀} \PS{v.~} \SN \GE{to sharpen e.g. knife} \SN \GE{to make smooth, to plaster} \XV{Ayuo saale la o dipaalaa na o serɛ naŋ ŋmaa.} \XE{Ayuo has plastered the new room her husband put up.} \PD{saalɛɛ, saala, saala, saaleba, saalaa, - } 
\LX{saale} \HM{} \PH{sààlɪ̀} \PS{v.~} \GE{to be sufficient} \XV{A bondirii ama eŋ na saale ba la.} \XE{As for this food, it will be sufficient for them.} \PD{saalɛɛ, saala, saala, saaleba, saalaa, - } 
\LX{saalesɔglaa} \HM{} \PH{sààlɪ̀sɔ́gláá} \PS{n.~} \GE{a type of tree with hard wood and medicinal properties} \OP{saalesɔglɔ} \TP{saalesɔglɛɛ} 
\LX{saalieni} \HM{} \PH{sààlìèní} \PS{n.~} \GE{a type of small plant with an edible, jellylike tuber} \OP{saalieme} \TP{saalienee} 
\LX{saalombugo} \HM{} \PH{sààlʊ̀mbùgò} \PS{n.~} \GE{a type of small plant whose leaves are used in place of okra} \TP{saalombugiree} \VA{saalompelaa}
\LX{saalommaane} \HM{} \PH{sààlʊ̌mmáánɪ̀ɪ̀} \PS{n.~} \GE{okra} \OP{saalommaana} \VA{maanee}
\LX{saalompelaa} \HM{} \PH{sààlʊ̀mpɪ̀làà} \PS{n.~} \GE{a type of small plant whose leaves are used in place of okra} \OP{saalompeɛle} \TP{saalompeɛlɛɛ} \VA{saalombugo}
\LX{saaloŋ} \HM{} \PH{sààlʊ̀ŋ} \PS{adj.~} \GE{slippery, slimy} \XV{A mui sa᷉a᷉ɛ᷉ la, a waa la saaloŋ lɛ.} \XE{The rice is spoilt; it is slimy.} 
\LX{saaloŋ} \HM{} \PH{sáálʊ́ŋ} \PS{n.~} \GE{okra or other slimy plant used in soup} 
\LX{saalɔre} \HM{} \PH{sààlɔ́rɪ́} \PS{n.~} \GE{ball of T.Z. (usually formed by putting T.Z in water)} \SG{saalɔroo} \OP{saalɔrɔ} \TP{saalɔrɛɛ} 
\LX{saaŋ} \HM{} \PH{sááŋ} \PS{n.~} \GE{see zambɛre} \OP{saane} \TP{saanɛɛ} \VA{sa᷉a}
\LX{saana} \HM{} \PH{sáánà} \PS{n.~} \GE{stranger, visitor} \OP{saama} \TP{saanɛɛ} 
\LX{saanaa} \HM{} \PH{sáánáá} \PS{n.~} \GE{fault, wrong} \OP{saane} 
\LX{saanweɛleŋ} \HM{} \PH{sáánwɪ́ɛ́lɪ́ŋ} \PS{n.~} \SN \GE{opposite, contradiction} \SN \GE{irony (literature)} \VA{saanweɛloŋ}
\LX{saapɛle} \HM{} \PH{sààpɛ́lɪ́} \PS{n.~} \GE{layer of T.Z} \OP{saapɛlɛ} \TP{saapɛlɛɛ} 
\LX{saapuori} \HM{} \PH{sàápúórì} \PS{n.~} \GE{the middle of T.Z} \OP{saapoe} \TP{saaporee} 
\LX{saapurimo} \HM{} \PH{sààpúrímó} \PS{n.~} \GE{pieces of flour paste (T.Z.)} \TP{saapurimee} 
\LX{saaraa} \HM{} \PH{sááráá} \PS{n.~} \GE{saw} \OP{saare} \TP{saarɛɛ} 
\LX{saare} \HM{} \PH{sáárɪ̀} \PS{n.~} \GE{broom} \OP{saɛ} \TP{saarɛɛ} 
\LX{saaremuri} \HM{} \PH{sáármúrí} \PS{n.~} \GE{broom straw} \OP{saaremue} \TP{saamuree} 
\LX{saareŋ} \HM{} \PH{sáárɪ́ŋ} \PS{adj.~} \GE{even if} \XV{Ka fooŋ saareŋ gaa kyɛ fo koŋ nyɛ o.} \XE{Even if you go, you will not see her.} \VA{saa}
\LX{saa-toloŋ} \HM{} \PH{sàá-tʊ́lʊ́ŋ} \PS{n.~} \GE{hot TZ} \OP{saa-tolɔ/saa-tola} 
\LX{saazu-kogilo} \HM{} \PH{sààzú-kógiló} \PS{n.~} \GE{dried top of T.Z} \SG{saazu-kogiluu} 
\LX{sabaabo} \HM{} \PH{sàbáábʊ́} \PS{adj.~} \GE{mysterious} 
\LX{sabeelɛ} \HM{} \PH{sàbɪ́ɪ́lɛ̀} \PS{n.~} \GE{civilian} \SG{sabeeloo} 
\LX{sabɛlemɛ} \HM{} \PH{sàbɛ́lɪ́mɛ́} \PS{n.~} \GE{earthworm} \OP{sabɛlemɛre} \TP{sabɛlemɛrɛɛ} \VA{sɛbɛlemɛ, short form of sabɛlmɛkyɔɔ}
\LX{sabɛlemɛkyɔɔ} \HM{} \PH{sàbɛ́lɪ́mɛ̀ʧɔ́ɔ́} \PS{n.~} \GE{earthworm} \OP{sabɛlemɛkyɔɔre} \TP{sabɛlemɛkyɔɔrɛɛ} \VA{sabɛlemɛ}
\LX{sabiri} \HM{} \PH{sàbírì} \PS{n.~} \GE{raindrop} \OP{sabie} \TP{sabiiree} 
\LX{sabogi} \HM{} \PH{sábógí} \PS{n.~} \GE{something that causes concern e.g incessant complain} \XV{A yi a pɔge bie naŋ kpi la ka o deɛ maŋ yele sabogi yɛlɛ.} \XE{Since the woman lost her child she has always been complaining.} 
\LX{sabogi} \HM{} \PH{sàbògí} \PS{n.~} \GE{a type of fetish believed to have healing power and protection especially for children} \OP{sabogri} \TP{sabogree} 
\LX{sabɛlemɛkyɔɔ} \HM{} \PH{sàbɛ́lɪ́mɛ́kyɔ́ɔ́} \PS{n.~} \GE{earthworm} \OP{sabɛlemɛkyɔɔre} \TP{sabɛlemɛkyɔɔrɛɛ} \VA{sabɛlemɛ; sɛbɛlemɛkyɔɔ,  sɛbɛlemɛ}\VA{sɛbɛlemɛkyɔɔ, sabɛlemɛ/ sɛbɛlemɛ.}
\LX{sadɛŋe} \HM{} \PH{sàdɛ́ŋ} \PS{n.~} \GE{first rain (first heavy rain after which the people sow their crops)} \OP{sadɛnne} \TP{sadɛnnɛɛ} \VA{sadɛŋɛ}
\LX{safeaa} \HM{} \PH{sàfɪ́áá} \PS{n.~} \GE{survey} \OP{safeere} \TP{safeerɛɛ} 
\LX{safekyeɛrɛ} \HM{} \PH{sàfɪ̀ʧɪ̀ɛ̀rɛ̀} \PS{n.~} \GE{surveyor} \OP{safekyeɛrebɛ} \TP{safekyeɛrebɛɛ} 
\LX{safoa} \HM{} \PH{sàfʋ̀á} \PS{n.~} \GE{lock} \OP{safoore} \TP{safoorɛɛ} \VA{safoɔ}
\LX{safobiri} \HM{} \PH{sàfʊ̀bírì} \PS{n.~} \GE{key} \OP{safobie} \TP{safobiiree} \VA{saafobiri, saafebiri}
\LX{safoɔ} \HM{} \PH{sàfʊ̀ɔ́} \PS{n.~} \GE{lock} \OP{safoore} \TP{safoorɛɛ} \VA{safoa, saafee}
\LX{sagane} \HM{} \PH{sàgánɪ̀} \PS{n.~} \GE{sky blue} \OP{sagama} \TP{sagannɛɛ} 
\LX{sagbini} \HM{} \PH{ságbíní} \PS{n.~} \GE{big bundle of traditional broomsticks. The local broom is usually a bundle of slim but strong straw} \OP{sagbime} \TP{sagbimee} 
\LX{sage} \HM{} \PH{sàgɪ̀} \PS{v.~} \SN \GE{to answer} \SN \GE{to agree, to accept, to obey} \SN \GE{to allow, to permit} \SN \GE{to admit guilt} \SN \GE{to receive} \SN \GE{to sing after a lead singer} \PD{sagɛɛ, sagra, sagra, sagreba, sagraa, sagreba } 
\LX{sage} \HM{} \PH{ságɪ́} \PS{v.~} \GE{to be piled up} \PD{sagɛɛ, sagera, sagera, sagereba, sageraa, sagereba } 
\LX{sagele} \HM{} \PH{sàgɪ̀lɪ̀} \PS{v.~} \GE{to be caught up; to be entangled e.g. stick thrown into a tree} \PD{sagelɛɛ, sagela, sagela, sageleba, sagelaa, - } 
\LX{sagere} \HM{} \PH{sàgɪ̀rɪ́} \PS{n.~} \GE{rubbish} \OP{saga} \TP{sagɛɛ} 
\LX{sagoŋgoŋŋo} \HM{} \PH{sàgòŋgóŋŋò} \PS{n.~} \GE{rainbow} \OP{sagoŋgonni} \TP{sagoŋgonnee} 
\LX{sakaliŋkuuree} \HM{} \PH{sàkálíŋkúúréé} \PS{n.~} \GE{whirlwind} \OP{sakaliŋkuuri} 
\LX{sakere (Eng)} \HM{} \PH{sákɪ̀rɪ̀} \PS{n.~} \GE{bicycle} \OP{sakeɛ} \TP{sakerɛɛ} \VA{baasakuuri}
\LX{sakoraa} \HM{} \PH{sàkóràá} \PS{n.~} \GE{gathered rain clouds; the condition of preparing to rain} \OP{sakorri} \TP{sakorree} 
\LX{sakɔre} \HM{} \PH{sàkɔ́rɪ́} \PS{n.~} \GE{scrap of saabo left in the pot} \TP{sakɔrɛɛ} 
\LX{sakɔre} \HM{} \PH{sákɔ́rɪ́} \PS{adj.~} \SN \GE{useless, lazy} \XV{Ba-sakɔre ba nɔŋ mɔyɔ.} \XE{A lazy dog does not like hunting.} \SN \GE{dirty, untidy} \XV{Nensaal-sakɔre zaa maŋ nyuuro la.} \XE{Every dirty person has an odour.} 
\LX{sako᷉ɔ} \HM{} \PH{sàkʊ̀᷉ɔ́᷉} \PS{n.~} \GE{heavy rain} \TP{sakoɔnɛɛ} 
\LX{sakperɛ} \HM{} \PH{sákpɪ́rɛ́} \PS{n.~} \GE{disused broom} \OP{sakperre} \TP{sakperɛɛ} 
\LX{sakubie} \HM{} \PH{sákúbìé} \PS{n.~} \GE{student} \OP{sakubiiri} \TP{sakubiiree} 
\LX{sakubo} \HM{} \PH{sàkúbò} \PS{n.~} \GE{hail} \OP{sakuuri} \TP{sakuuree} 
\LX{sakuuri} \HM{} \PH{sàkúúrí} \PS{n.~} \SN \GE{hailstone, ice cube} \OP{sakue} \TP{sakuuree} 
\LX{sakuuri} \HM{} \PH{sàkùùrì} \PS{n.~} \GE{school} \OP{sakue} \TP{sakuuree} 
\LX{sakuuribie} \HM{} \PH{sàkúúríbíé} \PS{n.~} \GE{student} \OP{sakuuribiiri} \VA{sakubie}
\LX{sakyeɛre} \HM{} \PH{sàʧɪ̀ɛ̀rɪ̀} \PS{n.~} \GE{a type of hat with a large brim e.g. cowboy} \OP{sakyeɛrɛ} 
\LX{sakyɔre} \HM{} \PH{sàʧɔ̀rɪ̀} \PS{adv.~} \GE{height of the rainy season} 
\LX{salala} \HM{} \PH{sàlàlà} \PS{adj.~} \GE{dull, passive} 
\LX{salamante} \HM{} \PH{sàlàmántɪ̀} \PS{n.~} \GE{cement} \OP{salmantere} \TP{salmanterɛɛ} 
\LX{Salaŋa} \HM{} \PH{sàláŋá} \PS{n.~} \GE{a person who works as a cleaner and orderly maintenance of the innermost part of a chief’s palace} \OP{salanne} \TP{salannɛɛ} 
\LX{salansalaŋ} \HM{} \PH{sálánsáláŋ} \PS{adj.~} \GE{smooth} 
\LX{sale} \HM{} \PH{sàlɪ̀} \PS{v.~} \SN \GE{to be dispirited} \XV{N woŋ la o fieluu yɛle, ka n eŋɛ zaa sale.} \XE{When I heard of her failure, I was disturbed.} \SN \GE{to cook a small amount soup} \XV{A biyaala seŋ ka o sale zeɛre ko a bie.} \XE{The babysitter should be able to cook some small food for the child.} \PD{salɛɛ, salla, salla, salba, sallaa, - } 
\LX{sale} \HM{} \PH{sàlɪ̀} \PS{v.~} \GE{to make sauce slippery} \PD{-, -, -, -, -, - } 
\LX{sale} \HM{} \PH{sálɪ́} \PS{n.~} \GE{a type of plant with medicinal properties, believed to provide strength and power} \TP{salɛɛ} 
\LX{salema} \HM{} \PH{sálɪ̀má} \PS{n.~} \GE{gold} \TP{sallɛɛ/salemɛɛ} 
\LX{salembogi} \HM{} \PH{sàlɪ̀mbògí} \PS{n.~} \GE{gold mine} \OP{salembogri} \TP{salembogree} 
\LX{salempelaa} \HM{} \PH{sàlɪ̀mpɪ́làá} \PS{n.~} \GE{diamond} \OP{salempeɛle} \TP{salempeɛlɛɛ} 
\LX{salendoge} \HM{} \PH{sàlɪ̀ndʊ̀gɪ̀} \PS{n.~} \GE{treasury (traditional)} \OP{salendogre} \TP{salendogrɛɛ} \VA{salendoglee}
\LX{salendogelee} \HM{} \PH{sàlɪ̀ndʊ́gléé} \PS{n.~} \GE{a pot made of gold} \GE{a pot in which gold or treasure is kept} \OP{salendogebilii} \VA{salendoge}
\LX{saleŋgoɔ} \HM{} \PH{sálɪ́ŋgʊ́ɔ́} \PS{n.~} \GE{a type of tree with long, spiky thorns} \OP{saleŋgoore} \TP{saleŋgoorɛɛ} \VA{sansaŋgoɔ}
\LX{saleŋkuuri} \HM{} \PH{sàlɪ̀ŋkúúrí} \PS{n.~} \SN \GE{gold nugget} \SN \GE{pearl} \OP{saleŋkubo} 
\LX{salinsoo-naadaperaa} \HM{} \PH{sàlìnsòò-náádàpɪ́ráà} \PS{n.~} \GE{large grasshopper} \OP{salinso-naadaperre} \TP{salinso-naadaperɛɛ} 
\LX{salinsoori} \HM{} \PH{sàlìnsóórí} \PS{n.~} \GE{grasshopper} \OP{salinsobo} \TP{salinsooree} 
\LX{salinsuuri} \HM{} \PH{sálínsúúrí} \PS{n.~} \GE{shadow} \VA{dasuoluŋ}
\LX{saluoni} \HM{} \PH{s̀alúóní} \PS{n.~} \GE{sky} \OP{saluomo} \TP{saluonee} 
\LX{samafuu} \HM{} \PH{sàmàfúù} \PS{n.~} \GE{worth nothing in the end} 
\LX{samane} \HM{} \PH{sámánɪ́} \PS{n.~} \GE{immediate surroundings of a house} \OP{samama} \TP{samanɛɛ} 
\LX{samane} \HM{} \PH{sámánɪ́} \PS{n.~} \GE{toilet} \OP{samama} \TP{samanɛɛ} 
\LX{samasama} \HM{} \PH{sàmàsámà} \PS{n.~} \GE{sanitary inspector} \OP{sansammine} 
\LX{samasama} \HM{} \PH{sámásámá} \PS{adj.~} \GE{occupied and filled to capacity} 
\LX{sambakelee} \HM{} \PH{sámbákéléé} \PS{n.~} \GE{pottery shard} \OP{sambakelli} \VA{semekelee}
\LX{sambala} \HM{} \PH{sámbálá} \PS{n.~} \GE{a type of long, hairy grass} \SG{sambaloo} \OP{sambala} \TP{sambalɛɛ} 
\LX{sa-miiraa} \HM{} \PH{sa-miiraa} \PS{n.~} \GE{rain that is possible to come} \OP{sa-miiri} 
\LX{sampa} \HM{} \PH{sámpá} \PS{n.~} \GE{a type of shed the roof of which is used to store corn and the under as a place to relax} \OP{sampare} \TP{samparɛɛ} 
\LX{sampa} \HM{} \PH{sàmpá} \PS{n.~} \GE{raised stand with grass/stalks on it on which corn is stored or resting place} \OP{sampare} \TP{samparɛɛ} 
\LX{sampane} \HM{} \PH{sámpánɪ́} \PS{n.~} \GE{wooden foot-shaped mallet for beating floor to plaster it} \OP{sampama} \TP{sampanɛɛ} 
\LX{sampire} \HM{} \PH{sàmpìré} \PS{n.~} \GE{clusters of grass} \TP{sampirree} \VA{sempire}
\LX{sampulo} \HM{} \PH{sámpúló} \PS{adv.~} \GE{several times} \XV{O wa la kyɛ sampulo bowola zaa kyɛ ba nyɛ fo.} \XE{S/he came here several times but did not see you.} 
\LX{saŋa} \HM{} \PH{sáŋɪ̀} \PS{n.~} \GE{time, season} \TP{sannɛɛ} \VA{saŋe}
\LX{sanaŋe} \HM{} \PH{sànáŋ} \PS{n.~} \GE{a type of large, black scorpion} \OP{sananne} \TP{sanannɛɛ} 
\LX{sandaana} \HM{} \PH{sándááná} \PS{n.~} \GE{creditor, one who is owed money} \OP{sandaama} \TP{sandaamɛɛ} 
\LX{sandire} \HM{} \PH{sándíré} \PS{n.~} \GE{debtor, one who owes money} \OP{sandiribe} \TP{sandiribee} 
\LX{sane} \HM{} \PH{sánɪ̀} \PS{n.~} \GE{debt} \OP{sama} \TP{sannɛɛ} 
\LX{saŋe} \HM{} \PH{sáŋɪ̀} \PS{v.~} \GE{to heal, to cure} \XV{Te᷉e᷉ ama la daare soŋ saŋ a baala.} \XE{This medicine helped in curing the patient.} \PD{saŋɛɛ, sanna/saŋena, sanena, sanema, sansaŋena, saŋenaa } 
\LX{saŋgara} \HM{} \PH{sàŋgàrá} \PS{n.~} \GE{a type of worm found in kola nuts} 
\LX{saŋgorema} \HM{} \PH{} \PS{n.~} \GE{foot yaws} 
\LX{sangye᷉a᷈a᷈} \HM{} \PH{sǎŋdzɪ̀᷉á᷈á᷈} \PS{n.~} \SN \GE{a type of long water beetle} \SN \GE{crab} \OP{saŋgyeɛne} \TP{saŋgyeɛnɛɛ} 
\LX{saniŋe} \HM{} \PH{sàníŋè} \PS{n.~} \GE{west} \VA{luou}
\LX{saŋkora} \HM{} \PH{sáŋkʊ́rá} \PS{n.~} \GE{old indebtedness} \XV{Ka fooŋ pɛŋe libie ba yɔɔ baare kyɛ la boɔrɔ a mine see ka fo yɔɔ a fo saŋkora na sɛre.} \XE{If you borrow money and have not paid back you will not get a new loan unless you pay up your old debt.} 
\LX{saŋkpalema} \HM{} \PH{sáŋkpálɪ́mà} \PS{n.~} \GE{a kind of large antelope} \OP{saŋkpalle} 
\LX{saŋkpana} \HM{} \PH{sǎŋkpáná} \PS{n.~} \GE{scabies} \TP{saŋkpanɛɛ} 
\LX{sanlɛre} \HM{} \PH{sànlɛ̀rɪ̀} \PS{n.~} \GE{straw put up together to form a mat on which harvested ears of guinea corn is kept to dry} \OP{sanlɛrre} \TP{sanlɛrɛɛ} \VA{sɛnlɛre}
\LX{saŋmene} \HM{} \PH{sáŋmɪ́nɪ̀} \PS{n.~} \GE{god of rain} \OP{saŋmemɛ} \TP{saŋmennɛɛ} 
\LX{sanna} \HM{} \PH{sànnà} \PS{v.~} \GE{to caution, to warn} \XV{Fo naŋ daare to a dɔɔ ŋa lɛ na aŋ sanna o yiri gaabo.} \XE{The way you insulted this man, you's better be warned not to go to his house.} \PD{sannɛɛ, sanna, sanna, sanema, sannaa, - } 
\LX{sanne} \HM{} \PH{sánnɪ̀} \PS{v.~} \GE{to attempt} \PD{sannɛɛ, sanna, sanna, sanema, sannaa, - } 
\LX{sansalaŋ} \HM{} \PH{sànsàlàŋ} \PS{adv.~} \GE{disorderly} \VA{sansaraŋ}
\LX{sansaŋgoɔ} \HM{} \PH{sá nsa ́ŋgʊ́ɔ̀} \PS{n.~} \GE{a type of thorn tree} \OP{sansaŋgoore} \TP{sansaŋgoorɛɛ} \VA{saleŋgoɔ}
\LX{sansaraŋ} \HM{} \PH{sánsáráŋ} \PS{adv.~} \GE{making moves to leave all of a sudden (without prior notice)} \VA{sansalaŋ}
\LX{sanyegeraa} \HM{} \PH{sànyɪ́gɪ́ráá} \PS{n.~} \GE{lightening (flashes of light in a storm)} \OP{sanyegere} \TP{sanyegerɛɛ} 
\LX{san-yɔɔ} \HM{} \PH{sán-yɔ́ɔ́} \PS{n.~} \GE{salary} \OP{san-yɔɔre} \TP{san-yɔɔrɛɛ} 
\LX{sapare} \HM{} \PH{sàpárɪ́} \PS{n.~} \GE{east} \OP{saparre} \TP{saparɛɛ} 
\LX{sapelaa} \HM{} \PH{sàpɪ̀làá} \PS{n.~} \SN \GE{rain from cumulus clouds} \SN \GE{a type of rain god (white)} \OP{sapeɛle} \TP{sapeɛlɛɛ} 
\LX{sapigi} \HM{} \PH{sápígí} \PS{n.~} \GE{large hat} \OP{sapige} \TP{sapgree} 
\LX{sapi᷈i} \HM{} \PH{sàpí᷉ì᷉} \PS{n.~} \GE{bolt of lightning} \OP{sapiime} \TP{sapiinee} 
\LX{sapɔ} \HM{} \PH{sàpɔ́} \PS{n.~} \GE{sponge} \OP{sapɔre} \TP{sapɔrɛɛ} 
\LX{sapooraa} \HM{} \PH{sàpʊ́ʊ́ráá} \PS{n.~} \GE{rain shower} \OP{sapoore} \TP{sapoorɛɛ} 
\LX{saraa} \HM{} \PH{sáràà} \PS{adj.~} \GE{young and female (human and nonhuman)} \OP{sarre} \TP{sarrɛɛ} 
\LX{sare} \HM{} \PH{sàrɪ̀} \PS{v.~} \GE{to slip} \PD{sarɛɛ, sara, sara, sareba, saraa, - } 
\LX{sare} \HM{} \PH{sárɪ̀} \PS{n.~} \GE{water trough (container from which animals drink)} \OP{saɛ} \TP{sarɛɛ} \VA{sɛre}
\LX{sareka} \HM{} \PH{sàrɪ́kà} \PS{n.~} \GE{a ritual in which one provides for the needy for atonement} \OP{sarekare} \TP{sakarɛɛ.} 
\LX{sarre} \HM{} \PH{sárrɪ̀} \PS{v.~} \GE{to develop into a young woman} \PD{sarrɛɛ, sarra, sarra, sareba, sarraa, - } 
\LX{saseɛ} \HM{} \PH{sàsɪ̀ɛ́} \PS{n.~} \SN \GE{air, wind} \SN \GE{storm} \OP{saseere} \TP{saseerɛɛ} 
\LX{sase-ɔɔre} \HM{} \PH{sásɪ̀-ɔ́ɔ́rɪ̀} \PS{n.~} \GE{harmattan} \TP{sase-ɔɔrɛɛ} 
\LX{sasepilimbaŋe} \HM{} \PH{sàsɪ̀pílímbɔ́ŋ} \PS{n.~} \GE{whirlwind} \OP{sasepilimbanne} \TP{sasepilimbannɛɛ} \VA{sasepilimbɔŋ}
\LX{sasere} \HM{} \PH{sàsɪ́rɪ̀} \PS{n.~} \GE{grasscutter, greater cane rat (large rodent that feeds mostly on grass)} \SG{saseroo} \OP{saseɛ} \TP{saserɛɛ} 
\LX{sasigoo} \HM{} \PH{sàsígóó} \PS{n.~} \GE{hiccough} \OP{sasigri} \TP{sasigree} 
\LX{sasɔgelaa} \HM{} \PH{sàsɔ́gláá} \PS{n.~} \SN \GE{rain from black clouds} \SN \GE{a type of fetish or deity connected with rain (black)} \OP{sasɔgelɔ/sasɔgele} \TP{sasɔgelɛɛ} 
\LX{sasusulee} \HM{} \PH{sàsúsúléé} \PS{n.~} \GE{a type of tree} \OP{sasusulli} \TP{sasusullee} 
\LX{satannaa} \HM{} \PH{sàtànnáà} \PS{v.~} \GE{to thunder} \PD{-, -, -, -, -, - } 
\LX{savaale} \HM{} \PH{sàváálɪ́} \PS{n.~} \GE{rainstorm (at night)} \OP{savaala} \TP{savaalɛɛ} 
\LX{sawarɛɛ} \HM{} \PH{sàwárɛ́ɛ́} \PS{n.~} \GE{white clouds (non-rain-bearing)} \TP{sawarrɛɛ/sawaremɛɛ} \VA{sawarema}
\LX{sawarema} \HM{} \PH{sàwárɪ́má} \PS{n.~} \GE{white-looking cloud} \SG{sawaroo} \TP{sawaremɛɛ} \VA{sawarɛɛ}
\LX{sawɛge} \HM{} \PH{sáwɛ́gɪ́} \PS{n.~} \GE{long cloud sprawling across the sky} \OP{sawɛgere} \TP{sawɛgerɛɛ} 
\LX{sawɔlɔ} \HM{} \PH{sáwɔ́lɔ́} \PS{n.~} \GE{fried dough made from flour e.g. bean, corn, millet} \TP{sawɔlɛɛ} 
\LX{sazeɛ} \HM{} \PH{sázɪ̀ɛ̀} \PS{n.~} \GE{a type of soft broom} \OP{sazeere} \TP{sazeerɛɛ} 
\LX{sazɔŋɔ} \HM{} \PH{sàzɔ́ŋɔ̀} \PS{n.~} \GE{snow} \TP{sazɔnne} 
\LX{sazɔɔ} \HM{} \PH{sàzɔ̀ɔ́} \PS{n.~} \GE{first rain which begins the farming season} \OP{sazɔɔre} \TP{sazɔɔrɛɛ} 
\LX{sazu} \HM{} \PH{sàzú} \PS{n.~} \GE{sky} \OP{sazuri} \TP{sazuree} 
\LX{sazu} \HM{} \PH{sàzú} \PS{adv.~} \GE{up} 
\LX{sazupi᷉i᷉} \HM{} \PH{sázúpí᷉ì᷈} \PS{n.~} \GE{danger lurking above} 
\LX{sa᷉a} \HM{} \PH{sá᷉á} \PS{n.~} \GE{blacksmith's shop} \VA{saaŋ}
\LX{sa᷉a} \HM{} \PH{sá᷉à᷉} \PS{v.~} \GE{to mash, to mix} \XV{Kɔŋe kpɛ la a balee, sa᷉a saabo eŋe o.} \XE{The puppy is hungry; mash some TZ for it.} \PD{-, -, -, -, -, - } 
\LX{sa᷉a} \HM{} \PH{sá᷉à᷉} \PS{v.~} \GE{to appear, to be visible} \PD{-, -, -, -, -, - } 
\LX{sa᷈a} \HM{} \PH{sà᷈à} \PS{v.~} \SN \GE{to spoil, to destroy, to damage} \SN \GE{to be guilty} \PD{sa᷈aɛ, saana, saana, saanema, saanaa, saanema } 
\LX{sa᷉abile} \HM{} \PH{sà᷉á᷉bìlé} \PS{n.~} \GE{uncle (father's younger brother)} \OP{saabilii/saabilimine} \TP{saabilee} \VA{saambile}
\LX{sa᷉adɔɔ} \HM{} \PH{sá᷉á᷉dɔ́ɔ́} \PS{n.~} \GE{stranger (male)} \OP{sa᷉adɔbɔ} 
\LX{sa᷉akoma} \HM{} \PH{sà᷉à᷉kʊ́má} \PS{n.~} \GE{grandfather} \OP{saakommine} \TP{saakomɛɛ/saakonnɛɛ} \VA{saaŋkoma, naa}
\LX{sa᷉akoŋkommuulitoori} \HM{} \PH{sà᷅àkʊ̀ŋkʊ́mmúúlítóórí} \PS{n.~} \GE{great-grandfather} \OP{sa᷉akoŋkommuulitoorimine} \VA{sa᷉akontulimo}
\LX{sa᷉akoŋkorema} \HM{} \PH{sà᷉àkʊ̌ŋkʊ́rɪ́má} \PS{n.~} \GE{great-great-grandfather, forefather} \OP{sa᷉akoŋkoremamine} 
\LX{sa᷉akoŋkoroma} \HM{} \PH{sá᷉á᷉kʊ́kʊ́rʊ́ma} \PS{n.~} \GE{great-great-grandfather} \VA{sa᷉akontulimotulimo}
\LX{sa᷉akonnoŋ} \HM{} \PH{sa ̃̀ã̀ kʊ́ nnʊ́ ŋ} \PS{n.~} \GE{tradition, custom} 
\LX{sa᷉akontulimo} \HM{} \PH{sa᷉ ̀ a᷉ ̀kʋ́ntúlímó} \PS{n.~} \GE{great-grandfather} \OP{sa᷉akontulimine} \VA{sa᷉akontulo}
\LX{sa᷉apɔge} \HM{} \PH{sá᷉á᷉pɔ́gɪ́} \PS{n.~} \GE{stranger (female)} 
\LX{se} \HM{} \PH{sɪ̀} \PS{v.~} \GE{to skin (slaughtered animal)} \PD{seɛɛ, seerɛ, seerɛ, seerebɛ, seeraa, - } 
\LX{sɛ} \HM{} \PH{sɛ̀} \PS{adv.~} \GE{already} \XV{O gaaɛ la sɛ.} \XE{S/he has gone already.} 
\LX{sɛ} \HM{} \PH{sɛ́} \PS{v.~} \GE{to roast} \PD{sɛɛ, seɛrɛ, seɛrɛ, seɛreba, seɛraa, seɛre } 
\LX{sɛ} \HM{} \PH{sɛ́} \PS{v.~} \GE{to sue} \PD{sɛɛ, seɛrɛ, seɛrɛ, seɛreba, seɛraa, seɛre } 
\LX{sɛ} \HM{} \PH{sɛ̀} \PS{v.~} \GE{to sew} \PD{-, seɛrɛ, -, -, -, - } 
\LX{sebaa} \HM{} \PH{sɪ́báá} \PS{n.~} \GE{a type of wild animal} \OP{sebre} \TP{sebrɛɛ} 
\LX{sɛbɛlemɛkyɔɔ} \HM{} \PH{sɛ̀bɛ̌lɪ́mɛ̀ʧɔ́ɔ̀} \PS{n.~} \GE{earthworm} \VA{sabɛlemɛkyɔɔ}
\LX{sebiri} \HM{} \PH{sɪ́bírì} \PS{n.~} \GE{bee} \OP{sebie} \TP{sebiiree} 
\LX{sebitetebaa} \HM{} \PH{sɪ́bítɪ́tɪ́báá} \PS{n.~} \GE{bee stinger} \OP{sebiteterre} \TP{sebiteterrɛɛ} \VA{tetebaa}
\LX{sebo} \HM{} \PH{sɪ̀bʊ̀} \PS{v.~} \SN \GE{to affect} \XV{Ka fooŋ zagere toma gaabo maa naane ka fo na sebo.} \XE{If you refuse to go to work it will not affect me.} \SN \GE{to refuse to give} \XV{N sɔre la o kuuri ka o sebo ma.} \XE{I begged for his hoe, but he refused me.} \PD{sebɛɛ, sebrɛ, sebrɛ, sebrebɛ, sebraa, - } 
\LX{sebogi} \HM{} \PH{sɪ́bògí} \PS{n.~} \GE{beehive,  hole (usually in a tree) where bees have made honey} \OP{sebogiri} \TP{sebogiree} 
\LX{sedanegere} \HM{} \PH{sɪ́dànɪ́gɪ́rɪ́} \PS{n.~} \GE{brown, sour-tasting part of honey} \OP{sedanegere} \TP{sedanegerɛɛ} \VA{danegɛ}
\LX{sedoge} \HM{} \PH{sɪ́dʊ́gɪ̀} \PS{n.~} \GE{swarm trap pot for bees} \OP{sedogre} \TP{sedogrɛɛ} 
\LX{see} \HM{} \PH{séé} \PS{conj.~} \GE{unless e.g having no option} \XV{See ka fo daŋe iruu ka lɛ naane fo koŋ nyɛ a baala.} \XE{Unless you leave early otherwise you will not see the patient.} 
\LX{seɛ} \HM{} \PH{sɪ́ɛ́} \PS{n.~} \GE{waist} \OP{seere} \TP{seerɛɛ} 
\LX{seɛ} \HM{} \PH{sɪ̀ɛ̀} \PS{v.~} \SN \GE{to wear (on the waist)} \SN \GE{to dance} \SN \GE{to be better than, to surpass} \XV{A fanfane ŋaŋ seɛ a ŋa.} \XE{This soap is better than this.} \XV{Ka fooŋ mare a bie anaŋ seɛ fo naŋ na pɛgele.} \XE{To carry the baby on your back is better than carrying by your side.} \PD{-, seɛrɛ, -, -, -, - } 
\LX{seɛ} \HM{} \PH{sɪ́ɛ̀} \PS{n.~} \GE{soul, spirit (of a living person)} \OP{seere} 
\LX{seɛ} \HM{} \PH{sɪ̀ɛ́} \PS{n.~} \SN \GE{joke, teasing} \XV{A pɔge taa la poɔ ka o taaba erɛ o seɛ.} \XE{The woman is pregnant and her peers are teasing her.} \SN \GE{fresh air, wind} 
\LX{seɛ} \HM{} \PH{sɪ́ɛ́} \PS{adj.~} \GE{by the side (wall)} \XV{A boɔ gaŋ la a dakyini seɛ.} \XE{The goat is lying at the foot of the wall.} 
\LX{seɛ kpelle} \HM{} \PH{sɪ̀ɛ̀ kpɪ́llɪ́} \PS{v.~} \GE{to get someone in trouble} \PD{seɛɛ kpelle, seɛrɛ kpelle, -, -, -, - } 
\LX{seɛbaa} \HM{} \PH{sɪ́ɛ́báá} \PS{adj.~} \GE{considerably big} \XV{De peɛ naŋ e seɛbaa, ka lɛ naane a kyi koŋ kpɛ.} \XE{Take a good-sized basket, or else the millet will be too much for a small one.} \OP{seɛbaare} 
\LX{seɛbaa} \HM{} \PH{̀sɪ̀ɛ́báà} \PS{n.~} \GE{twine from daw-daw fruit} \OP{ seɛre} \TP{seɛrɛɛ} 
\LX{seebo} \HM{} \PH{sɪ́ɪ́bʊ́} \PS{n.~} \SN \GE{honey} \SN \GE{beehive} \OP{seere} \TP{seerɛɛ} \VA{seeo}
\LX{seɛgoɔ} \HM{} \PH{sɪ̀ɛ́gʊ́ɔ́} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{seɛgoore} 
\LX{seɛlaa} \HM{} \PH{sɪ̀ɛ̀làá} \PS{n.~} \SN \GE{clay eaten by cattle} \SN \GE{pit where cattle eat this clay} \OP{seɛle} \TP{seɛlɛɛ} 
\LX{seɛlee} \HM{} \PH{sɪ́ɛ́lɪ́ɪ́} \PS{n.~} \GE{mystery} 
\LX{seɛlee} \HM{} \PH{sɪ́ɛ́lɪ́ɪ́} \PS{adj.~} \GE{mysterious, strange} 
\LX{seene} \HM{} \PH{sɪ́ɪ́nɪ̀} \PS{v.~} \GE{to touch} \PD{seenɛɛ, seenɛ, seenɛ, seenemɛ, seenaa, - } \VA{seere}
\LX{seɛne} \HM{} \PH{sɪ́ɛ́nɪ́} \PS{n.~} \GE{porcupine} \OP{seɛmɛ} \TP{seɛnɛɛ} 
\LX{seɛne} \HM{} \PH{sɪ̀ɛ̀nɪ̀} \PS{v.~} \SN \GE{to grab firmly with fingers around the object e.g. neck} \SN \GE{to cram e.g a bag with clothes} \PD{seɛnɛɛ, seɛnɛ, seɛnɛ, seɛnemɛ, seɛnaa, - } 
\LX{seɛnoɔre} \HM{} \PH{sɪ̀ɛ́nʊ́ɔ́rɪ́} \PS{n.~} \GE{the beginning of the farming season} 
\LX{seeraa} \HM{} \PH{sɪ́ɪ́ráá} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{seere} \TP{seerɛɛ} \VA{seetere}
\LX{seere} \HM{} \PH{sɪ́ɪ́rɪ̀} \PS{n.~} \GE{honey} \TP{seerɛɛ} 
\LX{seetere} \HM{} \PH{sɪ́ɪ́tɪ́rɪ́} \PS{n.~} \GE{a type of tree with hard word and is also medicinal} \SG{seeterɛɛ} \TP{seeteere} \VA{seeraa}
\LX{sefaa} \HM{} \PH{sɪ́fáá} \PS{n.~} \GE{demon, evil spirit} \OP{sefaare} \TP{sefaarɛɛ} 
\LX{segammo} \HM{} \PH{sɪ́gàmmʊ́} \PS{n.~} \GE{honeycomb} \OP{segama} \TP{segamɛɛ} 
\LX{segampelaa} \HM{} \PH{sɪ́gǎmpɪ̀làá} \PS{n.~} \GE{white honeycomb (usually contains the pure honey)} \OP{segampeɛle} \TP{segampeɛlɛɛ} 
\LX{seganfuu} \HM{} \PH{sɪ́gànfúù} \PS{n.~} \GE{honeycomb without honey} \OP{seganfuuri} \TP{seganfuuree} 
\LX{sege} \HM{} \PH{sɪ́gɪ́} \PS{n.~} \GE{farm hut; camp} \OP{segere} \TP{segerɛɛ} 
\LX{sege} \HM{} \PH{sɪ́gɪ́} \PS{v.~} \SN \GE{to resemble;} \SN \GE{to name a child after someone (usually an anscestor)} \PD{segɛɛ, segerɛ, segerɛ, segerebɛ, segeraa, - } 
\LX{sɛge} \HM{} \PH{sɛ́gɪ́} \PS{v.~} \GE{to write} \PD{sɛgɛɛ, sɛgrɛ, sɛgrɛ, sɛgrebɛ, sɛgraa, - } 
\LX{sɛgebinyaŋaa} \HM{} \PH{sɛ́gɪ́bíɲáŋáá} \PS{n.~} \GE{vowel} \OP{sɛgebinyaane} 
\LX{sɛgebiri} \HM{} \PH{sɛ́gɪ́bírì} \PS{n.~} \GE{amulet, charm, fetish} \OP{sɛgebie} \TP{sɛgebiiree} 
\LX{sɛgebiri} \HM{} \PH{sɛ́gɪ́bírì} \PS{n.~} \GE{letter (consonant or vowel)} \OP{sɛgebie} \TP{sɛgebiiree} 
\LX{sɛgelɛgɛ} \HM{} \PH{sɛ́gɪ́lɛ́gɛ̀} \PS{adv.~} \GE{unable to move e.g. lazy dog} \VA{hɛglɛgɛ}
\LX{segeraa} \HM{} \PH{sɪ́gɪ́ráá} \PS{n.~} \GE{ancestral name given to a baby} \OP{segere} \TP{segerɛɛ} 
\LX{sɛgeraa} \HM{} \PH{sɛ̀gɪ̀ráà} \PS{n.~} \GE{a piece of flat object fixed alongside a hoe blade into the handle to make the blade firm} \OP{sɛgere} \TP{sɛgerɛɛ} \VA{shortened form of bonsɛgeraa}
\LX{sɛgere} \HM{} \PH{sɛ́gɪ́rɪ́} \PS{v.~} \GE{to arrange} \PD{sɛgerɛɛ, sɛgerɛ, sɛgrɛ, sɛgrebɛ, sɛgeraa, sɛgerebɛ } 
\LX{sɛgere} \HM{} \PH{sɛ̀gɪ̀rɪ̀} \PS{v.~} \GE{to stuff, to plug (holes, spaces)} \PD{sɛgerɛɛ, sɛgerɛ, sɛgerɛ, sɛgerebɛ, sɛgeraa, - } 
\LX{segesege} \HM{} \PH{sɪ̀gɪ̀sɪ́gɪ́} \PS{v.~} \GE{to attempt to move but being constrained by one thing or the other} \PD{segesegɛɛ, segesegerɛ, segesegerɛ, segesegerebɛ, segesegeraa, - } \VA{segesege}
\LX{sɛgesɛge} \HM{} \PH{sɛ́gsɛ́g} \PS{adv.~} \GE{making moves; before one is aware} \XV{Saa koe la ka a koɔreba yeli ka ba erɛ sɛgesɛge ka ona la miire.} \XE{Clouds had gathered and before the farmers could make any move it started raining.} 
\LX{segilekele} \HM{} \PH{ségílékélé} \PS{adv.~} \GE{standing upright, tall, and solitary e.g. tower} 
\LX{sekãa} \HM{} \PH{sɪ́káà} \PS{n.~} \GE{pure honey} \SG{seka᷉ao} \OP{sekaanɛɛ} \VA{sekãaboli}
\LX{sekãaboli} \HM{} \PH{sɪ́ká᷉á᷉bólí} \PS{n.~} \GE{pure honey} \OP{seka᷉a᷉bolo} \TP{seka᷉a᷉bolee} \VA{sekãa}
\LX{sekalmanyuo} \HM{} \PH{sɪ́kálmáɲùód} \PS{n.~} \GE{swam of bees (usually clustering on a tree branch)} \OP{sekalmanyuuri} \TP{sekalmanyuuree} 
\LX{sekpɔgere} \HM{} \PH{sɪ́kpɔ́gɪ́rɪ̀} \PS{n.~} \GE{proverb, riddle} \SG{sekpɔgeroo} \OP{sekpɔgɔ} \TP{sekpɔgerɛɛ} 
\LX{selaa} \HM{} \PH{sɪ̀làà} \PS{n.~} \GE{hawk} \OP{selle} \TP{sellɛɛ} \VA{gbaara, gbaare, gbaarɛɛ}
\LX{sele} \HM{} \PH{sɪ́lɪ̀} \PS{v.~} \GE{to decant} \PD{selɛɛ, sellɛ, sellɛ, selbɛ, sellaa, - } 
\LX{sɛle} \HM{} \PH{sɛ̀lɪ̀} \PS{v.~} \GE{to transplant} \PD{sɛlɛɛ, sɛllɛ, sɛllɛ, sɛlebɛ, sɛllaa, - } 
\LX{sɛle} \HM{} \PH{sɛ̀lɪ̀} \PS{v.~} \GE{to sit, perch (derogatory use)} \XV{A dɔɔ tanne la a bie,”N toŋe fo la ka fo zagere kyɛ wa sɛle n niŋeŋ boɔrɔ boŋ?”} \XE{The man shouted at the boy, “I sent you and you refused, what do you want  perched in front of me?”.} \PD{sɛlɛɛ, sɛllɛ, sɛllɛ, sɛlebɛ, sɛllaa, - } 
\LX{sɛle} \HM{} \PH{sɛ́lɪ̀} \PS{n.~} \GE{the act of winning a game of bie outright} \OP{sɛlɛ} \TP{sɛlɛɛ} 
\LX{selee} \HM{} \PH{séléé} \PS{n.~} \GE{pottery shard} \OP{sebilii} \VA{semekelee/sambekelee}
\LX{seleŋkaa} \HM{} \PH{sɪ́lɪ́ŋkáá} \PS{n.~} \GE{a type of small wasp} \OP{seleŋkaare} \TP{seleŋkaarɛɛ} \VA{selɛŋkaa, seleŋkpaa}
\LX{selɛŋkaa} \HM{} \PH{sɪ́lɛ́ŋkáà} \PS{n.~} \GE{a small brown wasp with a painful sting} \OP{selɛŋkaare} \TP{selɛŋkaarɛɛ} \VA{selɛŋkpaa, seleŋkaa}
\LX{selɛŋkpaa} \HM{} \PH{sɪ́lɛ́ŋkpáá} \PS{n.~} \GE{a small brown wasp with a painful sting} \OP{selɛŋkpaare} \TP{selɛŋkpaarɛɛ} \VA{selɛŋkaa, seleŋkaa}
\LX{selgyuuni} \HM{} \PH{sɪ̀ldzùùnì} \PS{n.~} \GE{a type of eagle} \OP{selgyuumo} \TP{selgyuunee} 
\LX{selle} \HM{} \PH{sɪ́llɪ̀} \PS{v.~} \GE{to tell, to recount (story)} \PD{sellɛɛ, sellɛ, sellɛ, sellebɛ, sellaa, - } 
\LX{selle} \HM{} \PH{sɪ́llɪ̀} \PS{v.~} \GE{to reduce in size (body)} \PD{sellɛɛ, sellɛ, sellɛ, sellebɛ, sellaa, - } 
\LX{sɛma} \HM{} \PH{sɛ́mà} \PS{v.~} \GE{to sue in court} \PD{sɛmaɛ, seɛrɛmara, sɛmara, sɛmareba, -, - } \VA{sɛ, sɛmare}
\LX{semare} \HM{} \PH{sɪ́màrɪ́} \PS{n.~} \GE{a type of fetish to which there is a secret cult that has the power to identify evil people like wizards and witches} \OP{semarɛɛ} 
\LX{sɛmare} \HM{} \PH{sɛ́márɪ̀} \PS{v.~} \GE{to sue} \PD{sɛmarɛɛ, seɛrɛmara, seɛremara, seɛremareba, -, - } \VA{sɛ, sɛma}
\LX{semekelee} \HM{} \PH{sèmèkèlèè} \PS{n.~} \GE{broken piece from a clay molded object e.g a pot} \VA{sembakelee}
\LX{sɛmpili} \HM{} \PH{sɛ̌mpílí} \PS{n.~} \GE{a type of long mat} \OP{sɛmpile} \TP{sɛmpilee} \VA{sɛnwoɔ}
\LX{sɛŋ} \HM{} \PH{sɛ́ŋ} \PS{adv.~} \GE{actually} \XV{Fo sɛŋ digi la a bie bare wɛ?} \XE{So, you have actually driven the child away?.} 
\LX{sɛndɔɔ} \HM{} \PH{sɛ̀ndɔ́ɔ́} \PS{n.~} \GE{the boyfriend of one's wife} \OP{sɛndɔbɔ} \TP{sɛndɔɔrɛɛ} 
\LX{sɛndɔɔ-dogi-za} \HM{} \PH{sɛ̌ndɔ́ɔ́-dògì-zá} \PS{n.~} \GE{small lumps of earth on top of a wall for decoration} 
\LX{seŋe} \HM{} \PH{sɪ́ŋ} \PS{v.~} \SN \GE{to suffice, to be enough} \XV{A bondirii e la yaga; a na seŋe la a saama.} \XE{The food is plenty; it will do for the guests.} \SN \GE{to be equal to} \XV{A toma ŋa seŋe la a anaŋ fo naŋ daare toŋe.} \XE{This work is equal to what you did.} \PD{seŋɛɛ, sennɛ, sennɛ, -, sennaa, - } 
\LX{seŋe} \HM{} \PH{sɪ̂ŋ} \PS{adv.~} \GE{like} \XV{O e seŋ ŋa o ba nyɛ ma na.} \XE{S/he behaved like s/he did not see me.} 
\LX{seŋe} \HM{} \PH{sɪ̂ŋ} \PS{v.~} \SN \GE{to wake up (someone)} \SN \GE{to revive} \PD{seŋɛɛ, sennɛ, sennɛ, senemɛ, sennaa, senemɛ } 
\LX{seŋɛ} \HM{} \PH{sɪ́ŋ} \PS{n.~} \GE{large pot used to store local beer during preparation or to store water} \OP{senne} \TP{sennɛɛ} 
\LX{sɛne} \HM{} \PH{sɛ́nɪ́} \PS{n.~} \SN \GE{lover (of another sex)} \SN \GE{fiance} \OP{sɛnemɛ} \TP{sɛnemɛɛ} 
\LX{sɛŋe} \HM{} \PH{sɛ̀ŋ} \PS{v.~} \GE{to engage in friendship between people of opposite sexes} \PD{sɛŋɛɛ, sɛnɛ, sɛnɛ, sɛnemɛ, sɛnaa, sɛnemɛ } 
\LX{sɛŋe} \HM{} \PH{sɛ̂ŋ} \PS{adj.~} \SN \GE{one's side} \XV{A sane yɔɔbo be la fo sɛŋe.} \XE{Paying the debt rests with you.} \SN \GE{towards} \XV{A sakuuri are la a teŋɛ sapare sɛŋe.} \XE{The school is situated to the east of the town.} \OP{sɛnne} 
\LX{sɛŋɛ} \HM{} \PH{sɛ̂ŋɛ́} \PS{n.~} \GE{mat; bed} \OP{sɛnne} \TP{sɛnnɛɛ} 
\LX{seŋgelɛnte} \HM{} \PH{sɪ́ŋgɪ́lɛ́ntɪ́} \PS{n.~} \GE{singlet} \OP{seŋgelɛntere} \TP{seŋgelɛnterɛɛ} 
\LX{sengye} \HM{} \PH{sándzé} \PS{n.~} \GE{mosquito net} \OP{sengyeri} \TP{sengyeree} \VA{sangye}
\LX{seŋkaafa} \HM{} \PH{sɪ̀ŋkááfá} \PS{n.~} \GE{rice (Hausa)} \OP{seŋka᷉afare} \TP{sɪŋka᷈a᷈farɛɛ} \VA{mui}
\LX{seŋka᷉awerɛ} \HM{} \PH{sɪ́ŋká᷉á᷉wɪ́rɛ́} \PS{n.~} \GE{groundnut (a variety), has small nuts} 
\LX{seŋka᷈a᷈} \HM{} \PH{sɪ̀ŋkáá} \PS{n.~} \GE{groundnut, peanut} \SG{seŋkaanoo} \TP{seŋkaanɛɛ} 
\LX{seŋka᷉a᷉kpɔlɔ} \HM{} \PH{sɪ́ŋká᷉á᷉kpɔ́lɔ́} \PS{n.~} \GE{groundnuts (a variety), has big nuts} 
\LX{sɛŋkpɛdie} \HM{} \PH{sɛ̀ŋkpɛ̀dìé} \PS{n.~} \GE{deep love between lovers; relationship that involves sex} \OP{sɛŋkpɛderi} \TP{sɛŋkpɛderee} 
\LX{seŋkpegere} \HM{} \PH{sɪ́ŋkpɪ́gɪ́rɪ́} \PS{n.~} \GE{hip} \OP{seŋkpegɛ} \TP{seŋkpegerɛɛ} 
\LX{sɛŋkpɔgele} \HM{} \PH{sɛ̂ŋkpɔ̀glɪ̀} \PS{n.~} \GE{bedbug} \OP{sɛŋkpɔgelɔ} \TP{sɛŋkpɔgelɛɛ} \VA{dakpalori, dakpaloree}
\LX{sɛnlee} \HM{} \PH{sɛǹléè} \PS{n.~} \GE{small mat} \OP{sɛmbilii} 
\LX{senlɛre} \HM{} \PH{sɪ́nlɛ́rɪ́} \PS{n.~} \GE{sweet drink} \OP{senlerre} \TP{senlɛrɛɛ} 
\LX{sɛnlɛre} \HM{} \PH{sɛ̀nlɛ̀rɪ̀} \PS{n.~} \GE{a type mat made by putting straw together on which harvested guinea corn is kept to dry} \OP{sɛnlɛrre} \TP{sɛnlɛrɛɛ} \VA{sanlɛre}
\LX{senlogere} \HM{} \PH{sìnlʊ̀gɪ̀rì} \PS{n.~} \GE{bile} \OP{senlogera} 
\LX{sensɛlaa} \HM{} \PH{sɪ́nsɛ̀láá} \PS{n.~} \GE{earthenware for frying cakes} \OP{sensɛlaare} \TP{sensɛlaarɛɛ} 
\LX{sensɛlaabogiri} \HM{} \PH{sɪ́nsɛ̀lààbògrí} \PS{n.~} \GE{frying earthenware with hollow depressions} \TP{sensɛlaabogiree} 
\LX{senselloŋ} \HM{} \PH{sɪ́nsɪ́llʊ́ŋ} \PS{n.~} \GE{story, tale} \OP{sensellɛ} 
\LX{sɛnsɛmbie} \HM{} \PH{sɛ̀nsɛ̀mbíé} \PS{n.~} \GE{bastard} \OP{sɛnsɛmbiiri} \TP{sɛnsɛmbiiree} 
\LX{sɛnsɛŋ} \HM{} \PH{sɛnsɛŋ} \PS{n.~} \GE{the act of love making (friendship)} 
\LX{sensɛre} \HM{} \PH{sɪ́nsɛ́rɪ́} \PS{n.~} \GE{bean cake} \OP{sensɛɛ} \TP{sensɛrɛɛ} 
\LX{sensɔge} \HM{} \PH{sɪ́nsɔ́gɪ́} \PS{n.~} \GE{rattle used by babysitters to accompany the singing of a lullaby} \OP{sɔnsɔgre} \TP{sensɔgrɛɛ} 
\LX{sensogelensogɔ} \HM{} \PH{sɪ̀nsʊ́glɪ́nsʊ̀gɔ́} \PS{n.~} \GE{middle} \OP{sensogelensogere} \TP{sensogelensogerɛɛ} \VA{sensogelensoga}
\LX{sensogɔ} \HM{} \PH{sɪ̀sʊ́gɔ́} \PS{adj.~} \GE{middle} \XV{A yiri be la a teŋɛ sensogɔ.} \XE{The house is in the middle of the town.} \OP{sensogere} \TP{sensogerɛɛ} \VA{sensoglensogɔ, sensoga, soga}
\LX{sensogɔ} \HM{} \PH{sɪ́nsʊ́gɔ̀} \PS{n.~} \SN \GE{home drain e.g one leading out of the bathroom} \SN \GE{a track fixed to drain rainwater from the top of a roof} \OP{sensogere} \TP{sensogerɛɛ} \VA{sensoga}
\LX{sensɔɔre} \HM{} \PH{sɪ̀nsɔ́ɔ́rɪ́} \PS{n.~} \GE{tigernut} \SG{sensɔɔroo} \OP{sensɔbɔ} \TP{sensɔɔrɛɛ} 
\LX{sensɔraa} \HM{} \PH{sɪ̀nsɔ̀ráà} \PS{n.~} \GE{kitchen sponge} \OP{sensɔrre} \TP{sensɔrɛɛ/sensɔrrɛɛ} 
\LX{sentaa} \HM{} \PH{sɪ́ntáá} \PS{n.~} \GE{equality} 
\LX{sentaaloŋ} \HM{} \PH{sɪ́ntáá} \PS{n.~} \GE{the state of being equal} \GE{the state of being equal} \VA{sentaa}\VA{sentaa}\VA{sentaaloŋ}
\LX{senuulee} \HM{} \PH{sɪ́núúléé} \PS{n.~} \GE{honey bird} \OP{senuuli} 
\LX{sɛnwoɔ} \HM{} \PH{sɛ́nwʊ́ɔ́} \PS{n.~} \GE{a type of long mat} \OP{sɛnwoore} \TP{sɛnwoorɛɛ} \VA{sɛmpili}
\LX{senyoɔle} \HM{} \PH{sɪ́nyʊ́ɔ́lɪ́} \PS{n.~} \GE{a person or any creature that is fond of sweet things} \OP{senyoɔlɔ/senyoɔla} \TP{senyoɔlɛɛ} 
\LX{Sepaanase} \HM{} \PH{sɪ́páánásɪ̀} \PS{n.~} \GE{Spanish} \OP{Sepaanasedeme} 
\LX{sepelaa} \HM{} \PH{sɪ́pɪ́làá} \PS{n.~} \GE{a soul that has left a person but can be returned again} \OP{sepeɛle} 
\LX{sepɔpɔrɛɛ} \HM{} \PH{sɪ́pɔ́pɔ́rɛ́ɛ́} \PS{n.~} \GE{honeycomb where the larvae are found} \SG{sepɔpɔroo} \VA{pɔpɔrɛɛ}
\LX{sere} \HM{} \PH{sɪ́rɪ̀} \PS{v.~} \GE{to dry up, to evaporate} \PD{serɛɛ, serɛ, serɛ, serebɛ, seraa, - } 
\LX{sere} \HM{} \PH{sɪ́rɪ̀} \PS{v.~} \GE{to wash, to scrub (bathing)} \PD{serɛɛ, serɛ, serɛ, serebɛ, seraa, - } 
\LX{serɛ} \HM{} \PH{sɪ́rɛ́} \PS{n.~} \GE{husband} \OP{serebɛɛ} \TP{serrɛɛ/serɛɛ} 
\LX{serɛ} \HM{} \PH{sɪ̀rɛ̀} \PS{n.~} \GE{truth} \OP{serre} \TP{serɛɛ} 
\LX{sɛre} \HM{} \PH{sɛ́rɪ́} \PS{adv.~} \GE{before} \XV{Sɛre ka a karembiiri la gaa sakuuri see ka ba dɔgereba yɔɔ a sakuuri boma na zaa ba naŋ sãã.} \XE{Before the students will be re-admitted to the school their parents must pay for all the school item that were destroyed.} 
\LX{sɛre} \HM{} \PH{sɛ́rɪ̀} \PS{v.~} \GE{move, shift} \XV{Yɛ naŋ sɛre gaa ka N meŋ zeŋ a dakogi zu.} \XE{Could you shift so that I could also sit on the bench.} \PD{sɛrɛɛ, sɛrɛ, sɛrɛ, sɛrebɛ, sɛraa, - } 
\LX{sɛredire} \HM{} \PH{sɛ̀rɪ́díré} \PS{n.~} \GE{judge} \OP{sɛredirbe} \TP{sɛredirbee} 
\LX{sɛrɛɛ} \HM{} \PH{sɛ̀rɛ́ɛ̀} \PS{n.~} \GE{judgement} \OP{sɛrre} \TP{sɛrrɛɛ} 
\LX{serekuli} \HM{} \PH{sɪ̀rɪ̀kúlí} \PS{n.~} \GE{the act of marrying (on the part of woman)} 
\LX{serekulo} \HM{} \PH{sɪ̀rɪ̀kúló} \PS{n.~} \GE{bride} \OP{serkulbo} 
\LX{sɛremaalogi} \HM{} \PH{sɛ́rɪ́máálógí} \PS{n.~} \GE{pepper farm} \OP{sɛremaalogiri} \TP{sɛremaalogiree} 
\LX{sɛremaane} \HM{} \PH{sɛ́rɪ́máánɪ̀} \PS{n.~} \GE{pepper} \SG{sɛremaanoo} \OP{sɛremaama} \TP{sɛremaanɛɛ} 
\LX{seretara} \HM{} \PH{sɪ̀rɪ̀tárá} \PS{n.~} \GE{married woman} 
\LX{serete} \HM{} \PH{sɪ́rɪ́tɪ́} \PS{adv.~} \GE{very quiet} \XV{A bibiiri gbiree la ka a zie e serete ŋaa.} \XE{The children are asleep and the place is very quiet.} 
\LX{sesɔgelaa} \HM{} \PH{sɪ́sɔ́gláà} \PS{n.~} \GE{a soul that has left a person for good} \OP{sesɔgelɔ.} 
\LX{setaana} \HM{} \PH{sɪ̀táánà} \PS{n.~} \GE{devil} \OP{setaamine} \TP{setaanɛɛ} \VA{sitaana}
\LX{setampe} \HM{} \PH{sɪ̀támpɪ́} \PS{n.~} \GE{stamp} \OP{setampere} \TP{setamperɛɛ} 
\LX{setetebaa} \HM{} \PH{sɪ́tɪ́tɪ́báá} \PS{n.~} \GE{the stinger of a bee} \SG{setetebaarɛɛ} \TP{setetebaare} 
\LX{se᷉ɛ} \HM{} \PH{sɪ́᷉ɛ̀᷈} \PS{n.~} \GE{rainy season} \OP{seɛmɛ} \TP{seɛnɛɛ} 
\LX{se᷉ɛkpɛpaalaa} \HM{} \PH{sɪ̀᷉ɛ̀᷉kpɛ̀pááláá} \PS{n.~} \GE{early farming season} \OP{se᷉ɛkpɛpaala} 
\LX{sieree} \HM{} \PH{sìèrèè} \PS{n.~} \GE{benefit} 
\LX{sigi} \HM{} \PH{sígí} \PS{v.~} \SN \GE{to go down, to descend} \SN \GE{to land, to alight} \SN \GE{to decrease (intransitive)} \PD{sigee, sigre, sigre, sigribe, sigraa, sigribe } \SE{sigi kɔmpare} \GE{to sink} \SE{sigi puli} \GE{to be low} 
\LX{sigili} \HM{} \PH{sìgilì} \PS{v.~} \GE{to become calm, to become quiet} \XV{Zenɛ, saseɛ ba da lɛnso ka a zie sigili.} \XE{Today there is no wind and that is why the place is calm.} \PD{siglee, sigle, sigle, siglibe, siglaa, - } 
\LX{sigili} \HM{} \PH{sìgilì} \PS{v.~} \GE{to offer a gift or praise in honour of something well done} \XV{A pɔgelee seɛ la a seɛre ne toori ka a naa sigili o ne libie.} \XE{The young lady danced so well that the chief honoured her by offering her money.} \PD{siglee, sigle, sigle, siglibe, siglaa, - } 
\LX{sigili} \HM{} \PH{sìgilì} \PS{v.~} \GE{to get choked (on food)} \XV{A noɔ dire la a zɔŋ gborɔgborɔ ka a te sigili o.} \XE{The fowl was eating the feed very fast and choked on it.} \PD{siglee, sigle, sigle, siglibe, siglaa, - } 
\LX{sigili} \HM{} \PH{sìgilì} \PS{v.~} \GE{to gang up on (two or more people attacking one)} \XV{Bayuo ne o yɔɔ sigili la a dɔɔ kyɛ ba to᷉ɔ᷉ o.} \XE{Bayuo and his brother fought the man simutaneously but they could not beat him.} \PD{siglee, sigle, sigle, siglibe, siglaa, - } 
\LX{sigiri} \HM{} \PH{sígírì} \PS{v.~} \SN \GE{to lower, to reduce} \XV{Ka banaŋ ba sigiri a nɔɔteɛ daaroŋ N koŋ to᷉ɔ᷉ a da.} \XE{If they don't lower the price of the shoes, I can't buy them.} \SN \GE{to humble} \XV{A seŋ ka karembiiri  maŋ sigiri ba menne ka karemamine to᷉ɔ᷉ kanne ba.} \XE{Pupils should learn to humble themselves so that teachers can teach them.} \PD{sigiree, sigire, sigire, sigiribe, sigiraa, sigiribe } 
\LX{sikiri} \HM{} \PH{síkìrí} \PS{n.~} \GE{sugar} \TP{sikiree} \VA{sukyiri}
\LX{sile} \HM{} \PH{sílè} \PS{n.~} \GE{shilling} \OP{silli} \TP{sillee} 
\LX{simmie} \HM{} \PH{símmíé} \PS{n.~} \SN \GE{bambara beans} \SN \GE{groundnuts (Northern Dagaare dialect)} \SG{simmiruu} \TP{simmiiree} \VA{siŋgbule}
\LX{simmiri} \HM{} \PH{símmírì} \PS{n.~} \GE{thread} \SG{simmiruu} \OP{simmie} \TP{simmiiree} 
\LX{siŋgbule} \HM{} \PH{sìŋgbúlé} \PS{n.~} \GE{bambara beans (a type of round beans that bear nuts in the ground like groundnuts) (Northern Dagaare dialect)} \VA{simmie}
\LX{sinkyigiri} \HM{} \PH{sínʧígirí} \PS{n.~} \GE{bunch, head (of guinea corn)} \OP{sinkyige} \TP{sinkyigree} 
\LX{sinsuoluŋ} \HM{} \PH{sínsúólúŋ} \PS{n.~} \GE{ground bean} \SG{sinsuoluu} \TP{sinsuolee} 
\LX{sipirii} \HM{} \PH{sìpìríì} \PS{n.~} \GE{aspirin} \SG{sipiruu} \TP{sipirree} 
\LX{siremm} \HM{} \PH{sírémm} \PS{adv.~} \GE{serene} 
\LX{sisi} \HM{} \PH{sìsí} \PS{v.~} \GE{to cheat} \PD{sisiee, sisiire, sisiire, sisiiribe, sisiiraa, - } 
\LX{sitime} \HM{} \PH{sìtìmè} \PS{n.~} \GE{ship, steamer} \OP{sitimeri} \TP{sitimeree} 
\LX{so} \HM{} \PH{sò} \PS{v.~} \GE{to bathe} \PD{soe, suoro, suoro, suoribo, suoraa, suori } 
\LX{so} \HM{} \PH{sʊ́} \PS{v.~} \GE{to own} \PD{soɛ, soora, soora, sooreba, sooraa, - } 
\LX{so} \HM{} \PH{sʊ́} \PS{conj.~} \GE{as a result} 
\LX{sɔ} \HM{} \PH{sɔ́} \PS{v.~} \GE{to smear, to apply} \PD{sɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } 
\LX{sɔ} \HM{} \PH{sɔ̀} \PS{v.~} \GE{to add vegetables to groundnut sauce} \PD{sɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } 
\LX{sɔ kpaare} \HM{} \PH{sɔ́ kpáárɪ́} \PS{v.~} \GE{to forget to deliver (a message)} \PD{sɔɛ kpaare, soɔrɔ kpaare, -, -, -, - } \VA{kpaare soɔbo}
\LX{sobile} \HM{} \PH{sóbílé} \PS{n.~} \GE{path} \OP{sobilii} \VA{solee}
\LX{sobiri} \HM{} \PH{sóbírì} \PS{n.~} \SN \GE{road, way} \SN \GE{journey} \OP{sobie} \TP{sobiiree} \VA{sori}
\LX{sɔbiri} \HM{} \PH{sɔ̀bírì} \PS{n.~} \GE{knife blade} \OP{sɔbie} \TP{sɔbiiree} 
\LX{sobɔ} \HM{} \PH{sʊ́bɔ́} \PS{n.~} \SN \GE{owner} \XV{N ba baŋe a gane ŋa sobɔ a na de te ko o.} \XE{I don't know the owner of this book well enough that I could take it to him/her.} \SN \GE{the one, the person} \XV{A sobɔ na naŋ la yuo a pane, onɔ la de a libie.} \XE{The person who opened the door is the one who has taken the money.} \OP{demee} \VA{soba}
\LX{so-eŋe} \HM{} \PH{sò-ìŋ} \PS{v.~} \SN \GE{to initiate} \PD{so-eŋɛɛ, suoro-ennɛ, so-ennɛ, so-enemɛ, so-ennaa, - } 
\LX{so-eŋe} \HM{} \PH{sò-ìŋ} \PS{n.~} \GE{metaphor (literature)} 
\LX{sofoɔlaa} \HM{} \PH{sófʊ̀ɔ́láà} \PS{n.~} \GE{street} \OP{sofoɔlɪ} 
\LX{sɔgbale} \HM{} \PH{sɔ̀gbàlɪ́} \PS{n.~} \GE{blunt knife} \OP{sɔgbala} \TP{sɔgbalɛɛ} 
\LX{sɔge} \HM{} \PH{sɔ́gɪ́} \PS{v.~} \GE{to shell nuts} \PD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } 
\LX{sɔge} \HM{} \PH{sɔ́gɪ̀} \PS{v.~} \GE{to rinse} \PD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } \SE{sɔge noɔre} \GE{to eat breakfast} 
\LX{sɔge} \HM{} \PH{sɔ̀gɪ̀} \PS{adj.~} \GE{bushy, littered} \XV{A dendɔre sɔgɛɛ la, a seŋ ka ba peere o.} \XE{The couryard is litered with rubbish, so it needs to be swept.} 
\LX{sɔge} \HM{1} \PH{sɔ̀gɪ̀} \PS{v.~} \GE{to respond, to reply (in a conversation)} \PD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } 
\LX{sɔge} \HM{2} \PH{sɔ̀gɪ̀} \PS{v.~} \GE{to be bushy} \PD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } 
\LX{sɔgelaa} \HM{} \PH{sɔ́glàá} \PS{adj.~} \SN \GE{black} \SN \GE{dark (colour)} \OP{sɔgelɔ} 
\LX{sɔgelaa} \HM{} \PH{sɔ̀glàà} \PS{n.~} \GE{neck of a pot} \OP{sɔgele} \TP{sɔgelɛɛ} 
\LX{sɔgele} \HM{} \PH{sɔ́glɪ̀} \PS{v.~} \GE{to hide} \PD{sɔgelɛɛ, sɔgelɔ, sɔgelɔ, sɔgeleba, sɔgelaa, sɔgeleba } 
\LX{sogelɔŋ} \HM{} \PH{sʊ̀glɔ̀ŋ} \PS{adv.~} \GE{heavy and dull} 
\LX{sogi} \HM{} \PH{sògì} \PS{v.~} \GE{to catch in midair} \PD{sogee, sogro, sogro, sogribo, sograa, sogribo } 
\LX{sogi} \HM{} \PH{sògì} \PS{v.~} \GE{to pant, to breathe with difficulty} \PD{sogee, sogro, sogro, sogribo, sograa, sogribo } 
\LX{sogi} \HM{} \PH{sògì} \PS{v.~} \GE{to chirp (guinea fowl)} \PD{sogee, sogro, sogro, sogribo, sograa, sogribo } 
\LX{sogɔ} \HM{} \PH{sʊ́gɔ́} \PS{adj.~} \GE{middle} \OP{sogere} \TP{sogerɛɛ} \VA{soga, shortened form of sensogelensogɔ}
\LX{sogya} \HM{} \PH{sóɲà} \PS{n.~} \GE{soldier} \OP{sogyare} \TP{sogyarɛɛ} \VA{sogyɛ}
\LX{sokonsuo} \HM{} \PH{sókónsúó} \PS{n.~} \GE{storey} \OP{sokonsori} \TP{sokonsoree} 
\LX{sokoɔraa} \HM{} \PH{sókʊ́ɔ́ràá} \PS{n.~} \GE{large road} \OP{sokoɔre} \TP{sokoɔrɛɛ} 
\LX{sɔkpoŋi} \HM{} \PH{sɔ̀kpóŋ} \PS{n.~} \GE{cutlass, large knife} \OP{sɔkponni} \TP{sɔkponnee} 
\LX{sokyara} \HM{} \PH{sóʧárá} \PS{n.~} \GE{crossroads, intersection} \OP{sokyarre} \TP{sokyarɛɛ} 
\LX{sɔkyeɛraa} \HM{} \PH{sɔ̀ʧɪ̀ɛ̀ráà} \PS{n.~} \GE{cutlass} \OP{ sɔkyeɛrre} 
\LX{sokyɛnɛ} \HM{} \PH{sóʧɛ́nɛ̀} \PS{n.~} \GE{traveler} \OP{sokyɛnemɛ} \TP{sokyɛnemɛɛ} 
\LX{sɔlare} \HM{} \PH{sɔ̀lárɪ̀} \PS{n.~} \GE{cutlass} \OP{sɔlaɛ} \TP{sɔlarɛɛ} 
\LX{solee} \HM{} \PH{sóléé} \PS{n.~} \GE{path} \VA{sobile}
\LX{sɔlee} \HM{} \PH{sɔ̀léé} \PS{n.~} \GE{small knife} \OP{sɔbilii} 
\LX{solgbolo} \HM{} \PH{sʊ̀lgbʊ̀lʊ̀} \PS{adj.~} \GE{inactive} 
\LX{soli} \HM{} \PH{sólí} \PS{adj.~} \GE{in a neat manner} \XV{a pɔgelee vare gaŋe la a miri lɛ soli.} \XE{The girlcleared the high jump bar neat.} 
\LX{solli} \HM{} \PH{sòllì} \PS{v.~} \GE{to be well polished} \PD{sollee, sollo, sollo, solibo, sollaa, - } 
\LX{solli} \HM{} \PH{sóllí} \PS{v.~} \GE{to narrate/repeat incessantly} \PD{sollee, sollo, sollo, solibo, sollaa, - } 
\LX{solom} \HM{} \PH{sólóm} \PS{adv.~} \GE{completely e.g without being touched} \XV{a nannyige dgereba iri la ba nuuri solom kyɛ bare o ka ba gaa te pɔge.} \XE{The parents of the thief withdrew their support for him completely and he was sent to jail.} 
\LX{soma} \HM{} \PH{sʊ́má} \PS{n.~} \GE{a person who is good to one} 
\LX{somasoma} \HM{} \PH{sʊ́másʊ́má} \PS{adv.~} \GE{quietly} 
\LX{somɔ} \HM{} \PH{sʊ́mɔ́} \PS{adj.~} \GE{good} \GE{beautiful, attractive} \GE{nice} \XV{A pɔge na yɛlɛ somɔ la o ba baŋe yelfaare yeluu.} \XE{That woman is nice, she never uses unsavory language.} \OP{sonne} \TP{sonnɛɛ} \VA{soma}
\LX{somɔ} \HM{} \PH{sʊ́mɔ́} \PS{adj.~} \GE{holy} 
\LX{sompoɔ} \HM{} \PH{sʊ́mpʊ́ɔ́} \PS{n.~} \GE{three pence} \OP{sompoore} \TP{sompoorɛɛ} 
\LX{soŋ} \HM{} \PH{sʊ̌ŋ} \PS{adj.~} \GE{good, fine} \OP{sonne} \TP{sonnɛɛ} 
\LX{soŋe} \HM{} \PH{sʊ̀ŋ} \PS{v.~} \GE{to help} \XV{Dagaaba nɔŋe ka ba maŋ soŋe taa ba tomɔ poɔ.} \XE{The Dagaaba like to help each other in their work.} \PD{soŋɛɛ, sonna, sonna, sonema, sonnaa, sonema } 
\LX{sɔŋɔ} \HM{} \PH{sɔ́ŋ} \PS{n.~} \GE{wood carrier} \OP{sɔnne} \TP{sɔnnɛɛ} \VA{sɔŋa}
\LX{sonono} \HM{} \PH{sʊ́nʊ́nʊ́} \PS{adv.~} \GE{extremely cold} 
\LX{soo} \HM{} \PH{sóó} \PS{adv.~} \GE{pure, unmixed} 
\LX{soɔ} \HM{} \PH{sʊ́ɔ́} \PS{n.~} \GE{knife} \OP{sɔre} \TP{sɔrɛɛ} \VA{sɔlee,sɔbilii}
\LX{soɔ} \HM{} \PH{sʊ́ɔ̀} \PS{v.~} \GE{to blacken, to darken} \PD{soɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } \VA{sɔɔ}
\LX{soɔ} \HM{} \PH{sʊ́ɔ̀} \PS{v.~} \GE{to blossom e.g. vegetation} \PD{soɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } \VA{sɔɔ}
\LX{sɔɔ} \HM{} \PH{sɔ̀ɔ̀} \PS{adv.~} \GE{in a little while} \XV{A na erɛ sɔɔ ka sakuuri iree.} \XE{In a little while schools will close.} 
\LX{sɔɔ} \HM{} \PH{sɔ́ɔ̀} \PS{v.~} \GE{to blacken, to darken} \PD{-, -, -, -, -, - } \VA{soɔ}
\LX{sɔɔ} \HM{} \PH{sɔ́ɔ̀} \PS{v.~} \GE{to blossom e.g. vegetation} \PD{-, -, -, -, -, - } \VA{soɔ}
\LX{soɔbɔ} \HM{} \PH{sʊ́ɔ́bɔ́} \PS{n.~} \GE{wizard, witch} \OP{soɔre} \TP{sɔrɛɛ} \VA{soɔba}
\LX{soɔloŋ} \HM{} \PH{sʊ̀ɔ̀lʊ́ŋ} \PS{n.~} \GE{witchcraft} 
\LX{soɔŋaa} \HM{} \PH{sʊ́ɔ́ŋàà} \PS{n.~} \GE{rabbit, hare} \OP{soɔne} \TP{soɔnɛɛ} 
\LX{soɔne} \HM{} \PH{sʊ̀ɔ̀nɪ̀} \PS{v.~} \GE{to praise, to glorify} \PD{soɔnɛɛ, soɔnɔ, soɔnɔ, soɔnema, soɔnaa, - } \VA{soane}
\LX{soore} \HM{} \PH{sʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{to question, to inquire} \PD{soorɛɛ, soorɔ, soorɔ, sooreba, sooraa, - } 
\LX{soore} \HM{} \PH{sʊ́ʊ́rɪ́} \PS{v.~} \SN \GE{to hiss to drive fowls away} \SN \GE{to hiss in pain e.g. after eating very spicy food} \PD{-, -, -, -, -, - } 
\LX{soɔre} \HM{} \PH{sʊ́ɔ́rɪ̀} \PS{n.~} \GE{dye} \TP{soɔrɛɛ} \VA{sɔɔre}
\LX{sɔɔre} \HM{1} \PH{sɔ́ɔ́rɪ́} \PS{n.~} \GE{dye} \TP{sɔɔrɛɛ} \VA{soɔre}
\LX{sɔɔre} \HM{2} \PH{sɔ́ɔ́rɪ́} \PS{n.~} \SN \GE{liver} \SN \GE{a type of dark brown bush mouse that lives along river banks} \OP{sɔbɔ} \TP{sɔɔrɛɛ} \VA{shortened form of ɔŋ-sɔɔre}
\LX{soori} \HM{} \PH{sòòrí} \PS{n.~} \GE{locusts} \OP{soori} \TP{sooree} 
\LX{soɔsobɔ} \HM{} \PH{sʊ́ɔ́sʊ́bɔ́} \PS{n.~} \GE{chief celebrant (traditional worship)} \OP{sɔdemee} \VA{soɔsoba}
\LX{sɔre} \HM{} \PH{sɔ̀rɪ̀} \PS{v.~} \SN \GE{to beg} \SN \GE{to ask, to request} \SN \GE{to plead, to implore} \PD{sɔrɛɛ, sɔrɔ, sɔrɔ, sɔrebɔ, sɔraa, sɔrebɔ } 
\LX{sɔre} \HM{} \PH{sɔ̀rɪ̀} \PS{v.~} \GE{to cook soup made with groundnuts and vegetables} \PD{sɔrɛɛ, sɔrɔ, sɔrɔ, sɔrebɔ, sɔraa, sɔrebɔ } 
\LX{sɔre} \HM{} \PH{sɔ́rɪ́} \PS{v.~} \SN \GE{to count} \SN \GE{to read} \PD{-, -, -, -, -, - } 
\LX{sɔrebiri} \HM{} \PH{sɔ́rɪ́bírì} \PS{n.~} \GE{counter, pebble} \SG{sɔrebiruu} \OP{sɔrebie} \TP{sɔrebiiree} 
\LX{sɔresɔreŋmanlee} \HM{} \PH{sɔ̀rɪ̀sɔ̀rɪ̀ŋmànléé} \PS{n.~} \GE{a beggar's calabash} \OP{sɔresɔreŋmamblii} \TP{sɔresɔreŋmambilee} 
\LX{sɔresɔrɔ} \HM{} \PH{sɔ̀rsɔ́rɔ́} \PS{n.~} \GE{beggar} \OP{sɔrsɔreba} \TP{sɔrsɔrebɛɛ} 
\LX{sɔresɔrɔ} \HM{} \PH{sɔ̀rsɔ́rɔ́} \PS{n.~} \GE{beggar} \OP{sɔresɔrebɔ} \TP{sɔresɔrebɛɛ} \VA{sɔresɔra}
\LX{sori} \HM{} \PH{sórì} \PS{n.~} \GE{way, road} \OP{soe} \TP{soree} 
\LX{sɔteɛ} \HM{} \PH{sɔ̀tɪ́ɛ́} \PS{n.~} \GE{scabbard for a knife} \OP{sɔteere} \TP{sɔteerɛɛ} 
\LX{sotugane} \HM{} \PH{sótùgánɪ́} \PS{n.~} \GE{map} \OP{sotugama} \TP{sotugannɛɛ} 
\LX{sotuuraa} \HM{} \PH{sótúúráà} \PS{n.~} \GE{pathway} \OP{sotuuri} 
\LX{sɔvu᷉u} \HM{} \PH{sɔ̀vù᷉ú᷉} \PS{n.~} \GE{light associated with witchcraft} \SG{sɔvuunuu} \TP{sɔvuunee} 
\LX{so᷉o} \HM{} \PH{sʊ́ʊ́} \PS{adv.~} \GE{confused} 
\LX{so᷉ɔ} \HM{} \PH{sʊ̀ɔ̀} \PS{v.~} \GE{to flatter} \PD{so᷉ɔɛ, soɔnɔ, soɔnɔ, soɔnema, soɔnaa, - } 
\LX{so᷉ɔ} \HM{} \PH{sʊ̀᷉ɔ́᷉} \PS{n.~} \GE{witchcraft, supernatural power} \XV{N teɛre ka a bie ŋa taa la so᷉ɔ᷉; wolɔ ka o e do a teɛ ŋa?} \XE{I think this boy has supernatural power; how did he manage to climb this tree?.} 
\LX{so᷉ɔ} \HM{} \PH{sʊ̀᷉ɔ̀᷉} \PS{v.~} \GE{to draw a line on the ground with the foot} \PD{soɔɛ, soɔnɔ, soɔnɔ, soɔnema, soɔnaa, - } 
\LX{so᷉ɔgo᷉ɔo} \HM{} \PH{sʊ́᷉ɔ́᷉gʊ́᷉ɔ́᷉ʊ́} \PS{n.~} \GE{pretence} 
\LX{so᷉ɔzeɛnagbare} \HM{} \PH{sʊ́᷉ɔ́zɪ́ɛ́nàgbárɪ́} \PS{n.~} \GE{small but old or strong} \OP{so᷉ɔzeɛnagbarɛɛ} 
\LX{su} \HM{} \PH{sù} \PS{v.~} \GE{to put on a top (shirt)} \PD{sue, suuro, -, -, -, - } \SE{su bonsuuri} \GE{to wear clothes} \SE{su kaayaa} \GE{to wear clothes} 
\LX{su} \HM{} \PH{sù} \PS{v.~} \GE{to put inside (a hole)} \PD{sue, suuro, -, -, -, - } \SE{su bonsuuri} \GE{to wear clothes} \SE{su kaayaa} \GE{to wear clothes} 
\LX{su} \HM{} \PH{sù} \PS{v.~} \GE{to spoon feed e.g. baby, patient} \PD{sue, suuro, -, -, -, - } \SE{su bonsuuri} \GE{to wear clothes} \SE{su kaayaa} \GE{to wear clothes} 
\LX{subaaloŋ} \HM{} \PH{sùbààlʊ́ŋ} \PS{n.~} \GE{tenderheartedness} 
\LX{sudire} \HM{} \PH{sùdíré} \PS{n.~} \GE{one who forgives, one who pardons} \OP{ surdiribe} \TP{sudiribee} 
\LX{sugaraate} \HM{} \PH{sùgàráátɪ́} \PS{n.~} \GE{cigarette} \OP{sugaraatere} \TP{sugaraaterɛɛ} \VA{segarate}
\LX{sugi} \HM{} \PH{sùgì} \PS{v.~} \SN \GE{to dislocate e.g. joint} \SN \GE{to dislodge e.g. a pot from the top of another} \PD{sugee, sugro, sugro, sugribo, sugraa, - } 
\LX{sugi} \HM{} \PH{sùgì} \PS{v.~} \GE{to ladel soup from a pot} \PD{sugee, sugro, sugro, sugribo, sugraa, - } 
\LX{sugili} \HM{} \PH{sùgìlì} \PS{v.~} \SN \GE{to come next to e.g. in a line} \XV{Ba naŋ tɔnɔ a lae maa la sugili o zu.} \XE{When they were queueing, I came next to him.} \SN \GE{to place one on top of the other e.g. pots} \XV{Ba pɛge la a laare wuo sugili taa.} \XE{They washed the bowls and placed them one on top of the other.} \PD{suglee, suglo, suglo, suglibo, suglaa, - } 
\LX{Sugilo} \HM{} \PH{súgíló} \PS{n.~} \GE{name traditionally given to a child immediately after twins} 
\LX{sugo} \HM{} \PH{súgò} \PS{n.~} \GE{herd; group of many e.g. troop of animals} \XV{ŋmaa-sugo kpɛ la n weɛ sa᷉a a kamaana zaa.} \XE{A troop of monkeys invaded my farm and destoyed all the maize.} \OP{sugiri} \TP{sugiree} 
\LX{sugo} \HM{} \PH{súgò} \PS{n.~} \GE{a type of tiny white ant that destroys yams} \XV{Sugo saana waare.} \XE{White ants destroys yams.} \TP{sugee} \VA{wasugo}
\LX{sukaale} \HM{} \PH{sùkààlɪ̀} \PS{n.~} \GE{a type of large dress with wide sleeves that are thrown over the shoulder} \TP{sukaalɛɛ} 
\LX{sukyiri} \HM{} \PH{sùʧírí} \PS{n.~} \SN \GE{heart} \XV{N sukyiri maŋ zaara la, a seŋ ka N gaa dɔɔta.} \XE{I keep experiencing heart palpitations; I need to go to the hospital.} \SN \GE{courage} \XV{Ka fooŋ ba taa sukyiri fo koŋ baŋ toŋ asibiti toma.} \XE{If you have no courage, you can't work in the hospital.} \OP{sukyie} \TP{sukyiree} \VA{sikyiri}
\LX{sukyiribaaloŋ} \HM{} \PH{sùʧìrìbààlʊ̀ŋ} \PS{n.~} \SN \GE{kind heart} \SN \GE{heart disease} 
\LX{sukyirifo᷉ɔ᷉o᷉} \HM{} \PH{súʧírífʊ́᷉ɔ́᷉ʊ́᷉} \PS{n.~} \GE{lightheartedness} \OP{sukyirifoɔne} \TP{sukyirifoɔnɛɛ} 
\LX{sukyiriliŋe} \HM{} \PH{sùʧìrìlíŋé} \PS{n.~} \GE{end of the ribcage} \OP{sukyirilinni} \TP{sukyirilinnee} 
\LX{sukyiritage} \HM{} \PH{súʧírítágɪ́} \PS{n.~} \GE{bravery} \OP{sukyiritagre} 
\LX{sulee} \HM{} \PH{sùlèè} \PS{n.~} \GE{chimney} \OP{sulli} \TP{sullee} 
\LX{suli} \HM{} \PH{sùlì} \PS{v.~} \GE{to hang one's head, to look down} \PD{sulee, sullo, sullo, sulbo, sullaa, - } 
\LX{sumaaroŋ} \HM{} \PH{sùmáárʊ́ŋ} \PS{n.~} \GE{coolheaded} \XV{A pɔge taa la sumaaroŋ ka lɛ naane o da na zɔɔ ne la a dɔɔ.} \XE{The woman is coolheaded otherwise she would have fought the man.} 
\LX{sumiime} \HM{} \PH{súmíímé} \PS{n.~} \GE{heartburn} 
\LX{sumosumo} \HM{} \PH{sùmòsúmó} \PS{n.~} \GE{a type of fried food made from the flour of yam} 
\LX{sundaa} \HM{} \PH{súndàá} \PS{n.~} \GE{a stick to which birdlime is applied to trap birds} \OP{sundaare} \TP{sundarɛɛ} 
\LX{sunni} \HM{} \PH{súnnì} \PS{n.~} \SN \GE{glue, gum, sticky substance} \SN \GE{a plant that produces gum} \SN \GE{birdlime} \SG{sunnuu} \OP{sumo} \TP{sunnee} \VA{sun (saanchi)}
\LX{sunni} \HM{} \PH{súnní} \PS{n.~} \GE{blue powder used in dyeing clothes during washing} 
\LX{sunsugiri} \HM{} \PH{súnsúgrí} \PS{n.~} \GE{white growth within a sore} \OP{sunsugiri} \TP{sunsugiree} 
\LX{suori} \HM{} \PH{súórì} \PS{v.~} \GE{to fall off in great numbers (leaves)} \PD{suoree, suoro, suoro, suoribo, suoraa, - } 
\LX{supaana} \HM{} \PH{sùpáánà} \PS{n.~} \GE{spanner} \SG{supaanoo} \OP{supaanare} \TP{supaanarɛɛ} 
\LX{suree} \HM{} \PH{sùréé} \PS{n.~} \GE{ladle} \OP{surri} 
\LX{suromm} \HM{} \PH{súrómm} \PS{adv.~} \GE{quiet} 
\LX{susu} \HM{} \PH{súsú} \PS{n.~} \GE{a cycle in which contributions of members are given to others} \OP{susuri} \TP{susuree} 
\LX{susu} \HM{} \PH{sùsú} \PS{v.~} \SN \GE{to measure} \SN \GE{to be careful, to not overdo something} \SN \GE{to think about carefully} \PD{susue, susuuro, susuuro, susuribo, susuuraa, - } 
\LX{susu} \HM{} \PH{sùsù} \PS{adj.~} \GE{fast and without stopping e.g. eating} 
\LX{susu} \HM{} \PH{sùsù} \PS{interjection.~} \GE{utterance to urge a dog on to pursue (animal, thief)} 
\LX{su-su-su} \HM{} \PH{sù-sù-sù} \PS{interjection.~} \GE{an utterance to urge a dog on to catch game} 
\LX{sutoloŋ} \HM{} \PH{sʊ̀tʊ́lʊ́ŋ} \PS{n.~} \GE{the state of being quick tempered} \XV{Daŋ ta wa deɛnɛ ne a dɔɔ na, o taa la sutoloŋ.} \XE{Don't you ever play with that man; he is quick tempered.} 
\LX{suu} \HM{} \PH{súú} \PS{v.~} \GE{to doze off for a short time} \PD{-, -, -, -, -, - } 
\LX{suuri} \HM{} \PH{súúrí} \PS{n.~} \GE{anger} \OP{sue} \TP{suuree} 
\LX{suuri} \HM{} \PH{súúrì} \PS{v.~} \GE{to coax, to encourage, to push to pursue e.g. dog for game} \PD{suuree, suuro, suuro, suuribo, suuraa, - } 
\LX{suuri-di-teroo} \HM{} \PH{súúr-dì-tɪ́rʊ́ʊ́} \PS{n.~} \GE{forgiveness} 
\LX{sɛnwoɔ} \HM{} \PH{sɛ́nwʋ́ɔ́} \PS{n.~} \GE{a type of long mat} \VA{sɛmpili}
\LX{ta} \HM{} \PH{tá} \PS{adj.~} \GE{not (negative marker)} \XV{A pɔge kpa᷉a la o bie ka o ta gaa a sini.} \XE{The woman cautioned her son not to go to the cinema.} 
\LX{ta} \HM{} \PH{tá} \PS{v.~} \SN \GE{to reach, to arrive} \XV{Ka fooŋ wa ta fo yiri foŋ yeli ko ma.} \XE{When you arrive at your house, tell me.} \SN \GE{to approach} \XV{Fo maŋ gaa ka fo ta o yiri ka baare  vuoli paale ziezaa.} \XE{As soon as you approach the house, dogs bark all over.} \PD{taɛ, tara, tara, tareba, taraa, tareba } 
\LX{ta kpɛ} \HM{} \PH{tá kpɛ́} \PS{v.~} \GE{to run away, to flee, to take to one's heels} \XV{A bie gɛrɛ la te nyɛ waabo, a leɛ takpɛ.} \XE{As the girl walked along, she came across a snake and took to her heels.} \PD{ta kpɛɛ, tara kpeɛrɛ, takpeɛrɛ, takpeɛrebɛ, takpeɛraa, - } 
\LX{taa} \HM{} \PH{tàá} \PS{v.~} \SN \GE{to have} \XV{A bie taa la maŋgori ayi.} \XE{The child has two mangoes.} \SN \GE{to keep, to hold onto} \XV{N ba de la o zupili ka N taa.} \XE{My father has given his hat to me to keep.} \XV{A pɔge taa la poɔ.} \XE{The woman is pregnant.} \XV{o taa la yelmeŋɛ} \XE{S/he is honest.} \XV{A bibiiri taa la gɔnne.} \XE{The children are noisy.} \PD{taaɛ, tara, tara, tareba, taara, tareba } \SE{taa kɔŋ} \GE{to be hungry} 
\LX{taa} \HM{} \PH{táá} \PS{pron.~} \GE{each other} 
\LX{taa kɔnnyuuri} \HM{} \PH{tàá kɔ̀nɲúúrì} \PS{v.~} \GE{to be thirsty} \PD{-, -, -, -, -, - } 
\LX{taa laare} \HM{} \PH{tàá láárɪ́} \PS{adj.~} \GE{amusing, funny} 
\LX{taa nimibaaloŋ} \HM{} \PH{} \PS{adj.~} \GE{kind} 
\LX{taa nimiri} \HM{} \PH{tàá nímírì} \PS{adj.~} \GE{covetous} 
\LX{taa noɔre} \HM{} \PH{tàá nʊ́ɔ́rɪ̀} \PS{adj.~} \GE{talkative, noisy} \XV{A balee taa la noɔre te zuo, o maŋ vuolo la tegitegi lɛ.} \XE{The puppy is noisy; it always barks.} 
\LX{taa nyonono} \HM{} \PH{tàá ɲʋ́nʋ́nʋ́} \PS{n.~} \SN \GE{selfishness} \SN \GE{sympathy} 
\LX{taa poɔ} \HM{} \PH{tàá pʊ́ɔ́} \PS{v.~} \GE{to be pregnant} \PD{-, -, -, -, -, - } 
\LX{taa toori} \HM{} \PH{táá tòòrí} \PS{adj.~} \GE{stubborn (lit. be hard-eared)} 
\LX{taa yelmeŋɛ} \HM{} \PH{taa yelmeŋɛ} \PS{adj.~} \GE{is honest} 
\LX{taala} \HM{} \PH{táálá} \PS{n.~} \GE{ingredients e.g. sauce} \TP{taalɛɛ} 
\LX{taama} \HM{} \PH{táámá} \PS{n.~} \GE{suffering} 
\LX{taambaa} \HM{} \PH{tàámbàá} \PS{n.~} \GE{young man} \OP{tambaare} \TP{taambaarɛɛ} 
\LX{taambeene} \HM{} \PH{tàámbɪ́ɪ́nɪ́} \PS{n.~} \GE{ripe shea tree fruit} \OP{taambeemɛ} \TP{taambeenɛɛ} 
\LX{taane} \HM{} \PH{tàànɪ̀} \PS{v.~} \GE{to stretch a great deal} \PD{taanɛɛ, taana, taana, taanema, taanaa, - } 
\LX{taaŋkaraa} \HM{} \PH{tàáŋkàráá} \PS{n.~} \GE{unripe shea fruit} \OP{taaŋkare} \TP{taaŋkarɛɛ} 
\LX{taaŋkpara} \HM{} \PH{tááŋkpárá} \PS{n.~} \GE{shea fruit with two nuts} \OP{taaŋkparre} \TP{taaŋkparɛɛ} 
\LX{taannuulee} \HM{} \PH{tàánnúúléé} \PS{n.~} \GE{canary} \OP{taannuuli} 
\LX{taantumma} \HM{} \PH{tá᷉â᷉ntú᷉mmá᷉} \PS{n.~} \GE{a type of caterpillar in hibernation} \OP{taantummamine.} 
\LX{taantunnuu} \HM{} \PH{tàântúnnì} \PS{n.~} \GE{shea tree caterpillar} \SG{taantunnuu} \OP{taantunni} \TP{taantunnee} 
\LX{taare} \HM{} \PH{tààrɪ̀} \PS{v.~} \GE{to press hard e.g. plaster, to crush e.g. an ant} \PD{taarɛɛ, taara, taara, taareba, taaraa, - } 
\LX{taare} \HM{} \PH{táárɪ̀} \PS{v.~} \GE{to tighten, to screw} \PD{taarɛɛ, taara, taara, taareba, taaraa, - } 
\LX{taayɛ} \HM{} \PH{tááyɛ̀} \PS{n.~} \SN \GE{tyre (rubber)} \SN \GE{catapult} \OP{ taayɛre} \TP{taayɛrɛɛ} 
\LX{tabole} \HM{} \PH{tábʊ̀l} \PS{n.~} \GE{table} \OP{tabolɔ} \TP{tabolɛɛ/tabollɛɛ} 
\LX{tagaremiri} \HM{} \PH{táŋgàrɪ̀mírì} \PS{n.~} \GE{telegram cable} \OP{tagaremie} \TP{tagaremiiree} 
\LX{tage} \HM{} \PH{tàgɪ̀} \PS{v.~} \SN \GE{to pull, to draw in or out} \SN \GE{to make thick (liquid)} \PD{tagɛɛ, tagra, tagra, tagreba, tagraa, tagreba } 
\LX{tage} \HM{} \PH{tágɪ́} \PS{adj.~} \GE{thick} \OP{tagre} 
\LX{tagekori} \HM{} \PH{tàgɪ̀kòrì} \PS{n.~} \GE{a type of thorn tree with sharp, curved thorns} \TP{tagekoree} 
\LX{tagekyoore} \HM{} \PH{tágɪ̀ʧʊ́ʊ́rɪ́} \PS{n.~} \GE{measles} \TP{tagekyoorɛɛ} 
\LX{tagetaa} \HM{} \PH{tàgɪ̀táá} \PS{n.~} \SN \GE{dispute} \SN \GE{conflict (literature)} 
\LX{tagetage} \HM{} \PH{tàgtàg} \PS{n.~} \GE{ability} \XV{A baala gbɛɛ kpie la, o ba taa tagetage zaa.} \XE{The patient's legs are paralysed; s/he cannot do anything.} 
\LX{tagevama} \HM{} \PH{tàgɪ̀vàmà} \PS{n.~} \SN \GE{smallpox} \SN \GE{chicken pox} \VA{teɛgara}
\LX{tahɛlɛ} \HM{} \PH{táhɛ́lɛ́} \PS{n.~} \SN \GE{extreme suffering} \XV{Kɔmbuli saŋa pɔgeba maŋ nyɛrɛ la tahɛlɛ ko᷉ɔ᷉ yɛlɛ.} \XE{During drought, women suffer severely for lack of water.} \SN \GE{carefree speech, behaviour; nonchalence} \XV{A bie ŋa eŋ ba naŋ ba feɛ o o yelee tahɛlɛ.} \XE{Since this boy was not caned he speaks with airs.} \OP{tahɛlle} \TP{tahɛlɛɛ/tahɛllɛɛ} 
\LX{takaare} \HM{} \PH{tákáárɪ̀} \PS{adj.~} \GE{vain, proud of oneself} 
\LX{takaraa} \HM{} \PH{tákáråá} \PS{n.~} \GE{small lizard} \OP{takarrre} \TP{takarrɛɛ} 
\LX{takɔdaa} \HM{} \PH{tákɔ́dàá} \PS{n.~} \GE{Sabbath} \OP{takɔdaare} \TP{takɔdarɛɛ} \VA{takɔraa}
\LX{takoŋi} \HM{} \PH{tákòŋ} \PS{n.~} \GE{ban on performance of funeral rites} 
\LX{takɔraa} \HM{} \PH{tákɔ́ráá} \PS{n.~} \GE{Sabbath} \VA{takɔdaa}
\LX{takɔraa} \HM{} \PH{tàkɔ̀ràà} \PS{n.~} \SN \GE{fin (fish)} \SN \GE{a hook or spike on an object e.g. arrow} \OP{takɔrre} \TP{takɔrrɛɛ} \VA{kɔraa}
\LX{takoriko} \HM{} \PH{tàkòríkò} \PS{n.~} \GE{bridge} \OP{takorikori} \TP{takorikoree} \VA{tokoriko}
\LX{takoroo} \HM{} \PH{tákóróó} \PS{n.~} \GE{window} \OP{takorri} \TP{takoree/takorree} \VA{tokoroo}
\LX{takpala} \HM{} \PH{tákpálá} \PS{n.~} \GE{great suffering} \TP{takpallɛɛ} 
\LX{takpala} \HM{n} \PH{tàkpálà} \PS{n.~} \GE{area between the throat and the chin} \OP{takpalle} \TP{takpallɛɛ} \VA{takpaleɛne}
\LX{takpaleɛne} \HM{} \PH{tàkpàlɪ̀ɛ̀nɪ́} \PS{n.~} \GE{area between the throat and chin} \OP{takpaleɛmɛ} \TP{takpaleɛnɛɛ/takpaleɛmɛɛ} \VA{takpalan}
\LX{takyeŋi} \HM{} \PH{táʧêŋ} \PS{n.~} \GE{the absence of argument} \OP{takyenni} 
\LX{takyeraa} \HM{} \PH{táʧɪ́ràá} \PS{n.~} \GE{gecko} \OP{takyɪrrɪ} \TP{takyɪrrɛɛ} 
\LX{takyere} \HM{} \PH{táʧɪ̀rɪ̀} \PS{n.~} \GE{temple} \OP{takyeɛ} \TP{takyerɛɛ} 
\LX{talaa} \HM{} \PH{tàláà} \PS{n.~} \GE{metal bowl or plate} \OP{talaare} \TP{talaarɛɛ} 
\LX{talane} \HM{} \PH{tàlánɪ̀} \PS{n.~} \GE{palate} \OP{talama} \TP{talannɛɛ} 
\LX{talantalaŋ} \HM{} \PH{tàlàntáláŋ} \PS{adv.~} \GE{haphazard} 
\LX{tale} \HM{} \PH{tálɪ́} \PS{n.~} \GE{lobe (of a kola nut)} \OP{tala} \TP{talɛɛ/tallɛɛ} 
\LX{tale} \HM{} \PH{tálɪ́} \PS{adj.~} \GE{flat open e.g. palm of the hand} \OP{tala} \TP{talɛɛ/tallɛɛ} 
\LX{taleri} \HM{} \PH{tálérí} \PS{n.~} \GE{small bush lizard} \OP{talerri} \TP{taleree} 
\LX{talle} \HM{} \PH{tàllɪ̀} \PS{v.~} \GE{to make long, quick strides; to walk fast} \PD{tallɛɛ, talla, talla, taleba, tallaa, - } 
\LX{taltal} \HM{} \PH{tàtál} \PS{v.~} \GE{to move with great pain} \PD{taltalɛɛ, taltala, taltala, taltaleba, taltalaa, - } 
\LX{tam} \HM{} \PH{tàm} \PS{v.~} \SN \GE{to be numb} \SN \GE{shocked; not able to react} \PD{tamɛɛ, tamena, tamena, tamenema, tamenaa, - } 
\LX{tamaa} \HM{} \PH{tàmáà} \PS{n.~} \GE{hope} 
\LX{tamanyini} \HM{} \PH{tàmàɲíní} \PS{n.~} \GE{gum} \OP{tamanyime} \TP{tamanyimee} \VA{tanyini}
\LX{tamaraa} \HM{} \PH{támáráà} \PS{n.~} \GE{small quiver} \OP{tamarre} \TP{tamarrɛɛ} 
\LX{tamaregye} \HM{} \PH{támárɪ́dzé} \PS{n.~} \GE{small pantaloons; short, tight trousers} \OP{tamaregyeri} \TP{tamaregyeree} \VA{tamaragye}
\LX{tambaa} \HM{} \PH{tǎmbáá} \PS{n.~} \GE{bow bell (a hollow, carved object that is tied to the beam of the bow and makes a sound when the bow is shot)} \OP{tambaare} \TP{tambaarɛɛ} 
\LX{tambalaaraa} \HM{} \PH{támbálåáráá} \PS{n.~} \GE{honey made by the yoyogibie} \OP{tambalaare} \TP{tambalaarɛɛ} \VA{tammalaaraa}
\LX{tambiri} \HM{} \PH{támbírì} \PS{n.~} \GE{bamboo} \OP{tambie} \TP{tambiiree} 
\LX{tambɔge} \HM{} \PH{tàmbɔ́gɪ̀} \PS{n.~} \GE{ditch or pit created by digging earth to build.(usually near the house)} \OP{tambɔgere} \TP{tambɔgerɛɛ} 
\LX{tambogi} \HM{} \PH{tàmbògí} \PS{n.~} \GE{cave} \OP{tambogiri} \TP{tambogiree} 
\LX{tammalaaraa} \HM{} \PH{támmáláåráá} \PS{n.~} \GE{} \VA{tambalaaraa}
\LX{tammiri} \HM{} \PH{tàmmírì} \PS{n.~} \GE{bowstring} \OP{tammie} \TP{tammiiree} 
\LX{tammo} \HM{} \PH{támmʊ́} \PS{n.~} \GE{bow} \OP{tama} \TP{tamɛɛ/tannɛɛ} 
\LX{tamogilo} \HM{} \PH{támógíló} \PS{adj.~} \GE{toothless} \TP{tamogilee} \VA{tomogilo}
\LX{tamogilo} \HM{} \PH{támógíló} \PS{n.~} \GE{one who is toothless} \TP{tamogilee} 
\LX{tamoni} \HM{} \PH{tàmónì} \PS{n.~} \GE{toothless gum} \OP{tamonni} \TP{tamonnee} \VA{tamoŋi, tomoŋi}
\LX{tampɛle} \HM{} \PH{tàmpɛ́lɪ́} \PS{n.~} \GE{a hoe blade size of a slice/slab of earth (usually cut from soft ground with a hoe)} \OP{tampɛlɛ} \TP{tampɛlɛɛ} 
\LX{tampɛloŋ} \HM{} \PH{tàmpɛ̀lʊ̀ŋ} \PS{n.~} \GE{ashes} \TP{tampɛllɛɛ} 
\LX{tampoli} \HM{} \PH{tàmpólí} \PS{n.~} \GE{top end of a bow} \OP{tampolo} \TP{tampolee} 
\LX{tampuori} \HM{} \PH{tampuori} \PS{n.~} \GE{rubbish or garbage dump} \OP{tampoe} \TP{tampuoree} 
\LX{taŋa} \HM{} \PH{táŋà} \PS{n.~} \GE{mountain, hill} \OP{tanne} \TP{tannɛɛ} 
\LX{taŋaa} \HM{} \PH{táŋàà} \PS{n.~} \SN \GE{shea tree} \SN \GE{shea fruit} \OP{taŋema} \TP{taŋemɛɛ} 
\LX{taŋaaweɛlɛ} \HM{} \PH{táŋááwɪ̀ɛ̀lɛ́} \PS{n.~} \GE{mistletoe of shea tree} \SG{taŋaaweɛloo} \TP{taŋaaweɛlɛɛ} 
\LX{tandaa} \HM{} \PH{tàndàá} \PS{n.~} \GE{bow stick} \OP{tandaare} \TP{tandarɛɛ} 
\LX{tandagere} \HM{} \PH{tàndàgɪ́rɪ́} \PS{n.~} \GE{piece of hard dried mud} \OP{tandaga} \TP{tandagɛɛ} 
\LX{tandɛre} \HM{} \PH{tàndɛ́rɪ̀} \PS{n.~} \GE{bow stick} \OP{tandɛɛ} \TP{tandɛrɛɛ} 
\LX{taŋe} \HM{} \PH{tâŋ} \PS{v.~} \GE{to keep silent, to refuse to respond (to a call)} \PD{taŋɛɛ, tanna, tanna, tanema, tannaa, - } 
\LX{tanfaraŋ} \HM{} \PH{tánfáráŋ} \PS{n.~} \GE{a group farming agreement in which the members alternate farms} 
\LX{tanfoɔlaa} \HM{} \PH{tânfʊ́ɔ́làà} \PS{n.~} \GE{valley between hills} \OP{tanfoɔle} \TP{tanfoɔlɛɛ} 
\LX{taŋgara} \HM{} \PH{táŋgárá} \PS{n.~} \GE{telegraph} \OP{taŋgarre} \TP{taŋgarɛɛ/taŋgarrɛɛ} 
\LX{taŋgboŋi} \HM{} \PH{tàŋgbôŋ} \PS{n.~} \GE{a type of short tree with broad leaves and fibrous bark} \OP{taŋgbonni} \TP{taŋgbonnee} 
\LX{taŋgɔɔtuo} \HM{} \PH{tàŋgɔ̀ɔ̀túó} \PS{n.~} \GE{a kind of tree} \OP{taŋgɔtuuri} \TP{taŋgɔɔtuoree} 
\LX{tangyugi} \HM{} \PH{tándzúgí} \PS{n.~} \GE{in the habit of being promiscuous e.g male goat} \OP{tangyugri} \TP{tangyugree} \VA{tangyuu}
\LX{tangyuu} \HM{} \PH{tágyùú} \PS{n.~} \GE{in the habit of being promiscuous e.g (male goat)} \VA{tangyugi}
\LX{taŋkama} \HM{} \PH{tàŋkàmà} \PS{n.~} \GE{vanity} 
\LX{tanlori} \HM{} \PH{tánlórí} \PS{n.~} \GE{miniature bow, children's bow} \OP{ tanloe} \TP{tanloree} 
\LX{tanne} \HM{} \PH{tànnɪ̀} \PS{v.~} \SN \GE{to thunder} \SN \GE{to shout at, to scold} \PD{tannɛɛ, tanna, tanna, tannema, tannaa, tannema } 
\LX{tannetaa} \HM{} \PH{tànnɪ̀táá} \PS{n.~} \GE{quarrel} 
\LX{tannoɔ} \HM{} \PH{tǎnnʊ́ɔ̀} \PS{n.~} \GE{a small, dark, henlike bird that lives in highlands} \OP{tannoore} \TP{tannoorɛɛ} 
\LX{tantammonaa} \HM{} \PH{tántámmʊ́náá} \PS{n.~} \GE{a type of red and blac canary} \OP{tantammonne} \TP{tantammonnɛɛ} 
\LX{tantammonaa} \HM{} \PH{tántámmʊ́náà} \PS{n.~} \GE{red weaver bird} \OP{tantammonne} \TP{tantammonɛɛ} 
\LX{tantaŋ} \HM{} \PH{tàntàŋ} \PS{adv.~} \GE{immediately} 
\LX{tanyini} \HM{} \PH{tànyíní} \PS{n.~} \GE{} \OP{tanyinni} \TP{tanyinee} \VA{tamanyini}
\LX{tanzuŋ} \HM{} \PH{tànzûŋ} \PS{n.~} \GE{carrying pad} \OP{tanzunni} \TP{tanzunnee} \VA{tasere}
\LX{tapɔgefa᷉a} \HM{} \PH{tápɔ́gɪ́fà᷉à᷉} \PS{v.~} \GE{finish very quickly} \PD{-, -, -, -, -, - } 
\LX{tapolae} \HM{} \PH{tápʊ́láɪ̀} \PS{n.~} \GE{touch-light} \OP{tapolaɛ} \TP{tapolɛɛrɛɛ} \VA{tapolɛɛ}
\LX{tapolii} \HM{} \PH{tápólíí} \PS{n.~} \GE{tarpaulin} \OP{tapolli} \TP{tapollee} 
\LX{taporo} \HM{} \PH{tàpòrò} \PS{n.~} \GE{brain (matter)} \OP{taporri} \TP{taporee} \VA{kyaporo}
\LX{taraa} \HM{} \PH{tàràà} \PS{adj.~} \GE{flat and thick} 
\LX{taramaŋ} \HM{} \PH{tàràmàŋ} \PS{adv.~} \GE{blindingly} 
\LX{tare} \HM{} \PH{tàrì} \PS{v.~} \SN \GE{to set sitting right on top of something e.g. pot on fire} \SN \GE{to face directly e.g. problem} \PD{tarɛɛ, tara, tara, tareba, taraa, - } 
\LX{taremana} \HM{} \PH{tàrɪ̀mànà} \PS{adv.~} \GE{completely blind} 
\LX{tasere} \HM{} \PH{tàsɪ́rɪ́} \PS{n.~} \GE{see tanzuŋi; a pad put on the head to carry load} \OP{taseɛ} \TP{taserɛɛ} \VA{tanzuŋ}
\LX{tasɔle} \HM{} \PH{tásɔ́lɪ́} \PS{n.~} \GE{a type of lizard with smooth pink skin} \OP{tasɔlɔ} \TP{tasɔlɛɛ} 
\LX{tata} \HM{} \PH{tátà} \PS{n.~} \GE{boastfulness, the act of attempting the impossible} 
\LX{tata yɛlɛ} \HM{} \PH{tátà yɛ́lɛ́} \PS{n.~} \GE{boastful behaviour or claim} 
\LX{tateŋɛ} \HM{} \PH{tátɪ́ŋɛ́} \PS{n.~} \GE{distant country} \OP{tatenne} \TP{tatennɛɛ} 
\LX{tatoɔroŋ} \HM{} \PH{tàtʊ́ɔ́rʊ́ŋ} \PS{n.~} \GE{carelessness} 
\LX{taworo} \HM{} \PH{táwòró} \PS{n.~} \GE{towel} \OP{taworri} \TP{taworee} 
\LX{tayeŋine} \HM{} \PH{táyéŋíné} \PS{n.~} \GE{the state of trembling e.g. from severe hunger} \OP{tayenni} \TP{tayennee} 
\LX{ta᷉a} \HM{} \PH{tà᷉à᷉} \PS{v.~} \GE{to serve very little e.g. soup} \XV{A zeɛre ba seŋe bone lɛnso ka ba ta᷉a᷉ a a lɛ.} \XE{There is very little soup left and that is why they have given out so little.} \PD{ta᷉aɛ, taana, taana, taanema, taanaa, - } 
\LX{ta᷉a} \HM{} \PH{tà᷉à᷉} \PS{v.~} \GE{to tap on the head with a finger} \PD{ta᷉aɛ, taana, taana, taanema, taanaa, - } 
\LX{ta᷉ate᷉e} \HM{} \PH{tà᷉á᷉tɪ̀᷉ɪ́᷉} \PS{n.~} \GE{a type of small bird} \OP{ta᷉ateene} \TP{ta᷉ateenɛɛ} 
\LX{te} \HM{} \PH{tɪ̀} \PS{pronoun.~} \SN \GE{we (first person plural)} \SN \GE{us (first person plural)} 
\LX{te} \HM{} \PH{té} \PS{adv.~} \GE{really} \VA{to}
\LX{tɛ} \HM{} \PH{tɛ́} \PS{v.~} \GE{to spread in an orderly manner} \XV{Dasɛrɛɛ ka Dagaaba maŋ tɛ ne ba die kyɛ lɔɔ.} \XE{The Dagaaba arrange small sticks on planks of wood and on which they put local mortar to roof a building.} \PD{tɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, - } 
\LX{tɛbaa} \HM{} \PH{tɛ́báá} \PS{n.~} \GE{snare} \OP{tɛrre} \TP{tɛrrɛɛ} 
\LX{teɛ} \HM{} \PH{tɪ̀ɛ̀} \PS{v.~} \GE{to support} \PD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛ} \HM{} \PH{tɪ̀ɛ̀} \PS{v.~} \GE{to shoot with a bow} \PD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛ} \HM{} \PH{tɪ̀ɛ̀} \PS{v.~} \GE{to gamble} \PD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛ} \HM{} \PH{tɪ́ɛ́} \PS{n.~} \GE{scabbard} \OP{teere} \TP{teerɛɛ} 
\LX{teɛ} \HM{} \PH{tɪ́ɛ̀} \PS{v.~} \GE{to kick} \PD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛ} \HM{} \PH{tɪ́ɛ̀} \PS{v.~} \GE{to unset, to release (snare)} \PD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛ} \HM{} \PH{} \PS{n.~} \GE{tree} \OP{teere} \TP{teerɛɛ} 
\LX{teɛ} \HM{} \PH{} \PS{n.~} \GE{a disease that causes a rash (usually on the legs)} \OP{teere} \TP{teerɛɛ} 
\LX{tẽɛ} \HM{} \PH{tɪ́᷉ɛ́} \PS{v.~} \SN \GE{to stretch out e.g. legs} \XV{A baala zeŋe dɛlle la a gado kyɛ te᷉ɛ o gbɛɛ.} \XE{The patient sat leaning against the bed with his legs stretched out.} \SN \GE{to point to} \XV{A pɔgelee da to᷉ɔ te᷉ɛ wuli la a nannyige.} \XE{The girl was able to point to the thief.} \SN \GE{to hold out to give} \XV{Ba de la a bondirii te᷉ɛ a pɔge ka o zagere.} \XE{They held out  the food to the woman, but she refused.} \PD{teɛ᷈ɛ, teɛnɛ, teɛnɛ, teɛnemɛ, teɛnaa, teɛnemɛ } 
\LX{teebo} \HM{} \PH{tɪ́ɪ́bʊ́} \PS{n.~} \GE{fetish} \OP{tebɛ} \TP{teerɛɛ} 
\LX{teelɛ} \HM{} \PH{tɪ́ɪ́lɛ́} \PS{n.~} \GE{sewing machine} \OP{teelɛre} \TP{teelɛrɛɛ} 
\LX{teene} \HM{} \PH{tɪ̀ɪ̀nɪ̀} \PS{v.~} \GE{improve upon} \PD{teenɛɛ, teenɛ, teenɛ, teenemɛ, teenaa, - } 
\LX{teene} \HM{} \PH{tɪ̀ɪ̀nɪ̀} \PS{v.~} \GE{delay, waste time} \PD{teenɛɛ, teenɛ, teenɛ, teenemɛ, teenaa, - } 
\LX{teene} \HM{} \PH{tɪ̀ɪ̀nɪ̀} \PS{v.~} \GE{to be the spokesman for a suitor in courtship} \PD{teenɛɛ, teenɛ, teenɛ, teenemɛ, teenaa, - } 
\LX{teɛne} \HM{} \PH{tɪ̀ɛ́nɪ́} \PS{n.~} \GE{chin} \OP{teɛmɛ} \TP{teɛnɛɛ} 
\LX{teɛne} \HM{} \PH{tɪ̀ɛ̀nɪ̀} \PS{v.~} \GE{to stuff e.g. pillow} \PD{teɛnɛɛ, teɛnɛ, teɛnɛ, teɛnemɛ, teɛnaa, - } 
\LX{teere} \HM{} \PH{tɪ̀ɪ̀rɪ̀} \PS{v.~} \SN \GE{to rub against} \SN \GE{to iron clothes} \PD{teerɛɛ, teerɛ, teerɛ, teerebɛ, teeraa, - } 
\LX{teɛre} \HM{} \PH{tɪ́ɛ́rɪ̀} \PS{v.~} \SN \GE{to think} \SN \GE{to remind} \PD{teɛrɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛre} \HM{} \PH{tɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to feel for texture (smoothness of flour, swelling)} \PD{teɛrɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\LX{teɛre tɔ} \HM{} \PH{tɪ́ɛ́rɪ́ tɔ̀} \PS{} \GE{hope} \XV{Ba de la a baala gaa ne asibiti kyɛ teɛre toɔbo kyebe.} \XE{The patient has been taken to the hospital but there is no hope.} 
\LX{tɛɛse} \HM{} \PH{tɛ́ɛ́sɪ́} \PS{n.~} \GE{test} \OP{tɛɛsɪrɪ} \TP{tɛɛsɪrɛɛ} 
\LX{tɛɛtɛɛ} \HM{} \PH{tɛ́ɛ́tɛ̀ɛ̀} \PS{adj.~} \GE{different} 
\LX{tee᷉-erɛ} \HM{} \PH{tɪ́ɪ́᷉-ɪ́rɛ́} \PS{n.~} \GE{medicine man, traditional healer} \OP{tee᷉-erebɛ} \TP{tee᷉-erebɛɛ} 
\LX{tege} \HM{} \PH{tɪ́gɪ́} \PS{v.~} \GE{to be satisfied} \PD{tegɛɛ, tegrɛ, tegrɛ, tegrebɛ, tegraa, - } 
\LX{tɛge} \HM{} \PH{tɛ̀gɪ̀} \PS{v.~} \SN \GE{to exchange, to barter} \SN \GE{to change (action, form)} \SN \GE{to stop doing something suddenly and for a short time} \PD{-, -, -, -, -, - } 
\LX{tɛge} \HM{} \PH{tɛ̀gɪ́} \PS{n.~} \GE{meaning} \TP{tɛgrɛɛ} \VA{tɛgɛ}
\LX{tɛgɛ} \HM{} \PH{tɛ́gɛ́} \PS{n.~} \GE{meaning} \VA{tɛge}
\LX{tegele} \HM{} \PH{tɪ̀glɪ̀} \PS{v.~} \GE{to shorten} \PD{tegelɛɛ, tegelɛ, tegelɛ, tegelebɛ, tegelaa, teglebɛ } 
\LX{tɛgele} \HM{} \PH{tɛ́glɪ̀} \PS{v.~} \SN \GE{to place a container with the bottom on the ground} \SN \GE{to collect in a small pool e.g water} \PD{tɛgelɛɛ, tɛgelɛ, tɛgelɛ, tɛgelebɛ, tɛgelaa, - } 
\LX{tegelɛŋ} \HM{} \PH{tɪ́glɛ́ŋ} \PS{adv.~} \SN \GE{little, small (amount of liquid)} \SN \GE{short (length, time)} 
\LX{tɛgere} \HM{} \PH{tɛ̀grɪ̀} \PS{v.~} \GE{to apply the sap of a leaf to the body (local method of tattooing)} \PD{tɛgerɛɛ, tɛgerɛ, tɛgerɛ, tɛgerebɛ, tɛgraa, - } 
\LX{tegeroŋ} \HM{} \PH{tɪ́grʊ́ŋ} \PS{adj.~} \GE{heavy (weight)} \OP{tegerɛ} 
\LX{tegeroŋ} \HM{} \PH{tɪ́grʊ́ŋ} \PS{n.~} \GE{a heavy thing} \OP{tegerɛ} 
\LX{tɛgɛtɛgɛ} \HM{} \PH{tɛ́gɛ́tɛ́gɛ́} \PS{adv.~} \GE{little e.g. liquid} 
\LX{tegitegi} \HM{} \PH{tégtég} \PS{adv.~} \GE{always} 
\LX{tegitegi lɛ} \HM{} \PH{tégtéglɛ̀} \PS{adj.~} \GE{always like that, usually like that} 
\LX{tegyeerɛ} \HM{} \PH{tɪ̀dzɪ́ɪ́rɛ́} \PS{n.~} \GE{woodpecker} \OP{tegyeere} \TP{tegyeerɛɛ} \VA{tekoliŋkuoraa}
\LX{tekoliŋkuoraa} \HM{} \PH{tɪ̀kólíŋkúóráá} \PS{n.~} \GE{woodpecker} \OP{tekoliŋkuori} \TP{tekoliŋkuoree} \VA{tegyeerɛ, dakyeɛrɛko}
\LX{tekonee} \HM{} \PH{tɪ́kʊ́nɪ́ɪ́} \PS{adj.~} \GE{tiny} 
\LX{tekoɔ᷉} \HM{} \PH{tɪ́kʋ̃́ɔ́} \PS{n.~} \GE{sap} \VA{tesunni}
\LX{tekpiri} \HM{} \PH{tɪ̀kpírì} \PS{n.~} \GE{stump} \SG{tekpiruu} \TP{tekpiree} 
\LX{tekyɛ} \HM{} \PH{tɪ́ʧɛ̀} \PS{n.~} \GE{teacher} \OP{tekyɛre} \TP{tekyɛrɛɛ} 
\LX{tekyɛ} \HM{} \PH{tɪ̀ʧɛ̀} \PS{n.~} \GE{the act of felling trees} 
\LX{telare} \HM{} \PH{tɪ́làrɪ̀} \PS{adv.~} \GE{too much} 
\LX{tɛlɛntɛlɛŋ} \HM{} \PH{tɛ̀lɛ̀ntɛ́lɛ́ŋ} \PS{adv.~} \GE{sprawled out} 
\LX{teli} \HM{} \PH{télí} \PS{v.~} \GE{to be in an upright position e.g. head} \PD{telee, tele, tele, telibe, telaa, - } 
\LX{tɛll} \HM{} \PH{tɛ̀ll} \PS{adj.~} \GE{long, straight, and rigid} 
\LX{teloɔraa} \HM{} \PH{tɪ̀lʊ́ɔ́ráá} \PS{n.~} \GE{animal that lives in trees e.g. squirrel} \OP{teloɔre} \TP{teloɔrɛɛ} 
\LX{tembaa} \HM{} \PH{tɪ̌mbåá} \PS{n.~} \GE{a type of black ant with a painful sting} \OP{tembaare} \TP{tembaarɛɛ} 
\LX{tembeene} \HM{} \PH{tɪ̂mbɪ́ɪ́nɪ́} \PS{n.~} \GE{late night} \OP{tembeemɛ} \TP{tembeemɛɛ} 
\LX{tembie} \HM{} \PH{tɪ̀mbíé} \PS{n.~} \GE{citizen} \OP{tembiiri} \TP{tembiiree} 
\LX{tembimeŋɛ} \HM{} \PH{tɪ̀mbìmɪ́ŋɛ́} \PS{n.~} \GE{patriot} \OP{tembimenne} \TP{tembimennɛɛ} 
\LX{tɛmbiri} \HM{} \PH{tɛ̌mbìrí} \PS{n.~} \SN \GE{white ants (termites)} \SN \GE{the habitat of these termites} \OP{tɛmbie} \TP{tɛmbiiree} 
\LX{tɛmbogi} \HM{} \PH{tɛ̀mbògí} \PS{n.~} \GE{ditch where mortar is prepared} \OP{tɛmbogiri} \TP{tɛmbogiree} 
\LX{tɛmburo} \HM{} \PH{tɛ̀mbùró} \PS{n.~} \GE{dry soil} \TP{tɛmburee} 
\LX{temm} \HM{} \PH{témm} \PS{adv.~} \GE{alone; lonely} 
\LX{tɛmpelaa} \HM{} \PH{tɛ̌mpɪ̀làá} \PS{n.~} \GE{white soil (used to plaster walls)} \OP{tɛmpeɛle} \TP{tɛmpeɛlɛɛ} 
\LX{tɛmpoɔroŋ} \HM{} \PH{tɛ̀mpʊ́ɔ́rʊ́ŋ} \PS{n.~} \GE{fertile soil} \OP{tɛmpoɔrɔ} 
\LX{tempoɔsori} \HM{} \PH{tîmpʊ́ɔ́sórì} \PS{n.~} \GE{street} \OP{tempoɔsoe} \TP{tempoɔsoree} 
\LX{tɛŋ} \HM{} \PH{tɛ̀ŋ} \PS{v.~} \SN \GE{to sweeten} \SN \GE{to ferment} \PD{tɛŋɛɛ, tɛnɛ, tɛnɛ, tɛnɛ, tɛnemɛ, tɛnaa } 
\LX{teŋa} \HM{} \PH{tɪ́ŋá} \PS{n.~} \SN \GE{ground} \SN \GE{town, village} \SN \GE{country} \VA{teŋɛ}
\LX{tendaa} \HM{} \PH{tɪ̀ndàá} \PS{adv.~} \GE{country wide} 
\LX{tendaana} \HM{} \PH{tìndááná} \PS{n.~} \GE{landowner} \OP{tendaama} \TP{tendaamɛɛ} 
\LX{teŋe} \HM{} \PH{tɪ̀ŋ} \PS{v.~} \GE{} \PD{teŋɛɛ, teŋenɛ, teŋenɛ, teŋenemɛ, teŋenaa, - } 
\LX{teŋɛ} \HM{} \PH{tɪ́ŋɛ́} \PS{adj.~} \GE{down} 
\LX{teŋɛ} \HM{} \PH{tíŋɛ́} \PS{n.~} \SN \GE{ground} \SN \GE{town, village} \SN \GE{country} \OP{tenne} \TP{tennɛɛ} \VA{teŋa}
\LX{tɛne} \HM{} \PH{tɛ́nɪ́} \PS{n.~} \SN \GE{soil, earth, mud} \SN \GE{the spirit of a land} \TP{tɛnɛɛ} 
\LX{tenee} \HM{} \PH{tɪ̀nɪ́ɪ̀} \PS{poss. pron.~} \GE{we} 
\LX{teŋɛsoga} \HM{} \PH{tɪ́ŋɛ́sʊ́gá} \PS{n.~} \GE{floor} \OP{teŋɛsogre} \VA{teŋɛsogɔ}
\LX{teŋɛzu} \HM{} \PH{tɪ́ŋɛ́zù} \PS{n.~} \GE{world} \OP{teŋɛzuri} \TP{teŋɛzuree} 
\LX{tenfaa} \HM{} \PH{tɪ̀nfáá} \PS{n.~} \SN \GE{bad country, bad land} \SN \GE{hell} \OP{tenfaare} \TP{tenfaarɛɛ} 
\LX{teŋgane} \HM{} \PH{tǐŋgánɪ̀} \PS{n.~} \SN \GE{top layer of soil} \SN \GE{god of the land} \OP{teŋgama} \TP{teŋgamɛɛ} 
\LX{teŋganezu waloŋ zannoo} \HM{} \PH{tɪ̀ŋgánzu wààlʊ̀ŋ zánnʊʊ} \PS{n.~} \GE{geography} 
\LX{teŋgbigiri} \HM{} \PH{tɪ́ŋɛ́ gbígrúú} \PS{n.~} \GE{earthquake} 
\LX{teŋko᷉ɔ᷉} \HM{} \PH{tɪ́ŋkʊ́᷉ɔ́᷉} \PS{n.~} \GE{tradition} \TP{teŋkoɔnɛɛ} \VA{tensɛre}
\LX{teŋkpoŋ} \HM{} \PH{tɪ̀ŋkpôŋ} \PS{n.~} \GE{town, city} \OP{teŋkponni} \TP{teŋkponnee} 
\LX{tɛŋkubo} \HM{} \PH{tɛ̌ŋkúbò} \PS{n.~} \GE{gravel} 
\LX{tɛnnɛ} \HM{} \PH{tɛ̌nnɛ̀} \PS{n.~} \GE{a type of grass usually found in front of houses} \OP{tɛnne} \TP{tɛnnɛɛ} 
\LX{tensaana} \HM{} \PH{tɪ̌nsáánà} \PS{n.~} \GE{alien land, strange country} \OP{tensaama} \TP{tensaanɛɛ} 
\LX{tensaana} \HM{} \PH{tɪ̀nsááná} \PS{n.~} \GE{bandit} \OP{tensaanema} \TP{tensaamɛɛ} 
\LX{tensane} \HM{} \PH{tɪ̀nsánɪ̀} \PS{n.~} \GE{state of owing the gods of the land} \OP{tensama} \TP{tensannɛɛ} 
\LX{tensa᷉a} \HM{} \PH{tɪ̀nsàà᷉} \PS{n.~} \GE{disruption of community, country} 
\LX{tensɛre} \HM{} \PH{tɪ́sɛ́rɪ̀} \PS{n.~} \GE{tradition} \VA{teŋko᷉ɔ}
\LX{tensoba} \HM{} \PH{tɪ̀nsʊ̀bá} \PS{n.~} \GE{landowner} \OP{tendemee} 
\LX{tensogezɔnne} \HM{} \PH{tɪ́sʊ́gɪ́zɔ́nnɪ́} \PS{n.~} \GE{night blindness} 
\LX{tensogɔ} \HM{} \PH{tɪ̂nsʊ́gɔ́} \PS{n.~} \GE{night} \OP{tensogere} \TP{tensogerɛɛ} \VA{tensoga, yensogɔ}
\LX{tensunni} \HM{} \PH{tɪ́nsúnní} \PS{n.~} \SN \GE{muddy ground} \SN \GE{fertile land} \OP{tensumo} 
\LX{tentɔle} \HM{} \PH{tɪ́ntɔ́lɪ́} \PS{n.~} \GE{under pants} \OP{tentɔlɔ} \TP{tentɔlɛɛ} 
\LX{tentoloŋ} \HM{} \PH{tɪ̂ntʊ́lʊ́ŋ} \PS{n.~} \SN \GE{hot ground} \SN \GE{hot seaon} 
\LX{tentɔlvaa} \HM{} \PH{tɪ́ntɔ̀lváà} \PS{n.~} \GE{triangular underwear} \OP{tentɔlvaare} \TP{tentɔlvaarɛɛ} 
\LX{tentɔɔre} \HM{} \PH{tɪ̀ntɔ́ɔ́rɪ́} \PS{n.~} \GE{distant country, town, or village} \OP{tentɔɔbɔ} 
\LX{tenvelaa} \HM{} \PH{tɪ̀nvɪ̀làá} \PS{n.~} \SN \GE{good country, good land} \SN \GE{heaven} \OP{tenveɛle} 
\LX{teraale} \HM{} \PH{tìrààlɪ́} \PS{n.~} \GE{perfume} \OP{teraalere} \TP{terelerɛɛ} 
\LX{tere} \HM{} \PH{téré} \PS{adj.~} \GE{only} 
\LX{tere} \HM{} \PH{tɪ̀rɪ́} \PS{n.~} \GE{spoon} \OP{teɛ} \TP{terɛɛ} 
\LX{tere} \HM{} \PH{tɪ̀rɪ̀} \PS{v.~} \GE{to give, to send} \PD{terɛɛ, terɛ, terɛ, terebɛ, teraa, - } 
\LX{tɛre} \HM{} \PH{tɛ̀rɪ̀} \PS{v.~} \GE{to strain one's legs} \XV{A mɔmoɔre da tɛre la gbɛɛ eŋe taa kyɛ prri wale.} \XE{The wrestlers strained their legs and sweated too.} \PD{tɛrɛɛ, tɛrɛ, tɛrɛ, tɛrebɛ, tɛraa, tɛrebɛ } 
\LX{tere noɔre} \HM{} \PH{tɪ̀rɪ̀ nʊ́ɔ́rɪ̀} \PS{v.~} \GE{to leave a message} \PD{-, -, -, -, -, - } \VA{bare noɔre}
\LX{tere sori} \HM{} \PH{tɪ́rɪ́ sórì} \PS{v.~} \GE{allow, permit} \PD{-, -, -, -, -, - } 
\LX{teree} \HM{} \PH{téréé} \PS{n.~} \GE{a statement made for comments} \XV{A koŋkono lɔɔ la teree ko a kotuo deme.} \XE{The dirge singer has presented a riddle to the bereaved family.} \OP{terri} \TP{terree} 
\LX{teree-teree} \HM{} \PH{térée-téréé} \PS{adj.~} \GE{spread out and roundish at the end, cone-shaped} 
\LX{tɛrɛɛtɛrɛɛ} \HM{} \PH{tɛ́rɛ́ɛ́tɛ́rɛ́ɛ́} \PS{adv.~} \GE{flat and even} 
\LX{terembelli} \HM{} \PH{térémbéllí} \PS{adv.~} \GE{straight and upright} 
\LX{tɛrɛsɛ} \HM{} \PH{tɛ̀rɛ̀sɛ̀} \PS{adv.~} \GE{clumsily} \XV{O tegɛɛ la a deɛ leɛ e tɛrɛsɛ lɛ o naŋ zeŋ.} \XE{He is so full with food that he can hardly move.} 
\LX{teretere} \HM{} \PH{tɪ̀rɪ̀tɪ́rɪ́} \PS{v.~} \GE{to crush with feet, to trample} \PD{tereterɛɛ, tereterɛ, tereterebɛ, tereteraa, -, - } 
\LX{tɛretɛre} \HM{} \PH{tɛ̀rɪ̀tɛ́rɪ́} \PS{v.~} \GE{to move in protest e.g. pampered child} \XV{A bie zagere ka kyɛŋe kyɛ maŋ tɛretɛre.} \XE{The child has refused to walk up and is moving under protest.} \PD{tɛretɛrɛɛ, tɛretɛrɛ, tɛretɛrɛ, tɛretɛrebɛ, tɛretɛraa, - } 
\LX{tɛrre} \HM{} \PH{tɛ́rrɪ̀} \PS{v.~} \GE{to level out} \PD{tɛrrɛɛ, tɛrrɛ, tɛrrɛ, tɛrebɛ, tɛrraa, - } 
\LX{tɛrre} \HM{} \PH{tɛ́rrɪ̀} \PS{v.~} \GE{to push a container away along the ground/floor usually in anger} \PD{tɛrrɛɛ, tɛrrɛ, tɛrrɛ, tɛrebɛ, tɛrraa, - } 
\LX{terri} \HM{} \PH{térrì} \PS{v.~} \GE{to smooth out the end e.g. pestle} \PD{terree, terre, terre, teribe, terraa, - } 
\LX{tesogɔ} \HM{} \PH{tɪ̀sʊ́gɔ́} \PS{n.~} \GE{trunk (of tree)} \OP{tesogere} \TP{tesogerɛɛ} \VA{tesoga}
\LX{tesunni} \HM{} \PH{tèsúnnì} \PS{n.~} \GE{sap} \SG{tesunnuu} \TP{tesunnee} \VA{tekoɔ᷉}
\LX{tetebaa} \HM{} \PH{tɪ́tɪ́báá} \PS{n.~} \GE{stinger; the tip of the tail of a bee usually used to sting} \OP{teterre} \TP{teterɛɛ} 
\LX{tɛtɛgere} \HM{} \PH{tɛ̀tɛ̀gɪ̀rɪ́} \PS{n.~} \GE{muddy thing} \OP{tɛtɛgɛ} 
\LX{tetere} \HM{} \PH{tɪ́tɪ̀rɪ̀} \PS{adj.~} \GE{small in quantity} 
\LX{tɛtɛrewɛrɛ} \HM{} \PH{tɛ̀tɛ̀rɪ́wɛ̌rɛ́} \PS{adv.~} \GE{daftly} 
\LX{tevaa} \HM{} \PH{tɪ̀váà} \PS{adj.~} \GE{green} \OP{tevaare} \TP{tevaarɛɛ} 
\LX{tewɔne} \HM{} \PH{tɪ̀wɔ́nɪ́} \PS{n.~} \SN \GE{berry} \SN \GE{fruit of a tree} \OP{tewɔmɔ} \TP{tewɔnnɛɛ} 
\LX{tewɔŋko᷈ɔ᷉} \HM{} \PH{tɪ́wɔ́ŋkʊ́ɔ́} \PS{n.~} \GE{juice} \TP{tewɔŋkoɔnɛɛ} 
\LX{te᷉e} \HM{} \PH{tɪ́᷉ɪ́᷉} \PS{n.~} \SN \GE{medicine} \XV{O beɛrɛ la kyɛ ba sage na nyu te᷉e.} \XE{She is sick but refuses to take medicine.} \SN \GE{solution} \XV{A pɔge taa la a bie ŋa te᷉e, o maŋ wa de o la ka o ba la kono.} \XE{The woman has some solution for this child, because any time she picks him up, he stops crying.} \SN \GE{poison} \XV{Ba de la gyaŋgbonni te᷉e ko ne a yiri gyaŋgbonni zaa.} \XE{The used mouse poison to kill all the mice in the house.} \SG{teenoo} \TP{teenɛɛ} 
\LX{te᷉eda} \HM{} \PH{t̀ɪ̀᷉ɪ̀᷉dà} \PS{n.~} \GE{the act of buying medicine} \XV{A dɔɔ ŋa ba nɔŋe asibiti gaabo, te᷉eda ka o dɛlle.} \XE{This man does not like going to the hospital; he relies on buying medicine.} 
\LX{te᷉e-daare} \HM{} \PH{tɪ́᷉ɪ́᷉-dáárɪ́} \PS{n.~} \GE{bought medicine} 
\LX{te᷉e-die} \HM{} \PH{tɪ́᷉ɪ́᷉-dìé} \PS{n.~} \SN \GE{clinic, hospital} \SN \GE{pharmacy} \OP{te᷉e-deri} \TP{te᷉e-deree} 
\LX{te᷉e-doge} \HM{} \PH{tɪ́᷉ɪ́᷉-dʊ́gɪ̀} \PS{n.~} \GE{medicine pot} \OP{te᷉e-dogere} \TP{te᷉e-dogerɛɛ} 
\LX{te᷈ɛkɔɔloŋ} \HM{} \PH{tɪ̀᷉ɛ́᷉kɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{beard} \OP{te᷈ɛkɔɔlɔ} \TP{te᷈ɛkɔɔlɛɛ} 
\LX{te᷉ekoɔre-die} \HM{} \PH{tɪ̀᷈ɪ́᷈kʊ̀ɔ́r-díé} \PS{n.~} \GE{pharmacy} \OP{te᷉e᷉koɔr-deri} \TP{te᷉e᷉koɔr-deree} 
\LX{te᷉eko᷉ɔ} \HM{} \PH{tɪ́᷈ɪ́᷉kʊ́᷉ɔ́᷉} \PS{n.~} \GE{syrup (medicine)} \TP{te᷉ekoɔnɛɛ} 
\LX{ti} \HM{} \PH{tì} \PS{v.~} \GE{to vomit} \PD{tiee, tiire, tiire, tiiribe, tiiraa, tiiribe } 
\LX{ti} \HM{} \PH{tí} \PS{v.~} \GE{to put hand on e.g wall} \XV{A baala ti la dankyini lɛnso ka o ba le.} \XE{The patient held the wall and that is why s/he did not fall.} \PD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\LX{ti} \HM{} \PH{tí} \PS{v.~} \GE{to support by the use of a walking stick} \PD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\LX{ti} \HM{} \PH{tí} \PS{v.~} \GE{to put up sticks with their tops meeting (making a farm hut)} \PD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\LX{ti} \HM{} \PH{tí} \PS{v.~} \GE{to begin, to start} \PD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\LX{tigi} \HM{} \PH{tìgì} \PS{v.~} \GE{to gather e.g. crowd} \PD{tigee, tigre, tigre, tigribe, tigraa, - } 
\LX{tigiki} \HM{} \PH{tìgìkì́} \PS{adj.~} \GE{dark and deep set (eyes)} \VA{tigtig}
\LX{tigikpoŋi} \HM{} \PH{tìgìkpóŋ} \PS{n.~} \GE{big feast, solemnity} \OP{tigikponni} \TP{tikponnee} 
\LX{tigiri} \HM{} \PH{tígrí} \PS{n.~} \GE{feast, festival} \OP{tige} \TP{tigee} 
\LX{tigiri} \HM{} \PH{tígrì} \PS{v.~} \GE{to grow fat} \PD{tigiree, tigire, tigire, tigiribe, tigiraa, - } 
\LX{tigiti} \HM{} \PH{tígítì} \PS{n.~} \GE{ticket} \OP{tigitiri} \TP{tigitiree} 
\LX{tigitigi} \HM{} \PH{tìgìtígí} \PS{v.~} \GE{to shiver violently e.g. when one has a fever} \PD{tigitigee, tigitigre, tigitigre, tigitigribe, tigitigraa, - } 
\LX{tiili} \HM{} \PH{tíílì} \PS{v.~} \GE{to place against an upright object e.g a plank against a tree/wall} \PD{tiilee, tiile, tiile, tiilibe, tiilaa, - } 
\LX{tiiri} \HM{} \PH{tíírí} \PS{n.~} \GE{vomit} \TP{tiiree} 
\LX{tiiri} \HM{} \PH{tììrì} \PS{v.~} \GE{to come very close to} \PD{tiiree, tiire, tiire, tiiribe, tiiraa, - } 
\LX{tiiri} \HM{} \PH{tììrì} \PS{v.~} \GE{to be fed up with} \PD{tiiree, tiire, tiire, tiiribe, tiiraa, - } 
\LX{tiiruŋ} \HM{} \PH{tíírǔŋ} \PS{n.~} \SN \GE{a type of plant with sour sap used for cooking} \SN \GE{a type of medicinal herbal concoction} 
\LX{tiiruŋ} \HM{} \PH{tììrǔŋ} \PS{adj.~} \GE{sour (taste)} 
\LX{tilento} \HM{} \PH{tíléntó} \PS{n.~} \SN \GE{hornbill} \SN \GE{pelican} \OP{tilentori} \TP{tilentoree} 
\LX{tiŋi} \HM{} \PH{tíŋ} \PS{v.~} \GE{to make firm (object fixed in the ground)} \PD{tiŋee, tiŋine, tiŋine, tiŋinime, tiŋinaa, - } 
\LX{tiŋimeŋ} \HM{} \PH{tìŋìméŋ} \PS{adj.~} \GE{dark} 
\LX{tiretire} \HM{} \PH{tírétíré} \PS{adv.~} \GE{tasteless} 
\LX{ti᷉i} \HM{} \PH{tí᷉í᷉} \PS{adv.~} \GE{closely and carefully e.g. look, observe} 
\LX{to} \HM{} \PH{tʊ́} \PS{v.~} \GE{to insult} \PD{toɛ, toora, toora, tooreba, tooraa, tooreba } 
\LX{to} \HM{} \PH{tó} \PS{adv.~} \GE{really} \XV{A bie ŋa zoro to.} \XE{This child runs really fast.} \VA{te}
\LX{tɔ} \HM{} \PH{tɔ́} \PS{n.~} \SN \GE{colleague, an equal} \XV{N baŋ ka n tɔ kaŋa la toŋ a tomɔ ŋa.} \XE{I guess this work is done by a neighbour.} \SN \GE{same type, same kind} \XV{N meŋ boɔrɔ la a fo kparoo ŋa tɔ.} \XE{I also want the same type of shirt that you have.} \OP{taaba} 
\LX{tɔ} \HM{} \PH{tɔ̀} \PS{v.~} \SN \GE{to pound} \SN \GE{to touch} \XV{O de la o nu tɔ ne a gane na o naŋ boɔrɔ.} \XE{She used her hand to touch the book she wanted.} \SN \GE{to head-butt} \XV{Ka fooŋ deɛnɛ ne naadɔgerɔ bile a naabo na tɔ fo la.} \XE{If you joke with the frehly born calf, its mother will head-butt you.} \SN \GE{to get to the end of} \XV{Te kyɛŋe te tɔ la a sori tɛgɛ.} \XE{We walked until we came to the end of the road.} \PD{tɔe, toɔrɔ, toɔrɔ, toɔreba, toɔraa, toɔreba } 
\LX{tɔ teere} \HM{} \PH{tɔ̀̀ tɪ̀ɪ̀rɪ̀} \PS{v.~} \SN \GE{without any hope/hit the rocks} \XV{A pɔge bie baaloŋ la so ka o sori na yiibu tɔ teere.} \XE{The woman's intended journey hit the rocks because of her child's sickness.} \SN \GE{to worsen} \XV{O gaa wɛre la o nimie na a deɛ tɔ teere.} \XE{The eye operation has worsened his sight.} \PD{tɔ teerɛɛ, toɔrɔ teerɛ, tɔteerɛ, tɔteerɛ, tɔteeraa, tɔteerebɛ } 
\LX{tɔ zɔŋɔ kyi} \HM{} \PH{tɔ̀ zɔ̀ŋɔ̀ kyî} \PS{v.~} \SN \GE{to beat about the bush} \SN \GE{to be ignorant} \PD{tɔɛ zɔŋ kyi, toɔrɔ zɔŋ kyi, -, -, -, - } 
\LX{tɔba} \HM{} \PH{tɔ̀bà} \PS{n.~} \GE{tobacco} \VA{tɔbɔ}
\LX{tobaa} \HM{} \PH{tóbáá} \PS{n.~} \GE{fox} \OP{tobaare} \TP{tobaarɛɛ} 
\LX{tɔbaa} \HM{} \PH{tɔ̀bàà} \PS{n.~} \GE{a type of sharp axe used for carving} \OP{tɔbre} \TP{tɔbrɛɛ} 
\LX{tobie sawɔlɔ} \HM{} \PH{tóbíè sáwɔ́lɔ́} \PS{n.~} \GE{fried food made from boabab seed} 
\LX{tobiri} \HM{} \PH{tóbírì} \PS{n.~} \GE{boabab seed} \SG{tobiruu} \OP{tobie} \TP{tobiiree} 
\LX{tɔbɔ} \HM{} \PH{tɔ̀bɔ̀} \PS{n.~} \GE{tobacco} \SG{tɔɔroo} \TP{tɔɔrɛɛ} \VA{tɔba}
\LX{tɔge} \HM{} \PH{tɔ́gɪ́} \PS{n.~} \SN \GE{an earthenware container with a hole in the bottom for sieving} \SN \GE{fish trap} \OP{tɔgere} \TP{tɔgerɛɛ} 
\LX{tɔge (Eng)} \HM{} \PH{tɔ́gɪ́} \PS{v.~} \GE{to stop} \PD{tɔgɛɛ, tɔgerɔ, tɔgerɔ, tɔgereba, tɔgeraa, - } 
\LX{tɔgele} \HM{} \PH{tɔ̀glɪ̀} \PS{v.~} \GE{to imitate} \XV{A bie tɔgele la a biwɔɔ kyɛnne a meŋ deɛ te kyɛnɛ wɔɔreŋ.} \XE{The boy imitated the way the lame boy walks and now he limps.} \PD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\LX{tɔgele} \HM{} \PH{tɔ̀glɪ̀} \PS{v.~} \GE{to collect from above e.g. rain water} \XV{Ka fooŋ tɔgele saa ko᷉ɔ᷉ a maŋ maaroŋ.} \XE{Gathered rain water is always cold.} \PD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\LX{tɔgele} \HM{} \PH{tɔ̀glɪ̀} \PS{v.~} \GE{to turn to face e.g. in a direction} \XV{O leɛ la o niŋe tɔgele sapare a na nyɛ a saa naŋ ko.} \XE{She has turned towards the east to see the clouds.} \PD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\LX{tɔgele} \HM{} \PH{tɔ̀glɪ̀} \PS{v.~} \GE{to help deliver a baby} \PD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\LX{togi} \HM{} \PH{tògì} \PS{v.~} \SN \GE{take down e.g. from head, fire} \SN \GE{to drip} \PD{togee, togiro, togiro, togiribo, togiraa, - } 
\LX{togifa} \HM{} \PH{tògífá} \PS{n.~} \GE{sixpence} \OP{togifare} \TP{togifarɛɛ} 
\LX{togilogo} \HM{} \PH{tóglógó} \PS{adv.~} \SN \GE{small (space)} \XV{Ba boma koŋ kpɛ a die, o deɛ waa la togilogoo lɛ.} \XE{The room cannot contain their belongings; it is very small.} \SN \GE{small and shallow (hole)} \XV{Ba tu la a bulee fa᷉a᷉ kyɛ deɛ e boŋkaŋa togilogoo lɛ.} \XE{They were unable to dig the well; they just etched a tiny hole.} 
\LX{togilogo} \HM{} \PH{tòglògò} \PS{adv.~} \GE{deep} \XV{Te tu la tambɔge togilogoo kaŋa a na kuire sagere ennɛ.} \XE{We dug a deep ditch into which we will dump refuse.} 
\LX{togiroo} \HM{} \PH{tógróó} \PS{n.~} \GE{raised wooded ground (Southern Dagaare dialect)} \OP{togiri} \TP{togiree} \VA{tuu}
\LX{togitogi} \HM{} \PH{tógtóg} \PS{adv.~} \GE{never, not at all} \XV{N gɛrɛ ka o gaaɛ N ba nyɛ o togitogi.} \XE{When I got there, he had left; I never saw him at all.} 
\LX{togoko} \HM{} \PH{tògòkò} \PS{adv.~} \GE{refusing to listen, hear; not minding} 
\LX{tokele} \HM{} \PH{tòkélé} \PS{n.~} \GE{hard dawdaw spice seed (not processable)} \OP{tokelee} \VA{dɔkele, kele}
\LX{tokokyile} \HM{} \PH{tòkóʧìlé} \PS{n.~} \GE{a bottle-like container made by removing the seeds from the baobab fruit} \OP{tokokyilli} \TP{tokokyilee} 
\LX{tokoroo} \HM{} \PH{tókóróó} \PS{n.~} \GE{window} \OP{tokorri} \TP{tokoree/tokorree} \VA{takoroo}
\LX{tokparaa} \HM{} \PH{tòkpáráà} \PS{n.~} \GE{boabab leaves (vegetable for soup)} \TP{tokparɛɛ} 
\LX{tɔle} \HM{} \PH{tɔ́lɪ́} \PS{v.~} \GE{to pass by} \PD{tɔlɛɛ, tɔlɔ, tɔlɔ, tɔlebɔ, tɔlaa, - } 
\LX{tolee} \HM{} \PH{tóléé} \PS{n.~} \GE{small mortar} \OP{tobilii} \TP{tobilee} 
\LX{tolema} \HM{} \PH{tʊ́lɪ́má} \PS{adv.~} \GE{avariciously} 
\LX{tolimala} \HM{} \PH{tòlmálá} \PS{n.~} \GE{great enjoyment, merrymaking (edibles)} \TP{tolimalɛɛ} \VA{tolomala}
\LX{toliŋka} \HM{} \PH{tólíŋkà} \PS{n.~} \GE{a car} \OP{toliŋkaɛ} \TP{toliŋkarɛɛ} 
\LX{toll} \HM{} \PH{tòll} \PS{adv.~} \GE{large (heap)} 
\LX{toll} \HM{} \PH{tóll} \PS{adj.~} \GE{small (heap)} 
\LX{tɔll} \HM{} \PH{tɔ̀ll} \PS{adj.~} \GE{long and cylindrical} 
\LX{tolomboll} \HM{} \PH{tólómbóll} \PS{adv.~} \GE{lonely-looking} 
\LX{toloŋ} \HM{} \PH{tʊ́lʊ́ŋ} \PS{n.~} \GE{heat} \XV{Toloŋ maŋ vɛŋ la ka te puri wale.} \XE{Heat induces swet.} \OP{tola} 
\LX{toloŋ} \HM{} \PH{tʊ́lʊ́ŋ} \PS{adj.~} \SN \GE{hot} \XV{A zeɛre ba taa diibu, a toloŋ.} \XE{The soup cannot be eaten just yet; it is very hot.} \SN \GE{fast, quick} \XV{Ka fooŋ ba kyɛŋ gbɛ toloŋ fo koŋ pɔge o.} \XE{If you don't hasten your steps, you won't meet her.} \OP{tola} 
\LX{tom} \HM{} \PH{tôm} \PS{v.~} \GE{to become sweet, to become enjoyable} \PD{tomee, tomino, tomino, tominimo, tominaa, - } 
\LX{tomomo} \HM{} \PH{tómómó} \PS{adv.~} \GE{very sweet} 
\LX{tone} \HM{} \PH{tʊ́nɪ́} \PS{n.~} \GE{work} \OP{toma} 
\LX{toŋe} \HM{} \PH{tʊ̀ŋɪ̀} \PS{v.~} \SN \GE{to work} \SN \GE{to send (someone to do something)} \PD{toŋɛɛ, tona, tona, tonema, tonaa, tonema } 
\LX{tɔŋe} \HM{} \PH{tɔ̀ŋ} \PS{v.~} \GE{to respond to, to answer, to reply} \PD{tɔŋɛɛ, tɔŋena/tɔnnɔ, tɔnnɔ, tɔnema, tɔnnaa, tɔnema } 
\LX{tɔŋe} \HM{} \PH{tɔ́ŋ} \PS{v.~} \SN \GE{to be in a queue} \SN \GE{to light a fire} \SN \GE{to fetch fire} \PD{tɔŋɛɛ, tɔnɔ, tɔnɔ, tɔnema, tɔnnaa, - } \VA{kyɔŋ}
\LX{tɔŋe} \HM{} \PH{tɔ̂ŋ} \PS{v.~} \GE{to light a fire from another fire} \PD{tɔŋɛɛ, tɔŋena/tɔnnɔ, tɔnnɔ, tɔnemɔ, tɔnnaa, tɔnema } 
\LX{tonemaa} \HM{} \PH{tʊ̊nɪ́máá} \PS{adj.~} \GE{at once, instantly} 
\LX{tɔŋemɔ} \HM{} \PH{tɔ́ŋɪ́mɔ̀} \PS{n.~} \GE{colleague} \OP{tɔŋemɔmine} \VA{tɔŋema}
\LX{toŋkpeɛne} \HM{} \PH{tʊ̂ŋkpɪ́ɛ́nɪ́} \PS{n.~} \GE{miracles, powerful deeds} \TP{toŋkpeɛnɛɛ} 
\LX{tɔnne} \HM{} \PH{tɔ́nnɪ̀} \PS{v.~} \GE{to join} \PD{tɔnnɛɛ, tɔnnɔ, tɔnnɔ, tɔnema, tɔnnaa, tɔnema } 
\LX{tɔnne} \HM{} \PH{tɔ̀nnɪ̀} \PS{v.~} \GE{to sell (usually at a low price)} \PD{tɔnnɛɛ, tɔnnɔ, tɔnnɔ, tɔnema, tɔnnaa, tɔnema } 
\LX{tonni} \HM{} \PH{tònnì} \PS{v.~} \GE{to move blindly; groping} \PD{tonnee, tonno, tonno, tonimo, tonnaa, - } 
\LX{tonni} \HM{} \PH{tónní} \PS{v.~} \GE{to drip} \PD{tonnee, tonno, tonno, tonimo, tonnaa, - } 
\LX{tɔnɔ} \HM{} \PH{tɔ̀nɔ́} \PS{n.~} \GE{profit, gain, benefit} \TP{tɔnɛɛ} \VA{tɔna}
\LX{tɔntɔbaa} \HM{} \PH{tɔ̀ntɔ́báá} \PS{adv.~} \GE{at once} 
\LX{tontombone} \HM{} \PH{tʊ̀ntʊ̀mbʊ́nɪ́} \PS{n.~} \GE{tool} \OP{tontombomai} 
\LX{tontoŋ} \HM{} \PH{tóntóŋ} \PS{adv.~} \SN \GE{very sour} \XV{A buuluŋ waa la tontoŋ zuo a nyuubu tuo.} \XE{The poridge is very sour; it is difficult to eat.} \SN \GE{very painful, sore e.g. over-night bruised body} \XV{A dɔɔ za᷉a de la o zu za ne a gbori, zenɛ a zaaŋ mii tontoŋ lɛ.} \XE{The man banged his head against the door yesterday and today it really hurts.} 
\LX{tontonɔ} \HM{} \PH{tʊ̌tʊ́ná} \PS{n.~} \SN \GE{worker} \SN \GE{industrious person} \OP{tontonema} \VA{tontona}
\LX{too} \HM{} \PH{tòò} \PS{use word.~} \GE{okay} \VA{tɔɔ}
\LX{tɔɔ} \HM{} \PH{tɔ̀ɔ̀} \PS{use word.~} \GE{all right, okay} \VA{too}
\LX{toobini} \HM{} \PH{tòòbíní} \PS{n.~} \GE{earwax (ear shit)} \OP{n/a} \TP{toobinee} 
\LX{tɔɔbɔle} \HM{} \PH{tɔ̀ɔ̀bɔ́lɪ́} \PS{n.~} \GE{cylinder-shaped pounded tobacco} \OP{tɔɔbɔlɔ} \TP{tɔɔbɔlɛɛ} 
\LX{too-gbuli} \HM{} \PH{tòò-gbúlí} \PS{n.~} \GE{unpierced ear} \OP{too-gbulo} \TP{too-gbulee} 
\LX{toogbulo} \HM{} \PH{tòògbùló} \PS{n.~} \SN \GE{hardness of hearing, deafness, blocked ears} \XV{Maa teɛre ka toogbulo ka taa, N maŋ yeli la a yɛlɛ ka a deɛ kaara n niŋeŋ.} \XE{I think she is hard of hearing, because when I talk to her she just stares into my face.} \SN \GE{an animal without a mark of identification on the ear} \XV{N peroo na naŋ bɔre lambori ŋmaa la pelaa kyɛ taa toogbulo.} \XE{My lost sheep has a white spot on its side but unmarked ears.} 
\LX{toogee} \HM{} \PH{tóógéé} \PS{interjection.~} \GE{a state indicating an incorrect utterance which is therefore withdrawn} \XV{Yeli ka toogee ka lɛ naane o na ŋmɛ fo la.} \XE{Better say it is a mistaken statement; otherwise he will beat you.} 
\LX{toogyɛle} \HM{} \PH{tòòdzɛ́lɪ́} \PS{n.~} \GE{earlobe} \OP{toogyɛlɛ} 
\LX{tookpane} \HM{} \PH{tòòkpànɪ̀} \PS{n.~} \GE{earring} \OP{tookpama} \TP{tookpanɛɛ/tookpannɛɛ} 
\LX{tookuriseɛ} \HM{} \PH{tòòkúrísɪ́ɛ́} \PS{n.~} \GE{the base of the ear (back)} \OP{tookuriseere} \TP{tookuriseerɛɛ} 
\LX{tɔɔlaa} \HM{} \PH{tɔ̀ɔ̀láá} \PS{n.~} \GE{tobacco pipe} \OP{tɔɔlaare} \TP{tɔɔlaarɛɛ} 
\LX{tɔɔlaka᷉a} \HM{} \PH{tɔ̀ɔ̀làká᷉à᷉} \PS{n.~} \GE{nicotine} \TP{tɔɔlaka᷉anɛɛ} 
\LX{toole} \HM{} \PH{tʊ̀ʊ̀lɪ̀} \PS{adj.~} \GE{used with ‘ba’ little e.g food} \XV{a bondirii ba toole a bie koŋ tege.} \XE{The food is little, the child will not eat to his fill.} 
\LX{toɔle} \HM{} \PH{tʊ́ɔ́lɪ̀} \PS{v.~} \GE{to send a message} \PD{toɔlɛɛ, toɔlɔ, toɔlɔ, toɔlbɔ, toɔlaa, - } 
\LX{toolere} \HM{} \PH{tòòléré} \PS{n.~} \GE{peak, climax, most enjoyable part (of music)} \OP{toolerri} \TP{tooleree/toolerree} 
\LX{toolii} \HM{} \PH{tòòlíí} \PS{n.~} \GE{joke} \OP{tooliiri} \TP{tooliiree} 
\LX{toɔlɔ} \HM{} \PH{tʊ́ɔ́lɔ̀} \PS{n.~} \GE{message} \VA{toɔla}
\LX{toɔne} \HM{} \PH{tʊ̀ɔ̀nɪ̀} \PS{v.~} \GE{to tear off in strips (fibre from the stalk)} \PD{toɔnɛɛ, toɔnɔ, toɔnɔ, toɔnema, toɔnaa, - } 
\LX{toopa᷉apere} \HM{} \PH{tòòpà᷉à᷉pɪ́rɪ̀} \PS{} \GE{eardrum} 
\LX{toopurimo} \HM{} \PH{tòópúrímó} \PS{n.~} \GE{draining ear} \TP{toopurimee} 
\LX{toɔraa} \HM{} \PH{tʊ́ɔ́ráá} \PS{n.~} \SN \GE{self} \XV{Ka fooŋ beɛrɛ kyɛ zagere te᷉e᷉ nyuubu fo toɔraa ka fo erɛ.} \XE{If you are sick but refuse medication, you will be hurting yourself.} \SN \GE{portion, share} \XV{Ba poŋ la nɛne ka N de n toɔraa ko n ma.} \XE{They shared the meat and I gave my share to my mother.} \OP{tɔɛ} \VA{toɔre}
\LX{toore} \HM{} \PH{tʊ́ʊ́rɪ̀} \PS{v.~} \GE{to spit} \PD{toorɛɛ, toorɔ, toorɔ, tooreba, tooraa, tooreba } 
\LX{toore} \HM{} \PH{tʊ̀ʊ̀rɪ̀} \PS{v.~} \SN \GE{to push into e.g foot into socks} \SN \GE{to push along a surface, to drag along} \PD{toorɛɛ, toorɔ, toorɔ, tooreba, tooraa, - } 
\LX{toɔre} \HM{} \PH{tʊ́ɔ́rɪ́} \PS{n.~} \GE{mortar, pounding pot} \TP{tɔɛ} \VA{toɔraa}
\LX{toɔre} \HM{} \PH{tʊ́ɔ́rɪ́} \PS{n.~} \GE{one's share of something} \XV{A seemaa yɛ naŋ maale, N toɔre bɛ?} \XE{Where is my share of the food you cooked?.} \TP{tɔɛ} \VA{toɔraa}
\LX{toɔre} \HM{} \PH{tʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to be careless; to not pay enough attention} \XV{O toɔre la a ne,ne bare ka baare de.} \XE{S/he left the meat carelessly and dogs ate it.} \PD{-, toɔrɔ, -, -, -, - } 
\LX{tɔɔre} \HM{} \PH{tɔ̀ɔ̀rɪ̀} \PS{n.~} \GE{ball of tobacco} \OP{tɔbɔ} \TP{tɔɔrɛɛ} 
\LX{tɔɔre} \HM{} \PH{tɔ̀ɔ̀rɪ́} \PS{adj.~} \GE{far} \OP{tɔbɔ} 
\LX{toori} \HM{} \PH{tòòrí} \PS{n.~} \GE{ear} \OP{tobo} \TP{tooree} 
\LX{toori} \HM{} \PH{tòòrí} \PS{adj.~} \GE{naughty e.g. child} \XV{A bie na sa᷉a᷉ la baaloŋ kyɛ o taa la toori yaga.} \XE{That child appears to be the quiet type but he is very naughty.} 
\LX{toori} \HM{} \PH{tòòrí} \PS{adj.~} \GE{musically exceptional} \XV{A bawaseɛrebɛ sereŋ taa la toori, ba gbɛɛ lanna lere la velaa.} \XE{The dancers have a good ear for music; their step falls on the beat.} 
\LX{toovaa} \HM{} \PH{tòòváà} \PS{n.~} \GE{ear flap; the visible part of the ear (pinna)} \OP{toovaare} \TP{toovaarɛɛ} 
\LX{tɔɔvaa} \HM{} \PH{tɔ̀ɔ̀váà} \PS{n.~} \GE{tobacco leaf} \OP{tɔɔvaare} 
\LX{toovoɔre} \HM{} \PH{tòòvʊ́ɔ́rɪ́} \PS{n.~} \GE{ear canal; ear piercing} \OP{toovɔ᷉ɛ} \TP{toovoɔrɛɛ} 
\LX{tɔɔwuoraa} \HM{} \PH{tɔ̀ɔ̀wúóráá} \PS{n.~} \GE{a shell used in gathering and collecting powdered tobacco (snuff)} \OP{tɔɔwuori} 
\LX{tɔɔzɔŋe} \HM{} \PH{tɔ̀ɔ̀zɔ̂ŋ} \PS{n.~} \GE{powdered tobacco; snuff} \TP{tɔɔzɔnnɛɛ} 
\LX{topɛge} \HM{} \PH{tòpɛ́gɪ́} \PS{n.~} \GE{boabab fruit shell} \OP{topɛgre} \TP{topɛgrɛɛ} 
\LX{toraza} \HM{} \PH{tʊ́rázà} \PS{n.~} \GE{trousers} \OP{torazare} \TP{torazarɛɛ} 
\LX{tore} \HM{} \PH{tʊ̀rɪ̀} \PS{v.~} \GE{to sprout, to break through the seed coat} \PD{torɛɛ, tora, tora, tora, toraa, - } 
\LX{tore} \HM{} \PH{tʊ̀rɪ̀} \PS{v.~} \GE{to close one's mouth tightly (usually in anger)} \PD{torɛɛ, tora, tora, tora, toraa, - } 
\LX{tɔre} \HM{} \PH{tɔ̀rɪ́} \PS{pronoun.~} \GE{self} \XV{N na de la a bondirii tte ko o N tɔre.} \XE{I will take the food to him myself.} 
\LX{tɔre} \HM{} \PH{tɔ̀rɪ̀} \PS{v.~} \GE{to share in pieces} \XV{A bondirii ba seŋe bone kyɛ te na de tɔre la a bibiiri.} \XE{We have very little food left but we shall share it in the little bits that we can.} \PD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\LX{tɔre} \HM{} \PH{tɔ́rɪ́} \PS{v.~} \SN \GE{to break easily e.g. weak twine} \XV{A miri poɔ̃ɛ la lɛnso ka O gaa ka o lene a tuobu ka o tɔre.} \XE{The rope was rotten and that was why when s/he attempted to tie up the luggage it gave way easily.} \SN \GE{to cut a small piece} \XV{N baŋ ka a nɛne ba seŋe bone kyɛ N na tɔre la fe᷉e ko fo.} \XE{I know the piece meat is small but I'll give you a little bit.} \SN \GE{die (ailing child)} \PD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\LX{tɔre} \HM{} \PH{tɔ̀rɪ̀} \PS{v.~} \GE{to cut a small piece} \XV{N baŋe ka a nɛne ba seŋe bone kyɛ N na tɔre la fe᷉e ko fo.} \XE{I know the piece meat is small but I'll give you a little bit.} \PD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\LX{tɔre} \HM{} \PH{tɔ́rɪ́} \PS{v.~} \GE{die e.g a person who has been sick for long} \PD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\LX{tɔre} \HM{} \PH{tɔ̀rɪ́} \PS{adj.~} \GE{valuable} \XV{O yeli ka ba zu la o libie kyɛ a ba e libi tɔre zaa.} \XE{She said they had stolen her money but it was not worth anything.} 
\LX{torebogi} \HM{} \PH{tʊ̀rɪ̀bógí} \PS{n.~} \GE{frontier, boundary} \OP{torebogri} \TP{torebogree} 
\LX{tori} \HM{} \PH{tórí} \PS{adv.~} \GE{continuously (doing one thing)} \XV{Ko᷉ɔ yoŋi ka a baala tori nyuuro.} \XE{The patient just goes on drinking water only.} 
\LX{tori} \HM{1} \PH{tórí} \PS{v.~} \SN \GE{to be straight} \SN \GE{to be right, to be correct} \PD{toree, toro, toro, toribo, toraa, toribo } 
\LX{tori} \HM{2} \PH{tórí} \PS{v.~} \GE{to speak with anger, to blame, to complain} \PD{toree, toro, toro, toribo, toraa, - } 
\LX{tori-buuluŋ} \HM{} \PH{tórí-búúlúŋ} \PS{n.~} \GE{porridge made from corn and boabab flour} 
\LX{toriko} \HM{} \PH{tòríkò} \PS{n.~} \GE{cart} \OP{torikori} \TP{torikoree} 
\LX{toriko} \HM{} \PH{tòríkò} \PS{n.~} \GE{large metal snare for trapping large animals} \OP{torikori} \TP{torikoree} 
\LX{torizeɛ} \HM{} \PH{tórízɪ́ɛ́} \PS{n.~} \GE{red beret} \OP{torizeere} \TP{torizerɛɛ} 
\LX{to-uri} \HM{} \PH{tò-ùrí} \PS{n.~} \GE{tiny spikes covering the fruit of boabab} \TP{to-uree} 
\LX{tozɔŋe} \HM{} \PH{tòzɔ́ŋ} \PS{n.~} \GE{boabab fruit powder} \TP{tozɔnnɛɛ} 
\LX{to᷉ɔ} \HM{} \PH{tʊ́᷉ɔ́᷉} \PS{v.~} \SN \GE{to subdue, to conquer} \XV{A bibiiri naŋ mɔ buo la to᷉ɔ o tɔ?} \XE{Which of boys overcame the other in the wrestling competition?.} \SN \GE{to be able, to be capable} \XV{N gaa la kyɛ N koŋ to᷉ɔ leɛ wa.} \XE{I'll go, but I'll not be able to come back.} \PD{to᷉ɔɛ, toɔnɔ, toɔnɔ, toɔnema, toɔnaa, - } 
\LX{tu} \HM{} \PH{tù} \PS{v.~} \GE{to follow a route, to take a route} \PD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\LX{tu} \HM{} \PH{tù} \PS{v.~} \GE{to recount, to narrate} \PD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\LX{tu} \HM{} \PH{tù} \PS{v.~} \GE{to bring back an attempted (female) divorcee} \PD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\LX{tu} \HM{} \PH{tú} \PS{v.~} \SN \GE{to dig} \XV{A sakuuri tu la bulee ba dabɔge poɔ.} \XE{The school has dug a well in their garden.} \SN \GE{to uproot} \XV{A koɔraa tu la waare paale o wasege.} \XE{The farmer has dug up yams and filled the yam barn.} \PD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\LX{tu ba} \HM{} \PH{tú bà} \PS{v.~} \GE{to take off quickly e.g. in running} \PD{tu baɛ, tuuro baara, tu baara, tu baareba, -, - } 
\LX{tu noɔre} \HM{} \PH{tù nʊ́ɔrɪ̀} \PS{v.~} \GE{to narrate, to give an account of} \XV{Ba yeli ka a bibiiri na naŋ zɔɔ taa ka kaŋa zaa maŋ tu o noɔre.} \XE{Each of the boys who fought has been asked to give an account of what led to the fight.} \PD{-, -, -, -, -, - } 
\LX{tu sori} \HM{} \PH{tù sórí} \PS{v.~} \GE{to be right} \XV{Te yɔɔ la a gama kyɛ ka dɔɔ wa wuo a zaa gaa ne, a lɛ ba tu sori.} \XE{We paid for the books but the man came and took them away; that is not right.} \PD{-, -, -, -, -, - } 
\LX{tuli} \HM{} \PH{tùlì} \PS{v.~} \GE{to make a mistake, to err, to be wrong} \PD{tulee, tullo, tullo, tulbo, tulaa, - } 
\LX{tulli} \HM{} \PH{tùllì} \PS{v.~} \SN \GE{to change (state), disguise} \XV{A bibiiri da tulli la ba menne a tigiri daare ka fo koŋ la baŋ ba.} \XE{The boys diguised themselves during the festival, so that you could not recognise them.} \SN \GE{to confuse} \XV{N bare la gyɛle ka a pɔge tulli ma kyɛ zo ne a bie.} \XE{I was so careless that the woman tricked me and escaped with the child.} \PD{tullee, tullo, tullo, tulibo, tullaa, - } 
\LX{tulli} \HM{} \PH{tùllì} \PS{v.~} \GE{to incriminate spiritually} \XV{O tulli la omeŋɛ lɛnso ka o bageboge dagoli ba la nyɛrɛ.} \XE{She has wronged the spirits and that is why she cannot see things in the spiritual world.} \PD{tullee, tullo, tullo, tulibo, tullaa, - } 
\LX{tulundaa} \HM{} \PH{túlúndàá} \PS{n.~} \GE{pestle} \OP{tulundaare} \TP{tunlundaarɛɛ} \VA{tulindaa, tundaa}
\LX{tumakuli} \HM{} \PH{tùmákúlí} \PS{n.~} \GE{burri; a type of small plant with thorny fruits} \SG{tumakuluu} \OP{tumakulli} \TP{tumakulee} 
\LX{tumo} \HM{} \PH{tùmó} \PS{n.~} \GE{one born with the feet coming first instead of the head} \TP{tumee} 
\LX{tumo} \HM{} \PH{tùmó} \PS{n.~} \GE{glutton} \TP{tumee} 
\LX{tumpaane} \HM{} \PH{túmpáánɪ́} \PS{n.~} \GE{large talking drum; a drum that gives messages} \OP{tumpaama} \TP{tumpaanɛɛ} \VA{tumpanne}
\LX{tumpaane} \HM{} \PH{túmpáánɪ́} \PS{n.~} \GE{slabs of bean flour boiled in leaves} \OP{tumpaama} \TP{tumpaanɛɛ} \VA{tumpanne}
\LX{tuŋi} \HM{} \PH{tùŋ} \PS{v.~} \GE{to thread (beads)} \PD{tuŋee, tuno, tuno, tunimo, tunaa, tunimo } 
\LX{tuŋi} \HM{} \PH{tùŋ} \PS{v.~} \SN \GE{to dip e.g. hand in pocket} \SN \GE{to insert a finger into the vagina (defile, rape)} \PD{tuŋee, tuno, tuno, tunimo, tunaa, tunimo } 
\LX{tuŋi} \HM{} \PH{tûŋ} \PS{v.~} \SN \GE{to stumble about to fall, to totter} \SN \GE{to err, to make a mistake} \PD{tuŋee, tuŋino, tuŋino, tuŋinimo, tuŋinaa, tuŋinimo } \VA{tummu (of -bu form)}
\LX{tuŋi} \HM{} \PH{túŋ} \PS{v.~} \GE{to overgrow e.g. hair, grass of lawn} \PD{-, -, -, -, -, - } 
\LX{tuntundaana} \HM{} \PH{túntúndááná} \PS{n.~} \GE{mosquito net} \OP{tuntundaane} \TP{tuntundaanɛɛ} 
\LX{tunturi} \HM{} \PH{túntúrì} \PS{n.~} \GE{a type of very large hog with powerful claws and tail for digging} \OP{tuntue} \TP{tuturee} 
\LX{tuo} \HM{} \PH{tùò} \PS{v.~} \SN \GE{to carry e.g. on head} \SN \GE{to swell e.g. blister} \PD{-, tuoro, tuoro, tuoribo, -, - } 
\LX{tuo} \HM{} \PH{tùó} \PS{n.~} \SN \GE{baobab tree} \SN \GE{baobab fruit} \OP{tori} \TP{toree} 
\LX{tuo} \HM{} \PH{túó} \PS{n.~} \SN \GE{suffering} \XV{Tuo yoŋ ka zɔɔre taa.} \XE{There is only suffering in war.} \SN \GE{difficulty} \XV{A daŋmaa ŋa ŋmaabo tuo la so ka o ba wa.} \XE{The difficulty involved in cutting this log has disuaded him from coming.} \OP{tuuri} \TP{tuuree} 
\LX{tuo} \HM{} \PH{túó} \PS{adj.~} \SN \GE{bitter e.g. quinine} \SN \GE{hot e.g. pepper} \OP{tuuri} \TP{tuuree} 
\LX{tuobu} \HM{} \PH{tùòbú} \PS{n.~} \SN \GE{load} \SN \GE{burden} \SN \GE{weight} 
\LX{tuori} \HM{} \PH{tùòrì} \PS{v.~} \SN \GE{to meet face to face} \SN \GE{to welcome} \PD{tuoree, tuoro, tuoro, tuoribo, tuoraa, tuoribo } 
\LX{tuosoba} \HM{} \PH{túósʊ́bá} \PS{n.~} \GE{person who does not want to make offers to any other person} 
\LX{turi} \HM{} \PH{túrì} \PS{n.~} \GE{thousand} \SG{turree} \TP{turee} 
\LX{turi} \HM{} \PH{tùrì} \PS{v.~} \SN \GE{to give from above or below} \XV{A pɔge wuo a kyi peɛ zu de turi o serɛ.} \XE{The woman gathered the millet from the top of the roof and passed it down to her husband.} \SN \GE{to offer one's hand} \PD{turee, turo, turo, turibo, turaa, - } 
\LX{tutɔ} \HM{} \PH{tútɔ́} \PS{n.~} \GE{companion} \OP{tutɔmine} 
\LX{tuu} \HM{} \PH{túú} \PS{n.~} \SN \GE{thicket, forest, wood} \SN \GE{raised ground (usually on a farm)} \SN \GE{eagle} \OP{tuuri} \TP{tuuree} \VA{togroo}
\LX{tuu} \HM{} \PH{túú} \PS{n.~} \SN \GE{thicket, forest, wood} \SN \GE{raised ground (usually on a farm)} \OP{tuuri} \TP{tuuree} \VA{togroo}
\LX{tuu} \HM{} \PH{túú} \PS{n.~} \GE{eagle} \OP{tuuri} \TP{tuuree} 
\LX{tuuli} \HM{} \PH{tùùlì} \PS{v.~} \SN \GE{to trace e.g. line} \SN \GE{to use a finger to apply e.g. ointment} \SN \GE{to go about, to roam around} \SN \GE{to carried away by water} \PD{tuulee, tuulo, tuulo, tuulbo, tuulaa, - } 
\LX{tuuli} \HM{} \PH{túúlì} \PS{v.~} \SN \GE{to give to drink e.g. a child} \SN \GE{not deeply emersed (arrow hitting game)} \PD{tuulee, tuulo, tuulo, tuulbo, tuulaa, - } 
\LX{tuuree} \HM{} \PH{túúréé} \PS{n.~} \GE{Europe or America} 
\LX{tuuri} \HM{} \PH{tùùrì} \PS{v.~} \GE{to pick out, to sort out, to separate} \PD{tuuree, tuuro, tuuro, tuuribo, tuuraa, - } 
\LX{tɛrewɛrɛ} \HM{} \PH{tɛ́rɪ́wɛ́rɛ́} \PS{adv.~} \GE{standing with legs bent backwards} 
\LX{ugi} \HM{} \PH{úgì} \PS{v.~} \GE{to expand e.g. cooked rice} \PD{ugee, ugro, ugro, ugribo, ugraa, - } 
\LX{ulee} \HM{} \PH{ùléè} \PS{n.~} \GE{tree branch} \OP{ulli} \TP{ullee} 
\LX{uli} \HM{} \PH{úlí} \PS{n.~} \GE{fog, mist} \OP{ulli} \TP{ullee} 
\LX{uli} \HM{} \PH{ùlì} \PS{v.~} \SN \GE{to be dusty} \SN \GE{to become moldy e.g. food} \XV{Ka boroboroŋ wa kɔɔre o maŋ uli la.} \XE{When bread is kept for long it becomes mouldy.} \PD{ulee, ullo, ullo, ulibo, ullaa, - } 
\LX{ulimo} \HM{} \PH{úlmó} \PS{adj.~} \GE{grey} 
\LX{unni} \HM{} \PH{ùnnì} \PS{v.~} \SN \GE{unburying something buried} \XV{Polisiri yeli ka ba ba baŋ a dɔɔ ku᷉u᷉, pare a gaa te unni a ku᷉u᷉.} \XE{The police said they did not know the cause of the man's death, so they went and exhumed the corpse.} \SN \GE{to bring back past memories} \XV{Ka Dɛre ne o pɔgeŋ wa zɔɔrɔ zaa a pɔge maŋ unni la yelikorɔ.} \XE{Each time Dɛre and his wife quarrel, the wife will always recount past events.} \PD{unnee, unno, unno, unimo, unnaa, - } 
\LX{unni} \HM{} \PH{únnì} \PS{v.~} \GE{to remind} \XV{N deɛmɛ za᷉a᷉ wa unni ma la a koɔbo.} \XE{My in-law reminded me about the day of the group farming.} \PD{unnee, unno, unno, unimo, unnaa, - } 
\LX{uoni} \HM{} \PH{úònì} \PS{n.~} \GE{dry season} \OP{uomo} \TP{uonee} 
\LX{uori} \HM{} \PH{úórì} \PS{v.~} \GE{to warm oneself (by fire, sun)} \PD{uoree, uoro, uoro, uoribo, uoraa, - } 
\LX{uori} \HM{} \PH{úórì} \PS{v.~} \GE{to mourn} \PD{uoree, uoro, uoro, uoribo, uoraa, - } 
\LX{uori} \HM{} \PH{úórì} \PS{v.~} \GE{to peer (shading one's face with the hand)} \PD{uoree, uoro, uoro, uoribo, uoraa, - } 
\LX{uori} \HM{} \PH{ùòrì} \PS{v.~} \SN \GE{to bellow (bull)} \SN \GE{to groan; a sound uttered in pain or grief} \PD{uoree, uoro, uoro, uoribo, uoraa, - } 
\LX{uri} \HM{} \PH{ùrí} \PS{n.~} \SN \GE{itchy stuff} \SN \GE{chaff} \OP{n/a} \TP{urree} 
\LX{uri} \HM{} \PH{ùrì} \PS{v.~} \SN \GE{to take unaware} \SN \GE{to be shocked} \XV{Maa ne a baala naŋ yeli yɛlɛ ba kɔɔre, o ku᷉u᷉ uri ma la.} \XE{I spoke with the patient not long ago; I'm shocked at his sudden death.} \PD{uree, uro, uro, uribo, uraa, - } 
\LX{uri} \HM{} \PH{úrì} \PS{v.~} \GE{to make burn e.g fire} \PD{-, -, -, -, -, - } 
\LX{uri} \HM{} \PH{ùrí} \PS{n.~} \GE{hair-like projections on the straw of the grass family and other plants that cause itchiness of the skin} 
\LX{uri} \HM{} \PH{úrí} \PS{v.~} \GE{to scrub a very dirty object e.g. cooking pot} \PD{uree, uro, uro, uribo, uraa, - } 
\LX{uuni} \HM{} \PH{úúnì} \PS{v.~} \GE{to crouch, to hide} \XV{Ba digi la a bie ka o zo te uuni tuu poɔ.} \XE{As they chased the boy, he ran and hid in a thicket.} \PD{uunee, uuno, uuno, uunimo, uunaa, - } 
\LX{uuni} \HM{} \PH{úúnì} \PS{v.~} \GE{to incubate, to brood, to sit on (a fowl on its eggs)} \XV{Noɔ maŋ uuni la daare ata kyɛ wɛge.} \XE{A fowl sits on its eggs before they hatch.} \PD{uunee, uuno, uuno, uunimo, uunaa, - } 
\LX{uuni} \HM{} \PH{úúnì} \PS{v.~} \GE{to form pus e.g. a swelling} \XV{Goɔ daare kyɔge la n nubiri ka o mɔre eŋ kõɔ ka ba wɛre.} \XE{A thorn pierced my finger and it got swollen, formed pus, and was operated upon.} \PD{uunee, uuno, uuno, uunimo, uunaa, - } 
\LX{uuri} \HM{} \PH{úúrí} \PS{v.~} \GE{to hoot, to call out e.g. lost person} \PD{uuree, uuro, uuro, uuribo, uuraa, - } 
\LX{uuri} \HM{} \PH{ùùrì} \PS{v.~} \SN \GE{to smoke out} \SN \GE{to warm a place with smoke} \PD{uuree, uuro, uuro, uuribo, uuraa, - } 
\LX{uuruŋ} \HM{} \PH{úúrúŋ} \PS{n.~} \GE{dust} \TP{uuurunnee} 
\LX{u᷉u} \HM{} \PH{ú᷉ú᷉} \PS{v.~} \GE{to close tightly (fist, eyes)} \PD{u᷉ue, uuno, uuno, uuimo, uunaa, - } 
\LX{u᷈u} \HM{} \PH{ù᷉ù᷈} \PS{v.~} \GE{to bury} \PD{u᷉ue, uuno, uuno, uuimo, uunaa, - } 
\LX{va} \HM{} \PH{và} \PS{v.~} \SN \GE{to whip} \SN \GE{to thresh (millet)} \PD{vaɛ, -, vaara, vareba, vaaraa, - } 
\LX{va} \HM{} \PH{vá} \PS{v.~} \GE{used with ‘-de’(va-de); snatch; to run directly and suddenly at someone seizing him} \PD{-, -, -, -, -, - } 
\LX{va} \HM{} \PH{và} \PS{v.~} \GE{to graze a target; miss narrowly} \PD{vaɛ, -, vaara, vareba, vaaraa, - } 
\LX{va} \HM{} \PH{vá} \PS{v.~} \GE{to rush, to dash} \XV{A dɔbɔ naŋ ŋmeɛrɛ la ka kaŋa wa va de o tɔ gaa lɔɔ.} \XE{As the men argued, one suddenly rushed on the other and threw him down.} \PD{vaɛ, vaara, vaara, vareba, vaaraa, - } 
\LX{vaa} \HM{} \PH{váà} \PS{n.~} \GE{leaf} \SG{vabo} \OP{vaare} \TP{vaarɛɛ} 
\LX{vaa} \HM{} \PH{vàà} \PS{v.~} \GE{collect or take everything from a container} \PD{vaaɛ, vaara, vaara, vaareba, vaaraa, - } 
\LX{vaa} \HM{} \PH{vàà} \PS{v.~} \GE{to draw e.g. bowstring} \PD{vaaɛ, vaara, vaara, vaareba, vaaraa, - } 
\LX{vaa} \HM{} \PH{vàà} \PS{v.~} \GE{to lift an opponent's leg (wrestling)} \PD{vaaɛ, vaara, vaara, vaareba, vaaraa, - } 
\LX{vaagaviige} \HM{} \PH{vààgàvíígè} \PS{v.~} \GE{I command you (to die)} \PD{-, -, -, -, -, - } 
\LX{vaa-iri} \HM{} \PH{vàà-ìrì} \PS{v.~} \GE{to stand up quickly} \PD{vaa-iree, vaara-ire, vaa-ire, vaa-iribe, -, - } 
\LX{vaale} \HM{} \PH{vààlɪ̀} \PS{v.~} \GE{to gather e.g. rubbish, garbage} \PD{-, -, -, -, -, - } 
\LX{vaalee} \HM{} \PH{váálɪ́ɪ́} \PS{n.~} \GE{speck} \SG{vaaloo} \OP{vaala} \TP{vaalɛɛ} \VA{vaale}
\LX{vaane} \HM{} \PH{vàànɪ̀} \PS{v.~} \SN \GE{to shrink, contract} \SN \GE{to wrinkle} \PD{vaanɛɛ, vaana, vaana, vaaneba, vaanaa, - } 
\LX{vaane} \HM{} \PH{váánɪ́} \PS{adj.~} \GE{shrunken} 
\LX{vaali} \HM{} \PH{vạ̀ạ̀lì} \PS{v.~} \GE{move fast e.g in standing up} \PD{-, -, -, -, -, - } 
\LX{vaare} \HM{} \PH{vààrɪ̀} \PS{v.~} \GE{to wrap oneself, usually with a cloth} \PD{vaarɛɛ, vaara, vaara, vaareba, vaaraa, - } 
\LX{vaare} \HM{} \PH{váárɪ̀} \PS{v.~} \GE{to tug, to pull strongly and suddenly towards oneself} \XV{A bie da nyɔge la a pedaa miri kyɛ ka a peroo vaare o lɔɔ.} \XE{The boy held onto the rope around the ram's neck, but it tugged him to the ground.} \PD{vaarɛɛ, vaara, vaara, vaareba, vaaraa, - } 
\LX{vaare} \HM{} \PH{váárɪ̀} \PS{v.~} \GE{to rinse} \XV{Fanfane fuuri naŋ be la laare poɔ a seŋ ka vaare.} \XE{There are still soapsuds in the bowls, so they need to be rinsed.} \PD{vaarɛɛ, vaara, vaara, vaareba, vaaraa, - } 
\LX{vãee} \HM{} \PH{vã̀ɪ̃́ɪ̃̀} \PS{adv.~} \GE{aromatically} 
\LX{vabɔre} \HM{} \PH{vábɔ́rɪ̀} \PS{n.~} \GE{spring} \TP{vabɔrrɛɛ} 
\LX{vage} \HM{} \PH{vágɪ́} \PS{v.~} \GE{onomatopoeia; sounding like a bird pecking at a soft object} \PD{vagɛɛ, vagra, vagra, vagreba, vagraa, - } 
\LX{vage} \HM{} \PH{vàgɪ̀} \PS{adv.~} \GE{sound of pecking} 
\LX{vagere} \HM{} \PH{vàgɪ̀rɪ̀} \PS{adv.~} \GE{pecking} 
\LX{vakogilo} \HM{} \PH{vákóglo} \PS{n.~} \GE{dried leaves (that once worn by women)} \SG{vakogiluu} 
\LX{vakuoŋaa} \HM{} \PH{vákúóŋáá} \PS{n.~} \GE{dry leaf} \OP{vakuoni} 
\LX{vakuoŋaa} \HM{} \PH{vákúóŋáá} \PS{n.~} \GE{bad luck} \OP{vakuoni} 
\LX{valenvo᷉ɔ} \HM{} \PH{válɪ́nvʊ́᷉ɔ̀᷉} \PS{n.~} \GE{mason wasp, potter wasp} \SG{valenvoɔnoo} \OP{valenvoɔne} \TP{valenvoɔnɛɛ} \VA{valenvõa}
\LX{valere} \HM{} \PH{vàlɪ́rɪ́} \PS{v.~} \GE{to intertwine e.g. weaving} \PD{-, -, -, -, -, - } 
\LX{valimmu} \HM{} \PH{vàlímmú} \PS{n.~} \GE{a type of small tree with soft large leaves which women used to wear to cover parts of their private parts} 
\LX{valimmu} \HM{} \PH{vàlǐmmú} \PS{n.~} \GE{a type of broad-leafed plant} \OP{valinni} \TP{valinnee} 
\LX{valinvaa} \HM{} \PH{vàlìnv́áà} \PS{n.~} \SN \GE{leaf of the palm tree} \SN \GE{the leaf of the valimmu tree} \OP{valinvaare} \TP{valinvaarɛɛ} 
\LX{valinvuuluŋ} \HM{} \PH{v̀àlìnvúúlúŋ} \PS{n.~} \GE{soot} \OP{valinvuulo/valinvuuli} \TP{valinvuulee} \VA{valanvuuluŋ, vuuluŋ}
\LX{valogema} \HM{} \PH{válʊ́gɪ́mà} \PS{n.~} \GE{green, fresh, and tender leaves} 
\LX{vama} \HM{} \PH{vàmà} \PS{adv.~} \GE{effortlessly} 
\LX{vamaaroŋ} \HM{} \PH{vámáárʊ́ŋ} \PS{n.~} \GE{green leaf} \OP{vamaara} \TP{vamaarɛɛ} 
\LX{vanne} \HM{} \PH{vánnɪ̀} \PS{v.~} \GE{to stretch, to grow longer} \PD{vannɛɛ, vanna, vanna, vannaa, -, - } 
\LX{vapeɛlegboro} \HM{} \PH{vápíɛ́lgbóró} \PS{n.~} \GE{a type of tree with medicinal properties} 
\LX{vapɛge} \HM{} \PH{vápɛ́gɪ́} \PS{n.~} \GE{fallen leaf} \OP{vapɛgre} \TP{vapgrɛɛ} 
\LX{vapielee} \HM{} \PH{vápìèlèè} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{vapieli} 
\LX{vapɔrɔ} \HM{} \PH{vápɔ̀rɔ̀} \PS{n.~} \GE{fresh green leaves} \TP{vapɔrrɛɛ} 
\LX{vare} \HM{} \PH{vàrɪ̀} \PS{v.~} \GE{to wrinkle} \PD{varɛɛ, vara, vara, vareba, varaa, vareba } 
\LX{vare} \HM{} \PH{vàrɪ̀} \PS{v.~} \GE{to make laundry not well washed e.g. clothes} \PD{varɛɛ, vara, vara, vareba, varaa, vareba } 
\LX{vare} \HM{} \PH{vàrɪ̀} \PS{v.~} \GE{to make the layout of a mound} \PD{varɛɛ, vara, vara, vareba, varaa, vareba } 
\LX{vare} \HM{} \PH{várɪ́} \PS{v.~} \GE{to jump, to leap, to hop} \PD{varoo, vara, vara, varibo, varaa, variba } 
\LX{vare} \HM{} \PH{várɪ́} \PS{v.~} \GE{to castrate} \PD{varoo, vara, vara, varibo, varaa, variba } 
\LX{vare niŋe} \HM{} \PH{vàrɪ̀ níŋé} \PS{v.~} \GE{make a face; usually in disgust or anger} \PD{-, -, -, -, -, - } 
\LX{varekpara} \HM{} \PH{vàrìkpàrà} \PS{adv.~} \GE{long and loose e.g. twine} 
\LX{varekpenɛ} \HM{} \PH{várɪ́kpɪ́nɛ́} \PS{n.~} \GE{thick paste from beans} \SG{varekpenoo} \TP{varekpennɛɛ} 
\LX{varema} \HM{} \PH{várɪ́má} \PS{adj.~} \SN \GE{tough e.g. meat, rope} \SN \GE{difficult} \XV{A pɔgelee na nɔŋe la laare kyɛ o boɔbo yɛlɛ varema.} \XE{That girl likes smiling but she is very difficult when it comes to dating her.} 
\LX{varevare} \HM{} \PH{vàrɪ̀várɪ́} \PS{v.~} \GE{to heckle} \PD{varevarɛɛ, varevara, varevara, varevareba, varevaraa, - } 
\LX{varevare} \HM{} \PH{vàrɪ̀várɪ́} \PS{v.~} \GE{to pull along} \PD{varevarɛɛ, varevara, varevara, varevareba, varevaraa, - } 
\LX{varoo} \HM{} \PH{várʊ̀ʊ̀} \PS{adj.~} \GE{tough e.g. meat} \VA{varema}
\LX{va᷉ao} \HM{} \PH{và᷉à᷈ʊ᷉} \PS{adv.~} \GE{courageously} 
\LX{ve} \HM{} \PH{vɪ́} \PS{v.~} \GE{to knit} \PD{veɛɛ, verɛ, verɛ, verebɛ, veraa, - } 
\LX{vebo} \HM{} \PH{vɪ̀bʊ̀} \PS{adj.~} \GE{striking directly e.g. arrow} 
\LX{vedeɛ} \HM{} \PH{vɪ̀dɪ̀ɛ́} \PS{n.~} \GE{sugarcane} \OP{vederre} \TP{vedeerɛɛ} 
\LX{veɛle} \HM{} \PH{vɪ̀ɛ̀lɪ̀} \PS{v.~} \GE{to be beautiful} \PD{veɛlɛɛ, veɛlɛ, veɛlɛ, veɛle, -, - } 
\LX{veɛlɛ} \HM{} \PH{vɪ̀ɛ̀lɛ́} \PS{adj.~} \GE{beautiful} 
\LX{veɛlɛyeŋe} \HM{} \PH{vɪ̀ɛ̀lɛ́yéŋé} \PS{n.~} \GE{one with physical beauty only} 
\LX{veɛloŋ} \HM{} \PH{vɪ̀ɛ́lʊ́ŋ} \PS{n.~} \SN \GE{beauty} \SN \GE{goodness} \SN \GE{light} 
\LX{veene} \HM{} \PH{vɪ̀ɪ̀nɪ̀} \PS{n.~} \GE{a type of small bird with very lean legs (short form of daveene)} \OP{veemɛ} \TP{veenɛɛ} \VA{daveene}
\LX{veene} \HM{} \PH{vɪ́ɪ́nɪ́} \PS{v.~} \GE{to tie tightly} \PD{veenɛɛ, veenɛ, veenɛ, veenemɛ, veenaa, - } 
\LX{veɛre} \HM{} \PH{vɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to push aside growth e.g in searching for something in an overgrown bush} \PD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa, - } 
\LX{veɛre} \HM{} \PH{vɪ́ɛ́rɪ̀} \PS{v.~} \GE{to fly low and slowly} \PD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa, - } 
\LX{veɛre} \HM{} \PH{vɪ́ɛ́rɪ̀} \PS{v.~} \GE{to evade e.g. a verbal attack} \PD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa, - } 
\LX{veɛrenyɔge} \HM{} \PH{vɪ̀ɛ̀rɪ̀ɲɔ́gɪ́} \PS{n.~} \GE{rodent} 
\LX{velaa} \HM{} \PH{vɪ̀làà} \PS{adj.~} \SN \GE{good, nice} \SN \GE{beautiful} \OP{veɛle} 
\LX{vɛle} \HM{} \PH{vɛ̀lɪ̀} \PS{v.~} \GE{to become clear e.g. day, mind} \PD{vɛlɛɛ, vɛllɛ, vɛllɛ, vɛlebɛ, vɛllaa, - } 
\LX{vɛlevɛle} \HM{} \PH{vɛ̀lvɛ̀l} \PS{adj.~} \GE{very long and and wavy} 
\LX{vɛmm} \HM{} \PH{vɛ́mm} \PS{adv.~} \GE{smoothly} 
\LX{vell} \HM{} \PH{vɪ̀ll} \PS{adj.~} \GE{full to the brim} 
\LX{velle} \HM{} \PH{vɪ̀llɪ̀} \PS{v.~} \GE{to hold the legs tightly in such a way as to make the person lose balance} \XV{A pɔge velle la a dɔɔ gbɛɛ ka o le.} \XE{The woman held the man’s legs so tightly that he lost balance and fell.} \PD{vellɛɛ, vellɛ, vellɛ, velbɛ, vellaa, - } 
\LX{velɔɔ} \HM{} \PH{vɪ̀lɔ̀ɔ̀} \PS{adj.~} \GE{turning instantly} 
\LX{vɛŋe} \HM{} \PH{vɛ́ŋ} \PS{v.~} \SN \GE{to leave} \SN \GE{to let go} \SN \GE{to stop} \PD{vɛŋɛɛ, vɛnɛ, vɛnɛ, vɛnemɛ, vɛnaa, - } 
\LX{vɛnɛvɛnɛ} \HM{} \PH{vɛ̀nɛ̀vɛ̀nɛ̀} \PS{adj.~} \SN \GE{sticky e.g. honey} \SN \GE{oily} 
\LX{vɛnlɛmm} \HM{} \PH{vɛ́nlɛ́mm} \PS{adj.~} \GE{slim and beautiful} 
\LX{venenne} \HM{} \PH{vɪ̀nɪ̀nnɪ̀} \PS{adj.~} \GE{elastic} 
\LX{venne} \HM{} \PH{vɪ́nnɪ́} \PS{v.~} \GE{to throw off (wrestling)} \PD{vennɛɛ, vennɛ, vennɛ, venemɛ, vennaa, - } 
\LX{venoo} \HM{} \PH{vɪ́nʊ̀ʊ̀} \PS{n.~} \GE{wine} 
\LX{venɔɔ} \HM{} \PH{vɪ̀nɔ̀ɔ̀} \PS{adj.~} \GE{thick and sticky e.g. grease} 
\LX{venveŋ} \HM{} \PH{vɪ́nvɪ́ŋ} \PS{n.~} \GE{climax} 
\LX{venveŋ} \HM{} \PH{vɪ́nvɪ́ŋ} \PS{adv.~} \GE{at the maximum} \XV{A daa naŋ da w ate kpɛ venveŋ saŋa la ka ba da moɔle a zuyɔɔ yɛlɛ.} \XE{It was at the climax of the market activities that they announced the payment of taxes.} 
\LX{vere} \HM{} \PH{vɪ̀rɪ̀} \PS{v.~} \GE{to shoot with a catapult} \PD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ } 
\LX{vere} \HM{} \PH{vɪ̀rɪ̀} \PS{v.~} \SN \GE{to sprain} \SN \GE{to dislocate (joint)} \PD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ } 
\LX{vere} \HM{} \PH{vɪ̀rɪ̀} \PS{v.~} \GE{to make an effort, to strain} \PD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ } 
\LX{vere} \HM{} \PH{vɪ́rɪ́} \PS{v.~} \GE{to fence e.g. garden} \PD{verɛɛ, verɛ, verɛ, verebɛ, veraa, - } 
\LX{vɛre} \HM{} \PH{vɛ̀rɪ̀} \PS{v.~} \GE{to prepare mortar} \PD{vɛrɛɛ, vɛrɛ, vɛrɛ, vɛrebɛ, vɛraa, - } 
\LX{vɛre} \HM{} \PH{vɛ́rɪ́} \PS{v.~} \GE{to scatter thick liquid e.g. mortar} \PD{vɛrɛɛ, vɛrɛ, vɛrɛ, vɛrebɛ, vɛraa, - } 
\LX{vɛre} \HM{} \PH{vɛ̀rɪ̀} \PS{v.~} \GE{mix water with soil e.g mortar} \PD{-, -, -, -, -, - } 
\LX{vɛre} \HM{} \PH{vɛ́rɪ́} \PS{n.~} \GE{stony farmland} \OP{vɛrɛɛ} 
\LX{vɛrɛɛ} \HM{} \PH{vɛ̀rɛ̀ɛ̀} \PS{adj.~} \GE{lying low e.g. dew-laden grass} 
\LX{verɛnverɛŋ} \HM{} \PH{vɪ̀rɛ̀nvɪ́rɛ́ŋ} \PS{adj.~} \GE{found everywhere} 
\LX{verevere} \HM{} \PH{vɪ̀rɪ̀vɪ́rɪ́} \PS{v.~} \GE{to pull very hard on e.g. tug of war} \PD{vereverɛɛ, vereverɛ, vereverɛ, vereverebɛ, vereveraa, - } 
\LX{verɛverɛ} \HM{} \PH{vɪ̀rɛ̀vɪ̀rɛ̀} \PS{adj.~} \GE{scattered all over} 
\LX{vɛvɛ} \HM{} \PH{vɛ́vɛ́} \PS{adj.~} \GE{glidingly e.g. a bird} 
\LX{vɛvɛre} \HM{} \PH{vɛ̀vɛ̀rɪ́} \PS{n.~} \GE{mud, marsh} \TP{vɛvɛrɛɛ} 
\LX{vɛ᷉ɛ} \HM{} \PH{vɛ̀᷉ɛ̀᷉} \PS{adv.~} \GE{less sunny, dusky} 
\LX{vi} \HM{} \PH{vî} \PS{n.~} \GE{shame, disgrace} \OP{viri} \TP{viree} 
\LX{vigi} \HM{} \PH{vígì} \PS{n.~} \GE{owl} \OP{vigiri} \TP{vigree} \VA{vig/vib}
\LX{vigi} \HM{} \PH{vígì} \PS{v.~} \GE{to throw out in a whole (solid waste)} \PD{vigee, vigre, vigre, vigribe, vigraa, - } 
\LX{viiri} \HM{} \PH{vììrì} \PS{v.~} \SN \GE{to turn around, to rotate} \SN \GE{to visit a place repeatedly} \PD{viiree, viire, viire, viiribe, viiraa, viiribe } 
\LX{vilenva} \HM{} \PH{vílénvà} \PS{n.~} \GE{going out of sight} \XV{A karema yeli ka o erɛ vilenva ka a karembiiri yaarɛɛ.} \XE{No sooner had the teacher left than the students dispersed.} 
\LX{vilenkyimm} \HM{} \PH{vílénkyímm} \PS{adj.~} \GE{twisted} 
\LX{vilenvileŋ} \HM{} \PH{vílénvíléŋ} \PS{adv.~} \GE{round, circular} 
\LX{vili} \HM{} \PH{vílí} \PS{v.~} \GE{to entwine, to tie, to wrap around} \PD{vilee, vile, vile, vilbe, vilaqa, - } 
\LX{vilime} \HM{} \PH{vílímé} \PS{adj.~} \GE{circular} \VA{vilenvileŋ}
\LX{virenvireŋ} \HM{} \PH{vírénvíréŋ} \PS{adv.~} \GE{totally e.g. refuse} 
\LX{viretaŋgbelli} \HM{} \PH{vìrèntáŋgbéllí} \PS{expression.~} \GE{your first and last} 
\LX{vitaŋkpulee} \HM{} \PH{vítáŋkpúléé} \PS{n.~} \GE{a type of small owl} \OP{vitaŋkpulli} 
\LX{vo} \HM{} \PH{vò} \PS{v.~} \GE{to make less crowded} \PD{voe, voro, voro, -, -, - } 
\LX{volavola} \HM{} \PH{vʋ̀làvʋ̀là} \PS{adv.~} \GE{slimly} \VA{volɔvolɔ}
\LX{volɔvolɔ} \HM{} \PH{vʋ̀lɔ̀vʋ̀lɔ̀} \PS{adv.~} \GE{slimly} \VA{volavola}
\LX{vorikporo} \HM{} \PH{vóríkpóró} \PS{adv.~} \GE{watery manner} 
\LX{vɔ} \HM{} \PH{vɔ́} \PS{v.~} \GE{to perform a type of ritual for a dead male} \PD{vɔɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\LX{vɔgaa} \HM{} \PH{vɔ̀gáá} \PS{n.~} \GE{silk cotton tree} \OP{vɔgre} \TP{vɔgrɛɛ} 
\LX{vɔge} \HM{} \PH{vɔ̀gɪ̀} \PS{v.~} \SN \GE{to take off} \SN \GE{to uncover} \PD{vɔgɛɛ, vɔgrɔ, vɔgrɔ, vɔgreba, vɔgraa, - } 
\LX{vɔge} \HM{} \PH{vɔ̀gɪ̀} \PS{v.~} \SN \GE{to scald} \SN \GE{to get burnt from hot substance} \PD{vɔgɛɛ, vɔgrɔ, vɔgrɔ, vɔgreba, vɔgraa, - } 
\LX{vogɛɛ} \HM{} \PH{vʊ̀gɛ̀ɛ̀} \PS{adv.~} \GE{voluminous e.g. pair of trousers} 
\LX{vɔgele} \HM{} \PH{vɔ̀glɪ̀} \PS{v.~} \SN \GE{to cover} \SN \GE{to put on top e.g. hat} \SN \GE{to overturn e.g. container} \PD{vɔglɛɛ, vɔglɔ, vɔglɔ, vɔgleba, vɔglaa, vɔgleba } 
\LX{vɔga} \HM{} \PH{vɔ́gà} \PS{n.~} \GE{track, trail} \VA{vɔgɔ}
\LX{vɔgɔ} \HM{} \PH{vɔ́gɔ̀} \PS{n.~} \GE{track, trail} \OP{vɔgre} \TP{vɔgrɛɛ} \VA{vɔga}
\LX{vokyɔ} \HM{} \PH{vʊ̀ʧɔ̀} \PS{n.~} \SN \GE{a type of tree with long fruit} \SN \GE{the fruit of this tree (used in kneading)} \OP{vokyɔre} \TP{vokyɔrɛɛ} 
\LX{vɔlaa} \HM{} \PH{vɔ́láà} \PS{n.~} \GE{great desire for} 
\LX{vɔle} \HM{} \PH{vɔ́lɪ̀} \PS{n.~} \GE{great desire for meat; carnivorousness} \OP{vɔlɛɛ} \TP{vɔllɛɛ} 
\LX{vɔle} \HM{} \PH{vɔ́lɪ́} \PS{v.~} \GE{to swallow} \PD{vɔlɛɛ, vɔlɔ, vɔlɔ, vɔlebɔ, vɔlaa, vɔlebɔ } 
\LX{voŋe} \HM{} \PH{vʊ́ŋ} \PS{v.~} \GE{to stay put and refuse to move} \PD{voŋɛɛ, vonɔ, vonɔ, vonema, vonaa, - } 
\LX{vonne} \HM{} \PH{vʊ̀nnɪ̀} \PS{v.~} \GE{to throw far away} \PD{vonnɛɛ, vonna, vonna, vonema, vonnaa, - } 
\LX{voɔ} \HM{} \PH{vʊ́ɔ̀} \PS{v.~} \GE{to pull out, to uproot} \PD{voɔɛ, voɔrɔ, voɔrɔ, voɔreba, vɔraa, - } 
\LX{voɔbale} \HM{} \PH{vʊ́ɔ́bàlɪ̀} \PS{n.~} \GE{arrow with many hooks} 
\LX{voɔloŋ} \HM{} \PH{vʊ̀ɔ̀lʊ́ŋ} \PS{n.~} \GE{slimy sap (from a type of tree) used to settle sediments in liquids} 
\LX{voɔloŋ} \HM{} \PH{vʊ̀ɔ̀lʊ́ŋ} \PS{adj.~} \GE{slimy} 
\LX{voone} \HM{} \PH{vʊ̀ʊ̀nɪ̀} \PS{v.~} \GE{to sound, to vibrate} \PD{voonɛɛ, voona/voonɔ, voona/voonɔ, -, -, - } 
\LX{vooraa} \HM{} \PH{vʊ́ʊ́ráá} \PS{n.~} \GE{a type of lumpy paste made from bean flour} \OP{voore} 
\LX{voore} \HM{} \PH{vʊ̀ʊ̀rɪ̀} \PS{v.~} \GE{echo, resound (noise)} \PD{voorɛɛ, voora, voora, vooreba, vooraa, - } 
\LX{voore} \HM{} \PH{vʊ́ʊ́rɪ́} \PS{v.~} \SN \GE{to breathe, to take a breath} \SN \GE{to be relieved, to be free, to rest} \XV{A saama kulee la ka N naa voore.} \XE{The guets have gone and now I'm relieved.} \PD{voorɛɛ, voora, voora, vooreba, vooraa, vooreba } 
\LX{voraa} \HM{} \PH{vʋ́ráá} \PS{n.~} \GE{a hand drill; see kyaleere} 
\LX{vorovoro} \HM{} \PH{vòròvòrò} \PS{adj.~} \GE{watery} 
\LX{voɔre} \HM{} \PH{vʊ́ɔ́rɪ́} \PS{n.~} \SN \GE{opening, hole (through an object)} \SN \GE{a small door, usually at the back of a compound house (comparable to an emergency exit)} \SN \GE{small space} \XV{N daare ba nyɛ voɔre gaa a kuori.} \XE{I did not get the chance to attend the funeral.} \OP{vɔ᷉ɛ} \TP{vɔrɛɛ} 
\LX{voɔre} \HM{} \PH{vʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to be worn out, to be spent, to be tired and weak} \PD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\LX{voɔre} \HM{} \PH{vʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to get scorched e.g. plants during drought} \PD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\LX{voɔre} \HM{} \PH{vʊ̀ɔ̀rɪ̀} \PS{v.~} \GE{to smear} \PD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\LX{voore fõõ} \HM{} \PH{vʊ́ʊ́rɪ́ fo᷉o᷉} \PS{v.~} \GE{to take a deep breath, to sigh} \PD{-, -, -, -, -, - } 
\LX{vooroŋ} \HM{} \PH{vʊ́ʊ́rʊ́ŋ} \PS{n.~} \SN \GE{life} \SN \GE{breathed air, breath} \OP{voora} \TP{voorɛɛ} 
\LX{Vooronsoŋ} \HM{} \PH{vʊ́ʊ́rʊ́nsʊ́ŋ} \PS{n.~} \GE{Holy Spirit} 
\LX{voowoɔ} \HM{} \PH{vʊ́ʊ́wʊ́ɔ́} \PS{n.~} \GE{womb} \OP{voowoore} \TP{voowoorɛɛ} 
\LX{vore} \HM{} \PH{vʊ́rɪ́} \PS{v.~} \GE{to make a hole through, to pierce e.g. ears, lips} \PD{vorɛɛ, vorɔ, vorɔ, voreba, voraa, voreba } 
\LX{vore} \HM{} \PH{vʊ́rɪ́} \PS{v.~} \GE{to weave} \PD{vorɛɛ, vorɔ, vorɔ, voreba, voraa, voreba } 
\LX{vore} \HM{} \PH{vʊ́rɪ̀} \PS{n.~} \GE{foetus} \OP{voɛ} \TP{vorɛɛ} 
\LX{vore} \HM{} \PH{vʊ́rɪ̀} \PS{adj.~} \GE{alive} \OP{voɛ} \TP{vorɛɛ} 
\LX{vɔrekpɔrɔ} \HM{} \PH{vɔ̀rɪ̀kpɔ̀rɔ̀} \PS{adv.~} \SN \GE{wet and unwieldy} \XV{Saseɛ fue la ka saa meŋ mi kõɔ lɔre a sagere ka a zie e lɛ vɔrekpɔrɔ.} \XE{The rain which followed the rainstorm has soaked the garbage in the yard making it wet and disorderly.} \SN \GE{sound (onomatopoeic); fall by a long winding thing e.g. snake, rope} 
\LX{vɔll} \HM{} \PH{vɔ̀ll} \PS{adj.~} \GE{beautifully; usually in the state of growing up} 
\LX{vori} \HM{} \PH{vórì} \PS{v.~} \GE{to rot, to spoil completely} \PD{voree, voro, voro, voribo, voraa, - } 
\LX{vo᷉o} \HM{} \PH{vʊ᷉ʊ᷉} \PS{v.~} \GE{to mix, to disturb e.g. water} \PD{voo᷉ɛ, voonɔ, voonɔ, voonema, voonaa, - } \VA{voŋe}
\LX{vu} \HM{} \PH{vú} \PS{v.~} \SN \GE{to crawl e.g. child} \SN \GE{to drag along the ground} \PD{vue, vuuro, vuuro, vuribo, vuuraa, vuribo } 
\LX{vubiri} \HM{} \PH{vúbìrí} \PS{n.~} \GE{mound for yam plant} \OP{vubie} \TP{vubiiree} 
\LX{vuduori} \HM{} \PH{vúdúórì} \PS{n.~} \GE{the act of making yam mounds} 
\LX{vuduoro} \HM{} \PH{vúdúóró} \PS{n.~} \GE{one who makes yam mounds} \OP{vuduoribo} 
\LX{vulaŋkaŋkpogili} \HM{} \PH{vùlàŋkǎŋkpóglí} \PS{n.~} \GE{final thing} \OP{vulaŋkaŋkpolo} \TP{vulaŋkaŋkpolee} 
\LX{vuloŋkpogili} \HM{} \PH{vùlòŋkpóglí} \PS{n.~} \GE{beetle} \OP{vuloŋkpogilo} \TP{vuloŋkpogilee} \VA{vulaŋkpogli}
\LX{vulotokpolo} \HM{} \PH{vùlòtókpóló} \PS{n.~} \GE{piece of work meant to be one’s work for the day} \OP{vulotokpolli} \TP{vulotokpolee} 
\LX{vulotokpolo} \HM{} \PH{vùlòtókpóló} \PS{n.~} \GE{stanza (literature)} \OP{vulotokpolli} \TP{vulotokpolee} 
\LX{vunni} \HM{} \PH{vùnnì} \PS{v.~} \GE{to throw far away} \PD{vunnee, vunno, vunno, vunimo, vunnaa, - } 
\LX{vuo} \HM{} \PH{vùó} \PS{n.~} \SN \GE{space,  passage} \SN \GE{chance, time} \SN \GE{occasion} 
\LX{vuo} \HM{} \PH{vùó} \PS{adj.~} \GE{hollow} \OP{vori} \TP{voree} 
\LX{vuoli} \HM{} \PH{vúólì} \PS{v.~} \SN \GE{to bark (dog)} \SN \GE{to hoot (to put to shame)} \PD{vuolee, vuolo, vuolo, vuolibo, vuolaa, - } \VA{wuoli}
\LX{vuoli kaa} \HM{} \PH{vúólì káá} \PS{v.~} \GE{to peer at} \PD{-, -, -, -, -, - } 
\LX{vutaŋkpulee} \HM{} \PH{vútàŋkpùlèè} \PS{n.~} \GE{gate pillar} \OP{vutaŋkpulli} \TP{vutaŋkpullee} 
\LX{vuu} \HM{} \PH{vúù} \PS{v.~} \GE{to drag} \XV{Ba vuu la a kamaana boɔraa boŋyeni na.} \XE{They dragged that one sack of maize.} \PD{vuue, vuuro, vuuro, vuuribo, vuuraa, - } 
\LX{vuuli} \HM{} \PH{vúúlí} \PS{v.~} \GE{to singe} \PD{vuulee, vuulo, vuulo, vuulibo, vuulaa, - } 
\LX{vuuluŋ} \HM{} \PH{vúúlúŋ} \PS{n.~} \GE{an animal believed to have supernatural power and usually attack people in the night} \VA{shortened form of valinvuuluŋ}
\LX{vuuluŋ} \HM{} \PH{vúúlúŋ} \PS{n.~} \GE{short form of valinvuuluŋ (soot)} \VA{shortened form of valinvuuluŋ}
\LX{vũulunsaa} \HM{} \PH{vùùlúnsáà} \PS{n.~} \GE{first rain after the dry season} 
\LX{vuunimiri} \HM{} \PH{vùùnímírì} \PS{n.~} \GE{lightbulb} \OP{vuunimie} \TP{vuunimiiree} \VA{vu᷈ugyɛle}
\LX{vuuri} \HM{} \PH{vùùrì} \PS{v.~} \GE{to feign; to pretend or attempt to do something (which doesn't happen)} \PD{vuuree, vuuro, vuuro, vuuribo, vuuraa, vuuribo } 
\LX{vuuri} \HM{} \PH{vúúrì} \PS{n.~} \GE{small oar-like stick for stirring porridge} \OP{vue} \TP{vuuree} 
\LX{vuuri} \HM{} \PH{vúúrì} \PS{n.~} \GE{yam farm} \OP{vue} \TP{vuuree} 
\LX{vu᷈u} \HM{} \PH{vù᷉ú᷉} \PS{n.~} \GE{fire} \SG{vuunuu} \TP{vuunee} 
\LX{vu᷉udindigili} \HM{} \PH{vù᷉ú᷉díndíglí} \PS{n.~} \SN \GE{electricity} \SN \GE{electric light} \VA{vu᷉udigili}
\LX{vu᷉udiraa} \HM{} \PH{vù᷉ù᷉dìráà} \PS{n.~} \GE{flame} \OP{vu᷉udirri} \TP{vu᷉udiree} 
\LX{vu᷉udirri} \HM{} \PH{vù᷉ùdìrrì} \PS{n.~} \GE{a stream of burning} 
\LX{vu᷈ugyɛle} \HM{} \PH{vù᷈ù᷈gyɛ́lɪ́} \PS{n.~} \GE{lightbulb} \OP{vù᷈ùgyɛlɛ} \TP{vù᷈ùgyɛlɛɛ} \VA{vu᷈u᷈nimiri}
\LX{vu᷉usaalbiri} \HM{} \PH{vù᷉ùsáálbírí} \PS{n.~} \GE{burning charcoal} \VA{vu᷉usaale}
\LX{vu᷉usaale} \HM{} \PH{vù᷉ù᷉sáálɪ́} \PS{n.~} \GE{burning coal} \OP{vu᷉usaala} \TP{vu᷉usaaloo} \VA{vu᷉usaalbiri}
\LX{vu᷉uvaale} \HM{} \PH{vù᷉úváálɪ́} \PS{n.~} \GE{a type of bird which hovers around bushes that are being burnt to catch fleeing insects} \OP{vu᷉uvaala} 
\LX{vu᷈uyɛlɛ} \HM{} \PH{vù᷈ù᷉gyɛ́lɛ́} \PS{n.~} \GE{lightbulb} \OP{vu᷈u᷈gyɛlɛ} \TP{vu᷈u᷈gyɛlɛɛ} 
\LX{vu᷈uzeɛle} \HM{} \PH{vù᷉ù᷉zɪ́ɛ́lɪ́} \PS{n.~} \GE{flying burnt particles from a fire e.g. grass} \SG{vu᷈uzeɛloo} \OP{vu᷈uzeɛlɛ} \TP{vu᷈uzeɛlɛɛ} 
\LX{vu᷈uzele} \HM{} \PH{vù᷈ù᷈zɪ̀lɪ́} \PS{n.~} \GE{flame, tongue of fire} \OP{vu᷈uzelɛ} \TP{vu᷈uzelɛɛ} 
\LX{Wa} \HM{} \PH{wǎ} \PS{n.~} \GE{name of the regional headquarters of the Upper West Region} 
\LX{wa} \HM{} \PH{wà} \PS{v.~} \GE{to come} \PD{waɛ, waana, waana, wanema, waanaa, wanema } 
\LX{wa} \HM{} \PH{wá} \PS{part.~} \GE{by the time; before the end of a given time} \XV{Te na wa tara ka zivɛlmɛ la.} \XE{By the time we arrive, it will be daybreak.} 
\LX{waa} \HM{} \PH{wàá} \PS{n.~} \GE{yam} \OP{waare} \TP{waarɛɛ} \VA{nyuuri}
\LX{waa} \HM{} \PH{wàá} \PS{auxiliary verb.~} \SN \GE{to be} \XV{A tomɔ waa la yaga.} \XE{The work is plenty.} \SN \GE{to resemble, to be like} \XV{A biiri waa ŋa yeere.} \XE{The children are like twins.} 
\LX{waabo} \HM{} \PH{wáábʊ́} \PS{n.~} \GE{snake} \OP{weere/wamine} \TP{weerɛɛ} 
\LX{waale} \HM{} \PH{wáálɪ̀} \PS{v.~} \GE{to whisper} \PD{waalɛɛ, waala, waala, waaleba, waalaa, waaleba } 
\LX{waale} \HM{} \PH{wáálɪ̀} \PS{v.~} \GE{to parboil} \PD{waalɛɛ, waala, waala, waaleba, waalaa, waaleba } 
\LX{waale} \HM{} \PH{wáálɪ̀} \PS{v.~} \GE{to bruise slightly} \PD{waalɛɛ, waala, waala, waaleba, waalaa, waaleba } 
\LX{waaleŋ} \HM{} \PH{wààlɪ̂ŋ} \PS{n.~} \GE{a type of gunpowder (imported)} 
\LX{waaleŋ} \HM{} \PH{wààlɪ̂ŋ} \PS{adj.~} \GE{like a native of Wa} 
\LX{waaloŋ} \HM{} \PH{wààlʊ̂ŋ} \PS{n.~} \SN \GE{nature, character} \SN \GE{resemblance} 
\LX{Waaloo} \HM{} \PH{wáálʊ̀ʊ̀} \PS{n.~} \GE{native of Wa} \OP{waala} 
\LX{waane} \HM{} \PH{wáánɪ̀} \PS{v.~} \SN \GE{to dress tightly (pull trousers to navel level)} \SN \GE{to propel, to spin} \PD{waanɛɛ, waana, waana, waanema, waanaa, - } 
\LX{waare} \HM{} \PH{wààrɪ̀} \PS{v.~} \GE{be touched by (liquid, powder)} \XV{A bie ɔŋ la tɛne waare o ma.} \XE{The child has smeared the mother with soil.} \PD{waarɛɛ, waara, waara, waareba, waaraa, - } 
\LX{waare} \HM{} \PH{wáárɪ́} \PS{v.~} \SN \GE{to burn the cover (when you attempt to burn green grass, it is only the blades that burn leaving the straw)} \XV{A moɔ da e la mɔmaara lɛnso ka ba di kyɛ waare.} \XE{The grass was green and that is why when they attempted burning it it did not burn.} \XV{A naŋkpaana nyege la a mɔmaara kyɛ a ba di, a waarɛɛ la.} \XE{The hunter tried to burn the green grass, but it did not burn well; it only burnt halfway.} \SN \GE{to be painful} \XV{Ba tɔ la a bie fele ka o eŋɛ zaa waare.} \XE{They hit on his sore and it was quite painful.} \PD{waarɛɛ, waara, waara, waareba, -, - } 
\LX{waaroŋ} \HM{} \PH{wààrʊ́ŋ} \PS{n.~} \GE{organised group hunting} 
\LX{wadɛre} \HM{} \PH{wàdɛ́rɪ́} \PS{n.~} \GE{} \OP{wadɛrɛɛ} \VA{nyudɛre}
\LX{wadɛrɛɛ} \HM{} \PH{wàdɛ́rɛ́ɛ́} \PS{n.~} \GE{} \TP{wadɛrrɛɛ} \VA{nyudɛrɛɛ}
\LX{wae} \HM{} \PH{wáɪ́} \PS{adj.~} \GE{nine} 
\LX{waee} \HM{} \PH{wáɪ́ɪ́} \PS{adv.~} \GE{at once} 
\LX{wagere} \HM{} \PH{wágrɪ̀} \PS{n.~} \GE{time for something, period} \OP{wagerɛɛ} 
\LX{wagere} \HM{} \PH{wàgrɪ̀} \PS{v.~} \GE{to bite on the surface (dog)} \PD{wagerɛɛ, wagera, wagera, wagereba, wageraa, - } 
\LX{wagyɛ} \HM{} \PH{wàdzɛ̀} \PS{n.~} \GE{blanket, cloth} \OP{wagyɛre} \TP{wagyɛrɛɛ} 
\LX{walaa} \HM{} \PH{wáláà} \PS{n.~} \GE{antelope} \OP{walle} \TP{wallɛɛ} 
\LX{walanse} \HM{} \PH{wálánsɪ̀} \PS{n.~} \GE{radio, wireless set} \OP{walansere} \TP{walanserɛɛ} 
\LX{wale} \HM{} \PH{wálɪ́} \PS{n.~} \GE{sweat} \TP{walɛɛ/wallɛɛ} 
\LX{walpelaa} \HM{} \PH{wàlpɪ̀làà} \PS{n.~} \GE{a type of antelope} \OP{walpeɛle} 
\LX{walpieli} \HM{} \PH{wàlpíélí} \PS{n.~} \GE{} \SG{walpieluu} \OP{walpiele} \TP{walpielee} \VA{walpieluu}
\LX{walpieluu} \HM{} \PH{wàpíélúú} \PS{n.~} \GE{} \OP{walpiele} \TP{walpielee} \VA{walpieli}
\LX{walsugo} \HM{} \PH{wálsúgó} \PS{n.~} \GE{group of deer} \OP{walsugri} \TP{walsugree} 
\LX{walsunni} \HM{} \PH{wàlsúnnì} \PS{n.~} \GE{perspiration, extreme sweat} \TP{wasunnee} 
\LX{walwoɔ} \HM{} \PH{wàlwʊ́ɔ́} \PS{n.~} \SN \GE{newcomer, novice} \SN \GE{a bag made of antelope skin} \OP{walwoore} \TP{walwooɛɛ} 
\LX{waŋene} \HM{} \PH{wáŋɪ́nɪ́} \PS{n.~} \GE{great one} \OP{wanne} \TP{wannɛɛ} 
\LX{waŋmane} \HM{} \PH{wàŋmánɪ́} \PS{n.~} \GE{a hundred yams} \OP{waŋmama} \TP{waŋmamɛɛ} 
\LX{wanwɛ} \HM{} \PH{wánwɛ́} \PS{v.~} \GE{to be gone forever; to fail to return as expected} \XV{Ba toŋ la a bie ka o deɛ gaa te wanwɛ.} \XE{When they sent the child, he left and never returned.} \PD{wanwɛɛ, wanaweɛrɛ, wanweɛrɛ, wanweɛrebɛ, wanweɛraa, - } 
\LX{ware} \HM{} \PH{wárɪ̀} \PS{n.~} \GE{coins; money} \OP{warre} \TP{warɛɛ} \VA{warebiri}
\LX{ware} \HM{} \PH{wárí} \PS{adv.~} \GE{all gone (people on a journey)} \XV{A teŋɛ dɔbɔ zaaŋ e weɛ, ware.} \XE{All the men in the town have gone to the farm; not even one was left behind.} 
\LX{ware} \HM{} \PH{wárɪ́} \PS{n.~} \GE{long draught} \TP{warɛɛ} 
\LX{warebiri} \HM{} \PH{wárɪ́bírì} \PS{n.~} \GE{see ware} \OP{warebie} \TP{warebiiree} \VA{ware}
\LX{warema} \HM{} \PH{wárɪ́má} \PS{n.~} \GE{brick, block} \OP{waremare} \TP{waremarɛɛ} 
\LX{wasaaloŋ} \HM{} \PH{wàsáálʊ́ŋ} \PS{n.~} \GE{water yam} \TP{wasaalɛɛ.} 
\LX{wasege} \HM{} \PH{wàsɪ́gɪ́} \PS{n.~} \GE{yam barn} \OP{wasegre} \TP{wasegrɛɛ} \VA{nyusege}
\LX{wasɔŋɔ} \HM{} \PH{wàsɔ́ŋ} \PS{n.~} \GE{a basket of yam} \OP{wasɔnne} \TP{wasɔnnɛɛ} 
\LX{wasugo} \HM{} \PH{wàsúgó} \PS{n.~} \GE{see sugo (termites)} \OP{wasugiri} \TP{wasugiree} 
\LX{waweɛlɛ} \HM{} \PH{wàwɪ́ɛ́lɛ́} \PS{n.~} \GE{seed yam} \SG{waweɛloo} \TP{waweɛlɛɛ} \VA{nyuweɛlɛ}
\LX{wazɪɛ} \HM{} \PH{wázɪ̀ɛ́} \PS{n.~} \GE{a type of brown snake} \OP{wazeere} \TP{wazeerɛɛ} 
\LX{we} \HM{} \PH{wé} \PS{interjection.~} \GE{expression of despair} 
\LX{wɛ} \HM{} \PH{wɛ̀} \PS{adv.~} \SN \GE{of cause} \XV{Ba nyɛ la a bonzuuri o dieŋ, ona la  zu wɛ.} \XE{They found the stolen items in his room, of cause he stole them.} \SN \GE{then} \XV{Ka fo wa pagra foŋ de we᷉ɛ᷉ wɛ.} \XE{If you are in a haste then take the lead.} 
\LX{wɛ} \HM{} \PH{wɛ́} \PS{v.~} \SN \GE{to go bad} \XV{Sugo kpɛ la a dɔɔ wasege, a waare zaaŋ wɛ.} \XE{Termites have invaded the man's yam barn; all the yams are destroyed.} \SN \GE{to be lost forever} \PD{wɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\LX{wɛ} \HM{} \PH{wɛ́} \PS{adj.~} \GE{not strong or courageous} \XV{Nyɛ dɔɔ naŋ wɛ, haali o koŋ to᷉ɔ᷉ zɛle a pelee.} \XE{See how useless the man is; he cannot even lift the small basket.} 
\LX{wɛbaa} \HM{} \PH{wɛ̀báá} \PS{n.~} \GE{fox} \OP{wɛbaare} \TP{wɛbaarɛɛ} 
\LX{webe} \HM{} \PH{wébé} \PS{interjection.~} \GE{exclamation of exhaustion} 
\LX{wɛdoŋɔ} \HM{} \PH{wɛ̀dʊ́ŋɔ́} \PS{n.~} \GE{a non-domesticated animal} \OP{wɛdonne.} \TP{wɛdonnɛɛ.} \VA{mɔdoŋɔ, wɛdoŋa,}
\LX{weɛ} \HM{} \PH{wɪ̀ɛ́} \PS{n.~} \SN \GE{farm} \SN \GE{wilderness} \OP{wɛre} \TP{wɛrɛɛ} 
\LX{weɛ} \HM{} \PH{wɪ̀ɛ̀} \PS{v.~} \SN \GE{to make a furrow/trench e.g for water to pass} \SN \GE{cut open; usually part of the body in a bid to remove a foreign body or pus} \SN \GE{to push growth aside e.g. in a search for something in a thick growth} \PD{weɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\LX{weɛ} \HM{} \PH{wɪ́ɛ́} \PS{v.~} \GE{to save} \XV{A kpogili la weɛ o ka lɛ naane o zu da na ŋmare la.} \XE{He would have broken his head but he was saved by the crash helmet.} \PD{weɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\LX{weɛ} \HM{} \PH{wɪ́ɛ́} \PS{adj.~} \GE{not related to} 
\LX{weɛle} \HM{} \PH{wɪ́ɛ́lɪ́} \PS{v.~} \GE{to branch out in lobes (yam seed)} \PD{weɛlɛɛ, weɛlɛ, weɛlɛ, weɛlbɛ, weɛlaa, - } 
\LX{weɛle} \HM{} \PH{wɪ́ɛ́lɪ́} \PS{v.~} \GE{to miss out, to lose} \PD{weɛlɛɛ, weɛlɛ, weɛlɛ, weɛlbɛ, weɛlaa, - } 
\LX{weɛle} \HM{} \PH{wɪ̀ɛ́lɪ́} \PS{n.~} \GE{twins} \TP{weɛlɛɛ} 
\LX{weɛlɛ} \HM{} \PH{wɪ̀ɛ̀lɛ́} \PS{n.~} \GE{mistletoe (a parasitic plant whose fruit is used to produce gum)} \TP{weɛlɛɛ} 
\LX{weɛlɛ} \HM{} \PH{wɪ́ɛ̀lɛ̀} \PS{adj.~} \SN \GE{in vain, for nothing} \XV{Fo gɛrɛ la be weɛlɛ, o koŋ leɛ ka fo.} \XE{You are going there for nothing; she will not mind you.} \SN \GE{free} \XV{A doroba de la a pɔgenaŋaa gaa ne a dɔɔta weɛlɛ.} \XE{The driver took the old woman to the hospital free of charge.} 
\LX{weɛmɛ} \HM{} \PH{wɪ̀ɛ̀mɛ́} \PS{adj.~} \GE{wild, untamed} 
\LX{weene} \HM{} \PH{wɪ̀ɪ̀nɪ̀} \PS{v.~} \GE{to begin to boil} \PD{weenɛɛ, weenɛ, weenɛ, -, -, - } 
\LX{weɛre} \HM{} \PH{wɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to search through e.g. brush} \PD{weɛrɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\LX{wɛge} \HM{} \PH{wɛ́gɪ̀} \PS{v.~} \GE{to hatch} \PD{wɛgɛɛ, wɛgrɛ, wɛgrɛ, wɛgrebɛ, wɛgraa, wɛgrebɛ } 
\LX{wɛge} \HM{} \PH{wɛ́gɪ̀} \PS{v.~} \GE{to multiply} \PD{wɛgɛɛ, wɛgrɛ, wɛgrɛ, wɛgrebɛ, wɛgraa, wɛgrebɛ } 
\LX{wɛge} \HM{} \PH{wɛ̀gɪ̀} \PS{v.~} \GE{to remove with a pointed implement e.g. corn from the cob} \PD{wɛgɛɛ, wɛgrɛ, wɛgrɛ, wɛgrebɛ, wɛgraa, - } 
\LX{wɛge} \HM{} \PH{wɛ́gɪ̀} \PS{n.~} \GE{beam, large plank used in making mud roofing} \OP{wɛgere} \TP{wɛgerɛɛ} \VA{wɛgɛ, dawɛge}
\LX{wɛgɛ} \HM{} \PH{wɛ́gɛ́} \PS{n.~} \GE{beam} \OP{wɛgre} \TP{wɛgɛɛ} \VA{wɛge}
\LX{wegere} \HM{} \PH{wɪ̀grɪ̀} \PS{v.~} \GE{to squeeze through e.g. through a thick crowd} \PD{wegerɛɛ, wegerɛ, wegerɛ, wegerebɛ, wegeraa, - } 
\LX{wɛgere} \HM{} \PH{wɛ̀grɪ̀} \PS{v.~} \GE{to force out, to squeeze out e.g. seed from pod} \PD{wɛgerɛɛ, wɛgerɛ, wɛgerɛ, wɛgerebɛ, wɛgeraa, - } 
\LX{wɛkaraa} \HM{} \PH{wɛ̀káráá} \PS{n.~} \GE{mounds of a farm made without first clearing the land} \OP{wɛkare} \TP{wɛkarɛɛ} 
\LX{welaa} \HM{} \PH{wɪ́láá} \PS{adj.~} \GE{harbouring evil intentions} 
\LX{wele} \HM{} \PH{wɪ́lɪ́} \PS{v.~} \GE{to rummage} \PD{welɛɛ, welɛ, welɛ, welebɛ, welaa, - } 
\LX{wele} \HM{} \PH{wɪ́lɪ́} \PS{v.~} \GE{to dismantle} \PD{welɛɛ, welɛ, welɛ, welebɛ, welaa, - } 
\LX{wele} \HM{} \PH{wɪ́lɪ́} \PS{v.~} \GE{to reduce in size e.g. obesity} \PD{welɛɛ, welɛ, welɛ, welebɛ, welaa, - } 
\LX{wele} \HM{} \PH{wɪ́lɪ́} \PS{n.~} \GE{a type of ornament for the arm} \OP{welɛ} \TP{welɛɛ} 
\LX{wɛle} \HM{} \PH{wɛ́lɪ̀} \PS{v.~} \SN \GE{to split or break into two} \SN \GE{to separate, to disperse} \PD{wɛlɛɛ, wɛlɛ, wɛlɛ, wɛlebɛ, wɛlaa, wɛlebɛ } 
\LX{wɛlɛmbɛll} \HM{} \PH{wɛ̀lɛ̀mbɛ̀ll} \PS{adv.~} \GE{massively moving} 
\LX{welɛnwelɛŋ} \HM{} \PH{wɪ̀lɛ́nwɪ̀lɛ́ŋ} \PS{adv.~} \GE{vigorous, rolling (boiling liquid)} 
\LX{welɛwelɛ} \HM{} \PH{wɪ̀lɛ̀wɪ̀lɛ̀} \PS{adv.~} \GE{intermittent spackle} 
\LX{weli} \HM{} \PH{wélí} \PS{n.~} \GE{snail} \OP{wele} \TP{welee} 
\LX{weliweli} \HM{} \PH{wélwél} \PS{adv.~} \GE{clearly visible} 
\LX{welle} \HM{} \PH{wɪ́llɪ̀} \PS{v.~} \SN \GE{to scrutinize, to examine closely} \SN \GE{to rummage} \PD{wellɛɛ, wellɛ, wellɛ, welebɛ, wellaa, - } 
\LX{welli} \HM{} \PH{wéllì} \PS{v.~} \GE{to make very clear, clean, or neat} \PD{wellee, welle, welle, welibe, wellaa, - } 
\LX{wɛnaabo} \HM{} \PH{wɛ̀náábʊ̀} \PS{n.~} \GE{buffalo} \OP{wɛnii} \TP{wɛniinee} 
\LX{wenne} \HM{} \PH{wɪ̀nnɪ̀} \PS{v.~} \GE{to go into the interior e.g. of bush or a crowd} \PD{wennɛɛ, wennɛ, wennɛ, wenemɛ, wennaa, - } 
\LX{wɛnoɔ} \HM{} \PH{wɛ́nʊ́ɔ́} \PS{n.~} \GE{dark grey hen} \OP{wɛnoore} \TP{wɛnoorɛɛ} 
\LX{wɛnu} \HM{} \PH{wɛ́nú} \PS{n.~} \GE{an arm of the main body of hunters} \OP{wɛnuuri.} \TP{wɛnuuree.} 
\LX{wɛnwɛrɛɛ} \HM{} \PH{wɛ̌nwɛ́rɛ́ɛ́} \PS{n.~} \GE{a type of small plant with edible leaves} \SG{wɛnwɛroo} \TP{wɛnwɛrrɛɛ} 
\LX{wɛnyuo} \HM{} \PH{wɛ́nyúó} \PS{n.~} \GE{leopard} \OP{wɛnyuuri} \TP{wɛnyuuree} 
\LX{werɛ} \HM{} \PH{wɪ̀rɛ́} \PS{adj.~} \GE{small species, not of high breed (groundnuts)} \TP{werɛɛ} 
\LX{werɛ} \HM{} \PH{wɪ́rɛ́} \PS{n.~} \GE{mountain} \TP{werɛɛ/werrɛɛ} 
\LX{werɛ} \HM{} \PH{wɪ́rɛ́} \PS{adj.~} \GE{in vain} \XV{O baloo na zaa kyɛ werɛ la, o ba nyɛ a libie.} \XE{All his suffering has been in vain; he did not get the money.} 
\LX{wɛre} \HM{} \PH{wɛ̀rɪ̀} \PS{v.~} \SN \GE{to cut open} \SN \GE{to operate} \SN \GE{to split, to saw} \SN \GE{to make a furrow or gutter} \PD{wɛrɛɛ, wɛrɛ, wɛrɛ, wɛrebɛ, wɛraa, - } 
\LX{were-bare} \HM{} \PH{wìrɪ̀-bàrì} \PS{v.~} \GE{to ignore} \PD{were-barɛɛ, werɛ-bara, were-bara, were-bareba, were-baraa, - } 
\LX{werekao} \HM{} \PH{wírɪ́káʊ̀} \PS{n.~} \GE{wound, mark, defect} \XV{Lɔɔre la le ne o ka a o gbɛre nyɛ werekao.} \XE{A car crash has caused a wound on his leg.} \OP{werekarre} \TP{werekarɛɛ} 
\LX{werewere} \HM{} \PH{wɪ́rɪ́wɪ́rɪ́} \PS{adv.~} \GE{smartly, quickly} 
\LX{werɛwerɛ} \HM{} \PH{wɪ́rɛ́wɪ́rɛ́} \PS{adj.~} \GE{tiny things} 
\LX{werɛwerɛ} \HM{} \PH{wɪ̀rɛ̀wɪ̀rɛ̀} \PS{adj.~} \GE{coarse (flour)} 
\LX{weri} \HM{} \PH{wérí} \PS{adv.~} \GE{valuable, good quantity} 
\LX{weroo} \HM{} \PH{wɪ́rʊ́ʊ́} \PS{adj.~} \GE{small in stature} \OP{werɛ} 
\LX{we᷉derɛ} \HM{} \PH{wɪ́᷉dɪ́rɛ́} \PS{n.~} \GE{leader, forerunner} \OP{we᷉derebɛ} \TP{we᷉derebɛɛ} 
\LX{we᷉ɛ} \HM{} \PH{wí᷉ɛ̀᷉} \PS{adv.~} \GE{in front, ahead} \XV{Dabuo la na de o sakuuri maakyeebo we᷉ɛ.} \XE{Dabuo will lead his school at the parade ground.} 
\LX{we᷉ɛ} \HM{} \PH{wɪ̀᷉ɛ̀᷉} \PS{v.~} \GE{to be wild, to be untamed, to be out of control} \PD{we᷉ɛɛ, weɛnɛ, weɛnɛ, weɛnemɛ, weɛnaa, - } 
\LX{wi} \HM{} \PH{wí} \PS{v.~} \GE{to carve; to engrave} \PD{wie, wiire, wiire, wiiribe, wiiraa, - } 
\LX{wigiri} \HM{} \PH{wígrì} \PS{v.~} \GE{to toss to turn over (grain in a basket)} \PD{wiigree, wigire, wigire, wigiribe, wigiraa, - } 
\LX{wiiruŋ} \HM{} \PH{wíírúŋ} \PS{n.~} \GE{design} 
\LX{willi} \HM{} \PH{wìllì} \PS{v.~} \SN \GE{to overturn to capsize} \XV{A saseɛ da waa la kpeɛŋaa, a da da a gboe zaa willi a mane poɔ.} \XE{The wind was so strong that it capsized all the boats on the sea.} \SN \GE{to heave over} \PD{willee, wille, wille, wilibe, willaa, wilibe } 
\LX{wire} \HM{} \PH{wíré} \PS{adv.~} \GE{gone far} 
\LX{wirewire} \HM{} \PH{wíréwíré} \PS{adj.~} \GE{far away} 
\LX{wiri} \HM{} \PH{wìrí} \PS{n.~} \GE{horse} \OP{wie} \TP{wiree/wirree} 
\LX{wiri} \HM{} \PH{wìrì} \PS{v.~} \GE{to turn inside out e.g. pocket} \PD{wiree, wire, wire, wiribe, wiraa, - } 
\LX{wiri-buuli} \HM{} \PH{wìrì-búúló} \PS{n.~} \GE{bangles for a horse} \OP{wiri-buulo} \TP{wiri-buulee} 
\LX{wiridaa} \HM{} \PH{wìrìdáá} \PS{n.~} \GE{male horse} \OP{wiridaare} \TP{wiridaarɛɛ} 
\LX{wiriduoruŋ} \HM{} \PH{wìrìdúórúŋ} \PS{n.~} \SN \GE{jaundice} \SN \GE{yellow fever} 
\LX{wiri-eɛraa} \HM{} \PH{wìrì-ɪ́ɛ́ráá} \PS{n.~} \GE{metal spur worn on the heel for prodding the horse} \OP{wiri-eɛre} 
\LX{wirigoɔ} \HM{} \PH{wìrìgʊ́ɔ́} \PS{n.~} \GE{a type of thorny plant} \OP{wirigoore} \TP{wirigoorɛɛ} 
\LX{wirikogi} \HM{} \PH{wìrìkógí} \PS{n.~} \GE{horse saddle} \OP{wirikogri} \TP{wirikogree} 
\LX{wirikuoluu} \HM{} \PH{wìrìkúólúú} \PS{n.~} \GE{a type of honey-making insect} \OP{wirikuolo} \TP{wirikuolee} 
\LX{wirilee} \HM{} \PH{wìrìléé} \PS{n.~} \GE{colt} \OP{wiribilii} 
\LX{wirinɔbɔŋ} \HM{} \PH{wìrìnɔ́bɔ́ŋ} \PS{n.~} \GE{metal ring placed arond the muzzle of a horse} \OP{wirinɔbɔnne} \VA{wirinɔbɔŋenɔ}
\LX{wirinyaŋaa} \HM{} \PH{wìrìɲáŋáá} \PS{n.~} \GE{female horse} \OP{wirinyaŋene} \TP{wirinyaŋenɛɛ} 
\LX{wirizɔɔnɔ} \HM{} \PH{wìrìzɔ́ɔ́nɔ́} \PS{n.~} \GE{horse rider} \OP{wirizɔɔnema} \TP{wirizɔɔmɛɛ} 
\LX{wirizoore} \HM{} \PH{wìrìzʊ́ʊ́rɪ́} \PS{n.~} \GE{horse tail} \OP{wirizoɛ} 
\LX{wirizoore} \HM{} \PH{wìrìzʊ́ʊ́rɪ́} \PS{n.~} \GE{a type of grass} \OP{wirizoɛ} 
\LX{wirizoore} \HM{} \PH{wìrìzʊ́ʊ́rɪ́} \PS{n.~} \GE{a type of sickness} \OP{wirizoɛ} 
\LX{wi᷉eu} \HM{} \PH{wí᷉é᷉ú᷉} \PS{adj.~} \GE{fast, quick} \VA{wieŋ}
\LX{wofeɛmaa} \HM{} \PH{wʊ́fɪ́ɛ́máá} \PS{n.~} \GE{bag made from animal skin without hair} \OP{wofeɛne} \TP{wofeɛnɛɛ/wofeɛmɛɛ} \VA{wofe᷉ɛ, wofuoraa, wokpele}
\LX{wofe᷉ɛ} \HM{} \PH{wʊ̀fɪ́ɛ́} \PS{n.~} \GE{a bag made from animal skin without hair} \OP{wofeɛne} \TP{wofeɛnɛɛ} \VA{wofeɛmaa}
\LX{wofuoraa} \HM{} \PH{wʊ̀fúóráá} \PS{n.~} \GE{} \OP{wofuori} \TP{wofuoree} \VA{wofeɛmaa}
\LX{wɔge} \HM{} \PH{wɔ̀gɪ̀} \PS{v.~} \GE{to treat fresh skin in preparation for making a bag} \PD{wɔgɛɛ, wɔgrɔ, wɔgrɔ, wɔgreba, wɔgraa, - } 
\LX{wɔge} \HM{} \PH{wɔ̀gɪ̀} \PS{v.~} \GE{to cause to suffer} \PD{wɔgɛɛ, wɔgrɔ, wɔgrɔ, wɔgreba, wɔgraa, - } 
\LX{wɔgele} \HM{} \PH{wɔ̀glɪ̀} \PS{v.~} \GE{to physically suffer inwardly intensely} \PD{wɔglɛɛ, wɔglɔ, wɔglɔ, wɔgleba, wɔglaa, - } 
\LX{wɔgele} \HM{} \PH{wɔ̀glɪ̀} \PS{v.~} \GE{to be burnt by touching a hot object} \PD{wɔglɛɛ, wɔglɔ, wɔglɔ, wɔgleba, wɔglaa, - } 
\LX{wɔgele} \HM{} \PH{wɔ̀glɪ̀} \PS{v.~} \GE{to treat beans (by soaking and drying)} \PD{wɔglɛɛ, wɔglɔ, wɔglɔ, wɔgleba, wɔglaa, - } 
\LX{wɔgere} \HM{} \PH{wɔ́grɪ́} \PS{v.~} \GE{to shake e.g. objects in a tin} \PD{wɔgrɛɛ, wɔgrɔ, wɔgrɔ, wɔgreba, wɔgraa, - } \VA{wogiri}
\LX{Wogi} \HM{} \PH{wògí} \PS{n.~} \GE{a town in the Upper West Region of Ghana} 
\LX{wogi} \HM{} \PH{wógì} \PS{adj.~} \GE{long, tall} 
\LX{wogiri} \HM{} \PH{wógrí} \PS{v.~} \GE{to cause to shake e.g. carried container of pebbles} \PD{wogiree, wogiro, wogiro, wogiribo, wogiraa, - } \VA{wɔgere}
\LX{wogiri} \HM{} \PH{wògrì} \PS{v.~} \GE{to steal cunningly} \PD{wogiree, wogiro, wogiro, wogiribo, wogiraa, - } 
\LX{wogiroŋ} \HM{} \PH{wògròŋ} \PS{adj.~} \GE{large, big e.g. container} \OP{wogiroŋ wogiroŋ} 
\LX{wogoko} \HM{} \PH{wógókó} \PS{adv.~} \GE{intense (painful bite)} 
\LX{wokpale} \HM{} \PH{wʊ̀kpálɪ́} \PS{n.~} \GE{} \OP{wokpala} \TP{wokpalɛɛ} \VA{wofeɛmaa}
\LX{wola} \HM{} \PH{wʊ́là} \PS{interrogative pronoun.~} \SN \GE{how} \XV{Wola ka te na e maale a pɔgebilii bayi bama yelbawontaa ŋa?} \XE{How are we going to settle the misunderstanding between the two girls?.} \SN \GE{why} \XV{Wola la ka fo la zeŋe zomm lɛ?} \XE{Why are you sitting so quietly?.} \SN \GE{what} \XV{O yeli ka wola?} \XE{What did s/he say?.} 
\LX{wolaa} \HM{} \PH{wʊ́láá} \PS{adj.~} \GE{dangerous} \OP{wolle} 
\LX{woli} \HM{} \PH{wólí} \PS{n.~} \GE{leather rope worn on the waist} \OP{wolo} \TP{wollee} 
\LX{woliwoli} \HM{} \PH{wòlwòlí} \PS{v.~} \GE{to be restless (of a person)} \PD{wolwolee, wolwolo, wolwolo, wolwolibo, wolwolaa, wolwolibo } 
\LX{wolle} \HM{} \PH{wʊ̀llɪ̀} \PS{v.~} \GE{to rattle, to speak fluently (langauge)} \PD{wollɛɛ, wolla, wolla, wolba, wollaa, - } 
\LX{wolle} \HM{} \PH{wʊ́llɪ́} \PS{v.~} \GE{to be mischievous} \PD{-, wolla, wolba, wollaa, -, - } 
\LX{wolɔ} \HM{} \PH{wʊ́lɔ̀} \PS{v.~} \GE{to deceive, to hoax, to humbug} \PD{wolɛɛ, wolɔ, wola, woleba, wolaa, - } 
\LX{wolomboll} \HM{} \PH{wòlòmbòll} \PS{adj.~} \GE{in disarray or confusion e.g. mob} 
\LX{woloŋ} \HM{} \PH{wʊ́lʊ́ŋ} \PS{n.~} \SN \GE{disease} \SN \GE{problem} 
\LX{wolonwoloŋ} \HM{} \PH{wólónwólóŋ} \PS{adv.~} \GE{haphazard, in a troubled state} 
\LX{wolowolo} \HM{} \PH{wólówóló} \PS{adj.~} \GE{restive} 
\LX{wonaa} \HM{} \PH{wónáá} \PS{n.~} \GE{that which pesters} 
\LX{wɔne} \HM{} \PH{wɔ́nɪ́} \PS{n.~} \SN \GE{small swelling} \SN \GE{pimple} \OP{wɔmɔ} \TP{wɔnnɛɛ} 
\LX{wɔne} \HM{} \PH{wɔ́nɪ́} \PS{n.~} \GE{fruit} \OP{wɔmɔ} \TP{wɔnnɛɛ} 
\LX{wɔŋe} \HM{} \PH{wɔ́ŋ} \PS{v.~} \SN \GE{to bear fruit} \SN \GE{to develop rashes, pimples} \PD{wɔŋɛɛ, wɔnɔ, wɔnɔ, wɔnema, wɔnaa, - } 
\LX{woŋi} \HM{} \PH{wòŋ} \PS{v.~} \GE{to hear} \XV{N ba teɛre ka o woŋ a yɛlɛ te naŋ yeli.} \XE{I don't think she has heard what we said.} \PD{woŋee, wono, wono, wonimo, wonaa, wonimo } \SE{woŋ noɔ} \GE{to enjoy, to be glad} \SE{woŋ tuo} \GE{to suffer} 
\LX{woŋi} \HM{} \PH{wòŋ} \PS{v.~} \GE{to be deaf} \XV{Ka fo toboŋ beɛrɛ ka fo ba kaa a zu velaa fo na woŋ la.} \XE{If you have a disease of the ears and you do not take good care of them, you will be deaf.} \PD{woŋee, wono, wono, wonimo, wonaa, wonimo } \SE{woŋ noɔ} \GE{to enjoy, to be glad} \SE{woŋ tuo} \GE{to suffer} 
\LX{woŋi} \HM{} \PH{wòŋ} \PS{v.~} \GE{to feel, to suffer} \XV{Ba ŋmɛ la baa ka a woŋi ka o nyɛ, bini.} \XE{When they hit the dog it felt so much pain that it defecated.} \PD{woŋee, wono, wono, wonimo, wonaa, wonimo } \SE{woŋ noɔ} \GE{to enjoy, to be glad} \SE{woŋ tuo} \GE{to suffer} 
\LX{woŋi} \HM{} \PH{wòŋ} \PS{v.~} \GE{to be of the correct quantity} \XV{Zenɛ, Ayuo zeɛre ŋa nyɛnnoo ba woŋ.} \XE{Today, there was not the right amount of salt in Ayuo's soup.} \PD{woŋee, wono, wono, wonimo, wonaa, wonimo } \SE{woŋ noɔ} \GE{to enjoy, to be glad} \SE{woŋ tuo} \GE{to suffer} 
\LX{wɔŋkampɔre} \HM{} \PH{wɔ́ŋkámpɔ́rɪ́} \PS{n.~} \SN \GE{a yamlike plant that bears fruit which is boiled and eaten} \SN \GE{the fruit of this plant} \OP{wɔŋkampɔrɔ} \TP{wɔŋkampɔrɛɛ} \VA{feree}
\LX{woŋo} \HM{} \PH{wòŋó} \PS{n.~} \GE{deaf person} \OP{wonni} \TP{wonnee} \VA{woŋi}
\LX{wontaa} \HM{} \PH{wǒntáá} \PS{n.~} \GE{understanding between people} 
\LX{woo} \HM{} \PH{wʊ́ʊ̀} \PS{n.~} \GE{water iguana} \OP{woore} \TP{woorɛɛ} 
\LX{woo} \HM{} \PH{wóó} \PS{interjection.~} \GE{a cry of pity} 
\LX{woɔ} \HM{} \PH{wʊ́ɔ́} \PS{n.~} \GE{bag} \OP{woore} \TP{woorɛɛ} 
\LX{woɔ} \HM{} \PH{wʊ̀ɔ̀} \PS{v.~} \SN \GE{to remove from the top of e.g. basket of grain} \XV{A seŋka᷉a᷉ paale la a pelee zuo, yɛ woɔ ka a ta yaara.} \XE{The basket is too full; remove some of the groundnuts so it does not spill.} \SN \GE{to become smaller} \XV{Dabuo da ba poɔrɔ kyɛ beri ayi ŋa o woɔ la o zaa.} \XE{Dabuo used to be plump, but recently he has gotten smaller.} \PD{woɔɛ, woɔrɔ, woɔrɔ, woɔreba, woɔraa, - } 
\LX{wɔɔ} \HM{} \PH{wɔ̀ɔ́} \PS{n.~} \GE{lame person} \OP{ wɔɔre} 
\LX{wɔɔ} \HM{} \PH{wɔ́ɔ́} \PS{v.~} \GE{to provide (food for the kitchen)} \PD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\LX{wɔɔ} \HM{} \PH{wɔ́ɔ́} \PS{v.~} \GE{to pluck all the fruits from a tree} \PD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\LX{wɔɔ} \HM{} \PH{wɔ́ɔ́} \PS{v.~} \GE{to destroy} \PD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\LX{wɔɔ} \HM{} \PH{wɔ́ɔ́} \PS{v.~} \GE{to whip mercilessly (figurative)} \PD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\LX{wɔɔ} \HM{} \PH{wɔ̀ɔ̀} \PS{v.~} \GE{to be lame} \PD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\LX{wɔɔ} \HM{} \PH{wɔ́ɔ̀} \PS{n.~} \GE{elephant} \OP{wɔɔre} \TP{wɔɔrɛɛ} 
\LX{wɔɔgbɛre} \HM{1} \PH{wɔ́ɔ̀gbɛ́rɪ̀} \PS{n.~} \GE{elephantiasis} 
\LX{wɔɔgbɛre} \HM{2} \PH{wɔ́ɔ̀gbɛ́rɪ̀} \PS{n.~} \SN \GE{leg of an elephant} \SN \GE{difficult task (figurative)} \OP{wɔɔgbɛɛ} \TP{wɔɔgbɛrɛɛ} 
\LX{wɔɔkye} \HM{} \PH{wɔ̀ɔ́ʧɪ̀} \PS{n.~} \GE{watch} \OP{wɔɔkyere} \TP{wɔɔkyerɛɛ} 
\LX{woɔle} \HM{} \PH{wʊ̀ɔ̀lɪ̀} \PS{v.~} \GE{to remove top matter; to skim e.g. yeast from pito} \PD{woɔlɛɛ, woɔla, woɔla, woɔleba, woɔlaa, - } 
\LX{wɔɔnu} \HM{} \PH{wɔ́ɔ́nù} \PS{n.~} \GE{trunk of an elephant} \OP{wɔɔnuuri} \TP{wɔɔnuuree} 
\LX{wɔɔnyene} \HM{} \PH{wɔ́ɔ́ɲɪ́nɪ̀} \PS{n.~} \GE{tusk of an elephant} \OP{wɔɔnyemɛ} \TP{wɔɔnyennɛɛ/wɔɔnyemɛɛ} 
\LX{woore} \HM{} \PH{wʊ̀ʊ̀rɪ́} \PS{n.~} \GE{suffering} 
\LX{wɔɔreŋ} \HM{} \PH{wɔ̀ɔ́rɪ́ŋ} \PS{n.~} \SN \GE{lameness} \SN \GE{the act of limping} \VA{wɔɔleŋ}\SE{} \GE{} 
\LX{wore} \HM{} \PH{wʊ́rɪ́} \PS{adv.~} \GE{too much; excessive} 
\LX{woroto} \HM{} \PH{wʊ̀rʊ̀tʊ̀} \PS{adv.~} \GE{crashingly} 
\LX{worɔworɔ} \HM{} \PH{wʊ̀rɔ̀wʊ̀rɔ̀} \PS{adv.~} \GE{full of lumps} 
\LX{woyaara} \HM{} \PH{wʊ̀yáárá} \PS{n.~} \GE{part of an animal-skin bag} \TP{woyaarɛɛ} 
\LX{wulee} \HM{} \PH{wúléé} \PS{n.~} \GE{whistle ; fife, flute} \OP{wubilii} \TP{wullee} 
\LX{wuli} \HM{} \PH{wùlì} \PS{v.~} \SN \GE{to show, to point to} \SN \GE{to direct, to guide} \SN \GE{to teach, to advise} \PD{wulee, wullo, wullo, wulibo, wulaa, wulibo } 
\LX{wuliwuli} \HM{} \PH{wùlwúlí} \PS{v.~} \GE{to rub over gently but fast} \PD{wuliwulee, wuliwulo, wuliwulo, wuliwulibo, wuliwulaa, - } 
\LX{wuliwullo} \HM{} \PH{wùlwúlló} \PS{n.~} \GE{teacher; one who directs} \OP{wulwulbo} \TP{wulwulbee} 
\LX{wuluu} \HM{} \PH{wúlúú} \PS{n.~} \SN \GE{teaching, education} \SN \GE{the act of showing} 
\LX{wuluu bonteɛre} \HM{} \PH{wúlúú bʊ́ntɪ́ɛ́rɪ́} \PS{n.~} \GE{teaching or learning materials} \SG{wuluu bonteɛraa} 
\LX{wuluu yelsɛgreminni} \HM{} \PH{wúlúú yèlsɛ́grɪ́bínní} \PS{n.~} \GE{lesson notes, lesson plan} 
\LX{wuo} \HM{} \PH{wùò} \PS{v.~} \GE{to gather, to collect} \PD{-, wuoro, wuoro, wuoribo, wuoraa, wuori } 
\LX{wuo} \HM{} \PH{wúò} \PS{v.~} \GE{to expose (a secret)} \PD{wuoe, wuoro, wuoro, wuoribo, wuoraa, - } 
\LX{wuo} \HM{} \PH{wúò} \PS{v.~} \GE{to formally introduce dirge singing at the funeral} \PD{wuoe, wuoro, wuoro, wuoribo, wuoraa, - } 
\LX{wuoli} \HM{} \PH{wúólì} \PS{v.~} \SN \GE{to despise, to hoot at} \SN \GE{to bark} \PD{wuolee, wuolo, wuolo, wuolibo, wuolaa, - } \VA{vuoli}
\LX{wuoraa} \HM{} \PH{wúóráá} \PS{n.~} \SN \GE{an implement for gathering} \SN \GE{a round-shaped piece of calabash with a hole in the middle used like a funnel in putting powdered tobacco into small containers} \OP{wuori} 
\LX{wurimo} \HM{} \PH{wúrímó} \PS{n.~} \GE{small, round cakes made from beans} \SG{wurimuu} \TP{wurimoree} 
\LX{wurowuro} \HM{} \PH{wùròwùrò} \PS{adv.~} \GE{not well cooked; underdone (round beans)} 
\LX{wuu} \HM{} \PH{wùù} \PS{adv.~} \SN \GE{continuously} \XV{Yuoni ŋa saa miire la wuu lɛ.} \XE{This year it has been raining unceasingly.} \SN \GE{having many instances} \XV{Zenɛ, koe deɛ waa la wuu lɛ.} \XE{Today, funerals are commonplace.} 
\LX{wuuri} \HM{} \PH{wúúrí} \PS{v.~} \SN \GE{to hoot (with laughter)} \SN \GE{to shout in excitement} \PD{wuuree, wuuro, wuuro, wuuribo, wuuraa, - } 
\LX{ya} \HM{} \PH{yá} \PS{v.~} \GE{to be mad, to be insane} \PD{yae,, yaara, yaara, yareba, yaaraa, - } 
\LX{yaa} \HM{} \PH{yàá} \PS{n.~} \GE{tomb} \OP{yaare} \TP{yarɛɛ} 
\LX{yaa} \HM{} \PH{yáà} \PS{v.~} \GE{used with 'sane'; to take revenge, to retaliate} \PD{yaaɛ, yaara, yaara, yaareba, yaaraa, yaareba } 
\LX{yaa} \HM{} \PH{yàà} \PS{interj.~} \GE{expression indicating disbelief} 
\LX{yaa} \HM{} \PH{yàà} \PS{v.~} \GE{remove from liquid} \PD{-, -, -, -, -, - } 
\LX{yaa} \HM{} \PH{yàà} \PS{v.~} \GE{to remove from liquid} \PD{yaaɛ, yaara, yaara, yaareba, yaaraa, - } 
\LX{yaa} \HM{} \PH{yàà} \PS{use word.~} \GE{please} \XV{Yaa, vɛŋe ka o meŋ tu o deme.} \XE{Oh, do allow him to tell his part of the story too.} 
\LX{yaa} \HM{} \PH{yáá} \PS{adv.~} \GE{really?} \XV{O yeli ka boŋ yaa?} \XE{What did she really say?.} 
\LX{yaa} \HM{} \PH{yáá} \PS{n.~} \GE{a strap; usually fixed to something for carrying e.g a bag} \OP{yaare} \TP{yaarɛɛ} 
\LX{yaale} \HM{} \PH{yáálɪ̀} \PS{v.~} \GE{to nurse (a baby)} \PD{yaalɛɛ, yaala, yaala, yaaleba, yaalaa, - } 
\LX{yaale} \HM{} \PH{yáálɪ̀} \PS{v.~} \GE{to swing a container in circles} \PD{yaalɛɛ, yaala, yaala, yaaleba, yaalaa, - } 
\LX{yaale} \HM{} \PH{yáálɪ̀} \PS{v.~} \GE{to get to hoplessness} \PD{yaalɛɛ, yaala, yaala, yaaleba, yaalaa, - } 
\LX{yaane} \HM{} \PH{yáánɪ́} \PS{n.~} \SN \GE{greeting(s)} \XV{Dagaaba ba deɛnɛ ne ba yaane puoruu.} \XE{The Dagaaba do not joke with their greetings.} \SN \GE{thanks, gratitude} \XV{Yɛ taa la yaane puoruu ne yɛ tonsonne yɛ naŋ toŋe.} \XE{You deserve our gratitude for your good work.} \TP{yaanɛɛ} 
\LX{yaane} \HM{} \PH{yáánɪ́} \PS{n.~} \GE{exhaustion} \XV{N koŋ la to᷉ɔ᷉ gaa yaane kpɛ ma la.} \XE{I can't go again; I'm tired.} \TP{yaanɛɛ} 
\LX{yaantulimo} \HM{} \PH{yáántúlímó} \PS{n.~} \GE{great-grandchild} \TP{yantulimee} 
\LX{yaara} \HM{} \PH{yààrá} \PS{adv.~} \GE{easily} \XV{Kyaanaa ba yaara ŋmaroo.} \XE{A mirror breaks easily.} 
\LX{yaare} \HM{} \PH{yààrɪ̀} \PS{v.~} \SN \GE{to spill, to scatter, to spread} \XV{A boɔraa la puri lɛnso ka a a kyi yaare a sori zu zaa.} \XE{The bag is burst and that is why the millet is scattered all over the road.} \SN \GE{to enlarge, to widen} \XV{A pelee ŋa noɔre yaare gaŋ la n boɔbo.} \XE{The mouth of this basket is too large for my liking.} \SN \GE{to perform final rites of widowhood} \XV{Ka banaŋ ba yaare kodɔɔ bee kopɔge o ba taa sori na toŋ sɛŋgaŋ toma.} \XE{Until the final widowhood rites are performed neither the widow nor the widower can engage in sex.} \SN \GE{to pester (typically used with noɔre)} \XV{Ba yaare la n noɔre ne ba pɔge na yɛlɛ ka N te bale.} \XE{They pestered me about their woman's case until I became fed up.} \PD{yaarɛɛ, yaara, yaara, yaareba, yaaraa, yaareba } 
\LX{yaaroŋ} \HM{} \PH{yààrʊ̂ŋ} \PS{n.~} \GE{salt} \VA{nyɛnnoŋ}
\LX{yabapa} \HM{} \PH{yàbàpà} \PS{adv.~} \GE{resoundingly e.g. noise of a large crowd} \XV{A pɔgebeɛlebɛ kyɛ la yieluŋ yabapa pa pa tu a sori gɛrɛ ne a pɔge.} \XE{The bridemaids sang earnestly along the road as they took the bride along.} 
\LX{Yaga} \HM{} \PH{yágá} \PS{n.~} \GE{name of a town in the Dagaaba land} 
\LX{yaga} \HM{} \PH{yágà} \PS{adj.~} \GE{many, much, plenty} \OP{yagere} \TP{yagerɛɛ} \VA{yage}
\LX{yage} \HM{} \PH{yágɪ̀} \PS{v.~} \SN \GE{to dredge to enlarge or deepen} \XV{Uoni saŋa Dagaaba maŋ yage la baa, a na nyɛ ko᷉ɔ᷉.} \XE{During the dry season, the Dagaaba dredge wells in order to get water.} \SN \GE{to increase in size e.g. a sore} \XV{Ka onaŋ ba gaa asibiti o fele na na yage la.} \XE{If she does not go to the hospital, her sore will develop into an ulcer.} \PD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } 
\LX{yage} \HM{} \PH{yàgɪ̀} \PS{v.~} \SN \GE{to take off e.g. a shirt} \SN \GE{to take down from a hanging position e.g. a bag from a hook} \PD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } \SE{yage bonsuuraa} \GE{to undress} 
\LX{yage} \HM{} \PH{yàgɪ̀} \PS{v.~} \GE{to show signs of butter forming e.g in the last stage in the preparation of shea butter} \PD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } \SE{yage bonsuuraa} \GE{to undress} 
\LX{yage} \HM{} \PH{yágɪ́} \PS{v.~} \GE{to remove objects from a container e.g. bag, hole} \XV{N ba boɔrɔ ka N yage a woɔ poɔ boma, a taa la dabe᷉ɛ.} \XE{I don't want to empty the contents of the bag because they are frightening.} \PD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } 
\LX{yage} \HM{} \PH{yágɪ́} \PS{v.~} \GE{to open one's eyes widely (in a concentrated look)} \XV{A bie nyɛ la a saana a deɛ yage nimie eŋ.} \XE{When the child saw the stranger, he gave him a fixed look.} \PD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } 
\LX{yagekɔɔlare} \HM{} \PH{yágɪ́kɔ́ɔ́lárɪ́} \PS{n.~} \GE{jawbone} \OP{yagekɔɔlaɛ} \TP{yagekɔɔlarɛɛ} \VA{yagekɔɔre}
\LX{yagekɔɔre} \HM{} \PH{yàgɪ̀kɔ́ɔ́rɪ́} \PS{n.~} \GE{cheek} \OP{yagekɔbɔ} \TP{yagekɔɔrɛɛ} \VA{yagekɔɔlare}
\LX{yagelaa} \HM{} \PH{yágɪ́láà} \PS{n.~} \SN \GE{a rope holder hung from the ceiling to hold objects} \SN \GE{a basket with calabashes that is hung on a yagelaa} \OP{yagele} \TP{yagelɛɛ} 
\LX{yagele} \HM{} \PH{yáglɪ̀} \PS{v.~} \GE{to hang (up)} \PD{yagelɛɛ, yagela, yagela, yageleba, yagelaa, - } 
\LX{yageraa} \HM{} \PH{} \PS{n.~} \GE{clay} \OP{yagere} \VA{yagere}
\LX{yagere} \HM{} \PH{yágrɪ́} \PS{v.~} \GE{to separate stones from grain} \PD{yagerɛɛ, yagera, yagera, yagereba, yageraa, - } 
\LX{yagere} \HM{} \PH{yágrɪ́} \PS{n.~} \GE{modelling clay} \TP{yagerɛɛ} \VA{yageraa}
\LX{yagere} \HM{} \PH{yàgrɪ́} \PS{n.~} \GE{cheek} \OP{yaga} \TP{yagrɛɛ} 
\LX{yakpolo} \HM{} \PH{yàkpólò} \PS{n.~} \GE{graveyard} \TP{yakpolee} 
\LX{yammo} \HM{} \PH{yámmʊ́} \PS{n.~} \SN \GE{bile} \SN \GE{gallbladder} \OP{yanne} \TP{yaanɛɛ} 
\LX{yaŋaa} \HM{} \PH{yáŋàá} \PS{n.~} \GE{grandchild} \OP{yaŋene/yanne, yaamine} \TP{yaŋenɛɛ/yannɛɛ} 
\LX{yan-yaa} \HM{} \PH{yán-yáà} \PS{n.~} \GE{insane person} \OP{yan-yaare} \TP{yan-yaarɛɛ} 
\LX{yanyaare} \HM{} \PH{yàɲáárɪ̀} \PS{n.~} \GE{madness} 
\LX{ya᷉ayiri} \HM{} \PH{yá᷉á᷉yírì} \PS{n.~} \GE{grandparent's house} \OP{ya᷉ayie} \TP{ya᷉ayiree} 
\LX{yɛ} \HM{} \PH{yɛ̀} \PS{pronoun.~} \GE{you (second person plural)} 
\LX{yɛ} \HM{} \PH{yɛ̀} \PS{v.~} \GE{to become exhausted, to get fed up} \PD{yɛɛ, yeɛrɛ, yeɛrɛ, yeɛrebɛ, yeɛraa, yɛyeɛrɛ } 
\LX{yɛɛ} \HM{} \PH{yɛ̀ɛ́} \PS{adj.~} \GE{spent out; reminants; not of substance (of crops that sprout from a previously cultivated field)} \OP{yɛre} \TP{yɛrɛɛ} 
\LX{yeebaa} \HM{} \PH{yɪ́ɪ́báá} \PS{n.~} \GE{twin} \OP{yeere} \TP{yeerɛɛ} 
\LX{yeelaa} \HM{} \PH{yɪ́ɪ́láá} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{yeele} 
\LX{yeele} \HM{} \PH{yɪ̀ɪ̀lɪ̀} \PS{v.~} \SN \GE{to make neat} \SN \GE{to use water to extract or separate wanted things from unwanted ones e.g. grain from stones} \PD{yeelɛɛ, yeelɛ, yeelɛ, yeelebɛ, yeelaa, - } 
\LX{yeele} \HM{} \PH{yɪ̀ɪ̀lɪ̀} \PS{adv.~} \GE{extremely clean or beautiful} 
\LX{yeɛle} \HM{} \PH{yɪɛlɪ} \PS{v.~} \GE{to winnow, to throw grain into the air to separate it from the husk} \XV{A dɔbɔ ŋmɛ la a kyi ka a pɔgeba te yeɛle.} \XE{The men threshed the millet and the women winnowed it.} \PD{-, yeɛlɛ, yeɛlɛ, yeɛlebɛ, -, - } 
\LX{yeɛle} \HM{} \PH{yɪɛlɪ} \PS{v.~} \GE{to come down in tiny bits e.g. a drizzle} \XV{A saa ba mi soŋ, yeɛloo ka o yeɛle.} \XE{It did not really rain; it just drizzled.} \PD{-, yeɛlɛ, yeɛlɛ, yeɛlebɛ, -, - } 
\LX{yeɛle} \HM{} \PH{yɪ́ɛ́lɪ̀} \PS{v.~} \GE{to follow in significant numbers} \XV{Ba nyɔge la nannyige ka noba deɛ yeɛle tu.} \XE{A thief was arrested and many people ran after the team.} \PD{-, yeɛlɛ, yeɛlɛ, yeɛlebɛ, -, - } 
\LX{yeere} \HM{} \PH{yɪ́ɪ́rɪ́} \PS{n.~} \GE{twins} 
\LX{yeere} \HM{} \PH{yɪ̀ɪ̀rɪ̀} \PS{v.~} \GE{to experience emotional distress, to suffer} \XV{O pɔge naŋ kpi zaa, o deɛ tori la yeeroo yeerɛ.} \XE{Since his wife's death he has been suffering emotional distress.} \PD{yeerɛɛ, yeerɛ, yeerɛ, yeerebɛ, yeeraa, - } 
\LX{yeɛre} \HM{} \PH{yɪ̀ɛ̀rɪ̀} \PS{v.~} \GE{to want attention, to act childishly} \PD{yeɛrɛɛ, yeɛrɛ, yeɛrɛ, yeɛrebɛ, yeɛraa, - } 
\LX{yeɛre} \HM{} \PH{yìɛ́rɪ́} \PS{n.~} \GE{underside of the jaw} \OP{yɛrɛɛ} 
\LX{yelbaaloŋ} \HM{} \PH{yèlbààlʊ́ŋ} \PS{n.~} \GE{an issue under control} 
\LX{yelbaboɔraa} \HM{} \PH{yèlbʊ́ɔ́ráà} \PS{n.~} \SN \GE{something that is sought, desire, wish} \SN \GE{theme (literature)} \OP{yelbaboɔre} 
\LX{yelbalee} \HM{} \PH{yèlbàlɪ̀ɪ̀} \PS{n.~} \GE{a particular issue} \XV{A yelbalee ŋa eŋɛ la ka o ne o pɔge dare zɔɔ taa.} \XE{It was because of this particular issue that he quarreled with his wife.} \VA{yelbalaa}
\LX{yelbalaa} \HM{} \PH{yèbáláá} \PS{n.~} \GE{a particular issue} 
\LX{yelbaŋ} \HM{} \PH{yèlbàŋ} \PS{n.~} \GE{the act of knowing} 
\LX{yelbaŋena} \HM{} \PH{yèbáŋɪ́ná} \PS{n.~} \GE{one who knows} \OP{yelbaŋenema} \TP{yelbaŋenemɛɛ} \VA{yelbanna}
\LX{yelbanna} \HM{} \PH{yèbánná} \PS{n.~} \GE{one who knows} 
\LX{yelbeɛ} \HM{} \PH{yèlbɪ́ɛ́} \PS{n.~} \GE{mysterious, miraculous thing} \OP{yelbeere} \TP{yelbeerɛɛ} 
\LX{yelbieri} \HM{} \PH{yèlbíérí} \PS{n.~} \SN \GE{evil thing} \SN \GE{sin} \OP{yelbebe} 
\LX{yelbiri} \HM{} \PH{yèlbírì} \PS{n.~} \SN \GE{word} \SN \GE{something that is alluded to} \OP{yelbie} 
\LX{yelbitɔnnaa} \HM{} \PH{yélbítɔ̀nnáà} \PS{n.~} \GE{conjunction} \OP{yelbitɔnne} \TP{yelbitɔnnɛɛ} 
\LX{yelboɔraa} \HM{} \PH{yèlbʊ́ɔ́ráá} \PS{n.~} \GE{something that is wanted} \OP{yelboɔre} 
\LX{yelboɔraa} \HM{} \PH{yèlbʋ́ráá} \PS{n.~} \SN \GE{that which one likes} \SN \GE{objective} \SN \GE{something negative that affects one but caused by one’s self} \VA{yeldaaraa}
\LX{yeldaaraa} \HM{} \PH{yèldàárà} \PS{n.~} \SN \GE{that which one likes} \SN \GE{objective} \SN \GE{something negative that affects one but caused by one’s self} \OP{yelboɔre} \TP{yeldaare} \VA{yelboɔraa}
\LX{yelbuli} \HM{} \PH{yèlbúlí} \PS{n.~} \GE{main point} \OP{yelbulo} 
\LX{yelgbɔge} \HM{} \PH{yèlgbɔ́gɪ́} \PS{n.~} \SN \GE{incomplete talk} \SN \GE{gr. clause} 
\LX{yeldaa} \HM{} \PH{yèldáà} \PS{n.~} \GE{simple sentence} \OP{yedaare} 
\LX{yeldeebo} \HM{} \PH{yéldɪ́ɪ́bʊ́} \PS{n.~} \GE{acceptance} 
\LX{yeldegere} \HM{} \PH{yèldɪ̀grɪ́} \PS{n.~} \GE{idiom} \OP{yeldegɛ} \TP{yeldɛgɛɛ} 
\LX{yeldɛgere} \HM{} \PH{yèldɛ́gɪ́rɪ̀} \PS{n.~} \GE{unpleasant or dirty talk} \OP{yeldɛgɛ} \TP{yeldɛgerɛɛ} 
\LX{yɛle} \HM{} \PH{yɛ́lɪ̀} \PS{n.~} \GE{cockroach} \OP{yɛlɛ} \TP{yɛlɛɛ} 
\LX{yɛle} \HM{} \PH{yɛ̀lɪ̀} \PS{v.~} \SN \GE{to multiply e.g. animals reproducing} \SN \GE{to enlarge} \PD{yɛlɛɛ, yɛllɛ, yɛllɛ, yɛlebɛ, yɛllaa, - } 
\LX{yɛlɛ} \HM{} \PH{yɛ́lɛ́} \PS{n.~} \SN \GE{an issue} \SN \GE{discussion, conversation} \SN \GE{serious matter} 
\LX{yel-eebo} \HM{} \PH{yèl-ɪ́ɪ́bʊ́} \PS{n.~} \GE{behaviour} 
\LX{yelɛnyelɛŋ} \HM{} \PH{yɪ́lɛɲɪ́lɛ́ŋ} \PS{adj.~} \GE{clear and clean (water)} 
\LX{yel-eraa} \HM{} \PH{yèl-ɪ́ráà} \PS{n.~} \GE{something to be done} \OP{yel-erre} 
\LX{yelfaa} \HM{} \PH{yèlfáà} \PS{n.~} \GE{evil} \OP{yelfaare} \TP{yelfaarɛɛ} 
\LX{yelfa-e} \HM{} \PH{yèlfà-ì} \PS{n.~} \GE{wrongdoing} \OP{yelfa-erre} 
\LX{yelfa-erɛ} \HM{} \PH{yèlfà-ɪ́rɛ́} \PS{n.~} \GE{evildoer} \OP{yelfa-erebɛ} \TP{yelfa-erebɛɛ} 
\LX{yelferaa} \HM{} \PH{yèlfɪ̀ráà} \PS{n.~} \GE{problem, something that matters} \OP{yelferre} \TP{yelferrɛɛ} 
\LX{yelfogilo} \HM{} \PH{yèlfógló} \PS{n.~} \GE{useless talk} \SG{yelfogiluu} 
\LX{yelfoli} \HM{} \PH{yèlfòlí} \PS{n.~} \GE{obscene talk} \OP{yelfolo} 
\LX{yelfuoni} \HM{} \PH{yèlfùònì} \PS{n.~} \GE{imagery (literature)} \OP{yelfuomo} \TP{yelfuomee} 
\LX{yelgalema} \HM{} \PH{yèlgálɪ́má} \PS{n.~} \GE{strange thing} \TP{yelgalemɛɛ} \VA{yelgalee}
\LX{yeli} \HM{} \PH{yélí} \PS{n.~} \SN \GE{something of concern} \XV{Ta bɔ yeli ko o, vɛŋe o ka o gɛrɛ.} \XE{Don't put him into any trouble; allow him to go.} \SN \GE{matter} \XV{Yeli ne o yelmeŋɛ N zãa ba gaa be.} \XE{As a matter of fact I did not go there.} \OP{yɛlɛ} 
\LX{yeli} \HM{} \PH{yélí} \PS{n.~} \GE{sentence} \XV{Fo na baŋ de la eronwuluu yelbiri maale ne yɛlɛ?} \XE{Can you construct a sentence with an adverb?.} \OP{yɛlɛ} 
\LX{yeli} \HM{} \PH{yèlì} \PS{v.~} \SN \GE{to speak, to talk, to inform} \XV{N teɛre ka o yeli la a yɛlɛ ko o ma.} \XE{I guess she has has informed her mother about it.} \SN \GE{to rebuke} \XV{O daare yeli la o pɔgeyaa lɛɛ ka o te kono.} \XE{He rebuked his daughter until she wept.} \PD{yelee, yele, yele, yelibe, yelaa, yelibe } \SE{waale yeli} \GE{to whisper} 
\LX{yelkaama} \HM{} \PH{yèlkáámá} \PS{n.~} \SN \GE{beautiful talk} \SN \GE{literature} \TP{yelkaamɛɛ} 
\LX{yelkabeɛ} \HM{} \PH{yélkábɪ́ɛ́} \PS{n.~} \GE{strange thing} \OP{yelkabeere} \TP{yelkabeerɛɛ} 
\LX{yelka᷉asɛgere} \HM{} \PH{yèlká᷉á᷉sɛ́grɪ́} \PS{n.~} \GE{written literature} 
\LX{yelka᷉a᷉yelli} \HM{} \PH{yèlká᷉á᷉yéllí} \PS{n.~} \GE{oral literature} 
\LX{yelkpeɛŋaa} \HM{} \PH{yèlkpɪ́ɛ́ŋàá} \PS{n.~} \GE{difficult thing} \OP{ yelkpeɛne} 
\LX{yel-lige} \HM{} \PH{yélígè} \PS{n.~} \GE{a mystery} \XV{Lɛ a pɔge naŋ e saŋe a baaloŋ e la ye-lige.} \XE{The way the woman was able to cure the patient is a mystery.} 
\LX{yel-lige} \HM{} \PH{yèl-lígè} \PS{adj.~} \GE{hidden} \XV{A dɔɔ na maŋ tona la yel-lige toma.} \XE{That man performs mysterious deeds.} \OP{yel-ligri} 
\LX{yelmaalaa} \HM{} \PH{yèlmààláà} \PS{n.~} \GE{rite, ritual} \OP{yelimaale} 
\LX{yelmaalaa} \HM{} \PH{yèlmààláà} \PS{n.~} \GE{something that needs to rectified} \OP{yelimaale} 
\LX{yelmaaroŋ} \HM{} \PH{yèlmáárʊ́ŋ} \PS{n.~} \GE{peaceful talk} 
\LX{yelmannaa} \HM{} \PH{yèlmànnáà} \PS{n.~} \SN \GE{comparison} \SN \GE{narrative, narration} \OP{yelmanne} \TP{yelmannɛɛ} 
\LX{yelmanne} \HM{} \PH{yélmánnɪ́} \PS{n.~} \GE{the act of narrating} 
\LX{yelmeŋɛ} \HM{} \PH{yèlmɪ́ŋɛ́} \PS{n.~} \GE{truth} \OP{yelimenne} \TP{yelimennɛɛ} 
\LX{yelnimiri} \HM{} \PH{yènímírì} \PS{n.~} \GE{main idea} 
\LX{yelnɔnaa} \HM{} \PH{yèlnɔ́náà} \PS{n.~} \GE{something that is liked} \OP{yelinɔnne} 
\LX{yelnoɔ} \HM{} \PH{yèlnʊ́ɔ́} \PS{n.~} \SN \GE{interesting thing} \SN \GE{good news} \OP{yelinoore} \TP{yelinoorɛɛ} 
\LX{yɛloŋ} \HM{} \PH{yɛ́lʊ́ŋ} \PS{adj.~} \GE{wide} \OP{yɛlle} 
\LX{yelpaala} \HM{} \PH{yèlpáálà} \PS{n.~} \GE{new thing} \OP{yelipaale} 
\LX{yelpɛgere} \HM{} \PH{yèlpɛ́grɪ́} \PS{n.~} \GE{sayings that have no value} \OP{yelpɛgɛ} 
\LX{yelsaana} \HM{} \PH{yèlsáánà} \PS{n.~} \GE{strange or new thing} \OP{yelisaama} \TP{yelisaamɛɛ} 
\LX{yelsaana} \HM{} \PH{yèlsááná} \PS{n.~} \SN \GE{destroyer} \SN \GE{one who meddles in issues} \OP{ yelisaanema} 
\LX{yelsaanaa} \HM{} \PH{yèlsáánáá} \PS{n.~} \GE{something that has gone wrong} \OP{yelisnaane} \TP{yelisaanɛɛ} 
\LX{yelsɔgelaa} \HM{} \PH{yèlsɔ́gláà} \PS{n.~} \GE{mysterious thing, secret} \OP{yelisɔgelɔ} 
\LX{yelsoŋ} \HM{} \PH{yèlsʊ̀ŋ} \PS{n.~} \GE{good deed} \OP{yelisonne} 
\LX{yeltegeroŋ} \HM{} \PH{yèltɪ́grʊ́ŋ} \PS{n.~} \GE{difficult position} \OP{yelitegerɛ} 
\LX{yeltɔgelaa} \HM{} \PH{yèltɔ̀gláà} \PS{n.~} \GE{that which can be imitated} \OP{yelitɔgele} 
\LX{yeltɔre} \HM{} \PH{yèltɔ̀rɪ́} \PS{n.~} \GE{something of worth or value} 
\LX{yeltoloŋ} \HM{} \PH{yèltʊ́lʊ́ŋ} \PS{n.~} \GE{serious matter, something that needs urgent attention} \OP{yelitolɔ} 
\LX{yeltulimo} \HM{} \PH{yèltúli̊mó} \PS{n.~} \GE{opposite} \VA{yelitulo}
\LX{yeltuluu} \HM{} \PH{yèltúlúú} \PS{n.~} \GE{mistaken talk, error} 
\LX{yeltuo} \HM{} \PH{yèltúó} \PS{n.~} \GE{difficult thing} \OP{yelituuri} \TP{yelituuree} 
\LX{yeltuuraa} \HM{} \PH{yèltùùráà} \PS{n.~} \GE{rite, tradition} \OP{yelituuri} 
\LX{yeluu} \HM{} \PH{yélúú} \PS{n.~} \GE{saying} 
\LX{yelwiiraa} \HM{} \PH{yèlwííráá} \PS{n.~} \GE{poem} \OP{yeliwiiri} 
\LX{yelwiiri} \HM{} \PH{yèlwíírí} \PS{n.~} \GE{poetry} 
\LX{yelwonaa} \HM{} \PH{yèlwònáà} \PS{n.~} \SN \GE{that which is a challenge; problem} \SN \GE{something that is heard} \OP{yeliwonni} 
\LX{yelwono} \HM{} \PH{yèlwónó} \PS{n.~} \SN \GE{listener, hearer} \SN \GE{obedient person} \OP{yeliwonimo} 
\LX{yelwontaa} \HM{} \PH{yèlwòntàá} \PS{n.~} \GE{compromise, understanding} 
\LX{yelyelbaraa} \HM{} \PH{yèlyèlbàráà} \PS{n.~} \GE{statement (grammar)} \OP{yelyelbarre} 
\LX{yelyeluu} \HM{} \PH{yèlyélúú} \PS{n.~} \GE{issues that cause anger} 
\LX{yelzagera} \HM{} \PH{yèlzågrá} \PS{n.~} \GE{one who refuses, one who disobeys} \OP{yelzagereba} 
\LX{yɛmbaŋ} \HM{} \PH{yɛ́mbàŋ} \PS{n.~} \GE{cleverness, wisdom} 
\LX{yɛmbaŋena} \HM{} \PH{yɛ́mbáŋɪ́ná} \PS{n.~} \GE{one who has good sense, wise person} \OP{yɛmbaŋnema} 
\LX{yɛmbile} \HM{} \PH{yɛ́mbílé} \PS{n.~} \GE{small intelligence} \OP{yɛmbilii} \TP{yɛmbillee} 
\LX{yeme} \HM{} \PH{yèmé} \PS{n.~} \GE{slave} \OP{yenni} \TP{yennee} \VA{yieme}
\LX{yieme} \HM{} \PH{yìèmé} \PS{n.~} \GE{slave} 
\LX{yempɔge} \HM{} \PH{yèmpɔ́gɪ́} \PS{n.~} \SN \GE{female slave} \SN \GE{name traditionally given to a female child born outside the house} \OP{yelmpɔgeba} \TP{yempɔgrɛɛ} 
\LX{yeŋ} \HM{} \PH{yêŋ} \PS{adv.~} \GE{where} 
\LX{yɛŋ} \HM{} \PH{yɛ̂ŋ} \PS{adv.~} \GE{kindly e.g. in an appeal} \XV{Yɛŋ de a bie gaa ne sakuuri.} \XE{Try and take the child to school.} 
\LX{yendɔɔ} \HM{} \PH{yèndɔ́ɔ́} \PS{n.~} \SN \GE{male slave} \SN \GE{name traditionally given to a male child born outside the house} 
\LX{yeŋi} \HM{} \PH{yèŋí} \PS{adv.~} \GE{outside} \VA{yeŋe}
\LX{yeŋe} \HM{} \PH{yèŋé} \PS{adv.~} \GE{outside} \VA{yeŋi}
\LX{yɛ} \HM{} \PH{yɛ̀} \PS{v.~} \GE{become  tired} \XV{Ka a tontonmaŋ wa yɛ ba iri.} \XE{When the workers are tired they can stop.} \PD{-, -, -, -, -, - } \VA{ yɛɛ}
\LX{yɛ} \HM{} \PH{yɛ̀} \PS{adj.~} \GE{tired} \XV{Maa eŋ yɛɛ la ne a zeɛre ŋa.} \XE{ I am tired of this sauce.} \VA{ yɛɛ}
\LX{yɛŋe} \HM{} \PH{yɛ̂ŋ} \PS{adj.~} \GE{intelligent} \XV{A pɔgelee na taa la yɛŋe.} \XE{That girl is intelligent.} 
\LX{yɛŋe} \HM{} \PH{yɛ̂ŋ} \PS{n.~} \GE{intelligence, knowledge} \XV{A bie yɛŋ ba seŋ bone.} \XE{The child has very little knowledge.} \OP{yɛnne} \TP{yɛnnɛɛ} 
\LX{yeni} \HM{} \PH{yénì} \PS{adj.~} \GE{one} \OP{yeme} \TP{yennee} 
\LX{yɛnkyeerɛ} \HM{} \PH{yɛ́nʧɪ́ɪ́rɛ́} \PS{n.~} \GE{little mind} 
\LX{yɛnkyeeroo} \HM{} \PH{yɛ́nʧɪ́ɪ́rʊ̊ʊ́} \PS{n.~} \GE{prudence} 
\LX{yɛnlige} \HM{} \PH{yɛ́nlígè} \PS{n.~} \GE{unclear mind} 
\LX{yɛnnyaa} \HM{} \PH{yɛ́nɲáà} \PS{n.~} \GE{the act of developing intelligence} 
\LX{yensogɔ} \HM{} \PH{yénsʊ́gɔ́} \PS{n.~} \GE{night} \OP{yensogere} \TP{yensogerɛɛ} \VA{yensoga, tensogɔ, tensoga}
\LX{yɛnwuli} \HM{} \PH{yɛ́nwúlì} \PS{n.~} \GE{the act of teaching sense} 
\LX{yɛnwuluu} \HM{} \PH{yɛ́nwúlúú} \PS{n.~} \GE{the act of teaching sense or conscience} 
\LX{yɛnyɛŋ} \HM{} \PH{yɛ́ɲɛ́ŋ} \PS{adj.~} \GE{near} 
\LX{yɛre} \HM{} \PH{yɛ̀rɪ̀} \PS{v.~} \SN \GE{to wear, to dress} \XV{Ba zaa yɛre la tigiri bonsuuri.} \XE{They all dressed their best for the feast.} \SN \GE{to hang around the neck and shoulders} \XV{O yɛre bɔnzɔl velaa kaŋa.} \XE{She wore a beautiful chain.} \PD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\LX{yɛre} \HM{} \PH{yɛ̀rɪ̀} \PS{v.~} \GE{to trade, to do business} \XV{Ka fooŋ na yɛre ferebo la ka fo taa libima.} \XE{If you want to engage in business, you must have some capital.} \PD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\LX{yɛre} \HM{} \PH{yɛ́rɪ̀} \PS{v.~} \GE{to spread out e.g. to dry} \PD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\LX{yɛre} \HM{} \PH{yɛ́rɪ̀} \PS{v.~} \GE{to hunt in a group} \PD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\LX{yɛre} \HM{} \PH{yɛ́rɪ́} \PS{adj.~} \GE{wild; used with crops e.g groundnuts} \XV{A bibiiri gaa pɛ la seŋkãa yɛɛ.} \XE{The children went harvesting wild groundnuts.} \OP{yɛɛ} \TP{yɛrɛɛ} 
\LX{yɛreyɛrɛ} \HM{} \PH{yɛ̀ryɛ́rɛ́} \PS{n.~} \GE{trader, businessperson, merchant} \OP{yɛreyɛrebɛ} \TP{yɛreyɛrebɛɛ} 
\LX{yɛroŋ} \HM{} \PH{yɛ́rʊ́ŋ} \PS{n.~} \GE{trade, business} 
\LX{yi} \HM{} \PH{yí} \PS{v.~} \SN \GE{to go out, to exit} \XV{Te naŋ yi ka te gaa a kpaaroŋ la ka saa deɛ wa miire.} \XE{As we went out to go to the farm, it started raining.} \SN \GE{to divorce (by a woman)} \XV{A dɔɔ nɔŋ la o pɔge ŋmeɛbo lɛnso ka o yi.} \XE{The man likes beating his wife and that is why she has divorced him.} \PD{yie, yire, yire, yiribe, yiraa, yiribe } 
\LX{yi} \HM{} \PH{yí} \PS{v.~} \GE{to be clean} \XV{Fo ba pɛge a laa ne fanfane a zuiŋ o ba yi.} \XE{You did not wash the bowl with soap and so it is not clean.} \PD{yie, yire, yire, yiribe, yiraa, yiribe } 
\LX{yi} \HM{} \PH{yí} \PS{v.~} \GE{to resemble} \XV{A bie ŋa yi la o ma yaga.} \XE{This child resembles the mother very much.} \PD{yie, yire, yire, yiribe, yiraa, yiribe } 
\LX{yi} \HM{} \PH{yí} \PS{v.~} \GE{of land;  become infertile, to be barren e.g land} \XV{Ba kɔ la a koɔlɔ ŋa ka o te yi, ka fooŋ kɔ o a ba eŋ kuoluŋ fo koŋ nyɛ bonezaa.} \XE{Several years of farming on this land has made it infertile and if you farm it without applying fertilizer to it you won’t get any good yield.} \PD{yie, yire, yire, yiribe, yiraa, yiribe } 
\LX{yi} \HM{} \PH{yí} \PS{v.~} \GE{to lose taste e.g. beer} \XV{Ba naŋ zaameŋ ba pɔge a da᷉a᷉ noɔre na la so ka o yi.} \XE{Because the beer was not covered, it has lost its taste.} \PD{yie, yire, yire, yiribe, yiraa, yiribe } 
\LX{yi kyeɛ} \HM{} \PH{yí ʧɪ́ɛ́} \PS{adj.~} \GE{uninteresting, boring} 
\LX{yi kyeɛ} \HM{} \PH{yí ʧɪ́ɛ́} \PS{v.~} \GE{to fade} \PD{-, -, -, -, -, - } 
\LX{yi sori} \HM{} \PH{yi23 sórí} \PS{v.~} \GE{to travel, to go on a trip} \PD{-, -, -, -, -, - } 
\LX{yidaampɔge} \HM{} \PH{yídààmpɔ́gɪ́} \PS{n.~} \GE{head woman in a family; mistress of a compound} \OP{yidaampɔgeba} \TP{yidaampɔgrɛɛ} 
\LX{yidaana} \HM{} \PH{yídááná} \PS{n.~} \GE{house owner} \OP{yidaama} \TP{yidaanɛɛ/yidaamɛɛ} 
\LX{yidaandɔɔ} \HM{} \PH{yídààndɔ́ɔ́} \PS{n.~} \GE{male head of a family; master of of compound} \OP{yidaandɔbɔ} \TP{yidaandɔɔrɛɛ} 
\LX{yideme} \HM{} \PH{yídèmé} \PS{n.~} \GE{family group; family members} \SG{yidemuu} \TP{yidemee} 
\LX{yieli} \HM{} \PH{yíélì} \PS{v.~} \GE{to sing} \PD{yielee, yiele, yiele, yielibe, -, - } 
\LX{yieluŋ} \HM{} \PH{yíélúŋ} \PS{n.~} \GE{song; music} \OP{yiele} 
\LX{yieri} \HM{} \PH{yìèrì} \PS{v.~} \GE{to extend, to spread out} \XV{Ka dabeɛŋ wa kpɛ karaa, o maŋ yieri la o zu ne o kɔkɔrɛɛ.} \XE{When a cobra is frightened, it spreads out its head and neck.} \PD{yieree, yiere, yiere, yieribe, yieraa, - } 
\LX{yigileŋ} \HM{} \PH{yígléŋ} \PS{adv.~} \GE{perched in a lonely manner} \XV{A bie dɔŋɛɛ la ka o zu deɛ waa yigileŋ lɛ.} \XE{The child has grown so lean that his head is literally perching on his neck.} 
\LX{yinni} \HM{} \PH{yìnnì} \PS{v.~} \SN \GE{to forget} \XV{N yeli la a kpaaroŋ yɛlɛ ko fo ta erɛ wa yinni.} \XE{I told you about the meeting, so don't forget.} \SN \GE{to glean (farm field)} \XV{Ka banaŋ wa pɛ seŋka᷉a᷉ baare bibiiri maŋ gaa te yinni la a gbannaa poɔ.} \XE{After groundnuts are harvested, children would normally go to the field to glean.} \PD{yinnee, yinne, yinne, yinime, yinnaa, yinime } \VA{inni}
\LX{yiri} \HM{} \PH{yírì} \PS{n.~} \SN \GE{house, home, compound} \SN \GE{family, family members} \OP{yie} \TP{yiree} 
\LX{yivilime} \HM{} \PH{yívílímé} \PS{n.~} \GE{round compound} 
\LX{yi᷉e} \HM{} \PH{yì᷉è᷉} \PS{v.~} \GE{to faint} \PD{yi᷉ee, yiene, yiene, yienime, yienaa, - } 
\LX{yi᷉eu} \HM{} \PH{yí᷉é᷉ú᷉} \PS{n.~} \GE{spectacle, wonders} \OP{yieme} 
\LX{yɔ} \HM{} \PH{yɔ́} \PS{v.~} \GE{to wander, to roam, to travel} \PD{yɔɛ, yoɔrɔ, yoɔrɔ, yoɔreba, -, - } 
\LX{yobie} \HM{} \PH{yóbíé} \PS{n.~} \GE{ill luck, bad name} 
\LX{yofaa} \HM{} \PH{yófáá} \PS{n.~} \GE{a name that stigmatizes} \OP{yofaare} \TP{yofaarɛɛ.} 
\LX{yofuu} \HM{} \PH{yófúù} \PS{n.~} \GE{abstract noun} \OP{yofuuri} \TP{yofuuree} 
\LX{yɔge} \HM{} \PH{yɔ̀gɪ̀} \PS{v.~} \GE{to loosen e.g. hard substance (soil)} \PD{yɔgɛɛ, yɔgrɔ, yɔgrɔ, yɔgrebɔ, yɔgraa, - } 
\LX{yɔgele} \HM{} \PH{yɔ́glɪ̀} \PS{v.~} \GE{to be near, to approach (time)} \XV{Se᷉ɛ᷉ yɔgelɔ la ka suŋkpulo kyoɔrɔ.} \XE{The farming season is coming round; hence, dawdaw trees are flowering.} \PD{yɔgelɛɛ, yɔgelɔ, yɔglɔ, yɔgeleba, yɔgelaa, - } 
\LX{yɔgere} \HM{} \PH{yɔ̀grɪ́} \PS{n.~} \GE{farm field around the compound} \OP{yɔgɔ} \TP{yɔgɛɛ} 
\LX{yɔgere} \HM{} \PH{yɔ́grɪ̀} \PS{v.~} \GE{to loosen e.g of a string tied to something} \PD{yɔgerɛɛ, yɔgerɔ, yɔgerɔ, yɔgereba, yɔgeraa, - } 
\LX{yɔgere} \HM{} \PH{yɔ́grɪ́} \PS{v.~} \GE{to be loose, to be slack} \XV{Ba leŋ la a tuobu kyɛ ka a leɛ yɔgere.} \XE{The load was tied but it is loose again.} \PD{-, -, -, -, -, - } 
\LX{yɔgevaa} \HM{} \PH{yɔ́gɪ̀váárɪ̀} \PS{n.~} \GE{pumpkin leaf} \OP{yɔgevaare} \TP{yɔgevaarɛɛ} 
\LX{yɔgevaare} \HM{} \PH{yɔ̀gɪ́váárɪ̀} \PS{n.~} \GE{pumpkin leaves used as vegetable} \TP{yɔgevaarɛɛ} 
\LX{yɔgevaare} \HM{} \PH{yɔ̀gváárɪ̀} \PS{n.~} \GE{leaves of pumpkin} \SG{yɔgevaabo} \OP{yɔgevaarɛɛ} 
\LX{yɔgevaare-waabo} \HM{} \PH{yɔ̀gɪ̀váárɪ̀-wáábʊ̀} \PS{n.~} \GE{green mamba} \OP{yɔgevaare-weere} \TP{yɔgevaare-weerɛɛ} 
\LX{yogi} \HM{} \PH{yógí} \PS{v.~} \GE{to eat, especially plenty food that is tasty} \PD{yogee, yogro, yogro, yogro, yogribo, yograa } \VA{mogi}
\LX{yogi} \HM{} \PH{yògì} \PS{adj.~} \GE{enough, of considerable quantity} \XV{A zɛvaare ba yogi.} \XE{There is not enough of the vegetable.} 
\LX{yogili} \HM{} \PH{yòglì} \PS{v.~} \GE{to pamper, to treat kindly or gently} \PD{yoglee, yoglo, yoglo, yoglibo, yoglaa, - } 
\LX{yogoko} \HM{} \PH{yógókó} \PS{adv.~} \GE{very sweet} 
\LX{yɔgoo} \HM{} \PH{yɔ̀gʊ́ʊ̊} \PS{n.~} \SN \GE{pumpkin} \SN \GE{the plant which bears the fruit} \OP{yɔgɔ} 
\LX{yɔgoo} \HM{} \PH{yɔ̀gʊ́ʊ̊} \PS{n.~} \SN \GE{a type of large, white ant} \SN \GE{the home of these ants} \OP{yɔgɔ} 
\LX{yo-iri} \HM{} \PH{yó-írì} \PS{n.~} \GE{glory, praise, honour,} 
\LX{yɔkɔɔloŋ} \HM{} \PH{yɔ̌kɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{pubic hair of a male} 
\LX{yɔle} \HM{} \PH{yɔ́lɪ́} \PS{v.~} \GE{to slip off as a result of being loose e.g. shorts} \PD{yɔlɛɛ, yɔlɔ, yɔlɔ, yɔlɔ, yɔlaa, - } 
\LX{yololo} \HM{} \PH{yólóló} \PS{adv.~} \GE{very cold} 
\LX{yɔlpelaa} \HM{} \PH{yɔ̀lpɪ̀láà} \PS{n.~} \GE{a type of tree with edible fruits} \OP{yɔlpeɛle} 
\LX{yoluŋ} \HM{} \PH{yólúŋ} \PS{n.~} \GE{cone-shaped cage} \OP{yolli} \TP{yollee} 
\LX{yoluŋ} \HM{} \PH{yólúŋ} \PS{adj.~} \GE{emptiness e.g wearing a blouse or shirt without a singlet or a vest under it} \OP{yolli} \TP{yolee} 
\LX{yɔlzeɛ} \HM{} \PH{yɔ̀lzɪ́ɛ́} \PS{n.~} \GE{a type of tree with grey bark} \OP{yɔlzeere} 
\LX{yombo} \HM{} \PH{yómbó} \PS{n.~} \GE{black hair dye} \OP{yombori} \TP{yomboree} 
\LX{yɔmɔ} \HM{} \PH{yɔ́mɔ́} \PS{adj.~} \GE{free and plentiful} \OP{yɔnne} 
\LX{yoŋ} \HM{} \PH{yôŋ} \PS{adj.~} \SN \GE{alone} \SN \GE{only} \OP{yonni} 
\LX{yontaa} \HM{} \PH{yòntáå} \PS{n.~} \SN \GE{wife in a polygamous marriage} \SN \GE{relationship between wives in a polygamous marriage} \SN \GE{enemy (figurative)} \OP{yontaare} \TP{yontaarɛɛ} 
\LX{yoo} \HM{} \PH{yóó} \PS{interjection.~} \SN \GE{okay} \XV{Yoo, bieo te na gaa la.} \XE{Okay tomorrow we shall go.} \SN \GE{be informed} \XV{A toma ŋa fo yoŋiŋ so a.} \XE{Be informed that that work is for you alone.} 
\LX{yɔɔ} \HM{} \PH{yɔ́ɔ̀} \PS{n.~} \GE{sibling (used between individuals of the same sex)} \OP{yɔɔmine/yɔɔre} \TP{yɔɔrɛɛ} 
\LX{yɔɔ} \HM{} \PH{yɔ́ɔ̀} \PS{n.~} \GE{strophantus (a type of tree with highly poisonous sap)} \OP{yɔɔre} \TP{yɔɔrɛɛ} 
\LX{yɔɔ} \HM{} \PH{yɔ́ɔ́} \PS{v.~} \GE{to pay, to redeem} \PD{yɔɔɛ, yɔɔrɔ, yɔɔrɔ, yɔɔreba, yɔɔraa, yɔɔre } 
\LX{yɔɔ} \HM{} \PH{yɔ̀ɔ̀} \PS{adv.~} \GE{freely e.g. of liquid being poured} 
\LX{yoɔbo} \HM{} \PH{yʊ́ɔ́bʊ̀} \PS{v.~} \SN \GE{the act of roaming, wandering} \SN \GE{hunting expedition} \PD{-, -, -, -, -, - } 
\LX{yoɔbo} \HM{} \PH{yʊ́ɔ́bʊ́} \PS{n.~} \GE{the stoppage of rain} 
\LX{yoɔbo} \HM{} \PH{yʊ̀ɔ̀bʊ́} \PS{adj.~} \GE{short form of six} 
\LX{yɔɔdɔɔ} \HM{} \PH{yɔ́ɔ́dɔ́ɔ́} \PS{n.~} \GE{brother (used by females to address male siblings)} \OP{yɔɔdɔɔbɔ} \TP{yɔɔdɔɔrɛɛ} 
\LX{yoɔle} \HM{} \PH{yʋ̀ɔ̀lɪ̀} \PS{v.~} \GE{to escape slowly e.g. from a punctured tyre} \PD{-, -, -, -, -, - } 
\LX{yoɔle} \HM{} \PH{yʋ́ɔ́lɪ́} \PS{v.~} \GE{to melt, to disintegrate slowly e.g salt or sugar in sauce or tea} \PD{-, -, -, -, -, - } 
\LX{yɔɔlee} \HM{} \PH{yɔ́ɔ́léé} \PS{n.~} \GE{younger sibling (used between individuals of the same sex)} \OP{yɔɔbilii} 
\LX{yoɔne} \HM{} \PH{yʊ́ɔ́nɪ́} \PS{n.~} \GE{a type of small, spotted amphibean that looks like a frog and is found in wet sand} \OP{yoɔmɔ/yoɔma} \TP{yoɔnɛɛ} 
\LX{yɔɔpɔge} \HM{} \PH{yɔ́ɔ́pɔ́gɪ́} \PS{n.~} \GE{sister (used by males to address female siblings)} \OP{yɔɔpɔgebɔ/yɔɔpɔgeba.} \VA{yɔɔpuulee}
\LX{yɔɔpuulee} \HM{} \PH{yɔ́ɔ̀púúléé} \PS{n.~} \GE{female sibling (usually less older) see kpɪ́ɛ́mɛ́} \OP{yɔɔpuuli} \VA{yɔɔpɔge}
\LX{Yoɔraa} \HM{} \PH{yʋ́ɔ́ráá} \PS{n.~} \GE{a traditional name given to a child believed to be reincarnated} 
\LX{yoɔraa} \HM{} \PH{yʊ́ɔ́ráà} \PS{n.~} \GE{tourist, one who roams} \OP{yoɔrebɔ/yoɔreba} 
\LX{yoɔre} \HM{} \PH{yʊ̀ɔ̀rɪ́} \PS{n.~} \GE{penis} \OP{yɔɛ} \TP{yɔrɛɛ} 
\LX{yoɔre} \HM{} \PH{yʊ̀ɔ̀rɪ̀} \PS{v.~} \SN \GE{to cause to be exhausted or emotionless} \SN \GE{to bother, irritate; to pester someone} \PD{yoɔrɛɛ, yoɔrɔ, yoɔrɔ, yoɔreba, yoɔraa, - } 
\LX{yɔɔre} \HM{} \PH{yɔ́ɔ́rɪ̀} \PS{n.~} \GE{payment} \TP{yɔɔrɛɛ} 
\LX{yopɔgeraa} \HM{} \PH{yòpɔ̀gràà} \PS{n.~} \GE{fishing pot} \OP{yopɔgere} \VA{yokyigiraa}
\LX{yokyigiraa} \HM{} \PH{yòkyìgìráà} \PS{n.~} \GE{fishing pot} 
\LX{yopɔgeraa} \HM{} \PH{yòpɔ́gɪ́ráá} \PS{n.~} \GE{lid, cover of a pot} 
\LX{yopo᷉i} \HM{} \PH{yòpó᷉i᷈} \PS{adj.~} \GE{short form of ayoɔpõi (seven)} 
\LX{yɔre} \HM{} \PH{yɔ́rɪ̀} \PS{v.~} \GE{to fall apart, to disintegrate} \PD{yɔrɛɛ, yɔrɔ, yɔrɔ, yɔreba, yɔraa, - } 
\LX{yɔre} \HM{} \PH{yɔ̀rɪ̀} \PS{v.~} \SN \GE{to become intense e.g. rainfall} \SN \GE{to become widespread (disease)} \SN \GE{that which is plenty} \PD{yɔrɛɛ, yɔrɔ, yɔrɔ, yɔreba, yɔraa, - } 
\LX{Yɔrebaa} \HM{} \PH{yɔ̀rɪ́báá} \PS{n.~} \GE{Yoruba (an ethnic group found in Nigeria)} \SG{yɔrebaaroo} \OP{yɔrebaare} \TP{yɔrebaarɛɛ} 
\LX{yɔrebɔnɔ} \HM{} \PH{yɔ̀rɪ̀bɔ̀nɔ́} \PS{adv.~} \GE{loosely} 
\LX{yoreyoree} \HM{} \PH{yʊ̀rɪ̀yʊ́rɪ́} \PS{v.~} \GE{to break into little bits} \PD{-, -, -, -, -, - } 
\LX{yori} \HM{} \PH{yòrì} \PS{v.~} \GE{to make a loop} \PD{yoree, yoro, yoro, yoribo, yoraa, - } 
\LX{yori} \HM{} \PH{yórì} \PS{v.~} \GE{to dismantle a loop} \PD{yoree, yoro, yoro, yoribo, yoraa, - } 
\LX{yorombiri} \HM{} \PH{yórómbírí} \PS{n.~} \GE{herring} \OP{yorombie} \TP{yorombireeyorombiiree} 
\LX{yorɔyorɔ} \HM{} \PH{yʊ́rɔ́yʊ́rɔ́} \PS{adv.~} \GE{smallish; usually in size} \VA{yorayora}
\LX{yɔrɔyɔrɔ} \HM{} \PH{yɔ̀rɔ̀yɔ̀rɔ̀} \PS{adj.~} \GE{not firm, loose e.g. mud} 
\LX{yosa᷈a} \HM{} \PH{yósà᷈à᷈} \PS{n.~} \GE{slander} \OP{yosa᷉a} 
\LX{yoyogibiri} \HM{} \PH{yóyògìbírì} \PS{n.~} \GE{a type of small flying insect that makes tambalaaraa (a type of honey)} \SG{yoyogibiruu} \OP{yoyogibie} \TP{yoyogibiree/yoyogibiiree} 
\LX{yuli} \HM{} \PH{yúlí} \PS{v.~} \GE{to leak slowly} \PD{yulee, yulo, yulo, yulibo, yulaa, yulibo } 
\LX{yuloo} \HM{} \PH{yúlóó} \PS{n.~} \GE{object on which bows are hung} \OP{yulli} \TP{yullee} 
\LX{yuo} \HM{} \PH{yùò} \PS{v.~} \GE{to open} \PD{yuoe, yuoro, yuoro, -, yuoraa, - } 
\LX{yuo} \HM{} \PH{yúó} \PS{adj.~} \GE{different, of another kind} \XV{Zenɛ kpare yuo la ka a baala su.} \XE{Today the patient is wearing a different shirt.} 
\LX{yuoni ŋa} \HM{} \PH{yúóní ŋá} \PS{adv.~} \GE{this year} 
\LX{yuonaa} \HM{} \PH{yùònáá} \PS{adv.~} \GE{this year} \VA{shortened form of yuoni ŋa}
\LX{yuoni} \HM{} \PH{yúóní} \PS{n.~} \GE{year} \OP{yuomo} \TP{yuonee} 
\LX{Yuoni} \HM{} \PH{yúóní} \PS{n.~} \GE{name given to a child born after nine months of pregnancy} 
\LX{yuoni} \HM{} \PH{yúóní} \PS{n.~} \GE{the crop of a bird} \OP{yuomo} \TP{yuonee} 
\LX{yuoni-yuoni} \HM{} \PH{yúóní-yúóní} \PS{adv.~} \GE{annually} 
\LX{yuori} \HM{} \PH{yúórí} \PS{n.~} \GE{jar} \OP{yoe} \TP{yoree} 
\LX{yuori} \HM{} \PH{yǔórì} \PS{n.~} \SN \GE{name} \SN \GE{reputation} \SN \GE{namesake} \OP{yoe} \TP{yoree} 
\LX{yuuli} \HM{} \PH{yùùlì} \PS{v.~} \GE{to spread small amount on top of e.g. dry cement} \PD{yuulee, yuulo, yuulo, yuulo, yuulaa, - } 
\LX{za} \HM{} \PH{zá} \PS{v.~} \GE{to throw} \PD{zaɛ, zaara, zaara, zareba, zaaraa, zareba } 
\LX{za} \HM{} \PH{zá} \PS{v.~} \GE{to palpitate (of heart)} \PD{zaɛ, zaara, zaara, zareba, zaaraa, zareba } 
\LX{zaa} \HM{} \PH{zàá} \PS{adj.~} \GE{all, every} \XV{A kõɔ zaa ka ba nyu.} \XE{They have drank all the water.} \VA{haa}
\LX{zaale} \HM{} \PH{záálɪ́} \PS{v.~} \GE{to move water around in a container by swinging the container around} \PD{zaalɛɛ, zaala, zaala, zaaleba, zaalaa, - } 
\LX{zaale} \HM{} \PH{záálɪ́} \PS{v.~} \GE{to commit suicide by hanging oneself} \PD{zaalɛɛ, zaala, zaala, zaaleba, zaalaa, - } 
\LX{zaameŋ} \HM{} \PH{zààmɪ̂ŋ} \PS{adv.~} \GE{yesterday} \VA{za᷉a}
\LX{zaayee} \HM{} \PH{zàáyéé} \PS{adv.~} \GE{surprisingly} \XV{N da teɛre ka o be la a dieŋ zaayee a bie yoŋ la.} \XE{I thought she was in the room, but to my surprise it was only the child that was there.} 
\LX{zabɔge} \HM{} \PH{zábàgɪ́} \PS{n.~} \GE{foe, enemy, invader} \OP{zabɔgre} \TP{zabɔgrɛɛ} \VA{zabage}
\LX{zabage} \HM{} \PH{zábágɪ́} \PS{n.~} \GE{foe, enemy, invader} \VA{zabɔge}
\LX{zaboraa} \HM{} \PH{zábʊ́ráá} \PS{n.~} \GE{person under review} \OP{zaborre} \TP{zaborɛɛ} 
\LX{zaga} \HM{} \PH{zágà} \PS{n.~} \SN \GE{animal or poultry pen} \SN \GE{mortuary} \OP{zagere} \TP{zagerɛɛ} \VA{zage}
\LX{zage} \HM{} \PH{zágɪ́} \PS{n.~} \GE{head cold} \TP{zagrɛɛ} 
\LX{zage} \HM{} \PH{zágɪ́} \PS{v.~} \GE{to be worn out} \PD{zagɛɛ, zagra, zagra, zagreba, zagraa, - } 
\LX{zage} \HM{} \PH{zàgɪ́} \PS{n.~} \GE{enclosure for animals, pen, kraal} \OP{zagere} \TP{zagerɛɛ} \VA{zaga}
\LX{zagedie} \HM{} \PH{zàgɪ̀dìè} \PS{n.~} \GE{main chamber in a rodent's burrow} \OP{zagederi} \TP{zagederee} 
\LX{zagela} \HM{} \PH{záglá} \PS{n.~} \GE{one being referred to} \XV{Zagela maŋ yeli ka o puoro la kyɛ zeɛre o taaba.} \XE{Some people claim to be godly but hate their neigbours.} 
\LX{zagelaa} \HM{} \PH{zágláà} \PS{adj.~} \GE{empty} \OP{zagele} \TP{zagelɛɛ} 
\LX{zagepɛmpɛlɛŋ} \HM{} \PH{zàgɪ̀pɛ́mpɛ́lɛ́ŋ} \PS{n.~} \GE{spleen} \OP{zagepɛmpɛlle} \TP{zagepɛmpɛlɛɛ} \VA{hɔɔpɛmpɛlɛŋ}
\LX{zagere} \HM{} \PH{zàgrɪ̀} \PS{v.~} \GE{to refuse, to reject} \PD{zagerɛɛ, zagera, zagera, zagereba, zageraa, zagreba } 
\LX{zagezage} \HM{} \PH{zàgzág} \PS{v.~} \GE{to harass} \PD{zagezagɛɛ, zagezagera, -, zagezagereba, zagezageraa, - } 
\LX{zalball} \HM{} \PH{zàlbàll} \PS{adv.~} \GE{airborne} 
\LX{zalle} \HM{} \PH{zállɪ̀} \PS{v.~} \SN \GE{to swing} \SN \GE{to commit suicide by hanging oneself} \PD{zallɛɛ, zalla, zalla, zaleba, zallaa, zaleba } \VA{zɛlle}
\LX{zɛlle} \HM{} \PH{zɛ́llɪ̀} \PS{v.~} \SN \GE{to swing} \SN \GE{to commit suicide by hanging oneself} \PD{-, -, -, -, -, - } \VA{zalle}
\LX{zamaa} \HM{} \PH{zàmàà} \PS{adj.~} \GE{confused} 
\LX{zamasoba} \HM{} \PH{zámásúbá} \PS{n.~} \GE{one that is homeless} 
\LX{zambaa} \HM{} \PH{zàmbáá} \PS{n.~} \GE{a type of small striped bush mouse} \OP{zambaare} \TP{zambaarɛɛ} \VA{zambateraa}
\LX{zambadaa} \HM{} \PH{zàmbàdàá} \PS{n.~} \GE{stilt} \OP{zambadaare} \TP{zambadaarɛɛ} 
\LX{zambateraa} \HM{} \PH{zàmbàtɪ̀ràà} \PS{n.~} \GE{a type of striped mouse} \OP{zambaterre} \TP{zambaterɛɛ} \VA{zamba}
\LX{zambayɛlɛ} \HM{} \PH{zámbáyɛ́lɛ̀} \PS{adv.~} \GE{lying in a manner with face up} \VA{zamayɛlɛ}
\LX{Zambɔ} \HM{} \PH{zàmbɔ́} \PS{n.~} \GE{name of a village near Lawra in the Upper West Region of Ghana} 
\LX{zambɛre} \HM{} \PH{zàmbɛ̀rɪ̀} \PS{n.~} \GE{blacksmith’s shop} \OP{zambɛrre} 
\LX{zamm} \HM{} \PH{zàmm} \PS{adv.~} \GE{huge but light} 
\LX{zammo} \HM{} \PH{zàmmʊ́} \PS{n.~} \GE{onion} \OP{zama} \TP{zannɛɛ} \VA{gaabu (Hausa)}\VA{zamadaare}
\LX{zammuli} \HM{} \PH{zàmmúlí} \PS{n.~} \GE{tree} 
\LX{zamparemuu} \HM{} \PH{zâmpárɪ̀mùú} \PS{n.~} \GE{wrist guard for archery} \OP{zamparemuuri} \TP{zamparemuuree} 
\LX{zampoŋ} \HM{} \PH{zàmpóŋ} \PS{n.~} \GE{hedgehog} \OP{zamponni} \TP{zamponnee} 
\LX{zampoone} \HM{} \PH{zàmpʊ́ʊ́nɪ́} \PS{adj.~} \GE{young and helpless e.g. chick} \OP{zampooma} \TP{zampoomɛɛ} 
\LX{zaŋaa} \HM{} \PH{záŋáá} \PS{n.~} \GE{shell} \OP{zanne} \TP{zannɛɛ} 
\LX{zane} \HM{} \PH{zánɪ́} \PS{n.~} \GE{wrist guard for archery} \OP{zama} \TP{zamɛɛ} 
\LX{zaŋe} \HM{} \PH{záŋ} \PS{v.~} \GE{to itch} \PD{zaŋɛɛ, zaŋena, zaŋena, zaŋeneba, zaŋenaa, - } 
\LX{zaŋgalaŋ} \HM{} \PH{záŋgáláŋ} \PS{n.~} \GE{veranda} \OP{zaŋgalle} \TP{zaŋgalɛɛ} 
\LX{zaŋgbaale} \HM{} \PH{zàŋgbààlɪ̀} \PS{n.~} \GE{dome-shaped cage} \OP{zaŋgbaala} \TP{zaŋgbaalɛɛ} 
\LX{zaŋkpegiri} \HM{} \PH{zàŋkpégrí} \PS{n.~} \GE{a type of hard pimple} \OP{zaŋkpege} \TP{zaŋkpegee} \VA{kpaŋkpazegele}
\LX{zanne} \HM{} \PH{zánnɪ́} \PS{v.~} \GE{to thresh neatly or completely} \PD{zannɛɛ, zanna, zanna, zanema, -, - } 
\LX{zanne} \HM{} \PH{zànnɪ́} \PS{v.~} \GE{to learn} \PD{zannɛɛ, zanna, zanna, zanema, zannaa, zanema } 
\LX{zanne} \HM{} \PH{zànnɪ́} \PS{v.~} \GE{to dream} \PD{zannɛɛ, zanna, zanna, zanema, zannaa, zanema } 
\LX{zannoŋ} \HM{} \PH{zànnʊ̌ŋ} \PS{n.~} \GE{dream} \OP{zanonne} \TP{zanonnɛɛ} \VA{zannoo}
\LX{zannoo} \HM{} \PH{zánnʋ́ʋ́} \PS{n.~} \GE{dream} \VA{zannoŋ}
\LX{zannoo} \HM{} \PH{zànnʊ́ʊ́} \PS{n.~} \GE{education, the act of learning} 
\LX{zannyuu} \HM{} \PH{zànɲúú} \PS{n.~} \GE{garlic} \OP{zannyuuri} \TP{zannyuuree} 
\LX{zannzanna} \HM{} \PH{zànnzánná} \PS{n.~} \GE{learner, pupil} \SG{zannzanemɛɛ} \TP{zannzanema} 
\LX{zanzakpiri-duoruŋ} \HM{} \PH{zánzákpìri-dúórúŋ} \PS{n.~} \GE{ringworm} 
\LX{zanzakpirime} \HM{} \PH{zánzákpìrìmé} \PS{n.~} \GE{small bat} \OP{zanzakpirri} \TP{zanzakpirree} \VA{zinzakpirime}
\LX{zanzaŋ} \HM{} \PH{zánzáŋ} \PS{adv.~} \GE{intensely e.g. heat} \XV{A yɛlɛ pa᷉a᷉ ta la a zanzaŋ.} \XE{The matter has now got to the climax.} 
\LX{zanzaŋaa} \HM{} \PH{zánzáŋàá} \PS{n.~} \GE{bat} \OP{zanzanne} \TP{zanzannɛɛ} 
\LX{zanzaŋaa} \HM{} \PH{zánzáŋàá} \PS{n.~} \SN \GE{a type of shellfish similar to a clam} \SN \GE{the shell of  this shelfish} \OP{zanzanne} \TP{zanzannɛɛ} 
\LX{zanzaŋkpara} \HM{} \PH{zànzáŋkpàrà} \PS{n.~} \GE{the act of jumping up and down in hysteria} \OP{zanzaŋkparre} \TP{zanzaŋkparɛɛ} 
\LX{zanzansɔglaa} \HM{} \PH{zánzánsɔ́glàá} \PS{n.~} \GE{a type of blackish-grey fruit bat} \OP{zanzansoglɔ} \TP{zanzansɔglɛɛ} 
\LX{zanzantokele} \HM{} \PH{zánzàntókélé} \PS{n.~} \GE{a type of bat} \OP{zanzantokelli} \TP{zanzantokelee} 
\LX{zanzanzeɛ} \HM{} \PH{zánzànzɪ́ɛ́} \PS{n.~} \GE{a type of reddish-brown bat} \OP{zanzanzeere} \TP{zanzanzeerɛɛ} 
\LX{zanzunmuli} \HM{} \PH{zaǹzùmmùlì} \PS{n.~} \GE{a type of tree with medicinal properties} 
\LX{zare} \HM{} \PH{zárɪ̀} \PS{n.~} \GE{large metal hammer used in the forge} \OP{ zaɛ} \TP{zarɛɛ} 
\LX{zarema} \HM{} \PH{zárɪ́má} \PS{n.~} \GE{rags} \TP{zaremɛɛ} \VA{zɛremɛ}
\LX{zɛremɛ} \HM{} \PH{zɛ́rɪ́mɛ́} \PS{n.~} \GE{rags} \VA{zarema}
\LX{zazaakɔɔ} \HM{} \PH{zàzáákɔ́ɔ́} \PS{adj.~} \GE{steady} \VA{dedeekɔɔ}
\LX{za᷉a} \HM{} \PH{zá᷉à} \PS{adv.~} \GE{yesterday; short form of zaameŋ} \VA{zaameŋ}
\LX{za᷉a} \HM{} \PH{zà᷉à᷉à} \PS{adv.~} \GE{hazy} 
\LX{zɛ} \HM{} \PH{zɛ̀} \PS{v.~} \GE{to smear} \PD{zɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, zeɛrebɛ } 
\LX{zɛ} \HM{} \PH{zɛ́} \PS{v.~} \GE{to sift} \PD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\LX{zɛ} \HM{} \PH{zɛ́} \PS{v.~} \GE{to hate} \PD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\LX{zɛdambo} \HM{} \PH{zɛ̀dàmbó} \PS{n.~} \GE{unseasoned soup} \OP{zɛdambori} \TP{zɛdamboree} 
\LX{zɛdogelee} \HM{} \PH{zɛ̀dʊ́gléé} \PS{n.~} \GE{earthenware cooking pot for soup} \OP{zɛdogebilii} \VA{saa-doge}
\LX{saa-doge} \HM{} \PH{sààdʋ́gɪ̀} \PS{n.~} \GE{earthenware cooking pot for soup} \VA{zɛdogelee}
\LX{zeɛ} \HM{} \PH{zɪ̀ɛ́} \PS{n.~} \GE{late millet} \SG{zeroo (for plant)} \OP{zeere} \TP{zeerɛɛ} 
\LX{zeɛ} \HM{} \PH{zɪ̀ɛ̀} \PS{v.~} \GE{to snatch, to seize} \PD{zeɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, zeɛrebɛ } 
\LX{zeɛ} \HM{} \PH{zɪ̀ɛ́} \PS{adj.~} \GE{reddish brown} \OP{zeere} \TP{zerɛɛ} 
\LX{zeɛ} \HM{} \PH{zɪ̀ɛ́} \PS{adj.~} \GE{serious} \XV{Yelzeɛ la ka ba yele, azuiŋ ta laara.} \XE{It is a serious case, so don't laugh.} \OP{zeere} \TP{zerɛɛ} 
\LX{zɛɛ} \HM{} \PH{zɛ̀ɛ̀} \PS{adv.~} \GE{sticking on to all over e.g. fruits on a tree} 
\LX{zeɛ-genegene} \HM{} \PH{zɪ̀ɛ́-gènègènè} \PS{adj.~} \GE{brown} 
\LX{zeɛmaa} \HM{} \PH{zɪ́ɛ́máá} \PS{n.~} \GE{open ground where millet is gathered, threshed, and winnowed out} \OP{zeɛne} \TP{zeɛnɛɛ/zeɛmɛɛ} \VA{gbeŋgbelaa}
\LX{Zeɛmɛ} \HM{} \PH{zɪ́ɛ́mɛ̀} \PS{n.~} \GE{name traditionally given to the firstborn of twins (male)} 
\LX{zeɛraa} \HM{} \PH{zɪ́ɛ́ráá} \PS{n.~} \GE{sieve} \OP{zeɛre} \TP{zeɛrɛɛ} 
\LX{zeɛraa} \HM{} \PH{zɪ́ɛ́ráá} \PS{n.~} \GE{hawk} \OP{zeɛre} \TP{zeɛrɛɛ} 
\LX{zeɛre} \HM{} \PH{zɪ̀ɛ̀rɪ́} \PS{n.~} \SN \GE{sauce} \SN \GE{soup} \OP{ zɛrɛɛ} 
\LX{zeɛre} \HM{} \PH{zɪ́ɛ́rɪ̀} \PS{v.~} \GE{to hate, to despise} \PD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\LX{zeɛre} \HM{} \PH{zɪ́ɛ́rɪ̀} \PS{v.~} \GE{to trot} \PD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\LX{zeɛre} \HM{} \PH{zɪ́ɛ́rɪ̀} \PS{v.~} \GE{to be in haste, to go after something eagerly} \PD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\LX{zege} \HM{} \PH{zɪ̀gɪ̀} \PS{v.~} \GE{to grow to maturity and full strength} \PD{zegɛɛ, zegerɛ, zegerɛ, zegerebɛ, zegeraa, - } 
\LX{zegɛ} \HM{} \PH{zɪ́gɛ́} \PS{n.~} \SN \GE{area behind and between the shoulders} \SN \GE{hump (of a cow)} \OP{zegere} \TP{zegerɛɛ} 
\LX{zɛge} \HM{} \PH{zɛ́gɪ̀} \PS{v.~} \SN \GE{to raise, to lift above} \SN \GE{to promote} \PD{zɛgɛɛ, zɛgrɛ, zɛgrɛ, zɛgrebɛ, zɛgraa, - } 
\LX{zɛgelɛkɛlɛ} \HM{} \PH{zɛ́glɛ́kɛ́lɛ́} \PS{adv.~} \GE{not missing out} \XV{A pɔgeba zaa ne ba biiri da la, zɛgelɛkɛlɛ.} \XE{Each of the women had a child.} 
\LX{zegere} \HM{} \PH{zɪ́grɪ̀} \PS{v.~} \GE{to shrug; raise one’s shoulders, indicating lack of concern} \PD{zegerɛɛ, zegerɛ, zegerɛ, zegerebɛ, zegeraa, - } 
\LX{zegere} \HM{} \PH{zɪ̀grɪ̀} \PS{v.~} \GE{to grow to maturity} \PD{zegerɛɛ, zegerɛ, zegerɛ, zegerebɛ, zegeraa, - } 
\LX{zɛgere} \HM{} \PH{zɛ́grɪ̀} \PS{v.~} \SN \GE{to shake (hands)} \SN \GE{to shake in the hand to weigh} \PD{zɛgerɛɛ, zɛgerɛ, zɛgerɛ, zɛgerebɛ, zɛgeraa, - } 
\LX{zɛgeroŋ} \HM{} \PH{zɛ̊grʊ́ŋ} \PS{n.~} \SN \GE{something that has worth} \SN \GE{something that has scent} 
\LX{zegezege} \HM{} \PH{zɪ̀gzɪ̀g} \PS{v.~} \GE{to be impatient} \PD{zegezegɛɛ, zegezegrɛ, zegezegerɛ, zegezegerebɛ, zegezegeraa, - } 
\LX{zegitege} \HM{} \PH{zègìtégé} \PS{adv.~} \GE{not moving at all, standing still in the way} 
\LX{zɛko᷉ɔ} \HM{} \PH{zɛ̀kʊ́᷉ɔ́᷉} \PS{n.~} \GE{soup that is not thick; broth-like} \OP{zɛkoɔnɛɛ} 
\LX{zele} \HM{} \PH{zɪ̀lɪ́} \PS{n.~} \GE{tongue} \OP{zelɛ} \TP{zelɛɛ/zellɛɛ} 
\LX{zɛle} \HM{} \PH{zɛ́lɪ́} \PS{v.~} \SN \GE{to apply poison; to be filled with venom} \SN \GE{to be furious} \SN \GE{to be bitter (taste)} \PD{-, -, -, -, -, - } 
\LX{zɛle} \HM{} \PH{zɛ́lɪ́} \PS{n.~} \GE{afterbirth (nonhuman)} \OP{zɛlɛ} \TP{zɛlɛɛ} 
\LX{zɛle} \HM{} \PH{zɛ̀lɪ̀} \PS{v.~} \GE{to lift, to pick up} \PD{zɛlɛɛ, zɛllɛ, zɛllɛ, zɛlebɛ, zɛllaa, zɛlebɛ } 
\LX{zɛle} \HM{} \PH{zɛ̀lɪ̀} \PS{v.~} \GE{to beg (Northern Dagaare dialect)} \PD{zɛlɛɛ, zɛllɛ, zɛllɛ, zɛlebɛ, zɛllaa, zɛlebɛ } 
\LX{zelegannoɔre} \HM{} \PH{zɪ̀lɪ́gànʊ́ɔ́rɪ̀} \PS{n.~} \GE{a type small bird with blue feathers and a long beak that is fond of flowers} \OP{zelegannɔɛ} 
\LX{zɛlɛŋkpɛ} \HM{} \PH{zɛ̀lɛ̀ŋkpɛ̀} \PS{adj.~} \GE{hanging loosely} 
\LX{zelɛzelɛ} \HM{} \PH{zɪ̀lɛ́zɪ̀lɛ́} \PS{adj.~} \GE{slim, at the point of giving way} 
\LX{zɛlezɛle} \HM{} \PH{zɛ̀lɪ̀zɛ́lɪ́} \PS{v.~} \GE{to move a heavy load with difficulty} \PD{zɛlezɛlɛɛ, zɛlezɛlɛ, zɛlezɛlɛ, zɛlezɛlebɛ, zɛlezɛlaa, - } 
\LX{zelgannoɔre} \HM{} \PH{zɪ̌lgǎnnʊ́ɔ́rɪ̀} \PS{n.~} \GE{a type of small bird with a very long curved beak} \OP{zelgannɔɛ} \TP{zelgannɔrɛɛ} 
\LX{zell} \HM{} \PH{zɪ́ll} \PS{adv.~} \GE{filled to the brim} 
\LX{zelpi᷉i} \HM{} \PH{zɪ̌pí᷉ì᷉} \PS{n.~} \GE{a type of disease that affects the tongue of dogs causing them to not eat} \OP{zelpiime} \TP{zelpiimee} 
\LX{zɛmbɛre} \HM{} \PH{zɛ̀mbɛ̀rɪ̀} \PS{n.~} \SN \GE{a blacksmith’s shop} \SN \GE{god of the forge} \OP{ zɛmbɛrɛɛ} \VA{zambɛre}
\LX{zambɛre} \HM{} \PH{zàmbɛ̀rɪ̀} \PS{n.~} \SN \GE{a blacksmith’s shop} \SN \GE{god of the forge} \VA{zɛmbɛre}
\LX{zemberemɛ} \HM{} \PH{zɪ́mbɪ́rɪ́mɛ́} \PS{n.~} \GE{a type of tree whose fruit gives red juice} \OP{zemberre} \TP{zemberɛɛ/zemberrɛɛ} 
\LX{zɛmɛzɛmɛ} \HM{} \PH{zɛ̀mɛ̀zɛ̀mɛ̀} \PS{adv.~} \GE{long and flexible e.g. cane} 
\LX{zɛmiime} \HM{} \PH{zɛ̀míímé} \PS{n.~} \GE{sour sauce} 
\LX{zɛmm} \HM{} \PH{zɛ̀mm} \PS{adv.~} \GE{quietly, gently, and slowly} 
\LX{Zempɛnne} \HM{} \PH{zɪ̀mpɛ́nnɪ́} \PS{n.~} \GE{name of a village} 
\LX{zɛŋ} \HM{} \PH{zɛ̀ŋ} \PS{v.~} \GE{to lift} \PD{zɛŋɛɛ, zɛnnɛ, zɛnnɛ, zɛnemɛ, zɛnnaa, - } 
\LX{zenɛ} \HM{} \PH{zɪ́nɛ̀} \PS{n.~} \GE{today} 
\LX{zeŋe} \HM{} \PH{zɪ̀ŋ} \PS{v.~} \SN \GE{to sit, to be seated} \SN \GE{to settle} \PD{zeŋɛɛ, zenɛ, zenɛ, zenemɛ, zenaa, zenemɛ } 
\LX{zɛŋe} \HM{} \PH{zɛ̂ŋ} \PS{n.~} \GE{to shake e.g. tree} \OP{ zenɛ  zenzenɛ  zenemɛ} 
\LX{zenɛbie} \HM{} \PH{zɪ́nɛ̀bìé} \PS{adv.~} \GE{nowadays, the present time} 
\LX{zeŋgbiri} \HM{} \PH{zɪ̀ŋgbírì} \PS{n.~} \GE{sleeping sickness} 
\LX{zɛnzɛŋ} \HM{} \PH{zɛ̀nzɛ́ŋ} \PS{v.~} \SN \GE{to be impatient} \SN \GE{to be eager, to be zealous} \PD{zɛnzɛŋɛɛ, zɛnzɛnɛ, zɛnzɛnɛ, zɛnzɛnemɛ, zɛnzɛnaa, zɛnzɛnemɛ } 
\LX{zɛnzɛŋ} \HM{} \PH{zɛ̀nzɛ̌ŋ} \PS{n.~} \GE{impatience} \OP{zɛnzɛnne} \TP{zɛnzɛnɛɛ} 
\LX{zenzie} \HM{} \PH{zɪ̀nzíé} \PS{n.~} \SN \GE{sitting place} \SN \GE{settlement} \OP{zenziiri} \TP{zenziiree} 
\LX{zɛrekpɛrɛ} \HM{} \PH{zɛ̀rɪ̀kpɛ̀rɛ̀} \PS{adv.~} \GE{hanging or swelling all over e.g. fruits} 
\LX{zɛroo} \HM{} \PH{zɛ̀rʊ̀ʊ̀} \PS{adv.~} \GE{fleshy and loose (hanging)} 
\LX{zɛvaa} \HM{} \PH{zɛ́váà} \PS{n.~} \GE{vegetable} \SG{zɛvaabo} \OP{zɛvaare} \TP{zɛvaarɛɛ} 
\LX{ze᷈e} \HM{} \PH{zɪ́᷈ɪ́᷈} \PS{n.~} \GE{blood} \TP{zeenɛɛ} 
\LX{ze᷉ɛ} \HM{} \PH{zɪ̀᷉ɛ́᷉} \PS{n.~} \GE{acid} \TP{zeɛnɛɛ} 
\LX{ze᷉ɛ} \HM{} \PH{zɪ́᷉ɛ́᷉} \PS{n.~} \GE{fortification for one who has commited murder or killed an animal believed to have supernatural powers} \TP{zeɛnɛɛ} 
\LX{ze᷉ɛ} \HM{} \PH{zɪ̀᷉ɛ̀} \PS{v.~} \GE{to be needed, to be necessary} \PD{ze᷉ɛɛ, zeɛnɛ, zeɛnɛ, zeɛnemɛ, zeɛnaa, - } 
\LX{ze᷉ɛda᷉a} \HM{} \PH{zí᷉ɛ́᷉dá᷉á᷉} \PS{n.~} \GE{local beer prepared solely for ritual purposes} \TP{ze᷉ɛdaanɛɛ} 
\LX{ze᷉ɛgoɔ} \HM{} \PH{zɪ́᷉ɛ́᷉gʊ́ɔ̀} \PS{n.~} \GE{a type of tree with medicinal properties} \OP{ze᷉ɛgoore} 
\LX{Ze᷉ɛpɔge} \HM{} \PH{zɪ̀᷉ɛ́᷉pɔ́gɪ́} \PS{n.~} \GE{name traditionally given to the firstborn of twins (female)} \OP{Ze᷉ɛpɔgemine} 
\LX{ze᷈ete᷉e} \HM{} \PH{zɪ́᷈ɪ̀᷈tɪ́᷈ɪ́᷈} \PS{n.~} \GE{medicine that helps improve the level of blood in the body} \TP{ze᷉eteenɛɛ} 
\LX{ze᷉ewoɔ} \HM{} \PH{zɪ́᷉ɪ́᷉wʊ́ɔ́} \PS{n.~} \GE{blood vessel} \OP{ze᷉ewoore} \TP{ze᷉ewoorɛɛ} 
\LX{zii} \HM{} \PH{zìì} \PS{adj.~} \GE{thick; foliage of a tree. (Used only in this situation)} 
\LX{zibalaa} \HM{} \PH{zìbàlàà} \PS{n.~} \GE{odd place} 
\LX{zibalaa} \HM{} \PH{zìbáláá} \PS{n.~} \GE{detour, diversion (road)} \OP{ziballe} 
\LX{ziberaa} \HM{} \PH{zìbèráà} \PS{n.~} \GE{the place where a person or something is} \OP{ ziberri} 
\LX{ziberaa} \HM{} \PH{zìbɪ̀ráà} \PS{n.~} \GE{late night} \OP{ziberre} 
\LX{zibiri} \HM{} \PH{zìbírì} \PS{n.~} \GE{clitoris} \OP{zibie} \TP{zibiiree} \VA{gyambiri}
\LX{gyambiri} \HM{} \PH{ʣàbírì} \PS{n.~} \GE{clitoris} \VA{zibiri}
\LX{zie} \HM{} \PH{zìé} \PS{n.~} \GE{place} \OP{ziiri} \TP{ziiree} 
\LX{zie} \HM{} \PH{zìé} \PS{n.~} \GE{weather} \OP{ziiri} \TP{ziiree} 
\LX{zie} \HM{} \PH{zìé} \PS{n.~} \GE{time} \OP{ziiri} \TP{ziiree} 
\LX{ziezaa} \HM{} \PH{zíézàá} \PS{adv.~} \GE{everywhere} 
\LX{zigi} \HM{} \PH{zígí} \PS{n.~} \GE{python} \OP{zigri} \TP{zigree/zigee} \VA{zibu}
\LX{zibu} \HM{} \PH{zíé} \PS{n.~} \GE{python} \VA{zigi}
\LX{zigi} \HM{} \PH{zígí} \PS{v.~} \SN \GE{to grow in size} \SN \GE{to grow more leaves (yam vine)} \PD{zigee, zigire, zigire, zigiribe, zigiraa, - } 
\LX{zigiree} \HM{} \PH{zígréé} \PS{n.~} \GE{gravel, laterite} \OP{zigiri} 
\LX{zigizukpara} \HM{} \PH{zìgìzúkpárá} \PS{n.~} \GE{a monster; in the form of a two-headed python} \OP{zigizukparre} 
\LX{zigoo} \HM{} \PH{zìgòò} \PS{adj.~} \GE{showing no concern for others} \XV{A bie kono  tuuro la o ma kyɛ ka o deɛ ba taa zigoo zaa.} \XE{The child is crying while running after the mother but woman did not show any concern.} 
\LX{zikpeɛraa} \HM{} \PH{zìkpɪ́ɛ́ráà} \PS{n.~} \GE{inhabitance, residence} \OP{zikpeɛre} 
\LX{zikuoŋaa} \HM{} \PH{zìkúóŋáá} \PS{n.~} \GE{dry place} \OP{zikuoni} \TP{zikuonee} 
\LX{zikyɛnfeɛle} \HM{} \PH{zìʧɛ̀nfɪ́ɛ́lɪ́} \PS{n.~} \GE{cold early morning} \OP{zikyɛnfeɛlɛ} 
\LX{zimaale} \HM{} \PH{zìmààlɪ̀} \PS{n.~} \GE{development} 
\LX{zimaaliire} \HM{} \PH{zìmáálììré} \PS{n.~} \GE{dusk, twilight (after sunset)} \OP{zimaaliiri} \TP{zimaaliiree} 
\LX{zimaane} \HM{} \PH{zìmáánɪ́} \PS{adv.~} \GE{evening} 
\LX{zimaaroŋ} \HM{} \PH{zìmàárʊ́ŋ} \PS{n.~} \GE{cool or wet place} \OP{zimaara} 
\LX{zimaavɛnɛvɛnɛ} \HM{} \PH{zìmáávɛ̀nɛ̀vɛ̀nɛ̀} \PS{n.~} \GE{dusk} 
\LX{zimaavɛ᷉ɛ} \HM{} \PH{zìmáávɛ̀᷉ɛ̀᷉} \PS{n.~} \GE{evening; between four-thirty and five o'clock} 
\LX{zimɛrɛmɛrɛ} \HM{} \PH{zìmɛ̀rɛ̀mɛ̀rɛ̀} \PS{n.~} \GE{muddy place} 
\LX{zimezime} \HM{} \PH{zímézímé} \PS{adv.~} \GE{of considerable weight} 
\LX{ziŋi} \HM{} \PH{zîŋ} \PS{v.~} \GE{to pound vigorously} \PD{ziŋee, ziŋine, ziŋine, ziŋime, ziŋinaa, - } 
\LX{zinyaane} \HM{} \PH{zìɲàànɪ́} \PS{n.~} \GE{dawn} \TP{zinyaanɛɛ} \VA{zivɛlemɛ}
\LX{zinziŋ} \HM{} \PH{zíŋzíŋ} \PS{adv.~} \GE{very heavy} 
\LX{zipigi} \HM{} \PH{zìpìgí} \PS{n.~} \GE{a type of large brown antelope with white spots on its sides} \OP{zipigri} \TP{zipigree} 
\LX{ziponsentaa} \HM{} \PH{zìpónsɪ́ntáá} \PS{n.~} \GE{midnight} 
\LX{zirii} \HM{} \PH{zìríì} \PS{n.~} \GE{a lie or lies} \TP{zirree} \VA{ziriŋ}
\LX{ziriŋ} \HM{} \PH{zìrǐŋ} \PS{n.~} \GE{lies} \TP{zirree} \VA{zirii}
\LX{ziriŋma} \HM{} \PH{zìríŋmà} \PS{n.~} \GE{the act of lying} 
\LX{ziriŋmaara} \HM{} \PH{zìríŋmáárá} \PS{n.~} \GE{liar} \OP{ ziriŋmaareba} 
\LX{zisɔgelaa} \HM{} \PH{zìsɔ́glåá} \PS{n.~} \SN \GE{dark place} \SN \GE{hiding place} \OP{ zisɔgele} 
\LX{zitɔɔre} \HM{} \PH{zìtɔ̀ɔ̀rɪ́} \PS{n.~} \GE{distant place} \OP{zitɔbɔ} 
\LX{zivelaa} \HM{} \PH{zìvɪ̀làá} \PS{n.~} \GE{good place, nice place} \OP{ziveɛle} \TP{ziveɛlɛɛ} 
\LX{zivɛlemɛ} \HM{} \PH{zìvɛ́lɪ́mɛ́} \PS{n.~} \GE{dawn} \TP{zivɛlemɛɛ} \VA{zinyaane}
\LX{ziyuo} \HM{} \PH{zìyùò} \PS{n.~} \GE{modernization} 
\LX{zi᷉i} \HM{} \PH{zì᷈ì᷈} \PS{adj.~} \GE{not clear, blurry} 
\LX{zo} \HM{} \PH{zò} \PS{v.~} \SN \GE{to run} \SN \GE{to run away, to escape} \SN \GE{to flow (river water)} \PD{zoe, zoro, zoro, zoribo, zoraa, zoribo } \VA{zoori (of nominal form)}
\LX{zɔ} \HM{} \PH{zɔ̌} \PS{n.~} \SN \GE{father} \SN \GE{friend (between males)} \OP{zɔmine/zɔre} \TP{zɔrɛɛ} 
\LX{zo nyonono} \HM{} \PH{zò ɲʊ́nʊ́nʊ́} \PS{v.~} \GE{to have mercy} \PD{-, -, -, -, -, - } 
\LX{zo vi} \HM{} \PH{zò ví} \PS{v.~} \SN \GE{to be shy} \SN \GE{to be ashamed} \PD{-, -, -, -, -, - } 
\LX{zɔbɔ} \HM{} \PH{zɔ̀bɔ́} \PS{n.~} \GE{the act of pestering} \VA{zɔba}
\LX{zɔba} \HM{} \PH{zɔ̀bá} \PS{n.~} \GE{the act of pestering} \VA{zɔbɔ}
\LX{zogboŋgboli} \HM{} \PH{zʊ̀gbòŋgbòlì} \PS{n.~} \GE{the thickest part of an animal’s tail that is long} \OP{zogboŋgbolo} \TP{zogboŋgbolee} 
\LX{zoge} \HM{} \PH{zʊ̀gɪ̀} \PS{v.~} \GE{to thicken, to become hard e.g. palm of a one who does rough/odd jobs} \PD{zogɛɛ, zogerɔ/zogera, zogerɔ/zogera, zogereba, zogeraa, - } 
\LX{zogele} \HM{} \PH{zʊ̀glɪ̀} \PS{v.~} \GE{to develop large rashes} \PD{zogelɛɛ, zogelɔ/zogela, zogelɔ/zogela, zogeleba, zogelaa, - } 
\LX{zogelɔŋ} \HM{} \PH{zʊ̀glɔŋ} \PS{adj.~} \GE{thick and attached to} \VA{zogelaŋ}
\LX{zogere} \HM{} \PH{zʊ́grɪ̀} \PS{v.~} \GE{to shake vigorously} \PD{zogrɛɛ, zogra, zogra, zoreba, zograa, - } 
\LX{zogere} \HM{} \PH{zʊ̀grɪ̀} \PS{v.~} \GE{to become thick or hard} \PD{zogerɛɛ, zogera, zogera, zogereba, zogeraa, - } 
\LX{zogetogo} \HM{} \PH{zʊ̀gtʊ̀gʊ̀} \PS{adv.~} \GE{thick and wool-like} 
\LX{zokpore} \HM{} \PH{zʊ́kpʊ́rɪ́} \PS{n.~} \GE{the bottom end of the spinal cord} \OP{zokpoɛ} \TP{zokporɛɛ} 
\LX{zolaa} \HM{} \PH{zʊ́láá} \PS{n.~} \GE{pot with a very small mouth} \OP{zolaare} \TP{zolaarɛɛ} 
\LX{zolaa} \HM{} \PH{zʊ̀làà} \PS{adj.~} \GE{big, large} 
\LX{zɔle} \HM{} \PH{zɔ́lɪ̀} \PS{n.~} \GE{fool} \OP{zɔlɔ} \TP{zɔlɛɛ} 
\LX{zɔle} \HM{} \PH{zɔ́lɪ́} \PS{v.~} \GE{to be related to people} \PD{zɔlɛɛ, zɔlɔ, zɔlɔ, zɔleba, zɔlaa, - } 
\LX{zɔlɔ} \HM{} \PH{zɔ̀lɔ́} \PS{n.~} \GE{a type of guinea corn} \TP{zɔllɛɛ} \VA{zɔla}
\LX{zɔla} \HM{} \PH{zɔ̀lá} \PS{n.~} \GE{a type of guinea corn} \VA{zɔlɔ}
\LX{zombaa} \HM{} \PH{zʊ́mbáá} \PS{n.~} \GE{catfish} \OP{zombaare} \TP{zombaarɛɛ} 
\LX{Zombɛnte} \HM{} \PH{zúmbɛ́ntɪ́} \PS{n.~} \GE{a religious festival of Southern Dagao} \OP{zombɛntere} \TP{zombɛnterɛɛ} 
\LX{zombɛraa} \HM{} \PH{zʊ́bɛ́ráá} \PS{n.~} \GE{fishhook} \OP{ zombɛrre} 
\LX{zomberemɛ} \HM{} \PH{zʊ́mbɪ́rɪ́mɛ́} \PS{n.~} \GE{a type of tree with medicinal properties} \TP{zomberemɛɛ} 
\LX{zombiri} \HM{} \PH{zʊ̀mbírì} \PS{n.~} \GE{insult, abuse} \SG{zombiruu} \OP{zombie} \TP{zombiiree} 
\LX{zomm} \HM{} \PH{zómm} \PS{adv.~} \GE{still, calm, unmoving} 
\LX{zommo} \HM{} \PH{zʊ́mmʊ́} \PS{n.~} \GE{fish} \OP{zoma} \TP{zonnɛɛ} 
\LX{zomɔ} \HM{} \PH{zʊ̀mɔ́} \PS{n.~} \GE{instinct} \VA{zoma}
\LX{zoma} \HM{} \PH{zʊ̀má} \PS{n.~} \GE{instinct} \VA{zomɔ}
\LX{zompɛge} \HM{} \PH{zʊ́mpɛ́gɪ̀} \PS{n.~} \GE{fish scale} \OP{zompɛgre/zompɛgɛ} \TP{zompɛgrɛɛ} 
\LX{zompi᷉i᷉} \HM{} \PH{zʊ́pí᷉ì᷉} \PS{n.~} \GE{fishhook} \OP{zompiime} \TP{zompiimee/zompiinee} 
\LX{zɔmpɔrɔŋ} \HM{} \PH{zɔ́mpɔ́rɔ́ŋ} \PS{adv.~} \GE{of considerable size, state} 
\LX{zone} \HM{} \PH{zʊ́nɪ́} \PS{n.~} \GE{insult, abuse} \OP{zoma} \TP{zonnɛɛ} 
\LX{zoŋe} \HM{} \PH{zʊ́ŋ} \PS{v.~} \GE{to shake (person)} \PD{zoŋɛɛ, zoŋenɔ, zoŋenɔ, zoŋenema, zoŋenaa, - } 
\LX{zɔŋe} \HM{} \PH{zɔ́ŋ} \PS{n.~} \SN \GE{flour} \SN \GE{pulp} \TP{zɔnnɛɛ} 
\LX{zɔŋe} \HM{} \PH{zɔ̀ŋ} \PS{v.~} \GE{to be blind} \PD{zɔŋɛɛ, zɔnnɔ, zɔnnɔ, zɔnemɔ, zɔnnaa, zɔnemɔ } 
\LX{zɔŋe} \HM{} \PH{zɔ̌ŋ} \PS{n.~} \GE{flour} \OP{zɔnne} \TP{zɔnnɛɛ} \VA{zɔŋɔ}
\LX{zɔŋkɔŋ-ɔnnɔ} \HM{} \PH{zɔ̌ŋkɔ́ŋ-ɔ́nnɔ́} \PS{n.~} \GE{one who cooks workers' food} \OP{zɔŋkɔŋ-ɔnemɔ} \TP{zɔŋkɔŋ-ɔnemɛɛ.} 
\LX{zoŋkɔɔre} \HM{} \PH{zʊ́ŋkɔ́ɔ̀ré} \PS{n.~} \GE{fish bone} \OP{zoŋkɔbɔ} \TP{zoŋkɔɔrɛɛ} 
\LX{zoŋkoroŋ} \HM{} \PH{zʊ̀ŋkʊ̀rʊ̀ŋ} \PS{adv.~} \GE{grotesque; hideous} 
\LX{zɔŋko᷉ɔ᷉} \HM{} \PH{zɔ̌ŋkʊ́᷉ɔ́} \PS{n.~} \SN \GE{food for workers at work} \SN \GE{water mixed with flour} \OP{zɔŋkoɔnɛɛ} 
\LX{zonkyoo} \HM{} \PH{zʊ́nʧóò} \PS{n.~} \GE{fishing net} \OP{zonkyoori} \TP{zonkyooree} 
\LX{zonne} \HM{} \PH{zúnnɪ́} \PS{v.~} \GE{to take short, heavy steps} \PD{zonnɛɛ, zonnɔ/zonna, zonnɔ/zonna, zonemɔ/zonema, zonnaa, - } 
\LX{zonne} \HM{} \PH{zúnnɪ́} \PS{v.~} \GE{to shake (from sudden shock)} \PD{zonnɛɛ, zonnɔ/zonna, zonnɔ/zonna, zonemɔ/zonema, zonnaa, - } 
\LX{zɔnne} \HM{} \PH{zɔ̀nnɪ́} \PS{n.~} \GE{blindness} 
\LX{zonni} \HM{} \PH{zònnì} \PS{v.~} \GE{to hesitate} \PD{zonnee, zonno, zonno, zonimo, zonnaa, zonimo } 
\LX{zɔnnoŋ} \HM{} \PH{zɔ̀nnʊ́ŋ} \PS{n.~} \GE{blindness} 
\LX{zonnyɔge} \HM{} \PH{zʊ́nɲɔ́gɪ́} \PS{n.~} \GE{the practice of fishing} 
\LX{zonnyɔgerɔ} \HM{} \PH{zʊ̌nɲɔ́grɔ̀} \PS{n.~} \GE{fisherman} \OP{zonnyɔgerebɔ/zonnyɔgereba} \TP{zonnyɔgerebɛɛ} 
\LX{zɔŋɔ} \HM{} \PH{zɔ̀ŋɔ́} \PS{n.~} \GE{blind person} \OP{zɔnne} \TP{zɔnnɛɛ} \VA{zɔŋa, zɔŋe}
\LX{zɔŋa} \HM{} \PH{zɔ̀ŋá} \PS{n.~} \GE{blind person} \VA{zɔŋɔ, zɔŋe}
\LX{zɔŋe} \HM{} \PH{zɔ̀ŋɪ́} \PS{n.~} \GE{blind person} \VA{zɔŋɔ, zɔŋa}
\LX{zonsaalaa} \HM{} \PH{zʊ́nsààglàá} \PS{n.~} \GE{a type of fish with a flat head and two fins and a very slippery body, found in rivers and ponds, referred to as mudfish} \OP{zonsaale} \TP{zonsaalɛɛ} 
\LX{zonsɔgelaa} \HM{} \PH{zʊ́nsɔ́gláá} \PS{n.~} \GE{a type of black mudfish with a flat head} \OP{zonsɔgelɔ} \TP{zonsɔgelɛɛ} 
\LX{zontalane} \HM{} \PH{zʊ̀ntàlánɪ̀} \PS{n.~} \GE{gill} \OP{zontalama} \TP{zontalanne} 
\LX{zonwaabo} \HM{} \PH{zʊ́nwàábʊ́} \PS{n.~} \GE{eel} \OP{zonweere} \TP{zonweerɛɛ} 
\LX{zɔnzɔremɔ} \HM{} \PH{zɔ̀nzɔ̀rɪ́mɔ́} \PS{n.~} \GE{cobweb, dirt hanging on walls and ceiling} \TP{zɔnzɔremɛɛ} \VA{zɔnzɔrema}
\LX{zɔnzɔrema} \HM{} \PH{zɔ̀nzɔ́rɪ́má} \PS{n.~} \GE{cobweb, dirt hanging on walls and ceiling} \VA{zɔnzɔremɔ}
\LX{zɔɔ} \HM{} \PH{zɔ̀ɔ́} \PS{n.~} \GE{ambush} \OP{zɔɔre} \TP{zɔɔrɛɛ} 
\LX{zɔɔ} \HM{} \PH{zɔ̀ɔ̀} \PS{adj.~} \GE{long and flowing e.g. hair} 
\LX{zɔɔ} \HM{} \PH{zɔ̀ɔ̀} \PS{v.~} \GE{to fight, to quarrel} \PD{zɔɔɛ, zɔɔrɔ, zɔɔrɔ, zɔɔrebɔ/zɔɔreba, -, - } 
\LX{zɔɔ-bɔ} \HM{} \PH{zɔ̀ɔ̀-bɔ́} \PS{n.~} \GE{the act of irritating others} 
\LX{zoɔle} \HM{} \PH{zʊ́ɔ̀lɪ̀} \PS{v.~} \GE{to wither} \PD{zoɔlɛɛ, zoɔlɔ, zoɔlɔ, zoɔlba, zoɔlaa, - } 
\LX{zoore} \HM{} \PH{zʊ́ʊ́ré} \PS{n.~} \GE{tail} \OP{zoɛ} \TP{zoorɛɛ} 
\LX{zoore} \HM{} \PH{zʊ́ʊ́rɪ̀} \PS{n.~} \GE{smoke} \VA{nyoore}
\LX{zɔɔre} \HM{} \PH{zɔ̀ɔ̀rɪ́} \PS{n.~} \GE{quarrel, fight} 
\LX{zɔɔzeele} \HM{} \PH{zɔ̀ɔ̀zɪ́ɪ́lɪ́} \PS{n.~} \GE{silky flower on a corncob} \SG{zɔɔzeeloo} \OP{zɔɔzeelɛ} \TP{zɔɔzeelɛɛ} 
\LX{zori} \HM{} \PH{zòrì} \PS{v.~} \GE{to hold the hand in anticipation for food} \PD{zoree, zoro, zoro, zoribo, zoraa, - } 
\LX{zorre} \HM{} \PH{zʊ́rrɪ̀} \PS{v.~} \GE{shudder} \PD{zorrɛɛ, zorra, zorra, zoreba, zorraa, - } 
\LX{zɔzɔ} \HM{} \PH{zɔ̀zɔ̀} \PS{adv.~} \GE{reduced  to minimal (rainfall)} 
\LX{zo᷉o} \HM{} \PH{zʊ̀᷈ʊ̀᷉} \PS{adv.~} \GE{dull} 
\LX{zo᷈ɔ} \HM{} \PH{zʊ́᷉ɔ́᷉} \PS{n.~} \GE{rust (typically used with le)} \XV{A soɔ lɔre la ko᷉ɔ poɔ ana la so ka o le zo᷉ɔ.} \XE{The knife got soaked in water and that is why it is rusted.} \TP{zoɔnɛɛ} 
\LX{zɔ᷉ɔ} \HM{} \PH{zɔ̀᷉ɔ̀᷉} \PS{v.~} \GE{to ride} \PD{zɔ᷉ɔɛ, zɔɔnɔ, zɔɔnɔ, zɔɔnemɔ, zɔɔnaa, - } 
\LX{zu} \HM{} \PH{zû} \PS{n.~} \GE{head} \OP{zuri} \TP{zuree} 
\LX{zu} \HM{} \PH{zú} \PS{v.~} \GE{to steal} \PD{zuee, zuuro, -, -, zuraa, - } 
\LX{zu} \HM{} \PH{zû} \PS{adv.~} \GE{on top of} \XV{A gane dɔgele la a tabol zu.} \XE{The book is on the table.} 
\LX{zu pozu} \HM{} \PH{zú pòzú} \PS{v.~} \GE{to gossip} \PD{-, -, -, -, -, - } 
\LX{zu zɔŋe} \HM{} \PH{zú zɔ̂ŋ} \PS{v.~} \GE{to be ashamed} \PD{zue zɔŋe, -, -, -, -, - } 
\LX{zubɛroŋ} \HM{} \PH{zúbɛ́rʊ́ŋ} \PS{n.~} \SN \GE{strong and dangerous animal} \SN \GE{leader (often political)} \OP{zubɛrɛ} \TP{zubɛrɛɛ} 
\LX{zubie} \HM{} \PH{zúbíé} \PS{n.~} \GE{misfortune, ill luck} \TP{zubiiree} 
\LX{zudaa} \HM{} \PH{zúdàá} \PS{n.~} \GE{backbone; that which one relies on} \OP{zudaare} \TP{zudaarɛɛ} 
\LX{zufaa} \HM{} \PH{zúfáá} \PS{n.~} \GE{ill fate} \OP{zufaare} \TP{zufaarɛɛ.} 
\LX{zugane} \HM{} \PH{zúgánɪ́} \PS{n.~} \GE{skin of the head} \OP{zugama} \TP{zugamɛɛ} 
\LX{zugi} \HM{} \PH{zùgì} \PS{v.~} \GE{to turn a container upside down to bring out all the contents} \PD{zugee, zugiro, zugiro, zugiribo, zugiraa, - } 
\LX{zugi} \HM{} \PH{zùgì} \PS{v.~} \GE{to press the bellows of a furnace} \PD{zugee, zugiro, zugiro, zugiribo, zugiraa, - } 
\LX{zugi zugi} \HM{} \PH{zùgzùg} \PS{adv.~} \GE{moving through in numbers e.g. spectators} \XV{Ba ŋmɛ la a bɔle baare ka a noba ta zugi zugi zugi ba naŋ yire.} \XE{The football match is over and the spectators are trouping out.} 
\LX{zugilaa} \HM{} \PH{zùgìláá} \PS{n.~} \GE{bellows} \OP{zugilaare} \TP{zugilaarɛɛ} 
\LX{zugiloŋ} \HM{} \PH{zúglóŋ} \PS{adj.~} \GE{unproportioned} 
\LX{zugiraa} \HM{} \PH{zúgráa} \PS{n.~} \GE{blacksmith's bellows} \OP{zugiri} \TP{zugiree} 
\LX{zugiri} \HM{} \PH{zúgrì} \PS{v.~} \GE{to shake in order to know whether there something inside e.g. closed tin} \PD{zugiree, zugiro, zugiro, zugiribo, zugiraa, - } 
\LX{zugizaga} \HM{} \PH{zùgzàgà} \PS{adj.~} \GE{huge and heavy e.g. mallet} 
\LX{zuiŋ} \HM{} \PH{zúíŋ} \PS{conj.~} \SN \GE{because, as a result} \XV{A bie baaloŋ zuiŋ o ma ba gaa toma.} \XE{As a result of the child's illness, the mother didn't go to work.} \SN \GE{so} \XV{A saama na waa la yaga a zuiŋ maale a bondirii yaga.} \XE{There will be many guests, so cook plenty of food.} 
\LX{zukɔgeraa} \HM{} \PH{zúkɔ́gráá} \PS{n.~} \GE{pillow} \OP{zukɔgere} \TP{zukɔgerɛɛ} 
\LX{zukɔɔloŋ} \HM{} \PH{zúkɔ́ɔ́lʊ́ŋ} \PS{n.~} \GE{hair} \OP{zukɔɔlɔ} \TP{zukɔɔlɛɛ} 
\LX{zukoɔre} \HM{} \PH{zúkʊ́ɔ́rɪ́} \PS{n.~} \GE{skull} \OP{zukɔɛ} \TP{zukɔrɛɛ} \VA{zuŋmane}
\LX{zukpala} \HM{} \PH{zúkpálà} \PS{n.~} \GE{bald head} \TP{zukpallɛɛ} 
\LX{zukpeɛŋaa} \HM{} \PH{zùkpɪ́ɛ́ŋáá} \PS{adj.~} \GE{hardheaded} \OP{zukpeɛne} 
\LX{zukpeɛne} \HM{} \PH{zúkpɪ́ɛ́nɪ́} \PS{n.~} \GE{hardheadedness, stubbornness} 
\LX{zukpoli} \HM{} \PH{zúkpólí} \PS{n.~} \SN \GE{a shaven head} \SN \GE{a head that is carrying something without a pad} \OP{zukpolo} \TP{zukpolee} \VA{zugboli}
\LX{zugboli} \HM{} \PH{zúgbòlí} \PS{n.~} \SN \GE{a shaven head} \SN \GE{a head that is carrying something without a pad} \VA{zukpoli}
\LX{zukuri} \HM{} \PH{zùkúrí} \PS{n.~} \GE{back (body)} \OP{zukurri} \TP{zukuree} 
\LX{zukyuu} \HM{} \PH{zúʧùú} \PS{n.~} \GE{ritual hair} \OP{zukyuuri} \TP{zukyuuree} 
\LX{zulee} \HM{} \PH{zùlèè} \PS{adv.~} \GE{large at one end} 
\LX{zuleɛrɛ} \HM{} \PH{zúlɪ́ɛ́rɛ́} \PS{n.~} \GE{successor} \OP{zuleɛrebɛ} \TP{zuleɛrebɛɛ} 
\LX{zuli} \HM{} \PH{zùlì} \PS{v.~} \SN \GE{to be unbalanced, to lean to one side} \SN \GE{to slope} \PD{zulee, zullo, zullo, zulbo, zullaa, - } 
\LX{zulibala} \HM{} \PH{zùlìbàlà} \PS{adj.~} \GE{disproportionately big} 
\LX{zulimo} \HM{} \PH{zúlímó} \PS{adj.~} \SN \GE{uneven, sloped} \SN \GE{unbalanced (weight)} 
\LX{zuluŋ} \HM{} \PH{zùlûŋ} \PS{adj.~} \GE{deep} \OP{zulli} \TP{zullee} 
\LX{zuluŋ} \HM{} \PH{zùlǔŋ} \PS{n.~} \GE{army ant, soldier ant} \TP{zullee} 
\LX{zu-maala} \HM{} \PH{zú-máálá} \PS{n.~} \GE{hairdresser} \OP{zu-maaleba} 
\LX{zumaaroŋ} \HM{} \PH{zúmáárʊ̌ŋ} \PS{n.~} \GE{headache (concentrated around the forehead)} \OP{zumaara} \TP{zumaarɛɛ} 
\LX{zuŋ} \HM{} \PH{zǔŋ} \PS{adv.~} \GE{on top of} \VA{shortened form of zu poɔ}
\LX{zuŋmaara} \HM{} \PH{zúŋmáárà} \PS{n.~} \GE{executioner} \OP{zuŋmaareba} 
\LX{zuŋmane} \HM{} \PH{zúŋmánɪ́} \PS{n.~} \GE{skull} \OP{zuŋmama} \TP{zuŋmamɛɛ} \VA{zukoɔre}
\LX{zuŋmane ko᷉ɔ} \HM{} \PH{zúŋmánɪ́ kʊ́᷉ɔ́᷉} \PS{n.~} \GE{great suffering (typically used with nyu)} \XV{Deyɛŋ boŋkoɔre ba maale ka nembaaleba da nyu ba zuŋmane ko᷉ɔ.} \XE{Last year, there was a poor crop yield and poor people suffered greatly.} 
\LX{zuŋmanekoɔ} \HM{} \PH{zúŋmánɪ́kʊ̀ɔ́} \PS{n.~} \GE{suffering} 
\LX{zunoɔ} \HM{} \PH{zúnʊ̀ɔ́} \PS{n.~} \GE{fortune, favour, luck} \OP{zunoore} \TP{zunoorɛɛ} 
\LX{zunzugiri} \HM{} \PH{zùnzùgrí} \PS{n.~} \GE{large isolated cloud} \SG{zunzugiruu} \OP{zunzugo} \TP{zunzugiree} 
\LX{zuo} \HM{} \PH{zùò} \PS{v.~} \SN \GE{to be more than} \SN \GE{to be too much} \SN \GE{to pass someone by} \PD{-, -, -, -, -, - } 
\LX{zu-ɔɔ} \HM{} \PH{zú-ɔ̀ɔ̀} \PS{n.~} \GE{the condition of having a headache} \OP{zu-ɔɔre} 
\LX{zupili} \HM{} \PH{zúpílí} \PS{n.~} \GE{hat, cap} \SG{zupiluu} \OP{zupile} \TP{zuplee} 
\LX{zupiltobo} \HM{} \PH{zúpìlìtòbó} \PS{n.~} \GE{a hat that covers the ears} \TP{zupiltobee} 
\LX{zupoli} \HM{} \PH{zúpólí} \PS{n.~} \GE{sinciput (the front of the skull)} \SG{zupollee} \OP{zupolo} \TP{zupolli} 
\LX{zupono} \HM{} \PH{zú-pónó} \PS{n.~} \GE{barber} \XV{zu-ponimo} \XE{} 
\LX{zusa᷉a} \HM{} \PH{zúsà᷉à᷉} \PS{n.~} \GE{immorality} 
\LX{zusaanaa} \HM{} \PH{zúsã́ánáá} \PS{n.~} \GE{the act of engaging in evil practices} 
\LX{zusaana} \HM{} \PH{zúsááná} \PS{n.~} \GE{one who preaches vice} 
\LX{zusãa} \HM{} \PH{zúsã̀ã̀} \PS{n.~} \GE{the practice of vice} 
\LX{zusogɔ} \HM{} \PH{zúsʊ́gá} \PS{adv.~} \GE{above, high up} \VA{zusoga}
\LX{zusoga} \HM{} \PH{zúsʋ́gá} \PS{adv.~} \GE{above, high up} \VA{zusogɔ}
\LX{zutage} \HM{} \PH{zú-tàgɪ́} \PS{n.~} \GE{hardheaded} \OP{zutagere} 
\LX{zutalaa} \HM{} \PH{zútálàà} \PS{n.~} \GE{headdress} \OP{zutalle} \TP{zutallɛɛ} 
\LX{zutuo} \HM{} \PH{zútúó} \PS{n.~} \GE{misfortune} \OP{zutuuri} \TP{zutuuree} 
\LX{zuu} \HM{} \PH{zùù} \PS{adv.~} \GE{huge and heavy} 
\LX{zuuni} \HM{} \PH{zúúní} \PS{v.~} \GE{to stoop} \PD{zuunee, zuuno, zuuno, -, -, - } 
\LX{zuyɔɔ} \HM{} \PH{zúyɔ́ɔ́} \PS{n.~} \GE{head tax} \TP{zuyɔɔrɛɛ} 
\LX{zuzeɛ} \HM{} \PH{zúzɪ̀ɛ́} \PS{n.~} \GE{leader, forerunner} \OP{zuzeere} \TP{zuzeerɛɛ} 
\LX{zuzeɛ} \HM{} \PH{zúzɪ̀ɛ́} \PS{n.~} \GE{red-haired person} \OP{zuzeere} \TP{zuzeerɛɛ} 
\LX{zuzuu} \HM{} \PH{zùzúú} \PS{adj.~} \GE{monstrous} 
\LX{zzzzzzz} \HM{} \PH{ajsdf} \PS{n.~} \GE{hopefully works} 
