\tbLX{a} \tbHM{} \tbPH{à} \tbPS{conj.\ } \tbSN \tbGE{and} \tbXV{O waɛ la a wa gaŋ.} \tbXE{S/he came and left.} \tbSN \tbGE{in order to} \tbXV{O waɛ la a na nyɛ te.} \tbXE{S/he has come in order to visit us.} 
\tbLX{a} \tbHM{} \tbPH{á} \tbPS{conj.~} \tbGE{but} \tbXV{A wola ka o ba wa?} \tbXE{But why didn't s/he come?.} \tbVA{ana}
\tbLX{a} \tbHM{} \tbPH{à} \tbPS{pron  (pl).~} \tbGE{they, them (nonhuman)} \tbXV{Vɛŋ ka a kpɛ a zage.} \tbXE{Let them go into the pen.} \tbXV{De a moɔ eŋ a.} \tbXE{Give them the grass.} 
\tbLX{a} \tbHM{} \tbPH{à} \tbPS{def article.~} \tbGE{the} \tbXV{A bie kpɛ la a die.} \tbXE{The child has gone into the room.} 
\tbLX{a} \tbHM{} \tbPH{á} \tbPS{v.~} \tbGE{to hate} \tbXV{Te a ba la.} \tbXE{We hate them.} \tbPD{aɛ, aara, aara, aareba, aaraa, aareba } 
\tbLX{aa} \tbHM{} \tbPH{áà} \tbPS{interj;.~} \tbGE{expression of surprise and regret} \tbXV{ A ŋmane ŋmarɛɛ la, aa.} \tbXE{The calabash is broken; what a pity!.} \tbXV{ Aa, ka N da baŋ ka lɛ N gaaɛ.} \tbXE{Oh, if I had known, I would have gone.} 
\tbLX{aa} \tbHM{} \tbPH{áá} \tbPS{interj.~} \tbGE{expression of regret} 
\tbLX{aa} \tbHM{} \tbPH{àà} \tbPS{interj.~} \tbGE{expression expressing "I told you so"} 
\tbLX{a-a} \tbHM{} \tbPH{á-à} \tbPS{interj.~} \tbSN \tbGE{expression of disapproval} \tbXV{A-a, ta de ona.} \tbXE{Oh no, don't go that way.} \tbSN \tbGE{ why; expression of doubt} \tbXV{A-a, fo naŋ ba gaa?} \tbXE{Well, you haven't gone yet?.} 
\tbLX{aae} \tbHM{} \tbPH{ááɪ́} \tbPS{interj.~} \tbGE{expression of pain} 
\tbLX{âa} \tbHM{} \tbPH{à᷉á᷉} \tbPS{adv.~} \tbSN \tbGE{yes (answer to a call)} \tbXV{N boɔle la a bie ka o sage ka "a᷉a."} \tbXE{I called the child and he answered, "Yes.".} \tbSN \tbGE{what? (exclamation of disbelief)} \tbXV{Ka Naa boɔle fo ka fo zagere, a᷉a.} \tbXE{You refuse to answere the chief's call, what?.} 
\tbLX{âaa} \tbHM{} \tbPH{à᷉á᷉́á᷉́} \tbPS{part.~} \tbGE{answer in agreement} \tbXV{Fo na gaa la? âaa.} \tbXE{Will you go? Yes, I will.} 
\tbLX{a᷉̀aa} \tbHM{} \tbPH{à᷉á᷉à᷉} \tbPS{interj.~} \tbGE{is that so? (expression of uncertain surprise)} \tbXV{Bayuoː A pɔgebil-pɛɛre na taa la poɔ. Ayuoː A᷉aa?} \tbXE{Bayuo: That teenage girl is pregnant. Ayuo: Is that true?.} 
\tbLX{aa-ee} \tbHM{} \tbPH{áá-ɪ̀ɪ́} \tbPS{interj.~} \tbGE{expression of lack of confidence in someone's ability to perform a task} \tbXV{Aa-ee, fo koŋ daŋ ma ta.} \tbXE{I don't think you can get there before me.} 
\tbLX{aa-hii} \tbHM{} \tbPH{áá-hì᷉ì} \tbPS{interj.~} \tbGE{expression of total despair} 
\tbLX{aane} \tbHM{} \tbPH{áánɪ̀} \tbPS{adj.~} \tbGE{infertile (woman )} \tbOP{aama} \tbTP{aanɛɛ} 
\tbLX{aaŋ} \tbHM{} \tbPH{aaŋ} \tbPS{interj.~} \tbGE{well then} 
\tbLX{aba} \tbHM{} \tbPH{ábà} \tbPS{interj.~} \tbSN \tbGE{expression of despair over a repeated mishap} \tbXV{A bie la lee la, aba.} \tbXE{The child has fallen again, oh no!.} \tbSN \tbGE{expression of satisfaction or agreement} \tbXV{Aba, lɛ ka N boɔrɔ.} \tbXE{Good, that is what I want.} \tbSN \tbGE{expression of empathy} \tbXV{Aba, mɔ kyɛ so a bie ko᷉ɔ} \tbXE{Try and bathe the child.} 
\tbLX{abaa} \tbHM{} \tbPH{ábàà} \tbPS{interj.~} \tbSN \tbGE{exclamation of satisfaction or commendation} \tbXV{Abaa, fo waa bie.} \tbXE{Well done, you are a good child.} \tbSN \tbGE{exclamation of despair or being distraught} \tbXV{Abaa, vɛŋ a zɔɔre lɛ.} \tbXE{Oh, stop the fight.} 
\tbLX{abɛ} \tbHM{} \tbPH{àbɛ́} \tbPS{n.~} \tbSN \tbGE{palm tree} \tbSN \tbGE{palm fruit} 
\tbLX{abɛbiri} \tbHM{} \tbPH{àbɛ̀bírì} \tbPS{n.~} \tbGE{palm fruit} \tbSG{abɛbiruu} \tbOP{abɛbie} \tbTP{abɛbiree} 
\tbLX{abɛda᷉a᷉} \tbHM{} \tbPH{àbɛ́dá᷉à᷉} \tbPS{n.~} \tbGE{palm wine} \tbSG{abɛdãão᷉} \tbTP{abɛdaanɛɛ} 
\tbLX{abɛka᷉a} \tbHM{} \tbPH{àbɛ́ká᷉à} \tbPS{n.~} \tbGE{palm oil} \tbTP{abɛkaanɛɛ} 
\tbLX{abɛsaare} \tbHM{} \tbPH{ábɛ́sáárɪ̀} \tbPS{n.~} \tbGE{palm-frond broom} \tbOP{abɛsaɛ} \tbTP{abɛsaarɛɛ} 
\tbLX{abɛteɛ} \tbHM{} \tbPH{àbɛ̀tɪ̀ɛ̀} \tbPS{n.~} \tbGE{palm tree} \tbOP{abɛteere} \tbTP{abɛteerɛɛ} 
\tbLX{abuobo} \tbHM{} \tbPH{ábúóbò} \tbPS{adv.~} \tbGE{which ones (nonhuman)} \tbXV{Abuobo ka fo boɔrɔ?} \tbXE{Which ones do you want?.} 
\tbLX{Afereka} \tbHM{} \tbPH{áfɪ́rɪ́kà} \tbPS{n.~} \tbGE{Africa} \tbOP{Aferekare} \tbTP{Aferekarɛɛ} 
\tbLX{Aferekaneɛ} \tbHM{} \tbPH{áfíríkànɪ́ɛ́} \tbPS{n.~} \tbGE{African} \tbOP{Aferekanoba} 
\tbLX{afu} \tbHM{} \tbPH{àfú} \tbPS{n.~} \tbGE{preface (of a book)} \tbOP{afuri} \tbTP{afuree} 
\tbLX{age} \tbHM{} \tbPH{ágɪ́} \tbPS{v.~} \tbSN \tbGE{to dance} \tbSN \tbGE{to jump up and down} \tbPD{agɛɛ, agra, agra, agreba, agraa, agreba } \tbSE{} \tbGE{} 
\tbLX{age} \tbHM{} \tbPH{ágɪ̀} \tbPS{v.~} \tbSN \tbGE{to fly} \tbSN \tbGE{errect, ejaculate (penis)} \tbSN \tbGE{to jump over} \tbSN \tbGE{to be frightened; to miss a heartbeat} \tbPD{agɛɛ, agra, agra, agreba, agraa, agreba } \tbSE{pɔlaa age} \tbGE{to be frightened} 
\tbLX{agoo} \tbHM{} \tbPH{àgòò} \tbPS{v.~} \tbGE{to ask for permission to join someone (knock door)} \tbPD{-, -, -, -, -, - } 
\tbLX{agoo} \tbHM{} \tbPH{àgòò} \tbPS{n.~} \tbGE{a type of silky, colourful fabric} \tbOP{agoori} \tbTP{agooree} 
\tbLX{agye} \tbHM{} \tbPH{ádzé} \tbPS{n.~} \tbGE{bad luck} \tbOP{agyeri} \tbTP{agyeree} 
\tbLX{ahâa} \tbHM{} \tbPH{àhá᷉á᷉} \tbPS{interj.~} \tbGE{that's right (expression of agreement)} \tbXV{Ahaa, o binzie la be.} \tbXE{That's right, that is its position.} 
\tbLX{ahaŋ} \tbHM{} \tbPH{àháŋ} \tbPS{interj.~} \tbGE{an expression that denotes "well?"} \tbXV{Ahaŋ a polisi la waana.} \tbXE{All is well, as the police are coming.} 
\tbLX{ai} \tbHM{} \tbPH{àí} \tbPS{use word.~} \tbGE{no} 
\tbLX{aii} \tbHM{} \tbPH{áìì} \tbPS{interj.~} \tbGE{an expression of despair} \tbXV{Aii, a nɔgyɛle ŋmarɛɛ.} \tbXE{Oh no; the egg is broken!.} 
\tbLX{akaase} \tbHM{} \tbPH{àkáásɪ̀} \tbPS{interj.~} \tbGE{expression of boastful challenge} \tbXV{Akaase, neɛ kyebe na daŋ ma kɔnoɔre.} \tbXE{I dare anyone to finish his portion of the weeding before me.} 
\tbLX{akaasemanakei} \tbHM{} \tbPH{ákáásɪ̀mànà} \tbPS{interj.~} \tbGE{expression of disappointment at failure} \tbXV{A bie ba toɔ tuo daga, akaasemanakei.} \tbXE{The child has failed to carry the box, what a pity.} \tbVA{akaase}
\tbLX{akai} \tbHM{} \tbPH{àkàí} \tbPS{interj.~} \tbGE{expression of disgust and disagreement} \tbXV{ Akai, fo deme kyebe.} \tbXE{Too bad, you have no case.} 
\tbLX{aku} \tbHM{} \tbPH{ákú} \tbPS{interj.~} \tbGE{expression of despair, loss, empathy} \tbXV{A nannyigri la maaleŋ fa la n zɔmeŋɛ na, aku.} \tbXE{The thieves have robbed my friend again, how awful!.} \tbVA{akuroku}
\tbLX{akuroku} \tbHM{} \tbPH{àkúrókú} \tbPS{interj.~} \tbGE{emphatic form of aku} \tbVA{akuraku; aku}
\tbLX{akuu} \tbHM{} \tbPH{àkúú} \tbPS{interj.~} \tbGE{expression indicating one's inability to do a task} \tbXV{Akuu, ta ko ma, N koŋ toɔ̂ a tuori tuo.} \tbXE{Ei, don't kill me; I can't carry the load.} 
\tbLX{Ala} \tbHM{} \tbPH{álà} \tbPS{n.~} \tbGE{God} 
\tbLX{ala} \tbHM{} \tbPH{álà} \tbPS{interj.~} \tbGE{I swear} \tbXV{Ala, a dɔɔ ŋa naane so a woɔ.} \tbXE{I swear, the bag does not belong to this man.} 
\tbLX{alaakosebaroo} \tbHM{} \tbPH{àláákosɪ́bárʋ́ʊ́} \tbPS{interj.~} \tbGE{in the name of God} 
\tbLX{ale} \tbHM{} \tbPH{álɪ̀} \tbPS{v.~} \tbGE{to open (ring), to stretch out} \tbXV{Ka fooŋ ba ale a fenetere, o koŋ kpɛ a fo nubiri.} \tbXE{If you don't open up the ring you can't put it on your finger.} \tbPD{alɛɛ, alla, alla, aleba, allaa, - } 
\tbLX{alee} \tbHM{} \tbPH{àléé} \tbPS{v.~} \tbGE{expression ordering one to get out; away with you} \tbXV{A bibiiri naŋ di a bondirii baare la ka ba ba yeli ka alee, ka ba wa yi gaa sakuuri.} \tbXE{When the children had finished eating, their father ordered them to go to school at once.} \tbPD{-, -, -, -, -, - } 
\tbLX{aleɛma} \tbHM{} \tbPH{àlɪ́ɛ́mà} \tbPS{n.~} \tbGE{metaphor (literature)} 
\tbLX{alɛ} \tbHM{} \tbPH{álɛ̀} \tbPS{interj.~} \tbGE{expression of surprise mixed with fear} \tbXV{Alɛ, fentel diglii ka fo deɛnɛ ne?} \tbXE{Whoa, you're playing with electricity?.} 
\tbLX{aloo} \tbHM{} \tbPH{álòó} \tbPS{interj.~} \tbGE{hello} 
\tbLX{aloopelee} \tbHM{} \tbPH{álóópèléè} \tbPS{n.~} \tbGE{aeroplane} \tbOP{aloopelli} \tbTP{aloopellee} \tbVA{ayɔɔmapelee [áyɔÕÅ:mápeÕÄleÕÄ:]}
\tbLX{aloopeleeduoraa} \tbHM{} \tbPH{álóópèléédúóráà} \tbPS{n.~} \tbGE{airport} \tbOP{aloopeleeduori} 
\tbLX{ama} \tbHM{} \tbPH{àmá} \tbPS{conj.~} \tbGE{but, however} \tbXV{ A bɔlŋmeɛrebɛ mɔɛ la ama ba ba di.} \tbXE{The football players played well, but they did not win.} 
\tbLX{ama} \tbHM{} \tbPH{ámà} \tbPS{demonstrative pronoun.~} \tbGE{these (nonhuman)} \tbXV{A dogere ama daaroŋ tuo.} \tbXE{These pots are expensive.} 
\tbLX{ambaŋenaa} \tbHM{} \tbPH{ámbáŋɪ́náá} \tbPS{n.~} \tbGE{a type of small tree with large leaves and pink fruit} \tbOP{ambaŋene} \tbTP{ambaŋenɛɛ} 
\tbLX{Ameleka} \tbHM{} \tbPH{ámɪ́lɪ́kà} \tbPS{n.~} \tbGE{ America} \tbOP{amelekare} \tbTP{amelekarɛɛ} \tbVA{Hameleka}
\tbLX{aminekaŋa} \tbHM{} \tbPH{àmínékàŋá} \tbPS{conj.~} \tbSN \tbGE{perhaps, maybe} \tbXV{Aminekaŋa o koŋ wa pampana.} \tbXE{Perhaps s/he will not come now.} \tbSN \tbGE{sometimes} \tbVA{minekaŋa}
\tbLX{amma} \tbHM{} \tbPH{ámmá} \tbPS{function word.~} \tbGE{much more than (exaggeration)} \tbXV{A bie zoro to, amma soɔŋaa.} \tbXE{The boy runs really fast, much faster than the hare.} 
\tbLX{ammeseɛrɛ} \tbHM{} \tbPH{ámmɪ̀sɪ́ɛ́rɛ̀} \tbPS{n.~} \tbGE{thanks} \tbXV{Fo taa la ammeseɛrɛ ne fo naŋ toŋ a toma.} \tbXE{I am thankful to you for doing the work.} \tbTP{ammeseɛrɛɛ} 
\tbLX{ammine} \tbHM{} \tbPH{ámmìné} \tbPS{rel. interg. pron;pl.~} \tbGE{which people} 
\tbLX{ampe} \tbHM{} \tbPH{ámpé} \tbPS{n.~} \tbGE{a game traditionally played by women} \tbOP{amperi} \tbTP{amperee} 
\tbLX{ana} \tbHM{} \tbPH{áná} \tbPS{third pers. pron. pl.~} \tbGE{those (nonhuman)} \tbXV{Ana ka te na wuo.} \tbXE{We'll collect those.} 
\tbLX{anaare} \tbHM{} \tbPH{ànáárɪ̀} \tbPS{adj.~} \tbGE{four} \tbOP{anaare} \tbTP{anaarɛɛ} \tbVA{naare}
\tbLX{ananso} \tbHM{} \tbPH{ánánsʋ́} \tbPS{conj.~} \tbGE{because} \tbXV{O naŋ taa libie ananso ka o ba mɛ yiri.} \tbXE{S/he built a house because s/he has money.} \tbVA{shortened form of ana la so}
\tbLX{ananso} \tbHM{} \tbPH{ánánsʊ́} \tbPS{n.~} \tbGE{reason} \tbXV{A fo toma zagere gaabo ananso la wola?} \tbXE{What is your reason for refusing to go to work?.} 
\tbLX{anaŋ} \tbHM{} \tbPH{ánǎŋ} \tbPS{pron.pl.~} \tbGE{those ones} \tbXV{Te na iri la anaŋ naŋ sa᷉a te wuli a boŋkoɔrɔ.} \tbXE{We'll select the ones that are bad and show them to the seller.} 
\tbLX{anaŋ} \tbHM{} \tbPH{ánáŋ} \tbPS{pron.~} \tbGE{it (a problem, a situation)} \tbXV{Ka anaŋ waa lɛ, ta di a seemaa.} \tbXE{If that is the case, don't eat the food.} 
\tbLX{andonɛɛ} \tbHM{} \tbPH{ándʋ́nɛ́ɛ́} \tbPS{n.~} \tbSN \tbGE{world} \tbXV{Andonɛɛ waa la yɛloŋ.} \tbXE{The world is large.} \tbSN \tbGE{people of the world} \tbXV{Andonɛɛ ba baŋ ka a wono ba mine.} \tbXE{People of the world do not kow that some people are suffering.} 
\tbLX{ane} \tbHM{} \tbPH{ànɪ̀} \tbPS{conj.~} \tbSN \tbGE{and} \tbXV{Dɛre ane Daŋyi maŋ daŋ la sakuuri gaabo.} \tbXE{Dɛre and Daŋyi always go to school early.} \tbSN \tbGE{although, despite} \tbXV{Ane ka zie e la kpeɛŋaa kyɛ noba mine naŋ ba sage dɔge seŋ lɛ puori tuubu.} \tbXE{Although times are hard, some people do not want to practise family planning.} \tbSN \tbGE{with} \tbXV{O pɔge daa la a dɔɔ ane o faŋa zaa ka o te le.} \tbXE{The woman pushed the man with all her strength and he fell over.} 
\tbLX{aneazaa} \tbHM{} \tbPH{ánɪ́ázàá} \tbPS{conj.~} \tbGE{despite, even though, in spite of} \tbXV{Aneazaa ka saa ba miire kyɛ kôɔ᷉ naŋ be la a baa poɔ.} \tbXE{Even though it has not been raining, there is still water in the pond.} 
\tbLX{aneɛ} \tbHM{} \tbPH{ánɪ́ɛ̀} \tbPS{n.~} \tbGE{effort, attempt} \tbXV{A tontonema da erɛ la aneɛ na baare a toma kyɛ ka saa ba sage.} \tbXE{The workers tried to finish the job but the rain would not allow them.} 
\tbLX{anesɛ} \tbHM{} \tbPH{ánɪ́sɛ̀} \tbPS{v.~} \tbGE{welcome; (an expression of goodwill to a stranger who has just arrived)} \tbXV{Anesɛ, kpeɛrɛ waana.} \tbXE{Welcome, come in!.} \tbPD{-, -, -, -, -, - } 
\tbLX{angyoɔso} \tbHM{} \tbPH{ándzʋ̀ɔ̀sʋ́} \tbPS{n.~} \tbGE{gratitude, thanks} \tbXV{Fo taa la angyoɔso ne a kyɔɔtaa fo naŋ ko ma.} \tbXE{Accept my thanks for the gift.} \tbOP{angyoɔsere} \tbTP{angyoɔserɛɛ} 
\tbLX{anii} \tbHM{} \tbPH{àníì} \tbPS{adj.~} \tbGE{eight} \tbOP{aniini} \tbTP{aniinee} 
\tbLX{anlee} \tbHM{} \tbPH{ánléé} \tbPS{n.~} \tbSN \tbGE{a traditional women's dance} \tbSN \tbGE{music for such a dance} \tbOP{anleeri} \tbTP{anlerri/anleree} \tbVA{an-yee [án-yé:]}
\tbLX{annoo} \tbHM{} \tbPH{ánnʋ́ʊ́} \tbPS{interrogative pronoun.~} \tbGE{who is s/he?; which person?} \tbVA{shortened form of aŋ la o}
\tbLX{ansaŋ} \tbHM{} \tbPH{ánsáŋ} \tbPS{conj.~} \tbGE{before} \tbXV{Vɛŋ ka saa mi ansaŋ ka fo bore a kamaana.} \tbXE{Let it rain before you sow the maize.} 
\tbLX{ansiŋinee} \tbHM{} \tbPH{ánsíŋínéé} \tbPS{n.~} \tbSN \tbGE{a type of edible berry that turns violet when ripe} \tbSN \tbGE{the tree of this fruit} \tbSG{ansiŋinuu} \tbOP{ansiŋini} \tbVA{ansiinee}
\tbLX{ansoɔnee} \tbHM{} \tbPH{ánsʋ́ɔ́nɪ́́ɪ́} \tbPS{n.~} \tbGE{a woman who has recently given birth} \tbOP{ansoɔne/ansoɔmɔ} \tbTP{ansoɔnɛɛ} 
\tbLX{antere} \tbHM{} \tbPH{ántèré} \tbPS{v.~} \tbGE{good afternoon (Southern Dagaare dialect)} \tbPD{-, -, -, -, -, - } 
\tbLX{anuu} \tbHM{} \tbPH{ànúú} \tbPS{adj.~} \tbGE{five} \tbOP{anuuri} \tbTP{anuuree} 
\tbLX{anwoŋ} \tbHM{} \tbPH{ánwóŋ} \tbPS{n.~} \tbGE{one who is great or strong (physically or spiritually)} \tbOP{annwonni} \tbTP{anwonnee} 
\tbLX{aŋ} \tbHM{} \tbPH{ǎŋ} \tbPS{aux.v.~} \tbGE{to be alerted; used to caution or persuade} \tbXV{Ka fo wa gɛrɛ a libidaana na yiri, aŋ baŋ ne baare.} \tbXE{If you are going to that rich man’s house you'd better beware of the dogs.} 
\tbLX{aŋ} \tbHM{} \tbPH{áŋ} \tbPS{interrogative pronoun.~} \tbGE{who} \tbXV{Aŋ la ba nɔŋ bompaala?} \tbXE{Who does not like new things?.} 
\tbLX{aŋa} \tbHM{} \tbPH{̪áŋá} \tbPS{n.~} \tbGE{no one (retort; implies the matter is none of the asker's business)} \tbXV{Aŋ la wa a kyɛ? Aŋa la wa.} \tbXE{Who came here? No one. (Why do you care?).} 
\tbLX{aŋa} \tbHM{} \tbPH{àŋá} \tbPS{adv.~} \tbGE{like} \tbXV{O kono la aŋa bibile na.} \tbXE{S/he cries like a baby.} \tbVA{ŋa}
\tbLX{aŋgaara} \tbHM{} \tbPH{áŋgáárá} \tbPS{adj.~} \tbGE{having the blades burnt away, leaving the stalks (used for grass)} \tbXV{A kyeɛ zo kpɛ la a mɔ-aŋgaara poɔ.} \tbXE{The squirrel has escaped into the stalky bush.} \tbOP{aŋgaare} \tbTP{aŋgaarɛɛ} 
\tbLX{aŋgeli} \tbHM{} \tbPH{áŋgélí} \tbPS{n.~} \tbGE{a type of long grass used to weave sleeping mats} \tbOP{aŋgelli} \tbTP{aŋgelee} 
\tbLX{aŋgoɔle} \tbHM{} \tbPH{áŋgʋ́ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to adorn, to decorate} \tbXV{Ba aŋgoɔle la a naa wiri.} \tbXE{They have decorated the chief's horse.} \tbPD{aŋgoɔlɛɛ, aŋgoɔlɔ, aŋgoɔlɔ, aŋgoɔleba, aŋgoɔlaa, - } 
\tbLX{aŋguoŋaa} \tbHM{} \tbPH{áŋgúóŋáà} \tbPS{n.~} \tbGE{a type of tree with round, pink fruit} \tbOP{aŋguoni/aŋguomo} \tbTP{aŋguonee} \tbVA{pl.var. aŋguôaa}
\tbLX{aŋkaa} \tbHM{} \tbPH{àŋkàá} \tbPS{n.~} \tbGE{orange (fruit)} \tbSG{aŋkaaroo} \tbOP{aŋkaare/aŋkaɛ} \tbTP{aŋkarɛɛ} 
\tbLX{Aŋkara} \tbHM{} \tbPH{àŋkàrà} \tbPS{n.~} \tbGE{Accra} \tbOP{Aŋkarre} \tbTP{Aŋkarrɛɛ} 
\tbLX{aŋko} \tbHM{} \tbPH{áŋkò} \tbPS{n.~} \tbGE{river blindness} \tbOP{aŋkori} \tbTP{aŋkoree} 
\tbLX{aŋkorɔ} \tbHM{} \tbPH{áŋkʋ̀rɔ́} \tbPS{n.~} \tbGE{barrel, drum} \tbOP{aŋkorre} \tbTP{aŋkorɛɛ/aŋkorrɛɛ} \tbVA{aŋkora}
\tbLX{âo} \tbHM{} \tbPH{a᷉ʋ᷉̀} \tbPS{interj.~} \tbGE{of course (expression of certainty)} \tbXV{A. Fo na kuli la be? B. âo.} \tbXE{A.Will you go home? B. Of course I willl.} 
\tbLX{apa} \tbHM{} \tbPH{ápà} \tbPS{interj.~} \tbGE{expression of unpleasant surprise} \tbXV{A soɔŋaa muri la a baa bare kyɛ zo, apa.} \tbXE{The rabbit dodged the dog and escaped, what a pity.} \tbVA{aparapa}
\tbLX{aparapa} \tbHM{} \tbPH{ápárápà} \tbPS{interj.~} \tbGE{emphatic expression of despair} \tbXV{Aparapa, a soɔŋaa ŋa meŋ la zoe?} \tbXE{Oh no, this hare too has escaped?.} \tbVA{apa}
\tbLX{ara} \tbHM{} \tbPH{àra} \tbPS{v.~} \tbGE{to gather up one's courage} \tbXV{A pɔge ara kyɛ daa la a dɔɔlee eŋ tambɔgeŋ.} \tbXE{The woman gathered up her courage and pushed the man into a ditch.} \tbPD{-, -, -, -, -, - } 
\tbLX{arambiri} \tbHM{} \tbPH{árámbírí} \tbPS{n.~} \tbGE{kidney} \tbXV{A dɔɔta yeli ka a baala arambiri kaŋa sa᷈a᷈ɛ la.} \tbXE{The doctor said one of the patient's kidneys had failed.} \tbSG{arambiruu} \tbOP{arembie} \tbTP{arambiiree} \tbVA{arambiri [árámbírí]; pl arambie [árambíè]}
\tbLX{are} \tbHM{} \tbPH{árɪ́} \tbPS{v. pl.~} \tbGE{to cut, harvest (maize)} 
\tbLX{are} \tbHM{} \tbPH{árɪ́} \tbPS{v.~} \tbGE{to be related through the mother's line} \tbXV{N are la Kusielee.} \tbXE{I am maternally related to the Kusielee clan.} \tbPD{arɛɛ, ara, ara, areba, araa, - } 
\tbLX{are} \tbHM{} \tbPH{àrɪ̀} \tbPS{v.~} \tbSN \tbGE{to stand} \tbSN \tbGE{to wait} \tbXV{N baŋ ka o na are la ka N gaa te pɔge o.} \tbXE{I know s/he will wait so that I go to meet him/her.} \tbSN \tbGE{to stop} \tbXV{A wɔɔkye kyɛŋɛɛ la tasoga tasoga leɛ are.} \tbXE{The watch worked for a while and then stopped.} \tbPD{arɛɛ, ara, ara, areba, araa, - } 
\tbLX{are} \tbHM{} \tbPH{árɪ̀} \tbPS{v.~} \tbGE{to tear off from the main body (branch or limb of an animal)} \tbXV{A maŋgo wɔŋ zue la ka a ulee kaŋa are le.} \tbXE{The mango tree has borne so many fruits that one of the branches tore off.} \tbPD{arɛɛ, ara, ara, areba, araa, areba } 
\tbLX{areba} \tbHM{} \tbPH{árɪ́bá} \tbPS{n.~} \tbGE{maternal uncle} \tbOP{aremine} \tbTP{arrɛɛ} 
\tbLX{arekpoŋ} \tbHM{} \tbPH{árɪ́kpóŋ} \tbPS{n.~} \tbGE{eldest niece} \tbOP{arekponni} \tbTP{arekponnee} 
\tbLX{arelee} \tbHM{} \tbPH{árɪ́lèé} \tbPS{n.~} \tbGE{nephew or niece} \tbOP{arebilii} \tbTP{arebilee/arebillee} 
\tbLX{aremateke} \tbHM{} \tbPH{àrɪ́mátɪ́kɪ̀} \tbPS{n.~} \tbGE{arithmetic} \tbOP{arematekere} \tbVA{kuntaŋ (Akan)}
\tbLX{arezie} \tbHM{} \tbPH{[àrɪ̀zíé]} \tbPS{n.~} \tbSN \tbGE{standing place, spot} \tbSN \tbGE{post, position, rank} \tbXV{O arezie kpoŋ.} \tbXE{S/he holds a high position.} \tbSN \tbGE{position, stand (for or against an idea)} \tbXV{O ba wuli o arezie a nɔkpeɛne poɔ.} \tbXE{She has not indicated her position with regard to the argument..} \tbOP{areziiri} \tbTP{areziiree} 
\tbLX{aroozie} \tbHM{} \tbPH{árʋ́ʊ́zíé} \tbPS{n.~} \tbSN \tbGE{standing place; position} \tbSN \tbGE{adverb of place (on, under)} \tbOP{arooziiri} \tbTP{arooziiree} 
\tbLX{aseŋ} \tbHM{} \tbPH{àsɪ́ŋ} \tbPS{adv.~} \tbGE{for instance, like} 
\tbLX{asibiti} \tbHM{} \tbPH{ásíbítì} \tbPS{n.~} \tbGE{hospital, clinic} \tbOP{asibitiri} \tbTP{asibitiree} \tbVA{asiiti, [ásí:tì], asimeti, [ásímétì], asebere.}
\tbLX{ata} \tbHM{} \tbPH{átà} \tbPS{adj.~} \tbGE{three} \tbOP{atare} \tbTP{atarɛɛ} \tbSE{ata ata} \tbGE{in threes} 
\tbLX{awae} \tbHM{} \tbPH{àwáɪ̀} \tbPS{adj.~} \tbGE{nine} \tbOP{awarre} \tbTP{awarɛɛ/awarrɛɛ} \tbVA{wae}\tbSE{awae awae} \tbGE{nine each} 
\tbLX{awola} \tbHM{} \tbPH{àwʊ́là} \tbPS{adv.~} \tbSN \tbGE{how much (price)} \tbXV{Awola ka fo da a kparoo?} \tbXE{How much did the shirt cost you?.} \tbSN \tbGE{how many (nonhuman)} \tbXV{Teere awola ka yɛ sɛle?} \tbXE{How many trees did you plant?.} 
\tbLX{awommu} \tbHM{} \tbPH{àwómmú} \tbPS{n.~} \tbGE{hardship, suffering} \tbXV{Ka kɔŋeŋ le awommu maŋ pɔge la noba yaga.} \tbXE{When there is a famine, many people experience a lot of suffering.} \tbTP{awonnee} \tbVA{awonaa [awona:]}
\tbLX{awulo} \tbHM{} \tbPH{áwúló} \tbPS{n.~} \tbGE{continuous, loud, sharp cry of pain} \tbXV{A pɔge na bie naŋ kpi la lɔɔrɔ awulo.} \tbXE{The woman who lost her son is crying hysterically.} \tbTP{awulee} 
\tbLX{ayi} \tbHM{} \tbPH{áyì} \tbPS{adj.~} \tbGE{two} \tbOP{ayiri} \tbTP{ayiree} \tbSE{ayi ayi} \tbGE{in twos} 
\tbLX{ayililili} \tbHM{} \tbPH{ayililili} \tbPS{interjection.~} \tbGE{joyous, shrill shouting in commendation (usually by women)} \tbXV{A pɔgeba ŋmɛ la kulinni, ayililili, eŋ a bawaseɛrebɛ.} \tbXE{The women made shrill noises of commendation for the dancers.} \tbVA{ayiyiyiyi}
\tbLX{ayoɔbo} \tbHM{} \tbPH{àyʊ̀ɔ̀b} \tbPS{adj.~} \tbGE{six} \tbOP{ayoɔre} \tbTP{ayoɔrɛɛ} 
\tbLX{Ayoree} \tbHM{} \tbPH{áyóréé} \tbPS{n.~} \tbGE{name of a clan} \tbOP{Ayoruu} 
\tbLX{Ayɔɔ} \tbHM{} \tbPH{áyɔ̀ɔ̀} \tbPS{n.~} \tbGE{name traditionallly given to a female child born on a market day} \tbOP{Ayɔɔmine} \tbTP{Ayɔɔrɛɛ} 
\tbLX{ayɔpo᷉i᷉} \tbHM{} \tbPH{àyʊ̀ɔ̀pó᷉ì᷉} \tbPS{adj.~} \tbGE{seven} \tbOP{ayoɔponni} \tbTP{ayoɔponnee} 
\tbLX{Ayuo} \tbHM{} \tbPH{áyúó} \tbPS{n.~} \tbGE{name traditionally given to a female child born on a funeral day} \tbOP{Ayumine} \tbTP{Ayuuree} 
\tbLX{Azaasoba} \tbHM{} \tbPH{àzàásʊ́bá} \tbPS{n.~} \tbGE{almighty} 
\tbLX{ba} \tbHM{} \tbPH{bá} \tbPS{adv.~} \tbGE{not (negative particle)} \tbXV{N ba nyɛ o zenɛ.} \tbXE{I have not seen him today.} 
\tbLX{ba} \tbHM{} \tbPH{bà} \tbPS{pron.~} \tbSN \tbGE{they (human)} \tbXV{Ba na wa la.} \tbXE{They will come.} \tbSN \tbGE{their (human)} \tbXV{Te de la ba bondirii biŋ.} \tbXE{We have reserved their food.} \tbSN \tbGE{them (human)} \tbXV{Te de la a bondirii ko ba.} \tbXE{We have given them the food.} 
\tbLX{ba} \tbHM{} \tbPH{bà} \tbPS{v.~} \tbGE{to fix a pointed object firmly in the ground} \tbXV{Ba ba la daare gɔlle a dabɔgɔ zaa.} \tbXE{They have put up planks around the whole garden.} \tbPD{baɛ, baara, baara, bareba, baaraa, - } 
\tbLX{ba} \tbHM{} \tbPH{bà} \tbPS{v.~} \tbGE{to take a long drink from a container} \tbXV{A saana de la a ko᷉ɔ ba lɛɛ ka a ŋmane wa te kyɛre.} \tbXE{The stranger drank deeply until the calabash was empty.} \tbPD{baɛ, baara, baara, bareba, baaraa, - } 
\tbLX{ba} \tbHM{} \tbPH{bá} \tbPS{v.~} \tbGE{to gallop} \tbPD{baɛ, baara, baara, bareba, baaraa, - } 
\tbLX{ba} \tbHM{} \tbPH{bâ} \tbPS{n.~} \tbSN \tbGE{father} \tbSN \tbGE{friend (between males)} \tbSN \tbGE{Mr} \tbOP{bamine} \tbTP{barɛɛ} 
\tbLX{baa} \tbHM{} \tbPH{báà} \tbPS{conj.~} \tbGE{or; (normally used in question tag)} \tbXV{Te na gaa la, baa wolɔ?} \tbXE{We'll go, won't we?.} \tbVA{bee}
\tbLX{baa} \tbHM{} \tbPH{báá} \tbPS{n.~} \tbSN \tbGE{dog} \tbSN \tbGE{a small black ground ant-like creature which burrows into the ground with the tip of its abdomen} \tbSN \tbGE{a type of children's game} \tbOP{baare} \tbTP{baarɛɛ} 
\tbLX{baa} \tbHM{} \tbPH{bàá} \tbPS{n.~} \tbGE{body of water} \tbOP{bare} \tbTP{barɛɛ} 
\tbLX{baa} \tbHM{} \tbPH{bàà} \tbPS{v.~} \tbSN \tbGE{to grow up, to flourish} \tbXV{Sêɛ saŋa, moɔ maŋ baa la wogri lɛ.} \tbXE{During the rainy season, grass normally grows tall.} \tbSN \tbGE{to mature} \tbXV{A pɔgelee naŋ ba baa azuiŋ o ba ta serɛ kuluu.} \tbXE{The girl has not yet matured enough, so she is not marriageable.} \tbSN \tbGE{to survive, to live} \tbXV{A baala naŋ ba kpi zenɛ o na baa la.} \tbXE{Since the patient has not died today, s/he will survive.} \tbPD{-, -, -, -, -, - } 
\tbLX{baabaa} \tbHM{} \tbPH{bààbáá} \tbPS{n.~} \tbGE{father (familiar term of address used by toddlers)} \tbOP{baabaamine} \tbTP{baabaarɛɛ} 
\tbLX{Baabol} \tbHM{} \tbPH{báábʋ̀l} \tbPS{n.~} \tbGE{Bible} \tbOP{baabola} \tbTP{baabolɛɛ/baabollɛɛ} 
\tbLX{ba᷉abo᷉o} \tbHM{} \tbPH{bà᷉à᷉bʊ́᷉ʊ́᷉} \tbPS{n.~} \tbGE{bustard} \tbOP{ba᷉aboomɔ/ba᷉aboone} \tbTP{ba᷉aboomɛɛ/ba᷉aboonɛɛ} 
\tbLX{baaebaae} \tbHM{} \tbPH{bááɪ́bááɪ́} \tbPS{interj.~} \tbGE{goodbye} 
\tbLX{baage} \tbHM{} \tbPH{báágɪ̀} \tbPS{n.~} \tbGE{bag} \tbOP{baagere} \tbTP{bagerɛɛ} \tbVA{baagye}
\tbLX{baala} \tbHM{} \tbPH{bààlá} \tbPS{n.~} \tbGE{patient, sick person} \tbOP{baalba} \tbTP{baalbɛɛ} 
\tbLX{baalaa} \tbHM{} \tbPH{bààlàà} \tbPS{adj.~} \tbSN \tbGE{slim, slender, thin} \tbXV{N da nyɛ la pɔgebil baalaa kaŋa a seɛre zie.} \tbXE{I saw a slender girl at the dance.} \tbSN \tbGE{narrow} \tbXV{A sori waa la baalaa lɛ.} \tbXE{The road is very narrow.} \tbOP{baale} \tbTP{baalɛɛ} 
\tbLX{baale} \tbHM{} \tbPH{báálɪ̀} \tbPS{v.~} \tbGE{to ache slightly (disconfort of the stomach)} \tbXV{Ka maaŋ iri ba gaa bangyeraa n poɔ maŋ baala la.} \tbXE{If I don’t go to toilet for the day I normally have a slight stomach ache.} \tbPD{baalɛɛ, baala, baala, baaleba, baalaa, - } 
\tbLX{baale} \tbHM{} \tbPH{báálɪ̀} \tbPS{v.~} \tbGE{to separate (grains from unwanted material)} \tbXV{A pɔge baale iri la a seŋkãabie yi a pɛgere poɔ.} \tbXE{The woman separated the groundnuts from the shells.} \tbPD{baalɛɛ, baala, baala, baaleba, baalaa, - } 
\tbLX{baalebayiri} \tbHM{} \tbPH{bààlbáyírì} \tbPS{n.~} \tbGE{hospital} \tbOP{baalebayie} \tbTP{baalebayiree} 
\tbLX{baalimbo} \tbHM{} \tbPH{bààlímbò} \tbPS{n.~} \tbGE{the act of supporting someone by holding them by the midsection} \tbXV{Ba kpɛ la a baala baalimbo gaa ne a asibiti.} \tbXE{They supported the patient by the midsection and took him to the hospital.} \tbOP{baalimbori} \tbTP{baalimboree} \tbVA{boolimbo}
\tbLX{baalombiri} \tbHM{} \tbPH{bààlʋ́mbírì} \tbPS{n.~} \tbGE{germ, bacterium} \tbOP{baalombie} \tbTP{baalombiiree} 
\tbLX{baalomboɔlaa} \tbHM{} \tbPH{báálʋ̀mbʋ̀ɔ̀làá} \tbPS{n.~} \tbGE{feast organised for a friend} \tbXV{Te daare gaa la baalomboɔlaa Loraa poɔ.} \tbXE{We attended a friendship feast in Lawra.} \tbXV{baalomboɔle} \tbXE{.} 
\tbLX{baalonlɔnnaa} \tbHM{} \tbPH{bààlʋ̀nlɔ́nnáá} \tbPS{n.~} \tbGE{contagious disease} \tbOP{baalonlɔnnɔ} \tbTP{baalonlɔnnɛɛ} 
\tbLX{baalonnooraa} \tbHM{} \tbPH{bààlʋ́nnʋ̀ʋ̀ráà} \tbPS{n.~} \tbGE{recurring illness} \tbXV{baalonnoore} \tbXE{} 
\tbLX{baaloŋ} \tbHM{} \tbPH{báálʋ́ŋ} \tbPS{n.~} \tbGE{friendship} \tbXV{Ba nyɔge la baaloŋ saŋa zaa.} \tbXE{They have been friends for a long time.} \tbOP{baalondeme} 
\tbLX{baaloŋ} \tbHM{} \tbPH{bààlʊ́ŋ} \tbPS{n.~} \tbGE{illness, sickness} \tbXV{A baala baaloŋ seɛ o la zenɛ.} \tbXE{The patient's illness has improved today.} 
\tbLX{baaloŋ} \tbHM{} \tbPH{bààlʊ̀ŋ} \tbPS{adj.~} \tbSN \tbGE{calm} \tbXV{Zenɛ, a mane ko᷉ɔ᷉ e la baaloŋ.} \tbXE{Today, the sea is calm.} \tbSN \tbGE{gentle, gently} \tbXV{A doɔ baaloŋ na wa la kyɛ.} \tbXE{That gentle man came here.} \tbXV{A doɔ wa la baaloŋ.} \tbXE{The man is gentle.} \tbXV{De a bileɛ baaloŋ.} \tbXE{Take the baby gently.} \tbSN \tbGE{soft (objects, sounds)} \tbXV{A kapuro ŋa e la baaloŋ yelmeŋɛ.} \tbXE{This pillow is really soft.} \tbSN \tbGE{smooth (powdery substances)} \tbXV{Bɛnzɔŋ maŋ waa la baaloŋ.} \tbXE{Bean flour is usually smooth.} 
\tbLX{baalugiri} \tbHM{} \tbPH{bààlùgìrí} \tbPS{n.~} \tbGE{armpit} \tbSG{baalugiruu} \tbOP{baalugo} \tbTP{baalugiree} \tbVA{bagelugri}
\tbLX{baaluu} \tbHM{} \tbPH{bààlúú} \tbPS{n.~} \tbSN \tbGE{gallon measure} \tbSN \tbGE{balloon} \tbSG{baaluuruu} \tbOP{baaluuri} \tbTP{baaluuree} 
\tbLX{baa-ne-ŋmaaŋa} \tbHM{} \tbPH{bàá-nɪ́-ŋmááŋà} \tbPS{n.~} \tbGE{Ursa Major (constellation); also known as Big Dipper, Plough, Great Bear} 
\tbLX{baapɛmpɛle} \tbHM{} \tbPH{bàápɛ̀mpɛ́lɛ̀} \tbPS{n. pl.~} \tbGE{lily pads} 
\tbLX{baaraa} \tbHM{} \tbPH{bááráá} \tbPS{n.~} \tbGE{end, last} \tbXV{A baaraa zaa te na nyɛ la a nannyige.} \tbXE{At the end of it all, we will find the thief.} 
\tbLX{baaraakyi} \tbHM{} \tbPH{báárááʧì} \tbPS{n.~} \tbGE{bridge} \tbOP{baaraakyiri} \tbTP{baaraakyiree} 
\tbLX{baare} \tbHM{} \tbPH{báárɪ̀} \tbPS{v.~} \tbSN \tbGE{to finish, to end, to complete} \tbXV{A tontonema baare la a toma.} \tbXE{The workers have finished the work.} \tbSN \tbGE{to be used up} \tbXV{A baaloŋ vɛŋɛɛ la ka a dɔɔ bɛroŋ na baare la o zaa.} \tbXE{Sickness has caused that man to lose weight.} \tbPD{baarɛɛ, baara, baara, baareba, baaraa, - } 
\tbLX{baasaabiri} \tbHM{} \tbPH{bààsààbírí} \tbPS{n.~} \tbSN \tbGE{a seed from a type of tree} \tbSN \tbGE{a type of large bead made from this seed} \tbSG{baasaabiruu} \tbOP{baasaabie} \tbTP{baasaabiiree} 
\tbLX{Baasaale} \tbHM{} \tbPH{bàásààlɪ̀} \tbPS{n.~} \tbGE{Baasare (ethnic group)} \tbSG{baasaaloo} 
\tbLX{baasakuuri} \tbHM{} \tbPH{báásàkúúrì} \tbPS{n.~} \tbGE{bicycle} \tbOP{baasakue} \tbTP{baasakuuree} 
\tbLX{baataŋaa} \tbHM{} \tbPH{báátáŋáá} \tbPS{n.~} \tbGE{a type of tree that resembles the shea nut tree} \tbOP{baataŋema} \tbTP{baataŋemɛɛ/baataanɛɛ} 
\tbLX{babammo} \tbHM{} \tbPH{bábámmʊ́} \tbPS{n.~} \tbGE{ignorance} 
\tbLX{babaŋena} \tbHM{} \tbPH{bábáŋɪ́ná} \tbPS{n.~} \tbGE{one who is innocent} \tbOP{babaŋenema} \tbTP{babaŋenemɛɛ} 
\tbLX{babaŋyeli} \tbHM{} \tbPH{bábàŋyélí} \tbPS{n.~} \tbGE{innocence} \tbVA{dakumo}
\tbLX{babe᷉e} \tbHM{} \tbPH{bàbɪ́᷉ɪ́᷉} \tbPS{n.~} \tbGE{broth from dog meat} \tbTP{babeenɛɛ} 
\tbLX{babeɛ} \tbHM{} \tbPH{bábɪ́ɛ́} \tbPS{n.~} \tbGE{a friend that cannot be trusted} \tbOP{babeere} \tbTP{babeerɛɛ} 
\tbLX{babeɛraa} \tbHM{} \tbPH{bàbɪ̀ɛ́ráà} \tbPS{n.~} \tbGE{bridge} \tbOP{babeɛre} \tbTP{babeɛrɛɛ} \tbVA{baaraakyi, takoriko.}
\tbLX{babere} \tbHM{} \tbPH{bàbɪ̀rɪ́} \tbPS{n.~} \tbGE{the breast of a dog} \tbOP{baberɛ} \tbTP{baberɛɛ} 
\tbLX{babini} \tbHM{} \tbPH{bàbíní} \tbPS{n.~} \tbGE{dog faeces} \tbOP{babinee} \tbTP{babinnee} 
\tbLX{babiŋ} \tbHM{} \tbPH{bábíŋ} \tbPS{n.~} \tbGE{a type of tree} \tbOP{babinni} \tbTP{babinnee} 
\tbLX{babiri} \tbHM{} \tbPH{bábírì} \tbPS{n.~} \tbGE{bosom friend} \tbOP{babie} \tbTP{babiiree} \tbVA{batuo}
\tbLX{babogi} \tbHM{} \tbPH{bàbógí} \tbPS{n.~} \tbGE{the deepest part of a body of water} \tbOP{babogri} \tbTP{babogree} 
\tbLX{babullaa} \tbHM{} \tbPH{bàbùlláà} \tbPS{n.~} \tbGE{spring, pond where water is drawn in small amounts (usually during dry season)} \tbOP{babulli} 
\tbLX{badaa} \tbHM{} \tbPH{bàdáá} \tbPS{n.~} \tbGE{male dog} \tbOP{badaare} \tbTP{badaarɛɛ} 
\tbLX{badaa-de-nyoɔre} \tbHM{} \tbPH{bàdáádíɲɛʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{the last day of a festival} \tbTP{badaadenyoɔrɛɛ} \tbVA{baraadenyoɔre}
\tbLX{badalane} \tbHM{} \tbPH{bàdàlánɪ́} \tbPS{n.~} \tbGE{uncastrated male dog} \tbOP{badalama} \tbTP{badalamɛɛ} 
\tbLX{badalee} \tbHM{} \tbPH{bàdàléé} \tbPS{n.~} \tbGE{male puppy} \tbOP{badabilii} \tbTP{badabilee/badabillee} 
\tbLX{badavaraa} \tbHM{} \tbPH{bàdàváràà} \tbPS{n.~} \tbGE{castrated dog} \tbOP{badavarre} \tbTP{badavarɛɛ/badavarrɛɛ} 
\tbLX{badɛre} \tbHM{} \tbPH{bádɛ́rɪ́} \tbPS{n.~} \tbSN \tbGE{spider} \tbSN \tbGE{cunning person (fig.)} \tbOP{badɛrre} \tbTP{badɛrɛɛ} 
\tbLX{badie} \tbHM{} \tbPH{bàdìé} \tbPS{n.~} \tbGE{the deepest part of a body of water} \tbOP{baderi} \tbTP{baderee} 
\tbLX{badoge} \tbHM{} \tbPH{bàdʊ̀gɪ̀} \tbPS{n.~} \tbGE{a plant with a sweet smell, used to season sauce} \tbOP{badogre} \tbTP{badogrɛɛ} 
\tbLX{badɔgerɔ} \tbHM{} \tbPH{bàdɔ̀grɔ́} \tbPS{n.~} \tbGE{female dog that has recently given birth} \tbOP{badɔgere} \tbTP{badɔgerɛɛ} \tbVA{badɔgera}
\tbLX{bafoɔlaa} \tbHM{} \tbPH{bàfʊ́ɔ́láá} \tbPS{n.~} \tbSN \tbGE{valley} \tbSN \tbGE{creek} \tbOP{bafoɔle} \tbTP{baroɔlɛɛ} 
\tbLX{baga} \tbHM{} \tbPH{bágà} \tbPS{n.~} \tbGE{a triangular garment worn by men as underwear} \tbOP{bagere} \tbTP{bagerɛɛ} \tbVA{lere, para}
\tbLX{baga} \tbHM{} \tbPH{bàgà} \tbPS{n.~} \tbGE{not caring, unconcerned} \tbXV{Ka onaŋ ba wa a pɔgebeɛloŋ kyɛ n baga kyebe.} \tbXE{I don't care if she doesn't attend the wedding.} 
\tbLX{baganaa} \tbHM{} \tbPH{bàgànàà} \tbPS{n.~} \tbGE{dog leash} \tbOP{baganne} \tbTP{baganɛɛ/bagannɛɛ} 
\tbLX{bagane} \tbHM{} \tbPH{bàgánɪ̀} \tbPS{n.~} \tbGE{dog skin, dog hide} \tbOP{bagama} \tbTP{bagannɛɛ} 
\tbLX{bagaŋ} \tbHM{} \tbPH{bágáŋ} \tbPS{n.~} \tbGE{a mad dog} \tbOP{baganne} \tbTP{baganɛɛ/bagannɛɛ} \tbVA{bagao}
\tbLX{bagao} \tbHM{} \tbPH{bàgáʋ̀} \tbPS{n.~} \tbGE{mad dog} \tbOP{bagagre} \tbTP{bagagrɛɛ} \tbVA{bagaŋ}
\tbLX{bageba} \tbHM{} \tbPH{bágɪ́bá} \tbPS{n.~} \tbGE{a type of sacred cult which initiates boys and girls into adulthood} \tbTP{bagebɛɛ} \tbVA{bɔgebɔ}
\tbLX{bageboge} \tbHM{} \tbPH{bàgɪ̀bʋ̀gɪ̀} \tbPS{n.~} \tbGE{the act of divining} \tbXV{Kiritabiiri kyiiruŋ la bageboge.} \tbXE{It is a taboo for Christians to participate  in divining.} \tbVA{bɔgeboge}
\tbLX{bageboge-dagoli} \tbHM{} \tbPH{bàgɪ̀bʊ̀g-dàgólí} \tbPS{n.~} \tbGE{divining stick} \tbOP{bageboge-dagolo} \tbTP{bageboge-dagolee} \tbVA{bɔgeboge-dagoli}
\tbLX{bageboge-karaa} \tbHM{} \tbPH{bàgɪ̀bʊ̀g-kàràá} \tbPS{n.~} \tbGE{chief diviner} \tbOP{bagebogekarre} \tbTP{bagebogekarɛɛ} \tbVA{bɔgeboge-karaa}
\tbLX{bageboge-woɔ} \tbHM{} \tbPH{bàgɪ̀bʊ̀gɪ̀-wʊ́ɔ́} \tbPS{n.~} \tbGE{divining bag} \tbOP{bageboge-woore} \tbTP{bageboge-woorɛɛ} \tbVA{bɔgeboge-woɔ}
\tbLX{bagere} \tbHM{} \tbPH{bàgrɪ́} \tbPS{n.~} \tbGE{sacrifice for gods} \tbOP{baga} \tbTP{bagɛɛ} \tbVA{bɔgere}
\tbLX{bagukyɔfeɛle} \tbHM{} \tbPH{bágùʧɔ̀nfɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{very early morning (usually cold)} 
\tbLX{baguo} \tbHM{} \tbPH{bágúó} \tbPS{n.~} \tbGE{morning} \tbOP{baguuri} \tbTP{baguuree} \tbVA{beguo}
\tbLX{bagupi} \tbHM{} \tbPH{bàgùpí} \tbPS{n.~} \tbGE{early morning} \tbOP{baguipiri} \tbTP{bagupiree} \tbVA{bebipi}
\tbLX{bagupipi} \tbHM{} \tbPH{bágùpípí} \tbPS{n.~} \tbGE{very early morning} \tbOP{bagupipiri} \tbTP{bagupipiree} \tbVA{bagupi}
\tbLX{bagbɛre} \tbHM{} \tbPH{bàgbɛ́rɪ̀} \tbPS{n.~} \tbGE{leg of a dog} \tbOP{bagbɛɛ} \tbTP{bagbɛɛrɛɛ} 
\tbLX{bagbɛre} \tbHM{} \tbPH{bàgbɛ́rɪ̀} \tbPS{n.~} \tbGE{valley} \tbOP{bagbɛɛ} \tbTP{bagbɛɛrɛɛ} 
\tbLX{bagboŋ} \tbHM{} \tbPH{bágbòŋ} \tbPS{n.~} \tbGE{fool} \tbOP{bagbonni} \tbTP{bagbonnee} 
\tbLX{bagbori} \tbHM{} \tbPH{bàgbòrí} \tbPS{n.~} \tbGE{trough for a dog} \tbOP{bagboe} \tbTP{bagboree} 
\tbLX{bagyeɛnɛ} \tbHM{} \tbPH{bàdzɪ̀ɛ̀nɛ́} \tbPS{n.~} \tbGE{a type of children's dance} \tbTP{bagyeɛnɛɛ} 
\tbLX{bagyoɔ} \tbHM{} \tbPH{bàdzʊ́ɔ́} \tbPS{n.~} \tbSN \tbGE{squirrel-like rodent (believed to be very sexually active)} \tbSN \tbGE{sex maniac (male)} \tbVA{bagyoa, bazoɔ}
\tbLX{bagyɔroo} \tbHM{} \tbPH{bàdzɔ̀rʊ́ʊ́} \tbPS{n.~} \tbGE{a thick, fleshy, poisonous creeping plant} 
\tbLX{ba-irime} \tbHM{} \tbPH{bá-írímé} \tbPS{n.~} \tbGE{a type of wild mouse found along river valleys which has tender skin} \tbOP{ba-irri} \tbTP{ba-iree} 
\tbLX{bakana} \tbHM{} \tbPH{bàkáná} \tbPS{n.~} \tbGE{broth from dog meat, made with flour} \tbTP{bakanɛɛ} 
\tbLX{bakore} \tbHM{} \tbPH{bákʊ́rɪ́} \tbPS{n.~} \tbGE{trusted friend} \tbOP{bakorɔ/bakoremine} \tbTP{bakorɛɛ/bakorrɛɛ} 
\tbLX{bakɔ} \tbHM{} \tbPH{bàkɔ̀} \tbPS{n.~} \tbGE{a type of yam (early maturing)} \tbOP{bakɔɛ} \tbTP{bakɔrɛɛ} 
\tbLX{bakpoŋ} \tbHM{} \tbPH{bàkpóŋ} \tbPS{n.~} \tbSN \tbGE{large stream, river} \tbSN \tbGE{(capitalized) the name of a tributary of the Black Volta River in the Upper West Region} \tbOP{bakponni} \tbTP{bakponnee} 
\tbLX{balaa} \tbHM{} \tbPH{báláà} \tbPS{n.~} \tbGE{waistband (pair of shorts)} \tbOP{balle} \tbTP{balɛɛ} 
\tbLX{balambaaraa} \tbHM{} \tbPH{bálámbááráá} \tbPS{n.~} \tbGE{peg used for tethering animals} \tbOP{balambaare} \tbTP{balambaarɛɛ} \tbVA{balambiri}
\tbLX{balambiri} \tbHM{} \tbPH{bálámbírí} \tbPS{n.~} \tbGE{peg used for tethering animals} \tbOP{balambie} \tbTP{balambiiree} \tbVA{balambaaraa}
\tbLX{balantamm} \tbHM{} \tbPH{bàlàntàmm} \tbPS{adv.~} \tbGE{large and heavy} \tbXV{Wɔɔ waa la kpoŋi balantamm.} \tbXE{The elephant is enormously large.} 
\tbLX{balante} \tbHM{} \tbPH{bálántɪ́} \tbPS{n.~} \tbGE{belt} \tbSG{balanteerɛɛ} \tbOP{balantere} \tbTP{balantere} \tbVA{bɛlɛnte}
\tbLX{bale} \tbHM{} \tbPH{bàlɪ̀} \tbPS{v.~} \tbSN \tbGE{to be tired} \tbXV{O toŋ la a toma te zuo lɛnso ka o bale.} \tbXE{He has overworked himself; that's why he is tired.} \tbSN \tbGE{to be fed up} \tbXV{Bebiri zaa ka o maŋ wa sɔre libie, maa eŋ bale la o yɛlɛ.} \tbXE{He comes to beg for money every day; I'm fed up with him.} \tbSN \tbGE{to be weak} \tbXV{Ka neɛŋ kore o maŋ bale la.} \tbXE{One becomes weak when one is old.} \tbSN \tbGE{to be soft} \tbXV{A gane koe la, yɛ mere ko᷉ɔ ka o bale.} \tbXE{The animal skin is dry and hard, so sprinkle water on it to make it soft.} \tbPD{balɛɛ, balla, balla, baleba, balaa, baleba } 
\tbLX{bale} \tbHM{} \tbPH{bálɪ́} \tbPS{n.~} \tbGE{tribe, ethnic group} \tbXV{Dagaaba bale kaŋa yuori la Kowɛrɛɛ.} \tbXE{Kowɛrɛɛ is one of the tribes among the Dagaaba.} \tbOP{balmine} \tbTP{balɛɛ} 
\tbLX{bale} \tbHM{} \tbPH{bálɪ́} \tbPS{v.~} \tbSN \tbGE{avoid, evade} \tbXV{O daare to ma lɛnso ka o maŋ wa nyɛ ma bale.} \tbXE{She insulted me the last time, and that's why she avoids me when she sees me.} \tbSN \tbGE{divert} \tbXV{Ba naŋ da beɛrɛ a baa ba bale la a sori.} \tbXE{When they were constucting the bridge, they diverted the road.} \tbPD{balɛɛ, bala, bala, baleba, balaa, baleba } 
\tbLX{bale} \tbHM{} \tbPH{bálɪ́} \tbPS{adj.~} \tbGE{of a kind} \tbXV{A toma ŋa bale ba kye kyɛ.} \tbXE{We do not have his kind of work here.} 
\tbLX{balee} \tbHM{} \tbPH{bàléé} \tbPS{n.~} \tbGE{brook, stream} \tbOP{babilii} \tbTP{babillee} 
\tbLX{balee} \tbHM{} \tbPH{bàléé} \tbPS{n.~} \tbGE{puppy} \tbOP{babilii} \tbTP{babillee} 
\tbLX{balee} \tbHM{} \tbPH{bàlɪ̀ɪ̀} \tbPS{adj.~} \tbGE{of a dubious and dangerous nature} \tbXV{Pɔge koŋ sage toŋ tombalee ŋa tɔ.} \tbXE{A woman will not do this sort of shady thing.} 
\tbLX{baleŋkpɔgre} \tbHM{} \tbPH{bàlɪ́ŋkpɔ̀grì} \tbPS{n.~} \tbGE{tick} \tbOP{baleŋkpɔgɔ} \tbTP{baleŋkpɔgrɛɛ} 
\tbLX{balɔre} \tbHM{} \tbPH{bálɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{ugly person or thing} \tbSN \tbGE{one without strength} \tbOP{balɔɛ} \tbTP{balɔrɛɛ} 
\tbLX{balɔrroŋ} \tbHM{} \tbPH{bálɔ̀rrʊ́ŋ} \tbPS{n.~} \tbGE{not being strong} 
\tbLX{balɔrroŋ} \tbHM{} \tbPH{bálɔ̀rrʊ́ŋ} \tbPS{n.~} \tbGE{act of being ugly} 
\tbLX{bam} \tbHM{} \tbPH{bǎm} \tbPS{n.~} \tbGE{a type of large yam} \tbOP{bama} \tbTP{bamɛɛ} 
\tbLX{bama} \tbHM{} \tbPH{bámà} \tbPS{demonstrative pronoun.~} \tbGE{these (human)} \tbXV{A noba bama la maŋ toŋ a toma wie᷉u.} \tbXE{These are the people who normally do the work quickly.} \tbXV{Bibiiri bama la ka ba iri.} \tbXE{These are the children that have been chosen.} 
\tbLX{bambaala} \tbHM{} \tbPH{bàmbààlá} \tbPS{n.~} \tbGE{unassuming person; poor person} \tbOP{bambaaleba} \tbTP{bambaalebɛɛ} 
\tbLX{bambaalaa} \tbHM{} \tbPH{bámbááláá} \tbPS{n.~} \tbGE{small fish, fingerling} \tbOP{bambaale} \tbTP{bambaalɛɛ} 
\tbLX{bambalaaraa} \tbHM{} \tbPH{bámbálááráá} \tbPS{n.~} \tbGE{swallow (bird)} \tbOP{bambalaare} \tbTP{bambalaarɛɛ} 
\tbLX{bambaŋ} \tbHM{} \tbPH{bámbáŋ} \tbPS{adv.~} \tbGE{hot, steamy} \tbXV{A zee, re ba taa diibu, a toloŋ bambaŋ.} \tbXE{The soup is too hot; it can't be eaten.} \tbXV{Tentoloŋ saŋa a ŋmenaa maŋ toloŋ bambaŋ.} \tbXE{During the hot season, the sun is always extremely hot.} 
\tbLX{bambiŋi} \tbHM{} \tbPH{bàmbìŋ} \tbPS{n.~} \tbGE{an arrangement made in advance by two parties} 
\tbLX{bambugo} \tbHM{} \tbPH{bámbúgó} \tbPS{n.~} \tbGE{confused person; fool} \tbOP{bambugri} \tbTP{ bambugree} 
\tbLX{bambuŋi} \tbHM{} \tbPH{bàmbúŋ} \tbPS{n.~} \tbGE{salp (fish of the class Thaliacea)} \tbOP{babunni} \tbTP{bambunnee} 
\tbLX{bambuŋi} \tbHM{} \tbPH{bàmbúŋ} \tbPS{n.~} \tbGE{locally made gunpowder} \tbOP{babunni} \tbTP{bambunnee} 
\tbLX{bamm} \tbHM{} \tbPH{bàmm} \tbPS{adv.~} \tbGE{big and heavy} \tbXV{A nannyige zu mɔrɛɛ bamm lɛ.} \tbXE{The thief's head is badly swollen.} 
\tbLX{bammo} \tbHM{} \tbPH{bàmmʊ́} \tbPS{n.~} \tbGE{intelligence, wit} \tbXV{Kɔkɔlaa taa la bammo gaŋ baa.} \tbXE{The cat is more intelligent than the dog.} 
\tbLX{bamoɔ} \tbHM{} \tbPH{bàmʊ́ɔ́} \tbPS{n.~} \tbGE{a type of tall, strong grass found around bodies of water, whose dry stalks are used for weaving} \tbSG{bamoɔnoo} \tbOP{bamoore} \tbTP{bamoorɛɛ} 
\tbLX{bandaa} \tbHM{} \tbPH{bándáá} \tbPS{n.~} \tbGE{lizard (male)} \tbOP{bandaare} \tbTP{bandarɛɛ/bandaarɛɛ} \tbVA{baŋa, bɔndaa}
\tbLX{bandaakpogilo} \tbHM{} \tbPH{bàndáákpóglò} \tbPS{n.~} \tbGE{the act of hitting the opponent's armpits with one's knees to weaken the opponent (a wrestling move)} \tbXV{Bayuo to la Dɛre ka Dɛre nyɔge o lɔɔ tɔ bandaakpogilo.} \tbXE{Bayuo insulted Dɛre, and Dɛre threw Bayuo down and used his knees to hit his armpits.} \tbTP{bandaakpogilee} 
\tbLX{bandakarekpele} \tbHM{} \tbPH{bàndàkárkpɪ́lɪ́} \tbPS{n.~} \tbGE{Agama agama; also known as red-headed rock agama} \tbOP{bandakarekpelɛ} \tbTP{bandakarekpelɛɛ} \tbVA{karekpele}
\tbLX{bandakarema} \tbHM{} \tbPH{bàndàkàrɪ̀mà} \tbPS{n.~} \tbGE{large male lizard} \tbTP{bandakaremɛɛ} 
\tbLX{bandanyene} \tbHM{} \tbPH{bàndàɲɛɪ́nɪ́} \tbPS{n.~} \tbGE{tooth of a lizard} \tbOP{bandanyemɛ} \tbTP{bandanyenɛɛ} 
\tbLX{bandapoleɛne} \tbHM{} \tbPH{bàndàpʊ́lɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{female lizard} \tbOP{bandapoleɛmɛ} \tbTP{bandapoleɛmɛɛ/bandapoleɛnɛɛ} \tbVA{poleɛne}
\tbLX{bandazeɛraa} \tbHM{} \tbPH{bàndàzɪ̀ɛ́ráá} \tbPS{n.~} \tbGE{kite; small bird of prey (family Accipitridae)} \tbOP{bandazeɛre} \tbTP{bandazeɛrɛɛ} 
\tbLX{bandazoore} \tbHM{} \tbPH{bàndàzʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{tail of a lizard} \tbOP{bandazoɛ} \tbTP{bandazoorɛɛ} 
\tbLX{banfane} \tbHM{} \tbPH{bànfánɪ́} \tbPS{n.~} \tbGE{a bangle that is believed to have medicinal value and may be worn to cure a disease} \tbOP{banfama} \tbTP{banfanɛɛ} 
\tbLX{bangye} \tbHM{} \tbPH{bándzɪ́} \tbPS{n.~} \tbGE{bank} \tbOP{bangyere} \tbTP{bangyerɛɛ} 
\tbLX{bangyeraa} \tbHM{} \tbPH{bándzɪ́ráá} \tbPS{n.~} \tbSN \tbGE{latrine; washroom} \tbXV{Pampana ka fooŋ mɛ yiri a seŋ ka bagyeraa bebe.} \tbXE{Modern houses are supposed to have built-in washrooms.} \tbSN \tbGE{feces} \tbXV{A bie nyɛ la bini meele a sɛŋ.} \tbXE{The child has soiled the mat with feces.} \tbOP{bangyerre} \tbTP{bangyerɛɛ/bangyerrɛɛ} 
\tbLX{ba-nimie-anaare} \tbHM{} \tbPH{bà-nímíè-ànàárɪ̀} \tbPS{n.~} \tbGE{a dog, usually with big spots above the eyes that look like a second pair of eyes, believed to have supernatural powers} \tbOP{ba-nimie-anaare} \tbTP{ba-nimie-anaarɛɛ} 
\tbLX{bantakperee} \tbHM{} \tbPH{bàntákpérèé} \tbPS{n.~} \tbGE{a type of spotted lizard (believed to be poisonous)} \tbOP{bantakperri} \tbTP{bantakperree} 
\tbLX{bantoɔ} \tbHM{} \tbPH{bántʊ́ɔ́} \tbPS{n.~} \tbGE{rectal bulb syringe} \tbOP{bantoore} \tbTP{bantoorɛɛ} \tbVA{bɔntoɔ}
\tbLX{banweebaa} \tbHM{} \tbPH{bànwɪ́ɪ́báá} \tbPS{n.~} \tbGE{carnivorous bird (bigger than a kite but strong and fast)} \tbOP{banweere} \tbTP{banweerɛɛ} 
\tbLX{banzeɛ} \tbHM{} \tbPH{bànzɪ̀ɛ̀} \tbPS{n.~} \tbGE{copper bangle} \tbOP{banzeere} \tbTP{banzeerɛɛ} 
\tbLX{banzeɛraa} \tbHM{} \tbPH{bǎnzɪ́ɛ́ráá} \tbPS{n.~} \tbGE{bird that eats lizards} \tbOP{banzeɛre} \tbTP{banzeɛrɛɛ} 
\tbLX{banzɔlɔ} \tbHM{} \tbPH{bànzɔ̀lɔ̀} \tbPS{n.~} \tbGE{chain} \tbOP{banzɔlle} \tbTP{banzɔllɛɛ} \tbVA{bɔnzɔlɔ}
\tbLX{banyaa} \tbHM{} \tbPH{bàɲáà} \tbPS{n.~} \tbGE{chest of a dog} \tbOP{banyaɛ/banyaane} \tbTP{banyaanɛɛ} 
\tbLX{banyagere} \tbHM{} \tbPH{bàɲágrɪ́} \tbPS{n.~} \tbGE{intestine of a dog} \tbOP{banyaga} \tbTP{banyagɛɛ/banyagerɛɛ} 
\tbLX{banyaŋaa} \tbHM{} \tbPH{bàɲáŋáá} \tbPS{n.~} \tbGE{female dog} \tbOP{banyaŋene /banyanne} \tbTP{banyaŋenɛɛ/banyannɛɛ} 
\tbLX{Banyene} \tbHM{} \tbPH{bàɲɪ̀nɪ̀} \tbPS{n.~} \tbGE{homeland of the Banyenɛɛ clan} 
\tbLX{Banyenɛɛ} \tbHM{} \tbPH{bàɲɪ́nɛ́ɛ́} \tbPS{n.~} \tbGE{name of a clan} \tbSG{banyenoo} 
\tbLX{banyɛtɔgle} \tbHM{} \tbPH{báɲɛtɔ́glɪ́} \tbPS{n.~} \tbGE{imitation} \tbXV{Bibiiri taa la banyɛtɔgle, azuiŋ neŋkponni seŋ ka ba taa ba menne velaa.} \tbXE{Children like to immitate, so adults should behave well.} 
\tbLX{baŋa} \tbHM{} \tbPH{báŋá} \tbPS{n.~} \tbGE{bangle} \tbOP{banne} \tbTP{bannɛɛ} \tbVA{bɔŋɔ}
\tbLX{baŋa} \tbHM{} \tbPH{bàŋá} \tbPS{n.~} \tbGE{lizard} \tbOP{banne} \tbTP{bannɛɛ} \tbVA{shortened form of bandaa}
\tbLX{baŋe} \tbHM{} \tbPH{bàŋ} \tbPS{v.~} \tbSN \tbGE{to know (something or someone)} \tbXV{N baŋe la a pɔge ŋa serɛ, o yuori la Dabuo.} \tbXE{I know this woman's husband; he is called Dabuo.} \tbSN \tbGE{to be aware of} \tbXV{N baŋe ne la o waabo.} \tbXE{I know he's coming.} \tbPD{baŋɛɛ, banna, banna, banema, bannaa, banema } \tbVA{bɔŋe}
\tbLX{baŋena} \tbHM{} \tbPH{bàŋɪ́ná} \tbPS{n.~} \tbSN \tbGE{area between the shoulder and neck} \tbSN \tbGE{trickster} \tbOP{baŋema/banema} \tbTP{baŋemɛɛ/banemɛɛ} \tbVA{bɔŋenɔ}
\tbLX{baŋenaa} \tbHM{} \tbPH{báŋɪ́náá} \tbPS{n.~} \tbGE{a type of tree that has medicinal properties} \tbOP{baŋene} \tbTP{baŋenɛɛ} \tbVA{bɔŋenaa}
\tbLX{baŋene} \tbHM{} \tbPH{bàŋɪ̀nɪ́} \tbPS{n.~} \tbGE{cleverness} \tbXV{Nannyige maŋ taa la baŋene, ka o nyɔgebo maŋ e tuo.} \tbXE{A thief usually has such cleverness that it is difficult to catch him.} \tbVA{bɔŋene}
\tbLX{baŋgori} \tbHM{} \tbPH{báŋgórí} \tbPS{n.~} \tbGE{high-heeled shoes} \tbSG{baŋgoruu} \tbOP{baŋgorri} \tbTP{baŋgoree/baŋgorree.} 
\tbLX{baŋkaroŋ} \tbHM{} \tbPH{bàŋkàrʊ́ŋ} \tbPS{n.~} \tbGE{a type of brown bird with long, thin legs} \tbOP{baŋkarre} \tbTP{baŋkarɛɛ/baŋkarrɛɛ} 
\tbLX{baŋkoroŋ} \tbHM{} \tbPH{bàŋkʊ́rʊ́ŋ} \tbPS{n.~} \tbGE{foreknowledge} 
\tbLX{baŋŋmeɛle} \tbHM{} \tbPH{bàŋŋmɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{twisted bangle, often with a spiritual or medicinal significance} \tbOP{ baŋŋmeɛlɛ} \tbTP{baŋŋmeɛlɛɛ} 
\tbLX{ba-ɔge} \tbHM{} \tbPH{bà-ɔ̀gɪ́} \tbPS{n.~} \tbGE{a dog's nest, usually for a female that has recently given birth} \tbOP{ba-ɔgre} \tbTP{ba-ɔgɛɛ/ba-ɔgrɛɛ} 
\tbLX{bapare} \tbHM{} \tbPH{bápárɪ́} \tbPS{n.~} \tbGE{close friend} \tbOP{baparemine} \tbTP{baparɛɛ} \tbVA{batuo}
\tbLX{bapare} \tbHM{} \tbPH{bàpárɪ́} \tbPS{n.~} \tbGE{end of a valley} \tbOP{baparɛɛ} 
\tbLX{bapiiraa} \tbHM{} \tbPH{bàpííráá} \tbPS{n.~} \tbGE{a puppy earmarked for someone} \tbOP{bapiiri} \tbTP{bapiiree} 
\tbLX{bapimbaare} \tbHM{} \tbPH{bápímbáárɪ̀} \tbPS{n.~} \tbGE{a small part of the stomach of a ruminant that has many compartments} \tbOP{bapimbaare} \tbTP{bapimbaarɛɛ} 
\tbLX{bara} \tbHM{} \tbPH{bárá} \tbPS{n.~} \tbGE{boastful threat, often considered inappropriate or rude} \tbXV{Dakoraa naŋ baŋ ka dɔɔ zaa kye a yiri poɔ la so ka o wa fare a bara ama taaba.} \tbXE{Knowing that all the men had left the house, Dakoraa went to make boastful threats.} \tbTP{barɛɛ} 
\tbLX{baraa} \tbHM{} \tbPH{báráá} \tbPS{n.~} \tbGE{not making any effort; helpless} 
\tbLX{bare} \tbHM{} \tbPH{bàrɪ̀} \tbPS{v.~} \tbSN \tbGE{to leave, to release, to let go} \tbXV{Ba bare la sakubiiri.} \tbXE{They have let the schoolchildren go for vacation.} \tbSN \tbGE{to discontinue, to abstain, to cease, to stop} \tbXV{A baala bare la a dɔɔta gaabo.} \tbXE{The patient has stayed away from the hospital.} \tbSN \tbGE{to divorce (initiated by husband)} \tbXV{A dɔɔ gaa bare la o pɔge.} \tbXE{The man has divorced his wife.} \tbSN \tbGE{not having power} \tbXV{A dɔɔ ŋa eŋ barɛɛ la te zuo, o koŋ to᷉ɔ a dɛre do.} \tbXE{This man is so useless that he cannot climb the ladder.} \tbPD{barɛɛ, bara, bara, bareba, baraa, bareba } 
\tbLX{bare} \tbHM{} \tbPH{bàrɪ̀} \tbPS{v.~} \tbSN \tbGE{to cultivate seeds} \tbXV{A pɔge bare la kommie a dabɔge poɔ.} \tbXE{The woman has cultivated tomato seeds in the garden.} \tbSN \tbGE{to offer firstfruits to the gods} \tbXV{Bɔgebogereba maŋ bare la bompaala kyɛ baŋ di.} \tbXE{Diviners normally offer firstfruits to their gods before eating.} \tbPD{barɛɛ, bara, bara, bareba, baraa, bareba } 
\tbLX{bare gyɛle} \tbHM{} \tbPH{} \tbPS{v.~} \tbGE{to be careless; to not pay attention} \tbPD{barɛɛ gyɛle, bara gyɛle, gyɛle bara, -, -, - } 
\tbLX{barekanneɛbaa} \tbHM{} \tbPH{bàrɪ̀kánnɪ́ɛ́báá} \tbPS{n.~} \tbGE{local spirits (alcohol)} \tbOP{barekanneɛbaare} \tbTP{barekanneɛbaarɛɛ} 
\tbLX{bare yɛŋe} \tbHM{} \tbPH{bàrɪ̀ yɛ́ŋ} \tbPS{v.~} \tbGE{to feel free, to relax} \tbPD{-, -, -, -, -, - } 
\tbLX{barre} \tbHM{} \tbPH{bárrɪ́} \tbPS{v.~} \tbGE{to haggle} \tbPD{barrɛɛ, barra, barra, bareba, barraa, - } 
\tbLX{ba-sakɔre} \tbHM{} \tbPH{bà-sakɔrɪ́} \tbPS{} \tbGE{lazy and useless dog} 
\tbLX{basasale} \tbHM{} \tbPH{bàsàsàlɪ́} \tbPS{n.~} \tbGE{slippery waterway} \tbTP{basasalɛɛ} 
\tbLX{ba-soɔmee} \tbHM{} \tbPH{bà-sʊ́ɔ́mɪ́ɪ́} \tbPS{n.~} \tbGE{weak and sickly dog} \tbVA{ba-sakɔre}
\tbLX{bataabo} \tbHM{} \tbPH{bátáábʊ́} \tbPS{n.~} \tbGE{the state of not having anything} 
\tbLX{batara} \tbHM{} \tbPH{bátárá} \tbPS{n.~} \tbGE{poor or needy person} \tbOP{batareba} \tbTP{batarebɛɛ} 
\tbLX{batuo} \tbHM{} \tbPH{bátúó} \tbPS{n.~} \tbGE{bosom friend} \tbOP{batuuri} \tbTP{batuuree} \tbVA{bakore bapare, bameŋɛ, babiri}
\tbLX{bavilime} \tbHM{} \tbPH{bàvílímé} \tbPS{n.~} \tbGE{lake} \tbOP{bavilimeri} \tbTP{bavilimee} 
\tbLX{bawaa} \tbHM{} \tbPH{báwáá} \tbPS{n.~} \tbGE{a type of dance} \tbOP{bawaare} \tbTP{bawaarɛɛ} 
\tbLX{Bayaayiree} \tbHM{} \tbPH{báyààyíréé} \tbPS{n.~} \tbGE{people of Bayaayiri} \tbSG{bayaayiruu} 
\tbLX{Bayaayiri} \tbHM{} \tbPH{báyààyírí} \tbPS{n.~} \tbSN \tbGE{a settlement in the Upper West Region} \tbSN \tbGE{ancestors' dwelling; next world} 
\tbLX{Bayuo} \tbHM{} \tbPH{báyúó} \tbPS{n.~} \tbGE{name traditionally given to a male child born on a funeral day} \tbTP{bayumine} 
\tbLX{baze᷉e} \tbHM{} \tbPH{bàzɪ́᷉ɪ̀᷉} \tbPS{n.~} \tbGE{blood of a dog} \tbTP{bazeenɛɛ} 
\tbLX{bazele} \tbHM{} \tbPH{bàzɪ̀lɪ́} \tbPS{n.~} \tbGE{tongue of a dog} \tbOP{bazelɛ} \tbTP{bazelɛɛ} 
\tbLX{bazoɔ} \tbHM{} \tbPH{bàzʊ́ɔ̀} \tbPS{n.~} \tbGE{a type of wild fox} \tbTP{bazoorɛɛ} \tbVA{bazoa, bagyoɔ}
\tbLX{bazoraa} \tbHM{} \tbPH{bàzóráà} \tbPS{n.~} \tbGE{fast dog} \tbOP{bazorri} \tbTP{bazorree} 
\tbLX{bazugiri} \tbHM{} \tbPH{bázúgrí} \tbPS{n.~} \tbGE{a type of wild mouse that has thick fur} \tbOP{bazugo} \tbTP{bazugee/bazugiree} 
\tbLX{be} \tbHM{} \tbPH{bě} \tbPS{v.~} \tbGE{to exist, to be there} \tbXV{A baala be la asibiti poɔ.} \tbXE{The patient is in the hospital.} \tbPD{bee, bere, bere, beribe, beraa, beribe } \tbSE{be nimiri} \tbGE{to be awake, to be alert} 
\tbLX{be} \tbHM{} \tbPH{bě} \tbPS{loc.~} \tbGE{there (place)} \tbXV{N ba nyɛ fo be.} \tbXE{I did not see you there.} 
\tbLX{be} \tbHM{} \tbPH{bɪ̀} \tbPS{v.~} \tbSN \tbGE{to become boiled or cooked} \tbXV{A nɛne ba be, o varema.} \tbXE{The meat is not well boiled, it is hard.} \tbSN \tbGE{to be smooth (flour)} \tbXV{A mansinni ɛre la a bɛnzɔŋ ka a be.} \tbXE{The machine has ground the bean flour very smooth.} \tbSN \tbGE{to mature, to ripen (fruit, millet)} \tbXV{Ka taŋaaŋ wa be o maŋ pɔre le la.} \tbXE{A ripe shea fruit normally falls off the tree.} \tbSN \tbGE{to ferment (alcohol)} \tbXV{A da᷉a naŋ ba be azuiŋ a ba bugiro.} \tbXE{The pito is not yet fermented, so it is not alcholic.} \tbSN \tbGE{to fortify (spiritually)} \tbXV{Pɔlebilii yaga maŋ boɔrɔ ka ba tegere ba menne a be.} \tbXE{Today, most young men want to fortify themselves through spritual powers.} \tbPD{beɛ,  berɛ, beerɛ, berebɛ, beraa, berebɛ } 
\tbLX{be} \tbHM{} \tbPH{bɪ́} \tbPS{conj.~} \tbGE{or} \tbXV{Fo na gaa la be?} \tbXE{Will you go or not?.} \tbVA{shortened form of bee}
\tbLX{bebe} \tbHM{} \tbPH{bèbé} \tbPS{adj.~} \tbGE{present, available} 
\tbLX{bebelaa} \tbHM{} \tbPH{bɪ́bɪ́láá} \tbPS{n.~} \tbGE{termite hill} \tbOP{bebelle} \tbTP{bebellɛɛ} \tbVA{belaa}
\tbLX{bebelebiri} \tbHM{} \tbPH{bɪ́bɪ̀lɪ̀bírì} \tbPS{n.~} \tbGE{white ant} \tbSG{bebelebiruu} \tbOP{bebelebie} \tbTP{bebelebiiree} 
\tbLX{beberaa} \tbHM{} \tbPH{bébéráà} \tbPS{n.~} \tbGE{a persimist} \tbOP{beberri} \tbVA{beberee}
\tbLX{beberee} \tbHM{} \tbPH{bébéréè} \tbPS{n.~} \tbGE{one who is malicious} \tbOP{beberri} \tbTP{beberree} \tbVA{bebberaa}
\tbLX{beberiyɛlɛ} \tbHM{} \tbPH{bébèrìyɛ́lɛ́} \tbPS{n.~} \tbGE{malicious talk} \tbSG{beberiyeli} 
\tbLX{beberuŋ} \tbHM{} \tbPH{bébérúŋ} \tbPS{n.~} \tbGE{the act of causing trouble} 
\tbLX{bebibaaraa} \tbHM{} \tbPH{bèbíbááráá} \tbPS{n.~} \tbGE{last day} \tbOP{bebibaare} \tbTP{bebibaarɛɛ} 
\tbLX{bebibalaa} \tbHM{} \tbPH{bèbíbàlàà} \tbPS{n.~} \tbGE{particular day} 
\tbLX{bebidɛŋe} \tbHM{} \tbPH{bèbídɛ́ŋ} \tbPS{n.~} \tbGE{first day} \tbOP{bebidɛnne} \tbTP{bebidɛnnɛɛ} \tbVA{bebidɛndɛŋ}
\tbLX{bebifaa} \tbHM{} \tbPH{bèbífáá} \tbPS{n.~} \tbGE{sabbath; sacred day} \tbOP{bebifaare} \tbTP{bebifaarɛɛ} 
\tbLX{bebigbuli} \tbHM{} \tbPH{bèbígbùlì} \tbPS{n.~} \tbGE{whole day} \tbOP{bebigbulo} \tbTP{bebigbulee} 
\tbLX{bebikore} \tbHM{} \tbPH{bèbíkʊ́rɪ́} \tbPS{n.~} \tbGE{a former day, a day in the past} \tbOP{bebikorɔ/bebikora} \tbTP{bebikorɛɛ} 
\tbLX{bebipaalaa} \tbHM{} \tbPH{bèbípááláá} \tbPS{n.~} \tbGE{new day; different day} \tbOP{bebipaala} \tbTP{bebipaalɛɛ} 
\tbLX{bebiri} \tbHM{} \tbPH{bèbírì} \tbPS{n.~} \tbGE{date, day} \tbOP{bebie} \tbTP{bebiiri/bebiiree} 
\tbLX{bebisaana} \tbHM{} \tbPH{bèbísááná} \tbPS{n.~} \tbGE{strange day} \tbOP{bebisaama} 
\tbLX{bebisoŋ} \tbHM{} \tbPH{bèbísʊ́ŋ} \tbPS{n.~} \tbGE{good day, holy day} \tbOP{bebisonne} \tbTP{bebisonnɛɛ} 
\tbLX{bebitɔɔre} \tbHM{} \tbPH{bèbítɔ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{distant date} \tbOP{bebitɔbɔ} \tbTP{bebitɔɔrɛɛ} 
\tbLX{bebivuo} \tbHM{} \tbPH{bèbívúó} \tbPS{n.~} \tbGE{free day} \tbOP{bebivori} \tbTP{bebivoree} 
\tbLX{bebiyeni} \tbHM{} \tbPH{bèbíyéní} \tbPS{n.~} \tbGE{one day, a single day} \tbOP{bebiyaga} \tbVA{bebiyenaa}
\tbLX{bee} \tbHM{} \tbPH{bèé} \tbPS{v.~} \tbGE{to exist; to be present} \tbXV{O bee be.} \tbXE{S/he is there.} \tbPD{-, -, -, -, -, - } \tbVA{be; shortened form of be la}
\tbLX{bee} \tbHM{} \tbPH{bɪ́ɪ́} \tbPS{conj.~} \tbGE{or} \tbXV{O gaaɛ la bee o ba gaa?} \tbXE{Has s/he gone or not?.} \tbVA{beɛ, baa}
\tbLX{be᷉e} \tbHM{} \tbPH{bɪ́᷉ɪ́᷉} \tbPS{n.~} \tbSN \tbGE{concentrated extraction from something (tasty, interesting)} \tbSN \tbGE{broth from meat} \tbTP{beenɛɛ} 
\tbLX{beele} \tbHM{} \tbPH{bɪ́ɪ́lɪ̀} \tbPS{v.~} \tbGE{to envy (with evil intent)} \tbPD{beelɛɛ, beelɛ, beelɛ, beelebɛ, beelaa, beelebɛ } 
\tbLX{beele} \tbHM{} \tbPH{bɪ́ɪ́lɪ̀} \tbPS{v.~} \tbGE{to move slightly away from a spot} \tbPD{beelɛɛ, beelɛ, beelɛ, beelebɛ, beelaa, beelebɛ } 
\tbLX{beene} \tbHM{} \tbPH{bɪ́ɪ́nɪ́} \tbPS{n.~} \tbSN \tbGE{strong person} \tbSN \tbGE{one with supernatural power} \tbOP{beemɛ} \tbTP{beemɛɛ} 
\tbLX{beene} \tbHM{} \tbPH{bɪ́ɪ́nɪ́} \tbPS{adj.~} \tbSN \tbGE{well cooked, boiled} \tbSN \tbGE{ripe (fruit)} \tbOP{beemɛ} \tbTP{beenɛɛ} 
\tbLX{beene} \tbHM{} \tbPH{bɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbSN \tbGE{to move majestically} \tbSN \tbGE{to roam round leisurely} \tbPD{beenɛɛ, beenɛ, beenɛ, beenemɛ, beenaa, - } 
\tbLX{beere} \tbHM{} \tbPH{bɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to kill, to finish off, to beat mercilessly} \tbSN \tbGE{to boil pito in the final stage of production} \tbPD{beerɛɛ, beerɛ, beerɛ, beerebɛ, beeraa, - } 
\tbLX{beɛ} \tbHM{} \tbPH{bɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to move out of the way (allowing another to pass)} \tbSN \tbGE{to divert} \tbSN \tbGE{to pass someone by, to overtake} \tbPD{beɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, - } 
\tbLX{beɛ} \tbHM{} \tbPH{bɪ̀ɛ̀} \tbPS{v.~} \tbGE{to slice (meat)} \tbXV{A nɛŋŋmaara beɛ la a nɛne eŋ a doge poɔ.} \tbXE{The butcher sliced the meat and put it into the pot.} \tbPD{beɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, - } \tbVA{beɛle}
\tbLX{beɛ} \tbHM{} \tbPH{bíɛ́} \tbPS{adj.~} \tbGE{funny; strange} \tbOP{beere} \tbTP{beerɛɛ} \tbVA{beɛbeɛ}
\tbLX{be᷉ɛ} \tbHM{} \tbPH{bɪ́᷉ɛ̀᷉} \tbPS{n.~} \tbGE{beer} \tbSG{beɛnoo} \tbTP{beɛnɛɛ} 
\tbLX{beɛbeɛ} \tbHM{} \tbPH{bɪ́ɛ́bɪ́ɛ́} \tbPS{adj.~} \tbGE{strange} \tbVA{beɛ}
\tbLX{beɛlaa} \tbHM{} \tbPH{bɪ́ɛ́láá} \tbPS{adj.~} \tbGE{small, little (quantity, time)} \tbOP{beɛlɛ} 
\tbLX{beɛle} \tbHM{} \tbPH{bɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{to accompany} \tbPD{beɛlɛɛ, beɛlɛ, beɛlɛ, beɛlebɛ, beɛlaa, beɛlebɛ } \tbVA{beɛre}
\tbLX{beɛle} \tbHM{} \tbPH{bɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{to cut in strips  (meat, hide)} \tbPD{beɛlɛɛ, beɛlɛ, beɛlɛ, beɛlebɛ, beɛlaa, beɛlebɛ } \tbVA{beɛre}
\tbLX{beɛlɛbeɛlɛ} \tbHM{} \tbPH{bɪ́ɛ́lɛ́bɪ́ɛ́lɛ́} \tbPS{adv.~} \tbGE{slowly; bit by bit} 
\tbLX{be᷉ɛo} \tbHM{} \tbPH{bɪ̀᷉ɛ̀᷉ʊ́᷉} \tbPS{n.~} \tbGE{multitude, crowd} \tbOP{beɛnɛɛ} 
\tbLX{beɛre} \tbHM{} \tbPH{bɪ́ɛ́rɪ̀} \tbPS{n.~} \tbGE{older brother} \tbOP{beɛremine} \tbTP{beɛrɛɛ} 
\tbLX{beɛre} \tbHM{} \tbPH{bɪ́ɛ́rɪ́} \tbPS{v.~} \tbGE{to follow up} \tbPD{beɛrɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ } 
\tbLX{beɛre} \tbHM{} \tbPH{bɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to cut into strips} \tbPD{beɛrɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebeɛraa } \tbVA{beɛle}
\tbLX{bege} \tbHM{} \tbPH{bɪ̀gɪ̀} \tbPS{v.~} \tbGE{to stop up, to close a passage} \tbPD{begɛɛ, begerɛ, begeraa, begerebɛ, begeraa, begerebɛ } 
\tbLX{bege} \tbHM{} \tbPH{bɪ̀gɪ́} \tbPS{n.~} \tbGE{law} \tbOP{begɛ} \tbTP{begɛɛ} \tbVA{merɛ (Akan)}
\tbLX{begebinne} \tbHM{} \tbPH{bɪ̀gɪ̀bínné} \tbPS{n.~} \tbSN \tbGE{lawmaker} \tbSN \tbGE{member of parliament} \tbOP{begebinime} \tbTP{begebinimee} 
\tbLX{begebiŋi-duoraa} \tbHM{} \tbPH{bɪ̀gɪ̀bíŋ-duoraa} \tbPS{n.~} \tbGE{parliament house} \tbOP{begebiŋi-duori} \tbTP{begebiŋi-duoree} 
\tbLX{begere} \tbHM{} \tbPH{bɪ̀gɪ̀rɪ́} \tbPS{n.~} \tbGE{wing} \tbOP{begɛ} \tbTP{begerɛɛ} 
\tbLX{belaa} \tbHM{} \tbPH{bɪ́láá} \tbPS{n.~} \tbGE{anthill} \tbOP{belle} \tbTP{bellɛɛ} \tbVA{shortened form of bebelaa}
\tbLX{belaa} \tbHM{} \tbPH{bìláà} \tbPS{adv.~} \tbGE{interog. what} \tbXV{Belaa ka fo erɛ?} \tbXE{What are you doing?.} 
\tbLX{belaa} \tbHM{} \tbPH{bɪ́láá} \tbPS{adv.~} \tbSN \tbGE{small (quantity), few} \tbXV{A laa poɔ bondirii waa la belaa lɛ.} \tbXE{The amount of food is very small.} \tbXV{Zenɛ a tontonema waa la belaa.} \tbXE{Today the workers are few.} \tbSN \tbGE{short while (time)} \tbXV{Vɛŋ ka belaa ka fo gaa.} \tbXE{Wait for a short while before you go.} 
\tbLX{bele} \tbHM{} \tbPH{bɪ́lɪ́} \tbPS{v.~} \tbGE{to do extraneous work} \tbPD{belɛɛ, belɛ, belɛ, belbɛ, belaa, - } 
\tbLX{bele} \tbHM{} \tbPH{bɪ́lɪ́} \tbPS{n.~} \tbGE{yeast} \tbOP{belɛ} \tbTP{belɛɛ} 
\tbLX{belefele} \tbHM{} \tbPH{bɪ̀lɪ̀fɪ̀lɪ̀} \tbPS{adv.~} \tbGE{big and fat (person)} 
\tbLX{benee} \tbHM{} \tbPH{bènéé} \tbPS{adv.~} \tbGE{that place} 
\tbLX{benne} \tbHM{} \tbPH{bɪ̀nnɪ̀} \tbPS{v.~} \tbGE{to intrude} \tbPD{bennɛɛ, bennɛ, bennɛ, benemɛ, bennaa, benemɛ } 
\tbLX{bennɛ} \tbHM{} \tbPH{bɪ́nnɛ́} \tbPS{n.~} \tbGE{intruder (especially to meals)} \tbOP{benemɛ} \tbTP{benemɛɛ} 
\tbLX{benni} \tbHM{} \tbPH{benni} \tbPS{v.~} \tbGE{to doubt} \tbPD{bennee, benne, benne, benime, bennaa, - } 
\tbLX{bentere} \tbHM{} \tbPH{bɪ̀ntɪ́rɪ̀} \tbPS{n.~} \tbGE{a large gourd drum} \tbOP{benteɛ} \tbTP{benterɛɛ} \tbVA{bontere}
\tbLX{beo} \tbHM{} \tbPH{bɪ̀ʊ̀} \tbPS{adv.~} \tbGE{big and stagnant} 
\tbLX{bere} \tbHM{} \tbPH{bɪ́rɪ́} \tbPS{n.~} \tbSN \tbGE{a type of small plant whose leaves and flowers are used as vegetables} \tbSN \tbGE{the leaves of this plant} \tbTP{berrɛɛ} \tbVA{berenɛmɛ}
\tbLX{bere} \tbHM{} \tbPH{bɪ̀rɪ́} \tbPS{n.~} \tbSN \tbGE{breast} \tbOP{berɛ} \tbTP{berɛɛ} 
\tbLX{bere} \tbHM{} \tbPH{bɪ̀rɪ̀} \tbPS{v.~} \tbGE{protect by supernatural means} \tbPD{berɛɛ, berɛ, berɛ, berebɛ, beraa, beremɛ } 
\tbLX{bere} \tbHM{} \tbPH{bɪ̀rɪ̀} \tbPS{adj.~} \tbGE{powerful (supernaturally)} 
\tbLX{berebaaraa} \tbHM{} \tbPH{bɪ̀rɪ̀bááráá} \tbPS{n.~} \tbGE{last born (of humans and other mammals)} \tbOP{berebaare} 
\tbLX{berekaroo} \tbHM{} \tbPH{bɪ̀rìkàrʊ́ʊ́} \tbPS{n.~} \tbSN \tbGE{a type of disease of the breast} \tbSN \tbGE{a type of charm believed to inflict this disease} \tbOP{berekarre} \tbTP{berekarɛɛ} 
\tbLX{berekuuri} \tbHM{} \tbPH{bìrɪ̀kúúrí} \tbPS{n.~} \tbGE{thick centre of the breast} \tbSG{berekuuree} \tbOP{berekubo} 
\tbLX{berekpulo} \tbHM{} \tbPH{bɪ̀rɪ̀kpúló} \tbPS{n.~} \tbGE{hard, firm breasts (of a young woman)} \tbTP{berkpulee} 
\tbLX{beremɛ} \tbHM{} \tbPH{bɪ́rɪ́mɛ́} \tbPS{n.~} \tbGE{one who possesses supernatural powers} \tbTP{beremɛɛ} 
\tbLX{berenɛmɛ} \tbHM{} \tbPH{bɪ̀rɪ̀nɛ́mɛ́} \tbPS{n.~} \tbGE{the flower of the bere plant} \tbSG{berenɛnoo} \tbTP{berenɛmɛɛ} \tbVA{bere}
\tbLX{berenoɔ} \tbHM{} \tbPH{bɪ̀rɪ̀nʊ́ɔ̀} \tbPS{n.~} \tbGE{a bird used in rituals to induce breast milk in a nursing mother} \tbOP{berenoore} \tbTP{berenoorɛɛ} 
\tbLX{berenoɔre} \tbHM{} \tbPH{bɪ̀rɪ̀nʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{nipple (of a breast)} \tbOP{berenɔɛ} \tbTP{berenɔrɛɛ} 
\tbLX{berese} \tbHM{} \tbPH{bɪ̀rɪ̀sɪ̊̀} \tbPS{adv.~} \tbGE{extremely large and fat} 
\tbLX{berewoɔ} \tbHM{} \tbPH{bɪ̀rɪ̀wʊ́ɔ́} \tbPS{n.~} \tbGE{slang term for female breast} \tbOP{berewoore} \tbTP{berewoorɛɛ} 
\tbLX{bereyɛre} \tbHM{} \tbPH{bɪ̀rɪ̀yɛ́rɪ́} \tbPS{n.~} \tbGE{a type of wild plant that is used as a vegetable} \tbTP{bereyɛrɛɛ} 
\tbLX{berɛ} \tbHM{} \tbPH{bìrɛ́} \tbPS{v.~} \tbSN \tbGE{to wish} \tbXV{O berɛ ŋa zie da koŋ nyaa.} \tbXE{S/he wished day would never come.} \tbXV{O berɛ ona la da na gaa a tigiri.} \tbXE{S/he wished s/he would be the one to attend the feast.} \tbSN \tbGE{Before he realized it, a stranger arrived} \tbPD{-, -, -, -, -, - } 
\tbLX{berondɔndɔre} \tbHM{} \tbPH{bɪ̀rʊ́ndɔ̀ndɔ̀rɪ́} \tbPS{n.~} \tbGE{curdled milk} \tbTP{berondɔndɔrɛɛ} 
\tbLX{beroŋ} \tbHM{} \tbPH{bɪ̀rʊ́ŋ} \tbPS{n.~} \tbGE{milk} 
\tbLX{beroŋka᷈a} \tbHM{} \tbPH{bɪ́rʊ́ŋká᷈à᷈} \tbPS{n.~} \tbGE{fat from fresh milk; butter} \tbSG{beroŋkaamo} \tbTP{beroŋkaanɛɛ} 
\tbLX{berroŋ} \tbHM{} \tbPH{bɪ̀rrʊ́ŋ} \tbPS{n.~} \tbGE{the state of being supernaturally powerful} 
\tbLX{bɛ} \tbHM{} \tbPH{bɛ̀} \tbPS{v.~} \tbGE{to be sick, to be ill} \tbXV{Ka saaŋ poɔ a bie o na bɛ la.} \tbXE{If the child is out in the downpour he will get sick.} \tbPD{bɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ } 
\tbLX{bɛ} \tbHM{} \tbPH{bɛ̀} \tbPS{v.~} \tbGE{to seal, to close a hole with a muddy substance} \tbXV{Ka fooŋ ba bɛ a peɛ voɔre ka saaŋ wa miire a die na kyire la.} \tbXE{If you do not patch the hole in the roof, the room will leak when there is rain.} \tbPD{bɛɛ, beɛrɛ, beɛrɛ, beɛrebɛ, beɛraa, beɛrebɛ } 
\tbLX{bɛ} \tbHM{} \tbPH{bɛ́} \tbPS{interrogative.~} \tbGE{where} \tbXV{A libie bɛ?} \tbXE{Where is the money?.} \tbVA{bɛɛ}
\tbLX{bɛbɛgere} \tbHM{} \tbPH{bɛ̀bɛ̌grɪ̀} \tbPS{n.~} \tbGE{mud} \tbOP{bɛbɛgɛ} \tbTP{bɛbɛgerɛɛ} 
\tbLX{bɛɛ} \tbHM{} \tbPH{bɛ́ɛ̀} \tbPS{adv.~} \tbGE{where (emphatic)} \tbXV{A libie bɛɛ?} \tbXE{Where the hell is the money?.} \tbVA{bɛ}
\tbLX{bɛɛ} \tbHM{} \tbPH{bɛ̀ɛ́ɛ̀} \tbPS{n.~} \tbGE{sheep} 
\tbLX{bɛɛne} \tbHM{} \tbPH{bɛ́ɛ́nɪ́} \tbPS{n.~} \tbGE{hemp seeds} 
\tbLX{bɛge} \tbHM{} \tbPH{bɛ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to embank} \tbSN \tbGE{to block with a dam} \tbPD{bɛgɛɛ, bɛgerɛ, bɛgerɛ, bɛgerebɛ, bɛgeraa, bɛgerebɛ } 
\tbLX{bɛgelaa} \tbHM{} \tbPH{bɛ̀gɪ̀làá} \tbPS{n.~} \tbGE{a shell of a particular type of seed, used to scrape inside of cooking utensils} \tbOP{bɛgele} \tbTP{bɛgelɛɛ} 
\tbLX{bɛlɛmbɛgere} \tbHM{} \tbPH{bɛ̀lɛ̀mbɛ́grɪ́} \tbPS{n.~} \tbGE{nightengale} \tbOP{bɛlɛmbɛgɛ} \tbTP{bɛlɛmbɛgerɛɛ} \tbVA{bɛlembɛgre}
\tbLX{bɛlɛnte} \tbHM{} \tbPH{bɛ́lɛ́ntɪ̀} \tbPS{n.~} \tbGE{belt} \tbOP{bɛlɛntere} \tbTP{bɛlɛnteerɛɛ} 
\tbLX{bɛlle} \tbHM{} \tbPH{bɛ̀llɪ̀} \tbPS{v.~} \tbGE{to deceive} \tbPD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa, - } 
\tbLX{bɛlle} \tbHM{} \tbPH{bɛ̀llɪ̀} \tbPS{v.~} \tbGE{to coax (a baby)} \tbPD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa, - } 
\tbLX{bɛlle} \tbHM{} \tbPH{bɛ́llɪ̀} \tbPS{v.~} \tbGE{to look at closely, to inspect} \tbPD{bɛllɛɛ, bɛllɛ, bɛllɛ, bɛlbɛ, bɛllaa, - } 
\tbLX{bɛllebɛllɛ} \tbHM{} \tbPH{bɛ̀lbɛllɛ} \tbPS{n.~} \tbGE{bean flour shaped into slabs and steamed} \tbTP{bɛllebɛllɛɛ} \tbVA{tumpanne}
\tbLX{bɛmbɛŋ} \tbHM{} \tbPH{bɛ́mbɛ́ŋ} \tbPS{adv.~} \tbGE{extremely black} \tbXV{A peroo e la sɔgelaa bɛmbɛŋ.} \tbXE{The sheep is extremely black.} 
\tbLX{bɛmbɛŋ} \tbHM{} \tbPH{bɛ̀mbɛ̌ŋ} \tbPS{v.~} \tbGE{to move in a shaky manner} \tbPD{bɛmbɛŋɛɛ, bɛmbɛnɛ, bɛmbɛnɛ, bɛmbɛnemɛ, bɛmbɛnaa, - } 
\tbLX{bɛmbiri} \tbHM{} \tbPH{bɛ́mbírì} \tbPS{n.~} \tbGE{bean seed} \tbOP{bɛmbie} \tbTP{bɛmbiiree} 
\tbLX{bɛmbugo} \tbHM{} \tbPH{bɛ́mbùgó} \tbPS{n.~} \tbGE{barn for storing beans} \tbOP{bɛmbugri} \tbTP{bɛmbugree} 
\tbLX{bɛmbuli} \tbHM{} \tbPH{bɛ́mbúlí} \tbPS{n.~} \tbGE{bean sprout} \tbOP{bɛmbulo} \tbTP{bɛmbullee} 
\tbLX{bɛmpaala} \tbHM{} \tbPH{bɛ́mpáálà} \tbPS{n.~} \tbGE{new, fresh beans} \tbTP{bɛmpaalɛɛ} 
\tbLX{bɛmpelaa} \tbHM{} \tbPH{bɛ́mpɪ́láá} \tbPS{n.~} \tbGE{white beans} \tbOP{bɛmpeɛle} \tbTP{bɛmpeɛlɛɛ} 
\tbLX{bɛmpɛɛre} \tbHM{} \tbPH{bɛ́mpɛ̀ɛ̀rɪ́} \tbPS{n.~} \tbGE{unripened bean seed} \tbOP{bɛmpɛbɛ} \tbTP{bɛmpɛbɛɛ} 
\tbLX{bɛmpɛge} \tbHM{} \tbPH{bɛ́mpɛ́gɪ́} \tbPS{n.~} \tbGE{bean husk} \tbOP{bɛmpɛgre/bɛmpɛgɛ} \tbTP{bɛmpɛgrɛɛ} 
\tbLX{bɛndaa} \tbHM{} \tbPH{bɛ́ndàà} \tbPS{n.~} \tbGE{bean stalk} \tbOP{bɛndaare} \tbTP{bɛndarɛɛ/bɛndaarɛɛ} 
\tbLX{bɛndaa} \tbHM{} \tbPH{bɛ́ndáá} \tbPS{n.~} \tbGE{a cooked bean dish (usually not seasoned)} \tbOP{bɛndaare} \tbTP{bɛndarɛɛ/bɛndaarɛɛ} 
\tbLX{bɛnwɔne} \tbHM{} \tbPH{bɛ́nwɔ́nɪ́} \tbPS{n.~} \tbGE{the fruit of a bean plant} \tbOP{bɛnwɔmɔ} \tbTP{bɛnwɔnnɛɛ} 
\tbLX{bɛnzɔŋe} \tbHM{} \tbPH{bɛ́nzɔ́ŋ} \tbPS{n.~} \tbGE{bean flour} \tbOP{bɛnzɔnne} \tbTP{bɛnzɔnnɛɛ} 
\tbLX{bɛnzɔɔre} \tbHM{} \tbPH{bɛ́nzɔ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{tender (seedless) fruit of a bean plant} \tbOP{bɛnzɔbɔ} \tbTP{bɛnzɔrɛɛ} 
\tbLX{bɛŋe} \tbHM{} \tbPH{bɛ̂ŋ} \tbPS{v.~} \tbSN \tbGE{to sift, to separate grains from unwanted material} \tbSN \tbGE{to skim (usually by tapping water with the back of the hand or a container to push away debris)} \tbSN \tbGE{to move unsteadily, to sway} \tbPD{bɛŋɛɛ, bɛnnɛ, bɛnnɛ, bɛnemɛ, bɛnnaa, bɛnemɛ } \tbVA{pɛŋe}
\tbLX{bɛŋe} \tbHM{} \tbPH{bɛ̀ŋ} \tbPS{v.~} \tbGE{to pretend to invite someone to share something (but not whole-heartedly)} \tbPD{bɛŋɛɛ, bɛnnɛ, bɛnnɛ, bɛnnaa, bɛnemɛ, - } 
\tbLX{bɛŋɛ} \tbHM{} \tbPH{bɛ́ŋɛ̀} \tbPS{n.~} \tbGE{beans} \tbSG{bɛmmo} \tbTP{bɛnnɛɛ} 
\tbLX{bɛŋgbɛre} \tbHM{} \tbPH{bɛ̀ŋgbɛ̀rɪ́} \tbPS{n.~} \tbGE{a variety of beans} \tbOP{bɛŋgbɛrre} \tbTP{bɛŋgbɛrɛɛ} 
\tbLX{bɛŋkorɔ} \tbHM{} \tbPH{bɛ́ŋkʊ́rɔ́} \tbPS{n.~} \tbGE{old beans} \tbTP{bɛŋkorɛɛ} \tbVA{bɛŋkora}
\tbLX{bɛŋkuoŋaa} \tbHM{} \tbPH{bɛ́ŋkúóŋáá} \tbPS{n.~} \tbGE{dry bean} \tbOP{bɛŋkuoni} \tbTP{bɛŋkuonee} 
\tbLX{bɛraa} \tbHM{} \tbPH{bɛ́ráá} \tbPS{n.~} \tbSN \tbGE{trap, snare} \tbSN \tbGE{bait (fishhook)} \tbSN \tbGE{poison} \tbOP{bɛrre} \tbTP{bɛrɛɛ} 
\tbLX{bɛraa} \tbHM{} \tbPH{bɛ̀ràá} \tbPS{n.~} \tbGE{razor} \tbOP{bɛrre} \tbTP{bɛrrɛɛ} 
\tbLX{bɛre} \tbHM{} \tbPH{bɛ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to set (a trap)} \tbSN \tbGE{to poison} \tbXV{Dabe᷉ɛ deme maŋ boɔrɔ ka ba bɛre ba dendɔmɔ ko.} \tbXE{Cowards always try to kill their enemies by poisoning them.} \tbPD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ } 
\tbLX{bɛre} \tbHM{} \tbPH{bɛ̀rɪ̀} \tbPS{v.~} \tbGE{to ponder, to wonder} \tbXV{O da di la a bondirii tasoga leɛ zeŋe bɛre.} \tbXE{After eating for some time, she sat back and thought.} \tbPD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ } 
\tbLX{bɛre} \tbHM{} \tbPH{bɛ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to exclude, to separate from the herd} \tbXV{A naakyeenɛ bɛre la naadɔgerɔ na bare.} \tbXE{The cattle herder has separated the cow that had recently calved from the herd.} \tbSN \tbGE{to detach or cut off (a large piece)} \tbXV{A naŋkpaana bɛre la a naabo gbɛre koɔre.} \tbXE{The butcher has cut off and sold the thigh of the cow.} \tbPD{bɛrɛɛ, bɛrɛ, bɛrɛ, bɛrebɛ, bɛraa, bɛrebɛ } 
\tbLX{bɛren-uri} \tbHM{} \tbPH{bɛ̀rɪ̀n-ùrí} \tbPS{n.~} \tbGE{pollen from the kenaf plant (a fibre-producing plant similar to hibiscus cannabinus) which causes itching} \tbTP{bɛren-uree} \tbVA{bɛren-uri}
\tbLX{bɛroŋ} \tbHM{} \tbPH{bɛ́rʊ́ŋ} \tbPS{adj.~} \tbGE{fat, thick} \tbOP{bɛrɛ} \tbTP{bɛrɛɛ/bɛrrɛɛ} 
\tbLX{bibile} \tbHM{} \tbPH{bìbílé} \tbPS{n.~} \tbGE{small child} \tbOP{bibilii} 
\tbLX{bibile-deɛnaa} \tbHM{} \tbPH{bìbílédɪ́ɛ́náá} \tbPS{n.~} \tbGE{toy} \tbOP{bibiledeɛne} \tbTP{bibiledeɛnɛɛ} 
\tbLX{bibilpɔɔ} \tbHM{} \tbPH{bìbìlpɔ̀ɔ́} \tbPS{n.~} \tbGE{young child} \tbOP{bibilpɔɔre} \tbTP{bibilpɔɔrɛɛ} 
\tbLX{bibore} \tbHM{} \tbPH{bíbʊ̀rɪ̀} \tbPS{n.~} \tbGE{the act of playing chess} 
\tbLX{bidɛndɛŋe} \tbHM{} \tbPH{bìdɛ̀ndɛ́ŋ} \tbPS{n.~} \tbGE{firstborn (human)} \tbOP{bidɛndɛnne} \tbTP{bidɛndɛnnɛɛ} \tbVA{bidɛŋ, bidɛnne, bidɛnnɛɛ}
\tbLX{bidɛŋe} \tbHM{} \tbPH{bìdɛ́ŋ} \tbPS{n.~} \tbGE{firstborn (human)} \tbOP{bidɛnne} \tbTP{bidɛnnɛɛ} \tbVA{shortened form of bidɛndɛŋ}
\tbLX{bidɔŋe} \tbHM{} \tbPH{bìdɔ̂ŋ} \tbPS{n.~} \tbGE{sibling (clan level)} \tbOP{bidɔnne} \tbTP{bidɔnnɛɛ} 
\tbLX{bidɔɔ} \tbHM{} \tbPH{bìdɔ́ɔ́} \tbPS{n.~} \tbGE{son} \tbOP{bidɔbɔ} \tbTP{bidɔɔrɛɛ} 
\tbLX{bie} \tbHM{} \tbPH{bíé} \tbPS{n.~} \tbGE{child} \tbOP{biiri} \tbTP{biiree} 
\tbLX{bie} \tbHM{} \tbPH{bíé} \tbPS{n.~} \tbGE{day, date} \tbXV{Fo sori na yiibu bie la dabuo?} \tbXE{What date have you set for your journey?.} \tbOP{biiri} \tbTP{biiree} 
\tbLX{bie} \tbHM{} \tbPH{bíè} \tbPS{n.~} \tbGE{a type of game played using marbles} \tbOP{biree/biiree} 
\tbLX{bie} \tbHM{} \tbPH{bíé} \tbPS{adv.~} \tbGE{tomorrow} \tbVA{bieo}
\tbLX{bie} \tbHM{} \tbPH{bìè} \tbPS{v.~} \tbGE{to come to pass} \tbXV{A pɔlbile na de la pɔge, a bebiri la naŋ ba bie.} \tbXE{The young man will marry, it just has not happened yet.} \tbPD{-, biere, biere, bieribe, bieraa, bieribe } 
\tbLX{biebie} \tbHM{} \tbPH{bíébíé} \tbPS{adv.~} \tbGE{daily; every day} 
\tbLX{bieo} \tbHM{} \tbPH{bíèó} \tbPS{adv.~} \tbGE{tomorrow} \tbXV{Bieo, ba na wa la.} \tbXE{Tomorrow, they will come.} \tbVA{bie}
\tbLX{bieri} \tbHM{} \tbPH{bíérí} \tbPS{n.~} \tbGE{stinginess, unwillingness to share or open up} 
\tbLX{bifaa} \tbHM{} \tbPH{bìfáá} \tbPS{n.~} \tbGE{bad child} \tbOP{bifaare} \tbTP{bifaarɛɛ} 
\tbLX{bifaa} \tbHM{} \tbPH{bífáá} \tbPS{n.~} \tbGE{bad pebble} \tbOP{bifaare} \tbTP{bifaarɛɛ} 
\tbLX{bife᷉e} \tbHM{} \tbPH{bífɪ́᷉ɪ́᷉} \tbPS{n.~} \tbGE{small quantity of pebbles} \tbOP{bifeene} \tbTP{bifeenɛɛ} 
\tbLX{bifo᷉ɔo} \tbHM{} \tbPH{bífʊ́᷉ɔ́᷉ʊ́᷉} \tbPS{n.~} \tbGE{light pebble} \tbOP{bifoɔne} \tbTP{bifoɔnɛɛ} 
\tbLX{bigaalaa} \tbHM{} \tbPH{bìgáálàá} \tbPS{n.~} \tbGE{any bedding for a child} \tbOP{bigaale} 
\tbLX{bigaale} \tbHM{} \tbPH{bìgààlɪ̀} \tbPS{n.~} \tbGE{postnatal care} \tbOP{bigaale ̪[bìgáálɪ́]} 
\tbLX{bigane} \tbHM{} \tbPH{bìgánɪ̀} \tbPS{n.~} \tbGE{leather spread for a baby cot} \tbOP{bigama} \tbTP{bigamɛɛ} 
\tbLX{bigeŋ} \tbHM{} \tbPH{bìgèŋ} \tbPS{adv.~} \tbGE{all over (area)} 
\tbLX{bigeŋ} \tbHM{} \tbPH{bìgèŋ} \tbPS{adv.~} \tbGE{big heavy sound (fall of a big wall, gunshot)} 
\tbLX{bigileŋ} \tbHM{} \tbPH{̪bìgìlèŋ} \tbPS{adj.~} \tbGE{large and heavy} \tbXV{A kuuri waa la bigileŋ lɛ.} \tbXE{The stone is large and heavy.} 
\tbLX{bigiri} \tbHM{} \tbPH{bígírí} \tbPS{n.~} \tbSN \tbGE{fibre} \tbSN \tbGE{kenaf; a plant that produces fibre} \tbOP{bige} \tbTP{bigee} 
\tbLX{bigiri} \tbHM{} \tbPH{bígírí} \tbPS{adj.~} \tbGE{fibre-like} \tbXV{A maŋgo deɛ waa la bigiri.} \tbXE{The mango is fibrous.} 
\tbLX{bigiri} \tbHM{} \tbPH{bìgìrì} \tbPS{v.~} \tbGE{to describe} \tbPD{bigiree, bigire, bigire, bigiribe, bigiraa, bigiribe } 
\tbLX{biiree} \tbHM{} \tbPH{bììréé} \tbPS{n.~} \tbGE{sand} \tbOP{biiri} \tbTP{biiree} 
\tbLX{biiridalempoɔ} \tbHM{} \tbPH{bììrìdàlɪ̀mpʊ̀ɔ̀} \tbPS{n.~} \tbGE{desert} \tbOP{biiridalempoore} \tbTP{biiridalemporɛɛ} \tbVA{biirikpoŋ}
\tbLX{biirikpoŋi} \tbHM{} \tbPH{bíírìkpôŋ} \tbPS{n.~} \tbGE{desert} \tbOP{biirikponni} \tbTP{biirikponnee} \tbVA{biiridalempoɔ}
\tbLX{bikaa} \tbHM{} \tbPH{bìkáà} \tbPS{n.~} \tbGE{the act of caring for children} 
\tbLX{bikaale} \tbHM{} \tbPH{bíkààlɪ̀} \tbPS{n.~} \tbGE{the act of gathering pebbles} 
\tbLX{bikaara} \tbHM{} \tbPH{bìkáárá} \tbPS{n.~} \tbGE{caretaker of a child} \tbOP{bikaareba} \tbTP{bikarebɛɛ} 
\tbLX{biko} \tbHM{} \tbPH{bìkʊ́} \tbPS{n.~} \tbGE{the act of killing children} 
\tbLX{bikoŋkonaa} \tbHM{} \tbPH{bìkóŋkónáá} \tbPS{n.~} \tbGE{a child that cries a lot} \tbOP{bikoŋkonni} \tbTP{bikoŋkonnee} 
\tbLX{bikoɔraa} \tbHM{} \tbPH{bìkʊ́ɔ́ráá} \tbPS{n.~} \tbGE{industrious person} \tbOP{bikoɔre} 
\tbLX{bikore} \tbHM{} \tbPH{bìkʊ́rɪ́} \tbPS{n.~} \tbGE{old chap, child} \tbOP{bikorɔ/bikoreba} \tbTP{bikorɛɛ} 
\tbLX{biku᷉u} \tbHM{1} \tbPH{bìkú᷉ù᷉} \tbPS{n.~} \tbGE{baby cot} \tbOP{bikuuni} \tbTP{bikuunee} 
\tbLX{biku᷉u} \tbHM{2} \tbPH{bìkú᷉ú᷉} \tbPS{n.~} \tbGE{dead child} \tbOP{bikuuni} \tbTP{bikuunee} 
\tbLX{bikpeebɛ} \tbHM{} \tbPH{bìkpɪ́ɪ́bɛ́} \tbPS{n.~} \tbGE{orphan} \tbOP{bikpeere/bikpeemine} \tbTP{bikpeerɛɛ} 
\tbLX{bikyelaa} \tbHM{} \tbPH{bìʧɪ́láà} \tbPS{n.~} \tbGE{smart child} \tbOP{bikelle} \tbTP{bikyelɛɛ} 
\tbLX{bikyɛnaa} \tbHM{} \tbPH{bìʧɛ́náá} \tbPS{n.~} \tbGE{toddler} \tbOP{bikyɛnne/bikyɛnemɛ} \tbTP{bikyɛnemɛɛ} 
\tbLX{bildɛŋe} \tbHM{} \tbPH{bìldɛ́ŋ} \tbPS{n.~} \tbGE{firstborn (nonhuman)} \tbOP{bildɛnne} \tbTP{bildɛnnɛɛ} 
\tbLX{bile} \tbHM{} \tbPH{bile} \tbPS{adj.~} \tbSN \tbGE{small} \tbXV{A kparoo waa la bile; o koŋ kpɛ a bie.} \tbXE{The shirt is small; it won't fit the child.} \tbSN \tbGE{younger} \tbOP{bilii} 
\tbLX{bile} \tbHM{} \tbPH{bílé} \tbPS{n.~} \tbGE{offspring (nonhuman)} \tbOP{bilii} \tbTP{bilee} 
\tbLX{bileɛ} \tbHM{} \tbPH{bìlɪ́ɛ́} \tbPS{n.~} \tbGE{young baby} \tbOP{bileɛre} \tbTP{bileɛrɛɛ} \tbVA{bilɛɛ}
\tbLX{bilembileŋ} \tbHM{} \tbPH{bílémbíléŋ} \tbPS{adj.~} \tbGE{round and smooth} 
\tbLX{bilembileŋ} \tbHM{} \tbPH{bìlèmbíléŋ} \tbPS{n.~} \tbGE{caterpillar track, continous track} \tbOP{bilembilenni} \tbTP{bilembilennee} 
\tbLX{bilembileŋ} \tbHM{} \tbPH{bìlèmbíléŋ} \tbPS{adj.~} \tbGE{large and heavy, with weight distributed unevenly} 
\tbLX{bilentimm} \tbHM{} \tbPH{bìlèntìmm} \tbPS{adj.~} \tbGE{bloated and large} 
\tbLX{bilɛɛ} \tbHM{} \tbPH{bìlìɛ̀} \tbPS{n.~} \tbGE{baby (newly born)} \tbOP{bilɛɛre} \tbTP{bilɛɛrɛɛ} \tbVA{bileɛ}
\tbLX{bilibala} \tbHM{} \tbPH{bìlìbálá} \tbPS{adv.~} \tbGE{grabbing on to a huge object by throwing both arms around it} \tbXV{O le nyɔge la a bombɛroŋ na lɛ bilibala.} \tbXE{S/he firmly grabbed the fat one.} 
\tbLX{billi} \tbHM{} \tbPH{bìllì} \tbPS{v.~} \tbGE{to roll (a heavy object along the ground)} \tbPD{billee, bille, bille, bilibe, billaa, - } 
\tbLX{billi} \tbHM{} \tbPH{bíllí} \tbPS{v.~} \tbGE{to roll between the palms; to mould} \tbPD{billee, bille, bille, bilibe, billaa, - } 
\tbLX{bimaraa} \tbHM{} \tbPH{bìmáráá} \tbPS{n.~} \tbGE{baby sling} \tbOP{bimarre} \tbVA{maraa}
\tbLX{bimbiŋ} \tbHM{} \tbPH{bímbíŋ} \tbPS{adv.~} \tbGE{extremely dark (rain clouds)} \tbXV{A saa koe la bimbiŋ lɛ.} \tbXE{The rain clouds have gotten awfully thick.} 
\tbLX{bimbiŋe} \tbHM{} \tbPH{bìmbìŋé} \tbPS{n.~} \tbGE{pen, poultry shed} \tbOP{bimbinni} \tbTP{bimbinnee} \tbVA{nɔzage}
\tbLX{bimeŋɛ} \tbHM{} \tbPH{bìmɪ́ŋɛ́} \tbPS{n.~} \tbGE{good child} \tbOP{bimenne} \tbTP{bimennɛɛ} 
\tbLX{bimm} \tbHM{} \tbPH{bìmm} \tbPS{adv.~} \tbGE{large and solid} \tbXV{A bugo e la kpoŋ bimm lɛ.} \tbXE{The granary is big and solid.} 
\tbLX{bimm} \tbHM{} \tbPH{} \tbPS{adj.~} \tbGE{solid} \tbXV{A taabo daa waa la bimm lɛ.} \tbXE{The timber log is solid.} 
\tbLX{bimpeɛle} \tbHM{} \tbPH{bìmpɪ́ɛ́lɪ́} \tbPS{n.~} \tbSN \tbGE{white faeces} \tbSN \tbGE{fear (used with nyɛ...)} \tbXV{Ba toŋ yeli ka n pɔge ma beɛrɛ la ka o deɛ yɔ nyeɛrɛ bimpeɛle.} \tbXE{A message was sent to inform my wife that her mother was sick and she is in utter distress and fear.} \tbTP{bimpeɛlɛɛ} 
\tbLX{bimporɔ} \tbHM{} \tbPH{bìmpʊ́rɔ́} \tbPS{n.~} \tbGE{flatulence} \tbOP{bimporre} \tbTP{bimporɛɛ/bimporrɛɛ} \tbVA{bimpora}
\tbLX{bin-eɛraa} \tbHM{} \tbPH{bìn-ɪ́ɛ́ráá} \tbPS{n.~} \tbGE{cloth used for cleaning faeces (of a baby)} \tbOP{bin-eɛre} \tbTP{bin-eɛrɛɛ} 
\tbLX{bini} \tbHM{} \tbPH{bíní} \tbPS{n.~} \tbGE{excrement} \tbOP{n/a} \tbTP{binee} 
\tbLX{binkyoo} \tbHM{} \tbPH{bìnʧóó} \tbPS{n.~} \tbGE{open place full of defecation} \tbOP{binkyoori} \tbTP{binkyooree} 
\tbLX{binnyeɛbo} \tbHM{} \tbPH{bìnɲɛɪ́ɛ́bʊ́} \tbPS{n.~} \tbGE{act of misbehavior} 
\tbLX{binɔnaa} \tbHM{} \tbPH{bìnɔ̀náá} \tbPS{n.~} \tbGE{beloved child; favorite} \tbOP{binɔnne} 
\tbLX{binɔŋe} \tbHM{} \tbPH{bìnɔ̀ŋ} \tbPS{n.~} \tbGE{the act of loving children; the act of pampering} \tbOP{binɔnema} 
\tbLX{bintanyeɛrɛ} \tbHM{} \tbPH{bìntàɲɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{a type of diarrhoea} \tbOP{bintanyeɛre} \tbTP{bintanyeɛrɛɛ} 
\tbLX{binze᷉e} \tbHM{} \tbPH{bìnzɪ́᷉ɪ́} \tbPS{n.~} \tbGE{dysentery} \tbTP{binzeenɛɛ} 
\tbLX{biŋi} \tbHM{} \tbPH{bìŋ} \tbPS{v.~} \tbSN \tbGE{to put down, to place, to set out} \tbSN \tbGE{to store (up)} \tbPD{biŋee, binne, binne, binime, binnaa, binime } 
\tbLX{biŋibiŋi} \tbHM{} \tbPH{bìŋìbíŋí} \tbPS{v.~} \tbGE{to shift unsteadily, teeter} \tbXV{A tanzuŋ waa la bile, ka a doge biŋibiŋine.} \tbXE{The pot holder is small and so the pot is moving unsteadily.} \tbPD{biŋibiŋee, biŋibiŋine, biŋibiŋine, biŋibiŋinime, biŋibiŋinaa, biŋibiŋinime } 
\tbLX{biŋko᷉ɔ} \tbHM{} \tbPH{bǐŋkʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{watery faeces, diarrhoea} \tbTP{biŋkoɔnɛɛ} 
\tbLX{biŋkɔnnyeɛrɛ} \tbHM{} \tbPH{bìŋkɔ́nɲɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{one suffering from diarrhoea} \tbOP{biŋkɔnnyeɛrebɛ} \tbTP{biŋkɔnnyeɛrebɛɛ} 
\tbLX{bipeɛ} \tbHM{} \tbPH{bìpɪ́ɛ́} \tbPS{n.~} \tbGE{baby cot} \tbOP{bipɛre} \tbTP{bipɛrɛɛ} \tbVA{biku᷉u}
\tbLX{bipoɔnaa} \tbHM{} \tbPH{bìpʊ́ɔ́náá} \tbPS{n.~} \tbGE{pampered child} \tbOP{bipoɔne} \tbTP{bipoɔnɛɛ} 
\tbLX{bipɔlbile} \tbHM{} \tbPH{bìpɔ́lbílé} \tbPS{n.~} \tbGE{a lad; young man} 
\tbLX{bipɔlee} \tbHM{} \tbPH{bìpɔ́lɪ́ɪ́} \tbPS{n.~} \tbGE{young man, lad} \tbOP{bipɔlɔ} \tbTP{bipɔlɛɛ/bipɔllɛɛ} \tbVA{bipɔlee, bipɔlbile}
\tbLX{bipɔllee} \tbHM{} \tbPH{bìpɔ́lléé} \tbPS{n.~} \tbGE{young man} \tbOP{bipɔlɔ} \tbTP{bipɔlɛɛ} \tbVA{bipɔlee, bipɔlbile}
\tbLX{bipuulee} \tbHM{} \tbPH{bìpúúléé} \tbPS{n.~} \tbGE{young woman} \tbOP{bipuuli} 
\tbLX{biraa} \tbHM{} \tbPH{bíráá} \tbPS{n.~} \tbGE{one who stammers} \tbOP{biribe} \tbTP{biribee} 
\tbLX{biri} \tbHM{} \tbPH{bírí} \tbPS{v.~} \tbGE{cause debris to drop (down from a height)} \tbXV{Gyaŋgbonni biri la tɛne a dasɛrɛɛ poɔ eŋe a ko᷉ɔ poɔ.} \tbXE{The mice broke debris from the ceiling into the water.} \tbPD{biree, bire, bire, biribe, biraa, biribe } 
\tbLX{biri} \tbHM{} \tbPH{bírì} \tbPS{v.~} \tbGE{to stutter, to stammer} \tbPD{biree, bire, bire, biribe, biraa, - } 
\tbLX{biri} \tbHM{} \tbPH{bìrì} \tbPS{v.~} \tbGE{to pass by narrowly, to slip past} \tbXV{O deɛ biri la a taŋa kyɛ le.} \tbXE{S/he just passed the hill a little and fell.} \tbPD{biree, bire, bire, biribe, biraa, biribe } 
\tbLX{biri} \tbHM{1} \tbPH{bírì} \tbPS{n.~} \tbGE{chaff from malt} \tbOP{biree} 
\tbLX{biri} \tbHM{2} \tbPH{bírì} \tbPS{n.~} \tbGE{seed, nut} \tbOP{bie} \tbTP{biiree} 
\tbLX{biri} \tbHM{3} \tbPH{bírì} \tbPS{n.~} \tbGE{something of worth} \tbXV{Te zenɛ toma ŋa eŋ yi la biri.} \tbXE{Our work today has been fruitful.} 
\tbLX{biribiri} \tbHM{} \tbPH{bìrìbìrì} \tbPS{adv.~} \tbGE{rushing in large numbers e.g. in a stampede} 
\tbLX{biribiri} \tbHM{} \tbPH{bíríbírí} \tbPS{adv.~} \tbSN \tbGE{completely, totally} \tbXV{O zagerɛɛ biribiri ka o koŋ gaa.} \tbXE{He has completely refused to go.} \tbSN \tbGE{completely black} 
\tbLX{biribiri} \tbHM{} \tbPH{bìrìbírí} \tbPS{v.~} \tbGE{to slip, to not fit in (joint)} \tbXV{A sori na taa la kusiwara te zuo ka fo maŋ kyɛnɛ ka fo gbɛɛ bibire.} \tbXE{That road has too much gravel, so your feet slip around as you walk on it.} \tbPD{biribiree, biribire, biribire, biribiribe, biribiraa, - } 
\tbLX{birikyi} \tbHM{} \tbPH{bìríʧì} \tbPS{n.~} \tbSN \tbGE{brake} \tbSN \tbGE{brick} \tbOP{birikyiri} \tbTP{birikyiree} \tbVA{birisi}
\tbLX{birikyilaa} \tbHM{} \tbPH{bìríʧìláá} \tbPS{n.~} \tbGE{mould (for bricks)} \tbOP{birikyilaa} \tbTP{birikyilaarɛɛ} 
\tbLX{birikyisaabo} \tbHM{} \tbPH{bìríʧìsáábʊ́} \tbPS{n.~} \tbGE{brake pad} \tbTP{birikyisaarɛɛ} 
\tbLX{birisi} \tbHM{} \tbPH{bìrísì} \tbPS{n.~} \tbGE{brick} \tbSG{birisiruu} \tbOP{birisiri} \tbTP{birisiree} \tbVA{birikyi}
\tbLX{biriyokpekperaa} \tbHM{} \tbPH{bìríyǒkpɪ́kpìráà} \tbPS{n.~} \tbGE{wild duck-like water bird} \tbOP{biriyokpekperre} \tbTP{biriyokpekperɛɛ} 
\tbLX{biriyonaameɛo} \tbHM{} \tbPH{bíríyònààmɪ̀ɛ́ʊ} \tbPS{n.~} \tbGE{very large duck} \tbOP{biriyonaameɛne} \tbTP{biriyonaameɛnɛɛ} 
\tbLX{biriyuo} \tbHM{} \tbPH{bìríyúó} \tbPS{n.~} \tbGE{duck} \tbSG{biriyoruu} \tbOP{biriyori/biriyoe} \tbTP{biriyoree} \tbVA{biriyori}
\tbLX{bisaa} \tbHM{} \tbPH{bìsàá} \tbPS{n.~} \tbGE{father (human)} \tbOP{bisaamine} \tbTP{bisaarɛɛ} 
\tbLX{bisoŋ} \tbHM{} \tbPH{bìsʊ́ŋ} \tbPS{n.~} \tbGE{good child, well-behaved child} \tbOP{bisonne} \tbTP{bisonnɛɛ} 
\tbLX{bitoroo} \tbHM{} \tbPH{bìtòróò} \tbPS{n.~} \tbGE{virgin} \tbOP{bitorri} \tbTP{bitoree} 
\tbLX{bitɔgelɔ} \tbHM{} \tbPH{bìtɔ́glɔ̀} \tbPS{n.~} \tbGE{midwife} \tbOP{bitɔgelebɔ/bitɔgeleba} \tbTP{bitɔgelebɛɛ} \tbVA{bitɔgela}
\tbLX{bituuro} \tbHM{} \tbPH{bìtúúró} \tbPS{n.~} \tbGE{baby adopter, nanny} \tbOP{bituuribo} \tbTP{bituuribee} 
\tbLX{biweɛmɛ} \tbHM{} \tbPH{bìwɪ́ɛ́mɛ́} \tbPS{n.~} \tbGE{strayed or wayward child} \tbOP{biweɛne} \tbTP{biweɛnɛɛ} 
\tbLX{biweɛraa} \tbHM{} \tbPH{bìwɪ́ɛ́ráá} \tbPS{n.~} \tbGE{spoilt child} \tbOP{biweɛre} 
\tbLX{biwolaa} \tbHM{} \tbPH{bìwʊ́láá} \tbPS{n.~} \tbSN \tbGE{problem child} \tbSN \tbGE{beloved child} \tbOP{biwolle} \tbTP{biwolɛɛ} 
\tbLX{biyaalaa} \tbHM{} \tbPH{bìyááláá} \tbPS{n.~} \tbGE{babysitter} \tbOP{biyaalba} \tbTP{biyaalbɛɛ} 
\tbLX{biyaale} \tbHM{} \tbPH{bìyáálɪ́} \tbPS{n.~} \tbGE{the act of babysitting} 
\tbLX{biyaaleyieluŋ} \tbHM{} \tbPH{bìyáálɪ̀yíélúŋ} \tbPS{n.~} \tbGE{lullaby} \tbOP{biyaaleyiele} \tbVA{biyagleyieluŋ}
\tbLX{biyoɔraa} \tbHM{} \tbPH{bìyʊ́ɔ́ráá} \tbPS{n.~} \tbGE{well travelled person; one who roams a lot} \tbOP{biyoɔre} \tbTP{biyoɔrɛɛ} 
\tbLX{biyopore} \tbHM{} \tbPH{bìyópʊ̀rɪ̀} \tbPS{} \tbGE{naming ceremony (baby)} 
\tbLX{biyuori} \tbHM{} \tbPH{bìyúórí} \tbPS{n.~} \tbGE{popular person} \tbOP{biyoe} \tbTP{biyoree} 
\tbLX{bizɛgɛ} \tbHM{} \tbPH{bìzɛ̀gɛ́} \tbPS{n.~} \tbGE{young baby} \tbOP{bizɛgre} \tbTP{bizɛgɛɛ} 
\tbLX{bizɔnzɔɔ} \tbHM{} \tbPH{bìzɔ̀nzɔ́ɔ́} \tbPS{n.~} \tbGE{young baby} \tbOP{bizɔnzɔɔbɔ} \tbTP{bizɔnzɔɔrɛɛ} 
\tbLX{bo} \tbHM{} \tbPH{bʋ́} \tbPS{v.~} \tbGE{to gather momentum (rain seen coming from afar in full scale)} \tbPD{boɛ, boorɔ, boorɔ, booreba, booraa, booreba } 
\tbLX{bo} \tbHM{} \tbPH{bʊ̀} \tbPS{v.~} \tbSN \tbGE{to confer, to deliberate, to plan together} \tbXV{Ba na bo la sɛre kyɛ gaa a kɔɔto.} \tbXE{They will confer before they go to court.} \tbSN \tbGE{to suppose} \tbXV{Fo bo ka boŋ ka a saana na taa wa ne?} \tbXE{What do you suppose the stranger will bring along?.} \tbPD{boɛ, boorɔ, boorɔ, booreba, booraa, booreba } 
\tbLX{bobɛllɛ} \tbHM{} \tbPH{bʊ̀bɛ́llɛ́} \tbPS{n.~} \tbGE{a type of carnivorous bird} \tbOP{bobɛlle} 
\tbLX{boboo} \tbHM{} \tbPH{bʊ̀bʊ́ʊ́} \tbPS{adj.~} \tbGE{extremely voluminous} 
\tbLX{boboruu} \tbHM{} \tbPH{bóbórúú} \tbPS{n.~} \tbGE{a type of termite} \tbOP{boborri/boboree} \tbTP{boborree} 
\tbLX{bobɔge} \tbHM{} \tbPH{bʊ̀bɔ̀gɪ́} \tbPS{n.~} \tbGE{huge carnivorous bird} \tbOP{bobɔgre} \tbTP{bobɔgrɛɛ} 
\tbLX{bodaa} \tbHM{} \tbPH{bʊ̀dáá} \tbPS{n.~} \tbGE{male goat} \tbOP{bodaare} \tbTP{bodaarɛɛ/bodaɛ} 
\tbLX{bodaa-ansiinee} \tbHM{} \tbPH{bʊ̀dáá-ánsíínéé} \tbPS{n.~} \tbGE{a type of tree with edible fruit} \tbOP{bodaa-ansiini} \tbVA{boraa-ansiinee}
\tbLX{bodaanonnoonaa} \tbHM{} \tbPH{bʊ̀dáánʊ́nnʊ́ʊ́náá} \tbPS{n.~} \tbGE{a type of tree with edible fruit} \tbOP{bodaanonnoone} \tbTP{bodaanonnoonɛɛ} \tbVA{bononnoonaa}
\tbLX{bodaasensɔɔre} \tbHM{} \tbPH{bʊ̀dáásɪ̌nsɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{wild tigernut (not edible)} \tbOP{bodaasensɔbɔ} \tbTP{bodaasensɔɔrɛɛ} 
\tbLX{bodalane} \tbHM{} \tbPH{bʊ́dàlánɪ́} \tbPS{n.~} \tbGE{male goat (uncastrated)} \tbOP{bodalama} \tbTP{bodalamɛɛ} 
\tbLX{bodalee} \tbHM{} \tbPH{bʊ̀dàléé} \tbPS{n.~} \tbGE{ young male goat} \tbOP{bodabilii .} \tbTP{bodabilee} 
\tbLX{bodateɛne} \tbHM{} \tbPH{bʊ̀dàtɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{male goat with long hair on the chin; bearded male goat} 
\tbLX{bodavaraa} \tbHM{} \tbPH{bʊ̀dàváràà} \tbPS{n.~} \tbGE{castrated male goat} \tbOP{bodavarre} \tbTP{bodavarɛɛ/bodavarrɛɛ} 
\tbLX{bogere} \tbHM{} \tbPH{bʊ́grɪ̀} \tbPS{v.~} \tbGE{to break into little bits or pulp} \tbPD{bogerɛɛ, bogerɔ, bogerɔ, bogerebɔ, bogeraa, bogerebɔ } 
\tbLX{bogi} \tbHM{} \tbPH{bògí} \tbPS{n.~} \tbSN \tbGE{hole} \tbSN \tbGE{grave, tomb} \tbSN \tbGE{pit} \tbSN \tbGE{ditch} \tbSN \tbGE{limit} \tbOP{bogiri} \tbTP{bogree} 
\tbLX{bogigbɛre-katakye} \tbHM{} \tbPH{bògìgbɛre-kátákyè} \tbPS{n.~} \tbGE{subway, underground train} \tbOP{bogigbɛre-katakyeri} \tbTP{bogigbɛre-katakyeree} 
\tbLX{bogikantoo} \tbHM{} \tbPH{bògíkántóó} \tbPS{n.~} \tbGE{a type of rodent (not edible)} \tbOP{bogikantori} \tbTP{bogikantoree} 
\tbLX{bogisulee} \tbHM{} \tbPH{bògsùlèè} \tbPS{n.~} \tbGE{round, deep tomb (like a well)} \tbOP{bogisulli} \tbTP{bogisullee} 
\tbLX{bogiti} \tbHM{} \tbPH{bógítì} \tbPS{n.~} \tbGE{bucket} \tbOP{bogitiri} \tbTP{bogitiree} 
\tbLX{bogi-u᷉} \tbHM{} \tbPH{bògì-ù᷉ù᷉} \tbPS{n.~} \tbGE{the act of burying a human corpse} 
\tbLX{bogi-u᷉u-faare} \tbHM{} \tbPH{bògì-ù᷉ù᷉-fáárɪ́} \tbPS{n.~} \tbGE{xylophone music signaling the completion of the burial of the dead} \tbTP{bogi-u᷉u-faarɛɛ} 
\tbLX{bogi-u᷉u᷉-libie} \tbHM{} \tbPH{bògì-ù᷉ù᷉-líbíè} \tbPS{n.~} \tbGE{fee for those who bury the dead} \tbTP{bogi-u᷉u᷉-libiiree} 
\tbLX{bogi-u᷉u-naazoɔ} \tbHM{} \tbPH{bògì-ù᷉ù᷉-náázʊ́ɔ́} \tbPS{n.~} \tbGE{chief undertaker (one who buries the dead)} \tbOP{bogi-u᷉u-naazoore} \tbTP{bogi-u᷉u-naazoorɛɛ} 
\tbLX{bogi-uuno} \tbHM{} \tbPH{bògì-ù᷉ù᷉nó} \tbPS{n.~} \tbGE{one who buries the dead} \tbOP{bogi-uunimo} \tbTP{bogi-uunimee} 
\tbLX{bogi-u᷉u-saabo} \tbHM{} \tbPH{bògì-ù᷉ù᷉-sáábʊ́} \tbPS{n.~} \tbGE{ritual TZ for those who bury the dead} \tbTP{bogi-u᷉u-saarɛɛ} 
\tbLX{bogi-u᷉u-sawɔlɔ} \tbHM{} \tbPH{bògì-ù᷉ù᷉-sáwɔ́lɔ́} \tbPS{n.~} \tbGE{ritual dish for those who bury the dead} \tbTP{bogi-u᷉u-sawɔlɛɛ} 
\tbLX{bogi-u᷉u-zanna} \tbHM{} \tbPH{bògì-ù᷉ù᷉-zànná} \tbPS{n.~} \tbGE{amateur undertaker} \tbOP{bogi-u᷉u-zanema} \tbTP{bogi-u᷉u-zanemɛɛ} 
\tbLX{bogivuo} \tbHM{} \tbPH{bògìvùò} \tbPS{n.~} \tbGE{uncovered tomb} \tbOP{bogivori} \tbTP{bogivoree} 
\tbLX{bogiwɛraa} \tbHM{} \tbPH{bògwɛ́ráà} \tbPS{n.~} \tbGE{oblong tomb} \tbOP{bogiwɛrre} 
\tbLX{bogizeɛ} \tbHM{} \tbPH{bògzɪ̀ɛ̀} \tbPS{n.~} \tbGE{grave, deep pit} \tbOP{bogizeere} \tbTP{bogizeerɛɛ} 
\tbLX{bogɔbogɔ} \tbHM{} \tbPH{bʊ́gɔ́bʊ́gɔ́} \tbPS{adv.~} \tbGE{ground into pulp} \tbVA{bogaboga}
\tbLX{boguolo} \tbHM{} \tbPH{bʊ̀gúóló} \tbPS{n.~} \tbGE{goats given to one to raise in return for some} \tbTP{boguolee} 
\tbLX{bog-uuno} \tbHM{} \tbPH{bòg-úúnó} \tbPS{n.~} \tbGE{one who buries the dead; undertaker} \tbOP{bog-uunimo} \tbTP{bog-uunimee} 
\tbLX{bokolɔ} \tbHM{} \tbPH{bʊ̀kʊ̀lɔ̀} \tbPS{n.~} \tbGE{tasteless powdered stuff, tobacco} \tbOP{bokolle} \tbTP{bokolɛɛ} \tbVA{bokola}
\tbLX{bokontawɛ} \tbHM{} \tbPH{bʊ̀kòntáwɛ́} \tbPS{n.~} \tbGE{a daily marketplace} \tbOP{bokontawɛre} \tbTP{bokontawɛrɛɛ} 
\tbLX{bokoŋ} \tbHM{} \tbPH{bʊ̀kóŋ} \tbPS{n.~} \tbGE{lean or thin goat} \tbOP{bokonni} \tbTP{bokonnee} 
\tbLX{bokore} \tbHM{} \tbPH{bʊ̀kʊ́rɪ́} \tbPS{n.~} \tbGE{old goat} \tbOP{bokora} \tbTP{bokorɛɛ} 
\tbLX{bokpɛ} \tbHM{} \tbPH{bʊ̀kpɛ̀} \tbPS{v.~} \tbGE{to sympathise with} \tbPD{bokpɛɛ, boorɔkpeɛrɛ, bokpeɛrɛ, bokpeɛrebɛ, bokpeɛraa, - } 
\tbLX{bolaakaatoo} \tbHM{} \tbPH{bʊ̀lààkààtʊ̀ʊ̀} \tbPS{adv.~} \tbGE{talking continuously, noisy} 
\tbLX{bolaare} \tbHM{} \tbPH{bʊ̀láárɪ́} \tbPS{v.~} \tbGE{to burst strongly} \tbXV{A lɔɔre kɔbɔ da bolaarɛɛ la.} \tbXE{The lorry tyre blasted with a loud noise.} \tbPD{bolaarɛɛ, bolaara, bolaara, bolaareba, bolaaraa, bolaareba } \tbVA{belaare}
\tbLX{bole} \tbHM{} \tbPH{bʊ́lɪ́} \tbPS{v.~} \tbGE{to smear all over} \tbPD{bolɛɛ, bola, bola, bolba, bolaa, - } 
\tbLX{bolebole} \tbHM{} \tbPH{bʊ̀lbʊ́lɪ́} \tbPS{v.~} \tbGE{to speak unintelligibly, to mumble} \tbPD{bolebolɛɛ, bolebolɔ, bolebolɔ, bolebolebɔ, bolebolaa, bolebolebɔ } 
\tbLX{bolee} \tbHM{} \tbPH{bʊ̀léé} \tbPS{n.~} \tbGE{kid (young goat)} \tbOP{bobili} \tbTP{bobillee} 
\tbLX{bolenaa} \tbHM{} \tbPH{bʊ̀lénáá} \tbPS{n.~} \tbSN \tbGE{tethered goat} \tbSN \tbGE{one restricted from moving} \tbOP{bolenni} \tbTP{bolennee} 
\tbLX{bo-leŋi} \tbHM{} \tbPH{bʊ̀-léŋ} \tbPS{n.~} \tbGE{the act of tethering goats} 
\tbLX{boli} \tbHM{} \tbPH{bólí} \tbPS{v.~} \tbGE{to be foolish} \tbPD{bolee, bolo, bolo, bolibo, bolaa, bolibo } 
\tbLX{boli} \tbHM{} \tbPH{'bòlí} \tbPS{adj.~} \tbGE{pure (honey)} 
\tbLX{bolle} \tbHM{} \tbPH{bʊ̀llɪ} \tbPS{v.~} \tbGE{to murmur; to talk inaudibly} \tbPD{bollɛɛ, bolla, bolla, boleba, bollaa, boleba } 
\tbLX{bolɔ} \tbHM{} \tbPH{bʋ̀lɔ́} \tbPS{n.~} \tbGE{a type of vegetable} \tbVA{bola}
\tbLX{bolɔbolɔ} \tbHM{} \tbPH{bʋ́lɔ́bʋ́lɔ́} \tbPS{adv.~} \tbGE{lukewarm  (water)} \tbVA{bolabola}
\tbLX{boluu} \tbHM{} \tbPH{bòlúú} \tbPS{pron. interj.~} \tbGE{what is it?} 
\tbLX{bombaala} \tbHM{} \tbPH{bʋ́mbààlá} \tbPS{n.~} \tbGE{a sick thing} \tbOP{bombaaleba} \tbTP{bombaalɛɛ} 
\tbLX{bombaalaa} \tbHM{} \tbPH{bʋ́mbààlàà} \tbPS{n.~} \tbGE{a slim or narrow thing} \tbOP{bombaale} 
\tbLX{bombaaraa} \tbHM{} \tbPH{bʋ́mbààráà} \tbPS{n.~} \tbGE{something that can grow} \tbOP{bombaare} 
\tbLX{bombaaraa} \tbHM{} \tbPH{bʋ́mbááráá} \tbPS{n.~} \tbGE{the last one} \tbOP{bombaare} 
\tbLX{bombalaa} \tbHM{} \tbPH{bʋ́mbáláá} \tbPS{n.~} \tbGE{something which is avoided} \tbOP{bomballe} 
\tbLX{bombalaa} \tbHM{} \tbPH{bʊ́mbàlàà} \tbPS{n.~} \tbGE{kind of thing (referring to something huge or strange)} \tbXV{Yeŋ ka fo ta pɔge a bombalaa ŋa?} \tbXE{Where did you get this kind of thing?.} 
\tbLX{bombaltakyelmɛ} \tbHM{} \tbPH{bòmbàltáʧɪ́mɛ́} \tbPS{n.~} \tbGE{things of all sorts} \tbTP{bombaltakyelmɛɛ} 
\tbLX{bombannaa} \tbHM{} \tbPH{bʊ́mbànnáà} \tbPS{n.~} \tbGE{something which is known} \tbOP{bombanne} \tbVA{bombaŋenaaːbombaŋene}
\tbLX{bombaŋe} \tbHM{} \tbPH{bʋ́mbàŋ} \tbPS{n.~} \tbSN \tbGE{the act of knowing} \tbSN \tbGE{knowledge} 
\tbLX{bombare} \tbHM{} \tbPH{bʋ̌mbàrɪ̀} \tbPS{n.~} \tbSN \tbGE{the act of releasing; the act of setting animals free} \tbSN \tbGE{the act of offering firstfruits to gods} \tbOP{bombarre} 
\tbLX{bombeene} \tbHM{} \tbPH{bʋ́mbɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{something that is cooked or ripe} \tbOP{bombeemɛ} \tbTP{bombeenɛɛ} 
\tbLX{bombeɛ} \tbHM{} \tbPH{bʋ́mbɪ́ɛ́} \tbPS{n.~} \tbGE{something strange or mysterious} \tbOP{bombeere} \tbTP{bombeerɛɛ} 
\tbLX{bombeɛraa} \tbHM{} \tbPH{bʋ́mbɪ̀ɛ́ráà} \tbPS{n.~} \tbSN \tbGE{sliced strips (of leather)} \tbSN \tbGE{that which is patched or to be patched (as in car tyre)} \tbOP{bombeɛre} \tbTP{bombɛrɛɛ} 
\tbLX{bombeɛre} \tbHM{} \tbPH{bʋ́mbɪ̀ɛ̀rɪ́} \tbPS{n.~} \tbGE{boils} \tbOP{bombɛbɛ} \tbTP{bombɛrɛɛ} 
\tbLX{bombiiri} \tbHM{} \tbPH{bʊ́mbíírí} \tbPS{n.~} \tbGE{very young ones; unhatched embryos in eggs} \tbTP{bombiiree} 
\tbLX{bombile} \tbHM{} \tbPH{bòmbílé} \tbPS{n.~} \tbGE{ young donkey; foal} \tbOP{bombilii} \tbTP{bombillee} 
\tbLX{bombile} \tbHM{} \tbPH{bʊ́mbílé} \tbPS{n.~} \tbGE{small thing} \tbOP{bombilii} 
\tbLX{bombiri} \tbHM{} \tbPH{bʊ́mbírì} \tbPS{n.~} \tbGE{seed; cereal} \tbSG{bombiruu} \tbOP{bombie} \tbTP{bombiiree} 
\tbLX{bombo} \tbHM{} \tbPH{bǒmbò} \tbPS{n.~} \tbSN \tbGE{a small plant with a jellylike tuber} \tbSN \tbGE{the tuber or flour made from the tuber of this plant} \tbSN \tbGE{fried dish made from the flour of this plant} \tbSG{bomboruu (for plant or tuber)} \tbOP{bombori} \tbTP{bomboree} 
\tbLX{bombolaa} \tbHM{} \tbPH{bʋ́mbʋ́láá} \tbPS{n.~} \tbGE{a thing that is not of the same species; a strange thing} \tbOP{bombolle} \tbTP{bombollɛɛ} 
\tbLX{bombone} \tbHM{} \tbPH{bʋ́mbʋ́nɪ̀} \tbPS{n.~} \tbGE{a useless thing} \tbOP{bomboma} \tbTP{bombonnɛɛ} 
\tbLX{bombonoone} \tbHM{} \tbPH{bǒmbònʋ̀ʋ̀nɪ̀} \tbPS{n.~} \tbGE{a type of tree} \tbOP{bombonooma} \tbTP{ bombonoonɛɛ.} 
\tbLX{bomboŋɔ} \tbHM{} \tbPH{bʋ́mbʋ̀ŋɔ́} \tbPS{n.~} \tbGE{something that cannot be classified} \tbOP{bombonne} \tbTP{bombonnɛɛ} 
\tbLX{bombooree} \tbHM{} \tbPH{bʋ́mbʋ́ʋ́rɪ́} \tbPS{n.~} \tbSN \tbGE{something that is kept for posterity} \tbSN \tbGE{seed for sowing} \tbOP{bomboora} \tbTP{bomboorɛɛ} 
\tbLX{bomboorɔ} \tbHM{} \tbPH{bʊ́mbʊ́ʊ́rɔ́} \tbPS{n.~} \tbGE{seed for planting} \tbSG{bombooroo} \tbTP{bomboorɛɛ} \tbVA{bomboora}
\tbLX{bomboɔraa} \tbHM{} \tbPH{bʋ́mbʋ́ɔ́ráá} \tbPS{n.~} \tbGE{something that is sought ; something that is wanted or liked} \tbOP{bomboɔre} \tbTP{bomboɔrɛɛ.} 
\tbLX{bomboɔre} \tbHM{} \tbPH{bʊ́mbʊ́ɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{things one wishes to have} \tbSN \tbGE{wealth} 
\tbLX{bomboraa} \tbHM{} \tbPH{bʋ́mbʋ̀ráà} \tbPS{n.~} \tbGE{something that is sown; a crop} \tbOP{bomborre} \tbTP{ bomborrɛɛ.} 
\tbLX{bombore} \tbHM{} \tbPH{bʋ́mbʋ̀rɪ̀} \tbPS{n.~} \tbGE{the act of sowing} 
\tbLX{bombo sawɔlɔ} \tbHM{} \tbPH{bòmbò sáwɔ́lɔ́} \tbPS{n.~} \tbGE{fried duff from bombo (flour)} \tbTP{bombo sawɔlɛɛ} \tbVA{bombo sawala}
\tbLX{bombɔ} \tbHM{} \tbPH{bʋ́mbɔ́} \tbPS{n.~} \tbGE{quest for wealth} 
\tbLX{bombɔge} \tbHM{} \tbPH{bʋ́mbɔ́gɪ́} \tbPS{n.~} \tbGE{the act of grinding into fine pulp} \tbOP{bombɔgre} \tbTP{bombɔgrɛɛ} 
\tbLX{bombɔgɔ} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{the forearm or leg of an animal} \tbOP{bombɔgerɔ} \tbTP{bombɔgerɛɛ} \tbVA{bombɔge}
\tbLX{bombugilaa} \tbHM{} \tbPH{bʋ́mbúglàá} \tbPS{n.~} \tbGE{something that is nursed} \tbOP{bombugili} 
\tbLX{bombulaa} \tbHM{} \tbPH{bʋmbulaa} \tbPS{n.~} \tbGE{something that can sprout from a seed} \tbOP{bombulli} 
\tbLX{bombuli} \tbHM{} \tbPH{bʊ́mbúlí} \tbPS{n.~} \tbGE{shoot of a new plant} \tbOP{bombulo} \tbTP{bombullee} 
\tbLX{bombuuraa} \tbHM{} \tbPH{bʋ́mbúúràá} \tbPS{n.~} \tbSN \tbGE{a measure} \tbSN \tbGE{something that is respected} \tbOP{bombuuri} 
\tbLX{bommaalaa} \tbHM{} \tbPH{bʋ́mmààláà} \tbPS{n.~} \tbGE{something that is made} \tbOP{bommaale} 
\tbLX{bommaale} \tbHM{} \tbPH{bʋ́mmààlɪ̀} \tbPS{n.~} \tbGE{the act of making} \tbOP{bommaaleba} 
\tbLX{bommaane} \tbHM{} \tbPH{bʋ́mmàànɪ́} \tbPS{n.~} \tbSN \tbGE{an egg that is immature} \tbSN \tbGE{something that is tiny} \tbOP{bommaama} 
\tbLX{bommaaroŋ} \tbHM{} \tbPH{bʋ́mmáárʋ́ŋ} \tbPS{n.~} \tbGE{a wet, cold, or fresh thing} \tbOP{bommaara} 
\tbLX{bommannaa} \tbHM{} \tbPH{bʋ́mmánnáà} \tbPS{n.~} \tbGE{a measure; something that is measured} 
\tbLX{bompaala} \tbHM{} \tbPH{bʋ́mpáálà} \tbPS{n.~} \tbSN \tbGE{new or fresh things} \tbSN \tbGE{firstfruits of crops} \tbOP{bompaala} \tbTP{bompalɛɛ} 
\tbLX{bompaare} \tbHM{} \tbPH{bʋ́mpáárɪ̀} \tbPS{n.~} \tbGE{the vulva of any animal} \tbOP{bompɛɛrɛɛ} 
\tbLX{bompaare} \tbHM{} \tbPH{bǒmpáárɪ̀} \tbPS{n.~} \tbGE{the vulva of a donkey} \tbOP{bompaɛ} \tbTP{ bompɛɛrɛɛ} 
\tbLX{bompaare} \tbHM{} \tbPH{bǒmpáárɪ̀} \tbPS{n.~} \tbGE{foot rot} \tbOP{bompaɛ} \tbTP{ bompɛɛrɛɛ} 
\tbLX{bompare} \tbHM{} \tbPH{bʋ́mpárɪ̀} \tbPS{n.~} \tbSN \tbGE{residue} \tbSN \tbGE{bottom} \tbOP{bomparɛɛ} 
\tbLX{bompeelaa} \tbHM{} \tbPH{bʋ́mpɪ́ɪ́láá} \tbPS{n.~} \tbGE{a very young animal} \tbOP{bompeele} \tbTP{bompeelɛɛ} 
\tbLX{bompeelaa} \tbHM{} \tbPH{bʋ́mpɪ́ɪ́láá} \tbPS{n.~} \tbGE{a piece torn from a piece of cloth} \tbOP{bompeele} \tbTP{bompeelɛɛ} 
\tbLX{bompege} \tbHM{} \tbPH{bʋ́mpɪ̀gɪ̀} \tbPS{n.~} \tbGE{the act of decorating, improving or adding worth to something} 
\tbLX{bompegelaa} \tbHM{} \tbPH{bʋ́mpɪ́gláà} \tbPS{n.~} \tbGE{something that is moving and must be intercepted} \tbOP{bompegele} \tbTP{bompegelɛɛ} 
\tbLX{bompegeraa} \tbHM{} \tbPH{bʋ́mpɪ́gráà} \tbPS{n.~} \tbGE{something that is adorned} \tbOP{bompegere} \tbTP{bompegerɛɛ} 
\tbLX{bompegɛ} \tbHM{} \tbPH{bʋ́mpɪ́gɛ́} \tbPS{n.~} \tbGE{a piece of a broken container (calabash)} \tbOP{bompegerre} \tbTP{ bompegerɛɛ.} 
\tbLX{bompelaa} \tbHM{} \tbPH{bʋ́mpɪ̀làá} \tbPS{n.~} \tbGE{a white thing} \tbOP{bompeɛle} \tbTP{bompeɛlɛɛ} 
\tbLX{bompere} \tbHM{} \tbPH{bʋ́mpɪ̀rɪ̀} \tbPS{n.~} \tbGE{the act of dehusking (removing the dry outer cover of a grain)} 
\tbLX{bompere} \tbHM{} \tbPH{bʊ́pɪ́rɪ́} \tbPS{n.~} \tbGE{the act of scratching (as by fowl)} 
\tbLX{bompɛɛre} \tbHM{} \tbPH{bʋ́mpɛ̀ɛ̀rɪ́} \tbPS{n.~} \tbGE{an immature thing} \tbOP{bompɛbɛ} 
\tbLX{bompɛge} \tbHM{} \tbPH{bʋ́mpɛ́gɪ̀} \tbPS{n.~} \tbGE{the act of washing the shell or bark of something} 
\tbLX{bompɛgelaa} \tbHM{} \tbPH{bʋ́mpɛ́gláà} \tbPS{n.~} \tbGE{something that is carried by holding it to one's side} \tbOP{bompɛgele} \tbTP{bompɛgelɛɛ} 
\tbLX{bompɛgeraa} \tbHM{} \tbPH{bʊ́pɛ́gráá} \tbPS{n.~} \tbSN \tbGE{something that has been washed} \tbSN \tbGE{something for washing} \tbOP{bompɛgere} \tbTP{bompɛgerɛɛ} 
\tbLX{bompɛlaa} \tbHM{} \tbPH{bʋ́mpɛ́láá} \tbPS{n.~} \tbGE{something that can be peeled} \tbOP{bompɛlle} \tbTP{ bompɛllɛɛ} 
\tbLX{bompɛle} \tbHM{} \tbPH{bʋ́mpɛ́lɪ́} \tbPS{n.~} \tbGE{the act of peeling} 
\tbLX{bompɛle} \tbHM{} \tbPH{bʊḿpɛ̀lì} \tbPS{n.~} \tbGE{peel} \tbOP{bompɛlɛ} \tbTP{bompɛllɛɛ} 
\tbLX{bompɛnnaa} \tbHM{} \tbPH{bʊ́mpɛ̀nnáá} \tbPS{n.~} \tbGE{something that is carved} \tbOP{bompɛnne} \tbTP{bompɛnnɛɛ} 
\tbLX{bompɛnne} \tbHM{} \tbPH{bʋ́mpɛ̀nnɪ̀} \tbPS{n.~} \tbGE{the act of carving} 
\tbLX{bompi} \tbHM{} \tbPH{bʋ́mpí} \tbPS{n.~} \tbSN \tbGE{the process of becoming distended or bloated} \tbSN \tbGE{forerunning for something} 
\tbLX{bompigiraa} \tbHM{} \tbPH{bʋ́mpǐgráà} \tbPS{n.~} \tbGE{something that is to be watched or ought to be watched} \tbOP{bompigiri} 
\tbLX{bompiiraa} \tbHM{} \tbPH{bʋ́mpììráà} \tbPS{n.~} \tbSN \tbGE{a lost thing that is found; an unexpected find} \tbSN \tbGE{an easy find} \tbOP{bompiiri} 
\tbLX{bompiiraa} \tbHM{} \tbPH{bʊ́mpííráá} \tbPS{n.~} \tbGE{a distended thing} \tbOP{bompiiri} 
\tbLX{bompiiraa} \tbHM{} \tbPH{bʊ́mpííráá} \tbPS{n.~} \tbGE{something that has been earmarked for someone} \tbOP{bompiiri} 
\tbLX{bompiire} \tbHM{} \tbPH{bʋ́mpííré} \tbPS{n.~} \tbGE{one who finds a lost item} \tbOP{bompiiribe} \tbTP{bompiiribee} 
\tbLX{bompollaa} \tbHM{} \tbPH{bʊ́mpʊ́lláá} \tbPS{n.~} \tbGE{something on which an oath is taken} \tbOP{bompolle} 
\tbLX{bompoɔ} \tbHM{} \tbPH{bʋ́mpʋ́ɔ́} \tbPS{n.~} \tbGE{a pregnant thing (non human)} \tbOP{bompoore} \tbTP{ bompoorɛɛ.} 
\tbLX{bompoɔraa} \tbHM{} \tbPH{bʋ́mpʋ́ɔ́ráá} \tbPS{n.~} \tbGE{something by which one swears} \tbOP{bompoɔre} \tbTP{ bompoɔrɛɛ} 
\tbLX{bomporaa} \tbHM{} \tbPH{bʋ́mpʋ́ráá} \tbPS{n.~} \tbGE{something that is poured} \tbOP{bomporre} \tbTP{ bomporrɛɛ} 
\tbLX{bompore} \tbHM{} \tbPH{bʋ́mpʋ̀rɪ̀} \tbPS{n.~} \tbGE{plentiful things} \tbOP{bomporre} 
\tbLX{bompore} \tbHM{} \tbPH{bʋ́mpʋ́rɪ́} \tbPS{n.~} \tbGE{something that is small and short; a little thing} \tbOP{bompora} \tbTP{bomporɛɛ} 
\tbLX{bompore} \tbHM{} \tbPH{bʋ́mpʋ́rɪ̀} \tbPS{n.~} \tbGE{the act of pouring} 
\tbLX{bompɔ} \tbHM{} \tbPH{bʋ́mpɔ́} \tbPS{n.~} \tbGE{the act of swearing or taking an oath on something} 
\tbLX{bompɔge} \tbHM{} \tbPH{bʋ́mpɔ́gɪ́} \tbPS{n.~} \tbGE{a female thing} \tbOP{bompɔgeba} \tbTP{bompɔgrɛɛ} 
\tbLX{bompɔge} \tbHM{} \tbPH{bʋ́mpɔ̀gɪ̀} \tbPS{n.~} \tbGE{the act of confining animals in enclosures} 
\tbLX{bompɔgeraa} \tbHM{} \tbPH{bʋ́mpɔ̀gráà} \tbPS{n.~} \tbGE{something that is confined in an enclosure} \tbOP{bompɔgere} 
\tbLX{bompɔle} \tbHM{} \tbPH{bʋ́mpɔ́lɪ̀} \tbPS{n.~} \tbGE{a young thing} \tbOP{bompɔlɔ} \tbTP{ bompɔllɛɛ} 
\tbLX{bompɔre} \tbHM{} \tbPH{bʋ́mpɔ́rɪ́} \tbPS{n.~} \tbGE{the act of plucking something; (fruits or groundnuts)} 
\tbLX{bompɔrɔ} \tbHM{} \tbPH{bʊ́mpɔ́rɔ̀} \tbPS{n.~} \tbGE{harvester} \tbOP{bompɔreba} \tbTP{bompɔrebɛɛ} 
\tbLX{bompugo} \tbHM{} \tbPH{bʋ́mpúgó} \tbPS{n.~} \tbGE{an ash-coloured thing} \tbOP{bompugri} \tbTP{ bompugree} 
\tbLX{bompulonpuloŋ} \tbHM{} \tbPH{bʋ́mpùlònpúlóŋ} \tbPS{n.~} \tbGE{various things mixed up} 
\tbLX{bompuuraa} \tbHM{} \tbPH{bʋ́mpùùráà} \tbPS{n.~} \tbGE{something that has flowered} \tbOP{bompuuri} \tbTP{bompuuree.} 
\tbLX{bompuuraa} \tbHM{} \tbPH{bʋ́mpùùráà} \tbPS{n.~} \tbGE{something that is or can be praised} \tbOP{bompuuri} \tbTP{bompuuree.} 
\tbLX{bompuuri} \tbHM{} \tbPH{bʋ́mpúúrí} \tbPS{n.~} \tbGE{flowers of things} \tbTP{bompuuree} 
\tbLX{bonaare} \tbHM{} \tbPH{bʋ̀náárɪ̀} \tbPS{adj.~} \tbGE{four times} 
\tbLX{bon-agraa} \tbHM{} \tbPH{bʋ́ŋágráà} \tbPS{n.~} \tbSN \tbGE{winged creature} \tbSN \tbGE{ insect} \tbOP{boŋ-agre} 
\tbLX{bonda} \tbHM{} \tbPH{bʋ́ndà} \tbPS{n.~} \tbGE{the act of buying} 
\tbLX{bondaa} \tbHM{} \tbPH{bòndáá} \tbPS{n.~} \tbGE{male donkey} \tbOP{bondaare} \tbTP{bondarɛɛ} 
\tbLX{bondaa} \tbHM{} \tbPH{bʋ́ndáá} \tbPS{n.~} \tbGE{a male thing} \tbOP{bondaare} \tbTP{bondarɛɛ} 
\tbLX{bondaa} \tbHM{} \tbPH{bʋ́ndáà} \tbPS{n.~} \tbGE{the act of pushing} \tbOP{bonda} 
\tbLX{bondaana} \tbHM{} \tbPH{bʋ́ndááná} \tbPS{n.~} \tbGE{rich person; one who has wealth} \tbXV{.} \tbXE{} \tbOP{bondaanema/bondaama} \tbTP{ bondaamɛɛ/bondaanɛɛ} 
\tbLX{bondaaraa} \tbHM{} \tbPH{bʋ́ndààráà} \tbPS{n.~} \tbGE{something that is bought; something that is meant for sale} \tbOP{bondaare} 
\tbLX{bondaga} \tbHM{} \tbPH{bòndàgà} \tbPS{n.~} \tbGE{carefree} \tbOP{bondagre} \tbTP{ bondagrɛɛ.} 
\tbLX{bonde} \tbHM{} \tbPH{bʋ́ndɪ́} \tbPS{n.~} \tbSN \tbGE{the act of taking} \tbSN \tbGE{the act of harvesting} 
\tbLX{bondegere} \tbHM{} \tbPH{bʋ́ndɪ̀grɪ} \tbPS{n.~} \tbGE{a dwarflike thing} \tbOP{bondegɛ} 
\tbLX{bondemannewullaa} \tbHM{} \tbPH{bʊ́ndɪ́mánnɪ́wúlláá} \tbPS{n.~} \tbGE{example} \tbOP{bondemannewulli} 
\tbLX{bondɛgere} \tbHM{} \tbPH{bʋ́ndɛ̌grɪ̀} \tbPS{n.~} \tbGE{a dirty thing} \tbOP{bondɛgɛ} 
\tbLX{bondi} \tbHM{} \tbPH{bʋ́ndì} \tbPS{n.~} \tbGE{the act of eating} 
\tbLX{bondikaama} \tbHM{} \tbPH{bʊ̀ndìkáámá} \tbPS{n.~} \tbGE{delicious and assorted food; luncheon} \tbTP{bondikaanɛɛ} 
\tbLX{bondikaraa} \tbHM{} \tbPH{bʊ́ndìkáráá} \tbPS{n.~} \tbGE{raw food; food that is not well cooked} \tbOP{bondikare} \tbTP{bondikarɛɛ} 
\tbLX{bondi-kpaa} \tbHM{} \tbPH{bondi-kpaa} \tbPS{n.~} \tbGE{leftover food} \tbOP{bondi-kpaare} 
\tbLX{bondimaala} \tbHM{} \tbPH{bʊ́ndímáálá} \tbPS{n.~} \tbGE{cook, one who prepares food} 
\tbLX{bondiraa} \tbHM{} \tbPH{bʊ́ndìráà} \tbPS{n.~} \tbSN \tbGE{food; something that is edible} \tbSN \tbGE{a sharp thing (blade)} \tbOP{bondirri} \tbTP{bondirree} 
\tbLX{bondire} \tbHM{} \tbPH{bʊ́ndíré} \tbPS{n.~} \tbSN \tbGE{one who eats} \tbSN \tbGE{rightful owner} \tbOP{bondiribe} \tbTP{bondiribee} 
\tbLX{bondirii} \tbHM{} \tbPH{bʊ́ndìríì} \tbPS{n.~} \tbGE{food, things to eat} \tbTP{bondirree} 
\tbLX{bondi-sa᷉a᷉} \tbHM{} \tbPH{bʊ́ndì-sà᷉à} \tbPS{n.~} \tbGE{the act of wasting food} 
\tbLX{bondizie} \tbHM{} \tbPH{bʊ́ndìzíé} \tbPS{n.~} \tbGE{restaurant} \tbOP{bondiziiri} \tbTP{bondiziiree} 
\tbLX{bondogelaa} \tbHM{} \tbPH{bʋ́ndʋ́gláá} \tbPS{n.~} \tbGE{something that is used for cooking} \tbOP{bondogele} 
\tbLX{bondogeraa} \tbHM{} \tbPH{bʋ́ndʋ́gráá} \tbPS{n.~} \tbGE{a boiled thing} \tbOP{bondogere} 
\tbLX{bondonaa} \tbHM{} \tbPH{bʊ́ndʊ̀náà} \tbPS{n.~} \tbSN \tbGE{something that bites} \tbSN \tbGE{insect, bug} \tbOP{bondonne} \tbTP{bondonnɛɛ} \tbVA{kyibe}
\tbLX{bondori} \tbHM{} \tbPH{bʋ́ndórí} \tbPS{n.~} \tbGE{ heap} \tbOP{bondoree} \tbTP{bondorree} 
\tbLX{bondɔgelaa} \tbHM{} \tbPH{bʋ́ndɔ́gláá} \tbPS{n.~} \tbGE{an extra thing} \tbOP{bondɔgele} 
\tbLX{bondɔgerɔ} \tbHM{} \tbPH{bʊ́ndɔ̀gerɔ̀} \tbPS{n.~} \tbGE{an animal that has recently littered} \tbOP{bondɔgere} \tbTP{bondɔgerɛɛ} 
\tbLX{bone} \tbHM{} \tbPH{bʊ́nɪ̀} \tbPS{n.~} \tbGE{thing} \tbOP{boma} \tbTP{bonnɛɛ} 
\tbLX{bone} \tbHM{} \tbPH{bʋ́nɪ́} \tbPS{adj.~} \tbGE{worthless or useless} \tbOP{boma} 
\tbLX{bon-ennaa} \tbHM{} \tbPH{bʊ́n-ɪ́nnáà} \tbPS{n.~} \tbGE{container} \tbOP{bon-enne} \tbTP{bon-ennɛɛ} 
\tbLX{bonfa} \tbHM{} \tbPH{́bʋ́nfá} \tbPS{n.~} \tbGE{the act of robbing} 
\tbLX{bonfaa} \tbHM{} \tbPH{bʋ́nfáá} \tbPS{n.~} \tbGE{a bad or evil thing} \tbOP{bonfaare} \tbTP{ bonfaarɛɛ} 
\tbLX{bonfaa} \tbHM{} \tbPH{bʊ́nfáà} \tbPS{n.~} \tbGE{robbery} 
\tbLX{bonfaara} \tbHM{} \tbPH{bʋ́nfáárá} \tbPS{n.~} \tbGE{robber, thief} \tbOP{bonfareba} \tbTP{ bonfabarɛɛ.} 
\tbLX{bonfaaraa} \tbHM{} \tbPH{bʋ́nfááráá} \tbPS{n.~} \tbGE{something that is robbed} \tbOP{bonfaare} \tbTP{ bonfaarɛɛ} 
\tbLX{bonfane} \tbHM{} \tbPH{bʋ́nfànɪ́} \tbPS{n.~} \tbGE{a rough thing} \tbOP{bonfama} \tbTP{ bonfannɛɛ.} 
\tbLX{bonfarafara} \tbHM{} \tbPH{bʋ́nfàràfàrà} \tbPS{n.~} \tbGE{a rough, scaly thing} 
\tbLX{bonfe᷉ê} \tbHM{} \tbPH{bʋ́nfɪ́᷉ɪ́᷉} \tbPS{n.~} \tbGE{a small quantity} \tbOP{bonfeene} \tbTP{bonfeenɛɛ} 
\tbLX{bonfeeloo} \tbHM{} \tbPH{bʋ́fɪ́ɪ́lʋ́ʋ́} \tbPS{n.~} \tbGE{an opening that is small and tight} \tbOP{bonfeelɛ} \tbTP{ bonfeelɛɛ.} 
\tbLX{bonfegɛ} \tbHM{} \tbPH{bʋ́nfɪ̀gɛ́} \tbPS{n.~} \tbGE{narrow space} \tbXV{see [bʊ́nfɔ́glʊ́ʊ]́} \tbXE{} 
\tbLX{bonfogeloo} \tbHM{} \tbPH{bʋ́nfʋ́glʋ́ʋ́} \tbPS{n.~} \tbGE{a narrow thing} \tbOP{bofogela} \tbTP{ bonfogelɛɛ} 
\tbLX{bonfogiluu} \tbHM{} \tbPH{bʋ́nfóglúú} \tbPS{n.~} \tbGE{a seedless pod} \tbOP{bonfogilo} \tbTP{ bonfogilee} 
\tbLX{bonfole} \tbHM{} \tbPH{bʋ́nfʋ̀lɪ́} \tbPS{n.~} \tbGE{an immature thing} \tbOP{bonfola} \tbTP{ bonfolɛɛ.} 
\tbLX{bonfolo} \tbHM{} \tbPH{bʋ́nfòló} \tbPS{n.~} \tbGE{powerful spiritual beings} \tbOP{bonfolo} \tbTP{ bonfolee.} 
\tbLX{bonfo᷉ɔo} \tbHM{} \tbPH{bʋ́nfʋ́᷉ɔ́̀᷉́ʋ́᷉} \tbPS{n.~} \tbGE{a light thing} \tbOP{bonfoɔne} \tbTP{bonfoɔnɛɛ} 
\tbLX{bonfɔraa} \tbHM{} \tbPH{bʊ́nfɔ́ráá} \tbPS{n.~} \tbGE{a narrow thing} \tbOP{bonfɔrre} \tbTP{bonfɔrɛɛ} 
\tbLX{bonfuraa} \tbHM{} \tbPH{bʋ́nfùráà} \tbPS{n.~} \tbGE{something that is used to plug an opening} \tbOP{bonfurri} \tbTP{ bonfurree.} 
\tbLX{bonfuuraa} \tbHM{} \tbPH{bʋ́nfùùráá} \tbPS{n.~} \tbSN \tbGE{something that is cool} \tbSN \tbGE{something that threatens} \tbOP{bonfuuri} \tbTP{bonfuuree.} 
\tbLX{bonii} \tbHM{} \tbPH{bʋ̀níì} \tbPS{adj.~} \tbGE{eight times} 
\tbLX{bonkyaaraa} \tbHM{} \tbPH{bʊ́nʧááráá} \tbPS{n.~} \tbGE{gift} \tbOP{bonkyaare} \tbTP{bonkyaarɛɛ} 
\tbLX{bonkyêɛ᷉} \tbHM{} \tbPH{bʋ́nʧɪ́᷉ɛ́᷉} \tbPS{n.~} \tbGE{a torn thing} \tbOP{bonkyeɛne/bonkyeɛnɛ} \tbTP{ bonkyeɛnɛɛ} 
\tbLX{bonkyeɛraa} \tbHM{} \tbPH{bʋ́nʧɪ́ɛ́ráá} \tbPS{n.~} \tbSN \tbGE{something that is contributed} \tbSN \tbGE{something that is cut} \tbSN \tbGE{something used for cutting} \tbOP{bonkyeɛre} \tbTP{ bonkyeɛrɛɛ} 
\tbLX{bonkyɛlee} \tbHM{} \tbPH{bʋ́nʧɛ́lɪ́ɪ́} \tbPS{n.~} \tbGE{a half thing} \tbOP{bonkyɛlle} 
\tbLX{bonlaane} \tbHM{} \tbPH{bʋ́nláánɪ́} \tbPS{n.~} \tbGE{a worthless thing} \tbOP{bonlaama} \tbTP{ bonlaanɛɛ.} 
\tbLX{bonlane} \tbHM{} \tbPH{bʋ́nlánɪ́} \tbPS{n.~} \tbGE{uncastrated animal} \tbOP{bonlama} \tbTP{ bonlamɛɛ} 
\tbLX{bonlee} \tbHM{} \tbPH{bʊ́nléé} \tbPS{n.~} \tbGE{a small thing e.g. object/non human} \tbVA{bʊ́mbílé}
\tbLX{bonlee} \tbHM{} \tbPH{bònléé} \tbPS{n.~} \tbGE{colt, young donkey} \tbOP{bombilii} 
\tbLX{bonnaakale} \tbHM{} \tbPH{bǒnnààkálɪ́} \tbPS{n.~} \tbGE{the act of rolling over and over. (donkey having a sun bath)} \tbTP{bonnakalɛɛ} 
\tbLX{bonnaŋ} \tbHM{} \tbPH{bʋ́nnáŋ} \tbPS{n.~} \tbGE{scarcity} \tbOP{bonnanne} \tbTP{ bonnannnɛɛ.} 
\tbLX{bonne} \tbHM{} \tbPH{bʋ̀nnɪ̀, bʊ́nnɪ́} \tbPS{v.~} \tbGE{to doubt; to speculate, to guess} \tbPD{bonnɛɛ, bonnɔ, bonnɔ, bonema, bonnaa, bonema } 
\tbLX{bonneɛ} \tbHM{} \tbPH{bʋ́nnɪ́ɛ́} \tbPS{n.~} \tbSN \tbGE{the act of trampling on things} \tbSN \tbGE{the act of ironing clothes} \tbOP{bonneɛne} \tbTP{bonneɛnɛɛ} 
\tbLX{bonni} \tbHM{} \tbPH{bónní} \tbPS{v.~} \tbSN \tbGE{to burn (pain)} \tbSN \tbGE{to burn witth fire} \tbPD{bonnee, bonno, bonno, bonnaa, bonimo, - } 
\tbLX{bonnimiri} \tbHM{} \tbPH{bʋ́nnímírì} \tbPS{n.~} \tbGE{something that is very valuable} \tbOP{bonnimie} \tbTP{ bonimiiree.} 
\tbLX{bonnoɔ} \tbHM{} \tbPH{bʋ́nnʋ̀ɔ̀} \tbPS{n.~} \tbGE{a sweet thing} \tbOP{bonnoore} \tbTP{ bonnoorɛɛ} 
\tbLX{bonnyaa} \tbHM{} \tbPH{bʋ́nɲáà} \tbPS{n.~} \tbSN \tbGE{chest of;} \tbSN \tbGE{most important part of} \tbOP{bonnyaane} \tbTP{bonnyaanɛɛ} 
\tbLX{bonnyaanaa} \tbHM{} \tbPH{bʋ́nɲáànáà} \tbPS{n.~} \tbGE{a clever or wise animal} \tbOP{bonnyaane} 
\tbLX{bonnyaara} \tbHM{} \tbPH{bʋ́nɲàárɪ̀} \tbPS{n.~} \tbGE{coarseness (flour)} \tbTP{bonnyaarɛɛ} 
\tbLX{bonnyagere} \tbHM{} \tbPH{bʊ́nɲágá} \tbPS{n.~} \tbGE{intestines of nonhuman creatures} \tbOP{bonnyaga} \tbTP{bonnyagerɛɛ} 
\tbLX{bonnyaŋaa} \tbHM{} \tbPH{bʋ́nɲáŋáá} \tbPS{n.~} \tbGE{female} \tbOP{bonnyaŋene/bonnyaane} \tbTP{bonnyaanɛɛ} 
\tbLX{bonnyaŋaa} \tbHM{} \tbPH{bònɲàŋáà} \tbPS{n.~} \tbGE{female donkey} \tbOP{boŋnyaŋene} \tbTP{ boŋnyaŋenɛɛ.} 
\tbLX{bonnyuu} \tbHM{} \tbPH{bʋ́nɲúú} \tbPS{n.~} \tbSN \tbGE{something that smells} \tbSN \tbGE{smelling medicinal lotion} \tbOP{bonnyuuri} \tbTP{bonyuuree} 
\tbLX{bonnyuu} \tbHM{} \tbPH{bʊ́nɲúú} \tbPS{n.~} \tbGE{the neck of something} \tbOP{bonnyuuri} \tbTP{bonyuuree} 
\tbLX{bonnyuuraa} \tbHM{} \tbPH{bʋ́nɲúúráá} \tbPS{n.~} \tbSN \tbGE{something that is drunk; a drink} \tbSN \tbGE{something that is used for drinking} \tbOP{bonnyuuri} 
\tbLX{bononnoonaa} \tbHM{} \tbPH{bʊ̀nʊ́nnʊ́ʊ́náá} \tbPS{n.~} \tbGE{a type of tree with juicy fruits} \tbVA{bodaanonnoonaa}
\tbLX{bonoone} \tbHM{} \tbPH{bʋ̀nʋ̀ʋ̀nɪ̀} \tbPS{n.~} \tbGE{a type of tree with fleshy edible fruits; medicinal} \tbOP{bonooma} \tbTP{ bonoonɛɛ} 
\tbLX{bonɔbonɔ} \tbHM{} \tbPH{bʊ̀nɔ̀bʊ̀nɔ̀} \tbPS{adv.~} \tbGE{oily-looking} \tbXV{A zeɛre zu deɛ e la bonɔbonɔ lɛ.} \tbXE{The surface of the soup looks oily.} \tbVA{bonabona}
\tbLX{bonsâa} \tbHM{} \tbPH{bʋ́nsà̂à᷉} \tbPS{n.~} \tbGE{the act of being wasteful, destructive} 
\tbLX{bonsaalaa} \tbHM{} \tbPH{bʋ́nsààlàà} \tbPS{n.~} \tbGE{something that is smooth} \tbOP{bonsaale} \tbTP{ bonsaalɛɛ.} 
\tbLX{bonsaaloŋ} \tbHM{} \tbPH{bʊ́nsààlʊ́ŋ} \tbPS{n.~} \tbGE{a slippery thing} \tbOP{bonsaale} \tbTP{bonsaalɛɛ} 
\tbLX{bonsaana} \tbHM{} \tbPH{bʋ́nsáánà} \tbPS{n.~} \tbGE{a new or strange thing} \tbOP{bonsaama} \tbTP{ bonsaanɛɛ} 
\tbLX{bonsaana} \tbHM{} \tbPH{bʋ́nsááná} \tbPS{n.~} \tbGE{one who destroys} \tbOP{bonsaanema} 
\tbLX{bonsaanaa} \tbHM{} \tbPH{bʋ́nsáánáá} \tbPS{n.~} \tbGE{a spoilt thing} \tbOP{bonsaane} \tbTP{ bonsaanɛɛ} 
\tbLX{bonsammo} \tbHM{} \tbPH{bʊ́nsàmmʊ} \tbPS{n.~} \tbGE{unshelled nut} \tbOP{bonsama} \tbTP{bonsamɛɛ} 
\tbLX{bonsaraa} \tbHM{} \tbPH{bʋ́nsáràà} \tbPS{n.~} \tbGE{a young female} \tbOP{bonsarre} \tbTP{ bonsarɛɛ} 
\tbLX{bonseenaa} \tbHM{} \tbPH{bʋ́nsɪ́ɪ́náà} \tbPS{n.~} \tbGE{something that is touched} \tbOP{boseene} \tbTP{ bonseenɛɛ.} 
\tbLX{bonseɛraa} \tbHM{1} \tbPH{bʊ́nsɪ́ɛ́ráá} \tbPS{n.~} \tbSN \tbGE{something that is worn on the waist} \tbSN \tbGE{something that is sewn} \tbOP{bonseɛre} \tbTP{ bonseɛrɛɛ} 
\tbLX{bonseɛraa} \tbHM{1} \tbPH{bʊ́nsɪ́ɛ́ráá} \tbPS{n.~} \tbGE{that which is better than} \tbOP{bonseɛre} \tbTP{ bonseɛrɛɛ} 
\tbLX{bonseɛraa} \tbHM{2} \tbPH{bʊ́nsɪ́ɛ́ráá} \tbPS{n.~} \tbGE{something that is roasted} \tbOP{bonseɛre} \tbTP{bonseɛrɛɛ} 
\tbLX{bonsennaa} \tbHM{} \tbPH{bʊ̀nsɪ́nnáà} \tbPS{n.~} \tbGE{something that is the right size} \tbOP{bonsenne} \tbTP{bonsennɛɛ} 
\tbLX{bonserɛ} \tbHM{} \tbPH{bʋ́nsɪ̀rɛ́} \tbPS{n.~} \tbGE{a valuable thing} \tbOP{bonserre} \tbTP{ bonserɛɛ/bonserrɛɛ} 
\tbLX{bonsɛgeraa} \tbHM{} \tbPH{bʊ́nsɛ́gráá} \tbPS{n.~} \tbGE{something that is written} \tbOP{bonsɛgere} \tbTP{ bonsɛgerɛɛ} 
\tbLX{bonsɛgeraa} \tbHM{} \tbPH{bʊ́nsɛ̀gráà} \tbPS{n.~} \tbGE{something that is used to make another thing sturdy} \tbOP{bonsɛgere} 
\tbLX{bonsigiraa} \tbHM{} \tbPH{bʋ́nsìgráà} \tbPS{n.~} \tbSN \tbGE{something whose price is reduced} \tbSN \tbGE{something that is lowered (height)} \tbOP{bonsigiri} \tbTP{bonsigiree} 
\tbLX{bonso} \tbHM{} \tbPH{bʋ́nsʋ́} \tbPS{n.~} \tbGE{the act of owning} 
\tbLX{bonso} \tbHM{} \tbPH{bónsó} \tbPS{adv.~} \tbGE{why?} \tbXV{Bonso ka fo ba gaa a daa?} \tbXE{Why have you not gone to the market?.} 
\tbLX{bonso} \tbHM{} \tbPH{bónsó} \tbPS{conj.~} \tbGE{because} \tbXV{Ba ba gaa a sori bonso a lɔɔre sa᷉aɛ la.} \tbXE{They could not go on the trip because the car broke down.} 
\tbLX{bonsoŋ} \tbHM{} \tbPH{bʋ́nsʋ́ŋ} \tbPS{n.~} \tbGE{a good or blessed thing} \tbOP{bonsonne} \tbTP{bonsonnɛɛ} 
\tbLX{bonsooraa} \tbHM{} \tbPH{bʊ́nsʊ́ʊ́ráá} \tbPS{n.~} \tbGE{a thing that is owned; a belonging} \tbOP{bonsoore} \tbTP{bonsoorɛɛ} 
\tbLX{bonsooraa} \tbHM{} \tbPH{bʊ́nsʊ́ʊ́ráà} \tbPS{n.~} \tbGE{something that is asked for or wanted} 
\tbLX{bonsɔgelaa} \tbHM{} \tbPH{bʋ́nsɔ́gláà} \tbPS{n.~} \tbSN \tbGE{a black thing} \tbSN \tbGE{a hidden thing} \tbOP{bonsɔgelɔ} \tbTP{ bonsɔgelɛɛ} 
\tbLX{bonsɔgele} \tbHM{} \tbPH{bʊ́nsɔ́glɪ́} \tbPS{n.~} \tbGE{the act of hiding things} 
\tbLX{bonsɔraa} \tbHM{} \tbPH{bʋ́nsɔ̀ráà} \tbPS{n.~} \tbGE{something that is begged for} \tbOP{bonsɔrre} \tbTP{bonsɔrɛɛ} 
\tbLX{bonsɔraa} \tbHM{} \tbPH{bʊ́sɔ́ráá} \tbPS{n.~} \tbGE{something that has been counted} \tbXV{A pɔge iri la a bonsɔraa kaŋa.} \tbXE{The woman took one of the already counted ones.} \tbOP{bonsɔrre} 
\tbLX{bonsu} \tbHM{} \tbPH{bʋ́nsù} \tbPS{n.~} \tbGE{the act of wearing (clothes)} 
\tbLX{bonsu} \tbHM{} \tbPH{bʋ́nsù} \tbPS{n.~} \tbGE{the act of pushing things into an opening} 
\tbLX{bonsugo} \tbHM{} \tbPH{bʋ́nsúgò} \tbPS{n.~} \tbGE{things in great numbers;} \tbOP{bonsugri} \tbTP{ bonsugree} 
\tbLX{bonsuuraa} \tbHM{} \tbPH{bʋ́nsúů̀ráà} \tbPS{n.~} \tbGE{garment} \tbOP{bonsuuri} \tbTP{ bonsuuree} 
\tbLX{bontaa} \tbHM{} \tbPH{bʋ́ntáà} \tbPS{n.~} \tbGE{the act of owning} 
\tbLX{bontage} \tbHM{} \tbPH{bʋ́ntàgɪ́} \tbPS{n.~} \tbGE{a thick thing} \tbOP{bontagre} \tbTP{ bontagrɛɛ} 
\tbLX{bontage} \tbHM{} \tbPH{bʊ́ntágɪ̀} \tbPS{n.~} \tbGE{the act of pulling things} 
\tbLX{bontale} \tbHM{} \tbPH{bʋ́ntálɪ̀} \tbPS{n.~} \tbGE{a lobe of  (kola)} \tbOP{bontala} \tbTP{bontalɛɛ} 
\tbLX{bontaŋa} \tbHM{} \tbPH{bòtáŋá} \tbPS{n.~} \tbGE{camel} \tbOP{bontanne} \tbTP{bontannɛɛ} 
\tbLX{bontara} \tbHM{} \tbPH{bʋ́ntárá} \tbPS{n.~} \tbGE{rich man; one who owns many things} \tbOP{bontareba} \tbTP{bontarebɛɛ.} 
\tbLX{bonteɛ} \tbHM{} \tbPH{bʋ́ntɪ̀ɛ̀} \tbPS{n.~} \tbSN \tbGE{the act of shooting} \tbSN \tbGE{the act of supporting things} 
\tbLX{bonteɛraa} \tbHM{} \tbPH{bʋ́ntɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{something that supports something else} \tbOP{bonteɛre} \tbTP{bonteɛrɛɛ} 
\tbLX{bonteɛraa} \tbHM{} \tbPH{bʋ́ntɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{a boil} \tbOP{bonteɛre} \tbTP{bonteɛrɛɛ} 
\tbLX{bonteɛraa} \tbHM{} \tbPH{bʋ́ntɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{something that is shot at} \tbOP{bonteɛre} \tbTP{bonteɛrɛɛ} 
\tbLX{bonteɛrɛ} \tbHM{} \tbPH{bʊ́ntɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{shooter} \tbOP{bonteɛrebɛ} \tbTP{bonteɛrebɛɛ} 
\tbLX{bontegeroŋ} \tbHM{} \tbPH{bʋ́ntɪ́grʋ́ŋ} \tbPS{n.~} \tbGE{a heavy thing} \tbOP{ bontegerɛ} \tbTP{bontegerɛɛ} 
\tbLX{bonteraa} \tbHM{} \tbPH{bʋ́ntɪ́ráà} \tbPS{n.~} \tbGE{something that is given out or offered} \tbOP{bonterre} 
\tbLX{bontere} \tbHM{} \tbPH{bʋ́ntɪ́rɪ́} \tbPS{n.~} \tbGE{the act of giving out things} 
\tbLX{bontere} \tbHM{} \tbPH{bʊ̀ntɪ́rɪ} \tbPS{n.~} \tbGE{a big drum made of animal skin and a gourd} \tbVA{bentere}
\tbLX{bontɛge} \tbHM{} \tbPH{bʊ́tɛ́gɪ̀} \tbPS{n.~} \tbGE{the act of exchanging} 
\tbLX{bontɛgeraa} \tbHM{} \tbPH{bʊ́ntɛ̀gráà} \tbPS{n.~} \tbGE{something that is traded for} \tbOP{bontɛgere} \tbTP{bontɛgerɛɛ} 
\tbLX{bontɛgerɛ} \tbHM{} \tbPH{bʊ́ntɛ́grɛ́} \tbPS{n.~} \tbGE{one who exchanges} \tbOP{bontɛgerebɛ} 
\tbLX{bonto} \tbHM{} \tbPH{bʋ́ntʋ́} \tbPS{n.~} \tbGE{act of insaulting of things} 
\tbLX{bonto} \tbHM{} \tbPH{bòntó} \tbPS{n.~} \tbGE{type of cigar made be the Dagaabe} \tbOP{bontori} \tbTP{bontoree} 
\tbLX{bontobo} \tbHM{} \tbPH{bʋ́ntòbó} \tbPS{n.~} \tbSN \tbGE{a type of plant; medicinal} \tbSN \tbGE{the immediate tender leaves of a plant (bean leaves as vegetable for soup)} \tbSN \tbGE{thing with prominent ears} 
\tbLX{bontoɔlaa} \tbHM{} \tbPH{bʋ́ntʋ́ɔ́láá} \tbPS{n.~} \tbGE{that which is sent by a messenger} \tbOP{bontoɔle/bontoɔlɔ} \tbTP{bontoɔlɛɛ} 
\tbLX{bontɔraa} \tbHM{} \tbPH{bʊ́ntɔ̀ráà} \tbPS{n.~} \tbGE{something that is shared} \tbOP{bontɔrre} 
\tbLX{bontɔre} \tbHM{} \tbPH{bʋ́ntɔ̀rɪ́} \tbPS{n.~} \tbGE{a worthy thing} \tbOP{bontɔɛ} \tbTP{bontɔrɛɛ} 
\tbLX{bontɔre} \tbHM{} \tbPH{bʊ́ntɔ̀rɪ̀} \tbPS{n.~} \tbGE{the act of giving out or sharing to as many as possible} 
\tbLX{bontuo} \tbHM{} \tbPH{bʋ́ntúó} \tbPS{n.~} \tbSN \tbGE{a bitter thing} \tbSN \tbGE{a poisonous or dangerous thing} \tbOP{bontuuri} \tbTP{bonturee.} 
\tbLX{bonuu} \tbHM{} \tbPH{bʋ̀núú} \tbPS{adj.~} \tbGE{five times} \tbOP{bonuuri} \tbTP{bonuuree} 
\tbLX{bonvaane} \tbHM{} \tbPH{bʋ́nvàànɪ́} \tbPS{n.~} \tbGE{a shrivelled thing} \tbOP{bonvaama} \tbTP{ bonvaanɛɛ.} 
\tbLX{bonvaraa} \tbHM{} \tbPH{bʋ́nváràà} \tbPS{n.~} \tbGE{castrated animal} \tbOP{bonvarre} \tbTP{ bonvarrɛɛ.} 
\tbLX{bonvelaa} \tbHM{} \tbPH{bʋ́nvɪ̀làá} \tbPS{n.~} \tbGE{a good thing} \tbOP{bonveɛle} \tbTP{ bonveɛlɛɛ} 
\tbLX{bonveraa} \tbHM{} \tbPH{bʋ́nvɪ̀ráà} \tbPS{n.~} \tbSN \tbGE{an implement used in throwing stones (catapult)} \tbSN \tbGE{something that has been shot at by means of a catapult} \tbOP{bonverre} 
\tbLX{bonveraa} \tbHM{} \tbPH{bʋ́nvɪ̀ráà} \tbPS{n.~} \tbGE{a shift of a bone} \tbOP{bonverre} 
\tbLX{bonvere} \tbHM{} \tbPH{bʋ́nvɪ̀rɪ̀} \tbPS{n.~} \tbGE{the act of spraining (body part)} 
\tbLX{bonvere} \tbHM{} \tbPH{bʋ́nvɪ̀rɪ̀} \tbPS{n.~} \tbGE{the act of shooting with a catapult} 
\tbLX{bonvooraa} \tbHM{} \tbPH{bʋ́nvʋ́ʋ́ráá} \tbPS{n.~} \tbGE{something that breathes} \tbOP{bonvoore} \tbTP{ bonvoorɛɛ.} 
\tbLX{bonvoɔloŋ} \tbHM{} \tbPH{bʋ́nvʋ̀ɔ̀lʋ̌ŋ} \tbPS{n.~} \tbGE{a slimy thing} \tbOP{bonvoɔle} \tbTP{ bonvoɔlɛɛ.} 
\tbLX{bonvore} \tbHM{} \tbPH{bʋ́nvʋ́rɪ̀} \tbPS{n.~} \tbGE{a living thing} \tbOP{bonvoɛ/bonvoba} \tbTP{bonvorɛɛ} 
\tbLX{bonvuo} \tbHM{} \tbPH{bʋ́nvùó} \tbPS{n.~} \tbGE{an open thing; an uncovered enclosure or container} \tbOP{bonvori} \tbTP{ bonvoree} 
\tbLX{bonvuuraa} \tbHM{} \tbPH{bʋ́nvúúráá} \tbPS{n.~} \tbSN \tbGE{a thing that crawls (e.g. creature without legs)} \tbSN \tbGE{a thing that is dragged along the ground} \tbOP{bonvuuri} \tbTP{ bonvuuree.} 
\tbLX{bonweɛ} \tbHM{} \tbPH{bʋ́nwɪ́ɛ́} \tbPS{n.~} \tbGE{a common thing} \tbOP{bonweere} \tbTP{bonweerɛɛ.} 
\tbLX{bon-yaga} \tbHM{} \tbPH{bʋ́ŋ-yágà} \tbPS{n.~} \tbGE{many things} \tbTP{bon-yagrɛɛ} \tbVA{boŋyaga}
\tbLX{bon-yeɛle} \tbHM{} \tbPH{bʋ́n-yɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{winnowed thing} 
\tbLX{bon-yɛɛ} \tbHM{} \tbPH{bʋn-yɛɛ} \tbPS{n.~} \tbSN \tbGE{unimportant things} \tbSN \tbGE{wild crops} \tbSG{bon-yɛroo} \tbTP{bon-yɛrɛɛ} \tbVA{boŋyɛɛ}
\tbLX{bon-yɛraa} \tbHM{} \tbPH{bʋ́n-yɛ̀ráà} \tbPS{n.~} \tbSN \tbGE{something that is hung around the neck} \tbSN \tbGE{garment} \tbOP{bon-yɛrre} \tbTP{bon-yɛrrɛɛ} 
\tbLX{bon-yoɔraa} \tbHM{} \tbPH{bʋ́n-yʋ́ɔ́ráá} \tbPS{n.~} \tbSN \tbGE{something that roams or wanders about} \tbSN \tbGE{rheumatism} \tbOP{bon-yoɔre} \tbTP{ bon-yoɔrɛɛ} 
\tbLX{bon-yɔmɔ} \tbHM{} \tbPH{bʋ́n-yɔ̀mɔ́} \tbPS{n.~} \tbGE{plentiful and free thing(s)} \tbTP{ bon-yɔmɛɛ.} 
\tbLX{bon-yuo} \tbHM{} \tbPH{bʋ́n-yúó} \tbPS{n.~} \tbGE{different one} \tbXV{A bie yage la  a kparoo na bare kyɛ de bonyuo.} \tbXE{The boy took of that shirt and took a different one.} \tbOP{bon-yuobo} \tbVA{boŋyuo}
\tbLX{bon-yuori} \tbHM{} \tbPH{bún-yúórì} \tbPS{n.~} \tbGE{a popular or prominent thing} \tbOP{bon-yoe} \tbTP{ bon-yoree.} 
\tbLX{bonzaa} \tbHM{} \tbPH{bʊ̂nzàá} \tbPS{n.~} \tbSN \tbGE{anything, everything} \tbXV{Fo na nyɛ la bonzaa fo naŋ boɔrɔ.} \tbXE{You will get anything that you want.} \tbSN \tbGE{everyone} \tbXV{Bonzaa naŋ wa kyɛ taa la puoraa.} \tbXE{Thanks to everyone who has come here.} 
\tbLX{bonzagelaa} \tbHM{} \tbPH{bʊ́nzágláá} \tbPS{n.~} \tbGE{an empty thing} \tbOP{bonzagele} 
\tbLX{bonzeɛ} \tbHM{} \tbPH{bʊ́nzɪ̀ɛ̀} \tbPS{n.~} \tbGE{a red thing} \tbOP{bonzeere} \tbTP{bonzeerɛɛ} 
\tbLX{bonzoolɔ} \tbHM{} \tbPH{bʊ́nzʊ́ʊ́lɔ́} \tbPS{n.~} \tbGE{a thing with hairy ends (corncob)} \tbTP{bonzoolɛɛ} \tbVA{bonzoola}
\tbLX{bonyanlori} \tbHM{} \tbPH{bʊ̀ɲánlórí} \tbPS{n.~} \tbGE{very large female goat} \tbTP{bonyanloree} 
\tbLX{bonyaŋaa} \tbHM{} \tbPH{bʊ̀nɲáŋáá} \tbPS{n.~} \tbGE{female mother goat} \tbOP{bonyaŋene} \tbTP{bonyaŋenɛɛ} 
\tbLX{bonyɛ} \tbHM{} \tbPH{bʋ̀ɲɛ́} \tbPS{n.~} \tbGE{wicked ideas} \tbXV{A dɔɔ suuri da iri ne la o pɔge kyɛ o da ba de bonyɛ tuuro ne o.} \tbXE{The man was annoyed with his wife but he did not have any wicked ideas towards her.} 
\tbLX{bonyɛ} \tbHM{} \tbPH{bʋ̀ɲɛ́} \tbPS{adj.~} \tbGE{having wicked ideas} 
\tbLX{boŋ} \tbHM{} \tbPH{bóŋ} \tbPS{interrogative  pronoun.~} \tbGE{what?} \tbXV{Boŋ ka fo na e zenɛ?} \tbXE{What will you do today?.} 
\tbLX{boŋemaŋ} \tbHM{} \tbPH{bʋ́ŋɪ̀máŋ} \tbPS{adj.~} \tbGE{hazy} 
\tbLX{boŋganaa} \tbHM{} \tbPH{bʋŋganaa} \tbPS{n.~} \tbGE{bed, something that is slept on} \tbOP{boŋganne} \tbTP{ boŋganɛɛ.} 
\tbLX{boŋgoraa} \tbHM{} \tbPH{bʋ́ŋgʋ̀ráà} \tbPS{n.~} \tbGE{something that is abundant} \tbOP{boŋgorre} \tbTP{boŋgorɛɛ} 
\tbLX{boŋgori} \tbHM{} \tbPH{bʋ́ŋgòrí} \tbPS{n.~} \tbGE{half-ripe fruit} \tbOP{boŋgoe} \tbTP{ boŋgoree.} 
\tbLX{boŋgori} \tbHM{} \tbPH{bóŋgórí} \tbPS{n. pl.~} \tbGE{shoes} 
\tbLX{boŋgu} \tbHM{} \tbPH{bʋ́ŋgù} \tbPS{n.~} \tbGE{the act of watching over something (crops from animals)} 
\tbLX{boŋguolaa} \tbHM{} \tbPH{bʋ́ŋgúóláá} \tbPS{n.~} \tbSN \tbGE{something that is raised or reared} \tbSN \tbGE{an animal given to someone to be raised} \tbOP{boŋguoli} \tbTP{boŋguolee} 
\tbLX{boŋguoli} \tbHM{} \tbPH{bʊ́ŋgúólí} \tbPS{n.~} \tbGE{domestic animals and birds (poultry and animal husbandry)} \tbOP{boŋguolo} 
\tbLX{boŋguoli} \tbHM{} \tbPH{bʊ̌ŋgúólì} \tbPS{n.~} \tbGE{the rearing of animals} 
\tbLX{boŋguolo} \tbHM{} \tbPH{bʊ́ŋgúólò} \tbPS{n.~} \tbSN \tbGE{} \tbVA{doŋguolo}
\tbLX{boŋguraa} \tbHM{} \tbPH{bʋ́ŋgùráà} \tbPS{n.~} \tbGE{something jumped upon} \tbOP{boŋgurri} 
\tbLX{boŋguuro} \tbHM{} \tbPH{bʊ́ŋgúúro} \tbPS{n.~} \tbGE{one who watches over things} \tbOP{boŋguuribo} \tbTP{boŋguuribee} 
\tbLX{boŋgbuli} \tbHM{} \tbPH{bʊ́ŋgbùlí} \tbPS{n.~} \tbSN \tbGE{a whole thing} \tbSN \tbGE{tumour} \tbOP{boŋgbulo} \tbTP{boŋgbulee} 
\tbLX{boŋkaa} \tbHM{} \tbPH{bʋ́ŋkáá} \tbPS{n.~} \tbGE{invocation of spirits} \tbOP{boŋkaareba} \tbTP{boŋkaarebɛɛ} \tbVA{boŋkaale}
\tbLX{boŋkaa} \tbHM{} \tbPH{bʋ́ŋkàà} \tbPS{n.~} \tbSN \tbGE{the act of taking care of something} \tbSN \tbGE{the act of staring at something} 
\tbLX{boŋkâa} \tbHM{} \tbPH{bʋ́ŋká᷉à᷉} \tbPS{n.~} \tbSN \tbGE{oil or fat from an animal} \tbSN \tbGE{something oily or full of fat} \tbSG{boŋkaanoo} \tbTP{boŋkaanɛɛ} 
\tbLX{boŋkaara} \tbHM{} \tbPH{bʊ́ŋkáára} \tbPS{n.~} \tbGE{one who watches over something} \tbOP{boŋkaareba} \tbTP{boŋkaarebɛɛ} 
\tbLX{boŋkaŋa} \tbHM{} \tbPH{bʊ́ŋkáŋ} \tbPS{n.~} \tbGE{something} \tbOP{bommine} \tbVA{boŋkaŋa}
\tbLX{boŋkapala} \tbHM{} \tbPH{boŋkapala} \tbPS{n.~} \tbGE{a type of plant} \tbTP{boŋkapalɛɛ} 
\tbLX{boŋkaraa} \tbHM{} \tbPH{bʋ́ŋkáráá} \tbPS{n.~} \tbGE{a raw, uncooked, or unripe thing} \tbOP{boŋkare} \tbTP{ boŋkarɛɛ.} 
\tbLX{boŋkoŋi} \tbHM{} \tbPH{bʋ́ŋkǒŋ} \tbPS{n.~} \tbGE{a lean thing} \tbOP{boŋkonni} \tbTP{boŋkonnee} 
\tbLX{boŋkôɔ} \tbHM{} \tbPH{bʋ́ŋkʋ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{a watery thing} \tbTP{ boŋkoɔnɛɛ} 
\tbLX{boŋkoɔraa} \tbHM{} \tbPH{bʋ́ŋkʊ́ɔ́ráá} \tbPS{n.~} \tbSN \tbGE{crop; something that is cultivated} \tbSN \tbGE{something used for ploughing or farming} \tbOP{boŋkoɔre, boŋkoɔrɛɛ} 
\tbLX{boŋkoɔraa} \tbHM{} \tbPH{bʋ́ŋkʋ́ɔ́ráà} \tbPS{n.~} \tbGE{something meant for sale (goods)} \tbOP{boŋkoɔre} \tbTP{boŋkoɔrɛɛ} 
\tbLX{boŋkorabinzie} \tbHM{} \tbPH{bʊ́ŋkʊ́rábìnzíé} \tbPS{n.~} \tbGE{museum} \tbOP{boŋkorabinziiri} 
\tbLX{boŋkore} \tbHM{} \tbPH{bʋ́ŋkʊ́rɪ́} \tbPS{n.~} \tbGE{old thing} \tbOP{boŋkorɔ} \tbTP{boŋkorɛɛ} 
\tbLX{boŋkuoŋaa} \tbHM{} \tbPH{bʋ́ŋkúóŋáá} \tbPS{n.~} \tbGE{a dry thing} \tbOP{boŋkuoni} \tbTP{ boŋkuonee} 
\tbLX{boŋku᷉u} \tbHM{} \tbPH{bʊ́ŋkú᷉ú᷉} \tbPS{n.~} \tbGE{a dead thing} \tbOP{bokuuni} \tbTP{ boŋkuunee} 
\tbLX{boŋkpɔgɔ} \tbHM{} \tbPH{bʊ́ŋkpɔ́gɔ̀} \tbPS{n.~} \tbSN \tbGE{lumps} \tbSN \tbGE{a type of disease whose symptoms are swellings and boils} \tbTP{boŋkpɔgerɛɛ} \tbVA{boŋkpɔga/boŋkpaga}
\tbLX{Boŋŋo} \tbHM{} \tbPH{bóŋŋò} \tbPS{n.~} \tbGE{a traditional festival of Jirapa  traditional area} 
\tbLX{boŋŋmaa} \tbHM{} \tbPH{bʋ́ŋŋmáá} \tbPS{n.~} \tbGE{a piece of something} \tbOP{boŋŋmaara} \tbTP{ boŋŋmaarɛɛ} 
\tbLX{boŋŋmaa} \tbHM{} \tbPH{bʊ́ŋŋmàà} \tbPS{n.~} \tbGE{the practice of cutting things} 
\tbLX{boŋŋmaa} \tbHM{} \tbPH{bʊ́ŋŋmàà} \tbPS{n.~} \tbGE{confining of domestic animals that have been ranging freely (usually at the beginning of the farming season)} 
\tbLX{boŋŋmare} \tbHM{} \tbPH{bʋ́ŋŋmárɪ́} \tbPS{n.~} \tbGE{the act of breaking things} \tbOP{boŋŋmarre} \tbTP{ boŋŋmarɛɛ.} 
\tbLX{boŋŋmare} \tbHM{} \tbPH{bʊ́ŋŋmàrɪ́} \tbPS{n.~} \tbGE{something that is spoilt (food)} \tbTP{boŋŋmarɛɛ} 
\tbLX{boŋŋmeɛraa} \tbHM{} \tbPH{bʊ́ŋŋmɪ́ɛ́ráà} \tbPS{n.~} \tbGE{musical instrument} \tbOP{boŋŋmeɛre} 
\tbLX{boŋo} \tbHM{} \tbPH{̀bòŋó} \tbPS{n.~} \tbGE{donkey} \tbOP{bonni} \tbTP{ bonnee.} 
\tbLX{boŋɔ} \tbHM{} \tbPH{bʊ̀ŋɔ́} \tbPS{adv.~} \tbGE{done anyhow} \tbVA{boŋa}
\tbLX{boo} \tbHM{} \tbPH{bóò} \tbPS{pron.~} \tbGE{} \tbVA{buori}
\tbLX{boo} \tbHM{} \tbPH{bʊ̀ʊ̀} \tbPS{adj.~} \tbGE{voluminous (dress)} 
\tbLX{bo᷉o} \tbHM{} \tbPH{bʋ̀᷉ʊ̀᷉} \tbPS{adv.~} \tbGE{oily-looking (top of sauce)} 
\tbLX{boolimbo} \tbHM{} \tbPH{bòòlímbò} \tbPS{n.~} \tbGE{taking one by the shoulders and chest (patient) (typically used with kpɛ̧)} \tbOP{boolimbori} \tbTP{boolimboree} 
\tbLX{boore} \tbHM{} \tbPH{bʋ̀ʋ̀rɪ̀} \tbPS{v.~} \tbGE{to attack suddenly and fiercely, to rush on} \tbPD{boorɛɛ, boorɔ, boorɔ, boorebɔ, booraa, boorebɔ } 
\tbLX{boore} \tbHM{} \tbPH{bʊ̌ʊ́rɪ} \tbPS{n.~} \tbGE{of same stalk, species, descendant} \tbOP{boora} \tbTP{boorɛɛ} \tbVA{booree}
\tbLX{boore} \tbHM{} \tbPH{bʊ̀ʊ̀rɪ́} \tbPS{n.~} \tbGE{aggrement, plan} \tbTP{boorɛɛ} 
\tbLX{boorɔ} \tbHM{} \tbPH{bʊ́ʊ́rɔ́} \tbPS{n.~} \tbSN \tbGE{off same type, family} \tbSN \tbGE{seed for planting} \tbSG{booroo} \tbVA{boora}
\tbLX{boɔ} \tbHM{} \tbPH{bʊ́ɔ́} \tbPS{n.~} \tbGE{goat} \tbOP{boore} \tbTP{boorɛɛ} 
\tbLX{bo᷉ɔ} \tbHM{} \tbPH{bʊ̀᷉ɔ̀᷉} \tbPS{v.~} \tbGE{to scoop up a large amount of something} \tbXV{ O bo᷉ɔ la a ka᷉a poli o eŋɛ zaa.} \tbXE{She scooped up a large amount of lotion and rubbed it all over her body.} \tbPD{bo᷉ɔɛ, boɔnɔ, boɔnɔ, boɔnema, boɔnaa, - } 
\tbLX{bo᷉ɔ} \tbHM{} \tbPH{bʊ̀᷉ɔ̀᷉} \tbPS{v.~} \tbGE{snub} \tbXV{A dɔɔ suuri iree la ka o maŋ deɛ bo᷉ɔ neɛ te lɔɔ.} \tbXE{The man is so annoyed that he snubs anybody he comes across.} \tbPD{bo᷉ɔɛ, boɔnɔ, boɔnɔ, boɔnema, boɔnaa, - } 
\tbLX{boɔbo} \tbHM{} \tbPH{bʋ́ɔ́bʋ̀} \tbPS{n.~} \tbSN \tbGE{something that is preferred} \tbSN \tbGE{the act of liking or searching for something} 
\tbLX{boɔlaa} \tbHM{} \tbPH{bʋ̀ɔ̀làá} \tbPS{n.~} \tbSN \tbGE{call} \tbSN \tbGE{invitation} 
\tbLX{boɔle} \tbHM{} \tbPH{bʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to call} \tbSN \tbGE{to invite} \tbPD{boɔlɛɛ, boɔlɔ, boɔlɔ, boɔleba, boɔlaa, boɔleba } 
\tbLX{boɔraa} \tbHM{} \tbPH{bʋ̀ɔ̀ráà} \tbPS{n.~} \tbSN \tbGE{sack} \tbSN \tbGE{one hundred pounds} \tbOP{boɔre} \tbTP{boɔrɛɛ} 
\tbLX{boɔre} \tbHM{} \tbPH{bʋ̀ɔ̀rɪ́} \tbPS{v.~} \tbGE{to smear with thick paste} \tbPD{boɔrɛɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ } 
\tbLX{boɔrɔ} \tbHM{} \tbPH{bʊ́ɔ́rɔ̀} \tbPS{v.~} \tbSN \tbGE{to like, to love; to desire, to want} \tbSN \tbGE{to need} \tbPD{bɔɛ, boɔrɔ, boɔrɔ, boɔreba, boɔraa, - } 
\tbLX{boɔso} \tbHM{} \tbPH{bʋ́ɔ́sʋ́} \tbPS{n.~} \tbGE{bus} \tbOP{boɔsere} \tbTP{ boɔserɛɛ} 
\tbLX{bopeɛ} \tbHM{} \tbPH{bʋ̀pɪ́ɛ́} \tbPS{n.~} \tbGE{traditional storey building} \tbOP{bopeere} \tbTP{bopeerɛɛ} \tbVA{peɛzudie}
\tbLX{boraa} \tbHM{} \tbPH{bʊ́ráá} \tbPS{n.~} \tbGE{a stick used in sowing grains} \tbOP{borre} \tbTP{borrɛɛ} \tbVA{boredaa}
\tbLX{bore} \tbHM{} \tbPH{bʊ̀rɪ̀} \tbPS{v.~} \tbGE{to break part of a lump off in big chunks} \tbPD{borɛɛ, bora, bora, boreba, boraa, - } 
\tbLX{bore} \tbHM{} \tbPH{bʊ̀rɪ̀} \tbPS{v.~} \tbGE{to sow, to plant} \tbPD{borɛɛ, bora, bora, boreba, boraa, boreba } 
\tbLX{boredaa} \tbHM{} \tbPH{bʋ̀rɪ̀dàá} \tbPS{n.~} \tbGE{a stick for sowing} \tbOP{boredaare} \tbTP{ boredarɛɛ/boredaarɛɛ.} 
\tbLX{borenyɛ} \tbHM{} \tbPH{bʋ́rɪ́ɲɛ̀} \tbPS{n.~} \tbGE{Christmas} \tbOP{borenyɛre} \tbTP{ borenyɛrɛɛ} 
\tbLX{boreŋmaakpantoli} \tbHM{} \tbPH{bʊ̀rɪ̀ŋmàáakpántólí} \tbPS{n.~} \tbGE{a type of large, beautiful bird with a sweet voice} \tbOP{boreŋmaakpantolo} \tbTP{boreŋmaakpantolee} 
\tbLX{boreŋmaao} \tbHM{} \tbPH{bʋ̀rɪ̊ŋmààʋ́} \tbPS{n.~} \tbGE{a large bird with a strong sweet voice} \tbOP{boreŋmaane} \tbTP{ boreŋmaanɛɛ.} 
\tbLX{boreŋmanlee} \tbHM{} \tbPH{bʊ̀rɪ̀ŋmánléé} \tbPS{n.~} \tbGE{calabash for sowing} \tbOP{boreŋmambili} \tbTP{boreŋmambilee} 
\tbLX{boribara} \tbHM{} \tbPH{bòrìbàrà} \tbPS{adv.~} \tbGE{done poorly} 
\tbLX{boribori} \tbHM{} \tbPH{bòrìbórí} \tbPS{v.~} \tbGE{to carry out poorly (cleaning, talk)} \tbPD{boriboree, boriboro, boriboro, boriboribo, boriboraa, - } 
\tbLX{boroboro} \tbHM{} \tbPH{bòròbòrò} \tbPS{n.~} \tbGE{bread} \tbOP{boroborri} \tbTP{boroboree} 
\tbLX{boroborozɔŋ} \tbHM{} \tbPH{bòròbòròzɔ̂ŋ} \tbPS{n.~} \tbGE{bread flour} \tbTP{boroborozɔnnɛɛ} 
\tbLX{boroto} \tbHM{} \tbPH{bòròtò} \tbPS{adj.~} \tbGE{ big and round} 
\tbLX{boroto} \tbHM{} \tbPH{bʊ̀rʊ̀tʊ̊} \tbPS{adv.~} \tbGE{fall flat} \tbXV{A bie zoro gɛrɛ te ŋmɛ gbɛre a gaa te boroto o naŋ le.} \tbXE{As the boy ran along he stumbled on a stone and crashed to the ground.} 
\tbLX{borɔbɛ} \tbHM{} \tbPH{bʊ̀rɔ̀bɛ́} \tbPS{n.~} \tbGE{pineapple fruit or plant} \tbOP{borɔbɛre} \tbTP{borɔbɛrɛɛ} 
\tbLX{borɔborɔ} \tbHM{} \tbPH{bʋ̀rɔ̀bʋ̀rɔ̀} \tbPS{adj.~} \tbGE{multicoloured} 
\tbLX{borɔso} \tbHM{} \tbPH{bʋ̀rɔ́sʋ̀} \tbPS{n.~} \tbGE{ brush} \tbOP{borɔsore} \tbTP{ borɔsorɛɛ} 
\tbLX{bosaraa} \tbHM{} \tbPH{bʋ̀sáràà} \tbPS{n.~} \tbGE{young female goat} \tbOP{bosarre} \tbTP{bosarrɛɛ} 
\tbLX{bosere} \tbHM{} \tbPH{bʋ̀sɪ́rɪ̀} \tbPS{n.~} \tbGE{a necklace made of huge beads} \tbOP{boseɛ/boserre} \tbTP{boserɛɛ} 
\tbLX{bovammo} \tbHM{} \tbPH{bʋ̀vàmmʋ́} \tbPS{n.~} \tbGE{a type of bow, usually not made from bamboo} \tbOP{bovama} \tbTP{ bovannɛɛ.} 
\tbLX{bovaraa} \tbHM{} \tbPH{bʋ̀váràá} \tbPS{n.~} \tbGE{castrated goat} \tbOP{bovarre} \tbTP{ bovarrɛɛ} 
\tbLX{bowola} \tbHM{} \tbPH{bʊ́wʊ́lá} \tbPS{adv.~} \tbGE{how many times} 
\tbLX{boyaa} \tbHM{} \tbPH{[bóyáá} \tbPS{interj.~} \tbGE{expression of surprise} \tbVA{boyaaŋ}
\tbLX{boyaaŋ} \tbHM{} \tbPH{bòyááŋ} \tbPS{interj.~} \tbGE{expression of surprise} 
\tbLX{boyoɔbo} \tbHM{} \tbPH{bʋ̀yʋ̀ɔ́bʋ́} \tbPS{adj.~} \tbGE{six times} 
\tbLX{boyɔpôi} \tbHM{} \tbPH{bʋ̀yɔ̀pó̂í} \tbPS{adj.~} \tbGE{seven times} 
\tbLX{bozeɛ} \tbHM{} \tbPH{bʊ̀zɪ́ɛ́} \tbPS{n.~} \tbGE{dog} \tbOP{bozeere} \tbTP{bozeerɛɛ} 
\tbLX{bɔ} \tbHM{} \tbPH{bɔ} \tbPS{v.~} \tbSN \tbGE{to search; to look for} \tbXV{N bie bɔrɛɛ la soŋ ma ka N bɔ.} \tbXE{My child is lost; help me look for her.} \tbSN \tbGE{to court, to woo} \tbXV{Dagaaba maŋ bɔ la pɔge kyɛ de.} \tbXE{The Dagaaba usally court a woman before marriage.} \tbSN \tbGE{to engage in sexual intercourse, to make love} \tbXV{Ka fooŋ bɔ neɛ pɔge ba na ŋmɛ fo la sane yaga.} \tbXE{If you make love to a married woman, you will be heavily fined.} \tbPD{bɔɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ } 
\tbLX{bɔ} \tbHM{} \tbPH{bɔ̀} \tbPS{v.~} \tbGE{to clean a very dirty place or thing} \tbPD{bɔɛ, boɔrɔ, boɔrɔ, boɔrebɔ, boɔraa, boɔrebɔ } 
\tbLX{bɔbɔe} \tbHM{} \tbPH{bɔ́bɔ́ɪ̀} \tbPS{n.~} \tbGE{duck} \tbOP{bɔbɔɛ} \tbTP{bɔbɔrɛɛ} 
\tbLX{bɔbɔlaa} \tbHM{} \tbPH{bɔ́bɔ́láá} \tbPS{n.~} \tbGE{a wild animal like a dog} \tbOP{bɔbɔlle} \tbTP{bɔbɔlɛɛ} 
\tbLX{bɔe} \tbHM{} \tbPH{bɔ̀ɪ́} \tbPS{n.~} \tbGE{servant} \tbOP{bɔemine} 
\tbLX{bɔ᷉ee} \tbHM{} \tbPH{bɔ̀᷉ɪ̀᷉̂ɪ̀᷉} \tbPS{adv.~} \tbGE{calmly, slowly} \tbXV{O yeli la ayɛlɛ bɔ᷉ee lɛ.} \tbXE{She spoke gently.} 
\tbLX{bɔge} \tbHM{} \tbPH{bɔ́gɪ́} \tbPS{v.~} \tbGE{to grind into smooth flour} \tbPD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \tbVA{bage}
\tbLX{bɔge} \tbHM{} \tbPH{bɔ́gɪ́} \tbPS{v.~} \tbSN \tbGE{to hit with many arrows; to sting or bite many times (insects)} \tbSN \tbGE{to apply poisonous lotion to arrows} \tbPD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \tbVA{bage}
\tbLX{bɔge} \tbHM{} \tbPH{bɔ́gɪ́} \tbPS{v.~} \tbGE{to put a fence around something} \tbPD{bɔgɛɛ, bɔgerɔ, bɔgerɔ, bɔgerebɔ, bɔgeraa, bɔgerebɔ } \tbVA{bage}
\tbLX{bɔge} \tbHM{} \tbPH{bɔ́gɪ́} \tbPS{n.~} \tbSN \tbGE{shoulder} \tbSN \tbGE{fore limb of an animal} \tbOP{bɔgere} \tbTP{bɔgerɛɛ} \tbVA{bɔga/baga}
\tbLX{bɔgebie} \tbHM{} \tbPH{bɔ̀gìbíé} \tbPS{n.~} \tbGE{misfortune} \tbOP{bɔgebiiri} \tbTP{bɔgebiiree} \tbVA{bɔɔbie/bagebie}
\tbLX{bɔgebiesoba} \tbHM{} \tbPH{bɔ̀gɪ̀bíésʋ́bá} \tbPS{n.~} \tbGE{unfortunate person; unlucky person} \tbOP{ bɔgebiedeme} \tbVA{bagebiesoba}
\tbLX{bɔgebogerɔ} \tbHM{} \tbPH{bɔ̀gɪ̀bʊ́grɔ́} \tbPS{n.~} \tbGE{sorcerer, soothsayer} \tbOP{bɔgebogerebɔ} \tbTP{bɔgebogerebɛɛ} \tbVA{bagebogera}
\tbLX{bɔgedie} \tbHM{} \tbPH{bɔ̀gɪ̀dìè} \tbPS{n.~} \tbGE{room for divining} \tbOP{bɔgederi} \tbTP{bɔgederee} 
\tbLX{bɔgele} \tbHM{} \tbPH{bɔ̀glɪ̀} \tbPS{v.~} \tbSN \tbGE{to stalk (hunting)} \tbXV{Naŋkpaana zaa maŋ bɔgele la wɛdoŋɔ a pɔge o ŋmɛ.} \tbXE{A hunter will normally stalk a wild animal to get close enough to shoot.} \tbSN \tbGE{to feel, to move one's hand over} \tbXV{A zɔŋɔ bɔgele la a gado zu te nyɛ a o kparoo.} \tbXE{The blind man felt along the bed and found his shirt.} \tbPD{bɔgelɛɛ, bɔgelɔ, bɔgelɔ, bɔgeleba, bɔgelaa, - } 
\tbLX{bɔgemaale} \tbHM{} \tbPH{bɔ̀gɪ̀mààlɪ̀} \tbPS{n.~} \tbGE{the act of making sacrifice} \tbOP{bɔgemaaleba} \tbTP{bɔgemaalbɛɛ} \tbVA{bagemaale}
\tbLX{bɔgeŋmampegere} \tbHM{} \tbPH{bɔ̀gɪ̀ŋmǎmpɪ́gɪ́rɪ́} \tbPS{n.~} \tbGE{shoulder} \tbVA{bageŋmampere}
\tbLX{bɔgeŋmampere} \tbHM{} \tbPH{bɔ̀gɪ̀ŋmàmpɪ́rɪ́} \tbPS{n.~} \tbGE{shoulder blade} \tbOP{bɔgeŋmamperre} \tbTP{bɔgeŋmamperɛɛ} \tbVA{bɔgeŋmampegere}
\tbLX{bɔgeŋmane} \tbHM{} \tbPH{bɔ̀gɪ̀ŋmánɪ́} \tbPS{n.~} \tbGE{forearm} \tbOP{bɔgeŋmama} \tbTP{bɔgeŋmamɛɛ} \tbVA{bageŋmane}
\tbLX{bɔgere} \tbHM{} \tbPH{bɔ̌grɪ́} \tbPS{n.~} \tbGE{part, portion of a whole} \tbTP{bɔgerɛɛ} \tbVA{bɔgre}
\tbLX{bɔgere} \tbHM{} \tbPH{bɔ̀gɪ̀rɪ́} \tbPS{n.~} \tbSN \tbGE{fetish, object believed to embody a spirit} \tbSN \tbGE{sacrifice, offering to a divinity} \tbOP{bɔgɔ/bɔga} \tbTP{bɔgerɛɛ/bɔgɛɛ} \tbVA{bagere}
\tbLX{bɔgesaa} \tbHM{} \tbPH{bɔ̀gsàà} \tbPS{n.~} \tbGE{counsellor for a person being initiated into the bagre cult} \tbOP{bɔgesaamine} \tbTP{bɔgesaarɛɛ} \tbVA{bagesaa}
\tbLX{bɔgesoŋ} \tbHM{} \tbPH{bɔ̀gɪ̀sʋ̂ŋ} \tbPS{n.~} \tbGE{luck} \tbOP{bɔgesonne} \tbTP{bɔgesonnɛɛ} \tbVA{bagesoŋ}
\tbLX{bɔgetaŋkpolo} \tbHM{} \tbPH{bɔ̀gɪ̀táŋkpóló} \tbPS{n.~} \tbGE{cult members, those who have gone through the initiation process for a cult} \tbOP{bɔgetaŋkpolli} \tbTP{bɔgetaŋkpolee} \tbVA{bagetaŋkpolo}
\tbLX{bɔgɔ} \tbHM{} \tbPH{bɔ́gɔ̀} \tbPS{n.~} \tbGE{track, path (usually left by passing through grassy bush)} \tbOP{bɔgere} \tbTP{bɔgerɛɛ} \tbVA{bɔga}
\tbLX{bɔgɔ} \tbHM{} \tbPH{bɔ́gɔ́} \tbPS{n.~} \tbGE{shoulder} \tbOP{bɔgere} \tbTP{bɔgerɛɛ} \tbVA{bɔge/bɔga/bage}
\tbLX{bɔlaa} \tbHM{} \tbPH{bɔ̀làà} \tbPS{adv.~} \tbGE{ long, round and smooth} 
\tbLX{bɔle} \tbHM{} \tbPH{bɔ́lɪ́} \tbPS{n.~} \tbGE{ball} \tbOP{bɔlɔ} \tbTP{bɔlɛɛ/bɔllɛɛ} 
\tbLX{bɔle} \tbHM{} \tbPH{bɔ̀lɪ̀} \tbPS{n.~} \tbGE{field rat} \tbSG{bɔloo} \tbOP{bɔlɔ} \tbTP{bɔllɛɛ} 
\tbLX{bɔlembɔ} \tbHM{} \tbPH{bɔ̀lɪ̀mbɔ́} \tbPS{n.~} \tbGE{bottle} \tbSG{bɔlembɔroo} \tbOP{ bɔlembɔre} \tbTP{bɔlembɔrɛɛ} \tbVA{pilintoɔ}
\tbLX{bɔll} \tbHM{} \tbPH{bɔ̀ll} \tbPS{adj.~} \tbGE{long and oblong} 
\tbLX{bɔlle} \tbHM{} \tbPH{bɔ̀llɪ̀} \tbPS{v.~} \tbSN \tbGE{to move all over an object (insects)} \tbSN \tbGE{to smear liquid on a surface in an untidy way} \tbPD{bɔllɛɛ, bɔllɔ, bɔllɔ, bɔlbɔ, bɔllaa, - } 
\tbLX{bɔlɔmbɔlɔŋ} \tbHM{} \tbPH{bɔ́lɔ́mbɔ́lɔ́ŋ} \tbPS{adj.~} \tbGE{smooth and shiny} 
\tbLX{bɔmbɔrezeɛ} \tbHM{} \tbPH{bɔ̀mbɔ̀rzɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of snake} \tbOP{bɔmbɔrezeere} \tbTP{bɔmbɔrezeerɛɛ} 
\tbLX{bɔmbɔroŋ} \tbHM{} \tbPH{bɔ̌mbɔ̀rʋ́ŋ} \tbPS{n.~} \tbGE{a type of snake} \tbOP{bɔmbɔrre} \tbTP{bɔmbɔrɛɛ} 
\tbLX{bɔmpɔrɔ} \tbHM{} \tbPH{bɔ̀mpɔ́rɔ́} \tbPS{n.~} \tbGE{mumps} \tbXV{ .} \tbXE{} \tbOP{bɔmpɔrre} \tbTP{ bɔmpɔrɛɛ.} 
\tbLX{bɔndaa} \tbHM{} \tbPH{bɔ̀ndáá} \tbPS{n.~} \tbGE{lizard (variation of ‘bandaa’(individual) and ‘baŋa’(dialectal))} \tbOP{bɔndaare} \tbTP{bɔndarɛɛ} \tbVA{bandaa}
\tbLX{bɔŋenaa} \tbHM{} \tbPH{bɔ́ŋɪ́náá} \tbPS{n.~} \tbGE{type of tree, medicinal} \tbOP{bɔŋene} \tbTP{bɔŋenɛɛ} 
\tbLX{bɔŋenɔ} \tbHM{} \tbPH{bɔ̀ŋɪ́nɔ́} \tbPS{n.~} \tbGE{witty person} \tbOP{bɔŋenema} \tbTP{bɔŋenemɛɛ} 
\tbLX{bɔŋɔ} \tbHM{} \tbPH{bɔ̀ŋɔ́} \tbPS{n.~} \tbGE{lizard} \tbOP{bɔnne} \tbTP{bɔnnɛɛ} \tbVA{baŋa/bɔŋa}
\tbLX{bɔŋɔ} \tbHM{} \tbPH{bɔ́ŋɔ́} \tbPS{n.~} \tbGE{bangle} \tbOP{bɔnne} \tbTP{bɔnnɛɛ} \tbVA{baŋa}
\tbLX{bɔɔ} \tbHM{} \tbPH{bɔ̀ɔ̀} \tbPS{v.~} \tbGE{to dress very well} \tbPD{bɔɔɛ, bɔɔrɔ, bɔɔrɔ, bɔɔrebɔ, bɔɔraa, bɔɔrebɔ } 
\tbLX{bɔɔderoo} \tbHM{} \tbPH{bɔ̀ɔ̀dɪ́rʊ́ʊ́} \tbPS{n.~} \tbGE{plantain} \tbSG{bɔɔderoo} \tbOP{bɔɔdeɛ} \tbTP{bɔɔdeerɛɛ} 
\tbLX{bɔɔdoa} \tbHM{} \tbPH{bɔ̀ɔ̀dʊ́áá} \tbPS{n.~} \tbGE{towel} \tbOP{bɔɔdoore} \tbTP{bɔɔdoorɛɛ} 
\tbLX{bɔɔre} \tbHM{} \tbPH{bɔ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{sacrifice, deity} \tbOP{bɔbɔ} \tbVA{bɔgre}
\tbLX{bɔre} \tbHM{} \tbPH{bɔ́rɪ̀} \tbPS{v.~} \tbGE{to break off a large chunk (from a bigger mass)} \tbPD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔreba, bɔraa, bɔreba } 
\tbLX{bɔre} \tbHM{} \tbPH{bɔ̀rɪ̀} \tbPS{v.~} \tbGE{to blossom, to flourish (plants)} \tbPD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔrebɔ, bɔraa, - } 
\tbLX{bɔre} \tbHM{} \tbPH{bɔ̀rɪ̀} \tbPS{v.~} \tbGE{to get lost, to go astray; to make a mistake} \tbPD{bɔrɛɛ, bɔrɔ, bɔrɔ, bɔrebɔ, bɔraa, - } 
\tbLX{bɔre} \tbHM{} \tbPH{bɔ́rɪ́} \tbPS{n.~} \tbGE{puff adder; a thick large poisonous snake} \tbOP{bɔrre} \tbTP{bɔrɛɛ} \tbVA{bɔsere}
\tbLX{bɔreferoo} \tbHM{} \tbPH{bɔ̀rɪ́fɪ́rʊ́ʊ́} \tbPS{n.~} \tbGE{pawpaw} \tbOP{bɔreferre} \tbTP{bɔreferrɛɛ} 
\tbLX{Bɔrefɔ} \tbHM{} \tbPH{bɔ̀rɪ̀fɔ́} \tbPS{n.~} \tbGE{English} \tbOP{bɔrefɔre} \tbTP{ bɔrefɔrɛɛ} \tbVA{bɔrɔ/naasaaleŋ}
\tbLX{bu} \tbHM{} \tbPH{bù} \tbPS{v.~} \tbSN \tbGE{to get wet} \tbXV{Saa poɔ la a tontonema ka ba bonsuuri zaa bu.} \tbXE{The rain beat down on the workers and soaked all their clothes.} \tbSN \tbGE{to mix dry powder with liquid or oil (mortar)} \tbXV{A demeɛrɛ bu la tɛne na mɛ a dankyini.} \tbXE{The mason has prepared mortar to build the wall with.} \tbPD{bue, buuro, buuro, buuribo, buuraa, buuribo } 
\tbLX{bu} \tbHM{} \tbPH{bú} \tbPS{v.~} \tbGE{to measure (liquid), to calculate} \tbXV{Ba bu la ka᷉a᷉ bɔlembɔre ata eŋ a doge poɔ.} \tbXE{The have measured out three bottles of oil into the pot.} \tbPD{bue, buuro, buuro, buuribo, buuraa, buuribo } 
\tbLX{bu} \tbHM{} \tbPH{bú} \tbPS{v.~} \tbSN \tbGE{to ponder, to reflect} \tbXV{N bu la a yɛlɛ n zu poɔ lɛɛ kyɛ N naŋ ba baŋ a pare.} \tbXE{I've pondered over the issue for some time now, but I can't seem to come to terms with it.} \tbSN \tbGE{to respect , to value} \tbXV{Ka fooŋ ba bu ma N meŋ koŋ bu fo.} \tbXE{If you don't respect me, I wont respect you either.} \tbPD{bue, buuro, buuro, buuribo, buuraa, buuribo } 
\tbLX{bufuli} \tbHM{} \tbPH{bùfúlí} \tbPS{n.~} \tbGE{a blood-sucking maggot-like worm, usually found under floors} \tbOP{bufulo} \tbTP{bufulee} 
\tbLX{bugagu} \tbHM{} \tbPH{bùgàgú} \tbPS{n.~} \tbGE{a type of tree, medicinal} \tbOP{bugaguri} \tbTP{bugaguree} 
\tbLX{bugi} \tbHM{} \tbPH{bùgì} \tbPS{v.~} \tbSN \tbGE{to be confused, to be perplexed} \tbSN \tbGE{to be drunk} \tbXV{O ya᷉a᷉ nyu la da᷉a᷉ bugi.} \tbXE{As usual, he has been drinking and is drunk.} \tbPD{bugee., bugiro, bugiro, bugiribo, bugiraa, - } 
\tbLX{bugi} \tbHM{} \tbPH{búgí} \tbPS{v.~} \tbGE{to pound into powder} \tbXV{Ba bugi a kyooma na tɔ ka᷉a᷉.} \tbXE{The shea nuts have been pounded into powder in order to prepare shea butter.} \tbPD{bugee, bugro, bugro, bugribo, bugraa, - } 
\tbLX{bugi} \tbHM{} \tbPH{búgí} \tbPS{v.~} \tbGE{to make fresh mounds (farming)} \tbXV{A koɔraa bugi la o weɛ yuoni ŋa azuiŋ o bɛŋɛ wɔŋɛɛ la soŋ.} \tbXE{The farmer made fresh mounds; consequently, he had a good yield of beans.} \tbPD{bugee, bugro, bugro, bugribo, bugraa, - } 
\tbLX{bugikpara} \tbHM{} \tbPH{bùgìkpárá} \tbPS{n.~} \tbGE{a moulded granary with two apartments} \tbOP{bugkparre} \tbTP{bugkparɛɛ} 
\tbLX{bugili} \tbHM{} \tbPH{búgílì} \tbPS{v.~} \tbGE{to tend (seed)} \tbPD{bugilee, bugilo, bugilo, buglibo, bugilaa, - } \tbVA{buuli/buoli}
\tbLX{bugiliŋe} \tbHM{} \tbPH{bùglíŋé} \tbPS{n.~} \tbGE{lid for a barn or granary} \tbOP{bugilinni} \tbTP{bugilinnee} 
\tbLX{bugimoɔ} \tbHM{} \tbPH{bùgìmʊ́ɔ́} \tbPS{n.~} \tbGE{a type of grass, the straw of which is mixed with clay to build a granary} \tbSG{bugimoɔnoo} \tbTP{bugimoɔnɛɛ} 
\tbLX{bugipelaa} \tbHM{} \tbPH{bùgìpɪ̀làá} \tbPS{n.~} \tbGE{a small granary} \tbOP{bugipeɛle} \tbTP{ bugipeɛlɛɛ} 
\tbLX{bugita} \tbHM{} \tbPH{búgítà} \tbPS{n.~} \tbGE{kettle} \tbOP{bugitare} \tbTP{ bugitarɛɛ} 
\tbLX{bugo} \tbHM{} \tbPH{bùgó} \tbPS{n.~} \tbGE{granary} \tbOP{bugri} \tbTP{bugree} 
\tbLX{bukulo} \tbHM{1} \tbPH{búkúló} \tbPS{adv.~} \tbGE{perish completely} \tbXV{ A vu᷉u᷉ di la  yiri zaa bukulo.} \tbXE{Fire has burnt the house completely.} 
\tbLX{bulee} \tbHM{} \tbPH{búléé} \tbPS{n.~} \tbGE{well} \tbOP{bulli} \tbTP{ bullee} 
\tbLX{buli} \tbHM{} \tbPH{búlí} \tbPS{v.~} \tbSN \tbGE{to germinate, to sprout, to grow out from} \tbXV{Ba bore la kyi kyɛ a naŋ ba buli sɛre.} \tbXE{They have sown millet, but it has not germinated yet.} \tbSN \tbGE{to spring up, to ooze (water)} \tbXV{Te tu la a bulee ka o sigi kyɛ o naŋ ba buli ko᷉ɔ᷉.} \tbXE{We have dug the well quite deep, but water has not started to flow.} \tbPD{bulee, bulo, bulo, bulibo, bulaa, bulibo } 
\tbLX{buli} \tbHM{} \tbPH{búlí} \tbPS{n.~} \tbGE{sprout} \tbOP{bulo} 
\tbLX{buliba} \tbHM{} \tbPH{búlíbà} \tbPS{v.~} \tbGE{to run fast} \tbPD{bulibaɛ, bulobaara, bulibaara, bulibaareba, bulibaaraa, - } 
\tbLX{bulinimiri} \tbHM{} \tbPH{bùlnímírì} \tbPS{n.~} \tbGE{the point in a well from which water flows out} \tbOP{bulinimie} \tbTP{bulinimiiree} 
\tbLX{bulli} \tbHM{} \tbPH{bùllì} \tbPS{v.~} \tbGE{to collect bits of water flowing from a well or hole (e.g. during draught)} \tbPD{bullee, bullo, bullo, bulibo, bullaa, bulibo } 
\tbLX{bulo} \tbHM{} \tbPH{bùlò} \tbPS{n.~} \tbGE{soup from seed or nuts (beans, groundnuts)} \tbTP{bulee/bullee} 
\tbLX{bulugiri} \tbHM{} \tbPH{bùlùgrí} \tbPS{n.~} \tbGE{a container made from a gourd} \tbOP{bulugo} \tbTP{bulugiree} \tbVA{buluuri}
\tbLX{buluuri} \tbHM{} \tbPH{bùlùùrí} \tbPS{n.~} \tbGE{a fat goat} \tbTP{buluuree.} \tbVA{bulugiri}
\tbLX{bumbunturi} \tbHM{} \tbPH{bùmbùntùrì} \tbPS{n.~} \tbGE{a type of tree with edible fruits} \tbOP{bumbunturri} \tbTP{bumbunturee} 
\tbLX{buŋi} \tbHM{} \tbPH{búŋ} \tbPS{v.~} \tbGE{to get mouldy (seed)} \tbPD{buŋee, buŋino, buŋino, buŋinimo, buŋinaa, buŋinimo } 
\tbLX{buo} \tbHM{} \tbPH{búò} \tbPS{pron.~} \tbGE{which one} \tbVA{buori}
\tbLX{buoli} \tbHM{} \tbPH{búólì} \tbPS{v.~} \tbGE{to nurse (a seedling)} \tbPD{buolee, buolo, buolo, buolibo, buolaa, buolibo } \tbVA{buuli/bugili}
\tbLX{buori} \tbHM{} \tbPH{bùòrì} \tbPS{v.~} \tbGE{to hurt} \tbPD{buoree, buoro, buoro, buoribo, buoraa, buoribo } 
\tbLX{buori} \tbHM{} \tbPH{bùórì} \tbPS{pron.~} \tbGE{which} \tbVA{buo/booː abuo/abobo}
\tbLX{buri} \tbHM{} \tbPH{bùrì} \tbPS{v.~} \tbGE{to soften by moistening} \tbXV{O buri la a gane a na to᷉ɔ᷉sɛ a woɔ.} \tbXE{He softened the leather so he could sew the bag.} \tbPD{buree, buro, buro, buribo, buraa, buribo } 
\tbLX{buri} \tbHM{} \tbPH{bùrì} \tbPS{v.~} \tbGE{to gain weight} \tbXV{Beri ayi ŋa a dɔɔ ŋmeene na leɛ buri la o zaa.} \tbXE{Lately, that lean man has been putting on weght.} \tbPD{buree, buro, buro, buribo, buraa, buribo } 
\tbLX{buri} \tbHM{} \tbPH{búrí} \tbPS{v.~} \tbGE{to immerse a container in liquid or powder to fill it} \tbPD{buree, buro, buro, buribo, buraa, buro } 
\tbLX{buriburi} \tbHM{} \tbPH{bùrìbúrí} \tbPS{v.~} \tbGE{to be confused} \tbPD{buriburee, buriburo, buriburo, buriburibo, buriburaa, buriburibo } 
\tbLX{buriburi} \tbHM{} \tbPH{búríbúrí} \tbPS{n.~} \tbGE{state of confusion} 
\tbLX{burimo} \tbHM{} \tbPH{bùrìmó} \tbPS{n.~} \tbGE{a type of wild animal the size of a zebra} \tbOP{burri} \tbTP{burree} 
\tbLX{buro} \tbHM{} \tbPH{bùró} \tbPS{adj.~} \tbGE{in dry powdered form (flour)} \tbXV{O nɔŋ la zɔmburo mogebo.} \tbXE{He likes eating unsoaked flour.} \tbOP{burri} \tbTP{burree} 
\tbLX{buroni} \tbHM{} \tbPH{bùròní} \tbPS{n.~} \tbGE{white man} \tbOP{buroniri} \tbTP{buroniree} 
\tbLX{buroŋ-buroŋ} \tbHM{} \tbPH{búróŋ-búróŋ} \tbPS{adv.~} \tbGE{done haphazardly (without order)} \tbXV{A yɛlɛ deɛ waa la buromburoŋ lɛ.} \tbXE{The case is just in a confused state.} \tbVA{/bùròmbùròŋ/}
\tbLX{burri} \tbHM{} \tbPH{búrrì} \tbPS{v.~} \tbGE{to fool, to outwit} \tbXV{imperf. .perf. . agent:} \tbXE{} \tbPD{burree, burro, burro, buribo, burraa, burro } 
\tbLX{buseli} \tbHM{} \tbPH{búsélí} \tbPS{n.~} \tbGE{very heavy object} \tbOP{busele} \tbTP{ buselee} 
\tbLX{buulaa} \tbHM{} \tbPH{bùùláá} \tbPS{n.~} \tbGE{playground} \tbOP{buuli} \tbTP{buulee} 
\tbLX{buuli} \tbHM{} \tbPH{bùùlí} \tbPS{n.~} \tbGE{an insect that makes honey in the ground} \tbOP{buulo} \tbTP{buulee} \tbVA{wirikuoluŋ}
\tbLX{buuli} \tbHM{} \tbPH{búúlí} \tbPS{n.~} \tbGE{bells worn on the wrist to play xylophone} \tbOP{buulo} \tbTP{buulee} 
\tbLX{buuli} \tbHM{} \tbPH{búúlì} \tbPS{v.~} \tbGE{nurse (seed)} \tbXV{O buuli la kommie a dabɔge poɔ.} \tbXE{She has nursed tomoto seed in the garden.} \tbPD{buulee, buulo, buulo, buulibo, buulaa, buulibo } \tbVA{buoli/bugili}
\tbLX{buuli} \tbHM{} \tbPH{bùùlì} \tbPS{v.~} \tbGE{to boil (first stage in making porridge paste; saabo)} \tbPD{buulee, buulo, buulo, buulibo, buulaa, buulo } 
\tbLX{buuli} \tbHM{} \tbPH{bùùlì} \tbPS{v.~} \tbGE{to murmur, to grumble} \tbXV{Ka o poɔŋ sa᷉a᷉ o maŋ buuli la yɛlɛ yaga.} \tbXE{When she is sad, she grumbles a lot.} \tbPD{buulee, buulo, buulo, buulibo, buulaa, buulo } 
\tbLX{buulo} \tbHM{} \tbPH{búúló} \tbPS{n.~} \tbGE{something that complements or adorns another thing} \tbXV{Nɔtoro e la dɔɔ noɔre buulo.} \tbXE{A mustache complements a man's mouth.} 
\tbLX{buulunwala} \tbHM{} \tbPH{bùùlùnwálá} \tbPS{n.~} \tbGE{light porridge} \tbTP{buulunwalɛɛ} 
\tbLX{buulunwoɔlɔ} \tbHM{} \tbPH{bùùlùnwʊ́ɔ́lɔ́} \tbPS{n.~} \tbGE{porridge used in the preparation of TZ} \tbTP{buulunwoɔlɛɛ} \tbVA{buulunwoɔla}
\tbLX{buuluŋ} \tbHM{} \tbPH{búúlúŋ} \tbPS{n.~} \tbGE{porridge} \tbTP{buulunee} 
\tbLX{buuluŋ} \tbHM{} \tbPH{bùúlúŋ} \tbPS{n.~} \tbGE{fleshy portion (meat)} \tbOP{buulo} \tbTP{buulee} 
\tbLX{da} \tbHM{} \tbPH{dà} \tbPS{part.~} \tbGE{past tense marker} \tbXV{Ba da wa la kyɛ.} \tbXE{They came here.} 
\tbLX{da} \tbHM{} \tbPH{dà} \tbPS{n.~} \tbGE{time in the past; olden days} 
\tbLX{da} \tbHM{} \tbPH{dá} \tbPS{v.~} \tbSN \tbGE{to push (pl)} \tbSN \tbGE{to blow (wind)} \tbSN \tbGE{to wander (in the bush)} \tbXV{A naŋkpaana da la moɔ yaga kyɛ ba nyɛ nɛne.} \tbXE{The hunter wandered all day but did not get any game.} \tbPD{daɛ, daara, daara, daareba, daaraa, daareba } \tbSE{da lɔɔ} \tbGE{to push or blow down} 
\tbLX{da} \tbHM{} \tbPH{dà} \tbPS{v.~} \tbGE{to buy} \tbPD{daɛ, daara, daara, dareba, daaraa, daareba } 
\tbLX{daa} \tbHM{} \tbPH{dàá} \tbPS{n.~} \tbGE{stick, piece of wood} \tbOP{daare} \tbTP{darɛɛ} 
\tbLX{daa} \tbHM{} \tbPH{dàá} \tbPS{n.~} \tbGE{market; place for buying and selling} \tbXV{Ba maŋ ti la sampare daa poɔ.} \tbXE{Sheds are often put up at the marketplace.} \tbSN \tbGE{day or occasion for buying and selling} \tbXV{Wa daa ane Kaane daa la maŋ laŋ kpɛ.} \tbXE{The Wa market day and Kaane market day fall on the same day.} \tbOP{daare} \tbTP{darɛɛ} 
\tbLX{daa} \tbHM{} \tbPH{dàá} \tbPS{n.~} \tbGE{week (six-day)} \tbXV{O naŋ wa kyɛ, o daare ayi la.} \tbXE{It is two weeks since he came here.} \tbOP{daare} \tbTP{darɛɛ} 
\tbLX{daa} \tbHM{} \tbPH{dàá} \tbPS{n.~} \tbGE{cost, price} \tbXV{Se᷉ɛ᷉ saŋa bondirii daa maŋ do la.} \tbXE{The cost of food rises during the middle of the farming season.} \tbOP{daare} \tbTP{darɛɛ} \tbVA{daaroŋ}
\tbLX{daa} \tbHM{} \tbPH{dàá} \tbPS{n.~} \tbGE{gun} \tbOP{daare} \tbTP{darɛɛ} 
\tbLX{daa} \tbHM{} \tbPH{dáá} \tbPS{adv.~} \tbGE{last time} \tbXV{O daa wa la kyɛ?} \tbXE{Did s/he come here last time?.} \tbVA{daare}
\tbLX{daa} \tbHM{} \tbPH{dáà} \tbPS{v.~} \tbGE{to push} \tbPD{daaɛ, daara, daara, dareba, daaraa, dareba } 
\tbLX{daa} \tbHM{} \tbPH{dáá} \tbPS{adj.~} \tbGE{male (nonhuman)} \tbOP{daare/daɛ} \tbTP{darɛɛ/daarɛɛ} 
\tbLX{da᷉a} \tbHM{} \tbPH{dá᷉à᷉} \tbPS{n.~} \tbSN \tbGE{brewed drink, alcoholic beverage (pito)} \tbSN \tbGE{container of drink} \tbSG{daamo} \tbOP{daanɛɛ} 
\tbLX{dâabeene} \tbHM{} \tbPH{dá᷉á᷉bɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{fermented pito} \tbOP{dâabeemɛ} 
\tbLX{dâabeerɛ} \tbHM{} \tbPH{dá᷉á᷉bɪ̀ɪ̀rɛ́} \tbPS{n.~} \tbGE{second stage in brewing pito} \tbTP{dâabeerɛɛ} 
\tbLX{dâabele} \tbHM{} \tbPH{dá᷉ábɪ́lɪ́} \tbPS{n.~} \tbGE{pito yeast} \tbOP{dâabelɛ} \tbTP{dâabelɛɛ} 
\tbLX{dâabinnaa} \tbHM{} \tbPH{dá᷉á᷉bìnnáà} \tbPS{n.~} \tbGE{pito reserved for a regular customer} \tbOP{dâabinni} \tbTP{dâabinnee} 
\tbLX{da᷉abiri} \tbHM{} \tbPH{dá᷉á᷉bírí} \tbPS{n.~} \tbGE{residue from pito (local beer)} \tbTP{da᷉abiree} 
\tbLX{dâabiri} \tbHM{} \tbPH{dá᷉á᷉bírì} \tbPS{n.~} \tbGE{chaff of pito malt} \tbTP{dâabiree/dâabirree.} 
\tbLX{daadaa} \tbHM{} \tbPH{dáádáá} \tbPS{adv.~} \tbGE{straight ahead} 
\tbLX{dâada} \tbHM{} \tbPH{dá᷉á᷉dà} \tbPS{n.~} \tbGE{the act of buying alcohol} 
\tbLX{daa-daa} \tbHM{} \tbPH{dàá-dàá} \tbPS{adv.~} \tbGE{weekly; from time to time} \tbXV{Daa-daa ka a dɔɔta maŋ wa.} \tbXE{The doctor comes every week.} 
\tbLX{dâadaara} \tbHM{} \tbPH{dá᷉á᷉dáárá} \tbPS{n.~} \tbGE{one who buys alcohol} \tbOP{dâadareba} \tbTP{dâadarebɛɛ} 
\tbLX{dâadagoli} \tbHM{} \tbPH{dá᷉á᷉dàgólí} \tbPS{n.~} \tbGE{one who loves drinking alcholic beverages} \tbOP{dâadagolo} \tbTP{dâadagolee} 
\tbLX{dâadie} \tbHM{} \tbPH{dá᷉á᷉dìè} \tbPS{n.~} \tbGE{pito shop} \tbOP{dâaderi} \tbTP{dâaderee} 
\tbLX{dâadoge} \tbHM{} \tbPH{dá᷉á᷉dʋ́gɪ́} \tbPS{n.~} \tbGE{pito pot or pot of pito} \tbOP{dâadoggre} \tbTP{dâadogrɛɛ} 
\tbLX{dâadogerɔ} \tbHM{} \tbPH{dá᷉á᷉dʊ́grɔ́} \tbPS{n.~} \tbGE{brewer, one who brews alcohol} \tbOP{dâadogerebɔ} \tbTP{dâadogerebɛɛ} 
\tbLX{da᷉afaare} \tbHM{} \tbPH{dá᷉áfáárɪ́} \tbPS{n.~} \tbGE{bad alcoholic beverage} \tbTP{da᷉a᷉faarɛɛ} 
\tbLX{da᷈afaraa} \tbHM{} \tbPH{dá᷈áfáráá} \tbPS{n.~} \tbGE{well-fermented pito} \tbOP{dââfarre} \tbTP{dââfarrɛɛ} 
\tbLX{da᷉agyemɛ} \tbHM{} \tbPH{dá᷉ágyɪ́mɛ́} \tbPS{n.~} \tbGE{pito soda} 
\tbLX{dâagyemɛ} \tbHM{} \tbPH{dá᷉á᷉̂gyɪ̀mɛ̀} \tbPS{n.~} \tbGE{soda from pito brewing} \tbTP{dâagyemɛɛ/dâagyennɛɛ} 
\tbLX{dâakaraa} \tbHM{} \tbPH{dá᷉á᷈káráá} \tbPS{n.~} \tbGE{unfermented pito} \tbOP{dâakare} \tbTP{ dâakarɛɛ.} 
\tbLX{dâakonni} \tbHM{} \tbPH{dá᷉á᷉̂kònnì} \tbPS{n.~} \tbGE{pito that is fermenting} 
\tbLX{da᷉akoɔre} \tbHM{} \tbPH{dá᷉á᷉kʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{residue of pito from the last brewing stage} \tbTP{da᷉akoɔrɛɛ} 
\tbLX{dâakɔgere} \tbHM{} \tbPH{dá᷉á᷉̂kɔ́grɪ́} \tbPS{n.~} \tbGE{the first stage of pito brewing} \tbTP{dâakɔgerɛɛ} 
\tbLX{da᷉akuoli} \tbHM{} \tbPH{dá᷉á᷉kúólì} \tbPS{n.~} \tbGE{drunkenness, condition of being drunk} 
\tbLX{da᷈akuolo} \tbHM{} \tbPH{dá᷈á᷈kúóló} \tbPS{n.~} \tbGE{drunk} \tbOP{da᷈akuolibo} \tbTP{da᷈akuolibee} 
\tbLX{daale} \tbHM{} \tbPH{dààlɪ̀} \tbPS{v.~} \tbGE{to float; to be suspended (on liquid)} \tbXV{Ka fooŋ kyiri ka᷉a᷉ eŋ ko᷉ɔ᷉ poɔ a maŋ iri daale la.} \tbXE{If oil is poured into water it will float on the top.} \tbPD{daalɛɛ, daala, daala, daaleba, daalaa, daaleba } 
\tbLX{daama} \tbHM{} \tbPH{dáámá} \tbPS{n.~} \tbGE{state of ease; freedom} \tbXV{Ka a toma ŋaŋ wa baare N na nyɛ la daama fẽẽ.} \tbXE{When this work is over I shall be free a bit.} \tbTP{daamɛɛ} 
\tbLX{Daana} \tbHM{} \tbPH{dááná} \tbPS{n.~} \tbGE{God} 
\tbLX{daana} \tbHM{} \tbPH{dááná} \tbPS{n.~} \tbGE{owner, lord, master} \tbOP{daanema} \tbTP{daanemɛɛ} \tbVA{daŋena}
\tbLX{daane} \tbHM{} \tbPH{dàànɪ̀} \tbPS{v.~} \tbSN \tbGE{to change, to make different} \tbXV{O naŋ gaa o sori na te  waana la ka o daane (la) o zaa.} \tbXE{Returning from his trip he had changed completely.} \tbSN \tbGE{to steer, to redirect, to turn away} \tbXV{O poɔ sa᷉aɛ la ka o daane o niŋe bare ziyuo.} \tbXE{She was sad and turned her face away.} \tbPD{daanɛɛ, daana, daana, daanema, daanaa, daanema } 
\tbLX{daanyu} \tbHM{} \tbPH{dá᷉á᷉̂ɲú} \tbPS{n.~} \tbGE{the act of drinking alcohol} 
\tbLX{daanyuuro} \tbHM{} \tbPH{dá᷉á᷉́̂ɲúúrò} \tbPS{n.~} \tbSN \tbGE{an alcoholic, a habitual drinker of alcohol} \tbSN \tbGE{a person who is drinking alcohol} \tbOP{daanyuuribo} \tbTP{daanyuribee} 
\tbLX{da᷉aŋmane} \tbHM{} \tbPH{dá᷉á᷉ŋmánɪ́} \tbPS{n.~} \tbSN \tbGE{calabash used for drinking pito} \tbSN \tbGE{a calabash of pito} \tbOP{da᷉aŋmama} \tbTP{da᷉aŋmamɛɛ} 
\tbLX{da᷉apare} \tbHM{} \tbPH{dá᷉á᷉párɪ́} \tbPS{n.~} \tbGE{residue of pito from the pot} \tbTP{da᷉aparɛɛ} 
\tbLX{da᷈apelaa} \tbHM{} \tbPH{dá᷈á᷈pɪ̀làá} \tbPS{n.~} \tbGE{palm wine} \tbOP{da᷈apeɛle} \tbTP{da᷈apeɛlɛɛ} 
\tbLX{dâapɔge} \tbHM{} \tbPH{dá᷉á᷉̂́pɔ́gɪ́} \tbPS{n.~} \tbGE{ pito brewer or seller} \tbOP{dâapɔgeba} \tbTP{dâapɔgebɛɛ/dâapɔgrɛɛ} 
\tbLX{daare} \tbHM{} \tbPH{dààrɪ̀} \tbPS{v.~} \tbGE{to get out of control. (bad behaviour)} \tbXV{Ka bibileŋ ba nyɛ wuluu velaa o bile saŋa, o maŋ daare la.} \tbXE{If a child does not get good guidance during his early life, he gets out of hand.} \tbPD{daarɛɛ, daara, daara, daareba, daaraa, daareba } 
\tbLX{daare} \tbHM{} \tbPH{dààrɪ́} \tbPS{n.~} \tbGE{firewood (collective)} \tbTP{daarɛɛ} 
\tbLX{daare} \tbHM{} \tbPH{dáárɪ́} \tbPS{adv.~} \tbSN \tbGE{day before yesterday;} \tbSN \tbGE{last time} \tbVA{daa}
\tbLX{daaroŋ} \tbHM{} \tbPH{dààrʋ́ŋ} \tbPS{n.~} \tbGE{cost of} \tbXV{Uoni saŋa kommie daaroŋ maŋ e la kpeɛŋaa.} \tbXE{During the dry season the cost of tomatoes goes high.} 
\tbLX{dâasenyoɔle} \tbHM{} \tbPH{dá᷉á᷉sɪ́ɲʋ́ɔ́lɪ́} \tbPS{n.~} \tbSN \tbGE{ sugar ant} \tbSN \tbGE{one who likes drinking alcohol} \tbOP{dâasenyoɔlɔ/da᷉asenyoɔla} \tbTP{dâasenyoɔlɛɛ} 
\tbLX{dâa-so} \tbHM{} \tbPH{dá᷉á᷉-sò} \tbPS{n.~} \tbGE{ a ritual that involves bathing a corpse with pito before burial} 
\tbLX{da᷉asunni} \tbHM{} \tbPH{dá᷉á᷉súnní} \tbPS{n.~} \tbGE{sticky substance from pito} \tbSG{da᷉asunnuu} \tbTP{da᷉asunnee} 
\tbLX{da᷉asuoro} \tbHM{} \tbPH{dá᷉á᷉súóró} \tbPS{n.~} \tbGE{a corpse that is to be bathed with pito before burial} \tbOP{da᷉asuoribo} \tbTP{da᷉asuoribee} 
\tbLX{da᷉atambalaaraa} \tbHM{} \tbPH{dá᷉á᷉támbálááráá} \tbPS{n.~} \tbGE{sweet pito} \tbTP{da᷉atambalaarɛɛ} 
\tbLX{da᷈ateɛ} \tbHM{} \tbPH{dá᷈á᷈tɪ́ɛ́} \tbPS{n.~} \tbGE{grapevine} \tbOP{da᷈ateere} \tbTP{dâateerɛɛ} 
\tbLX{dâatuo} \tbHM{} \tbPH{dá᷉á᷉túó} \tbPS{n.~} \tbGE{pito meant for a ritual} \tbOP{dâatuuri} \tbTP{dâatuuree} 
\tbLX{daayi} \tbHM{} \tbPH{dááyí} \tbPS{v.~} \tbGE{to stand out} \tbPD{daa yiee, daara yire, daayire, daayiribe, daayiraa, - } 
\tbLX{da᷉azeɛ} \tbHM{} \tbPH{dá᷉á᷉zìɛ́} \tbPS{n.~} \tbGE{locally brewed beer (pito)} \tbTP{da᷉azerɛɛ/da᷉azeerɛɛ} 
\tbLX{dâazu} \tbHM{} \tbPH{dá᷉á᷉zú} \tbPS{n.~} \tbGE{the first pito to be poured from the pot} \tbOP{dâazuri} \tbTP{dâazuree} 
\tbLX{da᷉azukpala} \tbHM{} \tbPH{dá᷉á᷉zúkpálá} \tbPS{n.~} \tbGE{pito without yeast froth} \tbTP{da᷉azukpalɛɛ} \tbVA{da᷉a᷉zukpoli}
\tbLX{dabaala} \tbHM{} \tbPH{dàbáálà} \tbPS{n.~} \tbGE{effort, struggle} \tbTP{dabaalɛɛ} \tbVA{dabagla}
\tbLX{dabaaraa} \tbHM{} \tbPH{dàbááráá} \tbPS{n.~} \tbGE{a three-forked stick on which pots or calabashes are placed} \tbOP{dabaare} \tbTP{dabaarɛɛ.} 
\tbLX{dabagelakala} \tbHM{} \tbPH{dàbáglàkálà} \tbPS{interj.~} \tbGE{an expression of surprise at something great} \tbXV{A pɔge to᷉ɔ lɔɔ la a dɔɔ, dabagelakala.} \tbXE{The woman was able to throw the man down; that's great!.} 
\tbLX{dabaŋkalɛ} \tbHM{} \tbPH{dàbàŋkálɛ́} \tbPS{n.~} \tbGE{looking back with dissatisfaction; had I known, regret, remorse} \tbXV{Selaa de zeɛ la a nɔlee kaŋa, ka N da baŋ ka lɛ N pɔge a.} \tbXE{A hawk has snatched one of the chickens; if I had known, I would have put them in a pen.} 
\tbLX{dabaŋkoroŋ} \tbHM{} \tbPH{dàbàŋkʊ́rʊ́ŋ} \tbPS{n.~} \tbGE{prior knowledge} \tbXV{A seŋ ka karemamine maŋ de karembiiri dabaŋkoroŋ wuli ba.} \tbXE{Teachers should base their teaching on pupils' prior knowledge.} 
\tbLX{dabe᷈e} \tbHM{} \tbPH{dàbɪ́᷉ɪ́᷉} \tbPS{n.~} \tbSN \tbGE{juice} \tbSN \tbGE{semen} \tbOP{dabeene} \tbTP{dabeemɛɛ/dabeenɛɛ} 
\tbLX{dabe᷈ɛ} \tbHM{} \tbPH{dàbɪ́᷈ɛ́} \tbPS{n.~} \tbSN \tbGE{fear} \tbSN \tbGE{danger, fearsome} \tbXV{A vu᷉uzele taa la dabe᷉ɛ.} \tbXE{The burning flame is looking dangerous.} \tbOP{dabeɛmɛ} \tbTP{dabeɛnɛɛ} 
\tbLX{dabeɛmɛ} \tbHM{} \tbPH{dàbɪ̀ɛ́mɛ̀} \tbPS{n.~} \tbGE{coward} \tbXV{Nyɛ, a dɔɔŋ leɛ zoro, dabeɛmɛ la.} \tbXE{Look, the man is running away, he is a coward.} \tbTP{dabeɛmɛɛ} 
\tbLX{dabegere} \tbHM{} \tbPH{dábɪ́gɪ́rɪ́} \tbPS{n.~} \tbSN \tbGE{piece of, portion} \tbSN \tbGE{paragraph} \tbOP{dabegɛ} \tbTP{dabegerɛɛ} 
\tbLX{dabegere} \tbHM{} \tbPH{dábɪ́gɪ́rɪ́} \tbPS{n.~} \tbSN \tbGE{a type of white ant} \tbSN \tbGE{the nest of such ants} \tbOP{dabegɛ} \tbTP{dabegerɛɛ} 
\tbLX{dabilikpeo} \tbHM{} \tbPH{dàbìlkpɪ́ʋ́} \tbPS{n.~} \tbGE{a type of tree with medicinal properties and edible fruits} 
\tbLX{dabilkaŋkanyogelɔ} \tbHM{} \tbPH{dàbìlkáŋkáɲʊ́glɔ́} \tbPS{n.~} \tbGE{small twigs used as firewood, kindling} \tbTP{dabilkaŋkanyoglɛɛ} \tbVA{dabilkaŋkanyogla}
\tbLX{dabiltuo} \tbHM{} \tbPH{dàbìltúó} \tbPS{n.~} \tbGE{a bitter herb used to treat abdominal problems} \tbOP{dabiltuuri} \tbTP{dabiltuuree.} 
\tbLX{dabodabo} \tbHM{} \tbPH{dábʋ́dábʋ́} \tbPS{n.~} \tbGE{duck} \tbOP{dabodabore} \tbTP{dabodaborɛɛ.} 
\tbLX{dabogi} \tbHM{} \tbPH{dàbògí} \tbPS{n.~} \tbGE{hole in a tree} \tbOP{dabogri} \tbTP{dabogree.} 
\tbLX{dabolore} \tbHM{} \tbPH{dàbʋ́lʋ̀rɪ̀} \tbPS{n.~} \tbGE{a type of tree with medicinal properties and edible fruits} \tbTP{dabolorɛɛ} 
\tbLX{daboorabɔge} \tbHM{} \tbPH{dàbʊ̀ʊ̀rábɔ́gɪ́} \tbPS{n.~} \tbGE{praying mantis} \tbOP{daboorabɔgere} \tbTP{daboorabɔgerɛɛ} 
\tbLX{daboze᷈e} \tbHM{} \tbPH{dàbòzɪ́᷉ɪ́᷉̂} \tbPS{n.~} \tbGE{a type of small brown bird} \tbOP{dobazeene} \tbTP{dobazeenɛɛ} 
\tbLX{dabɔge} \tbHM{} \tbPH{dàbɔ̀gɪ́} \tbPS{n.~} \tbGE{a type of large bird (carnivorous)} \tbOP{dabɔgere} \tbTP{dabɔgerɛɛ} 
\tbLX{dabɔgɔ} \tbHM{} \tbPH{dàbɔ́gɔ̀} \tbPS{n.~} \tbGE{fenced enclosure; garden} \tbOP{dabɔgre} \tbTP{dabɔgrɛɛ} \tbVA{dabɔga/dabɔge}
\tbLX{dabɔle} \tbHM{} \tbPH{dàbɔ́lɪ́} \tbPS{n.~} \tbGE{round smooth stick, club} \tbOP{dabɔlɔ} \tbTP{dabɔlɛɛ} 
\tbLX{dabɔlmakasogi} \tbHM{} \tbPH{dábɔ́lmákásògí} \tbPS{n.~} \tbGE{a type of tall straw hat} \tbOP{dabɔlmakasogiri} \tbTP{dabɔlmakasogiree} \tbVA{dabɔlkasogi/dabɔlkosogi}
\tbLX{Dabuo} \tbHM{} \tbPH{dàbúó} \tbPS{n.~} \tbGE{name traditionally given to one born after the death of the father} \tbOP{Dabumine} \tbTP{Dabuuree} 
\tbLX{dabuo} \tbHM{} \tbPH{dàbúó} \tbPS{n.~} \tbGE{abandoned house} \tbOP{dabori} \tbTP{daboree} 
\tbLX{̀dabuori} \tbHM{} \tbPH{dàbúóri} \tbPS{interrogative pronoun.~} \tbGE{when?} \tbXV{Dabuori ka fo tu fo sori na?} \tbXE{When will you make your trip?.} 
\tbLX{dadagere} \tbHM{} \tbPH{dàdàgɪ́rɪ́} \tbPS{n.~} \tbGE{a kind of fish} \tbOP{dadaga} \tbTP{dadagerɛɛ} 
\tbLX{daduŋi} \tbHM{} \tbPH{dàdúŋí} \tbPS{n.~} \tbGE{wood weevil} \tbOP{dadunni} \tbTP{dadunnee} 
\tbLX{daga} \tbHM{} \tbPH{dágà} \tbPS{n.~} \tbGE{box} \tbOP{dagere} \tbTP{dagerɛɛ} 
\tbLX{Dagaare} \tbHM{} \tbPH{dàgàárɪ̀} \tbPS{n.~} \tbGE{Dagaare; a language of the Moore-Gurma family} \tbTP{dagaarɛɛ} \tbVA{Dagara}
\tbLX{dagaareŋ} \tbHM{} \tbPH{dàgààrɪ̀ŋ} \tbPS{adj.~} \tbGE{done according to the practices of the Dagaaba} \tbXV{Ziyuo vɛŋɛɛ la ka Dagaareŋ deɛ bɔrɔ.} \tbXE{Civilisation is killing most of the traditional practices of the Dagaaba.} 
\tbLX{dagalaa} \tbHM{} \tbPH{dàgàláà} \tbPS{n.~} \tbGE{a large, long plank used to support weak structures} \tbOP{dagalle} \tbTP{dagallɛɛ} 
\tbLX{dagaŋgaanaa} \tbHM{} \tbPH{dàgàŋgàànáà} \tbPS{n.~} \tbGE{large vine of a creeping or climbing plant} \tbOP{dagaŋgaane} \tbTP{dagaŋgaanɛɛ} \tbVA{gaŋgaanaa}
\tbLX{dagaŋgaŋ} \tbHM{} \tbPH{dàgàŋgáŋ} \tbPS{n.~} \tbGE{wit} \tbXV{A bie taa la dagaŋgaŋ, fo na yeli ka so᷉ɔ.} \tbXE{The child is so full of wit that it might be attributed to witchraft.} \tbOP{dagaŋganne} \tbTP{dagaŋgannɛɛ} 
\tbLX{Dagao} \tbHM{} \tbPH{dàgáʋ̀} \tbPS{n.~} \tbSN \tbGE{native speaker of Dagaare} \tbSN \tbGE{the homeland of the speakers of Dagaare} \tbOP{Dagaaba} \tbTP{Dagaabɛɛ} \tbVA{Dagara}
\tbLX{dagaraa} \tbHM{} \tbPH{dàgàráà} \tbPS{n.~} \tbSN \tbGE{cross; symbol of Christianity} \tbSN \tbGE{crossbeam} \tbOP{dagarre} \tbTP{dagarrɛɛ} 
\tbLX{Dagara (ND)} \tbHM{} \tbPH{dàgàrà} \tbPS{n.~} \tbGE{the Dagaare language} \tbVA{Dagao}
\tbLX{dageliŋe} \tbHM{} \tbPH{dágelíŋé} \tbPS{n.~} \tbGE{lid of a box} \tbOP{dagelinni} \tbTP{daglinnee.} 
\tbLX{dagoli} \tbHM{} \tbPH{dagoli} \tbPS{n.~} \tbGE{cudgel; a throwing stick (usually hook-shaped at one end)} \tbOP{dagolo} \tbTP{dagolee/dagollee} 
\tbLX{daguo} \tbHM{} \tbPH{dàgùó} \tbPS{n.~} \tbGE{club with a knobbed end; a short, large-headed piece of wood used for breaking mud-built walls and lumps of earth} \tbOP{dagori} \tbTP{dagoree/dagorree.} \tbVA{guo}
\tbLX{Dagbɛmmo} \tbHM{} \tbPH{dàgbɛ́mmʊ́} \tbPS{n.~} \tbGE{a native speaker of the Dagbaane language} \tbOP{Dagbɛnne} \tbTP{Dagbɛnnɛɛ} 
\tbLX{Dagbɛnne} \tbHM{} \tbPH{Dàgbɛ́nnɪ́} \tbPS{n.~} \tbGE{Dagbaane language} \tbTP{dagbɛnnɛɛ} 
\tbLX{Dagbɛnnoo} \tbHM{} \tbPH{dàgbɛ́nnʋ́ʋ́} \tbPS{n.~} \tbGE{a native speaker of the Dagbaane language} \tbOP{Dagbɛnne} \tbTP{Daggbɛnnɛɛ.} \tbVA{Dagbɛmmo}
\tbLX{dagboli} \tbHM{} \tbPH{dàgbólí} \tbPS{n.~} \tbGE{large piece of wood (firewood)} \tbOP{dagbolo} \tbTP{dagbolee} 
\tbLX{dagbollo} \tbHM{} \tbPH{dàgbólló} \tbPS{n.~} \tbGE{the fifth day from now} \tbOP{dagbolli} \tbTP{dagbollee} 
\tbLX{dagbuli} \tbHM{} \tbPH{dàgbùlí} \tbPS{n.~} \tbGE{the act of buying as a whole} \tbOP{dagbulo} \tbTP{dagbullee} 
\tbLX{dagbulo} \tbHM{} \tbPH{dàgbùló} \tbPS{n.~} \tbGE{appellation of a kinship group} 
\tbLX{dagyu} \tbHM{} \tbPH{dàdzú} \tbPS{n.~} \tbGE{a type of large, hairy spider; believed to be poisonous} \tbOP{dagyuri} \tbTP{dagyuree.} 
\tbLX{daha} \tbHM{} \tbPH{dáhá} \tbPS{n.~} \tbGE{a type of game similar to chess} \tbOP{dahare} \tbVA{dara}
\tbLX{dakala} \tbHM{} \tbPH{dàkálá} \tbPS{n.~} \tbGE{dried sperm} \tbTP{dakallɛɛ} 
\tbLX{dakogi} \tbHM{} \tbPH{dàkógí} \tbPS{n.~} \tbGE{chair, stool} \tbOP{dakogiri} \tbTP{dakogiree} 
\tbLX{dakogidɛllaa} \tbHM{} \tbPH{dàkògìdɛ́lláá} \tbPS{n.~} \tbGE{chair with a back} \tbOP{dakogidɛlle} 
\tbLX{dakogilee} \tbHM{} \tbPH{dàkògìléé} \tbPS{n.~} \tbGE{small stool} \tbOP{dakogibilii} \tbTP{dakogibilee} 
\tbLX{dakoɔre} \tbHM{} \tbPH{dàkʋ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{bachelor} \tbOP{dakɔbɔ} \tbTP{dakoɔrɛɛ.} 
\tbLX{dakoɔroŋ} \tbHM{} \tbPH{dàkʊ́ɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{celibacy} 
\tbLX{Dakoraa} \tbHM{} \tbPH{dàkʋ́ráá} \tbPS{n.~} \tbGE{name traditionally given to a reincarnated male child} \tbOP{dakormine} \tbTP{dakorɛɛ/dakorrɛɛ.} 
\tbLX{dakoraa} \tbHM{} \tbPH{dàkòràá} \tbPS{n.~} \tbGE{walking stick with a curved top} \tbOP{dakorri} \tbTP{dakoree/dakorree} 
\tbLX{dakoroŋ} \tbHM{} \tbPH{dàkʊ̀rʊ̀ŋ} \tbPS{n.~} \tbGE{former times; olden days} \tbVA{dasaŋa}
\tbLX{dakɔɔre} \tbHM{} \tbPH{dàkɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{armlet made from enamel} \tbOP{dakɔbɔ} \tbTP{dakɔɔrɛɛ} 
\tbLX{dakɔɔre} \tbHM{} \tbPH{dàkɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{cob (maize)} \tbOP{dakɔbɔ} \tbTP{dakɔɔrɛɛ} 
\tbLX{dakuoŋaa} \tbHM{} \tbPH{dàkúóŋáá} \tbPS{n.~} \tbGE{dry or dead wood} \tbOP{dakuoni/dakuomo} \tbTP{dakuonee} 
\tbLX{Dakpaa} \tbHM{} \tbPH{dàkpàá} \tbPS{n.~} \tbSN \tbGE{name of a village} \tbSN \tbGE{a kinship group} 
\tbLX{dakpaa} \tbHM{} \tbPH{dàkpáà} \tbPS{n.~} \tbGE{a type of wasp} \tbOP{dakpaare} \tbTP{dakpaarɛɛ} 
\tbLX{Dakpaaloo} \tbHM{} \tbPH{dàkpáálʊ́ʊ́} \tbPS{n.~} \tbGE{member of the Dakpaa kinship group} \tbOP{Dakpaalɛɛ} 
\tbLX{dakpalori} \tbHM{1} \tbPH{dàkpàlórí} \tbPS{n.~} \tbGE{a type of small bug that sucks blood (usually found in crevices of walls)} \tbSG{dakpaloruu} \tbOP{dakpaloree} \tbTP{dakpalorree} 
\tbLX{dakpele} \tbHM{} \tbPH{dàkpɪ́lɪ́} \tbPS{n.~} \tbGE{hard wood} \tbOP{dakpelɛ} \tbTP{dakpelɛɛ} 
\tbLX{dakpiri} \tbHM{} \tbPH{dàkpíri} \tbPS{n.~} \tbGE{stump} \tbSG{dakpirruu} \tbOP{dakpirri} \tbTP{dakpiree/dakpirree} \tbVA{dakpaa}\tbVA{daŋmaparekpiri}
\tbLX{Dakpɔge} \tbHM{} \tbPH{dàkpɔ́gɪ́} \tbPS{n.~} \tbGE{a kinship group} \tbOP{dakpɔgɔ} 
\tbLX{dakyage} \tbHM{} \tbPH{dàʧàgɪ́} \tbPS{n.~} \tbGE{lazy chair (for men), deckchair or sun lounger} \tbOP{dakyagre} \tbTP{dakyagrɛɛ} 
\tbLX{dakyara} \tbHM{} \tbPH{dàʧárà} \tbPS{n.~} \tbGE{forked piece of wood} \tbOP{dakyɛrre} \tbTP{dakyɛrrɛɛ} 
\tbLX{Dakyeeloo} \tbHM{} \tbPH{dàʧɪ́ɪ́lʋ́ʋ́} \tbPS{n.~} \tbGE{member of some type of kinship (Dakyeɛ) group} \tbOP{dakyeelɛɛ} 
\tbLX{dakyeerekpaglaga} \tbHM{} \tbPH{dàʧɪ̀ɪ̀rɪ̀kpàglágá} \tbPS{n.~} \tbGE{a type of tree} \tbOP{dakyeerekpaglagre} \tbTP{dakyeerekpaglagrɛɛ} 
\tbLX{dakyeɛ} \tbHM{} \tbPH{dàʧɪ́ɛ} \tbPS{n.~} \tbGE{brother- or sister-in-law} \tbOP{dakyeere/dakyemine} \tbTP{dakyeerɛɛ} 
\tbLX{dakyeɛraa} \tbHM{} \tbPH{dàʧɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{tree meant to be felled} \tbOP{dakyeɛre} 
\tbLX{dakyeɛraa} \tbHM{} \tbPH{dàʧɪ́ɛ́ráá} \tbPS{n.~} \tbGE{something used for cutting wood} \tbOP{dakyeɛre} 
\tbLX{dakyeɛrɛko} \tbHM{} \tbPH{dàʧɪ́ɛ́rɛ́kò} \tbPS{n.~} \tbGE{woodpecker} \tbOP{dakyeɛrɛkori} \tbTP{dakyeɛrɛkoree} \tbVA{tekoliŋkuoraa}
\tbLX{dakye-nɔguu} \tbHM{} \tbPH{dàʧɪ̀-nɔ́gùu]} \tbPS{n.~} \tbGE{parrot} \tbOP{dakye-nɔguuri} \tbTP{dakye-nɔguuree} 
\tbLX{dakyeraa} \tbHM{} \tbPH{dàʧɪ̀ráa} \tbPS{n.~} \tbSN \tbGE{canary} \tbSN \tbGE{parrot} \tbSN \tbGE{weaver (bird) or weaverbird} \tbOP{dakyerre} \tbTP{dakyerrɛɛ} \tbVA{dakyeɛ, dakyere, dakyerɛɛ}
\tbLX{dakyɛ} \tbHM{} \tbPH{dàʧɛ́} \tbPS{n.~} \tbGE{carelessness} \tbXV{N nɛɛse dakyɛ la so ka a bie kpi.} \tbXE{The child died because of the nurse's carelessness.} 
\tbLX{dala} \tbHM{} \tbPH{dàlà} \tbPS{n.~} \tbGE{conceited talk; arrogance, self-importance} \tbXV{A pɔgelee zagere la a toma kyɛ yele o dala.} \tbXE{The girl refused to do the work and put on airs.} 
\tbLX{dalaa} \tbHM{} \tbPH{dàlàá} \tbPS{n.~} \tbGE{small drum, made with hide and the rim of a pot} \tbOP{dalle} \tbTP{dallɛɛ} \tbVA{dalaraa}
\tbLX{dalaraa (ND)} \tbHM{} \tbPH{dàláràà} \tbPS{n.~} \tbGE{} \tbOP{dalarre} \tbTP{dalarɛɛ} \tbVA{dalaa}
\tbLX{dalee} \tbHM{} \tbPH{dàléé} \tbPS{n.~} \tbGE{ dry twig} \tbOP{dabili} \tbTP{dabilee} \tbVA{delee}
\tbLX{daleerɛ} \tbHM{} \tbPH{dàlɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{total knowledge of} \tbXV{A nannyige daleerɛ zaa ka a polisri baŋ.} \tbXE{The police have gotten to know all the thief's tricks.} 
\tbLX{dalege} \tbHM{} \tbPH{dàlɪ́gɪ́} \tbPS{n.~} \tbGE{stampede} \tbOP{dalegre} \tbTP{dalegrɛɛ} 
\tbLX{dalempoɔ} \tbHM{} \tbPH{dàlɪ́mpʋ́ɔ́} \tbPS{n.~} \tbGE{an area without vegetation cover} \tbOP{dalempoore} \tbTP{dalempoorɛɛ} \tbVA{dampoɔ}
\tbLX{dalendeɛloo} \tbHM{} \tbPH{dàlɪ́ndɪ́ɛ́lʊ̀ʊ́} \tbPS{n.~} \tbGE{leech; a blood-sucking aquatic worm} \tbOP{dalendeɛlɛ} \tbTP{dalendeɛlɛɛ} 
\tbLX{daliiri} \tbHM{} \tbPH{dàlíírì} \tbPS{n.~} \tbGE{substance} \tbXV{O yɛlɛ na o naŋ yeli ba taa daliiri.} \tbXE{There is no substance in his/her speech.(is useless).} \tbTP{daliiree} 
\tbLX{dalogeraa} \tbHM{} \tbPH{dàlʋ́gráá} \tbPS{n.~} \tbGE{wood bug} \tbOP{dalogere} \tbTP{dalogerɛɛ} 
\tbLX{dalogi} \tbHM{} \tbPH{dàlógì} \tbPS{n.~} \tbGE{an area where woman pile up firewood for use in the kitchen} \tbOP{dalogri} \tbTP{dalogree} 
\tbLX{daloraa} \tbHM{} \tbPH{dàlòràá} \tbPS{n.~} \tbGE{neat stack of firewood} \tbOP{dalorri} \tbTP{daloree} 
\tbLX{dalugiri} \tbHM{} \tbPH{dàlúgrí} \tbPS{n.~} \tbGE{Y-shaped plank that supports roof beams} \tbOP{dalugo} \tbTP{dalugiree} 
\tbLX{daluri} \tbHM{} \tbPH{dàlúri} \tbPS{n.~} \tbGE{a type of calabash} \tbOP{dalurri} \tbTP{daluree} 
\tbLX{damadama} \tbHM{} \tbPH{dàmàdàmà} \tbPS{adj.~} \tbGE{big, large, extensive} 
\tbLX{damagara} \tbHM{} \tbPH{dàmàgàrà} \tbPS{n.~} \tbGE{act of showing wit; adeptness} \tbOP{damagarre} \tbTP{damagarɛɛ.} 
\tbLX{damannaa} \tbHM{} \tbPH{dàmánnáá} \tbPS{n.~} \tbSN \tbGE{measuring stick} \tbSN \tbGE{example, comparison} \tbSN \tbGE{simile (lit.)} \tbOP{damanne} \tbTP{damaannɛɛ} 
\tbLX{dambo} \tbHM{} \tbPH{dàmbó} \tbPS{n.~} \tbGE{(zɛ---) of soup not spiced} \tbOP{dambori} \tbTP{ damboree} 
\tbLX{damboli} \tbHM{} \tbPH{dàmbólì} \tbPS{n.~} \tbGE{stupid or foolish person} \tbOP{dambolo} \tbTP{dambolee} 
\tbLX{damɔlɔ} \tbHM{} \tbPH{dàmɔ́l} \tbPS{n.~} \tbGE{cheap price (be inexpensive)} 
\tbLX{dampoɔ} \tbHM{} \tbPH{dàmpʊ́ɔ́} \tbPS{n.~} \tbGE{short form of  dalempoɔ} \tbOP{dampoore} \tbTP{dampoorɛɛ} 
\tbLX{dampuori} \tbHM{} \tbPH{dàmpúórí} \tbPS{n.~} \tbGE{the back of a house} \tbOP{dampoe} \tbTP{damporee} 
\tbLX{damui} \tbHM{} \tbPH{dàmùí} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} 
\tbLX{dandamiine} \tbHM{} \tbPH{dàndàmííné} \tbPS{n.~} \tbGE{a type of red ant (found in trees)} \tbOP{dandamiime} \tbTP{dandamiimee/dandamiinee} \tbVA{gyangyamiime}
\tbLX{dandayieni} \tbHM{1} \tbPH{dándáyíéní} \tbPS{n.~} \tbGE{a type of small toad} \tbOP{dandayieme} \tbTP{dandayiemee/dandayienee} \tbVA{dayieni}
\tbLX{dandɔɔ} \tbHM{} \tbPH{dàndɔ́ɔ́} \tbPS{n.~} \tbGE{owner (male)} \tbOP{dandɔbɔ} \tbTP{dandɔbɛɛ} 
\tbLX{danduli} \tbHM{} \tbPH{dàndúlí} \tbPS{n.~} \tbGE{caterpillar, worm} \tbOP{dandulo} \tbTP{dandulee} \tbVA{dunduli}
\tbLX{daneenɛ} \tbHM{} \tbPH{dànɪ́ɪ́nɛ́} \tbPS{n.~} \tbGE{the fifth day from today} \tbOP{daneene} \tbTP{daneenɛɛ} \tbVA{daneŋenɛ}
\tbLX{danegere} \tbHM{} \tbPH{dànɪ́grɪ́} \tbPS{n.~} \tbGE{a type of honeycomb where bees store their food} \tbOP{danegɛ} \tbTP{danegerɛɛ} \tbVA{sedanegre}
\tbLX{daneŋenɛ} \tbHM{} \tbPH{dànɪ́ŋɪ́nɛ́} \tbPS{n.~} \tbGE{the sixth day from today} \tbOP{daneŋene} \tbTP{daneŋenɛɛ} 
\tbLX{danfaraŋ} \tbHM{} \tbPH{dànfàràŋ} \tbPS{adj.~} \tbGE{numb} 
\tbLX{dankyimmiri} \tbHM{} \tbPH{dànʧǐmmírì} \tbPS{n.~} \tbGE{layer of a wall; horizontal layer of plaster} \tbOP{dankyimmie} \tbTP{dankyimmiree} 
\tbLX{dankyini} \tbHM{} \tbPH{dànʧíní} \tbPS{n.~} \tbGE{wall} \tbOP{dankyime} \tbTP{dankyimee/dankyinnee.} \tbVA{daŋkyini}
\tbLX{dankyin-ŋmaa} \tbHM{} \tbPH{dǎnʧǐn-ŋmáá} \tbPS{n.~} \tbGE{broken wall (reduced in height)} \tbOP{dankyin-ŋmaara} \tbTP{dankyin-ŋmaarɛɛ} 
\tbLX{dankyinsɛgeraa} \tbHM{} \tbPH{dǎnʧǐnsɛ́gráá} \tbPS{n.~} \tbGE{chalkboard (writing wall)} \tbOP{dankyinsɛgre} \tbTP{dankyinsɛgrɛɛ} 
\tbLX{dankyinsɔglaa} \tbHM{} \tbPH{dǎnʧǐnsɔ́gláá} \tbPS{n.~} \tbGE{blackboard (black wall)} \tbOP{dankyinsɔgle} 
\tbLX{dankyinsɔɔraa} \tbHM{} \tbPH{dǎnʧǐnsɔ́ɔ́ráá} \tbPS{n.~} \tbGE{blackboard (blackened wall)} \tbOP{dankyinsɔɔre} \tbTP{dankyinsɔɔrɛɛ} 
\tbLX{dankyiŋkogilaa} \tbHM{} \tbPH{dànʧìnkógláá} \tbPS{n.~} \tbGE{inner corner of a wall e.g. a room} \tbOP{dankyiŋkogli/dankyiŋkoglo} \tbTP{dankyiŋkoglee} 
\tbLX{dankyo} \tbHM{} \tbPH{dánʧó} \tbPS{n.~} \tbGE{single-barreled gun} \tbOP{dankyori} \tbTP{dankyoree} 
\tbLX{danne} \tbHM{} \tbPH{dánnɪ̀} \tbPS{v.~} \tbGE{to praise} \tbPD{dannɛɛ, danna, danna, danema, dannaa, danema } \tbVA{dɛnne}
\tbLX{dannoŋ} \tbHM{} \tbPH{dànnʊ́ŋ} \tbPS{n.~} \tbGE{praise} 
\tbLX{danooraa} \tbHM{} \tbPH{dànʊ́ʊ́ráá} \tbPS{n.~} \tbGE{a type of wood bug} \tbOP{danoore} \tbTP{danoorɛɛ} 
\tbLX{danse} \tbHM{} \tbPH{dánsɪ̀} \tbPS{n.~} \tbGE{a paid dance session} \tbOP{dansere} \tbTP{danserɛɛ} 
\tbLX{danseɛ} \tbHM{} \tbPH{dánsɪ́ɛ́} \tbPS{n.~} \tbGE{witness} \tbOP{danseere} \tbTP{danseerɛɛ} 
\tbLX{danseɛsoba} \tbHM{} \tbPH{dánsɪ́ɛ́sʊ́bá} \tbPS{n.~} \tbGE{witness} \tbOP{danseɛdeme} 
\tbLX{Danteɛ} \tbHM{} \tbPH{dàntɪ́ɛ́} \tbPS{n.~} \tbGE{a town in the Upper West Region} 
\tbLX{Danteɛloo} \tbHM{} \tbPH{dàntɪ́ɛ́lʋ́ʋ́} \tbPS{n.~} \tbGE{a member of the Danteɛ clan} \tbOP{Danteɛlɛɛ} 
\tbLX{dantiŋ} \tbHM{} \tbPH{dàtíŋ} \tbPS{n.~} \tbGE{a type of tree} \tbOP{dantinni} \tbTP{dantinnee} 
\tbLX{danwe᷉ɛ} \tbHM{} \tbPH{dànwé᷉ɛ̀᷉} \tbPS{adj.~} \tbGE{first, ahead} \tbXV{Bayuo la a zoobu danwe᷉ɛ soba.} \tbXE{Bayuo is the first in the race.} 
\tbLX{danyagela} \tbHM{} \tbPH{dàɲáglá} \tbPS{n.~} \tbGE{the sixth day coming} \tbOP{danyaegle} \tbTP{danyagelɛɛ} 
\tbLX{danyagere} \tbHM{} \tbPH{dàɲágrɪ} \tbPS{n.~} \tbGE{root} \tbOP{danyaga} \tbTP{danyagerɛɛ} 
\tbLX{danyaŋe} \tbHM{} \tbPH{dàɲáŋ} \tbPS{adj.~} \tbGE{quickly, fast} 
\tbLX{daŋ} \tbHM{} \tbPH{dàŋ} \tbPS{adj.~} \tbGE{before} \tbXV{O  daŋ kpɛ la aloopelee.} \tbXE{S/he has travelled by plane before.} 
\tbLX{daŋaa} \tbHM{} \tbPH{dàŋàá} \tbPS{n.~} \tbGE{hearth; fireplace for cooking} \tbOP{danne} \tbTP{dannɛɛ} 
\tbLX{daŋaa} \tbHM{} \tbPH{dàŋàà} \tbPS{adj.~} \tbGE{big and heavy} 
\tbLX{daŋe} \tbHM{} \tbPH{dáŋ} \tbPS{v.~} \tbGE{to be ahead of; to beat, to be the first to do something} \tbXV{N na daŋ fo la ta.} \tbXE{I will get there before you.} \tbPD{daŋɛɛ, danna, danna, danema, dannaa, danema } 
\tbLX{daŋe} \tbHM{} \tbPH{dàŋ} \tbPS{v.~} \tbSN \tbGE{be early} \tbXV{Zenɛ, o daŋe la toma waabo.} \tbXE{Today s/he came to work early.} \tbSN \tbGE{come earlier} \tbXV{Zenɛ, o daŋe la toma waabo gaŋe bebiri zaa.} \tbXE{Today s/he came to work earlier than all other days.} \tbPD{daŋɛɛ, danna, danna, danema, dannaa, danema } 
\tbLX{daŋena} \tbHM{} \tbPH{dáŋɪ́ná} \tbPS{n.~} \tbGE{stack of pots on top of one another} \tbOP{daŋene} \tbTP{daŋenɛɛ} 
\tbLX{daŋgoɔne} \tbHM{} \tbPH{dàŋgʊ́ɔ́nɪ́} \tbPS{n.~} \tbGE{cliff} \tbOP{daŋgoɔmɔ} \tbTP{daŋgoɔnɛɛ} 
\tbLX{daŋgbala} \tbHM{} \tbPH{dáŋgbála} \tbPS{n.~} \tbGE{bare and open place} \tbOP{daŋgballe} \tbTP{daŋgbalɛɛ} 
\tbLX{daŋgbale} \tbHM{} \tbPH{dàŋgbàlɪ́} \tbPS{n.~} \tbGE{walking stick} \tbOP{daŋgbala} \tbTP{daŋgbalɛɛ/daŋgballɛɛ} 
\tbLX{daŋkasaŋ} \tbHM{} \tbPH{dáŋkásáŋ} \tbPS{adj.~} \tbGE{bloated-looking} 
\tbLX{daŋkuuri} \tbHM{} \tbPH{dàŋkúúrí} \tbPS{n.~} \tbGE{stone used to support a cooking pot in the hearth} \tbOP{daŋkubo/daŋkurri} \tbTP{daŋkuree/daŋkuuree} 
\tbLX{daŋŋa} \tbHM{} \tbPH{dáŋŋá} \tbPS{n.~} \tbGE{mischief} 
\tbLX{Daŋyi} \tbHM{} \tbPH{dàŋyí} \tbPS{n.~} \tbGE{name traditionally given to the firstborn of a set of twins} \tbVA{Zeɛmɛ}
\tbLX{daŋmaa} \tbHM{} \tbPH{dàŋmáá} \tbPS{n.~} \tbGE{log} \tbOP{daŋmaara} \tbTP{daŋmaarɛɛ} 
\tbLX{daŋmeɛlɛ} \tbHM{} \tbPH{dàŋmɪ́ɛ́lɛ́} \tbPS{n.~} \tbGE{the sixth day from today} \tbOP{daŋmeɛlɛ} \tbTP{ daŋmeɛlɛɛ} 
\tbLX{daŋmeɛraa} \tbHM{} \tbPH{dàŋmɪ́ɛ́ráà} \tbPS{n.~} \tbGE{stick used to beat someone} \tbOP{daŋmeɛre} 
\tbLX{daŋmeɛraa} \tbHM{} \tbPH{dàŋmɪ́ɛ́ráà} \tbPS{n.~} \tbGE{solution, answer to a problem} \tbOP{daŋmeɛre} 
\tbLX{daoo} \tbHM{} \tbPH{dàʊ̀ʊ̀} \tbPS{adv.~} \tbGE{large and firm-looking} 
\tbLX{da-ɔɔraa} \tbHM{} \tbPH{dà-ɔ́ɔ́ráá} \tbPS{n.~} \tbGE{chewing stick (tooth cleaner)} \tbOP{da-ɔɔre} 
\tbLX{dapaaloŋ} \tbHM{} \tbPH{dàpáálʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{vanity} \tbSN \tbGE{splendour, glory} 
\tbLX{dapare} \tbHM{} \tbPH{dàpàrɪ́} \tbPS{n.~} \tbGE{dwelling place of the dead; heaven, eternity} 
\tbLX{daparedaa} \tbHM{} \tbPH{dàpárɪ́dàá} \tbPS{n.~} \tbGE{death, next world} 
\tbLX{daparekpeɛŋaa} \tbHM{} \tbPH{dàpàrɪ̀kpɪ́ɛ́ŋáá} \tbPS{n.~} \tbGE{strong spirit} \tbOP{daparekpeɛne} \tbTP{daparekpeɛnɛɛ} 
\tbLX{daparenuulee} \tbHM{} \tbPH{dàpàrɪ̀núúléé} \tbPS{n.~} \tbGE{cockalo} \tbOP{daparenuuli} 
\tbLX{daparoo} \tbHM{} \tbPH{dàpárʊ́ʊ́} \tbPS{n.~} \tbGE{soul or spirit of the dead} \tbSG{daparrɛɛ} \tbTP{daparɛɛ} 
\tbLX{dapeɛlkammo} \tbHM{} \tbPH{dàpɪ́ɛ́lkámmʊ́} \tbPS{n.~} \tbGE{a type of bean} \tbTP{dapeɛlkamɛɛ} 
\tbLX{daperaa} \tbHM{} \tbPH{dàpɪ́ràa} \tbPS{n.~} \tbGE{small dove or pigeon (grey and red)} \tbOP{daperre} \tbTP{daperɛɛ/daperrɛɛ} 
\tbLX{daperekalonni} \tbHM{} \tbPH{dàpɪ́rɪ́kálónnì} \tbPS{n.~} \tbGE{a type of trap for birds (usually made of bamboo and twine)} \tbTP{daperekalonnee} 
\tbLX{daperenyaa} \tbHM{} \tbPH{dàpɪ̀rɪ̀ɲáà} \tbPS{n.~} \tbGE{the breast of a type of dove} 
\tbLX{dapɛɛre} \tbHM{} \tbPH{dàpɛ̀ɛ̀rɪ́} \tbPS{n.~} \tbGE{immature tree or wood} \tbOP{dapɛbɛ} \tbTP{dapɛbɛɛ} 
\tbLX{dapɛge} \tbHM{} \tbPH{dàpɛ́gɪ́} \tbPS{n.~} \tbGE{bark (of tree)} \tbOP{dapɛgre/dapɛgɛ} \tbTP{dapɛgrɛɛ} 
\tbLX{dapɛle} \tbHM{} \tbPH{dàpɛ́lɪ́} \tbPS{n.~} \tbGE{splinter of wood} \tbOP{dapɛlɛ} \tbTP{dapɛlɛɛ/dapɛllɛɛ} 
\tbLX{dapɛlemɛ} \tbHM{} \tbPH{dàpɛ́lmɛ́} \tbPS{n.~} \tbGE{splinters of wood} \tbTP{dapɛlmɛɛ} 
\tbLX{dapɛlɛŋe} \tbHM{} \tbPH{dàpɛ̀lɛ̀ŋ} \tbPS{n.~} \tbGE{a type of tiny wild bat} \tbOP{dapɛlɛnne} \tbTP{dapɛlɛnnɛɛ} 
\tbLX{dapɛloo} \tbHM{} \tbPH{dàpɛ́lʋ́ʋ́} \tbPS{n.~} \tbGE{splinter of wood} 
\tbLX{dara} \tbHM{} \tbPH{dárá} \tbPS{n.~} \tbGE{a game played by moving small pins into marked-out holes} \tbOP{darre} \tbTP{ darɛɛ/darrɛɛ} 
\tbLX{daransaŋ} \tbHM{} \tbPH{dàránsáŋ} \tbPS{n.~} \tbGE{heavy, cast-aluminium cooking pot} \tbOP{daransanne} \tbTP{daransannɛɛ} 
\tbLX{dare} \tbHM{} \tbPH{dárɪ́} \tbPS{v.~} \tbGE{to be hysterical} \tbPD{darɛɛ, dara, dara, dareba, daraa, dareba } 
\tbLX{dare} \tbHM{} \tbPH{dàrɪ̀} \tbPS{v.~} \tbSN \tbGE{to harden} \tbSN \tbGE{to get acustomed to; to be used to something unpleasant} \tbPD{darɛɛ, dara, dara, dareba, dara, dareba } 
\tbLX{Dasa᷉a᷉} \tbHM{} \tbPH{dàsá᷉á᷉} \tbPS{n.~} \tbGE{name traditionally given to a child born on a rainy day} \tbOP{dasaamine} \tbTP{dasaanɛɛ} 
\tbLX{dasaŋa} \tbHM{} \tbPH{dàsáŋà} \tbPS{n.~} \tbGE{olden times} \tbTP{dasaŋɛɛ} \tbVA{dakoroŋ}
\tbLX{dasaraa} \tbHM{} \tbPH{dàsáràà} \tbPS{n.~} \tbGE{plank} \tbOP{dasarre} \tbTP{dasarrɛɛ} 
\tbLX{daseelɛ} \tbHM{} \tbPH{dàsɪ̀ɪ̀lɛ̀} \tbPS{n.~} \tbGE{suicide} \tbTP{daseelɛɛ} 
\tbLX{dasɛgeraa} \tbHM{} \tbPH{dàsɛ́gráà} \tbPS{n.~} \tbGE{wood for filling in gaps (especially in mud roofing)} \tbOP{dasɛgere} \tbTP{daseɛgerɛɛ} 
\tbLX{dasɛgeraa} \tbHM{} \tbPH{dàsɛ́gɪ́ráá} \tbPS{n.~} \tbGE{piece of wood used to cover gaps between bigger plunks of wood in an earth-constructed roof} \tbTP{dasɛgere, dasɛgerɛɛ} 
\tbLX{dasɛrɛɛ} \tbHM{} \tbPH{dàsɛ́rɛ́ɛ́} \tbPS{n.~} \tbGE{small sticks that are used to close a gap in a thatch earth-made roof of a building} \tbSG{dasɛroo} \tbTP{dasɛrrɛɛ} \tbVA{dasɛgeraa}
\tbLX{dasɛroo} \tbHM{} \tbPH{dàsɛ́rʋ́ʋ́} \tbPS{n.~} \tbGE{small stick that is used to close a gap in a thatch earth-made roof of a building} \tbOP{dasɛrɛɛ} \tbTP{dasɛrrɛɛ} 
\tbLX{dasugo} \tbHM{} \tbPH{dàsúgò} \tbPS{n.~} \tbGE{many kinds of trees} \tbOP{dasugiri} \tbTP{dasugree} 
\tbLX{dasuoluŋ} \tbHM{} \tbPH{dàsùólúŋ} \tbPS{n.~} \tbGE{shadow} \tbOP{dasuolo} \tbTP{dasuolee} \tbVA{salinsuuri}
\tbLX{dataa} \tbHM{} \tbPH{dàtáá} \tbPS{n.~} \tbSN \tbGE{rival (between males)} \tbSN \tbGE{antagonist (lit.)} \tbOP{dataare} \tbTP{ dataarɛɛ} 
\tbLX{datara} \tbHM{} \tbPH{dàtárá} \tbPS{n.~} \tbGE{third day from now} \tbOP{datarre} \tbTP{datarrɛɛ} 
\tbLX{datuo} \tbHM{1} \tbPH{dàtúó} \tbPS{n.~} \tbGE{a type of tree with medicinal properties and edible fruits} \tbOP{datuuri} \tbTP{datuuree} 
\tbLX{datuo} \tbHM{2} \tbPH{dàtúó} \tbPS{n.~} \tbGE{being expensive} \tbOP{datuuri} \tbTP{datuuree} 
\tbLX{dave} \tbHM{} \tbPH{dàvɪ́} \tbPS{n.~} \tbGE{spoke} \tbOP{davere} \tbTP{daverɛɛ} 
\tbLX{daveene} \tbHM{} \tbPH{dàvɪ̀ɪ̀nɪ̀} \tbPS{n.~} \tbGE{a type of small bird with very thin legs} \tbOP{daveemɛ} \tbTP{daveenɛɛ} \tbVA{veene}
\tbLX{daveraŋkantoll} \tbHM{} \tbPH{dàvɪ̀ràŋkántóll} \tbPS{n.~} \tbGE{last warning, last chance} 
\tbLX{davoro} \tbHM{} \tbPH{dàvòró} \tbPS{n.~} \tbGE{courtyard} \tbOP{davorri} \tbTP{davorree} 
\tbLX{davuo} \tbHM{} \tbPH{dàvùó} \tbPS{n.~} \tbGE{a hollowed out log} \tbOP{davori} \tbTP{davoree} 
\tbLX{dawɛge} \tbHM{} \tbPH{dàwɛ́gɪ́} \tbPS{n.~} \tbGE{a big plunk of wood used in the construction of earth-made roof} \tbOP{dawɛgre} \tbTP{dawɛgrɛɛ} \tbVA{wɛge}
\tbLX{dawɛrɛ} \tbHM{} \tbPH{dàwɛ̀rɪ̀} \tbPS{n.~} \tbGE{person who saws or splits wood} \tbOP{dawɛrebɛ} \tbTP{dawɛrebɛɛ} 
\tbLX{daworo} \tbHM{} \tbPH{dáwóró} \tbPS{n.~} \tbGE{a type of bell used in music and to call for attention} \tbOP{dawoe} \tbTP{ daworee.} 
\tbLX{dayeni} \tbHM{} \tbPH{dàyénì} \tbPS{adj.~} \tbGE{once} \tbOP{dayeme} \tbTP{dayennee} 
\tbLX{dayere} \tbHM{} \tbPH{dàyéré} \tbPS{n.~} \tbGE{the day after tomorrow} \tbTP{dayeree} 
\tbLX{dayieni} \tbHM{} \tbPH{dáyíéní} \tbPS{n.~} \tbGE{a type of small toad} \tbOP{dayieme} \tbTP{dayienee} \tbVA{dandayieni}
\tbLX{dayuu} \tbHM{} \tbPH{dàyúú} \tbPS{n.~} \tbGE{giant rat} \tbOP{dayuuri dayuuree} 
\tbLX{dazɛllaa} \tbHM{} \tbPH{dàzɛ́lláá} \tbPS{n.~} \tbGE{swing} \tbOP{dazɛlle} \tbTP{dazɛllɛɛ} 
\tbLX{dazugiri} \tbHM{} \tbPH{dàzùgrí} \tbPS{n.~} \tbGE{} \tbOP{dazugo} \tbTP{dazugiree} \tbVA{shortened form of kɔndazugri}
\tbLX{dazugivu᷉u} \tbHM{} \tbPH{dàzùgvù᷉ù} \tbPS{n.~} \tbGE{intense fire; hell; inferno} \tbTP{dazugvuunee} 
\tbLX{de} \tbHM{} \tbPH{dɪ̌} \tbPS{v.~} \tbSN \tbGE{to take} \tbSN \tbGE{to accept} \tbSN \tbGE{to believe} \tbPD{deɛɛ, derɛ, derɛ, derebɛ, deraa, derebɛ } 
\tbLX{dedee} \tbHM{} \tbPH{dèdéé} \tbPS{adv.~} \tbGE{see dedeekɔɔ} 
\tbLX{dedeekɔɔ} \tbHM{} \tbPH{dèdéékɔ́ɔ́} \tbPS{n.~} \tbGE{act of standing steadily e.g. a child beginning to walk} \tbVA{dedeekporo, zazaakɔɔ}
\tbLX{dêe} \tbHM{} \tbPH{dɪ̀̂ɪ̀̂} \tbPS{adv.~} \tbGE{unlimited} 
\tbLX{deedee} \tbHM{} \tbPH{déédéé} \tbPS{adj.~} \tbGE{straight, upright} 
\tbLX{deɛ} \tbHM{} \tbPH{dɪ́ɛ̀} \tbPS{adv.~} \tbGE{just} \tbXV{N deɛ ba baŋ a pare.} \tbXE{I just don't understand it.} 
\tbLX{deɛ} \tbHM{} \tbPH{dɪ́ɛ̀} \tbPS{v.~} \tbGE{to remain somewhere} \tbXV{O gaa la a tigiri te deɛ.} \tbXE{S/he went to the feast and never returned.} \tbPD{deɛɛ, deɛrɛ, deɛrɛ, deɛrebɛ, deɛraa, deɛrebɛ } 
\tbLX{de᷉ɛ} \tbHM{} \tbPH{dɪ̀᷉ɛ̀᷉} \tbPS{v.~} \tbGE{to play (child)} \tbPD{de᷉ɛ᷉ɛ᷉, deɛnɛ, deɛnɛ, deɛnemɛ, deɛnaa, - } 
\tbLX{dêɛ} \tbHM{} \tbPH{dɪ̂̀᷉ɛ̀᷈} \tbPS{v.~} \tbGE{to play, to joke} \tbPD{dêɛɛ, deɛnɛ, deɛnɛ, deɛnemɛ, deɛnaa, deɛnemɛ } 
\tbLX{de᷉ɛbaŋ} \tbHM{} \tbPH{dɪ̀᷉ɛ̀bàŋ} \tbPS{n.~} \tbGE{drama, play} 
\tbLX{de᷉ɛde᷉ɛkaraa} \tbHM{} \tbPH{dɪ̀᷉ɛ̀dɪ̀᷉ɛ̀kàràà} \tbPS{n.~} \tbSN \tbGE{lead player} \tbSN \tbGE{protagonist (lit.)} \tbOP{de᷉ɛde᷉ɛkarre} \tbTP{de᷉ɛde᷉ɛkarɛɛ} 
\tbLX{de᷉ɛdeɛnɛ} \tbHM{} \tbPH{dɪ̀᷉ɛ̀dɪ́ɛ́nɛ́} \tbPS{n.~} \tbSN \tbGE{one who plays e.g. of a child} \tbSN \tbGE{a character in a story, drama} \tbOP{de᷉ɛdeɛmɛ} \tbTP{de᷉ɛdeɛmɛɛ} \tbVA{deɛne}
\tbLX{dêɛkakalle} \tbHM{} \tbPH{dɪ̂̀ɛ̀kákàllɪ́} \tbPS{n.~} \tbGE{rough, dangerous play e.g. throwing hard objects at each other} \tbTP{  dêɛkakalɛɛ/dêɛkakallɛɛ.} 
\tbLX{de᷉ɛkoɔbo} \tbHM{} \tbPH{dɪ̀᷉ɛ̀kʊ́ɔ́bʊ́} \tbPS{n.~} \tbGE{farmwork done for one's in-law} \tbTP{de᷉ɛkɔrɛɛ} 
\tbLX{de᷉ɛkuori} \tbHM{} \tbPH{dɪ́᷉ɛ́᷉kúórí} \tbPS{n.~} \tbGE{funeral of an in-law} \tbOP{de᷉ɛkoe} \tbTP{de᷉ɛkoree} 
\tbLX{de᷉ɛkpa᷉ao} \tbHM{} \tbPH{dɪ̀ɛ̀kpá᷉áʊ́} \tbPS{n.~} \tbGE{guineafowl meant for one's in-law} \tbOP{de᷉ɛkpinni} \tbTP{de᷉ɛkpinnee} 
\tbLX{deɛlaa} \tbHM{} \tbPH{dɪ̀ɛ̀láá} \tbPS{n.~} \tbGE{nightjar} \tbOP{deɛle/deɛlɛ/deɛlɛɛ.} 
\tbLX{deɛle} \tbHM{} \tbPH{dɪ́ɛ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to glide away from the target, to deviate} \tbSN \tbGE{to deflect, to dodge} \tbPD{deɛlɛɛ, deɛlɛ, deɛlɛ, deɛlebɛ, deɛlaa, deɛlebɛ } 
\tbLX{deɛle} \tbHM{} \tbPH{dɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to spread out (to dry)} \tbSN \tbGE{to expose completely} \tbSN \tbGE{to float on liquid} \tbPD{deɛlɛɛ, deɛlɛ, deɛlɛ, deɛlebɛ, deɛlɛ, deɛlebɛ } 
\tbLX{deɛlɛɛpʊ᷉ʊ} \tbHM{} \tbPH{dɪ́ɛ́lɛ́ɛ́pʊ́᷉ʊ́} \tbPS{n.~} \tbGE{drizzle} 
\tbLX{deɛlɛɛpo᷈o᷈} \tbHM{} \tbPH{dɪ́ɛ́lɛ́ɛ́pʊ́᷈ʊ̀᷈} \tbPS{n.~} \tbGE{drizzle} \tbOP{deɛlɛɛpoone} \tbTP{deɛlɛɛpoonɛɛ} 
\tbLX{de᷉ɛmakoma} \tbHM{} \tbPH{dɪ̀᷉ɛ̀màkʊ́má} \tbPS{n.~} \tbGE{grandmother-in-law} \tbOP{de᷉ɛmakommine} 
\tbLX{deɛmɛ} \tbHM{} \tbPH{dɪ̀ɛ̀mɛ́} \tbPS{n.~} \tbSN \tbGE{in-law (father or mother)} \tbSN \tbGE{son-in-law} \tbOP{deɛmine} \tbTP{deɛmɛɛ} 
\tbLX{deɛmɛkonlɔŋ} \tbHM{} \tbPH{dɪ́ɛ́mɛ́kʊ́nlɔ́ŋ} \tbPS{n.~} \tbGE{the first layer of TZ in a bowl} \tbOP{deɛmɛkonlɔnne} \tbTP{deɛmɛkonlɔnnɛɛ} 
\tbLX{deɛmpɔge} \tbHM{} \tbPH{dɪ̀᷉ɛ̀pɔ́gɪ́} \tbPS{n.~} \tbGE{mother-in-law} \tbOP{deɛmpɔgeba/deɛmpɔgemine} \tbTP{deɛmpɔgrɛɛ} 
\tbLX{deɛnaa} \tbHM{} \tbPH{dɪ́ɛ́náá} \tbPS{n.~} \tbGE{plaything, toy} \tbOP{deɛne} \tbTP{deɛnɛɛ} 
\tbLX{deɛndɔɔ} \tbHM{} \tbPH{dè᷉ɛdɔ́ɔ́} \tbPS{n.~} \tbGE{father-in-law} \tbOP{deɛndɔbɔ/deɛndɔɔmine} \tbTP{deɛ᷉dɔɔrɛɛ} 
\tbLX{deɛne} \tbHM{} \tbPH{dɪ̀ɛ̀nɪ́} \tbPS{n.~} \tbSN \tbGE{play} \tbXV{Bibiiri nɔŋe la deɛne yaga.} \tbXE{Children love play.} \tbXV{A bibiiri ba nɔŋe a deɛne ŋa.} \tbXE{The children do not like this play.} \tbSN \tbGE{joke} \tbXV{Bonso ka fo de a yelnimize ŋa eŋe deɛne?} \tbXE{Why do you take this serious matter for a joke?.} \tbSN \tbGE{drama} \tbXV{Zenɛ te na kaa la karembiiri deɛne na naŋ kyaare ‘bibiiri emmaaroŋ yɛlɛ’.} \tbXE{Today we shall watch a drama piece by pupils on the title ‘children’s health’.} \tbTP{deɛnɛɛ} \tbVA{de᷉ɛdeɛnɛ}
\tbLX{dêɛne} \tbHM{} \tbPH{dɪ̂̀ɛ̀nɪ́} \tbPS{n.~} \tbGE{play, joke} \tbTP{deɛnɛɛ} 
\tbLX{deɛraa} \tbHM{} \tbPH{dɪ́ɛ́ráá} \tbPS{n.~} \tbGE{an instrument for scraping and smoothing the edge of objects} \tbOP{deɛre} \tbTP{deɛrɛɛ} 
\tbLX{deɛre} \tbHM{} \tbPH{dɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to trim, to cut to make even (mouth of calabash)} \tbPD{deɛrɛɛ, deɛrɛ, deɛrɛ, deɛrebɛ, deɛraa, deɛrebɛ } 
\tbLX{de᷉ɛsa᷉akoma} \tbHM{} \tbPH{dɪ̀᷉ɛ̀᷉sà᷉à᷉kʊ́má} \tbPS{n.~} \tbGE{grandfather-in-law} \tbOP{de᷉ɛsa᷉akommine} 
\tbLX{de᷉ɛyiri} \tbHM{} \tbPH{dɪ̀᷉ɛ̀᷉yírí} \tbPS{n.~} \tbGE{house of an in-law} \tbOP{de᷉ɛyie} \tbTP{de᷉ɛyiree} 
\tbLX{degedaga} \tbHM{} \tbPH{dɪ̀gɪ̀dàgà} \tbPS{adv.~} \tbGE{thick, heavy, and immovable} 
\tbLX{degere} \tbHM{} \tbPH{dɪ̀grɪ́} \tbPS{n.~} \tbGE{dwarf, one stunted in growth} \tbOP{degɛ} \tbTP{degerɛɛ} 
\tbLX{degili} \tbHM{} \tbPH{dèglì} \tbPS{v.~} \tbGE{not to be of substance} \tbPD{degilee, degile, degile, degilibe, degilaa, degilibe } 
\tbLX{degili} \tbHM{} \tbPH{dèglì} \tbPS{v.~} \tbGE{to marry one's sibling's husband} \tbPD{degilee, degile, degile, degilibe, degilaa, degilibe } 
\tbLX{deli} \tbHM{} \tbPH{délí} \tbPS{v.~} \tbGE{to have one's eyes dazzled; to look without seeing} \tbPD{delee, dele, dele, delibe, delaa, delibe } 
\tbLX{deme} \tbHM{} \tbPH{dèmé} \tbPS{pron.~} \tbGE{own pl. (yours/mine)} \tbXV{Fo deme la gaŋ n deme.} \tbXE{Yours are more than mine.} 
\tbLX{demee} \tbHM{} \tbPH{déméé} \tbPS{n.~} \tbGE{unidentified people} \tbXV{Yeŋ demee la wa?} \tbXE{Which people have come?.} 
\tbLX{demee} \tbHM{} \tbPH{déméé} \tbPS{n.~} \tbGE{owners} \tbXV{A noore taa la demee.} \tbXE{The fowls have owners.} 
\tbLX{demm} \tbHM{} \tbPH{dèmm} \tbPS{adv.~} \tbGE{large and upright} 
\tbLX{dendâee} \tbHM{} \tbPH{dɪ̀ndà᷉ɪ̀᷉ɪ̀᷈} \tbPS{adv.~} \tbGE{appearing suddenly and frighteningly} 
\tbLX{dendege} \tbHM{} \tbPH{dèndégé} \tbPS{n.~} \tbGE{a woman given in marriage to her sister's husband} \tbOP{dendegri} \tbTP{dedegree} 
\tbLX{dendeŋe} \tbHM{} \tbPH{d̀endèŋé} \tbPS{adj.~} \tbGE{upright} \tbOP{dendenni} \tbTP{dendennee} 
\tbLX{dendeŋi} \tbHM{} \tbPH{děndéŋ} \tbPS{adv.~} \tbGE{upright, straight} 
\tbLX{dendeŋŋe} \tbHM{} \tbPH{déndéŋŋé} \tbPS{n.~} \tbSN \tbGE{dagger} \tbSN \tbGE{the longest toe of a bird} \tbOP{dendenni} \tbTP{dendennee} 
\tbLX{dendɔgaŋe} \tbHM{} \tbPH{dɪ̀ndɔ́gǎŋ} \tbPS{n.~} \tbGE{bastard} \tbOP{dendɔganne} \tbTP{dedɔgannɛɛ} \tbVA{dendɔgaŋa}
\tbLX{dendɔlee} \tbHM{} \tbPH{dɪ̀ndɔ́léé} \tbPS{n.~} \tbGE{bastard} \tbOP{dendɔbilii} \tbTP{dendɔbillee} 
\tbLX{dendɔmɔ} \tbHM{} \tbPH{dɪ́ndɔ́mɔ́} \tbPS{n.~} \tbGE{enemy} \tbOP{dendɔnne} \tbTP{dendɔnnɛɛ} \tbVA{dendɔma, dɔndɔmɔ}
\tbLX{dendɔre} \tbHM{} \tbPH{dɪ̀ndɔ́rɪ̀} \tbPS{n.~} \tbSN \tbGE{doorway} \tbSN \tbGE{frontage of the main house} \tbOP{dendɔɛ} \tbTP{dendɔrɛɛ} 
\tbLX{deni} \tbHM{} \tbPH{dèní} \tbPS{pron.~} \tbGE{that which belongs to} \tbXV{A dagere poɔ n deni la bile.} \tbXE{Among the boxes mine(the one that belongs to me) is the smallest.} 
\tbLX{denni} \tbHM{} \tbPH{dénnì} \tbPS{v.~} \tbGE{to carry on one's head without the support of one's hands} \tbPD{dennee, denne, denne, denime, dennaa, denimee } \tbVA{delli}
\tbLX{deŋgeluu} \tbHM{} \tbPH{dèŋgélúú} \tbPS{n.~} \tbGE{sweet potato} \tbOP{deŋgele} \tbTP{ deŋgellee} 
\tbLX{deŋgereŋ} \tbHM{} \tbPH{dèŋgèrèŋ} \tbPS{adv.~} \tbGE{upright, stiff, unbending and strong} 
\tbLX{deŋo} \tbHM{} \tbPH{dɪ̀ŋʊ́} \tbPS{n.~} \tbGE{preference, wish (not caring about others)} 
\tbLX{de pɔge} \tbHM{} \tbPH{dɪ́ pɔ́gɪ́} \tbPS{v.~} \tbGE{to marry a woman} \tbPD{-, -, -, -, -, - } 
\tbLX{deraa} \tbHM{} \tbPH{dɪ̀ràà} \tbPS{adj.~} \tbGE{strong and hard, unbendable} 
\tbLX{deree} \tbHM{} \tbPH{déréé} \tbPS{n.~} \tbGE{negligence, carelessness} 
\tbLX{deri} \tbHM{} \tbPH{dèrì} \tbPS{v.~} \tbSN \tbGE{to harden, to dry (wet or muddy place or substance)} \tbXV{Pampana ŋa ka ba salle a dankyini o naŋ ba deri.} \tbXE{The wall has just been plastered and is not hardened yet.} \tbSN \tbGE{to go into a state of convulsion, to collapse in hysteria} \tbXV{Ka kɔmmaaroŋ waare pɔgenyaŋaa naŋ zoro ɔɔre o na deri la.} \tbXE{If cold water is accidently sprayed with water she will collapse in cold.} \tbPD{deree, dere, dere, deribe, deraa, deribe } 
\tbLX{detaareŋ} \tbHM{} \tbPH{dɪ̀táárɪ̌ŋ} \tbPS{adv.~} \tbGE{3 years ago} 
\tbLX{deyɛŋ} \tbHM{} \tbPH{dɪ̀yɛ̌ŋ} \tbPS{adv.~} \tbGE{last year} 
\tbLX{dɛ} \tbHM{} \tbPH{dɛ̀} \tbPS{v.~} \tbGE{pl. to cut to make even (fingernails)} \tbPD{dɛɛ, deɛrɛ, deɛrɛ, deɛrɛ, deɛraa, deɛrebɛ } 
\tbLX{dɛbiri} \tbHM{} \tbPH{dɛ́bírì} \tbPS{n.~} \tbGE{stair} \tbOP{dɛbie} \tbTP{dɛbiiree} \tbVA{dabiri}
\tbLX{dɛbogi} \tbHM{} \tbPH{dɛ́bògí} \tbPS{n.~} \tbGE{step on a ladder; stairs} \tbOP{dɛbogiri} \tbTP{dɛbogiree} 
\tbLX{dɛdɛrɛɛ} \tbHM{} \tbPH{dɛ́dɛ́rɛ́ɛ́} \tbPS{adv.~} \tbGE{shallow} 
\tbLX{dɛ᷉ɛ} \tbHM{} \tbPH{dɛ̀᷉ɛ̀᷉} \tbPS{adv.~} \tbGE{open and endless e.g. grassland, desert, ocean} 
\tbLX{dɛge} \tbHM{} \tbPH{dɛ́gɪ́} \tbPS{v.~} \tbGE{to be dirty} \tbPD{dɛgɛɛ, dɛgrɛ, dɛgrɛ, -, -, - } 
\tbLX{dɛgelɛkɛlɛ} \tbHM{} \tbPH{dɛ̀glɛ̀kɛ̀lɛ̀} \tbPS{adj.~} \tbGE{fatty and plentiful e.g of a big dinner (food) on a rich couple’s wedding day} 
\tbLX{dɛgere} \tbHM{} \tbPH{dɛ́grɪ̀} \tbPS{n.~} \tbGE{dirt} \tbOP{dɛgɛ} \tbTP{dɛgerɛɛ} 
\tbLX{dɛgere iribu} \tbHM{} \tbPH{dɛ̌grɪ́ íríbú} \tbPS{n.~} \tbSN \tbGE{circumcision (male)} \tbSN \tbGE{excision (female)} 
\tbLX{dɛlle} \tbHM{} \tbPH{dɛ́llɪ́} \tbPS{v.~} \tbGE{to lean against, to rely upon for support} \tbPD{dɛllɛɛ, dɛllɛ, dɛllɛ, dɛlebɛ, dɛllaa, dɛlebɛ } 
\tbLX{dɛlle} \tbHM{} \tbPH{dɛ́llɪ́} \tbPS{v.~} \tbGE{to stage a corpse} \tbPD{dɛllɛɛ, dɛllɛ, dɛllɛ, dɛlebɛ, dɛllaa, dɛlebɛ } 
\tbLX{dɛlle} \tbHM{} \tbPH{dɛ̀llɪ̀} \tbPS{v.~} \tbGE{to remove the top layer of liquid, to decant, to skim} \tbPD{dɛllɛɛ, dɛllɛ, dɛllɛ, dɛlebɛ, dɛllaa, dɛlebɛ } 
\tbLX{dɛmɛdɛmɛ} \tbHM{} \tbPH{dɛ̀mɛ̀dɛ́mɛ́} \tbPS{n.~} \tbGE{making an effort} 
\tbLX{dɛmɛdɛmɛ} \tbHM{} \tbPH{dɛ̀mɛ̀dɛ̀mɛ̀} \tbPS{adv.~} \tbSN \tbGE{full to capacity with liquid e.g. a swollen foot} \tbSN \tbGE{big, fat and tall or long e.g a huge python} 
\tbLX{dɛmmo} \tbHM{} \tbPH{dɛ̀mmʋ́} \tbPS{n.~} \tbGE{conversation, talk, a pending problem} \tbOP{dɛmɛ} \tbTP{ dɛmɛɛ} 
\tbLX{dɛndɛlɛŋ} \tbHM{} \tbPH{dɛ̀ndɛ̀lɛ̀ŋ} \tbPS{adv.~} \tbGE{long and wavy} 
\tbLX{dɛndɛŋe} \tbHM{} \tbPH{dɛ̀ndɛ́ŋ} \tbPS{adj.~} \tbGE{first, earlier} \tbVA{dɛŋ}
\tbLX{dɛnne} \tbHM{} \tbPH{dɛ̀nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to straighten} \tbXV{Da maaroŋ ka ba maŋ dɛnne, ba ba dɛnnɛ dakuoŋaa.} \tbXE{You can straighten a fresh stick, but not a dry one.} \tbSN \tbGE{to roam} \tbXV{A teŋɛ zaa ka o dɛnne kyɛ naŋ ba bale.} \tbXE{She has roamed the whole town but is not tired yet.} \tbSN \tbGE{to be full to capacity (with liquid)} \tbXV{Saa mie la ka a baa dɛnne o zaa.} \tbXE{It has rained and the stream is full to capacity.} \tbPD{dɛnnɛɛ, dɛnnɛ, dɛnnɛ, dɛnemɛ, dɛnnaa, dɛnemɛ } 
\tbLX{dɛnne} \tbHM{} \tbPH{dɛ́nnɪ́} \tbPS{v.~} \tbGE{praise} \tbPD{dɛnnɛɛ, dɛnnɛ, dɛnnɛ, dɛnemɛ, dɛnnaa, - } \tbVA{danne}
\tbLX{dɛŋe} \tbHM{} \tbPH{dɛ́ŋ} \tbPS{adj.~} \tbGE{first} 
\tbLX{dɛŋeneŋ} \tbHM{} \tbPH{dɛ̀ŋɪ́nɪ́ŋ} \tbPS{adj.~} \tbGE{first instance} 
\tbLX{dɛŋɛɛ} \tbHM{} \tbPH{dɛ̀ŋɛ̀ɛ̀} \tbPS{adj.~} \tbGE{plentiful} 
\tbLX{Dɛre} \tbHM{} \tbPH{dɛ́rɪ́} \tbPS{proper noun.~} \tbGE{name traditionally given to a male child born after the death of an infant boy of the same mother} 
\tbLX{dɛre} \tbHM{} \tbPH{dɛ́rɪ́} \tbPS{n.~} \tbGE{house mouse} \tbOP{dɛrɛɛ} 
\tbLX{dɛre} \tbHM{} \tbPH{dɛ́rɪ̀} \tbPS{n.~} \tbGE{ladder} \tbOP{dɛɛ} \tbTP{dɛrɛɛ} 
\tbLX{dɛre} \tbHM{} \tbPH{dɛ́rɪ́} \tbPS{v.~} \tbGE{to boil pito at the second stage of brewing} \tbPD{dɛrɛɛ, dɛrɛ, dɛrɛ, dɛrebɛ, dɛraa, - } 
\tbLX{dɛremaakonokono} \tbHM{} \tbPH{dɛ́rɪ́máákónókónó} \tbPS{n.~} \tbGE{a traditional women's dance} 
\tbLX{Dɛrepɔge} \tbHM{} \tbPH{dɛ̀rɪ̀pɔ́gɪ́} \tbPS{ n.~} \tbGE{name traditionally given to a female child born after the death of an infant girl of the same mother} 
\tbLX{di} \tbHM{} \tbPH{dì} \tbPS{v.~} \tbSN \tbGE{to eat, to consume} \tbXV{Bibiiri seŋ ka ba maŋ di bondirii kyɛ gaa sakuuri.} \tbXE{} \tbSN \tbGE{to burn,  to blaze} \tbXV{Vu᷉u᷉ di la a tuu zaa.} \tbXE{The whole shrub has been burnt down by fire.} \tbSN \tbGE{to erode, to make sore} \tbXV{A bie noɔre die la ka ba sage na di bonzaa.} \tbXE{The child has a sore mouth and refuses to eat anything.} \tbSN \tbGE{to overcome, to win} \tbSN \tbGE{to assume a throne} \tbSN \tbGE{to judge} \tbXV{Ba di la ba zɔɔre sɛrɛɛ na baare.} \tbXE{They have finished judging that case of fighting.} \tbPD{diee, dire, dire, diribe, diraa, diribe } \tbSE{di danseɛ} \tbGE{to give witness} \tbSE{di dapaaloŋ} \tbGE{to show elegance} \tbSE{di dɛmmo} \tbGE{to converse} \tbSE{di dɔɔloŋ} \tbGE{to show bravery} \tbSE{di dɔgɛɛ} \tbGE{to suffer} \tbSE{di kannyiri} \tbGE{to be patient} \tbSE{di naŋ} \tbGE{to be scarce} \tbSE{di naaloŋ} \tbGE{to gain wealth or authority} \tbSE{di neɛ} \tbGE{(of a person) to be worthless} \tbSE{di noɔ} \tbGE{to enjoy} \tbSE{di sane} \tbGE{to owe} \tbSE{di sɛrɛɛ} \tbGE{to judge} \tbSE{di suuri} \tbGE{to forgive} \tbSE{di vi} \tbGE{to be ashamed} \tbSE{di yuori} \tbGE{to have an unworthy name} \tbSE{kôɔ naŋ di neɛv} \tbGE{to drown} 
\tbLX{dibogi} \tbHM{} \tbPH{dìbògí} \tbPS{n.~} \tbGE{where the family line begins and where it ends} \tbOP{dibogri} \tbTP{dibogree} 
\tbLX{di daa} \tbHM{} \tbPH{dì dàá} \tbPS{v.~} \tbGE{to haggle, to negotiate a price} \tbPD{-, -, -, -, -, - } 
\tbLX{di dɛgere} \tbHM{} \tbPH{dì dɛ́gɪ́rɪ́} \tbPS{v.~} \tbGE{to suffer} \tbPD{die dɛgere, dire dɛgere, -, -, -, - } 
\tbLX{di dɛmɛ} \tbHM{} \tbPH{dì dɛ́mɛ́} \tbPS{v.~} \tbGE{to chat, to converse} \tbPD{-, -, -, -, -, - } 
\tbLX{di dɔgɛɛ} \tbHM{} \tbPH{di ̀dɔ́gɛ́ɛ́} \tbPS{v.~} \tbGE{to suffer} \tbPD{-, -, -, -, -, - } 
\tbLX{die} \tbHM{} \tbPH{dìè} \tbPS{n.~} \tbSN \tbGE{room} \tbSN \tbGE{compound, house} \tbOP{deri} \tbTP{deree} 
\tbLX{diebie} \tbHM{} \tbPH{dìébíè} \tbPS{n.~} \tbSN \tbGE{cat} \tbSN \tbGE{son of the house} \tbOP{diebiiri} \tbTP{diebiiree} \tbVA{yiribie}
\tbLX{diebugo} \tbHM{} \tbPH{dìèbùgó} \tbPS{n.~} \tbGE{one who knows nothing outside home (untravelled)} \tbOP{diebugri} \tbTP{diebugree} \tbVA{diene}
\tbLX{dieli} \tbHM{} \tbPH{dìèlì} \tbPS{v.~} \tbGE{to float} \tbPD{dielee, diele, diele, dielibe, dielaa, dielibe } 
\tbLX{dieneɛ} \tbHM{} \tbPH{dìènɪ́ɛ́} \tbPS{n.~} \tbGE{a close relative} \tbOP{dienoba} \tbVA{diebugo}
\tbLX{di faŋa} \tbHM{} \tbPH{dì fáŋá} \tbPS{v.~} \tbGE{wild power} \tbPD{-, -, -, -, -, - } 
\tbLX{difudeɛ} \tbHM{} \tbPH{dìfùdɪ́ɛ́} \tbPS{n.~} \tbGE{cheap} \tbTP{difudeerɛɛ} 
\tbLX{diganaa} \tbHM{} \tbPH{dìgánáà} \tbPS{} \tbGE{bedroom} 
\tbLX{digaŋene} \tbHM{} \tbPH{dìgàŋɪ̀nɪ́} \tbPS{n.~} \tbGE{gluttony} 
\tbLX{digaŋgaŋ} \tbHM{} \tbPH{dìgàŋgáŋ} \tbPS{n.~} \tbGE{craftiness} 
\tbLX{digi} \tbHM{} \tbPH{digi} \tbPS{v.~} \tbSN \tbGE{to chase, to run after} \tbSN \tbGE{to drive away} \tbPD{digee, digire, digre, digribe, digiraa, digiribe } 
\tbLX{digoo} \tbHM{} \tbPH{dìgòò} \tbPS{n.~} \tbGE{a disease affecting bean plants} 
\tbLX{digbogbo} \tbHM{} \tbPH{dìgbògbò} \tbPS{adj.~} \tbGE{gluttonous, eating too much} 
\tbLX{dii} \tbHM{} \tbPH{díì} \tbPS{v.~} \tbSN \tbGE{to press down e.g. with a heavier object} \tbXV{A teɛ da le dii la a lɔɔre ka noba perre.} \tbXE{The tree fell on the car and injured some people.} \tbSN \tbGE{to preserve (bean seed in ashes)} \tbXV{A yidaampɔge dii la a bɛŋɛ ka a na sa᷉a᷉.} \tbXE{The head of the women preserved the bean seed in ash to prevent it from getting destroyed by insects.} \tbSN \tbGE{to pile up} \tbXV{A boɔre la dii taa do teɛ a dasɛrɛɛ.} \tbXE{The bags are packed on top of each other up to the ceiling.} \tbPD{diiee, diire, dire, diiribe/diiribee, diiraa, - } 
\tbLX{dii} \tbHM{} \tbPH{díí} \tbPS{v.~} \tbSN \tbGE{to store grain in ashes} \tbXV{A yidaampɔge dii la a bɛŋɛ ne tampɛloŋ ka a na ta sa᷉a᷉.} \tbXE{The head of the women preserved the bean seed in ash to prevent it from getting destroyed by insects.} \tbSN \tbGE{plural of dii (to pile objects)} \tbPD{diiee, diire, dire, diiribe/diiribee, diiraa, diiribe } 
\tbLX{diibu} \tbHM{} \tbPH{dííbú} \tbPS{n.~} \tbGE{food} \tbTP{diiree} 
\tbLX{di kannyiri} \tbHM{} \tbPH{dì kànɲìrì} \tbPS{v.~} \tbGE{to be patient} \tbPD{-, -, -, -, -, - } 
\tbLX{dikpaare} \tbHM{} \tbPH{dìkpáárɪ́} \tbPS{n.~} \tbGE{back of a room} \tbOP{dikpaɛ} \tbTP{dikpaarɛɛ} 
\tbLX{dikpeɛraa} \tbHM{} \tbPH{dìkpɪ́ɛ́ráá} \tbPS{n.~} \tbGE{living room} \tbOP{dikpeɛre} 
\tbLX{dilɔɔraa} \tbHM{} \tbPH{dìlɔ́ɔ́ràá} \tbPS{n.~} \tbGE{a room thatched with wood and earth} \tbOP{dilɔɔre} \tbTP{ dilɔɔrɛ} 
\tbLX{di naaloŋ} \tbHM{} \tbPH{dì náálʊ́ŋ} \tbPS{v.~} \tbGE{to ascend to the throne} \tbPD{-, -, -, -, -, - } 
\tbLX{di naŋe} \tbHM{} \tbPH{dì náŋ} \tbPS{v.~} \tbGE{to be scarce} \tbPD{-, -, -, -, -, - } 
\tbLX{dindigili} \tbHM{} \tbPH{díndíglí} \tbPS{n.~} \tbGE{electric fish} \tbOP{dindigile} \tbTP{dindigilee} \tbVA{dindigiri}
\tbLX{dindigiri} \tbHM{} \tbPH{díndígirí} \tbPS{n.~} \tbGE{catfish, a type of fish that gives you an electric shock when touched} \tbOP{dindige/dindigire} \tbTP{dindigiree} \tbVA{dindigli}
\tbLX{di neɛ} \tbHM{} \tbPH{dì nɪ́ɛ́} \tbPS{v.~} \tbGE{to be hopeless} \tbPD{-, -, -, -, -, - } 
\tbLX{dipanaa} \tbHM{} \tbPH{dìpànáá} \tbPS{n.~} \tbGE{plastered room} \tbOP{dipanne} \tbTP{dipannɛɛ} 
\tbLX{dire} \tbHM{} \tbPH{dìré} \tbPS{v.~} \tbGE{to be sharp} \tbPD{diee, dire, dire, diribe, diraa, dire } 
\tbLX{diruŋ} \tbHM{} \tbPH{dìrúŋ} \tbPS{n.~} \tbSN \tbGE{food, plenty of food} \tbSN \tbGE{feast} 
\tbLX{di sane} \tbHM{} \tbPH{dì sánɪ̀} \tbPS{v.~} \tbGE{to owe, to be indebted to} \tbPD{-, -, -, -, -, - } 
\tbLX{di seebo} \tbHM{} \tbPH{dì sɪ́ɪ́bʊ́} \tbPS{v.~} \tbGE{to harvest, to collect (honey from a hive)} \tbPD{-, -, -, -, -, - } 
\tbLX{di sɛrɛɛ} \tbHM{} \tbPH{dì sɛ̀rɛ́ɛ́} \tbPS{v.~} \tbGE{to judge} \tbPD{-, -, -, -, -, - } 
\tbLX{disogɔ} \tbHM{} \tbPH{dìsʊ́gɔ́} \tbPS{n.~} \tbGE{middle room} \tbOP{disogere} \tbTP{disogerɛɛ} \tbVA{disoga}
\tbLX{di suuri} \tbHM{} \tbPH{dì súúrí} \tbPS{v.~} \tbGE{to forgive} \tbPD{-, -, -, -, -, - } 
\tbLX{diteenɛ} \tbHM{} \tbPH{dìtɪ́ɪ́nɛ́} \tbPS{n.~} \tbGE{mediator, spokesman in a marriage agreement} \tbOP{diteenemɛ} \tbTP{diteenɛɛ} 
\tbLX{di vi} \tbHM{} \tbPH{dì ví} \tbPS{v.~} \tbGE{to be ashamed} \tbPD{-, -, -, -, -, - } 
\tbLX{divuo} \tbHM{} \tbPH{dìvùó} \tbPS{n.~} \tbGE{abandoned room; room without a roof} \tbOP{divori} \tbTP{ divoree} 
\tbLX{do} \tbHM{} \tbPH{dò} \tbPS{v.~} \tbGE{to roll along the ground (circular object)} \tbPD{doe, duoro, duoro, duori, duoraa, duoro } 
\tbLX{do} \tbHM{} \tbPH{dó} \tbPS{v.~} \tbSN \tbGE{to climb up, to ascend} \tbSN \tbGE{to soar} \tbSN \tbGE{to mate, to crossbreed (nonhuman)} \tbSN \tbGE{to climb out of the ground into a new nest (trapped termites)} \tbPD{doe, duoro, duoro, duoribo, duoraa, duoribo } \tbSE{do saa} \tbGE{to be high} 
\tbLX{dõo} \tbHM{} \tbPH{dʋ̂̀ʋ} \tbPS{adv.~} \tbGE{hazy, misty} 
\tbLX{doba} \tbHM{} \tbPH{dʊ́bà} \tbPS{n.~} \tbGE{island} \tbOP{dobare} \tbTP{dobarɛɛ} 
\tbLX{dobaa} \tbHM{} \tbPH{dòbáá} \tbPS{n.~} \tbGE{pig (domesticated)} \tbOP{dobaare} \tbTP{dobaarɛɛ} 
\tbLX{dobadaa} \tbHM{} \tbPH{dòbàdáá} \tbPS{n.~} \tbGE{male pig} \tbOP{dobadaare} \tbTP{dobadaarɛɛ} \tbVA{dodaa}
\tbLX{dobagbɛre} \tbHM{} \tbPH{dòbàgbɛ́rɪ̀} \tbPS{n.~} \tbGE{the mane of a pig} \tbOP{dogbɛɛ} \tbTP{dogbɛɛrɛɛ} 
\tbLX{dobanɛne} \tbHM{} \tbPH{dòbànɛ́nɪ̀} \tbPS{n.~} \tbGE{pork} \tbOP{dobanɛmɛ} \tbTP{dobanɛmɛɛ/dobanɛnnɛɛ} 
\tbLX{dobapɔle} \tbHM{} \tbPH{dòbàpɔ́lɪ̀} \tbPS{n.~} \tbGE{young  pig} \tbOP{dobapɔlɔ} \tbTP{ dopɔllɛɛ} \tbVA{dopɔle}
\tbLX{dobasugo} \tbHM{} \tbPH{dòbàsúgò} \tbPS{n.~} \tbGE{herd of swine} \tbOP{dobasugri} \tbTP{dobasugree} 
\tbLX{dodaa} \tbHM{} \tbPH{dòdáá} \tbPS{n.~} \tbGE{wild male pig} \tbOP{dodaare} \tbTP{dodaarɛɛ} 
\tbLX{dodoo} \tbHM{} \tbPH{dʋ̀dʋ́ʋ́} \tbPS{adj.~} \tbGE{large and bloated} 
\tbLX{dodoo} \tbHM{} \tbPH{dʊ̀dʊ́ʊ́} \tbPS{n.~} \tbGE{an illness that causes bloating of the whole body} \tbOP{dodoore} \tbTP{doddoorɛɛ} 
\tbLX{doge} \tbHM{} \tbPH{dʋ́gɪ̀} \tbPS{n.~} \tbGE{large pot} \tbOP{dogre} \tbTP{dogrɛɛ.} 
\tbLX{doge} \tbHM{} \tbPH{dʋ́gɪ́} \tbPS{v.~} \tbSN \tbGE{to boil, to cook} \tbSN \tbGE{to brew} \tbSN \tbGE{to become very hot} \tbPD{dogɛɛ, dogerɔ, dogerɔ, dogereba, dogeraa, dogerɔ } 
\tbLX{dogebile} \tbHM{} \tbPH{dʋ́gɪ́bílé} \tbPS{n.~} \tbGE{small pot} \tbOP{dogebilii} \tbTP{dogebillee} 
\tbLX{dogekoma} \tbHM{} \tbPH{dʋ́gkʋ́má} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} 
\tbLX{dogele} \tbHM{} \tbPH{dʋ́glɪ̀} \tbPS{v.~} \tbGE{to put over a fire} \tbPD{dogelɛɛ, dogerɔ, dogelɔ, dogelebɔ, dogelɔ, dogelebɔ } 
\tbLX{dogelee} \tbHM{} \tbPH{dʊ́gléé} \tbPS{n.~} \tbGE{small pot} \tbOP{dugili} 
\tbLX{dogi} \tbHM{} \tbPH{dògì} \tbPS{v.~} \tbGE{to remove an object from the top of something} \tbXV{O dogi la a kuuri na naŋ pare a dankyini zu.} \tbXE{S/he has removed the stone from the top of the wall.} \tbPD{dogee, dogiro, dogiro, dogribo, dogiraa, dogiribo } 
\tbLX{dogo} \tbHM{} \tbPH{dʊ̀gʊ́} \tbPS{n.~} \tbGE{a type of dance} \tbTP{dogrɛɛ} 
\tbLX{dogodogo} \tbHM{} \tbPH{dógódògò} \tbPS{n.~} \tbGE{power of charm in magic} \tbOP{dogodogri} \tbTP{dogodogree} 
\tbLX{dole} \tbHM{} \tbPH{dʋ̀lɪ̀} \tbPS{v.~} \tbGE{to be unaware} \tbPD{dolɛɛ, dollɔ, dollɔ, dolebɔ, dollaa, dolebɔ } 
\tbLX{dole} \tbHM{} \tbPH{dʊ́lɪ́} \tbPS{v.~} \tbGE{to raise (one's head) to face upwards (when moving through tall grass)} \tbPD{dolɛɛ, dollɔ, dollɔ, dolebɔ, dolaa, dolebɔ } 
\tbLX{doledole} \tbHM{} \tbPH{dʊ̀lɪ̀dʊ́lɪ́} \tbPS{v.~} \tbGE{to swing one's head around} \tbPD{doledolɛɛ, doledolɔ, doledolɔ, doldolebɔ, doledolaa, doledolebɔ } 
\tbLX{dolee} \tbHM{} \tbPH{dolee} \tbPS{n.~} \tbGE{small wild pig} \tbOP{dobili} 
\tbLX{doli} \tbHM{} \tbPH{dólí} \tbPS{n.~} \tbGE{a type of tree whose bark is used as a disinfectant for washing pito pots} \tbOP{dolo} \tbTP{dolee/dollee} 
\tbLX{doloŋ} \tbHM{} \tbPH{dʊ̀lʊ̀ŋ} \tbPS{n.~} \tbGE{right side} \tbOP{dolle} \tbTP{dollɛɛ} 
\tbLX{domadoma} \tbHM{} \tbPH{dʊ̀màdʊ̀mà} \tbPS{adv.~} \tbGE{bloated and soft} 
\tbLX{dombaala} \tbHM{} \tbPH{dʊ̀mbààlá} \tbPS{n.~} \tbSN \tbGE{sick animal} \tbSN \tbGE{domesticated animal} \tbOP{dombaale} \tbTP{dombaalɛɛ.} 
\tbLX{dombɛroŋ} \tbHM{} \tbPH{dʊ̀mbɛ́rʊ̂ŋ} \tbPS{n.~} \tbSN \tbGE{large animal} \tbSN \tbGE{famous animal} \tbOP{dombɛrɛ} \tbTP{ dombɛrɛɛ/dombɛrrɛɛ.} 
\tbLX{dombɛroŋ} \tbHM{} \tbPH{dʊ̀mbɛ́rʊ́ŋ} \tbPS{n.~} \tbGE{large animal;} \tbOP{dombɛrɛ} \tbTP{dombɛrɛɛ/dombɛrrɛɛ} 
\tbLX{domm} \tbHM{} \tbPH{dòmm} \tbPS{adv.~} \tbGE{tall and big} 
\tbLX{domɔ} \tbHM{} \tbPH{dʊ̀mɔ́} \tbPS{n.~} \tbGE{totem; the sacred bird, animal, or insect of a clan} \tbOP{donne} \tbTP{domɛɛ} \tbVA{domɔ}
\tbLX{dompane} \tbHM{} \tbPH{dʋ̀mpánɪ̀} \tbPS{n.~} \tbGE{a type of snake that is short and poisonous} \tbOP{dompama} \tbTP{ dompannɛɛ.} 
\tbLX{dompelaa} \tbHM{} \tbPH{dʊ̀mpɪ̀làá} \tbPS{n.~} \tbSN \tbGE{white animal} \tbSN \tbGE{an ordinary animal (believed not to have any spiritual powers)} \tbOP{dompeɛle} \tbTP{dompeɛlɛɛ} 
\tbLX{dompɛge} \tbHM{} \tbPH{dʋ̀mpɛÕÅgɪ́} \tbPS{n.~} \tbGE{shell of an animal} \tbXV{ pl.} \tbXE{dompɛgere.} 
\tbLX{dompoll} \tbHM{} \tbPH{dʊ̀mpʊ̀ll} \tbPS{n.~} \tbGE{a material thing believed to be owed to a child from birth} \tbOP{dompolɔ} \tbTP{dompolɛɛ/dompollɛɛ} 
\tbLX{dompoɔ} \tbHM{} \tbPH{dʋ̀mpʋ́ɔ́} \tbPS{n.~} \tbGE{pregnant animal} \tbXV{pl.} \tbXE{} \tbOP{dompoore} \tbTP{ dompoorɛɛ.} 
\tbLX{dondaa} \tbHM{} \tbPH{dʋ̀ndáá} \tbPS{n.~} \tbGE{male animal} \tbXV{ pl.} \tbXE{} \tbOP{dondaare} \tbTP{ dondaarɛɛ.} 
\tbLX{dondigiraa} \tbHM{} \tbPH{dʊ̀ndígráá} \tbPS{n.~} \tbGE{something that drives mosquitoes away} \tbOP{dondigiri} \tbTP{dondigiree} 
\tbLX{dondigire} \tbHM{} \tbPH{dʊ̀ndígré} \tbPS{n.~} \tbGE{an insect-repellent plant used in keeping mosquitoes at bay} \tbOP{dondigiri} \tbTP{ dondigiree} 
\tbLX{dondoli} \tbHM{} \tbPH{dóndólí} \tbPS{n.~} \tbGE{dry land} \tbOP{dondolo} \tbTP{dondolee} 
\tbLX{dondolo} \tbHM{} \tbPH{dòndóló} \tbPS{n.~} \tbGE{planks used as seats (usually in front of a house)} 
\tbLX{donɛɛ} \tbHM{} \tbPH{dʊnɛɛ} \tbPS{n.~} \tbGE{humankind} \tbOP{n/a} \tbTP{n/a} \tbVA{andonɛɛ}
\tbLX{donfaa} \tbHM{} \tbPH{dʋnfaa} \tbPS{n.~} \tbGE{dangerous animal} \tbXV{ pl.} \tbXE{} \tbOP{donfaare} \tbTP{ donfaarɛɛ.} 
\tbLX{donforoŋ} \tbHM{} \tbPH{dʊ̀nfʊ̀rʊ̀ŋ} \tbPS{adv.~} \tbGE{large/big and puffed} \tbXV{A pɔge e la bɛroŋ donforoŋ lɛ.} \tbXE{The woman is big and puffed.} 
\tbLX{donkyeenɛ} \tbHM{} \tbPH{dùnʧɪ́ɪ́nɛ́} \tbPS{n.~} \tbGE{shepherd, herdsman} \tbOP{donkyeenemɛ} \tbTP{donkyeemɛɛ} 
\tbLX{donnyaŋaa} \tbHM{} \tbPH{dʋ́nɲáŋáá} \tbPS{n.~} \tbGE{adult female animal} \tbOP{donyaŋene} \tbTP{donyaŋenɛɛ} 
\tbLX{donnyaŋaa} \tbHM{} \tbPH{dòɲáŋáá} \tbPS{n.~} \tbGE{wild female boar} \tbOP{donnyaŋene} \tbTP{donnyaŋenɛɛ} 
\tbLX{donsaana} \tbHM{} \tbPH{dʋ́nsáánà} \tbPS{n.~} \tbGE{strange animal} \tbXV{ pl.} \tbXE{} \tbOP{donsaama} \tbTP{ donsaanɛɛ.} 
\tbLX{donseɛ} \tbHM{} \tbPH{donsɪɛ} \tbPS{n.~} \tbGE{the waist of an animal} \tbXV{ pl.} \tbXE{} \tbOP{donseere} \tbTP{ donseerɛɛ.} 
\tbLX{donsɔgelaa} \tbHM{} \tbPH{dʊ̀nsɔ́gláá} \tbPS{n.~} \tbSN \tbGE{black animal} \tbSN \tbGE{an animal believed to have spiritual powers} \tbOP{donsɔgelɔ} 
\tbLX{donsugo} \tbHM{} \tbPH{dʊnsugo} \tbPS{n.~} \tbGE{large herds of animals} \tbOP{donsugri} \tbTP{donsugree} 
\tbLX{doŋa} \tbHM{} \tbPH{dʊ́ŋá} \tbPS{n.~} \tbGE{animal} \tbOP{donne} \tbTP{donnɛɛ} \tbVA{doŋɔ}
\tbLX{doŋa} \tbHM{} \tbPH{dʊ́ŋá} \tbPS{n.~} \tbGE{mosquito} \tbOP{donne} \tbTP{donnɛɛ} \tbVA{doŋɔ}
\tbLX{doŋe} \tbHM{} \tbPH{dʊ̀ŋ} \tbPS{v.~} \tbGE{to bite} \tbSN \tbGE{to sting} \tbPD{doŋɛɛ, donɔ, donɔ, donemɔ, donaa, donemɔ } 
\tbLX{doŋe} \tbHM{} \tbPH{dʋ́ŋ} \tbPS{v.~} \tbGE{to be at close range} \tbXV{A saa da doŋe te la ka te zo fa᷉a᷉ kyɛ ka poɔ te.} \tbXE{The rain was so close we could not escape and therefore were caught in the downpour.} \tbPD{doŋɛɛ, donnɔ, donnɔ, donemɔ, donnaa, donemɔ } 
\tbLX{doŋenokono} \tbHM{} \tbPH{dʊ́ŋɪ́nʊ́kʊ́nʊ́} \tbPS{n.~} \tbSN \tbGE{a type of tree whose leaves are edible} \tbSN \tbGE{the leaves of this tree} \tbTP{doŋenokonɛɛ} 
\tbLX{doŋguolaa} \tbHM{} \tbPH{dʊ̀ŋgúóláà} \tbPS{n.~} \tbGE{an animal that has been raised, a domestic animal} \tbOP{doŋguoli} 
\tbLX{doŋguolo} \tbHM{} \tbPH{dʊ̀ŋgúólò} \tbPS{n.~} \tbGE{animals that are shared between an original owner and a caretaker who raises them} \tbVA{boŋguolo}
\tbLX{doŋi} \tbHM{} \tbPH{dǒŋ} \tbPS{n.~} \tbGE{a Muslim religious festival} \tbOP{donni} \tbTP{donnee} 
\tbLX{doŋkaraa} \tbHM{} \tbPH{dʋŋkaraa} \tbPS{n.~} \tbGE{a strong-smelling plant used for chasing away mosquitoes} 
\tbLX{doŋkommiruu} \tbHM{} \tbPH{dʊ̀ŋkómmírúú} \tbPS{n.~} \tbGE{a type of tiny mosquito (bites only during day)} \tbSG{doŋkommiruu} \tbOP{doŋkommie} \tbTP{doŋkommiree/doŋkommiiree} 
\tbLX{doŋkɔɔseɛ} \tbHM{} \tbPH{dʋ̀ŋkɔ̀ɔ̀sɪ́ɛ́} \tbPS{n.~} \tbGE{wasp; a stinging flying insect that resembles a bee but longer} \tbOP{doŋkɔɔseere} \tbTP{doŋkɔɔseerɛɛ} 
\tbLX{doŋkuuree} \tbHM{} \tbPH{dʋkuuree} \tbPS{n.~} \tbGE{a type of short, very poisonous snake} \tbXV{ pl.} \tbXE{doŋkuuri/doŋkuuree.} 
\tbLX{doogɔ} \tbHM{} \tbPH{dʊ́ʊ́gɔ́} \tbPS{n.~} \tbGE{plenty of enjoyment (food and drink)} \tbTP{doogɛɛ} \tbVA{dooga}
\tbLX{doore} \tbHM{} \tbPH{dʊ̀ʋ̀rɪ̀} \tbPS{v.~} \tbGE{to continuously heave chunks of load or matter} \tbXV{Ba doore la ko᷉ɔ̂ lɛɛ te bale ba zaa.} \tbXE{They heaved water until they became tired.} \tbPD{doorɛɛ, doorɔ, doorɔ, doorebɔ, dooraa, doorebɔ } 
\tbLX{doɔ} \tbHM{} \tbPH{dʋ́ɔ̀} \tbPS{v.~} \tbGE{to raise one's head, to look up} \tbXV{O doɔ la o zu dayeni kyɛ leɛ gaŋ} \tbXE{S/he raised his/her head once and laid back down.} \tbPD{doɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\tbLX{doɔ} \tbHM{} \tbPH{dʋ́ɔ̀} \tbPS{v.~} \tbGE{to cause to become worse (behaviour)} \tbXV{A dɔɔ ŋmɛ la a bie doɔ bare.} \tbXE{The man worsened the situation by beating the child.} \tbPD{doɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\tbLX{doɔ} \tbHM{} \tbPH{dʋ̀ɔ̀} \tbPS{v.~} \tbGE{to ooze; to overflow} \tbXV{A baa paalɛɛ la doɔrɔ.} \tbXE{The river is overflowing its banks.} \tbPD{doɔɛ, doɔrɔ, -, -, -, - } 
\tbLX{doɔ} \tbHM{} \tbPH{dʋ̀ɔ̀} \tbPS{v.~} \tbGE{to fetch water with a large container} \tbPD{doɔɛ, doɔrɔ, -, -, -, - } 
\tbLX{doɔ} \tbHM{} \tbPH{dʋ̀ɔ́} \tbPS{n.~} \tbSN \tbGE{dawadawa tree} \tbSN \tbGE{fruit of the dawadawa tree (edible)} \tbOP{dɔre} \tbTP{dɔrɛɛ} 
\tbLX{doɔbo} \tbHM{} \tbPH{dʋ́ɔ́bʋ̀} \tbPS{n.~} \tbGE{the act of weeding} 
\tbLX{doɔkpangyogele} \tbHM{} \tbPH{dʋ̀ɔ́kpá́ndzʋ̀glɪ̀} \tbPS{n.~} \tbGE{a large catapillar found on the leaves of the dawadawa tree} \tbOP{doɔkpangyoglɔ} \tbTP{ doɔkpangyoglɛɛ} \tbVA{doɔkpankyogele}
\tbLX{doɔle} \tbHM{} \tbPH{dʋ̀ɔ́lɪ́} \tbPS{n.~} \tbGE{xylophone mallet} \tbOP{doɔlɔ} \tbTP{doɔlɛɛ} 
\tbLX{doɔle} \tbHM{} \tbPH{dʋ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to walk carefully and majestically} \tbPD{doɔlɛɛ, doɔlɔ, doɔlɔ, doɔlebɔ, doɔlaa, doɔlebɔ } 
\tbLX{doɔloŋ} \tbHM{} \tbPH{dʊ̀ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{mudfish (lives in the mud during dry season)} \tbOP{doɔlɔ} \tbTP{doɔlɛɛ} 
\tbLX{doɔnyeraa} \tbHM{} \tbPH{dʊ̀ɔ̀ɲɪ́ráá} \tbPS{n.~} \tbGE{a type of small ant} \tbOP{doɔnyere} \tbTP{doɔnyerɛɛ} 
\tbLX{doɔre} \tbHM{} \tbPH{dʊ̀ɔ̀rɪ́} \tbPS{adj.~} \tbGE{yellow} \tbTP{doɔrɛɛ} 
\tbLX{dopɔge} \tbHM{} \tbPH{dòpɔ́gɪ́} \tbPS{n.~} \tbGE{female bush pig} \tbOP{dopɔgeba} \tbTP{dopɔgerɛɛ} 
\tbLX{dopɔle} \tbHM{} \tbPH{dòpɔ́lɪ̀} \tbPS{n.~} \tbGE{young bush pig} \tbOP{dopɔlɔ} \tbTP{dopɔlɛɛ/dopɔllɛɛ} 
\tbLX{dore} \tbHM{} \tbPH{dʊ̀rɪ̀} \tbPS{n.~} \tbGE{thud; sound of an impact e.g. a heavy log falling} 
\tbLX{doreba} \tbHM{} \tbPH{dʊ́rɪ́bà} \tbPS{n.~} \tbGE{driver} \tbOP{dorebare} \tbTP{dorebarɛɛ} 
\tbLX{doree} \tbHM{} \tbPH{dòrèè} \tbPS{adj.~} \tbGE{large and solid} 
\tbLX{dori} \tbHM{} \tbPH{dòrì} \tbPS{v.~} \tbGE{remove bits of the top} \tbPD{doree, doro, doro, doribo, doraa, doribo } 
\tbLX{dori} \tbHM{} \tbPH{dórí} \tbPS{n.~} \tbGE{heap} \tbOP{doe} \tbTP{doree/dorree} 
\tbLX{dori} \tbHM{} \tbPH{dórí} \tbPS{n.~} \tbGE{joint (stalk of grass)} \tbOP{doe} \tbTP{doree/dorree} 
\tbLX{dosoraa} \tbHM{} \tbPH{dòsáràà} \tbPS{n.~} \tbGE{young female bush pig} \tbOP{dosarre} \tbTP{ donsarɛɛ/donsarrɛɛ} 
\tbLX{dɔ} \tbHM{} \tbPH{dɔ̀} \tbPS{v.~} \tbGE{to spill over e.g. boiling water from yam being cooked over fire} \tbPD{dɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\tbLX{dɔ} \tbHM{} \tbPH{dɔ̀} \tbPS{v.~} \tbGE{to pay the last bride price} \tbPD{dɔɛ, doɔrɔ, doɔrɔ, doɔrebɔ, doɔraa, doɔrebɔ } 
\tbLX{dɔ} \tbHM{} \tbPH{dɔ́} \tbPS{v.~} \tbGE{to weed} \tbPD{dɔɛ, doɔrɔ, doɔrɔ, doɔreba, doɔraa, doɔrebɔ } 
\tbLX{dɔdɔgeroŋ} \tbHM{} \tbPH{dɔ́dɔ́grʊ̂ŋ} \tbPS{n.~} \tbGE{wretchedness, bad treatment} 
\tbLX{dɔɛ} \tbHM{} \tbPH{dɔ́ɛ́} \tbPS{n.~} \tbGE{last bride price} 
\tbLX{dɔge} \tbHM{} \tbPH{dɔ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to bear (child), to give birth} \tbSN \tbGE{to flower (grass pollen)} \tbSN \tbGE{take something from the top of another} \tbXV{A bie dɔge la a nɛne a saabo zu.} \tbXE{The child has taken the meat which was on top of the food.} \tbPD{dɔgɛɛ, dɔgrɔ, dɔgrɔ, dɔgreba, dɔgraa, dɔgreba } 
\tbLX{dɔgele} \tbHM{} \tbPH{dɔ̀glɪ̀} \tbPS{v.~} \tbSN \tbGE{to place on top of something} \tbXV{De a nɛne dɔgele dankyini zu ka baare ta wa de o.} \tbXE{Place the meat on the wall so that dogs do not eat it.} \tbSN \tbGE{to exceed, to be extra} \tbXV{A bie yɔɔ la a libie ka a dɔgele ka a seŋ ka fo leɛ de a dɔgelɔ ko o.} \tbXE{The child has overpaid and so the extra money should be returned to him.} \tbPD{dɔglɛɛ, dɔglɔ, dɔglɔ, dɔgleba, dɔglaa, - } 
\tbLX{dɔgele} \tbHM{} \tbPH{dɔ̀glɪ̀} \tbPS{v.~} \tbGE{to mix beans with flour} \tbPD{dɔglɛɛ, dɔglɔ, dɔglɔ, dɔgleba, dɔglaa, - } 
\tbLX{dɔgere} \tbHM{} \tbPH{dɔ́grɪ̀} \tbPS{v.~} \tbGE{to punish, to make to suffer} \tbPD{dɔgerɛɛ, dɔgerɔ, dɔgerɔ, dɔgerebɔ, dɔgeraa, dɔgerebɔ } 
\tbLX{dɔgeroŋ} \tbHM{} \tbPH{dɔ̀grʊ̂ŋ} \tbPS{n.~} \tbGE{descent} 
\tbLX{dɔgeroŋ} \tbHM{} \tbPH{dɔ́grʊ̂ŋ} \tbPS{n.~} \tbGE{penalty, punishment} 
\tbLX{dɔgerɔ} \tbHM{} \tbPH{dɔ́grɔ} \tbPS{n.~} \tbGE{relative (blood)} \tbOP{dɔgerebɔ} \tbTP{dɔgerebɛɛ} 
\tbLX{dɔgerɔlenne} \tbHM{} \tbPH{dɔ̀grɔ́lénné} \tbPS{n.~} \tbGE{ear of maize with multicoloured kernels} \tbTP{dɔgerɔlennee} 
\tbLX{dɔgɛɛ} \tbHM{} \tbPH{dɔ́gɛ́ɛ́} \tbPS{n.~} \tbGE{suffering} 
\tbLX{dɔkele} \tbHM{} \tbPH{dɔ̀kélé} \tbPS{n.~} \tbGE{kernels from dawdaw fruit} \tbTP{dɔkelee} \tbVA{tokele}
\tbLX{dɔkone} \tbHM{} \tbPH{dɔ̀kʊ́nɪ́} \tbPS{n.~} \tbGE{kenkey (thick corn paste shaped into balls and boiled in leaves)} \tbOP{dɔkona} \tbTP{dɔkonnɛɛ} \tbVA{dakone}
\tbLX{dɔle} \tbHM{} \tbPH{dɔ́lɪ́} \tbPS{adv.~} \tbGE{pop out smoothly e.g. from something slippery} 
\tbLX{dɔlle} \tbHM{} \tbPH{dɔ́llɪ̀} \tbPS{v.~} \tbGE{to stretch, to lengthen} \tbPD{dɔllɛɛ, dɔllɔ, dɔllɔ, dɔlebɔ, dɔllaa, dɔlebɔ } 
\tbLX{dɔlɔ} \tbHM{} \tbPH{dɔ́lɔ́́} \tbPS{n.~} \tbGE{way of doing something} \tbOP{dɔlɔdɔlɔ} 
\tbLX{dɔlɔdɔlɔ} \tbHM{} \tbPH{dɔ̀lɔ̀dɔ̀lɔ̀} \tbPS{adv.~} \tbGE{slimy and slippery} 
\tbLX{dɔndɔŋ-karaa} \tbHM{} \tbPH{dɔ́ndɔ̀ŋ-kàràá} \tbPS{n.~} \tbGE{chief enemy} \tbOP{dɔndɔŋ-karre} \tbTP{dɔndɔŋ-karɛɛ} 
\tbLX{dɔndɔre} \tbHM{} \tbPH{dɔ̀ndɔ̀rɪ́} \tbPS{n.~} \tbGE{curd; coagulated milk} \tbOP{dɔndɔrre} \tbTP{dɔndɔrɛɛ} 
\tbLX{dɔndɔre} \tbHM{} \tbPH{dɔ̀ndɔ̀rɪ́} \tbPS{n.~} \tbGE{shade caused by clouds coming between the earth and the sun} \tbOP{dɔndɔrre} \tbTP{dɔndɔrɛɛ} 
\tbLX{dɔŋe} \tbHM{} \tbPH{dɔ̌ŋ} \tbPS{n.~} \tbGE{enmity, hatred} \tbOP{dɔnne} \tbTP{dɔnnɛɛ} 
\tbLX{dɔŋe} \tbHM{} \tbPH{dɔ̀ŋ} \tbPS{v.~} \tbGE{to be lean, to be thin} \tbPD{dɔŋɛɛ, dɔnnɔ, dɔnnɔ, dɔnemɔ, dɔnnaa, dɔnemɔ } 
\tbLX{dɔɔ} \tbHM{} \tbPH{dɔ̀ɔ̀} \tbPS{v.~} \tbSN \tbGE{to squat} \tbSN \tbGE{to bow (as in greeting)} \tbSN \tbGE{to excrete} \tbSN \tbGE{to perch} \tbPD{dɔɔɛ, dɔɔrɔ, dɔɔrɔ, dɔɔrebɔ, dɔɔraa, dɔɔrebɔ } 
\tbLX{dɔɔ} \tbHM{} \tbPH{dɔ́ɔ́} \tbPS{n.~} \tbSN \tbGE{man} \tbSN \tbGE{friend of the same sex} \tbOP{dɔbɔ/dɔba} \tbTP{dɔɔrɛɛ} 
\tbLX{dɔ᷉ɔ} \tbHM{} \tbPH{dɔ̀᷈ɔ̀} \tbPS{v.~} \tbSN \tbGE{to disturb, to move (rustle)} \tbSN \tbGE{to stir, to mix} \tbSN \tbGE{to worry} \tbPD{dɔ᷉ɔɛ, dɔɔnɔ, dɔɔnɔ, dɔɔnema, dɔɔnaa, dɔɔnema } 
\tbLX{dɔ᷉ɔ} \tbHM{} \tbPH{dɔ̀᷈ɔ̀} \tbPS{v.~} \tbGE{to sprain e.g. a leg} \tbPD{dɔ᷉ɔɛ, dɔɔnɔ, dɔɔnɔ, dɔɔnema, dɔɔnaa, dɔɔnema } 
\tbLX{dɔɔbile} \tbHM{} \tbPH{dɔ̀ɔ̀bílé} \tbPS{n.~} \tbGE{boy} \tbOP{dɔɔbilii} 
\tbLX{Dɔɔda᷈a᷈} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{a male name} \tbOP{dɔɔda᷈a᷈nɪ᷈} \tbTP{dɔɔda᷈a᷈nɛ᷈ɛ᷈} 
\tbLX{dɔɔkɔɔteɛlɔɔ} \tbHM{} \tbPH{dɔ̀ɔ̀kɔ̀ɔ̀tɪ́ɛ́lɔ́ɔ́} \tbPS{n.~} \tbGE{a squatting kick game} 
\tbLX{dɔɔ-la-pɔge} \tbHM{} \tbPH{dɔ̀ɔ̀-la-́pɔ́gɪ́} \tbPS{n.~} \tbGE{a man with feminine features} \tbTP{dɔɔ-la-pɔgrɛɛ} 
\tbLX{dɔɔlee} \tbHM{} \tbPH{dɔ̀ɔ̀léé} \tbPS{n.~} \tbGE{boy} \tbOP{dɔɔbilii} \tbTP{dɔɔbilee} \tbVA{dɔɔbile}
\tbLX{dɔɔloŋ} \tbHM{} \tbPH{dɔ̀ɔ̀lʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{bravery, dexterity} \tbSN \tbGE{penis (figurative)} 
\tbLX{dɔɔne} \tbHM{} \tbPH{dɔ̀ɔ̀nɪ́} \tbPS{n.~} \tbSN \tbGE{noise} \tbXV{N woŋ la dɔɔne a die poɔ, neɛkaŋa be la be.} \tbXE{I have heard some noise in the room, there must be someone in there.} \tbSN \tbGE{movement} \tbXV{Te nyɛ la a baala nɔgbɛmɛ dɔɔne kyɛ te ba woŋi a yɛlɛ o naŋ yeli.} \tbXE{We saw some movement of the patient’s lips but we did not hear what s/he said.} \tbSN \tbGE{disturbance, worry} \tbXV{A pɔge yeli ka o ba boɔrɔ dɔɔne ka anaŋ so ka o yi.} \tbXE{The woman said she did not want trouble and that is why she divorced.} 
\tbLX{dɔɔnyaŋaa} \tbHM{} \tbPH{dɔ̀ɔ̀ɲáŋáá} \tbPS{n.~} \tbGE{old man} \tbOP{ dɔɔnyaŋene} \tbTP{dɔɔnyaŋenɛɛ} 
\tbLX{dɔɔ-ɔre} \tbHM{} \tbPH{dɔ̀ɔ̀-ɔ́rɪ́} \tbPS{n.~} \tbGE{infertile man} \tbTP{dɔɔ-ɔrɛɛ} \tbVA{dɔɔ-are}
\tbLX{dɔɔta} \tbHM{} \tbPH{dɔ́ɔ́tà} \tbPS{n.~} \tbSN \tbGE{doctor} \tbSN \tbGE{hospital, clinic} \tbOP{dɔɔtare} \tbTP{dɔɔtarɛɛ} 
\tbLX{dɔɔtayiri} \tbHM{} \tbPH{dɔ́ɔ́táyírí} \tbPS{n.~} \tbGE{hospital} \tbOP{dɔɔtayie} \tbTP{dɔɔtayiree} 
\tbLX{dɔre} \tbHM{} \tbPH{dɔ̀rɪ̀} \tbPS{v.~} \tbGE{to advise, to caution} \tbPD{dɔrɛɛ, dɔrɔ, dɔrɔ, dɔrebɔ, dɔraa, dɔrebɔ } 
\tbLX{dɔre} \tbHM{} \tbPH{dɔ́rɪ́} \tbPS{n.~} \tbGE{fruits of doɔ} \tbTP{dɔrɛɛ} 
\tbLX{dɔrekabakoma} \tbHM{} \tbPH{dɔ̀rɪ̀kábákʊ́má} \tbPS{n.~} \tbGE{one who incites hatred} \tbOP{dɔrekabakomamine} 
\tbLX{dɔsaŋkpare} \tbHM{} \tbPH{dɔ́sáŋkparɪ́} \tbPS{n.~} \tbGE{peels of dawadawa fruits} \tbTP{dɔsaŋkparɛɛ} \tbVA{dɔseŋkpare}
\tbLX{dɔsempeɛlɛ} \tbHM{} \tbPH{dɔ̀sɪ́mpɪ́ɛ́lɛ́} \tbPS{n.~} \tbGE{immature dawadawa fruits} \tbSG{dɔsempeɛloo} \tbOP{dɔsempeɛle} \tbTP{dɔsempeɛlɛɛ} 
\tbLX{dɔzɔŋe} \tbHM{} \tbPH{dɔ̀zɔ̂ŋ} \tbPS{n.~} \tbGE{daw-daw flour; pulp from daw-daw fruit} \tbOP{dɔzɔnne} \tbTP{dɔzɔnnɛɛ} \tbVA{dɔzɔŋa}
\tbLX{dɔzunni} \tbHM{} \tbPH{dɔ̀zúnnì} \tbPS{n.~} \tbGE{daw-daw seeds} \tbSG{dɔzunnuu} \tbOP{dɔzunni} \tbTP{dɔzunnee} 
\tbLX{du} \tbHM{} \tbPH{dú} \tbPS{v.~} \tbSN \tbGE{to rub against} \tbSN \tbGE{to iron (clothes)} \tbPD{due, duuro, duuro, duuribo, duuraa, duuribo } 
\tbLX{dudugiri} \tbHM{} \tbPH{dùdùgrí} \tbPS{n.~} \tbGE{a box made of grass} \tbOP{dudugo} \tbTP{dudugiree} 
\tbLX{dugi} \tbHM{} \tbPH{dúgí} \tbPS{n.~} \tbGE{swim} \tbVA{dúú}
\tbLX{dugi} \tbHM{} \tbPH{dùgì} \tbPS{v.~} \tbGE{to be eaten up to powder e.g. wood infected with bugs} \tbPD{dugee, dugiro, -, -, -, - } \tbVA{dúú}
\tbLX{dukuree} \tbHM{} \tbPH{dúkúréé} \tbPS{n.~} \tbGE{large sores caused by scabies} \tbTP{dukurree} \tbVA{dikuree}
\tbLX{duli} \tbHM{} \tbPH{dúlí} \tbPS{v.~} \tbSN \tbGE{to work strenuously on a difficult task} \tbSN \tbGE{to eat a large amount (fig.)} \tbPD{dulee, dulo, dulo, dulibo, dulaa, dulibo } 
\tbLX{duluŋ} \tbHM{} \tbPH{dùlûŋ} \tbPS{n.~} \tbGE{a type of large bird} \tbOP{ dulli} \tbTP{dullee} 
\tbLX{dunduli} \tbHM{} \tbPH{dùndùlí} \tbPS{n.~} \tbSN \tbGE{caterpillar} \tbSN \tbGE{worm} \tbSN \tbGE{maggot} \tbOP{dundulo} \tbTP{dundulee/dundullee} \tbVA{kpaŋkyori, kpaŋkyobo}
\tbLX{dundulivarekpɛnyoɔre} \tbHM{} \tbPH{dúndùlìváríkpɛ́ɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{small green caterpillar that springs from place to place} \tbOP{dundulivarekpɛnyɔɛ} 
\tbLX{dundulkyoo} \tbHM{} \tbPH{dùndùlʧóò} \tbPS{n.~} \tbGE{cocoon} \tbOP{dundulkyoori} \tbTP{dundulkyooree} 
\tbLX{dundulvarkpɛnyoɔre} \tbHM{} \tbPH{dùndùlvárkpɛ̀ɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{a type of small green caterpillar that moves by loops} \tbOP{dundulvarkpɛnyɔɛ} 
\tbLX{dunduŋkolo} \tbHM{} \tbPH{dùndǔŋkòlò} \tbPS{n.~} \tbGE{fried yam wedges} \tbSG{dunduŋkoluu} \tbTP{dunduŋkolee} 
\tbLX{duni} \tbHM{} \tbPH{dúni} \tbPS{n.~} \tbGE{knee} \tbOP{dumo} \tbTP{dunee/dunnee} 
\tbLX{dunselee} \tbHM{} \tbPH{dùnséléé} \tbPS{n.~} \tbGE{kneecap} \tbOP{ dunsebilii} \tbTP{dunsebilee} 
\tbLX{duŋi} \tbHM{} \tbPH{dúŋ} \tbPS{n.~} \tbGE{type of weevil (found in wood )} \tbOP{dunni} \tbTP{dunee/dunnee} 
\tbLX{duo} \tbHM{} \tbPH{dùó} \tbPS{n.~} \tbGE{wild pig, boar} \tbOP{dori} \tbTP{doree/dorree} 
\tbLX{duo} \tbHM{} \tbPH{dùò} \tbPS{v.~} \tbSN \tbGE{to be seated all over the place in great numbers (patients at consulting room)} \tbSN \tbGE{to prepare malt for brewing local beer;} \tbPD{duoe, duoro, duoro, duoribo, duoraa, duoribo } 
\tbLX{duogo} \tbHM{} \tbPH{dúógó} \tbPS{n.~} \tbGE{giant} \tbOP{duogori} \tbTP{duogoree} 
\tbLX{Duoŋ} \tbHM{} \tbPH{dúóŋ} \tbPS{n.~} \tbGE{town in the Upper West Region noted for treatment of bone fractures} 
\tbLX{duoraa} \tbHM{} \tbPH{dùóráa} \tbPS{n.~} \tbSN \tbGE{open space for a gathering;} \tbSN \tbGE{resting place for sheherds and their animals} \tbOP{duori} \tbTP{duoree} 
\tbLX{duori} \tbHM{} \tbPH{dúórì} \tbPS{v.~} \tbSN \tbGE{to urinate} \tbSN \tbGE{to ejaculate} \tbSN \tbGE{to set high up (child on shoulders)} \tbSN \tbGE{to raise yam mounds} \tbSN \tbGE{to set a bow ready for use} \tbPD{duoree, duoro, duoro, duoribo, duoraa, duoribo } 
\tbLX{duori} \tbHM{} \tbPH{dùòrì} \tbPS{v.~} \tbGE{to set out in heaps e.g. tomatoes for sale} \tbPD{duoree, duoro, duoro, duoribo, duoraa, duoribo } 
\tbLX{duoro} \tbHM{} \tbPH{dùòró} \tbPS{n.~} \tbGE{news} \tbTP{duoree} 
\tbLX{duor-tere} \tbHM{} \tbPH{dùòr-tɪ́rɪ́} \tbPS{n.~} \tbGE{annoucement, broadcast, information} \tbOP{duor-terre} \tbTP{duorterrɛɛ} 
\tbLX{duor-terɛ} \tbHM{} \tbPH{dùòr-tɪ́rɛ́} \tbPS{n.~} \tbGE{one who gives information, messenger, newscaster} \tbOP{duor-terebɛ} \tbTP{duor-terebɛɛ} 
\tbLX{duorunlaa} \tbHM{} \tbPH{dúórùnláá} \tbPS{n.~} \tbGE{chamber-pot} \tbOP{duorunlaare} \tbTP{duorunlaarɛɛ} 
\tbLX{duorunwoɔ} \tbHM{} \tbPH{dúórùnwʊ̀ɔ̀} \tbPS{n.~} \tbGE{bladder} \tbOP{duorunwoore} \tbTP{duorunwoorɛɛ} 
\tbLX{duorunze᷉e᷉} \tbHM{} \tbPH{dúórùnzɪ́᷉ɪ́᷉} \tbPS{n.~} \tbGE{bilharziasis} \tbTP{duorunzeenɛɛ} 
\tbLX{duoruŋ} \tbHM{} \tbPH{dúórǔŋ} \tbPS{n.~} \tbGE{urine} 
\tbLX{duoruŋkɔnsɔgerɔ} \tbHM{} \tbPH{dúórǔŋkɔ́nsɔ́grɔ́} \tbPS{n.~} \tbGE{the condition of having difficulty passing urine} \tbOP{duoruŋkɔnsɔgre} \tbTP{duoruŋkɔnsɔgrɛɛ} 
\tbLX{duulaa} \tbHM{} \tbPH{dùùláà} \tbPS{n.~} \tbSN \tbGE{row} \tbSN \tbGE{line (poetry)} \tbOP{duuli} 
\tbLX{e} \tbHM{} \tbPH{ɪ̀} \tbPS{v.~} \tbSN \tbGE{to act, to do} \tbXV{Te na e o la velaa.} \tbXE{We shall do him good.} \tbSN \tbGE{to let} \tbXV{E ka te gaa pampana.} \tbXE{Let us go now.} \tbSN \tbGE{to fetch} \tbXV{A dɔɔ e la a kyi ko a pɔge.} \tbXE{The man has fetched  millet for the woman.} \tbSN \tbGE{to be} \tbXV{A bie e la kpeɛŋaa.} \tbXE{The child is strong.} \tbPD{eɛɛ, erɛ, erɛ, erebɛ, eraa, erebɛ } 
\tbLX{eele} \tbHM{} \tbPH{ɪ́ɪ́lɪ́} \tbPS{n.~} \tbSN \tbGE{horn} \tbSN \tbGE{antenna} \tbOP{eelɛ} \tbTP{eelɛɛ} 
\tbLX{eɛ} \tbHM{} \tbPH{ìɛ} \tbPS{v.~} \tbSN \tbGE{to pinch} \tbSN \tbGE{to hint} \tbXV{Ba eɛ la a yɛlɛ fe᷉e᷉ ka N woŋ.} \tbXE{They have given me a hint of the situation.} \tbPD{eɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ } 
\tbLX{e᷉ɛ} \tbHM{} \tbPH{ɪ́᷉ɛ́᷉} \tbPS{n.~} \tbGE{} \tbSG{eɛnoo} \tbOP{e᷉ɛ} \tbTP{enɛɛ} \tbVA{eɛne}
\tbLX{e᷈ɛ} \tbHM{} \tbPH{ɪ́᷉ɛ̀᷉} \tbPS{v.~} \tbSN \tbGE{to mark the foundation of a building} \tbSN \tbGE{to cut a very small piece} \tbPD{eɛɛ, eɛnɛ, eɛnɛ, eɛnemɛ, eɛnaa, eɛnemɛ } 
\tbLX{eɛ eŋe} \tbHM{} \tbPH{ɪ̀ɛ̀ ɪ̀ŋ} \tbPS{v.~} \tbGE{to flee, to run away} \tbXV{A bie zu pɔre la maŋgo ka a daana waana ka o eɛ eŋ.} \tbXE{The boy stole a mango and when he saw the owner coming he took to his heels.} \tbPD{eɛ eŋɛɛ, eɛrɛ ennɛ, eɛ ennɛ, eɛ enemɛ, eɛ ennaa, - } 
\tbLX{eɛne} \tbHM{} \tbPH{ɪ̀ɛ̀nɪ̀} \tbPS{n.~} \tbGE{hippopotamus} \tbOP{eɛmɛ} \tbTP{eɛnɛɛ} 
\tbLX{eɛne} \tbHM{} \tbPH{ɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{claw} \tbSG{eɛnoo} \tbOP{eɛmɛ} \tbTP{eɛnɛɛ} 
\tbLX{eɛne} \tbHM{} \tbPH{ɪ̀ɛ̌nɪ́} \tbPS{n.~} \tbGE{a type of deer} \tbOP{eɛmɛ} \tbTP{eɛmɛɛr} 
\tbLX{eɛne} \tbHM{} \tbPH{ɪ̀ɛ́nɪ́} \tbPS{adj.~} \tbGE{tiny, e.g. fingernail} \tbOP{eɛmɛ} \tbTP{eɛmɛɛ} 
\tbLX{eɛne} \tbHM{} \tbPH{ɪ̀ɛ̀ɪ̀} \tbPS{v.~} \tbGE{to claw e.g a cat climbing a wall} \tbPD{-, -, -, -, -, - } 
\tbLX{e᷉ɛo} \tbHM{} \tbPH{ɪ̀᷉ɛ̀᷉ʊ́᷉} \tbPS{n.~} \tbGE{type of deer} \tbOP{eɛne} \tbTP{eɛnɛɛ} 
\tbLX{eɛraa} \tbHM{} \tbPH{ɪ́ɛ̀ráá} \tbPS{n.~} \tbGE{hook, e.g. fishhook} \tbOP{eɛre} \tbTP{eɛrɛɛ} 
\tbLX{eɛre} \tbHM{} \tbPH{ɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to wipe away e.g. dirt, tears} \tbPD{eɛrɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ } 
\tbLX{eɛre} \tbHM{} \tbPH{ɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to claw at} \tbPD{eɛrɛɛ, eɛrɛ, eɛrɛ, eɛrebɛ, eɛraa, eɛrebɛ } 
\tbLX{eɛre-miri} \tbHM{} \tbPH{ɪ̀ɛ̀rɪ̀-mírì} \tbPS{n.~} \tbGE{fishing line} \tbOP{eɛre-mie} \tbTP{eɛre-miiree} 
\tbLX{e-la} \tbHM{} \tbPH{ɪ̀-là} \tbPS{v.~} \tbGE{to tease} \tbPD{e-laɛ, erɛ laara, e-laara, e-laareba, e-laaraa, e-lareba } 
\tbLX{e-laara} \tbHM{} \tbPH{ɪ̀-láárá} \tbPS{n.~} \tbGE{one who teases} \tbOP{e-lareba} \tbTP{e-larebɛɛ} 
\tbLX{embaare} \tbHM{} \tbPH{imbaari} \tbPS{n.~} \tbGE{laziness} \tbTP{embarɛɛ} \tbVA{entuo}
\tbLX{emii} \tbHM{} \tbPH{ɪ́míì} \tbPS{n.~} \tbGE{act of desperation; impatience} \tbTP{emmiinee} 
\tbLX{emmaa} \tbHM{} \tbPH{ɪ̀máà} \tbPS{n.~} \tbGE{the condition of being in a good state} 
\tbLX{emmaaroŋ} \tbHM{} \tbPH{ɪ̀mmáárʊ́ŋ} \tbPS{n.~} \tbGE{health} \tbOP{emmaara} \tbTP{emmaarɛɛ} 
\tbLX{emmii} \tbHM{} \tbPH{ɪ̀mmìì} \tbPS{n.~} \tbGE{a body that is in a painful condition} \tbOP{emmiime} \tbTP{emmiinee} 
\tbLX{emmiiruŋ} \tbHM{} \tbPH{ɪ̀mmììrûŋ} \tbPS{n.~} \tbSN \tbGE{a body in pain} \tbSN \tbGE{stunted body} \tbTP{emmiiree} 
\tbLX{emmiiruŋ} \tbHM{} \tbPH{ɪ̀mmììrûŋ} \tbPS{adj.~} \tbGE{stunted} 
\tbLX{Emɔ} \tbHM{} \tbPH{ɪ́mɔ̀} \tbPS{n.~} \tbSN \tbGE{the name of a clan in Southern Dagao} \tbSN \tbGE{appellation of this group} \tbSG{emoɔloo} \tbOP{emoɔlɔ} \tbTP{emoɔlɛɛ} 
\tbLX{empoɔne} \tbHM{} \tbPH{ɪ́mpʊ́ɔ́nɪ́} \tbPS{n.~} \tbGE{ocean} \tbOP{mpoɔnɛɛ} \tbVA{po}
\tbLX{endaa} \tbHM{} \tbPH{ɪ̀ndàá} \tbPS{n.~} \tbGE{frame (of a person's body)} \tbOP{endaare} \tbTP{endarɛɛ/endaarɛɛ} 
\tbLX{endaa waaloŋ} \tbHM{} \tbPH{ɪ̌ndàá-wàálʊ̂ŋ} \tbPS{n.~} \tbGE{skin colour, appearance of the body} 
\tbLX{enfuomaala} \tbHM{} \tbPH{ɪ̀nfùòmáálá} \tbPS{n.~} \tbGE{one who makes pictures} \tbOP{enfuomaaleba} \tbTP{enfuomaalebɛɛ} 
\tbLX{enfuomaale} \tbHM{} \tbPH{ɪ̀nfùòmàálɪ̀} \tbPS{n.~} \tbGE{the act of making a picture (draw)} 
\tbLX{enfuoni} \tbHM{} \tbPH{ɪ̀nfúóní} \tbPS{n.~} \tbGE{image, photograph, picture} \tbOP{enfuomo} \tbTP{enfuonee} 
\tbLX{enfuori} \tbHM{} \tbPH{ɪ̀nfùòrì} \tbPS{n.~} \tbGE{the act of bewitching; the act of casting a spell} 
\tbLX{enne} \tbHM{} \tbPH{inni} \tbPS{n.~} \tbGE{honour} \tbTP{ennɛɛ} \tbVA{eŋene}
\tbLX{ennoɔ} \tbHM{} \tbPH{ɪ̀nnʊ̀ɔ́} \tbPS{n.~} \tbGE{happiness} \tbOP{ennoore} \tbTP{ennoorɛɛ} 
\tbLX{ensɔgelaa} \tbHM{} \tbPH{ɪ́nsɔ́gláá} \tbPS{n.~} \tbGE{black skin} \tbOP{ensɔglɔ} 
\tbLX{entaa} \tbHM{} \tbPH{ɪ̀ntáá} \tbPS{n.~} \tbGE{mutual respect} 
\tbLX{entage} \tbHM{} \tbPH{ɪ̌ntágɪ́} \tbPS{n.~} \tbGE{muscular body} \tbOP{entagre} \tbTP{entagrɛɛ} 
\tbLX{entole} \tbHM{} \tbPH{ɪ́ntʊ́lɪ̀} \tbPS{n.~} \tbGE{fever, rise in body temperature} \tbOP{entolɔ/entola} \tbTP{entolɛɛ} 
\tbLX{entoloŋ} \tbHM{} \tbPH{ɪ́ntʊ́lʊ̂ŋ} \tbPS{n.~} \tbGE{fever} \tbOP{entola} \tbTP{entollɛɛ} 
\tbLX{entukpalaa} \tbHM{} \tbPH{ɪ̂ntùkpàláà} \tbPS{n.~} \tbGE{extreme laziness} \tbOP{entukpalle} \tbTP{entukpalɛɛ} 
\tbLX{entuo} \tbHM{} \tbPH{ɪ̀ntúó} \tbPS{n.~} \tbGE{laziness} \tbOP{entuuri} \tbTP{entuuree} \tbVA{embaare}
\tbLX{enzeɛ} \tbHM{} \tbPH{ǐnzɪ̀ɛ́} \tbPS{n.~} \tbSN \tbGE{light skin} \tbSN \tbGE{leprosy} \tbOP{enzeere} \tbTP{enzeerɛɛ} 
\tbLX{eŋ} \tbHM{} \tbPH{ɪ́ŋ} \tbPS{part.~} \tbGE{as} \tbXV{Maa eŋ koŋ gaa.} \tbXE{As for me I wont go.} 
\tbLX{eŋe} \tbHM{} \tbPH{ɪ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to put in} \tbXV{o ɔŋ la kôɔ eŋe a doge poɔ.} \tbXE{S/he has put water into the pot.} \tbSN \tbGE{to put on (shoes, ring, spectacles)} \tbSN \tbGE{to respect} \tbXV{A seŋe ka bibile maŋ eŋe neŋkpoŋ.} \tbXE{The young should respect the elderly.} \tbPD{eŋɛɛ, ennɛ, ennɛ, enemɛ, ennaa, - } 
\tbLX{eŋe daa} \tbHM{} \tbPH{ɪ̀ŋ dàà} \tbPS{v.~} \tbGE{to set a date} \tbPD{-, -, -, -, -, - } 
\tbLX{eŋe daa} \tbHM{} \tbPH{ɪ̀ŋ dàà} \tbPS{v.~} \tbGE{to beat} \tbPD{-, -, -, -, -, - } 
\tbLX{eŋene} \tbHM{1} \tbPH{ɪ́ŋɪ́nɪ́} \tbPS{n.~} \tbGE{main hole leading to a rodent's living room in its burrow} \tbXV{A teŋɛ deme zaa taa la eŋene kora a naa.} \tbXE{All the people in town have respect for the chief.} \tbOP{enne} \tbTP{eŋenɛɛ} 
\tbLX{eŋene} \tbHM{2} \tbPH{ɪ́ŋɪ́nɪ́} \tbPS{n.~} \tbGE{respect, regard for, honour} \tbXV{N teɛre ka o deɛ la N eŋene.} \tbXE{I think he has just targeted me.} \tbTP{ennɛɛ} \tbVA{enne}
\tbLX{eŋɛ} \tbHM{} \tbPH{ɪ́ŋɛ́} \tbPS{pron.~} \tbGE{what, who} \tbXV{Boŋ eŋɛ ka fo wa?} \tbXE{What have you come for?.} 
\tbLX{eŋɛ} \tbHM{} \tbPH{ɪ́ŋɛ́} \tbPS{n.~} \tbGE{body (surface of)} \tbOP{enne} \tbTP{ennɛɛ} 
\tbLX{eŋgane} \tbHM{} \tbPH{ɪ̀ŋgánɪ̀} \tbPS{n.~} \tbGE{skin (human body)} \tbSG{eŋganoo} \tbOP{eŋgama} \tbTP{eŋgannɛɛ} 
\tbLX{Eŋgelɛ} \tbHM{} \tbPH{ɪ́ŋgɪ̀lɛ̀} \tbPS{n.~} \tbGE{England} \tbTP{eŋgelɛɛ} 
\tbLX{Eŋgilisi} \tbHM{} \tbPH{ɪ́ŋgìlìsì} \tbPS{n.~} \tbGE{English} \tbOP{ɪŋlisiri} \tbTP{ɪŋgilisiree} 
\tbLX{eŋkɔbɔ} \tbHM{} \tbPH{ɪ̀ŋkɔ́bɔ́} \tbPS{n.~} \tbGE{skeleton} \tbVA{eŋkɔba}
\tbLX{eŋkpaŋkpoli} \tbHM{} \tbPH{ɪ́ŋkpáŋkpòlì} \tbPS{n.~} \tbGE{naked body} \tbVA{eŋkpoŋkpoli, eŋkpoli}
\tbLX{eŋkpoli} \tbHM{} \tbPH{iŋkpoli} \tbPS{n.~} \tbGE{naked body} \tbOP{eŋkpolo} \tbTP{eŋkpolee} \tbVA{eŋkpoŋkpoli, eŋkpaŋkpoli}
\tbLX{eŋkpoŋkpoli} \tbHM{} \tbPH{ɪ̀ŋkpóŋkpólì} \tbPS{n.~} \tbGE{naked} \tbOP{eŋkpoŋkpolo} \tbTP{eŋkpoŋkpolee} \tbVA{eŋkpoli, eŋkpaŋkpoli}
\tbLX{eŋ noɔre} \tbHM{} \tbPH{ɪ̀ŋ nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to promise} \tbPD{-, -, -, -, -, - } 
\tbLX{eŋ poɔ} \tbHM{} \tbPH{ɪ̀ŋ pʊ́ɔ́} \tbPS{v.~} \tbGE{to empregnate} \tbPD{-, -, -, -, -, - } 
\tbLX{eŋyoɔre} \tbHM{} \tbPH{ɪ̀ŋyʊ̀ɔ̀rɪ̀} \tbPS{adj.~} \tbGE{that which does not produce excitement} \tbVA{enyoɔre}
\tbLX{eŋ ze᷉ɛ᷉} \tbHM{} \tbPH{ɪ̀ŋ zɪ᷉̀ɛ᷉} \tbPS{v.~} \tbGE{to punish severely} \tbPD{eŋɛɛ ze᷉ɛ᷉, ennɛ ze᷉ɛ᷉, -, -, -, - } 
\tbLX{eroŋ} \tbHM{} \tbPH{ɪ̀rʊ̂ŋ} \tbPS{n.~} \tbGE{conduct} \tbOP{erre} \tbTP{errɛɛ} 
\tbLX{eroŋ} \tbHM{} \tbPH{ɪ̀rʊ̂ŋ} \tbPS{n.~} \tbGE{verb} \tbOP{erre} \tbTP{errɛɛ} 
\tbLX{Etore} \tbHM{} \tbPH{ítʊ́rɪ̀} \tbPS{n.~} \tbGE{Easter} \tbOP{itorɛɛ} 
\tbLX{etɔ} \tbHM{} \tbPH{ɪ́tɔ̀} \tbPS{n.~} \tbGE{the name of a clan in Southern Dagao} \tbOP{etoɔlɔ} \tbTP{etoɔlɛɛ} 
\tbLX{ewa} \tbHM{} \tbPH{ɪ́wà} \tbPS{n.~} \tbSN \tbGE{a kinship group} \tbSN \tbGE{appellation of this group} \tbSG{ewaaloo} \tbOP{ewaala} \tbTP{ewaalɛɛ} 
\tbLX{ewɛ} \tbHM{} \tbPH{ɪ́wɛ̀} \tbPS{n.~} \tbGE{the name of a clan in Southern Dagao} \tbOP{eweɛlɛ} \tbTP{eweɛlɛɛ} 
\tbLX{ɛbaa} \tbHM{} \tbPH{ɛ́bàà} \tbPS{n.~} \tbGE{crocodile} \tbOP{ɛrre} \tbTP{ɛrrɛɛ} 
\tbLX{ɛɛ᷉} \tbHM{} \tbPH{ɛɛ᷉} \tbPS{use word.~} \tbGE{} \tbVA{o᷉o}
\tbLX{ɛlemɛ} \tbHM{} \tbPH{ɛ̀lɪ̀mɛ́} \tbPS{n.~} \tbGE{shellfish} \tbOP{ɛlemɛre} \tbTP{ɛlemɛɛ} 
\tbLX{ɛl-ɛloo} \tbHM{} \tbPH{ɛ̀l-ɛ̀lʊ́ʊ́} \tbPS{interjection.~} \tbGE{it serves you well} 
\tbLX{ɛnne} \tbHM{} \tbPH{ɛ̀nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to suckle, to nurse} \tbSN \tbGE{to suck e.g. breast or finger} \tbPD{ɛnnɛɛ, ɛnnɛ, ɛnnɛ, ɛnemɛ, ɛnnaa, ɛnemɛ } 
\tbLX{ɛre} \tbHM{} \tbPH{ɛ́rɪ́} \tbPS{v.~} \tbGE{to crawl; to slide along on the buttocks} \tbPD{ɛrɛɛ, ɛrɛ, ɛrɛ, ɛrebɛ, ɛraa, ɛrebɛ } 
\tbLX{ɛre} \tbHM{} \tbPH{ɛ́rɪ̀} \tbPS{v.~} \tbGE{to grind} \tbPD{ɛrɛɛ, ɛrɛ, ɛrɛ, ɛrebɛ, ɛraa, ɛrebɛ } 
\tbLX{ɛre} \tbHM{} \tbPH{ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to fill a hole or space tightly to disallow access} \tbPD{ɛrɛɛ, ɛrɛ, ɛrɛ, ɛrebɛ, ɛraa, ɛre-ɛrɛ } 
\tbLX{ɛre-eŋ} \tbHM{} \tbPH{ɛ̀rɪ̀-ɪ̀ŋ} \tbPS{n.~} \tbSN \tbGE{exaggeration} \tbSN \tbGE{metaphor (lit.)} 
\tbLX{ɛregbɛrɛ} \tbHM{} \tbPH{ɛ̀rɪ̀gbɛ̀rɛ̀} \tbPS{adv.~} \tbGE{fleshy and fatty (meat)} 
\tbLX{fa} \tbHM{} \tbPH{fà} \tbPS{v.~} \tbSN \tbGE{to be bad} \tbXV{Ka fooŋ ba kyɛlle a bie ŋa na fa la, bonso ka o ba maŋ kpɛ die wi᷉eu?} \tbXE{If you are not careful, this child will go bad, why doesn't s/he come early?.} \tbSN \tbGE{to become poor} \tbXV{A bie baaloŋ vɛŋɛɛ la ka o tuuli e te᷉e lɛɛ deɛ te fa o zaa.} \tbXE{The child's sickness has caused her to search so much for a cure that she has nothing left.} \tbPD{faɛ, faara, faaraa, fareba, -, fareba } 
\tbLX{fa} \tbHM{} \tbPH{fá} \tbPS{v.~} \tbSN \tbGE{to rob, to cheat} \tbXV{N dɔɔ, fo baŋ ka ka maaŋ yɔɔ libie zaa poɔ a anaŋ ka fo fa ma la eŋ?} \tbXE{My friend, do you know that if I paid anything more, you would have robbed me?.} \tbSN \tbGE{pl. of 'faa' (snatch from someone)} \tbXV{A seŋ ka ba pɔge a nannyige eŋ deŋɛ lɛ, N ba nyɛ teŋɛ na deme o naŋ ba fa.} \tbXE{It will be good if this thief is imprisoned for life, because there is no town he has not robbed.} \tbPD{faɛ, faara, faara, fareba, faaraa, fareba } 
\tbLX{faa} \tbHM{} \tbPH{fáá} \tbPS{adj.~} \tbSN \tbGE{bad} \tbXV{Saa naŋ piili miibu ŋaa, beri ayi ŋa zaa ka te teŋɛ sori ya᷉a na waa faa lɛ.} \tbXE{With the setting in of the rains, the road leading to our village will soon be very bad.} \tbSN \tbGE{ugly} \tbXV{A pɔge veɛlɛ la kyɛ o serɛ faa.} \tbXE{The woman is beautiful but her husband is ugly.} \tbOP{faare} \tbTP{faarɛɛ} 
\tbLX{faa} \tbHM{} \tbPH{fáà} \tbPS{v.~} \tbSN \tbGE{to take by force, to snatch} \tbXV{A bie ŋa yɛlɛ kpeɛŋaa, ka onaŋ sɔre o tɔ bone ka o mɔŋ o maŋ faa la.} \tbXE{This boy is ruthless, for if he begs from his colleague and is refused to give he takes the thing by force.} \tbSN \tbGE{to save} \tbXV{Kɔŋ da kpɛ la a bikpeebɛ ka o kono kyɛ ka pɔge-nimbaaloŋ kaŋ la da faa o ne bondirii.} \tbXE{The orphan was hungry and began crying, but a benevolent woman came to his aid with food.} \tbSN \tbGE{to get free, to get out of one's grip or hold} \tbXV{N nyɔge la a kpa᷉ao kyɛ ka o faa.} \tbXE{I caught the guinea fowl but it slipped free and escaped.} \tbPD{faaɛ, faara, faara, fareba, faaraa, faareba } 
\tbLX{fa᷉a} \tbHM{} \tbPH{fà᷉̀à̀᷉} \tbPS{adv.~} \tbGE{slowly (of a fall)} \tbXV{O kyeelɛɛ la gɛrɛ fa᷉a na te le.} \tbXE{S/he is falling slowly.} 
\tbLX{fa᷉a} \tbHM{} \tbPH{fà᷉à᷉} \tbPS{v.~} \tbSN \tbGE{to be unable} \tbXV{O gaa la o sori na fa᷉a.} \tbXE{S/he has failed to go on her trip.} \tbSN \tbGE{to have no choice} \tbXV{A pɔmpeŋ sa᷉aɛ la ka noba fa᷉a kyɛ sigi baa te ɔŋ ko᷉ɔ.} \tbXE{The water tap is spoilt, so people have no choice but to go the stream to fetch water.} \tbPD{faa᷉ɛ, faana, faana, faanema, faanaa, faa᷉faana } 
\tbLX{fa᷉afa᷉a} \tbHM{} \tbPH{fa᷉́a᷉́fa᷉́a᷉́} \tbPS{n.~} \tbGE{whistle used in times of war or danger} \tbOP{fa᷉afaane} \tbTP{fa᷉afaanɛɛ} 
\tbLX{faa le} \tbHM{} \tbPH{fáà lè} \tbPS{v.~} \tbGE{to drop, to fall} \tbPD{faa lee, faara lere, faa lere, faa leribe, faa leraa, faa leribe } 
\tbLX{Faaloŋ} \tbHM{} \tbPH{fààlʊ̂ŋ} \tbPS{n.~} \tbGE{the name of a village located near Lawra, a District headquarters in the UWR of Ghana} 
\tbLX{faaloŋ} \tbHM{} \tbPH{fààlʊ̂ŋ} \tbPS{n.~} \tbGE{the condition of being bad or wicked} 
\tbLX{faara} \tbHM{} \tbPH{fáárà} \tbPS{n.~} \tbGE{priest} \tbOP{faaramine} \tbTP{faarɛɛ} 
\tbLX{faara} \tbHM{} \tbPH{fáárá} \tbPS{n.~} \tbGE{saviour} \tbOP{faareba} 
\tbLX{faara} \tbHM{} \tbPH{fáárá} \tbPS{n.~} \tbGE{one who robs} \tbOP{faareba} 
\tbLX{faare} \tbHM{} \tbPH{fáárɪ̀} \tbPS{v.~} \tbGE{to intercede, to mediate (in a fight)} \tbPD{faarɛɛ, faara, faara, faareba, faaraa, faareba } 
\tbLX{faare} \tbHM{} \tbPH{fààrɪ̀} \tbPS{v.~} \tbGE{to stroke, to massage} \tbXV{Kɔlaa maŋ nɔŋ ka ba faara o tegitegi lɛ.} \tbXE{A cat likes to be stroded all the time.} \tbPD{faarɛɛ, faara, faara, faareba, faaraa, faareba } 
\tbLX{faare} \tbHM{} \tbPH{fààrɪ̀} \tbPS{n.~} \tbGE{a boys' game of snatching others' arrows} \tbTP{faarɛɛ} 
\tbLX{faaregyɔ} \tbHM{} \tbPH{fààrɪ́ɲɔ́} \tbPS{v.~} \tbGE{to not utter a word; to be quiet} \tbPD{-, -, -, -, -, - } 
\tbLX{fabo} \tbHM{} \tbPH{fàbʊ̀} \tbPS{adv.~} \tbGE{not responding to anything; speechless} 
\tbLX{fabo} \tbHM{} \tbPH{fàbʊ̀} \tbPS{adv.~} \tbGE{tasteless} 
\tbLX{faee} \tbHM{} \tbPH{fàɪ̀ɪ̀} \tbPS{adv.~} \tbGE{empty-looking} 
\tbLX{fafara} \tbHM{} \tbPH{fàfàrà} \tbPS{adv.~} \tbGE{suddenly, without permission (of entering)} 
\tbLX{fage} \tbHM{} \tbPH{fàgɪ̀} \tbPS{adv.~} \tbGE{plenty (and jumbled up), e.g. of different things} \tbXV{Naŋkoɔre boma da pale a daa lɛ fage.} \tbXE{There were many assorted goods in the market.} 
\tbLX{fage} \tbHM{} \tbPH{fágɪ́} \tbPS{v.~} \tbGE{to peck, to eat (of birds)} \tbPD{fagɛɛ, fagera, fagra, fagreba, fageraa, fagreba } 
\tbLX{fagere} \tbHM{} \tbPH{fàgrɪ̀} \tbPS{adv.~} \tbGE{sharp (of a bite)} 
\tbLX{fagoo} \tbHM{} \tbPH{fàgʊ̀ʊ̀} \tbPS{adv.~} \tbGE{full of rubbish, dirty} 
\tbLX{falala} \tbHM{} \tbPH{fàlàlà} \tbPS{adv.~} \tbGE{long, smooth, and straight} \tbVA{falalla}
\tbLX{falala} \tbHM{} \tbPH{fálálá} \tbPS{adj.~} \tbGE{smooth and straight} \tbVA{falalla}
\tbLX{falalla} \tbHM{} \tbPH{fálállá} \tbPS{adv.~} \tbGE{} \tbVA{falala}
\tbLX{falanfuuri} \tbHM{} \tbPH{fálánfúúrí} \tbPS{n.~} \tbGE{lungs} \tbOP{falanfugo} \tbTP{falanfugree} 
\tbLX{fale} \tbHM{} \tbPH{fálɪ̀} \tbPS{v.~} \tbGE{to slap} \tbPD{falɛɛ, falla, falla, faleba, falaa, faleba } 
\tbLX{fale pareŋ} \tbHM{} \tbPH{fálɪ̀ párɪ̀ŋ} \tbPS{v.~} \tbGE{to spank (a child)} \tbPD{-, -, -, -, -, - } 
\tbLX{fambataabo} \tbHM{} \tbPH{fǎmbátáábʊ́} \tbPS{n.~} \tbGE{weakness, powerlessness} 
\tbLX{fananna} \tbHM{} \tbPH{fánánná} \tbPS{adv.~} \tbGE{very clean} 
\tbLX{fane} \tbHM{} \tbPH{fànɪ̀} \tbPS{v.~} \tbGE{to become rough (skin)} \tbPD{fanɛɛ, fana, fana, fanema, fanaa, fanema } 
\tbLX{fane} \tbHM{} \tbPH{fànɪ́} \tbPS{adj.~} \tbGE{rough (skin)} \tbOP{fama} 
\tbLX{fanfane} \tbHM{} \tbPH{fànfánɪ̀} \tbPS{n.~} \tbGE{soap} \tbOP{fanfama} \tbTP{fanfanɛɛ/fanfannɛɛ} 
\tbLX{fanne} \tbHM{} \tbPH{fànnɪ̀} \tbPS{v.~} \tbGE{to sell one's wares at a low price"?} \tbPD{fannɛɛ, fanna, fanna, fanema, fannaa, fanema } 
\tbLX{fanne} \tbHM{} \tbPH{fánnɪ̀} \tbPS{v.~} \tbGE{to think} \tbPD{fannɛɛ, fanna, fanna, fanema, fannaa, - } 
\tbLX{faŋa} \tbHM{} \tbPH{fàŋá} \tbPS{n.~} \tbGE{strength; be strong (physically)} \tbOP{fanne} \tbTP{fannɛɛ} 
\tbLX{faŋa} \tbHM{} \tbPH{fáŋá} \tbPS{adj.~} \tbGE{young; used for calves} \tbTP{fanne} 
\tbLX{Farafara} \tbHM{} \tbPH{fàràfàrà} \tbPS{n.~} \tbSN \tbGE{a person of the Gurunɛ ethnicity} \tbSN \tbGE{the language spoken by this group} \tbSG{farafaroo} \tbOP{farafarɛɛ} 
\tbLX{farata} \tbHM{} \tbPH{fàràtà} \tbPS{} \tbGE{tall and straight} 
\tbLX{fare} \tbHM{} \tbPH{fárɪ́} \tbPS{v.~} \tbSN \tbGE{to foam (yeast in alcohol; discharge from eyes)} \tbSN \tbGE{to boil (porridge)} \tbPD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\tbLX{fare} \tbHM{} \tbPH{fárɪ́} \tbPS{v.~} \tbGE{to build mounds around crops (farming corn)} \tbPD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\tbLX{fare} \tbHM{} \tbPH{fárɪ́} \tbPS{v.~} \tbGE{to pluck leaves in large quantities} \tbPD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\tbLX{fare} \tbHM{} \tbPH{fárɪ̀} \tbPS{v.~} \tbGE{to make bare; (to shave hair off completely)} \tbPD{farɛɛ, fara, fara, fareba, faraa, fareba } 
\tbLX{fare} \tbHM{} \tbPH{fárɪ́} \tbPS{adv.~} \tbGE{completely} \tbXV{A bibiiri zaa zo kulee la fare.} \tbXE{All the children have gone without any left behind.} 
\tbLX{fare bini} \tbHM{} \tbPH{fárɪ́ bíní} \tbPS{v.~} \tbSN \tbGE{shit plenty} \tbSN \tbGE{say digusting things} \tbPD{-, -, -, -, -, - } 
\tbLX{fare yɛlɛ} \tbHM{} \tbPH{fárɪ́ yɛ́lɛ́} \tbPS{v.~} \tbSN \tbGE{to talk a great deal} \tbSN \tbGE{to complain} \tbPD{-, -, -, -, -, - } 
\tbLX{fasane} \tbHM{} \tbPH{fàsánɪ̀} \tbPS{n.~} \tbGE{(with ̀ŋmɛ') thought} \tbXV{A dɔɔ da ŋmɛ la fasane, a wa te yeli ka o ba la boɔrɔ a o pɔge.} \tbXE{The man thought for a while and finally said he did not want his wife.} \tbOP{fasama} \tbTP{fasamɛɛ/fasannɛɛ} 
\tbLX{febo} \tbHM{} \tbPH{fɪ́bʊ́} \tbPS{adv.~} \tbGE{piercingly; the manner a sharp object enters another suddenly and firmly (a thorn in the body)} 
\tbLX{fee} \tbHM{} \tbPH{féé} \tbPS{adv.~} \tbGE{completely (of drinking)} 
\tbLX{fe᷉e} \tbHM{} \tbPH{fɪ᷉́ɪ᷉́} \tbPS{adv.~} \tbSN \tbGE{for a short time; in a short time} \tbSN \tbGE{few} \tbSN \tbGE{small, little} 
\tbLX{feelaa} \tbHM{} \tbPH{fɪ́ɪ́láà} \tbPS{adj.~} \tbGE{young and tender, at the very infant age e.g. chicken, kid} \tbXV{A nɔnyaŋaa neɛ la nɔbil feelaa kaŋa ko.} \tbXE{The mother hen has stepped on one of the very young chicks and killed it.} \tbOP{feelɛ} 
\tbLX{feele} \tbHM{} \tbPH{fɪ̀ɪ̀lɪ̀} \tbPS{v.~} \tbGE{to make tight} \tbPD{feelɛɛ, feelɛ, feelɛ, feelbɛ, feelaa, feelbɛ } 
\tbLX{feele} \tbHM{} \tbPH{fɪ́ɪ́lɪ̀} \tbPS{v.~} \tbGE{to examine} \tbPD{feelɛɛ, feelɛ, feelɛ, feelbɛ, feelaa, feelbɛ } 
\tbLX{feele} \tbHM{} \tbPH{fɪ́ɪ́lɪ́} \tbPS{n.~} \tbGE{tsetse fly} \tbOP{feelɛ} \tbTP{feelɛɛ} 
\tbLX{feelee} \tbHM{} \tbPH{fɪ́ɪ́lɪ́ɪ́} \tbPS{n.~} \tbGE{chance, space to operate} \tbXV{Feelee kye a die poɔ, fo koŋ nyɛ zie biŋi fo sɛŋɛ.} \tbXE{There is no space in the room; you can't find space for your bed.} \tbOP{feelɛ} \tbTP{feelɛɛ} 
\tbLX{feeloo} \tbHM{} \tbPH{fɪ́ɪ́lʊ́ʊ́} \tbPS{adj.~} \tbGE{narrow, tight} \tbOP{feelɛ} 
\tbLX{feere} \tbHM{} \tbPH{fɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to blow one's nose} \tbSN \tbGE{to hold, press, or tie tightly} \tbPD{feerɛɛ, feerɛ, feerɛ, feerebɛ, feeraa, - } 
\tbLX{feerɔɔ} \tbHM{} \tbPH{fɪ̀ɪ̀rɔ̀ɔ̀} \tbPS{adv.~} \tbGE{suddenly disappearing into an opening e.g. into a hole or room} \tbXV{Bayuo to la o beɛre kyɛ zo gaa e die feerɔɔ, o naŋ kpɛ.} \tbXE{Bayuo insulted his elder brother and ran into the house at once.} 
\tbLX{feɛ} \tbHM{} \tbPH{fɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to whip} \tbSN \tbGE{to unshell e.g. rice or corn} \tbPD{feɛɛ, feɛrɛ, feɛrɛ, feɛrebɛ, feɛraa, feɛrebɛ } 
\tbLX{fe᷉ɛ} \tbHM{} \tbPH{fɪ́᷉ɛ́᷉} \tbPS{v.~} \tbGE{to be grown to some extent; to not be small} \tbXV{A da ba seŋ ka o la taa poɔ pampana, a bie na o naŋ taa naŋ ba fe᷉ɛ.} \tbXE{She should not be pregnant now because the baby she has is still too young.} \tbPD{fe᷉ɛɛ, feɛnɛ, feɛnɛ, feɛnemɛ, feɛnaa, - } 
\tbLX{feɛle} \tbHM{} \tbPH{fɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to pass closely by another e.g. a blow missing the target narrowly} \tbSN \tbGE{to pass gas silently} \tbPD{feɛlɛɛ, feɛlɛ, feɛlɛ, feɛlbɛ, feɛlaa, feɛlbɛ } 
\tbLX{feɛle} \tbHM{} \tbPH{fɪ́ɛ́lɪ́} \tbPS{v.~} \tbGE{to sneak out} \tbPD{feɛlɛɛ, feɛlɛ, feɛlɛ, feɛlbɛ, feɛlaa, feɛlbɛ } 
\tbLX{fefeɛ} \tbHM{} \tbPH{fɪ́fɪ̀ɛ́} \tbPS{n.~} \tbGE{crevice} \tbOP{fefeere} \tbTP{fefeerɛɛ} 
\tbLX{feferɛ} \tbHM{} \tbPH{fɪ́fɪ́rɛ́} \tbPS{adj.~} \tbGE{least} \tbXV{Ka fooŋ ba maale fo kyi ŋa biŋi,  fo yeli ka fo erɛ feferɛ ka noore wa la eŋɛ.} \tbXE{If you don't keep your millet well the least move away and fowls swoop on it.} \tbVA{fifire}
\tbLX{fege} \tbHM{} \tbPH{fɪ̀gɪ̀} \tbPS{v.~} \tbGE{to cut out a tiny opening} \tbXV{A daa kpeɛŋaa, a pɛnnepɛnɛ ba toɔ na fege o zie zaa.} \tbXE{The wood is so hard that the carver can hardly make any headway to carve out even a small hole.} \tbPD{fegɛɛ, fegrɛ, -, fegrebɛ, fegraa, fegrebɛ } 
\tbLX{fegele} \tbHM{} \tbPH{fɪ̀gɪ̀lɪ̀} \tbPS{v.~} \tbGE{to be tight} \tbPD{-, feglɛ, -, -, -, - } \tbVA{fogele}
\tbLX{fegere} \tbHM{} \tbPH{fɪ́gɪ́rɪ̀} \tbPS{v.~} \tbGE{sob} \tbPD{fegrɛɛ, fegrɛ, fegrɛ, fegrebɛ, fegraa, fegrebɛ } 
\tbLX{fegɛfegɛ} \tbHM{} \tbPH{fɪ̀gɛ̀fɪ̀gɛ̀} \tbPS{adv.~} \tbGE{flashing intermittently e.g. lightening} \tbVA{felɛfelɛ}
\tbLX{fele} \tbHM{} \tbPH{fɪ̀lɪ́} \tbPS{n.~} \tbGE{sore, wound} \tbOP{felɛ} \tbTP{felɛɛ} 
\tbLX{felɛfelɛ} \tbHM{} \tbPH{fɪ̀lɛ̀fɪ̀lɛ̀} \tbPS{adv.~} \tbGE{flash brightly} 
\tbLX{felɛfelɛ} \tbHM{} \tbPH{fɪ́lɛ́fɪ́lɛ́} \tbPS{adv.~} \tbGE{glittery and fashing with light} 
\tbLX{felko᷉ɔ} \tbHM{} \tbPH{fɪ̀lkʊ̀᷉ɔ́᷉} \tbPS{n.~} \tbGE{pus} \tbTP{felkoɔnɛɛ} 
\tbLX{felmaara} \tbHM{} \tbPH{filmaara} \tbPS{n.~} \tbGE{sores over the body thought to be incurable e.g leprosy} \tbVA{felmaaroŋ}
\tbLX{felmaaroŋ} \tbHM{} \tbPH{fɪ̀lmáárʊ́ŋ} \tbPS{n.~} \tbGE{sore caused by leprosy or yaws} \tbOP{felmaara} \tbTP{felmaarɛɛ} \tbVA{felmaara}
\tbLX{felɔɔ} \tbHM{} \tbPH{fɪ̀lɔ̀ɔ̀} \tbPS{adv.~} \tbGE{of a sparkle} 
\tbLX{felpɛmpɛle} \tbHM{} \tbPH{fɪ̀lpɛ̌mpɛ́lɪ́} \tbPS{n.~} \tbGE{scar} \tbSG{felpɛmpɛloo} \tbOP{felpɛmpɛlɛ} \tbTP{felpɛmpɛlɛɛ/felpɛmpɛllɛɛ} 
\tbLX{felzeɛ} \tbHM{} \tbPH{fɪ̀lzɪ̀ɛ̀} \tbPS{n.~} \tbGE{sore without pus} \tbOP{felzeere} \tbTP{felzeerɛɛ} 
\tbLX{femm} \tbHM{} \tbPH{fɪ́mm} \tbPS{adv.~} \tbGE{getting away fast and far e.g. bullet} 
\tbLX{fenlege} \tbHM{} \tbPH{fɪ́nlɪ́gɪ́} \tbPS{adv.~} \tbGE{laboured e.g. breathing} 
\tbLX{fenne} \tbHM{} \tbPH{fɪ̀nnɪ̀} \tbPS{v.~} \tbGE{to begin to ripen (first few fruits)} \tbPD{ fennɛɛ, fennɛ, fennɛ, fenemɛ, fennaa, - } 
\tbLX{fenne} \tbHM{} \tbPH{fɪ́nnɪ́} \tbPS{v.~} \tbGE{to cut a very small piece} \tbPD{-, -, -, -, -, - } 
\tbLX{fentelɛ} \tbHM{} \tbPH{fɪ̀ntɪ́lɛ̀} \tbPS{n.~} \tbGE{lamp} \tbOP{fentelle} \tbTP{fentelɛɛ/fentellɛɛ} 
\tbLX{fentellaa} \tbHM{} \tbPH{fɪ́ntɪ́lláá} \tbPS{n.~} \tbGE{local lamp} \tbOP{fentellaare} \tbTP{fentellaarɛɛ} 
\tbLX{fentere} \tbHM{} \tbPH{fɪ́ntɪ́rɪ́} \tbPS{n.~} \tbGE{ring e.g. that put on the finger} \tbOP{fenteɛ} \tbTP{fenterɛɛ/fenterrɛɛ} 
\tbLX{fentere} \tbHM{} \tbPH{fɪ́ntɪ́rɪ́} \tbPS{n.~} \tbGE{the act of standing at attention} \tbOP{fenteɛ} \tbTP{fenterɛɛ/fenterrɛɛ} 
\tbLX{fere} \tbHM{} \tbPH{fɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to force, to pester} \tbSN \tbGE{to be morally obliged to do something} \tbXV{A fere ka te sonnɔ nembaalba.} \tbXE{We are obliged to care for the helpless.} \tbPD{ferɛɛ, ferɛ, ferɛ, ferebɛ, feraa, - } \tbVA{fɛre}
\tbLX{fere} \tbHM{} \tbPH{fèrè} \tbPS{n.~} \tbGE{regret; bitter experience} \tbXV{Ka fooŋ wa deɛnɛ ne a badɔgrɔ bile fo na wa yeli ka fere.} \tbXE{If you play around that newborn puppy you will come to regret it.} \tbTP{feree/ferree} 
\tbLX{feree} \tbHM{} \tbPH{féréé} \tbPS{n.~} \tbGE{the fruit of the w??kamp?re plant} \tbVA{wɔŋkampɔre}
\tbLX{fereko} \tbHM{} \tbPH{fɪ́rɪ́kʊ́} \tbPS{n.~} \tbGE{a type of bawaa dance performed by moving to and fro to the sides (left and right)} \tbOP{ferekore} \tbTP{ferekorɛɛ} \tbVA{feroo}
\tbLX{feroo} \tbHM{} \tbPH{fɪ́rʊ́ʊ́} \tbPS{n.~} \tbGE{force} 
\tbLX{feroo} \tbHM{} \tbPH{fir??} \tbPS{n.~} \tbGE{} \tbVA{fereko}
\tbLX{feroo toma} \tbHM{} \tbPH{fɪ́rʊ́ʊ́ tʊ́má} \tbPS{n.~} \tbGE{duty, obligation} 
\tbLX{fɛɛ} \tbHM{} \tbPH{fɛ́ɛ́} \tbPS{adv.~} \tbSN \tbGE{happening in some manner e.g. completely} \tbXV{A baala nyu la a tẽẽ a de goɔoŋ e fɛɛ.} \tbXE{The patient took his medication and slept soundly.} \tbSN \tbGE{becoming airless instantly (tyre)} \tbXV{Pimbiri kyɔge la a kuriwiri kɔbɔ ka o e fɛɛ, o naŋ yi.} \tbXE{The tyre of the bicycle was pierced by a nail and deflated at once.} 
\tbLX{fɛɛ} \tbHM{} \tbPH{fɛ̀ɛ̀} \tbPS{adv.~} \tbGE{deflating completely of quiet sound of pent up air or gas escaping} 
\tbLX{fɛge} \tbHM{} \tbPH{fɛ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to indent} \tbSN \tbGE{to become one-eyed} \tbPD{fɛgɛɛ, fɛgrɛ, fɛgrɛ, fɛgrebɛ, fɛgraa, fɛgfɛgrɛ } 
\tbLX{fɛge} \tbHM{} \tbPH{fɛ́gɪ́} \tbPS{n.~} \tbGE{one-eyed person} \tbOP{fɛgre} \tbTP{fɛgrɛɛ} 
\tbLX{fɛge} \tbHM{} \tbPH{fɛ́gɪ́} \tbPS{adv.~} \tbGE{fall lightly, such that you don't hear anything e.g. dry leaf} 
\tbLX{fɛgelɛgɛ} \tbHM{} \tbPH{fɛgelɛgɛ} \tbPS{adv.~} \tbGE{out of energy} \tbVA{fɛgetɛgɛ}
\tbLX{fɛgetɛgɛ} \tbHM{} \tbPH{fɛ́gtɛ́gɛ́} \tbPS{adv.~} \tbGE{completely out of air or energy} 
\tbLX{fɛlefɛle} \tbHM{} \tbPH{fɛ́lfɛ́l} \tbPS{adj.~} \tbGE{tasteless} 
\tbLX{fɛlɛlɛ} \tbHM{} \tbPH{fɛ́lɛ́lɛ́} \tbPS{adv.~} \tbGE{utterly, completely} \tbXV{A dɔɔ ku᷉u maa la maa eŋ noɔre fɛlɛlɛ, ŋmenaa ŋa ka maa ne o puori taa daa poɔ} \tbXE{The man's death has utterly surprised me, because we greeted each other this afternoon in the marketplace.} 
\tbLX{fɛmm} \tbHM{} \tbPH{fɛ̀mm} \tbPS{adv.~} \tbGE{lying flat e.g. vegetation or hair} 
\tbLX{fɛraa} \tbHM{} \tbPH{fɛ́ráá} \tbPS{adj.~} \tbGE{narrow, tight} \tbOP{fɛrre} 
\tbLX{fɛre} \tbHM{} \tbPH{fɛ̀rɪ̀} \tbPS{v.~} \tbGE{to be sqeezed up in between} \tbPD{fɛrɛɛ, fɛrɛ, fɛrɛ, fɛrbɛ, fɛraa, fɛrbɛ } 
\tbLX{fɛre} \tbHM{} \tbPH{fɛ́rɪ́} \tbPS{adv.~} \tbGE{sound of drop of something small and light} 
\tbLX{fɛre-taa-kyɛ-moorɔ} \tbHM{} \tbPH{fɛ́rɪ̀tááʧɛ́mʊ́ʊ́rɔ́} \tbPS{n.~} \tbGE{one who has something but lies to others that one does not have it} \tbVA{fɛre-bin-kyɛ-moorɔ/fɛre-taa-kyɛ-moora}
\tbLX{fieli} \tbHM{} \tbPH{fíélí} \tbPS{adj.~} \tbGE{slim and flat (nape)} 
\tbLX{fieli} \tbHM{} \tbPH{fíélì} \tbPS{v.~} \tbGE{to fail} \tbPD{fielee, fiele, fiele, fielbe, fielaa, fielbe } 
\tbLX{fieluu} \tbHM{} \tbPH{fíélúú} \tbPS{n.~} \tbGE{the act of failing} 
\tbLX{fifire} \tbHM{} \tbPH{fífírè} \tbPS{adv.~} \tbGE{scarcely out of side than something happens} \tbXV{A dɔɔ yeli ka o erɛ fifire ka a kɔlaa de a nɛne.} \tbXE{No sooner had the man left than the cat snatched the meat.} \tbVA{feferɛ}
\tbLX{figi} \tbHM{} \tbPH{fígí} \tbPS{v.~} \tbGE{to throw out one's hand to signal to stop} \tbPD{figee, figire, figire, figiribe, figiraa, figiribe } 
\tbLX{figiraa} \tbHM{} \tbPH{fígràá} \tbPS{n.~} \tbGE{fan} \tbOP{figiri} \tbTP{figiree} 
\tbLX{figiri} \tbHM{} \tbPH{fígrí} \tbPS{v.~} \tbSN \tbGE{to fan} \tbSN \tbGE{to flap one's wings} \tbPD{figee, figire, figire, figiribe, figiraa, figiribe } 
\tbLX{fii} \tbHM{} \tbPH{fìì} \tbPS{adv.~} \tbGE{manner of movement e.g. of a heavy object in the air} 
\tbLX{filoo} \tbHM{} \tbPH{fílóó} \tbPS{adv.~} \tbGE{at glance} \tbXV{N nyɛ la a bie filoo kyɛ ka o kpɛ a lɔɔre.} \tbXE{I saw the child at a glance before he entered the lorry.} 
\tbLX{finni} \tbHM{} \tbPH{fìnnì} \tbPS{v.~} \tbGE{to cheat} \tbPD{finnee, finne, finne, finime, finnaa, finime } \tbVA{fuuli}
\tbLX{firimeŋ} \tbHM{} \tbPH{fìrìmèŋ} \tbPS{adv.~} \tbGE{appearing neatly cut e.g. growth} 
\tbLX{fo} \tbHM{} \tbPH{fʊ̀} \tbPS{pron.~} \tbSN \tbGE{you (second-person singular)} \tbXV{Fo na gaa la bee fo koŋ gaa?} \tbXE{Will you go or not?.} \tbSN \tbGE{your (second-person singular possessive)} \tbXV{Fo yuori la boŋ?} \tbXE{What is your name?.} 
\tbLX{foge} \tbHM{} \tbPH{fʊ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to unshell e.g. paddy rice} \tbSN \tbGE{to poke, to push quickly} \tbSN \tbGE{to foam (soap)} \tbPD{fogɛɛ, fogera, fogera, fogereba, fogeraa, fogereba } \tbVA{feɛ}
\tbLX{fogele} \tbHM{} \tbPH{fʊ́glɪ́} \tbPS{v.~} \tbGE{to be tight e.g. passengers in a crowded bus} \tbXV{A bibiiri fogele la taa a lɔɔre poɔ.} \tbXE{The children are squeezed tightly in the lorry.} \tbPD{fogelɛɛ, fogelɔ, fogelɔ, fogelebɔ, fogelaa, fogelebɔ } 
\tbLX{fogeloo} \tbHM{} \tbPH{fʊ́glʊ́ʊ́} \tbPS{adj.~} \tbGE{narrow} \tbXV{A die waa la fogeloo lɛ.} \tbXE{The room is very small.} \tbOP{fogelɔ} 
\tbLX{fogi} \tbHM{} \tbPH{fógí} \tbPS{v.~} \tbGE{to pound e.g. clay for modeling} \tbPD{fogee, fogiro, fogiro, fogiribo, fogiraa, fogiribo } 
\tbLX{fogi} \tbHM{} \tbPH{fógì} \tbPS{v.~} \tbGE{to pull out from a tight hold} \tbXV{Dasaŋa na ba maŋ fogiri la nannyige nubi-eɛmɛ.} \tbXE{In the olden days, thieves had their fingernails pulled out.} \tbPD{fogee, fogiro, fogiro, fogiribo, fogiraa, fogiribo } 
\tbLX{fogili} \tbHM{} \tbPH{fòglì} \tbPS{v.~} \tbGE{to be seedless e.g. bean pod} \tbXV{Ka saaŋ ba mi a bɛŋɛ na fogili la.} \tbXE{If it does not rain the beans will not form seed.} \tbPD{fogilee, fogilo, fogilo, fogilibo, fogilaa, fogilibo } 
\tbLX{fogiluu} \tbHM{} \tbPH{fóglúú} \tbPS{adj.~} \tbGE{seedless} \tbOP{fogilo} 
\tbLX{fole} \tbHM{} \tbPH{fʊ̀lɪ́} \tbPS{adj.~} \tbGE{immature, unripe} \tbOP{folɔ} \tbTP{folɛɛ/follɛɛ} 
\tbLX{foli} \tbHM{} \tbPH{fòlì} \tbPS{adj.~} \tbGE{not hidden} \tbOP{folo} \tbTP{folee} 
\tbLX{folinfoori} \tbHM{} \tbPH{fóli̊nfǒórí} \tbPS{n.~} \tbGE{a type of broadleaf grass used as pig feed} \tbOP{folinfobo} \tbTP{folinfooree} \tbVA{folonfoori}
\tbLX{Folo} \tbHM{} \tbPH{fóló} \tbPS{n.~} \tbGE{Fulani (an African ethnic group)} \tbOP{folibo} \tbTP{folee/follee} 
\tbLX{fomeŋɛ} \tbHM{} \tbPH{fʊ́mɪ́ŋɛ́} \tbPS{pron.~} \tbGE{yourself} 
\tbLX{fomfom} \tbHM{} \tbPH{fómfóm} \tbPS{n.~} \tbGE{type of small cake-like food prepared from flour} 
\tbLX{fomfom} \tbHM{} \tbPH{fómfóm} \tbPS{adl.~} \tbGE{deepset (eyes)} 
\tbLX{fonaano} \tbHM{} \tbPH{fʊ́náánʊ́} \tbPS{n.~} \tbGE{sleeveless singlet} \tbOP{fonaanore} \tbTP{fonaanorɛɛ/fonaanorrɛɛ} \tbVA{fenaano}
\tbLX{fonne} \tbHM{} \tbPH{fʊ̀nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to sniff (dog on a trail)} \tbSN \tbGE{to snuff} \tbPD{fonnɛɛ, fonnɔ, fonnɔ, fonnemɔ, fonnaa, fonnemɔ } 
\tbLX{fonɔfonɔ} \tbHM{} \tbPH{fʊ́nɔ́fʊ́nɔ́} \tbPS{adj.~} \tbGE{soft, silky, and  light (cloth)} \tbVA{fonafona}
\tbLX{foŋ} \tbHM{} \tbPH{fʊ́ŋ} \tbPS{pron.~} \tbGE{then you (used in conditional forms)} \tbXV{Ka saaŋ wa mi  foŋ bore a kamaana.} \tbXE{If it rains, you should sow the maize.} 
\tbLX{foo} \tbHM{} \tbPH{fòò} \tbPS{adv.~} \tbGE{looking open and endless e.g. a deep hole} 
\tbLX{fo᷉o} \tbHM{} \tbPH{fʊ̀᷉ʊ̀᷉} \tbPS{adj.~} \tbGE{dry and light e.g. grass} 
\tbLX{fo᷉o} \tbHM{} \tbPH{fʊ̌᷉ʊ᷉́} \tbPS{pron.~} \tbGE{you} \tbXV{Fo᷉o naane ka ba boɔrɔ.} \tbXE{They are not looking for you.} \tbVA{fooŋ}
\tbLX{foone} \tbHM{} \tbPH{fʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbSN \tbGE{to shrink e.g. a fruit} \tbSN \tbGE{to deflate e.g. a tyre} \tbPD{foonɛɛ, foona, foona, foonema, foonaa, foonema } 
\tbLX{foone} \tbHM{} \tbPH{fʊ́ʊ́nɪ́} \tbPS{adj.~} \tbGE{shrinked} \tbOP{fooma} \tbTP{foonɛɛ/foomɛɛ} 
\tbLX{fooŋ} \tbHM{} \tbPH{fʊ́ʊ́ŋ} \tbPS{pron.~} \tbGE{you (emphatic)} \tbXV{Fo yɔɔpɔge balɛɛ la fooŋ na gaa te ɛre a zɔŋe?} \tbXE{Your sister is tired, you will have to go and grind the flour.} 
\tbLX{foore} \tbHM{} \tbPH{fʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to hiss e.g. a snake} \tbPD{foorɛɛ, foora, foora, fooreba, fooraa, fooreba } 
\tbLX{foore} \tbHM{} \tbPH{fʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to become bloated} \tbPD{foorɛɛ, foora, foora, fooreba, fooraa, fooreba } 
\tbLX{foore} \tbHM{} \tbPH{fʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to rage} \tbPD{foorɛɛ, foora, foora, fooreba, fooraa, fooreba } 
\tbLX{foori} \tbHM{} \tbPH{fòòrí} \tbPS{adj.~} \tbSN \tbGE{empty} \tbSN \tbGE{empty-handed (typically used with nu)} \tbXV{O bɛlle ma la ne nufoori.} \tbXE{S/he deceived me with an empty hand.} 
\tbLX{foɔ} \tbHM{} \tbPH{fʊ́ɔ́} \tbPS{v.~} \tbGE{to fill holes in a wall with mortar} \tbPD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\tbLX{foɔ} \tbHM{} \tbPH{fʊ́ɔ́} \tbPS{v.~} \tbGE{to pile up a heap of guinea corn heads} \tbPD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\tbLX{foɔ} \tbHM{} \tbPH{fʊ́ɔ́} \tbPS{v.~} \tbGE{to place branches or thorns around an object for protection against animals} \tbPD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\tbLX{foɔ} \tbHM{} \tbPH{fʊ́ɔ́} \tbPS{v.~} \tbGE{to whip} \tbPD{foɔɛ, foɔrɔ, foɔrɔ, foɔreba/foɔrebɔ, foɔraa, - } 
\tbLX{fo᷉ɔ} \tbHM{} \tbPH{fʊ́᷉ɔ́᷉} \tbPS{v.~} \tbGE{to make lighter} \tbXV{E ka a tuobu fo᷉ɔ fe᷉e ka lɛ naane N koŋ to᷉ɔ a tuo.} \tbXE{Make the load a little light otherwise I cannot carry it.} \tbPD{fo᷉ɔɛ, foɔnɔ, foɔnɔ, foɔnemɔ, foɔnaa, foɔnemɔ } 
\tbLX{foɔlaa} \tbHM{} \tbPH{fʊ́ɔ́láá} \tbPS{n.~} \tbGE{passageway} \tbOP{foɔle} \tbTP{foɔlɛɛ} 
\tbLX{foɔle} \tbHM{} \tbPH{fʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to draw one's appendages against the body; to stand at attention} \tbPD{foɔlɛɛ, foɔlɔ, foɔlɔ, foɔlbɔ, foɔlaa, - } 
\tbLX{foɔle} \tbHM{} \tbPH{fʊ́ɔ́lɪ̀} \tbPS{v.~} \tbGE{to sneak, to move stealthily} \tbPD{foɔlɛɛ, foɔlɔ, foɔlɔ, foɔlbɔ, foɔlaa, - } 
\tbLX{fo᷉ɔo} \tbHM{} \tbPH{fʊ́᷉ɔ́᷉ʊ̀᷉} \tbPS{adj.~} \tbGE{light (weight)} \tbOP{foɔne} \tbTP{foɔnɛɛ} 
\tbLX{foɔre} \tbHM{} \tbPH{fʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to smear with thick liquid} \tbPD{foɔrɛɛ, foɔrɔ, foɔrɔ, foɔrba, foɔraa, foɔrba } 
\tbLX{fori} \tbHM{} \tbPH{fórì} \tbPS{v.~} \tbGE{to loosen, to slip off, to put off a dress around the waist} \tbPD{foree, foro, foro, forbo, foraa, forbo } 
\tbLX{fori} \tbHM{} \tbPH{fórí} \tbPS{use word.~} \tbGE{since} \tbXV{Fo naŋ fori yi, gaa kyɛ wa.} \tbXE{Since you have already prepared, go and come.} 
\tbLX{fori} \tbHM{} \tbPH{fòrì} \tbPS{v.~} \tbSN \tbGE{to force out from its cover or shell. (boiled bambara beans)} \tbSN \tbGE{to treat hide by removing the fur. (bleeching of the skin)} \tbSN \tbGE{to develop to an appreciable level; to mature} \tbXV{A pɔgelee ŋa naŋ ba fori kyɛ o pɔloo zuoe la.} \tbXE{This girl is not yet worth a woman but she is quite proud.} \tbPD{foree, foro, foro, forbo, foraa, forbo } 
\tbLX{forɔforɔ} \tbHM{} \tbPH{fʊ́rɔ̀fʊ́rɔ̀} \tbPS{adj.~} \tbGE{rough} \tbVA{forafora}
\tbLX{forre} \tbHM{} \tbPH{fʊ́rrɪ́} \tbPS{v.~} \tbSN \tbGE{to scrape leaves off a branch} \tbSN \tbGE{to run very fast (through vegetation)} \tbPD{forrɛɛ, forra, forra, foreba, forraa, foreba } 
\tbLX{foto} \tbHM{} \tbPH{fótó} \tbPS{n.~} \tbGE{photograph} \tbOP{fotori} \tbTP{fotoree/fotorree} 
\tbLX{fɔbo} \tbHM{} \tbPH{fɔ̀b} \tbPS{adv.~} \tbGE{very wet} 
\tbLX{fɔgele} \tbHM{} \tbPH{fɔ̀glɪ̀} \tbPS{v.~} \tbSN \tbGE{to press together} \tbSN \tbGE{to strangle} \tbPD{fɔgelɛɛ, fɔgelɔ, fɔgelɔ, fɔgelebɔ, fɔgelaa, fɔgeleba } 
\tbLX{fɔlle} \tbHM{} \tbPH{fɔ̀llɪ̀} \tbPS{v.~} \tbGE{to smear with} \tbPD{fɔllɛɛ, fɔllɔ, fɔllɔ, fɔlebɔ, fɔllaa, fɔlebɔ } 
\tbLX{fɔnne} \tbHM{} \tbPH{fɔ̀nnɪ̀} \tbPS{v.~} \tbGE{to grab, to hold} \tbPD{fɔnnɛɛ, fɔnnɔ, fɔnnɔ, fɔnemɔ, fɔnnaa, fɔnemɔ } 
\tbLX{fɔŋe} \tbHM{} \tbPH{fɔ́ŋ} \tbPS{v.~} \tbGE{to become lean, to shrink} \tbPD{fɔŋɛɛ, fɔŋenɔ/fɔnnɔ, fɔŋenɔ/fɔnnɔ, fɔŋemɔ/fɔnemɔ, fɔŋenaa/fɔnnaa, fɔŋenemɔ/fɔnemɔ } 
\tbLX{fɔŋɔ} \tbHM{} \tbPH{fɔ́ŋɔ̀} \tbPS{n.~} \tbGE{suburb, section} \tbOP{fɔnne} \tbTP{fɔnnɛɛ} \tbVA{fɔŋa}
\tbLX{fɔɔre} \tbHM{} \tbPH{fɔ̀ɔ̀rɪ̀} \tbPS{n.~} \tbGE{greed} \tbTP{fɔɔrɛɛ} 
\tbLX{fɔre} \tbHM{} \tbPH{fɔ̀rɪ̀} \tbPS{v.~} \tbGE{to hold, to squeeze tightly (by the throat)} \tbPD{fɔrɛɛ, fɔrɔ, fɔrɔ, fɔrbɔ, fɔraa, fɔrbɔ } 
\tbLX{fɔrɔ} \tbHM{} \tbPH{fɔ̀rɔ̀} \tbPS{n.~} \tbGE{yaws} \tbTP{fɔrrɛɛ} 
\tbLX{fu} \tbHM{} \tbPH{fú} \tbPS{n.~} \tbGE{rumour, hint} \tbXV{N woŋee fu kyɛ N ba baŋ ka serɛ la be.} \tbXE{I have heard a rumour but I am not suer whether it is true.} \tbTP{furee} 
\tbLX{fu} \tbHM{} \tbPH{fú} \tbPS{v.~} \tbGE{to blow (wind)} \tbPD{fue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\tbLX{fu} \tbHM{} \tbPH{fù} \tbPS{v.~} \tbSN \tbGE{to parboil, to steam (meat)} \tbSN \tbGE{to slough} \tbSN \tbGE{to make lather} \tbPD{fue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\tbLX{fugi} \tbHM{} \tbPH{fùgì} \tbPS{v.~} \tbGE{to threaten} \tbPD{fugee, fugiro, fugiro, fugiribo, fugiraa, fugiribo } 
\tbLX{fugi} \tbHM{} \tbPH{fùgì} \tbPS{v.~} \tbGE{to bluff} \tbPD{fugee, fugiro, fugiro, fugiribo, fugiraa, fugiribo } 
\tbLX{fugoŋ} \tbHM{} \tbPH{fúgóŋ} \tbPS{adv.~} \tbGE{falling suddenly into e.g. a hole} \tbXV{O faa le a bogiŋ lɛ fogoŋ.} \tbXE{S/he fell into the hole at once.} 
\tbLX{fulee} \tbHM{} \tbPH{fúléé} \tbPS{adv.~} \tbGE{eat gulpingly e.g of a hungry person} 
\tbLX{fuli} \tbHM{} \tbPH{fùlì} \tbPS{adv.~} \tbGE{exit suddenly} 
\tbLX{fulo} \tbHM{} \tbPH{fúló} \tbPS{adv.~} \tbGE{entering neatly, without friction} \tbXV{Naazoɔ deɛ e la n nyɔbogiriŋ lɛ fulo.} \tbXE{A fly shot into my nose at once.} 
\tbLX{fulofulo} \tbHM{} \tbPH{fúlófúló} \tbPS{adj.~} \tbGE{roundish e.g of aperture like a hole} 
\tbLX{fulonfugiri} \tbHM{} \tbPH{fúlónfúgrí} \tbPS{n.~} \tbGE{lung} \tbSG{fulonfuuruu} \tbOP{fulonfugo} \tbTP{fulonfugiree} 
\tbLX{fululu} \tbHM{} \tbPH{fúlúlú} \tbPS{adv.~} \tbGE{appearing round and deep e.g. a hole} 
\tbLX{fumini} \tbHM{} \tbPH{fúmìnì} \tbPS{n.~} \tbGE{needle} \tbOP{fumime} \tbTP{fuminee/fuminnee} 
\tbLX{fuoli} \tbHM{} \tbPH{fúólì} \tbPS{v.~} \tbSN \tbGE{to whistle} \tbSN \tbGE{to squeak (of a wheel)} \tbPD{fuolee, fuolo, fuolo, fuolibo, fuolaa, fuolibo } 
\tbLX{fuomii} \tbHM{} \tbPH{fùòmíì} \tbPS{n.~} \tbSN \tbGE{gluttony, voracious eating, wanting cheap things} \tbXV{A dɔɔ ŋa eŋ daŋ ba iri bonzaa, fuomii yoŋ ka o nɔŋ.} \tbXE{This man has never contributed anything, he only likes taking things for free.} \tbSN \tbGE{freeloading; wanting cheap things} \tbVA{fuomiŋ}
\tbLX{fuomiŋ} \tbHM{} \tbPH{fuomiŋ} \tbPS{n.~} \tbGE{} \tbVA{fuomii}
\tbLX{fuori} \tbHM{} \tbPH{fùòrì} \tbPS{v.~} \tbGE{to bewitch, to take the soul from the body} \tbPD{fuoree, fuoro, fuoro, fuoribo, fuoraa, fuoribo } 
\tbLX{fuori} \tbHM{} \tbPH{fúórì} \tbPS{v.~} \tbGE{to sip} \tbPD{fuoree, fuoro, fuoro, fuoribo, fuoraa, fuoribo } 
\tbLX{furi} \tbHM{} \tbPH{fùrì} \tbPS{v.~} \tbGE{to block an opening with an object, to seal} \tbPD{furee, furo, furo, furibo, furaa, furibo } 
\tbLX{furi} \tbHM{} \tbPH{fùrì} \tbPS{v.~} \tbGE{to open one's eyes widely e.g. when surprised} \tbPD{furee, furo, furo, furibo, furaa, furibo } 
\tbLX{furi} \tbHM{} \tbPH{fùrì} \tbPS{v.~} \tbGE{to stoop} \tbPD{furee, furo, furo, furibo, furaa, furibo } 
\tbLX{furi} \tbHM{} \tbPH{fúrí} \tbPS{v.~} \tbGE{to cave in} \tbPD{furee, furo, furo, furibo, furaa, furibo } 
\tbLX{furi} \tbHM{} \tbPH{fúrí} \tbPS{n.~} \tbGE{Islamic marriage rite} \tbOP{furri} \tbTP{furee/furree} 
\tbLX{furi kaa} \tbHM{} \tbPH{fùrì kàà} \tbPS{v.~} \tbGE{to stare, to glare} \tbPD{furi kaaɛ, furi kaaraa, furi kaaraa, furi kaareba, furi kaara, furi kaareba } 
\tbLX{fuu} \tbHM{} \tbPH{fúù} \tbPS{n.~} \tbGE{the cast-off skin of a snake} \tbOP{fuuri} \tbTP{fuuree} 
\tbLX{fuu} \tbHM{} \tbPH{fùù} \tbPS{v.~} \tbGE{to frighten, to threaten} \tbPD{fuue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\tbLX{fuu} \tbHM{} \tbPH{fúù} \tbPS{v.~} \tbGE{to speed} \tbPD{fuue, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\tbLX{fuu} \tbHM{} \tbPH{fúú} \tbPS{n.~} \tbGE{a game where bare arrows are used} \tbOP{fuuri} \tbTP{fuuree} 
\tbLX{fuuli} \tbHM{} \tbPH{fúúlí} \tbPS{v.~} \tbGE{to cheat; to make a fool of} \tbPD{fuulee, fuulo, fuulo, fuulibo, fuulaa, fuulibo } \tbVA{finni}
\tbLX{fuuli} \tbHM{} \tbPH{fùùlì} \tbPS{v.~} \tbGE{to constrict, to make rounded (opening)} \tbPD{fuulee, fuulo, fuulo, fuulibo, fuulibo, fuulfuulo } 
\tbLX{fuuri} \tbHM{} \tbPH{fúúrí} \tbPS{n.~} \tbSN \tbGE{foam, bubbles, lather, froth} \tbSN \tbGE{rest , cool off} \tbOP{fue} \tbTP{fuuree} 
\tbLX{fuuri} \tbHM{} \tbPH{fùùrì} \tbPS{v.~} \tbGE{to cool off} \tbPD{fuuree, fuuro, fuuro, fuuribo, fuuraa, fuuribo } 
\tbLX{fuuri} \tbHM{} \tbPH{fùúrí} \tbPS{n.~} \tbGE{treachery} 
\tbLX{fuurofuuro} \tbHM{} \tbPH{fúúrófúúró} \tbPS{n.~} \tbGE{bits and pieces (of information)} 
\tbLX{ga} \tbHM{} \tbPH{gà} \tbPS{v.~} \tbGE{to eat TZ without sorce/soup} \tbPD{gaɛ, gaara, gaara, garba, gaaraa, gaareba } 
\tbLX{ga} \tbHM{} \tbPH{gá} \tbPS{v.~} \tbGE{to run wildly or in a frenzy e.g. cow stung by tsetse fly} \tbPD{gagaɛ, gaara, gaara, gareba, gaaraa, gareba } 
\tbLX{gaa} \tbHM{} \tbPH{gáà} \tbPS{n.~} \tbGE{ebony (tree or fruit)} \tbOP{gaare} \tbTP{garɛɛ/gaarɛɛ} 
\tbLX{gaa} \tbHM{} \tbPH{gàà} \tbPS{v.~} \tbGE{to leave} \tbPD{gaaɛ, gɛrɛ, gɛrɛ, gɛrebɛ, gɛraa, gɛrebɛ } 
\tbLX{gaa} \tbHM{} \tbPH{gáà} \tbPS{v.~} \tbGE{to scoop pulp e.g. soil} \tbPD{gaaɛ, gaara, gaara, gaareba, gaaraa, gaareba } 
\tbLX{ga᷉a} \tbHM{} \tbPH{gà᷉à᷉} \tbPS{adv.~} \tbGE{outstretched and long} \tbXV{A sɔrsɔrɔ te᷉ɛ la o nu ga᷉a ka nembaaloŋ erɛ a noba ka ba korɔ o libie.} \tbXE{The beggar stretched out his hand ernestly and the onlookers were moved and began offering him money.} 
\tbLX{gaabu} \tbHM{} \tbPH{gáábú} \tbPS{n.~} \tbGE{onion} \tbOP{gaaburi} \tbTP{gaaburee/gaaburree} \tbVA{zammo}
\tbLX{gaafara} \tbHM{} \tbPH{gááfárà} \tbPS{n.~} \tbGE{excuse, sorry} \tbSG{gaafaroo} \tbOP{gaafarre} \tbTP{gaafarɛɛ} 
\tbLX{gaale} \tbHM{} \tbPH{gáálɪ̀} \tbPS{v.~} \tbSN \tbGE{to place flat on the ground, to put to bed} \tbSN \tbGE{to attend prenatal or postnatal care} \tbPD{gaalɛɛ, gaala, gaala, gaalba, gaalaa, gaaleba } 
\tbLX{Gaana} \tbHM{} \tbPH{gáånà} \tbPS{n.~} \tbGE{Ghana} \tbOP{gaanarɪ} \tbTP{gaanɛɛ} 
\tbLX{gaane} \tbHM{} \tbPH{gàànɪ̀} \tbPS{v.~} \tbSN \tbGE{to stride; to take strides of equal length with energy and confidence} \tbSN \tbGE{to wear by passing a strip between the buttocks (triangular underwear)} \tbXV{Neŋkponni mine taa la parre a maŋ de gaane yelmaale daare.} \tbXE{Some elderly men have triangular underwear which they wear on festive occasions.} \tbPD{gaanɛɛ, gaana, gaana, gaanema, gaanema, gaangaana } 
\tbLX{gaanoo} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{stride} \tbVA{gaane}
\tbLX{gabaa} \tbHM{} \tbPH{gàbàà} \tbPS{adv.~} \tbGE{open and extensive} 
\tbLX{gabadayaa (hsa)} \tbHM{} \tbPH{gàbádáyáá} \tbPS{adj.~} \tbGE{straightaway} \tbVA{gaba}
\tbLX{gabadii} \tbHM{} \tbPH{gàbàdìì} \tbPS{n.~} \tbGE{a type of woolen cloth used to make suits} 
\tbLX{gabaŋgabaŋ} \tbHM{} \tbPH{gàbàŋgàbàŋ} \tbPS{adv.~} \tbGE{anyhow} 
\tbLX{gadaa} \tbHM{} \tbPH{gàdàà} \tbPS{n.~} \tbSN \tbGE{ebony stick} \tbSN \tbGE{sign of extradition} \tbXV{A tendaana ba la nɔgraa gadaa.} \tbXE{The landlord has ordered the sojourner to leave.} 
\tbLX{gado} \tbHM{} \tbPH{gádó} \tbPS{n.~} \tbGE{bed} \tbOP{gadori} \tbTP{gadoree/gadorree} \tbVA{godo}
\tbLX{gagaraa} \tbHM{} \tbPH{gágáráà} \tbPS{n.~} \tbSN \tbGE{traitor} \tbSN \tbGE{One who denies the truth and treads on falsehood, inconsistent} \tbOP{gagarre} \tbTP{gagarɛɛ} \tbVA{minaafigi, minaafigri, minaafigree}
\tbLX{gakogili} \tbHM{} \tbPH{gákóglíí} \tbPS{n.~} \tbGE{dried berry of the ebony tree} \tbSG{gakogiluu} \tbOP{gakogilo} \tbTP{gakogilee} 
\tbLX{galaa} \tbHM{} \tbPH{gàláà} \tbPS{n.~} \tbGE{fool} \tbOP{galle} \tbTP{gallɛɛ} 
\tbLX{galbie} \tbHM{} \tbPH{gálbíé} \tbPS{n.~} \tbGE{a great person} \tbOP{galbiiri} \tbTP{galbiiree} 
\tbLX{galbie} \tbHM{} \tbPH{gálbíé} \tbPS{n.~} \tbGE{a child born after more than 9 months of pregnancy} \tbOP{galbiiri} \tbTP{galbiiree} 
\tbLX{gale} \tbHM{} \tbPH{gálɪ́} \tbPS{v.~} \tbGE{to speak in anger; to reproach} \tbPD{galɛɛ, gala, gala, galeba, galaa, galeba } 
\tbLX{gale} \tbHM{} \tbPH{gàlɪ̀} \tbPS{v.~} \tbGE{to miss the regular time} \tbXV{O daŋe ba gale toma waabo bebiyeni gba.} \tbXE{S/he never missed going to work even for a day.} \tbPD{galɛɛ, gala, gala, galba, galaa, galba } \tbVA{galle}
\tbLX{galegale} \tbHM{} \tbPH{gálgál} \tbPS{adv.~} \tbGE{of quick fast steps (run)} 
\tbLX{galegale} \tbHM{} \tbPH{gálɪ́gálɪ́} \tbPS{n.~} \tbGE{dog} 
\tbLX{galema} \tbHM{} \tbPH{gálɪmá́} \tbPS{adj.~} \tbGE{strange} \tbOP{galemalema} \tbVA{galee}
\tbLX{galeŋgaa} \tbHM{} \tbPH{gàlɪ́ŋgáà} \tbPS{n.~} \tbGE{crow} \tbOP{galeŋgaare} \tbTP{galeŋgaarɛɛ} 
\tbLX{galeŋgoɔnaa} \tbHM{} \tbPH{gàlɪ̀ŋgʊ́ɔ́náá} \tbPS{n.~} \tbGE{heron} \tbOP{galeŋgoɔne} \tbTP{galeŋgoɔnɛɛ} \tbVA{gɔleŋgoɔnaa}
\tbLX{galiŋkoori} \tbHM{} \tbPH{gàlíŋkóórì} \tbPS{n.~} \tbGE{millipede} \tbSG{galiŋkooruu} \tbOP{galiŋkoro} \tbTP{galiŋkoree} 
\tbLX{galle} \tbHM{} \tbPH{gàllɪ̀} \tbPS{v.~} \tbGE{to pass, to miss} \tbXV{Ka pɔgeŋ wa galle o pɛgebo kaapɔge poɔ la ka o taa.} \tbXE{If a woman misses her period, then she may be pregnant.} \tbPD{gallɛɛ, galla, galla, galeba, gallaa, - } 
\tbLX{galle} \tbHM{} \tbPH{gàllɪ̀} \tbPS{v.~} \tbGE{manoevre with skill and expertise} \tbXV{Te karema maŋ galle la a kɔkɔre ka noɔ kpɛ fo.} \tbXE{Our teacher manipulates the language such that you enjoy his talk.} \tbPD{gallɛɛ, galla, galla, galeba, gallaa, - } 
\tbLX{Galyuoni} \tbHM{} \tbPH{gàlyúóní} \tbPS{n.~} \tbGE{name traditionally given to a child born after 9 months of pregnancy} \tbVA{Yuoni}
\tbLX{gambanna} \tbHM{} \tbPH{gámbánná} \tbPS{n.~} \tbGE{one who is literate} \tbOP{gambanema} \tbTP{gambannɛɛ} 
\tbLX{gammiri} \tbHM{} \tbPH{gámmírì} \tbPS{n.~} \tbGE{woven twine string worn by women around the waist} \tbSG{gammiruu} \tbOP{gammie} \tbTP{gammiiree} 
\tbLX{gammisɔgelaa} \tbHM{} \tbPH{gámmísɔ́gláà} \tbPS{n.~} \tbGE{woven leather waist string} \tbOP{gammisɔgelɔ} 
\tbLX{gammizoolɔ} \tbHM{} \tbPH{gámmízóólɔ̀} \tbPS{n.~} \tbGE{waist strings for adolescent girl} \tbOP{gammizoolɛɛ} \tbVA{gammizoola}
\tbLX{gampɛle} \tbHM{} \tbPH{gámpɛ́lɪ̀} \tbPS{n.~} \tbSN \tbGE{sheet of paper} \tbSN \tbGE{page of a book} \tbOP{ gampɛlɛ} 
\tbLX{gamporrizie} \tbHM{} \tbPH{gámpórrìzíé} \tbPS{n.~} \tbGE{post office} \tbOP{gamporriziiri} \tbTP{gamporriziiree} 
\tbLX{ganaa} \tbHM{} \tbPH{gànàá} \tbPS{n.~} \tbSN \tbGE{strip of leather} \tbSN \tbGE{twine for chief mourners of a deceased person} \tbOP{ganne} \tbTP{gannɛɛ} 
\tbLX{ganaaganaa} \tbHM{} \tbPH{gánáágánáá} \tbPS{adv.~} \tbGE{small and stretched like a string, thin} 
\tbLX{ganaaganaa} \tbHM{} \tbPH{gànààgànàà} \tbPS{adv.~} \tbGE{large and stretched} \tbXV{A dɔbɔ ti la taa ganaaganaa, ka ba faara faana.} \tbXE{The men hooked each other tightly and fiercely such that people were finding it difficult to separate them.} 
\tbLX{gandaa} \tbHM{} \tbPH{gàndáá} \tbPS{n.~} \tbSN \tbGE{hero, great person, protagonist} \tbSN \tbGE{hero in literature} \tbOP{gandaare} \tbTP{gandaarɛɛ} 
\tbLX{gandaa} \tbHM{} \tbPH{gàndáá} \tbPS{n.~} \tbGE{placenta} \tbOP{gandaare} \tbTP{gandaarɛɛ} 
\tbLX{gane} \tbHM{} \tbPH{gánɪ̀} \tbPS{n.~} \tbSN \tbGE{book} \tbSN \tbGE{hide, skin} \tbOP{gama} \tbTP{gannɛɛ/gamɛɛ} 
\tbLX{gangyinni} \tbHM{} \tbPH{gàndzínní} \tbPS{n.~} \tbGE{the act of doing things in excess e.g. eating} \tbTP{gangyinnee} \tbVA{gangyinnuŋ}
\tbLX{gangyinnuŋ} \tbHM{} \tbPH{gàndzínnúŋ} \tbPS{n.~} \tbGE{the act of doing things in excess} \tbVA{gangyinnuu, gangyinni}
\tbLX{ganlaa} \tbHM{} \tbPH{gánláá} \tbPS{n.~} \tbGE{a type of dance} \tbOP{ganlaare} \tbTP{ganlaarɛɛ} 
\tbLX{ganlɔgɔ} \tbHM{} \tbPH{gànlɔ̀gɔ̀} \tbPS{n.~} \tbGE{small leather bag worn on the shoulder} \tbOP{ganlɔgre} \tbTP{ganlɔgrɛɛ} 
\tbLX{gansɛgerɛ} \tbHM{} \tbPH{gánsɛ́grɛ̀} \tbPS{n.~} \tbGE{writer, letter writer} \tbOP{gansegerebɛ} \tbTP{ganɛgerebɛɛ} \tbVA{gaŋŋmaara}
\tbLX{gansɔraa} \tbHM{} \tbPH{gánsɔ́ráá} \tbPS{n.~} \tbGE{a book for reading} \tbOP{gansɔrre} \tbTP{gansɔrrɛɛ} 
\tbLX{gantoɔle} \tbHM{} \tbPH{gántʊ́ɔ́lɪ̀} \tbPS{n.~} \tbGE{mail} \tbSG{gantoɔlaa} \tbTP{gantoɔlɛɛ} 
\tbLX{gaŋa} \tbHM{} \tbPH{gàŋá} \tbPS{n.~} \tbGE{An enclosure for livestock e.g. cattle} \tbOP{ganne} \tbTP{gannɛɛ} \tbVA{gaŋe}
\tbLX{gaŋe} \tbHM{} \tbPH{gáŋ} \tbPS{adj.~} \tbGE{mad or insane (dogs)} \tbXV{Bagaŋe zaameŋ doŋe la n baa.} \tbXE{A mad dog bit my dog yesterday.} 
\tbLX{gaŋe} \tbHM{} \tbPH{gâŋ} \tbPS{v.~} \tbSN \tbGE{to lie down} \tbSN \tbGE{to sleep} \tbPD{gaŋɛɛ, gana, gana, ganema, ganaa, ganema } 
\tbLX{gaŋe} \tbHM{} \tbPH{gàŋɪ̀} \tbPS{v.~} \tbSN \tbGE{to step over} \tbXV{A baa age gaŋ la a bogi.} \tbXE{The dog jumped over the hole.} \tbSN \tbGE{to overtake,} \tbXV{Ba lɔɔre zo wa gaŋ la te deni na bare.} \tbXE{Their lorry overtook ours.} \tbSN \tbGE{to be more than, to exceed} \tbXV{A kparoo waa la kpoŋ gaŋ a bie.} \tbXE{The shirt is bigger than the child.} \tbPD{gaŋɛɛ, gaŋena/gana, gana, gaŋenema/ganema, gaŋenaa/ganaa, gaŋenema/ganema } 
\tbLX{gaŋene} \tbHM{1} \tbPH{gàŋɪ̀nɪ́} \tbPS{n.~} \tbGE{next one} \tbXV{N daga maŋ biŋ la a gaŋene die na poɔ.} \tbXE{My box is kept in the room after the first one.} 
\tbLX{gaŋene} \tbHM{2} \tbPH{gàŋɪ̀nɪ́} \tbPS{n.~} \tbGE{that which is too much} 
\tbLX{gaŋgaa} \tbHM{} \tbPH{gàŋgàá} \tbPS{n.~} \tbGE{drum (general)} \tbOP{gaŋgarɪ} \tbTP{gaŋgarɛɛ} 
\tbLX{gaŋgaanaa} \tbHM{} \tbPH{gǎŋgàànáà} \tbPS{n.~} \tbSN \tbGE{reed} \tbSN \tbGE{vine} \tbOP{gaŋgaane} \tbTP{gaŋgaanɛɛ} 
\tbLX{gaŋgadaa} \tbHM{} \tbPH{gàŋgàdáá} \tbPS{n.~} \tbGE{small drum with a quieter sound} \tbOP{gaŋgadaare} \tbTP{gaŋgadaarɛɛ} 
\tbLX{gaŋgadalee} \tbHM{} \tbPH{gàŋgàdàléé} \tbPS{n.~} \tbGE{drumstick} \tbOP{gaŋgadabilii} \tbVA{gaŋgadabili}
\tbLX{gaŋgakpoŋ} \tbHM{} \tbPH{gàŋgàkpóŋ} \tbPS{n.~} \tbGE{bass drum} \tbOP{gaŋgakponni} \tbTP{gaŋgakponnee} 
\tbLX{gaŋgalee} \tbHM{} \tbPH{gàŋgàléé} \tbPS{n.~} \tbGE{small drum; alto drum} \tbOP{gaŋgabilii} \tbTP{gaŋgabilee} 
\tbLX{gaŋganyaŋaa} \tbHM{} \tbPH{gàŋgàɲáŋáá} \tbPS{n.~} \tbGE{drum with a deep, loud sound} \tbOP{gaŋganyaŋene} \tbTP{gaŋganyaŋenɛɛ} 
\tbLX{gaŋkanna} \tbHM{} \tbPH{gáŋkánná} \tbPS{n.~} \tbGE{reader, letter or book reader} \tbOP{gaŋkanema} \tbTP{gaŋkanemɛɛ} 
\tbLX{gaŋkoŋi} \tbHM{} \tbPH{gàŋkǒŋ} \tbPS{n.~} \tbGE{funeral string/strip given without any accompanying funeral donation} \tbOP{gaŋkonni} \tbTP{gaŋkonnee} 
\tbLX{gaŋkoŋi} \tbHM{} \tbPH{gáŋkóŋ} \tbPS{n.~} \tbGE{dry and crumpled animal hide} \tbOP{gaŋkonni} \tbTP{gaŋkonnee} 
\tbLX{gaŋkuoŋaa} \tbHM{} \tbPH{gàŋkúóŋáá} \tbPS{n.~} \tbGE{dried strip of leather} \tbOP{gaŋkuoni} \tbTP{gaŋkuonee} 
\tbLX{gaŋŋmaara} \tbHM{} \tbPH{gáŋŋmáárá} \tbPS{n.~} \tbGE{writer} \tbOP{gaŋŋmaareba} \tbTP{gaŋŋmaarebɛɛ} \tbVA{gansɛgrɛ}
\tbLX{gara} \tbHM{} \tbPH{gàrá} \tbPS{n.~} \tbGE{tapeworm} \tbOP{garre} \tbTP{garɛɛ/garrɛɛ} 
\tbLX{gara} \tbHM{} \tbPH{gárá} \tbPS{n.~} \tbGE{hobble (rope for holding two legs of an animal)} \tbOP{garre} \tbTP{garɛɛ/garrɛɛ} \tbVA{gbara}
\tbLX{garaa} \tbHM{} \tbPH{gáráá} \tbPS{adj.~} \tbGE{small and pointed} 
\tbLX{garaa} \tbHM{} \tbPH{gàráà} \tbPS{n.~} \tbGE{ridge} \tbOP{garre} \tbTP{garrɛɛ} 
\tbLX{gare} \tbHM{} \tbPH{gárɪ́} \tbPS{v.~} \tbSN \tbGE{to stumble, to trip over} \tbSN \tbGE{to tie or hold legs together} \tbPD{garɛɛ, gara, gara, gareba, garaa, gareba } 
\tbLX{gare} \tbHM{} \tbPH{gàrɪ̀} \tbPS{v.~} \tbGE{to place across, to close (road)} \tbXV{A dɔɔ de la daa gare a sori.} \tbXE{The man has placed a plank across the road.} \tbPD{garɛɛ, gara, gara, gareba, garaa, gareba } 
\tbLX{gare} \tbHM{} \tbPH{gárɪ́} \tbPS{v. adj.~} \tbGE{go bad, sour (food)} 
\tbLX{garegyerɛ} \tbHM{} \tbPH{gárɪ́yɪ̀rɛ} \tbPS{n.~} \tbGE{a shrub with milky sap, poisonous} \tbSG{garegyeroo} \tbTP{gargyerɛɛ} 
\tbLX{Garegyerɛbaa} \tbHM{} \tbPH{gàrɪ́dzìrɛ́bàá} \tbPS{n.~} \tbGE{One of the main district headquarters of the Dagaare speaking area whose dialeɔt is often referred to as the ‘central dialect’} \tbVA{Gyerɛbaa}
\tbLX{garepegele} \tbHM{} \tbPH{gàrɪ̀pɪ̀glɪ̀} \tbPS{v.~} \tbGE{to crisscross} \tbXV{A teɛ le la garepegele a sori.} \tbXE{The tree fell and crossed/bocked the road.} \tbPD{garepegelɛɛ, garapegelɛ, garapegelɛ, garepegelebɛ, garepegelaa, garepeglebɛ } 
\tbLX{garewaa} \tbHM{} \tbPH{gárɪ́wáá} \tbPS{n.~} \tbGE{four-gallon tin} \tbOP{garewaare} \tbTP{garewaarɛɛ} 
\tbLX{garoŋ} \tbHM{} \tbPH{gárʊ́ŋ} \tbPS{n.~} \tbGE{sour-tasting substance} 
\tbLX{garre} \tbHM{} \tbPH{gárrɪ́} \tbPS{v.~} \tbGE{to deny} \tbPD{garrɛɛ, garra, garra, gareba, garaa, gareba } 
\tbLX{gasalaa} \tbHM{} \tbPH{gàsàlàà} \tbPS{adj.~} \tbGE{gaping} 
\tbLX{gegili} \tbHM{} \tbPH{gèglì} \tbPS{v.~} \tbGE{be half-closed} \tbPD{gegilee, gegile, gegile, gegilibe, gegilaa, gegilibe } \tbVA{ogili}
\tbLX{geli} \tbHM{} \tbPH{} \tbPS{v.~} \tbGE{to lift over something heavy lying flat on the ground} \tbXV{Ba geli la a kuri gbori tiili a dankyini poɔ.} \tbXE{They have lifted over the iron gate unto the wall.} \tbPD{gelee, gelle, gelle, gelibe, -, gelibe } 
\tbLX{geli} \tbHM{} \tbPH{} \tbPS{v.~} \tbGE{to burn badly (food)} \tbPD{gelee, gelle, gelle, gelibe, -, gelibe } 
\tbLX{geli} \tbHM{} \tbPH{} \tbPS{v.~} \tbGE{to be mad, to behave abnormally or foolishly} \tbPD{gelee, gelle, gelle, gelibe, -, gelibe } 
\tbLX{gelime} \tbHM{} \tbPH{gélímè} \tbPS{adj/n.~} \tbGE{insanity} 
\tbLX{gemee} \tbHM{} \tbPH{géméé} \tbPS{adj.~} \tbGE{mad, insane} 
\tbLX{genegene} \tbHM{} \tbPH{gènègènè} \tbPS{adv.~} \tbSN \tbGE{tinted (colour)} \tbXV{A boɔ e la zeɛ genegene.} \tbXE{The goat is tinted brown.} \tbSN \tbGE{not appearing clearly} \tbXV{Zie naŋ da wa nyaana genegene la ka o wa.} \tbXE{S/he came when dawn had not clearly set in.} 
\tbLX{geŋgeŋ} \tbHM{} \tbPH{géŋgéŋ} \tbPS{adj.~} \tbGE{firmly fixed} 
\tbLX{geŋi} \tbHM{} \tbPH{gêŋ} \tbPS{v.~} \tbGE{to stretch out to reach e.g. to look over a wall} \tbPD{geŋee, geŋine/genne, geŋine/genne, geŋinime/genime, geŋinaa/gennaa, genime } 
\tbLX{geŋi} \tbHM{} \tbPH{gêŋ} \tbPS{v.~} \tbGE{to be mad, to be angry} \tbPD{geŋee, geŋine/genne, geŋine/genne, geŋinime/genime, geŋinaa/gennaa, genime } 
\tbLX{geregere} \tbHM{} \tbPH{gèrègèrè} \tbPS{adv.~} \tbGE{moderately hard e.g. half-cooked yam} 
\tbLX{gɛbɛɛ} \tbHM{} \tbPH{gɛ̀bɛ́ɛ́} \tbPS{n.~} \tbGE{fool, lunatic} 
\tbLX{gieri} \tbHM{} \tbPH{gíérì} \tbPS{v.~} \tbGE{to belch} \tbPD{gieree, giere, giere, gieribe, gieraa, gieribe } 
\tbLX{go} \tbHM{} \tbPH{gó} \tbPS{v.~} \tbGE{to stoop, to bend down} \tbPD{goe, guoro, guoro, guoribo, goraa, guoribo } 
\tbLX{go} \tbHM{} \tbPH{gò} \tbPS{v.~} \tbGE{watch over} \tbPD{goe, goro, goro, goribo, goraa, - } \tbVA{gu}
\tbLX{gobaa} \tbHM{} \tbPH{gòbàà} \tbPS{n.~} \tbGE{shame, disgrace} 
\tbLX{gogo} \tbHM{} \tbPH{gógó} \tbPS{n.~} \tbSN \tbGE{clock} \tbSN \tbGE{hour} \tbOP{gogori} \tbTP{gogoree} 
\tbLX{gogobiri} \tbHM{} \tbPH{gògòbírí} \tbPS{n.~} \tbGE{zebu (bull with a humped back)} \tbOP{gogobie} \tbTP{gogobiiree} 
\tbLX{gogoo} \tbHM{} \tbPH{gògóó} \tbPS{adj.~} \tbGE{standing out with long appendages e.g. pointing} 
\tbLX{gogoɔ} \tbHM{} \tbPH{gʊ̀gʊ̀ɔ́} \tbPS{n.~} \tbGE{sawfish} \tbOP{gogoɔre} \tbTP{gogoɔrɛɛ} 
\tbLX{golebɛlɛ} \tbHM{} \tbPH{gʊ̀lɪ̀bɛ̀lɛ̀} \tbPS{adj.~} \tbGE{lethargic and unsteady (very sick person)} 
\tbLX{goligoli} \tbHM{} \tbPH{gòlgólí} \tbPS{v.~} \tbGE{to scribble} \tbPD{goligolee, goligolo, goligolo, golgolibo, golgolaa, golgolibo } 
\tbLX{gollaa} \tbHM{} \tbPH{gólláà} \tbPS{n.~} \tbGE{a game in which players hop on one leg} \tbOP{golli} \tbTP{gollee} 
\tbLX{golle} \tbHM{} \tbPH{gʊ̀llɪ̀} \tbPS{v.~} \tbGE{to mix up any which way (clothes)} \tbPD{gollɛɛ, gollɔ/golla, gollɔ/golla, golba, gollaa, - } 
\tbLX{golli} \tbHM{} \tbPH{góllì} \tbPS{v.~} \tbGE{to hop on one leg} \tbPD{gollee, gollo, gollo, golibo, gollaa, golibo } 
\tbLX{goloŋgoloŋ} \tbHM{} \tbPH{gòlòŋgólóŋ} \tbPS{adv.~} \tbGE{long and twisted} 
\tbLX{golɔgolɔ} \tbHM{} \tbPH{gʊ̊lɔ́gʊ́lɔ́} \tbPS{adj.~} \tbGE{soft and wavy (hair)} 
\tbLX{golɔnsomm} \tbHM{} \tbPH{gʊ̀lɔ̀nsʊ̀mm} \tbPS{adv.~} \tbGE{roundish and big} 
\tbLX{golɔnsomm} \tbHM{} \tbPH{gʊ́lɔ́nsʊ́mm} \tbPS{adv.~} \tbGE{roundish and small} 
\tbLX{gombɔle} \tbHM{} \tbPH{gòmbɔ́lɪ́} \tbPS{n.~} \tbGE{kapok fruit} \tbOP{gombɔlɔ} \tbTP{gombɔlɛɛ} 
\tbLX{gomɔgomɔ} \tbHM{} \tbPH{gʊ̀mɔ̀gʊ̀mɔ̀} \tbPS{adv.~} \tbGE{not straight forward} \tbVA{gomagoma}
\tbLX{gondoŋ} \tbHM{} \tbPH{gòndóŋ} \tbPS{n.~} \tbGE{skirt} \tbOP{gondonni} \tbTP{gondonnee} 
\tbLX{gone} \tbHM{} \tbPH{gʊ́nɪ́} \tbPS{n.~} \tbGE{structural flaw} \tbOP{gonne} \tbTP{gonɛɛ/gonnɛɛ} 
\tbLX{gonnaabo} \tbHM{} \tbPH{gónnáábʊ́} \tbPS{n.~} \tbGE{plant louse} \tbOP{gonnii} \tbTP{gonniinee} 
\tbLX{gonni} \tbHM{} \tbPH{gònní} \tbPS{n.~} \tbGE{cotton} \tbSG{gonnuu} \tbOP{n/a} \tbTP{gonnee} 
\tbLX{gonseɛ} \tbHM{} \tbPH{gònsɪ̀ɛ́} \tbPS{n.~} \tbGE{fall; season of intermitent wind} 
\tbLX{goŋo} \tbHM{} \tbPH{gòŋó} \tbPS{n.~} \tbGE{cotton tree} \tbOP{gonni} \tbTP{gonnee} 
\tbLX{goŋmane} \tbHM{} \tbPH{góŋmánɪ́} \tbPS{n.~} \tbGE{one hundred kola nuts} \tbOP{goŋmama} \tbTP{goŋmannɛɛ} 
\tbLX{goole} \tbHM{} \tbPH{gʊ̀ʊ̀lɪ̀} \tbPS{v.~} \tbGE{to mix, to stir (thick paste)} \tbPD{goolɛɛ, goolɔ, goolɔ, gooleba, goolaa, gooleba } 
\tbLX{goone} \tbHM{} \tbPH{gʊ̀ʊ̀nì} \tbPS{v.~} \tbGE{to walk majestically with an up-and-down body movement} \tbXV{A daanedann da maŋe la a naa niŋŋ kyɛ goone iri.} \tbXE{The praise singer would walk majestically with up-and-down-movement of the body in front of the King.} \tbPD{goonɛɛ, goonɔ, goonɔ, goonema, goonaa, goonema } 
\tbLX{goɔ} \tbHM{} \tbPH{gʊ́ɔ̀} \tbPS{n.~} \tbGE{thorn} \tbOP{goore} \tbTP{goorɛɛ} 
\tbLX{goɔ} \tbHM{} \tbPH{gʊ̀ɔ̀} \tbPS{v.~} \tbGE{to bend at the trunk} \tbXV{Pɔgeba maŋ goɔ la kaabo kyɛ ŋma.} \tbXE{Women usually hold down guinea corn to cut the head.} \tbPD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ } 
\tbLX{goɔ} \tbHM{} \tbPH{gʊ̀ɔ̀} \tbPS{v.~} \tbGE{to prevent from progressing; to cause to return or refrain from something} \tbPD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ } 
\tbLX{goɔ} \tbHM{} \tbPH{gʊ̀ɔ̀} \tbPS{v.~} \tbGE{to bite (snake)} \tbPD{goɔɛ, goɔrɔ, goɔrɔ, goɔrbɔ, goɔraa, goɔrbɔ } 
\tbLX{go᷉ɔ} \tbHM{} \tbPH{gʊ́᷉ɔ́} \tbPS{n.~} \tbSN \tbGE{kiln (place for baking clay items to make them hard)} \tbSN \tbGE{arena, grouping point (ko-go᷉ɔ)} \tbOP{goɔne} \tbTP{goɔnɛɛ} 
\tbLX{go᷉ɔo} \tbHM{} \tbPH{gʊ́᷉ɔ̀ʊ́} \tbPS{n.~} \tbGE{sleep} \tbOP{goɔnɛɛ} 
\tbLX{goɔraa} \tbHM{} \tbPH{gʊ̀ɔ̀ràá} \tbPS{n.~} \tbGE{a type of small tree with medicinal properties} \tbOP{goɔre} \tbTP{goɔrɛɛ} 
\tbLX{goɔre} \tbHM{} \tbPH{gʊ́ɔ́rɪ́} \tbPS{v.~} \tbGE{to walk or run half stooping} \tbPD{goɔrɛɛ, goɔrɔ, goɔrɔ, goɔreba, goɔraa, goɔreba } 
\tbLX{goɔre} \tbHM{} \tbPH{gʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to go round and round} \tbSN \tbGE{to splash mashed ashes round the outer wall of a house (ritual to rid the house of activities of witchraft)} \tbPD{goɔrɛɛ, goɔrɔ, goɔrɔ, goɔreba, goɔraa, - } 
\tbLX{gora} \tbHM{} \tbPH{gʊ́rà} \tbPS{n.~} \tbSN \tbGE{hunchback} \tbSN \tbGE{disease of the back} \tbOP{gorre} \tbTP{gorrɛɛ} \tbVA{gorɔ}
\tbLX{goraa} \tbHM{} \tbPH{gòràà} \tbPS{adj.~} \tbGE{curved} 
\tbLX{goraa} \tbHM{} \tbPH{gòràà} \tbPS{adj.~} \tbGE{ashamed} 
\tbLX{gore} \tbHM{} \tbPH{gʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to be abundant} \tbSN \tbGE{to be overburdened with weight} \tbPD{gorɛɛ, gora, gora, goreba, goraa, - } 
\tbLX{gore} \tbHM{} \tbPH{gʊ́rɪ́} \tbPS{v.~} \tbGE{to pluck feathers or fur} \tbPD{gorɛɛ, gora, gora, gorba, goraa, - } 
\tbLX{gorgore} \tbHM{} \tbPH{gʊ̀rgʊ́rɪ́} \tbPS{v.~} \tbGE{to ruffle} \tbPD{gorgorɛɛ, gorgora, gorgora, gorgorba, gorgoraa, gorgorba } 
\tbLX{gori} \tbHM{} \tbPH{górí} \tbPS{n.~} \tbGE{dowry, bride price} \tbOP{gorri} \tbTP{goree} 
\tbLX{gori} \tbHM{} \tbPH{gòrì} \tbPS{v.~} \tbGE{to bend an object (into a sharp curve)} \tbPD{goree, goro, goro, gorbo, goraa, gorbo } 
\tbLX{gori} \tbHM{} \tbPH{górí} \tbPS{n.~} \tbGE{big pile of; load} \tbOP{goe} \tbTP{goree} 
\tbLX{gori-gori-gori} \tbHM{} \tbPH{gór-gór-gór} \tbPS{adv.~} \tbGE{motionless} 
\tbLX{gozambuuli} \tbHM{} \tbPH{gòzàmbúúlí} \tbPS{n.~} \tbGE{the fruit of a type of thorn tree} \tbOP{gozambuulo} \tbTP{gozambuulee} \tbVA{gɔzambuuli}
\tbLX{gozaŋaa} \tbHM{} \tbPH{gózàŋàà} \tbPS{n.~} \tbGE{a type of thorn tree with yellow fruit} \tbOP{gozanne} \tbTP{gozannɛɛ} \tbVA{gɔzaŋaa}
\tbLX{gɔbaa} \tbHM{} \tbPH{gɔ̀báà} \tbPS{n.~} \tbSN \tbGE{skillful person; one full of dexterity} \tbSN \tbGE{artist (music, visual arts)} \tbOP{gɔɔre} \tbTP{gɔɔrɛɛ} 
\tbLX{gɔbenɛ} \tbHM{} \tbPH{gɔ́bɪ̀nɛ̀} \tbPS{n.~} \tbGE{governor, government} \tbOP{gɔbenne} \tbTP{gɔbenɛɛ} 
\tbLX{gɔbenɛnte} \tbHM{} \tbPH{gɔbɪnɛntɪ} \tbPS{n.~} \tbGE{government} \tbOP{gɔbɪnɛntɪrɪ} \tbTP{gɔbɪnɛntɪrɛɛ} 
\tbLX{gɔfuuraa} \tbHM{} \tbPH{gɔ́fúúráá} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{gɔfuuri} \tbTP{gɔfuuree} 
\tbLX{gɔge} \tbHM{} \tbPH{gɔ́gɪ́} \tbPS{v.~} \tbGE{to nod} \tbPD{gɔgɛɛ, gɔgerɔ, gɔgerɔ, gɔgerebɔ, gɔgraa, gɔgerebɔ } 
\tbLX{gɔgɔ} \tbHM{} \tbPH{gɔ́gɔ̀} \tbPS{n.~} \tbSN \tbGE{a type of tree whose bark is used in tanning} \tbSN \tbGE{to type of black bird with a long tail} \tbOP{gɔgɔre} \tbTP{gɔgɔrɛɛ} 
\tbLX{gɔgɔ} \tbHM{} \tbPH{gɔ̀gɔ́} \tbPS{v.~} \tbGE{to bend halfway; to stoop} \tbPD{gɔgɔɛ, gɔgɔɔrɔ, gɔgɔɔrɔ, gɔgɔrebɔ, gɔgɔraa, gɔgɔrebɔ } 
\tbLX{gɔllaa} \tbHM{} \tbPH{gɔ́lláá} \tbPS{n.~} \tbGE{a type of small, fatty field mouse} \tbOP{gɔlle} \tbTP{gɔllɛɛ} \tbVA{mɔɔtere}
\tbLX{gɔlle} \tbHM{} \tbPH{gɔ̀llɪ̀} \tbPS{v.~} \tbSN \tbGE{to go round, to detour} \tbSN \tbGE{to surround, to encircle} \tbPD{gɔllɛɛ, gɔllɔ, gɔllɔ, gɔlebɔ, gɔllaa, gɔlebɔ } 
\tbLX{gɔlɔnteebo} \tbHM{} \tbPH{gɔ́lɔ́ntɪ̀ɪ̀bʊ́} \tbPS{n.~} \tbGE{chameleon} \tbOP{gɔlɔnteere} \tbTP{gɔlɔnteerɛɛ} \tbVA{gɔmɔ/gɔmɔteebo}
\tbLX{gɔmberaa} \tbHM{} \tbPH{gɔ́mbíráá} \tbPS{n.~} \tbGE{state of sleeplessness} \tbOP{gɔmberre} \tbTP{gɔmberɛɛ} 
\tbLX{gɔmbulo} \tbHM{} \tbPH{gɔ̀mbùlò} \tbPS{n.~} \tbGE{unique sound in poetry, sound that is pleasing to the ear} 
\tbLX{gɔmɔ} \tbHM{} \tbPH{gɔ́mɔ̀} \tbPS{n.~} \tbGE{noise, disturbance} \tbOP{gɔnne} \tbTP{gɔnnɛɛ} \tbVA{gɔnne}
\tbLX{gɔnda} \tbHM{} \tbPH{gɔ̌ndà} \tbPS{n.~} \tbGE{pawpaw, papaya} \tbSG{gɔndaroo} \tbOP{gɔndare} \tbTP{gɔndarɛɛ} 
\tbLX{gɔnne} \tbHM{} \tbPH{gɔ́nnɪ̀} \tbPS{n.~} \tbGE{noise} \tbOP{gɔmɔ} \tbTP{gɔnnɛɛ} 
\tbLX{gɔntɛre} \tbHM{} \tbPH{gɔ̀ntɛ́rɪ́} \tbPS{v.~} \tbGE{to talk incessantly} \tbPD{gɔntɛrɛɛ, gɔnɔtɛrɛ, gɔnɔtɛrɛ, gɔntɛrebɛ, -, - } 
\tbLX{gɔntɛre} \tbHM{} \tbPH{gɔ̀ntɛ́rɪ́} \tbPS{v.~} \tbGE{to burst out, especially when angry} \tbPD{gɔntɛrɛɛ, gɔnɔtɛrɛ, gɔnɔtɛrɛ, gɔntɛrebɛ, -, - } 
\tbLX{gɔnzuuro} \tbHM{} \tbPH{gɔ́nzúúrò} \tbPS{n.~} \tbGE{sleepiness} \tbXV{O taa la gɔnzuuro.} \tbXE{He/she is feeling sleepy.} \tbTP{gɔnzuuree} 
\tbLX{gɔŋe} \tbHM{} \tbPH{gɔ́ŋ} \tbPS{adj.~} \tbGE{crooked, bent} 
\tbLX{gɔŋe} \tbHM{} \tbPH{gɔ̀ŋ} \tbPS{v.~} \tbGE{to make crooked, to incline toward} \tbPD{gɔŋɛɛ, gɔnnɔ, gɔnnɔ, gɔnema, gɔnnaa, gɔnema } 
\tbLX{gɔŋe} \tbHM{} \tbPH{gɔ́ŋ} \tbPS{v.~} \tbGE{to make noise, to quarrel} \tbPD{gɔŋɛɛ, gɔnnɔ, gɔnɔ, gɔnemɔ, gɔnnaa, gɔnemɔ } 
\tbLX{gɔŋ-eŋ} \tbHM{} \tbPH{gɔ̀ŋ-ɪ̀ŋ} \tbPS{v.~} \tbGE{to redirect or state to suit one's purpose} \tbXV{A yɛlɛ ba waa lɛ a pɔge gɔŋ-eŋɛɛ la.} \tbXE{That is not the true state of the matter; the woman has tried to add something to make it look like that.} \tbPD{gɔŋ-eŋɛɛ, gɔnnɔ-ennɛ, gɔŋ-ennɛ, gɔŋ-ennaa, gɔŋ-enemɛ, - } 
\tbLX{gɔŋgɔlɔŋ} \tbHM{} \tbPH{gɔ̀ŋgɔ̀lɔ̀ŋ} \tbPS{adj.~} \tbGE{long and winding} 
\tbLX{gɔŋgɔŋe} \tbHM{} \tbPH{gɔ̀ŋgɔ́ŋ} \tbPS{v.~} \tbGE{to walk wearily and unsteadily; to trudge} \tbPD{gɔŋgɔŋɛɛ, gɔŋgɔnɔ, gɔŋgɔnɔ, gɔŋgɔnema, gɔŋgɔnaa, gɔŋgɔnema } 
\tbLX{gɔŋgbire} \tbHM{} \tbPH{gɔ́ŋgbírè} \tbPS{n.~} \tbGE{one who is asleep} \tbOP{gɔŋgbiribe} \tbTP{gɔŋgbiribee} 
\tbLX{gɔŋɔ} \tbHM{} \tbPH{gɔ́ŋɔ́} \tbPS{n.~} \tbGE{type of guinea corn} \tbTP{gɔnnɛɛ} 
\tbLX{gɔɔ} \tbHM{} \tbPH{gɔ̀ɔ́} \tbPS{n.~} \tbGE{left hand, left side} \tbOP{gɔɔe} \tbTP{gɔɔrɛɛ} 
\tbLX{gɔɔ} \tbHM{} \tbPH{gɔ̀ɔ́} \tbPS{adj.~} \tbGE{left-handed} \tbOP{gɔɔe} \tbTP{gɔɔrɛɛ} 
\tbLX{gɔɔ} \tbHM{} \tbPH{gɔ̀ɔ̀} \tbPS{v.~} \tbGE{to go round and round} \tbPD{gɔɔɛ, gɔɔrɔ, gɔɔrɔ, gɔɔrebɔ, gɔɔraa, gɔɔrebɔ } \tbVA{goɔre}
\tbLX{gɔɔrombiri} \tbHM{} \tbPH{gɔ́ɔ́rʊ́mbírí} \tbPS{n.~} \tbGE{literary device (lit.)} \tbSG{gɔɔrombiruu} \tbOP{gɔɔrombie} \tbTP{gɔɔrombiree} 
\tbLX{gɔpelaa} \tbHM{} \tbPH{gɔ̀pɪ̀làá} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{gɔpeɛle} \tbTP{gɔpeɛlɛɛ} 
\tbLX{gɔre} \tbHM{} \tbPH{gɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to bend} \tbSN \tbGE{to mould a hoe into a curved shape} \tbPD{gɔrɛɛ, gɔrɔ, gɔrɔ, gɔrebɔ, gɔraa, gɔrebɔ } 
\tbLX{gɔre} \tbHM{} \tbPH{gɔ́rɪ́} \tbPS{n.~} \tbGE{scythe} \tbOP{gɔrre} \tbTP{gɔrɛɛ} 
\tbLX{gɔsɔgelaa} \tbHM{} \tbPH{gɔ́sɔ́glàá} \tbPS{n.~} \tbGE{a type of thorn tree} \tbOP{gɔsɔgelɔ} \tbTP{gɔsɔgelɛɛ} 
\tbLX{gɔteɛ} \tbHM{} \tbPH{gɔ̂tɪ̀ɛ́} \tbPS{n.~} \tbGE{a tree with thorns} \tbOP{gɔteere} \tbTP{gɔteerɛɛ} 
\tbLX{gɔzeɛ} \tbHM{} \tbPH{gɔ̀zɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} 
\tbLX{gu} \tbHM{} \tbPH{gù} \tbPS{v.~} \tbSN \tbGE{to watch out for} \tbSN \tbGE{to look after} \tbSN \tbGE{to protect, to defend} \tbSN \tbGE{to wait for (someone),} \tbSN \tbGE{to waylay (someone)} \tbPD{gue, guuro, guuro, guribo, guuraa, guribo } \tbVA{go}\tbSE{gu zɔɔ} \tbGE{to ambush} 
\tbLX{gufu} \tbHM{} \tbPH{gúfù} \tbPS{n.~} \tbGE{interpreter} \tbOP{gufuri} \tbTP{gufuree/gufurree} 
\tbLX{Gumo} \tbHM{} \tbPH{gúmó} \tbPS{n.~} \tbGE{a town in the Sisaalaland noted for soothsaying and divining} \tbTP{gumee} 
\tbLX{gunni} \tbHM{} \tbPH{gúnní} \tbPS{v.~} \tbGE{to stalk (hunter tracking game)} \tbPD{gunnee, gunno, gunno, gunimo, gunnaa, - } 
\tbLX{gunni} \tbHM{} \tbPH{gùnnì} \tbPS{v.~} \tbGE{to pick tiny leaves from plants} \tbPD{gunnee, gunno, gunno, gunimo, gunnaa, - } 
\tbLX{guo} \tbHM{} \tbPH{gùò} \tbPS{n.~} \tbGE{heavy club with knobbed end} \tbOP{gori} \tbTP{goree} \tbVA{daguo}
\tbLX{guoli} \tbHM{} \tbPH{gùòlì} \tbPS{v.~} \tbGE{to coil (rope)} \tbPD{guolee, guolo, guolo, guolibo, guolaa, guolibo } 
\tbLX{guoli} \tbHM{} \tbPH{gúólì} \tbPS{v.~} \tbSN \tbGE{to bring up (child)} \tbSN \tbGE{to raise (animals)} \tbPD{guolee, guolo, guolo, guolibo, guolaa, guolibo } 
\tbLX{guoli} \tbHM{} \tbPH{gúólí} \tbPS{n.~} \tbGE{pot rest; pad on which pots are placed} \tbOP{guolo} \tbTP{guolee} 
\tbLX{guori} \tbHM{} \tbPH{gúórí} \tbPS{n.~} \tbGE{kola nuts} \tbSG{guoruu} \tbOP{goe} \tbTP{goree} 
\tbLX{guri} \tbHM{} \tbPH{gùri} \tbPS{v.~} \tbSN \tbGE{to throw one's body onto another (wrestling, catching a fowl)} \tbSN \tbGE{to fold up (leg)} \tbPD{guree, guro, guro, guribo, guraa, - } 
\tbLX{guri pɔge} \tbHM{} \tbPH{gùrì pɔ́gɪ́} \tbPS{v.~} \tbGE{to rape} \tbPD{-, -, -, -, -, - } 
\tbLX{gurri} \tbHM{} \tbPH{gúrrí} \tbPS{n.~} \tbGE{mushroom} \tbSG{gurruu} \tbTP{gurree} 
\tbLX{guruŋ} \tbHM{} \tbPH{gùrúŋ} \tbPS{n.~} \tbGE{mushroom} \tbSG{gurruu} \tbOP{gurri} \tbTP{gurree} 
\tbLX{guu} \tbHM{} \tbPH{gúú} \tbPS{v.~} \tbGE{to be unable, to fail} \tbPD{guue, guuro, guuro, guuribo, guuraa, guuribo } 
\tbLX{guu} \tbHM{} \tbPH{gùú} \tbPS{adj.~} \tbGE{blunt, not sharp} \tbOP{guuri} \tbTP{guuree} 
\tbLX{guu} \tbHM{} \tbPH{gùù} \tbPS{v.~} \tbGE{to be blunt} \tbPD{guue, guuro, guuro, guurbo, guuraa, guurbo } 
\tbLX{guuli} \tbHM{} \tbPH{gúúlì} \tbPS{v.~} \tbGE{to delay someone} \tbPD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \tbVA{gbuuli}
\tbLX{guuli} \tbHM{} \tbPH{gúúlì} \tbPS{v.~} \tbGE{to be not yet healed of a sprain} \tbPD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \tbVA{gbuuli}
\tbLX{guuli} \tbHM{} \tbPH{gúúlì} \tbPS{v.~} \tbGE{to delay someone} \tbPD{guulee, guulo, guulo, guulibo, guulaa, guulibo } \tbVA{gbuuli}
\tbLX{guuli} \tbHM{} \tbPH{gúúlí} \tbPS{n.~} \tbGE{knot} \tbOP{guulo} \tbTP{guulee} 
\tbLX{gu zɔɔ} \tbHM{} \tbPH{gù zɔ̀ɔ́} \tbPS{v.~} \tbGE{to ambush} \tbPD{gue  zɔɔ, guuro zɔɔ, zɔɔ guuro, -, -, - } 
\tbLX{gba} \tbHM{} \tbPH{gbá} \tbPS{v.~} \tbSN \tbGE{to gather substantive debris} \tbSN \tbGE{to beat someone completely (gambling)} \tbPD{gbaɛ, gbaara, gbaara, gbarba, gbaaraa, gbarba } 
\tbLX{gba} \tbHM{} \tbPH{gbà} \tbPS{v.~} \tbGE{to close tightly, to lock} \tbPD{gbaɛ, gbaara, gbaara, gbareba, gbaaraa, gbareba } 
\tbLX{gbaa} \tbHM{} \tbPH{gbáà} \tbPS{n.~} \tbGE{viper} \tbOP{gbaare} \tbTP{gbaarɛɛ} 
\tbLX{gbaalaa} \tbHM{} \tbPH{gbááláá} \tbPS{n.~} \tbGE{large, sturdy broom for sweeping pavement} \tbOP{gbaale} \tbTP{gbaalɛɛ} 
\tbLX{gbaale} \tbHM{} \tbPH{gbáálɪ́} \tbPS{n.~} \tbGE{large earthenware basin} \tbOP{gbaala} \tbTP{gbaalɛɛ} 
\tbLX{gbaale} \tbHM{} \tbPH{gbààlɪ̀} \tbPS{v.~} \tbGE{to gather extensively} \tbXV{A koɔraa gbaale la o weɛ kɛkɛrɛɛ baare.} \tbXE{The farmer has finished gathering the stalks on his farm.} \tbPD{gbaalɛɛ, gbaala, gbaala, gbaaleba, gbaalaa, gbaaleba } 
\tbLX{gbaale} \tbHM{} \tbPH{gbáálɪ̀} \tbPS{v.~} \tbGE{to follow in large numbers (typically used with tu)} \tbXV{Polsiri nyɔge la a nannyige ka bibiiri gbaale tu.} \tbXE{The police arrested the thief and children crowded behind them.} \tbPD{gbaalɛɛ, gbaala, gbaala, gbaalba, gbaalaa, gbaalba } 
\tbLX{gbaama} \tbHM{} \tbPH{gbààmà} \tbPS{n.~} \tbGE{weak appendages} \tbTP{gbaamɛɛ} 
\tbLX{gbaane} \tbHM{} \tbPH{gbàànɪ́} \tbPS{adj.~} \tbGE{lean} \tbOP{gbaama} 
\tbLX{gbaga} \tbHM{} \tbPH{gbágá} \tbPS{n.~} \tbGE{animal cage} \tbOP{gbagre} \tbTP{gbagrɛɛ} \tbVA{gbɔɡɔ}
\tbLX{gbage} \tbHM{} \tbPH{gbàgɪ̀} \tbPS{v.~} \tbGE{to face} \tbXV{A peroo yie la a de o niŋe gbage sapare kyɛ deɛ zoro.} \tbXE{When the sheep left the pen it just went running toward the east.} \tbPD{gbagɛɛ, gbagra, gbagra, gbagreba, gbagraa, gbagreba } \tbVA{gbɔge}
\tbLX{gbagelaŋ} \tbHM{} \tbPH{gbàglàŋ} \tbPS{adj.~} \tbGE{large and strong e.g. cob of corn} 
\tbLX{gbagele} \tbHM{} \tbPH{gbáglɪ́} \tbPS{n.~} \tbGE{bone with meat} \tbOP{gbagla} \tbTP{gbaglɛɛ} 
\tbLX{gbala} \tbHM{} \tbPH{gbàlà} \tbPS{n.~} \tbGE{a type of women's hairstyle} \tbXV{A pɔgepaalaa poŋi la gbala.} \tbXE{The bride has shaved gbala on her head.} \tbOP{gballe} \tbTP{gbalɛɛ} 
\tbLX{gbalaa} \tbHM{} \tbPH{gbàláà} \tbPS{n.~} \tbSN \tbGE{pole for plucking} \tbSN \tbGE{a type of tree that has aerial roots, milky sap, and (usually) a large amount of shade} \tbOP{gballe} \tbTP{gballɛɛ} \tbVA{laŋgbaraa, kaŋkaŋgbalaa}
\tbLX{gbale} \tbHM{} \tbPH{gbàlɪ́} \tbPS{adj.~} \tbGE{It is not for nothing that the child is crying} \tbOP{gbala} \tbTP{gbalɛɛ} 
\tbLX{gbale} \tbHM{} \tbPH{gbàlɪ́} \tbPS{adj.~} \tbGE{not wearing anything arround the waist} \tbXV{Maria taa la se-gbale.} \tbXE{Mary does not have any beads around her waist.} \tbOP{gbala} \tbTP{gbalɛɛ} 
\tbLX{gbale} \tbHM{} \tbPH{gbàlɪ́} \tbPS{adj.~} \tbGE{blunt, not sharp} \tbXV{A sɔ-gbale na koŋ to᷉ɔ ŋmaa a nɛne.} \tbXE{That blunt knife can't cut the meat.} \tbOP{gbala} \tbTP{gbalɛɛ} 
\tbLX{gbale} \tbHM{} \tbPH{gbàlɪ̀} \tbPS{v.~} \tbGE{to make blunt} \tbPD{gbalɛɛ, gballa, gballa, gbaleba, gballaa, gbaleba } 
\tbLX{gbambelɛ} \tbHM{} \tbPH{gbàmbɪ̀lɛ̀} \tbPS{n.~} \tbGE{albino} \tbOP{gbambelle} \tbTP{gbambelɛɛ} 
\tbLX{gbambiri} \tbHM{} \tbPH{gbàmbírì} \tbPS{n.~} \tbGE{an object with two distinct sides (cowrie) thrown in gambling to decide the winner} \tbSG{gbambiruu} \tbOP{gbambie} \tbTP{gbambiiree} 
\tbLX{gbambiri} \tbHM{} \tbPH{gbàmbìrì} \tbPS{n.~} \tbGE{a type of musical instrument, made of metal tin, a wooden board and spokes. Spoke-like metals are fixed on the board and fixed into the tin. It is played like the guitar} \tbTP{gbambiree} 
\tbLX{gbammo} \tbHM{} \tbPH{gbàmmʊ́} \tbPS{n.~} \tbSN \tbGE{comb (of a rooster)} \tbSN \tbGE{a variety of maize that usually has big cobs and pure white kernels} \tbOP{gbama/gbanne} \tbTP{gbamɛɛ} \tbVA{gyampegre}
\tbLX{gbandige} \tbHM{} \tbPH{gbǎndígé} \tbPS{n.~} \tbGE{area of jurisdiction} \tbOP{gbandigri} 
\tbLX{gbandiruŋ} \tbHM{} \tbPH{gbǎndírúŋ} \tbPS{n.~} \tbGE{inheritance} \tbTP{gbandiree} 
\tbLX{gbanfolɔ} \tbHM{} \tbPH{gbànfʊ̀là} \tbPS{n.~} \tbGE{a dish made from mashed fresh corn} \tbTP{gbanfolɛɛ/gbanfollɛɛ} \tbVA{gbanfola}
\tbLX{gbanna} \tbHM{} \tbPH{gbánná} \tbPS{n.~} \tbGE{small farm near the house} \tbOP{gbanne} \tbTP{gbannɛɛ} \tbVA{gbannaa}
\tbLX{gbannaa} \tbHM{} \tbPH{gbánnáá} \tbPS{n.~} \tbGE{crop field} \tbOP{gbanne} \tbTP{gbannɛɛ} \tbVA{gbanna}
\tbLX{gbannaa} \tbHM{} \tbPH{gbànnà} \tbPS{n.~} \tbGE{see gbannaa (crop field)} \tbOP{gbanne} \tbTP{gbannɛɛ} \tbVA{gbanna}
\tbLX{gbantieloo} \tbHM{} \tbPH{gbàntíélóó} \tbPS{n.~} \tbGE{the act of shooting an arrow as high as one can. (used with "saluoni')} \tbXV{O gbaŋ la saluoni gbantieloo.} \tbXE{He has shot an arrow into the sky.} 
\tbLX{gbantori} \tbHM{} \tbPH{gbàntòrí} \tbPS{n.~} \tbGE{ulcer} \tbOP{gbantoe} \tbTP{gbantoree} \tbVA{gbontori}
\tbLX{gbanzeɛ} \tbHM{} \tbPH{gbànzɪ̀ɛ́} \tbPS{n.~} \tbGE{very light-skinned person} \tbOP{gbanzeere} \tbTP{gbanzeerɛɛ} \tbVA{gbaŋ}
\tbLX{gbaŋe} \tbHM{} \tbPH{gbáŋ} \tbPS{n.~} \tbGE{gambling} \tbOP{gbanne} \tbTP{gbannɛɛ} \tbVA{gbaŋa}
\tbLX{gbaŋe} \tbHM{} \tbPH{gbǎŋ} \tbPS{n.~} \tbGE{albino} \tbOP{gbanne} \tbTP{gbannɛɛ} \tbVA{gbanzeɛ}
\tbLX{gbaŋe} \tbHM{} \tbPH{gbáŋ} \tbPS{v.~} \tbGE{to shoot an arrow as far as one can} \tbPD{gbaŋɛɛ, gbana, gbana, gbanema, gbanaa, gbanema } 
\tbLX{gbaŋgbaa} \tbHM{} \tbPH{gbàŋgbáà} \tbPS{n.~} \tbGE{slave} \tbOP{gbaŋgbaare} \tbTP{gbaŋgbaarɛɛ} 
\tbLX{gbaŋgbaaloŋ} \tbHM{} \tbPH{gbàŋgbààlʊ́ŋ} \tbPS{n.~} \tbGE{slavery} 
\tbLX{gbaŋgbakyeraa} \tbHM{} \tbPH{gbáŋgbákyɪ́ráà} \tbPS{n.~} \tbGE{stone axe (chopper?)} \tbOP{gbaŋgbakyerre} \tbTP{gbaŋgbakyerɛɛ/gbaŋgbakyerrɛɛ} 
\tbLX{gbaŋgbalaa} \tbHM{} \tbPH{gbáŋgbáláà} \tbPS{n.~} \tbSN \tbGE{a clearing} \tbSN \tbGE{spot for threshing grain} \tbOP{gbaŋgballe} \tbTP{gbaŋgbalɛɛ} \tbVA{gbeŋgbelaa}
\tbLX{gbaŋgbaŋ} \tbHM{} \tbPH{gbáŋgbáŋ} \tbPS{n.~} \tbGE{midday} \tbOP{gbaŋgbanne} \tbTP{gbaŋgbannɛɛ} 
\tbLX{gbaŋgbaŋ} \tbHM{} \tbPH{gbàŋgbàŋ} \tbPS{adv.~} \tbGE{wide, open, and endless} 
\tbLX{gbaŋgbele} \tbHM{} \tbPH{gbàŋgbɪ́lɪ́} \tbPS{adv.~} \tbGE{raw and hard (fruit)} 
\tbLX{gbaŋkeroo} \tbHM{} \tbPH{gbàŋkéróó} \tbPS{n.~} \tbGE{a type of wild plant with edible fruit} \tbOP{gbaŋkerri} \tbTP{gbaŋkeree/gbaŋkerree} 
\tbLX{gbara} \tbHM{} \tbPH{gbàrá} \tbPS{n.~} \tbGE{a deaf or mute person} \tbOP{gbarre} \tbTP{gbarɛɛ} 
\tbLX{gbare} \tbHM{} \tbPH{gbárɪ́} \tbPS{v.~} \tbGE{to struggle in agony e.g. dying} \tbXV{Lɔɔre ŋmɛ la naabo ka o le a sori zu deɛ gbara.} \tbXE{The lorry hit the cow and it fell on the road struggling.} \tbPD{gbarɛɛ, gbara, gbara, gbarba, gabaraa, gabrba } 
\tbLX{gbare} \tbHM{} \tbPH{gbárɪ́} \tbPS{v.~} \tbGE{to tie legs together firmly} \tbXV{Ba naneŋ nyɔge la a naaweɛmɛ na, a nyɔge gbare.} \tbXE{When they eventually caught that wild cow, they tied it down by the legs.} \tbPD{gbarɛɛ, gbara, gbara, gbarba, gabaraa, gabrba } 
\tbLX{gbare} \tbHM{} \tbPH{gbàrɪ̀} \tbPS{v.~} \tbGE{to shut off from hearing} \tbPD{gbarɛɛ, gbara, gbara, gbareba, gbaraa, gbareba } 
\tbLX{gbarema} \tbHM{} \tbPH{gbàrɪ̀má} \tbPS{n.~} \tbGE{small child} \tbOP{gbarre} \tbTP{gbaremɛɛ} 
\tbLX{gbateele} \tbHM{} \tbPH{gbàtɪ́ɪ́lɪ́} \tbPS{n.~} \tbGE{a kind of grass which is heated and smashed to explode like knockout} \tbSG{gbateeloo} \tbOP{gbateelɛ} \tbTP{gbateelɛɛ} 
\tbLX{gbatɛre} \tbHM{} \tbPH{gbàtɛ́rɪ́} \tbPS{n.~} \tbGE{small, shallow pond (usually in the bush)} \tbOP{gbatɛrre} \tbTP{gbatɛrɛɛ} 
\tbLX{gbe} \tbHM{} \tbPH{gbè} \tbPS{v.~} \tbSN \tbGE{to grind coarsely} \tbSN \tbGE{to knaw at something hard; (unripe fruit)} \tbPD{gbee, gbiere, gbiere, gbieribe, gbieraa, gbegbiere } 
\tbLX{gbe᷉ɛ} \tbHM{} \tbPH{gbɪ́᷉ɛ́} \tbPS{v.~} \tbGE{to cut a bit tiny} \tbPD{gbeɛ᷉ɛ, gbeɛnɛ, gbeɛnɛ, gbeɛnemɛ, gbeɛnaa, gbeɛnemɛ } 
\tbLX{Gbeɛdomoo} \tbHM{} \tbPH{gbɪ̀ɛ̀dʊ́mʊ́ʊ́} \tbPS{n.~} \tbGE{a member of the gbeɛdoma clan} \tbOP{Gbeɛdomɛɛ} 
\tbLX{Gbeɛdomɔ} \tbHM{} \tbPH{gbɪ̀ɛ̀dʊmɔ́} \tbPS{n.~} \tbGE{land area of the clan members (Gbeɛdomɛɛ)} \tbVA{Gbeɛdoma}
\tbLX{gbeɛle} \tbHM{} \tbPH{gbɪ̀ɛ̀lɪ̀} \tbPS{n.~} \tbGE{a type of fetish manifested in a stone; a stone believed to have supernatural power and treated as a god. The clan god of the Gbeɛdomɛɛ} \tbOP{gbeɛlɛ} \tbTP{gbeɛlɛɛ} 
\tbLX{gbeɛle} \tbHM{} \tbPH{gbɪ́ɛ́lɪ́} \tbPS{v.~} \tbGE{to eye with suspicion, to look angrily at} \tbPD{gbeɛle, gbeɛlɛ, gbeɛlɛ, gbeɛlebɛ, gbeɛlaa, gbeɛlebɛ } 
\tbLX{gbeɛle;} \tbHM{} \tbPH{gbɪ́ɛ́lɪ̀} \tbPS{v.~} \tbGE{to look at sternly} \tbPD{-, -, -, -, -, - } 
\tbLX{gbeɛle kuuri} \tbHM{} \tbPH{gbɪ̀ɛ́lɪ̀ kúúrí} \tbPS{n.~} \tbSN \tbGE{round, smooth stone} \tbSN \tbGE{such a stone believed to be a symbol of a fetish} \tbOP{gbeɛle kubo} \tbTP{gbeɛle kuuree} \tbVA{gbeɛle}
\tbLX{gbeɛre} \tbHM{} \tbPH{gbɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to be completely out of energy; to be spent} \tbPD{gbeɛre, gbeɛrɛ, gbeɛrɛ, gbeɛrebɛ, gbeɛraa, gbeɛrebɛ } 
\tbLX{gbeɛre} \tbHM{} \tbPH{gbɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to be felled by a rainstorm (of grass)} \tbPD{gbeɛre, gbeɛrɛ, gbeɛrɛ, gbeɛrebɛ, gbeɛraa, gbeɛrebɛ } 
\tbLX{gbeɛre} \tbHM{} \tbPH{gbɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to cut bits and pieces of meat} \tbPD{gbeɛre, gbeɛrɛ, gbeɛrɛ, gbeɛrebɛ, gbeɛraa, gbeɛrebɛ } 
\tbLX{gbeɛrɛ} \tbHM{} \tbPH{gbɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{bits and pieces of meat} \tbTP{gbeɛrɛɛ} 
\tbLX{gbegi} \tbHM{} \tbPH{gbégí} \tbPS{v.~} \tbGE{to pile up e.g. firewood} \tbPD{gbegee, gbegre, gbegire, gbegiribe, gbegiraa, gbegire } 
\tbLX{gbegili} \tbHM{} \tbPH{gbèglì} \tbPS{adj.~} \tbGE{hard and rigid} 
\tbLX{gbegili} \tbHM{} \tbPH{gbèglì} \tbPS{v.~} \tbGE{to become dry and hard} \tbPD{gbegilee, gbegile, gbegile, gbeglibe, gbegilaa, gbegilibe } 
\tbLX{gbelaa} \tbHM{} \tbPH{gbèláá} \tbPS{n.~} \tbGE{the bone of the forhead} \tbOP{gbelaare} \tbTP{gbelaarɛɛ} 
\tbLX{gbelaa} \tbHM{} \tbPH{gbɪ̀làà} \tbPS{adv.~} \tbGE{protruding, straight and hard} 
\tbLX{gbele} \tbHM{} \tbPH{gbɪ́lɪ́} \tbPS{n.~} \tbGE{a type of small fox with bushy fur} \tbOP{gbelɛ} \tbTP{gbelɛɛ/gbellɛɛ} 
\tbLX{gbele} \tbHM{} \tbPH{gbɪ́lɪ́} \tbPS{v.~} \tbGE{to work very hard} \tbPD{gbelɛɛ, gbelɛ, gbelɛ, gbelbɛ, gbelaa, - } 
\tbLX{gbeleŋgbeleŋ} \tbHM{} \tbPH{gbèlèŋgbéléŋ} \tbPS{n.~} \tbGE{flat metal musical instrument; a cymbal} \tbOP{gbeleŋgbelli} \tbTP{gbeleŋgbelee} \tbVA{lato/nato}
\tbLX{gbelime} \tbHM{} \tbPH{gbèlìmé} \tbPS{n.~} \tbGE{bell} \tbOP{gbelli} \tbTP{gbellee} 
\tbLX{gbelime} \tbHM{} \tbPH{gbèlìmé} \tbPS{n.~} \tbGE{hour} \tbOP{gbelli} \tbTP{gbellee} 
\tbLX{gbelli} \tbHM{} \tbPH{gbéllì} \tbPS{v.~} \tbGE{to turn the head away abruptly} \tbPD{gbellee, gbelle, gbelle, gbelibe, gbellaa, gbelibe } 
\tbLX{gbeŋgbegri} \tbHM{} \tbPH{gbèŋgbégrí} \tbPS{n.~} \tbSN \tbGE{river bank} \tbSN \tbGE{riverbed (dry)} \tbOP{gbeŋgbege} \tbTP{gbeŋgbegree} 
\tbLX{gbeŋgbelaa} \tbHM{} \tbPH{gbɪ́ŋgbɪ́láá} \tbPS{n.~} \tbGE{ground where millet is threshed and winnowed out} \tbOP{gbeŋgbelle} \tbTP{gbeŋgbelɛɛ/gbeŋgbellɛɛ} \tbVA{gbaŋgbalaa}
\tbLX{gbeŋgbeŋ} \tbHM{} \tbPH{gbéŋgbéŋ} \tbPS{adv.~} \tbGE{firmly fixed} 
\tbLX{gbeŋini} \tbHM{} \tbPH{gbéŋìní} \tbPS{n.~} \tbGE{lion} \tbOP{gbeŋime} \tbTP{gbennee} 
\tbLX{gbeŋkeroo} \tbHM{} \tbPH{gbèŋkéróó} \tbPS{n.~} \tbGE{a shrub with brown hard fruits that are edible. Also the fruit of such a tree} \tbOP{gbeŋkerri} \tbTP{gbeŋkeree} \tbVA{naakyeemɛgbeŋkeroo}
\tbLX{gberi} \tbHM{} \tbPH{gbèrì} \tbPS{adv.~} \tbGE{of a twine snapping sound} 
\tbLX{gberi} \tbHM{} \tbPH{gbérì} \tbPS{adv.~} \tbGE{stopping suddenly} 
\tbLX{gbɛbaŋa} \tbHM{} \tbPH{gbɛ́báŋ} \tbPS{n.~} \tbGE{ankle bangle} \tbOP{gbɛbanne} \tbTP{gbɛbannɛɛ} \tbVA{gbɛbɔŋɔ/gbɛbaŋe}
\tbLX{gbɛbi-eɛne} \tbHM{} \tbPH{gbɛ́bì-ɪ́ɛ́nɪ́} \tbPS{n.~} \tbSN \tbGE{toenail} \tbSN \tbGE{claw} \tbOP{gbɛbi-eɛmɛ} \tbTP{gbɛbi-eɛnɛɛ} 
\tbLX{gbɛbinyoɔre} \tbHM{} \tbPH{gbɛ́bìɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{tip of the toe} \tbOP{gbɛbinyɔɛ} \tbTP{gbɛbinyɔrɛɛ} 
\tbLX{gbɛbiri} \tbHM{} \tbPH{gbɛ́bìrí} \tbPS{n.~} \tbGE{toe} \tbOP{gbɛbie} \tbTP{gbɛbiree/gbɛbiiree} 
\tbLX{gbɛbogi} \tbHM{} \tbPH{gbɛ́bògí} \tbPS{n.~} \tbSN \tbGE{footprint} \tbSN \tbGE{spot} \tbOP{gbɛbogri} \tbTP{gbɛbogree} 
\tbLX{gbɛ᷉ɛ} \tbHM{} \tbPH{gbɛ̀᷉ɛ̀᷉} \tbPS{adv.~} \tbGE{dull, sleepy} 
\tbLX{gbɛɛ yaga} \tbHM{} \tbPH{gbɛ́ɛ́ yágá} \tbPS{adj.~} \tbGE{often} \tbVA{buyibuyi}
\tbLX{gbɛfaa} \tbHM{} \tbPH{gbɛ́fáá} \tbPS{n.~} \tbGE{ill luck; unluckiness} \tbOP{gbɛfaare} \tbTP{gbɛfaarɛɛ} 
\tbLX{gbɛgere} \tbHM{} \tbPH{gbɛ̀grɪ̀} \tbPS{v.~} \tbGE{to be miserly, to economize} \tbPD{gbɛgerɛɛ, gbɛgerɛ, gbɛgerɛ, gbɛgerebɛ, gbɛgeraa, gbɛgerebɛ } 
\tbLX{gbɛgere} \tbHM{} \tbPH{gbɛ́grɪ́} \tbPS{v.~} \tbGE{to be loose, to be shaky (object in a hole)} \tbXV{A daa bogi ba sigi yaga lɛnso ka o gbɛgrɛ.} \tbXE{The hole for the pole is not deep enough; that is why it is shaky.} \tbPD{gbɛgrɛɛ, gbɛgrɛ, gbɛgrɛ, gbɛgrebɛ, gbɛgraa, gbɛgrebɛ } 
\tbLX{gbɛgbɛ} \tbHM{} \tbPH{gbɛ̀gbɛ́} \tbPS{n.~} \tbGE{wide, shallow basket} \tbOP{gbɛgbɛre} \tbTP{gbɛgbɛrɛɛ} 
\tbLX{gbɛkpara} \tbHM{} \tbPH{gbɛ́kpárá} \tbPS{n.~} \tbGE{double legged} \tbOP{gbɛkparre} \tbTP{gbɛkparɛɛ} 
\tbLX{gbɛkpeɛŋaa} \tbHM{} \tbPH{gbɛ́kpɪ́ɛ́ŋáá} \tbPS{n.~} \tbGE{fleet footed} \tbOP{gbɛkpeɛne} 
\tbLX{gbɛkpoŋi} \tbHM{} \tbPH{gbɛ́kpóŋ} \tbPS{n.~} \tbGE{thigh} \tbOP{gbɛkponni} \tbTP{gbɛkponnee} 
\tbLX{gbɛle} \tbHM{} \tbPH{gbɛ́lɪ̀} \tbPS{n.~} \tbGE{a long, thin strip of material e.g. rubber} \tbOP{gbɛlɛ} \tbTP{gbɛlɛɛ/gbɛllɛɛ} 
\tbLX{gbɛle} \tbHM{} \tbPH{gbɛ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to wriggle one's trunk (like in dancing)} \tbSN \tbGE{way of begging where one wriggles one's wrist} \tbPD{gbɛlɛɛ, gbɛlɛ, gbɛlɛ, gbɛlbɛ, gbɛlaa, gbɛlbɛ } 
\tbLX{gbɛlee} \tbHM{} \tbPH{gbɛ́léé} \tbPS{n.~} \tbGE{foot} \tbOP{gbɛbilii} \tbTP{gbɛbilee} 
\tbLX{gbɛlɛɛ} \tbHM{} \tbPH{gbɛ̀lɛ̀ɛ̀} \tbPS{adj.~} \tbGE{loose and sagging e.g. a rope} 
\tbLX{gbɛlgbɛle} \tbHM{} \tbPH{gbɛ̀lgbɛ̀l} \tbPS{adv.~} \tbGE{long and flowing} 
\tbLX{gbɛloŋkoraa} \tbHM{} \tbPH{gbɛ́lòŋkòràá} \tbPS{n.~} \tbGE{the region at the back of the knee joint} \tbOP{gbɛloŋkorri} \tbTP{gbɛloŋkoree} 
\tbLX{gbɛlɔlɔraa} \tbHM{} \tbPH{gbɛ́lɔ̀lɔ̀ráà} \tbPS{n.~} \tbGE{calf (human body)} \tbOP{gbɛlɔlɔrre} \tbTP{gbɛlɔlɔrɛɛ} 
\tbLX{gbɛneɛraa} \tbHM{} \tbPH{gbɛ́níɛ́ráá} \tbPS{n.~} \tbGE{footprint} \tbOP{gbɛneɛre} \tbTP{gbɛneɛrɛɛ} 
\tbLX{gbɛnne} \tbHM{} \tbPH{gbɛ̀nnɪ̀} \tbPS{v.~} \tbGE{to pluck (leaves)} \tbPD{gbɛnnɛɛ, gbɛnnɛ, gbɛnnɛ, gbɛnemɛ, gbɛnnaa, gbɛnemɛ } 
\tbLX{gbɛnne} \tbHM{} \tbPH{gbɛ́nnɪ́} \tbPS{v.~} \tbGE{to chip the rim or edge of something} \tbPD{gbɛnnɛɛ, gbɛnnɛ, gbɛnnɛ, gbɛnemɛ, gbɛnnaa, gbɛnemɛ } 
\tbLX{gbɛre} \tbHM{} \tbPH{gbɛ́rɪ̀} \tbPS{n.~} \tbSN \tbGE{leg} \tbSN \tbGE{foot} \tbOP{gbɛɛ} \tbTP{gbɛɛrɛɛ} 
\tbLX{gbɛre} \tbHM{} \tbPH{gbɛ́rɪ́} \tbPS{n.~} \tbGE{person with a physical disability that prevents them from walking} \tbOP{gbɛrre} \tbTP{gbɛrɛɛ} \tbVA{gbɛɛ (1pl)}
\tbLX{gbɛre} \tbHM{} \tbPH{gbɛ̀rɪ́} \tbPS{n.~} \tbGE{mane} \tbOP{gbɛrre} \tbTP{gbɛrɛɛ} 
\tbLX{gbɛre} \tbHM{} \tbPH{gbɛ̀rɪ̀} \tbPS{v.~} \tbGE{to be unable to walk due to a physical disability} \tbPD{gbɛrɛɛ, gbɛrɛ, gbɛrɛ, gbɛrebɛ, gbɛraa, gbɛrebɛ } 
\tbLX{gbɛregbɛre} \tbHM{} \tbPH{gbɛ̀rɪ̀gbɛ́rɪ́} \tbPS{v.~} \tbGE{crawl, walk with serious difficulty} \tbPD{gbɛregbɛrɛɛ, gbɛregbɛrɛ, gbɛregbɛrɛ, gbɛregbɛrbɛ, gbɛregbɛraa, gbɛregbɛrbɛ } 
\tbLX{gbɛrɛ} \tbHM{} \tbPH{gbɛ̀rɛ̀} \tbPS{n.~} \tbGE{sponge} \tbSG{gbɛrrɛɛ} \tbTP{gbɛrɛɛ} 
\tbLX{gbɛroo} \tbHM{} \tbPH{gbɛ̀rʊ̀ʊ̀} \tbPS{n.~} \tbGE{weak and dejected; paralysis} 
\tbLX{gbɛsampane} \tbHM{} \tbPH{gbɛ́sámpánɪ́} \tbPS{n.~} \tbGE{top of foot} \tbOP{gbɛsampama} \tbTP{gbɛsampanɛɛ/gbɛsampannɛɛ} 
\tbLX{gbɛsugili} \tbHM{} \tbPH{gbɛ́súglí} \tbPS{n.~} \tbGE{ankle} \tbOP{gbɛsugilo} \tbTP{gbɛsugilee} 
\tbLX{gbɛtanlori} \tbHM{} \tbPH{gbɛ́tánlórí} \tbPS{n.~} \tbGE{half-bent leg on the other} \tbOP{gbɛtanloe} \tbTP{gbɛtanloree} 
\tbLX{gbɛtɛre} \tbHM{} \tbPH{gbɛ̀tɛ́rɪ́} \tbPS{n.~} \tbGE{small pond, usually in the bush} \tbOP{gbɛtɛrre} \tbTP{gbɛtɛrɛɛ/gbɛtɛrrɛɛ} \tbVA{gbatare}
\tbLX{gbɛtoloŋ} \tbHM{} \tbPH{gbɛ́tʊ́lʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{the condition of being fast-footed; quickness} \tbSN \tbGE{intelligence} \tbOP{gbɛtola.} 
\tbLX{gbɛtooraa} \tbHM{} \tbPH{gbɛ́tʊ́ʊ́ráá} \tbPS{n.~} \tbGE{sock} \tbOP{gbɛtoore} \tbTP{gbɛtoorɛɛ} 
\tbLX{gbɛvelaa} \tbHM{} \tbPH{gbɛ́vɪ́làá} \tbPS{n.~} \tbGE{good luck, good fortune} \tbOP{gbɛveɛle} 
\tbLX{gbɛyeni} \tbHM{} \tbPH{gbɛ́yénì} \tbPS{n.~} \tbGE{one-legged thing} 
\tbLX{gbi} \tbHM{} \tbPH{gbí} \tbPS{v.~} \tbGE{to kneel} \tbPD{gbie, gbiire, gbiire, gbiribe, gbiiraa, gbiribe } 
\tbLX{gbi} \tbHM{} \tbPH{gbí} \tbPS{v.~} \tbGE{to break down a mud or earth-made roof} \tbPD{gbie, gbiire, gbiire, gbiribe, gbiiraa, gbiribe } 
\tbLX{gbi} \tbHM{} \tbPH{gbí} \tbPS{v.~} \tbGE{to harvest yams to make room for seed yams to grow} \tbPD{gbie, gbiire, gbiire, gbiribe, gbiiraa, gbiribe } 
\tbLX{gbie} \tbHM{} \tbPH{gbìé} \tbPS{n.~} \tbGE{forehead} \tbSG{gberuu} \tbOP{gberi} \tbTP{gberee} 
\tbLX{gbiele} \tbHM{} \tbPH{gbíélé} \tbPS{n.~} \tbSN \tbGE{hands and arms} \tbXV{N ba taa gbiele derɛ biiri.} \tbXE{My arms are too week to curdle children.} 
\tbLX{gbieli} \tbHM{} \tbPH{gbíélí} \tbPS{n.~} \tbGE{elbow} \tbOP{gbiele} \tbTP{gbielee} \tbVA{shortened form of kpaŋkpannyugbieli, kpaŋkpaŋgbieli}
\tbLX{gbieni} \tbHM{} \tbPH{gbìènì} \tbPS{v.~} \tbGE{to crawl in pain, to crawl painfully} \tbPD{gbienee, gbiene, gbiene, gbienime, gbienaa, gbienime } 
\tbLX{gbigiri} \tbHM{} \tbPH{gbígrì} \tbPS{v.~} \tbGE{to ruminate, to chew cud} \tbPD{gbigiree, gbigire, gbigire, gbigiribe, gbigiraa, gbigiribe } 
\tbLX{gbigiri} \tbHM{} \tbPH{gbígrì} \tbPS{v.~} \tbGE{to shake to test firmness e.g. a stem} \tbPD{gbigiree, gbigire, gbigire, gbigiribe, gbigiraa, gbigiribe } 
\tbLX{gbili} \tbHM{} \tbPH{gbìlí} \tbPS{adj.~} \tbGE{silent, quiet} \tbOP{gbile} \tbTP{gbilee} 
\tbLX{gbini} \tbHM{} \tbPH{gbíní} \tbPS{n.~} \tbGE{bundle} \tbOP{gbime} 
\tbLX{gbiŋ-gbiŋ} \tbHM{} \tbPH{gbîŋ-gbîŋ} \tbPS{adv.~} \tbGE{manner of running that is uninterrupted} 
\tbLX{gbiŋi} \tbHM{} \tbPH{gbíŋ} \tbPS{v.~} \tbGE{to cause to tremble or shake} \tbXV{Bibiiri deɛnɛ la a peɛ lɛnso ka o gbine.} \tbXE{Children are playing on the roof; that is why the roof is shaking.} \tbPD{gbiŋee, gbine, gbine, gbinime, gbinaa, gbinime } 
\tbLX{gbiŋi} \tbHM{} \tbPH{gbíŋ} \tbPS{v.~} \tbGE{to hypnotise; to put into utter shock such that the person or thing cannot react} \tbXV{A dɔɔta kyɔge o la pimbiri ka o deɛ gbiŋi.} \tbXE{The doctor gave him an injection and he was put out.} \tbPD{gbiŋee, gbine, gbine, gbinime, gbinaa, gbinime } 
\tbLX{gbireŋgbireŋ} \tbHM{} \tbPH{gbíréŋgbíréŋ} \tbPS{adv.~} \tbGE{solid and firm} 
\tbLX{gbiri} \tbHM{} \tbPH{gbírì} \tbPS{v.~} \tbSN \tbGE{to sleep} \tbSN \tbGE{to coagulate, to clot} \tbPD{gbiree, gbire, gbire, gbiribe, gbiraa, gbiribe } 
\tbLX{gbo} \tbHM{} \tbPH{gbô} \tbPS{n.~} \tbGE{heart} \tbSG{gboruu} \tbOP{gbori} \tbTP{gboree} 
\tbLX{gbo} \tbHM{} \tbPH{gbò} \tbPS{adv.~} \tbGE{a loud sound, usually made when two solid things collide, e.g two bulls in a fight when their heads crash} 
\tbLX{gbogili} \tbHM{} \tbPH{gbóglì} \tbPS{v.~} \tbSN \tbGE{to pack many objects on top of one another} \tbSN \tbGE{to stand up against each other (quarrel)} \tbPD{gbogilee, gbogilo, gbogilo, gbogilibo, gbogilaa, - } 
\tbLX{gbogiri} \tbHM{} \tbPH{gbògrì} \tbPS{v.~} \tbSN \tbGE{to gnaw flesh from a bone} \tbSN \tbGE{to lick a container clean e.g. a palatable soup pot} \tbPD{gbogree, gbogiro, gbogiro, gbogiribo, gbogiraa, - } 
\tbLX{gbogbo} \tbHM{} \tbPH{gbògbò} \tbPS{adj.~} \tbGE{gluttonous} 
\tbLX{gbole} \tbHM{} \tbPH{gbʊ̀lɪ̀} \tbPS{v.~} \tbGE{to look up towards the sky} \tbXV{A naŋkpaana gbole kaa la a nuulee naŋ dɔɔ a teɛ zu.} \tbXE{The hunter looked up to see the bird in the tree.} \tbPD{gbolɛɛ, gbollɔ, gbollɔ, gboleba, gbollaa, gboleba } 
\tbLX{gboli} \tbHM{} \tbPH{gbòlí} \tbPS{adj.~} \tbGE{bare-headed (carrying a load on the head without a pad); without attachment} \tbOP{gbolo} \tbTP{gbolee} 
\tbLX{gboli} \tbHM{} \tbPH{gbólí} \tbPS{n.~} \tbGE{leather string worn around the waist} \tbSG{gbolee/gbollee} \tbOP{gbolo} \tbTP{gbollli} 
\tbLX{gboli} \tbHM{} \tbPH{gbólí} \tbPS{v.~} \tbGE{to search ernestly; to rummage} \tbPD{gbolee, gbolo, gbolo, gbolibo, gbolaa, gbolibo } 
\tbLX{gboligboli} \tbHM{} \tbPH{gbólgból} \tbPS{adj.~} \tbGE{sound of a loud echoing and persistent call} 
\tbLX{gbolli} \tbHM{} \tbPH{gbòllì} \tbPS{v.~} \tbGE{to harass; to pursue something or someone eagerly} \tbPD{gbollee, gbollo, gbollo, gbolbo, gbollaa, gbolibo } 
\tbLX{gbomo} \tbHM{} \tbPH{gbómò} \tbPS{n.~} \tbGE{lumps of matter used by a rodent to block its hole against intruders} \tbTP{gbonnee} 
\tbLX{gbontori} \tbHM{} \tbPH{gbòntòrí} \tbPS{n.~} \tbGE{ulcer} \tbOP{gbontoe} \tbTP{gbontoree} 
\tbLX{gboŋe} \tbHM{} \tbPH{gbʊ́ŋ} \tbPS{v.~} \tbGE{to put pressure on (demanding action)} \tbPD{gboŋɛɛ, gbonnɔ, gbonnɔ, gbonema, gbonnaa, gbonema } 
\tbLX{gboŋgboli} \tbHM{} \tbPH{gbóŋgbólí} \tbPS{n.~} \tbGE{big rodent} \tbOP{gboŋgbolo} \tbTP{gboŋgbolee} 
\tbLX{gboŋgboŋ} \tbHM{} \tbPH{gbóŋgbóŋ} \tbPS{adj.~} \tbGE{all of, completely} 
\tbLX{gboŋgboŋ} \tbHM{} \tbPH{gbòŋgbôŋ} \tbPS{n.~} \tbGE{bucket; metal container} \tbOP{gboŋgbonni} \tbTP{gboŋgbonee} 
\tbLX{gboŋgbopelaa} \tbHM{} \tbPH{gbòŋgbòpɪ̀làá} \tbPS{n.~} \tbGE{jackal} \tbOP{gboŋgbopeɛle} \tbTP{gboŋgbopeɛlɛɛ} 
\tbLX{gboŋgbori} \tbHM{} \tbPH{gbòŋgbòrí} \tbPS{n.~} \tbGE{hyena} \tbOP{gboŋgboe} \tbTP{gboŋgboree} 
\tbLX{gboŋi} \tbHM{1} \tbPH{gbòŋ} \tbPS{v.~} \tbGE{to press against between two things.(finger between two objects)} \tbPD{gboŋee, gbonno, gbonno, gbonimo, gbonnaa, gbonimo } 
\tbLX{gboŋi} \tbHM{2} \tbPH{gbòŋ} \tbPS{v.~} \tbSN \tbGE{to pile up; to pack together} \tbSN \tbGE{to block an entrance (of a rodent's hole)} \tbPD{gboŋee, gbonno, gbonno, gbonimo, gbonnaa, gbonimo } 
\tbLX{gbore} \tbHM{} \tbPH{gbʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to hold water in the mouth} \tbSN \tbGE{to conceal a hole (of a rodent)} \tbPD{gborɛɛ, gborɔ, gborɔ, gboreba, gboraa, gboreba } 
\tbLX{gboregbore} \tbHM{} \tbPH{gbʊ̀rgbʊ̀rɪ́} \tbPS{v.~} \tbGE{to gargle} \tbPD{gborgborɛɛ, gborgborɔ, gborgborɔ, gborgboreba, gboregboraa, gborgboreba } 
\tbLX{gbori} \tbHM{} \tbPH{gbórí} \tbPS{v.~} \tbGE{to move suddenly and strongly, to struggle} \tbXV{Vũũ nyege la a bie ka o gbori  na tage yi a daa a doge lɔɔ.} \tbXE{The fire burnt the child and as he struggled to move away he pushed the pot over.} \tbPD{gboree, gboro, gboro, gboribo, gboraa, gboribo } \tbVA{gbori}
\tbLX{gbori} \tbHM{} \tbPH{gbòrí} \tbPS{n.~} \tbGE{ship, boat, canoe} \tbOP{gboe} \tbTP{gboree/gborree} 
\tbLX{gbori} \tbHM{} \tbPH{gbòrí} \tbPS{n.~} \tbGE{door flap} \tbOP{gboe} \tbTP{gboree/gborree} 
\tbLX{gbori} \tbHM{} \tbPH{gbòrí} \tbPS{n.~} \tbGE{trough} \tbOP{gboe} \tbTP{gboree/gborree} 
\tbLX{gbori} \tbHM{} \tbPH{gbòrí} \tbPS{n.~} \tbGE{board on which meat is dressed and cut} \tbOP{gboe} \tbTP{gboree/gborree} 
\tbLX{gboribiri} \tbHM{} \tbPH{gbòrbírì} \tbPS{n.~} \tbGE{key} \tbOP{gboribie} \tbTP{gboribiree/gboribiiree} 
\tbLX{gborigbori} \tbHM{} \tbPH{gbòrgbórí} \tbPS{v.~} \tbGE{to struggle, especially on the ground} \tbPD{gborgboree, gborgboro, gborgboro, gborgborbo, gborgboraa, - } 
\tbLX{gborikuŋkuni} \tbHM{} \tbPH{gbòrìkúŋkúní} \tbPS{n.~} \tbGE{padlock} \tbOP{saafemamine} \tbVA{saafema}
\tbLX{gborivuuri} \tbHM{} \tbPH{gbòrvúúrì} \tbPS{n.~} \tbGE{paddle} \tbOP{gborivue} \tbTP{gborivuuree} 
\tbLX{gborogboro} \tbHM{} \tbPH{gbòrògbòrò} \tbPS{adj.~} \tbGE{very large e.g. grains} 
\tbLX{gboroto} \tbHM{} \tbPH{gbórótó} \tbPS{adv.~} \tbGE{the manner of meeting face-to-face suddenly} 
\tbLX{gbotole} \tbHM{} \tbPH{gbʊ̀tʊ́lɪ́} \tbPS{n.~} \tbGE{bush baby (animal)} \tbOP{gbotola} \tbTP{gbotolɛɛ/gbotollɛɛ} 
\tbLX{gbɔɡe} \tbHM{} \tbPH{gbɔ̀ɡɪ̀} \tbPS{v.~} \tbSN \tbGE{to be close to, to be next to} \tbXV{O zeŋ gbɔge la a baala.} \tbXE{He is seated close to the patient.} \tbSN \tbGE{to drink as a couple from one calabash at the same time (traditional kiss)} \tbPD{gbɔɡɛɛ, gbɔɡerɔ, gbɔɡerɔ, gbɔɡerebɔ, gbɔɡeraa, - } 
\tbLX{gbɔge} \tbHM{} \tbPH{gbɔ̀gɪ́} \tbPS{adj.~} \tbGE{not full e.g. sack of grain} \tbOP{gbɔgere} 
\tbLX{gbɔgele} \tbHM{} \tbPH{gbɔ̀glɪ̀} \tbPS{v.~} \tbGE{to hold too many things at a time} \tbPD{gbɔgelɛɛ, gbɔgelɔ, gbɔgelɔ, gbɔgelebɔ, gbɔgelaa, - } 
\tbLX{gbɔgele} \tbHM{} \tbPH{gbɔ̀glɪ̀} \tbPS{v.~} \tbGE{to assemble poorly e.g. broken-down machine} \tbPD{gbɔgelɛɛ, gbɔgelɔ, gbɔgelɔ, gbɔgelebɔ, gbɔgelaa, - } 
\tbLX{gbɔgɔ} \tbHM{} \tbPH{gbɔ̀gɔ̀} \tbPS{n.~} \tbGE{a type of reptile similar to the monitor lizard} \tbOP{gbɔgre} \tbTP{gbɔgrɛɛ} 
\tbLX{gbɔgɔ} \tbHM{} \tbPH{gbɔ́gɔ́} \tbPS{n.~} \tbGE{strong open cage for carrying animals} \tbOP{gbɔgre} \tbTP{gbɔgrɛɛ} \tbVA{gbaga}
\tbLX{gbugbulee} \tbHM{} \tbPH{gbúgbúlèè} \tbPS{n.~} \tbGE{tadpole} \tbOP{gbugbulli} \tbTP{gbugbullee} 
\tbLX{gbuli} \tbHM{} \tbPH{gbùlí} \tbPS{adj.~} \tbGE{whole} \tbOP{gbulo} \tbTP{gbulo gbulo} 
\tbLX{gbulli} \tbHM{} \tbPH{gbùllì} \tbPS{v.~} \tbSN \tbGE{to make into a ball} \tbSN \tbGE{to come together in large numbers} \tbXV{A pɔge mɔŋe la a fanfane baare a nyɔge gbulli gbulli sidi sidi.} \tbXE{When the woman finished making the soap she made it into balls of one cedi each.} \tbPD{-,  gbullo,  gbullo, gbulibo, gbullaa, gbulibo } 
\tbLX{gbutuli} \tbHM{} \tbPH{gbùtùlí} \tbPS{n.~} \tbGE{uninformed person; fool} \tbOP{gbutulo} \tbTP{gbutulee} 
\tbLX{gbuuli} \tbHM{} \tbPH{gbúúlí} \tbPS{v.~} \tbGE{see guuli (make a knot)} \tbPD{gbuulee, gbuulo, gbuulo, gbuulibo, gbuulaa, - } 
\tbLX{gyaaŋkoli} \tbHM{} \tbPH{dzàáŋkólí} \tbPS{n.~} \tbGE{the sacrum and coccyx} \tbOP{gyaaŋkolo} \tbTP{gyaaŋkolee} \tbVA{gyikoli}
\tbLX{gyaare} \tbHM{} \tbPH{dzààrɪ̀} \tbPS{v.~} \tbGE{to rake, to sweep roughly} \tbPD{gyaarɛɛ, gyaara, gyaara, gyaareba, gyaaraa, gyaareba } 
\tbLX{gyaaroo} \tbHM{} \tbPH{dzáárʊ́ʊ́} \tbPS{n.~} \tbGE{a type of wild mouse} \tbOP{gyaara} \tbTP{gyaarɛɛ} 
\tbLX{gya᷉ayo} \tbHM{} \tbPH{dza᷉ayo} \tbPS{n.~} \tbGE{thanks} \tbOP{gya᷉ayori} \tbTP{gya᷉ayoree} 
\tbLX{Gyabuni} \tbHM{} \tbPH{dzàbùnì} \tbPS{n.~} \tbGE{a type of idol with magical powers among the Dagaaba} \tbOP{gyabunni/gyabumo} \tbTP{gyabunee} 
\tbLX{gyabuno} \tbHM{} \tbPH{dzàbùnò} \tbPS{n.~} \tbGE{fetish} \tbOP{gyabunni/gyabumo} \tbTP{gyabunnee/gyabumee} \tbVA{gyabuni}
\tbLX{gyaga} \tbHM{} \tbPH{dzàgà} \tbPS{adj.~} \tbGE{useless} 
\tbLX{gyage} \tbHM{} \tbPH{dzàgɪ̀} \tbPS{v.~} \tbSN \tbGE{to face a huge task} \tbSN \tbGE{jack} \tbPD{gyagɛɛ, gyagera, gyagera, gyagereba, gyageraa, gyagereba } 
\tbLX{gyagele} \tbHM{} \tbPH{dzàglɪ̀} \tbPS{v.~} \tbGE{to hang on to, to support lightly} \tbPD{gyaglɛɛ, gyagela, gyagela, gyageleba, gyagelaa, gyageleba } 
\tbLX{gyambiri} \tbHM{} \tbPH{ʣàbírì} \tbPS{n.~} \tbGE{clitoris} \tbVA{zibiri}
\tbLX{gyammo} \tbHM{} \tbPH{dzàmmʊ́} \tbPS{n.~} \tbGE{clitoris} \tbOP{gyama} \tbTP{gyamɛɛ/gyannɛɛ} \tbVA{gyàmbírì,gyambie,gyambiiree}
\tbLX{gyampegere} \tbHM{} \tbPH{dzàmpɪ́grɪ́} \tbPS{n.~} \tbGE{comb (of a rooster)} \tbOP{gyampegɛ} \tbTP{gyampegɛɛ/gyampegerɛɛ} \tbVA{gbammo}
\tbLX{gyangyagera} \tbHM{} \tbPH{dzàndzágrá} \tbPS{n.~} \tbGE{one who behaves abnormally; insane person} \tbSG{gyangyagerebɛɛ} \tbTP{gyangyagereba} 
\tbLX{gyangyagera} \tbHM{} \tbPH{dzàndzágrá} \tbPS{n.~} \tbGE{fried corn or bean flour mixed with vegetables} \tbTP{gyangyagrɛɛ} 
\tbLX{gyaŋe} \tbHM{} \tbPH{dzáŋ} \tbPS{n.~} \tbGE{see gyɛŋ} 
\tbLX{gyaŋgbaa} \tbHM{} \tbPH{dzáŋgbàá} \tbPS{n.~} \tbGE{something that is acidic; usually harmful} \tbOP{gyaŋgbaare} \tbTP{gyaŋgbaarɛɛ} 
\tbLX{gyaŋgbeɛ} \tbHM{} \tbPH{dzáŋgbɪ́ɛ́} \tbPS{n.~} \tbGE{small child} \tbOP{gyaŋgbeere} \tbTP{gyaŋgbeerɛɛ} 
\tbLX{gyaŋgboŋi} \tbHM{} \tbPH{dzàŋgbòŋ} \tbPS{n.~} \tbGE{house mouse} \tbOP{gyaŋgbonni} \tbTP{gyaŋgbonnee} \tbVA{gyoŋgboŋi}
\tbLX{gyaŋgboro} \tbHM{} \tbPH{dzàŋgbòrò} \tbPS{n.~} \tbGE{fried dough made from the leaves of a small plant} \tbTP{gyaŋgboree} 
\tbLX{gyaŋgboro} \tbHM{} \tbPH{dzáŋgbóró} \tbPS{n.~} \tbGE{a type of small plant with edible leaves} \tbSG{gyaŋgboruu} \tbTP{gyaŋgboree} 
\tbLX{gyaŋgbɔgɔ} \tbHM{} \tbPH{dzǎŋgbɔ̀gɔ̀} \tbPS{n.~} \tbGE{a type of monitor} \tbOP{gyaŋgbɔgre} \tbTP{gyaŋgbɔgrɛɛ} 
\tbLX{gyarema} \tbHM{} \tbPH{dzàrɪ̀má} \tbPS{n.~} \tbGE{a type of yam farm where sticks with branches are used to support the vine of the yam} \tbTP{gyaremɛɛ} 
\tbLX{gyarima} \tbHM{} \tbPH{dzárɪ́má} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbTP{gyaremɛɛ} 
\tbLX{gyee} \tbHM{} \tbPH{dzɪ̀ɪ̀} \tbPS{v.~} \tbGE{short form of 'gyeere' to move away from; to change environment} \tbPD{gyeeɛ, gyeerɛ, gyeerɛ, gyeerebɛ, -, - } 
\tbLX{gyeere} \tbHM{} \tbPH{dzɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbGE{to circle around from front to back of an object} \tbPD{gyeerɛɛ, gyeerɛ, gyeerɛ, gyeerebɛ, gyeeraa, gyeerebɛ } 
\tbLX{gyeɛne} \tbHM{} \tbPH{dzɪ̀ɛ̀nɪ̀} \tbPS{v.~} \tbGE{to lurk; wishing to do something} \tbXV{A noɔ gyeɛnɛ la a kyi.} \tbXE{The fowl is lurking around the millet.} \tbPD{gyeɛnɛɛ, gyeɛnɛ, gyeɛnɛ, gyeɛnemɛ, gyeɛnaa, gyeɛnemɛ } 
\tbLX{gyeɛne} \tbHM{} \tbPH{dzɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{see gyɛngyeɛne (protrusion)} \tbOP{gyeɛmɛ} \tbTP{gyeɛmɛɛ} 
\tbLX{gyeɛraa} \tbHM{} \tbPH{dzɪ̀ɛ̀ràà} \tbPS{n.~} \tbGE{plume, long feather (of a rooster)} \tbOP{gyeɛre} \tbTP{gyeɛrɛɛ} 
\tbLX{gyefa} \tbHM{} \tbPH{dzɪ́fà} \tbPS{n.~} \tbGE{pocket} \tbOP{gyefare} \tbTP{gyefarɛɛ} 
\tbLX{gyefa} \tbHM{} \tbPH{dzɪ́fà} \tbPS{n.~} \tbGE{unclean thing} \tbOP{gyefare} \tbTP{gyefarɛɛ} 
\tbLX{gyege} \tbHM{} \tbPH{dzɪ̀gɪ̀} \tbPS{v.~} \tbGE{to be fully grown, to be strong, to be powerful} \tbPD{gyegɛɛ, gyegerɛ, gyegerɛ, gyegerebɛ, gyegeraa, gyegerebɛ } 
\tbLX{gyegele} \tbHM{} \tbPH{dzɪ́glɪ́} \tbPS{n.~} \tbGE{a spiritually powerful person} \tbOP{gyegelle} \tbTP{gyegelɛɛ} 
\tbLX{gyegere} \tbHM{} \tbPH{dzɪ̀grɪ̀} \tbPS{v.~} \tbGE{to display one's supernatural power, to challenge} \tbPD{gyegerɛɛ, gyegerɛ, gyegerɛ, gyegerebɛ, gyegeraa, gyegerebɛ } 
\tbLX{gyegiri} \tbHM{} \tbPH{dzégrí} \tbPS{v.~} \tbGE{to mix by placing things in between others} \tbPD{gyegiree, gyegire, gyegire, gyegiribe, gyegiraa, gyegiribe } 
\tbLX{gyegiri} \tbHM{} \tbPH{dzégrí} \tbPS{v.~} \tbGE{to counter someone's speech; to challenge} \tbPD{gyegiree, gyegire, gyegire, gyegiribe, gyegiraa, gyegiribe } 
\tbLX{gyelaa} \tbHM{} \tbPH{dzɪ̀làá} \tbPS{n.~} \tbGE{carnivorous animal} \tbOP{gyelaare} \tbTP{gyelaarɛɛ} 
\tbLX{gyell} \tbHM{} \tbPH{dzɪ̀ll} \tbPS{adj.~} \tbGE{protuding e.g. arrow stuck in an animal} 
\tbLX{gyemɛ} \tbHM{} \tbPH{dzɪ̀mɛ̀} \tbPS{n.~} \tbSN \tbGE{convulsion, disease of children} \tbSN \tbGE{short form of "da᷉a᷉gyemɛ"} \tbTP{gyemɛɛ} 
\tbLX{gyena} \tbHM{} \tbPH{dzɪ́nà} \tbPS{n.~} \tbGE{fairy} \tbOP{gyenne} \tbTP{gyennɛɛ} 
\tbLX{gyenne} \tbHM{} \tbPH{dzɪ̀nnɪ̀} \tbPS{v.~} \tbGE{to test} \tbPD{gyennɛɛ, gyennɛ, gyennɛ, gyenemɛ, gyennaa, gyenemɛ } 
\tbLX{gyenne} \tbHM{} \tbPH{dzɪ̀nnɪ̀} \tbPS{v.~} \tbGE{to waylay, to plan evil against} \tbPD{gyennɛɛ, gyennɛ, gyennɛ, gyenemɛ, gyennaa, gyenemɛ } 
\tbLX{gyennoo} \tbHM{} \tbPH{dzɪ́nnʊ́ʊ́} \tbPS{n.~} \tbGE{test, examination} \tbTP{gyennɛɛ} 
\tbLX{gyeraa} \tbHM{} \tbPH{dzɪ̀ráà} \tbPS{n.~} \tbGE{woven grass bracelet worn on the wrist} \tbOP{ gyerre} \tbTP{gyerrɛɛ} 
\tbLX{gyeraa} \tbHM{} \tbPH{dzɪ́ráá} \tbPS{adj.~} \tbGE{short (distance)} 
\tbLX{gyere} \tbHM{} \tbPH{dzɪ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to cross between (two objects)} \tbSN \tbGE{to stumble} \tbPD{gyerɛɛ, gyerɛ, gyerɛ, gyerebɛ, gyeraa, gyerebɛ } 
\tbLX{gyere} \tbHM{} \tbPH{dzɪ́rɪ̀} \tbPS{v.~} \tbGE{to subside (of rain)} \tbPD{gyerɛɛ, gyerɛ, -, -, -, - } \tbVA{yoɔ}
\tbLX{gyere} \tbHM{} \tbPH{dzɪ̀rɪ̀} \tbPS{v.~} \tbGE{to nock; to place across e.g. arrow on the bow} \tbPD{gyerɛɛ, gyerɛ, gyerɛ, gyerebɛ, gyeraa, gyerebɛ } 
\tbLX{gyere} \tbHM{} \tbPH{dzɪ̀rɪ̀} \tbPS{v.~} \tbGE{to pierce through to the other side} \tbPD{gyerɛɛ, gyerɛ, gyerɛ, gyerebɛ, gyeraa, gyerebɛ } 
\tbLX{gyere} \tbHM{} \tbPH{dzɪ́rɪ́} \tbPS{n.~} \tbGE{portion unattended to (part of farm)} \tbOP{gyerre} \tbTP{gyerɛɛ/gyerrɛɛ} 
\tbLX{gyeremɛ} \tbHM{} \tbPH{dzɪ́rɪ́mɛ́} \tbPS{n.~} \tbSN \tbGE{respect, honour, homage} \tbSN \tbGE{fifty fifty (gambling)} \tbTP{gyeremɛɛ} 
\tbLX{gyeremiri} \tbHM{} \tbPH{dzɪ́rɪ́mírí} \tbPS{n.~} \tbGE{rope made from a type of grass (kantɛrɛɛ)} \tbOP{gyeremie} \tbTP{gyeremiiree/gyeremiree} 
\tbLX{gyeremuri} \tbHM{} \tbPH{dzɪ̀rìmúrí} \tbPS{n.~} \tbGE{a type of river grass stalk} \tbOP{gyeremue} \tbTP{gyeremuree} 
\tbLX{Gyerɛbaa} \tbHM{} \tbPH{dzɪ̀rɛ̀bàá} \tbPS{n.~} \tbGE{Jirapa (the main town in Central Dagao)} 
\tbLX{gyerɛgyerɛ} \tbHM{} \tbPH{dzɪ́rɛ́dzɪ̀rɛ̀} \tbPS{n.~} \tbGE{a type of skin disease} 
\tbLX{gyerre} \tbHM{} \tbPH{dzɪ̀rrɪ̀} \tbPS{v.~} \tbGE{used with ...taa; to crisscross; lying across one another} \tbXV{A bibiiri le gyerre la taa.} \tbXE{The children fell crisscrossing each other.} \tbXV{A teere le gyerre la taa.} \tbXE{The trees fell across one another.} \tbPD{gyerrɛɛ, gyerrɛ, gyerrɛ, gyerbɛ, gyerraa, - } 
\tbLX{gyɛ} \tbHM{} \tbPH{dzɛ́} \tbPS{v.~} \tbSN \tbGE{to reject, to refuse to accept (in protest or annoyance)} \tbSN \tbGE{to abandon} \tbPD{gyɛɛ, gyeɛrɛ, gyeɛrɛ, gyeɛrebɛ, gyeɛraa, gyeɛrebɛ } 
\tbLX{Gyɛɛmaan} \tbHM{} \tbPH{dzɛ́ɛ́máán} \tbPS{n.~} \tbGE{Germany} \tbTP{gyɛɛmanɛɛ} 
\tbLX{Gyɛle} \tbHM{} \tbPH{dzɛ̀lɨ̀} \tbPS{n.~} \tbSN \tbGE{a stream in Kaane} \tbSN \tbGE{a native named after this stream} \tbOP{gyɛlɛ} \tbTP{gyɛlɛɛ} 
\tbLX{gyɛle} \tbHM{} \tbPH{dzɛ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to rummage; to look through searching} \tbSN \tbGE{to analyse} \tbPD{gyɛlɛɛ, gyɛlɛ, gyɛlɛ, gyɛlebɛ, gyɛlaa, gyɛlebɛ } 
\tbLX{gyɛle} \tbHM{} \tbPH{dzɛ́lɪ̀} \tbPS{n.~} \tbGE{egg} \tbSG{gyɛlee} \tbOP{gyɛlɛ} \tbTP{gyɛllɛɛ} 
\tbLX{gyɛle} \tbHM{} \tbPH{dzɛ̀lɪ́} \tbPS{n.~} \tbGE{thought (used with bare---) not giving thought; absent minded} \tbXV{A porisi bare la gyɛle ka a nannyige zo.} \tbXE{The police did not pay attention, so the thief escaped.} 
\tbLX{gyɛlemɛ} \tbHM{} \tbPH{dzɛ̀lɪ̀mɛ́} \tbPS{n.~} \tbGE{smallpox} \tbTP{gyɛlmɛɛ} 
\tbLX{gyɛleŋmɛne} \tbHM{} \tbPH{dzɛ́lɪ́ŋmɛ́nɪ́} \tbPS{n.~} \tbSN \tbGE{snail} \tbSN \tbGE{shellfish} \tbOP{gyɛleŋmɛmɛ} \tbTP{gyɛleŋmɛmɛɛ} 
\tbLX{gyɛlpɛge} \tbHM{} \tbPH{dzɛ́lpɛ́gɪ̀} \tbPS{n.~} \tbGE{eggshell} \tbOP{gyɛlpɛgre/gyɛlpɛgɛ} \tbTP{gyɛlpɛgrɛɛ} 
\tbLX{gyɛlzizeɛ} \tbHM{} \tbPH{dzɛ̀lzìzɪ̀ɛ́} \tbPS{n.~} \tbGE{yolk} \tbOP{gyɛlzeere} \tbTP{gyɛlzeerɛɛ} 
\tbLX{gyɛmɛgyɛmɛ} \tbHM{} \tbPH{dzɛ́mɛ́dzɛ́mɛ́} \tbPS{adj.~} \tbGE{slowly and unnoticed, jokingly} \tbXV{A kɔlaa e la gyɛmɛgyɛmɛlɛ kyɛ ɔɔ a nɛne zaa.} \tbXE{The cat ate all the meat without anyone noticing it.} 
\tbLX{gyɛndaa} \tbHM{} \tbPH{dzɛ̀ndàá} \tbPS{n.~} \tbGE{spindle} \tbOP{gyɛndaare} \tbTP{gyɛndarɛɛ} 
\tbLX{gyɛnee} \tbHM{} \tbPH{dzɛ̀nɪ́ɪ́} \tbPS{n.~} \tbGE{groundnut, peanut (Southern Dagaare dialect)} \tbTP{gyɛnnɛɛ} 
\tbLX{gyɛngyeɛne} \tbHM{} \tbPH{dzɛ̀ndzɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{protrusion, part sticking out} \tbOP{gyɛngyeɛmɛ} \tbTP{gyɛngyeɛmɛɛ} 
\tbLX{gyɛngyɛŋ} \tbHM{} \tbPH{dzɛ́ndzɛ́ŋ} \tbPS{n.~} \tbGE{circle arrangement for advancement. (war)} 
\tbLX{gyɛngyɛre} \tbHM{} \tbPH{dzɛ̀ngyɛ̀rɪ́} \tbPS{n.~} \tbGE{a climbing plant} \tbOP{gyɛngyɛrre} \tbTP{gyɛngyɛrɛɛ} 
\tbLX{gyɛnne} \tbHM{} \tbPH{dzɛ̀nnɪ́} \tbPS{v.~} \tbGE{please, beg} \tbXV{Gyɛnne, ɔŋe kõɔ ko ma ka N nyu.} \tbXE{Please, give me some water to drink.or I beg you, give me some water to drink.} \tbPD{-, -, -, -, -, - } 
\tbLX{gyɛnterɛ} \tbHM{} \tbPH{dzɛ́ntɪ́rɛ̀} \tbPS{n.~} \tbSN \tbGE{prostitute} \tbSN \tbGE{not sticking to one thing, jumping from one thing to another} \tbOP{gyɛnterre} \tbTP{gyɛnterrɛɛ} 
\tbLX{gyɛŋe} \tbHM{} \tbPH{dzɛ̌ŋ} \tbPS{adj.~} \tbGE{extensive; deep} \tbXV{A naŋkpaana kpɛ la mɔ-gyɛŋe.} \tbXE{The hunter has gone into the deep bush.} \tbOP{gyɛnne} \tbTP{gyɛnnɛɛ} \tbVA{gyaŋ}
\tbLX{gyɛŋe} \tbHM{} \tbPH{dzɛ́ŋ} \tbPS{v.~} \tbSN \tbGE{to move unsteadily} \tbSN \tbGE{to move aside} \tbPD{gyɛŋɛɛ, gyɛŋenɛ/gyɛnnɛ, gyɛŋenɛ/gyɛnnɛ, gyɛŋenaa/gyɛnnaa, gyɛŋenemɛ/gyɛnemɛ, - } 
\tbLX{gyɛŋe} \tbHM{} \tbPH{dzɛ̀ŋ} \tbPS{n.~} \tbGE{thought, mind} \tbOP{gyɛnne} \tbTP{gyɛnnɛɛ} \tbVA{gyaŋe}
\tbLX{gyɛŋgɛlɛ} \tbHM{} \tbPH{dzɛ́ŋgɛ́lɛ́} \tbPS{n.~} \tbGE{measles} \tbTP{gyɛŋgɛlɛɛ} 
\tbLX{gyɛŋgyɛŋ} \tbHM{} \tbPH{dzɛ̌ŋdzɛ̌ŋ} \tbPS{n.~} \tbGE{mind, thought} 
\tbLX{gyɛre} \tbHM{} \tbPH{dzɛ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to be mixed up (different types)} \tbSN \tbGE{to be confused, to be disturbed (mad)} \tbSN \tbGE{to not be in good health} \tbPD{gyɛrɛɛ, gyɛrɛ, gyɛrɛ, gyɛrbɛ, gyɛraa, gyɛrbɛ } 
\tbLX{gyɛre} \tbHM{} \tbPH{dzɛ̀rɪ̀} \tbPS{v.~} \tbGE{to stand firm, to not fall e.g. after a stumble} \tbPD{gyɛrɛɛ, gyɛrɛ, gyɛrɛ, gyɛrebɛ, gyɛraa, gyɛrebɛ } 
\tbLX{gyieli} \tbHM{} \tbPH{dzìèlì} \tbPS{v.~} \tbSN \tbGE{to note, to examine} \tbSN \tbGE{to count} \tbPD{gyielee, gyiele, gyiele, gyielbe, gyielaa, - } 
\tbLX{gyieri} \tbHM{} \tbPH{dzìèrì} \tbPS{v.~} \tbGE{to grind into coarse powder} \tbPD{gyieree, gyiere, gyiere, gyieribe, gyieraa, - } 
\tbLX{gyiili} \tbHM{} \tbPH{dzííĺí} \tbPS{n.~} \tbSN \tbGE{tendon} \tbSN \tbGE{artery} \tbSN \tbGE{nerve} \tbSG{gyiiluu} \tbOP{gyiile} \tbTP{gyiilee} 
\tbLX{gyikoli} \tbHM{} \tbPH{dzìkólí} \tbPS{n.~} \tbGE{part of body} \tbOP{gyikolo} \tbTP{gyikolee/gyikollee} \tbVA{gyaaŋkoli}
\tbLX{gyildaa} \tbHM{} \tbPH{dzìldáá} \tbPS{n.~} \tbGE{less soudning one of the pair of xylophones} \tbOP{gyidaare} \tbTP{gyildaarɛɛ} 
\tbLX{gyildoɔle} \tbHM{} \tbPH{dzìldʊ́ɔ́lɪ́} \tbPS{n.~} \tbGE{xylophone mallet} \tbSG{gyildoɔloo} \tbOP{gyildoɔlɔ} \tbTP{gyildoɔlɛɛ} 
\tbLX{gyili} \tbHM{} \tbPH{dzìlí} \tbPS{n.~} \tbGE{xylophone} \tbOP{gyile} \tbTP{gyilee/gyillee} 
\tbLX{gyili} \tbHM{} \tbPH{dzílí} \tbPS{v.~} \tbGE{to surround, to go behind} \tbXV{Teere are gyili la a bulee.} \tbXE{The well is surrounded by trees.} \tbPD{gyilee, gyile, gyile, gyilibe, gyilaa, gyilibe } 
\tbLX{gyilkoɔre} \tbHM{} \tbPH{dzìlkʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{gourd used in making the xylophone} \tbOP{gyilkɔɛ} \tbTP{gyilkɔrɛɛ} 
\tbLX{gyilnyaa} \tbHM{} \tbPH{dzìlɲáà} \tbPS{n.~} \tbGE{the base chord area of the xylophone} \tbOP{gyilnyaane} \tbTP{gyilnyaanɛɛ} 
\tbLX{gyilnyaŋaa} \tbHM{} \tbPH{dzìlɲáŋáá} \tbPS{n.~} \tbGE{the more sounding between a pair of xylophones} \tbOP{gyilnyanne} \tbTP{gyilnyannɛɛ} 
\tbLX{gyiri} \tbHM{} \tbPH{dzírí} \tbPS{n.~} \tbGE{cordlike talisman worn on the waist, believed to have supernatural power} \tbOP{gyirri} \tbTP{gyiree} 
\tbLX{gyiri} \tbHM{} \tbPH{dzírí} \tbPS{v.~} \tbGE{to bring about a change that makes life difficult} \tbXV{A zie gyiri ka donɛɛ kpeɛbo e kpeɛŋaa.} \tbXE{Society has declined and life has become difficult.} \tbPD{gyiree, gyire, gyire, gyirbe, gyiraa, - } 
\tbLX{gyirimuri} \tbHM{} \tbPH{dzìrìmúrí} \tbPS{n.~} \tbGE{sulphur} \tbOP{gyirimue} \tbTP{gyirimuree} 
\tbLX{gyoboraa} \tbHM{} \tbPH{dzʊ̀bʊ̀ràà} \tbPS{n.~} \tbGE{carnivore, e.g. a leopard} \tbOP{gyoborre.} \tbTP{gyoborɛɛ/gyoborrɛɛ} 
\tbLX{gyoge} \tbHM{} \tbPH{dzʊ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to pull mouth or lips in front to show anger. (typically used with noɔre)} \tbSN \tbGE{to weed badly} \tbPD{gyogrɛɛ, gyogra, gyogra, gyogreba, gyograa, - } \tbVA{gyogere}
\tbLX{gyogere} \tbHM{} \tbPH{dzʊ̀grɪ̀} \tbPS{v.~} \tbGE{to clear growth badly (barbing or hoeing)} \tbPD{gyogerɛɛ, gyogera, gyogera, gyogereba, gyogeraa, gyogereba } \tbVA{gyoge}
\tbLX{gyogili} \tbHM{} \tbPH{dzòglì} \tbPS{v.~} \tbGE{to hold on to lightly and gently} \tbPD{gyogilee, gyogilo, gyogilo, gyogilibo, gyogilaa, gyogilibo } 
\tbLX{gyogo} \tbHM{} \tbPH{dzógó} \tbPS{n.~} \tbGE{behaviour, character} \tbOP{gyogiri} \tbTP{gyogiree} 
\tbLX{gyoll} \tbHM{} \tbPH{dzʊ̀ll} \tbPS{adv.~} \tbGE{appearing alone} 
\tbLX{gyolom} \tbHM{} \tbPH{dzólóm} \tbPS{adv.~} \tbGE{quietly} \tbVA{gyoloŋ, gyolomm}
\tbLX{gyolomm} \tbHM{} \tbPH{dzolomm} \tbPS{adv.~} \tbGE{quietly and slowly (on tiptoe)} \tbVA{gyolom}
\tbLX{gyongyolingyo} \tbHM{} \tbPH{dzǒngyólǐndzó} \tbPS{adv.~} \tbGE{squat squarely, with knees fully bent} 
\tbLX{gyongyoŋ} \tbHM{} \tbPH{dzòngyóŋ} \tbPS{v.~} \tbGE{to move on tiptoe} \tbPD{gyongyoŋee, gyongyono, gyongyono, gyongyonimo, gyongyonaa, - } \tbVA{gyɔngyɔŋ}
\tbLX{gyoŋene} \tbHM{} \tbPH{dzʊ̀ŋɪ̀nɪ̀} \tbPS{v.~} \tbGE{to make unlively, to make dull} \tbPD{gyoŋenɛɛ, gyoŋena, gyoŋena, gyoŋenema, gyoŋenaa, gyoŋenema } \tbVA{gyoone}
\tbLX{gyoŋgoroŋ} \tbHM{} \tbPH{dzóŋgóróŋ} \tbPS{adj.~} \tbGE{a little higher or taller than} 
\tbLX{gyoŋgboŋi} \tbHM{} \tbPH{dzòŋgbǒŋ} \tbPS{n.~} \tbGE{house mouse} \tbOP{gyoŋgbonni} \tbTP{gyoŋgbonee} \tbVA{gyaŋgboŋi}
\tbLX{gyoone} \tbHM{} \tbPH{dzʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbGE{to become inactive} \tbPD{gyoonɛɛ, gyoona, gyoona, gyoonema, gyoonaa, - } \tbVA{gyoŋene}
\tbLX{gyore} \tbHM{} \tbPH{dzʊ́rɪ́} \tbPS{v.~} \tbGE{to work at or make badly, to spoil e.g. haircut} \tbPD{gyorɛɛ, gyora, gyora, gyoreba, gyoraa, gyoreba } 
\tbLX{gyɔŋgyɔŋ} \tbHM{} \tbPH{dzɔ̀ŋdzɔ́ŋ} \tbPS{n.~} \tbGE{to move haphazardly on one's toes} \tbVA{gyongyoŋ}
\tbLX{gyɔ᷉ɔ} \tbHM{} \tbPH{dzɔ̀᷉ɔ̀} \tbPS{adv.~} \tbGE{tall and lanky} \tbVA{gyɔgyɔ᷉ɔ}
\tbLX{gyɔreba} \tbHM{} \tbPH{dzɔ́rɪ́bá} \tbPS{n.~} \tbGE{long, loose dress worn by women; tunic} \tbOP{gyɔrebare} \tbTP{gyɔrebarɛɛ} 
\tbLX{gyuuni} \tbHM{} \tbPH{dzùùnì} \tbPS{n.~} \tbGE{vulture} \tbOP{gyuumo} \tbTP{gyuunee} 
\tbLX{ha} \tbHM{} \tbPH{há} \tbPS{v.~} \tbGE{to hire} \tbPD{haɛ, haara, haara, hareba, haaraa, - } 
\tbLX{haa} \tbHM{} \tbPH{háà} \tbPS{v.~} \tbGE{to open by stretching out two ends e.g. lips} \tbPD{haaɛ, haara, haara, haareba, haaraa, haareba } 
\tbLX{haa (cd)} \tbHM{} \tbPH{hàá} \tbPS{adv.~} \tbGE{all} \tbVA{zaa}
\tbLX{haale} \tbHM{} \tbPH{háalì́} \tbPS{adv.~} \tbGE{even, extremely} \tbVA{haali}
\tbLX{haanaa} \tbHM{} \tbPH{háánáá} \tbPS{n.~} \tbGE{blackberry} \tbOP{haane} \tbTP{haanɛɛ} 
\tbLX{haare} \tbHM{} \tbPH{háárɪ̀} \tbPS{v.~} \tbGE{to shout at to frighten or drive away} \tbPD{haarɛɛ, haara, haara, haareba, haaraa, haareba } 
\tbLX{hage} \tbHM{} \tbPH{hhágɪ́} \tbPS{v.~} \tbSN \tbGE{tight} \tbXV{A dɔɔ hage la o pɔge nyuu taa lɛ.} \tbXE{The man held tight to his wife’s neck.} \tbSN \tbGE{tighten} \tbXV{Ka fooŋ wa nyɔge foŋ hage taa lɛ.} \tbXE{When you get hold of it tighten your grip.} \tbPD{hagɛɛ, hagera, hagera, hagereba, hageraa, hagereba } 
\tbLX{ha-haa} \tbHM{} \tbPH{hà-háá} \tbPS{adv.~} \tbGE{wide open} 
\tbLX{hai} \tbHM{} \tbPH{hháí} \tbPS{interjection.~} \tbGE{so?} 
\tbLX{hale} \tbHM{} \tbPH{hálɪ̀} \tbPS{v.~} \tbGE{to smell strongly and unpleasantly, to reek} \tbPD{halɛɛ, hala, halla, haleba, halaa, haleba } 
\tbLX{hale} \tbHM{} \tbPH{hálɪ̀} \tbPS{v.~} \tbGE{to speak in anger} \tbPD{halɛɛ, hala, halla, haleba, halaa, haleba } 
\tbLX{halhale} \tbHM{} \tbPH{hàlhálɪ́} \tbPS{v.~} \tbGE{to strive with, to struggle} \tbPD{halhalɛɛ, halhala, halhala, halhaleba, halhalaa, - } 
\tbLX{haligba} \tbHM{} \tbPH{hálìgbà} \tbPS{conjunction.~} \tbGE{even} 
\tbLX{hama} \tbHM{} \tbPH{hámà} \tbPS{n.~} \tbGE{hammer} \tbOP{hamare} \tbTP{hamarɛɛ} 
\tbLX{hama} \tbHM{} \tbPH{hàmá} \tbPS{adv.~} \tbGE{plenty} 
\tbLX{hamama} \tbHM{} \tbPH{hàmàmà} \tbPS{adj.~} \tbGE{resounding (singing)} 
\tbLX{hamamma} \tbHM{} \tbPH{hàmàmmà} \tbPS{adj.~} \tbGE{loudly and continuously} 
\tbLX{hamm} \tbHM{} \tbPH{hàmm} \tbPS{adv.~} \tbGE{large} 
\tbLX{hankyere} \tbHM{} \tbPH{hánʧɪ́rɪ́} \tbPS{n.~} \tbGE{handkerchief} \tbOP{hankyeɛ} \tbTP{hankyerɛɛ} 
\tbLX{haŋe} \tbHM{} \tbPH{hháŋ} \tbPS{v.~} \tbSN \tbGE{to suffer} \tbSN \tbGE{to perspire, sweat (nd)} \tbPD{haŋɛɛ, hana, hana, hanema, hanaa, hanema } 
\tbLX{hap (nd)} \tbHM{} \tbPH{hháp} \tbPS{v.~} \tbGE{to clutch tightly} \tbPD{habɛɛ, hapra, hapra, hapreba, hapraa, hapreba } 
\tbLX{harahara} \tbHM{} \tbPH{hháráhhárá} \tbPS{adv.~} \tbGE{rough} 
\tbLX{hare} \tbHM{} \tbPH{hàrɪ̀} \tbPS{v.~} \tbGE{of the stomach (overwhelmed by hunger)} \tbPD{hanrɛɛ,, hara, hara, hareba, haraa, hareba } 
\tbLX{helli} \tbHM{} \tbPH{hhéllí} \tbPS{v.~} \tbGE{sharp shout e.g. during hearty laughter; the sharp cry of a dog when hit} \tbPD{hellee, helle, helle, helibe, hellaa, - } 
\tbLX{heŋi} \tbHM{} \tbPH{hhéŋ} \tbPS{v.~} \tbGE{to draw in one's stomach} \tbPD{heŋee, heŋine/henne, heŋine/henne, henime, hennaa, henime } 
\tbLX{heremm} \tbHM{} \tbPH{hérèmm} \tbPS{adj.~} \tbGE{very tight} 
\tbLX{hɛɛ} \tbHM{} \tbPH{hhɛ́ɛ̀} \tbPS{interjection.~} \tbGE{caution, be careful} 
\tbLX{hɛɛ} \tbHM{} \tbPH{hɛ́ɛ́} \tbPS{adv.~} \tbGE{overtaken completely e.g. sleep} 
\tbLX{hɛge} \tbHM{} \tbPH{hɛ̀gɪ̀} \tbPS{v.~} \tbGE{to increase e.g. fire} \tbPD{hɛgɛɛ, hɛgerɛ, hɛgerɛ, hɛgerebɛ, hɛgeraa, hɛgerebɛ } 
\tbLX{hɛgele} \tbHM{} \tbPH{hhɛ́glɪ́} \tbPS{v.~} \tbGE{to cramp} \tbPD{hɛgelɛɛ, hɛgelɛ, hɛgelɛ, hɛgelebɛ, hɛgelaa, hɛgelebɛ } 
\tbLX{hɛgelɛgɛ} \tbHM{} \tbPH{hhɛ́gɪ́lɛ́gɛ́} \tbPS{adv.~} \tbGE{tired and helpless} \tbVA{sɛgelɛgɛ}
\tbLX{hɛgere} \tbHM{} \tbPH{hhɛ́grɪ̀} \tbPS{v.~} \tbGE{to pant e.g. dog} \tbPD{hɛgerɛɛ, hɛgerɛ, hɛgerɛ, hɛgerebɛ, hɛgeraa, hɛgerebɛ } 
\tbLX{hɛgɛkɛ} \tbHM{} \tbPH{hɛ̀gɛ̀kɛ̀} \tbPS{adv.~} \tbGE{large and red e.g. bonfire} 
\tbLX{hɛmɛmmɛ} \tbHM{} \tbPH{hɛ̀mɛ̀mmɛ̀} \tbPS{adj.~} \tbGE{very sweet} 
\tbLX{hɛrɛ} \tbHM{} \tbPH{hɛ̀rɛ̀} \tbPS{adv.~} \tbGE{manner of taking a big sip of fluid} 
\tbLX{hieri} \tbHM{} \tbPH{hhíérì} \tbPS{v.~} \tbGE{to yawn} \tbPD{hieree, hiere, hiere, hieribe, hieraa, hieribe } 
\tbLX{hii} \tbHM{} \tbPH{hì᷉ì} \tbPS{interj.~} \tbGE{expression of sorrow (sobbing)} 
\tbLX{hiini} \tbHM{} \tbPH{hììnì} \tbPS{v.~} \tbGE{to cry earnestly, to wail, to sob, to moan} \tbPD{hiinee, hiine, hiine, hiinime, hiinaa, hiinime } 
\tbLX{hiri} \tbHM{} \tbPH{hìrì} \tbPS{v.~} \tbGE{to get to the climax} \tbPD{hiree, hire, -, -, -, - } 
\tbLX{hiri} \tbHM{} \tbPH{hìrì} \tbPS{v.~} \tbGE{to become hot} \tbPD{hiree, hire, -, -, -, - } 
\tbLX{hiri} \tbHM{} \tbPH{hìrì} \tbPS{v.~} \tbGE{to become serious} \tbPD{hiree, hire, -, -, -, - } 
\tbLX{hạạli} \tbHM{} \tbPH{hạ́ạ́lì} \tbPS{adv.~} \tbGE{ greatly, very much} 
\tbLX{hạạli} \tbHM{} \tbPH{hạ́ạ́lì} \tbPS{conj.~} \tbGE{even if} \tbXV{Haali ka a bieŋ wa kyɛ te koŋ o a gane.} \tbXE{Even if the child comes we will not give him/her the book.} 
\tbLX{hogoko} \tbHM{} \tbPH{hògòkò} \tbPS{adv.~} \tbGE{intensely red} 
\tbLX{holi} \tbHM{} \tbPH{hólí} \tbPS{n.~} \tbGE{triangular blade} \tbOP{holo} \tbTP{hollee} 
\tbLX{hoo} \tbHM{} \tbPH{hòò} \tbPS{adv.~} \tbGE{open and endless} 
\tbLX{hoo} \tbHM{} \tbPH{hóó} \tbPS{interj.~} \tbGE{hooting} 
\tbLX{ho᷉o} \tbHM{} \tbPH{hʊ̀᷉ʊ̀} \tbPS{interj.~} \tbGE{expression of dispair, depression} 
\tbLX{hoone} \tbHM{} \tbPH{hʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbGE{to growl} \tbPD{hoonɛɛ, hoonɔ, hoonɔ, honema, hoonaa, honema } 
\tbLX{hɔɔpɛmpɛlɛŋ} \tbHM{} \tbPH{hɔ̀ɔ̀pɛ́mpɛ́lɛ́ŋ} \tbPS{n.~} \tbGE{spleen} \tbOP{hɔɔpɛmpɛlle} \tbTP{hɔɔpɛmpɛlɛɛ} \tbVA{zagepɛmpɛlɛŋ}
\tbLX{hunihuni} \tbHM{} \tbPH{hùnìhúní} \tbPS{v.~} \tbGE{to speak undertone} \tbPD{hunihunee, hunihuno, hunihuno, hunihunimo, -, - } \tbVA{vunnii}
\tbLX{hunihuni} \tbHM{} \tbPH{hùnìhúní} \tbPS{v.~} \tbGE{to gossip} \tbPD{hunihunee, hunihuno, hunihuno, hunihunimo, -, - } \tbVA{vunnii}
\tbLX{hunni} \tbHM{} \tbPH{hhúnnì} \tbPS{v.~} \tbGE{whisper into the ear, gossip} \tbPD{hunnee, hunno, hunno, hunimo, hunnaaa, - } 
\tbLX{huuni} \tbHM{} \tbPH{hhúúní} \tbPS{v.~} \tbSN \tbGE{to hum (dispirited)} \tbSN \tbGE{of a dog's expression of a familiar person; welcome} \tbPD{huunee, huuno, huuno, huunimo, huunaa, - } 
\tbLX{iini} \tbHM{} \tbPH{ììnì} \tbPS{v.~} \tbGE{to get charged up} \tbXV{Ba to la a dɔɔ ka o iini gbie kyɛ sɔrɔ yɛlɛ.} \tbXE{They insulted the man and he got charged up and started grumbling.} \tbPD{-, -, -, -, -, - } 
\tbLX{ingyi} \tbHM{} \tbPH{índzì} \tbPS{n.~} \tbGE{engine} \tbOP{ingyime} \tbTP{ingyinnee} 
\tbLX{inni} \tbHM{} \tbPH{ìnnì} \tbPS{v.~} \tbGE{to forget} \tbPD{innee, inne, inne, inime, innaa, - } \tbVA{yinni}
\tbLX{iree} \tbHM{} \tbPH{ìréè} \tbPS{n.~} \tbGE{duiker (small antelope)} \tbOP{irri} \tbTP{irree} 
\tbLX{iri} \tbHM{} \tbPH{ìrì} \tbPS{v.~} \tbGE{to rise, to stand up} \tbPD{iree, ire, ire, iribe, iraa, iribe } 
\tbLX{iri} \tbHM{} \tbPH{írì} \tbPS{v.~} \tbSN \tbGE{to choose, to pick} \tbSN \tbGE{to take out (from a container)} \tbSN \tbGE{to subtract, to take away} \tbSN \tbGE{to create} \tbPD{iree, ire, ire, iribe, iraa, iribe } 
\tbLX{iribare} \tbHM{} \tbPH{i̊́rì bàrɪ̀} \tbPS{n.~} \tbGE{subtraction} 
\tbLX{iri noɔre} \tbHM{} \tbPH{írí nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to answer, to respond (to a question)} \tbSN \tbGE{to explain} \tbSN \tbGE{to fulfil a promise} \tbPD{iree noɔre, ire noɔre, -, -, -, - } 
\tbLX{iriŋ} \tbHM{} \tbPH{íríŋ} \tbPS{n.~} \tbGE{kind, race, species} \tbTP{irree} 
\tbLX{iriŋ-iriŋ} \tbHM{} \tbPH{írìŋ-írìŋ} \tbPS{adv.~} \tbGE{of all kinds, types} \tbVA{iruŋ-iruŋ}
\tbLX{iri pɔge} \tbHM{} \tbPH{írì pɔ́gɪ̀} \tbPS{v.~} \tbGE{to elope with a woman} \tbPD{-, -, -, -, -, - } 
\tbLX{iri suuri} \tbHM{} \tbPH{ìrì súúrí} \tbPS{v.~} \tbGE{to annoy} \tbPD{iree suuri, ire suuri, -, -, -, - } 
\tbLX{iri yuori} \tbHM{} \tbPH{írí yúórì} \tbPS{v.~} \tbGE{to praise} \tbPD{iree yuori, ire yuori, -, -, -, - } 
\tbLX{irizeɛ} \tbHM{} \tbPH{ìrìzìɛ̀} \tbPS{n.~} \tbGE{a type of small but fast and clever antelope} \tbOP{irizeere} \tbTP{irizeerɛɛ} 
\tbLX{ka} \tbHM{} \tbPH{ká} \tbPS{rel. pron.~} \tbGE{that} \tbXV{O yeli ka ba gɛrɛ.} \tbXE{S/he said s/he would not go.} 
\tbLX{ka} \tbHM{} \tbPH{kà} \tbPS{conj.~} \tbSN \tbGE{ if} \tbXV{Ka saaŋ mi ba na bore la.} \tbXE{If it rains, they will sow their crops.} \tbSN \tbGE{and} \tbXV{O maale la diibu de wa ko te.} \tbXE{S/he cooked and brought it to us.} 
\tbLX{ka} \tbHM{} \tbPH{ká} \tbPS{v.~} \tbGE{to bail out (water)} \tbPD{kaɛ, kaara, kaara, kaareba, kaaraa, - } 
\tbLX{kaa} \tbHM{} \tbPH{káà} \tbPS{v.~} \tbGE{to curse} \tbPD{kaaɛ, kaara, kaara, kaareba, kaaraa, - } \tbVA{kaale}
\tbLX{kaa} \tbHM{} \tbPH{káà} \tbPS{v.~} \tbGE{to invoke} \tbPD{kaaɛ, kaara, kaara, kaareba, kaaraa, - } \tbVA{kaale}
\tbLX{kaa} \tbHM{} \tbPH{kàà} \tbPS{v.~} \tbSN \tbGE{to look, to watch} \tbSN \tbGE{to take care of} \tbPD{kaaɛ, kaara, kaara, kaareba, kaaraa, - } 
\tbLX{ka᷈a} \tbHM{} \tbPH{ká᷉á} \tbPS{n.~} \tbSN \tbGE{oil; butter} \tbSN \tbGE{fat} \tbSG{kaanoo} \tbTP{kaanɛɛ} 
\tbLX{ka᷈a} \tbHM{} \tbPH{ká᷉á} \tbPS{n.~} \tbGE{nectar e.g. in honey} \tbSG{kaanoo} \tbTP{kaanɛɛ} 
\tbLX{kaabanyɛ} \tbHM{} \tbPH{káábányɛ́} \tbPS{n.~} \tbGE{direspect} 
\tbLX{kaabo} \tbHM{} \tbPH{káábʊ̀} \tbPS{n.~} \tbGE{millet plant} \tbOP{kyi} \tbTP{kyiree} 
\tbLX{ka᷉aboli} \tbHM{} \tbPH{ká᷉ábòlí} \tbPS{n.~} \tbGE{pure liquid oil} \tbOP{ka᷉abolo} \tbTP{ka᷉abolee} 
\tbLX{ka᷈aburi} \tbHM{} \tbPH{ká᷉ábúrí} \tbPS{n.~} \tbGE{burnt residue from shea butter} \tbOP{ka᷈aburri} \tbTP{ka᷈aburee} 
\tbLX{ka᷉aburituo} \tbHM{} \tbPH{ká᷉ábùrìtúó} \tbPS{n.~} \tbGE{type of residue from the last stage of shea butter preparation} \tbOP{ka᷉aburituuri} \tbTP{ka᷉aburituuree} 
\tbLX{kaale} \tbHM{} \tbPH{káálɪ́} \tbPS{v.~} \tbSN \tbGE{to invoke} \tbSN \tbGE{to perform a libation} \tbPD{kaalɛɛ, kaala, kaala, kaaleba, kaalaa, kaaleba } \tbVA{kaa}
\tbLX{kaale} \tbHM{} \tbPH{kààlɪ̀} \tbPS{v.~} \tbGE{to gather e.g. soil, grains} \tbPD{kaalɛɛ, kaala, kaala, kaaleba, kaalaa, kaaleba } 
\tbLX{kaalɔɔ} \tbHM{} \tbPH{káálɔ̀ɔ́} \tbPS{adv.~} \tbGE{little thing eaten instantly} 
\tbLX{kaama} \tbHM{} \tbPH{káámá} \tbPS{n.~} \tbGE{lively state} \tbTP{kaamɛɛ} 
\tbLX{kaamaako} \tbHM{} \tbPH{káámáákò} \tbPS{n.~} \tbGE{persistence} 
\tbLX{kaapɔge} \tbHM{} \tbPH{kàápɔ́gɪ̀} \tbPS{adv.~} \tbGE{maybe, perhaps} \tbXV{Saa koe la kaapɔge o na mi la.} \tbXE{Rain clouds have gathered, it may rain.} 
\tbLX{kaaraa} \tbHM{} \tbPH{kááráá} \tbPS{n.~} \tbGE{lumpy paste made from bean flour} \tbOP{kaarɛɛ} 
\tbLX{kaare} \tbHM{} \tbPH{káárɪ̀} \tbPS{n.~} \tbGE{car} \tbOP{kaɛ/kɛɛ} \tbTP{kaarɛɛ} 
\tbLX{kaare} \tbHM{} \tbPH{kààrɪ̀} \tbPS{v.~} \tbGE{to drag e.g. feet on the ground} \tbPD{kaarɛɛ, kaara, kaara, kaareba, kaaraa, kaareba } 
\tbLX{kaare} \tbHM{1} \tbPH{káárɪ́} \tbPS{v.~} \tbGE{to clear one's throat; to cough up (phlegm)} \tbPD{kaarɛɛ, kaara, kaara, kaareba, kaaraa, kaareba } 
\tbLX{kaare} \tbHM{2} \tbPH{káárɪ́} \tbPS{v.~} \tbSN \tbGE{to move away} \tbSN \tbGE{to move away immediately, not to return (typically used with gaba)} \tbXV{Ba iri la kparoo ko a bie ka o deɛ kaare gaba.} \tbXE{A shirt was given to the boy and he left at once.} \tbPD{kaarɛɛ, kaara, kaara, kaareba, kaaraa, kaareba } 
\tbLX{kaarebiri} \tbHM{} \tbPH{kàárɪ̀bírí} \tbPS{n.~} \tbGE{sputum} \tbSG{kaarebiruu} \tbOP{kaarebie} \tbTP{kaarebiree/kaarebiiree} 
\tbLX{kaaterɛ} \tbHM{} \tbPH{káàtɪ̀rɛ́} \tbPS{n.~} \tbGE{cursed one} \tbOP{kaaterre} \tbTP{kaaterɛɛ} 
\tbLX{kaa ti᷉i} \tbHM{} \tbPH{kàà tí᷉í} \tbPS{v.~} \tbGE{look closely} \tbPD{kaaɛ ti᷉i, kaara ti᷉i, -, -, -, - } 
\tbLX{ka᷉aze᷉ɛ} \tbHM{} \tbPH{ká᷉ázɪ́᷉ɛ́} \tbPS{n.~} \tbGE{debris from final preparation of shea butter oil} \tbTP{ka᷉aze᷉ɛnɛɛ} 
\tbLX{kabiri} \tbHM{} \tbPH{kábírì} \tbPS{n.~} \tbGE{millet grain} \tbSG{kabiruu} \tbOP{kabie} \tbTP{kabiiree} 
\tbLX{kabooroo} \tbHM{} \tbPH{kábʊ́ʊ́rʊ́ʊ́} \tbPS{n.~} \tbGE{seed millet} \tbOP{kaboora} \tbTP{kaboorɛɛ} 
\tbLX{kabɔrɔ} \tbHM{} \tbPH{kábɔ́rɔ̀} \tbPS{n.~} \tbGE{fresh, mature guinea corn} \tbTP{kabɔrɛɛ} 
\tbLX{kabulo} \tbHM{} \tbPH{kábúló} \tbPS{n.~} \tbGE{sown sprouts of millet} \tbTP{kabulee} 
\tbLX{kadaa} \tbHM{} \tbPH{[kádàá]} \tbPS{n.~} \tbSN \tbGE{millet stalk} \tbSN \tbGE{long stick for threshing millet} \tbOP{kadaare} \tbTP{kadarɛɛ/kadaarɛɛ} 
\tbLX{kadaara} \tbHM{} \tbPH{kádààrà} \tbPS{n.~} \tbGE{millet or guinea corn not that is properly matured} \tbTP{kadaarɛɛ} 
\tbLX{kadori} \tbHM{} \tbPH{kádórí} \tbPS{n.~} \tbGE{heap of millet} \tbOP{kadorri} \tbTP{kadoree} 
\tbLX{kafini} \tbHM{} \tbPH{kàfìní} \tbPS{n.~} \tbGE{lance} \tbOP{kafime} \tbTP{kafinnee} 
\tbLX{kage} \tbHM{} \tbPH{kàgɪ̀} \tbPS{v.~} \tbGE{to hold firmly on to} \tbPD{kagɛɛ, kagera, kagera, kagereba, kageraa, kagereba } 
\tbLX{kage} \tbHM{} \tbPH{kágɪ́} \tbPS{adv.~} \tbGE{stay intently} 
\tbLX{kagelaŋ} \tbHM{} \tbPH{kàglàŋ} \tbPS{adv.~} \tbGE{large and long} 
\tbLX{kagere} \tbHM{} \tbPH{kágrɪ̀} \tbPS{v.~} \tbGE{to beckon, to touch to call the attention of} \tbPD{kagerɛɛ, kagera, kagera, kagereba, kageraa, kagereba } 
\tbLX{kagere} \tbHM{} \tbPH{kágrɪ̀} \tbPS{v.~} \tbGE{to bite or cut lightly} \tbPD{kagerɛɛ, kagera, kagera, kagereba, kageraa, kagereba } 
\tbLX{kagesaga} \tbHM{} \tbPH{kágɪ́ságá} \tbPS{adv.~} \tbGE{small and dry} 
\tbLX{kagesaga} \tbHM{} \tbPH{kàgɪ̀sàgà} \tbPS{adv.~} \tbGE{large and dry} 
\tbLX{kagbin} \tbHM{} \tbPH{kágbín} \tbPS{n.~} \tbGE{ears of corn or millet tied together} \tbOP{ kagbime} \tbTP{kagbinnee} 
\tbLX{kagbulo} \tbHM{} \tbPH{kágbùlò} \tbPS{n.~} \tbGE{unbroken millet seeds} \tbTP{kagbulee} 
\tbLX{kaka} \tbHM{} \tbPH{kàkà} \tbPS{n.~} \tbGE{hippopotamus} \tbOP{kakare} \tbTP{kakarɛɛ} 
\tbLX{kakaduro} \tbHM{} \tbPH{kákádúró} \tbPS{n.~} \tbGE{ginger} \tbSG{kakaduruu} \tbOP{kakadurri} \tbTP{kakaduree} 
\tbLX{kakalaa} \tbHM{} \tbPH{kákáláá} \tbPS{n.~} \tbGE{a type of tree; has extremely hard wood} \tbOP{ kakalle} \tbTP{kakalɛɛ} 
\tbLX{Kakube} \tbHM{} \tbPH{kákúbé} \tbPS{n.~} \tbGE{a farming festival of the Nandom people} \tbTP{kakubee} 
\tbLX{kakubo} \tbHM{} \tbPH{kákúbò} \tbPS{n.~} \tbGE{millet mixed with stones} \tbTP{kakubee} 
\tbLX{kakumo} \tbHM{} \tbPH{kǎkúmò} \tbPS{n.~} \tbGE{donkey} \tbOP{kakunni} \tbTP{kakunnee} 
\tbLX{kakpennɛ} \tbHM{} \tbPH{kákpɪ́nnɛ̀} \tbPS{n.~} \tbGE{untidy millet} \tbTP{kakpennɛɛ} 
\tbLX{kakyeerɛ} \tbHM{} \tbPH{káʧɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{second-grade millet} \tbTP{kakyeerɛɛ} 
\tbLX{kakyeɛrɛ} \tbHM{} \tbPH{káʧɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{one who harvests millet} \tbOP{kakyeɛrebɛ} \tbTP{kakyeɛrebɛɛ} 
\tbLX{kakyo} \tbHM{} \tbPH{káʧó} \tbPS{n.~} \tbGE{basket strainer} \tbOP{ kakyori} \tbTP{kakyoree} 
\tbLX{Kala} \tbHM{} \tbPH{kàlà} \tbPS{n.~} \tbGE{a type of fetish} 
\tbLX{kala} \tbHM{} \tbPH{kàlà} \tbPS{n.~} \tbGE{dog} \tbOP{kalle} \tbTP{kalɛɛ/kallɛɛ} 
\tbLX{kalakala} \tbHM{} \tbPH{kàlàkàlà} \tbPS{adv.~} \tbGE{slightly bitter} 
\tbLX{kalbiri} \tbHM{} \tbPH{kalbiri} \tbPS{n.~} \tbGE{dawdaw spice (seed)} \tbSG{kalbiruu} \tbOP{kalbie} \tbTP{kalbire/kalbiiree} 
\tbLX{kale} \tbHM{} \tbPH{kálɪ́} \tbPS{n.~} \tbGE{suffering, woe as a result of another pesrson's action} \tbTP{kalɛɛ} 
\tbLX{kale} \tbHM{} \tbPH{kálɪ̀} \tbPS{n.~} \tbGE{spice made from dawdaw seed} \tbSG{kaloo} \tbTP{kallɛɛ} 
\tbLX{kale} \tbHM{} \tbPH{kálɪ̀} \tbPS{v.~} \tbGE{to scoop water in small amounts e.g. washing hands} \tbPD{kalɛɛ, kalla, kalla, kaleba, kallaa, kaleba } 
\tbLX{kale} \tbHM{} \tbPH{kálɪ̀} \tbPS{v.~} \tbGE{to make a short comment on another person’s speech} \tbPD{kalɛɛ, kalla, kalla, kaleba, kallaa, kaleba } 
\tbLX{kalekale} \tbHM{} \tbPH{kákál} \tbPS{adv.~} \tbGE{foul-smelling} 
\tbLX{kalekale} \tbHM{} \tbPH{kàlkàlɪ̀} \tbPS{v.~} \tbGE{to move in a jerky manner} \tbPD{kalekalɛɛ, kalekala, kalkala, kalekaleba, kalekalaa, kalekaleba } 
\tbLX{kalenyaa} \tbHM{} \tbPH{kàlɲàá} \tbPS{n.~} \tbGE{a type of spiky grass that causes skin iritation} \tbOP{kalenyaane} \tbTP{kalenyaanɛɛ} 
\tbLX{kaleŋkaa} \tbHM{} \tbPH{kálɪ̀ŋkáá} \tbPS{n.~} \tbGE{gall} \tbOP{kaleŋkaare} \tbTP{kalen.kaarɛɛ} \tbVA{yamm}
\tbLX{kaleŋkaloɔraa} \tbHM{} \tbPH{kálɪ́ŋkálʊ́ɔ́ráá} \tbPS{n.~} \tbGE{a round smooth stone used in grinding} \tbOP{kaleŋkaloɔre} \tbTP{kaleŋkaloɔrɛɛ} \tbVA{kaleŋkoɔraa}
\tbLX{kaleŋkawerɛ} \tbHM{} \tbPH{kálɪ́ŋkàwɪ̀rɛ̀} \tbPS{n.~} \tbGE{a type of wasp} \tbSG{kaleŋkaweroo} \tbTP{kaleŋkawerɛɛ} 
\tbLX{kaleŋkoɔraa} \tbHM{} \tbPH{kálɪ́ŋkʊ́ɔ̀ráá} \tbPS{n.~} \tbGE{round stone} \tbOP{kaleŋkoɔre} \tbTP{kaleŋkoɔrɛɛ} \tbVA{kaleŋkaloɔraa}
\tbLX{kaleŋmeɛ} \tbHM{} \tbPH{kálŋmɪ́ɛ̀} \tbPS{n.~} \tbGE{a type of bird} \tbOP{kaleŋmeɛne} \tbTP{kaleŋmeɛnɛɛ} \tbVA{kareŋmeɛ}
\tbLX{kalere} \tbHM{} \tbPH{kàlɪ́rɪ́} \tbPS{n.~} \tbGE{oath in which one gives away one's food if one fails to twist a finger on another while eating. (children)} \tbOP{kalerre} \tbTP{kalerɛɛ} 
\tbLX{kalinfuoro} \tbHM{} \tbPH{kálínfúóró} \tbPS{n.~} \tbGE{bit of the sign} \tbTP{kalinfuoree} 
\tbLX{kalpigiraa} \tbHM{} \tbPH{kálpígráà} \tbPS{n.~} \tbGE{ball of dawdaw spice} \tbOP{kalpigiri.} 
\tbLX{kalsoro} \tbHM{} \tbPH{kálesóró} \tbPS{n.~} \tbGE{water used to wash dawdaw seed for spice} \tbTP{kalesoree} 
\tbLX{kama} \tbHM{} \tbPH{kámá} \tbPS{adj.~} \tbGE{close eyes shortly} \tbXV{O poɔ zaa᷉ bɛ o la tensoga ka o daŋ gaabo kyɛ o nimri zaa᷉ ba kyɛ kama.} \tbXE{Last night, s/he had an upset stomch and went to bed early, but s/he did not get a wink of sleep.} 
\tbLX{kama} \tbHM{} \tbPH{kàmà} \tbPS{adj.~} \tbGE{cut sharply} 
\tbLX{kamaabiri} \tbHM{} \tbPH{kámáábírì} \tbPS{n.~} \tbGE{kernel (of corn, maize)} \tbSG{kamaabiruu} \tbOP{kamaabie} \tbTP{kamaabiiree} 
\tbLX{kamaadaa} \tbHM{} \tbPH{kámáádàá} \tbPS{n.~} \tbGE{maize stalk} \tbOP{kamaadaare} \tbTP{kamaadaarɛɛ} 
\tbLX{kamaadakɔɔre} \tbHM{} \tbPH{kámáádàkɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{corncob} \tbOP{kamaadakɔbɔ} \tbTP{kamaadakɔɔrɛɛ} 
\tbLX{kamaalɔnzeere} \tbHM{} \tbPH{kámáálɔ̌nzɪ́ɪ́rɪ́} \tbPS{n.~} \tbGE{corn silk} \tbOP{kamaalɔnzeerɛɛ} 
\tbLX{kamaane} \tbHM{} \tbPH{kàmáánɪ̀} \tbPS{n.~} \tbGE{maize} \tbSG{kamaanoo} \tbOP{kamaana} \tbTP{kamaanɛɛ} 
\tbLX{kamaapɛge} \tbHM{} \tbPH{kámáápɛ́gɪ́} \tbPS{n.~} \tbGE{corn husk} \tbOP{kamaapɛgre} \tbTP{kamaapɛgrɛɛ} 
\tbLX{kamakama} \tbHM{} \tbPH{kámákámá} \tbPS{adj.~} \tbGE{brittle and light} 
\tbLX{kamalaɛ} \tbHM{} \tbPH{kámàláɪ̀} \tbPS{n.~} \tbGE{carbide lamp; headlamp used by hunters at night} \tbOP{kamalaare} \tbTP{kamalaarɛɛ} 
\tbLX{kamasɛle} \tbHM{} \tbPH{kàmàsɛ́lɪ̀} \tbPS{n.~} \tbGE{a type of thorny, poisonous tree} \tbTP{kamasɛlɛɛ} 
\tbLX{kamɔnɔ} \tbHM{} \tbPH{kámɔ́nɔ́} \tbPS{n.~} \tbGE{driver} \tbOP{kamɔnema} \tbTP{kamɔnnɛɛ} 
\tbLX{kampala} \tbHM{} \tbPH{kàmpàlà} \tbPS{n.~} \tbGE{a type of smart hunting dog} \tbOP{kampalle} \tbTP{kampalɛɛ/kampallɛɛ} 
\tbLX{kampele} \tbHM{} \tbPH{kámpɪ̀l} \tbPS{n.~} \tbGE{city, town} \tbOP{kampelɛ} \tbTP{kampelɛɛ} 
\tbLX{kampili} \tbHM{} \tbPH{kámpílí} \tbPS{n.~} \tbGE{thatched hut} \tbOP{kampile} \tbTP{kampilee} \tbVA{kampiluu}
\tbLX{kampoŋi} \tbHM{} \tbPH{kámpʊ́ŋ} \tbPS{n.~} \tbGE{camp} \tbOP{kamponne} \tbTP{kamponɛɛ} 
\tbLX{kana} \tbHM{} \tbPH{káná} \tbPS{n.~} \tbGE{thick paste made from flour and broth} \tbTP{kanɛɛ/kannɛɛ} 
\tbLX{kanaakanaa} \tbHM{} \tbPH{kànààkànàà} \tbPS{adj.~} \tbGE{stretched to the limit; tight} 
\tbLX{kananna} \tbHM{} \tbPH{kánánná} \tbPS{adj.~} \tbGE{not coming forth} \tbXV{A saana wa la kyɛ deɛ e kananna.} \tbXE{The stranger came here and made a permanent stay.} 
\tbLX{kanaŋe} \tbHM{} \tbPH{kánáŋ} \tbPS{n.~} \tbGE{a type of red insect with a split tail} \tbOP{kananne} \tbTP{kanannɛɛ} \tbVA{kanaŋa}
\tbLX{kanatɛgɛ} \tbHM{} \tbPH{kánátɛ́gɛ́} \tbPS{n.~} \tbGE{reward, advantage} \tbOP{kanatɛgre} \tbTP{kanatɛgrɛɛ} 
\tbLX{kandazugidaa} \tbHM{} \tbPH{kàndàzùgdáá} \tbPS{n.~} \tbGE{wood of the kandazugiri tree} \tbOP{kandazugdaare} \tbTP{kandazugdarɛɛ} 
\tbLX{kandazugiri} \tbHM{} \tbPH{kàndàzùgdáá} \tbPS{n.~} \tbSN \tbGE{a type of small tree with multiple branches and yellow fruits that are edible} \tbSN \tbGE{the fruits of this tree} \tbOP{kandazugo} \tbTP{kandazugiree} \tbVA{kɔndazugiri}
\tbLX{kane} \tbHM{} \tbPH{kànɪ̀} \tbPS{v.~} \tbGE{to be overmature e.g. okro fruits} \tbPD{kanɛɛ, kanna, kanna, kanema, kannaa, kanema } 
\tbLX{kanea} \tbHM{} \tbPH{kànìá} \tbPS{n.~} \tbGE{hurricane lamp} \tbOP{kaneɛne} \tbTP{kaneɛnɛɛ} 
\tbLX{kaneɛbiri} \tbHM{} \tbPH{kànɪ́ɛ́bírí} \tbPS{n.~} \tbGE{lampshade} \tbOP{kaneɛbie} \tbTP{kaneɛbiree} 
\tbLX{kanfaraŋ} \tbHM{} \tbPH{káfáráŋ} \tbPS{adv.~} \tbGE{light (weight)} 
\tbLX{kankyeemɛ} \tbHM{} \tbPH{kánʧɪ́ɪ́mɛ́} \tbPS{n.~} \tbGE{one hired to cook for another} \tbTP{kankyeemɛɛ} 
\tbLX{kankyɛle} \tbHM{} \tbPH{kánʧɛ́lɪ́} \tbPS{n.~} \tbGE{a type of large tree with large, shiny leaves} \tbOP{kankyɛlɛ} \tbTP{kankyɛlɛɛ} 
\tbLX{kankyɛlpelaa} \tbHM{} \tbPH{kánʧɛ̀lpɪ̀làá} \tbPS{n.~} \tbGE{a type of tree} \tbOP{kankyɛlpeɛle} \tbTP{kankyɛlpeɛlɛɛ} 
\tbLX{kankyiri} \tbHM{} \tbPH{kànʧírí} \tbPS{n.~} \tbGE{a type of small plant with edible leaves} \tbTP{kankyiree} 
\tbLX{kankyirikpoŋ} \tbHM{} \tbPH{kánʧíríkpóŋ} \tbPS{n.~} \tbGE{variety of (see kankyiri)} \tbOP{kankyirikponni} \tbTP{kankyirikponnee} 
\tbLX{kanne} \tbHM{} \tbPH{kànnɪ̀} \tbPS{v.~} \tbSN \tbGE{to teach} \tbSN \tbGE{to read} \tbPD{kannɛɛ, kanna, kanna, kanema, kannaa, kanema } 
\tbLX{kanne} \tbHM{} \tbPH{kánnɪ̀} \tbPS{v.~} \tbGE{to doubt} \tbPD{kannɛɛ, kanna, kanna, kanema, kannaa, kanema } 
\tbLX{kannyiri} \tbHM{} \tbPH{kànɲìrì} \tbPS{n.~} \tbGE{be patient} \tbTP{kannyiree} 
\tbLX{kansele} \tbHM{} \tbPH{kánsɪ̀l} \tbPS{n.~} \tbGE{council} \tbOP{kanselɛ} \tbTP{kanselɛɛ} 
\tbLX{kansɛ} \tbHM{} \tbPH{kánsɛ́} \tbPS{n.~} \tbGE{cakes} \tbOP{kansɛre} \tbTP{kansɛrɛɛ} \tbVA{sɛnsɛre, sensɛ}
\tbLX{kansɔge} \tbHM{} \tbPH{kánsɔ́gɪ́} \tbPS{n.~} \tbGE{foreign (not locally made) hoe} \tbOP{kansɔgre} \tbTP{kansɔgrɛɛ} 
\tbLX{kantakye} \tbHM{} \tbPH{kántáʧè} \tbPS{n.~} \tbGE{train} \tbOP{kantakyeri} \tbTP{kantakyeree} 
\tbLX{kantɛroo} \tbHM{} \tbPH{kántɛ́rʊ́ʊ́} \tbPS{n.~} \tbGE{a type of strong grass that grows on riverbeds} \tbOP{kantɛrɛɛ} \tbTP{kantɛroo} 
\tbLX{kanwaa} \tbHM{} \tbPH{kánwàà} \tbPS{adv.~} \tbGE{supposed} \tbXV{O kanwaa libie na baarɛɛ.} \tbXE{His supposed money is finished.} 
\tbLX{kanwaa} \tbHM{} \tbPH{kánwáá} \tbPS{n.~} \tbGE{hard pebble of acid} \tbSG{kanwaarɛɛ} \tbOP{kanwaare} 
\tbLX{kanwaa-ŋmen-yelmanna} \tbHM{} \tbPH{kánwàá-ŋmɪ́n-yèlmǎnnà} \tbPS{n.~} \tbGE{false prophet} \tbOP{kanwaa-ŋmen-yelmanema} \tbTP{kanwaa-ŋmen-yelmanemɛɛ} 
\tbLX{kaŋazaa} \tbHM{} \tbPH{káŋázáá} \tbPS{adj.~} \tbGE{anyone, each one} 
\tbLX{kaŋemaŋ} \tbHM{} \tbPH{káŋmáŋ} \tbPS{adv.~} \tbGE{light (weight)} 
\tbLX{kaŋkaaraamoorɛɛ} \tbHM{} \tbPH{káŋkááráámʊ́ʊ́rɛ́ɛ́} \tbPS{n.~} \tbGE{expression signifying the end of a folktale} 
\tbLX{kaŋkalaŋ} \tbHM{} \tbPH{kàŋkàlàŋ} \tbPS{adv.~} \tbGE{appearing suddenly} \tbXV{Te zeŋe dire seemaa wa berɛ ka saana la kaŋkalaŋ.} \tbXE{As we sat down eating a stranger suddenly appeared.} 
\tbLX{kaŋkalare} \tbHM{} \tbPH{káŋkálàrɪ̀} \tbPS{n.~} \tbGE{persistence in doing something despite knowing it is beyond one's ability} \tbTP{kaŋkalarɛɛ} 
\tbLX{kaŋkale} \tbHM{} \tbPH{kǎŋkál} \tbPS{n.~} \tbGE{a type of medicinal stew food made from flour and broth} \tbTP{kaŋkalɛɛ} 
\tbLX{kaŋkambulo} \tbHM{} \tbPH{kaŋkambulo} \tbPS{n.~} \tbGE{chronic ulcer} 
\tbLX{kaŋkandoraa} \tbHM{} \tbPH{káŋkándòráá} \tbPS{n.~} \tbGE{a type of bird with tan plumage and a thick beak} \tbOP{kaŋkandori} \tbTP{kaŋkandoree} \tbVA{kaŋkanduo}
\tbLX{kaŋkanduo} \tbHM{} \tbPH{káŋkǎndúó} \tbPS{n.~} \tbGE{a type of small bird with a thick beak that eats fruits of trees} \tbOP{kaŋkandori, kaŋkandorri, kaŋkandoree} \tbVA{kaŋkandobaa/kaŋkandoraa}
\tbLX{kaŋkankyekyeɛ} \tbHM{} \tbPH{kákánʧɪ́ʧɪ́ɛ́} \tbPS{n.~} \tbGE{a type of tree with milky sap} \tbOP{kaŋkankyekyeere} \tbTP{kaŋkankyekyeerɛɛ} 
\tbLX{kaŋkansaalaa} \tbHM{} \tbPH{káŋkánsààláà} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{kaŋkansaale} \tbTP{kaŋkansaalɛɛ} 
\tbLX{kaŋkansɔglaa} \tbHM{} \tbPH{káŋkánɔ́gláá} \tbPS{n.~} \tbGE{a type of fig tree with aerial roots} \tbOP{kaŋkansɔglɔ} \tbTP{kaŋkansɔglɛɛ} 
\tbLX{kaŋkanzeɛ} \tbHM{} \tbPH{ káŋkánzɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of fig tree with medicinal properties and edible fruits} \tbOP{ kaŋkanzeere} \tbTP{ kaŋkanzeerɛɛ} 
\tbLX{kaŋkaŋ} \tbHM{} \tbPH{káŋkáŋ} \tbPS{adv.~} \tbGE{bad-smelling} 
\tbLX{kaŋkaŋaa} \tbHM{} \tbPH{kǎŋkáŋàá} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{kaŋkanne} \tbTP{kaŋkannɛɛ} 
\tbLX{kaŋkaŋgbalaa} \tbHM{} \tbPH{kàŋkàŋgbàláà} \tbPS{n.~} \tbGE{a type of shady tree} \tbOP{kaŋkaŋgballe} \tbTP{kaŋkaŋgbalɛɛ} \tbVA{gbalaa}
\tbLX{kaŋkaŋkyekyeɛ} \tbHM{} \tbPH{káŋkáŋʧɪ̀ʧɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of fig tree} \tbOP{kaŋkaŋkyekyeere} \tbTP{kaŋkaŋkyekyeerɛɛ} 
\tbLX{kaŋmeɛrɛ} \tbHM{} \tbPH{káŋmɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{one engaged in the threshing of millet} \tbOP{kaŋmeɛrebɛ} \tbTP{kaŋmeɛrebɛɛ} 
\tbLX{kaŋmɛ} \tbHM{} \tbPH{káŋmɛ̀} \tbPS{n.~} \tbGE{act of threshing millet} 
\tbLX{kaŋmɛdaa} \tbHM{} \tbPH{káŋmɛ̀dàá} \tbPS{n.~} \tbGE{stick for threshing millet} \tbOP{kaŋmɛdaare} \tbTP{kaŋmɛdarɛɛ/kaŋmɛdaarɛɛ} 
\tbLX{kapaalaa} \tbHM{} \tbPH{kápááláá} \tbPS{n.~} \tbGE{raised shed for guinea corn or millet} \tbOP{kapaale} \tbTP{kapaalɛɛ} 
\tbLX{kapala} \tbHM{} \tbPH{kàpálà} \tbPS{n.~} \tbGE{mashed yam, plantain or cassava} \tbSG{kapaloo} \tbTP{kapalɛɛ/kapallɛɛ} \tbVA{kapalo (Saanchi) -mashed yams, pl. kapala}
\tbLX{kapenta} \tbHM{} \tbPH{kápɪ́ntà} \tbPS{n.~} \tbGE{carpenter} \tbOP{kapentarɪ} 
\tbLX{kapile} \tbHM{} \tbPH{kàpílè} \tbPS{adj.~} \tbGE{crafty} 
\tbLX{kapoŋi} \tbHM{} \tbPH{kápòŋ} \tbPS{n.~} \tbGE{millet flower} \tbOP{kaponni} \tbTP{kaponnee} 
\tbLX{kapuro} \tbHM{} \tbPH{kàpùrò} \tbPS{n.~} \tbGE{pillow} \tbOP{kapurri} \tbTP{kapuree/kapurree} 
\tbLX{kara} \tbHM{} \tbPH{kárà} \tbPS{v.~} \tbGE{to prepare, to get ready} \tbXV{O kara la a saama karoo.} \tbXE{S/he is preparing for the visitors.} \tbPD{karɛɛ, kara, kara, kareba, karaa, kareba } 
\tbLX{kara} \tbHM{} \tbPH{kárá} \tbPS{n.~} \tbGE{a type of grass used for thatching} 
\tbLX{karaa} \tbHM{} \tbPH{káráá} \tbPS{adj.~} \tbSN \tbGE{unripe} \tbSN \tbGE{not cooked through} \tbSN \tbGE{not hot enough (of iron in fire)} \tbOP{kare} \tbTP{karrɛɛ} 
\tbLX{karaa} \tbHM{} \tbPH{káráá} \tbPS{adj.~} \tbGE{uncompromising} \tbOP{kare} \tbTP{karrɛɛ} 
\tbLX{karaa} \tbHM{} \tbPH{kàráà} \tbPS{n.~} \tbGE{cobra} \tbOP{karre} \tbTP{karrɛɛ} 
\tbLX{karaa} \tbHM{} \tbPH{kàráà} \tbPS{n.~} \tbGE{clerk} \tbOP{karre} \tbTP{karrɛɛ} 
\tbLX{karaa} \tbHM{} \tbPH{káráá} \tbPS{n.~} \tbGE{file} \tbOP{karre} \tbTP{karɛɛ/karrɛɛ} 
\tbLX{karaa} \tbHM{} \tbPH{kàràà} \tbPS{adj.~} \tbGE{large} \tbOP{karre} 
\tbLX{karambanne} \tbHM{} \tbPH{kárámbánnɪ́} \tbPS{n.~} \tbGE{wit} \tbTP{karambannɛɛ} 
\tbLX{karansunni} \tbHM{} \tbPH{kàrànsúnnì} \tbPS{n.~} \tbGE{kerosene} \tbTP{karansunnee} 
\tbLX{karasa} \tbHM{} \tbPH{kàràsà} \tbPS{adj.~} \tbGE{bad-smelling} 
\tbLX{karawaara} \tbHM{} \tbPH{káráwáárá} \tbPS{n.~} \tbGE{green grass that has not been burnt completyely} \tbSG{karawaaroo} \tbTP{karawaarɛɛ} 
\tbLX{kare} \tbHM{} \tbPH{kàrɪ̀} \tbPS{v.~} \tbSN \tbGE{to drive away} \tbSN \tbGE{to chase, to run after} \tbSN \tbGE{to sharpen or polish a weapon or tool (arrow or knife)} \tbSN \tbGE{to grow tall; to cause to rise tall or high} \tbSN \tbGE{to get overgrown (okro)} \tbPD{karɛɛ, kara, kara, kareba, karaa, kareba } 
\tbLX{kare} \tbHM{} \tbPH{kárɪ́} \tbPS{adv.~} \tbGE{completely (of people going away)} \tbXV{A bibiiri zaaŋ zo, kare.} \tbXE{All the children have run away, leaving none behind.} 
\tbLX{kare} \tbHM{} \tbPH{kárɪ́} \tbPS{v.~} \tbGE{to break (small objects)} \tbPD{karɛɛ, kara, kara, kareba, karaa, kareba } 
\tbLX{karekare} \tbHM{} \tbPH{kàrɪ̀kárɪ́} \tbPS{v.~} \tbGE{to attempt to make a move} \tbXV{Meŋ iri karekare ka te nyɛ ka o na sage la me.} \tbXE{Also make an effort and let's see whether he will agree.} \tbPD{karekarɛɛ, karekara, karekara, karekareba, karekaraa, - } 
\tbLX{karekpe} \tbHM{} \tbPH{kàrɪ́kpé} \tbPS{n.~} \tbGE{spitting cobra} \tbOP{karekperi} \tbTP{karekperee} 
\tbLX{karekpele} \tbHM{} \tbPH{kárɪ́kpɪ́lɪ́} \tbPS{n.~} \tbGE{male lizard; a small reptile of the squamata family that is red-headed} \tbOP{karekpelɛ} \tbTP{karekpelɛɛ} \tbVA{shortened form of bandakarekpele}
\tbLX{karekye} \tbHM{} \tbPH{kárɪ́ʧɪ̀} \tbPS{n.~} \tbGE{clerk} \tbOP{karekyere} \tbTP{karekyerɛɛ} 
\tbLX{karema} \tbHM{} \tbPH{kárɪ́má} \tbPS{n.~} \tbGE{teacher, one who deals with studies} \tbOP{karemamine} \tbTP{karemɛɛ} 
\tbLX{karemogɔ} \tbHM{} \tbPH{kárɪ̀mʊ́gá} \tbPS{n.~} \tbGE{fortune-teller, diviner} \tbOP{karemogre} \tbTP{karemogrɛɛ} 
\tbLX{karenteɛ} \tbHM{} \tbPH{kàrɪ́ntɪ́ɛ́} \tbPS{n.~} \tbGE{machete, cutlass} \tbOP{karenteere} \tbTP{karenteerɛɛ} 
\tbLX{kareŋmeɛ} \tbHM{} \tbPH{kárɪ́ŋmɪ́ɛ̀} \tbPS{n.~} \tbGE{a type of bird with long, thin legs} \tbOP{kareŋmeɛne} \tbTP{kareŋmeɛnɛɛ} \tbVA{kalŋmeɛ}
\tbLX{karingyoŋ} \tbHM{} \tbPH{káríndzóŋ} \tbPS{n.~} \tbGE{a type of musical instrument} \tbOP{karingyonni} \tbTP{karingyonnee} \tbVA{karangyoŋ/karengyoŋ}
\tbLX{kasɛɛte} \tbHM{} \tbPH{kàsɛ́ɛ́tɪ́} \tbPS{n.~} \tbGE{cassette tape} \tbOP{kasɛɛtere} \tbTP{kasɛɛterɛɛ} 
\tbLX{kasɛlemɛ} \tbHM{} \tbPH{kásɛ́lɪ́mɛ́} \tbPS{n.~} \tbGE{millet seedling} \tbVA{kasɛlle}
\tbLX{kasɛlle} \tbHM{} \tbPH{kásɛ́llɪ́} \tbPS{n.~} \tbGE{millet seedling meant for planting} \tbVA{kasɛlemɛ}
\tbLX{kasogi} \tbHM{} \tbPH{kásògí} \tbPS{n.~} \tbGE{a small cage used as a nest for young chicks} \tbOP{kasogri} \tbTP{kasogree} \tbVA{kosogi}
\tbLX{katakye} \tbHM{} \tbPH{kátáʧè} \tbPS{n.~} \tbGE{train} \tbOP{katakyeri} \tbTP{katakyeree} 
\tbLX{kataŋa} \tbHM{} \tbPH{kátáŋà} \tbPS{n.~} \tbGE{a large granary for storage of foodstuff} \tbVA{katɛgɛ}
\tbLX{kataweɛ} \tbHM{} \tbPH{kàtàwɪ́ɛ́} \tbPS{n.~} \tbGE{umbrella} \tbOP{kataweere} \tbTP{kataweerɛɛ} 
\tbLX{katɛgɛ} \tbHM{} \tbPH{kàtɛ́gɛ̀} \tbPS{n.~} \tbGE{large granary} \tbOP{katɛgre} \tbTP{katɛgrɛɛ} \tbVA{kataŋa}
\tbLX{katoŋe} \tbHM{} \tbPH{kàtʊ̂ŋ} \tbPS{n.~} \tbGE{civet} \tbOP{katonne} \tbTP{katonnɛɛ} 
\tbLX{ka-uri} \tbHM{} \tbPH{ká-úrí} \tbPS{n.~} \tbGE{dusty chaff from millet (irritates the skin)} \tbTP{ka-uree} 
\tbLX{kayeɛle} \tbHM{} \tbPH{káyɪ́ɛ́lɪ̀} \tbPS{n.~} \tbGE{the act of winnowing millet} 
\tbLX{kayeɛlɛ} \tbHM{} \tbPH{káyɪ́ɛ́lɛ́} \tbPS{n.~} \tbGE{one engaged in winnowing millet} \tbOP{kayeɛlbɛ} \tbTP{kayeɛlbɛɛ} 
\tbLX{kayɛre} \tbHM{} \tbPH{káyɛ́rɪ́} \tbPS{n.~} \tbGE{half-wild millet} \tbOP{kayɛɛ} \tbTP{kayɛɛrɛɛ} 
\tbLX{kazama} \tbHM{} \tbPH{kázàmá} \tbPS{n.~} \tbGE{poorly matured millet} \tbTP{kazamɛɛ} 
\tbLX{kazeɛ} \tbHM{} \tbPH{kázɪ̀ɛ́} \tbPS{n.~} \tbGE{guinea corn} \tbSG{kazeroo (plant)} \tbTP{kazeerɛɛ} 
\tbLX{kazu} \tbHM{} \tbPH{kázú} \tbPS{n.~} \tbGE{ear of millet} \tbOP{kazuri} \tbTP{kazuree} 
\tbLX{keɛre} \tbHM{} \tbPH{kɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to make a cluckling noise to call fowl or drive animals away} \tbPD{keɛrɛɛ, keɛrɛ, keɛrɛ, keɛrebɛ, keɛraa, keɛrebɛ } 
\tbLX{kegireŋ} \tbHM{} \tbPH{kégréŋ} \tbPS{adj.~} \tbGE{small and hanging} \tbXV{A bie yɛre la wo-kegireŋ kaŋa.} \tbXE{The child carried some small animal skin bag.} 
\tbLX{kegireŋ} \tbHM{} \tbPH{kègrèŋ} \tbPS{adj.~} \tbGE{large and hanging} 
\tbLX{kegisege} \tbHM{} \tbPH{kégségé} \tbPS{adv.~} \tbGE{small but compact} \tbXV{A noɔɛ la a e kegisegee lɛ.} \tbXE{the child has grown and is plump.} 
\tbLX{kegisege} \tbHM{} \tbPH{kègsègè} \tbPS{adv.~} \tbGE{large and compact} \tbXV{Wɛnaabo maŋ waa la kegisegee lɛ.} \tbXE{The buffalo is usually big and plump.} 
\tbLX{kei} \tbHM{} \tbPH{kéì} \tbPS{interjection.~} \tbGE{stop it} 
\tbLX{kei} \tbHM{} \tbPH{kèí} \tbPS{interjection.~} \tbGE{expression of disapproval} 
\tbLX{kele} \tbHM{} \tbPH{kélé} \tbPS{n.~} \tbGE{very hard seeds of some fruits; e.g baobab (adansonia digitata) gourd} \tbVA{tokele}
\tbLX{kelewele} \tbHM{} \tbPH{kéléwèlé} \tbPS{n.~} \tbGE{fried plantain chips} \tbTP{kelewellee} 
\tbLX{kelifelli} \tbHM{} \tbPH{kéféll} \tbPS{adj.~} \tbGE{light (weight)} 
\tbLX{keloo} \tbHM{} \tbPH{kélòó} \tbPS{adv.~} \tbGE{eat or chew instantly} 
\tbLX{keloo} \tbHM{} \tbPH{kèlòò} \tbPS{adv.~} \tbGE{glance at once} 
\tbLX{kenekene} \tbHM{} \tbPH{kɪ́nkɪ̀n} \tbPS{adv.~} \tbGE{full to capacity liquid} 
\tbLX{keŋkeleŋ} \tbHM{} \tbPH{kéŋkéléŋ} \tbPS{adj.~} \tbGE{light and dangling} \tbXV{A dɔɔ de la ŋmampege keŋkeleŋ kaŋa yɛre a baa.} \tbXE{The man has hung some light broken calabash on the neck of the dog.} 
\tbLX{keŋkeleŋ} \tbHM{} \tbPH{kèŋkèlèŋ} \tbPS{adj.~} \tbGE{large and dangling} \tbXV{Ka a bieŋ yɛre a wo-keŋkeleŋ ŋa o koŋ tõɔ kyɛŋe ne.} \tbXE{If the boy hands this large skin bag on his shoulder he cannot walk with it.} 
\tbLX{keŋkeŋ} \tbHM{} \tbPH{kéŋkéŋ} \tbPS{adj.~} \tbGE{very full (stomach, balloon)} 
\tbLX{kere} \tbHM{} \tbPH{kèré} \tbPS{n.~} \tbGE{dog} \tbOP{kerri} \tbTP{keree} 
\tbLX{kerekere} \tbHM{} \tbPH{kérékéré} \tbPS{adv.~} \tbGE{lightweight (thickness)} 
\tbLX{kerekere} \tbHM{} \tbPH{kɪ́rɪ́kɪ́rɪ́} \tbPS{n.~} \tbGE{wheel from the bucket} \tbOP{kerekerre} \tbTP{kerekerɛɛ} 
\tbLX{kerekere} \tbHM{} \tbPH{kɪ̀rɪ̀kɪ́rɪ́} \tbPS{v.~} \tbGE{to struggle} \tbPD{kerekerɛe,, kerekerɛ, kerekerɛ, kerekerebɛ, kerekeraa, - } 
\tbLX{kerekerekeree} \tbHM{} \tbPH{kérékérékèrèé} \tbPS{n.~} \tbGE{nursery rhyme} \tbOP{kerekerri.} 
\tbLX{kereŋkereŋ} \tbHM{} \tbPH{kéréŋkéréŋ} \tbPS{adv.~} \tbGE{full to capacity (tyre, balloon, or stomach)} 
\tbLX{kerese} \tbHM{} \tbPH{kérésé} \tbPS{adj.~} \tbGE{small and distended} \tbXV{A bibaala puori waa la kerese lɛ.} \tbXE{The sick child’s stomach is distended.} 
\tbLX{kerese} \tbHM{} \tbPH{kèrèsè} \tbPS{adj.~} \tbGE{big and distended (stomach)} 
\tbLX{keri} \tbHM{} \tbPH{kèrì} \tbPS{v.~} \tbGE{come round clumsiness} \tbXV{Zenɛ ka a boɔ dɔge a bile naŋ ba keri} \tbXE{The goat just literred today the kid has not come round yet.} \tbPD{keree, kere, kere, keribe, keraa, keribe } 
\tbLX{keri} \tbHM{} \tbPH{kérí} \tbPS{v.~} \tbSN \tbGE{to measure, to weigh} \tbSN \tbGE{to face a difficult task} \tbXV{O keri la toma ko ma ka N puro ne wale.} \tbXE{She has burdened me with a heavy job, and I am sweating.} \tbPD{keree, kere, kere, keribe, keraa, keribe } 
\tbLX{keri} \tbHM{} \tbPH{kèrì} \tbPS{adv.~} \tbGE{onomatopoeic; the way a type of sound is produced when a small object hits a bigger (stiff) object} 
\tbLX{kɛkɛre} \tbHM{} \tbPH{kɛ̀kɛ́rɪ̀} \tbPS{n.~} \tbSN \tbGE{stalk (of maize, millet)} \tbSN \tbGE{small bundle of stalks used for gathering threshed millet} \tbSG{kɛkɛroo} \tbOP{kɛkɛrɛɛ} \tbTP{kɛkɛrrɛɛ} 
\tbLX{kɛlɛkɛlɛ} \tbHM{} \tbPH{kɛ́lɛ́kɛ́lɛ́} \tbPS{adj.~} \tbGE{lightweight (thickness)} 
\tbLX{kieluu} \tbHM{} \tbPH{kíélúú} \tbPS{adj.~} \tbGE{burnt-smelling} 
\tbLX{kieluu} \tbHM{} \tbPH{kíélúú} \tbPS{adj.~} \tbGE{high-pitched and distorted (singing)} 
\tbLX{kieree} \tbHM{} \tbPH{kíéréé} \tbPS{n.~} \tbGE{brim of hat e.g. cowboy hat, bowler hat} \tbOP{kieri} 
\tbLX{ko} \tbHM{} \tbPH{kʊ́} \tbPS{v.~} \tbGE{to kill} \tbPD{koɛ, koora, koora, kooreba, kooraa, kooreba } 
\tbLX{ko} \tbHM{} \tbPH{kʊ́} \tbPS{v.~} \tbGE{to put out (fire)} \tbPD{koɛ, koora, koora, kooreba, kooraa, kooreba } 
\tbLX{ko} \tbHM{} \tbPH{kʊ́} \tbPS{v.~} \tbGE{to fold (cloth)} \tbPD{koɛ, koora, koora, kooreba, kooraa, kooreba } 
\tbLX{ko} \tbHM{} \tbPH{kò} \tbPS{v.~} \tbGE{to dry} \tbPD{ koe,  koro, koro, koribo, koraa, koribo } 
\tbLX{ko} \tbHM{} \tbPH{kó} \tbPS{v.~} \tbGE{to gather e.g rain clouds} \tbPD{ koe,  koro, -, -, -, - } 
\tbLX{ko} \tbHM{} \tbPH{kʊ̀} \tbPS{v.~} \tbGE{to give} \tbPD{koɛ, kora, kora, koreba, koraa, koreba } 
\tbLX{kobeɛre} \tbHM{} \tbPH{kòbɪ́ɛ́rɪ̀} \tbPS{n.~} \tbGE{first weekday of post-funeral rites} \tbOP{kobeɛre /kóbɪ́ɛ́rɪ́/} \tbTP{kobeɛrɛɛ} 
\tbLX{kobo} \tbHM{} \tbPH{ko̊bò} \tbPS{n.~} \tbGE{penny} \tbOP{kogri} \tbTP{kogree} 
\tbLX{koboŋkoboŋ} \tbHM{} \tbPH{kòbòŋkóbóŋ} \tbPS{adj.~} \tbGE{rugged, uneven e.g. road} 
\tbLX{koboɔ} \tbHM{} \tbPH{kòbʊ́ɔ́} \tbPS{n.~} \tbGE{goat for funeral rites} \tbOP{koboore} \tbTP{koboorɛɛ} 
\tbLX{kobora} \tbHM{} \tbPH{kʊ̀bʊ̀rà} \tbPS{n.~} \tbGE{leader, greatest one among a group} \tbOP{koborre} \tbTP{koborɛɛ} 
\tbLX{kodaa} \tbHM{} \tbPH{kòdàá} \tbPS{n.~} \tbGE{funeral week} \tbOP{kodare} \tbTP{kodarɛɛ} 
\tbLX{koda᷉a} \tbHM{} \tbPH{kòdá᷉á} \tbPS{n.~} \tbGE{funeral drink} \tbSG{koda᷉ao} \tbTP{kodaanɛɛ} 
\tbLX{kodie} \tbHM{} \tbPH{kòdìè} \tbPS{n.~} \tbGE{first weekend of post-funeral rites celebration} \tbOP{koderi} \tbTP{koderee} 
\tbLX{kodɔgerɔ} \tbHM{} \tbPH{kʊ̊dɔ̀grɔ̀} \tbPS{n.~} \tbGE{foetus (that causes the death of a parent during labour)} \tbOP{kodɔgere} \tbTP{koɖgerɛɛ} 
\tbLX{kodɔɔ} \tbHM{} \tbPH{kòdɔ́ɔ́} \tbPS{n.~} \tbGE{widower} \tbOP{ko-dɔbɔ} \tbTP{ko-dɔɔrɛɛ} 
\tbLX{kodu} \tbHM{} \tbPH{kòdú} \tbPS{n.~} \tbGE{banana} \tbSG{koduruu} \tbOP{koduri} \tbTP{koduree} 
\tbLX{kofi} \tbHM{} \tbPH{kófì} \tbPS{n.~} \tbGE{coffee} \tbOP{kofiri} \tbTP{kofiree} 
\tbLX{koganaa} \tbHM{} \tbPH{kògànàá} \tbPS{n.~} \tbGE{strip of leather, cloth, or twine for a bereaved person} \tbOP{koganne} \tbTP{kogannɛɛ} 
\tbLX{kogi} \tbHM{} \tbPH{kógí} \tbPS{n.~} \tbSN \tbGE{seat, stool, chair} \tbSN \tbGE{throne} \tbOP{kogree} 
\tbLX{kogi} \tbHM{} \tbPH{kògí} \tbPS{adl.~} \tbGE{thick and hard} 
\tbLX{kogilii} \tbHM{} \tbPH{kógilíí} \tbPS{adj.~} \tbGE{dry and wrinkled; used on humans, animals and objects like containers} \tbOP{kogilo} \tbVA{kogiluu}
\tbLX{kogilimala} \tbHM{} \tbPH{kógílímálá} \tbPS{adj.~} \tbGE{very lean and worn out} 
\tbLX{kogiloŋ} \tbHM{} \tbPH{kóglóŋ} \tbPS{adj.~} \tbGE{small and curved} 
\tbLX{kogiluu} \tbHM{} \tbPH{kógílúú} \tbPS{adj.~} \tbGE{lean e.g a person, objects like containers, animal and humans} \tbOP{kogilo} 
\tbLX{Kogiri} \tbHM{} \tbPH{kógrì} \tbPS{n.~} \tbGE{name of a village in Central Dagao} \tbTP{kogiree} 
\tbLX{kogiri} \tbHM{} \tbPH{kógrí} \tbPS{adj.~} \tbGE{near the side of} \tbXV{A teɛ peɛle la mɔkogiri.} \tbXE{The tree is near the edge of the bush.} 
\tbLX{kogiri} \tbHM{} \tbPH{kógrì} \tbPS{v.~} \tbGE{to wink} \tbPD{kogree, kogiro, kogiro, kogiribo, kogiraa, kogiribo } 
\tbLX{kogiro} \tbHM{} \tbPH{kógró} \tbPS{n.~} \tbGE{basket seat} \tbOP{kogiri} \tbTP{kogiree} 
\tbLX{kogiroŋ} \tbHM{} \tbPH{kògròŋ} \tbPS{adj.~} \tbGE{oblong in shape} 
\tbLX{kogluu} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{} \tbVA{kogilii}
\tbLX{kogo} \tbHM{} \tbPH{kógò} \tbPS{n.~} \tbGE{mahogany} \tbOP{kogri} \tbTP{kogree} \tbVA{kogi}
\tbLX{kogo᷉ɔ} \tbHM{} \tbPH{kógʊ́᷉ɔ́} \tbPS{n.~} \tbGE{funeral gound (arena of dirge singers)} \tbOP{kogoɔne} \tbTP{kogoɔnɛɛ} 
\tbLX{kokolaa} \tbHM{} \tbPH{kókóláà} \tbPS{n.~} \tbGE{shuttle (weaving)} \tbOP{kokolli} \tbTP{kokolee/kokollee} 
\tbLX{kokolaa} \tbHM{} \tbPH{kókóláà} \tbPS{n.~} \tbGE{a type of river god or fetish} \tbOP{kokolli} \tbTP{kokolee/kokollee} 
\tbLX{kokolo} \tbHM{} \tbPH{kòkòlò} \tbPS{n.~} \tbGE{a type of tree whose ashes are used as salt} \tbOP{kokolli} \tbTP{kokollee} 
\tbLX{kokomo} \tbHM{} \tbPH{kòkòmò} \tbPS{adj.~} \tbGE{the nature of the sound of when one takes a bite at something hard. e.g. a bone} 
\tbLX{kokoori} \tbHM{} \tbPH{kòkòòrí} \tbPS{n.~} \tbGE{pan flute} \tbOP{kokoe} \tbTP{kokoree/kokorree} 
\tbLX{koko᷉ɔ} \tbHM{} \tbPH{kòkʊ́᷉ɔ́} \tbPS{n.~} \tbGE{food or drink given to mourners at a funeral} \tbTP{kokoɔnɛɛ} 
\tbLX{kokyi} \tbHM{} \tbPH{kòʧí} \tbPS{n.~} \tbGE{millet for funeral rites} \tbTP{kokyiree} 
\tbLX{kolaa} \tbHM{} \tbPH{kʊ́láá} \tbPS{n.~} \tbSN \tbGE{river, source of water people must fetch} \tbSN \tbGE{river god} \tbOP{kolle} \tbTP{kollɛɛ} 
\tbLX{kolaanoɔre} \tbHM{} \tbPH{kʊ́láánʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{south of Ghana, Kumasi} 
\tbLX{kolaa nyɔge} \tbHM{} \tbPH{kʊ́láá ɲɔ́gɪ̀} \tbPS{intransitive v.~} \tbGE{to drown} 
\tbLX{kolakola} \tbHM{} \tbPH{kʊ́låkʊ́lå} \tbPS{adv.~} \tbGE{soft e.g. cloth, leather} \tbVA{kolɔkolɔ}
\tbLX{kolaŋene} \tbHM{} \tbPH{kòláŋɪ́nɪ́} \tbPS{n.~} \tbGE{dirge} \tbTP{laŋenɛɛ} 
\tbLX{kolboll} \tbHM{} \tbPH{kʊ̀bʊ̀ll} \tbPS{adv.~} \tbGE{rising suddenly and in great quantity e.g. dust, smoke} 
\tbLX{kolee} \tbHM{} \tbPH{kóléé} \tbPS{n.~} \tbGE{tin ; cup} \tbOP{kobilii} \tbTP{kobilee/kollee} 
\tbLX{kolemɛleŋ} \tbHM{} \tbPH{kʊ́lmɛ́lɪ́ŋ} \tbPS{adj.~} \tbGE{match or correspond very well e.g. rice and beans} 
\tbLX{kolezare} \tbHM{} \tbPH{kòlɪ̀zárɪ́} \tbPS{n.~} \tbGE{funeral donation} \tbOP{kolezaɛ} \tbTP{kolezɛrɛɛ} 
\tbLX{kolimasogilo} \tbHM{} \tbPH{kòlìmásògìló} \tbPS{n.~} \tbGE{a legendary bird believed to be as big as the ostrich and wicked} \tbOP{kolimasogili} \tbTP{kolimasoglee} 
\tbLX{kolkol} \tbHM{} \tbPH{kʊ̀lkʊ̀l} \tbPS{adv.~} \tbGE{in great numbers} 
\tbLX{kolkole} \tbHM{} \tbPH{kʊ̀lkʊ́l} \tbPS{v.~} \tbGE{to engage in affectionate play; usually between lovers} \tbPD{kolkolɛɛ, kolkolɔ/kolkola, kolkolɔ/kolkola, kolkoleba, kolkollaa, - } 
\tbLX{kolkoli} \tbHM{} \tbPH{kòlkól} \tbPS{v.~} \tbGE{to be all over another e.g. a dog welcoming an acquaintance} \tbPD{kololee, kolkolo, kolkolo, kolkolaa, kolkolibo, kolkolo } 
\tbLX{koll} \tbHM{} \tbPH{kʊ̀ll} \tbPS{adv.~} \tbGE{rising into the air e.g. smoke, dust} \tbVA{kull}
\tbLX{kolle} \tbHM{} \tbPH{kʊ̀llɪ̀} \tbPS{v.~} \tbGE{to gather or cut vegetation in great quantities} \tbPD{kollɛɛ, kolla, kolla, koleba, kollaa, koleba } 
\tbLX{kolnoɔ} \tbHM{} \tbPH{kʊ́lnʊ́ɔ̀} \tbPS{n.~} \tbGE{moorhen} \tbOP{kolnoore} \tbTP{kolnoorɛɛ} 
\tbLX{kolokolo} \tbHM{} \tbPH{kòlókòló} \tbPS{n.~} \tbGE{turkey} \tbOP{kolokolli} \tbTP{kolokollee} 
\tbLX{kolokolo} \tbHM{} \tbPH{kólókóló} \tbPS{adv.~} \tbGE{light} \tbVA{koloŋkoloŋ}
\tbLX{koloŋkoloŋ} \tbHM{} \tbPH{kólóŋkólóŋ} \tbPS{adv.~} \tbGE{watery; diluted e.g porridge} \tbVA{kolokolo}
\tbLX{kolɔ} \tbHM{} \tbPH{kʊ́là} \tbPS{v.~} \tbGE{to be unperturbed; to exhibit coolness. (used with negative marker)} \tbXV{A bie kaŋa ba kolɔ kyɛ kaŋa nyegnyege.} \tbXE{One of the children is calm but one is the boisterous type.} \tbPD{-, -, -, -, -, - } \tbVA{kola}
\tbLX{koltaŋaa} \tbHM{} \tbPH{kʊ̀ltáŋáà} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{koltaŋema} \tbTP{koltaŋemɛɛ} 
\tbLX{kolyaaraa} \tbHM{} \tbPH{kôlyááráá} \tbPS{n.~} \tbGE{a type of bird of the suborder alcedines that feeds on fish} \tbOP{kolyaare} \tbTP{kolyaarɛɛ} \tbVA{kɔn-yaara}
\tbLX{komaale} \tbHM{} \tbPH{kòmààlɪ̀} \tbPS{n.~} \tbGE{funeral rites} 
\tbLX{komaaroŋ} \tbHM{} \tbPH{kòmáárʊ́ŋ} \tbPS{n.~} \tbGE{a funeral where the corpse is not yet buried} \tbOP{komaara} \tbTP{komaarɛɛ} 
\tbLX{Komaase} \tbHM{} \tbPH{kʊ̀máásɪ́} \tbPS{n.~} \tbGE{main city in the Ashanti Region of Ghana} \tbOP{Komaasere} \tbTP{Komaaserɛɛ} 
\tbLX{komakoma} \tbHM{} \tbPH{kʊ̀màkʊ̀mà} \tbPS{adj.~} \tbGE{mild e.g. drink, alcohol} 
\tbLX{komboroŋ} \tbHM{} \tbPH{kʊ̀mbʊ̀rʊ̀ŋ} \tbPS{adj.~} \tbGE{wrinkled} 
\tbLX{komiri} \tbHM{} \tbPH{kòmírí} \tbPS{n.~} \tbGE{a string or strip of cloth tied round the wrist or waist of the bereaved during the funeral rites} \tbVA{koganaa}
\tbLX{kommiruu} \tbHM{} \tbPH{kómmírúú} \tbPS{n.~} \tbSN \tbGE{tomato} \tbSN \tbGE{eggplant} \tbSG{kommiruu} \tbOP{kommie} \tbTP{kommiiree} 
\tbLX{komoɔreŋ} \tbHM{} \tbPH{kómʊ́ɔ́rɪ́ŋ} \tbPS{n.~} \tbGE{a type of funeral dirge} 
\tbLX{kona} \tbHM{} \tbPH{kʊ́ná} \tbPS{adv.~} \tbGE{in large quantity, immeasurable} \tbXV{A kyi ba taa kona, a koŋ kpɛ a bugo.} \tbXE{There is a great deal of millet; the granary cannot contain it.} 
\tbLX{konaabo} \tbHM{} \tbPH{kònáábʊ́} \tbPS{n.~} \tbGE{cow meant for funeral rites} \tbOP{konii} \tbTP{koniinee} 
\tbLX{konɛɛ konɛɛ} \tbHM{} \tbPH{kʊ́nɛ́ɛ́kʊ́nɛ́ɛ́} \tbPS{adv.~} \tbGE{on and on in the same direction} \tbXV{A pɔgelee baŋe la seɛre, o deɛ taaɛ konɛɛ, konɛɛ o naŋ erɛ.} \tbXE{The girl knows how to dance, she is just dancing gracefully along.} 
\tbLX{konkon} \tbHM{} \tbPH{kʊ́nkʊ́n} \tbPS{adv.~} \tbGE{completely, flat (of falling)} 
\tbLX{kon-kon-kon} \tbHM{} \tbPH{kʊ́n-kʊ́n-kʊ́n} \tbPS{adv.~} \tbGE{to fall helplessly} 
\tbLX{konne} \tbHM{} \tbPH{kʊ́nnɪ̀} \tbPS{v.~} \tbGE{to make a short audible exhalation through the nose in to stop inhaling bad smell} \tbPD{konnɛɛ, konna, konna, konema, konnaa, konema } 
\tbLX{konni} \tbHM{} \tbPH{kònní} \tbPS{n.~} \tbGE{cry} \tbTP{konnee} 
\tbLX{konni} \tbHM{} \tbPH{kónnì} \tbPS{v.~} \tbGE{to catch a falling object in midair (to prevent from falling)} \tbPD{konnee, konno, konno, konimo, konnaa, konimo } 
\tbLX{konno} \tbHM{} \tbPH{kónnò} \tbPS{n.~} \tbGE{mouse nest} \tbVA{konzagera}
\tbLX{konobara} \tbHM{} \tbPH{kónóbàrá} \tbPS{n.~} \tbGE{lamentation} 
\tbLX{konono} \tbHM{} \tbPH{kʊ́nʊ́nʊ́} \tbPS{adv.~} \tbSN \tbGE{smooth (flour)} \tbSN \tbGE{soft (body)} \tbSN \tbGE{unasuming (behaviour)} 
\tbLX{konoɔ} \tbHM{} \tbPH{kónʊ́ɔ́} \tbPS{n.~} \tbGE{final funeral of the Sisala (an ethnic group in the UWR)} \tbOP{konoore} \tbTP{konoorɛɛ} 
\tbLX{kon-ɔɔnɛnesoba} \tbHM{} \tbPH{kʊ́nɔ́ɔ́nɛ́nɪ̀sʊ́bá} \tbPS{n.~} \tbGE{vegetarian} \tbOP{konɔɔnɛnedeme} \tbTP{konɔɔnɛnedemee} \tbVA{kon-ɔɔnɛne}
\tbLX{konzagera} \tbHM{} \tbPH{kónzàgrà} \tbPS{n.~} \tbGE{mouse nest} \tbTP{konzagerɛɛ} \tbVA{konno, konzage}
\tbLX{koŋ} \tbHM{} \tbPH{kʊ̀ŋ} \tbPS{negative marker.~} \tbGE{not} \tbXV{N koŋ gaa be zenɛ.} \tbXE{I will not go there today.} 
\tbLX{koŋi} \tbHM{} \tbPH{kǒŋ} \tbPS{adj.~} \tbGE{mangled} \tbOP{konni} 
\tbLX{koŋi} \tbHM{} \tbPH{kòŋ} \tbPS{v.~} \tbSN \tbGE{to cry, to lament} \tbSN \tbGE{to crow (rooster)} \tbPD{koŋee, kono, kono, konimo, konaa, konimo } 
\tbLX{koŋi} \tbHM{} \tbPH{kòŋ} \tbPS{v.~} \tbGE{to ferment} \tbPD{koŋee, kono, kono, konimo, konaa, konimo } 
\tbLX{koŋkogiri} \tbHM{} \tbPH{kóŋkógrí} \tbPS{n.~} \tbGE{boundary (of a field)} \tbOP{koŋkobo} \tbTP{koŋkogiree} 
\tbLX{koŋkola} \tbHM{} \tbPH{kʊ́ŋkʊ́lá} \tbPS{n.~} \tbGE{farming field} \tbOP{koŋkolle} \tbTP{koŋkolɛɛ/koŋkollɛɛ} \tbVA{koŋkolɔ}
\tbLX{koŋkombiri} \tbHM{} \tbPH{kǒŋkǒmbírì} \tbPS{n.~} \tbGE{funeral dirge text} \tbSG{koŋkombiruu} \tbOP{koŋkombie} \tbTP{koŋkombiiree} 
\tbLX{koŋkommoola} \tbHM{} \tbPH{kòŋkǒmmʊ́ʊ́lá} \tbPS{n.~} \tbGE{incessant crying and for no apparent reason e.g a child} \tbVA{koŋkommoolɔ}
\tbLX{koŋŋmɛbaa} \tbHM{} \tbPH{kʊ́ŋŋmɛ́báá} \tbPS{n.~} \tbGE{back of the head of the dog} 
\tbLX{koŋ-ɔɔ-moɔ} \tbHM{} \tbPH{kʊ́ŋ-ɔ́ɔ́-mʊ́ɔ́} \tbPS{n.~} \tbGE{dog} 
\tbLX{kookeree} \tbHM{} \tbPH{kòòkéréé} \tbPS{n.~} \tbGE{one's best attempt; effort} 
\tbLX{koone} \tbHM{} \tbPH{kʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbGE{to groan (in pain)} \tbPD{koonɛɛ, koonɔ/koona, koonɔ/koona, koonema, koonaa, - } 
\tbLX{kooraa} \tbHM{} \tbPH{kʊ́ʊ́ràà} \tbPS{n.~} \tbGE{umbilical hernia} \tbOP{koore} \tbTP{koorɛɛ} \tbVA{kpograa, kpogre, kpogrɛɛ}
\tbLX{koore} \tbHM{} \tbPH{kʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to snore} \tbPD{koorɛɛ, koorɔ/koora, koorɔ/koora, kooreba, kooraa, kooreba } \tbVA{korŋzaa}
\tbLX{koore} \tbHM{} \tbPH{kʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to collect or move a load at once} \tbPD{koorɛɛ, koorɔ/koora, koorɔ/koora, kooreba, kooraa, kooreba } \tbVA{korŋzaa}
\tbLX{koɔ} \tbHM{} \tbPH{kʊ̀ɔ̀} \tbPS{v.~} \tbGE{to have an upset stomach, intestinal illness} \tbPD{koɔɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\tbLX{ko᷉ɔ} \tbHM{} \tbPH{kʊ́ɔ́} \tbPS{n.~} \tbSN \tbGE{water} \tbSN \tbGE{pus} \tbOP{koɔnɛɛ} \tbTP{koɔnoo} 
\tbLX{ko᷉ɔ} \tbHM{} \tbPH{kʊ́᷅ɔ́} \tbPS{adj.~} \tbGE{water-like, liquid} 
\tbLX{koɔbo} \tbHM{} \tbPH{kʊ́ɔ́bʊ̀} \tbPS{n.~} \tbGE{the act of farming} \tbTP{kɔrɛɛ} 
\tbLX{koɔle} \tbHM{} \tbPH{kʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to gather; to bring together} \tbPD{koɔlɛɛ, koɔlɔ, koɔlɔ, koɔleba, koɔlaa, koɔlɔ } 
\tbLX{koɔle} \tbHM{} \tbPH{kʊ́ɔ́lɪ̀} \tbPS{v.~} \tbGE{to get sick} \tbPD{koɔlɛɛ, koɔlɔ, koɔlɔ, koɔleba, koɔlaa, koɔleba } 
\tbLX{koɔle sɔgelaa} \tbHM{} \tbPH{kʋ́ɔ́lɪ́sɔ̀gɪ́làá} \tbPS{n.~} \tbGE{fertile land} 
\tbLX{koɔlɔ} \tbHM{} \tbPH{kʊ̀ɔ̀lɔ̀} \tbPS{n.~} \tbSN \tbGE{farmland} \tbSN \tbGE{environment} \tbOP{koɔle} \tbTP{koɔlɛɛ} \tbVA{koɔla}
\tbLX{koɔraa} \tbHM{} \tbPH{kʊ́ɔ́ràà} \tbPS{n.~} \tbGE{farmer} \tbOP{koɔrebɔ} \tbTP{n/a} 
\tbLX{koɔre} \tbHM{} \tbPH{kʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to sell} \tbPD{koɔrɛɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\tbLX{koɔre} \tbHM{} \tbPH{kʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to scrape, to scratch} \tbPD{koɔrɛɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\tbLX{koɔre} \tbHM{} \tbPH{kʊ́ɔ́rɪ́} \tbPS{v.~} \tbGE{to tap} \tbPD{koɔrɛɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\tbLX{koɔre} \tbHM{} \tbPH{kʊ̀ɔ́rɪ̀} \tbPS{n.~} \tbGE{group of stars; galaxy} \tbSG{koɔroo} \tbOP{koɔrɛɛ} 
\tbLX{koɔre} \tbHM{1} \tbPH{kʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{gourd} \tbOP{kɔɛ} \tbTP{kɔrɛɛ} 
\tbLX{koɔre} \tbHM{2} \tbPH{kʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{residue of final beer} \tbTP{koɔrɛɛ} 
\tbLX{kopeɛ} \tbHM{} \tbPH{kòpɪ́ɛ́} \tbPS{n.~} \tbGE{ground immediately surrounding the stage of the corpse} \tbOP{kopeere} \tbTP{kopeerɛɛ} 
\tbLX{ko-pɔge} \tbHM{} \tbPH{kò-pɔ́gɪ́} \tbPS{n.~} \tbGE{widow} \tbOP{ko-pɔgeba} \tbTP{ko-pɔgrɛɛ} 
\tbLX{ko-puoraa} \tbHM{} \tbPH{kò-pùòràá} \tbPS{n.~} \tbGE{belated condolence (usually from someone who misses a funeral)} 
\tbLX{koraa} \tbHM{} \tbPH{kòràá} \tbPS{n.~} \tbGE{village; bush settlement} \tbOP{korri} \tbTP{korree} 
\tbLX{kore} \tbHM{} \tbPH{kʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to be old (not young)} \tbSN \tbGE{to be old (not new)} \tbPD{korɛɛ, kora, kora, koreba, koraa, koreba } 
\tbLX{kore} \tbHM{} \tbPH{kʊ́rɪ́} \tbPS{adj.~} \tbGE{old} \tbOP{kora} \tbTP{korɛɛ} 
\tbLX{korekore} \tbHM{} \tbPH{kʊ̀rɪ̀kʊ́rɪ́} \tbPS{v.~} \tbGE{to break into bits and pieces (grass)} \tbPD{korkorɛɛ, korekora, korekora, korkoreba, korkoraa, - } 
\tbLX{korekore} \tbHM{} \tbPH{kʊ̀rɪ̀kʊ̀rɪ̀} \tbPS{adv.~} \tbGE{many, in great numbers} 
\tbLX{koremaŋ} \tbHM{} \tbPH{kʊ̀rɪ̀màŋ} \tbPS{adj.~} \tbGE{unkempt} \tbVA{koremɔŋ}
\tbLX{korepere} \tbHM{} \tbPH{kʊ́rɪ́pɪ́rɪ́} \tbPS{n.~} \tbGE{a type of monkey with thick fur} \tbOP{korepeɛ} \tbTP{koreperɛɛ} 
\tbLX{korewayiri} \tbHM{} \tbPH{kʊ̀rɪ̀wàyírí} \tbPS{n.~} \tbGE{one who has has spent all their youth outside their original homeland} 
\tbLX{kori} \tbHM{} \tbPH{kòrì} \tbPS{v.~} \tbGE{to embrace, to hug} \tbPD{koree, koro, koro, koro, koraa, koro } 
\tbLX{kori} \tbHM{} \tbPH{kórí} \tbPS{n.~} \tbGE{laziness} \tbTP{koree} 
\tbLX{kori} \tbHM{} \tbPH{kórí} \tbPS{n.~} \tbGE{difficulty, suffering} \tbTP{koree} 
\tbLX{kori} \tbHM{} \tbPH{kórí} \tbPS{adv.~} \tbGE{the manner of a sound from hitting a hard surface} 
\tbLX{korimo} \tbHM{} \tbPH{kòrímò} \tbPS{n.~} \tbGE{a type of lizard believed to be poisonous} \tbOP{korri} \tbTP{korree} 
\tbLX{koringyoŋ} \tbHM{} \tbPH{kóríndzóŋ} \tbPS{n.~} \tbGE{violin} \tbOP{koringyonni} \tbTP{koringyonnee} 
\tbLX{koringyoŋ} \tbHM{} \tbPH{koríngyóŋ} \tbPS{n.~} \tbGE{violin, (harp)} \tbOP{koringyonni} \tbTP{koringyonnee} 
\tbLX{koro} \tbHM{} \tbPH{kòró} \tbPS{n.~} \tbGE{pot for storing TZ (thick paste made from flour)} \tbOP{korri} \tbTP{koree/korree} 
\tbLX{korodie} \tbHM{} \tbPH{kòródìé} \tbPS{n.~} \tbGE{kitchen} \tbOP{koroderi} \tbTP{koroderee/korodrree} 
\tbLX{korombo} \tbHM{} \tbPH{kórómbò} \tbPS{adv.~} \tbGE{to empty totally, to finish with completely} 
\tbLX{koroŋ} \tbHM{} \tbPH{kʊ́rʊ́ŋ} \tbPS{adj.~} \tbGE{a long time ago} 
\tbLX{koroso} \tbHM{} \tbPH{kórósó} \tbPS{adv.~} \tbGE{lean and dry} \tbXV{A balee ba dire lɛnso ka o waa a koroso lɛ.} \tbXE{The puppy does not eat well and that is why it is so lean and bony.} 
\tbLX{ko suuri} \tbHM{} \tbPH{kʊ́ súúrí} \tbPS{v.~} \tbGE{to arouse someone's anger} \tbPD{koɛ suuri, koora suuri, -, -, -, - } 
\tbLX{kotoloŋ} \tbHM{} \tbPH{kòtʊ́lʊ́ŋ} \tbPS{n.~} \tbGE{the early part of mourning at the funeral} \tbOP{kotolɔ/kotola} 
\tbLX{kotontona} \tbHM{} \tbPH{kòtʊ́ntʊ́ná} \tbPS{n.~} \tbGE{messenger who announces funeral rites} \tbOP{kotontonema} \tbTP{kotontomɛɛ} 
\tbLX{kotuo} \tbHM{} \tbPH{kòtúó} \tbPS{n.~} \tbGE{a funeral where mourning is in a frenzy} \tbOP{kotuuri} \tbTP{kotuuree} 
\tbLX{kotuosobɔ} \tbHM{} \tbPH{kòtúósʊ́bá} \tbPS{n.~} \tbGE{bereaved person (close relative)} \tbOP{kotuodeme} \tbVA{kotuosoba}
\tbLX{kovuoro} \tbHM{} \tbPH{kòvúóró} \tbPS{n.~} \tbGE{condolences by well wishers who could not attend the funeral rites} \tbTP{kovuoree} 
\tbLX{Koware} \tbHM{} \tbPH{kʊ́wárɪ} \tbPS{n.~} \tbGE{homeland of the Kowarɛe clan} 
\tbLX{Kowaroo} \tbHM{} \tbPH{kʊ́wárʊ́ʊ̊} \tbPS{n.~} \tbGE{a member of the Koware clan} \tbOP{Kowarɛɛ} 
\tbLX{koyeŋi} \tbHM{} \tbPH{kòyěŋ} \tbPS{n.~} \tbGE{wake keeping (funeral)} \tbOP{koyenni} \tbTP{koyenee} 
\tbLX{koyiri} \tbHM{} \tbPH{kóyírí} \tbPS{n.~} \tbGE{house of a bereaved family} \tbOP{koyie} \tbTP{koyiree} 
\tbLX{kɔ} \tbHM{} \tbPH{kɔ̌} \tbPS{v.~} \tbSN \tbGE{to hoe} \tbSN \tbGE{to cultivate, to farm} \tbPD{kɔɛ, koɔrɔ, koɔrɔ, koɔreba, koɔraa, koɔreba } 
\tbLX{Kɔbenɛ} \tbHM{} \tbPH{kɔ́bɪ́nɛ́} \tbPS{n.~} \tbGE{an annual festival celebrated by the people of Lawra} \tbTP{kɔbenɛɛ} 
\tbLX{kɔge} \tbHM{} \tbPH{kɔ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to be near, to be close to} \tbSN \tbGE{to put by the side to support (pillow)} \tbSN \tbGE{to be lean} \tbPD{kɔgɛɛ, kɔgrɔ, kɔgrɔ, kɔgreba, kɔgraa, - } 
\tbLX{kɔge} \tbHM{} \tbPH{kɔ́gɪ́} \tbPS{v.~} \tbGE{to boil ground malt at the the first stage in the preparation of pito} \tbPD{kɔgɛɛ, kɔgera, kɔgera, kɔgereba, kɔgeraa, kɔgereba } 
\tbLX{kɔgele} \tbHM{} \tbPH{kɔ̀glɪ̀} \tbPS{v.~} \tbSN \tbGE{to keep company} \tbSN \tbGE{to give aid} \tbPD{kɔgelɛɛ, kɔgela, kɔgela, kɔgeleba, kɔgelaa, kɔgeleba } 
\tbLX{kɔgelɔ} \tbHM{} \tbPH{kɔ̀glɔ̀} \tbPS{n.~} \tbGE{a type of fetish or god} \tbTP{kɔgelɛɛ} 
\tbLX{kɔgeraa} \tbHM{} \tbPH{kɔ́gráá} \tbPS{n.~} \tbGE{raised lump on the floor of a room used as a seat} \tbOP{kɔgere} \tbTP{kɔgerɛɛ} 
\tbLX{kɔkɔbaŋ} \tbHM{} \tbPH{kɔ̀kɔ́báŋ} \tbPS{n.~} \tbGE{neck bangle} \tbOP{kɔkɔbanne} \tbTP{kɔkɔbannɛɛ} \tbVA{kɔkɔbɔŋ}
\tbLX{kɔkɔkpɔgere} \tbHM{} \tbPH{kɔ̀kɔ́kpɔ́grɪ̀} \tbPS{n.~} \tbGE{goitre} \tbOP{kɔkɔkpɔgɔ} \tbTP{kɔkɔkpɔgerɛɛ} 
\tbLX{kɔkɔlaa} \tbHM{} \tbPH{kɔ̀kɔ̀láà} \tbPS{n.~} \tbGE{cat} \tbOP{kɔkɔlle} \tbTP{kɔkɔllɛɛ} \tbVA{kɔlaa}
\tbLX{kɔkɔlɛre} \tbHM{} \tbPH{kɔ̀kɔ́lɛ̀rɪ́} \tbPS{n.~} \tbGE{necklace} \tbSG{kɔkɔlɛroo} \tbOP{kɔkɔlɛɛ} 
\tbLX{kɔkɔliŋe} \tbHM{} \tbPH{kɔ̀kɔ́líŋé} \tbPS{n.~} \tbGE{velum} \tbOP{kɔkɔlinni} \tbTP{kɔkɔlinnee} 
\tbLX{kɔkɔllee} \tbHM{} \tbPH{kɔ̀kɔ̀llée} \tbPS{n.~} \tbGE{kitten} \tbOP{kɔkɔlbilii} \tbTP{kɔkɔlbillee} 
\tbLX{kɔkɔlɔŋgoɔne} \tbHM{} \tbPH{kɔ̀kɔ́lɔ̀ŋgʊ́ɔ́nɪ́} \tbPS{n.~} \tbGE{the trachea} \tbOP{kɔkɔlɔŋgoɔmɔ} \tbTP{kɔkɔlɔŋgoɔnɛɛ} 
\tbLX{kɔkɔlweɛmɛ} \tbHM{} \tbPH{kɔ̀kɔ̀lwɪ́ɛ̀mɛ́} \tbPS{n.~} \tbGE{wild cat} \tbOP{kɔkɔlweɛne} \tbTP{kɔkɔlweɛnɛɛ} \tbVA{kɔlweɛmɛ}
\tbLX{kɔkɔlwoɔ} \tbHM{} \tbPH{kɔ̀kɔ̌lwʊ́ɔ́} \tbPS{n.~} \tbGE{bag made from cat hide} \tbOP{kɔkɔlwoore} \tbTP{kɔkɔlwoorɛɛ} 
\tbLX{kɔkɔmaraa} \tbHM{} \tbPH{kɔ̀kɔ́máráá} \tbPS{n.~} \tbGE{tight-fitting necklace} \tbOP{kɔkɔmarre} \tbTP{kɔkɔmarrɛɛ} 
\tbLX{kɔkɔraa} \tbHM{} \tbPH{kɔ́kɔ́ràà} \tbPS{n.~} \tbGE{one who does dirty things (obscene language)} \tbOP{kɔkɔrre} 
\tbLX{kɔkɔre} \tbHM{} \tbPH{kɔ́kɔ́rɪ̀} \tbPS{n.~} \tbSN \tbGE{throat} \tbSN \tbGE{voice} \tbSN \tbGE{language} \tbOP{kɔkɔrɛɛ/kɔkɔɛ} \tbTP{kɔkɔrrɛɛ} 
\tbLX{kɔkɔreyɛlɛ} \tbHM{} \tbPH{kɔ́kɔ̀rɪ̀yɛ́lɛ́} \tbPS{n.~} \tbGE{obscene things} 
\tbLX{kɔkɔroŋ} \tbHM{} \tbPH{kɔ́kɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{the act of being uncouth} 
\tbLX{kɔkɔsɔgelaa} \tbHM{} \tbPH{kɔ̀kɔ́sɔ̀glàà} \tbPS{n.~} \tbGE{breastbone} \tbOP{kɔkɔsɔgele} \tbTP{kɔkɔsɔgelɛɛ} 
\tbLX{kɔkɔvɔlaa} \tbHM{} \tbPH{kɔ́kɔ́vɔ́làà} \tbPS{n.~} \tbGE{esophagus, gullet} \tbOP{kɔkɔvɔlle} \tbTP{kɔkɔvɔllɛɛ} 
\tbLX{kɔkɔyoraa} \tbHM{} \tbPH{kɔ̀kɔ́yòráá} \tbPS{n.~} \tbGE{larynx, voice box, Adam's apple} \tbSG{kɔkɔyoruu} \tbOP{kɔkɔyorri} \tbTP{kɔkɔyoree} 
\tbLX{kɔlaa} \tbHM{} \tbPH{kɔ̀láà} \tbPS{n.~} \tbGE{cat} \tbOP{kɔlle} \tbTP{kɔllɛɛ} \tbVA{kɔkɔlaa}
\tbLX{kɔlbaa} \tbHM{} \tbPH{kɔ́lbáá} \tbPS{n.~} \tbGE{bottle} \tbOP{kɔlbaare} \tbTP{kɔlbaarɛɛ} 
\tbLX{kɔle} \tbHM{} \tbPH{kɔ́lɪ́} \tbPS{v.~} \tbGE{to be loose within something; to be too small for} \tbXV{A dɔɔ kuree kɔlɔ la a o bidɔɔ seɛ.} \tbXE{The man's pair of shorts is too large for his son.} \tbPD{kɔlɛɛ, kɔlɔ, kɔlɔ, kɔlebɔ, kɔlaa, kɔlebɔ } 
\tbLX{kɔlema} \tbHM{} \tbPH{kɔ̀lɪ̀má} \tbPS{n.~} \tbGE{well} \tbOP{kɔlle} \tbTP{kɔlɛɛ} \tbVA{kolaa (kʊlaa), kolli}
\tbLX{kɔleŋkɔ} \tbHM{} \tbPH{kɔ́lɪ́ŋkɔ́} \tbPS{n.~} \tbGE{a type of gourd with a long neck} \tbOP{kɔleŋkɔre} \tbTP{kɔleŋkɔrɛɛ} 
\tbLX{kɔl-kɔl-kɔl} \tbHM{} \tbPH{kɔ̀l-kɔ̀l-kɔ̀l} \tbPS{adj.~} \tbGE{in a long line e.g a column of school children} 
\tbLX{kɔlle} \tbHM{} \tbPH{kɔ̀llɪ̀} \tbPS{v.~} \tbGE{to cluck} \tbPD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlebɔ, kɔllaa, kɔlebɔ } 
\tbLX{kɔlle} \tbHM{} \tbPH{kɔ̀llɪ̀} \tbPS{v.~} \tbGE{to roam} \tbPD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlebɔ, kɔllaa, kɔlebɔ } 
\tbLX{kɔlle} \tbHM{} \tbPH{kɔ́llɪ́} \tbPS{v.~} \tbGE{to gallop gracefully} \tbPD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlbɔ, kɔllaa, - } 
\tbLX{kɔlle} \tbHM{} \tbPH{kɔ́llɪ́} \tbPS{v.~} \tbGE{to smooth e.g. potter's pot} \tbPD{kɔllɛɛ, kɔllɔ, kɔllɔ, kɔlbɔ, kɔllaa, - } 
\tbLX{kɔlɔkɔlɔ} \tbHM{1} \tbPH{kɔ̀lɔ̀kɔ̀lɔ̀} \tbPS{adv.~} \tbGE{to begin to brighten e.g day breaking} 
\tbLX{kɔlɔkɔlɔ} \tbHM{2} \tbPH{kɔ̀lɔ̀kɔ̀lɔ̀} \tbPS{adv.~} \tbGE{not firm (object smaller than its fitting)} 
\tbLX{kɔmaaroŋ} \tbHM{} \tbPH{kɔ̀mààrʊ́ŋ} \tbPS{n.~} \tbGE{freshly weeded or hoed-up place} \tbOP{kɔmaara} \tbTP{kɔmaarɛɛ} 
\tbLX{kɔmbanna} \tbHM{} \tbPH{kɔ̀mbánná} \tbPS{n.~} \tbGE{swimmer} \tbOP{kɔmbanema} \tbTP{kɔmbanemɛɛ} 
\tbLX{kɔmbiri} \tbHM{} \tbPH{kɔ́mbírí} \tbPS{n.~} \tbGE{drop of water} \tbOP{kɔmbie} \tbTP{kɔbiiree} 
\tbLX{kɔmbɔ} \tbHM{} \tbPH{kɔ̀mbɔ́} \tbPS{n.~} \tbGE{dialog, conversation} \tbTP{kɔmbɔrɛɛ} 
\tbLX{kɔmbɔge} \tbHM{} \tbPH{kɔ̀mbɔ̀gɪ̀} \tbPS{n.~} \tbGE{water-way} \tbVA{kɔmbɔgeraa}
\tbLX{kɔmbɔgeraa} \tbHM{} \tbPH{kɔ̀mbɔ́gráá} \tbPS{n.~} \tbGE{waterway} \tbOP{kɔmbɔgre} \tbTP{kɔmbɔgrɛɛ} \tbVA{kɔmbɔge}
\tbLX{kɔmbɔnne} \tbHM{} \tbPH{kɔ̀mbɔnnɪ́̀} \tbPS{n.~} \tbGE{Akan language} \tbTP{kɔmbɔnnɛɛ} 
\tbLX{kɔmbɔŋenɔ} \tbHM{} \tbPH{kɔ́mbɔ́ŋɪ́nɔ́} \tbPS{n.~} \tbGE{swimmer} \tbOP{kɔmbɔŋenemɔ} \tbTP{kɔmbɔŋenemɛɛ} \tbVA{kɔmbɔŋena}
\tbLX{kɔmbɔŋɔ} \tbHM{} \tbPH{kɔ̀mbɔ̀ŋɔ́} \tbPS{n.~} \tbGE{Akan person (one who can swim)} \tbOP{kɔmbɔnne} \tbTP{kɔmbɔnnɛɛ} 
\tbLX{kɔmbuli} \tbHM{} \tbPH{kɔ̀mbúlí} \tbPS{n.~} \tbGE{period of water shortage e.g. during drought} \tbTP{kɔmbulee} 
\tbLX{kɔmmaaroŋ} \tbHM{} \tbPH{kɔ̌mmáárʊ́ŋ} \tbPS{n.~} \tbGE{cold water} 
\tbLX{kɔmmaaroŋ} \tbHM{} \tbPH{kɔ̌mmáárʊ́ŋ} \tbPS{n.~} \tbGE{trouble free; peace} 
\tbLX{kɔmmiiiruŋ} \tbHM{} \tbPH{kɔ̀mmíírûŋ} \tbPS{n.~} \tbGE{sour water} \tbTP{kɔmmiiree} 
\tbLX{kɔmmiri} \tbHM{} \tbPH{kɔ̀mmírì} \tbPS{n.~} \tbGE{current (running water)} \tbOP{kɔmmie} \tbTP{kɔmmiiree} \tbVA{kombiri}
\tbLX{kɔmɔkɔmɔ} \tbHM{} \tbPH{kɔ̀mɔ̀kɔ̀mɔ̀} \tbPS{adv.~} \tbGE{not hard} 
\tbLX{kɔmpaala} \tbHM{} \tbPH{kɔ́mpáálá} \tbPS{n.~} \tbGE{new water e.g bottled water for sale} \tbTP{kɔmpaalɛɛ} 
\tbLX{kɔmpare} \tbHM{} \tbPH{kɔ́mpárɪ́} \tbPS{n.~} \tbSN \tbGE{the under part of water, residue} \tbSN \tbGE{the southern part of Ghana} \tbOP{kɔmparɛɛ} 
\tbLX{kɔmpiiri} \tbHM{} \tbPH{kɔ́mpíírí} \tbPS{n.~} \tbGE{floodwaters} \tbOP{kɔmpie} \tbTP{kɔmpiree/kɔmpiiree} 
\tbLX{kɔmpɔge} \tbHM{} \tbPH{kɔ̀mpɔ̀gɪ̀} \tbPS{n.~} \tbGE{abundant waters, flood} \tbOP{kɔmpɔgre} \tbTP{kɔmpɔgrɛɛ} \tbVA{kɔmpu᷉o}
\tbLX{kɔmpu᷉o} \tbHM{} \tbPH{kɔmpu᷉o} \tbPS{adv.~} \tbGE{period of excessive water e.g during the peak of heavy rains} \tbVA{kɔmpɔge}
\tbLX{kɔndaare} \tbHM{} \tbPH{kɔ̀ndáárɪ̀} \tbPS{n.~} \tbGE{waves} \tbTP{kɔndaarɛɛ} 
\tbLX{kɔndaare} \tbHM{} \tbPH{kɔ̀ndàárɪ̀} \tbPS{n.~} \tbGE{water for sale} \tbTP{kɔndaarɛɛ} 
\tbLX{kɔndazugiri} \tbHM{} \tbPH{kɔ̀ndàzùgrí} \tbPS{n.~} \tbSN \tbGE{a type of wild dwarf tree, short and stout with complex branches} \tbSN \tbGE{the yellow fruit of this plant (edible)} \tbSG{kɔndazugiruu} \tbOP{kɔndazugo} \tbTP{kɔndazugiree} \tbVA{dazugri}
\tbLX{kɔndazugovu᷈u᷈} \tbHM{} \tbPH{kɔ̀ndàzúgóvù᷈ú᷈} \tbPS{n.~} \tbSN \tbGE{furnace} \tbSN \tbGE{hell} \tbTP{kɔndazugovuunee} 
\tbLX{kɔndoge} \tbHM{} \tbPH{kɔ̌ndʊ́gɪ̀} \tbPS{n.~} \tbGE{water pot} \tbOP{kɔndogre} \tbTP{kɔndogrɛɛ} 
\tbLX{kɔndugzie} \tbHM{} \tbPH{kɔ̌ndúgzíé} \tbPS{n.~} \tbGE{swimming pool} \tbOP{kɔndugziiri} \tbTP{kɔndugziiree} 
\tbLX{Kɔnduoraa} \tbHM{} \tbPH{kɔ̌ndúóràá} \tbPS{n.~} \tbGE{a river in the Upper West Region of Ghana} \tbOP{kɔnduori} \tbTP{kɔnduoree} 
\tbLX{kɔnkyɔkyɔgere} \tbHM{} \tbPH{kɔ̀nʧɔ́ʧɔ̀grɪ́} \tbPS{n.~} \tbGE{swamp} \tbTP{kɔnkyɔkyɔgerɛɛ} 
\tbLX{kɔnkyɔrɛɛ} \tbHM{} \tbPH{kɔ̌nkyɔ̀rɛ́ɛ́} \tbPS{n.~} \tbGE{a type of bath where soap and sponge are not used} \tbOP{kɔnkyɔrre} \tbTP{kɔnkyɔrrɛɛ} 
\tbLX{kɔnkyuuri} \tbHM{} \tbPH{kɔ̀nʧúúrì} \tbPS{n.~} \tbGE{waterfall} \tbTP{kɔnkyuuree} 
\tbLX{kɔnnaŋ} \tbHM{} \tbPH{kɔ́nnáŋ} \tbPS{n.~} \tbGE{water scarcity} \tbTP{kɔnnannɛɛ} 
\tbLX{kɔnne} \tbHM{} \tbPH{kɔ̀nnɪ́} \tbPS{n.~} \tbGE{leprosy} \tbOP{kɔnne} \tbTP{kɔnnɛɛ} 
\tbLX{kɔnnimiri} \tbHM{} \tbPH{kɔ̂nnímírì} \tbPS{n.~} \tbSN \tbGE{spring} \tbSN \tbGE{water hole} \tbSG{kɔnnimiiruu} \tbOP{kɔnnimie} \tbTP{kɔnnimiiree} \tbVA{bulnimiri}
\tbLX{kɔnnoŋ} \tbHM{} \tbPH{kɔ́nnʊ́ŋ} \tbPS{n.~} \tbGE{the condition of having leprosy} \tbVA{kɔnnoo}
\tbLX{kɔnnoɔre} \tbHM{} \tbPH{kɔ̀nnʊ́ɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{beach, coastline} \tbSN \tbGE{edge} \tbOP{kɔnnɔɛ} \tbTP{kɔnnɔrɛɛ} 
\tbLX{kɔnnyuuri} \tbHM{} \tbPH{kɔ̀nɲúúrì} \tbPS{n.~} \tbGE{thirst} \tbSG{kɔnyuuree} 
\tbLX{kɔnnyuuri} \tbHM{} \tbPH{kɔ́nɲúúrí} \tbPS{n.~} \tbGE{potable water, water meant for drinking} 
\tbLX{kɔnoɔre} \tbHM{} \tbPH{kɔ́nʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{a farmer's portion during farming} \tbOP{kɔnɔɛ} \tbTP{kɔnɔrɛɛ} 
\tbLX{kɔnɔŋmaara} \tbHM{} \tbPH{kɔ́nɔ́ŋmáárá} \tbPS{n.~} \tbGE{leader of working farmers in the farm} \tbOP{kɔnɔŋmaareba} \tbTP{kɔnɔŋmaarebɛɛ} 
\tbLX{kɔnɔtɛne} \tbHM{} \tbPH{kɔ́nɔ́tɛ́nɪ́} \tbPS{n.~} \tbGE{earth or soil used to mark out the boundaries of portions of land for weeding} \tbTP{kɔnɔtɛnɛɛ} 
\tbLX{kɔnɔtuuro} \tbHM{} \tbPH{kɔ́nɔ́túúró} \tbPS{n.~} \tbGE{deputy leader of working farmers in the farm} \tbOP{kɔnɔtuuribo} \tbTP{kɔnɔtuuribee} 
\tbLX{kɔnsa᷈a} \tbHM{} \tbPH{kɔ̀nsà᷈à} \tbPS{n.~} \tbGE{the act of polluting water} 
\tbLX{kɔnsaaloŋ} \tbHM{} \tbPH{kɔ́nsáálʊ́ŋ} \tbPS{n.~} \tbGE{plain fluid from a swelling (may be mixed with blood)} \tbTP{kɔnsaalɛɛ} 
\tbLX{kɔnsaane} \tbHM{} \tbPH{kɔ̌nsáánɪ́} \tbPS{n.~} \tbGE{water with marshed poridge} \tbTP{kɔnsaanɛɛ} 
\tbLX{kɔnso} \tbHM{} \tbPH{kɔ̀nsò} \tbPS{n.~} \tbGE{the act of bathing} \tbTP{kɔnsoree} 
\tbLX{kɔnsobɔɔdoaa} \tbHM{} \tbPH{kɔ̀nsòbɔ̀ɔ̀dʊ́áá} \tbPS{n.~} \tbGE{bath towel} \tbOP{kɔnsobɔɔdoore} \tbTP{kɔnsobɔɔdorɛɛ} 
\tbLX{kɔnsogɔ} \tbHM{} \tbPH{kɔ́nsʊ́gɔ́} \tbPS{n.~} \tbSN \tbGE{the middle of water e.g. a stream} \tbSN \tbGE{the act of keeping afloat in water with body upright} \tbOP{kɔnsogere} \tbTP{kɔnsogerɛɛ} \tbVA{kɔnsoga}
\tbLX{kɔnsolaa} \tbHM{} \tbPH{kɔ̀nsòláá} \tbPS{n.~} \tbGE{large jar used for bathing} \tbOP{kɔnsolaare} \tbTP{kɔnsolaarɛɛ} 
\tbLX{kɔnsoŋ} \tbHM{} \tbPH{kɔ̀nsʊ́ŋ} \tbPS{n.~} \tbGE{holy water} \tbOP{kɔnsonne} \tbTP{kɔnsonnɛɛ} 
\tbLX{kɔnsɔgerɔ} \tbHM{} \tbPH{kɔ̀nsɔ́grɔ́} \tbPS{n.~} \tbGE{dirty water} \tbTP{kɔnsɔgerɛɛ} \tbVA{kɔnsɔgera}
\tbLX{kɔnsuori} \tbHM{} \tbPH{kɔ̀nsúórì} \tbPS{n.~} \tbGE{bathwater} \tbTP{kɔnsuoree} 
\tbLX{kɔnsuoro} \tbHM{} \tbPH{kɔ̀nsúóró} \tbPS{n.~} \tbGE{one who is bathing} \tbOP{kɔnsuoribo} \tbTP{kɔnsuoribee} 
\tbLX{kɔntere} \tbHM{} \tbPH{kɔ́ntɪ́r} \tbPS{n.~} \tbGE{cheek} \tbOP{kɔnteɛ} \tbTP{kɔnteerɛɛ} 
\tbLX{kɔntɛtɛgere} \tbHM{} \tbPH{kɔ̀ntɛ́tɛ̀grɪ́} \tbPS{n.~} \tbGE{puddles} \tbOP{kɔntɛtɛgerɛɛ} 
\tbLX{kɔntɔmbile} \tbHM{} \tbPH{kɔ́ntɔ́mbílé} \tbPS{n.~} \tbGE{fairy} \tbOP{kɔntɔmbilii} \tbTP{kɔntɔmbilee} \tbVA{kɔntɔmɔ, kɔntɔnzuzeɛ}
\tbLX{kɔntɔmbɔgɔ} \tbHM{} \tbPH{kɔ̀ntɔ̀mbɔ́gɔ́} \tbPS{n.~} \tbGE{the act of divining through the power of fairy spirits} \tbOP{kɔntɔmbɔgɔrɔ} \tbTP{kɔntɔmbɔgrɛɛ} 
\tbLX{kɔntɔmɔ} \tbHM{} \tbPH{kɔ̀ntɔ́mɔ̀} \tbPS{n.~} \tbGE{fairy, dwarf (spiritual being)} \tbOP{kɔtɔnne} \tbTP{kɔtɔnnɛɛ} \tbVA{kɔntɔmbile}
\tbLX{kɔntɔn-ire} \tbHM{} \tbPH{kɔ̀ntɔ̀n-íré} \tbPS{n.~} \tbGE{one possessed by a spirit} \tbOP{kɔntɔn-iribe} \tbTP{kɔntɔn-iribee} \tbVA{kɔntɔŋ-ire}
\tbLX{kɔntɔnkyi} \tbHM{} \tbPH{kɔ̀ntɔ̀nʧì} \tbPS{n.~} \tbGE{a type of guinea corn with short thick bunch of seed} \tbTP{kɔntɔnkyiree} 
\tbLX{kɔntɔnnaa} \tbHM{} \tbPH{kɔ̀ntɔ̀nnáá} \tbPS{n.~} \tbGE{two-shilling coin} \tbOP{kɔntɔnnanne} \tbTP{kɔntɔnnannɛɛ} 
\tbLX{kɔntɔn-woɔ} \tbHM{} \tbPH{kɔ̀ntɔ̀n-wʊ́ɔ́} \tbPS{n.~} \tbGE{soothsayer's bag} \tbOP{kɔntɔn-woore} \tbTP{kɔntɔn-woorɛɛ} 
\tbLX{kɔntɔŋgbelime} \tbHM{} \tbPH{kɔ̀ntɔ̀ŋgbèlìmé} \tbPS{n.~} \tbGE{diviner's bell} \tbOP{kɔntɔŋgbelli} \tbTP{kɔntɔŋgbellee} 
\tbLX{kɔnvɔlɔ} \tbHM{} \tbPH{kɔ̀nvɔ̀lɔ́} \tbPS{n.~} \tbGE{liquid residue of shea butter} \tbOP{kɔnvɔlle} \tbTP{kɔnvɔlɛɛ/kɔnvɔllɛɛ} \tbVA{kɔvɔlɔ}
\tbLX{kɔn-yaaraa} \tbHM{} \tbPH{kɔ̀n-yáárá} \tbPS{n.~} \tbGE{a type of bird that feeds on fish} \tbOP{kɔn-yaare} \tbTP{kɔn-yaarɛɛ} \tbVA{kolyaaraa}
\tbLX{kɔn-yuori} \tbHM{} \tbPH{kɔn-yuori} \tbPS{n.~} \tbGE{pot for water} \tbOP{kɔn-yoe} \tbTP{kɔn-yoree} 
\tbLX{kɔnzuluŋ} \tbHM{} \tbPH{kɔ̀nzùlúŋ} \tbPS{n.~} \tbGE{deep water} \tbOP{kɔnzululli} \tbTP{kɔnzulullee} 
\tbLX{kɔnzunni} \tbHM{} \tbPH{kɔ́nzúnní} \tbPS{n.~} \tbGE{debris of water} \tbTP{kɔnzunnee} 
\tbLX{kɔŋe} \tbHM{} \tbPH{kɔ̀ŋ} \tbPS{v.~} \tbGE{to contract leprosy} \tbPD{kɔŋɛɛ, kɔnnɔ, kɔnnɔ, kɔnema, kɔnnaa, kɔnema } 
\tbLX{kɔŋe} \tbHM{} \tbPH{kɔ̀ŋ} \tbPS{v.~} \tbGE{to be unavailable} \tbXV{Ka saaŋ ba miire te maŋ kɔŋe la ko᷉ɔ.} \tbXE{If it does not rain, we run out of water.} \tbPD{kɔŋɛɛ, kɔnnɔ, kɔnnɔ, kɔnema, kɔnnaa, kɔnema } 
\tbLX{kɔŋe} \tbHM{} \tbPH{kɔ̂ŋ} \tbPS{n.~} \tbGE{hunger} \tbTP{kɔnnɛɛ} \tbVA{koŋ (saanchi)}
\tbLX{kɔŋgaŋene} \tbHM{} \tbPH{kɔ̀ŋgáŋɪ́nɪ́} \tbPS{adj.~} \tbGE{a place beyond the waters} 
\tbLX{kɔŋkaare} \tbHM{} \tbPH{kɔ̌ŋkáárɪ́} \tbPS{n.~} \tbSN \tbGE{water that has been bailed out} \tbSN \tbGE{water that must be bailed out} 
\tbLX{kɔŋkare} \tbHM{} \tbPH{kɔ̂ŋkárɪ́} \tbPS{n.~} \tbGE{plain drinking water} \tbTP{kɔŋkarɛɛ} 
\tbLX{kɔŋkɔŋ} \tbHM{} \tbPH{kɔ́ŋkɔ́ŋ} \tbPS{n.~} \tbGE{palm tree} \tbOP{kɔŋkɔnne} \tbTP{kɔŋkɔnnɛɛ} 
\tbLX{kɔŋkɔŋɔ} \tbHM{} \tbPH{kɔ̀ŋkɔ̀ŋɔ́} \tbPS{n.~} \tbGE{leper} \tbOP{kɔŋkɔnne} \tbTP{kɔŋkɔnnɛɛ} 
\tbLX{kɔŋkɔre} \tbHM{} \tbPH{kɔ̀ŋkɔ̀rɪ́} \tbPS{n.~} \tbGE{forest} \tbVA{tetige}
\tbLX{kɔŋkpaare} \tbHM{} \tbPH{kɔ̀ŋkpáárí} \tbPS{n.~} \tbGE{backstroke (swimming)} \tbOP{kɔŋkpaɛ} 
\tbLX{kɔɔ} \tbHM{} \tbPH{kɔ̀ɔ̀} \tbPS{v.~} \tbGE{to break} \tbPD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\tbLX{kɔɔ} \tbHM{} \tbPH{kɔ̀ɔ̀} \tbPS{v.~} \tbGE{to dance} \tbPD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\tbLX{kɔɔ} \tbHM{} \tbPH{kɔ̀ɔ̀} \tbPS{v.~} \tbGE{to become charred} \tbPD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\tbLX{kɔɔ} \tbHM{} \tbPH{kɔ̀ɔ̀} \tbPS{v.~} \tbGE{to stampede, sudden flight or dispersion} \tbPD{kɔɔɛ, kɔɔrɔ, kɔɔrɔ, kɔreba, kɔɔraa, kɔreba } 
\tbLX{kɔɔ} \tbHM{} \tbPH{kɔ̀ɔ́} \tbPS{adv.~} \tbGE{one hundred} 
\tbLX{kɔɔ} \tbHM{} \tbPH{kɔ́ɔ́} \tbPS{adj.~} \tbGE{appear broken} 
\tbLX{kɔɔkyaporo} \tbHM{} \tbPH{kɔ̀ɔ̀ʧàpòrò} \tbPS{n.~} \tbGE{bone marrow} \tbSG{kɔɔkyaporuu} \tbOP{kɔɔkyaporri} \tbTP{kɔɔkyaporree} 
\tbLX{kɔɔle} \tbHM{} \tbPH{kɔ́ɔ́lɪ́} \tbPS{adj.~} \tbGE{burnt, charred} \tbXV{A vu᷉u toloŋ, ka fooŋ deɛnɛ kamaa-kɔɔle ka fo na sɛ.} \tbXE{The fire is quite hot, and if you are not careful the maize will get burnt.} \tbOP{kɔɔlɔ} 
\tbLX{kɔɔlmare} \tbHM{} \tbPH{kɔ̀ɔ̀lmàrɪ̀} \tbPS{v.~} \tbGE{to hurt verbally e.g. insult, chastise} \tbPD{kɔɔlmarɛɛ, kɔɔlɔmara, kɔɔlɔmara, kɔɔlemareba, kɔɔlɔmaraa, - } 
\tbLX{kɔɔloŋ} \tbHM{} \tbPH{kɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{hair, fur} \tbOP{kɔɔlɔ} \tbTP{kɔɔlɛɛ} 
\tbLX{kɔɔre} \tbHM{} \tbPH{kɔ́ɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{bone} \tbSN \tbGE{armlet made from elephant tusk} \tbOP{kɔbɔ} \tbTP{kɔɔrɛɛ} 
\tbLX{kɔɔre} \tbHM{} \tbPH{kɔ̀ɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to delay} \tbXV{Ta kɔɔre ma, N boɔrɛɛ wĩẽũ.} \tbXE{Don’t delay me I am in a haste.} \tbSN \tbGE{keep long} \tbXV{Zenɛ a bie kɔɔre la a sakuuri poɔ.} \tbXE{The child has kept long in school today.} \tbSN \tbGE{to be late} \tbXV{N karembie kɔɔre la sakuuri gaabo ka ba ko o dɔgeroŋ.} \tbXE{The pupil was late to school and they gave him some punishment.} \tbPD{kɔɔrɛɛ, kɔɔrɔ, kɔɔrɔ, kɔɔreba, kɔɔraa, kɔɔreba } 
\tbLX{kɔɔto} \tbHM{} \tbPH{kɔ́ɔ́tʊ̀} \tbPS{n.~} \tbGE{court} \tbOP{kɔɔtere} \tbTP{kɔɔterɛɛ} 
\tbLX{kɔɔyɛ} \tbHM{} \tbPH{kɔ́ɔ́yɛ̀} \tbPS{n.~} \tbGE{forest} \tbOP{kɔɔyɛre} \tbTP{kɔɔyɛrɛɛ} 
\tbLX{kɔɔzii} \tbHM{} \tbPH{kɔ̀ɔ̀zìì} \tbPS{n.~} \tbGE{a type of tiny bird that moves in large groups} \tbOP{kɔɔziiri} \tbTP{kɔɔziiree} 
\tbLX{kɔpo} \tbHM{} \tbPH{kɔ́pʊ̀} \tbPS{n.~} \tbGE{cup} \tbOP{kɔpore} \tbTP{kɔporɛɛ} 
\tbLX{kɔpuori} \tbHM{} \tbPH{kɔ́pùòrí} \tbPS{n.~} \tbGE{the act of weeding again} \tbOP{kopoe} \tbTP{koporee} 
\tbLX{kɔraa} \tbHM{} \tbPH{kɔ̀ràà} \tbPS{n.~} \tbGE{hook} \tbOP{kɔrre} \tbTP{kɔrɛɛ} \tbVA{takɔraa}
\tbLX{kɔre} \tbHM{} \tbPH{kɔ̀rɪ́} \tbPS{n.~} \tbSN \tbGE{pocket} \tbSN \tbGE{bag} \tbOP{kɔɛ} \tbTP{kɔrɛɛ} 
\tbLX{kɔre} \tbHM{} \tbPH{kɔ̀rɪ̀} \tbPS{v.~} \tbGE{to scoop out; to empty} \tbXV{A yuori ko᷉ɔ zaa ka ba kɔre.} \tbXE{They have taken all the water from the pot.} \tbPD{kɔrɛɛ, kɔrɔ, kɔrɔ, kɔreba, kɔraa, - } 
\tbLX{kɔre} \tbHM{} \tbPH{kɔ̀rɪ̀} \tbPS{adv.~} \tbGE{how it sounds when a stone hits an object} 
\tbLX{kɔre} \tbHM{} \tbPH{kɔ́rɪ̀} \tbPS{v.~} \tbGE{to cough} \tbPD{kɔrɛɛ, kɔrɔ, kɔrɔ, kɔreba, kɔraa, kɔreba } 
\tbLX{kɔrema} \tbHM{} \tbPH{kɔ̀rɪ́má} \tbPS{n.~} \tbGE{partridge} \tbOP{kɔrre} \tbTP{kɔrrɛɛ} 
\tbLX{kɔretɔ} \tbHM{} \tbPH{kɔ́rɪ́tɔ́} \tbPS{n.~} \tbGE{sickness of animals (fur falls off)} \tbOP{kɔretɔre} \tbTP{kɔretɔrɛɛ} 
\tbLX{kɔrɛɛ} \tbHM{} \tbPH{kɔ́rɛ́ɛ́} \tbPS{adv.~} \tbGE{moving at once nearer another e.g. child to mother} 
\tbLX{kɔroŋ} \tbHM{} \tbPH{kɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{cough} 
\tbLX{kɔroŋkpoŋ} \tbHM{} \tbPH{kɔ́rʊ́ŋkpóŋ} \tbPS{n.~} \tbGE{mycobacterium tuberculosis} \tbOP{kɔroŋkponni} \tbTP{kɔroŋkponnee} 
\tbLX{kɔrre} \tbHM{} \tbPH{kɔ́rrɪ̀} \tbPS{v.~} \tbGE{to slaughter, to cut the throat of} \tbPD{kɔrrɛɛ, kɔrrɔ, kɔrrɔ, kɔreba, kɔrraa, kɔreba } 
\tbLX{kɔsɔrre} \tbHM{} \tbPH{kɔ́sɔ́rrɪ́} \tbPS{n.~} \tbGE{request for farm help} \tbTP{kɔsɔrɛɛ} 
\tbLX{kɔtɔ} \tbHM{} \tbPH{kɔ́tɔ́} \tbPS{n.~} \tbGE{crab} \tbOP{kɔtɔre} \tbTP{kɔtɔrɛɛ} 
\tbLX{kɔvɔlɔ} \tbHM{} \tbPH{kɔ̀vɔ̀lɔ́} \tbPS{n.~} \tbGE{} \tbTP{kɔvɔlɛɛ} \tbVA{kɔnvɔlɔ}
\tbLX{kubogi} \tbHM{} \tbPH{kúbògí} \tbPS{n.~} \tbGE{stone cave} \tbOP{kubogiri} \tbTP{kubogiree} 
\tbLX{kugbe} \tbHM{} \tbPH{kùgbé} \tbPS{n.~} \tbGE{coconut} \tbOP{kugberi} \tbTP{kugberee} 
\tbLX{kugbeɛ} \tbHM{} \tbPH{kùgbɪ̀ɛ́} \tbPS{n.~} \tbGE{worn-out hoe blade} \tbOP{kugbeere} \tbTP{kugbeerɛɛ} 
\tbLX{kukurbompaala} \tbHM{} \tbPH{kùkúrbʊ́mpáálà} \tbPS{n.~} \tbGE{firstfruits} \tbSG{kukurbompaalɛɛ} \tbVA{bompaala}
\tbLX{kukuri} \tbHM{} \tbPH{kùkúrì} \tbPS{n.~} \tbGE{hoe handle} \tbOP{kukue} \tbTP{kukuree} 
\tbLX{kukuribone} \tbHM{} \tbPH{kùkúrìbʊ̀n} \tbPS{n.~} \tbGE{farm produce} \tbOP{kukuriboma} 
\tbLX{kukurkpaare} \tbHM{} \tbPH{kúkúkpáárɪ́} \tbPS{n.~} \tbGE{nape of the neck of a hoe handle} \tbOP{kukukpaɛ} \tbTP{kukukpaarɛɛ} 
\tbLX{kukpoŋi} \tbHM{} \tbPH{kùkpôŋ} \tbPS{n.~} \tbGE{large hoe} \tbOP{kukponni} \tbTP{kukpomo/kukponnee} 
\tbLX{kuli} \tbHM{} \tbPH{kúlí} \tbPS{v.~} \tbGE{to go home} \tbPD{kulee, kulo, kulo, kulbo, kulaa, - } 
\tbLX{kulikuli} \tbHM{} \tbPH{kúlkúl} \tbPS{adv.~} \tbGE{very bad-smelling} 
\tbLX{kulinni} \tbHM{} \tbPH{kúlínní} \tbPS{n.~} \tbGE{shrill noise of praise} \tbTP{kulinnee} 
\tbLX{kull} \tbHM{} \tbPH{kùll} \tbPS{adv.~} \tbGE{rising in a cloud} \tbVA{koll}
\tbLX{kulli} \tbHM{} \tbPH{kùllì} \tbPS{v.~} \tbSN \tbGE{to mind} \tbSN \tbGE{to not regard, to ignore (typically used with ba)} \tbXV{A baa da vuolo yaga lɛ kyɛ a nannyige deɛ da ba kullo a.} \tbXE{The dog was barking intensely, but the thief just ignored it.} \tbPD{kullee, kullo, kullo, kulibo, kullaa, - } \tbVA{kpulli/kpullo}
\tbLX{kultaa} \tbHM{} \tbPH{kúltáá} \tbPS{n.~} \tbGE{marriage} \tbOP{kultaarɛɛ} 
\tbLX{kultaa wɛltaa} \tbHM{} \tbPH{kúltáá wɛ́ltáá} \tbPS{n.~} \tbGE{divorce} 
\tbLX{kunoɔre} \tbHM{} \tbPH{kùnʊ́ɔ̀rɪ̀} \tbPS{n.~} \tbGE{edge of hoe blade} \tbOP{kunɔɛ} \tbTP{kunɔrɛɛ} 
\tbLX{kuntuŋi} \tbHM{} \tbPH{kúntúŋ} \tbPS{n.~} \tbGE{blanket} \tbOP{kuntunni} \tbTP{kuntunnee} 
\tbLX{kuntuŋi} \tbHM{} \tbPH{kùntùŋ} \tbPS{n.~} \tbGE{wolf} \tbOP{kuntunni} \tbTP{kuntunnee} 
\tbLX{kuntuŋi} \tbHM{} \tbPH{kùntùŋ} \tbPS{n.~} \tbGE{blanket} \tbOP{kuntunni} \tbTP{kuntunnee} 
\tbLX{kunyoɔre} \tbHM{} \tbPH{kùɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{hoe, plough tip} \tbOP{kunyɔɛ} \tbTP{kùnyɔrɛɛ} 
\tbLX{kuŋkunee} \tbHM{} \tbPH{kúŋkúnéè} \tbPS{n.~} \tbGE{hillock, raised ground} \tbOP{kuŋkunni} \tbTP{kuŋkunnee} 
\tbLX{kuŋkunni} \tbHM{} \tbPH{kùŋkúnnì} \tbPS{n.~} \tbGE{epilepsy} \tbTP{kuŋkunnee} 
\tbLX{kuŋkuŋ} \tbHM{} \tbPH{kúŋkúŋ} \tbPS{adv.~} \tbGE{dark and threatening e.g. black storm clouds} 
\tbLX{kuŋkuŋlere} \tbHM{} \tbPH{kúŋkúnléré} \tbPS{n.~} \tbGE{one who has epilepsy} \tbOP{kuŋkunlerbe} \tbTP{kuŋkunlerbee} 
\tbLX{kuo} \tbHM{} \tbPH{kúó} \tbPS{n.~} \tbGE{a type of a wild mouse} \tbOP{kuuri} \tbTP{kuuree} 
\tbLX{kuoli} \tbHM{} \tbPH{kùòlì} \tbPS{v.~} \tbGE{to get drunk} \tbPD{kuolee, kuolo, kuolo, kuolbo, kuolaa, - } 
\tbLX{kuoli} \tbHM{} \tbPH{kùòlì} \tbPS{v.~} \tbGE{to apply fertilizer to crops} \tbPD{kuolee, kuolo, kuolo, kuolbo, kuolaa, - } 
\tbLX{kuoluŋ} \tbHM{} \tbPH{kùólǔŋ} \tbPS{n.~} \tbGE{manure} \tbTP{kuoklunnee} 
\tbLX{kuomo} \tbHM{} \tbPH{kúómó} \tbPS{n.~} \tbGE{a type of water plant} \tbOP{kuomee} 
\tbLX{kuoŋaa} \tbHM{} \tbPH{kúóŋáá} \tbPS{adj.~} \tbGE{dry} \tbOP{kuoni} \tbTP{kuonee} 
\tbLX{kuori} \tbHM{} \tbPH{kùórí} \tbPS{n.~} \tbGE{funeral} \tbOP{koe} \tbTP{koree} 
\tbLX{kuori} \tbHM{} \tbPH{kúórì} \tbPS{v.~} \tbGE{to tap with an object} \tbPD{kuoree, kuoro, kuoro, kuoribo, kuoraa, - } 
\tbLX{kuree} \tbHM{} \tbPH{kùrèè} \tbPS{n.~} \tbGE{shorts} \tbOP{kurri} \tbTP{kurree} 
\tbLX{kuri} \tbHM{} \tbPH{kúrí} \tbPS{n.~} \tbSN \tbGE{metal} \tbSN \tbGE{iron} \tbSG{kuruu} \tbOP{kurri} \tbTP{kuree} 
\tbLX{kuri} \tbHM{} \tbPH{kùrì} \tbPS{v.~} \tbGE{to forge, to shape metal} \tbPD{kuree, kuro, kuro, kuribo, kuraa, kuribo } 
\tbLX{kuri} \tbHM{} \tbPH{kùrì} \tbPS{v.~} \tbGE{to plant calabash seeds} \tbPD{kuree, kuro, kuro, kuribo, kuraa, kuribo } 
\tbLX{kuri} \tbHM{} \tbPH{kúrí} \tbPS{v.~} \tbGE{to scrape} \tbPD{kuree, kuro, kuro, kuribo, kuraa, kuribo } 
\tbLX{kuri} \tbHM{} \tbPH{kùrì} \tbPS{adv.~} \tbGE{onomatopoeia (thud of a heavy solid object falling)} 
\tbLX{kuribini} \tbHM{} \tbPH{kùrìbíní} \tbPS{n.~} \tbGE{slag (ore after iron is extracted)} \tbSG{kuribinuu} \tbTP{kuribinee/kurbinnee} 
\tbLX{kuridaa} \tbHM{} \tbPH{kùrìdàá} \tbPS{n.~} \tbGE{iron bar} \tbOP{kuridaare} \tbTP{kuriaarɛɛ} 
\tbLX{kurigbɛɛgbori} \tbHM{} \tbPH{kúrígbɛ́ɛ́gbòrí} \tbPS{n.~} \tbGE{train} \tbOP{kurigbɛɛgboe} \tbTP{kurigbɛɛgboree} 
\tbLX{kurigbɛɛsori} \tbHM{} \tbPH{kúrígbɛ́ɛ́sórí} \tbPS{n.~} \tbGE{rail line} \tbOP{kurigbɛɛsoe} \tbTP{kurigbɛɛsoree} 
\tbLX{kurigbɛwogri} \tbHM{} \tbPH{kúrígbɛ́wógrì} \tbPS{n.~} \tbGE{trousers} \tbTP{kurigbɛwogree} 
\tbLX{kurikuni} \tbHM{} \tbPH{kùrìkúní} \tbPS{n.~} \tbSN \tbGE{tortoise} \tbSN \tbGE{ball of seeds in the calabash fruit} \tbSN \tbGE{padlock} \tbOP{kurikumo} \tbTP{kurikunee} \tbVA{kuri/kuŋkuni}
\tbLX{kurikuri} \tbHM{} \tbPH{kùrìkúrí} \tbPS{v.~} \tbGE{to search} \tbPD{kurikuree, kurikuro, kurikuro, kurikuribo, kurikuraa, - } 
\tbLX{kurikuro} \tbHM{} \tbPH{kùrìkúró} \tbPS{n.~} \tbGE{blacksmith} \tbOP{kurikuribo} \tbTP{kurikuribee} \tbVA{kurkuro, kurkuribo}
\tbLX{kurimoŋ} \tbHM{} \tbPH{kúrímóŋ} \tbPS{adv.~} \tbGE{stunted e.g. malnourished child} 
\tbLX{kuriwiri} \tbHM{} \tbPH{kúríwìrí} \tbPS{n.~} \tbGE{bicycle} \tbOP{kuriwie} \tbTP{kurwiree} 
\tbLX{kuriyoluŋ} \tbHM{} \tbPH{kúríyólúŋ} \tbPS{n.~} \tbGE{the act of wearing shorts or trousers without underwear} \tbOP{kuriyolli} \tbTP{kuriyolee} 
\tbLX{kurizeɛ} \tbHM{} \tbPH{kùrìzɪ̀ɛ́} \tbPS{n.~} \tbSN \tbGE{brass} \tbSN \tbGE{copper} \tbOP{kurizeere} \tbTP{kurizeerɛɛ} 
\tbLX{kuri zu mare} \tbHM{} \tbPH{} \tbPS{v.~} \tbGE{to persevere} \tbPD{-, -, -, -, -, - } 
\tbLX{kurkumpɛge} \tbHM{} \tbPH{kùrkûmpɛ́gɪ̀} \tbPS{n.~} \tbGE{shell (of turtle)} \tbOP{kurkumpɛgre} \tbTP{kurkumpɛgrɛɛ} 
\tbLX{kurusu} \tbHM{} \tbPH{kúrúsú} \tbPS{adv.~} \tbGE{small and stunted} 
\tbLX{kusaalaa} \tbHM{} \tbPH{kúsààlàà} \tbPS{n.~} \tbGE{smooth stone used by a potter to smooth pots} \tbOP{kusaale} \tbTP{kusaalɛɛ} 
\tbLX{kusaalaa} \tbHM{} \tbPH{kúsàálàá} \tbPS{n.~} \tbGE{sharpening stone} \tbOP{kusaale} \tbTP{kusaalɛɛ} 
\tbLX{kuseɛbaa} \tbHM{} \tbPH{kúsɪ̀ɛ̀bàá} \tbPS{n.~} \tbGE{very large hoe} \tbOP{kuseɛbaare} \tbTP{kuseɛbaarɛɛ} 
\tbLX{kusibiri} \tbHM{} \tbPH{kúsíbírì} \tbPS{n.~} \tbGE{a very small stone} \tbOP{kusibie} \tbTP{kusibiree/kusibiiree} 
\tbLX{Kusieli} \tbHM{} \tbPH{kúsíélí} \tbPS{n.~} \tbGE{homeland of the Kusielee clan} 
\tbLX{Kusieluu} \tbHM{} \tbPH{kúsíélúů} \tbPS{n.~} \tbGE{a member of the Kusielee clan} \tbOP{Kusielee} 
\tbLX{kusiri} \tbHM{} \tbPH{kúsírì} \tbPS{n.~} \tbGE{stone} \tbOP{kusie} \tbTP{kusiruu} 
\tbLX{kusiwara} \tbHM{} \tbPH{kúsíwàrà} \tbPS{n.~} \tbGE{gravel} \tbTP{kusiwarɛɛ} 
\tbLX{kutaayi} \tbHM{} \tbPH{kútááyí} \tbPS{n.~} \tbGE{file} \tbOP{kutaayiri} \tbTP{kutaayiree} 
\tbLX{kutobo} \tbHM{} \tbPH{kùtòbó} \tbPS{n.~} \tbGE{hoe (traditional Dagaaba hoe)} \tbTP{kutooree} 
\tbLX{kutoori} \tbHM{} \tbPH{kùtóórí} \tbPS{n.~} \tbGE{plough breast} \tbOP{kutooree} 
\tbLX{ku᷉u} \tbHM{1} \tbPH{kú᷉ú} \tbPS{n.~} \tbSN \tbGE{death} \tbSN \tbGE{corpse} \tbOP{kuuni} \tbTP{kuunee} 
\tbLX{ku᷉u} \tbHM{2} \tbPH{kú᷉ù} \tbPS{n.~} \tbGE{} \tbOP{kuuni} \tbTP{kuunee} \tbVA{shortened form of biku᷉u}
\tbLX{ku᷉ubogi} \tbHM{} \tbPH{kú᷉úbògí} \tbPS{n.~} \tbGE{grave} \tbOP{ku᷉ubogri} \tbTP{ku᷉ubogree} 
\tbLX{kuunileɛiruu} \tbHM{} \tbPH{kúúnílɪ́ɛ́-írúú} \tbPS{n.~} \tbGE{resurrection} 
\tbLX{kuuri} \tbHM{} \tbPH{kùùrí} \tbPS{n.~} \tbGE{hoe} \tbOP{kue} \tbTP{kuree} 
\tbLX{kuuri} \tbHM{} \tbPH{kùùrì} \tbPS{v.~} \tbGE{to dredge; to remove from the bottom of a container of liquid} \tbPD{kuuree, kuuro, kuuro, kuuribo, kuuraa, - } 
\tbLX{kuuri} \tbHM{} \tbPH{kúúrì} \tbPS{v.~} \tbGE{to quickly dodge an object in the air} \tbPD{kuuree, kuuro, kuuro, kuuribo, kuuraa, - } 
\tbLX{kuuri} \tbHM{} \tbPH{kúúrì} \tbPS{n.~} \tbSN \tbGE{stone} \tbSN \tbGE{fetish stone} \tbOP{kubo} \tbTP{kuuree} \tbVA{kusiri, kusibe, kusirree, kusiruu}
\tbLX{ku᷉uyee} \tbHM{} \tbPH{kú᷉ú᷉yéè} \tbPS{n.~} \tbGE{a type of small owl believed to foresee evil happenings} \tbOP{ku᷉uyeeri} \tbTP{ku᷉uyeeree} 
\tbLX{kpa} \tbHM{} \tbPH{kpá} \tbPS{v.~} \tbSN \tbGE{to boil} \tbSN \tbGE{to dismantle (usually at the joints)} \tbSN \tbGE{to lament; to mourn} \tbPD{kpaɛ, kpaara, kpaara, kpaareba, kpaaraa, - } 
\tbLX{kpa} \tbHM{} \tbPH{kpà} \tbPS{v.~} \tbGE{to lock} \tbPD{kpaɛ, kpaara, kpaara, kpareba, kpaara, - } 
\tbLX{kpaa} \tbHM{} \tbPH{kpàà} \tbPS{v.~} \tbSN \tbGE{to be unsuccessful} \tbXV{A karema goɔ la a bie kpaa.} \tbXE{The teacher was unable to stop the child.} \tbSN \tbGE{to leave some of} \tbXV{Ba di la a bondirii kyɛ ka a kpaa.} \tbXE{They ate the food and some was left over.} \tbPD{kpaaɛ, kpaara, kpaara, kpaareba, kpaaraa, - } 
\tbLX{kpaa} \tbHM{} \tbPH{kpáà} \tbPS{v.~} \tbGE{to remove by pulling out} \tbPD{kpaaɛ, kpaara, kpaara, kpaareba, kpaaraa, - } 
\tbLX{kpaa} \tbHM{} \tbPH{kpàá} \tbPS{adj.~} \tbGE{leftover} \tbXV{A baa ŋa ba nɔŋ bondi-kpaa.} \tbXE{This dog does not like leftover food.} \tbOP{kpaare} \tbTP{kpaarɛɛ} 
\tbLX{kpaa} \tbHM{} \tbPH{kpáá} \tbPS{interj.~} \tbGE{surprised} 
\tbLX{kpa᷉a} \tbHM{} \tbPH{kpà᷉à᷉} \tbPS{v.~} \tbGE{to warn} \tbPD{kpaa᷉ɛ, kpaana, kpaana, kpaanema, kpaanaa, kpaanema } 
\tbLX{kpa᷉a} \tbHM{} \tbPH{kpá᷉à᷉} \tbPS{v.~} \tbGE{cook very palatable food, fry (saucetew)} \tbPD{kpa᷉aɛ, kpaana, kpaana, kpaanema, kpaanaa, - } 
\tbLX{kpaagyiri} \tbHM{} \tbPH{kpáádzrí} \tbPS{n.~} \tbGE{strong whip (made of tendon and leather)} \tbTP{kpaagyiree} \tbVA{gyiri}
\tbLX{kpaakpaa} \tbHM{} \tbPH{kpàákpàá} \tbPS{n.~} \tbGE{palm kernel} \tbOP{kpaakpaare} \tbTP{kpaakpaarɛɛ} 
\tbLX{kpaakpaabiri} \tbHM{} \tbPH{kpàákpàábírì} \tbPS{n.~} \tbGE{palm kernel seed} \tbSG{kpaakpaabiruu} \tbOP{kpaakpaabie} \tbTP{kpaakpaabiree/kpaakpaabiiree} \tbVA{kpaakpaa}
\tbLX{kpaakyinni} \tbHM{} \tbPH{kpááʧìnnì} \tbPS{v.~} \tbGE{to leap from a height to land} \tbPD{kpaakyinnee, kpaarakyinne, kpaakyinne, kpaakyinime, kpaakyinnaa, - } 
\tbLX{kpaale} \tbHM{} \tbPH{kpáálɪ̀} \tbPS{v.~} \tbGE{to remind} \tbPD{kpaalɛɛ, kpaala, kpaala, kpaaleba, kpaalaa, - } 
\tbLX{kpa᷉aloɔraa} \tbHM{} \tbPH{kpá᷉álʊ́ɔ́ráá} \tbPS{n.~} \tbGE{a type of small brown bird} \tbOP{kpa᷉aloɔre} \tbTP{kpa᷉aloɔrɛɛ} 
\tbLX{kpaalzeɛ} \tbHM{} \tbPH{kpààlzɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{kpaalzeere} \tbTP{kpaalzeerɛɛ} 
\tbLX{kpaamoɔ} \tbHM{} \tbPH{kpáámʊ́ɔ́} \tbPS{n.~} \tbGE{a type of grass} \tbSG{kpaamoɔnoo} \tbTP{kpaamoɔnɛɛ} 
\tbLX{kpaane} \tbHM{} \tbPH{kpàànɪ̀} \tbPS{v.~} \tbSN \tbGE{to become tight} \tbSN \tbGE{to go hunting} \tbPD{kpaanɛɛ, kpaana, kpaana, kpaanema, kpaanaa, - } 
\tbLX{kpaanyo} \tbHM{} \tbPH{kpàányʊ̀} \tbPS{n.~} \tbGE{nothing} \tbVA{kpaanyɔ}
\tbLX{kpa᷈ao} \tbHM{} \tbPH{kpá᷉áʊ́} \tbPS{n.~} \tbGE{guinea fowl} \tbOP{kpinni} \tbTP{kpinnee} 
\tbLX{kpaaraa} \tbHM{} \tbPH{kpááràá} \tbPS{n.~} \tbGE{whip} \tbOP{kpaare} \tbTP{kpaarɛɛ} \tbVA{kpeeraa, kpeere}
\tbLX{kpaare} \tbHM{} \tbPH{kpáárɪ́} \tbPS{n.~} \tbGE{nape of neck} \tbOP{kpaɛ} \tbTP{kpaarɛɛ} 
\tbLX{kpaare} \tbHM{} \tbPH{kpáárɪ̀} \tbPS{v.~} \tbSN \tbGE{to tap (on a door)} \tbSN \tbGE{to roof by knocking nails to hold roof} \tbSN \tbGE{to harvest termites (at the end of the process one knocks on the pot containing termites)} \tbSN \tbGE{to fly at top speed} \tbPD{-, -, -, -, -, - } 
\tbLX{kpaare} \tbHM{} \tbPH{kpààrɪ̀} \tbPS{v.~} \tbSN \tbGE{to come together} \tbSN \tbGE{to assemble, to meet together} \tbPD{kpaarɛɛ, kpaara, kpaara, kpaareba, kpaaraa, kpaareba } 
\tbLX{kpaasere} \tbHM{} \tbPH{kpáásɪ́rɪ́} \tbPS{v.~} \tbGE{to end up at a point, not returning} \tbXV{Ba toŋ la a bie ka o deɛ te kpaasere.} \tbXE{The child they sent has ended up not returning.} \tbPD{kpaaserɛɛ, kpaaserɛ, kpaaserɛ, kpaaserebɛ, kpaaseraa, - } 
\tbLX{kpa᷉a toori} \tbHM{} \tbPH{kpà᷉à᷉ tòòrí} \tbPS{v.~} \tbSN \tbGE{to insist} \tbSN \tbGE{to harden one's heart} \tbPD{-, -, -, -, -, - } 
\tbLX{kpaba} \tbHM{} \tbPH{kpàbà} \tbPS{adv.~} \tbGE{sweating profusely} 
\tbLX{kpabogi} \tbHM{} \tbPH{kpàbògí} \tbPS{n.~} \tbGE{point between the neck and the head} \tbOP{kpabogri} \tbTP{kpabogree} 
\tbLX{kpaga} \tbHM{} \tbPH{kpágà} \tbPS{adv.~} \tbGE{refuse to do anything} 
\tbLX{kpage} \tbHM{} \tbPH{kpàgɪ́} \tbPS{adj.~} \tbGE{thick, strong} 
\tbLX{kpagelaga} \tbHM{} \tbPH{kpàglàgà} \tbPS{adv.~} \tbSN \tbGE{thick} \tbSN \tbGE{motionless} 
\tbLX{kpagele} \tbHM{} \tbPH{kpàglɪ̀} \tbPS{v.~} \tbGE{to harden} \tbPD{kpagelɛɛ, kpagela, kpagela, kpageleba, kpagelaa, - } 
\tbLX{kpageloo} \tbHM{} \tbPH{kpáglʊ́ʊ́} \tbPS{adj.~} \tbSN \tbGE{hard, firm} \tbSN \tbGE{difficult} \tbOP{kpagela} 
\tbLX{kpagyɛle} \tbHM{} \tbPH{kpádzɛ́lɪ́} \tbPS{n.~} \tbGE{the pointed part of the back of the head} \tbOP{kpagyɛlɛ} \tbTP{kpagyɛlɛɛ} 
\tbLX{kpagyiili} \tbHM{} \tbPH{kpàdzíílí} \tbPS{n.~} \tbGE{main vein connecting the back of the head and the neck} \tbOP{kpagyiile} \tbTP{kpagyiilee} 
\tbLX{kpakyaga} \tbHM{} \tbPH{kpáʧàgá} \tbPS{adj.~} \tbGE{between} \tbOP{kpakyagre} \tbTP{kpakyagrɛɛ} 
\tbLX{kpala} \tbHM{} \tbPH{kpálà} \tbPS{n.~} \tbGE{a type of powder for the eyelashes} \tbOP{kpalɛɛ} 
\tbLX{kpala} \tbHM{} \tbPH{kpàlá} \tbPS{n.~} \tbSN \tbGE{nakedness, nudity} \tbSN \tbGE{the act of engaging in a fight (wrestling)} \tbTP{kpalɛɛ} 
\tbLX{kpalaa} \tbHM{} \tbPH{kpàlàá} \tbPS{n.~} \tbGE{sling} \tbOP{kpalle} \tbTP{kpallɛɛ} 
\tbLX{kpalaŋ} \tbHM{} \tbPH{kpàlàŋ} \tbPS{adj.~} \tbGE{quickly, at once} 
\tbLX{kpalaŋkpalaŋ} \tbHM{} \tbPH{kpáláŋkpáláŋ} \tbPS{adv.~} \tbGE{in a uneventful manner} \tbXV{A dɔɔ entuo na toŋ la a toma kpalaŋkpalaŋ lɛ.} \tbXE{That lazy man did the work haphazardly.} 
\tbLX{kpale} \tbHM{} \tbPH{kpálɪ́} \tbPS{v.~} \tbGE{to become bare (bald)} \tbXV{N sa᷉akoma zu zaaŋ kpale.} \tbXE{My granfather is completely bald.} \tbPD{kpalɛɛ, kpala, kpala, kpaleba, kpalaa, kpaleba } 
\tbLX{kpale} \tbHM{} \tbPH{kpàlɪ̀} \tbPS{adj.~} \tbSN \tbGE{bare; empty} \tbSN \tbGE{for nothing} \tbXV{O wa la kyɛ kpale lɛ, o ba nyɛ bonzaa.} \tbXE{She came here for nothing; she did not get anything.} \tbOP{kpala} \tbTP{kpalɛɛ} 
\tbLX{kpale} \tbHM{} \tbPH{kpàlɪ̀} \tbPS{v.~} \tbGE{to make firm (loose soil)} \tbPD{kpalɛɛ, kpalla, kpalla,  kpalba, kpallaa, - } 
\tbLX{kpaleŋkpaa} \tbHM{} \tbPH{kpàlɪ̀ŋkpàá} \tbPS{n.~} \tbGE{louse} \tbOP{kpaleŋkpaare} \tbTP{kpaleŋkpaarɛɛ} 
\tbLX{kpalle} \tbHM{} \tbPH{kpállɪ̀} \tbPS{v.~} \tbSN \tbGE{to remind} \tbSN \tbGE{to caution} \tbSN \tbGE{to remould the blade of an overused implement (axe)} \tbPD{kpallɛɛ, kpalla, kpalla, kpalba, kpallaa, - } 
\tbLX{kpambeɛ} \tbHM{} \tbPH{kpǎmbɪ́ɛ̀} \tbPS{n.~} \tbGE{spokesman} \tbOP{kpambeere} \tbTP{kpambeerɛɛ} 
\tbLX{kpambiri} \tbHM{} \tbPH{kpàmbírí} \tbPS{n.~} \tbGE{wooden bell} \tbOP{kpambie} \tbTP{kpambiiree} 
\tbLX{kpammo} \tbHM{} \tbPH{kpàmmó} \tbPS{adv.~} \tbGE{thick and hard} 
\tbLX{kpane} \tbHM{} \tbPH{kpànɪ̀} \tbPS{n.~} \tbSN \tbGE{lead} \tbSN \tbGE{spear, lance} \tbOP{ kpama} 
\tbLX{kpanne} \tbHM{} \tbPH{kpànnɪ̀} \tbPS{v.~} \tbSN \tbGE{to hypnotize} \tbSN \tbGE{to be overwhelmed by cold} \tbPD{kpannɛɛ, kpanna, kpanna, kpanema, kpannaa, - } 
\tbLX{kpantoli} \tbHM{} \tbPH{kpántólí} \tbPS{n.~} \tbSN \tbGE{patch of hair as a hairstyle} \tbSN \tbGE{crest (of a bird)} \tbOP{kpantolo} \tbTP{kpantolee} \tbVA{kpontoli; kpontolo' kpontolee}
\tbLX{kpaŋaa} \tbHM{} \tbPH{kpàŋàà} \tbPS{adj.~} \tbGE{large and hollow} 
\tbLX{kpaŋe} \tbHM{} \tbPH{kpáŋ} \tbPS{v.~} \tbGE{to become rigid and hard (of a surface)} \tbPD{kpaŋɛɛ, kpana, kpana, kpanema, kpanaa, - } 
\tbLX{kpaŋe} \tbHM{} \tbPH{kpàŋɪ́} \tbPS{adj.~} \tbGE{rigid and hard (surface, object)} \tbOP{kpama} 
\tbLX{kpaŋenyaŋa} \tbHM{} \tbPH{kpàŋɪ̀nyàŋà} \tbPS{adv.~} \tbGE{large and dirty (container)} 
\tbLX{kpaŋkpane} \tbHM{} \tbPH{kpáŋkpánɪ́} \tbPS{n.~} \tbGE{arm} \tbOP{kpaŋkpama} \tbTP{kpaŋkpannɛɛ} 
\tbLX{kpaŋkpannyugbieli} \tbHM{} \tbPH{kpàŋkpànnyúgbíélí} \tbPS{n.~} \tbGE{elbow} \tbOP{kpaŋkpannyugbiele} \tbTP{kpaŋkpannyugbielee} \tbVA{gbieli}
\tbLX{kpaŋkpaŋkpe᷉ɛ} \tbHM{} \tbPH{kpǎŋkpǎŋkpɪ́᷈ɛ́᷉} \tbPS{n.~} \tbGE{force} \tbOP{kpaŋkpaŋkpeɛne} \tbTP{kpaŋkpaŋkpeɛnɛɛ} 
\tbLX{kpaŋkpazegle} \tbHM{} \tbPH{kpáŋkpázɪ́glí} \tbPS{n.~} \tbGE{hard, pimple-like swelling} \tbOP{kpaŋkpazeglɛ} \tbTP{kpaŋkpazeglɛɛ} \tbVA{zaŋkpegiri}
\tbLX{kpaŋkpele} \tbHM{} \tbPH{kpáŋkpɪ́lɪ́} \tbPS{adj.~} \tbGE{extremely hard} 
\tbLX{kpaŋkpire} \tbHM{} \tbPH{kpǎŋkpírè} \tbPS{n.~} \tbGE{race, running competition ( typically used with ŋmɛ)} \tbXV{Ba ŋmɛ la kpaŋkpire te le dɔgele taa.} \tbXE{They raced neck and neck and fell over each other.} \tbOP{kpaŋkpirri} \tbTP{kpaŋkpiree/kpaŋkpirree} 
\tbLX{kpaŋkpoli} \tbHM{} \tbPH{kpáŋkpòlì} \tbPS{n.~} \tbGE{chicken without feathers} \tbOP{kpaŋkpolo} \tbTP{kpaŋkpolee} \tbVA{kpoŋkpoli}
\tbLX{kparata} \tbHM{} \tbPH{kpárátá} \tbPS{adv.~} \tbGE{lie prostrate; flat with face upwards} 
\tbLX{kpardiiraa} \tbHM{} \tbPH{kpǎrdííráá} \tbPS{n.~} \tbGE{smock with multiple folds} \tbOP{kpardiiri} \tbTP{kpardiiree} 
\tbLX{Kpare} \tbHM{} \tbPH{kpárɪ́} \tbPS{n.~} \tbGE{name of a town in the Upper West Region} 
\tbLX{kpare} \tbHM{} \tbPH{kpárɪ́} \tbPS{v.~} \tbSN \tbGE{to tie together (to make a bow)} \tbSN \tbGE{to suffer, to stuggle on the ground} \tbPD{kparɛɛ, kpara, kpara, kpareba, kparaa, - } 
\tbLX{kpare} \tbHM{} \tbPH{kpàrɪ̀} \tbPS{v.~} \tbSN \tbGE{to pair up} \tbSN \tbGE{closed to} \tbSN \tbGE{to ride one bike or animal together (two or more people)} \tbPD{kparɛɛ, kpara, kpara, kpareba, kparaa, - } 
\tbLX{kparoo} \tbHM{} \tbPH{kpàrʊ́ʊ̀} \tbPS{n.~} \tbGE{top (clothing)} \tbOP{kparre} \tbTP{kparrɛɛ} 
\tbLX{kparpi᷉i} \tbHM{} \tbPH{kpàrpí᷉ì᷉} \tbPS{n.~} \tbGE{needle} \tbOP{kparpiime} \tbTP{kparpiimee/kparpiinee} 
\tbLX{kparsirikyi} \tbHM{} \tbPH{kpǎrsíríʧì} \tbPS{n.~} \tbGE{expensive beautiful gown} \tbOP{kparsirikyiri} \tbTP{kparsirikyiree} 
\tbLX{kparyoluŋ} \tbHM{} \tbPH{kpǎryòlǔŋ} \tbPS{n.~} \tbGE{top wear without a supporting underwear} \tbOP{kparyolli} \tbTP{kparyollee} 
\tbLX{kpateɛrɛ} \tbHM{} \tbPH{kpàtɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{supporter, backer} \tbOP{kpateɛrebɛ} \tbTP{kpateɛrebɛɛ} 
\tbLX{kpatere} \tbHM{} \tbPH{kpàtɪ́rɪ́} \tbPS{n.~} \tbSN \tbGE{gorilla} \tbSN \tbGE{chimpanzee} \tbOP{kpateɛ} \tbTP{kpateerɛɛ} 
\tbLX{kpawerɛ} \tbHM{} \tbPH{kpàwɪ́rɛ́} \tbPS{n.~} \tbGE{back case; not concerned} 
\tbLX{kpawoɔ} \tbHM{} \tbPH{kpàwʊ́ɔ} \tbPS{n.~} \tbGE{hat (especially one made from cloth) (northern Dagaare dialect)} \tbOP{kpawoore} \tbTP{kpawoorɛɛ} 
\tbLX{kpe} \tbHM{} \tbPH{kpɪ́} \tbPS{v.~} \tbSN \tbGE{to shell a bit of (maize seed from cob)} \tbSN \tbGE{to wink (to give someone a sign)} \tbPD{kpeɛ, kpeerɛ, kpeerɛ, kpeerebɛ, kpeeraa, - } 
\tbLX{kpeebɛ} \tbHM{} \tbPH{kpɪ́ɪ́bɛ̀} \tbPS{adj.~} \tbGE{orphaned} \tbOP{kpeere} \tbTP{kpeerɛɛ} 
\tbLX{kpeebo} \tbHM{} \tbPH{kpɪ̀ɪ̀bʊ́} \tbPS{n.~} \tbGE{orphan} \tbOP{kpeere} \tbTP{kpeerɛɛ} 
\tbLX{kpe᷉ekpe᷉e} \tbHM{} \tbPH{kpɪ́᷉ɪ́kpɪ́᷉ɪ́} \tbPS{adv.~} \tbGE{not bright (fire)} 
\tbLX{kpeele} \tbHM{} \tbPH{kpɪ́ɪ́lɪ̀} \tbPS{v.~} \tbGE{to bend slightly to the side} \tbPD{kpeelɛɛ, kpeelɛ, kpeelɛ, kpeelbɛ, kpeelaa, - } 
\tbLX{kpeele} \tbHM{} \tbPH{kpɪ̀ɪ́lɪ̀} \tbPS{adj.~} \tbGE{small, tiny, immature} \tbOP{kpeelɛ} 
\tbLX{kpeeloŋ} \tbHM{} \tbPH{kpɪ̀ɪ̀lʊ́ŋ} \tbPS{n.~} \tbGE{the state of being an orphan} 
\tbLX{kpeeloŋ} \tbHM{} \tbPH{kpɪ́ɪ́lʊ́ŋ} \tbPS{n.~} \tbGE{hip joint} \tbOP{kpeelɛ} \tbTP{kpeelɛɛ} 
\tbLX{kpeemɛ} \tbHM{} \tbPH{kpɪ́ɪ́mɛ́} \tbPS{n.~} \tbGE{departed soul, the dead} \tbOP{kpenne} \tbTP{kpeemɛɛ/kpennɛɛ} 
\tbLX{kpeeraa} \tbHM{} \tbPH{kpɪ́ɪ́ráà} \tbPS{n.~} \tbGE{whip} \tbOP{kpeere} \tbTP{kpeerɛɛ} \tbVA{kpaaraa}
\tbLX{kpe᷉ɛ} \tbHM{} \tbPH{kpɪ́᷉ɛ́} \tbPS{n.~} \tbGE{strength, power, force} \tbOP{kpeɛne} \tbTP{kpeɛnɛɛ} \tbVA{kpeɛ᷉o}
\tbLX{kpe᷉ɛ} \tbHM{} \tbPH{kpɪ̀᷉ɛ̀} \tbPS{v.~} \tbGE{to be older than} \tbPD{kpeɛ᷉ɛ, kpeɛnɛ, kpeɛnɛ, kpeɛnemɛ, kpeɛnaa, - } 
\tbLX{kpeɛmɛ} \tbHM{} \tbPH{kpɪ́ɛ́mɛ́} \tbPS{n.~} \tbGE{older sister} \tbOP{kpeɛmine} \tbTP{kpeɛmɛɛ} \tbVA{kpe᷉ɛ᷉ɛ}
\tbLX{kpeɛŋaa} \tbHM{} \tbPH{kpɪ́ɛ́ŋáá} \tbPS{adj.~} \tbSN \tbGE{hard, strong, difficult} \tbSN \tbGE{well, healthy} \tbOP{kpeɛne} 
\tbLX{kpe᷉ɛo} \tbHM{} \tbPH{kpɪ́᷉ɛ́ʊ̀} \tbPS{n.~} \tbGE{strength} \tbOP{kpeɛne} \tbVA{kpe᷉ɛ}
\tbLX{kpeɛre} \tbHM{} \tbPH{kpɪ̀ɛ̀rɪ̀} \tbPS{n.~} \tbGE{popular entrance of a rodent's hole} \tbTP{kpeɛrɛɛ} 
\tbLX{kpeɛrentɔ} \tbHM{} \tbPH{kpɪ́ɛ́rɪ́ntɔ̀} \tbPS{n.~} \tbGE{neighbour} 
\tbLX{kpegeke} \tbHM{} \tbPH{kpegeke} \tbPS{adv.~} \tbGE{direct to the point (usually unleashing pain)} 
\tbLX{kpegere} \tbHM{} \tbPH{kpɪ́grɪ̀} \tbPS{v.~} \tbGE{to be curious (asking questions to find information)} \tbPD{kpegerɛɛ, kpegerɛ, kpegerɛ, kpegerebɛ, kpegeraa, kpegerebɛ } 
\tbLX{kpegileŋ} \tbHM{} \tbPH{kpégléŋ} \tbPS{adv.~} \tbGE{small but compact} 
\tbLX{kpegiri} \tbHM{} \tbPH{kpégrí} \tbPS{n.~} \tbGE{gizzard} \tbOP{kpege} \tbTP{kpegiree} 
\tbLX{kpegiri} \tbHM{} \tbPH{kpégrì} \tbPS{v.~} \tbSN \tbGE{to shiver (from cold)} \tbSN \tbGE{to move one's torso quickly and vigorously (usually in dancing)} \tbPD{kpegiree, kpegire, kpegire, kpegiribe, kpegiraa, - } 
\tbLX{kpegiruu} \tbHM{} \tbPH{kpégrúú} \tbPS{n.~} \tbGE{a type of dance} 
\tbLX{kpekpe} \tbHM{} \tbPH{kpɪ́kpɪ́} \tbPS{n.~} \tbGE{pair of scissors} \tbOP{kpekpere} \tbTP{kpekperɛɛ} 
\tbLX{kpele} \tbHM{} \tbPH{kpɪ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to develop a bad eye (black of eye looking whitish)} \tbSN \tbGE{to cause to be bare (growth)} \tbPD{kpelɛɛ, kpelɛ, kpelɛ, kpelbɛ, kpelaa, - } 
\tbLX{kpele} \tbHM{} \tbPH{kpɪ̀lɪ̀} \tbPS{adj.~} \tbGE{hard, difficult} \tbOP{kpelɛ} \tbTP{kpelɛɛ} 
\tbLX{kpele} \tbHM{} \tbPH{kpélé} \tbPS{adv.~} \tbGE{break completely} \tbVA{kpelekpele}
\tbLX{kpeleŋkpeleŋ} \tbHM{} \tbPH{kpèlèŋkpéléŋ} \tbPS{adj.~} \tbGE{uncompromising state (quarel)} 
\tbLX{kpelɛ} \tbHM{} \tbPH{kpɪ́lɛ́} \tbPS{n.~} \tbGE{nothing (typically used with negative forms 'ba', koŋ)} \tbXV{Ka fooŋ gaa a bondaana ŋa zie kyɛ fo koŋ nyɛ kpelɛ zaa.} \tbXE{Even if you go to this this rich you will not get anything.} 
\tbLX{kpeli} \tbHM{} \tbPH{kpélí} \tbPS{n.~} \tbGE{throw of dice in the game bie} \tbOP{kpele} \tbTP{kpelee} \tbVA{kpoli}
\tbLX{kpell} \tbHM{} \tbPH{kpéll} \tbPS{adv.~} \tbGE{breaking of a hard objecct} 
\tbLX{kpell} \tbHM{} \tbPH{kpɪ́ll} \tbPS{adv.~} \tbGE{spread out openly and in great quantity} 
\tbLX{kpelle} \tbHM{} \tbPH{kpɪ́llɪ̀} \tbPS{n.~} \tbGE{grass strings worn on the waist} \tbTP{kpellɛɛ} 
\tbLX{kpeme} \tbHM{} \tbPH{kpèmé} \tbPS{n.~} \tbGE{grass used for making hat} \tbOP{kpemee} \tbTP{kpennee} 
\tbLX{kpeme} \tbHM{} \tbPH{kpémé} \tbPS{adv.~} \tbGE{exerting something hefty (high price)} 
\tbLX{kpememe} \tbHM{} \tbPH{kpɪ́mɪ́mɪ́} \tbPS{adv.~} \tbGE{very sweet} 
\tbLX{kpenne} \tbHM{} \tbPH{kpɪ́nnɪ̀} \tbPS{v.~} \tbGE{to break off bits of the tip (of a fire log)} \tbPD{kpennɛɛ, kpennɛ, kpennɛ, kpenemɛ, kpennaa, kpenemɛ } 
\tbLX{kpenne} \tbHM{} \tbPH{kpɪ́nnɪ̀} \tbPS{n.~} \tbGE{ancestor} \tbOP{kpemɛ} \tbTP{kpennɛɛ} 
\tbLX{kpeŋkpegiri} \tbHM{} \tbPH{kpéŋkpégrí} \tbPS{n.~} \tbGE{highest point (xylophone sound)} \tbOP{kpeŋkpege} \tbTP{kpeŋkpegiree} 
\tbLX{kpeŋkperɛŋ} \tbHM{} \tbPH{kpɪ́ŋkpɪ́rɛ́ŋ} \tbPS{adv.~} \tbGE{smallish but strong (short person)} 
\tbLX{kperaa} \tbHM{} \tbPH{kpɪ́ráà} \tbPS{n.~} \tbGE{disease of the hip} \tbOP{kperre} 
\tbLX{kperaa} \tbHM{} \tbPH{kpɪ̀ràà} \tbPS{adv.~} \tbGE{exposed, seen clearly} 
\tbLX{kpere} \tbHM{} \tbPH{kpɪ̀rɪ̀} \tbPS{v.~} \tbGE{to be stunted in growth or development} \tbPD{kperɛɛ, kperɛ, kperɛ, kperebɛ, kperaa, - } 
\tbLX{kpere} \tbHM{} \tbPH{kpɪ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to break off an attached bit from the main body} \tbSN \tbGE{to sneeze} \tbPD{kperɛɛ, kperɛ, kperɛ, kperebɛ, kperaa, - } 
\tbLX{kpere} \tbHM{} \tbPH{kpɪ́rɪ́} \tbPS{v.~} \tbGE{to litter} \tbPD{kperɛɛ, kperɛ, kperɛ, kperebɛ, kperaa, - } 
\tbLX{kperefene} \tbHM{} \tbPH{kpɪ́rɪ́fɪ́nɪ́} \tbPS{adj.~} \tbGE{muscular and stocky} 
\tbLX{kperekpere} \tbHM{} \tbPH{kpɪ̀rɪ̀kpɪ́rɪ́} \tbPS{v.~} \tbGE{to struggle} \tbPD{kperekperɛɛ, kperekperɛ, kperekperɛ, kperekperebɛ, kperekperaa, - } 
\tbLX{kperekpere} \tbHM{} \tbPH{kpɪ̀rɪ̀kpɪ̀rɪ̀} \tbPS{adv.~} \tbGE{rough, forceful} \tbXV{O bie ŋa waa la kperekpere zuo, o ŋmare la a kyaanaa.} \tbXE{This child is too rough; he will break the mirror.} 
\tbLX{kperemene} \tbHM{} \tbPH{kpɪ̀rìmɪ̀nɪ̀} \tbPS{adj.~} \tbGE{all over (fruits on a tree or scratches on the body)} \tbXV{A maŋgo wɔŋe,ɛ la kperemenee lɛ.} \tbXE{The mango tree has borne a lot of fruit.} \tbVA{kpurimunu}
\tbLX{kperemɛŋ} \tbHM{} \tbPH{kpɪ́rɪ̀mɛ́ŋ} \tbPS{adv.~} \tbGE{stunted, short, and small} 
\tbLX{kperese} \tbHM{} \tbPH{kpírɪ́sɪ́} \tbPS{adv.~} \tbGE{very sweet} 
\tbLX{kperɛ} \tbHM{} \tbPH{kpɪ́rɛ̀} \tbPS{adj.~} \tbSN \tbGE{small} \tbSN \tbGE{worn out (broom)} 
\tbLX{kperɛkperɛ} \tbHM{} \tbPH{kpɪ́rɛ̀kpírɛ̀} \tbPS{adj.~} \tbGE{little bits} 
\tbLX{kperi} \tbHM{} \tbPH{kpérí} \tbPS{v.~} \tbGE{to chop (meat, tree branches)} \tbPD{kperee, kpere, kpere, kperibee, kperaa, - } 
\tbLX{kperime} \tbHM{} \tbPH{kpérímé} \tbPS{n.~} \tbGE{type of food (full of chopped meat)} \tbTP{kperimee} 
\tbLX{kperiwere} \tbHM{} \tbPH{kpéríwéré} \tbPS{adv.~} \tbGE{few, not much} 
\tbLX{kpɛ} \tbHM{} \tbPH{kpɛ̀} \tbPS{v.~} \tbSN \tbGE{to enter} \tbSN \tbGE{to dwell in, to inhabit} \tbSN \tbGE{to gain fertility (farmland)} \tbSN \tbGE{to fit (dress)} \tbSN \tbGE{to be well attended} \tbSN \tbGE{to feel hurt} \tbSN \tbGE{to come around (of the a season)} \tbPD{kpɛɛ, kpeɛrɛ, kpeɛrɛ, kpeɛrebɛ, kpeɛraa, - } 
\tbLX{kpɛ bire} \tbHM{} \tbPH{kpɛ̀ bìrè} \tbPS{v.~} \tbGE{to struggle, to wrestle, to contend} \tbPD{-, -, -, -, -, - } 
\tbLX{kpɛde} \tbHM{} \tbPH{kpɛ́dɪ́} \tbPS{n.~} \tbGE{ownership (wealth)} 
\tbLX{kpɛɛ} \tbHM{} \tbPH{kpɛ́ɛ̀} \tbPS{n.~} \tbGE{malt} \tbTP{kpɛɛrɛɛ} 
\tbLX{kpɛgelaa} \tbHM{} \tbPH{kpɛ́gláá} \tbPS{n.~} \tbSN \tbGE{supporting object (for a boiling pot on a hearth.)} \tbSN \tbGE{attachment} \tbOP{kpɛgele} \tbTP{kpɛgelɛɛ} 
\tbLX{kpɛgele} \tbHM{} \tbPH{kpɛ̀glɪ̀} \tbPS{v.~} \tbSN \tbGE{to place by the side (provide support for a cooking pot on the hearth)} \tbSN \tbGE{to adjoin} \tbPD{kpɛgelɛɛ, kpɛgelɛ, kpɛgelɛ, kpɛgelebɛ, kpɛgelaa, - } 
\tbLX{kpɛgelɛŋ} \tbHM{} \tbPH{kpɛ̀glɛ̀ŋ} \tbPS{adv.~} \tbGE{loose and flabby (diseased arm)} 
\tbLX{kpɛgere} \tbHM{} \tbPH{kpɛ́grì} \tbPS{v.~} \tbGE{to disturb or worry someone} \tbPD{kpɛgerɛɛ, kpɛgerɛ, kpɛgerɛ, kpɛgerebɛ, kpɛgeraa, - } 
\tbLX{kpɛle} \tbHM{} \tbPH{kpɛ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to faint} \tbSN \tbGE{to crave} \tbXV{O nyɛ la pegebil velaa kaŋa a deɛ kpɛlɛ o yɛlɛ.} \tbXE{He is just crazy about a beautiful girl he saw.} \tbPD{kpɛlɛɛ, kpɛlɛ, kpɛlɛ, kpɛlebɛ, kpɛlaa, kpɛlebɛ } 
\tbLX{kpɛle} \tbHM{} \tbPH{kpɛ́lɪ̀} \tbPS{v.~} \tbGE{to pluck} \tbPD{kpɛlɛɛ, kpɛlɛ, kpɛlɛ, kpɛlebɛ, kpɛlaa, kpɛlebɛ } 
\tbLX{kpɛlemɛ} \tbHM{} \tbPH{kpɛ́lɪ́mɛ̀} \tbPS{adj.~} \tbGE{snatch, do or act at once (hungry dog snatching a bone)} 
\tbLX{kpɛlewɛlɛ} \tbHM{} \tbPH{kpɛ́lɪ́wɛ́lɛ́} \tbPS{adv.~} \tbGE{not firm or strong} 
\tbLX{kpɛlewɛlɛ} \tbHM{} \tbPH{kpɛ́lwɛ́lɛ́} \tbPS{adv.~} \tbGE{weak and loose (pair of scissors)} 
\tbLX{kpɛlɛŋ} \tbHM{} \tbPH{kpɛ́lɛ́ŋ} \tbPS{adv.~} \tbGE{everything} 
\tbLX{kpɛlɛŋkpɛlɛŋ} \tbHM{} \tbPH{kpɛ́lɛ́ŋkpɛ́lɛ́ŋ} \tbPS{adv.~} \tbGE{finished copletely} \tbVA{[kpɛ́lɛ́ŋkpɛ̀lɛ̀ŋ]}
\tbLX{kpɛlɛŋ kpɛlɛŋ} \tbHM{} \tbPH{kpɛ́lɛ́ŋkpɛ̀lɛ̀ŋ} \tbPS{adv.~} \tbGE{finish completely} 
\tbLX{kpɛlle} \tbHM{} \tbPH{kpɛ́llɪ́} \tbPS{n.~} \tbGE{loud and sharp cry} \tbTP{kpɛllɛɛ} 
\tbLX{kpɛnezu} \tbHM{} \tbPH{kpɛ́nɪ́zú} \tbPS{n.~} \tbGE{a small black blood-sucking fly; causes river blindness} \tbOP{kpɛnezuri} \tbTP{kpɛnezurree} 
\tbLX{kpɛ nyuo} \tbHM{} \tbPH{kpɛ̀ ɲúó} \tbPS{v.~} \tbGE{to seek protection from} \tbPD{-, -, -, -, -, - } 
\tbLX{kpɛŋkpɛlemɛ} \tbHM{} \tbPH{kpɛ̌ŋkpɛ́lɪ́mɛ́} \tbPS{n.~} \tbGE{bits and pieces} \tbXV{A bɔgebogera die kpɛŋkpɛlemɛ ba di ba.} \tbXE{There are very many pieces of items in the diviner's room.} \tbTP{kpɛŋkpɛlemɛɛ} 
\tbLX{kpɛŋkpɛlɛŋ} \tbHM{} \tbPH{kpɛ́ŋkpɛ́lɛ́ŋ} \tbPS{adv.~} \tbGE{of a hanging bit} 
\tbLX{kpɛŋkpɛrɛŋ} \tbHM{} \tbPH{kpɛ́ŋkpɛ́rɛ́ŋ} \tbPS{adj.~} \tbGE{spilled all over} 
\tbLX{kpɛre} \tbHM{} \tbPH{kpɛ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to spill} \tbSN \tbGE{to cook a small quantity of} \tbPD{kpɛrɛɛ, kpɛrɛ, kpɛrɛ, kpɛrebɛ, kpɛraa, - } 
\tbLX{kpɛrewɛrɛ} \tbHM{} \tbPH{kpɛ́rɪ́wɛ́rɛ́} \tbPS{adv.~} \tbGE{small quantity} \tbXV{Zenɛ N koŋ to᷉ɔ nɛne da, n libie deɛ waa la kpɛrewɛrɛɛ lɛ.} \tbXE{Today I cannot buy meat; my money is very small.} \tbVA{kpɛrewɛrɛɛ}
\tbLX{kpɛrɛkpɛrɛ} \tbHM{} \tbPH{kpɛ́rɛ́kpɛ́rɛ́} \tbPS{adv.~} \tbGE{bits and pieces} 
\tbLX{kpɛrɛndɛŋ} \tbHM{} \tbPH{kpɛ́rɛ́ndɛ́ŋ} \tbPS{adv.~} \tbGE{best of (deed)} \tbXV{O maale la a bondirii de wa biŋi kpɛndɛŋ.} \tbXE{She cooked the food and served it neatly.} 
\tbLX{kpɛrɛŋkpɛrɛŋ} \tbHM{} \tbPH{kpɛ́rɛ́ŋkpɛ́rɛ́ŋ} \tbPS{adv.~} \tbGE{smeared all over, in a mess} \tbXV{A bie le la mɛmɛrɛɛ poɔ a e o zaa kpɛrɛŋkpɛrɛŋ lɛ} \tbXE{The boy fell in the mud and is soiled all over.} 
\tbLX{kpɛsenneŋenɛ} \tbHM{} \tbPH{kpɛ́sɪ́nnɪ́ŋɪ́nɛ́} \tbPS{n.~} \tbGE{pulp from pito malt} 
\tbLX{kpɛtaa} \tbHM{} \tbPH{kpɛ́táá} \tbPS{n.~} \tbSN \tbGE{be suited to; fit well} \tbSN \tbGE{a union in which the group members work for each other in turns} \tbOP{kpɛtaare} \tbTP{kpɛtaarɛɛ} 
\tbLX{kpɛwe᷉ɛ} \tbHM{} \tbPH{kpɛ́wɪ́᷉ɛ́} \tbPS{n.~} \tbGE{water from prepared malt} \tbOP{kpɛweɛne} \tbTP{kpɛweɛnɛɛ} 
\tbLX{kpɛzie} \tbHM{} \tbPH{kpɛ̀zíé} \tbPS{n.~} \tbGE{abode, living place} \tbOP{kpɛziiri} \tbTP{kpɛziiree} 
\tbLX{kpi} \tbHM{} \tbPH{kpî} \tbPS{v.~} \tbSN \tbGE{to die, to be dead} \tbSN \tbGE{to wither (plant)} \tbPD{kpie, kpiire, kpiire, kpiiribe, kpiiraa, kpiiribe } 
\tbLX{kpibifi} \tbHM{} \tbPH{kpíbífí} \tbPS{adj.~} \tbGE{small, tiny} 
\tbLX{kpieni} \tbHM{} \tbPH{kpíéní} \tbPS{adj.~} \tbGE{snatch (hawk picking a chicken)} 
\tbLX{kpiiri} \tbHM{} \tbPH{kpììrì} \tbPS{v.~} \tbGE{to sew the end of; to hem (dress)} \tbPD{kpiiree, kpiire, kpiire, kpiiribe, kpiiraa, kpiiribe } 
\tbLX{kpimare} \tbHM{} \tbPH{kpìmàrɪ̀} \tbPS{v.~} \tbGE{to be glued to (work)} \tbPD{-, -, -, -, -, - } 
\tbLX{kpi mare} \tbHM{} \tbPH{kpì màrɪ̀} \tbPS{v.~} \tbGE{to cling to, to persist} \tbPD{kpi marɛɛ, kpiire mara, kpimara, kpimareba, kpimaraa, - } 
\tbLX{kpinni} \tbHM{} \tbPH{kpìnnì} \tbPS{v.~} \tbGE{to quench, to extinguish} \tbPD{kpinnee, kpinne, kpinne, kpinime, kpinnaa, kpinime } 
\tbLX{kpinni} \tbHM{} \tbPH{kpínnì} \tbPS{n.~} \tbGE{remind} \tbVA{pinni}
\tbLX{kpirikpiri} \tbHM{} \tbPH{kpíríkpírí} \tbPS{adv.~} \tbGE{very black} \tbVA{malmal, bɛmbɛŋ}
\tbLX{kpiroo} \tbHM{} \tbPH{kpíróó} \tbPS{adv.~} \tbGE{protruding in a small manner} 
\tbLX{kpo} \tbHM{} \tbPH{kpo} \tbPS{v.~} \tbSN \tbGE{to break open an egg} \tbSN \tbGE{to remove the bark of a tree} \tbPD{kpoe, kpuoro, kpuoro, kpuoribo, kpuoraa, - } 
\tbLX{kpo} \tbHM{} \tbPH{kpô} \tbPS{n.~} \tbSN \tbGE{funeral donation} \tbSN \tbGE{meeting place} \tbOP{kpori} \tbTP{kporee} 
\tbLX{kpogeraa} \tbHM{} \tbPH{kpʊ́gráá} \tbPS{n.~} \tbGE{haenia} \tbOP{kpogere} \tbTP{kpogerɛɛ} 
\tbLX{kpogi} \tbHM{} \tbPH{kpògì} \tbPS{v.~} \tbGE{to bully} \tbPD{kpogee, kpogro, kpogro, kpogribo, kpograa, - } 
\tbLX{kpogili} \tbHM{} \tbPH{kpòglì} \tbPS{n.~} \tbGE{helmet} \tbOP{kpogilo} \tbTP{kpogilee} 
\tbLX{kpogili} \tbHM{} \tbPH{kpòglì} \tbPS{v.~} \tbGE{to mature, to grow big} \tbPD{kpogilee, kpogilo, kpogilo, kpogilibo, kpogilaa, - } 
\tbLX{kpogili} \tbHM{} \tbPH{kpóglí} \tbPS{v.~} \tbGE{to wait without getting attention; to sit aimlessly} \tbPD{kpogilee, kpogilo, kpogilo, kpogilibo, kpogilaa, - } 
\tbLX{kpogilo} \tbHM{} \tbPH{kpóglò} \tbPS{n.~} \tbGE{boiled balls of bean and corn flour} \tbSG{kpogiluu} \tbTP{kpogilee} 
\tbLX{kpogilokolo} \tbHM{} \tbPH{kpòglòkòlò} \tbPS{adv.~} \tbGE{appearance (many big items)} 
\tbLX{kpogiri} \tbHM{} \tbPH{kpógrí} \tbPS{v.~} \tbGE{to tap the inner side of a mortar (creating a beat)} \tbPD{kpogiree, kpogiro, kpogiro, kpogiribo, kpogiraa, - } 
\tbLX{kpogo} \tbHM{} \tbPH{kpógó} \tbPS{adv.~} \tbGE{refusing utterly to do something; (run an erand)} 
\tbLX{kpokpo} \tbHM{} \tbPH{kpókpó} \tbPS{adj.~} \tbGE{permanent} 
\tbLX{kpokpori} \tbHM{} \tbPH{kpòkpórí} \tbPS{n.~} \tbGE{mound} \tbOP{kpokporee} \tbTP{kpokporuu} 
\tbLX{kpolaa} \tbHM{} \tbPH{kpʊ̀làà} \tbPS{adj.~} \tbGE{large and blunt} 
\tbLX{kpolaa} \tbHM{} \tbPH{kpòlàà} \tbPS{adv.~} \tbGE{not hidden (truth)} 
\tbLX{kpolee} \tbHM{} \tbPH{kpóléé} \tbPS{adj.~} \tbGE{small and round} \tbOP{kpolee kpolee} 
\tbLX{kpoli} \tbHM{} \tbPH{kpólí} \tbPS{v.~} \tbSN \tbGE{to get to extreme; (red hot metal)} \tbSN \tbGE{to go bad, to spoil (rotten eggs)} \tbPD{kpolee, kpolo, kpolo, kpolibo, kpolaa, - } 
\tbLX{kpoli} \tbHM{} \tbPH{kpòlì} \tbPS{adj.~} \tbGE{bare, naked} \tbOP{kpolo} \tbTP{kpolee/kpollee} 
\tbLX{kpoli} \tbHM{} \tbPH{kpólí} \tbPS{n.~} \tbGE{a player letting go a dice of the opponent in the game of "bie" as a sign of total control of the game} \tbOP{kpoli} \tbTP{kpolee} \tbVA{kpeli}
\tbLX{kpolikpala} \tbHM{} \tbPH{kpòlìkpàlå} \tbPS{adv.~} \tbGE{appearance of something big with appendages} 
\tbLX{kpoliŋkpoori} \tbHM{} \tbPH{kpólíŋkpòòrí} \tbPS{n.~} \tbGE{a type of small ants} \tbOP{kpoliŋkpogo} \tbTP{kpoliŋkpooree} 
\tbLX{kpolo} \tbHM{} \tbPH{kpòló} \tbPS{n.~} \tbGE{termites (trapped for chicken feed.)} \tbSG{kpoluu} \tbTP{kpolee} 
\tbLX{kpoloŋkpoloŋ} \tbHM{} \tbPH{kpólóŋkpólóŋ} \tbPS{adj.~} \tbGE{bare and clean (shaven head)} 
\tbLX{kpolore} \tbHM{} \tbPH{kpʊ́lʊ́rɪ́} \tbPS{n.~} \tbGE{small, gourdlike flute} \tbOP{kpoloɛ} \tbTP{kpolorɛɛ} 
\tbLX{kpolɔ} \tbHM{} \tbPH{kpʊ̀lɔ̀} \tbPS{n.~} \tbGE{inner feeling; anger} \tbVA{kpola}
\tbLX{kpommu} \tbHM{} \tbPH{kpòmmú} \tbPS{n.~} \tbGE{cassava; plant} \tbOP{kpoŋŋo} \tbTP{kponnee} 
\tbLX{kponni} \tbHM{} \tbPH{kpónnì} \tbPS{v.~} \tbSN \tbGE{to hit a rigid object to clean of dirt (dust)} \tbSN \tbGE{to empty powdery stuff completely from a container} \tbPD{kponnee, kponno, kponno, kponimo, kponnaa, - } 
\tbLX{kpontoli} \tbHM{} \tbPH{kpóntólí} \tbPS{n.~} \tbGE{} \tbVA{kpantoli}
\tbLX{kpoŋi} \tbHM{} \tbPH{kpôŋ} \tbPS{adj.~} \tbGE{big, large, great} \tbOP{kponni} \tbTP{kponnee} 
\tbLX{kpoŋi} \tbHM{} \tbPH{kpôŋ} \tbPS{v.~} \tbGE{to knock on the head slightly} \tbPD{kpoŋee, kponno, kponno, kponimo, kponnaa, - } 
\tbLX{kpoŋkpogiri} \tbHM{} \tbPH{kpóŋkpógrì} \tbPS{n.~} \tbGE{wickedness; deliberate unfriendly behaviour} 
\tbLX{kpoŋkpoli} \tbHM{} \tbPH{kpókpòlì} \tbPS{n.~} \tbGE{} \tbOP{kpoŋkpolo} \tbTP{kpoŋkpolee} \tbVA{kpaŋkpoli}
\tbLX{kpoŋkpolimo} \tbHM{} \tbPH{kpǒŋkpólímó} \tbPS{n.~} \tbGE{coop for fowl} \tbOP{kpoŋkpolli} \tbTP{kpoŋkpolee} 
\tbLX{kpoŋkpolo} \tbHM{} \tbPH{kpóŋkpóló} \tbPS{n.~} \tbGE{second-rate xylophone notes (usually laid out on the ground)} \tbSG{kpoŋkpoluu} \tbTP{kpoŋkpolee} 
\tbLX{kpoŋkponti} \tbHM{} \tbPH{kpòŋkpòntí} \tbPS{n.~} \tbSN \tbGE{dried cassava tubers} \tbSN \tbGE{food made from cassava flour} \tbOP{kpoŋkpontiri} \tbTP{kpoŋkpontiree} 
\tbLX{kpoŋŋo} \tbHM{} \tbPH{kpǒŋŋó} \tbPS{n.~} \tbSN \tbGE{cassava plant} \tbSN \tbGE{dried cassava tubers} \tbOP{kponnee} 
\tbLX{kpoo} \tbHM{} \tbPH{kpóó} \tbPS{adj.~} \tbGE{expression of surprise and dismay} \tbVA{kpaa}
\tbLX{kpore} \tbHM{} \tbPH{kpʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to crack one's knuckles} \tbSN \tbGE{to seal completely (hole)} \tbPD{kporɛɛ, kpora, kporɔ, kporeba, kporaa, - } 
\tbLX{kpore} \tbHM{1} \tbPH{kpʊ́rɪ́} \tbPS{n.~} \tbGE{snuffbox (small container for snuff)} \tbOP{ kporɛɛ} 
\tbLX{kpore} \tbHM{2} \tbPH{kpʊ́rɪ́} \tbPS{n.~} \tbGE{lump (on wood)} \tbOP{kporɛɛ} \tbTP{kporrɛɛ} 
\tbLX{kpori} \tbHM{} \tbPH{kpórì} \tbPS{v.~} \tbGE{to empty} \tbXV{O kpori la o woɔ libie zaa da ne yiri.} \tbXE{He emptied his purse to buy the house.} \tbPD{kporee, kporo, kporo, kporibo, kporaa, - } 
\tbLX{kpoteɛ} \tbHM{} \tbPH{kpʊ́tɪ́ɛ́} \tbPS{n.~} \tbGE{lumps, protrusions} 
\tbLX{kpɔge} \tbHM{} \tbPH{kpɔ́gɪ́} \tbPS{v.~} \tbGE{to develop a lump; to swell up} \tbPD{kpɔgɛɛ, kpɔgrɔ, kpɔgrɔ, kpɔgreba, kpɔgraa, - } 
\tbLX{kpɔge} \tbHM{} \tbPH{kpɔ́gɪ̀} \tbPS{v.~} \tbGE{to carry, to pick up wholly (mud)} \tbPD{kpɔgɛɛ, kpɔgera, kpɔgera, kpɔgereba, kpɔgeraa, - } 
\tbLX{kpɔge} \tbHM{} \tbPH{kpɔ́gɪ́} \tbPS{v.~} \tbGE{carry (pl)  carrying muddy stuff severally} \tbPD{kpɔgɛɛ, -, -, -, -, - } 
\tbLX{kpɔgeraa} \tbHM{} \tbPH{kpɔ́gràá} \tbPS{n.~} \tbGE{a type of tree with hard but edible fruit} \tbOP{kpɔgere} \tbTP{kpɔgerɛɛ} \tbVA{kpɔgetere}
\tbLX{kpɔgere} \tbHM{} \tbPH{kpɔ́grɪ́} \tbPS{n.~} \tbGE{lump} \tbOP{kpɔgɔ} \tbTP{kpɔgerɛɛ} 
\tbLX{kpɔgetere} \tbHM{} \tbPH{kpɔ́gɪ́tɪ́rɪ́} \tbPS{n.~} \tbGE{} \tbVA{kpɔgeraa}
\tbLX{kpɔlɛɛ} \tbHM{} \tbPH{kpɔ̀lɛ̀ɛ̀} \tbPS{adv.~} \tbGE{blump} 
\tbLX{kpɔlɛɛkpɔlɛɛ} \tbHM{} \tbPH{kpɔ́lɛ́ɛ́kpɔ́lɛ́ɛ́} \tbPS{adv.~} \tbGE{appearing with bulging parts} 
\tbLX{kpɔŋkɔrɔŋ} \tbHM{} \tbPH{kpɔ́ŋkɔ́rɔ́ŋ} \tbPS{adv.~} \tbGE{small and dented} 
\tbLX{kpɔŋkɔrɔŋ} \tbHM{} \tbPH{kpɔ̀ŋkɔ̀rɔ̀ŋ} \tbPS{adv.~} \tbGE{large and dented} 
\tbLX{kpɔre} \tbHM{} \tbPH{kpɔ́rɪ̀} \tbPS{adj.~} \tbGE{weak, broken (pot, calabash)} \tbOP{kpɔrɔ} \tbTP{kpɔrɛɛ} 
\tbLX{kpugiloŋ} \tbHM{} \tbPH{kpúglóŋ} \tbPS{adj.~} \tbGE{small, solid, and round} 
\tbLX{kpugiloŋ} \tbHM{} \tbPH{kpùglòŋ} \tbPS{adj.~} \tbGE{large, solid, and round} 
\tbLX{kpulee} \tbHM{} \tbPH{kpúléé} \tbPS{adv.~} \tbGE{small and rounded on top} 
\tbLX{kpulee} \tbHM{} \tbPH{kpùlèè} \tbPS{adv.~} \tbGE{large and rounded on top} 
\tbLX{kpuleekpulee} \tbHM{} \tbPH{kpúléékpúléé} \tbPS{n.~} \tbGE{groundnut cake} \tbVA{kpulikpuli}
\tbLX{kpuli} \tbHM{} \tbPH{kpúlí} \tbPS{adj.~} \tbSN \tbGE{round-headed} \tbSN \tbGE{hornless (ram)} \tbOP{kpullo} \tbTP{kpulee} 
\tbLX{kpulikpuli} \tbHM{} \tbPH{kpùlìkpúlí} \tbPS{v.~} \tbGE{to walk with difficulty; to stutter (leper)} \tbPD{kpulikpulee, kpulikpulo, kpulikpulo, kpulikpulbo, kpulikpulaa, - } 
\tbLX{kpulli} \tbHM{} \tbPH{kpùllì} \tbPS{v.~} \tbGE{to conclude; to bring to the end} \tbPD{kpullee, kpullo, kpullo, kpulbo, kpullaa, - } 
\tbLX{kpuŋkpuli} \tbHM{} \tbPH{kpúŋkpúlí} \tbPS{n.~} \tbGE{round headed truncheon} \tbOP{kpuŋkpulo} \tbTP{kpuŋkpulee} 
\tbLX{kpurimunu} \tbHM{} \tbPH{kpùrìmùnù} \tbPS{adv.~} \tbSN \tbGE{in great numbers (fruit)} \tbSN \tbGE{of the fall of fruits} \tbVA{kperemene}
\tbLX{kya} \tbHM{} \tbPH{ʧá} \tbPS{v.~} \tbSN \tbGE{to make a gift} \tbSN \tbGE{to be divided, to be separated} \tbSN \tbGE{to split; to gnaw roughly} \tbPD{kyaɛ, kyaara, kyaara, kyareba, kyaaraa, - } 
\tbLX{kya᷉a} \tbHM{} \tbPH{ʧà᷉à} \tbPS{adj.~} \tbSN \tbGE{lit  up} \tbXV{Zie waa la kya᷉a dabe᷉ɛ kyebe.} \tbXE{It is day; there is no need to fear.} \tbSN \tbGE{full of tiny holes so that one can see through (light cloth)} 
\tbLX{kyaakyee} \tbHM{} \tbPH{ʧááʧèè} \tbPS{interjection.~} \tbGE{expression announcing arrival} 
\tbLX{kyaaleŋ} \tbHM{} \tbPH{ʧáálɪ́ŋ} \tbPS{adv.~} \tbGE{no use going back; just do it to put an end to} 
\tbLX{Kyaana} \tbHM{} \tbPH{ʧáánà} \tbPS{n.~} \tbGE{China} \tbTP{Kyaannɛɛ} 
\tbLX{kyaanaa} \tbHM{} \tbPH{ʧáánáá} \tbPS{n.~} \tbGE{a mirror, that which can be seen through} \tbOP{ kyaane} \tbTP{kyaanɛɛ} 
\tbLX{kyaane} \tbHM{} \tbPH{ʧàànɪ̀} \tbPS{v.~} \tbGE{to give light, to make clear} \tbXV{A lɔɔre vu᷉u᷉ kyaane la a sori zaa.} \tbXE{The car's lights lit up the whole road.} \tbPD{kyaanɛɛ, kyaana, kyaana, kyaanema, kyaanaa, - } 
\tbLX{kyaaraa} \tbHM{} \tbPH{ʧááráá} \tbPS{n.~} \tbGE{comb} \tbOP{kyaare} \tbTP{kyaarɛɛ} 
\tbLX{kyaaraa} \tbHM{} \tbPH{ʧààráà} \tbPS{n.~} \tbGE{hall, large room} \tbOP{kyaare} \tbTP{kyaarɛɛ} 
\tbLX{kyaare} \tbHM{} \tbPH{ʧààrɪ̀} \tbPS{v.~} \tbSN \tbGE{to face} \tbXV{Ba zeŋ kyaare la a dendɔre.} \tbXE{They sat facing the door.} \tbSN \tbGE{to have diarrhoea} \tbXV{Ka fooŋ di sɛremaane yaga fo na baŋ kyaare la.} \tbXE{If you eat too much pepper, you may get diarrhoea.} \tbSN \tbGE{to shade} \tbXV{A ŋmenaa tuo de kataweɛ kyaare a bie.} \tbXE{The sun is very hot; please use an umbrella over the child.} \tbSN \tbGE{to comb} \tbXV{A seŋ ka te maŋ kyaare bibiiri zuri baaloŋ.} \tbXE{We should learn to comb children's hair gently.} \tbSN \tbGE{to give in place of; to pay} \tbXV{O de la bɛŋɛ kookoolaa wa kyaare ne zeɛ deni.} \tbXE{She gave a bowl of beans in place of a bowl of millet.} \tbPD{-, kyaara, kyaara, kyaareba, -, - } 
\tbLX{kyaare} \tbHM{} \tbPH{ʧáárɪ́} \tbPS{v.~} \tbSN \tbGE{to exclaim, to shout loudly (in joy or victory)} \tbSN \tbGE{extremely boiled (oil)} \tbPD{kyaarɛɛ, kyaara, kyaara, kyaareba, kyaaraa, - } 
\tbLX{kyaare} \tbHM{} \tbPH{ʧààrɪ̀} \tbPS{adj.~} \tbGE{really} \tbXV{O kyaare boɔrɔ la a pɔge.} \tbXE{He really loves the woman.} 
\tbLX{kyabaa} \tbHM{} \tbPH{ʧábáá} \tbPS{n.~} \tbGE{pincers, something that clamps} \tbOP{kyaebre} \tbTP{kyaberɛɛ} 
\tbLX{kyabo} \tbHM{} \tbPH{ʧábʊ́} \tbPS{adv.~} \tbGE{} \tbVA{kyape᷉}
\tbLX{kya᷈e} \tbHM{} \tbPH{ʧà᷉ɪ̀} \tbPS{adj.~} \tbGE{new, shining} 
\tbLX{kyafo᷉o᷉} \tbHM{} \tbPH{ʧàfʊ̀᷉ʊ̀᷉} \tbPS{adj.~} \tbGE{like foam; ruffled (husks in water)} 
\tbLX{kyagaaloŋ} \tbHM{} \tbPH{ʧágààlʊ́ŋ} \tbPS{n.~} \tbGE{carelessness} \tbXV{O kyagaaloŋ la so ka a zeɛre kɔɔ.} \tbXE{It is because of your carelessness that the soup got burnt.} \tbVA{kyagaleŋ, kyagaloŋ}
\tbLX{kyagalaa} \tbHM{} \tbPH{ʧágálåà} \tbPS{n.~} \tbGE{careless person} 
\tbLX{kyagaleŋ} \tbHM{} \tbPH{ʧágálɪ́ŋ} \tbPS{n.~} \tbGE{} \tbVA{kyagaaloŋ}
\tbLX{kyagaloŋ} \tbHM{} \tbPH{ʧágàlʊ́ŋ} \tbPS{n.~} \tbGE{} \tbVA{kyagaaloŋ}
\tbLX{kyage} \tbHM{} \tbPH{ʧágɪ́} \tbPS{v.~} \tbGE{to check} \tbXV{Kyage a libie noɔre ka te nyɛ.} \tbXE{Check the amount of money; let's see.} \tbPD{kyagɛɛ, kyagera, kyagera, kyagereba, kyageraa, - } 
\tbLX{kyage} \tbHM{} \tbPH{ʧàgɪ́} \tbPS{adj.~} \tbGE{not straight, crooked, twisted} \tbXV{A o lɔɔre gbɛŋmɛ werekao na poɔ la ka o seɛ deɛ kyage.} \tbXE{His waist got twisted as result of the injury he sustained from the lorry accident.} \tbOP{kyagere} \tbTP{kyagerɛɛ} 
\tbLX{kyage} \tbHM{} \tbPH{ʧàgɪ̀} \tbPS{v.~} \tbSN \tbGE{supported in between; (Y-shaped wood)} \tbSN \tbGE{grab firmly with fingers (around the neck)} \tbPD{-, -, -, -, -, - } 
\tbLX{kyagele} \tbHM{} \tbPH{ʧàgɪ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to hold gently} \tbXV{O kyagele la a bie nubaala poɔ gaa ne asibiti.} \tbXE{He held the child's wounded hand gently as he took him to the hospital.} \tbSN \tbGE{to be well} \tbXV{O yeli ka a yiri kyagelɛɛ la.} \tbXE{She said the house was peaceful.} \tbSN \tbGE{to settle temporarily} \tbXV{Vɛŋ ka N kyagele kyɛ ka fo gaa kyɛ wa.} \tbXE{Let me rest here while you go and come back.} \tbPD{kyagelɛɛ, kyagela, kyagela, kyageleba, kyagelaa, - } 
\tbLX{kyagere} \tbHM{} \tbPH{ʧágɪ́rɪ̀} \tbPS{v.~} \tbGE{to hobble, to limp} \tbPD{kyagerɛɛ, kyagera, kyagera, kyagereba, kyageraa, kyagereba } 
\tbLX{kyakoli} \tbHM{} \tbPH{ʧákòlí} \tbPS{n.~} \tbGE{spoon} \tbOP{kyakolo/kyakolli} \tbTP{kyakolee/kyakollee} \tbVA{kyɛkoli}
\tbLX{kyakol-kyaaraa} \tbHM{} \tbPH{ʧákòl-kyááráá} \tbPS{n.~} \tbGE{fork} \tbOP{kyakol-kyaare} \tbTP{kyakol-kyaarɛɛ} 
\tbLX{kyakya} \tbHM{} \tbPH{ʧàʧá} \tbPS{v.~} \tbGE{to talk incessantly, to chirp} \tbPD{kyakyaɛ, kyakyaara, kyakyaara, kyakyareba, kyakyaaraa, - } 
\tbLX{kyåkyá} \tbHM{} \tbPH{ʧáʧá} \tbPS{n.~} \tbGE{betting, gambling, competition} \tbXV{Loto e la e la kyakya naŋ bɛllɛ faara noba libie.} \tbXE{The Lotto is a form of gambling that extracts money from people.} \tbOP{kyakyare} \tbTP{kyakyarɛɛ} 
\tbLX{kyakya᷉a} \tbHM{} \tbPH{ʧàʧá᷉á} \tbPS{adv.~} \tbGE{perforated, full of holes such that you can see through} \tbXV{A pelee waa la kyakya᷉a lɛ o koŋ to᷉ɔ eŋ a kyi.} \tbXE{The basket has sizeable spaces in between its straws and so cannot hold the millet.} 
\tbLX{kyalakyala} \tbHM{} \tbPH{ʧàlàʧàlà} \tbPS{adv.~} \tbSN \tbGE{not thick; having some gaps (vegetation)} \tbXV{A tuu poɔ waa la kyalakyala, a kpa᷉a᷉o᷉ koŋ baŋ sɔgle be.} \tbXE{The thicket is not thick, so the guinea fowl cannot hide there.} \tbSN \tbGE{dark but having some amount of visibility (dawn)} \tbXV{Zie waa la kyalakyala o na nyɛrɛ la a sori.} \tbXE{It is dark but light enough for him to find his way.} 
\tbLX{kyamakyama} \tbHM{} \tbPH{ʧámáʧámá} \tbPS{adv.~} \tbGE{brittle} \tbXV{Nɔgyɛlepɛgere man waa la kyamakyama lɛ.} \tbXE{Eggshells are normally brittle.} 
\tbLX{kyankyampelaa} \tbHM{} \tbPH{ʧánʧámpɪ̀làá} \tbPS{n.~} \tbGE{cattle egret} \tbOP{kyankyampeɛle} \tbTP{kyankyampeɛlɛɛ} 
\tbLX{kyankyamponaa} \tbHM{} \tbPH{ʧánʧámpʊ́nàá} \tbPS{n.~} \tbGE{a type of dark blue insect that emits light at night} \tbOP{kyankyamponne} \tbTP{kyankyamponnɛɛ} 
\tbLX{kyaperɛpe᷉} \tbHM{} \tbPH{ʧàpɪ́rɛ́pɪ́᷉} \tbPS{adv.~} \tbGE{gripping firmly and suddenly} \tbVA{kyape᷉}
\tbLX{Kyapore} \tbHM{} \tbPH{ʧápʊ́rɪ̀} \tbPS{n.~} \tbGE{a village in the Upper West Region of Ghana} \tbOP{kyaporre} \tbTP{kyaporrɛɛ} 
\tbLX{kyaporo} \tbHM{} \tbPH{ʧàpórò} \tbPS{n.~} \tbGE{brain mass} \tbOP{kyaporri} \tbTP{kyaporree} \tbVA{taporo}
\tbLX{kyara} \tbHM{} \tbPH{ʧárá} \tbPS{adj.~} \tbGE{fork shape (intersection)} \tbOP{kyarre} \tbTP{kyarɛɛ/kyarrɛɛ} 
\tbLX{kyara} \tbHM{} \tbPH{ʧárà} \tbPS{n.~} \tbGE{woman (nickname)} \tbOP{kyarre} \tbTP{kyarrɛɛ} 
\tbLX{kyaraakyaraa} \tbHM{} \tbPH{ʧárááʧáráá} \tbPS{adj.~} \tbGE{forked looking} 
\tbLX{kyarakyara} \tbHM{} \tbPH{ʧàràʧàra} \tbPS{n.~} \tbGE{liquid mixed roughage} 
\tbLX{kyarewara} \tbHM{} \tbPH{ʧárɪ́wárá} \tbPS{adv.~} \tbGE{few (items), little (money)} 
\tbLX{kye} \tbHM{} \tbPH{ʧɪ̀} \tbPS{v.~} \tbSN \tbGE{to sprout} \tbSN \tbGE{to make a small cut on (tattooing)} \tbSN \tbGE{to spilt open a gourd to make calabashes} \tbSN \tbGE{to mark out the mouth of a hole to be dug (grave)} \tbPD{kyeɛɛ, kyeerɛ, kyeerɛ, kyeerebɛ, kyeeraa, kyeerebɛ } 
\tbLX{kye} \tbHM{} \tbPH{ʧɪ́} \tbPS{v.~} \tbGE{to be smart; to awaken} \tbPD{kyeɛɛ, kyeerɛ, kyeerɛ, kyeerebɛ, kyeeraa, kyeerebɛ } 
\tbLX{kye} \tbHM{} \tbPH{ʧé} \tbPS{particle.~} \tbGE{not} \tbXV{Zenɛ o kye be.} \tbXE{Today he is not there.} 
\tbLX{kyea} \tbHM{} \tbPH{ʧɪ́á} \tbPS{interjection.~} \tbGE{no (expressing disagreement)} 
\tbLX{kyebɛ} \tbHM{} \tbPH{ʧɪ́bɛ́} \tbPS{adj.~} \tbGE{fast and cunning} \tbXV{A bie taa la kyebɛ te gaŋ ona ka lɔɔre ŋmɛ.} \tbXE{The child is just too restless, he has been hit by a car.} \tbVA{kyebɛkyebɛ}
\tbLX{kye᷉e} \tbHM{} \tbPH{ʧɪ̀ɪ̀} \tbPS{v.~} \tbGE{to follow around, to herd} \tbXV{A pɔge daare kye᷉ la serɛ a tigiri zie lɛɛ ka zie te soɔ.} \tbXE{The woman followed her husband all day at the feast.} \tbPD{kye᷉eɛ, kye᷉enɛ, kyeenɛ, kyeenemɛ, kyeenaa, kyeenemɛ } 
\tbLX{kye᷉e} \tbHM{} \tbPH{ʧɪ́ɪ̀} \tbPS{v.~} \tbGE{to fry} \tbXV{N nɔŋ la zoma ba naŋ maŋ kye᷉e.} \tbXE{I like fish that is fried.} \tbPD{kye᷉eɛ, kye᷉enɛ, kyeenɛ, kyeenemɛ, kyeenaa, kyeenemɛ } 
\tbLX{kyeele} \tbHM{} \tbPH{ʧɪ́ɪ́lɪ́} \tbPS{v.~} \tbGE{to incline} \tbPD{kyeelɛɛ, kyeelɛ,, kyeelɛ,, kyeelbɛ, kyeelaa, - } 
\tbLX{kyeele} \tbHM{} \tbPH{ʧɪ̀ɪ̀lɪ̀} \tbPS{v.~} \tbGE{to scratch with a pointed object} \tbPD{kyeelɛɛ, kyeelɛ,, kyeelɛ,, kyeelbɛ, kyeelaa, - } 
\tbLX{kyeene} \tbHM{} \tbPH{ʧɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbGE{to wipe (sweat from forehead with finger); to lick out a food container (usually with finger)} \tbPD{kyeenɛɛ, kyeenɛ, kyeenɛ, kyeenemɛ, kyeenaa, - } 
\tbLX{kyeene} \tbHM{} \tbPH{ʧɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{metal ankle gingle for dancing} \tbOP{kyeemɛ} \tbTP{kyeemɛɛ/kyeenɛɛ} 
\tbLX{kyeenɛ} \tbHM{} \tbPH{ʧɪ́ɪ́nɛ́} \tbPS{v.~} \tbGE{one who shepherds} \tbPD{kye᷉e᷉ɛ᷉, kyeenɛ, kyeenɛ, kyeenemɛ, kyeenaa, - } 
\tbLX{kyeere} \tbHM{} \tbPH{ʧɪ́ɪ́rɪ́} \tbPS{v.~} \tbGE{to cry loudly (from pain)} \tbPD{kyeerɛɛ, kyeerɛ, kyeerɛ, kyeerebe, kyeeraa, kyeerebɛ } \tbVA{kyeere}
\tbLX{kyeɛ} \tbHM{} \tbPH{ʧɪ́ɛ̀} \tbPS{n.~} \tbGE{ground squirrel} \tbOP{kyeere} \tbTP{kyeerɛɛ} 
\tbLX{kyeɛ} \tbHM{} \tbPH{ʧɪ̀ɛ́} \tbPS{adj.~} \tbSN \tbGE{faded} \tbXV{A kpaaroo ŋa da waa la zeɛ kyɛ o yi la kyeɛ a tɔgele bompelaa.} \tbXE{This shirt was red but it has faded and now looks white.} \tbSN \tbGE{outmoded} \tbXV{Fo yieli la yieluŋ ŋa lɛɛ ka o te yi kyeɛ.} \tbXE{You have sung this song so much that it has lost its sweetness.} 
\tbLX{kye᷉ɛ} \tbHM{} \tbPH{ʧɪ̀᷉ɛ̀} \tbPS{v.~} \tbSN \tbGE{to tear} \tbXV{Ka fooŋ tage a kparoo o na kye᷉ɛ la.} \tbXE{If you pull the shirt, it will tear.} \tbSN \tbGE{to dash, to run quickly} \tbXV{A yan-yaa da zeɛ la kparoo kaŋa deɛ leɛ kyeɛ ne.} \tbXE{The lunatic snatched one of the shirts and dashed away with it.} \tbPD{kye᷉ɛɛ, kyeɛnɛ, kyeɛnɛ, kyeɛnemɛ, kyeɛnaa, kyeɛnemɛ } 
\tbLX{kye᷉ɛ} \tbHM{} \tbPH{ʧɪ́ɛ́} \tbPS{adj.~} \tbGE{torn (shirt)} \tbXV{Kpare kye᷉ɛ la ka a sɔresɔrɔ su.} \tbXE{The beggar wore a torn shirt.} 
\tbLX{kyeɛle} \tbHM{} \tbPH{ʧɪ̀ɛ́lɪ̀} \tbPS{v.~} \tbGE{to spread across in some order (lines)} \tbPD{kyeɛlɛɛ, kyeɛlɛ, kyeɛlɛ, kyeɛlbɛ, kyeɛlaa, - } 
\tbLX{kyeɛlɛ} \tbHM{} \tbPH{ʧɪ́ɛ́lɛ́} \tbPS{n.~} \tbGE{faulty arrangement in the game of marbles; usually a wrong move where four or more marbles of same ??? are in a continuous row} \tbTP{kyeɛlɛɛ} 
\tbLX{kyeɛraa} \tbHM{} \tbPH{ʧɪ́ɛ́ráá} \tbPS{n.~} \tbGE{lighter, something used to create a spark to produce fire} \tbOP{kyeɛre} \tbTP{kyeɛrɛɛ} 
\tbLX{kyeɛraa} \tbHM{} \tbPH{ʧɪ̀ɛ̀ràà} \tbPS{n.~} \tbGE{contribution} \tbXV{N teɛre ka ba zaaŋ yɔɔ ba kyeɛraa.} \tbXE{I think everyone has paid his contribution.} \tbOP{kyeɛre} 
\tbLX{kyeɛre} \tbHM{} \tbPH{ʧɪɛrɪ} \tbPS{v.~} \tbSN \tbGE{to contribute} \tbSN \tbGE{to harvest yams} \tbPD{-, kyeɛrɛ, -, -, -, - } 
\tbLX{kyegi} \tbHM{} \tbPH{ʧègì} \tbPS{v.~} \tbSN \tbGE{to pay, to make a ritual offering (for herbal treatment)} \tbSN \tbGE{to pay the dowry or brideprice for} \tbPD{kyegee, kyegiri, kyeire, kyegiribe, kyegiraa, - } 
\tbLX{kyegiri} \tbHM{} \tbPH{ʧégírí} \tbPS{n.~} \tbGE{payment; ritual payment (pride price)} 
\tbLX{kyekye} \tbHM{} \tbPH{ʧèʧé} \tbPS{n.~} \tbGE{sewing machine} \tbOP{kyekyeri} \tbTP{kyekyeree} 
\tbLX{kyelaa} \tbHM{} \tbPH{ʧɪ̀làà} \tbPS{adv.~} \tbGE{pointed and large} 
\tbLX{kyelaa} \tbHM{} \tbPH{ʧɪ́láá} \tbPS{adj.~} \tbGE{pointed and small} 
\tbLX{kyele} \tbHM{} \tbPH{ʧɪ́lɪ́} \tbPS{v.~} \tbGE{to awaken; to come around} \tbXV{Pampana ŋa ka o iri go᷉ɔ᷉o᷉ poɔ vɛŋ ka niŋe kyele.} \tbXE{He has just woken up from sleep; allow him to come around.} \tbPD{kyelɛɛ, kyelɛ, kyelɛ, kyelbɛ, kyelaa, - } 
\tbLX{kyelenkyooraa} \tbHM{} \tbPH{ʧɪ́lɪ́nʧʊ́ʊ̊ráà} \tbPS{n.~} \tbGE{mole} \tbOP{kyelenkyoore} \tbTP{kyelenkyoorɛɛ} \tbVA{kyenkyooraa/kyore}
\tbLX{kyeli} \tbHM{} \tbPH{ʧélì} \tbPS{v.~} \tbGE{to gaze at with mind completely drawn, dumbfounded} \tbXV{O nyɛ la a pɔge velaa na a deɛ kyeli eŋ.} \tbXE{He saw that beautiful woman and just fixed both his eyes and mind on her.} \tbPD{kyelee, kyelle, kyelle, kyelbe, kyellaa, - } 
\tbLX{kyell} \tbHM{} \tbPH{ʧɪ́ll} \tbPS{adv.~} \tbGE{protruding with a pointed end} 
\tbLX{kyelle} \tbHM{} \tbPH{ʧɪ̀llɪ̀} \tbPS{v.~} \tbGE{to rejoice, to hope} \tbPD{kyellɛɛ, kyellɛ, kyellɛ, kyellebɛ, kyellaa, kyellebɛ } 
\tbLX{kyempereŋ} \tbHM{} \tbPH{ʧɪ̀mpɪ̀rɪ̀ŋ} \tbPS{adv.~} \tbGE{utter darkness} 
\tbLX{kyempereŋ} \tbHM{} \tbPH{ʧɪ́mpɪ́rɪ́ŋ} \tbPS{adv.~} \tbGE{small but sturdy} 
\tbLX{kyenekɔnɔ} \tbHM{} \tbPH{ʧɪ̀nɪ̀kɔ̀nɔ} \tbPS{adj.~} \tbGE{tall and twisted} 
\tbLX{kyenkyamm} \tbHM{} \tbPH{ʧɪ́nʧámm} \tbPS{adv.~} \tbGE{stupified, lost as to what move to make} 
\tbLX{kyenkyeremɛ} \tbHM{} \tbPH{ʧɪ̌nʧɪ́rɪ́mɛ́} \tbPS{adj.~} \tbGE{get lost, disappear} \tbXV{Ba yɔ lɔnne la a kuuri lɛɛ ka o deɛ te kpɛ kyenkyeremɛ.} \tbXE{The hoe was passed from one person to another until it disappeared.} 
\tbLX{kyenkyoɔlaa} \tbHM{} \tbPH{ʧɪ́nʧʊ̀ɔ̀láà} \tbPS{n.~} \tbGE{morning star} \tbOP{kyenkyoɔle} \tbTP{kyenkyoɔlɛɛ} 
\tbLX{kyeŋi} \tbHM{} \tbPH{êŋ} \tbPS{v.~} \tbGE{to stand up against in argument} \tbXV{Karemamine mine ba nɔŋ ka karembibiiri kyenne ba.} \tbXE{Some teachers dislike students arguing with them.} \tbPD{kyennee, kyenne, kyenne, kyenime, kyennaa, - } 
\tbLX{kyeŋkɔleŋ} \tbHM{} \tbPH{ʧɪ́ŋkɔ̀lɪ́ŋ} \tbPS{n.~} \tbGE{a type of small bird that goes around in groups of two or three} \tbOP{kyeŋkɔlle} \tbTP{kyeŋkɔllɛɛ} 
\tbLX{kyeraa} \tbHM{} \tbPH{ʧɪ̀ràà} \tbPS{n.~} \tbSN \tbGE{a type of tree that bears pink fruits with white edible flesh} \tbSN \tbGE{the edible part of this tree's fruit} \tbOP{kyerre} \tbTP{kyerɛɛ/kyerrɛɛ} 
\tbLX{kyeraa} \tbHM{} \tbPH{ʧɪ̀ráà} \tbPS{n.~} \tbGE{chisel} \tbOP{kyerre} \tbTP{kyerɛɛ/kyerrɛɛ} 
\tbLX{kyere} \tbHM{} \tbPH{ʧɪ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to sneeze} \tbSN \tbGE{to spit forcefully} \tbSN \tbGE{to roll a small stone in the game baa} \tbPD{kyerɛɛ, kyerɛ, kyerɛ, kyerebɛ, kyeraa, kyerebɛ } 
\tbLX{kyere} \tbHM{} \tbPH{ʧɪ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to hasten} \tbSN \tbGE{of the body; shake; irrestive} \tbPD{kyerɛɛ, kyerɛ, kyerɛ, kyerebɛ, kyeraa, kyerebɛ } 
\tbLX{kyere} \tbHM{} \tbPH{ʧɪ̀rɪ̀} \tbPS{v.~} \tbGE{to lance (to cut through the skin to let out pus)} \tbPD{kyerɛɛ, kyerɛ, kyerɛ, kyerebɛ, kyeraa, kyerebɛ } 
\tbLX{kyerebaa} \tbHM{} \tbPH{ʧɪ̀rɪ̀báá} \tbPS{n.~} \tbGE{a type of brown insect usually found on the tree and fruit} \tbOP{kyerbaare} \tbTP{kyerbaarɛɛ} 
\tbLX{kyerekyere} \tbHM{} \tbPH{ʧɪ˥rɪ́ʧɪ́rɪ́} \tbPS{adj.~} \tbGE{utter silence or quiet (empty house)} 
\tbLX{kyerelaane} \tbHM{} \tbPH{ʧɪ́rɪ́láánɪ́} \tbPS{n.~} \tbGE{ripe fruit of the archie tree} \tbOP{kyerelaama} \tbTP{kyerelaamɛɛ} 
\tbLX{kyeremɛ} \tbHM{} \tbPH{ʧɪ́rɪ́mɛ́} \tbPS{n.~} \tbGE{facial incision(s), tattoo(s)} \tbOP{kyeremɛɛ} \tbVA{yaakɔɔre}
\tbLX{kyerenaabo} \tbHM{} \tbPH{ʧɪ̀rɪ̀náábʊ́} \tbPS{n.~} \tbGE{a type of red insect} \tbOP{kyerenii} \tbTP{kyereniinee} 
\tbLX{kyerɔɔ} \tbHM{} \tbPH{ʧɪ̀rɔ̀ɔ̀} \tbPS{adv.~} \tbGE{disappearing instantly} 
\tbLX{kyerri} \tbHM{} \tbPH{ʧérrì} \tbPS{v.~} \tbGE{to scamper} \tbPD{kyerree, kyerre, kyerre, kyeribe, kyerraa, - } 
\tbLX{kyesel} \tbHM{} \tbPH{ʧɪ́sɪ̀l} \tbPS{n.~} \tbGE{chisel} \tbOP{kyeselɛ} \tbTP{kyeselɛɛ} 
\tbLX{kyɛ} \tbHM{} \tbPH{ʧɛ̀} \tbPS{v.~} \tbSN \tbGE{to fell (tree)} \tbSN \tbGE{to strike (matches)} \tbSN \tbGE{to peck (of fowl)} \tbSN \tbGE{to make small cuts on the skin (herbal treatment)} \tbSN \tbGE{to start or lead a song} \tbPD{kyɛɛ, kyeɛrɛ, kyeɛrɛ, kyeɛrebɛ, kyeɛraa, kyeɛrebɛ } 
\tbLX{kyɛ} \tbHM{} \tbPH{ʧɛ́} \tbPS{conj.~} \tbSN \tbGE{but} \tbXV{Wagre gɛrɛ la kyɛ o naŋ ba wa.} \tbXE{Time is running out, but s/he has not yet come.} \tbSN \tbGE{and} \tbXV{O zoro la bangyeraa kyɛ tiire.} \tbXE{S/he has diarrhoea and is vomiting.} 
\tbLX{kyɛ} \tbHM{} \tbPH{ʧɛ̂} \tbPS{adj.~} \tbGE{here} \tbXV{Sazɔŋ ba miire Ghana kyɛ.} \tbXE{Snow does not fall here in Ghana.} 
\tbLX{kyɛbaa} \tbHM{} \tbPH{ʧɛ́báá} \tbPS{n.~} \tbGE{trap, pincers} \tbOP{kyɛrre} \tbTP{kyɛrɛɛ} \tbVA{kyabaa}
\tbLX{kyɛbɛɛ} \tbHM{} \tbPH{ʧɛ́bɛ́ɛ́} \tbPS{n.~} \tbGE{fool} 
\tbLX{kyɛbɛɛ} \tbHM{} \tbPH{ʧɛ̀bɛ̀ɛ̀} \tbPS{adv.~} \tbGE{loose and hanging} \tbXV{A wono la a baala nyɛ o naŋ e o noɔre kyɛbɛɛ lɛ.} \tbXE{The patient is suffering; see how his mouth is hanging open.} 
\tbLX{kyɛgaleŋ} \tbHM{} \tbPH{ʧɛ́gálɪ̀ŋ} \tbPS{} \tbGE{careless} \tbVA{kyagaloŋ/kyagaleŋ}
\tbLX{kyɛkuuri} \tbHM{} \tbPH{ʧɛ́kúúrí} \tbPS{v.~} \tbGE{to duck} \tbPD{kyɛkuuree, kyeɛrɛkuuro, kyeɛrekuuro, kyeɛrekuuribo, kyeɛrekuuraa, - } 
\tbLX{kyɛkyɛre} \tbHM{} \tbPH{ʧɛ̀ʧɛ̀rɪ́} \tbPS{adj.~} \tbGE{striped design (black and white)} \tbTP{kyɛkyɛrɛɛ} 
\tbLX{kyɛle} \tbHM{} \tbPH{ʧɛ́lɪ̀} \tbPS{v.~} \tbGE{to stretch out the hand earnestly to receive} \tbPD{kyɛlɛɛ, kyɛllɛ, kyɛllɛ, kyɛlbɛ, kyɛllaa, - } 
\tbLX{kyɛlee} \tbHM{} \tbPH{ʧɛ́lɪ́ɪ́} \tbPS{adj.~} \tbSN \tbGE{half of} \tbSN \tbGE{part of} \tbOP{kyɛlle} \tbVA{kyɛlɛɛ}
\tbLX{kyɛleere} \tbHM{} \tbPH{ʧɛ̀lɪ́ɪ́rɪ́l} \tbPS{n.~} \tbGE{pointed iron bar for boring holes} \tbTP{kyɛleerɛɛ} 
\tbLX{kyɛlemɛŋ} \tbHM{} \tbPH{ʧɛ́lɪ́mɛ́ŋ} \tbPS{adv.~} \tbGE{narrow (strip of cloth)} \tbXV{A naŋkpaana kyeele la pɛnne kyɛlemɛŋ lɛ leŋ ne a o gbɛre naŋ ŋmaa.} \tbXE{The hunter tore a narrow piece of cloth and used it to tie his wounded leg.} 
\tbLX{kyɛlɛɛ} \tbHM{} \tbPH{ʧɛ́lɛ̀ɛ́} \tbPS{adj.~} \tbGE{} \tbOP{kyɛlle} \tbVA{kyɛlee}
\tbLX{kyɛlɛɛkyɛlɛɛ} \tbHM{} \tbPH{ʧɛ́lɛ́ɛ́ʧɛ̀lɛ̀ɛ̀} \tbPS{adj.~} \tbGE{done half and half} 
\tbLX{kyɛlle} \tbHM{} \tbPH{ʧɛ́llɪ̀} \tbPS{v.~} \tbSN \tbGE{to listen} \tbXV{Ka a karemaŋ wa yele a yɛlɛ yɛ kyɛlle velaa.} \tbXE{When the teacher is talking, listen attentively.} \tbSN \tbGE{to notice, to be aware} \tbXV{N ba kyɛlle ka daa la zenɛ.} \tbXE{I'm not aware that today is the market day.} \tbSN \tbGE{I'm in a hurry; I can't wait for you} \tbSN \tbGE{to stay away; to be irretrievable} \tbXV{A bie gaa nyɛ deɛne a deɛ kyɛlle a be.} \tbXE{The child went and saw there was a play going on, so s/he never showed up again.} \tbPD{kyɛllɛɛ, kyɛllɛ, kyɛllɛ, kyɛlebɛ, kyɛllaa, kyɛlebɛ } 
\tbLX{kyɛloo} \tbHM{} \tbPH{ʧɛ̀lʊ́ʊ́} \tbPS{n.~} \tbGE{striped non-venomous snake} \tbOP{kyɛlle} \tbTP{kyɛlɛɛ} 
\tbLX{kyɛnaa} \tbHM{} \tbPH{ʧɛ́nàá} \tbPS{adv.~} \tbGE{this place} \tbVA{shortened form of kyɛ ŋa}
\tbLX{kyɛnɛ} \tbHM{} \tbPH{ʧɛ̀nɛ́} \tbPS{adv.~} \tbGE{still} \tbXV{O naŋ kyɛnɛ ba iri.} \tbXE{S/he is still not up.} 
\tbLX{kyɛnɛ} \tbHM{} \tbPH{ʧɛ̀nɛ̀} \tbPS{n.~} \tbGE{friend (used among females)} \tbOP{kyɛmine} \tbTP{kyɛnnɛɛ} 
\tbLX{kyɛnkyɛmpelaa} \tbHM{} \tbPH{ʧɛnʧɛ́mpɪ̀làá} \tbPS{n.~} \tbGE{egret} \tbOP{ kyɛnkyɛmpeɛle} \tbTP{kyɛnkyɛmpeɛlɛɛ} \tbVA{kyankyampelaa}
\tbLX{kyɛnkyɛmpelaanagbare} \tbHM{} \tbPH{ʧɛ́nʧɛ́mpɪ́láánàgbárɪ́} \tbPS{n.~} \tbGE{a type of plant with medicinal properties} \tbOP{kyɛnkyɛmpelaanagbarre} \tbTP{kyɛnkyɛmpelaanagbarɛɛ} 
\tbLX{kyɛnkyɛŋ} \tbHM{} \tbPH{ʧɛ̀nʧɛ̀ŋ} \tbPS{adj.~} \tbGE{often} \tbXV{Saa ba la miire kyɛnkyɛŋ.} \tbXE{It does not rain often, lately.} 
\tbLX{kyɛnkyɛrɛŋ} \tbHM{} \tbPH{ʧɛ́nʧɛ́rɛ́ŋ} \tbPS{n.~} \tbGE{confused state; crossroads} \tbXV{N boɔrɔ ka N gaa tomɔ kyɛ ka n bie yoŋ la die poɔ, N deɛ kpɛ la kyɛnkyɛrɛŋ poɔ.} \tbXE{I want to go to work but my child is alone in the house; I'm completely confused.} 
\tbLX{kyɛnkyɛweerɛ} \tbHM{} \tbPH{ʧɛ́nʧɛ́wɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{belt of pearls (cowries and beads)} \tbSG{kyɛnkyɛweeroo} \tbTP{kyɛnkyɛweerɛɛ} \tbVA{(kyɛweerɛ)}
\tbLX{kyɛnne} \tbHM{} \tbPH{ʧɛ́nnɪ́} \tbPS{n.~} \tbGE{inheritance} \tbTP{kyɛnnɛɛ} 
\tbLX{kyɛnne} \tbHM{} \tbPH{ʧɛ́nnɪ̀} \tbPS{n.~} \tbSN \tbGE{the act of walking} \tbSN \tbGE{journey} \tbXV{Fo kyɛnne be soŋ?} \tbXE{How was your journey?.} \tbTP{kyɛnnɛɛ} 
\tbLX{kyɛnne} \tbHM{} \tbPH{ʧɛ̀nnɪ̀} \tbPS{n.~} \tbGE{friendship between females} \tbTP{kyɛnnɛɛ} 
\tbLX{kyɛnse} \tbHM{} \tbPH{ʧɛ́nsɪ̀} \tbPS{n.~} \tbGE{iron sheet} \tbOP{kyɛnsere} \tbTP{kyɛnserɛɛ} \tbVA{kyanse}
\tbLX{kyɛŋ} \tbHM{} \tbPH{ʧɛ̂ŋ} \tbPS{v.~} \tbGE{to walk} \tbPD{kyɛŋɛɛ, kyɛnɛ, kyɛnɛ, kyɛnemɛ, kyɛnaa, - } 
\tbLX{kyɛparepeɛ} \tbHM{} \tbPH{ʧɛ́párɪ́pɪ́ɛ́} \tbPS{n.~} \tbGE{a type of brown, spotted nocturnal bird} \tbOP{kyɛparepeere} \tbTP{kyɛparepeerɛɛ} 
\tbLX{kyɛre} \tbHM{} \tbPH{ʧɛ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to pay a dowry or bride price} \tbSN \tbGE{to untie, to set free} \tbSN \tbGE{to come across (road)} \tbPD{kyɛrɛɛ, kyɛrɛ, kyɛrɛ, kyɛrebɛ, kyɛraa, kyerebɛ } 
\tbLX{kyɛre} \tbHM{} \tbPH{ʧɛ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to left over, to remain} \tbSN \tbGE{to come after (in birth order)} \tbXV{Maa la kyɛre N yɔɔpɔge.} \tbXE{My sister comes after me.} \tbPD{kyɛrɛɛ, kyɛrɛ, kyɛrɛ, kyɛrebɛ, kyɛraa, - } 
\tbLX{kyɛre} \tbHM{} \tbPH{ʧɛ̀rɪ́} \tbPS{n.~} \tbGE{small body of water on a farm} \tbOP{kyɛrre} \tbTP{kyɛrɛɛ/kyɛrrɛɛ} 
\tbLX{kyɛrebo} \tbHM{} \tbPH{ʧɛ́rɪ́bʊ́} \tbPS{n.~} \tbGE{bride price} 
\tbLX{kyɛremɛ} \tbHM{} \tbPH{ʧɛ̀rɪ̀mɛ́} \tbPS{n.~} \tbGE{stick for stirring soup} \tbTP{kyɛremɛɛ} 
\tbLX{kyɛrɛ} \tbHM{} \tbPH{ʧɛ̀rɛ̀} \tbPS{n.~} \tbGE{a type of guinea corn} \tbTP{kyɛrɛɛ} 
\tbLX{kyɛrre} \tbHM{} \tbPH{ʧɛ̀rrɪ̀} \tbPS{v.~} \tbGE{to scoop liquid to the last bit (water)} \tbPD{kyɛrrɛɛ, kyɛrrɛ, kyɛrrɛ, kyɛrebɛ, kyɛrraa, - } 
\tbLX{kyi} \tbHM{} \tbPH{ʧî} \tbPS{n.~} \tbGE{millet} \tbSG{kyiruu (bag of) kaboo (for plant)} \tbTP{kyiree} 
\tbLX{kyi} \tbHM{} \tbPH{ʧí} \tbPS{v.~} \tbSN \tbGE{to direct toward, to pinpoint} \tbXV{Maa la ka o kyi kyɛ yele a yɛlɛ.} \tbXE{S/he directed the talk toward me.} \tbSN \tbGE{to hate} \tbXV{O deɛ kyi ma, o yeli ka o koŋ wuli ma a pɔge yiri zie.} \tbXE{She just hates me; she refuses to show me her house.} \tbPD{kyiee, kyire, kyire, kyiribe, kyiraa, kyiribe } 
\tbLX{kyibe} \tbHM{} \tbPH{ʧìbé} \tbPS{n.~} \tbGE{lice} \tbSG{kyibuu} \tbTP{kyibee} 
\tbLX{kyibuu} \tbHM{} \tbPH{ʧìbúú} \tbPS{n.~} \tbGE{flea} \tbOP{kyibe} \tbTP{kyibee} 
\tbLX{kyieree} \tbHM{} \tbPH{ʧìèréè} \tbPS{v.~} \tbGE{a large black bird with red tail feathers} \tbPD{-, -, -, -, -, - } 
\tbLX{kyieri} \tbHM{} \tbPH{ʧíèri} \tbPS{v.~} \tbSN \tbGE{to argue with} \tbSN \tbGE{to shout with joy} \tbPD{kyieree, kyiere, kyiere, kyieribe, kyieraa, kyieribe } 
\tbLX{kyieruŋ} \tbHM{} \tbPH{ʧíérúŋ} \tbPS{n.~} \tbGE{the ability to remember or recall information} 
\tbLX{kyigi} \tbHM{} \tbPH{ʧìgì} \tbPS{v.~} \tbGE{to cover entirely with a cuplike object} \tbPD{kyigee, kyigre, kyigre, kyigribe, kyigraa, - } 
\tbLX{kyiiri} \tbHM{} \tbPH{ʧíírì} \tbPS{v.~} \tbSN \tbGE{to forbid} \tbSN \tbGE{to deny} \tbPD{kyiiree, kyiire, kyiire, kyiiribe, kyiiraa, kyiiribe } 
\tbLX{kyiiruŋ} \tbHM{} \tbPH{ʧíírúŋ} \tbPS{n.~} \tbGE{taboo, prohibition} \tbOP{kyiire} \tbTP{kyiiree} 
\tbLX{kyileŋko} \tbHM{} \tbPH{ʧíléŋkó} \tbPS{adj.~} \tbGE{expert} \tbOP{kyileŋkori} \tbTP{kyileŋkoree} 
\tbLX{kyilli} \tbHM{} \tbPH{ʧíllì} \tbPS{v.~} \tbGE{to alter, to change} \tbPD{kyillee, kyille, kyille, kyilibe, kyillaa, kyilibe } 
\tbLX{kyinni} \tbHM{} \tbPH{ʧìnnì} \tbPS{v.~} \tbGE{to land suddenly and stongly (from a leap)} \tbPD{kyinnee, kyinne, kyinne, kyinime, kyinnaa, kyinnaa } 
\tbLX{kyiri} \tbHM{} \tbPH{ʧírì} \tbPS{v.~} \tbGE{to pour} \tbSN \tbGE{to leak} \tbPD{kyiree, kyire, kyire, kyiribe, kyiraa, kyirre } \tbSE{kyiri zee} \tbGE{to bleed} 
\tbLX{kyo} \tbHM{} \tbPH{ʧó} \tbPS{v.~} \tbGE{to aim at} \tbPD{kyoe, kyuoro, kyuoro, kyuoribo, kyuoraa, - } 
\tbLX{kyo} \tbHM{} \tbPH{ʧʊ̀} \tbPS{v.~} \tbSN \tbGE{to suck} \tbSN \tbGE{shooting contest (gambling with arrows)} \tbPD{kyoɛ, kyoora, kyoora, kyooreba, kyooraa, - } 
\tbLX{kyo᷉} \tbHM{} \tbPH{ʧó᷉} \tbPS{adv.~} \tbGE{silent, not uttering a word} 
\tbLX{kyogi} \tbHM{} \tbPH{ʧógí} \tbPS{v.~} \tbSN \tbGE{to continue, to resume} \tbSN \tbGE{to be in line with} \tbPD{kyogee, kyogro, kyogro, kyogribo, kyogaa, kyogribo } 
\tbLX{kyogi} \tbHM{} \tbPH{ʧògì} \tbPS{v.~} \tbGE{to pay a bride price} \tbPD{kyogee, kyogro, kyogro, kyogribo, kyograa, - } 
\tbLX{kyogili} \tbHM{} \tbPH{ʧóglì} \tbPS{v.~} \tbSN \tbGE{to stack delicately} \tbSN \tbGE{pick ears} \tbPD{kyogilee, kyogilo, kyoglo, kyogilibo, kyogilaa, - } 
\tbLX{kyogilokolo} \tbHM{} \tbPH{ʧóglókóló} \tbPS{adj.~} \tbGE{tall and high up into the sky} \tbVA{kyogilokolokolo}
\tbLX{kyogilokolo} \tbHM{} \tbPH{ʧóglókóló} \tbPS{n.~} \tbSN \tbGE{tower} \tbSN \tbGE{storey building} \tbVA{kyogilokolokolo}
\tbLX{kyogiri} \tbHM{} \tbPH{ʧógrì} \tbPS{v.~} \tbGE{to trot (of a horse)} \tbPD{kyogiree, kyogiro, kyogiro, kyogiribo, kyogiraa, - } 
\tbLX{kyolenkyooraa} \tbHM{} \tbPH{ʧʊ́lɪ́nʧʊ́ʊ́ráá} \tbPS{n.~} \tbGE{mole; small black mouse with a strong smell} \tbOP{kyolenkyoore} \tbTP{kyolenkyoorɛɛ} \tbVA{kyore}
\tbLX{kyoli} \tbHM{} \tbPH{ʧólí} \tbPS{v.~} \tbSN \tbGE{to pierce through} \tbSN \tbGE{to pass by} \tbSN \tbGE{to forget} \tbSN \tbGE{to overdo; to do wrong} \tbPD{kyolee, kyolo, kyolo, kyolbo, kyolaa, - } 
\tbLX{kyololo} \tbHM{} \tbPH{ʧólóló} \tbPS{adv.~} \tbGE{of light beam; gentle lighting} 
\tbLX{kyone} \tbHM{} \tbPH{ʧʊ́nɪ́} \tbPS{adj.~} \tbGE{bulging} 
\tbLX{kyonɛɛkyonɛɛ} \tbHM{} \tbPH{ʧʊ́nɛ́ɛ́ʧʊ́nɛ́ɛ́} \tbPS{adj.~} \tbGE{lighting in sparks} 
\tbLX{kyonkyolo} \tbHM{} \tbPH{ʧònʧóló} \tbPS{n.~} \tbGE{loose skirt} \tbTP{kyonkyolee} 
\tbLX{kyonkyoŋ} \tbHM{} \tbPH{ʧonʧoŋ} \tbPS{v.~} \tbGE{to jump around and over a spot} \tbPD{kyonkyoŋee, kyonkyono, kyonkyono, kyonkyonimo, kyonkyonaa, - } \tbVA{kyoŋkyoŋ}
\tbLX{kyonni} \tbHM{} \tbPH{ʧónnì} \tbPS{v.~} \tbGE{to pound fried shea nuts into a thick paste to make shea butter} \tbPD{kyonnee, kyonno, kyonno, kyonimo, kyonnaa, - } 
\tbLX{kyonni} \tbHM{} \tbPH{ʧónní} \tbPS{v.~} \tbSN \tbGE{to drip} \tbSN \tbGE{to trim} \tbPD{kyonnee, kyonno, kyonno, kyonimo, kyonnaa, kyonimo } 
\tbLX{kyoŋi} \tbHM{} \tbPH{ʧôŋ} \tbPS{v.~} \tbSN \tbGE{to take over (piece of work)} \tbSN \tbGE{to catch in midair} \tbPD{kyoŋee, kyonno, kyonno, kyonimo, kyonnaa, - } 
\tbLX{kyoo} \tbHM{} \tbPH{ʧóò} \tbPS{n.~} \tbGE{nest (in tree)} \tbOP{kyoori} \tbTP{kyooree} \tbVA{kyogi, kyogri, kyogree}
\tbLX{kyo᷉o} \tbHM{} \tbPH{ʧʊ̀᷉ʊ̀᷉} \tbPS{adv.~} \tbGE{of the splash of water, when one jumps into it} 
\tbLX{kyoone} \tbHM{} \tbPH{ʧʊ́ʊ́nɪ́} \tbPS{n.~} \tbGE{shea nut} \tbOP{kyoomɔ} \tbTP{kyoonɛɛ} 
\tbLX{kyooraa} \tbHM{} \tbPH{ʧʊ́ʊ́ráá} \tbPS{v.~} \tbGE{hernia} \tbPD{-, -, -, -, -, - } 
\tbLX{kyoore} \tbHM{} \tbPH{ʧʊ́ʊ́rɪ́} \tbPS{v.~} \tbSN \tbGE{a clicking sound used to call a dog} \tbSN \tbGE{expression of strong dislike} \tbPD{kyoorɛɛ, kyoora, kyoora, kyooreba, kyooraa, - } 
\tbLX{kyoore} \tbHM{} \tbPH{ʧʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to suck to the last bit} \tbPD{kyoorɛɛ, kyoora, kyoora, kyooreba, kyooraa, - } 
\tbLX{kyo᷉ɔ} \tbHM{} \tbPH{ʧʊ́ɔ̀} \tbPS{v.~} \tbGE{to leap} \tbPD{kyo᷉ɔɛ, kyoɔnɔ, kyoɔnɔ, kyoɔnema, kyoɔnaa, - } 
\tbLX{kyo᷉ɔ} \tbHM{} \tbPH{ʧʊ̀ɔ̀} \tbPS{v.~} \tbGE{to spring, to jump} \tbPD{kyo᷉ɔɛ, kyoɔnɔ, kyoɔnɔ, kyoɔnemɔ, kyoɔnaa, - } 
\tbLX{kyo᷉ɔ} \tbHM{} \tbPH{ʧʊ̀ɔ̀} \tbPS{v.~} \tbGE{to leap, to sprint} \tbPD{kyo᷉ɔɛ, kyoɔnɔ, kyoɔnɔ, kyoɔnema, kyoɔnaa, - } 
\tbLX{kyoɔne} \tbHM{} \tbPH{ʧʊ̀ɔ̀nɪ̀} \tbPS{adj.~} \tbGE{stripped} 
\tbLX{kyoɔnɔkyoɔnɔ} \tbHM{} \tbPH{ʧʊ̀ɔ̀nɔ̀ʧʊ́ɔ́nɔ̀} \tbPS{adj.~} \tbGE{stripped in many colours} 
\tbLX{kyoɔre} \tbHM{} \tbPH{ʧʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to filter, to strain} \tbPD{kyoɔrɛɛ, kyoɔrɔ, -, -, kyoɔraa, kyoɔre } 
\tbLX{kyoɔre} \tbHM{} \tbPH{ʧʊ̀ɔ̀rɪ̀} \tbPS{v-adj.~} \tbGE{stand loosely, helplessly (in a confused state)} \tbXV{o kyoɔre are la a teɛ pare kyɛ kaara a zɔɔre.} \tbXE{S/he stood under the tree, perplexed, watching the fight.} 
\tbLX{kyore} \tbHM{} \tbPH{ʧʊ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to be worn} \tbSN \tbGE{to be totally destroyed} \tbPD{kyorɛɛ, kyorɔ, kyorɔ, kyoreba, kyoraa, - } 
\tbLX{kyore} \tbHM{} \tbPH{ʧʊ́rɪ́} \tbPS{n.~} \tbGE{} \tbOP{kyorre} \tbTP{kyorɛɛ} \tbVA{shortened form of kyolenkyooraa}
\tbLX{kyori} \tbHM{} \tbPH{ʧórí} \tbPS{v.~} \tbGE{to cut or chop into pieces} \tbPD{kyoree, kyoro, kyoro, kyoribo, kyoraa, - } 
\tbLX{kyoriforo} \tbHM{} \tbPH{ʧórífóró} \tbPS{adv.~} \tbGE{squating position (open legs)} 
\tbLX{kyoroso} \tbHM{} \tbPH{ʧʊ́rʊ́sʊ́} \tbPS{adj.~} \tbSN \tbGE{of worth} \tbSN \tbGE{appearance like a twisted mouth} 
\tbLX{kyɔ} \tbHM{} \tbPH{ʧɔ́} \tbPS{v.~} \tbSN \tbGE{to hide behind an object} \tbSN \tbGE{to start a tune} \tbSN \tbGE{to bear dawdaw fruits} \tbSN \tbGE{to tune up to play xylophone text} \tbPD{kyɔɛ, kyoɔrɔ, kyoɔrɔ, kyoɔreba, kyoɔraa, - } 
\tbLX{kyɔ} \tbHM{} \tbPH{ʧɔ́} \tbPS{adj.~} \tbGE{at first; in the beginning} \tbXV{o naŋ kyɔ wa o ba soore te.} \tbXE{When s/he first came, s/he did not ask us.} 
\tbLX{kyɔ-de} \tbHM{} \tbPH{ʧɔ́-dɪ́} \tbPS{v.~} \tbGE{to take by surprise; to attack suddenly} \tbPD{kyɔ-deɛ, kyoɔrɔ-derɛ, kyɔ-derɛ, kyɔ-derebɛ, kyɔ-deraa, - } 
\tbLX{kyɔge} \tbHM{} \tbPH{ʧɔ́gɪ́} \tbPS{v.~} \tbGE{to sprain} \tbPD{kyɔgɛɛ, kyɔgrɔ, kyɔgrɔ, kyɔgreba, kyɔgraa, kyɔgreba } 
\tbLX{kyɔge} \tbHM{} \tbPH{ʧɔ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to chook, to stab} \tbSN \tbGE{to pierce, to inject} \tbPD{kyɔgɛɛ, kyɔgrɔ, kyɔgrɔ, kyɔgreba, kyɔgraa, kyɔgreba } 
\tbLX{kyɔge a mɛrɛɛ poɔ} \tbHM{} \tbPH{ʧɔ̀gɪ̀ à mɛ̀rɛ́ɛ pʊ́ɔ́} \tbPS{v.~} \tbGE{straight to the point; hitting the nail on the head} \tbPD{-, -, -, -, -, - } 
\tbLX{kyɔgelɔkɔlɔ} \tbHM{} \tbPH{ʧɔ́gɪ́lɔ́kɔ́lɔ́} \tbPS{adj.~} \tbGE{slim, (appendages)} 
\tbLX{kyɔkyɔɔ} \tbHM{} \tbPH{ʧɔʧɔɔ} \tbPS{adv.~} \tbGE{appearing very long (dresσ} 
\tbLX{kyɔlaa} \tbHM{} \tbPH{ʧɔ̀làà} \tbPS{adj.~} \tbGE{long and pointed} 
\tbLX{kyɔle} \tbHM{} \tbPH{ʧɔ́lɪ́} \tbPS{v.~} \tbGE{to roam, to go around} \tbPD{kyɔlɛɛ, kyɔlɔ, kyɔlɔ, kyɔlbɔ, kyɔlaa, - } 
\tbLX{kyɔlemɔ} \tbHM{} \tbPH{ʧɔ́lɪ́mɔ́} \tbPS{n.~} \tbGE{spear; lance} \tbOP{kyɔlle} \tbTP{kyɔllɛɛ} \tbVA{kpane}
\tbLX{kyɔlkyɔl} \tbHM{} \tbPH{ʧɔ̀lʧɔ̀l} \tbPS{adv.~} \tbGE{long (snake)} 
\tbLX{kyɔmpoore} \tbHM{} \tbPH{ʧɔ̀mpʊ̀ʊ̀rɪ̀} \tbPS{n.~} \tbGE{a challenge between two people in which one who wakes up first sprays the counterpart with water} \tbTP{kyɔpoorɛɛ} 
\tbLX{kyɔmpuuri} \tbHM{} \tbPH{ʧɔ̀mpúúrí} \tbPS{n.~} \tbGE{betting in which the earlier to wake up sprays water on his counterpart} \tbTP{kyɔmpuuree} 
\tbLX{kyɔnfeɛle} \tbHM{} \tbPH{ʧɔ̀nfɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{very early (morning)} \tbOP{kyɔnfeɛlɛ} \tbTP{kyɔnfeɛlɛɛ} 
\tbLX{kyɔnkyɔlemo} \tbHM{} \tbPH{ʧɔ̀nʧɔ́lɪ́mɔ́} \tbPS{n.~} \tbGE{hanging bits of an object} 
\tbLX{kyɔnkyɔlɔŋ} \tbHM{} \tbPH{ʧɔ̀nʧɔ̀lɔ̀ŋ} \tbPS{adv.~} \tbGE{long looking (rope)} 
\tbLX{kyɔŋ} \tbHM{} \tbPH{ʧɔ̀ŋ} \tbPS{adv.~} \tbGE{first thing done} \tbXV{Ko᷉ɔ nyegrɛ la see ka te kyɔŋ iri gaa a baa.} \tbXE{Water is scarce so we have to get up early and go to the stream.} 
\tbLX{kyɔŋe} \tbHM{} \tbPH{ʧɔ́ŋ} \tbPS{v.~} \tbGE{to file along in line; to follow one another} \tbPD{kyɔŋɛɛ, kyɔnɔ, kyɔnɔ, kyɔnemɔ, kyɔnaa, - } \tbVA{tɔŋ}
\tbLX{kyɔŋe} \tbHM{} \tbPH{ʧɔ̂ŋ} \tbPS{v.~} \tbGE{fetch or light from a fire} \tbPD{kyɔŋɛɛ, kyɔnnɔ, kyɔnnɔ, kyɔnema, kyɔnnaa, - } \tbVA{tɔŋ}
\tbLX{kyɔɔ} \tbHM{} \tbPH{ʧɔ̀ɔ̀} \tbPS{v.~} \tbGE{to hold gingerly, hanging from the tips of fingers, teeth, claws, or beak} \tbPD{kyɔɔɛ, kyɔɔrɔ, kyɔɔrɔ, kyɔɔrebɔ, kyɔɔraa, - } 
\tbLX{kyɔɔ} \tbHM{} \tbPH{ʧɔ̀ɔ̀} \tbPS{adj.~} \tbGE{long (dress)} 
\tbLX{kyɔɔɔ} \tbHM{} \tbPH{ʧɔ̀ɔ̀ɔ̀} \tbPS{adv.~} \tbGE{long and draping (of a hanging object)} 
\tbLX{kyɔɔtaa} \tbHM{} \tbPH{ʧɔ̀ɔ̀tàá} \tbPS{n.~} \tbGE{offer, gift} \tbOP{kyɔɔtaare} \tbTP{kyɔɔtaarɛɛ} 
\tbLX{kyɔre} \tbHM{} \tbPH{ʧɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to be saturated with water} \tbSN \tbGE{to pound in order to shell (nuts)} \tbSN \tbGE{to loosen soil} \tbSN \tbGE{to do things anyhow (speaking)} \tbPD{kyɔrɛɛ, kyɔrɔ, kyɔrɔ, kyɔreba, kyɔraa, kyɔreba } \tbSE{saa kyore} \tbGE{period of intense rain; downpour} 
\tbLX{kyɔre} \tbHM{1} \tbPH{ʧɔ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to slice; to chop into bits} \tbSN \tbGE{to sprinkle water on} \tbPD{kyɔrɛɛ, kyɔrɔ, kyɔrɔ, kyɔreba, kyɔraa, kyɔreba } 
\tbLX{kyɔre} \tbHM{2} \tbPH{ʧɔ́rɪ́} \tbPS{v.~} \tbGE{to do things anyhow (speaking)} \tbPD{-, -, -, -, -, - } 
\tbLX{kyugiloŋ} \tbHM{} \tbPH{ʧùglòŋ} \tbPS{adj.~} \tbGE{heavy and hung out (head)} 
\tbLX{kyugiloŋ} \tbHM{} \tbPH{ʧúglóŋ} \tbPS{adv.~} \tbGE{small and hung down (head)} 
\tbLX{kyuli} \tbHM{} \tbPH{ʧùlì} \tbPS{v.~} \tbGE{to bend one's head down} \tbPD{kyulee, kyullo, kyullo, kyulibo, kyullaa, - } 
\tbLX{kyulokyulo} \tbHM{} \tbPH{ʧúlóʧúló} \tbPS{adj.~} \tbGE{confined within a small area (giving assistance)} 
\tbLX{kyulom} \tbHM{} \tbPH{ʧúlóm} \tbPS{adv.~} \tbGE{disappearing at once into water or a hole} \tbVA{kyuloŋ}
\tbLX{kyuloŋ} \tbHM{} \tbPH{ʧúlóŋ} \tbPS{adv.~} \tbGE{} \tbVA{kyulom}
\tbLX{kyumm} \tbHM{} \tbPH{ʧúmm} \tbPS{adj.~} \tbGE{going straight down} 
\tbLX{kyunkyuŋ} \tbHM{} \tbPH{ʧúnʧúŋ} \tbPS{adj.~} \tbGE{very deep} 
\tbLX{kyuoli} \tbHM{} \tbPH{ʧùòlì} \tbPS{v.~} \tbGE{to dazzle the eyes (of light)} \tbPD{kyuolee, kyuolo, kyuolo, kyuolbo, kyuolaa, - } 
\tbLX{kyuori} \tbHM{} \tbPH{ʧúórì} \tbPS{v.~} \tbGE{empty (stomach)} \tbPD{kyuoree, kyuoro, kyuoro, kyuoribo, kyuoraa, - } 
\tbLX{kyuri} \tbHM{} \tbPH{ʧúrí} \tbPS{v.~} \tbSN \tbGE{to fall through} \tbSN \tbGE{to abort (pregnancy)} \tbPD{kyuree, kyuro, kyuro, kyuribo, kyuraa, - } 
\tbLX{kyuu} \tbHM{} \tbPH{ʧúù} \tbPS{n.~} \tbSN \tbGE{moon} \tbSN \tbGE{month} \tbSG{kyuuruu} \tbOP{kyuuri} \tbTP{kyuuree} 
\tbLX{kyuu} \tbHM{} \tbPH{ʧùú} \tbPS{n.~} \tbGE{poultry lice} \tbOP{kyuuri} \tbTP{kyuuree} 
\tbLX{kyuu} \tbHM{} \tbPH{ʧúú} \tbPS{adv.~} \tbGE{sleepy} 
\tbLX{kyuukyuu} \tbHM{} \tbPH{ʧúùʧúù} \tbPS{adv.~} \tbGE{monthly} 
\tbLX{kyuupaalaa} \tbHM{} \tbPH{ʧúúpáálàà} \tbPS{n.~} \tbSN \tbGE{new month} \tbSN \tbGE{crescent moon, new moon} 
\tbLX{kyuupaaleŋ} \tbHM{} \tbPH{ʧúúpààlɪ̀ŋ} \tbPS{n.~} \tbGE{full moon} 
\tbLX{kyuupelaa} \tbHM{} \tbPH{ʧúúpɪ̀làá} \tbPS{n.~} \tbGE{moonlight} \tbOP{kyuupeɛle} \tbTP{kyuupeɛlɛɛ} 
\tbLX{kyuuri} \tbHM{} \tbPH{ʧùùrì} \tbPS{v.~} \tbGE{to pour out from a height onto (of a waterfall)} \tbPD{kyuuree, kyuuro, kyuuro, kyuuribo, kyuuraa, - } 
\tbLX{la} \tbHM{} \tbPH{lá} \tbPS{particle.~} \tbSN \tbGE{emphatic particle be; is, are} \tbXV{Ona la na gaa.} \tbXE{S/he is the one who will go.} \tbSN \tbGE{helping verb} \tbXV{O na gaa la.} \tbXE{S/he will go.} 
\tbLX{la} \tbHM{} \tbPH{là} \tbPS{v.~} \tbGE{to laugh} \tbPD{laɛ, laara, laara, laareba, laaraa, laareba } 
\tbLX{la} \tbHM{} \tbPH{là} \tbPS{adv.~} \tbGE{again} \tbXV{O la gaaɛ la.} \tbXE{S/he has gone again.} 
\tbLX{la} \tbHM{} \tbPH{lá} \tbPS{v.~} \tbGE{to open or break an object made of lobes. (kola nut, fruit)} \tbPD{laɛ, laara, laara, lareba, laaraa, - } 
\tbLX{laa} \tbHM{} \tbPH{láá} \tbPS{n.~} \tbGE{dish, bowl} \tbOP{laare} \tbTP{laarɛɛ} 
\tbLX{laadimee} \tbHM{} \tbPH{làadíméé́} \tbPS{n.~} \tbGE{great suffering} \tbTP{laadinnee} 
\tbLX{laafeɛ} \tbHM{} \tbPH{làafɪ́ɛ́̀} \tbPS{n.~} \tbGE{health} \tbOP{laafeere} \tbTP{laafeerɛɛ} 
\tbLX{laalee} \tbHM{} \tbPH{lààléè} \tbPS{n.~} \tbGE{small eartheware dish for serving soup} \tbOP{laabilii} \tbTP{laabilee} 
\tbLX{laalere} \tbHM{} \tbPH{lààlɪ́rɪ̀} \tbPS{n.~} \tbGE{flute made from stalk} \tbOP{laaleɛ} \tbTP{laalerɛɛ/laalerrɛɛ} 
\tbLX{laalɛŋ} \tbHM{} \tbPH{lààlɛ́ŋ} \tbPS{n.~} \tbGE{medium-size bowl} 
\tbLX{laane} \tbHM{} \tbPH{láánɪ́} \tbPS{adj.~} \tbGE{useless} \tbOP{laama} \tbTP{laamɛɛ} 
\tbLX{laane} \tbHM{} \tbPH{láánɪ́} \tbPS{adj.~} \tbGE{ripe archie-fruit (showing a split)} \tbOP{laama} \tbTP{laamɛɛ} 
\tbLX{laane} \tbHM{} \tbPH{làànɪ̀} \tbPS{v.~} \tbGE{to grope, to move one's hand over to feel} \tbPD{laanɛɛ, laana, laana, laaneba, laanaa, - } 
\tbLX{laare} \tbHM{} \tbPH{láárɪ̀} \tbPS{v.~} \tbSN \tbGE{to wave e.g. greeting} \tbSN \tbGE{to move or spread out e.g. people in a circle} \tbSN \tbGE{to brush aside} \tbPD{laarɛɛ, laara, laara, laareba, laaraa, laareba } 
\tbLX{laare} \tbHM{} \tbPH{láárɪ̀} \tbPS{v.~} \tbGE{to hover above e.g. bird} \tbPD{laarɛɛ, laara, laara, laareba, laaraa, laareba } 
\tbLX{laare} \tbHM{} \tbPH{lààrɪ̀} \tbPS{v.~} \tbSN \tbGE{to spread all over} \tbXV{Saa mie la ka kõɔ laare a bombulo zaa.} \tbXE{The rain has caused a near flooding of all the crops.} \tbSN \tbGE{to move a finger gently over} \tbXV{A dɔɔta de la tẽẽ kãã mine laare ne a bie nimie.} \tbXE{The doctor applied some creamy medicine to the child’s eyes.} \tbPD{laarɛɛ, laara, laara, laareba, laaraa, - } 
\tbLX{laare} \tbHM{} \tbPH{lààrɪ̀} \tbPS{v.~} \tbGE{to assemble; to bring together} \tbPD{laarɛɛ, laara, laara, laareba, laaraa, - } 
\tbLX{laatereke} \tbHM{} \tbPH{láátɪ́rɪ́kɪ̀} \tbPS{n.~} \tbGE{electricity} \tbOP{laaterekere} \tbTP{laaterekerɛɛ} 
\tbLX{labiri} \tbHM{} \tbPH{lábírì} \tbPS{n.~} \tbGE{axe blade} \tbOP{labie} \tbTP{labiree} 
\tbLX{ladiraa} \tbHM{} \tbPH{ládìráà} \tbPS{n.~} \tbGE{sharp axe} \tbOP{ladirri} \tbTP{ladiree} 
\tbLX{laga} \tbHM{} \tbPH{lágà} \tbPS{n.~} \tbGE{spoon} \tbOP{lagre} \tbTP{lagrɛɛ} 
\tbLX{laga} \tbHM{} \tbPH{làgá} \tbPS{n.~} \tbGE{groin} \tbOP{lagre} \tbTP{lagɛɛ} 
\tbLX{lagara} \tbHM{} \tbPH{lágàrà} \tbPS{n.~} \tbGE{a type of creeping grass} \tbSG{lagaroo} \tbOP{lagarre} \tbTP{lagarɛɛ} 
\tbLX{lage} \tbHM{} \tbPH{lágɪ̀} \tbPS{v.~} \tbGE{to strip off hard material e.g. bark} \tbPD{lagɛɛ, lagra, lagra, lagreba, lagraa, lagreba } \tbVA{geli}
\tbLX{lage} \tbHM{} \tbPH{lágɪ̀} \tbPS{v.~} \tbGE{to remove malted grain e.g guinea corn from the place it was left to sprout} \tbPD{lagɛɛ, lagra, lagra, lagreba, lagraa, lagreba } \tbVA{geli}
\tbLX{lage} \tbHM{} \tbPH{làgɪ̀} \tbPS{v.~} \tbGE{to surround} \tbPD{lagɛɛ, lagra, lagra, lagreba, lagraa, lagreba } 
\tbLX{lagere} \tbHM{} \tbPH{lágrɪ́} \tbPS{v.~} \tbGE{to meet, to move across} \tbXV{A dataare na lagere la taa a naayiri poɔ'} \tbXE{The opposing parties will meet at the chief's palace.} \tbPD{lagerɛɛ, lagera, lagera, lagereba, lagraa, - } 
\tbLX{laguu} \tbHM{} \tbPH{lágùú} \tbPS{n.~} \tbGE{blunt axe} \tbOP{laguuri} \tbTP{laguuree} 
\tbLX{lakpoŋi} \tbHM{} \tbPH{lákpôŋ} \tbPS{n.~} \tbGE{large axe} \tbOP{lakponni} \tbTP{lakponnee} 
\tbLX{lalee} \tbHM{} \tbPH{lál̊èé} \tbPS{n.~} \tbGE{small axe} \tbOP{labilii} \tbTP{labilee} 
\tbLX{laloɔrɔ} \tbHM{} \tbPH{lálʊ́ɔ́rɔ́} \tbPS{n.~} \tbGE{playmate} \tbOP{laloɔreba} \tbTP{laloɔrɛɛ} 
\tbLX{lambiri} \tbHM{} \tbPH{lǎmbírì} \tbPS{n.~} \tbGE{testicle} \tbSG{lambiruu} \tbOP{lambie} \tbTP{lambiiree} 
\tbLX{lambori} \tbHM{} \tbPH{làmbórí} \tbPS{n.~} \tbSN \tbGE{side} \tbSN \tbPS{postposition.~} \tbGE{side (of something)} \tbOP{lamboe/lambobo} \tbTP{lamboree} \tbVA{lambobo (1pl)}
\tbLX{lamboriŋ} \tbHM{} \tbPH{làmbórîŋ} \tbPS{loc.~} \tbGE{beside} 
\tbLX{lambosentaa-naare} \tbHM{} \tbPH{làmbósɪ́ntáá-náárɪ̀} \tbPS{n.~} \tbGE{square} 
\tbLX{lammo} \tbHM{} \tbPH{làmmʊ́} \tbPS{n.~} \tbGE{gathering} 
\tbLX{lampare} \tbHM{} \tbPH{làmpàrɪ̀} \tbPS{n.~} \tbGE{groin} \tbOP{lampare} \tbTP{lamparɛɛ} 
\tbLX{lampɛ} \tbHM{} \tbPH{lâmpɛ́} \tbPS{n.~} \tbGE{cradle cover} \tbOP{lampɛre} \tbTP{lampɛrɛɛ} 
\tbLX{lampɛgere} \tbHM{} \tbPH{lámpɛ́gɪ́rɪ́} \tbPS{n.~} \tbGE{cheek} \tbOP{lampɛgɛ} \tbTP{lampɛgɛɛ/lampɛgerɛɛ} 
\tbLX{lamponderɛ} \tbHM{} \tbPH{lǎmpóndɪ́rɛ́} \tbPS{n.~} \tbGE{tax collector} \tbOP{lamponderebɛ} \tbTP{lamponderebɛɛ} 
\tbLX{lampoŋ} \tbHM{} \tbPH{lǎmpóŋ} \tbPS{n.~} \tbGE{levy, tax} \tbOP{lamponni} \tbTP{lamponnee} 
\tbLX{lane} \tbHM{} \tbPH{lánɪ́} \tbPS{n.~} \tbGE{scrotum} \tbOP{lama} \tbTP{lannɛɛ} \tbVA{lambiri, lambie}
\tbLX{lanleebu} \tbHM{} \tbPH{lànléébú} \tbPS{n.~} \tbGE{unison} 
\tbLX{lanleebu} \tbHM{} \tbPH{lànléébú} \tbPS{n.~} \tbGE{rhythm} 
\tbLX{lanna} \tbHM{} \tbPH{lànná} \tbPS{n.~} \tbGE{tree squirrel} \tbOP{lanne} \tbTP{lannɛɛ} \tbVA{lɔnnɔ}
\tbLX{lanne} \tbHM{} \tbPH{lánnɪ̀} \tbPS{v.~} \tbGE{to lean against} \tbPD{lannɛɛ, lanna, lanna, lanema, lannaa, lanema } 
\tbLX{laŋa} \tbHM{} \tbPH{láŋá} \tbPS{n.~} \tbGE{a tribe (Sisaale)} \tbOP{laŋema} \tbTP{laŋemɛɛ} 
\tbLX{Laŋe} \tbHM{} \tbPH{làŋɪ́} \tbPS{n.~} \tbGE{area inhabited by the Sisaala} 
\tbLX{laŋe} \tbHM{} \tbPH{làŋɪ̀} \tbPS{v.~} \tbSN \tbGE{to gather, to get together} \tbSN \tbGE{to add} \tbSN \tbGE{to be jointly owned} \tbPD{laŋɛɛ, lanna, -, -, -, - } 
\tbLX{laŋe} \tbHM{} \tbPH{láŋɪ́} \tbPS{v.~} \tbGE{to slosh} \tbPD{laŋɛɛ, laŋena/lanna, lanna, laŋenema/lanema, laŋenaa, - } 
\tbLX{laŋgbaraa} \tbHM{} \tbPH{láŋgbáráà} \tbPS{n.~} \tbGE{a long pole with a hook for plucking things high up} \tbOP{laŋgbarre} \tbTP{laŋgbarɛɛ} \tbVA{laŋgbalaa, gbalaa}
\tbLX{laŋkoroo} \tbHM{} \tbPH{láŋkóróò} \tbPS{n.~} \tbGE{corner} \tbOP{laŋkorri} \tbTP{laŋkorree} 
\tbLX{laŋkpaa} \tbHM{} \tbPH{lâŋkpáà} \tbPS{n.~} \tbGE{groin gland} \tbOP{laŋkpaare} \tbTP{laŋkpaarɛɛ} \tbVA{loŋkoroo}
\tbLX{laŋkpeɛrɛ} \tbHM{} \tbPH{làŋkpɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{neighbour} \tbOP{laŋkpeɛrebɛ} \tbTP{laŋkpeɛrebɛɛ} 
\tbLX{larata} \tbHM{} \tbPH{lárátá} \tbPS{adj.~} \tbGE{white, very clean e.g. cloudless full moon} 
\tbLX{lare} \tbHM{} \tbPH{lárì} \tbPS{n.~} \tbGE{axe} \tbOP{laɛ} \tbTP{larɛɛ} 
\tbLX{lare} \tbHM{} \tbPH{lárɪ́} \tbPS{v.~} \tbGE{to become very hot} \tbPD{larɛɛ, lara, lara, lareba, laraa, - } 
\tbLX{lare} \tbHM{} \tbPH{làrì} \tbPS{v.~} \tbGE{to lock} \tbPD{larɛɛ, lara, lara, lareba, laraa, - } 
\tbLX{lare} \tbHM{} \tbPH{làrì} \tbPS{v.~} \tbGE{to stack one container in another} \tbPD{larɛɛ, lara, lara, lareba, laraa, - } 
\tbLX{le} \tbHM{} \tbPH{lè} \tbPS{v.~} \tbSN \tbGE{to fall (down)} \tbSN \tbGE{to be guilty} \tbSN \tbGE{to settle e.g. sediment in liquid} \tbSN \tbGE{to reduce in pain e.g. stomach ache} \tbPD{lee, lere, lere, leribe, leraa, leribe } 
\tbLX{le} \tbHM{} \tbPH{lé} \tbPS{v.~} \tbGE{to be bent on} \tbXV{A porisi le ka o gaa ne a doreba kɔɔto.} \tbXE{The police officer is bent on prosecuting the driver in court.} \tbPD{-, -, -, -, -, - } 
\tbLX{leere} \tbHM{} \tbPH{lɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to manouvre, to weave through} \tbXV{O de la a bɔle leere o taaba zaa kyɛ eŋ gool.} \tbXE{He dribbled the ball through the other players and scored a goal.} \tbSN \tbGE{to go through and select from many objects leaving some behind} \tbXV{O dɔ la a moɔ kyɛ leere a bombulo bare.} \tbXE{He removed the weeds, leaving the crops.} \tbSN \tbGE{to exclude} \tbXV{Ba poŋi la a bibiiri diibu kyɛ te leere a entuo soba bare.} \tbXE{They shared the food among the children, excluding the lazy one.} \tbSN \tbGE{to get to the bottom of (hole), knooks (cheeks)} \tbXV{O naŋ ɔɔ a seŋka᷉a baare la ka o leere o yagekɔbɔ.} \tbXE{It is after she had finished eating the groudnuts that she brushed her cheeks using her tongue.} \tbPD{leerɛɛ, leerɛ, leerebɛ, leerebɛ, leeraa, - } 
\tbLX{leɛ} \tbHM{} \tbPH{lɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to return, to go back} \tbSN \tbGE{to transfer to a different container} \tbSN \tbGE{to change to something different e.g. behaviour} \tbPD{leɛɛ, leɛrɛ, leɛrɛ, leɛrebɛ, leɛraa, leɛrebɛ } \tbSE{leɛ gaa} \tbGE{} \tbSE{leɛ puori} \tbGE{} 
\tbLX{leɛ} \tbHM{} \tbPH{lɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to prepare self in readiness e.g. going on a journey} \tbSN \tbGE{to prepare an offering for a guest to take away} \tbPD{leɛɛ, leɛrɛ, leɛrɛ, leɛrebɛ, leɛraa, leɛrebɛ } \tbSE{leɛ gaa} \tbGE{} \tbSE{leɛ puori} \tbGE{} 
\tbLX{leɛ} \tbHM{} \tbPH{lɪ́ɛ̀} \tbPS{v.~} \tbGE{to replace, to take the place of} \tbPD{leɛɛ, leɛrɛ, leɛrɛ, leɛrebɛ, leɛraa, - } 
\tbLX{leɛ-eebo} \tbHM{} \tbPH{lìɛ̀-ɪ́ɪ́bʊ́} \tbPS{v.~} \tbGE{to redo; to go over} \tbPD{-, -, -, -, -, - } 
\tbLX{leɛgɔɔtɔbaa} \tbHM{} \tbPH{lɪ́ɛ́gɔ̀ɔ́tɔ̀báá} \tbPS{n.~} \tbGE{a local dish made from fried corn and vegetable} \tbTP{leɛgɔɔtɔbaarɛɛ} 
\tbLX{leɛne} \tbHM{} \tbPH{lɪ́ɛ́nɪ̀} \tbPS{n.~} \tbGE{a type of tree with sweet and sour fruits and medicinal properties} \tbOP{leɛmɛ} \tbTP{leɛmɛɛ} 
\tbLX{leɛne} \tbHM{} \tbPH{lɪ́ɛ́nɪ̀} \tbPS{n.~} \tbGE{large umbrella} \tbOP{leɛmɛ} \tbTP{leɛmɛɛ} 
\tbLX{leɛre} \tbHM{} \tbPH{lɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to change (appointment, time, place)} \tbXV{Ba leɛre la a kpaaroŋ biŋ bieo.} \tbXE{They have rescheduled the meeting for tomorrow.} \tbSN \tbGE{to return, to give back} \tbXV{o leɛre la a o gane ko o.} \tbXE{She has returned his book to him.} \tbSN \tbGE{to repudiate, to turn round} \tbXV{A nannyige boɔrɔ ka o leɛre a yɛlɛ tuo a o tutɔ.} \tbXE{The thief is trying to shift the theft onto his friend.} \tbPD{leɛ rɛɛ, leɛrɛ, leɛre terɛ, leɛre terebɛ, leɛ teraa, - } 
\tbLX{leɛre} \tbHM{} \tbPH{lɪ́ɛ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to exchange} \tbXV{O de la bɛŋɛ leɛre ne kyi.} \tbXE{S/he substituted beans for millet.} \tbSN \tbGE{change position} \tbXV{A dɔɔ leɛre la a boore kpɛ ne maaroŋ} \tbXE{The man moved the goats from the sun into the shade.} \tbSN \tbGE{substitute, take the place of} \tbXV{A bɔlŋmeɛrɛ kaŋa perɛɛ la ka bonyuo kpɛ leɛre o zu.} \tbXE{As one of the football players was injured they brought in another player to substitute him.} \tbPD{-, -, -, -, -, - } 
\tbLX{leɛre} \tbHM{} \tbPH{lɪ́ɛ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to wave one's hands or other objects in the air or in the face of another} \tbXV{A zɔɔzɔɔreba da leɛrɛ la nuuri ennɛ taa.} \tbXE{The fighters were throwing their hands in each other's face.} \tbXV{A lɔɔre naŋ da wa ire n ma da leɛre la o nu ko ma ka n poɔ zaa sãã.} \tbXE{As the lorry took off my mother waved me but I was very sad.} \tbSN \tbGE{wave hand vigorously in the face of another, usually in anger} \tbXV{Dɔbɔ mine bayi da toora la taa kyɛ leɛrɛ nuuri ennɛ taa.} \tbXE{Two men were waving their hands angrily in each other's face.} \tbPD{-, -, -, -, -, - } 
\tbLX{lege} \tbHM{} \tbPH{lɪ̀gɪ̀} \tbPS{v.~} \tbGE{to apply to a spot} \tbXV{A te᷉e-erɛ de la te᷉e lege a baala tobo ne o gbɛɛ.} \tbXE{The herbalist applied some ointment to the patient's ears and legs.} \tbPD{legɛɛ, legerɛ, legerɛ, legerebɛ, legeraa, - } 
\tbLX{lege} \tbHM{} \tbPH{lɪ̀gɪ̀} \tbPS{v.~} \tbGE{to fortify oneself} \tbXV{Ba lege la a naŋkpaana ŋa mɔyɔ te᷉e.} \tbXE{This hunter has been fortified against the attack of dangerous animals.} \tbPD{legɛɛ, legerɛ, legerɛ, legerebɛ, legeraa, - } 
\tbLX{lege} \tbHM{} \tbPH{lɪ́gɪ́} \tbPS{v.~} \tbSN \tbGE{to wag (dog's tail)} \tbXV{A baa da zo tuori la o daana a legerɛ o zoore.} \tbXE{The dog ran to meet its master and wagged its tail.} \tbSN \tbGE{to struggle} \tbPD{legɛɛ, legerɛ, legerɛ, legerɛ, legeraa, - } 
\tbLX{legelege} \tbHM{} \tbPH{lɪ̀gɪ̀lɪ́gɪ́} \tbPS{v.~} \tbGE{to struggle under a tight grip} \tbXV{A dɔɔ da feere la a nannyige kɔkɔre ka o te legelegerɛ.} \tbXE{The man held the thief's throat and he struggled to breathe.} \tbPD{legelegɛɛ, legelegɛ, legelegɛ, legelegerebɛ, legelegeraa, - } 
\tbLX{legere} \tbHM{} \tbPH{lɪ́gɪ́rɪ̀} \tbPS{v.~} \tbGE{to tickle} \tbPD{legrɛɛ, legrɛ, legrɛ, legrebɛ, legraa, legrebɛ } 
\tbLX{legere} \tbHM{} \tbPH{lɪ́gɪ́rɪ̀} \tbPS{v.~} \tbGE{to stalk} \tbPD{legrɛɛ, legrɛ, legrɛ, legrebɛ, legraa, legrebɛ } 
\tbLX{legere} \tbHM{} \tbPH{lɪ́gɪ́rɪ̀} \tbPS{v.~} \tbGE{to eat a great deal} \tbPD{legrɛɛ, legrɛ, legrɛ, legrebɛ, legraa, legrebɛ } 
\tbLX{legere} \tbHM{} \tbPH{lìgɪ̀rɪ̀} \tbPS{v.~} \tbGE{to move noiselessly to avoid being noticed} \tbPD{legerɛɛ, legerɛ, legerɛ, legerebɛ, legeraa, - } 
\tbLX{legiri} \tbHM{} \tbPH{légirí} \tbPS{v.~} \tbGE{to fill in what is missing e.g. sowing of grains} \tbXV{Kɔrre da pere la kyi yaga, ka a koɔreba la te legiri.} \tbXE{Partridges have scatched most of the millet that was sown and the farmers had to resow.} \tbPD{legiree, legire, legire, legiribe, legiraa, - } 
\tbLX{legiri} \tbHM{} \tbPH{légirí} \tbPS{v.~} \tbGE{to argue} \tbXV{Dagaaba ba nɔŋ bibile naŋ legire neŋkponni.} \tbXE{The Dagaaba do not take kindly to children who argue with elderly people.} \tbPD{legiree, legire, legire, legiribe, legiraa, - } 
\tbLX{lekere} \tbHM{} \tbPH{lɪ́kɪ́rɪ́} \tbPS{n.~} \tbGE{boastfulness; assuming (typically used with yɛlɛ)} \tbXV{A bikpoŋ na erɛ la lekere yɛlɛ yelmeŋɛ.} \tbXE{That big boy behaves in a boastful manner.} 
\tbLX{le ko᷉ɔ} \tbHM{} \tbPH{lè kʊ̀᷉ɔ́} \tbPS{adj.~} \tbGE{small} 
\tbLX{le ko᷉ɔ} \tbHM{} \tbPH{lè kʊ̀᷉ɔ́} \tbPS{adj.~} \tbGE{watery} 
\tbLX{lelaare} \tbHM{} \tbPH{lèlààrɪ̀} \tbPS{v.~} \tbGE{to take complete control;} \tbPD{lelaarɛɛ, lerelaara, lelaara, lelaareba, lelaaraa, - } 
\tbLX{le lɔɔ} \tbHM{} \tbPH{lè lɔ́g} \tbPS{adj.~} \tbGE{guilty of the law} 
\tbLX{le nimiri} \tbHM{} \tbPH{lè nímírì} \tbPS{adj.~} \tbGE{pleasant} 
\tbLX{lenlenni} \tbHM{} \tbPH{lènlènní} \tbPS{n.~} \tbGE{a type of skin disease} \tbTP{lenlennee} 
\tbLX{lenni} \tbHM{} \tbPH{lénní} \tbPS{v.~} \tbGE{to be spotted} \tbPD{lennee, lenne, lenne, lenime, lennaa, - } 
\tbLX{lenni} \tbHM{} \tbPH{lénní} \tbPS{v.~} \tbGE{to be interpersed} \tbPD{lennee, lenne, lenne, lenime, lennaa, - } 
\tbLX{lenni} \tbHM{} \tbPH{lénní} \tbPS{v.~} \tbSN \tbGE{to augment the main drummer's beat (of support xylophonist)} \tbSN \tbGE{play a musical instrument in support of the lead player} \tbPD{lennee, lenne, lenne, lenime, lennaa, - } 
\tbLX{leŋi} \tbHM{} \tbPH{lèŋ} \tbPS{v.~} \tbSN \tbGE{to fasten, to bind e.g. load} \tbSN \tbGE{to tether} \tbPD{leŋee, lene, lene, lenime, lenaa, - } 
\tbLX{leŋi miri} \tbHM{} \tbPH{léŋí mírì} \tbPS{v.~} \tbGE{to commit suicide} \tbPD{leŋee miri, lene miri, mi-lene, -, -, - } 
\tbLX{leŋi noɔre} \tbHM{} \tbPH{léŋ nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to fast; to refrain from eating} \tbPD{leŋee noɔre, lene noɔre, nɔ-lene, -, -, - } 
\tbLX{leŋi woɔ} \tbHM{} \tbPH{léŋ wʊ́ɔ́} \tbPS{v.~} \tbGE{to pack bags} \tbPD{leŋee woɔ, lene woɔ, wo-lene, -, -, - } 
\tbLX{leraa} \tbHM{} \tbPH{lɪ́ráá} \tbPS{n.~} \tbGE{safety pin} \tbOP{lerre} \tbTP{lerɛɛ/lerrɛɛ} 
\tbLX{lere} \tbHM{} \tbPH{lérè} \tbPS{n.~} \tbGE{a type of loincloth for men} \tbOP{lerri} \tbTP{leree/lerree} 
\tbLX{lere} \tbHM{} \tbPH{lɪ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to interweave} \tbSN \tbGE{crisscross e.g sitting with legs crossed} \tbPD{lerɛɛ, lerɛ, lerɛ, lerebɛ, leraa, - } 
\tbLX{leree} \tbHM{} \tbPH{lèrèè} \tbPS{adj.~} \tbGE{naked, bare bodied} 
\tbLX{leremɛ} \tbHM{} \tbPH{lɪ́rɪ́mɛ́} \tbPS{adj.~} \tbGE{interwoven} 
\tbLX{lerese} \tbHM{} \tbPH{lɪ́rɪ́sɪ́} \tbPS{adv.~} \tbGE{small but intact (person)} 
\tbLX{le tobo} \tbHM{} \tbPH{lè tòbó} \tbPS{adj.~} \tbGE{beautiful} 
\tbLX{lezaɛ anaare} \tbHM{} \tbPH{lɪ̀záɛ̀ ánáárɪ̀} \tbPS{adv.~} \tbGE{eighty} 
\tbLX{lezaɛ ata} \tbHM{} \tbPH{lɪ̀záɛ̀ átà} \tbPS{adv.~} \tbGE{sixty} 
\tbLX{lezaɛ ata ne pie} \tbHM{} \tbPH{lɪ̀záɛ̀ átà nɪ́ píé} \tbPS{adv.~} \tbGE{seventy} 
\tbLX{lezaɛ ayi} \tbHM{} \tbPH{lɪ̀záɛ̀ áyì} \tbPS{adv.~} \tbGE{forty} 
\tbLX{lezaɛ ayi ne pie} \tbHM{} \tbPH{lɪ̀záɛ̀ áyì nɪ́ píé} \tbPS{adv.~} \tbGE{fifty} 
\tbLX{lezaɛ ayi ŋmaa anuu} \tbHM{} \tbPH{lɪ̀záɛ́ áyì ŋmàà ànúú} \tbPS{adv.~} \tbGE{thirty-five} 
\tbLX{lezare} \tbHM{} \tbPH{lɪ̀zárɪ̀} \tbPS{adv.~} \tbGE{twenty} 
\tbLX{lezare ne ayi} \tbHM{} \tbPH{lɪ̀zárɪ̀ nɪ́ píé nɪ́ yéní} \tbPS{adv.~} \tbGE{thirty-one} 
\tbLX{lezare ne yeni} \tbHM{} \tbPH{lɪ̀zárɪ̀ nɪ́ yénì} \tbPS{adv.~} \tbGE{twenty-one} 
\tbLX{lɛ} \tbHM{} \tbPH{lɛ̌} \tbPS{adv.~} \tbSN \tbGE{so, usually} \tbXV{Lɛ ka a waa.} \tbXE{So it is.} \tbSN \tbGE{as it is, at that point} \tbXV{Meŋ bare a zoma lɛ.} \tbXE{Also, stop there with the insults.} \tbSN \tbGE{what} \tbXV{N na te woŋi la lɛ o naŋ na yeli.} \tbXE{I will go and listen to what s/he will say.} 
\tbLX{lɛbiri} \tbHM{} \tbPH{lɛ̀bírì} \tbPS{n.~} \tbGE{bead} \tbSG{lɛbiruu} \tbOP{lɛbie} \tbTP{lɛbiiree} 
\tbLX{lɛɛ} \tbHM{} \tbPH{lɛ̀ɛ̂} \tbPS{use word.~} \tbGE{is that so?} \tbXV{a. A dɔɔ ŋmaa na gaa la dansareka kyɛ yi. b. Lɛɛ?} \tbXE{a. That short man is an ex-convict. b. Is that so?.} 
\tbLX{lɛɛ} \tbHM{} \tbPH{lɛ́ɛ́} \tbPS{adv.~} \tbSN \tbGE{for a while (wait)} \tbXV{o are la be lɛɛ wa te bale.} \tbXE{S/he waited until s/he became tired.} \tbSN \tbGE{continuous and for quite sometime} \tbXV{Ba yeli ka a sori e la ŋmaa kyɛ ka te kyɛŋe o lɛɛ ka ŋmenaa te kpɛ.} \tbXE{We were told the journey was not long but walked and walked until dusk.} \tbVA{lɛɛŋ}
\tbLX{lɛgerɛŋ} \tbHM{} \tbPH{lɛ́gɪ̀rɛ́ŋ} \tbPS{adv.~} \tbGE{long, loose, and hanging} 
\tbLX{lɛmɛlɛmɛ} \tbHM{} \tbPH{lɛ́mɛ́lɛ́mɛ́} \tbPS{adv.~} \tbGE{slightly sweet} 
\tbLX{lɛmpɛrɛŋ} \tbHM{} \tbPH{lɛ́mpɛ́rɛ́ŋ} \tbPS{adv.~} \tbGE{short and even at the top} 
\tbLX{lɛnɛɛ} \tbHM{} \tbPH{lɛ̀nɛ́ɛ̂} \tbPS{conj.~} \tbSN \tbGE{what} \tbXV{N ba baŋe lɛnɛɛ N naŋ na e pampana.} \tbXE{I don’t know what to do next.} \tbSN \tbGE{how} \tbXV{Fo baŋe la lɛnɛɛ ba naŋ maŋ sɛ kparoo?} \tbXE{Do you know how a shirt is sewn?.} 
\tbLX{lɛnlɛŋko} \tbHM{} \tbPH{lɛ́nlɛ́ŋkò} \tbPS{use word.~} \tbGE{expression to soothe a child; usual tune in a cradle song} 
\tbLX{lɛnne} \tbHM{} \tbPH{lɛ̀nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to lick, to taste} \tbSN \tbGE{to offer a sample of food for sale} \tbPD{lɛnnɛɛ, lɛnnɛ, lɛnnɛ, lɛnemɛ, lɛnnaa, lɛnemɛ } 
\tbLX{lɛntɛre} \tbHM{} \tbPH{lɛ́ntɛ́rɪ́} \tbPS{n.~} \tbGE{thinking too much of oneself} \tbTP{lɛntɛrɛɛ} 
\tbLX{lɛŋŋɛlɛŋŋɛ} \tbHM{} \tbPH{lɛ́ŋŋɛ́lɛ́ŋŋɛ́} \tbPS{n.~} \tbGE{sickle} \tbOP{lɛŋŋɛlɛnne} \tbVA{sɔgɔraa}
\tbLX{lɛre} \tbHM{} \tbPH{lɛ̀rɪ́} \tbPS{n.~} \tbGE{bead} \tbOP{lɛɛ} \tbTP{lɛrɛɛ} 
\tbLX{lɛre} \tbHM{} \tbPH{lɛ́rɪ́} \tbPS{v.~} \tbGE{to behave badly to get attention e.g. of a child} \tbPD{lɛrɛɛ, lɛrɛ, lɛrɛ, lɛrebɛ, lɛraa, - } \tbVA{lɛroo}
\tbLX{lɛre} \tbHM{} \tbPH{lɛ́rɪ̀} \tbPS{v.~} \tbGE{to replace, to put in the place of} \tbXV{A bɔlŋmeɛrɛ kaŋa perɛɛ la ka boŋyuo kpɛ lɛre o zu.} \tbXE{One of the football players got injured and another replaced him.} \tbPD{lɛrɛɛ, lɛrɛ, lɛrɛ, lɛrebɛ, lɛraa, - } 
\tbLX{libibinzie} \tbHM{} \tbPH{lìbíbìnzíé} \tbPS{n.~} \tbGE{bank} \tbOP{libibinziiri} \tbTP{libibinziiree} 
\tbLX{libibɔ} \tbHM{} \tbPH{líbíbɔ́} \tbPS{n.~} \tbGE{quest for money} 
\tbLX{libibugo} \tbHM{} \tbPH{libibùgó} \tbPS{n.~} \tbGE{safe for money} \tbOP{libibugiri} \tbTP{libibugiree} 
\tbLX{libidaana} \tbHM{} \tbPH{líbídááná} \tbPS{n.~} \tbGE{rich person} \tbOP{libidaanema} \tbTP{libidaanemɛɛ} 
\tbLX{libidi} \tbHM{} \tbPH{líbídì} \tbPS{n.~} \tbGE{the act of spending money} 
\tbLX{libima} \tbHM{} \tbPH{líbímá} \tbPS{n.~} \tbGE{capital} 
\tbLX{libipelaa} \tbHM{} \tbPH{lìbípɪ̀làá} \tbPS{n.~} \tbGE{cowry} \tbOP{libipeɛle} \tbTP{libipeɛlɛɛ} 
\tbLX{libiri} \tbHM{} \tbPH{lìbírì} \tbPS{n.~} \tbGE{money} \tbSG{libiruu} \tbOP{libie} \tbTP{libiiree} 
\tbLX{libizɔŋe} \tbHM{} \tbPH{líbízɔ́ŋ} \tbPS{n.~} \tbSN \tbGE{gold dust} \tbSN \tbGE{plentiful money} \tbTP{libizɔnnɛɛ} 
\tbLX{libu} \tbHM{} \tbPH{líbú} \tbPS{n.~} \tbGE{a funeral rite for an adult woman} \tbTP{libiree} \tbVA{ligi}
\tbLX{liemu} \tbHM{} \tbPH{lìèmú} \tbPS{n.~} \tbGE{orange (fruit)} \tbOP{liemuri} \tbTP{liemuree} 
\tbLX{lige} \tbHM{} \tbPH{lìgè} \tbPS{adj.~} \tbGE{dark} \tbTP{ligiree} 
\tbLX{ligi} \tbHM{} \tbPH{lígí} \tbPS{v.~} \tbGE{to get dark} \tbPD{ligee, ligre, ligre, ligraa, -, - } 
\tbLX{ligi} \tbHM{} \tbPH{lígí} \tbPS{v.~} \tbGE{to shroud} \tbPD{ligee, ligre, ligre, ligraa, -, - } 
\tbLX{ligi} \tbHM{} \tbPH{lìgì} \tbPS{v.~} \tbGE{to withhold information, to leave someone in the dark} \tbPD{ligee, ligire, ligire, ligiraa, -, - } 
\tbLX{ligi} \tbHM{} \tbPH{lìgì} \tbPS{v.~} \tbGE{to perform a burial rite for a woman} \tbPD{ligee, ligire, ligire, ligiraa, -, - } 
\tbLX{ligi} \tbHM{} \tbPH{lígí} \tbPS{n.~} \tbGE{a burial rite for women} \tbVA{libu}
\tbLX{ligiki} \tbHM{} \tbPH{lìgìkì} \tbPS{adv.~} \tbGE{deep and dark} 
\tbLX{ligiligi} \tbHM{} \tbPH{ligiligi} \tbPS{adj.~} \tbGE{silent} 
\tbLX{ligiruŋ} \tbHM{} \tbPH{lígirúŋ} \tbPS{n.~} \tbGE{darkness} \tbOP{ligire} \tbTP{ligiree} 
\tbLX{liiri} \tbHM{} \tbPH{líírí} \tbPS{v.~} \tbGE{to patch, to mend} \tbPD{liiree, liire, liire, liiraa, -, - } 
\tbLX{liiri} \tbHM{} \tbPH{lììrì} \tbPS{v.~} \tbGE{to get out of sight} \tbXV{Pampana ŋa ka ba deɛ kyɛŋ liiri la a taŋa puori} \tbXE{They just went behind the hill and out of sight a short while ago.} \tbPD{liiree, liire, liire, liiribe, liiraa, - } 
\tbLX{limbo} \tbHM{} \tbPH{límbò} \tbPS{n.~} \tbSN \tbGE{a transit place to heaven} \tbSN \tbGE{a state of confusion} \tbOP{limbori} \tbTP{limboree} 
\tbLX{limelime} \tbHM{} \tbPH{lìmèlìmè} \tbPS{adv.~} \tbGE{between light and darkness} \tbXV{Zimaa limelime ka noore maŋ gaŋe.} \tbXE{Fowls sleep when darkness is setting in.} 
\tbLX{limm} \tbHM{} \tbPH{lìmm} \tbPS{adv.~} \tbGE{dark, not clear} \tbXV{Ka zieŋ wa soɔrɔ vaare poɔ maŋ waa la limm lɛ.} \tbXE{When it is dusk, it is dark in the thick vegetation.} 
\tbLX{limpiriŋ} \tbHM{} \tbPH{lìmpìrìŋ} \tbPS{adj.~} \tbGE{oval shape and dark, not discernable} 
\tbLX{linni} \tbHM{} \tbPH{lìnnì} \tbPS{v.~} \tbGE{to sink, to immerse e.g. as in water} \tbPD{linnee, linne, linne, linime, linnaa, - } 
\tbLX{linni} \tbHM{} \tbPH{lìnnì} \tbPS{v.~} \tbGE{to glean} \tbPD{linnee, linne, linne, linime, linnaa, - } 
\tbLX{linni} \tbHM{} \tbPH{línnì} \tbPS{v.~} \tbGE{to cover up someone e.g. from being embarrassed} \tbXV{A bie na la zu a libie kyɛ ka o ma linni o.} \tbXE{That boy stole the money but his mother covered him up.} \tbPD{linnee, linne, linne, linime, linnaa, - } 
\tbLX{liŋe} \tbHM{} \tbPH{líŋé} \tbPS{n.~} \tbGE{lid, cover} \tbOP{linni} \tbTP{linnee} 
\tbLX{liŋgbiraa} \tbHM{} \tbPH{lìŋgbíráá} \tbPS{n.~} \tbGE{a type of tree with medicinal properties and edible fruits} \tbOP{liŋgbirri} \tbTP{liŋgbiree} 
\tbLX{liroo} \tbHM{} \tbPH{líróó} \tbPS{adv.~} \tbGE{small and protruding} 
\tbLX{Lobaa} \tbHM{} \tbPH{lòbàà} \tbPS{n.~} \tbGE{a person from the Lobi ethnic group} \tbOP{luori} \tbTP{lobree} 
\tbLX{loge} \tbHM{} \tbPH{lʊ̀gɪ̀} \tbPS{v.~} \tbGE{to penetrate soft ground} \tbPD{logɛɛ, logra, logra, logreba, lograa, - } 
\tbLX{logere} \tbHM{} \tbPH{lʊ́gɪ́rɪ̀} \tbPS{v.~} \tbGE{to cause to move up and down e.g. muscle under the skin} \tbPD{logerɛɛ, logera, logera, logereba, logeraa, - } 
\tbLX{logerɔŋ} \tbHM{} \tbPH{lʊ̀gɪ̀rɔ̀ŋ} \tbPS{adv.~} \tbGE{round and fat} \tbXV{Nyɛ bufuli naŋ nyu ze᷉e pa᷉a waa logerɔŋ lɛ.} \tbXE{See how round and fat the ground maggot looks after sucking blood.} 
\tbLX{logɛɛ} \tbHM{} \tbPH{lʊ̊gɛɛ} \tbPS{adj.~} \tbGE{powdery} 
\tbLX{logi} \tbHM{} \tbPH{lògí} \tbPS{n.~} \tbSN \tbGE{poison} \tbSN \tbGE{venom e.g. of snake} \tbOP{logiri} \tbTP{logiree} 
\tbLX{logi} \tbHM{} \tbPH{lògí} \tbPS{n.~} \tbGE{quiver; sheath for arrows} \tbOP{logiri} \tbTP{logiree} 
\tbLX{logi} \tbHM{} \tbPH{lógì} \tbPS{n.~} \tbGE{rest stop, encampment} \tbXV{Moɔre da maŋ ŋmɛ logi Kaleo poɔ kyɛ gaa kɔmpare.} \tbXE{The Mossi used to camp at Kaleo on their way to southern Ghana.} \tbOP{logiri} \tbTP{logiree} 
\tbLX{logi} \tbHM{} \tbPH{lógì} \tbPS{n.~} \tbGE{farm for crops that produce nuts} \tbOP{logiri} \tbTP{logiree} 
\tbLX{logi} \tbHM{} \tbPH{lògì} \tbPS{v.~} \tbGE{to make a small hole} \tbXV{A bie logi la bogi a dankyini seɛ de a libiri u᷉u᷉.} \tbXE{The child  made a hole at the base of the wall and buried the money there.} \tbPD{logee, logro, logro, logribo, lograa, - } \tbVA{loguu}
\tbLX{logi} \tbHM{} \tbPH{lógí} \tbPS{adj.~} \tbGE{deep} \tbVA{logoko}
\tbLX{logiri} \tbHM{} \tbPH{lògirì} \tbPS{v.~} \tbGE{to remove forcibly, to take out from a hidden or tight position e.g. crevice, hole} \tbXV{A kɔrre logiri la a bɛŋɛ ba naŋ bore.} \tbXE{The partridges have scratched to remove the beans that were sown.} \tbPD{logiree, logiro, logiro, logribo, logiraa, - } 
\tbLX{logiri noɔre} \tbHM{} \tbPH{lógírí nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to pester, to bother (for information)} \tbXV{O logiri la o serɛ noɔre ka o na yeli a zɔɔre pare ko o.} \tbXE{She pestered her husband to tell her the cause of the fight.} \tbPD{logiree, logiro, logiro, logribo, logiraa, - } 
\tbLX{logitogo} \tbHM{} \tbPH{lògìtògò} \tbPS{adv.~} \tbGE{deep and open (large)} \tbXV{Ba yage la a baa lɛ logitogo.} \tbXE{They have dug the pond very big.} 
\tbLX{logitogo} \tbHM{} \tbPH{lógítógó} \tbPS{adv.~} \tbGE{shallow and small} \tbXV{A kparoo zomɔ deɛ waa la lɛ logitogo.} \tbXE{The pocket of the shirt is very shallow.} 
\tbLX{logoko} \tbHM{} \tbPH{lògòkò} \tbPS{adv.~} \tbGE{deep set and large} \tbXV{A dɔɔ nimie logoko.} \tbXE{The man's eyes are deep set.} 
\tbLX{logoko} \tbHM{} \tbPH{lógókó} \tbPS{adv.~} \tbGE{deep set and small} \tbXV{A bie nimie logoko.} \tbXE{The child's eyes are deep set.} 
\tbLX{lohoo} \tbHM{} \tbPH{lòhòó} \tbPS{n.~} \tbGE{a type of tree with edible fruits} \tbOP{lohoori} \tbTP{lohooree} 
\tbLX{lo᷉ilo᷉i} \tbHM{} \tbPH{ló᷉íló᷉í} \tbPS{adv.~} \tbGE{bright red} 
\tbLX{lomm} \tbHM{} \tbPH{lʊ̀mm} \tbPS{adv.~} \tbGE{round and big} 
\tbLX{lomm} \tbHM{} \tbPH{lòmm} \tbPS{adv.~} \tbGE{balloon-like} 
\tbLX{lomolomo} \tbHM{} \tbPH{lómólómó} \tbPS{adv.~} \tbGE{jellylike; soft} \tbXV{Duoronwoɔ maŋ waa la lomolomo lɛ.} \tbXE{The urinary bladder is soft and jellylike.} 
\tbLX{lomɔlomɔ} \tbHM{} \tbPH{lʊ̀mɔ̀lʊ̀mɔ̀} \tbPS{adv.~} \tbGE{soft e.g. flesh, fruit} 
\tbLX{lonlommu} \tbHM{} \tbPH{lǒnlǒmú} \tbPS{n.~} \tbGE{a type of fish that is not edible} \tbOP{lonlonni} \tbTP{lonlonnee} 
\tbLX{lonni} \tbHM{} \tbPH{lònnì} \tbPS{v.~} \tbGE{to hurry (Southern Dagaare dialect)} \tbPD{lonnee, lonno, lonno, lonimo, lonaa, - } 
\tbLX{loŋgoroŋ} \tbHM{} \tbPH{lʊ̀ŋgʊ̀rʊ̀ŋ} \tbPS{adj.~} \tbGE{cloudy} 
\tbLX{loŋkoroo} \tbHM{} \tbPH{lóŋkóróò} \tbPS{n.~} \tbGE{hidden corner} \tbXV{A waabo zo kpɛ la a die loŋkoroo kaŋa poɔ.} \tbXE{The sake ran into one of the dark corners of the room.} \tbOP{loŋkorri} \tbTP{loŋkoree} 
\tbLX{loŋŋo} \tbHM{} \tbPH{lǒŋŋò} \tbPS{n.~} \tbGE{a type of X-shaped small drum. Usually held under the armpit and played with a stick} \tbOP{lonni} \tbTP{lonnee} 
\tbLX{loɔ} \tbHM{} \tbPH{lʊ́ɔ́} \tbPS{n.~} \tbGE{the immediate front, under of one's self} \tbOP{loɔre} \tbTP{loɔrɛɛ} 
\tbLX{loɔraa} \tbHM{} \tbPH{lʊ́ɔ̀ráá} \tbPS{n.~} \tbGE{leopard} \tbOP{loɔre} \tbTP{loɔrɛɛ} \tbVA{wɛnyuo, nyuo}
\tbLX{loɔre} \tbHM{} \tbPH{lʊ́ɔ́rɪ́} \tbPS{v.~} \tbGE{to move from one to the other e.g. branches of a tree} \tbXV{A lɔnnɔ lɔnne la a ulee te ta o manzelɛ.} \tbXE{The squirrel moved along the branch up to the tip.} \tbXV{A bie dɔgereba taa la ba bibaal loɔre ne a asibitiri yaga kyɛ baŋ nyɛ sammo.} \tbXE{The sick child’s parents took him through several hospitals before he was cured.} \tbPD{loɔrɛɛ, loɔrɔ, loɔrɔ, loɔreba, loɔraa, - } 
\tbLX{loɔre} \tbHM{1} \tbPH{lʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to joke} \tbPD{loɔrɛɛ, loɔrɔ, loɔrɔ, loɔreba, loɔraa, - } 
\tbLX{loɔre} \tbHM{2} \tbPH{lʊ̀ɔ̀rì} \tbPS{v.~} \tbGE{to bend into a half-circle} \tbPD{loɔrɛɛ, loɔrɔ, loɔrɔ, loɔreba, loɔraa, - } 
\tbLX{loregboro} \tbHM{} \tbPH{lʊ̀rɪ̀gbʊ̀rʊ̀} \tbPS{adv.~} \tbGE{crumbled} 
\tbLX{lori} \tbHM{} \tbPH{lòrí} \tbPS{n.~} \tbGE{pidgeon} \tbSG{loruu} \tbOP{loe} \tbTP{loree} 
\tbLX{lori} \tbHM{} \tbPH{lórì} \tbPS{v.~} \tbSN \tbGE{to untie} \tbSN \tbGE{to put on weight} \tbXV{Se᷉ɛ᷉ saŋa don-lenni maŋ ŋmeene la kyɛ ka ba wa bare a uoni ka lee, lori a zaa.} \tbXE{During the farming season, tethered animals are lean, but they bounce back in weight when ranging free during the dry season.} \tbPD{-, loro, loro, loribo, -, - } 
\tbLX{lori} \tbHM{} \tbPH{lórí} \tbPS{n.~} \tbGE{meteorite} \tbOP{loe} \tbTP{loree} 
\tbLX{lori} \tbHM{} \tbPH{lòrì} \tbPS{adv.~} \tbGE{very full (from eating)} \tbXV{A diibu noma la ka a bie di tege te lori o zaa.} \tbXE{The food was palatable and the child ate and became very full.} 
\tbLX{lori noɔre} \tbHM{} \tbPH{lórì nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to break a fast} \tbPD{-, -, -, -, -, - } 
\tbLX{lori pɛnne} \tbHM{} \tbPH{lórì pɛ́nnɪ̀} \tbPS{v.~} \tbGE{to tell lies} \tbPD{-, -, -, -, -, - } 
\tbLX{loroto} \tbHM{} \tbPH{lórótó} \tbPS{adv.~} \tbGE{jellylike e.g. blister on the palm} 
\tbLX{loroto} \tbHM{} \tbPH{lórótó} \tbPS{adv.~} \tbGE{on continuous terrain (smooth-moving vehicle)} 
\tbLX{lɔbɔ} \tbHM{} \tbPH{lɔ́bɔ́} \tbPS{n.~} \tbSN \tbGE{riddle} \tbSN \tbGE{A charm believed to be needle-like and has super natural power that can inflict sickness on someone} \tbVA{lɔba, lɔɔbiri}
\tbLX{lɔɛ duoraa} \tbHM{} \tbPH{lɔ́ɛ̀ dúóráá} \tbPS{n.~} \tbGE{parking lot} \tbOP{lɔɛ duori} \tbTP{lɔɛ duoree} 
\tbLX{lɔge} \tbHM{} \tbPH{lɔ́gì} \tbPS{n.~} \tbGE{law} \tbTP{lɔgerɛɛ} 
\tbLX{lɔge} \tbHM{} \tbPH{lɔ̀gɪ́} \tbPS{n.~} \tbGE{a type of wild mouse} \tbOP{lɔgere} \tbTP{lɔgerɛɛ} \tbVA{lɔɔ}
\tbLX{lɔgetɔgɔ} \tbHM{} \tbPH{lɔ̀gɪ̀tɔ̀gɔ̀} \tbPS{adv.~} \tbGE{fat, flabby, and heavy} 
\tbLX{lɔgɔ} \tbHM{} \tbPH{lɔ̀gɔ̀} \tbPS{n.~} \tbGE{a type of small bag in which personal documents are kept and usually carried about} \tbOP{lɔgre} \tbTP{lɔgrɛɛ} 
\tbLX{lɔgɔ} \tbHM{} \tbPH{lɔ́gɔ́} \tbPS{n.~} \tbGE{state of helplessness} \tbXV{A bie bebiri zaa baaloŋ vɛŋɛɛ ka o dɔgerebɔ lɔgɔ zaa yi.} \tbXE{The child's frequent illness has made his parents helpless.} \tbOP{lɔgere} \tbTP{lɔgerɛɛ} 
\tbLX{lɔlɔraa} \tbHM{} \tbPH{lɔ̀lɔ̀ráà} \tbPS{n.~} \tbGE{calf (part of leg)} \tbOP{lɔlɔrre} \tbTP{lɔlɔrɛɛ} 
\tbLX{lɔlɔrezeɛ} \tbHM{} \tbPH{lɔ̀lɔ̀rɪ̀zɪ̀ɛ́} \tbPS{n.~} \tbGE{tree, is medicinal} \tbOP{lɔlɔrezeere} \tbTP{lɔlɔrezeerɛɛ} 
\tbLX{lɔmɛɛ} \tbHM{} \tbPH{lɔ̀mɛ̀ɛ̀} \tbPS{adv.~} \tbGE{fat and soft} 
\tbLX{lɔnne} \tbHM{} \tbPH{lɔ́nnɪ̀} \tbPS{v.~} \tbGE{to move from one point to another e.g. on a tree} \tbXV{A ŋmaaŋa lɔnne la a teɛ ulli zaa kyɛ baŋ sigi.} \tbXE{The monkey climbed through all the branches of the tree before coming down.} \tbPD{lɔŋɛɛ, lɔnnɔ, lɔnnɔ, lɔnema, lɔnnaa, lɔnema } \tbVA{loɔre}
\tbLX{lɔnnɔ} \tbHM{} \tbPH{lɔ̀nnɔ́} \tbPS{n.~} \tbGE{squirrel} \tbOP{lɔnne} \tbTP{lɔnnɛɛ} \tbVA{lanna}
\tbLX{lɔŋe} \tbHM{} \tbPH{lɔ̂ŋ} \tbPS{v.~} \tbGE{to spread, to move from one person or area to another e.g. disease, fire} \tbPD{lɔŋɛɛ, lɔnnɔ, lɔnnɔ, lɔnema, lɔnnaa, lɔnema } \tbVA{lɔnne}
\tbLX{lɔŋe} \tbHM{} \tbPH{lɔ́ŋ} \tbPS{v.~} \tbGE{to sag, to be slack e.g. drying line} \tbPD{lɔŋɛɛ, lɔnnɔ, lɔnnɔ, lɔnemɔ, lɔnnaa, - } 
\tbLX{lɔŋɔ} \tbHM{} \tbPH{lɔ́ŋɔ́} \tbPS{n.~} \tbGE{frog} \tbOP{lɔnne} \tbTP{lɔnnɛɛ} \tbVA{lɔŋa}
\tbLX{lɔɔ} \tbHM{} \tbPH{lɔ̀ɔ́} \tbPS{n.~} \tbGE{mouse} \tbOP{lɔɔre} \tbTP{lɔɔrɛɛ} \tbVA{lɔge}
\tbLX{lɔɔ} \tbHM{} \tbPH{lɔ́ɔ́} \tbPS{v.~} \tbSN \tbGE{to drop, to fall (from one's hands)} \tbSN \tbGE{to throw down (wrestling)} \tbSN \tbGE{to subdue, to overpower (argument)} \tbSN \tbGE{to be greater than, to be more than} \tbXV{A pɔgeba veɛlɛ la kyɛ a bonsɔgelaa naŋ lɔɔ ba zaa.} \tbXE{The women are beautiful, but the black one beats them all.} \tbSN \tbGE{to put drops into (eyes)} \tbSN \tbGE{to dash a buyer for an item bought; (tomatoes)} \tbPD{lɔɔɛ, lɔɔrɔ, lɔɔrɔ, lɔreba, lɔɔraa, lɔreba } \tbSE{lɔɔ baare} \tbGE{to throw away, to get rid of, to discard} 
\tbLX{lɔɔbiri} \tbHM{} \tbPH{lɔ́ɔ́bírí} \tbPS{n.~} \tbGE{a charm believed to be mobile and have destructive powers} \tbOP{lɔɔbie} \tbTP{lɔɔbiiree} 
\tbLX{lɔɔlage} \tbHM{} \tbPH{lɔ́ɔ́làgɪ̀} \tbPS{v.~} \tbGE{to surround} \tbPD{lɔɔlagɛɛ, lɔɔrɔlagra, lɔɔlagra, lɔɔlagreba, lɔɔlagraa, - } 
\tbLX{lɔɔre} \tbHM{1} \tbPH{lɔ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{lorry} \tbOP{lɔɛ} \tbTP{lɔrrɛɛ} 
\tbLX{lɔɔre} \tbHM{2} \tbPH{lɔ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{items given to a buyer for free e.g pepper} \tbXV{lɔbɔ} \tbXE{} 
\tbLX{lɔɔ-wullo} \tbHM{} \tbPH{lɔ́ɔ-́wúlló} \tbPS{n.~} \tbGE{scribe} \tbOP{lɔɔ-wulibo} 
\tbLX{lɔre} \tbHM{} \tbPH{lɔ̀rɪ̀} \tbPS{v.~} \tbGE{to soak, to immerse} \tbXV{A pɔge lɔre la  muri a na wuo peɛ.} \tbXE{The woman has soaked straw to weave a basket.} \tbPD{lɔrɛɛ, lɔrɔ, lɔrɔ, lɔreba, lɔraa, lɔreba } 
\tbLX{lɔre} \tbHM{} \tbPH{lɔ̀rɪ̀} \tbPS{v.~} \tbGE{to be ugly} \tbXV{A pɔge bonsuuri la so ka o lɔre.} \tbXE{The woman's clothing has made her ugly.} \tbPD{lɔrɛɛ, lɔrɔ, lɔrɔ, lɔreba, lɔraa, lɔreba } 
\tbLX{lɔre} \tbHM{} \tbPH{lɔ̀rɪ̀} \tbPS{adj.~} \tbGE{weak, useless} \tbXV{Ka a bieŋ ba lɔre o da na tuo la a daga ŋa eŋ.} \tbXE{If the child had not been weak, he could, at least, have carried this box.} 
\tbLX{lɔre} \tbHM{} \tbPH{lɔ́rɪ̀} \tbPS{v.~} \tbGE{to be lucky} \tbXV{N zaameŋ ba yeli ka fo na lɔre la zenɛ, ana naane ka fo ba toɔle fo libie.} \tbXE{Didn't I say you would be lucky today? See, your father has sent you money.} \tbPD{lɔrɛɛ, lɔrɔ, lɔrɔ, lɔreba, lɔraa, lɔreba } \tbVA{nɔre}
\tbLX{lɔre} \tbHM{} \tbPH{lɔ̀rɪ̀} \tbPS{adv.~} \tbGE{onomatopoeiac, sounding like a hard object hitting a soft object} 
\tbLX{lɔrehazie} \tbHM{} \tbPH{lɔ́rɪ́házìé} \tbPS{n.~} \tbGE{car rental company} \tbOP{lɔrehaziiri} \tbTP{lɔrehaʊziiree} 
\tbLX{lugi} \tbHM{} \tbPH{lùgì} \tbPS{v.~} \tbGE{to induce} \tbPD{lugee, lugiro, lugiro, lugiribo, lugiraa, - } 
\tbLX{lugi} \tbHM{} \tbPH{lùgì} \tbPS{v.~} \tbGE{to swell a little e.g. armpit, groin} \tbPD{lugee, lugiro, lugiro, lugiribo, lugiraa, - } 
\tbLX{lugiri} \tbHM{} \tbPH{lúgirí} \tbPS{n.~} \tbGE{a forked beam that supports the main beam in a swish-thatched roof} \tbOP{lugo} \tbTP{lugiree} 
\tbLX{lugiri} \tbHM{} \tbPH{lúgirí} \tbPS{v.~} \tbGE{to loosen (a nut on a bolt), to ease} \tbPD{lugiree, lugiro, lugiro, lugiribo, lugiribo, lugiraa } 
\tbLX{luŋguri} \tbHM{} \tbPH{lùŋgúrì} \tbPS{n.~} \tbGE{lemon} \tbSG{luŋguruu} \tbOP{luŋgue} \tbTP{luŋgurree} 
\tbLX{luori} \tbHM{} \tbPH{lùòrì} \tbPS{v.~} \tbSN \tbGE{to become bowed e.g. legs} \tbXV{A bie baaloŋ eɛ la kao gbɛe luori.} \tbXE{The child's sickness has caused his legs to become bowed.} \tbSN \tbGE{to extend the size of a vessel} \tbXV{A pɔgebeɛlebɛ da luori la poore na di dooga.} \tbXE{The bridemaids had prepared their stomachs to feast.} \tbPD{luoree, luoro, luoro, luoribo, luoraa, - } 
\tbLX{luou} \tbHM{} \tbPH{lùòú} \tbPS{n.~} \tbGE{west} \tbTP{luoree} \tbVA{saniŋe}
\tbLX{luri} \tbHM{} \tbPH{lùrì} \tbPS{v.~} \tbGE{to place underneath to hide e.g. armpit} \tbPD{luree, luro, luro, luribo, luraa, - } 
\tbLX{luri} \tbHM{} \tbPH{lúrì} \tbPS{v.~} \tbGE{to be filled with e.g. smoke} \tbXV{A vu᷉u᷉ ba dire ka nyoore luri a korodie zaa.} \tbXE{The fire is not burning cleanly, so the kitchen is choked with smoke.} \tbPD{luree, luro, luro, luribo, luraa, - } 
\tbLX{luri} \tbHM{} \tbPH{lúrì} \tbPS{v.~} \tbGE{to play a gourd drum vigorously} \tbXV{A kɔŋŋmeɛrɛ luri la a koɔre eŋ a gyile poɔ ne toori.} \tbXE{The gourd drummer played the drum alongside the xylophone with artistry.} \tbPD{luree, luro, luro, luribo, luraa, - } 
\tbLX{luroluro} \tbHM{} \tbPH{lúrólúró} \tbPS{adj.~} \tbGE{in a circular shape (usually within a small enclosure)} \tbXV{A nɔzage waa la luroluro lɛ.} \tbXE{The hen coop is roudish within.} 
\tbLX{luroluro} \tbHM{} \tbPH{lùròlùrò} \tbPS{adv.~} \tbGE{in a circular shape (usually within a big enclosure)} \tbXV{A taŋa bogi e la luroluro.} \tbXE{The inner part of the cave is large and round.} 
\tbLX{lurri} \tbHM{} \tbPH{lúrrì} \tbPS{v.~} \tbGE{to smooth out a mound by covering all protruding matter with soil} \tbXV{Ba maŋ lurri la vubiri ka o naa ta ona ka waa to᷉o᷉ nyɛ.} \tbXE{A yam mound is usually enlarged by adding more soil to enable the yam bear a good tuber.} \tbPD{lurree, lurro, luribo, lurraa, -, - } 
\tbLX{luuraa} \tbHM{} \tbPH{lùùráà} \tbPS{n.~} \tbGE{denial} \tbOP{luuri} \tbTP{luuree} 
\tbLX{luuri} \tbHM{} \tbPH{lùùrì} \tbPS{v.~} \tbGE{to duck, to bend to pass under a hanging object e.g. beam} \tbPD{-, luuro, luuro, luuribo, luuraa, lure } 
\tbLX{luuri} \tbHM{} \tbPH{lùùrì} \tbPS{v.~} \tbSN \tbGE{to renounce one's word or promise, to deny} \tbXV{O daare yeli ka o baŋe la a nannyige kyɛ ka zenɛ ka leɛ luuri o noɔre.} \tbXE{S/he said s/he knew the thief, but today s/he denies that s/he said so.} \tbSN \tbGE{to return something in exchange for something else} \tbXV{A dɔɔ luuri la n libie ko ma kyɛ de o noɔ.} \tbXE{The man has returned my money and taken back his fowl.} \tbPD{-, luuro, luuro, luuribo, luuraa, lure } 
\tbLX{ma} \tbHM{} \tbPH{mǎ} \tbPS{n.~} \tbGE{mother, female that has reproduced} \tbOP{mamine} \tbTP{marɛɛ} 
\tbLX{ma} \tbHM{} \tbPH{mà} \tbPS{v. pl.~} \tbGE{to stick to} \tbXV{Goore ma la a kparoo.} \tbXE{Thorns are stuck to the shirt.} 
\tbLX{ma} \tbHM{} \tbPH{má} \tbPS{particle.~} \tbGE{suppose, hope; it is used only in question form} \tbXV{Ma fo na gaa la?} \tbXE{I suppose/hope you will go?.} 
\tbLX{ma} \tbHM{} \tbPH{má} \tbPS{pronoun.~} \tbGE{me (first person singular)} \tbXV{O de la a gane ko ma.} \tbXE{S/he has given the book to me.} 
\tbLX{maa} \tbHM{} \tbPH{máâ} \tbPS{pronoun.~} \tbSN \tbGE{I (first person singular personal emphatic)} \tbXV{O yeli ka maa la sa᷉a.} \tbXE{S/he said it was I who was wrong.} \tbSN \tbGE{me (first person singular personal emphatic)} \tbXV{Maa la so a gane.} \tbXE{The book belongs to me.} 
\tbLX{maa} \tbHM{} \tbPH{máà} \tbPS{v.~} \tbSN \tbGE{to be wet} \tbXV{Saa mie la ka a teŋɛ zaa maa.} \tbXE{It has rained, making the ground wet.} \tbSN \tbGE{to be cold (objects)} \tbXV{A baala nyu la te᷉e᷉ ka o eŋɛ leɛ maa a zaa.} \tbXE{The patient has taken medicine and his temperature has gone down.} \tbSN \tbGE{to make peaceful} \tbXV{Zenɛ a zɔɔzɔɔreba na ba wa lɛnso ka a zie maa a ŋaa.} \tbXE{Today, those quarrelsome men did not come and that is why the place is peaceful.} \tbPD{maaɛ, maana, maana, maanema, maanaa, maanema } 
\tbLX{maa} \tbHM{} \tbPH{màà̀} \tbPS{adv.~} \tbSN \tbGE{move streamingly (used only with liquid substances)} \tbXV{O ma naŋ da kpi o deɛ kono la nentaŋ maaa.} \tbXE{When her mother died, she was shedding tears uncontrolably.} \tbSN \tbGE{slimy} 
\tbLX{maa gbɛre} \tbHM{} \tbPH{máá gbɛ́rɪ̀} \tbPS{v.~} \tbGE{to delay, to be late} \tbXV{Ba naŋ boɔle o o ba maa gbɛre kyɛ wa.} \tbXE{When he was called, he did not waste time but responded promptly.} \tbPD{-, -, -, -, -, - } 
\tbLX{maalam} \tbHM{} \tbPH{máálàm} \tbPS{n.~} \tbGE{Islamic cleric} \tbOP{maalammine} \tbTP{maalonnɛɛ} 
\tbLX{maale} \tbHM{} \tbPH{mààlɪ̀} \tbPS{v.~} \tbSN \tbGE{to make, to construct, to draw} \tbXV{A kɔnarata maale la baraakyi ŋmaa a baa.} \tbXE{The contractor has built a bridge across the river.} \tbSN \tbGE{to cook} \tbXV{Boŋ bondirii ka a bondimaala na maale zenɛ?} \tbXE{What food will the cook cook today?.} \tbPD{maalɛɛ, maala, maala, maaleba, maalaa, - } 
\tbLX{maale} \tbHM{} \tbPH{mààlɪ̀} \tbPS{v.~} \tbSN \tbGE{to bless, to make good, to clean, to make beautiful} \tbXV{Ba maale la a die poɔ veɛlaa.} \tbXE{They have cleaned the room well.} \tbSN \tbGE{to pacify, to offer sacrifice} \tbXV{Dagaaba mine maŋ maale la bagere.} \tbXE{Some Dagaaba sacrifice to their gods.} \tbPD{maalɛɛ, maala, maala, maaleba, maalaa, - } 
\tbLX{maale} \tbHM{} \tbPH{máálɪ̀} \tbPS{v.~} \tbGE{to cause to calm down, to reduce the intensity of} \tbXV{A yidaandɔɔ maale la a yɛlɛ ka a zɔɔre baare.} \tbXE{The family head calmed tempers and ended the quarrel.} \tbPD{-, -, -, -, -, - } 
\tbLX{maalekondi} \tbHM{} \tbPH{mààlɪ̀kʊ́ndí} \tbPS{n.~} \tbGE{a type of ritual food that is eaten only by specific people i.e. those who are members of the ritual} 
\tbLX{maaleŋ} \tbHM{} \tbPH{mààlɪ̀ŋ} \tbPS{adv.~} \tbGE{just too much} \tbXV{Saa la maaleŋ miire tu.} \tbXE{The rain just pours too much.} 
\tbLX{maaleŋ} \tbHM{} \tbPH{mààlɪ̀ŋ} \tbPS{adv.~} \tbGE{again} \tbXV{Saa la maaleŋ miire tu.} \tbXE{The rain just pours too much.} 
\tbLX{maane} \tbHM{} \tbPH{màànɪ̀} \tbPS{adj.~} \tbSN \tbGE{flattened} \tbXV{A kuree feele o la ka o pare deɛ waa maane lɛ} \tbXE{The undersized pair of knickers has made his buttocks flat.} \tbSN \tbGE{premature (egg)} \tbOP{maama} \tbTP{maanɛɛ} 
\tbLX{maanee} \tbHM{} \tbPH{máánɪ́ɪ́} \tbPS{n.~} \tbGE{okra} \tbSG{maanoo} \tbOP{maana} \tbTP{maanɛɛ} \tbVA{saalommaane}
\tbLX{maa noɔre} \tbHM{} \tbPH{máá nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to be surprised} \tbPD{-, -, -, -, -, - } 
\tbLX{maaŋmeɛ} \tbHM{} \tbPH{máàŋmɪ́ɛ̀} \tbPS{n.~} \tbGE{a kind of tall, brown, lean bird} \tbOP{maaŋmeere} 
\tbLX{maare} \tbHM{} \tbPH{máárɪ̀} \tbPS{v.~} \tbGE{to hold, to support with open palms facing upwards e.g. baby} \tbPD{maarɛɛ, maara, maara, maareba, maaraa, - } 
\tbLX{maaroŋ} \tbHM{} \tbPH{máárʊ́ŋ} \tbPS{adj.~} \tbGE{slow} \tbXV{O taa la maaroŋ zuo lɛnso ka ba wa pɔge kpaaroŋ.} \tbXE{He is very slow and that is why he did not come to the meeting.} 
\tbLX{maaroŋ} \tbHM{} \tbPH{máárʊ́ŋ} \tbPS{adj.~} \tbSN \tbGE{cold, cool} \tbXV{Firigyi ko᷉ɔ᷉ maŋ e la maaroŋ.} \tbXE{Water stored in the fridge is always cold.} \tbSN \tbGE{wet} \tbXV{Ka saaŋ mi a zie zaa maŋ e la maaroŋ lɛ.} \tbXE{When it rains, the whole place becomes wet.} 
\tbLX{maaroŋ} \tbHM{} \tbPH{máárʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{cold, coolness} \tbXV{Yuoni ŋa a maaroŋ zuoe la.} \tbXE{This year, the cold is excessive.} \tbSN \tbGE{shade} \tbXV{Tentoloŋ saŋa dendɔre teɛ maaroŋ maŋ tere la zemmo.} \tbXE{During the hot season, the shade of trees is always inviting.} 
\tbLX{maaroŋ} \tbHM{} \tbPH{máárʊ́ŋ} \tbPS{n.~} \tbGE{peace} \tbXV{Ka zieŋ na maale ferebo ka maaroŋ be be.} \tbXE{If development is to take place there must be peace.} 
\tbLX{maaroŋ} \tbHM{} \tbPH{máárʊ́ŋ} \tbPS{n.~} \tbGE{fever} \tbXV{Maaroŋ kpɛ la dɔɔ ka o gaa asibiti.} \tbXE{The man has a fever and has gone to the hospital.} 
\tbLX{maasa} \tbHM{} \tbPH{máásà} \tbPS{n.~} \tbGE{a type of cake made from millet flour} 
\tbLX{maa suuri} \tbHM{} \tbPH{máà súúrí} \tbPS{v.~} \tbGE{to appease, to pacify} \tbPD{-, -, -, -, -, - } 
\tbLX{maayee} \tbHM{} \tbPH{mááyéé} \tbPS{adv.~} \tbGE{empty} \tbXV{Nannyigiri da ŋmaa la a woɔ pare iri a libie, ka a dɔɔ da wa tuno a woɔ ka maayee.} \tbXE{Thieves had cut the bottom of the bag and removed all the man's money, and when he later put his hand into the bag, it was empty.} 
\tbLX{mabie} \tbHM{} \tbPH{mábíé} \tbPS{n.~} \tbGE{sibling, relative} \tbOP{mabiiri} \tbTP{mabiiree} 
\tbLX{mabiekontapɔgema} \tbHM{} \tbPH{mábíékʊ́ntápɔ́gɪ́má} \tbPS{n.~} \tbGE{a type of small, short snake that is highly venomous} \tbTP{mabiekontapɔgemɛɛ} 
\tbLX{mabile} \tbHM{} \tbPH{màbílé} \tbPS{n.~} \tbGE{maternal aunt, mother's sister} \tbOP{mabilii, mabilimine} \tbTP{mabilee} 
\tbLX{mabo} \tbHM{} \tbPH{màbʊ̀} \tbPS{adv.~} \tbGE{marshy and messy} 
\tbLX{maee} \tbHM{} \tbPH{màɪ̀ɪ̀} \tbPS{adj.~} \tbGE{red} \tbXV{A pɔgelee niŋ waa la maee lɛ.} \tbXE{The girl's face is looking red.} 
\tbLX{maemae} \tbHM{} \tbPH{máɪ́máɪ́} \tbPS{adj.~} \tbGE{bright red} \tbXV{A fele e la zeɛ maemae lɛ.} \tbXE{The saw is bright red.} 
\tbLX{magaka} \tbHM{} \tbPH{màgakà̀} \tbPS{adv.~} \tbGE{deep red} 
\tbLX{mage} \tbHM{} \tbPH{mágɪ́} \tbPS{v.~} \tbSN \tbGE{to write} \tbXV{Karekyere maŋ mage la gane velaa.} \tbXE{Literate people write well.} \tbSN \tbGE{to make beautiful patterns; usually on some clothes} \tbXV{A teelɛ mage la a kparoo niŋe ne puori zaa.} \tbXE{The tailor has designed both the front and back of the shirt.} \tbPD{magɛɛ, magra, magra, magreba, magraa, - } 
\tbLX{magedalee} \tbHM{} \tbPH{mágɪ́dàléé} \tbPS{n.~} \tbGE{writing stick, pen} \tbOP{magedabilii} \tbTP{magedabillee} 
\tbLX{magelakala} \tbHM{} \tbPH{màgɪ̀làkàlà} \tbPS{adj.~} \tbGE{skinny} \tbXV{A dɔɔ wogi na bɛɛ la a te waa magelakala lɛ.} \tbXE{That tall man has been sick for a long time hence his being skinny.} 
\tbLX{magelaŋ} \tbHM{} \tbPH{màgɪ̀làŋ} \tbPS{adj.~} \tbGE{large and flat} \tbVA{maŋgalaŋ}
\tbLX{magere} \tbHM{} \tbPH{mágɪ́rɪ̀} \tbPS{n.~} \tbGE{sign, mark} \tbTP{magrɛɛ} 
\tbLX{magetaga} \tbHM{} \tbPH{mágɪ̀tàgà} \tbPS{adv.~} \tbGE{thick and marshy} \tbVA{magtaga}
\tbLX{makoma} \tbHM{} \tbPH{màkʊ́má} \tbPS{n.~} \tbGE{grandmother} \tbOP{makommine} \tbTP{makomɛɛ/makonnɛɛ} 
\tbLX{makontulo} \tbHM{} \tbPH{màkʊ́ntúló} \tbPS{n.~} \tbGE{great-grandmother} 
\tbLX{makoŋkommuulitoori} \tbHM{} \tbPH{màkʊ́ŋkʊ́mmúúlítòòrí} \tbPS{n.~} \tbGE{great-grandmother} \tbOP{makoŋkommuulitobo} \tbTP{makoŋkommuulitooree} 
\tbLX{makpe᷉ɛ᷉} \tbHM{} \tbPH{màkpɪ́᷈ɛ́᷉} \tbPS{n.~} \tbGE{stepmother; older wife of one’s biological father or father’s elder brother’s wife} \tbOP{makpeɛmine} \tbTP{makpeɛnɛɛ/makpeɛmɛɛ} 
\tbLX{makpoŋ} \tbHM{} \tbPH{màkpóŋ} \tbPS{n.~} \tbGE{} \tbVA{makpe᷉ɛ᷉}
\tbLX{malaa} \tbHM{} \tbPH{màláà} \tbPS{n.~} \tbGE{a type of bird with deep blue and red eyes} \tbOP{ malle} \tbTP{mallɛɛ} 
\tbLX{malammalam} \tbHM{} \tbPH{màlàmmálám} \tbPS{adj.~} \tbGE{soiled e.g. mud} 
\tbLX{malantamm} \tbHM{} \tbPH{màlàntàmm} \tbPS{adj.~} \tbGE{bundled up} 
\tbLX{male} \tbHM{} \tbPH{màlɪ̀} \tbPS{v.~} \tbGE{to claim or pretend to} \tbXV{O male ka o so la ko᷉ɔ᷉} \tbXE{S/he claims to have taken his/her bath.} \tbXV{O male ka o taa la libie kyɛ ka o bie kaŋa zaa ba gaa sakuuri.} \tbXE{S/he claims to be rich but none of his /her children has been to school.} \tbPD{malɛɛ, malla, malla, maleba, malaa, maleba } 
\tbLX{male} \tbHM{} \tbPH{màlɪ̀} \tbPS{v.~} \tbGE{to suffer repercussions, to punish} \tbXV{Fomeŋɛ ka na male.} \tbXE{You will only end up punishing yourself.} \tbXV{Ka fo saaŋ yeli ka fo gaa sakuuri ka fo zagere fomeŋɛ ka male.} \tbXE{If your father sends you to school and you refuse you only punish yourself.} \tbPD{malɛɛ, malla, malla, maleba, malaa, maleba } 
\tbLX{male} \tbHM{} \tbPH{màlì} \tbPS{adv.~} \tbGE{anomatopoeic word expressing the way the soud of a cut} 
\tbLX{malefa} \tbHM{} \tbPH{málɪ́fà} \tbPS{n.~} \tbGE{gun} \tbOP{malefare} \tbTP{malefarɛɛ} 
\tbLX{malefabiri} \tbHM{} \tbPH{málɪ́fàbírì} \tbPS{n.~} \tbGE{bullet} \tbSG{malefabiruu} \tbOP{malefabie} \tbTP{malefabiiree} 
\tbLX{malefala} \tbHM{} \tbPH{màlɪ̀fàlà} \tbPS{adv.~} \tbGE{tasteless} 
\tbLX{malefate᷉e} \tbHM{} \tbPH{málɪ́fàtɪ̀᷉ɪ́᷉} \tbPS{n.~} \tbGE{gunpowder} \tbOP{malefateenɛɛ} 
\tbLX{malemale} \tbHM{} \tbPH{málɪ́málɪ́} \tbPS{adv.~} \tbGE{very black} \tbVA{kpirikpiri}
\tbLX{malemale} \tbHM{} \tbPH{màlɪ̀málɪ́} \tbPS{v.~} \tbGE{to struggle to do} \tbPD{malemalɛɛ, malemala, malemala, malemaleba, malemalaa, - } 
\tbLX{mana} \tbHM{} \tbPH{máná} \tbPS{n.~} \tbGE{rubber, elastic} \tbSG{manoo} \tbOP{manne} \tbTP{mannɛɛ} 
\tbLX{mana (Hausa)} \tbHM{} \tbPH{mànà} \tbPS{particle.~} \tbGE{of course} \tbXV{kɔlaa la kpɛ a die ona la ɔɔ a nɛne mana.} \tbXE{A cat went into the room; of course it is the one that has eaten the meat.} 
\tbLX{manamana} \tbHM{} \tbPH{mànàmànà} \tbPS{adj.~} \tbGE{sticky} 
\tbLX{mane} \tbHM{} \tbPH{mànɪ́} \tbPS{n.~} \tbGE{large body of water} \tbOP{mama} \tbTP{mannɛɛ} 
\tbLX{mankyere} \tbHM{} \tbPH{mánʧɪ̀rɪ̀} \tbPS{n.~} \tbGE{match} \tbOP{ mankyeɛ} \tbTP{mankyerɛɛ} 
\tbLX{mankyɛlɛ} \tbHM{} \tbPH{mánʧɛ̀lɛ̀} \tbPS{n.~} \tbGE{diarrhoea} 
\tbLX{manne} \tbHM{} \tbPH{mánnɪ̀} \tbPS{v.~} \tbSN \tbGE{to measure} \tbXV{Kookoolaa ka o manne ne a kyi.} \tbXE{They used a measuring bowl to measure the millet.} \tbXV{Ba de la kɛkɛre manne ne a dendɔre wogiruŋ.} \tbXE{They used a stalk to measure the height of the door.} \tbSN \tbGE{to compare} \tbXV{Sakubie koŋ tõɔ de omeŋɛ te manne ne karema.} \tbXE{A pupil cannot compare herself to a teacher.} \tbPD{mannɛɛ, manna, manna, manema, mannaa, - } 
\tbLX{manne} \tbHM{} \tbPH{mánnɪ̀} \tbPS{v.~} \tbGE{to narrate, to explain} \tbXV{O manne la a zɔɔre pare.} \tbXE{S/he explained the cause of the fight.} \tbXV{N naŋ da gaa n zɔmeŋɛ dɔgoo bebiri tigiri N da kuli te manne la a zaa ko n dɔgereba.} \tbXE{After attending my best friend’s birthday party, I went home and narrated everything to my parents.} \tbPD{mannɛɛ, manna, manna, manema, mannaa, - } 
\tbLX{mannoɔre} \tbHM{} \tbPH{mànnʊ́ɔ́r} \tbPS{n.~} \tbGE{beach} \tbOP{mannɔɛ} \tbTP{mannɔrɛɛ} 
\tbLX{mansugiri} \tbHM{} \tbPH{mànsùgirì} \tbPS{n.~} \tbGE{a type of wild mushroom with medicinal properties} \tbSG{mansugruu} \tbOP{mansugo} \tbTP{mansugree} 
\tbLX{mansugo} \tbHM{} \tbPH{mànsùgò} \tbPS{n.~} \tbSN \tbGE{a type of stomach ailment; hernia} \tbSN \tbGE{type of medicine for hernia} \tbOP{mansugiri} \tbTP{mansugiree} 
\tbLX{mantare} \tbHM{} \tbPH{màntàrɪ̀} \tbPS{n.~} \tbSN \tbGE{Southern Dagaare dialect} \tbSN \tbGE{the area where the Southern Dagaare dialect is spoken} \tbOP{mantarɛɛ} 
\tbLX{mantaroo} \tbHM{} \tbPH{màntàrʊ́ʊ́} \tbPS{n.~} \tbGE{speaker of Southern Dagaare dialect} \tbOP{mantarɛɛ} 
\tbLX{manteɛ} \tbHM{} \tbPH{mǎntɪ̀ɛ́} \tbPS{n.~} \tbGE{rubber tree} \tbOP{manteere} \tbTP{manteerɛɛ} 
\tbLX{manzelɛ} \tbHM{} \tbPH{mànzɪ́lɛ́} \tbPS{n.~} \tbGE{tip of (branch of tree)} \tbOP{manzelle} \tbTP{manzelɛɛ} 
\tbLX{maŋgalaŋ} \tbHM{} \tbPH{maŋgalaŋ} \tbPS{adj.~} \tbGE{flattened} \tbVA{magelaŋ}
\tbLX{maŋgaŋene} \tbHM{} \tbPH{màŋgáŋɪ́nɪ́} \tbPS{n.~} \tbGE{the other side of the sea; overseas regions} \tbTP{maŋgaŋenɛɛ} 
\tbLX{maŋgbɛre} \tbHM{} \tbPH{máŋgbɛ́rɪ́} \tbPS{n.~} \tbGE{tributary (stream)} \tbOP{maŋgbɛɛ} \tbTP{maŋgbɛɛrɛɛ} 
\tbLX{maŋkeni} \tbHM{} \tbPH{mâŋkèní} \tbPS{n.~} \tbGE{cocoyam} \tbSG{maŋkenuu} \tbOP{maŋkeme} \tbTP{maŋkenee/maŋkennee} 
\tbLX{mapo} \tbHM{} \tbPH{mápʊ̀} \tbPS{n.~} \tbGE{map} \tbOP{mapore} \tbTP{maporɛɛ} 
\tbLX{mara} \tbHM{} \tbPH{màrà} \tbPS{n.~} \tbGE{courage, persistence, steadfastness} \tbTP{marɛɛ} 
\tbLX{mara} \tbHM{} \tbPH{màrà} \tbPS{adv.~} \tbGE{onomatopoeia, sounding like the sound coming from one stepping into muddy matter} 
\tbLX{maraa} \tbHM{} \tbPH{máráá} \tbPS{n.~} \tbGE{necklace} \tbOP{marre} \tbTP{marrɛɛ} 
\tbLX{maraa} \tbHM{} \tbPH{máráá} \tbPS{n.~} \tbGE{glue} \tbOP{marre} \tbTP{marrɛɛ} 
\tbLX{maraa} \tbHM{} \tbPH{máráá} \tbPS{n.~} \tbGE{cloth used to carry a baby on the back} \tbOP{marre} \tbTP{marrɛɛ} 
\tbLX{maraa} \tbHM{} \tbPH{màràà} \tbPS{adj.~} \tbGE{not responding to anything, quiet, mute} \tbXV{A pɔge suuri naŋ iri na fo maŋ yeli la yɛlɛ ko o ka e maraa.} \tbXE{The woman is annoyed so when you talk to her she stays mute.} 
\tbLX{maraawa} \tbHM{} \tbPH{màrááwà} \tbPS{n.~} \tbGE{a type of tree animal} \tbOP{maraaware} \tbTP{maraawarɛɛ} 
\tbLX{maramara} \tbHM{} \tbPH{màràmàrà} \tbPS{adj.~} \tbGE{marshy} \tbXV{Saa mie la ko᷉ɔ laare a tɛne ka a waa maramara lɛ.} \tbXE{The soil has been beaten by rain and has become marshy.} 
\tbLX{marata} \tbHM{} \tbPH{màràtà} \tbPS{adv.~} \tbGE{refusing to respond to anything} 
\tbLX{mare} \tbHM{} \tbPH{màrɪ̀} \tbPS{v.~} \tbGE{to stick to, to attach to} \tbXV{Dunduli mare la a vaa poɔ.} \tbXE{There is a caterpillar hanging on the leaf.} \tbPD{marɛɛ, mara, mara, mareba, maraa, - } 
\tbLX{mare} \tbHM{} \tbPH{màrɪ̀} \tbPS{v.~} \tbGE{to carry on one's back e.g. a baby} \tbXV{A pɔge mare la a bie ka na gbiri.} \tbXE{The woman has carried the baby on her back in order to make it sleep.} \tbPD{marɛɛ, mara, mara, mareba, maraa, - } 
\tbLX{mare} \tbHM{} \tbPH{màrɪ̀} \tbPS{v.~} \tbGE{to light, to build a fire} \tbXV{Te maŋ mare vu᷉u᷉ maale seemaa.} \tbXE{We build a fire before we cook food.} \tbPD{marɛɛ, mara, mara, mareba, maraa, - } 
\tbLX{mare} \tbHM{} \tbPH{márɪ́} \tbPS{v.~} \tbGE{to detatch (a limb from the body)} \tbPD{marɛɛ, mara, mara, mareba, maraa, - } 
\tbLX{marekpɛ} \tbHM{} \tbPH{màrɪ̀kpɛ̀} \tbPS{v.~} \tbGE{to tease} \tbXV{Koŋkombie mine maŋ marekpɛ la a kotuo deme.} \tbXE{Some lyrics of funeral dirges tease the bereaved family.} \tbPD{-, -, -, -, -, - } 
\tbLX{meele} \tbHM{} \tbPH{mɪ̀ɪ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to wipe off (dirt, water)} \tbSN \tbGE{to soil, to make dirty} \tbPD{meelɛɛ, meelɛ, meelɛ, meelebɛ, -, - } 
\tbLX{meele} \tbHM{} \tbPH{mɪ́ɪ́lɪ́} \tbPS{v.~} \tbGE{to try to incriminate someone; to challenge} \tbXV{O sɔgele gaa toŋ la a toma o yoŋ, o deɛ boɔrɔ ka o meele ma.} \tbXE{S/he did the work alone behind me; I think this is aimed tarnishing my image.} \tbPD{meelɛɛ, meelɛ, meelɛ, meelebɛ, meelaa, - } 
\tbLX{meele eŋ} \tbHM{} \tbPH{mɪ̀ɪ̀lɪ̀ ɪ̀ŋ} \tbPS{v.~} \tbGE{to turn something to replace something else} \tbXV{Waabo doŋ la a nɔnyaŋaa ko ka ba nyɔge nɔdaa meele eŋ a nɔbilii.} \tbXE{The mother fowl was killed by a snake and so a cock was tamed to replace it.} \tbPD{-, -, -, -, -, - } 
\tbLX{meel-eŋ} \tbHM{} \tbPH{mɪ̀ɪ̀l-ɪ̀ŋ} \tbPS{n.~} \tbGE{metaphor (literature)} \tbXV{A bie zoobu wuli ka soɔŋaa la.} \tbXE{The child's speed makes him a rabbit.} 
\tbLX{meelɛ} \tbHM{} \tbPH{mɪ̀ɪ́lɛ̀} \tbPS{n.~} \tbSN \tbGE{a type of grass that is thin but strong} \tbSN \tbGE{decorative ribbons worn across the chest} \tbSN \tbGE{string scale; a locally made scale made from a string and a flat piece of wood or container} \tbTP{meelɛɛ} 
\tbLX{meeloo} \tbHM{} \tbPH{mɪ́ɪ́lʊ́ʊ́} \tbPS{adj.~} \tbGE{thin} \tbOP{meelɛ} 
\tbLX{meemee} \tbHM{} \tbPH{mɪ̀ɪ̀mɪ́ɪ́} \tbPS{v.~} \tbGE{to swing, to turn round (buttocks in a chair)} \tbXV{A bie meemee la pare ka o koŋ gaa.} \tbXE{The child swung his bottom, indicating that he would not go.} \tbPD{meemeeɛ, meemeenɛ, meemeenɛ, meemeenemɛ, meemeenaa, - } 
\tbLX{meɛ} \tbHM{} \tbPH{mɪ́ɛ̀} \tbPS{v.~} \tbGE{to incriminate, to make an accomplice} \tbXV{A bibarewono na zu la o ba libie kyɛ de meɛ ne o yɔɔlee.} \tbXE{That naughty boy stole his father's money and incriminated his younger brother.} \tbPD{meɛɛ, meɛnɛ, meɛnɛ, meɛnemɛ, meɛnaa, - } 
\tbLX{meɛbo} \tbHM{} \tbPH{mɪ́ɛ́bʊ́} \tbPS{n.~} \tbSN \tbGE{structure} \tbXV{A pɔge meɛbo wuli ka o taa la seɛ yelmeŋɛ.} \tbXE{The woman’s structure shows she has a really large waist.} \tbSN \tbGE{character; a distinguishing feature} \tbXV{A pɔlebile na meɛbo ba veɛlɛ.} \tbXE{That young man's character is not good.} 
\tbLX{meɛloŋ} \tbHM{} \tbPH{mɪ̀ɛ̀lʊ́ŋ} \tbPS{n.~} \tbGE{dew; drops of water on vegetation} \tbOP{meɛlɛ} \tbTP{meɛlɛɛ} 
\tbLX{meɛroŋ} \tbHM{} \tbPH{mɪ̀ɛ́rʊ̂ŋ} \tbPS{n.~} \tbSN \tbGE{structure, makeup} \tbSN \tbGE{character, behaviour} 
\tbLX{megemege} \tbHM{} \tbPH{mìgɪ̀mɪ́gɪ́} \tbPS{v.~} \tbGE{to wriggle e.g. fish in muddy water} \tbPD{megemegɛɛ, megemegrɛ, megemegrɛ, megemegrebɛ, megemegraa, - } 
\tbLX{melaa} \tbHM{1} \tbPH{mɪ́láá} \tbPS{n.~} \tbGE{fence} \tbOP{melle} \tbTP{mellɛɛ} 
\tbLX{melaa} \tbHM{2} \tbPH{mɪ́láá} \tbPS{n.~} \tbGE{talebearer} \tbOP{melebɛ} \tbTP{melebɛɛ} 
\tbLX{mele} \tbHM{} \tbPH{mɪ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to fence in} \tbXV{Ba maŋ mele la dabɔgɔ ka donne na ta sa᷉a᷉ o poɔ bombulo.} \tbXE{Gardens are fenced to prevent animals from destroying the plants.} \tbSN \tbGE{to entangle} \tbXV{Ba ba leŋ a boɔ lɛnso ka o mele kpi.} \tbXE{The goat was not tethered well and so it got entangled and died.} \tbSN \tbGE{to spread all over an object (ants)} \tbXV{Zuluŋ mele la boku᷉u᷉.} \tbXE{Soldier ants are all over the carcas.} \tbSN \tbGE{to tattle, to gossip} \tbXV{Bibiiri nɔŋ la mele taa.} \tbXE{Children like tattling on each other.} \tbPD{melɛɛ, melɛ, melɛ, melbɛ, melaa, - } 
\tbLX{melekala} \tbHM{} \tbPH{mɪ̀lɪ̀kàlà} \tbPS{adj.~} \tbGE{large and twisted} 
\tbLX{melekala} \tbHM{} \tbPH{mɪ́lɪ́kálá} \tbPS{adj.~} \tbGE{small and twisted} 
\tbLX{melele} \tbHM{} \tbPH{mɪ́lɪ́lɪ́} \tbPS{adv.~} \tbGE{thin, very slim} 
\tbLX{melemele} \tbHM{} \tbPH{mɪ̀lɪ̀mɪ́lɪ́} \tbPS{v.~} \tbGE{to move with unsteady legs} \tbPD{melemelɛɛ, melemelɛ, melemelɛ, melemelbɛ, melemelaa, - } 
\tbLX{meletaa} \tbHM{} \tbPH{mɪ́lɪ́táá} \tbPS{n.~} \tbGE{the practice of talebearing} \tbXV{Ka noba laŋkpeɛboŋ na noma a seŋ ka ba bare meletaa.} \tbXE{If people want to live in harmony, they must put an end to talebearing.} 
\tbLX{melle} \tbHM{} \tbPH{mìllɪ̀} \tbPS{v.~} \tbGE{to twist} \tbXV{Soɔ da kyebe kyɛ a dɔɔ da de la o nu melle ne a miri ŋmaa.} \tbXE{There was no knife, but the man twisted the rope until it snapped.} \tbPD{mellɛɛ, mellɛ, mellɛ, melbɛ, mellaa, - } 
\tbLX{meloo} \tbHM{} \tbPH{mélóó} \tbPS{adv.~} \tbGE{eating instantly and hurriedly; especially with tasty food} \tbXV{O nɔnoɔ zuoe la, a be᷉e᷉ ka o de e meloo.} \tbXE{S/he has a taste for delicious things, so s/he licked the broth at once.} 
\tbLX{menekana} \tbHM{} \tbPH{mɪ̀nɪ̀kànà} \tbPS{adj.~} \tbGE{crooked, twisted} 
\tbLX{menɛ} \tbHM{} \tbPH{mɪ́nɛ́} \tbPS{n.~} \tbGE{a type of thin grass} \tbTP{menɛɛ} 
\tbLX{menne} \tbHM{} \tbPH{mɪ́nnɪ́} \tbPS{v.~} \tbGE{to throw off something heavy e.g. wrestling} \tbXV{A dɔɔkpeɛŋaa na are la dendeŋ kyɛ menne a o tɔ na bare ka o te le.} \tbXE{The strong man heaved and threw his opponent away.} \tbPD{mennɛɛ, mennɛ, mennɛ, menemɛ, mennaa, - } 
\tbLX{menne} \tbHM{} \tbPH{mɪ́nnɪ́} \tbPS{v.~} \tbGE{to swallow food in large chunks} \tbXV{Kɔŋ da kpɛ o la te zuo ka o maŋ deɛ ŋmaa a saabo menne bare.} \tbXE{The man was so hungry that he swallowed the food in chunks.} \tbPD{mennɛɛ, mennɛ, mennɛ, menemɛ, mennaa, - } 
\tbLX{mennoŋ} \tbHM{} \tbPH{mɪ́nnʊ́ŋ} \tbPS{n.~} \tbGE{righteousness} \tbXV{O mennoŋ zuiŋ noba zaa nɔŋ la o yɛlɛ.} \tbXE{All people like him because of his righteousness.} 
\tbLX{menselɛ} \tbHM{} \tbPH{mɪ́nsɪ́lɛ́} \tbPS{n.~} \tbGE{well done (praise)} \tbXV{Gandaa menselɛ ka fooŋ da ba peɛle a bie kpie.} \tbXE{Well done; if you had not been around, the child would have died.} \tbOP{menselle} \tbTP{mensellɛɛ} \tbVA{nenselɛ}
\tbLX{meŋ} \tbHM{} \tbPH{mɪ̌ŋ} \tbPS{adv.~} \tbGE{also, too} \tbXV{Fo meŋ na gaa la be?} \tbXE{Will you go too?.} 
\tbLX{meŋɛ} \tbHM{} \tbPH{mɪ́ŋɛ́} \tbPS{n.~} \tbGE{self} \tbXV{Maa meŋɛ la na gaa a kuori.} \tbXE{I myself will attend the funeral.} \tbOP{menne} \tbVA{meŋa}
\tbLX{meŋɛ} \tbHM{} \tbPH{mɪ́ŋɛ́} \tbPS{adj.~} \tbGE{well behaved} \tbXV{O teɛre ka o waa la meŋɛ zaa kyɛ maa ba teɛre.} \tbXE{S/he claims to be good, but I doubt it.} 
\tbLX{mere} \tbHM{} \tbPH{mɪ́rɪ́} \tbPS{v.~} \tbGE{to shake in pain or distress e.g. hand in contact with fire} \tbXV{Vu᷉u nyege o la ka o mere nu.} \tbXE{Fire burned her hand and she shook it.} \tbPD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\tbLX{mere} \tbHM{} \tbPH{mɪ́rɪ́} \tbPS{v.~} \tbGE{to broadcast e.g rice seed during sowing} \tbXV{Ba kyɔre la baa mere a mui.} \tbXE{They have harrowed the river valley and broadcast the rice.} \tbPD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\tbLX{mere} \tbHM{} \tbPH{mɪ́rɪ́} \tbPS{v.~} \tbGE{to have a stillbirth (nonhumans)} \tbXV{A naapoɔ na da bɛɛ la lɛnso ka o mere bare.} \tbXE{That pregnant cow was sick and that is why it had a stillbirth.} \tbPD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\tbLX{mere} \tbHM{} \tbPH{mɪ̀rɪ̀} \tbPS{v.~} \tbGE{to stick the flap of an object into an opening} \tbXV{Pɔgeba maŋ mere la ba wagyɛ eŋ kyɛ zeŋ.} \tbXE{Women usually push the flaps of their cloth in between their legs as they sit.} \tbPD{merɛɛ, merɛ, merɛ, merebɛ, meraa, merebɛ } 
\tbLX{mere bare} \tbHM{} \tbPH{mɪ́rɪ́ bàrɪ̀} \tbPS{v.~} \tbSN \tbGE{to have a stillbirth (of animals)} \tbSN \tbGE{to desert unweaned young (of animals)} \tbXV{A peroo dɔgɛɛ la kyɛ mere bare ka a bilii kpi.} \tbXE{The sheep littered but deserted the lambs and they died.} \tbPD{mere barɛɛ, merɛbara, merebara, merebareba, merebaraa, - } 
\tbLX{mere pare} \tbHM{} \tbPH{mɪ̀rɪ̀ párɪ̀} \tbPS{v.~} \tbSN \tbGE{to lie helpless} \tbSN \tbGE{to die} \tbPD{merɛɛ pare, merɛ pare, pare merɛ, pare merebɛ, -, - } 
\tbLX{merete} \tbHM{} \tbPH{mèrètè} \tbPS{adj.~} \tbGE{unresponsive, mute} \tbXV{Ba yeli la a bie baaloŋ ko o saa ka o e merete.} \tbXE{When they reported the child's mother, he refused to do anything.} 
\tbLX{meri} \tbHM{} \tbPH{mérì} \tbPS{v.~} \tbGE{to twist one's lips (to show indifference); to refuse to care} \tbXV{A bie lee la ka leɛ kaa kyɛ meri o noɔre.} \tbXE{When the child fell, the brother looked at him but refused to bother.} \tbPD{meree, mere, mere, meribe, meraa, - } 
\tbLX{meri noɔre} \tbHM{} \tbPH{mérì nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to twist one's lips (to show indifference)} \tbPD{-, -, -, -, -, - } 
\tbLX{meri noɔre} \tbHM{} \tbPH{mérì nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to die} \tbPD{-, -, -, -, -, - } 
\tbLX{meruu} \tbHM{} \tbPH{mèrúú} \tbPS{adj.~} \tbGE{very small} 
\tbLX{mɛ} \tbHM{} \tbPH{mɛ̀} \tbPS{v.~} \tbSN \tbGE{to build e.g. a house} \tbSN \tbGE{to shape (pottery)} \tbPD{mɛɛ, meɛrɛ, meɛrɛ, meɛrebɛ, -, - } 
\tbLX{mɛɛ} \tbHM{} \tbPH{mɛ̀ɛ̀} \tbPS{adv.~} \tbGE{oozing (oily liquid)} \tbXV{Ba kyere la a gbɛre ka ko᷉o᷉ zoro mɛɛ.} \tbXE{They cut the swelling on his leg and pus was oozing.} 
\tbLX{mɛfaa} \tbHM{} \tbPH{mɛ́fáá} \tbPS{n.~} \tbGE{bad character} \tbOP{mɛfaare} \tbTP{mɛfaarɛɛ} 
\tbLX{mɛgelɛŋ} \tbHM{} \tbPH{mɛ́gɪ́lɛ́ŋ} \tbPS{adv.~} \tbGE{flat} \tbXV{Lɔɔre do la a bogiti ka o e ŋmɛgelɛŋ lɛ.} \tbXE{A car ran over the bucket and made it flat.} 
\tbLX{mɛgeyɛgɛ} \tbHM{} \tbPH{mɛ́gɪ́yɛ́gɛ́} \tbPS{adj.~} \tbGE{loose and flabby (of a fat body)} 
\tbLX{mɛle} \tbHM{} \tbPH{mɛ́lɪ̀} \tbPS{n.~} \tbGE{medal} \tbOP{mɛlɛ} \tbTP{mɛllɛɛ} 
\tbLX{mɛlefɛlɛ} \tbHM{} \tbPH{mɛ̀lɪ̀fɛ̀lɛ̀} \tbPS{adj.~} \tbGE{tasteless} \tbXV{N nɔŋ la zɛvaare zeɛre kyɛ ama deɛ waa la mɛlefɛlɛ lɛ.} \tbXE{I like vegetable soup but this one is tasteless.} 
\tbLX{mɛlemɛle} \tbHM{} \tbPH{mɛ̀lɪ́mɛ̀lɪ̀} \tbPS{adj.~} \tbGE{blowing severely with cold} \tbXV{Saseɛ saŋa baguo ɔɔre maŋ erɛ la mɛlemɛle lɛ, Dagao poɔ.} \tbXE{During hamattan, early mornings are normally very cold in the Dagaabaland.} 
\tbLX{mɛlɛmɛlɛ} \tbHM{} \tbPH{mɛ́lɛ́mɛ́lɛ́} \tbPS{adj.~} \tbGE{tasty; slightly sweet} 
\tbLX{mɛlɛmmɛlɛŋ} \tbHM{} \tbPH{mɛ́lɛ́mmɛ́lɛ́ŋ} \tbPS{adj.~} \tbGE{sweet or beautiful (kind words, deeds)} \tbXV{O yeli la a yɛlɛ mɛlɛmmɛlɛŋ lɛ, ka fo na teɛre ka yelimeŋɛ la.} \tbXE{She narrated the case so beautifully that you would think it was true.} 
\tbLX{mɛlɛntɛmm} \tbHM{} \tbPH{mɛ́lɛ́ntɛ́mm} \tbPS{adj.~} \tbSN \tbGE{exactly; precisely} \tbXV{A bie manne la a yɛlɛ na o naŋ nyɛ mɛlɛntɛmm wa gaa.} \tbXE{The boy narrated the story in a precise manner.} \tbSN \tbGE{meticulous} 
\tbLX{mɛmeɛrɛ} \tbHM{} \tbPH{mɛ̀mɪ́ɛ́rɛ́} \tbPS{n.~} \tbSN \tbGE{potter} \tbSN \tbGE{builder} \tbOP{mɛmeɛrebɛ} \tbTP{mɛmeɛrebɛɛ} 
\tbLX{mɛmɛrɛɛ} \tbHM{} \tbPH{mɛ̀mɛ́rɛ́ɛ́} \tbPS{n.~} \tbSN \tbGE{mud} \tbSN \tbGE{marsh} \tbTP{mɛmɛrrɛɛ} 
\tbLX{mɛre} \tbHM{} \tbPH{mɛ́rɪ́} \tbPS{v.~} \tbGE{to cook (struggling with)} \tbXV{A baala mɛre la seemaa fe᷉e᷉ meŋ di.} \tbXE{The patient has cooked some small food to eat.} \tbPD{mɛrɛɛ, mɛrɛ, mɛrɛ, mɛrebɛ, mɛraa, - } 
\tbLX{mɛrɛɛ} \tbHM{} \tbPH{mɛ̀rɛ́ɛ̀} \tbPS{n.~} \tbGE{pus} \tbOP{mɛrre} \tbTP{mɛrrɛɛ} 
\tbLX{mi} \tbHM{} \tbPH{mí} \tbPS{v.~} \tbGE{to rain} \tbPD{mie, miire, -, -, -, - } 
\tbLX{mi} \tbHM{} \tbPH{mí} \tbPS{v.~} \tbGE{to spin (thread)} \tbXV{A zɔŋɔ mi la miri ka o gbɛkpoŋ te zogere.} \tbXE{The blind person spun rope until her thigh became rough.} \tbPD{mie, miire, -, -, -, - } 
\tbLX{mii} \tbHM{} \tbPH{míì} \tbPS{v.~} \tbSN \tbGE{to shake, to shiver} \tbXV{Entole kpɛ la a bie lɛnso ka o maŋ mii a lɛ.} \tbXE{The child has a fever and that is why he is shivering like that.} \tbSN \tbGE{to become frightened, to tremble} \tbXV{A sogya tanne la a pɔge ka o mii.} \tbXE{The soldier shouted at the woman and she became fightened.} \tbSN \tbGE{to push, to shake e.g. the branch of a tree to knock down fruits} \tbXV{A pɔge do mii la a taŋaa ko o bibiiri ka ba di.} \tbXE{The woman climbed and shook the shea nut tree to knock down the fruits for her children.} \tbPD{miie, miine, miine, miinime, miinaa, - } \tbVA{miiu}
\tbLX{mii} \tbHM{} \tbPH{mìì} \tbPS{v.~} \tbSN \tbGE{to be sour (taste)} \tbXV{O za᷉a᷉ tuŋ la nu a zeɛre poɔ lɛnso ka a mii.} \tbXE{S/he put her hand into the soup and that is why it has become sour.} \tbSN \tbGE{to be sore} \tbXV{A bie za᷉a᷉ tɔ o nubiri, zenɛ o zaaŋ mii.} \tbXE{The child bruised his finger yesterday and today it is quite sore.} \tbSN \tbGE{to be angry, to be annoyed} \tbXV{Ba to o la lɛnso ka o mii.} \tbXE{They insulted him and that is why he is angry.} \tbSN \tbGE{to learn from past unpleasant experience} \tbXV{A tontona naŋ zagere toma gaabo na ba ba yɔɔ o, azuiŋ o miie la; zenɛ o ba nagera toma.} \tbXE{Since the worker refused to go to work, he was not paid, and as a result, he did not skip work today.} \tbPD{miie, miine, miine, miinime, miinaa, miinime } 
\tbLX{miii} \tbHM{} \tbPH{mììì} \tbPS{adv.~} \tbGE{swarming all over e.g. ants} \tbXV{A nyere deɛ da te paalɛɛ a korodie zaa, miloo lɛ.} \tbXE{The ants had invaded and covered the whole kitchen.} 
\tbLX{miire} \tbHM{1} \tbPH{mììré} \tbPS{v.~} \tbGE{to desist, to not dare} \tbXV{Ma fo to ma la, aŋ miire n yiri waabo.} \tbXE{Since you have insulted me, you dare not come to my house.} \tbPD{-, -, -, -, -, - } 
\tbLX{miire} \tbHM{2} \tbPH{mììré} \tbPS{v.~} \tbGE{to be careful; usually to warn or caution} \tbPD{miiree, miire, miire, miiribe, miiraa, - } 
\tbLX{miiri} \tbHM{} \tbPH{mììrì} \tbPS{v.~} \tbGE{to not attempt} \tbPD{miiree, miire, miire, miiribe, miiraa, - } 
\tbLX{miiri} \tbHM{} \tbPH{mììrì} \tbPS{v.~} \tbGE{to turn taste sour} \tbPD{miiree, miire, miire, miiribe, miiraa, - } 
\tbLX{miiruŋ} \tbHM{} \tbPH{mììrùŋ} \tbPS{n.~} \tbSN \tbGE{sour sap} \tbSN \tbGE{type of plant with sour sap} 
\tbLX{miiruŋ} \tbHM{} \tbPH{mììrùŋ} \tbPS{adj.~} \tbGE{sour} 
\tbLX{mikalaa} \tbHM{} \tbPH{mǐkàláà} \tbPS{n.~} \tbGE{thick, strong rope} \tbOP{mikalle} \tbTP{mikallɛɛ} 
\tbLX{milemmilem} \tbHM{} \tbPH{mílémílém} \tbPS{adv.~} \tbGE{intact, untouched} 
\tbLX{miliki} \tbHM{} \tbPH{mílíkì} \tbPS{n.~} \tbGE{milk} \tbSG{milikiruu} \tbOP{milikiri} \tbTP{milikiree} 
\tbLX{milikolo} \tbHM{} \tbPH{mìlìkòlò} \tbPS{adj.~} \tbGE{twisted} 
\tbLX{milli} \tbHM{} \tbPH{mìllì} \tbPS{v.~} \tbSN \tbGE{to get acquainted with; to be used to} \tbXV{Bie naŋ ba milli neɛ koŋ sage ka a soba de o.} \tbXE{A child that is not acquainted with someone will refuse to be cuddled by that person.} \tbSN \tbGE{all over} \tbPD{millee, mille, mille, milibe, millaa, - } \tbVA{milli pɔge}
\tbLX{milli pɔge} \tbHM{} \tbPH{mìllì-pɔ̀gɪ̀} \tbPS{v.~} \tbGE{to cover all over e.g. flies over a rotten carcass} \tbXV{Naazoore milli pɔge la a nɛmpoɔnaa zaa.} \tbXE{The rotten meat is completely covered with flies.} \tbPD{-, -, -, -, -, - } 
\tbLX{miloo} \tbHM{} \tbPH{mìlòò} \tbPS{adj.~} \tbGE{oily and dirty looking} 
\tbLX{miloo} \tbHM{} \tbPH{mìlòò} \tbPS{adj.~} \tbGE{existing in great numbers at a particular spot e.g. ants} 
\tbLX{mine} \tbHM{} \tbPH{mìné} \tbPS{adj.~} \tbSN \tbGE{some} \tbXV{A boore mine zoe la.} \tbXE{Some of the goats have escaped.} \tbSN \tbGE{pl. morpheme ( human type only)} \tbXV{Mamine la maŋ ko  biiri bere ka ba ɛnne.} \tbXE{Mothers provide their breasts to babies to suck.} 
\tbLX{minimini} \tbHM{} \tbPH{mìnìmìnì} \tbPS{adv.~} \tbGE{streaming continuously e.g. tears} \tbXV{A bie poɔ beɛrɛ la ka o kono minimini lɛ.} \tbXE{The child has a stomach ache and he is sheding tears uncontrollably.} 
\tbLX{mipelaa} \tbHM{} \tbPH{mípɪ́láá} \tbPS{n.~} \tbGE{a type of rope used in rituals} \tbOP{mipeɛle} 
\tbLX{miri} \tbHM{} \tbPH{mírì} \tbPS{n.~} \tbGE{rope, twine, string} \tbOP{mie} \tbTP{miiree} 
\tbLX{misiri} \tbHM{} \tbPH{mìsírì} \tbPS{n.~} \tbGE{mosque} \tbOP{misirri} \tbTP{misirree} \tbVA{musiri}
\tbLX{miti} \tbHM{} \tbPH{mítì} \tbPS{n.~} \tbGE{meeting} \tbXV{mitiri} \tbXE{mitiree.} \tbVA{English}
\tbLX{mizeɛ} \tbHM{} \tbPH{mízɪ̀ɛ́} \tbPS{n.~} \tbGE{type of very strong red twine used in making bird trap} \tbOP{mizeere} \tbTP{mizeerɛɛ} 
\tbLX{mạạli} \tbHM{} \tbPH{mạ́ạ́lì} \tbPS{v.~} \tbGE{get burnt e.g part of the body} \tbPD{-, -, -, -, -, - } 
\tbLX{mmeŋɛ} \tbHM{} \tbPH{mmɪ́ŋɛ́} \tbPS{pronoun.~} \tbGE{myself} \tbXV{Ka maaŋ beɛrɛ kyɛ ba gaa asibiti mmeŋɛ la ka N erɛ.} \tbXE{If I'm sick and refuse to go the hospital, I have only myself to blame.} \tbVA{mmeŋa}
\tbLX{moge} \tbHM{} \tbPH{mʊ́gɪ̀} \tbPS{v.~} \tbSN \tbGE{to suck e.g. sweets} \tbSN \tbGE{to eat powdery food} \tbSN \tbGE{to eat up e.g. wood, of white ants} \tbPD{mogɛɛ, mogrɔ/mogra, mogerɔ/mogra, mogreba, mograa, - } 
\tbLX{mogere} \tbHM{} \tbPH{mʊ̀gɪ̀rɪ̀} \tbPS{v.~} \tbGE{to suck at, to lick e.g. fingers} \tbPD{mogerɛɛ, mogera, mogera, mogereba, mogeraa, mogereba } \tbVA{moore}
\tbLX{mogere noɔre} \tbHM{} \tbPH{mʊ̀gɪ̀rɪ̀ nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to kiss (on the mouth)} \tbPD{-, -, -, -, -, - } 
\tbLX{mogi} \tbHM{} \tbPH{mógí} \tbPS{v.~} \tbGE{to eat delicious food voraciously} \tbPD{mogee, mogiro, mogiro, mogiribo, mogiraa, - } \tbVA{yogi}
\tbLX{moimoi} \tbHM{} \tbPH{móímóí} \tbPS{adj.~} \tbGE{very red} 
\tbLX{mokanzie} \tbHM{} \tbPH{mókánzíé} \tbPS{n.~} \tbGE{pawpaw} \tbOP{mokanziri} \tbTP{mokanziree} 
\tbLX{mole} \tbHM{} \tbPH{mʊ̀lɪ̀} \tbPS{n.~} \tbGE{become weak; used for metal that has become weak e.g. iron} 
\tbLX{mole} \tbHM{} \tbPH{mʊ̀lɪ̀} \tbPS{n.~} \tbGE{get wet} 
\tbLX{mollaa} \tbHM{} \tbPH{mʊ́lláà} \tbPS{n.~} \tbGE{a type of game in which a loop of string is hidden in soil to be found by other players using twigs} \tbOP{molle} \tbTP{mollɛɛ} 
\tbLX{molle} \tbHM{} \tbPH{mʊ́llɪ̀} \tbPS{v.~} \tbSN \tbGE{to deceive, to hide from} \tbSN \tbGE{to play the game mollaa} \tbPD{mollɛɛ, mollɔ/molla, mollɔ/molla, molbɔ/molba, mollaa, - } 
\tbLX{molɔmolɔ} \tbHM{} \tbPH{mʊ́lɔ́mʊ́lɔ́} \tbPS{adj.~} \tbGE{not stiff; flexible} \tbXV{A tammo waa la molɔmolɔ lɛ.} \tbXE{The bow is very flexible.} 
\tbLX{momm} \tbHM{} \tbPH{mómm} \tbPS{adv.~} \tbGE{wholly and completely} 
\tbLX{momone} \tbHM{} \tbPH{mʊ̀mùnɪ́} \tbPS{n.~} \tbGE{dried salted fish} \tbTP{momonɛɛ} 
\tbLX{monaŋe} \tbHM{} \tbPH{mʊ̀náŋ} \tbPS{n.~} \tbGE{hippopotamus} \tbOP{monanne} \tbTP{monannɛɛ} 
\tbLX{mone} \tbHM{} \tbPH{mʊ̀nɪ́} \tbPS{n.~} \tbGE{serious condition of piles} \tbOP{momɔ} \tbTP{monɛɛ/monnɛɛ} 
\tbLX{mone} \tbHM{} \tbPH{mʊ̀nɪ̀} \tbPS{v.~} \tbGE{to become serious} \tbPD{monɛɛ, mona, mona, monema, monaa, - } 
\tbLX{monɛɛ} \tbHM{} \tbPH{mʊ̀nɛ̀ɛ̀} \tbPS{adj.~} \tbGE{red} 
\tbLX{monɛɛmonɛɛ} \tbHM{} \tbPH{mʊ́nɛ́ɛ́mʊ́nɛ́ɛ́} \tbPS{adj.~} \tbGE{red looking; bloodshot (eyes)} 
\tbLX{monne} \tbHM{} \tbPH{mʊ́nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to lift and carry a heavy load without assisstance} \tbSN \tbGE{to dip to fill} \tbSN \tbGE{to gather up (smile to laugh)} \tbPD{monnɛɛ, monna, monna, monema, monnaa, - } 
\tbLX{montole} \tbHM{} \tbPH{mʊ̀ntʊ̀lɪ̀} \tbPS{n.~} \tbGE{a type of small round beans} \tbOP{montola} \tbTP{montollɛɛ} 
\tbLX{moŋgolɔ} \tbHM{} \tbPH{mʊ̀ŋgʊ̀lɔ̀} \tbPS{n.~} \tbGE{a type of fall in which the person rolls over} \tbXV{A da᷉a bugro lee la willi moŋgolɔ.} \tbXE{The drunk man fell and rolled over.} \tbVA{moŋgola}
\tbLX{moŋgbɛnɛɛ} \tbHM{} \tbPH{mʊ́ŋgbɛ́nɛ́ɛ́} \tbPS{n.~} \tbGE{infected millet seeds} \tbSG{moŋgbɛnnoo} 
\tbLX{moole} \tbHM{} \tbPH{mʊ̀ʊ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to be stunted, to be weak} \tbSN \tbGE{to cook in small quantity} \tbSN \tbGE{to cook poorly e.g. corn paste} \tbPD{moolɛɛ, moolɔ, moolɔ, mooleba, -, - } 
\tbLX{mooloo} \tbHM{} \tbPH{mʊ́ʊ́lʊ́ʊ́} \tbPS{adj.~} \tbGE{weak, stunted} 
\tbLX{moore} \tbHM{} \tbPH{mʊ́ʊ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to empathise; to sympathise with} \tbXV{Koŋkonimo maŋ kono moorɔ korɔ la kotuo deme.} \tbXE{Funeral dirge singers usually empathise with the bereaved family.} \tbSN \tbGE{to express regret for not being able to provide something} \tbXV{A da᷉a᷉pɔge moore la a da᷉a᷉.} \tbXE{The pito seller regrets that the pito is gone.} \tbPD{moorɛɛ, moorɔ, moorɔ, mooreba, mooraa, - } 
\tbLX{moore} \tbHM{} \tbPH{mʊ́ʊ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to dissolve, to diminish} \tbXV{A zeɛre nyɛnnoŋ ba moore.} \tbXE{The salt in the soup is not dissolved.} \tbSN \tbGE{to rain lightly} \tbXV{A saa ba mi soŋ mooroo ka o moore.} \tbXE{It did not rain hard; it just drizzled.} \tbSN \tbGE{to intercede, to suggest} \tbXV{N daare moore la o yiri kuluu ka o nyare.} \tbXE{When I suggested his going home, he flared up.} \tbPD{moorɛɛ, moorɔ, moorɔ, mooreba, mooraa, - } 
\tbLX{moore} \tbHM{} \tbPH{mʊ̀ʊ̀rɪ} \tbPS{v.~} \tbGE{to suck e.g. finger} \tbPD{moorɛɛ, moora, moora, mooreba, mooraa, - } \tbVA{mogre}
\tbLX{moore tere} \tbHM{} \tbPH{mʊ́ʊ́rɪ́ tɪ̀rɪ̀} \tbPS{v.~} \tbGE{to sympathise with} \tbPD{-, -, -, -, -, - } 
\tbLX{mooreterɛ} \tbHM{} \tbPH{mʊ́ʊ́rtɪ́rɛ́} \tbPS{n.~} \tbGE{consoler} \tbOP{moorterebɛ} \tbTP{mooreterebɛɛ} 
\tbLX{moorteroo} \tbHM{} \tbPH{mʊ́ʊ́rtɪ́rʊ́ʊ́} \tbPS{n.~} \tbGE{the act of consoling} 
\tbLX{moɔ} \tbHM{} \tbPH{mʊ̀ɔ̀} \tbPS{v.~} \tbSN \tbGE{to make red} \tbSN \tbGE{to be ripe} \tbSN \tbGE{to be serious} \tbXV{Ka fooŋ boɔrɔ ka fo paase gyennoo fo maŋ moɔ nimiri ne zannoo.} \tbXE{If you want to pass a test, you need to be serious about your studies.} \tbSN \tbGE{to fetch fire (red coals)} \tbPD{moɔɛ, moɔnɔ, moɔnɔ, moɔnema, moɔnaa, moɔnema } 
\tbLX{moɔ} \tbHM{} \tbPH{mʊ̀ɔ̀} \tbPS{adv.~} \tbGE{really} \tbXV{yeŋ moɔ ka a bie yi?} \tbXE{Really, where does the child come from?.} \tbVA{mɔɔ}
\tbLX{moɔ} \tbHM{} \tbPH{mʊ́ɔ́} \tbPS{n.~} \tbSN \tbGE{grass, weeds} \tbSN \tbGE{bush, wilderness} \tbSG{moɔnoo} \tbTP{moɔnɛɛ} 
\tbLX{moɔle} \tbHM{} \tbPH{mʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to stretch (hair)} \tbPD{moɔlɛɛ, moɔlɔ, moɔlɔ, moɔleba, moɔlaa, moɔleba } 
\tbLX{moɔle} \tbHM{} \tbPH{mʊ́ɔ́lɪ́} \tbPS{v.~} \tbGE{to announce} \tbXV{Ba moɔle ka neɛzaa ta wa gɛrɛ tomɔ bie.} \tbXE{It has been announced that no one should go to work tomorrow.} \tbPD{moɔlɛɛ, moɔlɔ, moɔlɔ, moɔleba, moɔlaa, moɔleba } 
\tbLX{moɔle} \tbHM{} \tbPH{mʊ́ɔ́lɪ́} \tbPS{v.~} \tbGE{to vanish, to sneak} \tbXV{Kɔkɔlaa wa la a yiri ka gyaŋgbonni zaa moɔle.} \tbXE{A cat has come to the house and all the mice have vanished.} \tbPD{moɔlɛɛ, moɔlɔ, moɔlɔ, moɔleba, moɔlaa, moɔleba } 
\tbLX{moɔlɔ} \tbHM{} \tbPH{mʊ́ɔ́lɔ̀} \tbPS{n.~} \tbGE{established behaviour, character; (past)} \tbXV{A bie de la o saa moɔlɔ.} \tbXE{The child has taken after his father.} \tbTP{moɔlɛɛ} \tbVA{moɔla}
\tbLX{moɔraa} \tbHM{} \tbPH{mʊ́ɔ́ráà} \tbPS{adv.~} \tbGE{like, similar manner} \tbVA{moɔreŋ}
\tbLX{moɔre} \tbHM{} \tbPH{mʊ́ɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to smear with thick liquid e.g. mud} \tbSN \tbGE{to camouflage} \tbXV{O ku᷉u deɛ moɔre ma la.} \tbXE{He camouflaged us before his death.} \tbPD{moɔrɛɛ, moɔrɔ, moɔrɔ, moɔreba, moɔraa, - } 
\tbLX{moɔreŋ} \tbHM{} \tbPH{mʊ́ɔ́rɪ́ŋ} \tbPS{adj.~} \tbGE{similar to how the Mosi people do it} \tbVA{komoɔreŋ}
\tbLX{moɔreŋ} \tbHM{} \tbPH{mʊ́ɔ́rɪ́ŋ} \tbPS{pp.~} \tbGE{like, according to} \tbXV{ŋmaaŋa moɔreŋ, dabe᷉ɛ sobaŋ taa yiri.} \tbXE{According to the monkey, he who runs away lives to fight another day.} \tbVA{moɔraa}
\tbLX{moɔrɔ} \tbHM{} \tbPH{mʊ́ɔ́rɔ́} \tbPS{n.~} \tbGE{illness of sexual misconduct; adultery} \tbTP{moɔrɛɛ} 
\tbLX{more} \tbHM{} \tbPH{mʊ̀rɪ̀} \tbPS{v.~} \tbGE{to disappear, to vanish} \tbPD{morɛɛ, morɔ, morɔ, morba, moraa, morba } 
\tbLX{more} \tbHM{} \tbPH{mʊ̀rɪ̀} \tbPS{v.~} \tbGE{to pop out e.g. seeds from a fruit in the mouth} \tbPD{morɛɛ, morɔ, morɔ, morba, moraa, morba } 
\tbLX{more} \tbHM{} \tbPH{mʊ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to smile} \tbSN \tbGE{to initiate (typically used with laare)} \tbPD{morɛɛ, morɔ, morɔ, moreba, moraa, moreba } 
\tbLX{moremore} \tbHM{} \tbPH{mʊ̀rɪ̀mʊ́rɪ́} \tbPS{v.~} \tbGE{to vanish at once} \tbPD{-, -, -, -, -, - } 
\tbLX{moromoro} \tbHM{} \tbPH{mórómóró} \tbPS{adj.~} \tbGE{soft, not solid} 
\tbLX{moroo} \tbHM{} \tbPH{mʊ́rʊ́ʊ́} \tbPS{n.~} \tbGE{local herbal medicine in tablet form used in massaging only} \tbOP{morre} \tbTP{morɛɛ/morrɛɛ} \tbVA{morɔ}
\tbLX{morɔ} \tbHM{} \tbPH{mʊ́rɔ́} \tbPS{n.~} \tbGE{herb tablet for external use} \tbOP{morre} \tbTP{morɛɛ} \tbVA{mora}
\tbLX{morɔmorɔ} \tbHM{} \tbPH{mʊ́rɔ́mʊ́rɔ́} \tbPS{adv.~} \tbSN \tbGE{showing slightly (smile)} \tbXV{O nyɛ la wagyɛ paalaa a monne laare morɔmorɔ lɛ.} \tbXE{She has got a new coth and this has made her beam with smiles.} \tbSN \tbGE{get lost mysteriously} \tbXV{A libie deɛ bɔrɛɛ la morɔmorɔ lɛ.} \tbXE{The money just got lost mysteriously.} 
\tbLX{mɔ} \tbHM{} \tbPH{mɔ́} \tbPS{v.~} \tbSN \tbGE{to wrestle} \tbXV{A pɔgebɔ da zɔɔ la taa, a mɔ te le.} \tbXE{The women fought, wrestled, and fell.} \tbSN \tbGE{to try hard} \tbXV{Ka fooŋ ba mɔ fo koŋ to᷉ɔ᷉ a teɛ do.} \tbXE{If you don't try hard, you can't climb the tree.} \tbPD{mɔɛ, moɔrɔ, moɔrɔ, moɔreba, moɔraa, - } 
\tbLX{mɔ-aŋgaara} \tbHM{} \tbPH{mɔ̀-áŋgáárá} \tbPS{n.~} \tbGE{half-burnt grass} \tbOP{mɔ-aŋgaare} \tbTP{mɔ-aŋgaarɛɛ} 
\tbLX{mɔdoŋɔ} \tbHM{} \tbPH{mɔ̀dʊ́ŋɔ́} \tbPS{n.~} \tbGE{wild animal} \tbOP{mɔdonne} \tbTP{mɔdonnɛɛ} \tbVA{mɔdoŋa, wɛdoŋɔ}
\tbLX{mɔgaabo} \tbHM{} \tbPH{mɔ̀gáábʊ̀} \tbPS{n.~} \tbGE{game hunting expedition from a pitched tent over a length of time} 
\tbLX{mɔge} \tbHM{} \tbPH{mɔ́gɪ́} \tbPS{v.~} \tbGE{to get dirty} \tbXV{Ka fooŋ kpɛ nɔzage fo nɔɔteɛ maŋ mɔge la nɔbini.} \tbXE{If you go into a hen coop, your shoes will get dirty with fowl droppings.} \tbPD{mɔgɛɛ, mɔgrɔ, mɔgrɔ, mɔgreba, -, - } 
\tbLX{mɔgele} \tbHM{} \tbPH{mɔ̀gɪ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to beseech, to supplicate} \tbSN \tbGE{to persuade} \tbSN \tbGE{to console, to comfort e.g. one in pain or distress} \tbSN \tbGE{to beg for forgiveness} \tbPD{mɔgelɛɛ, mɔgelɔ, mɔgelɔ, mɔgeleba, mɔgelaa, - } 
\tbLX{mɔgeraa} \tbHM{} \tbPH{mɔ̀gɪ̀ráá} \tbPS{n.~} \tbGE{a type of giant white ants} \tbOP{mɔgere} \tbTP{mɔgerɛɛ} 
\tbLX{mɔkangyi} \tbHM{} \tbPH{mɔ́kándzí} \tbPS{n.~} \tbSN \tbGE{a type of tree with medicial properties and edible fruits} \tbSN \tbGE{pawpaw} \tbOP{mɔkangyiri} \tbTP{mɔkangyiree} 
\tbLX{mɔkpeɛrɛ} \tbHM{} \tbPH{mɔ̀kpɪ́ɛ́rɛ́} \tbPS{n.~} \tbSN \tbGE{bush dweller} \tbSN \tbGE{one who lives in a faraway country} \tbOP{mɔkpeɛrebɛ} \tbTP{mɔkpeɛrebɛɛ} 
\tbLX{mɔll} \tbHM{} \tbPH{mɔ̀ll} \tbPS{adv.~} \tbGE{long and smooth} 
\tbLX{mɔloo} \tbHM{} \tbPH{mɔ̀lʊ́ʊ́} \tbPS{n.~} \tbGE{a type of large antelope} \tbOP{mɔlle} \tbTP{mɔlɛɛ} 
\tbLX{mɔlɔ} \tbHM{} \tbPH{mɔ̀lɔ́} \tbPS{adj.~} \tbSN \tbGE{cheap} \tbSN \tbGE{easy} \tbVA{mɔlɛɛ}
\tbLX{mɔlɔmɔlɔ} \tbHM{} \tbPH{mɔ́lɔ́mɔ́lɔ́} \tbPS{adj.~} \tbGE{seemingly easy, like a joke} 
\tbLX{mɔmaara} \tbHM{} \tbPH{mɔ̀màárà} \tbPS{n.~} \tbGE{fresh grass} \tbTP{mɔmaarɛɛ} 
\tbLX{mɔnaabo} \tbHM{} \tbPH{mɔ̀náábʊ́} \tbPS{n.~} \tbGE{bush cow} \tbOP{mɔnii} \tbTP{mɔniinee} \tbVA{wɛnaabo}
\tbLX{mɔneɛ} \tbHM{} \tbPH{mɔ̀nɪ́ɛ́} \tbPS{n.~} \tbGE{bushman, uncivilised person} \tbOP{mɔnoba} 
\tbLX{mɔnyuu} \tbHM{} \tbPH{mɔ̀ɲúú} \tbPS{n.~} \tbGE{a type of grass with a strong smell and has medicinal use} \tbOP{mɔnyuuri} \tbTP{mɔnyuuree} 
\tbLX{mɔŋe} \tbHM{} \tbPH{mɔ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to stir (food)} \tbSN \tbGE{to steer, to drive (car)} \tbSN \tbGE{to paddle (boat)} \tbPD{mɔŋɛɛ, mɔnɔ, mɔnɔ, mɔnemɔ, mɔnaa, mɔnemɔ } 
\tbLX{mɔŋe} \tbHM{} \tbPH{mɔ́ŋ} \tbPS{v.~} \tbGE{to refuse to give} \tbXV{A entuo daana sɔre la a bondaana libiri ka o mɔŋ o.} \tbXE{The lazy man begged for money from the rich man, but he refused him.} \tbPD{mɔŋɛɛ, mɔnɔ, mɔnnɔ, mɔnemɔ, mɔnaa, mɔnemɔ } 
\tbLX{mɔpelaa} \tbHM{} \tbPH{mɔ̀pɪ́láà} \tbPS{n.~} \tbGE{a stretch of bush made mainly of dry tall grass} \tbOP{mɔpeɛle} \tbTP{mɔpeɛlɛɛ} 
\tbLX{mɔre} \tbHM{} \tbPH{mɔ́rɪ́} \tbPS{n.~} \tbGE{a type of strong grass used for arrow shafts} \tbOP{mɔɛ} \tbTP{mɔrɛɛ} 
\tbLX{mɔre} \tbHM{} \tbPH{mɔ́rɪ́} \tbPS{v.~} \tbGE{to swell} \tbPD{mɔrɛɛ, mɔrɔ, mɔrɔ, mɔreba, mɔraa, mɔreba } 
\tbLX{mɔrɔmɔrɔ} \tbHM{} \tbPH{mɔ̀rɔ̀mɔ̀rɔ̀} \tbPS{adv.~} \tbGE{marshy, soft, and loose, so that it can't be held together} 
\tbLX{mɔrɔtɔ} \tbHM{} \tbPH{mɔrɔtɔ} \tbPS{adj.~} \tbGE{long and flat} 
\tbLX{mɔsogɔ} \tbHM{} \tbPH{mɔ̀sʊ́gɔ́} \tbPS{n.~} \tbGE{bush, the middle of the wilderness} \tbOP{mɔsogre} \tbTP{mɔsogrɛɛ} 
\tbLX{mɔsɔglaa} \tbHM{} \tbPH{mɔ̀sɔ́glàá} \tbPS{n.~} \tbGE{a type of grass whose presence indicates fertile land} \tbOP{mɔsɔglɔ} 
\tbLX{mɔtɔ} \tbHM{} \tbPH{mɔ́tɔ́} \tbPS{adv.~} \tbGE{very full} \tbXV{A boɔraa e la mɔtɔ o naŋ paale.} \tbXE{The sack is completely full.} 
\tbLX{mɔvɔ} \tbHM{} \tbPH{mɔ̀vɔ̀} \tbPS{n.~} \tbGE{part of the rainy season when there is thick growth of grass} \tbTP{mɔvɔrɛɛ} 
\tbLX{mɔwegi} \tbHM{} \tbPH{mɔ̀wégí} \tbPS{n.~} \tbGE{wild bush} \tbOP{mɔwegiri} \tbTP{mɔwegiree} 
\tbLX{mui} \tbHM{} \tbPH{mùí} \tbPS{n.~} \tbGE{rice} \tbTP{muunee} 
\tbLX{muli} \tbHM{} \tbPH{múlí} \tbPS{v.~} \tbGE{to remove hair or feathers from skin (usually by applying hot water)} \tbPD{mulee, mulo, mulo, mulibo, mulaa, - } 
\tbLX{mulli} \tbHM{} \tbPH{mùllì} \tbPS{v.~} \tbGE{to dive, to go underwater} \tbPD{mullee, mullo, mullo, mullibo, mullaa, mullibo } 
\tbLX{muni} \tbHM{} \tbPH{múní} \tbPS{n.~} \tbSN \tbGE{meaning} \tbXV{A dɔɔ ba baŋe a yelbiri ŋa muni.} \tbXE{the man does not know the meaning of this word.} \tbSN \tbGE{reason} \tbXV{A dɔɔ ba wuli a o polisiri ŋmeɛbo muni} \tbXE{The man does not know the reason why he was beaten by the police.} 
\tbLX{muni} \tbHM{} \tbPH{múní} \tbPS{adv.~} \tbSN \tbGE{bottom} \tbXV{A bie nu koŋ ta a doge muni.} \tbXE{The child’s hand cannot get to the bottom of the pot.} \tbSN \tbGE{under} \tbXV{A saama zeŋe la a maŋgo teɛ muni.} \tbXE{The strangers are sitting under the mango tree.} 
\tbLX{munii} \tbHM{} \tbPH{múníí} \tbPS{adj.~} \tbGE{tiny, small} \tbXV{A bɛmmo ŋa biri maŋ waa la munii lɛ.} \tbXE{This variety of bean has a tiny seed.} \tbOP{muno} \tbTP{munnee} 
\tbLX{munkyara} \tbHM{} \tbPH{mùnʧárá} \tbPS{n.~} \tbGE{a type of smock with an opening at the back} \tbOP{munkyarre} \tbTP{munkyarɛɛ} 
\tbLX{muŋkuri} \tbHM{} \tbPH{mùŋkúrí} \tbPS{n.~} \tbGE{skirt} \tbOP{muŋkurri} \tbTP{muŋkuree} 
\tbLX{mupelaa} \tbHM{} \tbPH{mùpílàá} \tbPS{n.~} \tbGE{a type of straw from tall grass used in lighting fires} \tbOP{mupeɛle} \tbTP{mupeɛlɛɛ} 
\tbLX{mupɛge} \tbHM{} \tbPH{múpɛ́gɪ́} \tbPS{n.~} \tbGE{grass husk} \tbOP{mupɛgre} \tbTP{mupɛgrɛɛ} 
\tbLX{muri} \tbHM{} \tbPH{múrí} \tbPS{n.~} \tbSN \tbGE{straw from a type of tall grass used in making mats} \tbSN \tbGE{a bundle of such straw} \tbOP{mie} \tbTP{miiree} 
\tbLX{muri} \tbHM{} \tbPH{múrí} \tbPS{n.~} \tbGE{snake} \tbOP{mie} \tbTP{miiree} 
\tbLX{muri} \tbHM{} \tbPH{mùrì} \tbPS{v.~} \tbSN \tbGE{to dodge, to escape} \tbSN \tbGE{to slide out e.g. rope around one's neck} \tbSN \tbGE{to deny an earlier statement} \tbPD{muree, muro, muro, muribu, muraa, muribo } 
\tbLX{muro} \tbHM{} \tbPH{mùró} \tbPS{n.~} \tbGE{the posture of lying facedown} \tbXV{Ba mɔɛ la te le vɔgele muro.} \tbXE{They wrestled and fell facedown.} 
\tbLX{muu} \tbHM{} \tbPH{múú} \tbPS{n.~} \tbGE{deaf and mute person} \tbOP{muuni} \tbTP{muunee} 
\tbLX{muu} \tbHM{} \tbPH{múú} \tbPS{adj.~} \tbGE{deaf and mute} 
\tbLX{muuli} \tbHM{} \tbPH{múúlí} \tbPS{v.~} \tbGE{to peep into or through} \tbPD{muulee, muulo, muulo, muulibo, muulaa, - } 
\tbLX{muuli} \tbHM{} \tbPH{mùùlì} \tbPS{v.~} \tbGE{to make a short visit (typically used with nyɛ)} \tbXV{N baŋ ka fo pagra la kyɛ fo naŋ muuli a baala kyɛ gɛrɛ.} \tbXE{I know you are in a hurry, but just pop in to visit the patient and then go.} \tbPD{muulee, muulo, muulo, muulibo, muulaa, - } 
\tbLX{muuu} \tbHM{} \tbPH{múúù} \tbPS{n.~} \tbSN \tbGE{onomatopoeia; the characteristic sound made by a cow} \tbSN \tbGE{cow} 
\tbLX{muu-u} \tbHM{} \tbPH{mùú-ú̀} \tbPS{n.~} \tbGE{moo (sound made by cows)} \tbOP{muuni} \tbTP{muunee} 
\tbLX{N} \tbHM{} \tbPH{ɪ̀n} \tbPS{pronoun.~} \tbGE{I (first person singular nominative)} \tbXV{N na gaa la daa.} \tbXE{I will go to the market.} \tbSN \tbGE{my (first person singular possessive)} \tbXV{N libie baarɛɛ la.} \tbXE{My money is gone.} 
\tbLX{na} \tbHM{} \tbPH{ná} \tbPS{pronoun.~} \tbGE{that (third person demonstrative)} \tbXV{A bi ŋmeene na beɛrɛ la.} \tbXE{That lean child is sick.} 
\tbLX{na} \tbHM{} \tbPH{nà} \tbPS{particle.~} \tbGE{will, shall (future tense marker)} \tbXV{A baala na gaa la asibiti.} \tbXE{The patient will go to the hospital.} 
\tbLX{Naa} \tbHM{} \tbPH{nàá} \tbPS{n.~} \tbGE{name traditionally given to a second male twin} \tbOP{Namine} 
\tbLX{naa} \tbHM{} \tbPH{nàá} \tbPS{n.~} \tbSN \tbGE{chief, king} \tbSN \tbGE{grandparent} \tbSN \tbGE{rich person} \tbOP{namine} \tbTP{narɛɛ} 
\tbLX{naa} \tbHM{} \tbPH{nàá} \tbPS{modal verb.~} \tbSN \tbGE{will/shall (emphatic 'na')} \tbXV{Te naa wa la.} \tbXE{We shall come.} \tbSN \tbGE{should have} \tbXV{Fo da naa taa la a bie gaa ne asibiti.} \tbXE{You should have taken the child to the hospital.} \tbSN \tbGE{would have} \tbXV{A boɔso ba wa wiẽu ka lɛ naane fo da naa wa waana ko o gaaɛ.} \tbXE{The bus came late, otherwise by the time you came it would have left.} 
\tbLX{naa} \tbHM{} \tbPH{náá} \tbPS{adv.~} \tbGE{not} \tbXV{Onɔ naa ka N yele.} \tbXE{I'm not referring to that one.} \tbVA{shortened form of naane}
\tbLX{naabaa} \tbHM{} \tbPH{nààbáá} \tbPS{n.~} \tbGE{a type of small black bird regarded a totem of a clan (Zagɛɛ)} \tbOP{naabaare} \tbTP{naabaarɛɛ} 
\tbLX{naabaaraa gane} \tbHM{} \tbPH{náábááráá gánɪ̀} \tbPS{n.~} \tbGE{bona fide property; status not changeable; that which naturally belongs to a specific set} \tbXV{Ka Dagabieŋ baa o areyiriŋ kyɛ o naŋ waa la naabaaraa gane.} \tbXE{Among the Dagaaba, even if a child is brought up in his mother's home, he still remains the kin of the father.} 
\tbLX{naabere} \tbHM{} \tbPH{náábɪ̀rɪ́} \tbPS{n.~} \tbGE{udder} \tbOP{naaberɛ} \tbTP{naaberɛɛ} 
\tbLX{naabileŋmene} \tbHM{} \tbPH{náábíléŋmɪ́nɪ̀} \tbPS{n.~} \tbGE{God, supreme being} \tbOP{naabileŋmemɛ} 
\tbLX{naabilfaŋ} \tbHM{} \tbPH{náábìlfáŋ} \tbPS{n.~} \tbGE{a large calf} \tbOP{naabilfanne} \tbTP{naabilfannɛɛ} 
\tbLX{naabil-paaŋaa} \tbHM{} \tbPH{náábìl-pááŋáá} \tbPS{n.~} \tbGE{calf, baby cow} \tbOP{naabil-paane} \tbTP{naabil-paanɛɛ} 
\tbLX{naabini} \tbHM{} \tbPH{náábíní} \tbPS{n.~} \tbGE{cow dung} \tbSG{naabinuu} \tbTP{naabinee} 
\tbLX{naabo} \tbHM{} \tbPH{náábʊ́} \tbPS{n.~} \tbGE{cow} \tbOP{nii} \tbTP{niinee} 
\tbLX{naada} \tbHM{} \tbPH{náádà} \tbPS{n.~} \tbGE{the act of buying cattle} 
\tbLX{naadaara} \tbHM{} \tbPH{náádáárá} \tbPS{n.~} \tbGE{one who buys cattle} \tbOP{naadareba} \tbTP{naadarebɛɛ} 
\tbLX{naadaaraa} \tbHM{} \tbPH{náádàáráà} \tbPS{n.~} \tbGE{a cow that has been bought} \tbOP{naadaare} 
\tbLX{naadalane} \tbHM{} \tbPH{náádalánɪ́̀} \tbPS{n.~} \tbGE{uncastrated bull} 
\tbLX{naadalee} \tbHM{} \tbPH{náádàléé} \tbPS{n.~} \tbGE{young bull, calf} \tbOP{naadabilii} \tbTP{naadabilee} 
\tbLX{naadaŋaa} \tbHM{} \tbPH{náádáŋáá} \tbPS{n.~} \tbGE{kraal for cattle} \tbOP{naadaŋene} \tbTP{naadaŋenɛɛ} 
\tbLX{naadavaraa} \tbHM{} \tbPH{náádàváràà} \tbPS{n.~} \tbGE{castrated bull} 
\tbLX{naadɛre} \tbHM{} \tbPH{náádɛ́rɪ́} \tbPS{n.~} \tbGE{bullock, bull} \tbOP{naadɛɛ} \tbTP{naadɛrɛɛ} 
\tbLX{naadɔgrɔ} \tbHM{} \tbPH{náádɔ̀grɔ́} \tbPS{n.~} \tbGE{a cow that has recently given birth} 
\tbLX{naaduoraa} \tbHM{} \tbPH{nåådúóráá} \tbPS{n.~} \tbGE{resting ground for cattle} \tbOP{naaduori} \tbTP{naaduoree} 
\tbLX{naafaa} \tbHM{} \tbPH{nààfáà} \tbPS{n.~} \tbGE{that which is advantageous} \tbXV{Ka anaŋ waa lɛ ka lɛ naafaa la kyaare te.} \tbXE{If that is what it has come to, then luck is on our side.} 
\tbLX{naaga} \tbHM{} \tbPH{nààgà} \tbPS{n.~} \tbGE{a type of guinea corn} \tbOP{naagare} \tbTP{naagarɛɛ} 
\tbLX{naagaaraa} \tbHM{} \tbPH{náágááráá} \tbPS{n.~} \tbGE{feral cow} \tbOP{naagaare} \tbTP{naagaarɛɛ} 
\tbLX{naagale} \tbHM{} \tbPH{náágálɪ́} \tbPS{n.~} \tbGE{a type of dance which is meant to invite other dancers to the stage} \tbOP{naagalle} \tbTP{naagalɛɛ} 
\tbLX{naagane} \tbHM{} \tbPH{náágánɪ̀} \tbPS{n.~} \tbGE{cowhide} \tbOP{naagama} \tbTP{naagannɛɛ} 
\tbLX{naagaŋa} \tbHM{} \tbPH{náágàŋá} \tbPS{n.~} \tbGE{kraal for cattle} \tbOP{naaganne} \tbTP{naagannɛɛ} 
\tbLX{naakoɔraa} \tbHM{} \tbPH{náákʊ́ɔ́ráá} \tbPS{n.~} \tbGE{working bullock} \tbOP{naakoɔre} \tbTP{naakoɔrɛɛ} 
\tbLX{naakoɔrɔ} \tbHM{} \tbPH{náákʊ́ɔ́rɔ́} \tbPS{n.~} \tbGE{cattle seller} \tbOP{naakoɔrebɔ} \tbTP{naakoɔrebɛɛ} 
\tbLX{naakyeemɛgbeŋkeroo} \tbHM{} \tbPH{nááʧɪ́ɪ́mɛ́gběŋkéróó} \tbPS{n.~} \tbGE{a type of wild small tree with edible fruit} \tbOP{naakyeemɛgbeŋkerri} \tbTP{naakyeemɛgbeŋkeree} \tbVA{gbeŋkeroo}
\tbLX{naakyeenɛ} \tbHM{} \tbPH{nááʧɪ́ɪ́nɛ́} \tbPS{n.~} \tbGE{cattle herdsman} \tbOP{naakyeemɛ} \tbTP{naakyeemɛɛ} 
\tbLX{naalee} \tbHM{} \tbPH{nááléé} \tbPS{n.~} \tbGE{calf, heifer; a young cow or bull} \tbOP{naabilii} \tbTP{naabillee} 
\tbLX{naamaaroŋ} \tbHM{} \tbPH{nàámáárʊ́ŋ} \tbPS{n.~} \tbGE{large umbrella} \tbOP{naamaara} \tbTP{naamaarɛɛ} 
\tbLX{naamanyeɛnɛ} \tbHM{} \tbPH{náámáɲɪ́ɛ́nɛ́} \tbPS{adj.~} \tbGE{common, unimportant} 
\tbLX{naameɛ} \tbHM{} \tbPH{nààmɪ̀ɛ́} \tbPS{n.~} \tbGE{water turtle} \tbOP{naameene} \tbTP{naameenɛɛ} \tbVA{naameɛo}
\tbLX{naameɛo} \tbHM{} \tbPH{nààmɪ̀ɛ̀ʊ̀} \tbPS{n.~} \tbGE{turtle} \tbOP{naameɛne} \tbTP{naameɛnɛɛ} \tbVA{naameɛ}
\tbLX{naamoɔ} \tbHM{} \tbPH{náámʊ́ɔ́} \tbPS{n.~} \tbGE{a type of grass} \tbSG{naamoɔnoo} \tbOP{naamoore} \tbTP{naamoɔnɛɛ} 
\tbLX{naamu} \tbHM{} \tbPH{nàámù} \tbPS{n.~} \tbGE{head cloth} \tbOP{naamuri} \tbTP{naamuree} 
\tbLX{naane} \tbHM{} \tbPH{nàànɪ̀} \tbPS{v.~} \tbGE{to prepare, to get ready, to gather momentum to do something} \tbPD{naanɛɛ, naana, naana, naanema, naanaa, - } 
\tbLX{naane} \tbHM{} \tbPH{náánɪ́} \tbPS{aux. v.~} \tbGE{not} \tbXV{Ona naane de a libie.} \tbXE{S/he is not the one who took the money.} 
\tbLX{naanɛne} \tbHM{} \tbPH{náánɛ́nɪ̀} \tbPS{n.~} \tbGE{beef} \tbOP{naanɛmɛ} \tbTP{naanɛnnɛɛ} 
\tbLX{naanyaŋaa} \tbHM{} \tbPH{nááɲáŋàá} \tbPS{n.~} \tbGE{cow that has had a calf} \tbOP{naanyaŋene} \tbTP{naanyaŋenɛɛ} 
\tbLX{Naaŋmene} \tbHM{} \tbPH{nààŋmɪ́nɪ̀} \tbPS{n.~} \tbGE{God} \tbXV{naaŋmemɛ} \tbXE{} \tbOP{naaŋmemɛ} \tbTP{naaŋmenɛɛ} 
\tbLX{naapɔge} \tbHM{} \tbPH{nààpɔ́gɪ́} \tbPS{n.~} \tbGE{daughter-in-law} \tbOP{naapɔgeba/naapɔgemine} \tbTP{naapɔgrɛɛ} 
\tbLX{naapɔge} \tbHM{} \tbPH{náápɔ́gɪ́} \tbPS{n.~} \tbGE{female cow} \tbOP{naapɔgebɔ} \tbTP{naapɔgebɛɛ} 
\tbLX{naapɔle} \tbHM{} \tbPH{náápɔ́lɪ̀} \tbPS{n.~} \tbGE{young cow} \tbOP{naapɔlɔ} 
\tbLX{naara} \tbHM{} \tbPH{náárà} \tbPS{n.~} \tbGE{a variety of sorghum} \tbSG{naaroo} \tbOP{naare} \tbTP{naarɛɛ} 
\tbLX{naare} \tbHM{} \tbPH{náárɪ̀} \tbPS{adj.~} \tbGE{four} \tbOP{naare} \tbTP{naarɛɛ} 
\tbLX{naare} \tbHM{} \tbPH{nààrɪ̀} \tbPS{v.~} \tbSN \tbGE{strife to snatch from the ground} \tbSN \tbGE{to attack, usually violently} \tbPD{naarɛɛ, naara, naara, naareba, naaraa, - } 
\tbLX{naare} \tbHM{} \tbPH{náárɪ̀} \tbPS{v.~} \tbGE{to delay, to be late} \tbXV{A dɔɔta naŋ woŋ a bie baaloŋ o ba naare fe᷉e zaa.} \tbXE{When the doctor learnt of the child's sickness, he did not delay at all.} \tbPD{naarɛɛ, naara, naara, naareba, naaraa, - } 
\tbLX{naasaa} \tbHM{} \tbPH{náásàà} \tbPS{n.~} \tbGE{dragonfly} \tbOP{naasaare} \tbTP{naasaarɛɛ} 
\tbLX{naasaalaa} \tbHM{} \tbPH{nààsááláá} \tbPS{n.~} \tbGE{white person, European} \tbOP{nasamine} \tbTP{naasaalɛɛ} 
\tbLX{naasaal-kuuri} \tbHM{} \tbPH{nààsààl-kúúrí} \tbPS{n.~} \tbGE{plough} \tbOP{naasaal-kue} \tbTP{naasaal-kuree} 
\tbLX{naasaalsensɛroo} \tbHM{} \tbPH{nààsàálsɪ́nsɛ́rʊ́ʊ́} \tbPS{n.~} \tbGE{biscuits} \tbSG{naasaalsensɛroo} \tbOP{naasaalsensɛ} \tbTP{naasaalsensɛrɛɛ} 
\tbLX{naasaalsori} \tbHM{} \tbPH{nààsààlsórí} \tbPS{n.~} \tbGE{road} \tbOP{naasaalsoe} \tbTP{naasaalsoree} 
\tbLX{naasaal-taŋaa} \tbHM{} \tbPH{nààsààl-tàŋáà} \tbPS{n.~} \tbGE{mango} \tbOP{naasaal-taŋema} \tbTP{naasaal-taŋemɛɛ} \tbVA{maŋgo}
\tbLX{naasaal-teŋɛ} \tbHM{} \tbPH{nààsààl-tɪ́ŋɛ́} \tbPS{n.~} \tbGE{Europe} \tbOP{naasaal-tenne} \tbTP{naasaal-tennɛɛ} \tbVA{nasa-teŋɛ}
\tbLX{naasaraa} \tbHM{} \tbPH{náásáràà} \tbPS{n.~} \tbGE{adult female cow that has not littered yet} \tbOP{naasarre} \tbTP{naasarrɛɛ} \tbVA{naasarelee}
\tbLX{naasarelee} \tbHM{} \tbPH{náásárɪ́léé} \tbPS{n.~} \tbGE{adult female cow that has not littered yet} \tbOP{naasarebilii} \tbTP{naasarebillee} \tbVA{'naasaraa}
\tbLX{naaseere} \tbHM{} \tbPH{nààsɪ̀ɪ̀rɪ̀} \tbPS{n.~} \tbGE{a diseaese of the gums and teeth that produces sores} \tbOP{naaseerɛɛ} 
\tbLX{naaweɛmɛ} \tbHM{} \tbPH{nàáwɪ́ɛ́mɛ́} \tbPS{n.~} \tbGE{strayed cow} \tbOP{naaweɛne} \tbTP{naaweɛnɛɛ} 
\tbLX{naaweɛzoore} \tbHM{} \tbPH{nááwɪ́ɛ́zʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{a type of grass} \tbOP{naaweɛzoɛ} \tbTP{naaweɛzoorɛɛ} 
\tbLX{naayiri} \tbHM{} \tbPH{nààyírì} \tbPS{n.~} \tbSN \tbGE{palace, chief's house} \tbSN \tbGE{place of refuge} \tbOP{naayie} \tbTP{naayiree} 
\tbLX{naayoɔ} \tbHM{} \tbPH{nààyʊ́ɔ} \tbPS{n.~} \tbGE{young one, child} \tbOP{naayoore} \tbTP{naayoorɛɛ} 
\tbLX{naayoɔre} \tbHM{} \tbPH{nààyʊ̀ɔ̀rɪ̀} \tbPS{n.~} \tbSN \tbGE{penis of a cow} \tbSN \tbGE{a type of whip made of leather} \tbOP{naayɔɛ} \tbTP{naayɔrɛɛ} 
\tbLX{naazage} \tbHM{} \tbPH{náázàgɪ́} \tbPS{n.~} \tbGE{cattle pen} \tbOP{naazagre} \tbTP{naazagrɛɛ} 
\tbLX{naazeɛraa} \tbHM{} \tbPH{náázɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{hostile cow (usually one that has recently given birth)} \tbOP{naazeɛre} 
\tbLX{naazoŋmaa} \tbHM{} \tbPH{náázʊ̀ŋmáá} \tbPS{n.~} \tbGE{a cow that has lost part or all its tail} \tbOP{naazoŋmaara} \tbTP{naazoŋmaarɛɛ} 
\tbLX{naazoɔ} \tbHM{} \tbPH{náázʊ́ɔ̀} \tbPS{n.~} \tbGE{fly, housefly} \tbOP{naazoore} \tbTP{naazoorrɛɛ} \tbVA{zoɔ}
\tbLX{naazo-sɔglaa} \tbHM{} \tbPH{náázʊ̀-sɔ́gláá} \tbPS{n.~} \tbGE{black fly} \tbOP{naazo-sɔglɔ} \tbTP{naazo-sɔglɛɛ} 
\tbLX{naazozuri} \tbHM{} \tbPH{náázʊ̀zúrí} \tbPS{n.~} \tbGE{type of beads that look like the heads of flies} \tbTP{naazozuree} 
\tbLX{nabɛllebie} \tbHM{} \tbPH{nàbɛ́llɪ́bíé} \tbPS{n.~} \tbGE{camouflage} \tbOP{nabɛllebiiri} \tbTP{nabɛllebiiree} 
\tbLX{nabie} \tbHM{} \tbPH{nàbíé} \tbPS{n.~} \tbGE{royal person, prince} \tbOP{nabiiri} \tbTP{nabiiree} 
\tbLX{nabiiluŋ} \tbHM{} \tbPH{nàbììlúŋ} \tbPS{n.~} \tbGE{the condition of being a prince} 
\tbLX{naboma} \tbHM{} \tbPH{nàbʊ́má} \tbPS{n.~} \tbGE{riches, treasure} \tbOP{nabomɛɛ} 
\tbLX{nadire} \tbHM{} \tbPH{nàdíré} \tbPS{n.~} \tbGE{one who is qualified to be chief, chief-elect} \tbOP{ nadiribe} \tbTP{nadiribee} 
\tbLX{naferemɛ} \tbHM{} \tbPH{náfɪ́rɪ́mɛ́} \tbPS{n.~} \tbSN \tbGE{heel} \tbSN \tbGE{hoof} \tbSN \tbGE{hind toe (of a bird)} \tbOP{naferre} \tbTP{naferɛɛ/naferemɛɛ} 
\tbLX{nage} \tbHM{} \tbPH{nágɪ́} \tbPS{v.~} \tbGE{to beat, to hit hard (especially on the back)} \tbPD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\tbLX{nage} \tbHM{} \tbPH{nágɪ́} \tbPS{v.~} \tbGE{to produce a lot of} \tbPD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\tbLX{nage} \tbHM{} \tbPH{nàgɪ̀} \tbPS{v.~} \tbGE{to bend backward} \tbPD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\tbLX{nage} \tbHM{} \tbPH{nàgɪ̀} \tbPS{v.~} \tbGE{to miss work} \tbXV{Daa zaa o maŋ nage la toma dayeni.} \tbXE{Every week she misses work once.} \tbPD{nagɛɛ, nagra, nagra, nagreba, nagraa, - } 
\tbLX{nage} \tbHM{} \tbPH{nàgɪ́} \tbPS{adj.~} \tbGE{crooked, bent backwards} \tbOP{nagre} 
\tbLX{nagenyuu} \tbHM{} \tbPH{nàgìɲúú} \tbPS{n.~} \tbGE{meningitis} 
\tbLX{nagbare} \tbHM{} \tbPH{nàgbárí} \tbPS{n.~} \tbGE{shin} \tbOP{nagbarɛɛ} \tbTP{nagharrɛɛ} 
\tbLX{nagbime} \tbHM{} \tbPH{nàgbímé} \tbPS{n.~} \tbGE{heavy footsteps} \tbTP{nagbimee} 
\tbLX{nakogi} \tbHM{} \tbPH{nàkógí} \tbPS{n.~} \tbGE{throne} \tbOP{nakogri} \tbTP{nakogree} 
\tbLX{nakpi} \tbHM{} \tbPH{nàkpí} \tbPS{adj.~} \tbGE{mortal} 
\tbLX{nakyige} \tbHM{} \tbPH{nàʧígé} \tbPS{n.~} \tbGE{that which is great} \tbOP{nakyigri} \tbTP{nakyigree} 
\tbLX{nakyige} \tbHM{} \tbPH{nàʧígé} \tbPS{n.~} \tbGE{four-legged animal} \tbOP{nakyigri} \tbTP{nakyigree} 
\tbLX{nakyige} \tbHM{} \tbPH{nàʧígé} \tbPS{n.~} \tbGE{long, heavy stride} \tbOP{nakyigri} \tbTP{nakyigree} 
\tbLX{nama} \tbHM{} \tbPH{nàmá} \tbPS{n.~} \tbGE{carnivorous animal e.g. hyena} \tbTP{namɛɛ} 
\tbLX{namba} \tbHM{} \tbPH{námbà} \tbPS{n.~} \tbGE{number} \tbOP{nambare} \tbTP{nambarɛɛ} 
\tbLX{nambaŋka} \tbHM{} \tbPH{nàmbàŋkà} \tbPS{n.~} \tbGE{state of being probable} 
\tbLX{nambiyaalaa} \tbHM{} \tbPH{námbìyáálàà} \tbPS{n.~} \tbGE{centipede} \tbOP{nambiyaale} \tbTP{nambiyaalɛɛ} 
\tbLX{nambonna} \tbHM{} \tbPH{nàmbʊ̀nná} \tbPS{n.~} \tbGE{doubtfulness} 
\tbLX{nambɔgelɔ} \tbHM{} \tbPH{nǎmbɔ́glɔ́} \tbPS{n.~} \tbGE{hunter} \tbOP{nambɔgele} \tbTP{nambɔgelɛɛ} 
\tbLX{namelemɛ} \tbHM{} \tbPH{nàmɪ́lɪ́mɛ́} \tbPS{n.~} \tbGE{knock-knee} \tbTP{namelemɛɛ} 
\tbLX{namonɔ} \tbHM{} \tbPH{nàmʊ̀nɔ̀} \tbPS{n.~} \tbGE{body decoration e.g. tattoo} \tbOP{namonne} \tbTP{namonɛɛ/namonnɛɛ} \tbVA{namona}
\tbLX{nana} \tbHM{} \tbPH{nàná} \tbPS{adv.~} \tbGE{slowly and painfully e.g. a very sick person} 
\tbLX{nandaa} \tbHM{} \tbPH{nàndàá} \tbPS{n.~} \tbGE{limb} \tbOP{nandaare} \tbTP{nandarɛɛ} 
\tbLX{nandaana} \tbHM{} \tbPH{nándááná} \tbPS{n.~} \tbGE{poor person, needy person} \tbOP{ nandaanema} 
\tbLX{naneŋ} \tbHM{} \tbPH{nànɪ́ŋ} \tbPS{adj.~} \tbGE{at last} \tbXV{Saa naneŋ mie la.} \tbXE{At last it has rained.} 
\tbLX{nanna-nanna} \tbHM{} \tbPH{nánná-nánná} \tbPS{adv.~} \tbGE{seldom} \tbXV{O da ba nagera yiri kyɛ o ma naŋ kpi o maŋ nanna-nanna la wa kuli wa.} \tbXE{She used to come home regularly but since her mother's death she seldom comes home.} 
\tbLX{nannyige} \tbHM{} \tbPH{nànnyìgé} \tbPS{n.~} \tbGE{thief} \tbOP{nannyigiri} \tbTP{nannyigiree} 
\tbLX{nansoba} \tbHM{} \tbPH{nánsʊ̀bá} \tbPS{n.~} \tbGE{poor person, needy person} 
\tbLX{nantɛŋ} \tbHM{} \tbPH{nàntɛ̀ŋɪ̀} \tbPS{n.~} \tbGE{very large gun} \tbOP{nantɛnne} \tbTP{nantɛnnɛɛ} \tbVA{nantɛŋ}
\tbLX{nanyuuri} \tbHM{} \tbPH{nàɲúúrí} \tbPS{n.~} \tbGE{potato; sweet potato} \tbSG{nanyuuruu} \tbOP{nanyie} \tbTP{nanyiiree} 
\tbLX{naŋ} \tbHM{1} \tbPH{nǎŋ} \tbPS{adv.~} \tbSN \tbGE{since} \tbXV{O naŋ gaa  a kɔɔrɛɛ la.} \tbXE{It has been a long time since she went.} \tbSN \tbGE{yet} \tbXV{O naŋ ba gaa.} \tbXE{She has not gone yet.} \tbXV{O suuri naŋ naŋ ba sigi na te koŋ baŋ maale a zɔɔre.} \tbXE{Since his anger has not cooled off yet, we cannot arbitrate.} 
\tbLX{naŋ} \tbHM{2} \tbPH{nǎŋ} \tbPS{adv.~} \tbSN \tbGE{still} \tbXV{O naŋ zeŋ la be.} \tbXE{She is still sitting there.} \tbSN \tbGE{kindly, please} \tbXV{Naŋ ko ma ko᷉ɔ ka N nyu.} \tbXE{Kindly give me some water to drink.} 
\tbLX{naŋa} \tbHM{} \tbPH{náŋá} \tbPS{n.~} \tbGE{scorpion} \tbOP{nanne} \tbTP{nannɛɛ} 
\tbLX{naŋe} \tbHM{} \tbPH{náŋɪ̀} \tbPS{n.~} \tbGE{poverty; scarcity} \tbXV{O toma lee la ka naŋ kpɛ o.} \tbXE{His business has collapsed and poverty has overtaken him.} \tbTP{nannɛɛ} 
\tbLX{naŋene} \tbHM{} \tbPH{náŋɪ̀nɪ̀} \tbPS{n.~} \tbGE{hippopotamus} \tbOP{naŋema} \tbTP{naŋenɛɛ} 
\tbLX{naŋgbagele} \tbHM{} \tbPH{nǎŋgbáglɪ́} \tbPS{n.~} \tbGE{bony meat; bone left over after the flesh has been removed} \tbOP{naŋgbagela} \tbTP{naŋgbagelɛɛ} 
\tbLX{naŋgbeɛnɛ} \tbHM{} \tbPH{nǎŋgbɪ́ɛ́rɛ́} \tbPS{n.~} \tbSN \tbGE{bits of meat cut from from different parts of the animal} \tbSN \tbGE{profit} \tbXV{Boŋ naŋgbeɛrɛ ka fo nyɛ a toma ŋa poɔ?} \tbXE{What profit have you made from this work?.} \tbOP{ naŋgbeɛmɛ} \tbTP{naŋgbeɛrɛɛ} \tbVA{naŋgbeɛnɛ}
\tbLX{naŋkoɔredie} \tbHM{} \tbPH{nàŋkʊ̀ɔ̀rɪ̀dìé} \tbPS{n.~} \tbGE{store} \tbOP{naŋkoɔrederi} \tbTP{nankoɔrederee} 
\tbLX{naŋkoɔrɔ} \tbHM{} \tbPH{nàŋkʊ́ɔ́rɔ́} \tbPS{n.~} \tbGE{trader} \tbOP{naŋkoɔreba} \tbTP{naŋkoɔrebɛɛ} \tbVA{naŋkoɔreba}
\tbLX{naŋkoɔse} \tbHM{} \tbPH{nàŋkʊ̀ɔ̀sɪ́} \tbPS{n.~} \tbGE{butcher} \tbOP{naŋkoɔsere} \tbTP{naŋkoɔserɛɛ} 
\tbLX{naŋkpaana} \tbHM{} \tbPH{nǎŋkpááná} \tbPS{n.~} \tbGE{hunter} \tbOP{naŋkpaama/naŋkpaane} \tbTP{naŋkpaanɛɛ} 
\tbLX{naŋmane} \tbHM{} \tbPH{nàŋmánɪ́} \tbPS{n.~} \tbGE{pigeon} \tbOP{naŋmama} \tbTP{naŋmannɛɛ} 
\tbLX{napoloŋ} \tbHM{} \tbPH{nàpʊ̀lʊ́ŋ} \tbPS{n.~} \tbGE{calf (part of the leg)} \tbOP{napolle} \tbTP{napollɛɛ} 
\tbLX{Napɔge} \tbHM{} \tbPH{nàpɔ́gɪ́} \tbPS{n.~} \tbGE{name traditionally given to a younger female twin} \tbOP{napɔgeba} \tbTP{napɔgebɛɛ} 
\tbLX{naraa} \tbHM{} \tbPH{nàràà} \tbPS{adv.~} \tbGE{large} 
\tbLX{nare} \tbHM{} \tbPH{nárɪ́} \tbPS{v.~} \tbGE{to prepare; to get ready} \tbPD{narɛɛ, nara, nara, nareba, naraa, - } 
\tbLX{nare} \tbHM{} \tbPH{nàrɪ̀} \tbPS{v.~} \tbGE{to increase, to get out of control} \tbXV{A vu᷉u narɛɛ la ka ba kpinne faana.} \tbXE{The fire has gotten out of control and they are struggling to put it out.} \tbPD{narɛɛ, nara, nara, nareba, naraa, - } 
\tbLX{nasa-kuuri} \tbHM{} \tbPH{nàsà-kúúrí} \tbPS{n.~} \tbGE{plough} \tbOP{nasa-kue} 
\tbLX{natɛbaa} \tbHM{} \tbPH{nátɛ́báá} \tbPS{n.~} \tbGE{large boil or abscess} \tbOP{natɛrre} \tbTP{natɛrɛɛ} 
\tbLX{nato} \tbHM{} \tbPH{nàtó} \tbPS{n.~} \tbGE{a pair of metal high sounding instrument, used to call people's attention and as a musical instrument} \tbOP{natori} \tbTP{natoree} \tbVA{lato; latori}
\tbLX{nawale} \tbHM{} \tbPH{náwálɪ́} \tbPS{n.~} \tbGE{ulcer} \tbOP{nawalle} \tbTP{nawalɛɛ} \tbVA{nayagere}
\tbLX{nayagere} \tbHM{} \tbPH{nàyàgɪ̀rɪ́} \tbPS{n.~} \tbGE{ulcer} \tbOP{nayaga} \tbTP{nayagerɛɛ} \tbVA{nawale}
\tbLX{ne} \tbHM{} \tbPH{nɪ̂} \tbPS{conj.~} \tbSN \tbGE{and} \tbXV{Ko᷉ɔ ne vu᷉u᷉ koŋ to᷉ɔ᷉ laŋ kpɛ.} \tbXE{Water and fire cannot coexist.} \tbSN \tbGE{with} \tbXV{Te maŋ kyɛ la daa ne lare.} \tbXE{We cut a tree with an axe.} \tbVA{shortened form of ane}
\tbLX{neere} \tbHM{} \tbPH{nɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbGE{to dodge by squeezing through} \tbXV{A nannyige neere tu la a lɔɔre takoroo zo.} \tbXE{The thief squeezed himself through the car window and escaped.} \tbPD{neerɛɛ, neerɛ, neerɛ, neerebɛ, neeraa, - } 
\tbLX{neɛ} \tbHM{} \tbPH{nɪ́ɛ́} \tbPS{v.~} \tbSN \tbGE{to trample, to stamp on} \tbSN \tbGE{to cheat} \tbPD{neɛɛ, neɛrɛ, neɛrɛ, neɛrebɛ, neɛraa, neɛrebɛ } 
\tbLX{neɛ} \tbHM{} \tbPH{nɪ́ɛ̀} \tbPS{v.~} \tbSN \tbGE{to step on} \tbXV{N ba baŋ kyɛ neɛ o gbɛre kyɛ ka o suuri iri.} \tbXE{I stepped on her foot unintentionally, but she was angry.} \tbSN \tbGE{to grind} \tbXV{A te᷉e᷉-erɛ neɛ la a te᷉e᷉ eŋ ko᷉ɔ᷉ poɔ ko a baala.} \tbXE{The herbal doctor ground the medicine and mixed it with water for the patient.} \tbPD{neɛɛ, neɛrɛ, neɛrɛ, neɛrebɛ, neɛraa, neɛrebɛ } 
\tbLX{neɛ} \tbHM{} \tbPH{nɪ̀ɛ̀} \tbPS{n.~} \tbGE{person} \tbOP{noba} \tbTP{nerrɛɛ} 
\tbLX{neɛne} \tbHM{} \tbPH{nɪ̀ɛ̀nɪ̀} \tbPS{v.~} \tbSN \tbGE{to push through or under} \tbXV{Kɛkɛpele neɛne la a bie nubi-e᷉ɛ.} \tbXE{A piece of stalk has lodged itself under the child's fingernail.} \tbSN \tbGE{to refer to indirectly, to gossip} \tbXV{A pɔge ŋa maŋ neɛnɛ la o yontaa.} \tbXE{This woman always gossips about her rival.} \tbSN \tbGE{to tease} \tbPD{neɛnɛɛ, neɛnɛ, neɛnɛ, neɛnemɛ, neɛnaa, - } 
\tbLX{neɛre} \tbHM{} \tbPH{nɪ́ɛ́rɪ̀} \tbPS{n.~} \tbGE{grindstone} \tbOP{nɛɛ} \tbTP{nɛrɛɛ} 
\tbLX{neɛre} \tbHM{} \tbPH{nɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to hide under cover} \tbXV{Kɔkɔlaa digi la a gyaŋgboŋo ka o zo kpɛ neɛre moɔ poɔ.} \tbXE{The cat chased the mouse and it ran into hiding under the grass.} \tbPD{-, -, -, -, -, - } 
\tbLX{neɛrekuuri} \tbHM{} \tbPH{nɪ́ɛ́rɪ́kúúrí} \tbPS{n.~} \tbGE{small stone used in grinding on the main grindstone} \tbOP{neɛrekubo} \tbTP{neɛrekuuree} \tbVA{nɛbiri}
\tbLX{neɛre-saare} \tbHM{} \tbPH{nɪ́ɛ́rsáárɪ́} \tbPS{n.~} \tbGE{small, soft broom used to gather flour} \tbOP{neɛrsaɛ} \tbTP{neɛrsaarɛɛ} 
\tbLX{neɛzaa} \tbHM{} \tbPH{nɪ́ɛ̀zàá} \tbPS{n.~} \tbGE{each person, everyone} \tbXV{Ba yeli ka neɛzaa gaa a kpaaroŋ.} \tbXE{It has been said that every person should attend the meeting.} 
\tbLX{negebaga} \tbHM{} \tbPH{nɪ̀gɪ̀bàgà} \tbPS{adj.~} \tbGE{heavily built; usually referring to people} 
\tbLX{negiri} \tbHM{} \tbPH{négirí} \tbPS{v.~} \tbGE{to bend the knees slightly as a sign of respect; usually by girls while the boys salute} \tbPD{negiree, negire, negire, negiribe, negiraa, - } 
\tbLX{nelebala} \tbHM{} \tbPH{nɪ̀lɪ̀bàlà} \tbPS{adj.~} \tbGE{large and bent backward} 
\tbLX{nembaala} \tbHM{} \tbPH{nɪ́mbààlá} \tbPS{n.~} \tbGE{destitute, weak} \tbOP{nembaalba} \tbTP{nembaalbɛɛ} 
\tbLX{nembaaloŋ} \tbHM{} \tbPH{nɪ́mbààlʊ́ŋ} \tbPS{n.~} \tbGE{pity} 
\tbLX{nembaal-zoro} \tbHM{} \tbPH{nɪ́mbààl-zóró} \tbPS{n.~} \tbGE{merciful person} \tbOP{nembaal-zoribo} \tbTP{nembaal-zoribee} 
\tbLX{nembeene} \tbHM{} \tbPH{nɪ́mbɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{someone with supernatural powers} \tbOP{nembeemɛ} \tbTP{nembeemɛɛ} 
\tbLX{nembeɛ} \tbHM{} \tbPH{nɪ́mbɪ̀ɛ́} \tbPS{n.~} \tbGE{a mysterious person} \tbOP{nembeere} \tbTP{nembeerɛɛ} 
\tbLX{nembe᷉ɛo} \tbHM{} \tbPH{nɪ́mbɪ̀᷉ɛ̀᷉ʊ́᷉} \tbPS{n.~} \tbGE{crowd of people; many people} \tbOP{nebeɛne/nembeɛmɛ} \tbTP{nembeɛnɛɛ/nembeɛmɛɛ} 
\tbLX{nembeleŋ} \tbHM{} \tbPH{némbéléŋ} \tbPS{adj.~} \tbGE{bent slightly backwards} 
\tbLX{nembieri} \tbHM{} \tbPH{nɪ́mbíérí} \tbPS{n.~} \tbGE{malicious person} \tbOP{nembebe} 
\tbLX{nembone} \tbHM{} \tbPH{nɪ́mbʊ́nɪ́} \tbPS{n.~} \tbGE{worthless person} \tbOP{nemboma} \tbTP{nembonnɛɛ} 
\tbLX{nembon-yɛlɛ} \tbHM{} \tbPH{nɪ́bʊ́n-yɛ́lɛ́} \tbPS{n.~} \tbGE{useless talk} 
\tbLX{nemboŋ} \tbHM{} \tbPH{nímbóŋ} \tbPS{n.~} \tbGE{what sort of person, who} \tbXV{Nemboŋ la na to᷉ɔ᷉ di naaloŋ Dagaaba laŋkpeɛbo poɔ?} \tbXE{Who qualifies to be a chief among the Dagaaba?.} 
\tbLX{nemboore} \tbHM{} \tbPH{nɪ́mbʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{type of person} \tbXV{Maa ba baŋ nemboore na zaa naŋ la o.} \tbXE{I don't know what type of person she is.} \tbOP{nemboorɔ} 
\tbLX{nembɔnnaa} \tbHM{} \tbPH{nɪ́mbɔ̀nnáà} \tbPS{n.~} \tbGE{acquaintance} \tbOP{nembɔnne} \tbTP{nembɔnnɛɛ} 
\tbLX{nembuo} \tbHM{} \tbPH{nɪ́mbúò} \tbPS{n.~} \tbGE{which person; who} \tbOP{nembuobo} 
\tbLX{nemburee} \tbHM{} \tbPH{nɪ́mbùréé} \tbPS{n.~} \tbGE{unintelligent person, confused person} \tbOP{nemburri} \tbTP{nemburree} 
\tbLX{nemmeŋɛ} \tbHM{} \tbPH{nɪ́mmɪ́ŋɛ́} \tbPS{n.~} \tbGE{good person} \tbOP{nemenne} \tbTP{nemmennɛɛ} 
\tbLX{nempaalaa} \tbHM{} \tbPH{nímpááláá} \tbPS{n.~} \tbGE{new person} \tbOP{nempaalba} \tbTP{nempaalɛɛ} 
\tbLX{nempelaa} \tbHM{} \tbPH{nɪ́mpɪ̀làá} \tbPS{n.~} \tbSN \tbGE{caucasian; white person} \tbSN \tbGE{simple person, straightforward person} \tbOP{nempeɛle} \tbTP{nempeɛlɛɛ} 
\tbLX{nempɔgeraa} \tbHM{} \tbPH{nɪ́mpɔ̀gɪ̀ráà} \tbPS{n.~} \tbGE{prisoner} \tbOP{nempɔgere} \tbTP{nempɔgerɛɛ} 
\tbLX{nempɔgere-die} \tbHM{} \tbPH{nɪ́mpɔ̀gɪ̀rdìé} \tbPS{n.~} \tbGE{prison} \tbOP{nempɔgere-deri} \tbTP{nempɔgere-deree} 
\tbLX{nendaare} \tbHM{} \tbPH{nɪ̀ndààrɪ̀} \tbPS{adv.~} \tbGE{one day; in the future} \tbXV{O zagere la sakuuri gaabo nendaare kaŋa o na koŋ la dabaŋkalɛ.} \tbXE{S/he has refused to go to school; one day she will regret it.} 
\tbLX{nendɛgere} \tbHM{} \tbPH{nɪ́ndɛ́grí} \tbPS{n.~} \tbSN \tbGE{dirty person} \tbSN \tbGE{unscruplous person} \tbOP{nendɛgɛ} 
\tbLX{nenfaa} \tbHM{} \tbPH{nɪ̀nfáà} \tbPS{n.~} \tbGE{bad person, wicked person} \tbOP{nenfaare} \tbTP{nenfaarɛɛ} 
\tbLX{nenfole} \tbHM{} \tbPH{nɪ́nfʊ́lɪ́} \tbPS{n.~} \tbGE{albino} \tbOP{nenfola} \tbTP{nenfolɛɛ} 
\tbLX{nenlaane} \tbHM{} \tbPH{nínlåånɪ́} \tbPS{n.~} \tbGE{worthless person} \tbOP{nenlaama} \tbTP{nelaanɛɛ} 
\tbLX{nenlaara} \tbHM{} \tbPH{nɪ́nláárá} \tbPS{n.~} \tbGE{person who laughs at others; scornful person} \tbOP{nenlaareba} \tbTP{nenlaarebɛɛ} 
\tbLX{nenlaaraa} \tbHM{} \tbPH{nɪ́nlààràà} \tbPS{n.~} \tbGE{person who is scorned, laughingstock} \tbOP{nenlaare} \tbVA{neŋ-e-laaraa}
\tbLX{nen-nyɔgera} \tbHM{} \tbPH{nɪ́n-ɲɔ́gɪ́rà} \tbPS{n.~} \tbGE{conqueror} \tbOP{ nen-nyɔgereba} \tbTP{nen-nyɔgerebɛɛ} \tbVA{nen-nyɔgerɔ}
\tbLX{nen-nyɔgeraa} \tbHM{} \tbPH{nɪ́n-ɲɔ́gɪ́ráá} \tbPS{n.~} \tbGE{captive} \tbOP{ nen-nyɔgere} \tbTP{nen-nyɔgerɛɛ} 
\tbLX{nensaala} \tbHM{} \tbPH{nɪ̌nsààlá} \tbPS{n.~} \tbGE{human being} \tbOP{nensaaleba} \tbTP{nensaalebɛɛ} 
\tbLX{nensaalabie} \tbHM{} \tbPH{nɪ́nsààlàbíé} \tbPS{n.~} \tbGE{son of man} \tbOP{nensaalabiiri} \tbTP{nensaalabiiree} 
\tbLX{nensaaloŋ} \tbHM{} \tbPH{nɪ́nsààlʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{humanity} \tbSN \tbGE{something that is given a human face} \tbXV{Ka a karemazuzeɛŋ da ba de nensaaloŋ kpɛ ne a bie dɔgeroŋ poɔ a ŋa a bie yi a sakuuriŋ.} \tbXE{But for the headmaster giving a human face to the boy’s punishment the boy would have been out of school.} \tbSN \tbGE{personification (literature)} 
\tbLX{nensaal-sɔgelaa} \tbHM{} \tbPH{nɪ́nsáál-sɔ́gláá} \tbPS{n.~} \tbGE{the typical African person} \tbOP{nensaal-sɔglɔ} 
\tbLX{nenseɛ} \tbHM{} \tbPH{nɪ́nsɪ́ɛ̀} \tbPS{n.~} \tbGE{ghost, evil spirit, apparition} \tbOP{nenseere} \tbTP{nenseerɛɛ} 
\tbLX{nenseɛbaa} \tbHM{} \tbPH{nɪ̀nsɪ̀ɛ̀báá} \tbPS{n.~} \tbSN \tbGE{a fairly grown person in age} \tbXV{A pɔgelee serɛ e la nenseɛbaa.} \tbXE{The young woman's husband is middle aged.} \tbSN \tbGE{a fairly big person (size)} \tbOP{nenseɛbaare} \tbTP{nenseɛbaarɛɛ} 
\tbLX{nenselɛ} \tbHM{} \tbPH{nɪ̀nsɪ̊lɛ} \tbPS{interj.~} \tbGE{great, well done} \tbVA{menselɛ}
\tbLX{nensenyɛrɛ} \tbHM{} \tbPH{nɪ́nsɪ́ɲɛ́rɛ̀} \tbPS{n.~} \tbGE{one who sees spirits} \tbOP{nensenyɛrebɛ} \tbTP{nensenyɛrebɛɛ} 
\tbLX{nensoŋ} \tbHM{} \tbPH{nɪ́nsʊ́ŋ} \tbPS{n.~} \tbGE{good person, holy person,} \tbOP{nensonne} \tbTP{nensonnɛɛ} 
\tbLX{nensɔglaa} \tbHM{} \tbPH{nɪ́nsɔ́gláá} \tbPS{n.~} \tbGE{black person} \tbOP{nensɔglɔ} 
\tbLX{nentaare} \tbHM{} \tbPH{nɪ́ntáárɪ̀} \tbPS{n.~} \tbGE{saliva} \tbTP{nentaarɛɛ} 
\tbLX{nentaboli} \tbHM{} \tbPH{nɪ́ntábòlí} \tbPS{n.~} \tbGE{large glob of saliva} \tbOP{nentabolo} \tbTP{nentabolee} 
\tbLX{nentage} \tbHM{} \tbPH{nɪ́ntágɪ́} \tbPS{n.~} \tbSN \tbGE{a thickset person} \tbSN \tbGE{person of substance; wealthy} \tbOP{nentagere} \tbTP{nentagerɛɛ} 
\tbLX{nentage} \tbHM{} \tbPH{nɪ́ntàgɪ̀} \tbPS{n.~} \tbSN \tbGE{act of tugging e.g in a tussle} \tbSN \tbGE{the act of taking people to court} 
\tbLX{nentakyeɛrɛ} \tbHM{} \tbPH{nɪ́ntáʧɪɛrɛ} \tbPS{n.~} \tbSN \tbGE{dried saliva around the mouth} \tbSN \tbGE{dried tears on the cheek} \tbOP{nentakyeɛre} \tbTP{nentakyeɛrɛɛ} 
\tbLX{nentaŋ} \tbHM{} \tbPH{nɪ́ntáŋ} \tbPS{n.~} \tbGE{tears} \tbOP{nentanne} \tbTP{nentannɛɛ} \tbVA{ŋmentɔŋ, ŋmentɔnnɛɛ}
\tbLX{nentegeroŋ} \tbHM{} \tbPH{nɪ́ntɪ́grʊ́ŋ} \tbPS{n.~} \tbGE{important person; person of worth} \tbOP{nentegerɛ} 
\tbLX{nenweɛ} \tbHM{} \tbPH{nɪ́nwɪ́ɛ́} \tbPS{n.~} \tbSN \tbGE{a person who is not related to another} \tbSN \tbGE{unwanted person} \tbOP{neweere} \tbTP{nenweerɛɛ} 
\tbLX{neŋ-iraa} \tbHM{} \tbPH{nɪ́ŋ-íráà} \tbPS{n.~} \tbGE{chosen person} \tbOP{neŋ-irri} \tbTP{neŋ-irree} 
\tbLX{neŋkaa} \tbHM{} \tbPH{nɪ́ŋkàà} \tbPS{n.~} \tbSN \tbGE{the act of looking at people} \tbSN \tbGE{the act of caring for people} \tbOP{neŋkaareba} \tbTP{neŋkaarebɛɛ} 
\tbLX{neŋkaara} \tbHM{} \tbPH{nɪ́ŋkáárá} \tbPS{n.~} \tbSN \tbGE{one who looks at people} \tbSN \tbGE{one who takes care of people, caregiver} \tbOP{neŋkaareba} \tbTP{neŋkaarebɛɛ} 
\tbLX{neŋkaaraa} \tbHM{} \tbPH{nɪ̌ŋkààráà} \tbPS{n.~} \tbGE{one who is cared for} \tbOP{neŋkaare} 
\tbLX{neŋkanoo} \tbHM{} \tbPH{nɪ́ŋkànʊ́ʊ̀} \tbPS{n.~} \tbGE{mature person, old person} \tbOP{neŋkana/neŋkama} \tbTP{neŋkanɛɛ} 
\tbLX{neŋkaraa} \tbHM{} \tbPH{nɪ́ŋkáráá} \tbPS{n.~} \tbGE{an unruly person} \tbOP{neŋkare} \tbTP{neŋkarɛɛ} 
\tbLX{neŋkaraa} \tbHM{} \tbPH{nɪ́kàràà} \tbPS{n.~} \tbGE{large person} 
\tbLX{neŋkoorɔ} \tbHM{} \tbPH{nɪ́ŋkʊ́ʊ́rɔ̀} \tbPS{n.~} \tbGE{murderer} \tbOP{ neŋkoorebɔ} \tbVA{neŋkoora}
\tbLX{neŋkore} \tbHM{} \tbPH{nɪ́ŋkʊ́rɪ́} \tbPS{n.~} \tbGE{old person} \tbOP{neŋkora} 
\tbLX{neŋkɔge} \tbHM{} \tbPH{nɪ́ŋkɔ́gɪ́} \tbPS{n.~} \tbGE{lean person} \tbOP{neŋkɔgre} \tbTP{neŋkɔgrɛɛ} 
\tbLX{neŋkpeɛŋaa} \tbHM{} \tbPH{nɪ́ŋkpɪ́ɛ́ŋáá} \tbPS{n.~} \tbSN \tbGE{healthy person} \tbSN \tbGE{strong person} \tbOP{ neŋkpeɛne} 
\tbLX{neŋkpoŋi} \tbHM{} \tbPH{nɪ́ŋkpôŋ} \tbPS{n.~} \tbSN \tbGE{adult, elder} \tbSN \tbGE{leader} \tbOP{neŋkponni} \tbTP{neŋkponnee} 
\tbLX{neŋkpɔre} \tbHM{} \tbPH{nɪ́ŋkpɔ́rɪ́} \tbPS{n.~} \tbGE{weak and sickly person} \tbOP{neŋkpɔɛ} \tbTP{neŋkpɔrɛɛ} 
\tbLX{nerroŋ} \tbHM{} \tbPH{nɪ́rrʊ́ŋ} \tbPS{n.~} \tbGE{consanguinity, blood ties, blood relationship} \tbXV{A ba seŋ ka naasaalaa toma kaara nerroŋ.} \tbXE{Government employment should not be based on blood ties.} 
\tbLX{nɛbiri} \tbHM{} \tbPH{nɛ́bìrí} \tbPS{n.~} \tbGE{smaller stone used in grinding on the main grindstone} \tbOP{nɛbie} \tbTP{nɛbiiree} 
\tbLX{nɛbo} \tbHM{} \tbPH{nɛ́bʊ́} \tbPS{v.~} \tbSN \tbGE{to dress well} \tbSN \tbGE{to tuck in one's shirt} \tbPD{nɛbɛɛ, nɛbrɛ, nɛbraa, nɛbrebɛ, nɛbraa, - } 
\tbLX{nɛ-bɔgeraa} \tbHM{} \tbPH{nɛ́-bɔ́gɪ́ràá} \tbPS{n.~} \tbGE{smooth grinding stone} \tbOP{nɛ-bɔgere} \tbTP{nɛ-bɔgerɛɛ} 
\tbLX{nɛgaa} \tbHM{} \tbPH{nɛ́gáá} \tbPS{n.~} \tbGE{a type of cedar-like tree with very hard wood} \tbOP{nɛgere} \tbTP{nɛgerɛɛ} 
\tbLX{nɛgetɛgɛɛ} \tbHM{} \tbPH{nɛ̀gɪ̀tɛ́gɛ́ɛ́} \tbPS{adj.~} \tbGE{curved inward} \tbXV{A a die peɛ deɛ waa la nɛgetɛgɛɛ lɛ.} \tbXE{The roof of the house is looking sunken in.} \tbVA{nɛgtɛgɛ}
\tbLX{nɛgetɛgɛɛ} \tbHM{} \tbPH{nɛ̀gɪ́tɛ́gɛ́ɛ́} \tbPS{n.~} \tbGE{depression, place that is curved inward} \tbXV{Saa mie la ka ko᷉ɔ are a peɛ nɛgetɛgɛɛ poɔ.} \tbXE{It has rained and water has collected in a depression in the roof.} \tbOP{nɛgtɛgere} \tbTP{nɛgtɛgerɛɛ} 
\tbLX{nɛguo} \tbHM{} \tbPH{nɛ́gùó} \tbPS{n.~} \tbGE{raised ground on which grindstones are placed} \tbOP{nɛgori} \tbTP{nɛgoree} 
\tbLX{nɛ-gyieraa} \tbHM{} \tbPH{nɛ́-dzèráà} \tbPS{n.~} \tbGE{rough grindstone} \tbOP{nɛ-gyieri} 
\tbLX{nɛku-bile} \tbHM{} \tbPH{nɛ́kú-bílé} \tbPS{n.~} \tbGE{smaller version of a neɛrekuuri} \tbOP{nɛku-bilii} \tbTP{nɛku--bilee} 
\tbLX{nɛku-kpoŋ} \tbHM{} \tbPH{nɛ̀kú-kpôŋ} \tbPS{n.~} \tbGE{larger version of neɛrekuuri} \tbOP{nɛku-kponni} \tbTP{nɛku-kponnee} 
\tbLX{nɛle} \tbHM{} \tbPH{nɛ́lɪ́} \tbPS{v.~} \tbGE{to fill completely} \tbXV{A bogi da nɛle ne la ɔnne.} \tbXE{The burrow was full of mice.} \tbPD{nɛlɛɛ, nɛlɛ, nɛlaa, nɛlebɛ, nɛlaa, - } 
\tbLX{nɛll} \tbHM{} \tbPH{nɛ̀ll} \tbPS{adv.~} \tbGE{long and unwieldy} 
\tbLX{nɛmbeene} \tbHM{} \tbPH{nɛ́mbɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{cooked meat} \tbOP{nɛmbeemɛ} \tbTP{nɛmbeenɛɛ} 
\tbLX{nɛmbɛlɛŋ} \tbHM{} \tbPH{nɛ̀mbɛ̀lɛ̀ŋ} \tbPS{adv.~} \tbGE{long and crooked} 
\tbLX{nɛmbiri} \tbHM{} \tbPH{nɛ́mbírì} \tbPS{n.~} \tbGE{boneless meat} \tbSG{nɛmbiruu} \tbOP{nɛmbie} \tbTP{nɛmbiiree} 
\tbLX{nɛmmaaroŋ} \tbHM{} \tbPH{nɛ́mmáárʊ́ŋ} \tbPS{n.~} \tbGE{fresh meat} \tbOP{nɛmmaara} \tbTP{nɛmmaarɛɛ} 
\tbLX{nɛmpoɔnaa} \tbHM{} \tbPH{nɛ́mpʊ́ɔ́náá} \tbPS{n.~} \tbGE{rotten meat} \tbOP{nɛmpoɔne} \tbTP{nɛmpoɔnɛɛ} 
\tbLX{nɛne} \tbHM{} \tbPH{nɛ́nɪ̀} \tbPS{n.~} \tbSN \tbGE{meat} \tbSN \tbGE{flesh} \tbOP{ nɛmɛ} 
\tbLX{nɛnefɛne} \tbHM{} \tbPH{nɛ̀nɪ̀fɛ́nɪ́} \tbPS{v.~} \tbGE{to try to get attention} \tbXV{O naŋ waa biyeni na ana la so ka o nɛnefɛnɛ.} \tbXE{She is an only child; that is why she wants so much attention.} \tbPD{nɛnefɛnɛɛ, nɛnefɛnɛ, nɛnefɛnɛ, nɛnefɛnemɛ, nɛnefɛnaa, - } 
\tbLX{nɛnɛ} \tbHM{} \tbPH{nɛ́nɛ́} \tbPS{adv.~} \tbGE{precisely} \tbXV{A sokyɛnɛ manne la o sori yɛlɛ nɛnɛ lɛɛŋ te baare.} \tbXE{The traveler narrated precisely what she/he encountered on the journey.} 
\tbLX{nɛnzeɛ} \tbHM{} \tbPH{nɛ́nzɪ̀ɛ́} \tbPS{n.~} \tbGE{lean meat} \tbSG{nɛnzeroo} \tbOP{nɛnzeere} \tbTP{nɛnzeerɛɛ} 
\tbLX{nɛŋɛɛ} \tbHM{} \tbPH{nɛ̀ŋɛ̀ɛ̀} \tbPS{adj.~} \tbGE{saggy} 
\tbLX{nɛŋɛɛ} \tbHM{} \tbPH{nɛ́ŋɛ́ɛ́} \tbPS{adj.~} \tbGE{saggy} 
\tbLX{nɛŋka᷉a} \tbHM{} \tbPH{nɛ́ŋká᷉à᷉} \tbPS{n.~} \tbGE{fatty meat} \tbSG{nɛŋkaao} \tbOP{nɛŋkaanɛɛ} 
\tbLX{nɛŋkaraa} \tbHM{} \tbPH{nɛ́ŋkáráá} \tbPS{n.~} \tbGE{raw meat} \tbOP{nɛŋkare} \tbTP{nɛŋkarɛɛ} 
\tbLX{nɛŋkoorɔ} \tbHM{} \tbPH{nɛ́ŋkʊ́ʊ́rɔ̀} \tbPS{n.~} \tbGE{butcher} \tbOP{nɛŋkooreba} 
\tbLX{nɛŋkuoŋaa} \tbHM{} \tbPH{nɛ́ŋkúóŋáá} \tbPS{n.~} \tbGE{smoked meat} \tbOP{nɛŋkuoni} \tbTP{nɛŋkuonee} \tbVA{nɛŋku᷉oaa}
\tbLX{nɛŋŋmaara} \tbHM{} \tbPH{nɛ́ŋŋmáárá} \tbPS{n.~} \tbGE{butcher} \tbOP{nɛŋŋmaareba} \tbVA{naŋkoɔse, naŋkoɔsere, naŋkoɔserɛɛ}
\tbLX{nɛre} \tbHM{} \tbPH{nɛ̀rɪ̀} \tbPS{v.~} \tbGE{to be buried underneath e.g. in the soil} \tbPD{nɛrɛɛ, nɛrɛ, nɛrɛ, nɛrebɛ, nɛraa, - } 
\tbLX{nibiri} \tbHM{} \tbPH{nìbirì̀} \tbPS{n.~} \tbGE{emotion} \tbTP{nibiree} 
\tbLX{nii} \tbHM{} \tbPH{níì} \tbPS{adj.~} \tbGE{eight} 
\tbLX{nimibaaloŋ} \tbHM{} \tbPH{nímíbààlʊ̂ŋ} \tbPS{n.~} \tbGE{kindness} 
\tbLX{nimibaare} \tbHM{} \tbPH{nímíbáárɪ́} \tbPS{n.~} \tbGE{bloodshot} \tbTP{nimibaarɛɛ} \tbVA{nimibeɛmɛ}
\tbLX{nimibiri} \tbHM{} \tbPH{nímíbírì} \tbPS{n.~} \tbGE{pupil (of eye)} \tbSG{nimibiruu} \tbOP{nimibie} \tbTP{nimibiiree} 
\tbLX{nimibuli} \tbHM{} \tbPH{nímíbúlí} \tbPS{n.~} \tbGE{eye with a cataract} \tbOP{nimibulo} \tbTP{nimibulee} 
\tbLX{nimifaa} \tbHM{} \tbPH{nímífáá} \tbPS{n.~} \tbGE{evil eye} \tbOP{nimifaare} \tbTP{nimifaarɛɛ} 
\tbLX{nimifaa} \tbHM{} \tbPH{nímífáá} \tbPS{n.~} \tbGE{wrong side (of fabric)} \tbOP{nimifaare} \tbTP{nimifaarɛɛ} 
\tbLX{nimifuri} \tbHM{} \tbPH{nímífùrì} \tbPS{v.~} \tbSN \tbGE{to be wide eyed} \tbSN \tbGE{to show wit} \tbPD{-, -, -, -, -, - } 
\tbLX{nimigane} \tbHM{} \tbPH{nímígánɪ̀} \tbPS{n.~} \tbGE{eyelid} \tbOP{nimigama} \tbTP{nimiganɛɛ} 
\tbLX{nimigbaŋkɔɔloŋ} \tbHM{} \tbPH{nímígbàŋkɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{eyebrow} 
\tbLX{nimigbɛlɛɛ} \tbHM{} \tbPH{nímígbɛ̀lɛ̀ɛ̀} \tbPS{n.~} \tbGE{the state of being considerate} \tbVA{nimimaaroŋ}
\tbLX{nimigboŋkogiri} \tbHM{} \tbPH{nímígbòŋkógrí} \tbPS{n.~} \tbGE{brow, the part of the face where the eyebrow is located} \tbOP{nimigboŋkogiro} \tbTP{nimigboŋkogiree} \tbVA{(1p) nimigboŋkobo}
\tbLX{nimikaare} \tbHM{} \tbPH{nímíkáárɪ̀} \tbPS{n.~} \tbGE{eyeballs} \tbOP{nimikaare} 
\tbLX{nimiko} \tbHM{} \tbPH{nímíkò} \tbPS{n.~} \tbGE{act of doing wicked things} 
\tbLX{nimiko᷉ɔ} \tbHM{} \tbPH{nímíkʊ́᷉ɔ́᷉} \tbPS{n.~} \tbSN \tbGE{watery eyes} \tbSN \tbGE{tears} \tbTP{nimikoɔnɛɛ} 
\tbLX{nimikɔɔloŋ} \tbHM{} \tbPH{nímíkɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{eyelash} 
\tbLX{nimikuoŋaa} \tbHM{} \tbPH{nímíkúóŋáá} \tbPS{n.~} \tbGE{wickedness} \tbOP{nimikuoni} \tbTP{nimikuonee} \tbVA{nimitiiruŋ}
\tbLX{nimikyaane} \tbHM{} \tbPH{nímíʧáánɪ́} \tbPS{n.~} \tbGE{eye glasses, spectacles} \tbOP{nimikyaane} \tbTP{nimikyaanaa} 
\tbLX{nimilaa} \tbHM{} \tbPH{nímíláá} \tbPS{n.~} \tbGE{eye socket} \tbOP{nimilaare} \tbTP{nimilaarɛɛ} 
\tbLX{nimimaaroŋ} \tbHM{} \tbPH{nímímáárʊ́ŋ} \tbPS{n.~} \tbGE{kindness, having pity on others} \tbOP{nimimaara} \tbTP{nimimaarɛɛ} 
\tbLX{niminyɛrɛ} \tbHM{} \tbPH{nímínyɛ́rɛ̀} \tbPS{n.~} \tbGE{visionary, seer} \tbOP{ niminyɛrebɛ} \tbTP{niminyɛrebɛɛ} 
\tbLX{nimipe᷉ɛ} \tbHM{} \tbPH{nímípɪ̀᷉ɛ̀᷉} \tbPS{n.~} \tbGE{the act of claiming to be wise} 
\tbLX{nimipeɛnɛ} \tbHM{} \tbPH{nímípɪ́ɛ́nɛ́} \tbPS{n.~} \tbGE{one who claims to be wise} \tbOP{nimipeɛnemɛ} \tbTP{nimipeɛnemɛɛ} 
\tbLX{nimipe᷉ɛ᷉} \tbHM{} \tbPH{nímípɪ́᷉ɛ́᷉} \tbPS{n.~} \tbGE{the act of showing oneself to be wiser than others} 
\tbLX{nimipoŋi} \tbHM{} \tbPH{nímípôŋ} \tbPS{n.~} \tbGE{puss-like fluid on the surface of the eyeball} \tbOP{nimiponni} \tbTP{nimiponnee} \tbVA{nimipuuri}
\tbLX{nimipɔge} \tbHM{} \tbPH{nímípɔ̀gɪ̀} \tbPS{n.~} \tbGE{the act of blindfolding or confusing others} 
\tbLX{nimipuuri} \tbHM{} \tbPH{nímípúúrí} \tbPS{n.~} \tbGE{puss-like fluid on the surface of the eyeball} \tbOP{nimipuuree} \tbVA{nimipoŋi}
\tbLX{nimiri} \tbHM{} \tbPH{nímírì} \tbPS{n.~} \tbSN \tbGE{eye} \tbSN \tbGE{stinginess; avarice} \tbXV{N ba taa la libie kyɛ nimiri koŋ sage ka o da tiivi paalaa.} \tbXE{My father has the money but his stinginess would not allow him to buy a new TV.} \tbSN \tbGE{covetousness} \tbXV{A bie ŋa taa la nimiri, o maŋ nyɛ la o tɔ zaa bone boɔrɔ o.} \tbXE{This child is covetous; he wants anything that another child has.} \tbSG{nimiruu} \tbOP{nimie} \tbTP{nimiiree} 
\tbLX{nimiri} \tbHM{} \tbPH{nímírì} \tbPS{n.~} \tbGE{success} \tbXV{O toma taa la nimiri pampana.} \tbXE{That job of his is becoming fruitful.} \tbSG{nimiruu} \tbOP{nimie} \tbTP{nimiiree} 
\tbLX{nimiri-soba} \tbHM{} \tbPH{nímírì-sʊ̀bá} \tbPS{n.~} \tbSN \tbGE{miser} \tbSN \tbGE{one who covets} \tbOP{nimiri-deme} \tbTP{nimi-demee} 
\tbLX{nimisogɔ} \tbHM{} \tbPH{nímísʊ́gɔ́} \tbPS{n.~} \tbGE{face} \tbXV{A pɔge kaa la a o bibaala nimisogɔ kyɛ viiri zu.} \tbXE{The woman looked into her sick child's face and shook her head.} \tbOP{nimisogere} \tbTP{nimisogerɛɛ} 
\tbLX{nimisogɔ} \tbHM{} \tbPH{nímísʊ́gɔ́} \tbPS{adv.~} \tbGE{in front of, ahead of} \tbXV{A gane biŋ la n nimisogɔ.} \tbXE{The book is placed in front of me.} 
\tbLX{nimisoɔ} \tbHM{} \tbPH{nímísʊ́ɔ̀} \tbPS{n.~} \tbGE{dizziness, confusion} \tbOP{nimisoɔre} \tbTP{nimisoɔrɛɛ} 
\tbLX{nimitiiruŋ} \tbHM{} \tbPH{nímítììrúŋ} \tbPS{n.~} \tbGE{wickedness} \tbVA{nimikuoŋaa}
\tbLX{nimitoloŋ} \tbHM{} \tbPH{nímítʊ́lʊ́ŋ} \tbPS{n.~} \tbGE{haste} 
\tbLX{nimitɔɔre} \tbHM{} \tbPH{nímítɔ́ɔ́rɪ́} \tbPS{prep.~} \tbSN \tbGE{ahead, in front of} \tbXV{Ka fooŋ mɔnɔ lɔɔre a seŋ ka fo kaara nimitɔɔre.} \tbXE{If you drive you must look ahead.} \tbSN \tbGE{forward (distance, direction)} \tbXV{O yeli ka N gaa nimitɔɔre fe᷉e᷉ kyɛ leɛ tu nudoloŋ.} \tbXE{She directed that I should go ahead a little and turn right.} 
\tbLX{nimiyuo} \tbHM{} \tbPH{nímíyúò} \tbPS{n.~} \tbGE{an action that creates the impression that one is more intelligent than another} \tbXV{Te naŋ gaa a nɛne pommu Bayuo da boɔrɔ ka o wuli nimiyuo kyɛ ka te ba sage.} \tbXE{When we went to share the meat Bayuo wanted to cheat but we did not give him the chance.} 
\tbLX{nimiyuoro} \tbHM{} \tbPH{nímíyúóró} \tbPS{n.~} \tbGE{well-informed person, one who is abreast of the times} \tbOP{nimiyuoribo} \tbTP{nimiyuoribee} 
\tbLX{nimizeɛ} \tbHM{} \tbPH{nímízɪ̀ɛ́} \tbPS{n.~} \tbGE{seriousness} \tbOP{nimizeere} \tbTP{nimizeerɛɛ} 
\tbLX{niŋe} \tbHM{} \tbPH{níŋè} \tbPS{n.~} \tbGE{face} \tbXV{A bi-entuo na ba maŋ boɔrɔ na pɛge o niŋe baguo.} \tbXE{That lazy child does not like washing his face in the morning.} \tbOP{ninni} \tbTP{ninnee} \tbSE{niŋe sógɔ́ --in front, before, niŋesogre, niŋesogrɛɛ} \tbGE{in front, before} 
\tbLX{niŋe} \tbHM{} \tbPH{níŋé} \tbPS{adv.~} \tbGE{in front of, ahead of} \tbXV{O de la a kogi wa biŋ a saana niŋeŋ.} \tbXE{He placed the seat in front of the visitor.} 
\tbLX{nirii} \tbHM{} \tbPH{nìríì} \tbPS{n.~} \tbGE{a type of creeping plant with edible seeds} \tbTP{nirree} 
\tbLX{no} \tbHM{} \tbPH{nʊ̀} \tbPS{v.~} \tbGE{to grind a object by rubbing it against a surface} \tbPD{noɛ, noona, noona, noonema, noonaa, - } 
\tbLX{nomɔ} \tbHM{} \tbPH{nʊ́má} \tbPS{adj.~} \tbSN \tbGE{sweet, tasty} \tbXV{A tigiri bondirii da nomɔ la yaga.} \tbXE{The food at the party was really tasty.} \tbSN \tbGE{interesting} \tbXV{A senselloŋ da nomɔ la ka a bibiiri la paale zie zaa.} \tbXE{The story was interesting and so the children laughed off their heads.} \tbVA{noma}
\tbLX{noore} \tbHM{} \tbPH{nʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to slither (snake)} \tbSN \tbGE{to move aside; usually backwards and away from} \tbPD{noorɛɛ, noora, noora, nooreba, nooraa, nooreba } 
\tbLX{noore} \tbHM{} \tbPH{nʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to search, to look for} \tbPD{noorɛɛ, noora, noora, nooreba, nooraa, nooreba } 
\tbLX{noɔ} \tbHM{} \tbPH{nʊ̀ɔ́} \tbPS{n.~} \tbGE{happiness, joy} \tbXV{Pɔgebilii yaga maŋ teɛre ka noɔ yoŋ la be serekuli poɔ.} \tbXE{Most young women think that marriage carries with it only joy.} \tbOP{noore} \tbTP{noorɛɛ} 
\tbLX{noɔ} \tbHM{} \tbPH{nʊ́ɔ̀} \tbPS{n.~} \tbGE{fowl} \tbOP{noore} \tbTP{noorɛɛ} 
\tbLX{noɔ} \tbHM{} \tbPH{nʊ̀ɔ̀} \tbPS{v.~} \tbGE{to grow fat or large} \tbPD{noɔɛɛ, noɔnɔ, noɔnɔ, noɔnema, noonaa, - } 
\tbLX{noɔ} \tbHM{} \tbPH{nʊ̀ɔ́} \tbPS{adj.~} \tbGE{sweet} \tbXV{N lɛnne la a tii ka a waa noɔ lɛ.} \tbXE{I tasted the tea and it was very sweet.} \tbOP{noɔne} \tbTP{noɔnɛɛ} 
\tbLX{noɔ-noɔre} \tbHM{} \tbPH{nʊ́ɔ́-nʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{the habit of always eating} 
\tbLX{noɔre} \tbHM{} \tbPH{nʊ́ɔ́rè} \tbPS{n.~} \tbSN \tbGE{mouth} \tbSN \tbGE{beak, bill} \tbOP{nɔɛ} \tbTP{nɔrɛɛ} 
\tbLX{noɔre} \tbHM{} \tbPH{nʊ́ɔ́rè} \tbPS{n.~} \tbGE{edge} \tbOP{nɔɛ} \tbTP{nɔrɛɛ} 
\tbLX{noɔre} \tbHM{} \tbPH{nʊ́ɔ́rè} \tbPS{n.~} \tbGE{exact quantity} \tbOP{nɔɛ} \tbTP{nɔrɛɛ} 
\tbLX{noɔre} \tbHM{} \tbPH{nʊ́ɔ́rè} \tbPS{n.~} \tbGE{statement} \tbOP{nɔɛ} \tbTP{nɔrɛɛ} 
\tbLX{noɔre} \tbHM{} \tbPH{nʊ́ɔ́rè} \tbPS{n.~} \tbGE{the habit of talking a lot} \tbOP{nɔɛ} \tbTP{nɔrɛɛ} 
\tbLX{noɔre-emmo} \tbHM{} \tbPH{nʊ́ɔ́rɪ́-ɪ́mmʊ́} \tbPS{n.~} \tbGE{promise} 
\tbLX{nore} \tbHM{} \tbPH{nʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to bend inward, usually with force} \tbSN \tbGE{to hide something} \tbSN \tbGE{to appear without warning e.g. forcing a seed out of a fruit} \tbPD{norɛɛ, norɔ, norɔ, noreba, noraa, - } 
\tbLX{nɔbɛraa} \tbHM{} \tbPH{nɔ́bɛ́ráá} \tbPS{n.~} \tbGE{irritableness} \tbOP{nɔbɛrre} \tbTP{nɔbɛrɛɛ} 
\tbLX{nɔbinnaa} \tbHM{} \tbPH{nɔ́bìnnáá} \tbPS{n.~} \tbGE{agreement} \tbOP{nɔbinni} 
\tbLX{nɔbirime} \tbHM{} \tbPH{nɔ́bìrìmé} \tbPS{n.~} \tbGE{fowl with stunted feathers} \tbOP{nɔbirimeri} \tbTP{nɔbirimee} 
\tbLX{nɔboɔroŋ} \tbHM{} \tbPH{nɔ́bʊ́ɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{troubling another person; irritation} 
\tbLX{nɔboyi} \tbHM{} \tbPH{nɔ́bʊ́yì} \tbPS{n.~} \tbGE{double standards} 
\tbLX{nɔbɔŋɔ} \tbHM{} \tbPH{nɔ́bɔ́ŋɔ́} \tbPS{n.~} \tbGE{bit e.g. of a horse} \tbOP{nɔbɔnne} \tbTP{nɔbɔnnɛɛ} \tbVA{nɔbaŋa}
\tbLX{nɔdaa} \tbHM{} \tbPH{nɔ́dáá} \tbPS{n.~} \tbGE{rooster} \tbOP{nɔdaare} \tbTP{nɔdaarɛɛ} \tbVA{(1p) nɔraa}
\tbLX{nɔdaakommiruu} \tbHM{} \tbPH{nɔ́dáákómmírúú} \tbPS{n.~} \tbGE{a type of caterpillar, similar in appearance to a maggot, with a red head; typically found in the ground} \tbOP{nɔdaakommie} \tbTP{nɔdaakommiree} 
\tbLX{nɔdakpoŋ} \tbHM{} \tbPH{nɔ̀dàkpóŋ} \tbPS{n.~} \tbGE{capon} \tbOP{nɔdakponni} \tbTP{nɔdakpomo/nɔdakponnee} 
\tbLX{nɔdalee} \tbHM{} \tbPH{nɔ̀dàléé} \tbPS{n.~} \tbGE{cockerel} \tbOP{nɔdabilii} \tbTP{nɔdabillee} 
\tbLX{nɔdo} \tbHM{} \tbPH{nɔ́dó} \tbPS{n.~} \tbGE{food} 
\tbLX{nɔdɔɔne} \tbHM{} \tbPH{nɔ́dɔ̀ɔ̀nɪ́} \tbPS{n.~} \tbGE{that which is too little to satisfy (food)} \tbTP{nɔdɔɔnɛɛ} 
\tbLX{nɔ-ennaa} \tbHM{} \tbPH{nɔ́-ɪ̀nnáà} \tbPS{n.~} \tbGE{vow, promise} \tbOP{nɔ-enne} \tbTP{nɔ-ennɛɛ} 
\tbLX{nɔfaa} \tbHM{} \tbPH{nɔ́fáá} \tbPS{n.~} \tbGE{rudeness; foul mouth} \tbOP{nɔfaare} \tbTP{nɔfaarɛɛ} 
\tbLX{nɔfolee} \tbHM{} \tbPH{nɔ́fóléé} \tbPS{n.~} \tbGE{inability to keep a secret} 
\tbLX{nɔfo᷉ɔo} \tbHM{} \tbPH{nɔ́fʊ᷉ɔ́᷉ʊ́᷉} \tbPS{n.~} \tbGE{quick at talking} \tbOP{nɔfoɔne} \tbTP{nɔfoɔnɛɛ} 
\tbLX{nɔgeraa} \tbHM{} \tbPH{nɔ́gráá} \tbPS{n.~} \tbGE{sojourner} \tbOP{nɔgere} \tbTP{nɔgerɛɛ} 
\tbLX{nɔgo᷉o} \tbHM{} \tbPH{nɔ́gʊ̀᷉ʊ́᷉} \tbPS{n.~} \tbGE{moustache} \tbOP{nɔgoone} \tbTP{nɔgoonɛɛ} 
\tbLX{nɔgoolɔ} \tbHM{} \tbPH{nɔ̀gʊ̀ʊ̀lɔ̀} \tbPS{n.~} \tbGE{moustache} \tbVA{nɔgoola}
\tbLX{nɔgɔɔ} \tbHM{} \tbPH{nɔ́gɔ̀ɔ́} \tbPS{n.~} \tbGE{one-winged fowl} \tbOP{nɔgɔɔre} \tbTP{nɔgɔɔrɛɛ} 
\tbLX{nɔgbɛmbebe} \tbHM{} \tbPH{nɔ́gbɛ́mbébé} \tbPS{n.~} \tbGE{curse, foul language} 
\tbLX{nɔgbɛne} \tbHM{} \tbPH{nɔ́gbɛ́nɪ̀} \tbPS{n.~} \tbGE{lip} \tbSG{nɔgbɛnnoo} \tbOP{ nɔgbɛmɛ} \tbTP{nɔgbɛnnɛɛ} 
\tbLX{nɔhaaroŋ} \tbHM{} \tbPH{nɔ́háárʊ́ŋ} \tbPS{n.~} \tbGE{troublesomeness, pestering behaviour} \tbVA{nɔyaaroŋ}
\tbLX{nɔkɔnnoɔ} \tbHM{} \tbPH{[nɔ́kɔ̀nnʊ́ɔ́]} \tbPS{n.~} \tbGE{sweet talk} \tbOP{nɔkɔnnoore} \tbTP{nɔkɔnnoorɛɛ} 
\tbLX{nɔkɔɔloŋ} \tbHM{} \tbPH{nɔ́kɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{feathers of fowl} \tbTP{nɔkɔɔlɛɛ} 
\tbLX{nɔkɔrema} \tbHM{} \tbPH{nɔ́kɔ́rɪ́má} \tbPS{n.~} \tbGE{a type of fowl with feathers like those of a partridge} \tbOP{nɔkɔrre} \tbTP{nɔkɔrrɛɛ} 
\tbLX{nɔkpage} \tbHM{} \tbPH{nɔ́kpágɪ́} \tbPS{n.~} \tbGE{gluttonism} \tbVA{nɔkpogi}
\tbLX{nɔkpankyoɔne} \tbHM{} \tbPH{nɔ́kpánʧʊ́ɔ́nɪ́} \tbPS{n.~} \tbGE{a type of fowl with ash-coloured feathers} \tbOP{nɔkpankyoɔmɔ} \tbTP{nɔkpankyoɔnɛɛ} 
\tbLX{nɔkpeɛne} \tbHM{} \tbPH{nɔ́kpɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{argument} \tbTP{nɔkpeɛnɛɛ} 
\tbLX{nɔkpogi} \tbHM{} \tbPH{nɔ́kpógí} \tbPS{n.~} \tbGE{gluttonism} \tbOP{nɔkpogiri} \tbTP{nɔkpogiree} \tbVA{nɔkpage}
\tbLX{nɔlaŋ} \tbHM{} \tbPH{nɔ́làŋ} \tbPS{n.~} \tbGE{unity; agreement} 
\tbLX{nɔlee} \tbHM{} \tbPH{nɔ̀léé} \tbPS{n.~} \tbGE{chick} \tbOP{nɔbilii} \tbTP{nɔbillee} 
\tbLX{nɔlibie} \tbHM{} \tbPH{nɔ́líbíé} \tbPS{n.~} \tbGE{a type of fowl with spotted feathers} 
\tbLX{nɔll} \tbHM{} \tbPH{nɔ̀ll} \tbPS{adj.~} \tbGE{oval and large} 
\tbLX{nɔll} \tbHM{} \tbPH{nɔ́ll} \tbPS{adj.~} \tbGE{oval and small} 
\tbLX{nɔluuri} \tbHM{} \tbPH{nɔ́lùùrì} \tbPS{n.~} \tbGE{the act of renouncing an oath} 
\tbLX{nɔmaa} \tbHM{} \tbPH{nɔ́máà} \tbPS{n.~} \tbGE{surprise} \tbOP{nɔmaama} \tbTP{nɔmaanɛɛ} 
\tbLX{nɔmelime} \tbHM{} \tbPH{nɔ́mélímé} \tbPS{n.~} \tbGE{sweets} \tbOP{nɔmelimee} 
\tbLX{nɔmelime} \tbHM{} \tbPH{nɔ́mélímé} \tbPS{adj.~} \tbGE{craving for sweet things} 
\tbLX{nɔmm} \tbHM{} \tbPH{nɔ̀mm} \tbPS{adv.~} \tbGE{at rest; without movement} 
\tbLX{nɔmmo} \tbHM{} \tbPH{nɔ́mmʊ́} \tbPS{n.~} \tbGE{love; strong affection or likeness for} 
\tbLX{nɔmoɔnaa} \tbHM{} \tbPH{nɔ́mʊ́ɔ́náá} \tbPS{n.~} \tbGE{lips decorated with red lipstick} \tbOP{nɔmoɔne/nɔmoɔmɔ} \tbTP{nɔmoɔnɛɛ} 
\tbLX{nɔmuri} \tbHM{} \tbPH{nɔ́múrí} \tbPS{n.~} \tbGE{a small, smooth stick fixed in a hole made in the lip} \tbOP{nɔmue} \tbTP{nɔmuuree} 
\tbLX{nɔmuu} \tbHM{} \tbPH{nɔ́mùú} \tbPS{n.~} \tbGE{dumbness} \tbOP{nɔmuuni} \tbTP{nɔmuunee} 
\tbLX{nɔnɛne} \tbHM{} \tbPH{nɔ́nɛ́nɪ̀} \tbPS{n.~} \tbGE{chicken (meat)} \tbSG{nɔnɛnoo} \tbOP{nɔnɛmɛ} \tbTP{nɔnɛnnɛɛ} 
\tbLX{nɔnne} \tbHM{} \tbPH{nɔ̀nnɪ́} \tbPS{n.~} \tbGE{love, affection} \tbTP{nɔnnɛɛ} 
\tbLX{nɔnne boore} \tbHM{} \tbPH{nɔ̀nnɪ́ bʊ̀ʊ̀rɪ́} \tbPS{n.~} \tbGE{the act of taking sides} 
\tbLX{nɔnoɔ} \tbHM{} \tbPH{nɔ́nʊ̀ɔ́} \tbPS{n.~} \tbGE{convincing talk; sweet mouth} \tbOP{nɔnoore} \tbTP{nɔnoorɛɛ} 
\tbLX{nɔnɔ} \tbHM{} \tbPH{nɔ́nɔ́} \tbPS{n.~} \tbGE{lover, one who likes} \tbOP{nɔnema} \tbTP{nɔnemɛɛ} \tbVA{nɔna}
\tbLX{nɔnɔmaa} \tbHM{} \tbPH{nɔ̀nɔ̀màà} \tbPS{n.~} \tbGE{a type of tree with medicinal properties and edible fruits} \tbOP{nɔnɔmaɛ} 
\tbLX{nɔnyaŋaa} \tbHM{} \tbPH{nɔ́ɲáŋáá} \tbPS{n.~} \tbGE{adult female hen (one that has hatched chicks)} \tbOP{nɔnyaŋene} \tbTP{nɔnyaŋenɛɛ} 
\tbLX{nɔnyeɛraa} \tbHM{} \tbPH{ńɔɲɪ́ɛ́ráà} \tbPS{n.~} \tbGE{laying fowl} \tbOP{nɔnyeɛre} \tbTP{nɔnyeɛrɛɛ} 
\tbLX{nɔnyemɛ} \tbHM{} \tbPH{nɔ́ɲɪ́mɛ́} \tbPS{n.~} \tbGE{disturbing sharp utterance e.g. barking puppy} \tbTP{nɔnyemɛɛ} 
\tbLX{nɔŋe} \tbHM{} \tbPH{nɔ́ŋ} \tbPS{v.~} \tbSN \tbGE{to massage} \tbSN \tbGE{to make soft by applying substances and wringing out (leather)} \tbSN \tbGE{to knead} \tbPD{nɔŋɛɛ, nɔnɔ, nɔnɔ, nɔnema, nɔnaa, - } 
\tbLX{nɔŋe} \tbHM{} \tbPH{nɔ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to like, to love} \tbSN \tbGE{to prefer} \tbPD{nɔŋɛɛ, nɔnɔ, nɔnɔ, nɔnema, nɔnaa, nɔnema } 
\tbLX{nɔɔ} \tbHM{} \tbPH{nɔ́ɔ́} \tbPS{adv.~} \tbGE{very one, exactly} \tbXV{O ne o zu nɔɔ la da wa a kyɛ.} \tbXE{He is the very one who came here.} 
\tbLX{nɔɔfeele} \tbHM{} \tbPH{nɔ̀ɔ̀fɪ́ɪ́lɪ́} \tbPS{n.~} \tbGE{a hard, sharp nail on the back of the leg of a full-grown male fowl} \tbOP{nɔɔfeele,} \tbTP{nɔɔfeelɛɛ} 
\tbLX{nɔɔganaa} \tbHM{} \tbPH{nɔ̀ɔ̀gànàá} \tbPS{n.~} \tbGE{big vein running from the heel to the calf} \tbOP{nɔɔganne} \tbTP{nɔɔganɛɛ} 
\tbLX{nɔɔkyoone} \tbHM{} \tbPH{nɔ̀ɔ̀kyʊ́ʊ́nɪ́} \tbPS{n.~} \tbGE{ankle} \tbOP{nɔɔkyooma} \tbTP{nɔɔkyoonɛɛ} 
\tbLX{nɔɔpelaa} \tbHM{} \tbPH{nɔ́ɔ́pɪ́láá} \tbPS{n.~} \tbGE{abscess on the sole of the foot} 
\tbLX{nɔɔsaŋkpaare} \tbHM{} \tbPH{nɔ̀ɔ̀sàŋkpáàrɪ́} \tbPS{n.~} \tbSN \tbGE{heel} \tbSN \tbGE{hind toe of a bird} \tbOP{nɔɔsaŋkpaɛ} \tbTP{nɔɔsaŋkpaarɛɛ} 
\tbLX{nɔɔtemiri} \tbHM{} \tbPH{nɔ̀ɔ̀tɪ́mírì} \tbPS{n.~} \tbGE{shoelaces} \tbOP{nɔɔtemie} \tbTP{nɔɔtemiiree} 
\tbLX{nɔɔtere} \tbHM{} \tbPH{nɔ̀ɔ̀tɪ́rɪ̀} \tbPS{n.~} \tbGE{footwear, sandal, shoe} \tbSG{nɔɔteroo} \tbOP{nɔɔteɛ} \tbTP{nɔɔteerɛɛ} 
\tbLX{nɔɔtere} \tbHM{} \tbPH{nɔ́ɔ́tɪ́rɪ́} \tbPS{n.~} \tbGE{boil, abscess} \tbSG{nɔɔteroo} \tbOP{nɔɔteɛ} \tbTP{nɔɔteerɛɛ} 
\tbLX{nɔpollaa} \tbHM{} \tbPH{nɔ́pʊ́lláà} \tbPS{n.~} \tbGE{vow, promise} \tbOP{nɔpolle} \tbTP{nɔpollɛɛ} 
\tbLX{nɔpɔllee} \tbHM{} \tbPH{nɔ́pɔ́lléé} \tbPS{n.~} \tbGE{young fowl} \tbOP{nɔpɔlbilii} \tbTP{nɔpɔlbillee} \tbVA{nɔpɔle; nɔpɔlɔ}
\tbLX{nɔre} \tbHM{} \tbPH{nɔ̀rɪ̀} \tbPS{v.~} \tbGE{to be lucky} \tbXV{Zenɛ karembiiri banaŋ naŋ ba wa sakuuri wiẽu nɔrɛɛ la ne a karema potuo na naŋ ba wa sakuuri, o da na eŋe  ba zẽɛ.} \tbXE{Today, those pupils who came to school late were lucky that wicked teacher did not come to school since he would have punished them severely.} \tbPD{nɔrɛɛ, nɔrɔ, nɔrɔ, nɔreba, nɔraa, - } 
\tbLX{nɔsare} \tbHM{} \tbPH{nɔ́sárɪ́} \tbPS{n.~} \tbGE{a hen trough} \tbVA{nɔsɛre}
\tbLX{nɔsɛre} \tbHM{} \tbPH{nɔ́sɛ́rɪ́} \tbPS{n.~} \tbGE{drinking water container for fowl} \tbOP{nɔsaɛ} \tbTP{nɔsɛrɛɛ} \tbVA{nɔsare}
\tbLX{nɔtɛɛtɛɛ} \tbHM{} \tbPH{nɔ́tɛ́ɛ́tɛ̀ɛ̀} \tbPS{} \tbGE{incompatible statements; usually made by the same person} 
\tbLX{nɔtɛnzuŋi} \tbHM{} \tbPH{nɔ́tɛ́nzúŋ} \tbPS{n.~} \tbGE{pursed lips; protruding lips to show anger, annoyance} \tbOP{nɔtɛnzunni} \tbTP{nɔtɛnzunee/nɔtɛnzunee} 
\tbLX{nɔtore} \tbHM{} \tbPH{nɔ́tʊ́rɪ̀} \tbPS{n.~} \tbGE{the act of pressing the lips together (sign of annoyance)} 
\tbLX{nɔtoro} \tbHM{} \tbPH{nɔ̀tòrò} \tbPS{n.~} \tbGE{moustache} \tbOP{nɔtorri} \tbTP{nɔtoree} 
\tbLX{nɔtuo} \tbHM{} \tbPH{nɔ́túó} \tbPS{n.~} \tbGE{abusiveness; rudeness of language} 
\tbLX{nɔtuo} \tbHM{} \tbPH{nɔ́túó} \tbPS{n.~} \tbGE{hungering} 
\tbLX{nɔ-uunaa} \tbHM{} \tbPH{nɔ́-úúnáá} \tbPS{n.~} \tbGE{brooding fowl} \tbOP{nɔ-uuni} \tbTP{nɔ-uunee} 
\tbLX{nɔvelaa} \tbHM{} \tbPH{nɔ́vɪ́láá} \tbPS{n.~} \tbGE{encouraging talk} 
\tbLX{nɔwogi} \tbHM{} \tbPH{nɔ́wógì} \tbPS{n.~} \tbGE{dog} \tbOP{nɔwogri} \tbTP{nɔwogree} 
\tbLX{nɔwɔɔ} \tbHM{} \tbPH{nɔ́wɔ̀ɔ́} \tbPS{n.~} \tbGE{abrassiveness} \tbOP{nɔwɔɔre} \tbTP{nɔwɔɔrɛɛ} \tbVA{nɔtuo}
\tbLX{nɔyaaroŋ} \tbHM{} \tbPH{nɔ́yáárʊ́ŋ} \tbPS{n.~} \tbGE{trouble causing, pestering} \tbVA{nɔyaaroŋ}
\tbLX{nɔyoɔre} \tbHM{} \tbPH{nɔ́yʊ̀ɔ̀rɪ̀} \tbPS{n.~} \tbGE{pestering e.g. a little child} 
\tbLX{nɔzage} \tbHM{} \tbPH{nɔ́zàgɪ́} \tbPS{n.~} \tbGE{pen for hens} \tbOP{nɔzagere} \tbTP{nɔzagerɛɛ} 
\tbLX{nɔzagere} \tbHM{} \tbPH{nɔ́zàgɪ̀rɪ̀} \tbPS{n.~} \tbGE{disobedience} 
\tbLX{nɔzeɛ} \tbHM{} \tbPH{nɔ́zɪ́ɛ́} \tbPS{n.~} \tbGE{red mouth} 
\tbLX{nɔzoolɔ} \tbHM{} \tbPH{nɔ́zʊ́ʊ́lɔ́} \tbPS{n.~} \tbGE{fringe} \tbOP{nɔzoolɛɛ} \tbVA{nɔzoola}
\tbLX{nɔzɔbɔ} \tbHM{} \tbPH{nɔ́zɔ̀bɔ́} \tbPS{n.~} \tbGE{the act of pestering someone to annoy that person} \tbVA{nɔzɔba}
\tbLX{nsɛ} \tbHM{} \tbPH{nsɛ̀} \tbPS{n.~} \tbSN \tbGE{Mr} \tbSN \tbGE{great one} 
\tbLX{nu} \tbHM{} \tbPH{nû} \tbPS{n.~} \tbGE{hand, arm} \tbSG{nuuruu} \tbOP{nuuri} \tbTP{nuuree} 
\tbLX{nubaŋa} \tbHM{} \tbPH{núbáŋá} \tbPS{n.~} \tbGE{bangle for the wrist} \tbOP{nubanne} \tbTP{ nubannɛɛ.} \tbVA{nu bɔŋɔ}
\tbLX{nubidaa} \tbHM{} \tbPH{núbìdáà} \tbPS{n.~} \tbGE{thumb} \tbOP{nubidaare} \tbTP{nubidaarɛɛ} 
\tbLX{nubidori} \tbHM{} \tbPH{núbìdórí} \tbPS{n.~} \tbGE{finger joint; knuckle} \tbOP{nubidorri} \tbTP{nubidoree} 
\tbLX{nubi-eɛne} \tbHM{} \tbPH{núbì-ɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{fingernail} \tbOP{nubi-eɛmɛ} \tbTP{nubi-eɛnɛɛ} \tbVA{nubi-e᷉ɛ/nubi-e᷉mɛ}
\tbLX{nubilee} \tbHM{} \tbPH{núbìléé} \tbPS{n.~} \tbGE{smallest finger, pinky} \tbOP{nubibilii} \tbTP{nubilbillee} 
\tbLX{nubiri} \tbHM{} \tbPH{núbìrì} \tbPS{n.~} \tbGE{finger} \tbOP{nubie/nubiiri} \tbTP{nubiiree} 
\tbLX{nudɛgere} \tbHM{} \tbPH{núdɛ́gɪ́rɪ́} \tbPS{n.~} \tbGE{dirty hand} \tbOP{nudɛgɛ} \tbTP{nudɛgrɛɛ} 
\tbLX{nudoloŋ} \tbHM{} \tbPH{núdʊ̀lʊ̂ŋ} \tbPS{n.~} \tbSN \tbGE{right hand} \tbSN \tbGE{to the right (direction)} \tbOP{nudolle} \tbTP{nudollɛɛ} \tbVA{doloŋ}
\tbLX{nufoori} \tbHM{} \tbPH{núfóórì} \tbPS{n.~} \tbGE{closed empty hand} \tbXV{A bie ũũ la libiri o nu kaŋa poɔ kyɛ yeli ka N iri kaŋa ka N ŋmɛ a nugɔɔ, ka o yuoro ka nufoori la.} \tbXE{The girl hid a coin in one of her closed hands and asked me to choose one, and when I chose her left hand, it was empty.} \tbOP{nufoe} \tbTP{nuforee} 
\tbLX{nufoɔ᷉o} \tbHM{} \tbPH{núfʊ́᷉ɔ́᷉ʊ́᷉} \tbPS{n.~} \tbGE{ability to act fast with the hand} \tbOP{nufoɔne} \tbTP{nufoɔnɛɛ} 
\tbLX{nugɔɔ} \tbHM{} \tbPH{núgɔ̀ɔ̀} \tbPS{n.~} \tbGE{left hand} \tbOP{nugɔɔre} \tbTP{nugɔɔrɛɛ} \tbVA{gɔɔ}
\tbLX{nugɔɔ} \tbHM{} \tbPH{núgɔ̀ɔ̀} \tbPS{postposition.~} \tbGE{towards the left direction} \tbVA{gɔɔ}
\tbLX{nugunno} \tbHM{} \tbPH{núgúnnó} \tbPS{n.~} \tbGE{very poor, helpless} \tbTP{nugunnee} 
\tbLX{nukuni} \tbHM{} \tbPH{núkúní} \tbPS{n.~} \tbGE{fist} \tbOP{nukumo} \tbTP{nukunee} 
\tbLX{nukuoŋaa} \tbHM{} \tbPH{núkúóŋáá} \tbPS{n.~} \tbGE{dry hand} \tbOP{nukuoni} \tbTP{nukuonee} 
\tbLX{nukpara} \tbHM{} \tbPH{núkpárá} \tbPS{n.~} \tbGE{both hands put together} \tbOP{nukparre} \tbTP{nukparrɛɛ} 
\tbLX{nulaama} \tbHM{} \tbPH{núláámá} \tbPS{n.~} \tbGE{idle hands} 
\tbLX{nulaayɛlɛ} \tbHM{} \tbPH{núlàáyɛ́lɛ} \tbPS{n.~} \tbGE{work of idle hands} 
\tbLX{nuperaa} \tbHM{} \tbPH{núpɪ́ràà} \tbPS{n.~} \tbGE{finger snap} \tbOP{nuperre} \tbTP{nuperrɛɛ} 
\tbLX{nupoɔ} \tbHM{} \tbPH{núpʊ́ɔ́} \tbPS{n.~} \tbGE{palm (of hand)} \tbOP{nupoore} \tbTP{nupoorɛɛ} 
\tbLX{nupuori} \tbHM{} \tbPH{núpúórì} \tbPS{n.~} \tbGE{back of the hand} \tbOP{nupoe} \tbTP{nuporee} 
\tbLX{nusampane} \tbHM{} \tbPH{núsámpánɪ́} \tbPS{n.~} \tbGE{palm (of hand)} \tbOP{nusampama} \tbTP{nusampanɛɛ} 
\tbLX{nusugili} \tbHM{} \tbPH{núsúgílí} \tbPS{n.~} \tbGE{wrist} \tbOP{nusugilo} \tbTP{nusugilee} 
\tbLX{nutaa} \tbHM{} \tbPH{nútàà} \tbPS{n.~} \tbGE{wealthiness} 
\tbLX{nutale} \tbHM{} \tbPH{nútálɪ̀} \tbPS{n.~} \tbGE{palm (of hand)} \tbOP{nutala} \tbTP{nutalɛɛ} 
\tbLX{nutara} \tbHM{} \tbPH{nútárá} \tbPS{n.~} \tbGE{wealthy person} \tbOP{nutareba} \tbTP{nutarebɛɛ} 
\tbLX{nute᷉ɛ} \tbHM{} \tbPH{nútɪ́᷉ɛ́᷉} \tbPS{n.~} \tbGE{the act of giving, benevolence} 
\tbLX{nuu} \tbHM{} \tbPH{núú} \tbPS{adj.~} \tbGE{five} \tbOP{nuuri} \tbTP{nuuree} 
\tbLX{nuulee} \tbHM{} \tbPH{núúléé} \tbPS{n.~} \tbGE{bird} \tbOP{nuuli} 
\tbLX{nuuleemako᷉ɔ᷉} \tbHM{} \tbPH{núúléémákʊ̀᷉ɔ̀᷉} \tbPS{n.~} \tbGE{a type of small creeping plant} \tbOP{nuuleemakoɔnɛɛ} \tbVA{nuuleemako᷉ɔ᷉ko᷉ɔ᷉}
\tbLX{nuuleemako᷉ɔko᷉ɔ} \tbHM{} \tbPH{núúléémákʊ́᷉ɔ́kʊ́᷉ɔ́} \tbPS{n.~} \tbGE{a type of plant with medicinal properties} 
\tbLX{nuulikɔɔloŋ} \tbHM{} \tbPH{nùúlìkɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{feather} \tbOP{nuulikɔɔlo} 
\tbLX{nuulikyoo} \tbHM{} \tbPH{nùúlìʧóò} \tbPS{n.~} \tbGE{bird nest} \tbOP{nuulikyoori} \tbTP{nuulikyooree} 
\tbLX{nuuliparezoore} \tbHM{} \tbPH{núúlìpárɪ̀zʊ́ʊ́rɪ́n} \tbPS{n.~} \tbGE{pintailed bird} \tbOP{nuuliparezoɛ} \tbTP{nuuliparezoorɛɛ} 
\tbLX{nuulitige} \tbHM{} \tbPH{núùlitígé} \tbPS{n.~} \tbGE{flock (of birds) -- singular or plural agreement} \tbTP{nuulitigree} 
\tbLX{nuwogi} \tbHM{} \tbPH{núwógí} \tbPS{n.~} \tbSN \tbGE{light-fingeredness} \tbSN \tbGE{thief} \tbOP{nuwogiri} \tbTP{nuwogiree} 
\tbLX{nuyoɔre} \tbHM{} \tbPH{núyʊ̀ɔ̀rɪ̀} \tbPS{n.~} \tbGE{something that causes boredom, inactivity} 
\tbLX{nu-yuo} \tbHM{} \tbPH{nú-yùò} \tbPS{n.~} \tbGE{generosity} \tbOP{nu-yuori} \tbTP{nu-yuoree} 
\tbLX{nuzagele-pie} \tbHM{} \tbPH{núzáglɪ̀-píé} \tbPS{n.~} \tbGE{empty handedness} 
\tbLX{nuzɛgere} \tbHM{} \tbPH{núzɛ́grɪ́} \tbPS{n.~} \tbGE{handshake} 
\tbLX{nyaa} \tbHM{} \tbPH{ɲáà} \tbPS{n.~} \tbGE{chest} \tbOP{nyaɛ} \tbTP{nyaane/nyaanɛɛ} 
\tbLX{nyaa} \tbHM{} \tbPH{ɲàà} \tbPS{v.~} \tbSN \tbGE{to break e.g. light of day} \tbSN \tbGE{to recover oneself; to come round} \tbSN \tbGE{to awaken} \tbPD{nyaaɛ, nyaana, nyaana, nyaanema, nyaanaa, - } 
\tbLX{nyaa} \tbHM{} \tbPH{ɲàà} \tbPS{adj.~} \tbGE{wise, crafty} 
\tbLX{nyaaba} \tbHM{} \tbPH{ɲáábà} \tbPS{n.~} \tbGE{disgrace, shame} \tbTP{nyaabɛɛ} \tbVA{nyaama}
\tbLX{nyaalbini} \tbHM{} \tbPH{ɲààlbíní} \tbPS{n.~} \tbGE{a disease that causes itchy rashes} \tbTP{nyaalbinee} 
\tbLX{nyaalbiri} \tbHM{} \tbPH{ɲààlbírí} \tbPS{n.~} \tbGE{guinea worm (the worm)} \tbOP{nyaalbie} \tbTP{nyaalbiiree} 
\tbLX{nyaale} \tbHM{} \tbPH{ɲáálɪ̀} \tbPS{v.~} \tbGE{to move fire closely over a surface or move an object closely over fire} \tbPD{nyaalɛɛ, nyaala, nyaala, nyaaleba, nyaalaa, - } 
\tbLX{nyaalenyieli} \tbHM{} \tbPH{ɲààlɪ̀nyíélì} \tbPS{v.~} \tbGE{to walk with buttocks thrust backward and legs apart (because of sores on the buttocks or groin)} \tbPD{nyaalenyielee, nyaalenyiele, nyaalenyiele, nyaalenyielibe, nyaalenyielaa, - } 
\tbLX{nyaalombiri} \tbHM{} \tbPH{ɲáàlʊ̂mbírì} \tbPS{n.~} \tbGE{guinea worm (the worm)} \tbOP{nyaalombie} \tbTP{nyaalombiiree} 
\tbLX{nyaalompoɔ} \tbHM{} \tbPH{ɲáálʊ́mpʊ́ɔ́} \tbPS{n.~} \tbGE{undeveloped or dormant guinea-worm disease} \tbOP{nyaalompoore} \tbTP{nyaalompoore} 
\tbLX{nyaaloŋ} \tbHM{} \tbPH{ɲààlʊ́ŋ} \tbPS{n.~} \tbGE{guinea worm disease} \tbOP{nyiile} \tbTP{nyiilee} 
\tbLX{nyaanaa} \tbHM{} \tbPH{ɲàànáà} \tbPS{n.~} \tbGE{craft, cunning} \tbOP{nyaane} 
\tbLX{nyaane} \tbHM{} \tbPH{ɲààné} \tbPS{n.~} \tbSN \tbGE{wit} \tbSN \tbGE{wisdom} \tbTP{nyaanɛɛ} 
\tbLX{nyaanema} \tbHM{} \tbPH{ɲàànɪ́mà} \tbPS{n.~} \tbGE{cat} \tbOP{nyaanemamine} \tbTP{nyaanemɛɛ} 
\tbLX{nyaanoo} \tbHM{} \tbPH{ɲáánɔ́ɔ́} \tbPS{adv.~} \tbGE{at once} \tbXV{A wuo la a boma nyaanɔɔ ka saa na poɔ a.} \tbXE{S/he collected the things at once to prevent them being soaked by the rain.} 
\tbLX{nyaanyieli} \tbHM{} \tbPH{ɲàànyíélì} \tbPS{v.~} \tbGE{to walk with legs apart (usually because of sores on the buttocks)} \tbPD{nyaanyielee, nyaalanyiele, nyaanyiele, nyaanyielibe, nyaanyielaa, - } 
\tbLX{nyaanyoo} \tbHM{} \tbPH{ɲáányʊ́ʊ́} \tbPS{adv.~} \tbSN \tbGE{dim} \tbSN \tbGE{light (weight)} 
\tbLX{nyaaraa} \tbHM{} \tbPH{ɲááráá} \tbPS{n.~} \tbSN \tbGE{rake, fork;} \tbSN \tbGE{a type of small plant that grows on waste land} \tbSN \tbGE{bean soup, soup without okra} \tbOP{nyaare} \tbTP{nyaarɛɛ} \tbVA{gbaalaa}
\tbLX{nyaaraa} \tbHM{} \tbPH{ɲááráá} \tbPS{n.~} \tbGE{broom made from this plant} \tbOP{nyaare} \tbTP{nyarɛɛ} 
\tbLX{nyaare} \tbHM{} \tbPH{ɲààrɪ̀} \tbPS{v.~} \tbGE{to grind roughly e.g. in order to peel the husks off seeds} \tbPD{nyaarɛɛ, nyaara, nyaara, nyaareba, nyaaraa, - } 
\tbLX{nyaatafale} \tbHM{} \tbPH{ɲáátáfálɪ́} \tbPS{n.~} \tbGE{emptiness} 
\tbLX{nyabipeɛ} \tbHM{} \tbPH{ɲábìpɪ́ɛ́} \tbPS{n.~} \tbGE{ribcage} \tbOP{nyabipɛre} \tbTP{nyabipɛrɛɛ} 
\tbLX{nyabiri} \tbHM{} \tbPH{ɲábìrí} \tbPS{n.~} \tbSN \tbGE{rib} \tbSN \tbGE{a disease that causes a cramp in the chest} \tbOP{nyabie} \tbTP{nyabiree/nyabiiree} 
\tbLX{nyage} \tbHM{} \tbPH{ɲágɪ́} \tbPS{v.~} \tbGE{to become very angry, to become wild} \tbPD{nyagɛɛ, nyagra, nyagra, nyagreba, nyagraa, - } 
\tbLX{nyagele} \tbHM{} \tbPH{ɲàgɪ̀lɪ̀} \tbPS{v.~} \tbGE{to be entangled in spiky material (thorns)} \tbPD{nyaglɛɛ, nyagla, nyagla, nyagleba, nyaglaa, - } 
\tbLX{nyagele} \tbHM{} \tbPH{ɲàgɪ̀lɪ̀} \tbPS{v.~} \tbGE{to draw arms in a bid to fight (typically used with kpaŋkpama)} \tbPD{nyaglɛɛ, nyagla, nyagla, nyagleba, nyaglaa, - } 
\tbLX{nyagere} \tbHM{} \tbPH{ɲágɪ́rɪ́} \tbPS{n.~} \tbSN \tbGE{bowel} \tbSN \tbGE{tree root} \tbOP{nyaga} \tbTP{nyagrɛɛ} 
\tbLX{nyagere} \tbHM{} \tbPH{ɲàgɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{bite suddenly and sharply (dog bite)} \tbXV{A baa digi te nyɔge la a nannyige gbɛre e nygere o naŋ doŋ.} \tbXE{The dog ran after the thief and bit tore at his leg instantly.} \tbSN \tbGE{get stimulated} \tbXV{Saa ba miire azuiŋ n eŋɛ ba nyagera ne weɛ gaabo.} \tbXE{It does not rain and so I am not encouraged to go to the farm.} \tbPD{nyagrɛɛ, nyagra, nyagra, nyagreba, nyagraa, - } 
\tbLX{nyageroŋ} \tbHM{} \tbPH{ɲágrʊ́ŋ} \tbPS{adv.~} \tbSN \tbGE{quickly} \tbXV{Kyɛŋe nyageroŋ ka te gaa wi᷉eu.} \tbXE{Walk fast so that we can get there quickly.} \tbSN \tbGE{intelligently} \tbXV{A pɔge da e la nyãgeroŋ ne o naŋ tage a dɔɔ yi a soriŋ, ka lɛ naane a lɔɔre da na ŋmɛ o la.} \tbXE{The woman was smart by pulling the man out of the way otherwise the lorry would have hit him.} 
\tbLX{nyagesaga} \tbHM{} \tbPH{ɲágɪ́ságá} \tbPS{adj.~} \tbGE{jumbled, dry, and bare e.g. metal scrap, bones} \tbXV{A kuridaare e la nyagesaga lɛ a davorikpoŋi poɔ.} \tbXE{The scrap of metal was jumbled up in the yard.} 
\tbLX{nyakpuli} \tbHM{} \tbPH{ɲákpúlí} \tbPS{n.~} \tbGE{sternum} \tbOP{nyakpulo} \tbTP{nyakpulee} 
\tbLX{nyale} \tbHM{} \tbPH{ɲàlɪ́} \tbPS{adv.~} \tbGE{suddenly, usually used to describe the sudden manner some animals excrete} \tbXV{Ba ŋmɛ la a bagao ka o nyɛ bini nyale.} \tbXE{They hit the mad dog and it defecated suddenly.} 
\tbLX{nyalle} \tbHM{} \tbPH{ɲàllɪ̀} \tbPS{v.~} \tbGE{to eat with very little sauce} \tbPD{nyallɛɛ, nyalla, nyalla, nyaleba, nyallaa, - } 
\tbLX{nyam} \tbHM{} \tbPH{ɲàm} \tbPS{v.~} \tbGE{to take heart, to be patient} \tbPD{nyamɛɛ, nyamena, nyamena, nyamena, nyamenaa, - } 
\tbLX{nyama} \tbHM{} \tbPH{ɲàmà} \tbPS{adv.~} \tbGE{gripping suddenly} \tbXV{A kɔkɔlaa vare guri la a gyaŋgboŋ lɛ nyama.} \tbXE{The cat jumped on the mouse, gripping it firmly.} 
\tbLX{nyamaa} \tbHM{} \tbPH{ɲàmàà} \tbPS{adv.~} \tbGE{mixed up, congested} 
\tbLX{nyamama} \tbHM{} \tbPH{ɲámámá} \tbPS{adv.~} \tbGE{rising in intensity (pain)} \tbXV{Ka gyaŋgboŋiŋ wa peɛle fo kyɛ doŋ a be maŋ buori nyamama lɛ.} \tbXE{When a mouse gnaws at your foot, the pain rises slowly.} 
\tbLX{nyambaraŋ} \tbHM{} \tbPH{ɲàmbàràŋ} \tbPS{adj.~} \tbGE{itchy} \tbXV{Ka neɛŋ puri wale eŋe ka-uri poɔ, o eŋɛ maŋ waa la nyambaraŋ lɛ.} \tbXE{If you sweat on top of millet chaff your body is normally itchy.} 
\tbLX{nyaŋ} \tbHM{} \tbPH{ɲáŋ} \tbPS{v.~} \tbGE{to age, to become old} \tbPD{nyaŋɛɛ, nyaŋena, nyaŋena, nyaŋenema, nyaŋenaa, - } 
\tbLX{nyaŋe} \tbHM{} \tbPH{ɲáŋ} \tbPS{adj.~} \tbGE{hasty} \tbXV{Kyɛŋ gbɛnyaŋe ka te gaa, zie soɔrɔ la.} \tbXE{Walk fast; let's go, for the night is falling on us.} 
\tbLX{nyaŋema} \tbHM{} \tbPH{ɲáŋímà} \tbPS{n.~} \tbGE{accident} \tbTP{nyaŋemɛɛ} 
\tbLX{nyaŋgalaŋ} \tbHM{} \tbPH{ɲàŋgàlàŋ} \tbPS{adv.~} \tbGE{thick and stiff} 
\tbLX{nyaraa-nyaraa} \tbHM{} \tbPH{ɲárááɲáráá} \tbPS{adj.~} \tbGE{with small pointed appendages e.g. crab} 
\tbLX{nyaranyara} \tbHM{} \tbPH{ɲàràɲàrà} \tbPS{adj.~} \tbGE{rough, rugged e.g. stony ground} 
\tbLX{nyarata} \tbHM{} \tbPH{ɲárátá} \tbPS{adv.~} \tbGE{swiftly e.g. movement of lizard} \tbVA{nyaratatata}
\tbLX{nyaremaŋ} \tbHM{} \tbPH{ɲàrɪ̀màŋ} \tbPS{adj.~} \tbGE{hot e.g. food} 
\tbLX{nyarewara} \tbHM{} \tbPH{ɲàrɪ̀wàrà} \tbPS{adj.~} \tbGE{grainy, stony e.g. unplastered floor} 
\tbLX{nyaroo} \tbHM{} \tbPH{ɲárʊ́ʊ́} \tbPS{adj.~} \tbGE{small and lean} 
\tbLX{nyawele} \tbHM{} \tbPH{ɲåwɪ́lɪ́} \tbPS{n.~} \tbGE{one that is the greatest} \tbOP{nyawelɛ} \tbTP{nyawelɛɛ} 
\tbLX{nyeɛ} \tbHM{} \tbPH{ɲɪ̀ɛ́} \tbPS{n.~} \tbGE{snot} \tbTP{nyeɛnɛɛ} 
\tbLX{nyeɛne} \tbHM{} \tbPH{ɲɪ̀ɛ́nɪ́} \tbPS{n.~} \tbSN \tbGE{bulb, tuber} \tbSN \tbGE{descendant of mother's line} \tbOP{nyeɛmɛ} \tbTP{nyeɛnɛɛ} 
\tbLX{nyeɛre} \tbHM{} \tbPH{ɲɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to test, to try out, to pilot} \tbPD{nyeɛrɛɛ, nyeɛrɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\tbLX{nyeɛre} \tbHM{} \tbPH{ɲɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to pay for the cost of herbs after treatment} \tbPD{nyeɛrɛɛ, nyeɛrɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\tbLX{nyegenyege} \tbHM{} \tbPH{ɲɪ̀gɪ̀ɲɪ̀gɪ̀} \tbPS{n.~} \tbGE{restlessness} \tbXV{O taa la nyegenyege zuo, o na ŋmare la ŋmane.} \tbXE{His restlessness is great; he will break the calabsh.} 
\tbLX{nyegenyege} \tbHM{} \tbPH{ɲɪ̀gɪ̀ɲɪ̀gɪ̀} \tbPS{adj.~} \tbGE{restless, restive} \tbXV{O waa la nyegenyege lɛnso noba yaga zoro o la.} \tbXE{He is very restive, so many people fear him.} 
\tbLX{nyegere} \tbHM{} \tbPH{ɲɪ̀grɪ̀} \tbPS{v.~} \tbGE{to be scarce} \tbXV{Yuoni ŋa boŋkoɔre yaga ba maale azuiŋ bondirii na nyegere la.} \tbXE{This year's crop yield was poor and so food will be scarce.} \tbPD{-, nyegrɛ, -, -, -, - } 
\tbLX{nyegere} \tbHM{} \tbPH{ɲɪ̀grɪ̀} \tbPS{v.~} \tbGE{to massage with a hot, dry object} \tbXV{Daa daare kyɔge la a koɔraa gbɛre ka ba voɔ kyɛ nyegere.} \tbXE{A thorn pierced and lodged itself in the farmer's foot, and after removing it they used hot water to massage it.} \tbPD{-, nyegrɛ, -, -, -, - } 
\tbLX{nyegere} \tbHM{} \tbPH{ɲɪ́grɪ́} \tbPS{v.~} \tbSN \tbGE{to flash e.g. lightening} \tbSN \tbGE{give way momentarily (limb as a result of sharp twist)} \tbXV{A dɔɔ gbɛbaala na nyegerɛɛ la ka o le.} \tbXE{The man's wounded leg made a snapping noise and he fell.} \tbPD{nyegrɛɛ, nyegrɛ, nyegrɛ, nyegrebɛ, nyegraa, - } 
\tbLX{nyegɛnyegɛ} \tbHM{} \tbPH{ɲɪ́gɛ́ɲɪ́gɛ́} \tbPS{adj.~} \tbGE{very sharp and shiny e.g. blade} 
\tbLX{nyelaa} \tbHM{} \tbPH{ɲɪ̀ràà} \tbPS{adj.~} \tbGE{not hidden, plainly visible} \tbXV{A goɔ kɔɔ eŋ la a gbɛre poɔ kyɛ o sa᷉a la kperaa lɛ} \tbXE{The thorn has pierced the foot, but it can be seen clearly.} 
\tbLX{nyelele} \tbHM{} \tbPH{ɲɪ́lɪ́lɪ́} \tbPS{adv.~} \tbGE{slightly} \tbXV{A vũũ da sããɛ̃ nyelele lɛ gaŋe zaameŋ.} \tbXE{The light appeared slightly less bright than it was yesterday.} 
\tbLX{nyelennyeleŋ} \tbHM{} \tbPH{ɲélénɲéléŋ} \tbPS{adj.~} \tbGE{bare and shiny e.g. razor shaved head} \tbXV{A bie zu pu᷉oe la ka ba fare ka o e nyelennyeleŋ lɛ.} \tbXE{The child has a cut on his head and they have shaven it clean.} 
\tbLX{nyelɔɔ} \tbHM{} \tbPH{ɲɪ́lɔ́ɔ́} \tbPS{adj.~} \tbGE{tiny} 
\tbLX{nyembini} \tbHM{} \tbPH{ɲɪ́bɪ́nɪ́} \tbPS{n.~} \tbGE{matter (dirt) in between the teeth} \tbTP{nyembinee} 
\tbLX{nyembulo} \tbHM{} \tbPH{ɲɪ́mbúló} \tbPS{n.~} \tbGE{adult} \tbOP{ nyembulibo} 
\tbLX{nyempara} \tbHM{} \tbPH{ɲɪ́mpàrá} \tbPS{n.~} \tbGE{unchiselled teeth} 
\tbLX{nyempɛge dalee} \tbHM{} \tbPH{ɲɪ́mpɛ́gɪ̀ dàléé} \tbPS{n.~} \tbGE{tooth stick, toothbrush} \tbOP{nyempɛge dabilii} \tbTP{nyempɛge dabilee} 
\tbLX{nyene} \tbHM{} \tbPH{ɲɪ́nɪ̀} \tbPS{n.~} \tbSN \tbGE{tooth, fang (snake)} \tbSN \tbGE{toothache} \tbOP{nyemɛ} \tbTP{nyennɛɛ} 
\tbLX{nyeŋgoŋe} \tbHM{} \tbPH{ɲíŋgʊ́ŋ} \tbPS{n.~} \tbGE{the state of grimacing (typically used with "vere")} \tbXV{A dɔɔ da tage la a miri te vere nyeŋgoŋ.} \tbXE{The man tugged at the rope until he grimaced.} \tbOP{nyeŋgonne} \tbTP{nyeŋgonnɛɛ} 
\tbLX{nyeŋkaane} \tbHM{} \tbPH{ɲɪ́ŋkáánɪ́} \tbPS{n.~} \tbGE{molar} \tbSG{nyeŋkaanoo} \tbOP{nyeŋkaama} \tbTP{nyeŋkaanɛɛ} 
\tbLX{nyeraa} \tbHM{} \tbPH{ɲìráá} \tbPS{n.~} \tbGE{ant} \tbOP{nyere} \tbTP{nyerɛɛ} 
\tbLX{nyeraa} \tbHM{} \tbPH{ɲɪ̀ràà} \tbPS{adj.~} \tbGE{not hidden; clearly visible} 
\tbLX{nyere} \tbHM{} \tbPH{ɲɪ̀rɪ̀} \tbPS{v.~} \tbGE{to bare one's teeth; to laugh (scornfully)} \tbPD{nyerɛɛ, nyerɛ, nyerɛ, nyerebɛ, nyeraa, - } 
\tbLX{nyerebaa} \tbHM{} \tbPH{ɲɪ̀rɪ̀bàá} \tbPS{n.~} \tbGE{a type of ant with a painful sting} \tbOP{nyerebaare} \tbTP{nyerebaarɛɛ} \tbVA{nyeresɔgelaa}
\tbLX{nyeregaŋguu} \tbHM{} \tbPH{ɲɪ̀rɪ̀gàŋgúú} \tbPS{n.~} \tbGE{a type of ant} \tbOP{nyeregaŋguuri} \tbTP{nyeregaŋguuree} 
\tbLX{nyeregbere} \tbHM{} \tbPH{ɲɪ̀rɪ̀gbɪ̀rɪ̀} \tbPS{adj.~} \tbGE{mixed} 
\tbLX{nyerekɔɔloŋ} \tbHM{} \tbPH{ɲɪ̀rɪ̀kɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{a type of ant that flies} \tbOP{nyerekɔɔlɔ} \tbTP{nyerekɔɔlɛɛ} 
\tbLX{nyerenyaga} \tbHM{} \tbPH{ɲɪ̀rɪ̀nyàgà} \tbPS{adv.~} \tbGE{uncovered, exposed completely} 
\tbLX{nyeresaare} \tbHM{} \tbPH{ɲɪ̀rɪ̀sáárɪ́} \tbPS{n.~} \tbGE{a type of small slim dark brown ant; usually in a swarm and on trees} \tbSG{nyeresaaroo} \tbOP{nyeresaare} \tbTP{nyeresaarɛɛ} \tbVA{nyeresalensaare}
\tbLX{nyeresalensaare} \tbHM{} \tbPH{ɲɪ̀rɪ̀sálɪ́nsáárɪ̀} \tbPS{n.~} \tbGE{a type of tiny ant that lives together in large numbers} \tbSG{nyeresalensaaroo} \tbTP{nyeresalensaarɛɛ} \tbVA{nyeresaare}
\tbLX{nyeresɔgelaa} \tbHM{} \tbPH{ɲɪ́rɪ́sɔ́gláá} \tbPS{n.~} \tbGE{a type of small black ant, usually found in the house; it has a painful sting} \tbOP{nyeresɔgelɔ} \tbVA{nyerebaa}
\tbLX{nyɛ} \tbHM{} \tbPH{ɲɛ́} \tbPS{v.~} \tbSN \tbGE{to see} \tbSN \tbGE{to find} \tbSN \tbGE{to get, to obtain} \tbPD{nyɛɛ, nyɛrɛ, nyɛrɛ, nyɛrebɛ, nyɛraa, - } 
\tbLX{nyɛ} \tbHM{} \tbPH{ɲɛ̀} \tbPS{v.~} \tbSN \tbGE{to defecate} \tbSN \tbGE{pass substance through the anus (lay eggs)} \tbPD{nyɛɛ, nyɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\tbLX{nyɛ} \tbHM{} \tbPH{ɲɛ̀} \tbPS{v.~} \tbGE{to produce tubers or nuts e.g. yam} \tbPD{nyɛɛ, nyɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\tbLX{nyɛ} \tbHM{} \tbPH{ɲɛ̀} \tbPS{v.~} \tbSN \tbGE{to drain a pond of its water; usually to enable people catch the fish in the pond} \tbSN \tbGE{to drill or scoop matter out to make a hole} \tbPD{nyɛɛ, nyɛ, nyeɛrɛ, nyeɛrebɛ, nyeɛraa, - } 
\tbLX{nyɛdanaao} \tbHM{} \tbPH{ɲɛ̀dànáábʊ́} \tbPS{n.~} \tbGE{a type of tiny brown bird that moves in large groups that lay eggs together} \tbOP{nyɛdanii} 
\tbLX{nyɛ eebo} \tbHM{} \tbPH{ɲɛ́ ɪ́ɪ́bʊ́} \tbPS{v.~} \tbGE{succeed} \tbPD{-, -, -, -, -, - } 
\tbLX{nyɛ ferɛ} \tbHM{} \tbPH{ɲɛ̀ fɪ́rɛ̀} \tbPS{v.~} \tbGE{to suffer} \tbPD{-, -, -, -, -, - } 
\tbLX{nyɛgɛkɛ} \tbHM{} \tbPH{ɲɛ́gɛ́kɛ́} \tbPS{adj.~} \tbGE{warmish} 
\tbLX{nyɛ kparɛɛ} \tbHM{} \tbPH{ɲɛ́ kpárɛ́ɛ́} \tbPS{v.~} \tbGE{to suffer} \tbPD{nyɛɛ kparɛɛ, nyɛrɛ kparɛɛ, -, -, -, - } 
\tbLX{nyɛle} \tbHM{} \tbPH{ɲɛ́lɪ̀} \tbPS{v.~} \tbGE{to melt} \tbXV{ŋmenaa ŋmɛ la a ka᷉a ka a nyɛle.} \tbXE{The oil melted as a result of the sun beating down on it.} \tbPD{nyɛllɛɛ, nyɛllɛ, nyɛllɛ, nyɛleba, nyɛllaa, nyɛleba } 
\tbLX{nyɛlle} \tbHM{} \tbPH{ɲɛ́llɪ́} \tbPS{v.~} \tbGE{to shine, to sparkle} \tbXV{A pɔge du la a talaa ka o te nyɛllɛ.} \tbXE{The womann scrubbed the pan until it shone.} \tbPD{nyɛllɛɛ, nyɛllɛ, nyɛllɛ, nyɛlebɛ, nyɛllaa, - } 
\tbLX{nyɛlle} \tbHM{} \tbPH{ɲɛ̀llɪ̀} \tbPS{v.~} \tbGE{to covet, to aim at getting something} \tbXV{N kuriwiri korɛɛ la N nyɛllɛ la bopaalaa.} \tbXE{My bicycle is old, so I'm aiming at getting a new one.} \tbPD{nyɛllɛɛ, nyɛllɛ, nyɛllɛ, nyɛlebɛ, nyɛllaa, - } 
\tbLX{nyɛlmaanyu} \tbHM{} \tbPH{ɲɛ́lmáányú} \tbPS{n.~} \tbGE{a type of small plant with a strong smell and medicinal properties} \tbOP{nyɛlmaanyuri/nyɛlmaanyuuri} \tbTP{nyɛlmaanyuree/nyɛlmaanyuuree} 
\tbLX{nyɛlugiri} \tbHM{} \tbPH{ɲɛ̀lúgrí} \tbPS{n.~} \tbGE{support stick for a yam plant} \tbOP{nyɛlugo} \tbTP{nyɛlugee/nyɛlugree} \tbVA{nyalugri}
\tbLX{nyɛnnoŋ} \tbHM{} \tbPH{ɲɛ̀nnʊ́ŋ} \tbPS{n.~} \tbGE{salt} \tbVA{nyɛnnoo, yaaroŋ}
\tbLX{Nyɛntɔge} \tbHM{} \tbPH{ɲɛ́ntɔ́gɪ̀} \tbPS{n.~} \tbGE{a meteorite revered to be a fetish in the Sisaalaland} \tbOP{nyɛntɔgre} \tbTP{nyɛntɔgrɛɛ} 
\tbLX{nyɛ pare} \tbHM{} \tbPH{ɲɛ́ párɪ́} \tbPS{v.~} \tbGE{to lie down helplessly, to crumple} \tbPD{nyɛɛ pare, nyeɛrɛ pare, -, -, -, - } 
\tbLX{nyɛraa} \tbHM{} \tbPH{ɲɛ́ráá} \tbPS{n.~} \tbGE{seer} \tbOP{nyɛrebɛ} \tbTP{nyɛrebɛɛ} 
\tbLX{nyɛre} \tbHM{} \tbPH{ɲɛ́rɪ̀} \tbPS{v.~} \tbGE{to miss (a target)} \tbPD{nyɛrɛɛ, nyɛrɛ, nyɛrɛ, nyɛrebɛ, nyɛraa, - } \tbVA{nyare}
\tbLX{nyɛre} \tbHM{} \tbPH{ɲɛ̀rɪ̀} \tbPS{v.~} \tbGE{to be fierce, to get very angry} \tbPD{nyɛrɛɛ, nyɛrɛ, nyɛrɛ, nyɛrebɛ, nyɛraa, - } \tbVA{nyare}
\tbLX{nyɛtaa} \tbHM{} \tbPH{ɲɛ́táá} \tbPS{v.~} \tbGE{confer} \tbPD{-, -, -, -, -, - } 
\tbLX{nyɛ yuori} \tbHM{} \tbPH{ɲɛ́ yúórí} \tbPS{v.~} \tbGE{get fame, (be) powerful} \tbPD{-, -, -, -, -, - } 
\tbLX{nyɛ zɛ} \tbHM{} \tbPH{ɲɛ̀ zɛ̀} \tbPS{v.~} \tbGE{to suffer} \tbPD{-, -, -, -, -, - } 
\tbLX{nyɛ zɛ} \tbHM{} \tbPH{ɲɛ̀ zɛ̀} \tbPS{v.~} \tbGE{to suffer greatly} \tbPD{nyɛ zɛɛ, nyeɛrɛ zeɛrɛ, nyɛ zeɛrɛ, nyɛ zeɛraa, -, - } 
\tbLX{nyieli} \tbHM{} \tbPH{ɲíélì} \tbPS{v.~} \tbGE{to tempt; to show something to someone but refuse to give it to him/her} \tbPD{nyíéliee, nyíéle, nyíélie, nyíélaa, -, - } 
\tbLX{nyigiri} \tbHM{} \tbPH{ɲígírì} \tbPS{v.~} \tbGE{to close both eyes tightly and simultaneously towards someone (insulting gesture)} \tbPD{nyigree, nyigre, nyigre, nyigiribe, nyigraa, nyigiribe } 
\tbLX{nyigiri} \tbHM{} \tbPH{ɲìgìrì} \tbPS{v.~} \tbGE{to steal} \tbPD{nyigiree, nyigire, nyigire, nyigiribe, nyigiraa, nyigiribe } 
\tbLX{nyiile} \tbHM{} \tbPH{ɲììlé} \tbPS{n.~} \tbGE{guinea worm disease} \tbOP{nyiili} \tbTP{nyiilee} 
\tbLX{nyiili} \tbHM{} \tbPH{ɲììlì} \tbPS{v.~} \tbGE{to touch gently} \tbPD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\tbLX{nyiili} \tbHM{} \tbPH{ɲììlì} \tbPS{v.~} \tbGE{to incense} \tbPD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\tbLX{nyiili} \tbHM{} \tbPH{ɲììlì} \tbPS{v.~} \tbGE{to shudder} \tbPD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\tbLX{nyiili} \tbHM{} \tbPH{ɲììlì} \tbPS{v.~} \tbGE{to swell (breast)} \tbPD{nyiilee, nyiile, nyiile, nyiilibe, nyiilaa, nyiile } 
\tbLX{nyiiri} \tbHM{} \tbPH{ɲììrì} \tbPS{v.~} \tbGE{an intransitive verb; used with ‘nyemɛ’ to bare teeth} \tbXV{Basɔgelaa kaŋa nyiiri la nyemɛ eŋe a nannyige.} \tbXE{A black dog bared its teeth at the thief.} \tbPD{nyiiree, nyiire, nyiire, -, nyiiraa, - } 
\tbLX{nyimenyime} \tbHM{} \tbPH{ɲíményímé} \tbPS{adj.~} \tbGE{tasteful} 
\tbLX{nyobogdaa} \tbHM{} \tbPH{ɲòbògìdáá} \tbPS{n.~} \tbGE{bridge of nose} \tbOP{nyobogdaare} \tbTP{nyobogdaarɛɛ} 
\tbLX{nyoge} \tbHM{} \tbPH{ɲʊ̀gɪ̀} \tbPS{v.~} \tbGE{to show off} \tbPD{nyogɛɛ, nyogrɔ/nyogra, nyogrɔ/nyogra, nyogreba, nyograa, - } 
\tbLX{nyoge} \tbHM{} \tbPH{ɲʊ̀gɪ̀} \tbPS{v.~} \tbGE{to become sour or bad e.g. sauce} \tbPD{nyogɛɛ, nyogrɔ/nyogra, nyogrɔ/nyogra, nyogreba, nyograa, - } 
\tbLX{nyoge} \tbHM{} \tbPH{ɲʊ́gɪ́} \tbPS{v.~} \tbGE{used with ‘tɔ’ (tɔ-nyoge) punish severely (physically)} \tbPD{-, -, -, -, -, - } 
\tbLX{nyoge} \tbHM{} \tbPH{ɲʊ́gɪ́} \tbPS{n.~} \tbGE{concoction believed to be powerful} 
\tbLX{nyoge} \tbHM{} \tbPH{ɲʊ́gɪ́} \tbPS{n.~} \tbGE{filth} 
\tbLX{nyoge-kaa-kpɛre} \tbHM{} \tbPH{ɲʊ́gɪ́-káá-kpɛ́rɪ́} \tbPS{n.~} \tbGE{over show off that ends badly} 
\tbLX{nyogele} \tbHM{} \tbPH{ɲʊ̀gɪ̀lɪ̀} \tbPS{v.~} \tbGE{to dress well} \tbPD{-, -, -, -, -, - } 
\tbLX{nyogele} \tbHM{} \tbPH{ɲʊ̀gɪ̀lɪ̀} \tbPS{v.~} \tbGE{to be congested} \tbPD{nyoglɛɛ, nyoglɔ, nyoglɔ, nyoglebɔ, nyoglaa, - } 
\tbLX{nyogere} \tbHM{} \tbPH{ɲʊ́gɪ́rɪ̀} \tbPS{n.~} \tbGE{pride} 
\tbLX{nyoll} \tbHM{} \tbPH{ɲʊ̀ll} \tbPS{adj.~} \tbGE{heavy} 
\tbLX{nyoll} \tbHM{} \tbPH{ɲʊ́ll} \tbPS{adj.~} \tbGE{small and light} 
\tbLX{nyonne} \tbHM{} \tbPH{ɲʊ́nnɪ́} \tbPS{v.~} \tbGE{to dry up, to evaporate} \tbPD{nyonnɛɛ, nyonna, nyonna, nyonnema, nyonnaa, nyonnema } 
\tbLX{nyonne} \tbHM{} \tbPH{ɲʊ̀nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to sniff} \tbSN \tbGE{to search; to delve into} \tbSN \tbGE{to test; to examine} \tbPD{nyonnɛɛ, nyonna, nyonna, nyonnema, nyonnaa, nyonnema } 
\tbLX{nyonono} \tbHM{} \tbPH{ɲʊ́nʊ́nʊ́} \tbPS{n.~} \tbGE{mercy, pity} 
\tbLX{nyonsomm} \tbHM{} \tbPH{ɲʊ̀nsʊ̀mm} \tbPS{adj.~} \tbGE{biggish} 
\tbLX{nyonyonne} \tbHM{} \tbPH{ɲʊ́nyʊ́nnɪ̀} \tbPS{n.~} \tbGE{nausea} 
\tbLX{nyonyonne} \tbHM{} \tbPH{ɲʊ́nyʊ́nnɪ̀} \tbPS{n.~} \tbGE{dirt} 
\tbLX{nyoŋgoloŋ} \tbHM{} \tbPH{ɲʊ̀ŋgʊ̀lʊ̀ŋ} \tbPS{adj.~} \tbGE{mixed up e.g. heap of things} 
\tbLX{nyoŋgoloŋ} \tbHM{} \tbPH{ɲʊ̀ŋgʊ̀lʊ̀ŋ} \tbPS{adj.~} \tbGE{thick and stiff e.g. paste} 
\tbLX{nyoore} \tbHM{} \tbPH{ɲʊ̀ʊ́rɪ̀} \tbPS{n.~} \tbGE{smoke} \tbTP{nyoorɛɛ} \tbVA{zoore}
\tbLX{nyoore} \tbHM{} \tbPH{ɲʊ̀ʊ́rɪ̀} \tbPS{n.~} \tbGE{pride} \tbTP{nyoorɛɛ} \tbVA{zoore}
\tbLX{nyoore} \tbHM{} \tbPH{ɲʊ̀ùrɪ̀} \tbPS{v.~} \tbSN \tbGE{to make hard} \tbXV{Zenɛ ka ba mɛ a dankyini ŋa bieo ka o na nyoore.} \tbXE{This wall was only built today, so by tomorrow it will get hard.} \tbSN \tbGE{to dry up} \tbXV{Uoni saŋa bare yaga maŋ nyoore la.} \tbXE{Many ponds get dried up during the dry season.} \tbPD{nyoorɛɛ, nyoora, nyoora, nyooreba, nyooraa, - } 
\tbLX{nyoore} \tbHM{} \tbPH{ɲʊ̀ùrɪ̀} \tbPS{v.~} \tbGE{to test the capability of e.g. a tool or a person} \tbPD{nyoorɛɛ, nyoora, nyoora, nyooreba, nyooraa, - } 
\tbLX{nyoɔ} \tbHM{} \tbPH{ɲʊ́ɔ̀} \tbPS{v.~} \tbGE{to fire, to bake to harden (clay)} \tbPD{nyoɛ, nyoɔnɔ, nyoɔnɔ, nyoɔnema, nyoɔnaa, nyoɔnema } 
\tbLX{nyoɔle} \tbHM{} \tbPH{ɲʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to begin rising} \tbPD{nyoɔlɛɛ, nyoɔlɔ, nyoɔlɔ, nyoɔlba, nyoɔlaa, - } 
\tbLX{nyoɔle} \tbHM{} \tbPH{ɲʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to close partially e.g. of eyes in strong light} \tbPD{nyoɔlɛɛ, nyoɔlɔ, nyoɔlɔ, nyoɔlba, nyoɔlaa, - } 
\tbLX{nyoɔnaa} \tbHM{} \tbPH{ɲʊ́ɔ́náá} \tbPS{n.~} \tbGE{oven} \tbOP{nyoɔne} \tbTP{nyoɔnɛɛ} 
\tbLX{nyoɔre} \tbHM{} \tbPH{ɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{nose} \tbOP{nyɔɛ} \tbTP{nyɔrɛɛ} 
\tbLX{nyoɔre} \tbHM{} \tbPH{ɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{life} \tbOP{nyɔɛ} \tbTP{nyɔrɛɛ} 
\tbLX{nyoɔre} \tbHM{} \tbPH{ɲʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{tip of e.g. stick} \tbOP{nyɔɛ} \tbTP{nyɔrɛɛ} 
\tbLX{nyore} \tbHM{} \tbPH{ɲʊ̀rɪ̀} \tbPS{v.~} \tbGE{to shrivel, to wrinkle (cloth)} \tbPD{nyorɛɛ, nyora, nyora, nyoreba, nyoraa, nyoreba } 
\tbLX{nyore} \tbHM{} \tbPH{ɲʊ̀rɪ̀} \tbPS{v.~} \tbGE{to take courage} \tbPD{nyorɛɛ, nyora, nyora, nyoreba, nyoraa, nyoreba } 
\tbLX{nyoremaŋ} \tbHM{} \tbPH{ɲʊ̀rɪ̀màŋ} \tbPS{adj.~} \tbGE{bushy, unkempt (hair)} 
\tbLX{nyɔbaare} \tbHM{} \tbPH{ɲɔ́báárɪ́} \tbPS{n.~} \tbGE{disease of the nose} \tbTP{nyɔbaarɛɛ} 
\tbLX{nyɔbogiri} \tbHM{} \tbPH{ɲɔ̀bògirí} \tbPS{n.~} \tbSN \tbGE{nostril} \tbSN \tbGE{nose} \tbOP{nyɔbogo} \tbTP{nyɔbogiree} \tbVA{nyobogi}
\tbLX{nyɔfeerɛ} \tbHM{} \tbPH{ɲɔ̀fɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{cold (illness)} \tbTP{nyɔfeerɛɛ} 
\tbLX{nyɔfuli} \tbHM{} \tbPH{ɲɔ̀fùlí} \tbPS{n.~} \tbGE{dried snot} \tbSG{nyɔfuluu} \tbOP{nyɔfulo} \tbTP{nyɔfulee} 
\tbLX{nyɔge} \tbHM{} \tbPH{ɲɔ́gɪ̀} \tbPS{v.~} \tbSN \tbGE{to hold} \tbXV{A bibiiri nyɔge la taa nuuri kyerre te gɛrɛ.} \tbXE{The children held each others' hands and scampered away.} \tbSN \tbGE{to catch, to arrest} \tbXV{A polisi nyɔge la a nanyige baguo ŋa.} \tbXE{The police arrested the thief this morning.} \tbPD{nyɔgɛɛ, nyɔgrɔ, nyɔgrɔ, nyɔgreba, nyɔgraa, nyɔgreba } 
\tbLX{nyɔge} \tbHM{} \tbPH{ɲɔ́gɪ̀} \tbPS{v.~} \tbGE{to choke} \tbXV{A baa vɔlɔ la a nɛne yaga lɛ ka a te nyɔge o.} \tbXE{The dog swallowed the meat so fast that the meat choked it.} \tbPD{nyɔgɛɛ, nyɔgrɔ, nyɔgrɔ, nyɔgreba, nyɔgraa, nyɔgreba } 
\tbLX{nyɔge} \tbHM{} \tbPH{ɲɔ́gɪ̀} \tbPS{v.~} \tbGE{to infect} \tbXV{A bie kɔroŋ na nyɔge la o yɔɔ.} \tbXE{The child's cold has infected his sister.} \tbPD{nyɔgɛɛ, nyɔgrɔ, nyɔgrɔ, nyɔgreba, nyɔgraa, nyɔgreba } 
\tbLX{nyɔge kukuri} \tbHM{} \tbPH{ɲɔ́gɪ́ kúkúrì} \tbPS{v.~} \tbGE{to farm} \tbPD{nyɔgɛɛ kukuri, nyɔgrɔ kukuri, -, -, -, - } 
\tbLX{nyɔge noɔre} \tbHM{} \tbPH{ɲɔ́gɪ́ nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to be surprised} \tbPD{-, -, -, -, -, - } 
\tbLX{nyɔgetaa} \tbHM{} \tbPH{ɲɔ́gɪ́táá} \tbPS{n.~} \tbGE{unity} 
\tbLX{nyɔgezie} \tbHM{} \tbPH{ɲɔ́gɪ̀zíé} \tbPS{n.~} \tbGE{handle, part of an object to hold} \tbOP{nyɔgeziiri} \tbTP{nyɔgeziiree} 
\tbLX{nyɔɔ} \tbHM{} \tbPH{ɲɔ̀ɔ̀} \tbPS{v.~} \tbGE{to have sexual intercourse (human)} \tbPD{nyɔɔɛ, nyɔɔnɔ/nyɔɔrɔ, nyɔɔnɔ/nyɔɔrɔ, nyɔɔnema, nyɔɔnaa/nyɔɔraa, nyɔɔnema } 
\tbLX{nyɔpolema} \tbHM{} \tbPH{ɲɔ̀pʊ́lɪ́má} \tbPS{n.~} \tbGE{summit, highest point} \tbOP{nyɔpolle} \tbTP{nyɔpollɛɛ} 
\tbLX{nyɔre} \tbHM{} \tbPH{ɲɔ̀rɪ̀} \tbPS{v.~} \tbGE{to despise, to hate} \tbPD{nyɔrɛe,, nyɔrɔ, nyɔrɔ, nyɔreba/nyɔrebɔ, nyɔraa, - } \tbVA{nyɔroo}
\tbLX{nyɔtage} \tbHM{} \tbPH{ɲɔ̀tàgɪ̀} \tbPS{n.~} \tbGE{sniveling} 
\tbLX{nyɔvore} \tbHM{} \tbPH{ɲɔ̀vʊ́rɪ̀} \tbPS{n.~} \tbGE{life} \tbOP{nyɔvoɛ} \tbTP{nyɔvorɛɛ} 
\tbLX{nyu} \tbHM{} \tbPH{ɲù} \tbPS{v.~} \tbGE{to smell (scent)} \tbXV{A zɛkpaane da nyu la noɔ lɛ.} \tbXE{The simmering sauce smelt nice.} \tbPD{-, -, nyue, nyuuro, nyuuro, nyuuraa } 
\tbLX{nyu} \tbHM{} \tbPH{ɲú} \tbPS{v.~} \tbSN \tbGE{to drink} \tbSN \tbGE{to suck (of a mosquito)} \tbPD{nyue, nyuuro, nyuuro, nyuuribo, nyuuraa, nyuuribo } 
\tbLX{nyu} \tbHM{} \tbPH{ɲú} \tbPS{v.~} \tbGE{to smoke (pipe)} \tbPD{nyue, nyuuro, nyuuro, nyuuribo, nyuuraa, nyuuribo } 
\tbLX{nyubuli} \tbHM{} \tbPH{ɲúbúlí} \tbPS{n.~} \tbGE{yam sprout} \tbOP{nyubulo} \tbTP{nyubulee} 
\tbLX{nyudɛre} \tbHM{} \tbPH{ɲúdɛ́rɪ́} \tbPS{n.~} \tbGE{large yam tuber} \tbOP{nyudɛɛ} \tbTP{nyudɛrɛɛ/nyudɛrrɛɛ} \tbVA{wadɛre}
\tbLX{nyumoɔ} \tbHM{} \tbPH{ɲúmʊ́ɔ́} \tbPS{n.~} \tbGE{field prepared to make yam mounds} \tbTP{nyumoɔnɛɛ} \tbVA{vutɛne}
\tbLX{nyunni} \tbHM{} \tbPH{ɲùnnì} \tbPS{v.~} \tbSN \tbGE{to smell} \tbXV{A zɔŋɔ nyunni la a kparoo, a yeli ka o deni ba e o.} \tbXE{The blind man smelt the shirt and said it was not his.} \tbSN \tbGE{to sniff e.g. dog} \tbPD{nyunnee, nyunno, nyunno, nyunimo, nyunnaa, nyunimo } 
\tbLX{nyuo} \tbHM{} \tbPH{ɲùó} \tbPS{n.~} \tbSN \tbGE{navel} \tbSN \tbGE{see wɛnyuo} \tbOP{nyuuri} \tbTP{nyuuree} 
\tbLX{nyusege} \tbHM{} \tbPH{ɲúsɪ́gɪ́} \tbPS{n.~} \tbGE{yam barn} \tbOP{nyusegre} \tbTP{nyusegrɛɛ} \tbVA{wasege}
\tbLX{nyuu} \tbHM{} \tbPH{ɲúú} \tbPS{n.~} \tbSN \tbGE{neck} \tbSN \tbGE{smell, odour} \tbOP{nyuuri} \tbTP{nyuuree} 
\tbLX{nyuu} \tbHM{} \tbPH{ɲùú} \tbPS{n.~} \tbGE{a sensation, good or bad, detected by inhaling} 
\tbLX{nyuu} \tbHM{} \tbPH{ɲúú} \tbPS{adj.~} \tbGE{stinky} \tbOP{nyuuri} \tbTP{nyuuree} 
\tbLX{nyuuli} \tbHM{} \tbPH{ɲúúlì} \tbPS{v.~} \tbGE{to empower supernaturally} \tbPD{nyuulee, nyuulo, nyuulo, nyuulibo, nyuulaa, - } 
\tbLX{nyu-uo} \tbHM{} \tbPH{ɲú-úó} \tbPS{n.~} \tbGE{yam barn} \tbOP{nyu-uori} \tbTP{nyu-uoree} \tbVA{nyu-sege}
\tbLX{nyuuri} \tbHM{} \tbPH{ɲúúrí} \tbPS{n.~} \tbGE{envy,  jealousy} \tbOP{n/a} \tbTP{nyuuree} 
\tbLX{nyuuri} \tbHM{} \tbPH{ɲúúrì} \tbPS{n.~} \tbGE{yam} \tbOP{nyie} \tbTP{nyiree/nyiiree} \tbVA{waa}
\tbLX{nyuweɛle} \tbHM{} \tbPH{ɲùwɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{yam seed} \tbSG{nyuweɛloo} \tbOP{nyuweɛlɛ} \tbTP{nyuweɛlɛɛ} \tbVA{waweɛlɛ}
\tbLX{nyuzɛle} \tbHM{} \tbPH{ɲùzɛ̀lɪ́} \tbPS{n.~} \tbGE{umbilical cord} \tbSG{nyuzɛloo} \tbOP{nyuzɛlɛ} \tbTP{nyuzɛlɛɛ/nyuzɛllɛɛ} 
\tbLX{ŋa} \tbHM{} \tbPH{ŋá} \tbPS{pronoun.~} \tbGE{this (third person demonstrative)} \tbXV{A dɔɔ ŋa zaameŋ wa la kyɛ.} \tbXE{This man came here yesterday.} 
\tbLX{ŋa} \tbHM{} \tbPH{ŋá} \tbPS{adv.~} \tbGE{similar to, like} \tbXV{A dɔɔ waa ŋa sogya na.} \tbXE{The man is like a soldier.} 
\tbLX{ŋma} \tbHM{} \tbPH{ŋmá} \tbPS{v.~} \tbGE{to conquer, to plunder} \tbPD{ŋmaɛ, ŋmaara, ŋmaara, ŋmaareba, ŋmaaraa, - } 
\tbLX{ŋmaa} \tbHM{} \tbPH{ŋmàà} \tbPS{v.~} \tbGE{to cut} \tbXV{Nɛne la ka a pɔge da ŋmaara ka a soɔ ŋmaa o nubiri.} \tbXE{The woman was cutting meat when the knife cut her finger.} \tbPD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\tbLX{ŋmaa} \tbHM{} \tbPH{ŋmàà} \tbPS{v.~} \tbGE{to write} \tbXV{Ayuo ŋmaa la sɛnsɛŋe gane gaa ko o sɛne.} \tbXE{Ayuo has written a love letter to her boyfriend.} \tbPD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\tbLX{ŋmaa} \tbHM{} \tbPH{ŋmàà} \tbPS{v.~} \tbSN \tbGE{to cross (river)} \tbXV{A mane paalɛɛ la ka ba maŋ de gbori ŋmaa ne.} \tbXE{The river is flooded and they are using a boat to cross it.} \tbSN \tbGE{to go across to meet} \tbXV{Polisiri ŋmaa la a sori nyɔgere a nannyigiri.} \tbXE{The police crossed the road and arrested the thieves.} \tbPD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\tbLX{ŋmaa} \tbHM{} \tbPH{ŋmàà} \tbPS{v.~} \tbGE{to prevent} \tbPD{ŋmaarɛɛ, ŋmaara, -, -, -, - } 
\tbLX{ŋmaa} \tbHM{} \tbPH{ŋmáá} \tbPS{adj.~} \tbGE{short} \tbXV{O serɛ la a dɔɔ ŋmaa na.} \tbXE{That short man is her husband.} \tbOP{ŋmaara} 
\tbLX{ŋmaa bare} \tbHM{} \tbPH{ŋmàà bàrɪ̀} \tbPS{v.~} \tbGE{to exclude; to put aside} \tbPD{ŋmaa barɛɛ, ŋmaara bara, ŋmaabara, ŋmaabareba, -, - } 
\tbLX{ŋmaabini} \tbHM{} \tbPH{ŋmàábíní} \tbPS{n.~} \tbGE{a type of plant} \tbOP{ŋmaabinee} 
\tbLX{ŋmaadaa} \tbHM{} \tbPH{ŋmààdáá} \tbPS{n.~} \tbGE{mature big male monkey} \tbOP{ŋmaadaare} \tbTP{ŋmaadarɛɛ} 
\tbLX{ŋmaadasaa} \tbHM{} \tbPH{ŋmààdàsàà} \tbPS{n.~} \tbGE{a huge monkey} \tbOP{ŋmaadasaare} \tbTP{ŋmaadasaarɛɛ} 
\tbLX{ŋmaakalaa} \tbHM{} \tbPH{ŋmààkàlàá} \tbPS{n.~} \tbGE{small but smart type of monkey} \tbOP{ŋmaakalle} \tbTP{ŋmaakalɛɛ} 
\tbLX{ŋmaakyeraa} \tbHM{} \tbPH{ŋmààʧɪ̀ràá} \tbPS{n.~} \tbSN \tbGE{a type of tree with medicinal properties} \tbSN \tbGE{a type of small monkey} \tbOP{ŋmaakyerre} \tbTP{ŋmaakyerɛɛ} 
\tbLX{ŋmaaŋa} \tbHM{} \tbPH{ŋmááŋà} \tbPS{n.~} \tbGE{monkey} \tbOP{ŋmaane} \tbTP{ŋmaanɛɛ} 
\tbLX{ŋmaapare} \tbHM{} \tbPH{ŋmààpárɪ́} \tbPS{n.~} \tbGE{name of a popular market in Teza, a village in the Upper West region of Ghana} \tbTP{ŋmaaparɛɛ} 
\tbLX{ŋmaaraa} \tbHM{} \tbPH{ŋmààráà} \tbPS{n.~} \tbGE{an evil spirit with powers of witchcraft, usually in the form of an animal that haunts people} \tbOP{ŋmaare} \tbTP{ŋmaarɛɛ} \tbVA{vuuluŋ}
\tbLX{ŋmaa seɛ} \tbHM{} \tbPH{ŋmàá sɪ́ɛ́} \tbPS{v.~} \tbGE{to conceive, to be pregnant} \tbPD{ŋmaaɛ seɛ, ŋmaara seɛ, se-ŋmaara, se-ŋmareba, -, - } 
\tbLX{ŋmaasugo} \tbHM{} \tbPH{ŋmààsúgó} \tbPS{n.~} \tbGE{band of monkeys} \tbOP{ŋmaasugri} \tbTP{ŋmaasugree} 
\tbLX{ŋmaataŋaa} \tbHM{} \tbPH{ŋmààtàŋáá} \tbPS{n.~} \tbGE{a type of tree whose fruits are eaten by birds and animals} \tbOP{ŋmaataŋema} \tbTP{ŋmaatannɛɛ} 
\tbLX{ŋmaateɛ} \tbHM{} \tbPH{ŋmààtɪ́ɛ́} \tbPS{n.~} \tbGE{a type of tree} \tbOP{ŋmaateere} \tbTP{ŋmaateerɛɛ} 
\tbLX{ŋmaa vooroŋ} \tbHM{} \tbPH{ŋmàà vʊ́ʊ́rʊ́ŋ} \tbPS{v.~} \tbGE{to stop breathing, to die} \tbPD{-, -, -, -, -, - } 
\tbLX{ŋmaazeɛ} \tbHM{} \tbPH{ŋmààzɪ̀ɛ̀} \tbPS{n.~} \tbGE{a type of monkey} \tbOP{ŋmaazeere} \tbTP{ŋmaazeerɛɛ} 
\tbLX{ŋmalaŋmeɛ} \tbHM{} \tbPH{ŋmáláŋmɪ́ɛ́} \tbPS{n.~} \tbGE{a type of  bird with thin long legs and a long beak} \tbOP{ŋmalaŋmeere} \tbTP{ŋmalaŋmeerɛɛ} \tbVA{ŋmalanweɛ}
\tbLX{ŋmamberɛ} \tbHM{} \tbPH{ŋmámbɪ́rɛ́} \tbPS{n.~} \tbGE{a type of small weed with white flowers and medicinal properties} \tbSG{ŋamberoo} \tbOP{ŋmambere/ŋmamberoo} \tbTP{ŋamberɛɛ} 
\tbLX{ŋmammo} \tbHM{} \tbPH{ŋmámmʊ́} \tbPS{n.~} \tbGE{a type of cake made from millet flour} \tbOP{ŋmama} \tbTP{ŋmamɛɛ} 
\tbLX{ŋmampapaa} \tbHM{} \tbPH{ŋmámpàpáà} \tbPS{n.~} \tbGE{a type of pigeon with red eyes} \tbOP{ŋmampapaare} \tbTP{ŋmampapaarɛɛ} 
\tbLX{ŋmampelaa} \tbHM{} \tbPH{ŋmámpɪ̀láá} \tbPS{n.~} \tbGE{wild grey pigeon} \tbOP{ŋmampeɛle} \tbTP{ŋmampeɛlɛɛ} 
\tbLX{ŋmampelaa} \tbHM{} \tbPH{ŋmàmpɪ̀làá} \tbPS{n.~} \tbGE{neat calabash used for serving water or poridge} \tbOP{ŋmampeɛle} \tbTP{ŋmampeɛlɛɛ} 
\tbLX{ŋmamporezupelaa} \tbHM{} \tbPH{ŋmǎmpʊ̀rzúpɪ́làá} \tbPS{n.~} \tbGE{a character in Dagaaba folktales which has supernatural power and which it uses to test the ego of other characters} \tbOP{ŋmamporzupeɛle} \tbVA{ŋmamporzuzeɛ}
\tbLX{ŋmamporɔ} \tbHM{} \tbPH{ŋmàmpʊ́rɔ́} \tbPS{n.~} \tbGE{a red round insect} \tbOP{ŋmamporre} \tbTP{ŋmamporɛɛ} \tbVA{ŋmampora}
\tbLX{ŋmamporzuzeɛ} \tbHM{} \tbPH{ŋmàmpʊ́rzúzɪ̀ɛ́} \tbPS{n.~} \tbGE{} \tbVA{ŋmamporzupelaa}
\tbLX{ŋmandɛre} \tbHM{} \tbPH{ŋmándɛ́rè} \tbPS{n.~} \tbGE{spider} \tbOP{ŋmandɛɛ} \tbTP{ŋmandɛrɛɛ} 
\tbLX{ŋmane} \tbHM{} \tbPH{ŋmánɪ́} \tbPS{n.~} \tbSN \tbGE{calabash, bottle gourd} \tbSN \tbGE{calabash plant} \tbOP{ŋmama} \tbTP{ŋmannɛɛ} 
\tbLX{ŋmanee} \tbHM{} \tbPH{ŋmánɪ́ɪ̀} \tbPS{n.~} \tbGE{dove} \tbOP{ŋmama} \tbTP{ŋmannɛɛ} 
\tbLX{ŋmantummu} \tbHM{} \tbPH{ŋmǎntúmmú} \tbPS{n.~} \tbGE{unpolished calabash} \tbOP{ŋmantumo} \tbTP{ŋmantumee} 
\tbLX{ŋmantuo} \tbHM{} \tbPH{ŋmàntúó} \tbPS{n.~} \tbGE{bitter calabash} \tbOP{ŋmantuuri} \tbTP{ŋmantuuree} 
\tbLX{ŋmanzeɛ} \tbHM{1} \tbPH{ŋmǎnzɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of dove with pink plumage} \tbSG{ŋmanzeroo} \tbOP{ŋmanzeere} \tbTP{ŋmanzeerɛɛ} 
\tbLX{ŋmanzeɛ} \tbHM{2} \tbPH{ŋmǎnzɪ̀ɛ́} \tbPS{n.~} \tbGE{calabash used in the preparation of oil} \tbOP{ŋmanzeere} \tbTP{ŋmanzeerɛɛ} 
\tbLX{ŋmaŋgbɛre} \tbHM{} \tbPH{ŋmàŋgbɛ́rɪ́} \tbPS{n.~} \tbGE{vine of the calabash plant} \tbOP{ŋmaŋgbɛɛ} \tbTP{ŋmaŋgbɛɛrɛɛ} 
\tbLX{ŋmaŋkuŋkummu} \tbHM{} \tbPH{ŋmàŋkùŋkùmmú} \tbPS{n.~} \tbGE{calabash seed in a ball form} \tbOP{ŋmaŋkuŋkumo} \tbTP{ŋmaŋkuŋkunni} 
\tbLX{ŋmaraa} \tbHM{} \tbPH{ŋmáràà} \tbPS{n.~} \tbSN \tbGE{moon} \tbSN \tbGE{month} \tbOP{ŋmarre} \tbTP{ŋmarrɛɛ} 
\tbLX{ŋmare} \tbHM{} \tbPH{ŋmárɪ̀} \tbPS{v.~} \tbSN \tbGE{to break e.g. pot} \tbSN \tbGE{to crush} \tbPD{ŋmarɛɛ, ŋmara, ŋmara, ŋmareba, ŋmaraa, ŋmareba } 
\tbLX{ŋmare} \tbHM{} \tbPH{ŋmárɪ́} \tbPS{v.~} \tbGE{to change money into smaller units} \tbPD{ŋmarɛɛ, ŋmara, ŋmara, ŋmareba, ŋmaraa, ŋmareba } 
\tbLX{ŋmarebiri} \tbHM{} \tbPH{ŋmárɪ́bírì} \tbPS{n.~} \tbGE{star} \tbSG{ŋmarebiruu} \tbOP{ŋmarebie} \tbTP{ŋmarebiiree} 
\tbLX{ŋmaroo} \tbHM{} \tbPH{ŋmárʊ̀ʊ̀} \tbPS{n.~} \tbGE{change (money)} \tbSG{ŋmarrɛɛ} \tbTP{ŋmarre} 
\tbLX{ŋma zirii} \tbHM{} \tbPH{ŋmàà zìríì} \tbPS{v.~} \tbGE{to lie} \tbXV{O ŋmaa la zirii ka o baare la a toma.} \tbXE{He lied that he had finished the work.} \tbPD{-, -, -, -, -, - } 
\tbLX{ŋmee} \tbHM{} \tbPH{ŋmɪ́ɪ̀} \tbPS{v.~} \tbGE{to crush grains out of fresh heads of millet (usually after roasting)} \tbPD{ŋmeɛɛ, ŋmeɛnɛ, ŋmeɛnɛ, ŋmeɛnemɛ, ŋmenaa, - } \tbVA{ŋmeɛ}
\tbLX{ŋmeene} \tbHM{} \tbPH{ŋmɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbGE{to become thin or small} \tbPD{ŋmeenɛɛ, ŋmeenɛ, ŋmeenɛ, ŋmeenemɛ, ŋmeenaa, - } 
\tbLX{ŋmeene} \tbHM{} \tbPH{ŋmɪ́ɪ́nɪ́} \tbPS{adj.~} \tbGE{lean} \tbOP{ŋmeemɛ} 
\tbLX{ŋmeeo} \tbHM{} \tbPH{ŋmɪ́ɪ́ʊ́} \tbPS{n.~} \tbGE{wicked and secret act} 
\tbLX{ŋmeɛ} \tbHM{} \tbPH{ŋmɪ́ɛ} \tbPS{v.~} \tbGE{to crush} \tbPD{ŋmeɛɛ, ŋmeɛnɛ, ŋmeɛnɛ, ŋmeɛnemɛ, ŋmeɛnaa, - } \tbVA{ŋmee}
\tbLX{ŋmeɛ} \tbHM{} \tbPH{ŋmɪ́ɛ} \tbPS{adj.~} \tbGE{scorched, withered e.g. green vegetation turning brown} 
\tbLX{ŋmeɛlaa} \tbHM{} \tbPH{ŋmɪ́ɛ́láá} \tbPS{n.~} \tbGE{skipping rope} \tbOP{ŋmeɛle} \tbTP{ŋmeɛlɛɛ} 
\tbLX{ŋmeɛle} \tbHM{} \tbPH{ŋmɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{to twist; to wring e.g. hand} \tbPD{ŋmeɛlɛɛ, ŋmeɛlɛ, ŋmeɛlɛ, ŋmeɛlebɛ, ŋmeɛlaa, ŋmeɛlebɛ } \tbVA{nyɛ}
\tbLX{ŋmeɛle} \tbHM{} \tbPH{ŋmɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{to curl out, to ooze (smoke)} \tbXV{Nyoore ŋmeɛlɛ la a sege poɔ, ba mare la vu᷉u᷉.} \tbXE{Smoke is oozing from the hut, an indication that they have made fire.} \tbPD{ŋmeɛlɛɛ, ŋmeɛlɛ, ŋmeɛlɛ, ŋmeɛlebɛ, ŋmeɛlaa, ŋmeɛlebɛ } \tbVA{nyɛ}
\tbLX{ŋmeɛle} \tbHM{} \tbPH{ŋmɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{to pass gas, to break wind, to fart} \tbXV{A bie poɔ pie la ka o deɛ maŋ ŋmeɛle bimporɔ bare.} \tbXE{The child has a bloated stomach and he has terrible gas.} \tbPD{ŋmeɛlɛɛ, ŋmeɛlɛ, ŋmeɛlɛ, ŋmeɛlebɛ, ŋmeɛlaa, ŋmeɛlebɛ } \tbVA{nyɛ}
\tbLX{ŋmeɛle} \tbHM{} \tbPH{ŋmɪ́ɛ́lɪ́} \tbPS{adj.~} \tbGE{twisted} \tbXV{A nɔ pare ŋmeɛle na koŋ la nyɛ.} \tbXE{That fowl with the twisted tail will not lay eggs again.} 
\tbLX{ŋmegere} \tbHM{} \tbPH{ŋmɪ́gɪrɪ́} \tbPS{v.~} \tbGE{to shrug one's shoulders} \tbXV{Ba toŋ la a bie ka o ŋmegere o bɔgeŋmampegɛ kyɛ zagere.} \tbXE{When they sent the child, he shrugged his shoulders and refused to go.} \tbPD{ŋmegrɛɛ, ŋmegrɛ, ŋmegrɛ, ŋmegrebɛ, ŋmegraa, - } 
\tbLX{ŋmegere} \tbHM{} \tbPH{ŋmɪ́gɪrɪ́} \tbPS{v.~} \tbGE{to swallow with difficulty} \tbXV{A bɔbɔe ŋmegere la a pɔntere eŋ.} \tbXE{The duck was able to swallow the toad.} \tbPD{ŋmegrɛɛ, ŋmegrɛ, ŋmegrɛ, ŋmegrebɛ, ŋmegraa, - } 
\tbLX{ŋmenaa} \tbHM{} \tbPH{ŋmɪ́nàá} \tbPS{n.~} \tbSN \tbGE{sun} \tbSN \tbGE{noon} \tbSN \tbGE{daytime} 
\tbLX{ŋmenaa-diibu} \tbHM{} \tbPH{ŋmɪ́náá-dííbú} \tbPS{n.~} \tbGE{lunch} 
\tbLX{ŋmendaare} \tbHM{} \tbPH{ŋmɪ́ndáárɪ́} \tbPS{n.~} \tbGE{Sabbath} \tbOP{ ŋmendɛɛ} 
\tbLX{ŋmendie} \tbHM{} \tbPH{ŋmɪ̀ndìè} \tbPS{n.~} \tbGE{church} \tbOP{ŋmenderi} \tbTP{ŋmenderee} 
\tbLX{ŋmene} \tbHM{} \tbPH{ŋmɪ́nɪ̀} \tbPS{n.~} \tbGE{god, spirit} \tbOP{ŋmemɛ} \tbTP{ŋmennɛɛ} 
\tbLX{ŋmenɛŋmenɛ} \tbHM{} \tbPH{ŋmɪ́nɛ́ŋmɪ́nɛ́} \tbPS{adj.~} \tbGE{firmly and beautifully woven} 
\tbLX{ŋmentantuo} \tbHM{} \tbPH{ŋmɪ́ntántúó} \tbPS{n.~} \tbGE{hot or bright afternoon} \tbOP{ŋmentantuuri} \tbTP{ŋmentantuuree} 
\tbLX{ŋmentɔmpɔle} \tbHM{} \tbPH{ŋmɪ́ntɔ̀mpɔ́lɪ̀} \tbPS{n.~} \tbGE{late morning} \tbOP{ŋmentɔmpɔlɔ} 
\tbLX{ŋmentɔŋ} \tbHM{} \tbPH{ŋmɪ́ntɔ̀ŋ} \tbPS{n.~} \tbGE{sunshine, daylight} \tbOP{ŋmentɔnne} \tbTP{ŋmentɔnnɛɛ} 
\tbLX{ŋmentɔŋgɔŋ} \tbHM{} \tbPH{ŋmɪ́ntɔ̀ŋgɔ́ŋ} \tbPS{n.~} \tbGE{early afternoon (between 1 and 2 pm)} \tbOP{ŋmentɔŋgɔnne} \tbTP{ŋmentɔŋgɔnnɛɛ} 
\tbLX{ŋmentɔŋkpeɛraa} \tbHM{} \tbPH{ŋmɪ́ntɔ̀ŋkpɪ́ɛ́ráá} \tbPS{n.~} \tbGE{sunset} \tbOP{ŋmentɔŋkpeɛre} \tbTP{ŋmentɔŋkpeɛrɛɛ} 
\tbLX{ŋmen-yelyelli} \tbHM{} \tbPH{ŋmɪ́n-yèlyèllì} \tbPS{v.~} \tbGE{to divine, to prophesy} \tbPD{-, -, -, -, -, - } 
\tbLX{ŋmen-yɛlɛ} \tbHM{} \tbPH{ŋmɪ́n-yɛ́lɛ́} \tbPS{n.~} \tbGE{work for God} 
\tbLX{ŋmenzone} \tbHM{} \tbPH{ŋmɪ́nzʊ́nɪ́} \tbPS{n.~} \tbGE{blasphemy} \tbOP{ ŋmenzomɔ} \tbVA{ŋmenzoma}
\tbLX{ŋmeregɛ} \tbHM{} \tbPH{ŋmírɪ́gɛ́} \tbPS{adv.~} \tbGE{totally} \tbXV{Ba toŋ la a bi entuo na ka o zagere ŋmeregɛ.} \tbXE{They sent the lazy boy and he refused totally.} 
\tbLX{ŋmɛ} \tbHM{} \tbPH{ŋmɛ̀} \tbPS{v.~} \tbSN \tbGE{to beat, to hit, to strike} \tbSN \tbGE{to clap, to applaud} \tbPD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \tbSE{ŋmɛ kpɛlle} \tbGE{shout; scream} \tbSE{ŋmɛ gbelime} \tbGE{ring (bell)} 
\tbLX{ŋmɛ} \tbHM{} \tbPH{ŋmɛ̀} \tbPS{v.~} \tbGE{to thresh} \tbXV{Dagaaba maŋ ŋmɛ kyi kyɛ baŋ yeɛle.} \tbXE{Among the Dagaaba, millet is threshed before being winnowed out.} \tbPD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \tbSE{ŋmɛ kpɛlle} \tbGE{shout; scream} \tbSE{ŋmɛ gbelime} \tbGE{ring (bell)} 
\tbLX{ŋmɛ} \tbHM{} \tbPH{ŋmɛ̀} \tbPS{v.~} \tbGE{to bump into, to knock against} \tbXV{A lɔɛ ama la za᷉a᷉ ŋmɛ taa na.} \tbXE{These are the two cars that collided yesterday.} \tbPD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \tbSE{ŋmɛ kpɛlle} \tbGE{shout; scream} \tbSE{ŋmɛ gbelime} \tbGE{ring (bell)} 
\tbLX{ŋmɛ} \tbHM{} \tbPH{ŋmɛ̀} \tbPS{v.~} \tbGE{to fire, to shoot (gun)} \tbXV{Ba ŋmɛ la malefa ka a nuuli iri agere.} \tbXE{When they fired the gun, the birds flew away.} \tbPD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \tbSE{ŋmɛ kpɛlle} \tbGE{shout; scream} \tbSE{ŋmɛ gbelime} \tbGE{ring (bell)} 
\tbLX{ŋmɛ} \tbHM{} \tbPH{ŋmɛ̀} \tbPS{v.~} \tbGE{to begin to hatch} \tbPD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \tbSE{ŋmɛ kpɛlle} \tbGE{shout; scream} \tbSE{ŋmɛ gbelime} \tbGE{ring (bell)} 
\tbLX{ŋmɛ} \tbHM{} \tbPH{ŋmɛ̀} \tbPS{v.~} \tbGE{to apply by dusting (powder)} \tbPD{ŋmɛɛ, ŋmeɛrɛ, ŋmeɛrɛ, ŋmeɛrebɛ, ŋmeɛraa, ŋmeɛrebɛ } \tbSE{ŋmɛ kpɛlle} \tbGE{shout; scream} \tbSE{ŋmɛ gbelime} \tbGE{ring (bell)} 
\tbLX{ŋmɛgele} \tbHM{} \tbPH{ŋmɛ̀gɪ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to flatten} \tbXV{Kuuri ka o de tɔ ne a koŋkoŋ te ŋmɛgele o bare.} \tbXE{She hit the tin with a stone until she flattened it out.} \tbSN \tbGE{to suffer from hunger} \tbXV{Zaameŋ zaa ba ba maale diibu ka a bibiiri deɛ ŋmɛgelɛ.} \tbXE{They have cooked food yesterday and the children are suffering from hunger.} \tbPD{ŋmɛglɛɛ, ŋmɛglɛ, ŋmɛglɛ, ŋmɛglebɛ, ŋmɛglaa, ŋmɛglebɛ } 
\tbLX{ŋmɛgelɛŋ} \tbHM{} \tbPH{ŋmɛ́gɪ́lɛ́ŋ} \tbPS{adj.~} \tbGE{flat e.g. run-over container} 
\tbLX{ŋmɛgeloo} \tbHM{} \tbPH{ŋmɛ́gɪ́lʊ́ʊ́} \tbPS{adj.~} \tbGE{flattened out e.g. tin run over by a car} \tbOP{ŋmɛgelɛ} 
\tbLX{ŋmɛgere} \tbHM{} \tbPH{ŋmɛ̀gɪ̀rɪ̀} \tbPS{v.~} \tbGE{to press, to squeeze out} \tbXV{Ba ŋmɛgere la sunni yi a teɛ ŋa poɔ.} \tbXE{They have been able to squeeze some gum out of this tree.} \tbPD{ŋmɛgrɛɛ, ŋmɛgrɛ, ŋmɛgrɛ, ŋmɛgrebɛ, ŋmɛgraa, - } 
\tbLX{ŋmɛgere} \tbHM{} \tbPH{ŋmɛ̀gɪ̀rɪ̀} \tbPS{v.~} \tbGE{to weep inwardly, to suffer inwardly} \tbXV{Ba ŋmɛ la a bie kyɛ dabe᷉ɛ᷉ ba sage ka o koŋ ka o fa᷉a᷉ kyɛ ŋmɛgerɛ.} \tbXE{The child was beaten and, being afraid to cry, he wept inwardly.} \tbPD{ŋmɛgrɛɛ, ŋmɛgrɛ, ŋmɛgrɛ, ŋmɛgrebɛ, ŋmɛgraa, - } 
\tbLX{ŋmɛleŋmɛle} \tbHM{} \tbPH{ŋmɛ̀lɪ̀ŋmɛ́lɪ́} \tbPS{v.~} \tbGE{to writhe in pain} \tbXV{O ŋmɛ la a waabo ka o ŋmɛleŋmɛle.} \tbXE{He hit the snake and it writhed.} \tbPD{-, -, -, -, -, - } 
\tbLX{ŋmɛmɛɛ} \tbHM{} \tbPH{ŋmɛ̀mɛ̀ɛ̀} \tbPS{adj.~} \tbGE{flat and twisted e.g. nose} 
\tbLX{ŋmɛmm} \tbHM{} \tbPH{ŋmɛ́mm} \tbPS{adv.~} \tbGE{stand against something quietly (hide)} 
\tbLX{ŋmɛnne} \tbHM{} \tbPH{ŋmɛ́nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to squeeze with the fingers e.g. breastfeeding} \tbSN \tbGE{to press to feel (swollen leg)} \tbPD{ŋmɛnnɛɛ, ŋmɛnnɛ, ŋmɛnnɛ, ŋmɛnemɛ, ŋmɛnnaa, ŋmɛnemɛ } 
\tbLX{ŋmɛŋgɛlɛŋ} \tbHM{} \tbPH{ŋmɛ́ŋgɛ́lɛ́ŋ} \tbPS{adj.~} \tbGE{flattened out of shape} 
\tbLX{ŋmɛ ɔge} \tbHM{} \tbPH{ŋmɛ̀ ɔ̀gɪ́} \tbPS{v.~} \tbSN \tbGE{to burrow to make a nest} \tbXV{Soɔŋaa maŋ ŋmɛ la ɔge kpɛ gaŋ be.} \tbXE{The rabbit digs a warren where it sleeps.} \tbSN \tbGE{to settle down in one place} \tbXV{A pɔlebile nyɛ la n yɔɔpuulee a deɛ ŋmɛ te yiriŋ.} \tbXE{The young man saw my sister and has settled down in our house.} \tbPD{-, -, -, -, -, - } 
\tbLX{ŋmɛ sane} \tbHM{} \tbPH{ŋmɛ̀ sánɪ̀} \tbPS{v.~} \tbSN \tbGE{to make a net loss} \tbXV{Ka fooŋ da kommie sidi kɔɔ kyɛ te koɔre sidi lezaɛ anaare fo ŋmɛ la sane sidi lezare.} \tbXE{If you buy tomatoes for one hundred cedis and sell them for eighty cedis, you have made a loss of twenty cedis.} \tbSN \tbGE{to spend money on something that is not of much worth to oneself} \tbXV{Ka kuoriŋ ko fo yiriŋ fo maŋ ŋmɛ la sane.} \tbXE{When there is a funeral in your house you normally spend a lot.} \tbPD{-, -, -, -, -, - } 
\tbLX{ŋmobofo} \tbHM{} \tbPH{ŋmʊ̀bʊ̀fʊ̀} \tbPS{adj.~} \tbGE{big and ruffled} 
\tbLX{ŋmolle} \tbHM{} \tbPH{ŋmʊ̀llɪ̀} \tbPS{v.~} \tbGE{to stuff into the mouth (food)} \tbPD{ŋmollɛɛ, ŋmolla, ŋmolla, ŋmoleba, ŋmollaa, - } 
\tbLX{ŋmolɔŋŋmooraa} \tbHM{} \tbPH{ŋmolɔŋŋmooraa} \tbPS{n.~} \tbGE{a type of insect that feeds on yam leaves} \tbOP{ŋmolɔŋŋmoore} \tbTP{ŋmolɔŋŋmoorɛɛ} \tbVA{ŋmolaŋŋmooraa}
\tbLX{ŋmonne} \tbHM{} \tbPH{ŋmʊ́nnɪ́} \tbPS{v.~} \tbSN \tbGE{to make spotted} \tbSN \tbGE{to beautify} \tbPD{ŋmonnɛɛ, ŋmonnɔ, ŋmonnɔ, ŋmoneba, ŋmonnaa, - } 
\tbLX{ŋmonɔŋmonɔ} \tbHM{} \tbPH{ŋmʊ́náŋmʊ́ná} \tbPS{adj.~} \tbSN \tbGE{spotted} \tbSN \tbGE{beautiful} \tbVA{ŋmonaŋmona}
\tbLX{ŋmoo} \tbHM{} \tbPH{ŋmʊ́ʊ̀} \tbPS{v.~} \tbGE{to become scorched (green vegetation)} \tbPD{ŋmooɛ, ŋmoona, -, ŋmoonema, ŋmoonaa, - } 
\tbLX{ŋmoone} \tbHM{} \tbPH{ŋmʊ̀ʊ̀ne} \tbPS{v.~} \tbSN \tbGE{to mutter, to grumble} \tbSN \tbGE{to growl e.g. dog} \tbPD{ŋmoonɛɛ, ŋmoona, ŋmoona, ŋmoonema, ŋmoonaa, ŋmoonema } 
\tbLX{ŋmore} \tbHM{} \tbPH{ŋmʊ́rɪ́} \tbPS{v.~} \tbGE{to crush} \tbXV{Ka aloopeleeŋ le o maŋ ŋmore la o zaa.} \tbXE{An aircraft that crashes is always crushed completely.} \tbPD{ŋmorɛɛ, ŋmora, ŋmora, ŋmoreba, ŋmoraa, ŋmoreba } 
\tbLX{ŋmore} \tbHM{} \tbPH{ŋmʊ́rɪ́} \tbPS{v.~} \tbGE{to rush; to move suddenly and fast} \tbXV{A bie lere la ka o ma ŋmore te nyɔge.} \tbXE{The child was going to fall, but the mother quickly took hold of him.} \tbPD{ŋmorɛɛ, ŋmora, ŋmora, ŋmoreba, ŋmoraa, ŋmoreba } 
\tbLX{o} \tbHM{} \tbPH{ʊ̀} \tbPS{pronoun.~} \tbGE{s/he, it (third person singular)} 
\tbLX{ogi} \tbHM{} \tbPH{ògì} \tbPS{v.~} \tbGE{to bring out of the mouth} \tbPD{ogee, ogro, ogro, ogribo, ograa, - } 
\tbLX{ogi} \tbHM{} \tbPH{ógì} \tbPS{v.~} \tbSN \tbGE{to segregate; to discriminate} \tbSN \tbGE{to dislocate (joint)} \tbPD{ogee, ogro, ogro, ogribo, ograa, - } 
\tbLX{ogili} \tbHM{} \tbPH{ògìlì} \tbPS{adj.~} \tbGE{not properly fitted or seated (lid, cover)} 
\tbLX{oii} \tbHM{} \tbPH{óìì} \tbPS{interjection.~} \tbGE{expression of despair} 
\tbLX{oli-oli} \tbHM{} \tbPH{òlì-ólí} \tbPS{v.~} \tbGE{to be restless, to be unsettled} \tbPD{oli-olee, oli-olo, oli-olo, oli-olibo, oli-olaa, oli-olibo } 
\tbLX{onaŋ} \tbHM{} \tbPH{ʊ́náŋ} \tbPS{pronoun.~} \tbGE{that one (nonhuman, third person demonstrative)} 
\tbLX{onɔ} \tbHM{} \tbPH{ʊ́nɔ́} \tbPS{pronoun.~} \tbGE{s/he, it (third person, demonstrative)} \tbXV{Onɔ la na yi.} \tbXE{S/he is the one that will go out.} \tbVA{ona/one}
\tbLX{oŋgoloŋ} \tbHM{} \tbPH{óŋgólóŋ} \tbPS{adj.~} \tbGE{circular, spherical in shape and small} 
\tbLX{oo} \tbHM{} \tbPH{òó} \tbPS{use word.~} \tbGE{yes (acceptance)} 
\tbLX{ooi} \tbHM{} \tbPH{óóì} \tbPS{use word.~} \tbGE{yes (response to call)} 
\tbLX{o᷉o} \tbHM{} \tbPH{ʊ́᷉ʊ́᷉} \tbPS{use word.~} \tbGE{yes (agreement)} \tbVA{ɛ᷉ɛ}
\tbLX{oone} \tbHM{} \tbPH{ʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbSN \tbGE{to clear one's throat} \tbSN \tbGE{to push (giving birth)} \tbPD{oonɛɛ, oona, oona, oonema, oonaa, - } 
\tbLX{oore} \tbHM{} \tbPH{ʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to dash up against} \tbPD{oorɛɛ, oora, oora, ooreba, ooraa, - } 
\tbLX{o᷉-o᷉} \tbHM{} \tbPH{ʊ́-ʊ̂} \tbPS{use word.~} \tbGE{why so?} 
\tbLX{o᷉-o᷉} \tbHM{} \tbPH{ʊ́-ʊ̀} \tbPS{use word.~} \tbGE{oh no} 
\tbLX{ó᷉̊ó} \tbHM{} \tbPH{ʊ́᷉-ʊ̀᷉} \tbPS{interjection.~} \tbGE{expression of surprise} 
\tbLX{o᷉-o᷉o} \tbHM{} \tbPH{ʊ́᷉-ʊ́᷉ʊ́᷉} \tbPS{interjection.~} \tbGE{no (disagreement)} 
\tbLX{o᷉ɔ} \tbHM{} \tbPH{ʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{a pot with small holes all over; colander} \tbOP{oɔne} \tbTP{oɔnɛɛ} 
\tbLX{oɔle} \tbHM{} \tbPH{ʊ́ɔ́lɪ́} \tbPS{v.~} \tbGE{to happen gradually e.g. stomach ache} \tbPD{oɔlɛɛ, oɔlɔ, oɔlɔ, oɔlba, oɔlaa, - } 
\tbLX{oɔne} \tbHM{} \tbPH{ʊ̀ɔ̀nɪ̀} \tbPS{v.~} \tbGE{to grip firmly with claws} \tbPD{oɔnɛɛ, oɔnɔ, oɔnɔ, oɔnema, oɔnaa, - } 
\tbLX{ore} \tbHM{} \tbPH{ʊ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to wash very dirty things e.g clothes} \tbSN \tbGE{to make one suffer severely e.g. beat} \tbPD{-, -, -, -, -, - } 
\tbLX{ore} \tbHM{} \tbPH{ʊ́rɪ́} \tbPS{adj.~} \tbGE{filled to the brim} 
\tbLX{oree} \tbHM{} \tbPH{óre̊e̊} \tbPS{adj.~} \tbGE{round and open looking e.g. a hole} 
\tbLX{orɔba} \tbHM{} \tbPH{ʊ̂rɔ́bà} \tbPS{n.~} \tbGE{rubber} \tbOP{orɔbare} \tbTP{orɔbarɛɛ} 
\tbLX{ɔge} \tbHM{} \tbPH{ɔ́gɪ̀} \tbPS{n.~} \tbGE{ground den, lair, hole} \tbOP{ɔgre} \tbTP{ɔgrɛɛ} 
\tbLX{ɔgere} \tbHM{} \tbPH{ɔ́gɪ́rɪ́} \tbPS{v.~} \tbGE{to struggle to vomit; to retch} \tbPD{ɔgrɛɛ, ɔgrɔ, ɔgrɔ, ɔgreba, ɔgraa, - } 
\tbLX{ɔgere} \tbHM{} \tbPH{ɔ́gɪ́rɪ̀} \tbPS{v.~} \tbGE{to get choked e.g during eating when a child is unable to swallow a muscle of food and strains to bring the food out} \tbPD{-, -, -, -, -, - } 
\tbLX{ɔlaa} \tbHM{} \tbPH{ɔ́làà} \tbPS{n.~} \tbGE{large black fish} \tbOP{ɔlle} \tbTP{ɔllɛɛ} 
\tbLX{ɔlɔ-ɔlɔ} \tbHM{} \tbPH{ɔ̀lɔ̀-ɔ̀lɔ̀} \tbPS{adj.~} \tbGE{slimy and dirty} 
\tbLX{ɔlɔ-ɔlɔ} \tbHM{} \tbPH{ɔ̀lɔ̀-ɔ̀lɔ̀} \tbPS{adj.~} \tbGE{purple} 
\tbLX{ɔmm} \tbHM{} \tbPH{ɔ̀mm} \tbPS{adv.~} \tbGE{idling} 
\tbLX{ɔnne} \tbHM{} \tbPH{ɔ̀nnɪ̀} \tbPS{v.~} \tbGE{to smoke, to put under extreme heat (meat)} \tbPD{ɔnnɛɛ, ɔnnɔ, ɔnnɔ, ɔnema, ɔnnaa, - } 
\tbLX{ɔnsɔɔre} \tbHM{} \tbPH{ɔ̀nsɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{} \tbOP{ɔnsɔbɔ} \tbVA{ɔŋ-sɔɔre}
\tbLX{ɔŋe} \tbHM{} \tbPH{ɔ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to draw something liquid e.g. water, to fetch something powdery e.g. flour, soil} \tbSN \tbGE{to harvest vegetables} \tbSN \tbGE{to harvest dawdaw fruits or leaves} \tbPD{ɔŋɛɛ, ɔnnɔ, ɔnnɔ, ɔnema, ɔnnaa, ɔnnema } 
\tbLX{ɔŋe} \tbHM{} \tbPH{ɔ̀ŋ} \tbPS{v.~} \tbGE{to smell (bad odour)} \tbPD{ɔŋɛɛ, ɔnnɔ, ɔnnɔ, ɔnema, ɔnnaa, ɔnnema } 
\tbLX{ɔŋ-eŋ} \tbHM{} \tbPH{ɔ̀ŋ-ɪ̀ŋ} \tbPS{v.~} \tbGE{to flee} \tbPD{ɔŋ-eŋɛɛ, ɔnnɔ-ennɛ, ɔŋ-ennɛ, ɔŋ-enemɛ, ɔŋ-ennaa, ɔŋ-enemɛ } 
\tbLX{ɔŋɔ} \tbHM{} \tbPH{ɔ́ŋɔ́} \tbPS{n.~} \tbGE{wild mouse} \tbOP{ɔnne} \tbTP{ɔnnɛɛ} \tbVA{ɔŋa}
\tbLX{ɔŋ-ɔɔbo} \tbHM{} \tbPH{ɔ̀ŋ-ɔ́ɔ́bʊ́} \tbPS{n.~} \tbGE{valuable, useful} 
\tbLX{ɔŋ-sɔɔre} \tbHM{} \tbPH{ɔ̌ŋ-sɔ̊́ɔ́rɪ́} \tbPS{n.~} \tbGE{a type of wild mouse that lives along river banks; has a dark brown colour} \tbOP{ɔŋ-sɔbɔ} \tbTP{ɔŋ-sɔɔrɛɛ} \tbVA{ɔnsɔɔre}
\tbLX{ɔɔ} \tbHM{} \tbPH{ɔ̀ɔ̀} \tbPS{v.~} \tbSN \tbGE{to chew} \tbSN \tbGE{to ache; to throb with pain} \tbPD{ɔɔɛ, ɔɔrɔ, ɔɔrɔ, ɔɔreba, ɔɔraa, ɔɔreba } 
\tbLX{ɔɔ-doŋɔ} \tbHM{} \tbPH{ɔ̀ɔ̀-dʊ́ŋó} \tbPS{n.~} \tbSN \tbGE{intestinal worm} \tbSN \tbGE{roundworm} \tbOP{ɔɔ-donne} \tbTP{ɔɔ-donnɛɛ} 
\tbLX{ɔɔkpoŋi} \tbHM{} \tbPH{ɔ́ɔ́kpóŋ} \tbPS{n.~} \tbGE{winter} \tbOP{ɔɔkponni} \tbTP{ɔɔkponnee} 
\tbLX{ɔɔkyi} \tbHM{} \tbPH{ɔ́ɔ́ʧí} \tbPS{n.~} \tbGE{goose bumps i.e. small bumps on skin caused by cold} \tbOP{ɔɔkyiree} 
\tbLX{ɔ-ɔlle} \tbHM{} \tbPH{ɔ́-ɔ́llɪ̀} \tbPS{n.~} \tbGE{moss (usually green matter found in unkempt wet places e.g bath house)} \tbTP{ɔ-ɔllɛɛ} 
\tbLX{ɔɔnyuko᷉ɔ} \tbHM{} \tbPH{ɔ́ɔ́nyúkʊ́᷉ɔ́} \tbPS{n.~} \tbGE{a type of tree with edible fruits} \tbOP{ɔɔnyuko᷉ɔnɛɛ} \tbTP{ɔɔnyukoɔnoo} 
\tbLX{ɔɔnyuko᷉ɔ} \tbHM{} \tbPH{ɔɔnyúkʊ̀᷉ɔ́᷉} \tbPS{n.~} \tbGE{a type of small tree whose leaves are edible and have medicinal properties} \tbTP{ɔɔnyukoɔnɛɛ} 
\tbLX{ɔɔraa} \tbHM{} \tbPH{ɔ́ɔ́ráá} \tbPS{n.~} \tbGE{chewer} \tbOP{ɔɔreba} \tbTP{ɔɔrebɛɛ} 
\tbLX{ɔɔraa-piiri} \tbHM{} \tbPH{ɔ́ɔ́ráá-pìírí} \tbPS{n.~} \tbGE{safe abode} \tbOP{ɔɔraa-pie} 
\tbLX{ɔɔre} \tbHM{} \tbPH{ɔ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{cold} \tbOP{ɔɔrɛɛ} 
\tbLX{ɔɔtoorebare} \tbHM{} \tbPH{ɔ̀ɔ̀tʊ́ʊ́rɪ́bàrɪ̀} \tbPS{n.~} \tbGE{October (month)} 
\tbLX{ɔraa} \tbHM{} \tbPH{ɔ́ràà} \tbPS{n.~} \tbGE{a type of climbing plant with edible fruit} \tbOP{ɔrrɪ} \tbTP{ɔrrɛɛ} 
\tbLX{ɔraa} \tbHM{} \tbPH{ɔ́ràà} \tbPS{n.~} \tbGE{portable cage (for carrying chicks)} \tbOP{ɔrrɪ} \tbTP{ɔrrɛɛ} 
\tbLX{ɔre} \tbHM{} \tbPH{ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to prepare bean leaf broth for soup} \tbPD{ɔrɛɛ, ɔrɔ, ɔrɔ, ɔreba, ɔraa, ɔreba } 
\tbLX{ɔre} \tbHM{} \tbPH{ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to throw hard against e.g. ball against a player} \tbPD{ɔrɛɛ, ɔrɔ, ɔrɔ, ɔreba, ɔraa, ɔreba } 
\tbLX{ɔre} \tbHM{} \tbPH{ɔ̀rɪ́} \tbPS{adj.~} \tbGE{impotent} \tbOP{ɔrɔ} \tbTP{ɔrɛɛ} 
\tbLX{ɔre} \tbHM{} \tbPH{ɔ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to begin to rot} \tbSN \tbGE{to stink} \tbSN \tbGE{to be useless (person)} \tbPD{ɔrɛɛ, ɔrɔ, ɔrɔ, ɔreba, ɔraa, ɔreba } 
\tbLX{ɔregbɔrɔ} \tbHM{} \tbPH{ɔ̀rɪ̀gbɔ̀rɔ̀} \tbPS{adv.~} \tbGE{messy looking} 
\tbLX{ɔresɔglaa} \tbHM{} \tbPH{ɔ́rɪ́sɔ́gláà} \tbPS{n.~} \tbGE{a type of climbing plant with edible fruit} \tbOP{ɔresɔglɔ} 
\tbLX{ɔrɔsɔ} \tbHM{} \tbPH{ɔ̀rɔ̀sɔ̀} \tbPS{adv.~} \tbGE{paying dearly for being obstinate} \tbXV{O yeli ka o ba wono yɛlɛ, o taraa be ɔrɔsɔ.} \tbXE{He says he is hardheaded; he is up to suffer for it.} 
\tbLX{pa} \tbHM{} \tbPH{pá} \tbPS{v.~} \tbGE{to set up the beginning of something e.g. weaving} \tbPD{paɛ, paara, paara, pareba, paaraa, - } 
\tbLX{pa} \tbHM{} \tbPH{pá} \tbPS{interjection.~} \tbGE{an expression of surprise} \tbVA{paa}
\tbLX{pa} \tbHM{} \tbPH{pà} \tbPS{v.~} \tbGE{to be full} \tbPD{paɛ, paara, paara, pareba, paaraa, - } 
\tbLX{paa} \tbHM{} \tbPH{páá} \tbPS{interj.~} \tbGE{expression of fear and surprise} \tbVA{pa}
\tbLX{pa᷉a} \tbHM{} \tbPH{pá᷉à᷉} \tbPS{conjunction.~} \tbGE{then, and now} \tbXV{O waɛ la pa᷉a wuli o a toma.} \tbXE{He has come, you should then show him the work.} 
\tbLX{pa᷉a} \tbHM{} \tbPH{pá᷉à} \tbPS{adj.~} \tbGE{what else} \tbXV{Libie kyebe N pa᷉a la e wolɔ?} \tbXE{There is no money, what else can I do?.} 
\tbLX{pa᷉a} \tbHM{} \tbPH{pà᷉à᷉} \tbPS{v.~} \tbGE{to warn, to caution} \tbPD{pa᷉aɛ, paana, paana, paanema, paanaa, - } 
\tbLX{paalaa} \tbHM{} \tbPH{pááláá} \tbPS{adj.~} \tbGE{new} \tbOP{paale} \tbTP{paalɛɛ} 
\tbLX{paalaa} \tbHM{} \tbPH{pááláá} \tbPS{n.~} \tbSN \tbGE{raised dais, platform} \tbSN \tbGE{stage for laying a corpse in state} \tbOP{paale} \tbTP{paalɛɛ} 
\tbLX{paalaa} \tbHM{} \tbPH{pààláà} \tbPS{n.~} \tbGE{space in between four mounds} \tbOP{paale} \tbTP{paalɛɛ} 
\tbLX{paale} \tbHM{} \tbPH{pààlɪ̀} \tbPS{v.~} \tbSN \tbGE{to fill} \tbSN \tbGE{to add} \tbPD{paalɛɛ, paala, paala, paaleba, -, - } 
\tbLX{paalonwuluugane} \tbHM{} \tbPH{páálʊ́nwúlúúgánɪ̀} \tbPS{n.~} \tbGE{map} \tbOP{paalonwuluugama} 
\tbLX{paaloŋ} \tbHM{} \tbPH{páálʊ́ŋ} \tbPS{n.~} \tbGE{area of jurisdiction (country, district, kingdom)} 
\tbLX{paaloŋ} \tbHM{} \tbPH{páálʊ́ŋ} \tbPS{adj.~} \tbGE{first, fresh instance} 
\tbLX{paanaa} \tbHM{} \tbPH{pàànáá} \tbPS{n.~} \tbGE{ground cricket} \tbOP{paane} \tbTP{paanɛɛ} \tbVA{paaŋaa - mole cricket (Durant)}
\tbLX{paane} \tbHM{} \tbPH{pàànɪ̀} \tbPS{v.~} \tbGE{to blame} \tbPD{paanɛɛ, paana, paana, paanema, paanaa, - } 
\tbLX{paantakyeraa} \tbHM{} \tbPH{pààntáʧɪ́ráá} \tbPS{n.~} \tbGE{wild flying cricket} \tbOP{paantakyerre} 
\tbLX{paapaa} \tbHM{} \tbPH{páápàà} \tbPS{n.~} \tbGE{pope} \tbOP{paapaamine/paapaare} \tbTP{paapaarɛɛ} 
\tbLX{pa᷉apere} \tbHM{} \tbPH{pa᷉a᷉pɪ́rɪ́} \tbPS{n.~} \tbGE{paperlike spider egg sac} \tbOP{pa᷉ape᷉ɛ} \tbTP{pa᷉apeerɛɛ} 
\tbLX{paare} \tbHM{} \tbPH{páárɪ̀} \tbPS{n.~} \tbGE{vagina, vulva} \tbOP{paɛ} \tbTP{pɛɛrɛɛ} 
\tbLX{paare} \tbHM{} \tbPH{pààrɪ̀} \tbPS{v.~} \tbGE{to scrape e.g. roasted yam} \tbPD{paarɛɛ, paara, paara, paareba, paaraa, - } 
\tbLX{paare} \tbHM{} \tbPH{pààrɪ̀} \tbPS{v.~} \tbGE{to weed e.g. grass} \tbPD{paarɛɛ, paara, paara, paareba, paaraa, - } 
\tbLX{paare} \tbHM{} \tbPH{páárɪ̀} \tbPS{v.~} \tbGE{to pass (a test)} \tbPD{paarɛɛ, paara, paara, paareba, paaraa, - } 
\tbLX{paasepɔɔto} \tbHM{} \tbPH{páásɪ́pɔ́ɔ́tʊ̀} \tbPS{n.~} \tbGE{passport} \tbOP{paasepɔɔtere} \tbTP{paasepɔɔterɛɛ} 
\tbLX{pa᷉atakyeraa} \tbHM{} \tbPH{pá᷉á᷉táʧɪ́ráá} \tbPS{n.~} \tbGE{a type of cricket} \tbOP{pa᷉atakyerre} \tbTP{pa᷉atakyerɛɛ} 
\tbLX{paate} \tbHM{} \tbPH{páátɪ̀} \tbPS{interjection.~} \tbGE{expression of surprise astonishment e.g. one found in a challenging situation} 
\tbLX{paawa} \tbHM{} \tbPH{pááwà} \tbPS{n.~} \tbGE{power, electric light} \tbOP{paaware} \tbTP{paawarɛɛ} 
\tbLX{paawerɛ} \tbHM{} \tbPH{pááwɪ́rɛ̀} \tbPS{n.~} \tbGE{powder} \tbOP{paawerre} \tbTP{paawerɛɛ/paawerrɛɛ} 
\tbLX{pabo} \tbHM{} \tbPH{pábʊ́} \tbPS{adj.~} \tbGE{completely} 
\tbLX{padaa} \tbHM{} \tbPH{pàdàá} \tbPS{n.~} \tbGE{wood used to make a stage for a corpse} \tbOP{padaare} \tbTP{padaarɛɛ} 
\tbLX{page} \tbHM{} \tbPH{págɪ́} \tbPS{v.~} \tbGE{to hasten, to hurry} \tbPD{pagɛɛ, pagera, pagereba, pageraa, pagereba, pagepagera } 
\tbLX{pagere} \tbHM{} \tbPH{págrɪ̀} \tbPS{v.~} \tbGE{to compete, to challenge} \tbPD{pagrɛɛ, pagra, pagra, pagreba, pagraa, - } 
\tbLX{pakɔɔloŋ} \tbHM{} \tbPH{pákɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{female pubic hair} 
\tbLX{pakyoŋ} \tbHM{} \tbPH{pàʧʋ́ŋ} \tbPS{n.~} \tbGE{a large black ant with a painful sting} \tbOP{pakyonne} \tbTP{ pakyonnɛɛ.} 
\tbLX{palampiiri} \tbHM{} \tbPH{pálámpììrí} \tbPS{n.~} \tbSN \tbGE{butterfly} \tbSN \tbGE{moth} \tbOP{palampie} \tbTP{palampiiree} \tbVA{palampigri/palimpiiri}
\tbLX{palle} \tbHM{} \tbPH{pàllɪ̀} \tbPS{v.~} \tbGE{to plait, to braid} \tbPD{pallɛɛ, palla, palla, paleba, pallaa, - } 
\tbLX{pamm} \tbHM{} \tbPH{pámm} \tbPS{adj.~} \tbGE{straight onto the target} 
\tbLX{pampana} \tbHM{} \tbPH{pámpánà} \tbPS{adv.~} \tbGE{now} 
\tbLX{pampuntɛgɛ} \tbHM{} \tbPH{pámpùntɛ̀gɛ́} \tbPS{n.~} \tbGE{the act of being inward looking; concerned with only immediate relation} \tbOP{pampuntɛgre} 
\tbLX{panana} \tbHM{} \tbPH{pànànà} \tbPS{adv.~} \tbGE{lit up e.g. car headlight in the dark} 
\tbLX{panapana} \tbHM{} \tbPH{pànàpànà} \tbPS{adj.~} \tbGE{colourful, beautiful} 
\tbLX{pane} \tbHM{} \tbPH{pánɪ́} \tbPS{n.~} \tbGE{doorway cover made of straw} \tbOP{pama} \tbTP{panɛɛ} 
\tbLX{paŋe} \tbHM{} \tbPH{pàŋ} \tbPS{v.~} \tbSN \tbGE{to beat down wet rough surface to make it even (local way to plaster) e.g floor} \tbSN \tbGE{fig. walk without any footwear on, barefooted} \tbPD{paŋɛɛ, pana, pana, panema, panaa, - } 
\tbLX{paŋe} \tbHM{} \tbPH{páŋ} \tbPS{v.~} \tbSN \tbGE{to beat mercilessily} \tbSN \tbGE{to gather plenty of leaves} \tbSN \tbGE{to sew (akn)} \tbPD{paŋɛɛ, pana, pana, panema, panaa, - } 
\tbLX{papeɛmɛ} \tbHM{} \tbPH{pápɪ́ɛ́mɛ́} \tbPS{n.~} \tbGE{waist string with cowries} \tbOP{papeɛne} \tbTP{papeɛnɛɛ} 
\tbLX{papoɔ} \tbHM{} \tbPH{pàpʊ́ɔ́} \tbPS{n.~} \tbGE{space in between mounds} \tbOP{papoore} \tbTP{papoorɛɛ} 
\tbLX{papoɔ} \tbHM{} \tbPH{pápʊ́ɔ́} \tbPS{n.~} \tbGE{inner vagina} 
\tbLX{para} \tbHM{} \tbPH{párá} \tbPS{n.~} \tbGE{triangular loincloth} \tbOP{parre} \tbTP{parɛɛ} 
\tbLX{para} \tbHM{} \tbPH{pàrá} \tbPS{v.~} \tbGE{to walk quickly} \tbPD{parɛɛ, para, para, pareba, paraa, - } 
\tbLX{parapara} \tbHM{} \tbPH{pàràpàrà} \tbPS{adv.~} \tbGE{hurried manner} \tbXV{A tontonema da nyɔge la a toma paraparapara, ka ba baare a.} \tbXE{The worker started the work hurriedly and soon finished.} 
\tbLX{parata} \tbHM{} \tbPH{párátá} \tbPS{adj.~} \tbGE{white and spotless} \tbVA{perete}
\tbLX{pare} \tbHM{} \tbPH{párɪ̀} \tbPS{n.~} \tbGE{buttocks, bottom} \tbOP{parɛɛ} \tbVA{tɛgɛ}
\tbLX{pare} \tbHM{} \tbPH{párɪ̀} \tbPS{n.~} \tbGE{meaning} \tbXV{A yɛlɛ ŋa pare la wola?} \tbXE{What is the meaning of this?.} \tbXV{A yɛlɛ pare da ba kyaane.} \tbXE{The meaning of what was said was not clear.} \tbOP{parɛɛ} \tbVA{tɛgɛ}
\tbLX{pare} \tbHM{} \tbPH{párɪ̀} \tbPS{v.~} \tbGE{to understand} \tbXV{A bie da ba woŋi a yɛlɛ pare.} \tbXE{The child did not understand what was said.} \tbPD{-, -, -, -, -, - } 
\tbLX{pare} \tbHM{} \tbPH{pàrɪ́} \tbPS{n.~} \tbGE{raised tomb} \tbOP{parɛɛ} 
\tbLX{pare} \tbHM{} \tbPH{párɪ̀} \tbPS{postposition.~} \tbGE{under} \tbXV{A tontonema balɛɛ la gaa zeŋe teɛ pare ka ba pɛnne.} \tbXE{The workers were tired and sat under the tree to rest.} 
\tbLX{pare} \tbHM{} \tbPH{pàrè} \tbPS{v.~} \tbSN \tbGE{to place on top} \tbSN \tbGE{to overtake, to pass} \tbPD{parɛɛ, para, para, pareba, paraa, pareba } 
\tbLX{pare} \tbHM{} \tbPH{pàrè} \tbPS{adj.~} \tbGE{more than, over} 
\tbLX{pare-araa} \tbHM{} \tbPH{párɪ̀-áráá} \tbPS{n.~} \tbGE{the act of having time} \tbXV{N ba taa pare-araa na gaa be zenɛ, n toma yaga.} \tbXE{I have no time to go there today, because I have a lot of work.} 
\tbLX{parebaaraa} \tbHM{} \tbPH{párbàáráà} \tbPS{n.~} \tbGE{last one} \tbOP{parbaare} 
\tbLX{parebiri} \tbHM{} \tbPH{párɪ̀bírì} \tbPS{n.~} \tbGE{anus} \tbOP{parebie} \tbTP{parebiiree} \tbVA{parebogi}
\tbLX{parebogi} \tbHM{} \tbPH{párbógì} \tbPS{n.~} \tbGE{anus} \tbOP{parebogiri} \tbTP{parebogiree} \tbVA{parebiri}
\tbLX{parefe᷉ɛ} \tbHM{} \tbPH{párɪ́fɪ́᷉ɛ́} \tbPS{adv.~} \tbGE{in vain} \tbXV{N viiri la o yiri te bale kyɛ ka parefẽɛ.} \tbXE{I have gone to his house several times but all without success.} 
\tbLX{parefuri} \tbHM{} \tbPH{párfùrì} \tbPS{n.~} \tbGE{act of dragging one's feet e.g. cowardice} 
\tbLX{parekolee} \tbHM{} \tbPH{pàrèkòlèé} \tbPS{n.~} \tbGE{piglet} \tbOP{parekobilii} \tbVA{dobalee, dobabilii}
\tbLX{parekonɛne} \tbHM{} \tbPH{pàrɪ́kònɛ́nɪ̀} \tbPS{n.~} \tbGE{pork} \tbOP{parekonɛmɛ} \tbTP{porkonɛnnɛɛ} 
\tbLX{parekyara} \tbHM{} \tbPH{párʧárá} \tbPS{n.~} \tbGE{The part in between the legs e.g. when in astride position} \tbOP{parekyarre} \tbTP{parekyarɛɛ} \tbVA{parekyɛrɛ}
\tbLX{parekyɛrɛ} \tbHM{} \tbPH{párʧɛ́rɛ̀} \tbPS{n.~} \tbGE{} \tbOP{ parekyɛrre} \tbTP{parekyɛrɛɛ} \tbVA{parekyara}
\tbLX{parema} \tbHM{} \tbPH{párɪ́má} \tbPS{n.~} \tbGE{extra} 
\tbLX{parenyɔge} \tbHM{} \tbPH{párɲɔ́gɪ́} \tbPS{n.~} \tbGE{courage} 
\tbLX{parepe᷉ɛ} \tbHM{} \tbPH{párɪ̀pɪ̀ɛ̀} \tbPS{n.~} \tbGE{act of struggling it out with another} \tbXV{Ba yɛlɛ na e la parepe᷉ɛ yɛlɛ.} \tbXE{That problem of theirs is a difficult one.} 
\tbLX{parepe᷉ɛ᷉} \tbHM{} \tbPH{párɪ̀pɪ̀᷉ɛ̀᷉} \tbPS{n.~} \tbGE{struggle} 
\tbLX{parepi᷉i} \tbHM{} \tbPH{párɪ́pí᷉ì} \tbPS{n.~} \tbGE{the end of the anus} \tbOP{parepiime} 
\tbLX{parepoɔroŋ} \tbHM{} \tbPH{párɪ́pʊ́ɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{a type of wild mouse that reproduces prolifically} \tbOP{parepoɔrɔ} \tbTP{parepoɔrɛɛ} 
\tbLX{parepu᷉o᷉} \tbHM{} \tbPH{párɪ̀pù᷉ò᷉} \tbPS{n.~} \tbGE{fear} 
\tbLX{paresɛgere} \tbHM{} \tbPH{párɪ̀sɛ́grɪ̀} \tbPS{n.~} \tbGE{buttock} \tbOP{paresɛgɛ} \tbTP{paresɛgerɛɛ} 
\tbLX{paresunsugiri} \tbHM{} \tbPH{párɪ̀súnsúgrí} \tbPS{n.~} \tbGE{the knob-like appendage attached to the area above the anus of birds} \tbOP{paresunsugo} \tbTP{paresunsugree} 
\tbLX{paretaa} \tbHM{} \tbPH{pàrɪ̀táà} \tbPS{n.~} \tbGE{game} \tbTP{paretaarɛɛ} 
\tbLX{paretɛleŋ} \tbHM{} \tbPH{párɪ́tɛ́lɪ́ŋ} \tbPS{adj.~} \tbGE{at once} 
\tbLX{parɛɛ} \tbHM{} \tbPH{párɛ́ɛ́} \tbPS{n.~} \tbGE{species; kind} \tbOP{parɛɛ parɛɛ} 
\tbLX{parikonyaŋaa} \tbHM{} \tbPH{pàríkóɲáŋáá} \tbPS{n.~} \tbGE{sow (pig)} \tbOP{parikonyaŋene} 
\tbLX{pataase} \tbHM{} \tbPH{pàtáásɪ́} \tbPS{n.~} \tbGE{type of local spirit (alcoholic drink)} \tbOP{pataasere} \tbTP{pataaserɛɛ} 
\tbLX{pazɔŋɔ} \tbHM{} \tbPH{pàzɔ̌ŋ} \tbPS{n.~} \tbGE{blood-sucking fly} \tbOP{pazɔnne} \tbTP{pazɔnnɛɛ} \tbVA{pazɔŋa}
\tbLX{pe} \tbHM{} \tbPH{pɪ́} \tbPS{v.~} \tbGE{to be fed up with} \tbPD{peɛ, peerɛ, peerɛ, peerebɛ, peeraa, peerebɛ } 
\tbLX{pedaa} \tbHM{} \tbPH{pɪ́dáá} \tbPS{n.~} \tbGE{ram} \tbOP{padaare} \tbTP{pedaɛ/padaarɛɛ} 
\tbLX{peda-eelɛ} \tbHM{} \tbPH{pɪ́dá-ɪ́ɪ́lɛ́} \tbPS{n.~} \tbGE{male sheep with horns} 
\tbLX{pedavaraa} \tbHM{} \tbPH{pídàváràà} \tbPS{n.~} \tbGE{castrated ram} \tbOP{pedavarre} \tbTP{pedavarɛɛ} 
\tbLX{pee} \tbHM{} \tbPH{pèè} \tbPS{adv.~} \tbGE{open and large} 
\tbLX{peelaa} \tbHM{} \tbPH{pɪ́ɪ́láà} \tbPS{adj.~} \tbGE{infant} \tbOP{peele} 
\tbLX{peele} \tbHM{} \tbPH{pɪ́ɪ́lɪ̀} \tbPS{v.~} \tbGE{to tear into shreds e.g. rag} \tbPD{peelɛɛ, peelɛ, peelɛ, peelbɛ, peelaa, peelbɛ } 
\tbLX{peere} \tbHM{} \tbPH{pɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to sweep, to dust, to brush off (shirt)} \tbSN \tbGE{to cleanse; to reinstate someone's soul believed to have deserted her/him by performing certain rituals to cleanse the person of an evil spirit, which is believed to have overshadowed the person’s actual soul} \tbPD{peerɛɛ, peerɛ, peerebɛ, peeraa, -, - } 
\tbLX{peɛ} \tbHM{} \tbPH{pɪ́ɛ́} \tbPS{n.~} \tbGE{roof} \tbOP{peere} \tbTP{peerɛɛ} 
\tbLX{peɛ} \tbHM{} \tbPH{pɪ̀ɛ́} \tbPS{n.~} \tbGE{basket} \tbOP{pɛre} \tbTP{pɛrɛɛ} 
\tbLX{peɛ} \tbHM{} \tbPH{pɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to drink (taking in large chunks)} \tbSN \tbGE{to take a wife or house maid from her parents} \tbPD{peɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{pe᷉} \tbHM{} \tbPH{pɪ̀᷉} \tbPS{adv.~} \tbGE{onomatopoeic word (weighty object smashing a smaller object)} 
\tbLX{pe᷉e} \tbHM{} \tbPH{pɪ́᷉ɪ́᷉} \tbPS{adj.~} \tbGE{unrelenting, steadfast, not giving way} 
\tbLX{pe᷉epe᷉e} \tbHM{} \tbPH{pɪ́᷉ɪ́᷉pɪ́᷉ɪ́᷉} \tbPS{adv.~} \tbGE{very sharp (taste, sting)} 
\tbLX{peeteree} \tbHM{} \tbPH{pɪ́ɪ́tɪ́rɪ́ɪ́} \tbPS{adj.~} \tbGE{everything about something, the whole truth} \tbXV{A polisiri da soore la noba yaga a na baŋe a yɛlɛ peeteree zaa.} \tbXE{The police questioned many people in a bid to get to the bottom of the matter.} \tbTP{peeterɛɛ} 
\tbLX{pe᷉ɛ} \tbHM{} \tbPH{pɪ́᷉ɛ́᷉} \tbPS{adv.~} \tbGE{completely e.g. finished} \tbXV{A ko᷉ɔ zaaŋ baare pe᷉ɛ.} \tbXE{All the water is finished completely.} 
\tbLX{pe᷉ɛ} \tbHM{} \tbPH{pɪ̀᷉ɛ̀᷉} \tbPS{v.~} \tbGE{pull apart in a bid to open e.g pulling the eyelid(s) apart} \tbPD{pe᷉ɛɛ, peɛnɛ, peɛnɛ, peɛnemɛ, peɛnaa, - } 
\tbLX{pe᷉ɛ} \tbHM{} \tbPH{pɪ̀᷉ɛ̀᷉} \tbPS{v.~} \tbSN \tbGE{to decorate well} \tbSN \tbGE{to dress gorgeously} \tbPD{pe᷉ɛɛ, peɛnɛ, peɛnɛ, peɛnemɛ, peɛnaa, - } 
\tbLX{pe᷉ɛ} \tbHM{} \tbPH{pɪ̀᷉ɛ̀᷉} \tbPS{v.~} \tbGE{to court} \tbPD{pe᷉ɛɛ, peɛnɛ, peɛnɛ, peɛnemɛ, peɛnaa, - } 
\tbLX{peɛle} \tbHM{} \tbPH{pɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{} \tbPD{peɛlɛɛ, peɛlɛ, peɛlɛ, peɛlebɛ, peɛlaa, peɛlebɛ } \tbSE{peɛle eelee} \tbGE{to blow a horn} 
\tbLX{peɛle} \tbHM{} \tbPH{pɪ́ɛ́lɪ̀} \tbPS{v.~} \tbGE{to be near; to be close to} \tbPD{peɛlɛɛ, peɛlɛ, peɛlɛ, peɛlebɛ, peɛlaa, peɛlebɛ } 
\tbLX{peɛle-peɛle} \tbHM{} \tbPH{pɪ́ɛ́lɪ́-pɪ́ɛ́lɪ́} \tbPS{adj.~} \tbGE{having white spots} 
\tbLX{peɛloŋ} \tbHM{} \tbPH{pɪ́ɛ́lʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{whiteness} \tbSN \tbGE{food} 
\tbLX{peɛmɛ} \tbHM{} \tbPH{pɪ́ɛ́mɛ́} \tbPS{n.~} \tbSN \tbGE{decorations} \tbSN \tbGE{beautiful clothing} \tbOP{peɛne} \tbTP{peɛnɛɛ/peɛmɛɛ} 
\tbLX{peɛmo} \tbHM{} \tbPH{pɪ́ɛ́mʊ́} \tbPS{n.~} \tbSN \tbGE{courtship} \tbSN \tbGE{sexual intercourse} 
\tbLX{pe᷉ɛo} \tbHM{} \tbPH{pɪ́᷉ɛ́᷉ʊ́᷉} \tbPS{n.~} \tbGE{a type of wild cat (dark spotted)} \tbOP{peɛne} \tbTP{peɛnɛɛ} 
\tbLX{peɛre} \tbHM{} \tbPH{pɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to investigate} \tbPD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{peɛre} \tbHM{} \tbPH{pɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to not be fussy about choices (typically used with negative marker ba)} \tbPD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{peɛre} \tbHM{1} \tbPH{pɪ́ɛ́rɪ́} \tbPS{v.~} \tbGE{to line up; to place side by side} \tbPD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{peɛre} \tbHM{2} \tbPH{pɪ́ɛ́rɪ́} \tbPS{v.~} \tbGE{to tap (to call attention)} \tbPD{peɛrɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{peɛrɛ} \tbHM{} \tbPH{pɪ́ɛ́rɛ̀} \tbPS{n.~} \tbGE{a type of small, black potato} \tbSG{peɛroo} \tbOP{peɛrɛ} \tbTP{peɛrɛɛ} 
\tbLX{pege} \tbHM{} \tbPH{pɪ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to decorate (dress)} \tbSN \tbGE{to make tasty, delicious} \tbPD{pegɛɛ, pegrɛ, pegrɛ, pegrebɛ, pegraa, - } 
\tbLX{pegelaa} \tbHM{} \tbPH{pìgláà} \tbPS{n.~} \tbGE{row (of farm mounds)} \tbOP{pegle} \tbTP{peglɛɛ} 
\tbLX{pegele} \tbHM{} \tbPH{pɪ̀glɪ̀} \tbPS{v.~} \tbGE{cut across, in order to meet at a point} \tbPD{peglɛɛ, peglɛ, peglɛ, peglebɛ, peglaa, - } 
\tbLX{pegele} \tbHM{} \tbPH{pɪ̀glɪ̀} \tbPS{adj.~} \tbGE{not staight on e.g. feet} 
\tbLX{pegerɛ} \tbHM{} \tbPH{pɪ́grɛ̀} \tbPS{n.~} \tbGE{wings} \tbSG{pegroo} 
\tbLX{pegɛ} \tbHM{} \tbPH{pɪ́gɛ́} \tbPS{adj.~} \tbGE{not whole; e.g. broken piece of calabash} 
\tbLX{pegi} \tbHM{} \tbPH{pègì} \tbPS{v.~} \tbGE{to flap one's wings} \tbPD{pegee, pegre, pegre, pegribe, pegraa, - } 
\tbLX{pekɔɔloŋ} \tbHM{} \tbPH{pɪ́kɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{wool} 
\tbLX{pekpoŋi} \tbHM{} \tbPH{pɪ̀kpôŋ} \tbPS{n.~} \tbGE{large roof} \tbOP{pekponni} \tbTP{pekponnee} 
\tbLX{pekyeenɛ} \tbHM{} \tbPH{ṕʧɪ́ɪ́nɛ́} \tbPS{n.~} \tbGE{shepherd} \tbOP{pekyeemɛ} \tbTP{pekyeenɛɛ} 
\tbLX{pelaa} \tbHM{} \tbPH{pɪ́làá} \tbPS{adj.~} \tbGE{white} \tbOP{peɛle} 
\tbLX{pelaaya (Eng)} \tbHM{} \tbPH{pɪ́lááyà} \tbPS{n.~} \tbGE{pliers} \tbOP{pelaayare} \tbTP{pelaayarɛɛ} 
\tbLX{pele} \tbHM{} \tbPH{pɪ́lɪ̀} \tbPS{v.~} \tbSN \tbGE{to split a piece from a stalk} \tbSN \tbGE{to leave and fail to return} \tbPD{pelɛɛ, pellɛ, pellɛ, pelebɛ, pelaa, - } 
\tbLX{pele} \tbHM{} \tbPH{pɪ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to tear, to break off (splinter)} \tbSN \tbGE{to discontinue (visits)} \tbSN \tbGE{to continue without going back to former place} \tbPD{pelɛɛ, pellɛ, pellɛ, pelbɛ, pellaa, - } 
\tbLX{pelee} \tbHM{} \tbPH{pɪ́léè} \tbPS{n.~} \tbGE{lamb} \tbOP{pebilii} 
\tbLX{pelee} \tbHM{} \tbPH{pèléè} \tbPS{n.~} \tbGE{small basket} \tbOP{ pebilii} \tbVA{pɛlee}
\tbLX{peleemaze᷈e᷈} \tbHM{} \tbPH{pɪ́léèmàzɪ́᷈ɪ̀᷈} \tbPS{n.~} \tbGE{a type of with pink coloured edible fruits} 
\tbLX{pelempeleŋ} \tbHM{} \tbPH{pélémpéléŋ} \tbPS{adv.~} \tbGE{bare, empty} 
\tbLX{pelewelee} \tbHM{} \tbPH{pɪ̀liɪ̀wɪ̀lɪ̀ɪ̀} \tbPS{adj.~} \tbGE{plenty of things found everywhere e.g. assorted items} \tbXV{A weɛ tomɔ deɛ waa la pelewelee lɛ.} \tbXE{There is plenty of work on the farm.} \tbVA{pelewele}
\tbLX{pelɛmpelɛŋ} \tbHM{} \tbPH{pɪ̀lɛ̀mpɪ́lɛ́ŋ} \tbPS{adv.~} \tbGE{all over the place, disorganised} 
\tbLX{peli} \tbHM{} \tbPH{pélì} \tbPS{n.~} \tbGE{spirit, heart} \tbOP{pele} \tbTP{pelee} 
\tbLX{peli} \tbHM{} \tbPH{pél} \tbPS{n.~} \tbGE{fright} 
\tbLX{pell} \tbHM{} \tbPH{pɪ́ll} \tbPS{adj.~} \tbGE{full to capacity} \tbVA{pelpel}
\tbLX{pell} \tbHM{} \tbPH{péll} \tbPS{adj.~} \tbGE{rise into the air e.g. flame} 
\tbLX{pelɔɔ} \tbHM{} \tbPH{pɪ̀lɔ̀ɔ̀} \tbPS{adv.~} \tbGE{suddenly and quickly e.g fish splashing in water} 
\tbLX{pelpel} \tbHM{} \tbPH{pɪ́lpɪ́l} \tbPS{adv.~} \tbGE{full to the brim} 
\tbLX{pelpele} \tbHM{} \tbPH{pìlpɪ́l} \tbPS{v.~} \tbGE{to litter with broken pieces e.g. of straw} \tbPD{pelpelɛɛ, pelpelɛ, pelpelɛ, pelpelebɛ, pelpelaa, - } 
\tbLX{pempeleŋ} \tbHM{} \tbPH{pémpéléŋ} \tbPS{adj.~} \tbGE{small e.g. piece of cloth} 
\tbLX{pemperi} \tbHM{} \tbPH{pèmpèrì} \tbPS{v.~} \tbGE{to bend or place something in an oblique position} \tbPD{pemperee, pempere, pempere, pemperibe, pemperaa, - } \tbVA{pɛmpɛre}
\tbLX{penɛne} \tbHM{} \tbPH{pɪ́nɛ́nɪ̀} \tbPS{n.~} \tbGE{mutton} \tbOP{penɛmɛ} \tbTP{penɛnnɛɛ} 
\tbLX{penne} \tbHM{} \tbPH{pɪ́nnɪ̀} \tbPS{v.~} \tbGE{to make delicious} \tbPD{pennɛɛ, pennɛ, pennɛ, penemɛ, pennaa, - } 
\tbLX{penti} \tbHM{} \tbPH{péntì} \tbPS{n.~} \tbGE{paint} \tbOP{pentiri} \tbTP{pentiree} 
\tbLX{penyaŋaa} \tbHM{} \tbPH{pɪ́ɲáŋàá} \tbPS{n.~} \tbGE{ewe, adult female sheep} \tbOP{penyaŋene/penyanne} \tbTP{penyaŋenɛɛ/penyannɛɛ} 
\tbLX{peraa} \tbHM{} \tbPH{pɪ́ráá} \tbPS{n.~} \tbSN \tbGE{deliberateness} \tbSN \tbGE{obstinacy} 
\tbLX{peraa} \tbHM{} \tbPH{pɪ́ráá} \tbPS{n.~} \tbGE{hyperbole (literature)} 
\tbLX{peraa} \tbHM{} \tbPH{pɪ́ràà} \tbPS{n.~} \tbGE{castanet; a pair of metal rings worn on the fingers and clicked together rhythmically} \tbOP{perre} \tbTP{perrɛɛ} 
\tbLX{pere} \tbHM{} \tbPH{pɪ́rɪ̀} \tbPS{v.~} \tbGE{push away; usually a small object with the foot} \tbPD{-, -, -, -, -, - } 
\tbLX{pere} \tbHM{} \tbPH{pɪ́rɪ́} \tbPS{n.~} \tbGE{a loop used in making the rim of a basket} \tbOP{perɛɛ} 
\tbLX{pere} \tbHM{} \tbPH{pɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{remove from; shell (groundnuts), unwrap (parcel),} \tbSN \tbGE{get wounded} \tbXV{O daare zɔ᷉ɔ᷉la kuriwiri te le ka o gbɛre pere.} \tbXE{He rode a bicycle, crashed and got his leg wounded.} \tbPD{perɛɛ, perɛ, perɛ, perebɛ, peraa, - } 
\tbLX{pere} \tbHM{} \tbPH{pɪ́rɪ̀} \tbPS{v.~} \tbGE{push away (small object), ignore} \tbXV{Ka fooŋ ba boɔrɔ a kuuri ŋa kyɛ ta pere o bare.} \tbXE{Even if you don't like this stone don't push it away.} \tbPD{perɛɛ, perɛ, perɛ, perebɛ, peraa, - } 
\tbLX{pere} \tbHM{} \tbPH{pɪ́rɪ̀} \tbPS{n.~} \tbGE{} \tbOP{peere} \tbVA{peroo}
\tbLX{perebare} \tbHM{} \tbPH{pɪ́rɪ́bàrɪ̀} \tbPS{n.~} \tbGE{outer toe of a fowl} \tbOP{perebarre} \tbTP{perebarɛɛ} 
\tbLX{perekara} \tbHM{} \tbPH{pɪ̀rɪ̀kàrà} \tbPS{adj.~} \tbGE{astride, firm and immovable} \tbXV{A sogya yaare la o gbɛɛ perekara ka tɔloo zie kyebe.} \tbXE{The soldier stood firmly with his feet astride such that no one can pass by him.} 
\tbLX{perekatɔɔre} \tbHM{} \tbPH{pɪ́rɪ́kátɔ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{cemetery} \tbOP{perekatɔɛ} \tbTP{perekatɔɔrɛɛ} 
\tbLX{perepere} \tbHM{} \tbPH{pìrɪ̀pɪ́rɪ́} \tbPS{v.~} \tbGE{to struggle e.g. insect on a hot object} \tbPD{pereperɛɛ, pereperɛ, pereperɛ, pereperebɛ, pereperaa, - } 
\tbLX{perete} \tbHM{} \tbPH{pɪ́rɪ́tɪ́} \tbPS{adv.~} \tbGE{spotlessly white} \tbVA{parata}
\tbLX{perɛperɛ} \tbHM{} \tbPH{pɪ́rɛ́pɪ́rɛ́} \tbPS{adj.~} \tbGE{not stuck together e.g. well-cooked rice} 
\tbLX{periperi} \tbHM{} \tbPH{pèrìpérí} \tbPS{v.~} \tbGE{to make an effort to overcome, to try hard} \tbXV{O ba taa a eebo kyɛ o meŋ aŋ peripere la.} \tbXE{She has not got the means, but she is also still trying hard.} \tbPD{periperee, peripere, peripere, periperibe, periperaa, - } 
\tbLX{peroo} \tbHM{} \tbPH{pɪ́rʊ́ʊ́} \tbPS{n.~} \tbGE{sheep} \tbOP{peere} \tbTP{peerɛɛ} \tbVA{pere}
\tbLX{pesaraa} \tbHM{} \tbPH{pɪ́sáráà} \tbPS{n.~} \tbGE{young adult female sheep} \tbSG{pesarrɛɛ} \tbOP{pesarre} 
\tbLX{petooŋmaa} \tbHM{} \tbPH{pɪ́tòòŋmáá} \tbPS{n.~} \tbGE{one who will not listen to anything} \tbOP{petooŋmaara} \tbTP{petooŋmaarɛɛ} 
\tbLX{pezage} \tbHM{} \tbPH{pɪ́zàgɪ́} \tbPS{n.~} \tbGE{sheep pen} \tbOP{pezagre} \tbTP{pezagrɛɛ} 
\tbLX{pɛ} \tbHM{} \tbPH{pɛ̀} \tbPS{v.~} \tbGE{to squeeze out e.g. milk from a cow} \tbPD{-, -, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{pɛ} \tbHM{} \tbPH{pɛ̀} \tbPS{v.~} \tbGE{to open one's eyes widely} \tbPD{-, -, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{pɛ} \tbHM{} \tbPH{pɛ̀} \tbPS{v.~} \tbGE{to spiritually deprive another of prowess} \tbPD{-, -, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{pɛ} \tbHM{} \tbPH{pɛ́} \tbPS{v.~} \tbGE{to harvest nuts from the ground} \tbPD{pɛɛ, peɛrɛ, peɛrɛ, peɛrebɛ, peɛraa, - } 
\tbLX{pɛ} \tbHM{} \tbPH{pɛ́} \tbPS{adj.~} \tbGE{same, equal} \tbVA{pɛpɛ}
\tbLX{pɛ᷉ɛ} \tbHM{} \tbPH{pɛ̀᷉ɛ̀᷉} \tbPS{adv.~} \tbGE{fatty; oily} 
\tbLX{pɛɛre} \tbHM{} \tbPH{pɛ́ɛ́rɪ́} \tbPS{adj.~} \tbGE{immature} \tbOP{pɛbɛ} 
\tbLX{pɛɛtɛɛre} \tbHM{} \tbPH{pɛ́ɛ́tɛ́ɛ́rɪ́} \tbPS{n.~} \tbGE{small, flat fish; tilapia} \tbSG{pɛɛtɛroo} \tbOP{pɛɛtɛbɛ} \tbTP{pɛɛtɛbɛɛ} \tbVA{pɛɛtere}
\tbLX{pɛge} \tbHM{} \tbPH{pɛ́gɪ̀} \tbPS{v.~} \tbGE{to wash e.g. clothes, utensils} \tbPD{-, pɛgrɛ, pɛgrɛ, pɛgrebɛ, -, - } 
\tbLX{pɛge} \tbHM{} \tbPH{pɛ́gɪ́} \tbPS{n.~} \tbGE{hard external covering of seeds, nuts, animals, trees} \tbOP{pɛgre/pɛgɛ} \tbTP{pɛgrɛɛ} 
\tbLX{pɛgele} \tbHM{} \tbPH{pɛ̀glɪ̀} \tbPS{v.~} \tbGE{to carry on one's side, shoulder, in arms} \tbPD{pɛglɛɛ, pɛglɛ, pɛglɛ, pɛglebɛ, pɛglaa, - } 
\tbLX{pɛge yi} \tbHM{} \tbPH{pɛ́gɪ̀ yí} \tbPS{adj.~} \tbGE{clean, bright} 
\tbLX{pɛgɛkɛ} \tbHM{} \tbPH{pɛ̀gɛ̀kɛ̀} \tbPS{adj.~} \tbGE{deep and wide} 
\tbLX{pɛglɛŋ} \tbHM{} \tbPH{pɛ́glɛ́ŋ} \tbPS{n.~} \tbGE{small and flat} \tbOP{pɛglɛŋ pɛglɛŋ} \tbVA{pɛŋgɛlɛŋ}
\tbLX{pɛkogo} \tbHM{} \tbPH{pɛ̀kógó} \tbPS{n.~} \tbGE{sticks put under a basket as a protective base} \tbOP{pɛkogri} \tbTP{pɛkogree} 
\tbLX{pɛkpoŋ} \tbHM{} \tbPH{pɛ̀kp̂oŋ} \tbPS{n.~} \tbGE{large basket} \tbOP{pɛkponni} \tbTP{pɛkponnee} 
\tbLX{pɛlaa} \tbHM{} \tbPH{pɛ́làà} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{pɛlle} \tbTP{pɛllɛɛ} 
\tbLX{pɛle} \tbHM{} \tbPH{pɛ̀lɪ̀} \tbPS{v.~} \tbGE{to be white} \tbPD{-, -, -, -, -, - } 
\tbLX{pɛle} \tbHM{} \tbPH{pɛ̀lɪ̀} \tbPS{v.~} \tbGE{to open up; be happy} \tbXV{Zenɛ o pɛle la poɔ manne a yɛlɛ ko te.} \tbXE{Today s/he open up by explaining the issues to us.} \tbPD{-, -, -, -, -, - } 
\tbLX{pɛle} \tbHM{} \tbPH{pɛ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to peel (yam)} \tbSN \tbGE{to clear grass by weeding with a hoe} \tbPD{-, -, -, -, -, - } 
\tbLX{pɛmpɛlɛŋ} \tbHM{} \tbPH{pɛ́mpɛ́lɛ́ŋ} \tbPS{n.~} \tbGE{flat metal (cover of a tin)} \tbOP{pɛmpɛlle} \tbTP{pɛmpɛlɛɛ} 
\tbLX{pɛmpɛnɛɛ} \tbHM{} \tbPH{pɛ́mpɛ́nɛ́ɛ́} \tbPS{n.~} \tbGE{wild gum; rubber} \tbOP{pɛmpɛnne} \tbTP{pɛmpɛnnɛɛ} 
\tbLX{pɛmpɛre} \tbHM{} \tbPH{pɛ̀mpɛ̀rɪ̀} \tbPS{n.~} \tbGE{oblique position (bending very low)} 
\tbLX{pɛmpɛre} \tbHM{} \tbPH{pɛ̀mpɛ̀rɪ̀} \tbPS{v.~} \tbGE{to bend very low and in an oblique position} \tbPD{pɛmpɛrɛɛ, pɛmpɛrɛ, pɛmpɛrɛ, pɛmpɛrebɛ, pɛmpɛraa, - } 
\tbLX{pɛnaa} \tbHM{} \tbPH{pɛ̀náà} \tbPS{n.~} \tbGE{musical instrument made from seasoned straw} \tbOP{pɛnne} \tbTP{pɛnnɛɛ} 
\tbLX{pɛnne} \tbHM{} \tbPH{pɛ̀nnɪ̀} \tbPS{v.~} \tbGE{to carve} \tbPD{pɛnnɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, - } 
\tbLX{pɛnne} \tbHM{} \tbPH{pɛ̀nnɪ̀} \tbPS{v.~} \tbGE{to rest} \tbPD{pɛnnɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, - } 
\tbLX{pɛnne} \tbHM{} \tbPH{pɛ̀nnɪ́} \tbPS{n.~} \tbGE{lower part of abdomen} \tbOP{pɛmɛ} \tbTP{pɛnnɛɛ} 
\tbLX{pɛnne} \tbHM{} \tbPH{pɛ́nnɪ̀} \tbPS{n.~} \tbGE{cloth} \tbOP{pɛmɛ} \tbTP{pɛnnɛɛ} 
\tbLX{pɛnnoŋ} \tbHM{} \tbPH{pɛ̀nnʊ̂ŋ} \tbPS{n.~} \tbGE{} \tbVA{pɛnnoo}
\tbLX{pɛnnoo} \tbHM{} \tbPH{pɛ̀nnʊ̀ʊ́} \tbPS{n.~} \tbGE{air, breeze} \tbVA{pɛnnoŋ}
\tbLX{pɛnnu} \tbHM{} \tbPH{pɛ́nnù} \tbPS{n.~} \tbGE{hand measure (about 9 inches) of woven strip of cloth} \tbOP{pɛnnuuri} \tbTP{pɛnnuuree} 
\tbLX{pɛnzage} \tbHM{} \tbPH{pɛ́nzágɪ́} \tbPS{n.~} \tbGE{rag} \tbOP{pɛnzagra} \tbTP{pɛnzagrɛɛ} 
\tbLX{pɛŋe} \tbHM{} \tbPH{pɛ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to groan} \tbSN \tbGE{to grunt (from effort)} \tbPD{pɛŋɛɛ, pɛnɛ, pɛnɛ, pɛnemɛ, pɛnaa, pɛnemɛ } 
\tbLX{pɛŋe} \tbHM{} \tbPH{pέŋì} \tbPS{} \tbGE{to skim (usually by tapping water with the back of the hand or a container to push away debris)} 
\tbLX{pɛŋe} \tbHM{} \tbPH{pɛ̂ŋ} \tbPS{v.~} \tbSN \tbGE{to borrow} \tbSN \tbGE{to lend} \tbPD{pɛŋɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, pɛnemɛ } 
\tbLX{pɛŋe} \tbHM{} \tbPH{pɛ̂ŋ} \tbPS{v.~} \tbGE{to tap on water to clear the top of unwanted material} \tbPD{pɛŋɛɛ, pɛnnɛ, pɛnnɛ, pɛnemɛ, pɛnnaa, pɛnemɛ } \tbVA{bɛŋ}
\tbLX{pɛŋkɔɔloŋ} \tbHM{} \tbPH{pɛ̌ŋkɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{pubic hair} 
\tbLX{pɛpɛlaa} \tbHM{} \tbPH{pɛ́pɛ́láà} \tbPS{n.~} \tbGE{a type of small, carnivorous bird} \tbOP{pɛpɛlle} 
\tbLX{pɛre} \tbHM{} \tbPH{pɛ́rɪ́} \tbPS{v.~} \tbGE{to stand abreast, to be equal e.g. height} \tbPD{pɛrɛɛ, pɛrɛ, pɛrɛ, pɛrebɛ, pɛraa, - } 
\tbLX{pɛre} \tbHM{} \tbPH{pɛ̀rɪ̀} \tbPS{v.~} \tbGE{to bend in an oblique manner} \tbPD{pɛrɛɛ, pɛrɛ, pɛrɛ, pɛrebɛ, pɛrɛ, - } 
\tbLX{pɛrɛpɛrɛ} \tbHM{} \tbPH{pɛ̀rɛ̀pɛ́rɛ́} \tbPS{n.~} \tbGE{food made of mashed yam} 
\tbLX{pɛrɛpɛrɛ} \tbHM{} \tbPH{pɛ́rɛ́pɛ́rɛ́} \tbPS{adv.~} \tbGE{completely} \tbXV{A nimibaaloŋ vɛŋɛɛ la ka a dɔɔ zɔŋe pɛrɛpɛrɛ.} \tbXE{The eye disease has caused him to be blind completely.} 
\tbLX{pɛteɛ} \tbHM{} \tbPH{pɛ̀tɪ́ɛ́} \tbPS{n.~} \tbGE{medium-sized basket} \tbOP{pɛteere} \tbTP{pɛteerɛɛ} \tbVA{pɛtelee}
\tbLX{pɛyagelaa} \tbHM{} \tbPH{pɛ̀yàgláà} \tbPS{n.~} \tbGE{a basket, usually hung, in which calabashes are stored} \tbOP{pɛyagle} \tbTP{pɛyaglɛɛ} 
\tbLX{pi} \tbHM{} \tbPH{pí} \tbPS{v.~} \tbGE{to become bloated, to swell up} \tbPD{pie, piire, piire, piiribe, piiraa, - } 
\tbLX{pi} \tbHM{} \tbPH{pí} \tbPS{v.~} \tbGE{to reserve, to claim in advance} \tbPD{pie, piire, piire, piiribe, piiraa, - } 
\tbLX{pi} \tbHM{} \tbPH{pì} \tbPS{v.~} \tbGE{to be in fear; to clutch self tightly in fear} \tbPD{pie, piire, piire, piiribe, piiraa, - } 
\tbLX{pibɔŋɔ} \tbHM{} \tbPH{pìbɔ̀ŋɔ́} \tbPS{n.~} \tbGE{a type of lizard believed to be poisonous} \tbOP{pibɔnne} \tbTP{pibɔnnɛɛ} \tbVA{pibaŋa}
\tbLX{pie} \tbHM{} \tbPH{píé} \tbPS{adj.~} \tbGE{ten} \tbOP{piiri} \tbTP{piiree} 
\tbLX{pie ane anaare} \tbHM{} \tbPH{pí é ánɪ́ ánáárɪ̀} \tbPS{adj.~} \tbGE{fourteen} \tbOP{pie ne anaare} 
\tbLX{pie ane anii} \tbHM{} \tbPH{pí é ánɪ́ áníì} \tbPS{adj.~} \tbGE{eighteen} \tbOP{pie ne aniini} 
\tbLX{pie ane anuu} \tbHM{} \tbPH{pí é áne anúú} \tbPS{adj.~} \tbGE{fifteen} \tbOP{pie ne anuuri} \tbTP{pie ne anuuree} 
\tbLX{pie ane ata} \tbHM{} \tbPH{pí é ánɪ́ átà} \tbPS{adj.~} \tbGE{thirteen} \tbOP{pie ne atare} 
\tbLX{pie ane ayi} \tbHM{} \tbPH{pí é ánɪ́ áyì} \tbPS{adj.~} \tbGE{twelve} \tbOP{pie ne ayiri} 
\tbLX{pie ane ayoɔbo} \tbHM{} \tbPH{píé  ánɪ́ àyʊ̀ɔ̀bʊ́} \tbPS{adj.~} \tbGE{sixteen} \tbOP{pie ne ayoɔre} 
\tbLX{pie ane ayɔpo᷉i} \tbHM{} \tbPH{píé ánɪ́ àyɔ̀po᷉́ì} \tbPS{adj.~} \tbGE{seventeen} \tbOP{pie ne ayɔponni} 
\tbLX{pie ane yeni} \tbHM{} \tbPH{píé ánɪ́ yénì} \tbPS{adj.~} \tbGE{eleven} \tbOP{pie ne yeme} 
\tbLX{pie ne awae} \tbHM{} \tbPH{pie nɪ awaɪ} \tbPS{adj.~} \tbGE{nineteen} 
\tbLX{pieri} \tbHM{} \tbPH{pìèrì} \tbPS{v.~} \tbSN \tbGE{to cause the skin to peel e.g as a result of scrapes on rough surfaces} \tbSN \tbGE{to cause to be bare (bush)} \tbPD{pieree, piere, piere, pieribe, pieraa, pieribe } 
\tbLX{pieto} \tbHM{} \tbPH{píétó} \tbPS{n.~} \tbGE{loincloth} \tbOP{pietori} \tbTP{pietoree} 
\tbLX{pigepige} \tbHM{} \tbPH{pígépígé} \tbPS{adj.~} \tbGE{moving around craftily} 
\tbLX{pigi} \tbHM{} \tbPH{pìgì} \tbPS{v.~} \tbSN \tbGE{to obstruct (view)} \tbSN \tbGE{to shield, to prevent} \tbSN \tbGE{to trap in an enclosure e.g. house, hole} \tbPD{pigee, pigre, pigre, pigribe, pigraa, pigribe } 
\tbLX{pigi} \tbHM{} \tbPH{pígí} \tbPS{v.~} \tbGE{to make a ball e.g. from mud} \tbPD{pigee, pigre, pigre, pigribe, pigraa, - } 
\tbLX{pigiri} \tbHM{} \tbPH{pìgirì} \tbPS{v.~} \tbGE{to look out for, to watch over} \tbPD{pigiree, pigire, pigire, pigiribe, pigiraa, - } 
\tbLX{pi᷈i} \tbHM{} \tbPH{pí᷈ì᷈} \tbPS{n.~} \tbGE{arrow} \tbSG{pi᷈iu} \tbOP{piime} \tbTP{piinee} 
\tbLX{pi᷈i biri} \tbHM{} \tbPH{pí᷈ì᷈ bírì} \tbPS{n.~} \tbGE{arrowhead} \tbOP{pi᷈i bie} \tbVA{pimbiri}
\tbLX{pi᷉ikaraa} \tbHM{} \tbPH{pí᷉í᷉kàràà} \tbPS{n.~} \tbGE{sharpened arrow} \tbOP{pi᷉ikarre} \tbTP{pi᷉ikarrɛɛ} 
\tbLX{piili} \tbHM{} \tbPH{píílì} \tbPS{v.~} \tbGE{to begin; to start} \tbPD{piilee, piile, piile, piilibe, piilaa, piilibe } 
\tbLX{pi᷉i-mɔre} \tbHM{} \tbPH{pí᷉í᷉-mɔ́rɪ́} \tbPS{n.~} \tbGE{arrow shaft} \tbOP{pi᷉i-mɔɛ} \tbTP{pi᷉i-mɔrɛɛ} 
\tbLX{piiri} \tbHM{} \tbPH{píírì} \tbPS{v.~} \tbGE{to massage with warm water e.g. on bruises} \tbPD{piiree, piire, piire, piiribe, piiraa, - } 
\tbLX{piiri} \tbHM{} \tbPH{píírí} \tbPS{n.~} \tbGE{rock} \tbOP{pie} \tbTP{piiree} \tbVA{pie (Southern Dagaare dialect)}
\tbLX{piiri} \tbHM{} \tbPH{pììrì} \tbPS{v.~} \tbGE{to recover a lost object by chance} \tbPD{-, piire, piiribe, piiraa, -, - } 
\tbLX{piirimpi᷉i᷉} \tbHM{} \tbPH{pììrímpí᷉ì᷉} \tbPS{n.~} \tbGE{a type of lizard} \tbOP{piirimpiime} 
\tbLX{pi᷉i-sabɔle} \tbHM{} \tbPH{pí᷉í᷉sàbɔ́l} \tbPS{n.~} \tbGE{a type of square-shaped arrow without spikes} \tbOP{pi᷉i-sabɔlɔ} \tbTP{pi᷉i-sabɔlɛɛ} 
\tbLX{pi᷉i-sabɔle} \tbHM{} \tbPH{pí᷉í᷉-sàbɔ́lɪ́} \tbPS{n.~} \tbGE{finless arrow} \tbOP{pi᷉i-sabɔlɔ} \tbTP{pi᷉i-sabɔlɛɛ} 
\tbLX{pi᷉iseɛbaa} \tbHM{} \tbPH{pí᷉í᷉sɪ́ɛ́báá} \tbPS{n.~} \tbGE{large, long, flat needle for weaving} \tbOP{pi᷉iseɛre} \tbVA{pi᷉iseɛraa}
\tbLX{pi᷉i-takɔraa} \tbHM{} \tbPH{pi̊́᷉í-tàkɔ̀ràá} \tbPS{n.~} \tbGE{a type of arrow with spikes on the sides} \tbOP{pi᷉i-takɔrre} \tbTP{pi᷉i-takɔrɛɛ} 
\tbLX{pi᷉iteɛraa} \tbHM{} \tbPH{pí᷉ìtɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{an arrow that is shot out; arrow for shooting} \tbOP{piiteɛre} 
\tbLX{pi᷉izɛle} \tbHM{} \tbPH{pí᷉í᷉zɛ́lɪ́} \tbPS{n.~} \tbGE{poisoned arrow} \tbOP{pi᷉izɛlɛ} 
\tbLX{pikuuri} \tbHM{} \tbPH{pìkúúrì} \tbPS{n.~} \tbGE{large piece of rock} \tbOP{pikubo} \tbTP{pikuuree} 
\tbLX{pileesekɔɔ} \tbHM{} \tbPH{píléésɪ́kɔ́ɔ́} \tbPS{n.~} \tbGE{a type of bird} \tbOP{pileesekɔɔre} 
\tbLX{pilempileŋ} \tbHM{} \tbPH{pìlèmpíléŋ} \tbPS{adv.~} \tbGE{gusty} 
\tbLX{pilepile} \tbHM{} \tbPH{pílépílé} \tbPS{adv.~} \tbGE{in a fast manner} 
\tbLX{pili} \tbHM{} \tbPH{pílí} \tbPS{n.~} \tbGE{a type of mat made of tiny straw and thread used to sleep on} \tbOP{pile} \tbTP{pilee} 
\tbLX{pilintoɔ} \tbHM{} \tbPH{pìlìntʊ́ɔ́} \tbPS{n.~} \tbGE{bottle} \tbOP{pilintoore} \tbTP{pilintoorɛɛ} 
\tbLX{pilipili} \tbHM{} \tbPH{pìlìpílí} \tbPS{v.~} \tbGE{to cover up or confuse by clever actions} \tbPD{pilipilee, pilipile, pilipile, pilipilibe, pilipilaa, - } 
\tbLX{pilli} \tbHM{} \tbPH{pìllì} \tbPS{v.~} \tbGE{to roll up (mat)} \tbPD{pillee, pille, pille, pilibe, pillaa, pilibe } 
\tbLX{pilli} \tbHM{} \tbPH{pìllì} \tbPS{v.~} \tbSN \tbGE{to cover (with a cloth)} \tbSN \tbGE{cover the top of a building} \tbPD{pillee, pille, pille, pilibe, pillaa, pilibe } 
\tbLX{pilli} \tbHM{} \tbPH{pìllì} \tbPS{v.~} \tbGE{to roll the tip of a feather in the ear to scratch it} \tbPD{pillee, pille, pille, pilibe, pillaa, pilibe } 
\tbLX{pilli pɔge} \tbHM{} \tbPH{pilli pɔgɪ} \tbPS{v.~} \tbGE{to overwhelm with a swarm (bee attack)} \tbPD{pilili pɔgɛɛ, pille pɔgrɔ, pilli-pɔgrɔ, -, -, - } 
\tbLX{pimbiri} \tbHM{} \tbPH{pímbírì} \tbPS{n.~} \tbSN \tbGE{nail} \tbSN \tbGE{arrowhead} \tbSN \tbGE{syringe} \tbSG{pimbiruu} \tbOP{pimbie} \tbTP{pimbiiree} 
\tbLX{pimbitaaraa} \tbHM{} \tbPH{pìmbítááráà} \tbPS{n.~} \tbGE{screw,  bolt} \tbOP{pimbitaare} \tbTP{pimbitaarɛɛ} 
\tbLX{pimpiŋ} \tbHM{} \tbPH{pímpíŋ} \tbPS{adj.~} \tbGE{far away} 
\tbLX{pimpiŋ} \tbHM{} \tbPH{pìmpíŋ} \tbPS{v.~} \tbGE{to attempt to confuse, to cover a deed} \tbPD{pimpiŋee, pimpine, pimpine, pimpinime, pimpinaa, - } 
\tbLX{pinni} \tbHM{} \tbPH{pínnì} \tbPS{v.~} \tbSN \tbGE{to tap on to empty e.g. pocket} \tbSN \tbGE{to shake e.g. to remove unwanted bodies (bugs from a mat)} \tbSN \tbGE{to perform the final threshing to remove bits of grain} \tbPD{pinnee, pinne, pinne, pinime, pinnaa, - } 
\tbLX{pinni} \tbHM{} \tbPH{pínnì} \tbPS{v.~} \tbGE{to remind} \tbPD{pinnee, pinne, pinne, pinime, pinnaa, - } 
\tbLX{piŋgaraa} \tbHM{} \tbPH{píŋgáráà} \tbPS{n.~} \tbGE{pickaxe} \tbOP{piŋgarre} \tbTP{piŋgarrɛɛ} 
\tbLX{piŋi} \tbHM{} \tbPH{píŋ} \tbPS{v.~} \tbSN \tbGE{to shake to remove e.g. dust from cloth} \tbSN \tbGE{to beat mercilessly (figurative)} \tbPD{piŋee, pine, pine, pinime, pinaa, pinime } 
\tbLX{piŋi} \tbHM{} \tbPH{píŋ} \tbPS{v.~} \tbGE{to hide or take cover behind an object} \tbPD{piŋee, pine, pine, pinime, pinaa, pinime } 
\tbLX{pipiiri} \tbHM{} \tbPH{pìpììrí} \tbPS{n.~} \tbGE{a type of tiny black flying insect that moves in groups} 
\tbLX{pipiree} \tbHM{} \tbPH{pìpíréé} \tbPS{n.~} \tbSN \tbGE{chaff} \tbSN \tbGE{residue of the matter left after chewing and extracting the liquid from some nuts like coconut} \tbOP{pipirri} \tbTP{pipirree} 
\tbLX{piraa} \tbHM{} \tbPH{píráá} \tbPS{n.~} \tbGE{button} \tbOP{pirri} 
\tbLX{pire} \tbHM{} \tbPH{pìré} \tbPS{n.~} \tbGE{new land cleared for farming} \tbOP{pirri} \tbTP{piree/pirree} 
\tbLX{pirempireŋ} \tbHM{} \tbPH{pírémpíréŋ} \tbPS{adv.~} \tbGE{full (tyre, stomach)} 
\tbLX{pirepire} \tbHM{} \tbPH{pírépíré} \tbPS{n.~} \tbGE{the state of complete emptiness; the state of having nothing} \tbXV{N da boɔrɔ ka N gaa sori kyɛ pirepire ba mare ma.} \tbXE{I had wanted to go on a trip, but I had nothing on me.} 
\tbLX{piri} \tbHM{} \tbPH{pírí} \tbPS{v.~} \tbSN \tbGE{to compete to win e.g. vacant seat} \tbSN \tbGE{to speak at the same time (two or more people)} \tbPD{piree, pire, pire, piribe, piraa, - } 
\tbLX{piri} \tbHM{} \tbPH{pírì} \tbPS{v.~} \tbGE{to lance, to make a small cut to release pus} \tbPD{piree, pire, pire, piribe, piraa, - } 
\tbLX{piri} \tbHM{} \tbPH{pìrì} \tbPS{v.~} \tbSN \tbGE{to button up e.g. shirt} \tbSN \tbGE{to strap on e.g. wristwatch} \tbSN \tbGE{to put a noose on or off e.g. neck or leg of an animal} \tbPD{piree, pire, pire, piribe, piraa, - } 
\tbLX{piriyoro} \tbHM{} \tbPH{píríyòró} \tbPS{adv.~} \tbGE{completely scattered in different directions e.g. members of a search party} 
\tbLX{po} \tbHM{} \tbPH{pʊ̀} \tbPS{v.~} \tbGE{to pull a grass stalk off the base} \tbPD{poɛ, poorɔ/poora, poorɔ/poora, pooreba, pooraa, - } 
\tbLX{po} \tbHM{} \tbPH{pò} \tbPS{v.~} \tbGE{to dig for termites (to feed young chicks)} \tbPD{poe, puoro, puoro, puoribo, puoraa, - } 
\tbLX{po} \tbHM{} \tbPH{pò} \tbPS{v.~} \tbGE{to ooze, to flow (water)} \tbPD{poe, puoro, puoro, puoribo, puoraa, - } 
\tbLX{pobeɛmɛ} \tbHM{} \tbPH{pʊ̀bɪ́ɛ́mɛ́} \tbPS{n.~} \tbGE{stomachache, upset stomach} \tbTP{pobeɛmɛɛ} 
\tbLX{pobieri} \tbHM{} \tbPH{pʊ̀bìèrí} \tbPS{n.~} \tbGE{wickedness} \tbTP{pobieree} \tbVA{powolaa}
\tbLX{poboɔbo} \tbHM{} \tbPH{pʊ̀bʊ́ɔ́bʊ̀} \tbPS{n.~} \tbGE{will, desire} 
\tbLX{pobu} \tbHM{} \tbPH{pòbú} \tbPS{adv.~} \tbGE{filled with (tears in the eyes)} 
\tbLX{pode᷉o} \tbHM{} \tbPH{pʊ̀dɪ̀᷉ʊ́᷉} \tbPS{n.~} \tbGE{preference} 
\tbLX{podɛgere} \tbHM{} \tbPH{pʊ̀dɛ́gɪ́rɪ̀} \tbPS{n.~} \tbGE{evil thought} 
\tbLX{pofaa} \tbHM{} \tbPH{pʊ̀fáá} \tbPS{n.~} \tbGE{state of being evil-minded} \tbOP{pofaare} \tbTP{pofaarɛɛ} 
\tbLX{pogaale} \tbHM{} \tbPH{pʊ̀gáálɪ̀} \tbPS{n.~} \tbGE{prenatal care} \tbOP{pogaaɭɛɛ} 
\tbLX{pogitogo} \tbHM{} \tbPH{pógítógó} \tbPS{adv.~} \tbGE{very short and small} \tbVA{pogitogoo}
\tbLX{pokuoŋaa} \tbHM{} \tbPH{pʊ̀kůóŋáá} \tbPS{n.~} \tbGE{sharp stomachache without diarrhoea} \tbOP{pokuoni} \tbTP{pokuonee} 
\tbLX{pokpaane} \tbHM{} \tbPH{pʊ̀kpàànɪ̀} \tbPS{n.~} \tbGE{fasting} \tbTP{pokpaanɛɛ} \tbVA{pokpegri}
\tbLX{pokpegiri} \tbHM{} \tbPH{pʊ̀kpégrí} \tbPS{n.~} \tbGE{closed-heartedness, without an open mind, wickedness} \tbOP{pokpege} \tbTP{pokpegree} \tbVA{pokpaane}
\tbLX{pokpɛlɛnkyɛ} \tbHM{} \tbPH{pʊ̀kpɛ́lɛ́nʧɛ́} \tbPS{n.~} \tbGE{the act of being potentially dangerous} 
\tbLX{pokpɔgere} \tbHM{} \tbPH{pʊ̀kpɔ́grí} \tbPS{n.~} \tbGE{tumor in the stomach} \tbOP{pokpɔgɔ} \tbTP{pokpɔgrɛɛ} 
\tbLX{pokyiiri} \tbHM{} \tbPH{pʊ́ʧíírí} \tbPS{n.~} \tbGE{confession} 
\tbLX{polee} \tbHM{} \tbPH{póléé} \tbPS{n.~} \tbGE{small farm} 
\tbLX{polee} \tbHM{} \tbPH{pʊ̀léé} \tbPS{n.~} \tbGE{early pregnancy} 
\tbLX{poleɛne} \tbHM{} \tbPH{pʊ́lɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{female lizard} \tbOP{poleɛmɛ} \tbTP{poleɛmɛɛ} 
\tbLX{polepole} \tbHM{} \tbPH{pʊ́lpʊ́lɪ́} \tbPS{adj.~} \tbGE{very tall and pointed summit. (church tower)} 
\tbLX{poli} \tbHM{} \tbPH{pólí} \tbPS{v.~} \tbGE{smear plenty of oil producing a shining effect} \tbPD{polee, polo, polo, polibo, polaa, polibo } 
\tbLX{poli} \tbHM{} \tbPH{pòlì} \tbPS{adv.~} \tbGE{flaring up abruptly (fire)} 
\tbLX{polige} \tbHM{} \tbPH{pʊ̀lígè} \tbPS{n.~} \tbGE{the state of not being open-minded} 
\tbLX{poll} \tbHM{} \tbPH{pʊ́ll} \tbPS{adj.~} \tbGE{small and sharp} \tbOP{poll poll} 
\tbLX{polle} \tbHM{} \tbPH{pʊ̌llɪ̀} \tbPS{v.~} \tbGE{to take an oath; to make a promise} \tbGE{an oath} \tbPD{pollɛɛ, polla, polla, poleba, pollaa, poleba } \tbSE{polle noɔre, polle teebo} \tbGE{to vow} 
\tbLX{polle} \tbHM{} \tbPH{pʊ̌llɪ̀} \tbPS{v.~} \tbGE{to sharpen (a point)} \tbGE{an oath} \tbPD{pollɛɛ, polla, polla, poleba, pollaa, poleba } \tbSE{polle noɔre, polle teebo} \tbGE{to vow} 
\tbLX{pololo} \tbHM{} \tbPH{pʊ́lʊ́lʊ́} \tbPS{adj.~} \tbGE{pointed and sharp} 
\tbLX{pomm} \tbHM{} \tbPH{pòmm} \tbPS{adv.~} \tbGE{bloated like a balloon} 
\tbLX{pompoluoree} \tbHM{} \tbPH{pómpólúóréé} \tbPS{n.~} \tbGE{a type of wild tree with edible fruit} \tbOP{pompoluori} \tbTP{ pompoluoree} \tbVA{polimpoluoree}
\tbLX{pompoŋɔ} \tbHM{} \tbPH{pʊ́mpʊ́ŋɔ̀} \tbPS{n.~} \tbGE{a type of poisonous green-trunked tree with white nectar} \tbOP{pomponne} \tbTP{pomponnɛɛ} \tbVA{pompoŋa}
\tbLX{ponaŋa} \tbHM{} \tbPH{pʊ́náŋ} \tbPS{n.~} \tbGE{severe stomachache} \tbOP{ponanne} \tbTP{ponannɛɛ} \tbVA{ponaŋe}
\tbLX{ponne} \tbHM{} \tbPH{pʊ́nnɪ́} \tbPS{v.~} \tbSN \tbGE{to drop in bits e.g. rain} \tbSN \tbGE{to test tightness e.g. string of bow, guitar} \tbPD{ponnɛɛ, ponnɔ, ponnɔ, ponema, ponnaa, - } 
\tbLX{ponne} \tbHM{} \tbPH{pʊ̀nnɪ̀} \tbPS{v.~} \tbGE{to covet; to make advances to mate with e.g. animals} \tbPD{-, -, -, -, -, - } 
\tbLX{ponne} \tbHM{} \tbPH{pʊ̀nnɪ̀} \tbPS{adj.~} \tbGE{not up to standard} 
\tbLX{poŋi} \tbHM{} \tbPH{pôŋ} \tbPS{n.~} \tbGE{pollen} \tbOP{ponni} \tbTP{ponnee} 
\tbLX{poŋi} \tbHM{} \tbPH{pòŋ} \tbPS{v.~} \tbSN \tbGE{to shave (hair)} \tbSN \tbGE{to remove corn from the cob} \tbPD{poŋee, pono, pono, ponimo, ponaa, ponimo } 
\tbLX{poŋi} \tbHM{} \tbPH{póŋ} \tbPS{v.~} \tbGE{to share, to divide} \tbPD{poŋee, pono, pono, ponimo, ponaa, - } 
\tbLX{poo} \tbHM{} \tbPH{pʊ̀ʊ̀} \tbPS{adv.~} \tbGE{onomatopoeia; like noise made when one wades into water} 
\tbLX{po-o} \tbHM{} \tbPH{pʊ́-ʊ́} \tbPS{n.~} \tbGE{ocean} \tbTP{poorɛɛ} \tbVA{mpoɔne, mpoɔnɛɛ (Akan)}
\tbLX{poomɔ} \tbHM{} \tbPH{pʊ́ʊ́mɔ̀} \tbPS{n.~} \tbGE{bean weevil} \tbTP{poomɛɛ} \tbVA{pooma}
\tbLX{poone} \tbHM{} \tbPH{pʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbGE{to persist e.g. talk, argue} \tbXV{O maŋ boɔle la ba woŋi yeli kyɛ meŋ aŋ poonɔ pare.} \tbXE{When he calls, there comes no answer, but he persists in calling earnestly.} \tbPD{poonɛɛ, poonɔ, poonɔ, poonema, poonaa, - } 
\tbLX{poore} \tbHM{} \tbPH{pʊ̌ʊ̌rɪ́} \tbPS{n.~} \tbGE{challenge} \tbTP{poorɛɛ} 
\tbLX{poore} \tbHM{} \tbPH{pʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to challenge, to compete} \tbPD{poorɛɛ, poora, poora, pooreba, pooraa, - } 
\tbLX{poore} \tbHM{} \tbPH{pʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to spray e.g. water onto someone} \tbPD{poorɛɛ, poora, poora, pooreba, pooraa, - } 
\tbLX{poore} \tbHM{} \tbPH{pʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to harvest grass stalks} \tbPD{poorɛɛ, poora, poora, pooreba, pooraa, - } 
\tbLX{poɔ} \tbHM{} \tbPH{pʊ́ɔ́} \tbPS{n.~} \tbGE{stomach, abdomen} \tbOP{poore} \tbTP{poorɛɛ} 
\tbLX{poɔ} \tbHM{} \tbPH{pʊ́ɔ́} \tbPS{n.~} \tbGE{part (of a group)} \tbXV{A pɔgelee ba poɔ a zɔɔreŋ.} \tbXE{The girl is not part of the fight.} \tbOP{poore} \tbTP{poorɛɛ} 
\tbLX{poɔ} \tbHM{} \tbPH{pʊ́ɔ́} \tbPS{n.~} \tbGE{pregnancy} \tbOP{poore} \tbTP{poorɛɛ} 
\tbLX{poɔ} \tbHM{} \tbPH{pʊ́ɔ́} \tbPS{v.~} \tbGE{to beat down on by rain} \tbPD{poɔɛɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\tbLX{poɔ} \tbHM{} \tbPH{pʊ́ɔ̀} \tbPS{v.~} \tbGE{to add to; to include} \tbPD{poɔɛɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\tbLX{poɔ} \tbHM{} \tbPH{pʊ́ɔ́} \tbPS{prep.~} \tbSN \tbGE{in, inside} \tbXV{Ka diibuŋ be fo noɔre poɔ fo koɳ to᷉ɔ᷉ yeli yɛlɛ velaa.} \tbXE{If there is food in your mouth, you cannot talk well.} \tbSN \tbGE{among, within} \tbXV{O poɔ la a bibiiri na naŋ daŋe waaboŋ.} \tbXE{S/he is among the children who came early.} 
\tbLX{po᷉ɔ} \tbHM{} \tbPH{pʊ̀᷉ɔ̀᷉} \tbPS{v.~} \tbSN \tbGE{to interrupt} \tbXV{A bie ba tu o noɔre baare kyɛ ka a karema po᷉ɔ kpɛ, yeli ka zirii la.} \tbXE{The child had not finished telling his story when the teacher interrupted and said it was a lie.} \tbXV{A pɔgebilii deɛnɛ la velaa zaa ka dɔɔbilii mine wa põɔ kpɛ ka a deɛne ŋmaa.} \tbXE{The girls were happily playing when some boys interrupted them and the play ended.} \tbSN \tbGE{to perform a task earlier than expected} \tbXV{A tontonema da põɔ iree la gaa o toma zie a na te de ba sanyɔɔ.} \tbXE{The workers left before dawn to the work place in order to receive their salaries.} \tbPD{po᷉ɔɛ, poɔnɔ, poɔnɔ, poɔnema, poɔnaa, - } 
\tbLX{po᷉ɔ} \tbHM{} \tbPH{pʊ́᷉ɔ́᷉} \tbPS{v.~} \tbGE{to rot} \tbPD{po᷉ɔ᷉ɛ᷉, poɔnɔ, poɔnɔ, poɔnema, poɔnaa, - } 
\tbLX{po᷉ɔ} \tbHM{} \tbPH{pʊ́᷉ɔ́᷉} \tbPS{v.~} \tbGE{to play a game of hide and seek} \tbPD{po᷉ɔ᷉ɛ᷉, poɔnɔ, poɔnɔ, poɔnema, poɔnaa, - } 
\tbLX{poɔle} \tbHM{} \tbPH{pʊ́ɔ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to cause wetness of e.g. floor} \tbSN \tbGE{to soak herbs (preparing medicine)} \tbPD{poɔlɛɛ, poɔlɔ, poɔlɔ, poɔlbɔ, poɔlaa, - } 
\tbLX{poɔle} \tbHM{} \tbPH{pʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to make the end of something pointed} \tbPD{poɔlɛɛ, poɔlɔ, poɔlɔ, poɔlbɔ, poɔlaa, - } 
\tbLX{poɔle} \tbHM{} \tbPH{pʊ́ɔ́lɪ́} \tbPS{adj.~} \tbGE{pointed end} \tbXV{A bugo pare waa la poɔle lɛ.} \tbXE{The bottom of the granary is pointed.} 
\tbLX{poɔnaa} \tbHM{} \tbPH{pʊ́ɔ́náá} \tbPS{n.~} \tbGE{hide and seek (children's game)} \tbOP{poɔne} \tbTP{poɔnɛɛ} 
\tbLX{poɔre} \tbHM{} \tbPH{pʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to rub to clear e.g. hand over face just from sleep} \tbSN \tbGE{to clutch with both hands and feet e.g in climbing a tree or wall} \tbPD{poɔrɛɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\tbLX{poɔroŋ} \tbHM{} \tbPH{pʊ́ɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{rot, rotten thing} \tbOP{poɔrɔ} 
\tbLX{popaale} \tbHM{} \tbPH{pʊ́páálɪ́} \tbPS{n.~} \tbGE{disease that causes swollen feet and a distended stomach} 
\tbLX{popeɛloŋ} \tbHM{} \tbPH{pʊ́pɪ́ɛ́lʊ́ŋ} \tbPS{n.~} \tbGE{happiness} 
\tbLX{popelaa} \tbHM{} \tbPH{pʊ́pɪ́láá} \tbPS{n.~} \tbGE{stomachache} 
\tbLX{popelaa} \tbHM{} \tbPH{pʊ́pɪ́láá} \tbPS{adv.~} \tbSN \tbGE{prostrate} \tbSN \tbGE{open minded} 
\tbLX{popɛge} \tbHM{} \tbPH{pòpɛ́gɪ́} \tbPS{n.~} \tbGE{outer shell} \tbOP{popɛgere/popɛgɛ} \tbTP{popɛgerɛɛ} 
\tbLX{popɛle} \tbHM{} \tbPH{pʊ̀pɛ̀lɪ̀} \tbPS{n.~} \tbSN \tbGE{happiness, joy (taa popɛle)} \tbSN \tbGE{choice, wish} 
\tbLX{popo} \tbHM{} \tbPH{pòpò} \tbPS{n.~} \tbGE{motorcycle} \tbOP{popori} \tbTP{poporee} 
\tbLX{pore} \tbHM{} \tbPH{pʊ́rɪ́} \tbPS{adj.~} \tbGE{tiny, small} \tbOP{porre} \tbTP{porrɛɛ} 
\tbLX{pore} \tbHM{} \tbPH{pʊ̀rɪ̀} \tbPS{v.~} \tbGE{to name, to mention} \tbPD{porɛɛ, porɔ, porɔ, poreba, poraa, - } 
\tbLX{pore} \tbHM{} \tbPH{pʊ́rɪ̀} \tbPS{v.~} \tbGE{to pour out} \tbPD{porɛɛ, porɔ, porɔ, poreba, poraa, - } 
\tbLX{pore} \tbHM{} \tbPH{pʊ́rɪ̀} \tbPS{v.~} \tbGE{to buy herbal medicine} \tbPD{porɛɛ, porɔ, porɔ, poreba, poraa, - } 
\tbLX{pore} \tbHM{} \tbPH{pʋ́rɪ́} \tbPS{adj.~} \tbGE{small in size} 
\tbLX{pore} \tbHM{} \tbPH{pʊ̀rɪ̀} \tbPS{adj.~} \tbGE{abundant in quantity} 
\tbLX{poree} \tbHM{} \tbPH{póréé} \tbPS{n.~} \tbGE{pair of pegs on which pots and calabashes are placed} \tbOP{porri} \tbTP{porree} 
\tbLX{pori} \tbHM{} \tbPH{pòrì} \tbPS{v.~} \tbGE{to cause to blister, to scald (of fire)} \tbPD{poree, poro, poro, poribo, poraa, - } 
\tbLX{pori} \tbHM{} \tbPH{pòrì} \tbPS{adj.~} \tbGE{sound of a big falling object from a height (a big fruit)} 
\tbLX{pori} \tbHM{} \tbPH{pórí} \tbPS{adj.~} \tbGE{sound of a small object falling from a height (small fruit)} 
\tbLX{pori} \tbHM{} \tbPH{pórí} \tbPS{v.~} \tbGE{to struggle e.g. dying fowl} \tbPD{poree, poro, poro, poribo, poraa, - } 
\tbLX{poriko} \tbHM{} \tbPH{póríkó} \tbPS{n.~} \tbGE{pig} \tbOP{porikori} \tbTP{porikoree} \tbVA{pariko}
\tbLX{porisi} \tbHM{} \tbPH{pórísì} \tbPS{n.~} \tbGE{police} \tbOP{porisiri} \tbTP{porisiree} 
\tbLX{poritiisan} \tbHM{} \tbPH{pòrítɪ́ɪ́sa} \tbPS{n.~} \tbGE{police station} \tbOP{poritiisanne} \tbTP{poritiisanɛɛ/poritiisannɛɛ} 
\tbLX{poroso} \tbHM{} \tbPH{pòròsò} \tbPS{adv.~} \tbGE{large and round e.g. balloon} 
\tbLX{poroworo} \tbHM{} \tbPH{pʊ̀rɪ̀wʊ̀rʊ̀} \tbPS{adj.~} \tbGE{countless and all over the place e.g. piglets of a sow} 
\tbLX{porɔ} \tbHM{} \tbPH{pʊ́rɔ́} \tbPS{n.~} \tbGE{fail to respond smoothly or click (archer's arrow)} \tbXV{A tammo ŋmɛ la porɔ ka a tanteɛrɛ teɛ a soɔŋaa nyɛre.} \tbXE{The bow failed to release the arrow and so the archer failed to hit the hare.} \tbOP{porre} \tbTP{porɛɛ} 
\tbLX{porre} \tbHM{} \tbPH{pʊ́rrɪ̀} \tbPS{v.~} \tbGE{to force air through closed lips (rude way of disagreeing; lie)} \tbPD{porrɛɛ, porra, porra, poreba, porraa, - } 
\tbLX{porre} \tbHM{} \tbPH{pʊ́rrɪ̀} \tbPS{v.~} \tbGE{to break into tiny bits} \tbPD{porrɛɛ, porra, porra, poreba, porraa, - } 
\tbLX{porri} \tbHM{} \tbPH{pórrì} \tbPS{v.~} \tbGE{to report} \tbPD{porree, porro, porro, poribo, porraa, - } 
\tbLX{posa᷉a} \tbHM{} \tbPH{pʊ̀sàà} \tbPS{n.~} \tbGE{sadness} 
\tbLX{posaanaa} \tbHM{} \tbPH{posaanaa} \tbPS{n.~} \tbGE{sorrow, unhappiness} \tbOP{posaane/posaama} \tbTP{posaanɛɛ} 
\tbLX{posogedaa} \tbHM{} \tbPH{pòsʊ̀gdàá} \tbPS{n.~} \tbGE{backbone} \tbOP{posogedaare} \tbTP{posogedaarɛɛ} 
\tbLX{posoge-kpoŋi} \tbHM{} \tbPH{pósʊ́g-kpôŋ} \tbPS{n.~} \tbGE{spinal tuberculosis} 
\tbLX{posogɔ} \tbHM{} \tbPH{pósʊ́gɔ́} \tbPS{n.~} \tbGE{mid-back} \tbOP{posogere} \tbTP{posogerɛɛ} \tbVA{posoga}
\tbLX{poteɛrɛ} \tbHM{} \tbPH{pʊ́tɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{thoughts, wishes} 
\tbLX{poteɛroŋ} \tbHM{} \tbPH{pʊ̀tɪ́ɛ́rʊ́ŋ} \tbPS{n.~} \tbGE{thought} \tbOP{poteɛrɛ} 
\tbLX{po-tu} \tbHM{} \tbPH{pò-tù} \tbPS{n.~} \tbGE{the act of following} 
\tbLX{potulimo} \tbHM{} \tbPH{pótúlímó} \tbPS{adj.~} \tbGE{backwards} \tbVA{potulo}
\tbLX{potulo} \tbHM{} \tbPH{pótúló} \tbPS{adj.~} \tbGE{backwards} \tbVA{potulimo}
\tbLX{potuo} \tbHM{} \tbPH{pòtúó} \tbPS{n.~} \tbGE{wickedness, malice} \tbOP{potuuri} \tbTP{potuuree} 
\tbLX{potuoluŋ} \tbHM{} \tbPH{po̊túólúŋ} \tbPS{n.~} \tbGE{wickedness} 
\tbLX{potuosobɔ} \tbHM{} \tbPH{pòtúósʊ́bá} \tbPS{n.~} \tbGE{wicked person} \tbOP{ potuodeme} \tbVA{potuosoba}
\tbLX{potuuro} \tbHM{} \tbPH{pòtúúró} \tbPS{n.~} \tbGE{follower, disciple} \tbOP{ potuuribo} 
\tbLX{pou} \tbHM{} \tbPH{póú} \tbPS{adv.~} \tbGE{onomatopoeia; sound of blast (balloon)} 
\tbLX{povelaa} \tbHM{} \tbPH{pʊ̀vɪ̀làá} \tbPS{adj.~} \tbGE{sincere, pure} \tbOP{poveɛle} 
\tbLX{powɛle} \tbHM{} \tbPH{pʊ́wɛ́lɪ́} \tbPS{n.~} \tbGE{confession} 
\tbLX{powolaa} \tbHM{} \tbPH{pʊ́wʊláá́} \tbPS{n.~} \tbGE{wickedness} \tbOP{powolle} \tbVA{pobieri}
\tbLX{po-yiraa} \tbHM{} \tbPH{pʊ̀-yíráá} \tbPS{n.~} \tbGE{miscarriage} \tbOP{po-yirri} \tbVA{po-yi}
\tbLX{pozagela} \tbHM{} \tbPH{pʊ́zágɪ́là} \tbPS{n.~} \tbGE{innocence} 
\tbLX{pozagelaa} \tbHM{} \tbPH{pʊ́zágɪ́láá} \tbPS{n.~} \tbGE{empty stomach} \tbOP{pozagele} 
\tbLX{pozagelaa} \tbHM{} \tbPH{pʊ́zágɪ́láá} \tbPS{n.~} \tbGE{the state of being free of malice toward another} \tbOP{pozagele} 
\tbLX{pozu} \tbHM{} \tbPH{pòzú} \tbPS{n.~} \tbGE{gossip} \tbOP{pozuri} \tbTP{pozuree} 
\tbLX{pɔ} \tbHM{} \tbPH{pɔ́} \tbPS{v.~} \tbGE{to swear; to make a vow} \tbPD{pɔɛ, poɔrɔ, poɔrɔ, poɔreba, poɔraa, - } 
\tbLX{pɔ-eŋe} \tbHM{} \tbPH{pɔ́-ɪ̀ŋ} \tbPS{v.~} \tbGE{to curse; putting a malicious spell on someone} \tbPD{pɔ eŋɛɛ, poɔrɔ ennɛ, pɔ-ennɛ, pɔ-enemɛ, pɔ-ennaa, - } 
\tbLX{pɔge} \tbHM{} \tbPH{pɔ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to close, to shut} \tbSN \tbGE{to cover} \tbSN \tbGE{to imprison} \tbPD{pɔgɛɛ, pɔgrɔ, pɔgrɔ, pɔgreba, pɔgraa, pɔgreba } 
\tbLX{pɔge} \tbHM{} \tbPH{pɔ́gɪ́} \tbPS{n.~} \tbSN \tbGE{female} \tbSN \tbGE{woman} \tbSN \tbGE{wife} \tbOP{pɔgebɔ} \tbTP{pɔgebɛɛ} 
\tbLX{pɔge} \tbHM{} \tbPH{pɔ́gɪ̀} \tbPS{v.~} \tbSN \tbGE{to meet the presence of} \tbXV{Daŋ iribu wa pɔge ma kyɛ ka N gaa.} \tbXE{Get up early to come and meet me before I go.} \tbSN \tbGE{to hit a target} \tbXV{A naŋkpaana ŋmɛ la walaa kyɛ o ba pɔge o.} \tbXE{The hunter shot an antelope but did not hit the target.} \tbSN \tbGE{to be possible} \tbXV{A baa kye a dieŋ; o naa pɔge la yeŋe poɔ.} \tbXE{The dog is not in the room; it may be outside.} \tbPD{pɔgɛɛ, pɔgrɔ, pɔgrɔ, pɔgreba, pɔgraa, pɔgreba } 
\tbLX{pɔgebaa} \tbHM{} \tbPH{pɔ̀gɪ̀báá} \tbPS{n.~} \tbSN \tbGE{a woman who has a sharp tongue} \tbSN \tbGE{prostitute} \tbOP{pɔgebaare} \tbTP{pɔgebaarɛɛ} 
\tbLX{pɔgebeɛle} \tbHM{} \tbPH{pɔ̀gɪ̀bɪ̀ɛ̀lɪ̀} \tbPS{n.~} \tbGE{the act of taking a bride to the bridegroom} 
\tbLX{pɔgebeɛlɛ} \tbHM{} \tbPH{pɔ̀gɪ̀bɪ́ɛ́lɛ} \tbPS{n.~} \tbGE{bridesmaid} \tbOP{pɔgebeɛlbɛ} \tbTP{pɔgebeɛloo} 
\tbLX{pɔgebeɛloŋ} \tbHM{} \tbPH{pɔ̀gɪ̀bɪ̀ɛ̀lʊ̌ŋ} \tbPS{n.~} \tbGE{marriage feast (taking the bride to the bridegroom)} 
\tbLX{pɔgebeɛre} \tbHM{} \tbPH{pɔ̀gɪ̀bɪ́ɛ́rɪ́} \tbPS{n.~} \tbGE{a stage in courtship} 
\tbLX{pɔge-bɔ} \tbHM{} \tbPH{pɔ̀gɪ̀bɔ́} \tbPS{n.~} \tbGE{courtship} 
\tbLX{Pɔgeda᷈a᷈} \tbHM{} \tbPH{pɔ̀gɪ̀dá᷉à} \tbPS{n.~} \tbGE{name traditionally given to a female child born on day of locally brewed beer} \tbOP{Pɔgedaamine} \tbTP{Pɔgedaanɛɛ} 
\tbLX{pɔge-de} \tbHM{} \tbPH{pɔ̀gɪ́-dɪ́} \tbPS{n.~} \tbGE{marriage} 
\tbLX{pɔgederaa} \tbHM{} \tbPH{pɔ̀gɪ̀dɪ́ráá} \tbPS{n.~} \tbGE{a woman that is qualified to be married} \tbOP{pɔgederre} \tbTP{pɔgederɛɛ} 
\tbLX{pɔgederɛ} \tbHM{} \tbPH{pɔ̀gɪ̀dɪ́rɛ́} \tbPS{n.~} \tbSN \tbGE{bridegroom} \tbSN \tbGE{a man who is eligible to marry} 
\tbLX{pɔgederoŋ} \tbHM{} \tbPH{pɔ̀gɪ̀dɪ́rʋ́ŋ} \tbPS{n.~} \tbGE{marriage (action of groom)} 
\tbLX{pɔgedɔgerɔ} \tbHM{} \tbPH{pɔ̀gɪ̀dɔ̀gɪ̀rɔ́} \tbPS{n.~} \tbGE{a woman who has just given birth to a child} \tbOP{pɔgedɔgereba} \tbTP{pɔgedɔgerebɛɛ} 
\tbLX{pɔgefaa-ŋmenaa} \tbHM{} \tbPH{pɔ́gɪ́fáá-ŋɪ́náá} \tbPS{n.~} \tbGE{early evening (4 pm)} 
\tbLX{pɔgegori} \tbHM{} \tbPH{pɔ̀gígórí} \tbPS{n.~} \tbGE{bride price} \tbOP{pɔgegoe} \tbTP{pɔgegoree} 
\tbLX{pɔgeguri} \tbHM{} \tbPH{pɔ̀gɪ̀gùrì} \tbPS{n.~} \tbGE{rape} 
\tbLX{pɔgegbeɛloo} \tbHM{} \tbPH{pɔ̀gɪ̀bɪ́ɛ́lʊ́ʊ́} \tbPS{n.~} \tbGE{one who leads the bride to the bridegroom's house} \tbOP{pɔgegbeɛlebɛ} \tbTP{pɔgegbeɛlebɛɛ} 
\tbLX{pɔge-iraa} \tbHM{} \tbPH{pɔ̀gɪ́-íráá} \tbPS{n.~} \tbGE{a woman who has eloped with her fiance} \tbOP{pɔge-irri} 
\tbLX{pɔgekoɔre} \tbHM{} \tbPH{pɔ̀gɪ̀kʊ́ɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{unmarried woman} \tbSN \tbGE{widower} \tbOP{pɔgekɔbɔ} \tbTP{pɔgekɔbɛɛ} 
\tbLX{pɔgekɔŋ} \tbHM{} \tbPH{pɔ̀gɪ̀kɔ̂ŋ} \tbPS{n.~} \tbGE{unavailability of mariagable women} 
\tbLX{pɔgekuree} \tbHM{} \tbPH{pɔ̀gɪ̀kúréé} \tbPS{n.~} \tbGE{skirt} \tbOP{ pɔgekurri} \tbTP{pɔgekurree} 
\tbLX{pɔge-kyaaraa} \tbHM{} \tbPH{pɔ̀gɪ̀-ʧàáráá} \tbPS{n.~} \tbGE{woman offered for marriage} \tbOP{pɔge-kyaare} \tbTP{pɔge-kyaarɛɛ} 
\tbLX{pɔgekyeɛ} \tbHM{} \tbPH{pɔ̀gɪ̀ʧɪ́ɛ́} \tbPS{n.~} \tbGE{sister-in-law} \tbOP{pɔgekyeere\\pɔgekyemine} \tbTP{pɔgekyeerɛɛ} 
\tbLX{pɔgekyɛre} \tbHM{} \tbPH{pɔ̀gɪ̀ʧɛ̀rɪ̀} \tbPS{n.~} \tbGE{payment of bride price} 
\tbLX{pɔge-la-dɔɔ} \tbHM{} \tbPH{pɔ̀gɪ̀-lá-dɔ́ɔ́} \tbPS{n.~} \tbGE{woman with manly features} \tbOP{pɔge-la-dɔbɔ} 
\tbLX{pɔgele} \tbHM{} \tbPH{pɔ̀gɪ̀lɪ̀} \tbPS{v.~} \tbGE{to tame} \tbPD{pɔgelɛɛ, pɔgelɔ,, pɔgelɔ,, pɔgeleba, pɔgelaa, - } 
\tbLX{pɔgelee} \tbHM{} \tbPH{pɔ̀gɪ̀léé} \tbPS{n.~} \tbSN \tbGE{girl} \tbSN \tbGE{junior wife (second of two wives)} \tbOP{pɔgebilii} \tbTP{pɔgebillee} 
\tbLX{pɔgeloŋ} \tbHM{} \tbPH{pɔ̀gɪ̀lʊ́ŋ} \tbPS{n.~} \tbGE{womanliness} 
\tbLX{pɔge-nɔgeraa} \tbHM{} \tbPH{pɔ̀gɪ̀-nɔ̀gɪ̀ràá} \tbPS{n.~} \tbGE{woman (a derogatory term implying that the woman is a sojourner and therefore has no power)} \tbOP{pɔge-nɔgere} \tbTP{pɔge-nɔgerɛɛ} 
\tbLX{pɔgenyaŋaa} \tbHM{} \tbPH{pɔ̀gɪ̀ɲáŋáá} \tbPS{n.~} \tbGE{old woman} \tbOP{pɔgnyaŋene/pɔgenyanne} \tbTP{pɔgenyannɛɛ} 
\tbLX{pɔgenyɔgeraa} \tbHM{} \tbPH{pɔgeɲɔgeraa} \tbPS{n.~} \tbGE{a woman who has eloped with a man} \tbOP{pɔgenyɔgerre} 
\tbLX{pɔgenyuo} \tbHM{} \tbPH{pɔ̀gɪ̀ɲúó} \tbPS{n.~} \tbGE{woman with a big navel (name)} \tbOP{pɔgenyuuri/pɔgenyumine} 
\tbLX{pɔgepe᷉ɛ᷉} \tbHM{} \tbPH{pɔ̀gɪ̀pɪ̀᷉ɛ̀᷉} \tbPS{n.~} \tbGE{courtship} 
\tbLX{pɔge-pɛgebo} \tbHM{} \tbPH{pɔ̀gɪ̀-pɛ́gɪ́bʊ́} \tbPS{n.~} \tbGE{menstruation} 
\tbLX{pɔgepiiraa} \tbHM{} \tbPH{pɔ̀gɪ̀pííráá} \tbPS{n.~} \tbSN \tbGE{fiancee, betrothed woman} \tbSN \tbGE{engaged woman for marriage} \tbOP{pɔgepiiri} \tbTP{pɔgepiiree} 
\tbLX{pɔgepoɔ} \tbHM{} \tbPH{pɔ̀gìpʊ́ɔ́} \tbPS{n.~} \tbGE{pregnant woman} \tbOP{pɔgepoore} \tbTP{pɔgepoorɛɛ} 
\tbLX{Pɔgesaa} \tbHM{} \tbPH{pɔ̀gɪ̀sáá} \tbPS{n.~} \tbGE{name traditionally given to a female child born on a rainy day} \tbOP{pɔgesamine} 
\tbLX{pɔgesaraa} \tbHM{} \tbPH{pɔ̀gɪ̀sáráà} \tbPS{n.~} \tbGE{young woman} \tbOP{pɔgesarre} \tbTP{pɔgesarrɛɛ} 
\tbLX{pɔgesarebie} \tbHM{} \tbPH{pɔ̀gɪ̀sárɪ́bíé} \tbPS{n.~} \tbGE{a woman's firstborn} \tbOP{pɔgesarebiiri} \tbTP{pɔgesarebiiree} 
\tbLX{pɔgesarelee} \tbHM{} \tbPH{pɔ̀gɪ̀sárɪ́léé} \tbPS{n.~} \tbGE{young woman} \tbOP{pɔgesarebilii} 
\tbLX{pɔgeseɛ} \tbHM{} \tbPH{pɔ̀gɪ̀sɪ́ɛ́} \tbPS{} \tbGE{woman who is menstruating} 
\tbLX{pɔgeseɛ} \tbHM{} \tbPH{pɔ̀gɪ̀sɪ́ɛ́} \tbPS{n.~} \tbGE{a type of yam} \tbOP{pɔgeseere} \tbTP{pɔgeseerɛɛ} 
\tbLX{pɔgeseɛ} \tbHM{} \tbPH{pɔ̀gɪ̀sɪ́ɛ́} \tbPS{n.~} \tbSN \tbGE{a woman’s cloth measuring about 72 x 45 inches worn on the waist} \tbSN \tbGE{any such size of cloth} \tbSG{seerɛɛ} \tbOP{pɔgeseere} \tbTP{pɔge} 
\tbLX{pɔgeserekulo} \tbHM{} \tbPH{pɔ̀gɪ̀sɪ̀rɪ̀kúlò} \tbPS{n.~} \tbGE{marriageable woman} \tbOP{ pɔgeserekulibo} \tbTP{pɔgeserekulibee} 
\tbLX{pɔgesoga} \tbHM{} \tbPH{pɔ̀gɪ̀sʊ́gá} \tbPS{n.~} \tbGE{middle wife (one in between the first and last of three or more wives)} \tbOP{pɔgesogre} \tbTP{pɔgesogrɛɛ} 
\tbLX{pɔgesoɔbɔ} \tbHM{} \tbPH{pɔ̀gɪ̀sʊ́ɔ́bɔ́} \tbPS{n.~} \tbGE{witch} \tbOP{pɔgesoɔre} \tbTP{pɔgesoɔrɛɛ} 
\tbLX{Pɔgesugilo} \tbHM{} \tbPH{pɔ̀gɪ̀súgiló} \tbPS{n.~} \tbGE{female name traditionally given to the second born after twins} 
\tbLX{pɔgesugo} \tbHM{} \tbPH{pɔ̀gɪ̀súgó} \tbPS{n.~} \tbGE{a large group of women} \tbOP{pɔgesugri} \tbTP{pɔgesugree} 
\tbLX{pɔge-tu} \tbHM{} \tbPH{pɔ̀gɪ̀-tù} \tbPS{n.~} \tbGE{taking home of a woman who failed to divorce} 
\tbLX{pɔge-tu} \tbHM{} \tbPH{pɔ̀gɪ̀-tú} \tbPS{n.~} \tbGE{final stage in courtship} 
\tbLX{pɔgetuuro} \tbHM{} \tbPH{pɔ̀gɪ̀túúró} \tbPS{n.~} \tbGE{one who takes a woman from her family to her husband's house} \tbOP{pɔgetuuribo} \tbTP{pɔgetuurbee} 
\tbLX{Pɔgeweɛ} \tbHM{} \tbPH{pɔ̀gɪ̀wɪ̀ɛ́} \tbPS{n.~} \tbGE{name traditionally given to a female child born in the farm/bush (outside of the normal living environment)} 
\tbLX{pɔgeyaa} \tbHM{} \tbPH{pɔ̀gɪ̀yáá} \tbPS{n.~} \tbGE{daughter} \tbOP{pɔgeyaare} \tbTP{pɔgeyaarɛɛ} 
\tbLX{pɔgeyagede} \tbHM{} \tbPH{pɔ̀gɪ̀yágɪ́dɪ́} \tbPS{n.~} \tbGE{polygamy} 
\tbLX{pɔgeyi-gane} \tbHM{} \tbPH{pɔ̀gɪ̀yí-gánɪ̀} \tbPS{n.~} \tbGE{writ of divorce} \tbOP{pɔgeyi-gama} \tbTP{pɔgeyi-gannɛɛ} 
\tbLX{pɔgeyiraa} \tbHM{} \tbPH{pɔ̀gɪ̀yíráá} \tbPS{n.~} \tbGE{divorced woman} \tbOP{pɔgeyirri} \tbTP{pɔgeyirree} 
\tbLX{Pɔgezeɛ} \tbHM{} \tbPH{pɔ̀gɪ̀zɪ̀ɛ́} \tbPS{n.~} \tbSN \tbGE{light-skinned woman} \tbSN \tbGE{female name} \tbOP{pɔgezeere/pɔgezeɛmine} 
\tbLX{pɔlaa} \tbHM{} \tbPH{pɔ̀làà} \tbPS{n.~} \tbGE{heartbeat, pulse} \tbOP{pɔlle} \tbTP{pɔllɛɛ} 
\tbLX{pɔlaare} \tbHM{} \tbPH{pɔ́láárɪ̀} \tbPS{v.~} \tbGE{to swear vehemently} \tbPD{pɔlaarɛɛ, poɔrɔlaara, pɔlaara, pɔlaareba, pɔlaaraa, - } 
\tbLX{pɔlbile} \tbHM{} \tbPH{pɔ́lbílé} \tbPS{n.~} \tbGE{young man} \tbXV{pɔlbilii} \tbXE{} \tbOP{pɔlbilee} \tbVA{pɔle, pɔlee}
\tbLX{pɔle} \tbHM{} \tbPH{pɔ́lɪ̀} \tbPS{n.~} \tbGE{young boy} \tbOP{pɔlbilii} \tbVA{pɔlee, pɔlbile}
\tbLX{pɔle} \tbHM{} \tbPH{pɔ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to grow to maturity} \tbSN \tbGE{to be full of life} \tbSN \tbGE{to become beautiful} \tbPD{pɔlɛɛ, pɔllɔ, pɔllɔ, pɔleba, -, - } 
\tbLX{pɔlee} \tbHM{} \tbPH{pɔ́lɪ́ɪ́} \tbPS{n.~} \tbGE{young man} \tbVA{pɔle, pɔlbile}
\tbLX{pɔlese} \tbHM{} \tbPH{pɔ́lɪ́sɪ̀} \tbPS{v.~} \tbGE{to polish} \tbPD{pɔleseɛ, pɔleseerɛ, pɔleseerɛ, pɔleseerebɛ, pɔleseeraa, - } 
\tbLX{pɔlkakyeerɛ} \tbHM{} \tbPH{pɔ̌lkáʧɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{the act of being overly proud of one’s self} 
\tbLX{pɔllɔ} \tbHM{} \tbPH{pɔ̀llɔ̀} \tbPS{n.~} \tbGE{pathway used by wild rodents} \tbOP{pɔlle} \tbTP{pɔllɛɛ} \tbVA{pɔle}
\tbLX{pɔloo} \tbHM{} \tbPH{pɔ́lʊ́ʊ́} \tbPS{n.~} \tbGE{liveliness} \tbXV{Yɛ yieluŋ na taa la pɔloo.} \tbXE{That song of yours is lively.} 
\tbLX{pɔloo} \tbHM{} \tbPH{pɔ́lʊ́ʊ́} \tbPS{n.~} \tbSN \tbGE{pride} \tbSN \tbGE{vanity} \tbXV{A pɔgelee na veɛlɛ la kyɛ o erɛ la pɔloo.} \tbXE{That young lady is beautiful but she is proud.} 
\tbLX{pɔlteɛ} \tbHM{} \tbPH{pɔ̀ltìɛ̀} \tbPS{n.~} \tbGE{frightening} 
\tbLX{pɔmɛɛ} \tbHM{} \tbPH{pɔ̀mɛ̀ɛ̀} \tbPS{adv.~} \tbGE{fat and round} 
\tbLX{pɔmpɔre} \tbHM{} \tbPH{pɔ̌mpɔ̀rɪ́} \tbPS{adj.~} \tbGE{very young and immature e.g. puppy} 
\tbLX{pɔnne} \tbHM{} \tbPH{pɔ̀nnɪ̀} \tbPS{v.~} \tbSN \tbGE{to hold onto someone's chest or belly, but not firmly} \tbXV{A saa naŋ iri la ka a bie zɛle o yɔɔlee pɔnne a zɛlzɛlɛ zoro ne.} \tbXE{As the rain started, the child picked his brother and half dragged him away.} \tbSN \tbGE{to be calm} \tbXV{Saseɛ daare erɛ la kyɛ saa mie la ka zie pɔnne.} \tbXE{It used to be windy but with the coming of rain, the weather is calm.} \tbPD{pɔnnɛɛ, pɔnnɔ, pɔnnɔ, pɔnema, pɔnnaa, - } 
\tbLX{pɔntenakaare} \tbHM{} \tbPH{pɔ̀ntɪ́nàkáárɪ́} \tbPS{n.~} \tbGE{one that is here and there} \tbTP{pɔntenakaarɛɛ} 
\tbLX{pɔntere} \tbHM{} \tbPH{pɔ̀ntɪ́rɪ̀} \tbPS{n.~} \tbGE{toad} \tbOP{pɔnteɛ} \tbTP{pɔnteerɛɛ} 
\tbLX{pɔɔ} \tbHM{} \tbPH{pɔ̀ɔ́} \tbPS{v.~} \tbGE{to be too small, to be too young} \tbPD{pɔɔɛ, pɔɔrɔ, pɔɔrɔ, pɔɔrebɔ, pɔɔraa, - } 
\tbLX{pɔɔre} \tbHM{} \tbPH{pɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{a short and poisonous snake} \tbVA{mabiekontapɔgema}
\tbLX{pɔpɔrɛɛ} \tbHM{} \tbPH{pɔ́pɔ́rɛ́ɛ́} \tbPS{n.~} \tbGE{} \tbSG{pɔpɔroo} \tbVA{shortened form of sepɔpɔrɛɛ}
\tbLX{pɔraa} \tbHM{} \tbPH{pɔ́ràà} \tbPS{n.~} \tbSN \tbGE{an ordeal to find a wrongdoer} \tbSN \tbGE{the condition of hanging in the balance} \tbOP{pɔrre} \tbTP{pɔrrɛɛ} 
\tbLX{pɔre} \tbHM{} \tbPH{pɔ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to pluck (fruit)} \tbSN \tbGE{to die (baby)} \tbPD{pɔrɛɛ, pɔrɔ, pɔrɔ, pɔreba, pɔraa, - } 
\tbLX{pɔre} \tbHM{} \tbPH{pɔ́rɪ́} \tbPS{v.~} \tbGE{to carry out a test to find a wrongdoer} \tbPD{pɔrɛɛ, pɔrɔ, pɔrɔ, pɔreba, pɔraa, - } 
\tbLX{pɔ saa} \tbHM{} \tbPH{pɔ́ sáá} \tbPS{v.~} \tbGE{to jump from a very high point e.g. lizard from the tip of a tree branch} \tbPD{-, -, -, -, -, - } 
\tbLX{pu} \tbHM{} \tbPH{pù} \tbPS{v.~} \tbGE{to flower} \tbPD{pue, puuro, puuro, puuribo, puuraa, - } \tbVA{puuri}
\tbLX{pu} \tbHM{} \tbPH{pù} \tbPS{v.~} \tbGE{to pluck very tender fruits e.g. okra} \tbPD{pue, puuro, puuro, puuribo, puuraa, - } \tbVA{puuri}
\tbLX{pu} \tbHM{} \tbPH{pú} \tbPS{v.~} \tbGE{direct to} \tbXV{O naŋ yeli a yɛlɛ o ba pore neɛ zaa yuori kyɛ te baŋe la neɛ na o naŋ pu.} \tbXE{S/he did not mention any person’s name in the talk but we know who it was directed to.} \tbPD{-, -, -, -, -, - } 
\tbLX{pugi} \tbHM{} \tbPH{pùgì} \tbPS{v.~} \tbGE{praise, adore} \tbPD{-, -, -, -, -, - } 
\tbLX{pugi} \tbHM{} \tbPH{púgí} \tbPS{v.~} \tbSN \tbGE{to disturb e.g. dust, water} \tbSN \tbGE{to knead e.g. flour or shea butter dough} \tbPD{pugee, pugiro, pugiro, pugiribo, pugiraa, - } 
\tbLX{pugi} \tbHM{} \tbPH{pùgì} \tbPS{v.~} \tbGE{to praise} \tbPD{pugee, pugiro, pugiro, pugiribo, pugiraa, - } 
\tbLX{pugo} \tbHM{} \tbPH{púgó} \tbPS{adj.~} \tbGE{gray} \tbOP{pugiri} \tbTP{pugiree} 
\tbLX{puli} \tbHM{} \tbPH{púlí} \tbPS{adv.~} \tbGE{under, below} 
\tbLX{puli} \tbHM{} \tbPH{pùlí} \tbPS{n.~} \tbGE{an emergency exit of rodents that live in holes} \tbOP{pulo} \tbTP{pulee} 
\tbLX{puli} \tbHM{} \tbPH{púlí} \tbPS{n.~} \tbGE{hind toe of a bird} \tbOP{pulli} \tbTP{pulee} 
\tbLX{pulli} \tbHM{} \tbPH{pùllì} \tbPS{v.~} \tbGE{to mix} \tbXV{Ba de la bɛŋɛ pulli ne a mui.} \tbXE{They mixed the rice with beans.} \tbPD{pulee, pullo, pullo, pulibo, pullaa, pulibo } 
\tbLX{pulli} \tbHM{} \tbPH{púllì} \tbPS{v.~} \tbSN \tbGE{to repeat} \tbSN \tbGE{to fold over} \tbPD{pullee, pullo, pullo, pulibo, pullaa, pulibo } 
\tbLX{pulluu} \tbHM{} \tbPH{púllúú} \tbPS{n.~} \tbSN \tbGE{repetition} \tbSN \tbGE{refrain (poem, song)} 
\tbLX{pulompuloŋ} \tbHM{} \tbPH{pùlòmpúlóŋ} \tbPS{adj.~} \tbGE{mixed up e.g. different items} 
\tbLX{puloŋ} \tbHM{} \tbPH{púlóŋ} \tbPS{n.~} \tbGE{sound of a small object falling into water} 
\tbLX{pulopulo} \tbHM{} \tbPH{púlópúló} \tbPS{adj.~} \tbGE{less serious than} \tbXV{Ka fooŋ la yi tensogɔ saŋa, a feɛbo ŋa fo ba naŋ feɛ fo zenɛ na waa la pulopulo.} \tbXE{If you go out in the night again, the punishment you have received today will be less serious than what you will get.} 
\tbLX{pulopulo} \tbHM{} \tbPH{púlópúló} \tbPS{adj.~} \tbGE{incomparable} \tbXV{Tentoŋ saŋa ŋmenaa maŋ tasoga puri la ka vũũ meŋ waa pulopulo.} \tbXE{During the hot season sometimes the heat of the sun is incomparable to fire.} 
\tbLX{pulopulo} \tbHM{} \tbPH{pùlópùló} \tbPS{adj.~} \tbGE{made of different varieties} \tbXV{A tigiri daare o da di la bondi pulopulo te zuo lɛnso ka o poɔ da beɛrɛ.} \tbXE{On the feast day s/he ate too many assorted dishes and that was why s/he had a stomach ache.} 
\tbLX{puluŋ} \tbHM{} \tbPH{pùlúŋ} \tbPS{n.~} \tbGE{a type of grass with sharp tips that can pierce soft or light objects e.g. feet} 
\tbLX{pumpuŋkɔrɔ} \tbHM{} \tbPH{púmpúŋkɔ̀rɔ̀} \tbPS{n.~} \tbGE{a type of poisonous plant} \tbOP{pumpuŋkɔrɛɛ} 
\tbLX{pumpuroŋ} \tbHM{} \tbPH{púmpúróŋ} \tbPS{adv.~} \tbGE{small, round, and hairy e.g. tail end of a squirrel} 
\tbLX{pumpuroŋ} \tbHM{} \tbPH{pùmpùròŋ} \tbPS{adv.~} \tbGE{large, round, and furry} 
\tbLX{puo} \tbHM{} \tbPH{púò} \tbPS{n.~} \tbSN \tbGE{small farm near the house} \tbSN \tbGE{farmland} \tbOP{poe} \tbTP{poree} 
\tbLX{pu᷉o} \tbHM{} \tbPH{pù᷉ò᷉} \tbPS{v.~} \tbGE{to make a hole through an object} \tbPD{pu᷉oe, puono, puono, puono, puonimo, puonaa } 
\tbLX{puoli} \tbHM{} \tbPH{púólí} \tbPS{v.~} \tbGE{to saturate with water} \tbPD{puolee, puolo, puolo, puolibo, puolaa, - } 
\tbLX{puoraa} \tbHM{} \tbPH{pùòràá} \tbPS{n.~} \tbSN \tbGE{greetings} \tbSN \tbGE{gratitude} \tbOP{puoro} 
\tbLX{puoraa karetaa} \tbHM{} \tbPH{pùòràá káretàá} \tbPS{n.~} \tbGE{postcard} \tbOP{puoraa karetaare} \tbTP{puoraa karetaarɛɛ} 
\tbLX{puori} \tbHM{} \tbPH{pùòrí} \tbPS{n.~} \tbGE{back (body)} \tbOP{poe} \tbTP{poree} 
\tbLX{puori} \tbHM{} \tbPH{púórì} \tbPS{n.~} \tbGE{stomach; the organ in humans and animals in which food is stored} \tbOP{poe} \tbTP{puoree} 
\tbLX{puori} \tbHM{} \tbPH{pùòrì} \tbPS{v.~} \tbSN \tbGE{to pray} \tbSN \tbGE{to greet} \tbSN \tbGE{to thank} \tbPD{puoree, puoro, puoro, puoribo, puoraa, puoribo } 
\tbLX{puori} \tbHM{} \tbPH{pùòrí} \tbPS{adv.~} \tbSN \tbGE{behind} \tbSN \tbGE{backward (direction)} 
\tbLX{puoriŋ} \tbHM{} \tbPH{púòríŋ} \tbPS{conj.~} \tbGE{after} 
\tbLX{puoruu} \tbHM{} \tbPH{púórúú} \tbPS{n.~} \tbGE{prayer} \tbOP{puoro} 
\tbLX{puree} \tbHM{} \tbPH{pùrèé} \tbPS{n.~} \tbGE{paternal aunt, father's sister} \tbOP{purri/purimine} \tbTP{purree} 
\tbLX{puri} \tbHM{} \tbPH{pùrì} \tbPS{v.~} \tbSN \tbGE{to burst, to break} \tbSN \tbGE{to appear suddenly} \tbSN \tbGE{to rise (sun)} \tbPD{puree, puro, puro, puribo, puraa, puribo } \tbSE{puri ŋmenaa} \tbGE{sunrise} 
\tbLX{puri wale} \tbHM{} \tbPH{pùrì wálɪ́} \tbPS{v.~} \tbSN \tbGE{to sweat, to perspire} \tbSN \tbGE{to struggle, to be in a difficult situation} \tbPD{-, -, -, -, -, - } 
\tbLX{puro} \tbHM{} \tbPH{púrò} \tbPS{n.~} \tbGE{a type of tree that has sour sap which is used in preparing porridge} \tbOP{purri} \tbTP{purree} 
\tbLX{puro} \tbHM{} \tbPH{púrò} \tbPS{n.~} \tbGE{a type of small tree squirrel} \tbOP{purri} \tbTP{purree} 
\tbLX{puu} \tbHM{} \tbPH{pùù} \tbPS{v.~} \tbGE{to praise, to congratulate} \tbPD{puue, puuro, puuro, puuribo, puuraa, - } 
\tbLX{puulee} \tbHM{} \tbPH{púúléé} \tbPS{n.~} \tbGE{lady (typically used with bi-)} \tbOP{puuli} 
\tbLX{puuli} \tbHM{} \tbPH{púúlì} \tbPS{n.~} \tbGE{a funeral rhythm; a type of musical performance at a funeral rites session} \tbOP{puulo} \tbTP{puulee} 
\tbLX{puuri} \tbHM{} \tbPH{púúrí} \tbPS{n.~} \tbGE{flowers} \tbSG{puuruu} \tbTP{puuree} \tbVA{tepuuri}
\tbLX{puuri} \tbHM{} \tbPH{pùùrì} \tbPS{v.~} \tbGE{to spray from the mouth (usually chewed material as medicine)} \tbPD{puuree, puuro, puuro, puuribo, puuraa, - } \tbVA{pu}
\tbLX{puuri} \tbHM{} \tbPH{pùùrì} \tbPS{v.~} \tbGE{to harvest okra prematurely} \tbPD{puuree, puuro, puuro, puuribo, puuraa, - } \tbVA{pu}
\tbLX{sa} \tbHM{} \tbPH{sá} \tbPS{v.~} \tbGE{to administer an enema} \tbPD{saɛ, saara, saara, sareba, saaraa, - } 
\tbLX{saa} \tbHM{} \tbPH{sàá} \tbPS{n.~} \tbGE{father} \tbOP{saamine} \tbTP{saarɛɛ} \tbVA{ba}
\tbLX{saa} \tbHM{} \tbPH{sáá} \tbPS{n.~} \tbGE{rain} \tbTP{saarɛɛ} 
\tbLX{saa} \tbHM{} \tbPH{sáá} \tbPS{adv.~} \tbGE{even if; short form of saareŋ} \tbXV{Ka fooŋ saa be gaa kyɛ fo koŋ nyɛ o.} \tbXE{Even if you go you wont see him/her.} \tbVA{shortened form of saareŋ}
\tbLX{sa᷉a} \tbHM{} \tbPH{sá᷉á} \tbPS{n.~} \tbGE{blacksmith's shop} \tbVA{saaŋ}
\tbLX{sa᷉a} \tbHM{} \tbPH{sá᷉à᷉} \tbPS{v.~} \tbGE{to mash, to mix} \tbXV{Kɔŋe kpɛ la a balee, sa᷉a saabo eŋe o.} \tbXE{The puppy is hungry; mash some TZ for it.} \tbPD{-, -, -, -, -, - } 
\tbLX{sa᷉a} \tbHM{} \tbPH{sá᷉à᷉} \tbPS{v.~} \tbGE{to appear, to be visible} \tbPD{-, -, -, -, -, - } 
\tbLX{sa᷈a} \tbHM{} \tbPH{sà᷈à} \tbPS{v.~} \tbSN \tbGE{to spoil, to destroy, to damage} \tbSN \tbGE{to be guilty} \tbPD{sa᷈aɛ, saana, saana, saanema, saanaa, saanema } 
\tbLX{sa᷉abile} \tbHM{} \tbPH{sà᷉á᷉bìlé} \tbPS{n.~} \tbGE{uncle (father's younger brother)} \tbOP{saabilii/saabilimine} \tbTP{saabilee} \tbVA{saambile}
\tbLX{saabo} \tbHM{} \tbPH{sáábʊ́} \tbPS{n.~} \tbGE{flour paste (staple food of the Dagaare-speaking people)} \tbTP{saarɛɛ} \tbVA{TZ}
\tbLX{saabuulaa} \tbHM{} \tbPH{sààbùùláà} \tbPS{n.~} \tbGE{porridge set for preparing saabo} \tbOP{saabuuli} 
\tbLX{saadaare} \tbHM{} \tbPH{sáádáárɪ́} \tbPS{adv.~} \tbGE{rainy, wet day} 
\tbLX{saadoge} \tbHM{} \tbPH{sààdʊ́gɪ́} \tbPS{n.~} \tbGE{large cooking pot} \tbOP{saadogere} \tbTP{saadogerɛɛ} 
\tbLX{saa-doge} \tbHM{} \tbPH{sààdʋ́gɪ̀} \tbPS{n.~} \tbGE{earthenware cooking pot for soup} \tbVA{zɛdogelee}
\tbLX{sa᷉adɔɔ} \tbHM{} \tbPH{sá᷉á᷉dɔ́ɔ́} \tbPS{n.~} \tbGE{stranger (male)} \tbOP{sa᷉adɔbɔ} 
\tbLX{saafebiri} \tbHM{} \tbPH{sààfɪ̀bírì} \tbPS{n.~} \tbGE{key} \tbOP{saafebie} \tbTP{saafebiruu} 
\tbLX{saafee} \tbHM{} \tbPH{sáàfɪ́ɪ́} \tbPS{n.~} \tbGE{lock} \tbOP{saafeere} \tbTP{saafeerɛɛ} \tbVA{safoa, safoɔ}
\tbLX{sa-ageraa} \tbHM{} \tbPH{sà-ágɪ́ráà} \tbPS{n.~} \tbGE{rainclouds that gather and suddenly give way to clear weather} \tbOP{sa-agere} \tbTP{sa-agerɛɛ} 
\tbLX{saakaŋkaŋaa} \tbHM{} \tbPH{sáákǎŋkàŋáá} \tbPS{n.~} \tbGE{a type of tree} \tbOP{saakaŋkanne} \tbTP{saakaŋkannɛɛ} \tbVA{kaŋkaŋaa}
\tbLX{saakaraa} \tbHM{} \tbPH{sààkáráá} \tbPS{n.~} \tbGE{flour paste not fully cooked, still part raw} \tbOP{saakarare} \tbTP{saakarɛɛ} 
\tbLX{sa᷉akoma} \tbHM{} \tbPH{sà᷉à᷉kʊ́má} \tbPS{n.~} \tbGE{grandfather} \tbOP{saakommine} \tbTP{saakomɛɛ/saakonnɛɛ} \tbVA{saaŋkoma, naa}
\tbLX{sa᷉akonnoŋ} \tbHM{} \tbPH{sa ̃̀ã̀ kʊ́ nnʊ́ ŋ} \tbPS{n.~} \tbGE{tradition, custom} 
\tbLX{sa᷉akontulimo} \tbHM{} \tbPH{sa᷉ ̀ a᷉ ̀kʋ́ntúlímó} \tbPS{n.~} \tbGE{great-grandfather} \tbOP{sa᷉akontulimine} \tbVA{sa᷉akontulo}
\tbLX{sa᷉akoŋkommuulitoori} \tbHM{} \tbPH{sà᷅àkʊ̀ŋkʊ́mmúúlítóórí} \tbPS{n.~} \tbGE{great-grandfather} \tbOP{sa᷉akoŋkommuulitoorimine} \tbVA{sa᷉akontulimo}
\tbLX{sa᷉akoŋkorema} \tbHM{} \tbPH{sà᷉àkʊ̌ŋkʊ́rɪ́má} \tbPS{n.~} \tbGE{great-great-grandfather, forefather} \tbOP{sa᷉akoŋkoremamine} 
\tbLX{sa᷉akoŋkoroma} \tbHM{} \tbPH{sá᷉á᷉kʊ́kʊ́rʊ́ma} \tbPS{n.~} \tbGE{great-great-grandfather} \tbVA{sa᷉akontulimotulimo}
\tbLX{saako᷉ɔ᷉} \tbHM{} \tbPH{sààkʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{water in mashed poridge} \tbTP{saakoɔnɛɛ} 
\tbLX{saakɔntuo} \tbHM{} \tbPH{sáákɔ̌ntúó} \tbPS{n.~} \tbGE{Water from the first rain of the onset of the rainy season} \tbOP{saakɔntuuri} \tbTP{saakɔntuuree} 
\tbLX{saala} \tbHM{} \tbPH{sáálá} \tbPS{n.~} \tbGE{charcoal} \tbSG{saaloo} \tbTP{saalɛɛ} 
\tbLX{saalaa} \tbHM{} \tbPH{sààlàà} \tbPS{adj.~} \tbSN \tbGE{smooth} \tbSN \tbGE{slippery} \tbOP{saale} 
\tbLX{saalaa} \tbHM{} \tbPH{sááláá} \tbPS{n.~} \tbGE{object used for sharpening implements} \tbOP{saale} \tbTP{saaɛɛ} 
\tbLX{saalaa} \tbHM{} \tbPH{sààlàà} \tbPS{n.~} \tbGE{a type of tree with hard wood} \tbOP{saale} 
\tbLX{saalaa} \tbHM{} \tbPH{sààláá} \tbPS{n.~} \tbGE{bowl of T.Z} \tbOP{saalaare} \tbTP{saalaarɛɛ} 
\tbLX{saale} \tbHM{} \tbPH{sààlɪ̀} \tbPS{v.~} \tbSN \tbGE{to sharpen e.g. knife} \tbSN \tbGE{to make smooth, to plaster} \tbXV{Ayuo saale la o dipaalaa na o serɛ naŋ ŋmaa.} \tbXE{Ayuo has plastered the new room her husband put up.} \tbPD{saalɛɛ, saala, saala, saaleba, saalaa, - } 
\tbLX{saale} \tbHM{} \tbPH{sààlɪ̀} \tbPS{v.~} \tbGE{to be sufficient} \tbXV{A bondirii ama eŋ na saale ba la.} \tbXE{As for this food, it will be sufficient for them.} \tbPD{saalɛɛ, saala, saala, saaleba, saalaa, - } 
\tbLX{saalesɔglaa} \tbHM{} \tbPH{sààlɪ̀sɔ́gláá} \tbPS{n.~} \tbGE{a type of tree with hard wood and medicinal properties} \tbOP{saalesɔglɔ} \tbTP{saalesɔglɛɛ} 
\tbLX{saalieni} \tbHM{} \tbPH{sààlìèní} \tbPS{n.~} \tbGE{a type of small plant with an edible, jellylike tuber} \tbOP{saalieme} \tbTP{saalienee} 
\tbLX{saalombugo} \tbHM{} \tbPH{sààlʊ̀mbùgò} \tbPS{n.~} \tbGE{a type of small plant whose leaves are used in place of okra} \tbTP{saalombugiree} \tbVA{saalompelaa}
\tbLX{saalommaane} \tbHM{} \tbPH{sààlʊ̌mmáánɪ̀ɪ̀} \tbPS{n.~} \tbGE{okra} \tbOP{saalommaana} \tbVA{maanee}
\tbLX{saalompelaa} \tbHM{} \tbPH{sààlʊ̀mpɪ̀làà} \tbPS{n.~} \tbGE{a type of small plant whose leaves are used in place of okra} \tbOP{saalompeɛle} \tbTP{saalompeɛlɛɛ} \tbVA{saalombugo}
\tbLX{saaloŋ} \tbHM{} \tbPH{sààlʊ̀ŋ} \tbPS{adj.~} \tbGE{slippery, slimy} \tbXV{A mui sa᷉a᷉ɛ᷉ la, a waa la saaloŋ lɛ.} \tbXE{The rice is spoilt; it is slimy.} 
\tbLX{saaloŋ} \tbHM{} \tbPH{sáálʊ́ŋ} \tbPS{n.~} \tbGE{okra or other slimy plant used in soup} 
\tbLX{saalɔre} \tbHM{} \tbPH{sààlɔ́rɪ́} \tbPS{n.~} \tbGE{ball of T.Z. (usually formed by putting T.Z in water)} \tbSG{saalɔroo} \tbOP{saalɔrɔ} \tbTP{saalɔrɛɛ} 
\tbLX{saana} \tbHM{} \tbPH{sáánà} \tbPS{n.~} \tbGE{stranger, visitor} \tbOP{saama} \tbTP{saanɛɛ} 
\tbLX{saanaa} \tbHM{} \tbPH{sáánáá} \tbPS{n.~} \tbGE{fault, wrong} \tbOP{saane} 
\tbLX{saanweɛleŋ} \tbHM{} \tbPH{sáánwɪ́ɛ́lɪ́ŋ} \tbPS{n.~} \tbSN \tbGE{opposite, contradiction} \tbSN \tbGE{irony (literature)} \tbVA{saanweɛloŋ}
\tbLX{saaŋ} \tbHM{} \tbPH{sááŋ} \tbPS{n.~} \tbGE{see zambɛre} \tbOP{saane} \tbTP{saanɛɛ} \tbVA{sa᷉a}
\tbLX{saapɛle} \tbHM{} \tbPH{sààpɛ́lɪ́} \tbPS{n.~} \tbGE{layer of T.Z} \tbOP{saapɛlɛ} \tbTP{saapɛlɛɛ} 
\tbLX{sa᷉apɔge} \tbHM{} \tbPH{sá᷉á᷉pɔ́gɪ́} \tbPS{n.~} \tbGE{stranger (female)} 
\tbLX{saapuori} \tbHM{} \tbPH{sàápúórì} \tbPS{n.~} \tbGE{the middle of T.Z} \tbOP{saapoe} \tbTP{saaporee} 
\tbLX{saapurimo} \tbHM{} \tbPH{sààpúrímó} \tbPS{n.~} \tbGE{pieces of flour paste (T.Z.)} \tbTP{saapurimee} 
\tbLX{saaraa} \tbHM{} \tbPH{sááráá} \tbPS{n.~} \tbGE{saw} \tbOP{saare} \tbTP{saarɛɛ} 
\tbLX{saare} \tbHM{} \tbPH{sáárɪ̀} \tbPS{n.~} \tbGE{broom} \tbOP{saɛ} \tbTP{saarɛɛ} 
\tbLX{saaremuri} \tbHM{} \tbPH{sáármúrí} \tbPS{n.~} \tbGE{broom straw} \tbOP{saaremue} \tbTP{saamuree} 
\tbLX{saareŋ} \tbHM{} \tbPH{sáárɪ́ŋ} \tbPS{adj.~} \tbGE{even if} \tbXV{Ka fooŋ saareŋ gaa kyɛ fo koŋ nyɛ o.} \tbXE{Even if you go, you will not see her.} \tbVA{saa}
\tbLX{saa-toloŋ} \tbHM{} \tbPH{sàá-tʊ́lʊ́ŋ} \tbPS{n.~} \tbGE{hot TZ} \tbOP{saa-tolɔ/saa-tola} 
\tbLX{saazu-kogilo} \tbHM{} \tbPH{sààzú-kógiló} \tbPS{n.~} \tbGE{dried top of T.Z} \tbSG{saazu-kogiluu} 
\tbLX{sabaabo} \tbHM{} \tbPH{sàbáábʊ́} \tbPS{adj.~} \tbGE{mysterious} 
\tbLX{sabeelɛ} \tbHM{} \tbPH{sàbɪ́ɪ́lɛ̀} \tbPS{n.~} \tbGE{civilian} \tbSG{sabeeloo} 
\tbLX{sabɛlemɛ} \tbHM{} \tbPH{sàbɛ́lɪ́mɛ́} \tbPS{n.~} \tbGE{earthworm} \tbOP{sabɛlemɛre} \tbTP{sabɛlemɛrɛɛ} \tbVA{sɛbɛlemɛ, short form of sabɛlmɛkyɔɔ}
\tbLX{sabɛlemɛkyɔɔ} \tbHM{} \tbPH{sàbɛ́lɪ́mɛ̀ʧɔ́ɔ́} \tbPS{n.~} \tbGE{earthworm} \tbOP{sabɛlemɛkyɔɔre} \tbTP{sabɛlemɛkyɔɔrɛɛ} \tbVA{sabɛlemɛ}
\tbLX{sabɛlemɛkyɔɔ} \tbHM{} \tbPH{sàbɛ́lɪ́mɛ́kyɔ́ɔ́} \tbPS{n.~} \tbGE{earthworm} \tbOP{sabɛlemɛkyɔɔre} \tbTP{sabɛlemɛkyɔɔrɛɛ} \tbVA{sabɛlemɛ; sɛbɛlemɛkyɔɔ,  sɛbɛlemɛ}\tbVA{sɛbɛlemɛkyɔɔ, sabɛlemɛ/ sɛbɛlemɛ.}
\tbLX{sabiri} \tbHM{} \tbPH{sàbírì} \tbPS{n.~} \tbGE{raindrop} \tbOP{sabie} \tbTP{sabiiree} 
\tbLX{sabogi} \tbHM{} \tbPH{sábógí} \tbPS{n.~} \tbGE{something that causes concern e.g incessant complain} \tbXV{A yi a pɔge bie naŋ kpi la ka o deɛ maŋ yele sabogi yɛlɛ.} \tbXE{Since the woman lost her child she has always been complaining.} 
\tbLX{sabogi} \tbHM{} \tbPH{sàbògí} \tbPS{n.~} \tbGE{a type of fetish believed to have healing power and protection especially for children} \tbOP{sabogri} \tbTP{sabogree} 
\tbLX{sadɛŋe} \tbHM{} \tbPH{sàdɛ́ŋ} \tbPS{n.~} \tbGE{first rain (first heavy rain after which the people sow their crops)} \tbOP{sadɛnne} \tbTP{sadɛnnɛɛ} \tbVA{sadɛŋɛ}
\tbLX{safeaa} \tbHM{} \tbPH{sàfɪ́áá} \tbPS{n.~} \tbGE{survey} \tbOP{safeere} \tbTP{safeerɛɛ} 
\tbLX{safekyeɛrɛ} \tbHM{} \tbPH{sàfɪ̀ʧɪ̀ɛ̀rɛ̀} \tbPS{n.~} \tbGE{surveyor} \tbOP{safekyeɛrebɛ} \tbTP{safekyeɛrebɛɛ} 
\tbLX{safoa} \tbHM{} \tbPH{sàfʋ̀á} \tbPS{n.~} \tbGE{lock} \tbOP{safoore} \tbTP{safoorɛɛ} \tbVA{safoɔ}
\tbLX{safobiri} \tbHM{} \tbPH{sàfʊ̀bírì} \tbPS{n.~} \tbGE{key} \tbOP{safobie} \tbTP{safobiiree} \tbVA{saafobiri, saafebiri}
\tbLX{safoɔ} \tbHM{} \tbPH{sàfʊ̀ɔ́} \tbPS{n.~} \tbGE{lock} \tbOP{safoore} \tbTP{safoorɛɛ} \tbVA{safoa, saafee}
\tbLX{sagane} \tbHM{} \tbPH{sàgánɪ̀} \tbPS{n.~} \tbGE{sky blue} \tbOP{sagama} \tbTP{sagannɛɛ} 
\tbLX{sage} \tbHM{} \tbPH{sàgɪ̀} \tbPS{v.~} \tbSN \tbGE{to answer} \tbSN \tbGE{to agree, to accept, to obey} \tbSN \tbGE{to allow, to permit} \tbSN \tbGE{to admit guilt} \tbSN \tbGE{to receive} \tbSN \tbGE{to sing after a lead singer} \tbPD{sagɛɛ, sagra, sagra, sagreba, sagraa, sagreba } 
\tbLX{sage} \tbHM{} \tbPH{ságɪ́} \tbPS{v.~} \tbGE{to be piled up} \tbPD{sagɛɛ, sagera, sagera, sagereba, sageraa, sagereba } 
\tbLX{sagele} \tbHM{} \tbPH{sàgɪ̀lɪ̀} \tbPS{v.~} \tbGE{to be caught up; to be entangled e.g. stick thrown into a tree} \tbPD{sagelɛɛ, sagela, sagela, sageleba, sagelaa, - } 
\tbLX{sagere} \tbHM{} \tbPH{sàgɪ̀rɪ́} \tbPS{n.~} \tbGE{rubbish} \tbOP{saga} \tbTP{sagɛɛ} 
\tbLX{sagoŋgoŋŋo} \tbHM{} \tbPH{sàgòŋgóŋŋò} \tbPS{n.~} \tbGE{rainbow} \tbOP{sagoŋgonni} \tbTP{sagoŋgonnee} 
\tbLX{sagbini} \tbHM{} \tbPH{ságbíní} \tbPS{n.~} \tbGE{big bundle of traditional broomsticks. The local broom is usually a bundle of slim but strong straw} \tbOP{sagbime} \tbTP{sagbimee} 
\tbLX{sakaliŋkuuree} \tbHM{} \tbPH{sàkálíŋkúúréé} \tbPS{n.~} \tbGE{whirlwind} \tbOP{sakaliŋkuuri} 
\tbLX{sakere (Eng)} \tbHM{} \tbPH{sákɪ̀rɪ̀} \tbPS{n.~} \tbGE{bicycle} \tbOP{sakeɛ} \tbTP{sakerɛɛ} \tbVA{baasakuuri}
\tbLX{sako᷉ɔ} \tbHM{} \tbPH{sàkʊ̀᷉ɔ́᷉} \tbPS{n.~} \tbGE{heavy rain} \tbTP{sakoɔnɛɛ} 
\tbLX{sakoraa} \tbHM{} \tbPH{sàkóràá} \tbPS{n.~} \tbGE{gathered rain clouds; the condition of preparing to rain} \tbOP{sakorri} \tbTP{sakorree} 
\tbLX{sakɔre} \tbHM{} \tbPH{sàkɔ́rɪ́} \tbPS{n.~} \tbGE{scrap of saabo left in the pot} \tbTP{sakɔrɛɛ} 
\tbLX{sakɔre} \tbHM{} \tbPH{sákɔ́rɪ́} \tbPS{adj.~} \tbSN \tbGE{useless, lazy} \tbXV{Ba-sakɔre ba nɔŋ mɔyɔ.} \tbXE{A lazy dog does not like hunting.} \tbSN \tbGE{dirty, untidy} \tbXV{Nensaal-sakɔre zaa maŋ nyuuro la.} \tbXE{Every dirty person has an odour.} 
\tbLX{sakubie} \tbHM{} \tbPH{sákúbìé} \tbPS{n.~} \tbGE{student} \tbOP{sakubiiri} \tbTP{sakubiiree} 
\tbLX{sakubo} \tbHM{} \tbPH{sàkúbò} \tbPS{n.~} \tbGE{hail} \tbOP{sakuuri} \tbTP{sakuuree} 
\tbLX{sakuuri} \tbHM{} \tbPH{sàkúúrí} \tbPS{n.~} \tbSN \tbGE{hailstone, ice cube} \tbOP{sakue} \tbTP{sakuuree} 
\tbLX{sakuuri} \tbHM{} \tbPH{sàkùùrì} \tbPS{n.~} \tbGE{school} \tbOP{sakue} \tbTP{sakuuree} 
\tbLX{sakuuribie} \tbHM{} \tbPH{sàkúúríbíé} \tbPS{n.~} \tbGE{student} \tbOP{sakuuribiiri} \tbVA{sakubie}
\tbLX{sakperɛ} \tbHM{} \tbPH{sákpɪ́rɛ́} \tbPS{n.~} \tbGE{disused broom} \tbOP{sakperre} \tbTP{sakperɛɛ} 
\tbLX{sakyeɛre} \tbHM{} \tbPH{sàʧɪ̀ɛ̀rɪ̀} \tbPS{n.~} \tbGE{a type of hat with a large brim e.g. cowboy} \tbOP{sakyeɛrɛ} 
\tbLX{sakyɔre} \tbHM{} \tbPH{sàʧɔ̀rɪ̀} \tbPS{adv.~} \tbGE{height of the rainy season} 
\tbLX{salala} \tbHM{} \tbPH{sàlàlà} \tbPS{adj.~} \tbGE{dull, passive} 
\tbLX{salamante} \tbHM{} \tbPH{sàlàmántɪ̀} \tbPS{n.~} \tbGE{cement} \tbOP{salmantere} \tbTP{salmanterɛɛ} 
\tbLX{salansalaŋ} \tbHM{} \tbPH{sálánsáláŋ} \tbPS{adj.~} \tbGE{smooth} 
\tbLX{Salaŋa} \tbHM{} \tbPH{sàláŋá} \tbPS{n.~} \tbGE{a person who works as a cleaner and orderly maintenance of the innermost part of a chief’s palace} \tbOP{salanne} \tbTP{salannɛɛ} 
\tbLX{sale} \tbHM{} \tbPH{sàlɪ̀} \tbPS{v.~} \tbSN \tbGE{to be dispirited} \tbXV{N woŋ la o fieluu yɛle, ka n eŋɛ zaa sale.} \tbXE{When I heard of her failure, I was disturbed.} \tbSN \tbGE{to cook a small amount soup} \tbXV{A biyaala seŋ ka o sale zeɛre ko a bie.} \tbXE{The babysitter should be able to cook some small food for the child.} \tbPD{salɛɛ, salla, salla, salba, sallaa, - } 
\tbLX{sale} \tbHM{} \tbPH{sàlɪ̀} \tbPS{v.~} \tbGE{to make sauce slippery} \tbPD{-, -, -, -, -, - } 
\tbLX{sale} \tbHM{} \tbPH{sálɪ́} \tbPS{n.~} \tbGE{a type of plant with medicinal properties, believed to provide strength and power} \tbTP{salɛɛ} 
\tbLX{salema} \tbHM{} \tbPH{sálɪ̀má} \tbPS{n.~} \tbGE{gold} \tbTP{sallɛɛ/salemɛɛ} 
\tbLX{salembogi} \tbHM{} \tbPH{sàlɪ̀mbògí} \tbPS{n.~} \tbGE{gold mine} \tbOP{salembogri} \tbTP{salembogree} 
\tbLX{salempelaa} \tbHM{} \tbPH{sàlɪ̀mpɪ́làá} \tbPS{n.~} \tbGE{diamond} \tbOP{salempeɛle} \tbTP{salempeɛlɛɛ} 
\tbLX{salendoge} \tbHM{} \tbPH{sàlɪ̀ndʊ̀gɪ̀} \tbPS{n.~} \tbGE{treasury (traditional)} \tbOP{salendogre} \tbTP{salendogrɛɛ} \tbVA{salendoglee}
\tbLX{salendogelee} \tbHM{} \tbPH{sàlɪ̀ndʊ́gléé} \tbPS{n.~} \tbGE{a pot made of gold} \tbGE{a pot in which gold or treasure is kept} \tbOP{salendogebilii} \tbVA{salendoge}
\tbLX{saleŋgoɔ} \tbHM{} \tbPH{sálɪ́ŋgʊ́ɔ́} \tbPS{n.~} \tbGE{a type of tree with long, spiky thorns} \tbOP{saleŋgoore} \tbTP{saleŋgoorɛɛ} \tbVA{sansaŋgoɔ}
\tbLX{saleŋkuuri} \tbHM{} \tbPH{sàlɪ̀ŋkúúrí} \tbPS{n.~} \tbSN \tbGE{gold nugget} \tbSN \tbGE{pearl} \tbOP{saleŋkubo} 
\tbLX{salinsoo-naadaperaa} \tbHM{} \tbPH{sàlìnsòò-náádàpɪ́ráà} \tbPS{n.~} \tbGE{large grasshopper} \tbOP{salinso-naadaperre} \tbTP{salinso-naadaperɛɛ} 
\tbLX{salinsoori} \tbHM{} \tbPH{sàlìnsóórí} \tbPS{n.~} \tbGE{grasshopper} \tbOP{salinsobo} \tbTP{salinsooree} 
\tbLX{salinsuuri} \tbHM{} \tbPH{sálínsúúrí} \tbPS{n.~} \tbGE{shadow} \tbVA{dasuoluŋ}
\tbLX{saluoni} \tbHM{} \tbPH{s̀alúóní} \tbPS{n.~} \tbGE{sky} \tbOP{saluomo} \tbTP{saluonee} 
\tbLX{samafuu} \tbHM{} \tbPH{sàmàfúù} \tbPS{n.~} \tbGE{worth nothing in the end} 
\tbLX{samane} \tbHM{} \tbPH{sámánɪ́} \tbPS{n.~} \tbGE{immediate surroundings of a house} \tbOP{samama} \tbTP{samanɛɛ} 
\tbLX{samane} \tbHM{} \tbPH{sámánɪ́} \tbPS{n.~} \tbGE{toilet} \tbOP{samama} \tbTP{samanɛɛ} 
\tbLX{samasama} \tbHM{} \tbPH{sàmàsámà} \tbPS{n.~} \tbGE{sanitary inspector} \tbOP{sansammine} 
\tbLX{samasama} \tbHM{} \tbPH{sámásámá} \tbPS{adj.~} \tbGE{occupied and filled to capacity} 
\tbLX{sambakelee} \tbHM{} \tbPH{sámbákéléé} \tbPS{n.~} \tbGE{pottery shard} \tbOP{sambakelli} \tbVA{semekelee}
\tbLX{sambala} \tbHM{} \tbPH{sámbálá} \tbPS{n.~} \tbGE{a type of long, hairy grass} \tbSG{sambaloo} \tbOP{sambala} \tbTP{sambalɛɛ} 
\tbLX{sa-miiraa} \tbHM{} \tbPH{sa-miiraa} \tbPS{n.~} \tbGE{rain that is possible to come} \tbOP{sa-miiri} 
\tbLX{sampa} \tbHM{} \tbPH{sámpá} \tbPS{n.~} \tbGE{a type of shed the roof of which is used to store corn and the under as a place to relax} \tbOP{sampare} \tbTP{samparɛɛ} 
\tbLX{sampa} \tbHM{} \tbPH{sàmpá} \tbPS{n.~} \tbGE{raised stand with grass/stalks on it on which corn is stored or resting place} \tbOP{sampare} \tbTP{samparɛɛ} 
\tbLX{sampane} \tbHM{} \tbPH{sámpánɪ́} \tbPS{n.~} \tbGE{wooden foot-shaped mallet for beating floor to plaster it} \tbOP{sampama} \tbTP{sampanɛɛ} 
\tbLX{sampire} \tbHM{} \tbPH{sàmpìré} \tbPS{n.~} \tbGE{clusters of grass} \tbTP{sampirree} \tbVA{sempire}
\tbLX{sampulo} \tbHM{} \tbPH{sámpúló} \tbPS{adv.~} \tbGE{several times} \tbXV{O wa la kyɛ sampulo bowola zaa kyɛ ba nyɛ fo.} \tbXE{S/he came here several times but did not see you.} 
\tbLX{sanaŋe} \tbHM{} \tbPH{sànáŋ} \tbPS{n.~} \tbGE{a type of large, black scorpion} \tbOP{sananne} \tbTP{sanannɛɛ} 
\tbLX{sandaana} \tbHM{} \tbPH{sándááná} \tbPS{n.~} \tbGE{creditor, one who is owed money} \tbOP{sandaama} \tbTP{sandaamɛɛ} 
\tbLX{sandire} \tbHM{} \tbPH{sándíré} \tbPS{n.~} \tbGE{debtor, one who owes money} \tbOP{sandiribe} \tbTP{sandiribee} 
\tbLX{sane} \tbHM{} \tbPH{sánɪ̀} \tbPS{n.~} \tbGE{debt} \tbOP{sama} \tbTP{sannɛɛ} 
\tbLX{sangye᷉a᷈a᷈} \tbHM{} \tbPH{sǎŋdzɪ̀᷉á᷈á᷈} \tbPS{n.~} \tbSN \tbGE{a type of long water beetle} \tbSN \tbGE{crab} \tbOP{saŋgyeɛne} \tbTP{saŋgyeɛnɛɛ} 
\tbLX{saniŋe} \tbHM{} \tbPH{sàníŋè} \tbPS{n.~} \tbGE{west} \tbVA{luou}
\tbLX{sanlɛre} \tbHM{} \tbPH{sànlɛ̀rɪ̀} \tbPS{n.~} \tbGE{straw put up together to form a mat on which harvested ears of guinea corn is kept to dry} \tbOP{sanlɛrre} \tbTP{sanlɛrɛɛ} \tbVA{sɛnlɛre}
\tbLX{sanna} \tbHM{} \tbPH{sànnà} \tbPS{v.~} \tbGE{to caution, to warn} \tbXV{Fo naŋ daare to a dɔɔ ŋa lɛ na aŋ sanna o yiri gaabo.} \tbXE{The way you insulted this man, you's better be warned not to go to his house.} \tbPD{sannɛɛ, sanna, sanna, sanema, sannaa, - } 
\tbLX{sanne} \tbHM{} \tbPH{sánnɪ̀} \tbPS{v.~} \tbGE{to attempt} \tbPD{sannɛɛ, sanna, sanna, sanema, sannaa, - } 
\tbLX{sansalaŋ} \tbHM{} \tbPH{sànsàlàŋ} \tbPS{adv.~} \tbGE{disorderly} \tbVA{sansaraŋ}
\tbLX{sansaŋgoɔ} \tbHM{} \tbPH{sá nsa ́ŋgʊ́ɔ̀} \tbPS{n.~} \tbGE{a type of thorn tree} \tbOP{sansaŋgoore} \tbTP{sansaŋgoorɛɛ} \tbVA{saleŋgoɔ}
\tbLX{sansaraŋ} \tbHM{} \tbPH{sánsáráŋ} \tbPS{adv.~} \tbGE{making moves to leave all of a sudden (without prior notice)} \tbVA{sansalaŋ}
\tbLX{san-yɔɔ} \tbHM{} \tbPH{sán-yɔ́ɔ́} \tbPS{n.~} \tbGE{salary} \tbOP{san-yɔɔre} \tbTP{san-yɔɔrɛɛ} 
\tbLX{sanyegeraa} \tbHM{} \tbPH{sànyɪ́gɪ́ráá} \tbPS{n.~} \tbGE{lightening (flashes of light in a storm)} \tbOP{sanyegere} \tbTP{sanyegerɛɛ} 
\tbLX{saŋa} \tbHM{} \tbPH{sáŋɪ̀} \tbPS{n.~} \tbGE{time, season} \tbTP{sannɛɛ} \tbVA{saŋe}
\tbLX{saŋe} \tbHM{} \tbPH{sáŋɪ̀} \tbPS{v.~} \tbGE{to heal, to cure} \tbXV{Te᷉e᷉ ama la daare soŋ saŋ a baala.} \tbXE{This medicine helped in curing the patient.} \tbPD{saŋɛɛ, sanna/saŋena, sanena, sanema, sansaŋena, saŋenaa } 
\tbLX{saŋgara} \tbHM{} \tbPH{sàŋgàrá} \tbPS{n.~} \tbGE{a type of worm found in kola nuts} 
\tbLX{saŋgorema} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{foot yaws} 
\tbLX{saŋkora} \tbHM{} \tbPH{sáŋkʊ́rá} \tbPS{n.~} \tbGE{old indebtedness} \tbXV{Ka fooŋ pɛŋe libie ba yɔɔ baare kyɛ la boɔrɔ a mine see ka fo yɔɔ a fo saŋkora na sɛre.} \tbXE{If you borrow money and have not paid back you will not get a new loan unless you pay up your old debt.} 
\tbLX{saŋkpalema} \tbHM{} \tbPH{sáŋkpálɪ́mà} \tbPS{n.~} \tbGE{a kind of large antelope} \tbOP{saŋkpalle} 
\tbLX{saŋkpana} \tbHM{} \tbPH{sǎŋkpáná} \tbPS{n.~} \tbGE{scabies} \tbTP{saŋkpanɛɛ} 
\tbLX{saŋmene} \tbHM{} \tbPH{sáŋmɪ́nɪ̀} \tbPS{n.~} \tbGE{god of rain} \tbOP{saŋmemɛ} \tbTP{saŋmennɛɛ} 
\tbLX{sapare} \tbHM{} \tbPH{sàpárɪ́} \tbPS{n.~} \tbGE{east} \tbOP{saparre} \tbTP{saparɛɛ} 
\tbLX{sapelaa} \tbHM{} \tbPH{sàpɪ̀làá} \tbPS{n.~} \tbSN \tbGE{rain from cumulus clouds} \tbSN \tbGE{a type of rain god (white)} \tbOP{sapeɛle} \tbTP{sapeɛlɛɛ} 
\tbLX{sapigi} \tbHM{} \tbPH{sápígí} \tbPS{n.~} \tbGE{large hat} \tbOP{sapige} \tbTP{sapgree} 
\tbLX{sapi᷈i} \tbHM{} \tbPH{sàpí᷉ì᷉} \tbPS{n.~} \tbGE{bolt of lightning} \tbOP{sapiime} \tbTP{sapiinee} 
\tbLX{sapooraa} \tbHM{} \tbPH{sàpʊ́ʊ́ráá} \tbPS{n.~} \tbGE{rain shower} \tbOP{sapoore} \tbTP{sapoorɛɛ} 
\tbLX{sapɔ} \tbHM{} \tbPH{sàpɔ́} \tbPS{n.~} \tbGE{sponge} \tbOP{sapɔre} \tbTP{sapɔrɛɛ} 
\tbLX{saraa} \tbHM{} \tbPH{sáràà} \tbPS{adj.~} \tbGE{young and female (human and nonhuman)} \tbOP{sarre} \tbTP{sarrɛɛ} 
\tbLX{sare} \tbHM{} \tbPH{sàrɪ̀} \tbPS{v.~} \tbGE{to slip} \tbPD{sarɛɛ, sara, sara, sareba, saraa, - } 
\tbLX{sare} \tbHM{} \tbPH{sárɪ̀} \tbPS{n.~} \tbGE{water trough (container from which animals drink)} \tbOP{saɛ} \tbTP{sarɛɛ} \tbVA{sɛre}
\tbLX{sareka} \tbHM{} \tbPH{sàrɪ́kà} \tbPS{n.~} \tbGE{a ritual in which one provides for the needy for atonement} \tbOP{sarekare} \tbTP{sakarɛɛ.} 
\tbLX{sarre} \tbHM{} \tbPH{sárrɪ̀} \tbPS{v.~} \tbGE{to develop into a young woman} \tbPD{sarrɛɛ, sarra, sarra, sareba, sarraa, - } 
\tbLX{saseɛ} \tbHM{} \tbPH{sàsɪ̀ɛ́} \tbPS{n.~} \tbSN \tbGE{air, wind} \tbSN \tbGE{storm} \tbOP{saseere} \tbTP{saseerɛɛ} 
\tbLX{sase-ɔɔre} \tbHM{} \tbPH{sásɪ̀-ɔ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{harmattan} \tbTP{sase-ɔɔrɛɛ} 
\tbLX{sasepilimbaŋe} \tbHM{} \tbPH{sàsɪ̀pílímbɔ́ŋ} \tbPS{n.~} \tbGE{whirlwind} \tbOP{sasepilimbanne} \tbTP{sasepilimbannɛɛ} \tbVA{sasepilimbɔŋ}
\tbLX{sasere} \tbHM{} \tbPH{sàsɪ́rɪ̀} \tbPS{n.~} \tbGE{grasscutter, greater cane rat (large rodent that feeds mostly on grass)} \tbSG{saseroo} \tbOP{saseɛ} \tbTP{saserɛɛ} 
\tbLX{sasigoo} \tbHM{} \tbPH{sàsígóó} \tbPS{n.~} \tbGE{hiccough} \tbOP{sasigri} \tbTP{sasigree} 
\tbLX{sasɔgelaa} \tbHM{} \tbPH{sàsɔ́gláá} \tbPS{n.~} \tbSN \tbGE{rain from black clouds} \tbSN \tbGE{a type of fetish or deity connected with rain (black)} \tbOP{sasɔgelɔ/sasɔgele} \tbTP{sasɔgelɛɛ} 
\tbLX{sasusulee} \tbHM{} \tbPH{sàsúsúléé} \tbPS{n.~} \tbGE{a type of tree} \tbOP{sasusulli} \tbTP{sasusullee} 
\tbLX{satannaa} \tbHM{} \tbPH{sàtànnáà} \tbPS{v.~} \tbGE{to thunder} \tbPD{-, -, -, -, -, - } 
\tbLX{savaale} \tbHM{} \tbPH{sàváálɪ́} \tbPS{n.~} \tbGE{rainstorm (at night)} \tbOP{savaala} \tbTP{savaalɛɛ} 
\tbLX{sawarema} \tbHM{} \tbPH{sàwárɪ́má} \tbPS{n.~} \tbGE{white-looking cloud} \tbSG{sawaroo} \tbTP{sawaremɛɛ} \tbVA{sawarɛɛ}
\tbLX{sawarɛɛ} \tbHM{} \tbPH{sàwárɛ́ɛ́} \tbPS{n.~} \tbGE{white clouds (non-rain-bearing)} \tbTP{sawarrɛɛ/sawaremɛɛ} \tbVA{sawarema}
\tbLX{sawɛge} \tbHM{} \tbPH{sáwɛ́gɪ́} \tbPS{n.~} \tbGE{long cloud sprawling across the sky} \tbOP{sawɛgere} \tbTP{sawɛgerɛɛ} 
\tbLX{sawɔlɔ} \tbHM{} \tbPH{sáwɔ́lɔ́} \tbPS{n.~} \tbGE{fried dough made from flour e.g. bean, corn, millet} \tbTP{sawɔlɛɛ} 
\tbLX{sazeɛ} \tbHM{} \tbPH{sázɪ̀ɛ̀} \tbPS{n.~} \tbGE{a type of soft broom} \tbOP{sazeere} \tbTP{sazeerɛɛ} 
\tbLX{sazɔŋɔ} \tbHM{} \tbPH{sàzɔ́ŋɔ̀} \tbPS{n.~} \tbGE{snow} \tbTP{sazɔnne} 
\tbLX{sazɔɔ} \tbHM{} \tbPH{sàzɔ̀ɔ́} \tbPS{n.~} \tbGE{first rain which begins the farming season} \tbOP{sazɔɔre} \tbTP{sazɔɔrɛɛ} 
\tbLX{sazu} \tbHM{} \tbPH{sàzú} \tbPS{n.~} \tbGE{sky} \tbOP{sazuri} \tbTP{sazuree} 
\tbLX{sazu} \tbHM{} \tbPH{sàzú} \tbPS{adv.~} \tbGE{up} 
\tbLX{sazupi᷉i᷉} \tbHM{} \tbPH{sázúpí᷉ì᷈} \tbPS{n.~} \tbGE{danger lurking above} 
\tbLX{se} \tbHM{} \tbPH{sɪ̀} \tbPS{v.~} \tbGE{to skin (slaughtered animal)} \tbPD{seɛɛ, seerɛ, seerɛ, seerebɛ, seeraa, - } 
\tbLX{sebaa} \tbHM{} \tbPH{sɪ́báá} \tbPS{n.~} \tbGE{a type of wild animal} \tbOP{sebre} \tbTP{sebrɛɛ} 
\tbLX{sebiri} \tbHM{} \tbPH{sɪ́bírì} \tbPS{n.~} \tbGE{bee} \tbOP{sebie} \tbTP{sebiiree} 
\tbLX{sebitetebaa} \tbHM{} \tbPH{sɪ́bítɪ́tɪ́báá} \tbPS{n.~} \tbGE{bee stinger} \tbOP{sebiteterre} \tbTP{sebiteterrɛɛ} \tbVA{tetebaa}
\tbLX{sebo} \tbHM{} \tbPH{sɪ̀bʊ̀} \tbPS{v.~} \tbSN \tbGE{to affect} \tbXV{Ka fooŋ zagere toma gaabo maa naane ka fo na sebo.} \tbXE{If you refuse to go to work it will not affect me.} \tbSN \tbGE{to refuse to give} \tbXV{N sɔre la o kuuri ka o sebo ma.} \tbXE{I begged for his hoe, but he refused me.} \tbPD{sebɛɛ, sebrɛ, sebrɛ, sebrebɛ, sebraa, - } 
\tbLX{sebogi} \tbHM{} \tbPH{sɪ́bògí} \tbPS{n.~} \tbGE{beehive,  hole (usually in a tree) where bees have made honey} \tbOP{sebogiri} \tbTP{sebogiree} 
\tbLX{sedanegere} \tbHM{} \tbPH{sɪ́dànɪ́gɪ́rɪ́} \tbPS{n.~} \tbGE{brown, sour-tasting part of honey} \tbOP{sedanegere} \tbTP{sedanegerɛɛ} \tbVA{danegɛ}
\tbLX{sedoge} \tbHM{} \tbPH{sɪ́dʊ́gɪ̀} \tbPS{n.~} \tbGE{swarm trap pot for bees} \tbOP{sedogre} \tbTP{sedogrɛɛ} 
\tbLX{see} \tbHM{} \tbPH{séé} \tbPS{conj.~} \tbGE{unless e.g having no option} \tbXV{See ka fo daŋe iruu ka lɛ naane fo koŋ nyɛ a baala.} \tbXE{Unless you leave early otherwise you will not see the patient.} 
\tbLX{seebo} \tbHM{} \tbPH{sɪ́ɪ́bʊ́} \tbPS{n.~} \tbSN \tbGE{honey} \tbSN \tbGE{beehive} \tbOP{seere} \tbTP{seerɛɛ} \tbVA{seeo}
\tbLX{seene} \tbHM{} \tbPH{sɪ́ɪ́nɪ̀} \tbPS{v.~} \tbGE{to touch} \tbPD{seenɛɛ, seenɛ, seenɛ, seenemɛ, seenaa, - } \tbVA{seere}
\tbLX{seeraa} \tbHM{} \tbPH{sɪ́ɪ́ráá} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{seere} \tbTP{seerɛɛ} \tbVA{seetere}
\tbLX{seere} \tbHM{} \tbPH{sɪ́ɪ́rɪ̀} \tbPS{n.~} \tbGE{honey} \tbTP{seerɛɛ} 
\tbLX{seetere} \tbHM{} \tbPH{sɪ́ɪ́tɪ́rɪ́} \tbPS{n.~} \tbGE{a type of tree with hard word and is also medicinal} \tbSG{seeterɛɛ} \tbTP{seeteere} \tbVA{seeraa}
\tbLX{seɛ} \tbHM{} \tbPH{sɪ́ɛ́} \tbPS{n.~} \tbGE{waist} \tbOP{seere} \tbTP{seerɛɛ} 
\tbLX{seɛ} \tbHM{} \tbPH{sɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to wear (on the waist)} \tbSN \tbGE{to dance} \tbSN \tbGE{to be better than, to surpass} \tbXV{A fanfane ŋaŋ seɛ a ŋa.} \tbXE{This soap is better than this.} \tbXV{Ka fooŋ mare a bie anaŋ seɛ fo naŋ na pɛgele.} \tbXE{To carry the baby on your back is better than carrying by your side.} \tbPD{-, seɛrɛ, -, -, -, - } 
\tbLX{seɛ} \tbHM{} \tbPH{sɪ́ɛ̀} \tbPS{n.~} \tbGE{soul, spirit (of a living person)} \tbOP{seere} 
\tbLX{seɛ} \tbHM{} \tbPH{sɪ̀ɛ́} \tbPS{n.~} \tbSN \tbGE{joke, teasing} \tbXV{A pɔge taa la poɔ ka o taaba erɛ o seɛ.} \tbXE{The woman is pregnant and her peers are teasing her.} \tbSN \tbGE{fresh air, wind} 
\tbLX{seɛ} \tbHM{} \tbPH{sɪ́ɛ́} \tbPS{adj.~} \tbGE{by the side (wall)} \tbXV{A boɔ gaŋ la a dakyini seɛ.} \tbXE{The goat is lying at the foot of the wall.} 
\tbLX{se᷉ɛ} \tbHM{} \tbPH{sɪ́᷉ɛ̀᷈} \tbPS{n.~} \tbGE{rainy season} \tbOP{seɛmɛ} \tbTP{seɛnɛɛ} 
\tbLX{seɛbaa} \tbHM{} \tbPH{sɪ́ɛ́báá} \tbPS{adj.~} \tbGE{considerably big} \tbXV{De peɛ naŋ e seɛbaa, ka lɛ naane a kyi koŋ kpɛ.} \tbXE{Take a good-sized basket, or else the millet will be too much for a small one.} \tbOP{seɛbaare} 
\tbLX{seɛbaa} \tbHM{} \tbPH{̀sɪ̀ɛ́báà} \tbPS{n.~} \tbGE{twine from daw-daw fruit} \tbOP{ seɛre} \tbTP{seɛrɛɛ} 
\tbLX{seɛgoɔ} \tbHM{} \tbPH{sɪ̀ɛ́gʊ́ɔ́} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{seɛgoore} 
\tbLX{seɛ kpelle} \tbHM{} \tbPH{sɪ̀ɛ̀ kpɪ́llɪ́} \tbPS{v.~} \tbGE{to get someone in trouble} \tbPD{seɛɛ kpelle, seɛrɛ kpelle, -, -, -, - } 
\tbLX{se᷉ɛkpɛpaalaa} \tbHM{} \tbPH{sɪ̀᷉ɛ̀᷉kpɛ̀pááláá} \tbPS{n.~} \tbGE{early farming season} \tbOP{se᷉ɛkpɛpaala} 
\tbLX{seɛlaa} \tbHM{} \tbPH{sɪ̀ɛ̀làá} \tbPS{n.~} \tbSN \tbGE{clay eaten by cattle} \tbSN \tbGE{pit where cattle eat this clay} \tbOP{seɛle} \tbTP{seɛlɛɛ} 
\tbLX{seɛlee} \tbHM{} \tbPH{sɪ́ɛ́lɪ́ɪ́} \tbPS{n.~} \tbGE{mystery} 
\tbLX{seɛlee} \tbHM{} \tbPH{sɪ́ɛ́lɪ́ɪ́} \tbPS{adj.~} \tbGE{mysterious, strange} 
\tbLX{seɛne} \tbHM{} \tbPH{sɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{porcupine} \tbOP{seɛmɛ} \tbTP{seɛnɛɛ} 
\tbLX{seɛne} \tbHM{} \tbPH{sɪ̀ɛ̀nɪ̀} \tbPS{v.~} \tbSN \tbGE{to grab firmly with fingers around the object e.g. neck} \tbSN \tbGE{to cram e.g a bag with clothes} \tbPD{seɛnɛɛ, seɛnɛ, seɛnɛ, seɛnemɛ, seɛnaa, - } 
\tbLX{seɛnoɔre} \tbHM{} \tbPH{sɪ̀ɛ́nʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{the beginning of the farming season} 
\tbLX{sefaa} \tbHM{} \tbPH{sɪ́fáá} \tbPS{n.~} \tbGE{demon, evil spirit} \tbOP{sefaare} \tbTP{sefaarɛɛ} 
\tbLX{segammo} \tbHM{} \tbPH{sɪ́gàmmʊ́} \tbPS{n.~} \tbGE{honeycomb} \tbOP{segama} \tbTP{segamɛɛ} 
\tbLX{segampelaa} \tbHM{} \tbPH{sɪ́gǎmpɪ̀làá} \tbPS{n.~} \tbGE{white honeycomb (usually contains the pure honey)} \tbOP{segampeɛle} \tbTP{segampeɛlɛɛ} 
\tbLX{seganfuu} \tbHM{} \tbPH{sɪ́gànfúù} \tbPS{n.~} \tbGE{honeycomb without honey} \tbOP{seganfuuri} \tbTP{seganfuuree} 
\tbLX{sege} \tbHM{} \tbPH{sɪ́gɪ́} \tbPS{n.~} \tbGE{farm hut; camp} \tbOP{segere} \tbTP{segerɛɛ} 
\tbLX{sege} \tbHM{} \tbPH{sɪ́gɪ́} \tbPS{v.~} \tbSN \tbGE{to resemble;} \tbSN \tbGE{to name a child after someone (usually an anscestor)} \tbPD{segɛɛ, segerɛ, segerɛ, segerebɛ, segeraa, - } 
\tbLX{segeraa} \tbHM{} \tbPH{sɪ́gɪ́ráá} \tbPS{n.~} \tbGE{ancestral name given to a baby} \tbOP{segere} \tbTP{segerɛɛ} 
\tbLX{segesege} \tbHM{} \tbPH{sɪ̀gɪ̀sɪ́gɪ́} \tbPS{v.~} \tbGE{to attempt to move but being constrained by one thing or the other} \tbPD{segesegɛɛ, segesegerɛ, segesegerɛ, segesegerebɛ, segesegeraa, - } \tbVA{segesege}
\tbLX{segilekele} \tbHM{} \tbPH{ségílékélé} \tbPS{adv.~} \tbGE{standing upright, tall, and solitary e.g. tower} 
\tbLX{sekãa} \tbHM{} \tbPH{sɪ́káà} \tbPS{n.~} \tbGE{pure honey} \tbSG{seka᷉ao} \tbOP{sekaanɛɛ} \tbVA{sekãaboli}
\tbLX{sekãaboli} \tbHM{} \tbPH{sɪ́ká᷉á᷉bólí} \tbPS{n.~} \tbGE{pure honey} \tbOP{seka᷉a᷉bolo} \tbTP{seka᷉a᷉bolee} \tbVA{sekãa}
\tbLX{sekalmanyuo} \tbHM{} \tbPH{sɪ́kálmáɲùód} \tbPS{n.~} \tbGE{swam of bees (usually clustering on a tree branch)} \tbOP{sekalmanyuuri} \tbTP{sekalmanyuuree} 
\tbLX{sekpɔgere} \tbHM{} \tbPH{sɪ́kpɔ́gɪ́rɪ̀} \tbPS{n.~} \tbGE{proverb, riddle} \tbSG{sekpɔgeroo} \tbOP{sekpɔgɔ} \tbTP{sekpɔgerɛɛ} 
\tbLX{selaa} \tbHM{} \tbPH{sɪ̀làà} \tbPS{n.~} \tbGE{hawk} \tbOP{selle} \tbTP{sellɛɛ} \tbVA{gbaara, gbaare, gbaarɛɛ}
\tbLX{sele} \tbHM{} \tbPH{sɪ́lɪ̀} \tbPS{v.~} \tbGE{to decant} \tbPD{selɛɛ, sellɛ, sellɛ, selbɛ, sellaa, - } 
\tbLX{selee} \tbHM{} \tbPH{séléé} \tbPS{n.~} \tbGE{pottery shard} \tbOP{sebilii} \tbVA{semekelee/sambekelee}
\tbLX{seleŋkaa} \tbHM{} \tbPH{sɪ́lɪ́ŋkáá} \tbPS{n.~} \tbGE{a type of small wasp} \tbOP{seleŋkaare} \tbTP{seleŋkaarɛɛ} \tbVA{selɛŋkaa, seleŋkpaa}
\tbLX{selɛŋkaa} \tbHM{} \tbPH{sɪ́lɛ́ŋkáà} \tbPS{n.~} \tbGE{a small brown wasp with a painful sting} \tbOP{selɛŋkaare} \tbTP{selɛŋkaarɛɛ} \tbVA{selɛŋkpaa, seleŋkaa}
\tbLX{selɛŋkpaa} \tbHM{} \tbPH{sɪ́lɛ́ŋkpáá} \tbPS{n.~} \tbGE{a small brown wasp with a painful sting} \tbOP{selɛŋkpaare} \tbTP{selɛŋkpaarɛɛ} \tbVA{selɛŋkaa, seleŋkaa}
\tbLX{selgyuuni} \tbHM{} \tbPH{sɪ̀ldzùùnì} \tbPS{n.~} \tbGE{a type of eagle} \tbOP{selgyuumo} \tbTP{selgyuunee} 
\tbLX{selle} \tbHM{} \tbPH{sɪ́llɪ̀} \tbPS{v.~} \tbGE{to tell, to recount (story)} \tbPD{sellɛɛ, sellɛ, sellɛ, sellebɛ, sellaa, - } 
\tbLX{selle} \tbHM{} \tbPH{sɪ́llɪ̀} \tbPS{v.~} \tbGE{to reduce in size (body)} \tbPD{sellɛɛ, sellɛ, sellɛ, sellebɛ, sellaa, - } 
\tbLX{semare} \tbHM{} \tbPH{sɪ́màrɪ́} \tbPS{n.~} \tbGE{a type of fetish to which there is a secret cult that has the power to identify evil people like wizards and witches} \tbOP{semarɛɛ} 
\tbLX{semekelee} \tbHM{} \tbPH{sèmèkèlèè} \tbPS{n.~} \tbGE{broken piece from a clay molded object e.g a pot} \tbVA{sembakelee}
\tbLX{sengye} \tbHM{} \tbPH{sándzé} \tbPS{n.~} \tbGE{mosquito net} \tbOP{sengyeri} \tbTP{sengyeree} \tbVA{sangye}
\tbLX{senlɛre} \tbHM{} \tbPH{sɪ́nlɛ́rɪ́} \tbPS{n.~} \tbGE{sweet drink} \tbOP{senlerre} \tbTP{senlɛrɛɛ} 
\tbLX{senlogere} \tbHM{} \tbPH{sìnlʊ̀gɪ̀rì} \tbPS{n.~} \tbGE{bile} \tbOP{senlogera} 
\tbLX{senselloŋ} \tbHM{} \tbPH{sɪ́nsɪ́llʊ́ŋ} \tbPS{n.~} \tbGE{story, tale} \tbOP{sensellɛ} 
\tbLX{sensɛlaa} \tbHM{} \tbPH{sɪ́nsɛ̀láá} \tbPS{n.~} \tbGE{earthenware for frying cakes} \tbOP{sensɛlaare} \tbTP{sensɛlaarɛɛ} 
\tbLX{sensɛlaabogiri} \tbHM{} \tbPH{sɪ́nsɛ̀lààbògrí} \tbPS{n.~} \tbGE{frying earthenware with hollow depressions} \tbTP{sensɛlaabogiree} 
\tbLX{sensɛre} \tbHM{} \tbPH{sɪ́nsɛ́rɪ́} \tbPS{n.~} \tbGE{bean cake} \tbOP{sensɛɛ} \tbTP{sensɛrɛɛ} 
\tbLX{sensogelensogɔ} \tbHM{} \tbPH{sɪ̀nsʊ́glɪ́nsʊ̀gɔ́} \tbPS{n.~} \tbGE{middle} \tbOP{sensogelensogere} \tbTP{sensogelensogerɛɛ} \tbVA{sensogelensoga}
\tbLX{sensogɔ} \tbHM{} \tbPH{sɪ̀sʊ́gɔ́} \tbPS{adj.~} \tbGE{middle} \tbXV{A yiri be la a teŋɛ sensogɔ.} \tbXE{The house is in the middle of the town.} \tbOP{sensogere} \tbTP{sensogerɛɛ} \tbVA{sensoglensogɔ, sensoga, soga}
\tbLX{sensogɔ} \tbHM{} \tbPH{sɪ́nsʊ́gɔ̀} \tbPS{n.~} \tbSN \tbGE{home drain e.g one leading out of the bathroom} \tbSN \tbGE{a track fixed to drain rainwater from the top of a roof} \tbOP{sensogere} \tbTP{sensogerɛɛ} \tbVA{sensoga}
\tbLX{sensɔge} \tbHM{} \tbPH{sɪ́nsɔ́gɪ́} \tbPS{n.~} \tbGE{rattle used by babysitters to accompany the singing of a lullaby} \tbOP{sɔnsɔgre} \tbTP{sensɔgrɛɛ} 
\tbLX{sensɔɔre} \tbHM{} \tbPH{sɪ̀nsɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{tigernut} \tbSG{sensɔɔroo} \tbOP{sensɔbɔ} \tbTP{sensɔɔrɛɛ} 
\tbLX{sensɔraa} \tbHM{} \tbPH{sɪ̀nsɔ̀ráà} \tbPS{n.~} \tbGE{kitchen sponge} \tbOP{sensɔrre} \tbTP{sensɔrɛɛ/sensɔrrɛɛ} 
\tbLX{sentaa} \tbHM{} \tbPH{sɪ́ntáá} \tbPS{n.~} \tbGE{equality} 
\tbLX{sentaaloŋ} \tbHM{} \tbPH{sɪ́ntáá} \tbPS{n.~} \tbGE{the state of being equal} \tbGE{the state of being equal} \tbVA{sentaa}\tbVA{sentaa}\tbVA{sentaaloŋ}
\tbLX{senuulee} \tbHM{} \tbPH{sɪ́núúléé} \tbPS{n.~} \tbGE{honey bird} \tbOP{senuuli} 
\tbLX{senyoɔle} \tbHM{} \tbPH{sɪ́nyʊ́ɔ́lɪ́} \tbPS{n.~} \tbGE{a person or any creature that is fond of sweet things} \tbOP{senyoɔlɔ/senyoɔla} \tbTP{senyoɔlɛɛ} 
\tbLX{seŋe} \tbHM{} \tbPH{sɪ́ŋ} \tbPS{v.~} \tbSN \tbGE{to suffice, to be enough} \tbXV{A bondirii e la yaga; a na seŋe la a saama.} \tbXE{The food is plenty; it will do for the guests.} \tbSN \tbGE{to be equal to} \tbXV{A toma ŋa seŋe la a anaŋ fo naŋ daare toŋe.} \tbXE{This work is equal to what you did.} \tbPD{seŋɛɛ, sennɛ, sennɛ, -, sennaa, - } 
\tbLX{seŋe} \tbHM{} \tbPH{sɪ̂ŋ} \tbPS{adv.~} \tbGE{like} \tbXV{O e seŋ ŋa o ba nyɛ ma na.} \tbXE{S/he behaved like s/he did not see me.} 
\tbLX{seŋe} \tbHM{} \tbPH{sɪ̂ŋ} \tbPS{v.~} \tbSN \tbGE{to wake up (someone)} \tbSN \tbGE{to revive} \tbPD{seŋɛɛ, sennɛ, sennɛ, senemɛ, sennaa, senemɛ } 
\tbLX{seŋɛ} \tbHM{} \tbPH{sɪ́ŋ} \tbPS{n.~} \tbGE{large pot used to store local beer during preparation or to store water} \tbOP{senne} \tbTP{sennɛɛ} 
\tbLX{seŋgelɛnte} \tbHM{} \tbPH{sɪ́ŋgɪ́lɛ́ntɪ́} \tbPS{n.~} \tbGE{singlet} \tbOP{seŋgelɛntere} \tbTP{seŋgelɛnterɛɛ} 
\tbLX{seŋkaafa} \tbHM{} \tbPH{sɪ̀ŋkááfá} \tbPS{n.~} \tbGE{rice (Hausa)} \tbOP{seŋka᷉afare} \tbTP{sɪŋka᷈a᷈farɛɛ} \tbVA{mui}
\tbLX{seŋka᷈a᷈} \tbHM{} \tbPH{sɪ̀ŋkáá} \tbPS{n.~} \tbGE{groundnut, peanut} \tbSG{seŋkaanoo} \tbTP{seŋkaanɛɛ} 
\tbLX{seŋka᷉a᷉kpɔlɔ} \tbHM{} \tbPH{sɪ́ŋká᷉á᷉kpɔ́lɔ́} \tbPS{n.~} \tbGE{groundnuts (a variety), has big nuts} 
\tbLX{seŋka᷉awerɛ} \tbHM{} \tbPH{sɪ́ŋká᷉á᷉wɪ́rɛ́} \tbPS{n.~} \tbGE{groundnut (a variety), has small nuts} 
\tbLX{seŋkpegere} \tbHM{} \tbPH{sɪ́ŋkpɪ́gɪ́rɪ́} \tbPS{n.~} \tbGE{hip} \tbOP{seŋkpegɛ} \tbTP{seŋkpegerɛɛ} 
\tbLX{Sepaanase} \tbHM{} \tbPH{sɪ́páánásɪ̀} \tbPS{n.~} \tbGE{Spanish} \tbOP{Sepaanasedeme} 
\tbLX{sepelaa} \tbHM{} \tbPH{sɪ́pɪ́làá} \tbPS{n.~} \tbGE{a soul that has left a person but can be returned again} \tbOP{sepeɛle} 
\tbLX{sepɔpɔrɛɛ} \tbHM{} \tbPH{sɪ́pɔ́pɔ́rɛ́ɛ́} \tbPS{n.~} \tbGE{honeycomb where the larvae are found} \tbSG{sepɔpɔroo} \tbVA{pɔpɔrɛɛ}
\tbLX{sere} \tbHM{} \tbPH{sɪ́rɪ̀} \tbPS{v.~} \tbGE{to dry up, to evaporate} \tbPD{serɛɛ, serɛ, serɛ, serebɛ, seraa, - } 
\tbLX{sere} \tbHM{} \tbPH{sɪ́rɪ̀} \tbPS{v.~} \tbGE{to wash, to scrub (bathing)} \tbPD{serɛɛ, serɛ, serɛ, serebɛ, seraa, - } 
\tbLX{serekuli} \tbHM{} \tbPH{sɪ̀rɪ̀kúlí} \tbPS{n.~} \tbGE{the act of marrying (on the part of woman)} 
\tbLX{serekulo} \tbHM{} \tbPH{sɪ̀rɪ̀kúló} \tbPS{n.~} \tbGE{bride} \tbOP{serkulbo} 
\tbLX{seretara} \tbHM{} \tbPH{sɪ̀rɪ̀tárá} \tbPS{n.~} \tbGE{married woman} 
\tbLX{serete} \tbHM{} \tbPH{sɪ́rɪ́tɪ́} \tbPS{adv.~} \tbGE{very quiet} \tbXV{A bibiiri gbiree la ka a zie e serete ŋaa.} \tbXE{The children are asleep and the place is very quiet.} 
\tbLX{serɛ} \tbHM{} \tbPH{sɪ́rɛ́} \tbPS{n.~} \tbGE{husband} \tbOP{serebɛɛ} \tbTP{serrɛɛ/serɛɛ} 
\tbLX{serɛ} \tbHM{} \tbPH{sɪ̀rɛ̀} \tbPS{n.~} \tbGE{truth} \tbOP{serre} \tbTP{serɛɛ} 
\tbLX{sesɔgelaa} \tbHM{} \tbPH{sɪ́sɔ́gláà} \tbPS{n.~} \tbGE{a soul that has left a person for good} \tbOP{sesɔgelɔ.} 
\tbLX{setaana} \tbHM{} \tbPH{sɪ̀táánà} \tbPS{n.~} \tbGE{devil} \tbOP{setaamine} \tbTP{setaanɛɛ} \tbVA{sitaana}
\tbLX{setampe} \tbHM{} \tbPH{sɪ̀támpɪ́} \tbPS{n.~} \tbGE{stamp} \tbOP{setampere} \tbTP{setamperɛɛ} 
\tbLX{setetebaa} \tbHM{} \tbPH{sɪ́tɪ́tɪ́báá} \tbPS{n.~} \tbGE{the stinger of a bee} \tbSG{setetebaarɛɛ} \tbTP{setetebaare} 
\tbLX{sɛ} \tbHM{} \tbPH{sɛ̀} \tbPS{adv.~} \tbGE{already} \tbXV{O gaaɛ la sɛ.} \tbXE{S/he has gone already.} 
\tbLX{sɛ} \tbHM{} \tbPH{sɛ́} \tbPS{v.~} \tbGE{to roast} \tbPD{sɛɛ, seɛrɛ, seɛrɛ, seɛreba, seɛraa, seɛre } 
\tbLX{sɛ} \tbHM{} \tbPH{sɛ́} \tbPS{v.~} \tbGE{to sue} \tbPD{sɛɛ, seɛrɛ, seɛrɛ, seɛreba, seɛraa, seɛre } 
\tbLX{sɛ} \tbHM{} \tbPH{sɛ̀} \tbPS{v.~} \tbGE{to sew} \tbPD{-, seɛrɛ, -, -, -, - } 
\tbLX{sɛbɛlemɛkyɔɔ} \tbHM{} \tbPH{sɛ̀bɛ̌lɪ́mɛ̀ʧɔ́ɔ̀} \tbPS{n.~} \tbGE{earthworm} \tbVA{sabɛlemɛkyɔɔ}
\tbLX{sɛge} \tbHM{} \tbPH{sɛ́gɪ́} \tbPS{v.~} \tbGE{to write} \tbPD{sɛgɛɛ, sɛgrɛ, sɛgrɛ, sɛgrebɛ, sɛgraa, - } 
\tbLX{sɛgebinyaŋaa} \tbHM{} \tbPH{sɛ́gɪ́bíɲáŋáá} \tbPS{n.~} \tbGE{vowel} \tbOP{sɛgebinyaane} 
\tbLX{sɛgebiri} \tbHM{} \tbPH{sɛ́gɪ́bírì} \tbPS{n.~} \tbGE{amulet, charm, fetish} \tbOP{sɛgebie} \tbTP{sɛgebiiree} 
\tbLX{sɛgebiri} \tbHM{} \tbPH{sɛ́gɪ́bírì} \tbPS{n.~} \tbGE{letter (consonant or vowel)} \tbOP{sɛgebie} \tbTP{sɛgebiiree} 
\tbLX{sɛgelɛgɛ} \tbHM{} \tbPH{sɛ́gɪ́lɛ́gɛ̀} \tbPS{adv.~} \tbGE{unable to move e.g. lazy dog} \tbVA{hɛglɛgɛ}
\tbLX{sɛgeraa} \tbHM{} \tbPH{sɛ̀gɪ̀ráà} \tbPS{n.~} \tbGE{a piece of flat object fixed alongside a hoe blade into the handle to make the blade firm} \tbOP{sɛgere} \tbTP{sɛgerɛɛ} \tbVA{shortened form of bonsɛgeraa}
\tbLX{sɛgere} \tbHM{} \tbPH{sɛ́gɪ́rɪ́} \tbPS{v.~} \tbGE{to arrange} \tbPD{sɛgerɛɛ, sɛgerɛ, sɛgrɛ, sɛgrebɛ, sɛgeraa, sɛgerebɛ } 
\tbLX{sɛgere} \tbHM{} \tbPH{sɛ̀gɪ̀rɪ̀} \tbPS{v.~} \tbGE{to stuff, to plug (holes, spaces)} \tbPD{sɛgerɛɛ, sɛgerɛ, sɛgerɛ, sɛgerebɛ, sɛgeraa, - } 
\tbLX{sɛgesɛge} \tbHM{} \tbPH{sɛ́gsɛ́g} \tbPS{adv.~} \tbGE{making moves; before one is aware} \tbXV{Saa koe la ka a koɔreba yeli ka ba erɛ sɛgesɛge ka ona la miire.} \tbXE{Clouds had gathered and before the farmers could make any move it started raining.} 
\tbLX{sɛle} \tbHM{} \tbPH{sɛ̀lɪ̀} \tbPS{v.~} \tbGE{to transplant} \tbPD{sɛlɛɛ, sɛllɛ, sɛllɛ, sɛlebɛ, sɛllaa, - } 
\tbLX{sɛle} \tbHM{} \tbPH{sɛ̀lɪ̀} \tbPS{v.~} \tbGE{to sit, perch (derogatory use)} \tbXV{A dɔɔ tanne la a bie,”N toŋe fo la ka fo zagere kyɛ wa sɛle n niŋeŋ boɔrɔ boŋ?”} \tbXE{The man shouted at the boy, “I sent you and you refused, what do you want  perched in front of me?”.} \tbPD{sɛlɛɛ, sɛllɛ, sɛllɛ, sɛlebɛ, sɛllaa, - } 
\tbLX{sɛle} \tbHM{} \tbPH{sɛ́lɪ̀} \tbPS{n.~} \tbGE{the act of winning a game of bie outright} \tbOP{sɛlɛ} \tbTP{sɛlɛɛ} 
\tbLX{sɛma} \tbHM{} \tbPH{sɛ́mà} \tbPS{v.~} \tbGE{to sue in court} \tbPD{sɛmaɛ, seɛrɛmara, sɛmara, sɛmareba, -, - } \tbVA{sɛ, sɛmare}
\tbLX{sɛmare} \tbHM{} \tbPH{sɛ́márɪ̀} \tbPS{v.~} \tbGE{to sue} \tbPD{sɛmarɛɛ, seɛrɛmara, seɛremara, seɛremareba, -, - } \tbVA{sɛ, sɛma}
\tbLX{sɛmpili} \tbHM{} \tbPH{sɛ̌mpílí} \tbPS{n.~} \tbGE{a type of long mat} \tbOP{sɛmpile} \tbTP{sɛmpilee} \tbVA{sɛnwoɔ}
\tbLX{sɛndɔɔ} \tbHM{} \tbPH{sɛ̀ndɔ́ɔ́} \tbPS{n.~} \tbGE{the boyfriend of one's wife} \tbOP{sɛndɔbɔ} \tbTP{sɛndɔɔrɛɛ} 
\tbLX{sɛndɔɔ-dogi-za} \tbHM{} \tbPH{sɛ̌ndɔ́ɔ́-dògì-zá} \tbPS{n.~} \tbGE{small lumps of earth on top of a wall for decoration} 
\tbLX{sɛne} \tbHM{} \tbPH{sɛ́nɪ́} \tbPS{n.~} \tbSN \tbGE{lover (of another sex)} \tbSN \tbGE{fiance} \tbOP{sɛnemɛ} \tbTP{sɛnemɛɛ} 
\tbLX{sɛnlee} \tbHM{} \tbPH{sɛǹléè} \tbPS{n.~} \tbGE{small mat} \tbOP{sɛmbilii} 
\tbLX{sɛnlɛre} \tbHM{} \tbPH{sɛ̀nlɛ̀rɪ̀} \tbPS{n.~} \tbGE{a type mat made by putting straw together on which harvested guinea corn is kept to dry} \tbOP{sɛnlɛrre} \tbTP{sɛnlɛrɛɛ} \tbVA{sanlɛre}
\tbLX{sɛnsɛmbie} \tbHM{} \tbPH{sɛ̀nsɛ̀mbíé} \tbPS{n.~} \tbGE{bastard} \tbOP{sɛnsɛmbiiri} \tbTP{sɛnsɛmbiiree} 
\tbLX{sɛnsɛŋ} \tbHM{} \tbPH{sɛnsɛŋ} \tbPS{n.~} \tbGE{the act of love making (friendship)} 
\tbLX{sɛnwoɔ} \tbHM{} \tbPH{sɛ́nwʊ́ɔ́} \tbPS{n.~} \tbGE{a type of long mat} \tbOP{sɛnwoore} \tbTP{sɛnwoorɛɛ} \tbVA{sɛmpili}
\tbLX{sɛnwoɔ} \tbHM{} \tbPH{sɛ́nwʋ́ɔ́} \tbPS{n.~} \tbGE{a type of long mat} \tbVA{sɛmpili}
\tbLX{sɛŋ} \tbHM{} \tbPH{sɛ́ŋ} \tbPS{adv.~} \tbGE{actually} \tbXV{Fo sɛŋ digi la a bie bare wɛ?} \tbXE{So, you have actually driven the child away?.} 
\tbLX{sɛŋe} \tbHM{} \tbPH{sɛ̀ŋ} \tbPS{v.~} \tbGE{to engage in friendship between people of opposite sexes} \tbPD{sɛŋɛɛ, sɛnɛ, sɛnɛ, sɛnemɛ, sɛnaa, sɛnemɛ } 
\tbLX{sɛŋe} \tbHM{} \tbPH{sɛ̂ŋ} \tbPS{adj.~} \tbSN \tbGE{one's side} \tbXV{A sane yɔɔbo be la fo sɛŋe.} \tbXE{Paying the debt rests with you.} \tbSN \tbGE{towards} \tbXV{A sakuuri are la a teŋɛ sapare sɛŋe.} \tbXE{The school is situated to the east of the town.} \tbOP{sɛnne} 
\tbLX{sɛŋɛ} \tbHM{} \tbPH{sɛ̂ŋɛ́} \tbPS{n.~} \tbGE{mat; bed} \tbOP{sɛnne} \tbTP{sɛnnɛɛ} 
\tbLX{sɛŋkpɛdie} \tbHM{} \tbPH{sɛ̀ŋkpɛ̀dìé} \tbPS{n.~} \tbGE{deep love between lovers; relationship that involves sex} \tbOP{sɛŋkpɛderi} \tbTP{sɛŋkpɛderee} 
\tbLX{sɛŋkpɔgele} \tbHM{} \tbPH{sɛ̂ŋkpɔ̀glɪ̀} \tbPS{n.~} \tbGE{bedbug} \tbOP{sɛŋkpɔgelɔ} \tbTP{sɛŋkpɔgelɛɛ} \tbVA{dakpalori, dakpaloree}
\tbLX{sɛre} \tbHM{} \tbPH{sɛ́rɪ́} \tbPS{adv.~} \tbGE{before} \tbXV{Sɛre ka a karembiiri la gaa sakuuri see ka ba dɔgereba yɔɔ a sakuuri boma na zaa ba naŋ sãã.} \tbXE{Before the students will be re-admitted to the school their parents must pay for all the school item that were destroyed.} 
\tbLX{sɛre} \tbHM{} \tbPH{sɛ́rɪ̀} \tbPS{v.~} \tbGE{move, shift} \tbXV{Yɛ naŋ sɛre gaa ka N meŋ zeŋ a dakogi zu.} \tbXE{Could you shift so that I could also sit on the bench.} \tbPD{sɛrɛɛ, sɛrɛ, sɛrɛ, sɛrebɛ, sɛraa, - } 
\tbLX{sɛredire} \tbHM{} \tbPH{sɛ̀rɪ́díré} \tbPS{n.~} \tbGE{judge} \tbOP{sɛredirbe} \tbTP{sɛredirbee} 
\tbLX{sɛremaalogi} \tbHM{} \tbPH{sɛ́rɪ́máálógí} \tbPS{n.~} \tbGE{pepper farm} \tbOP{sɛremaalogiri} \tbTP{sɛremaalogiree} 
\tbLX{sɛremaane} \tbHM{} \tbPH{sɛ́rɪ́máánɪ̀} \tbPS{n.~} \tbGE{pepper} \tbSG{sɛremaanoo} \tbOP{sɛremaama} \tbTP{sɛremaanɛɛ} 
\tbLX{sɛrɛɛ} \tbHM{} \tbPH{sɛ̀rɛ́ɛ̀} \tbPS{n.~} \tbGE{judgement} \tbOP{sɛrre} \tbTP{sɛrrɛɛ} 
\tbLX{sieree} \tbHM{} \tbPH{sìèrèè} \tbPS{n.~} \tbGE{benefit} 
\tbLX{sigi} \tbHM{} \tbPH{sígí} \tbPS{v.~} \tbSN \tbGE{to go down, to descend} \tbSN \tbGE{to land, to alight} \tbSN \tbGE{to decrease (intransitive)} \tbPD{sigee, sigre, sigre, sigribe, sigraa, sigribe } \tbSE{sigi kɔmpare} \tbGE{to sink} \tbSE{sigi puli} \tbGE{to be low} 
\tbLX{sigili} \tbHM{} \tbPH{sìgilì} \tbPS{v.~} \tbGE{to become calm, to become quiet} \tbXV{Zenɛ, saseɛ ba da lɛnso ka a zie sigili.} \tbXE{Today there is no wind and that is why the place is calm.} \tbPD{siglee, sigle, sigle, siglibe, siglaa, - } 
\tbLX{sigili} \tbHM{} \tbPH{sìgilì} \tbPS{v.~} \tbGE{to offer a gift or praise in honour of something well done} \tbXV{A pɔgelee seɛ la a seɛre ne toori ka a naa sigili o ne libie.} \tbXE{The young lady danced so well that the chief honoured her by offering her money.} \tbPD{siglee, sigle, sigle, siglibe, siglaa, - } 
\tbLX{sigili} \tbHM{} \tbPH{sìgilì} \tbPS{v.~} \tbGE{to get choked (on food)} \tbXV{A noɔ dire la a zɔŋ gborɔgborɔ ka a te sigili o.} \tbXE{The fowl was eating the feed very fast and choked on it.} \tbPD{siglee, sigle, sigle, siglibe, siglaa, - } 
\tbLX{sigili} \tbHM{} \tbPH{sìgilì} \tbPS{v.~} \tbGE{to gang up on (two or more people attacking one)} \tbXV{Bayuo ne o yɔɔ sigili la a dɔɔ kyɛ ba to᷉ɔ᷉ o.} \tbXE{Bayuo and his brother fought the man simutaneously but they could not beat him.} \tbPD{siglee, sigle, sigle, siglibe, siglaa, - } 
\tbLX{sigiri} \tbHM{} \tbPH{sígírì} \tbPS{v.~} \tbSN \tbGE{to lower, to reduce} \tbXV{Ka banaŋ ba sigiri a nɔɔteɛ daaroŋ N koŋ to᷉ɔ᷉ a da.} \tbXE{If they don't lower the price of the shoes, I can't buy them.} \tbSN \tbGE{to humble} \tbXV{A seŋ ka karembiiri  maŋ sigiri ba menne ka karemamine to᷉ɔ᷉ kanne ba.} \tbXE{Pupils should learn to humble themselves so that teachers can teach them.} \tbPD{sigiree, sigire, sigire, sigiribe, sigiraa, sigiribe } 
\tbLX{sikiri} \tbHM{} \tbPH{síkìrí} \tbPS{n.~} \tbGE{sugar} \tbTP{sikiree} \tbVA{sukyiri}
\tbLX{sile} \tbHM{} \tbPH{sílè} \tbPS{n.~} \tbGE{shilling} \tbOP{silli} \tbTP{sillee} 
\tbLX{simmie} \tbHM{} \tbPH{símmíé} \tbPS{n.~} \tbSN \tbGE{bambara beans} \tbSN \tbGE{groundnuts (Northern Dagaare dialect)} \tbSG{simmiruu} \tbTP{simmiiree} \tbVA{siŋgbule}
\tbLX{simmiri} \tbHM{} \tbPH{símmírì} \tbPS{n.~} \tbGE{thread} \tbSG{simmiruu} \tbOP{simmie} \tbTP{simmiiree} 
\tbLX{sinkyigiri} \tbHM{} \tbPH{sínʧígirí} \tbPS{n.~} \tbGE{bunch, head (of guinea corn)} \tbOP{sinkyige} \tbTP{sinkyigree} 
\tbLX{sinsuoluŋ} \tbHM{} \tbPH{sínsúólúŋ} \tbPS{n.~} \tbGE{ground bean} \tbSG{sinsuoluu} \tbTP{sinsuolee} 
\tbLX{siŋgbule} \tbHM{} \tbPH{sìŋgbúlé} \tbPS{n.~} \tbGE{bambara beans (a type of round beans that bear nuts in the ground like groundnuts) (Northern Dagaare dialect)} \tbVA{simmie}
\tbLX{sipirii} \tbHM{} \tbPH{sìpìríì} \tbPS{n.~} \tbGE{aspirin} \tbSG{sipiruu} \tbTP{sipirree} 
\tbLX{siremm} \tbHM{} \tbPH{sírémm} \tbPS{adv.~} \tbGE{serene} 
\tbLX{sisi} \tbHM{} \tbPH{sìsí} \tbPS{v.~} \tbGE{to cheat} \tbPD{sisiee, sisiire, sisiire, sisiiribe, sisiiraa, - } 
\tbLX{sitime} \tbHM{} \tbPH{sìtìmè} \tbPS{n.~} \tbGE{ship, steamer} \tbOP{sitimeri} \tbTP{sitimeree} 
\tbLX{so} \tbHM{} \tbPH{sò} \tbPS{v.~} \tbGE{to bathe} \tbPD{soe, suoro, suoro, suoribo, suoraa, suori } 
\tbLX{so} \tbHM{} \tbPH{sʊ́} \tbPS{v.~} \tbGE{to own} \tbPD{soɛ, soora, soora, sooreba, sooraa, - } 
\tbLX{so} \tbHM{} \tbPH{sʊ́} \tbPS{conj.~} \tbGE{as a result} 
\tbLX{sobile} \tbHM{} \tbPH{sóbílé} \tbPS{n.~} \tbGE{path} \tbOP{sobilii} \tbVA{solee}
\tbLX{sobiri} \tbHM{} \tbPH{sóbírì} \tbPS{n.~} \tbSN \tbGE{road, way} \tbSN \tbGE{journey} \tbOP{sobie} \tbTP{sobiiree} \tbVA{sori}
\tbLX{sobɔ} \tbHM{} \tbPH{sʊ́bɔ́} \tbPS{n.~} \tbSN \tbGE{owner} \tbXV{N ba baŋe a gane ŋa sobɔ a na de te ko o.} \tbXE{I don't know the owner of this book well enough that I could take it to him/her.} \tbSN \tbGE{the one, the person} \tbXV{A sobɔ na naŋ la yuo a pane, onɔ la de a libie.} \tbXE{The person who opened the door is the one who has taken the money.} \tbOP{demee} \tbVA{soba}
\tbLX{so-eŋe} \tbHM{} \tbPH{sò-ìŋ} \tbPS{v.~} \tbSN \tbGE{to initiate} \tbPD{so-eŋɛɛ, suoro-ennɛ, so-ennɛ, so-enemɛ, so-ennaa, - } 
\tbLX{so-eŋe} \tbHM{} \tbPH{sò-ìŋ} \tbPS{n.~} \tbGE{metaphor (literature)} 
\tbLX{sofoɔlaa} \tbHM{} \tbPH{sófʊ̀ɔ́láà} \tbPS{n.~} \tbGE{street} \tbOP{sofoɔlɪ} 
\tbLX{sogelɔŋ} \tbHM{} \tbPH{sʊ̀glɔ̀ŋ} \tbPS{adv.~} \tbGE{heavy and dull} 
\tbLX{sogi} \tbHM{} \tbPH{sògì} \tbPS{v.~} \tbGE{to catch in midair} \tbPD{sogee, sogro, sogro, sogribo, sograa, sogribo } 
\tbLX{sogi} \tbHM{} \tbPH{sògì} \tbPS{v.~} \tbGE{to pant, to breathe with difficulty} \tbPD{sogee, sogro, sogro, sogribo, sograa, sogribo } 
\tbLX{sogi} \tbHM{} \tbPH{sògì} \tbPS{v.~} \tbGE{to chirp (guinea fowl)} \tbPD{sogee, sogro, sogro, sogribo, sograa, sogribo } 
\tbLX{sogɔ} \tbHM{} \tbPH{sʊ́gɔ́} \tbPS{adj.~} \tbGE{middle} \tbOP{sogere} \tbTP{sogerɛɛ} \tbVA{soga, shortened form of sensogelensogɔ}
\tbLX{sogya} \tbHM{} \tbPH{sóɲà} \tbPS{n.~} \tbGE{soldier} \tbOP{sogyare} \tbTP{sogyarɛɛ} \tbVA{sogyɛ}
\tbLX{sokonsuo} \tbHM{} \tbPH{sókónsúó} \tbPS{n.~} \tbGE{storey} \tbOP{sokonsori} \tbTP{sokonsoree} 
\tbLX{sokoɔraa} \tbHM{} \tbPH{sókʊ́ɔ́ràá} \tbPS{n.~} \tbGE{large road} \tbOP{sokoɔre} \tbTP{sokoɔrɛɛ} 
\tbLX{sokyara} \tbHM{} \tbPH{sóʧárá} \tbPS{n.~} \tbGE{crossroads, intersection} \tbOP{sokyarre} \tbTP{sokyarɛɛ} 
\tbLX{sokyɛnɛ} \tbHM{} \tbPH{sóʧɛ́nɛ̀} \tbPS{n.~} \tbGE{traveler} \tbOP{sokyɛnemɛ} \tbTP{sokyɛnemɛɛ} 
\tbLX{solee} \tbHM{} \tbPH{sóléé} \tbPS{n.~} \tbGE{path} \tbVA{sobile}
\tbLX{solgbolo} \tbHM{} \tbPH{sʊ̀lgbʊ̀lʊ̀} \tbPS{adj.~} \tbGE{inactive} 
\tbLX{soli} \tbHM{} \tbPH{sólí} \tbPS{adj.~} \tbGE{in a neat manner} \tbXV{a pɔgelee vare gaŋe la a miri lɛ soli.} \tbXE{The girlcleared the high jump bar neat.} 
\tbLX{solli} \tbHM{} \tbPH{sòllì} \tbPS{v.~} \tbGE{to be well polished} \tbPD{sollee, sollo, sollo, solibo, sollaa, - } 
\tbLX{solli} \tbHM{} \tbPH{sóllí} \tbPS{v.~} \tbGE{to narrate/repeat incessantly} \tbPD{sollee, sollo, sollo, solibo, sollaa, - } 
\tbLX{solom} \tbHM{} \tbPH{sólóm} \tbPS{adv.~} \tbGE{completely e.g without being touched} \tbXV{a nannyige dgereba iri la ba nuuri solom kyɛ bare o ka ba gaa te pɔge.} \tbXE{The parents of the thief withdrew their support for him completely and he was sent to jail.} 
\tbLX{soma} \tbHM{} \tbPH{sʊ́má} \tbPS{n.~} \tbGE{a person who is good to one} 
\tbLX{somasoma} \tbHM{} \tbPH{sʊ́másʊ́má} \tbPS{adv.~} \tbGE{quietly} 
\tbLX{somɔ} \tbHM{} \tbPH{sʊ́mɔ́} \tbPS{adj.~} \tbGE{good} \tbGE{beautiful, attractive} \tbGE{nice} \tbXV{A pɔge na yɛlɛ somɔ la o ba baŋe yelfaare yeluu.} \tbXE{That woman is nice, she never uses unsavory language.} \tbOP{sonne} \tbTP{sonnɛɛ} \tbVA{soma}
\tbLX{somɔ} \tbHM{} \tbPH{sʊ́mɔ́} \tbPS{adj.~} \tbGE{holy} 
\tbLX{sompoɔ} \tbHM{} \tbPH{sʊ́mpʊ́ɔ́} \tbPS{n.~} \tbGE{three pence} \tbOP{sompoore} \tbTP{sompoorɛɛ} 
\tbLX{sonono} \tbHM{} \tbPH{sʊ́nʊ́nʊ́} \tbPS{adv.~} \tbGE{extremely cold} 
\tbLX{soŋ} \tbHM{} \tbPH{sʊ̌ŋ} \tbPS{adj.~} \tbGE{good, fine} \tbOP{sonne} \tbTP{sonnɛɛ} 
\tbLX{soŋe} \tbHM{} \tbPH{sʊ̀ŋ} \tbPS{v.~} \tbGE{to help} \tbXV{Dagaaba nɔŋe ka ba maŋ soŋe taa ba tomɔ poɔ.} \tbXE{The Dagaaba like to help each other in their work.} \tbPD{soŋɛɛ, sonna, sonna, sonema, sonnaa, sonema } 
\tbLX{soo} \tbHM{} \tbPH{sóó} \tbPS{adv.~} \tbGE{pure, unmixed} 
\tbLX{so᷉o} \tbHM{} \tbPH{sʊ́ʊ́} \tbPS{adv.~} \tbGE{confused} 
\tbLX{soore} \tbHM{} \tbPH{sʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{to question, to inquire} \tbPD{soorɛɛ, soorɔ, soorɔ, sooreba, sooraa, - } 
\tbLX{soore} \tbHM{} \tbPH{sʊ́ʊ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to hiss to drive fowls away} \tbSN \tbGE{to hiss in pain e.g. after eating very spicy food} \tbPD{-, -, -, -, -, - } 
\tbLX{soori} \tbHM{} \tbPH{sòòrí} \tbPS{n.~} \tbGE{locusts} \tbOP{soori} \tbTP{sooree} 
\tbLX{soɔ} \tbHM{} \tbPH{sʊ́ɔ́} \tbPS{n.~} \tbGE{knife} \tbOP{sɔre} \tbTP{sɔrɛɛ} \tbVA{sɔlee,sɔbilii}
\tbLX{soɔ} \tbHM{} \tbPH{sʊ́ɔ̀} \tbPS{v.~} \tbGE{to blacken, to darken} \tbPD{soɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } \tbVA{sɔɔ}
\tbLX{soɔ} \tbHM{} \tbPH{sʊ́ɔ̀} \tbPS{v.~} \tbGE{to blossom e.g. vegetation} \tbPD{soɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } \tbVA{sɔɔ}
\tbLX{so᷉ɔ} \tbHM{} \tbPH{sʊ̀ɔ̀} \tbPS{v.~} \tbGE{to flatter} \tbPD{so᷉ɔɛ, soɔnɔ, soɔnɔ, soɔnema, soɔnaa, - } 
\tbLX{so᷉ɔ} \tbHM{} \tbPH{sʊ̀᷉ɔ́᷉} \tbPS{n.~} \tbGE{witchcraft, supernatural power} \tbXV{N teɛre ka a bie ŋa taa la so᷉ɔ᷉; wolɔ ka o e do a teɛ ŋa?} \tbXE{I think this boy has supernatural power; how did he manage to climb this tree?.} 
\tbLX{so᷉ɔ} \tbHM{} \tbPH{sʊ̀᷉ɔ̀᷉} \tbPS{v.~} \tbGE{to draw a line on the ground with the foot} \tbPD{soɔɛ, soɔnɔ, soɔnɔ, soɔnema, soɔnaa, - } 
\tbLX{soɔbɔ} \tbHM{} \tbPH{sʊ́ɔ́bɔ́} \tbPS{n.~} \tbGE{wizard, witch} \tbOP{soɔre} \tbTP{sɔrɛɛ} \tbVA{soɔba}
\tbLX{so᷉ɔgo᷉ɔo} \tbHM{} \tbPH{sʊ́᷉ɔ́᷉gʊ́᷉ɔ́᷉ʊ́} \tbPS{n.~} \tbGE{pretence} 
\tbLX{soɔloŋ} \tbHM{} \tbPH{sʊ̀ɔ̀lʊ́ŋ} \tbPS{n.~} \tbGE{witchcraft} 
\tbLX{soɔne} \tbHM{} \tbPH{sʊ̀ɔ̀nɪ̀} \tbPS{v.~} \tbGE{to praise, to glorify} \tbPD{soɔnɛɛ, soɔnɔ, soɔnɔ, soɔnema, soɔnaa, - } \tbVA{soane}
\tbLX{soɔŋaa} \tbHM{} \tbPH{sʊ́ɔ́ŋàà} \tbPS{n.~} \tbGE{rabbit, hare} \tbOP{soɔne} \tbTP{soɔnɛɛ} 
\tbLX{soɔre} \tbHM{} \tbPH{sʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{dye} \tbTP{soɔrɛɛ} \tbVA{sɔɔre}
\tbLX{soɔsobɔ} \tbHM{} \tbPH{sʊ́ɔ́sʊ́bɔ́} \tbPS{n.~} \tbGE{chief celebrant (traditional worship)} \tbOP{sɔdemee} \tbVA{soɔsoba}
\tbLX{so᷉ɔzeɛnagbare} \tbHM{} \tbPH{sʊ́᷉ɔ́zɪ́ɛ́nàgbárɪ́} \tbPS{n.~} \tbGE{small but old or strong} \tbOP{so᷉ɔzeɛnagbarɛɛ} 
\tbLX{sori} \tbHM{} \tbPH{sórì} \tbPS{n.~} \tbGE{way, road} \tbOP{soe} \tbTP{soree} 
\tbLX{sotugane} \tbHM{} \tbPH{sótùgánɪ́} \tbPS{n.~} \tbGE{map} \tbOP{sotugama} \tbTP{sotugannɛɛ} 
\tbLX{sotuuraa} \tbHM{} \tbPH{sótúúráà} \tbPS{n.~} \tbGE{pathway} \tbOP{sotuuri} 
\tbLX{sɔ} \tbHM{} \tbPH{sɔ́} \tbPS{v.~} \tbGE{to smear, to apply} \tbPD{sɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } 
\tbLX{sɔ} \tbHM{} \tbPH{sɔ̀} \tbPS{v.~} \tbGE{to add vegetables to groundnut sauce} \tbPD{sɔɛ, soɔrɔ, soɔrɔ, soɔreba, soɔraa, - } 
\tbLX{sɔbiri} \tbHM{} \tbPH{sɔ̀bírì} \tbPS{n.~} \tbGE{knife blade} \tbOP{sɔbie} \tbTP{sɔbiiree} 
\tbLX{sɔge} \tbHM{} \tbPH{sɔ́gɪ́} \tbPS{v.~} \tbGE{to shell nuts} \tbPD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } 
\tbLX{sɔge} \tbHM{} \tbPH{sɔ́gɪ̀} \tbPS{v.~} \tbGE{to rinse} \tbPD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } \tbSE{sɔge noɔre} \tbGE{to eat breakfast} 
\tbLX{sɔge} \tbHM{} \tbPH{sɔ̀gɪ̀} \tbPS{adj.~} \tbGE{bushy, littered} \tbXV{A dendɔre sɔgɛɛ la, a seŋ ka ba peere o.} \tbXE{The couryard is litered with rubbish, so it needs to be swept.} 
\tbLX{sɔge} \tbHM{1} \tbPH{sɔ̀gɪ̀} \tbPS{v.~} \tbGE{to respond, to reply (in a conversation)} \tbPD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } 
\tbLX{sɔge} \tbHM{2} \tbPH{sɔ̀gɪ̀} \tbPS{v.~} \tbGE{to be bushy} \tbPD{sɔgɛɛ, sɔgerɔ, sɔgerɔ, sɔgereba, sɔgeraa, sɔgereba } 
\tbLX{sɔgelaa} \tbHM{} \tbPH{sɔ́glàá} \tbPS{adj.~} \tbSN \tbGE{black} \tbSN \tbGE{dark (colour)} \tbOP{sɔgelɔ} 
\tbLX{sɔgelaa} \tbHM{} \tbPH{sɔ̀glàà} \tbPS{n.~} \tbGE{neck of a pot} \tbOP{sɔgele} \tbTP{sɔgelɛɛ} 
\tbLX{sɔgele} \tbHM{} \tbPH{sɔ́glɪ̀} \tbPS{v.~} \tbGE{to hide} \tbPD{sɔgelɛɛ, sɔgelɔ, sɔgelɔ, sɔgeleba, sɔgelaa, sɔgeleba } 
\tbLX{sɔgbale} \tbHM{} \tbPH{sɔ̀gbàlɪ́} \tbPS{n.~} \tbGE{blunt knife} \tbOP{sɔgbala} \tbTP{sɔgbalɛɛ} 
\tbLX{sɔ kpaare} \tbHM{} \tbPH{sɔ́ kpáárɪ́} \tbPS{v.~} \tbGE{to forget to deliver (a message)} \tbPD{sɔɛ kpaare, soɔrɔ kpaare, -, -, -, - } \tbVA{kpaare soɔbo}
\tbLX{sɔkpoŋi} \tbHM{} \tbPH{sɔ̀kpóŋ} \tbPS{n.~} \tbGE{cutlass, large knife} \tbOP{sɔkponni} \tbTP{sɔkponnee} 
\tbLX{sɔkyeɛraa} \tbHM{} \tbPH{sɔ̀ʧɪ̀ɛ̀ráà} \tbPS{n.~} \tbGE{cutlass} \tbOP{ sɔkyeɛrre} 
\tbLX{sɔlare} \tbHM{} \tbPH{sɔ̀lárɪ̀} \tbPS{n.~} \tbGE{cutlass} \tbOP{sɔlaɛ} \tbTP{sɔlarɛɛ} 
\tbLX{sɔlee} \tbHM{} \tbPH{sɔ̀léé} \tbPS{n.~} \tbGE{small knife} \tbOP{sɔbilii} 
\tbLX{sɔŋɔ} \tbHM{} \tbPH{sɔ́ŋ} \tbPS{n.~} \tbGE{wood carrier} \tbOP{sɔnne} \tbTP{sɔnnɛɛ} \tbVA{sɔŋa}
\tbLX{sɔɔ} \tbHM{} \tbPH{sɔ̀ɔ̀} \tbPS{adv.~} \tbGE{in a little while} \tbXV{A na erɛ sɔɔ ka sakuuri iree.} \tbXE{In a little while schools will close.} 
\tbLX{sɔɔ} \tbHM{} \tbPH{sɔ́ɔ̀} \tbPS{v.~} \tbGE{to blacken, to darken} \tbPD{-, -, -, -, -, - } \tbVA{soɔ}
\tbLX{sɔɔ} \tbHM{} \tbPH{sɔ́ɔ̀} \tbPS{v.~} \tbGE{to blossom e.g. vegetation} \tbPD{-, -, -, -, -, - } \tbVA{soɔ}
\tbLX{sɔɔre} \tbHM{1} \tbPH{sɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{dye} \tbTP{sɔɔrɛɛ} \tbVA{soɔre}
\tbLX{sɔɔre} \tbHM{2} \tbPH{sɔ́ɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{liver} \tbSN \tbGE{a type of dark brown bush mouse that lives along river banks} \tbOP{sɔbɔ} \tbTP{sɔɔrɛɛ} \tbVA{shortened form of ɔŋ-sɔɔre}
\tbLX{sɔre} \tbHM{} \tbPH{sɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to beg} \tbSN \tbGE{to ask, to request} \tbSN \tbGE{to plead, to implore} \tbPD{sɔrɛɛ, sɔrɔ, sɔrɔ, sɔrebɔ, sɔraa, sɔrebɔ } 
\tbLX{sɔre} \tbHM{} \tbPH{sɔ̀rɪ̀} \tbPS{v.~} \tbGE{to cook soup made with groundnuts and vegetables} \tbPD{sɔrɛɛ, sɔrɔ, sɔrɔ, sɔrebɔ, sɔraa, sɔrebɔ } 
\tbLX{sɔre} \tbHM{} \tbPH{sɔ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to count} \tbSN \tbGE{to read} \tbPD{-, -, -, -, -, - } 
\tbLX{sɔrebiri} \tbHM{} \tbPH{sɔ́rɪ́bírì} \tbPS{n.~} \tbGE{counter, pebble} \tbSG{sɔrebiruu} \tbOP{sɔrebie} \tbTP{sɔrebiiree} 
\tbLX{sɔresɔreŋmanlee} \tbHM{} \tbPH{sɔ̀rɪ̀sɔ̀rɪ̀ŋmànléé} \tbPS{n.~} \tbGE{a beggar's calabash} \tbOP{sɔresɔreŋmamblii} \tbTP{sɔresɔreŋmambilee} 
\tbLX{sɔresɔrɔ} \tbHM{} \tbPH{sɔ̀rsɔ́rɔ́} \tbPS{n.~} \tbGE{beggar} \tbOP{sɔrsɔreba} \tbTP{sɔrsɔrebɛɛ} 
\tbLX{sɔresɔrɔ} \tbHM{} \tbPH{sɔ̀rsɔ́rɔ́} \tbPS{n.~} \tbGE{beggar} \tbOP{sɔresɔrebɔ} \tbTP{sɔresɔrebɛɛ} \tbVA{sɔresɔra}
\tbLX{sɔteɛ} \tbHM{} \tbPH{sɔ̀tɪ́ɛ́} \tbPS{n.~} \tbGE{scabbard for a knife} \tbOP{sɔteere} \tbTP{sɔteerɛɛ} 
\tbLX{sɔvu᷉u} \tbHM{} \tbPH{sɔ̀vù᷉ú᷉} \tbPS{n.~} \tbGE{light associated with witchcraft} \tbSG{sɔvuunuu} \tbTP{sɔvuunee} 
\tbLX{su} \tbHM{} \tbPH{sù} \tbPS{v.~} \tbGE{to put on a top (shirt)} \tbPD{sue, suuro, -, -, -, - } \tbSE{su bonsuuri} \tbGE{to wear clothes} \tbSE{su kaayaa} \tbGE{to wear clothes} 
\tbLX{su} \tbHM{} \tbPH{sù} \tbPS{v.~} \tbGE{to put inside (a hole)} \tbPD{sue, suuro, -, -, -, - } \tbSE{su bonsuuri} \tbGE{to wear clothes} \tbSE{su kaayaa} \tbGE{to wear clothes} 
\tbLX{su} \tbHM{} \tbPH{sù} \tbPS{v.~} \tbGE{to spoon feed e.g. baby, patient} \tbPD{sue, suuro, -, -, -, - } \tbSE{su bonsuuri} \tbGE{to wear clothes} \tbSE{su kaayaa} \tbGE{to wear clothes} 
\tbLX{subaaloŋ} \tbHM{} \tbPH{sùbààlʊ́ŋ} \tbPS{n.~} \tbGE{tenderheartedness} 
\tbLX{sudire} \tbHM{} \tbPH{sùdíré} \tbPS{n.~} \tbGE{one who forgives, one who pardons} \tbOP{ surdiribe} \tbTP{sudiribee} 
\tbLX{sugaraate} \tbHM{} \tbPH{sùgàráátɪ́} \tbPS{n.~} \tbGE{cigarette} \tbOP{sugaraatere} \tbTP{sugaraaterɛɛ} \tbVA{segarate}
\tbLX{sugi} \tbHM{} \tbPH{sùgì} \tbPS{v.~} \tbSN \tbGE{to dislocate e.g. joint} \tbSN \tbGE{to dislodge e.g. a pot from the top of another} \tbPD{sugee, sugro, sugro, sugribo, sugraa, - } 
\tbLX{sugi} \tbHM{} \tbPH{sùgì} \tbPS{v.~} \tbGE{to ladel soup from a pot} \tbPD{sugee, sugro, sugro, sugribo, sugraa, - } 
\tbLX{sugili} \tbHM{} \tbPH{sùgìlì} \tbPS{v.~} \tbSN \tbGE{to come next to e.g. in a line} \tbXV{Ba naŋ tɔnɔ a lae maa la sugili o zu.} \tbXE{When they were queueing, I came next to him.} \tbSN \tbGE{to place one on top of the other e.g. pots} \tbXV{Ba pɛge la a laare wuo sugili taa.} \tbXE{They washed the bowls and placed them one on top of the other.} \tbPD{suglee, suglo, suglo, suglibo, suglaa, - } 
\tbLX{Sugilo} \tbHM{} \tbPH{súgíló} \tbPS{n.~} \tbGE{name traditionally given to a child immediately after twins} 
\tbLX{sugo} \tbHM{} \tbPH{súgò} \tbPS{n.~} \tbGE{herd; group of many e.g. troop of animals} \tbXV{ŋmaa-sugo kpɛ la n weɛ sa᷉a a kamaana zaa.} \tbXE{A troop of monkeys invaded my farm and destoyed all the maize.} \tbOP{sugiri} \tbTP{sugiree} 
\tbLX{sugo} \tbHM{} \tbPH{súgò} \tbPS{n.~} \tbGE{a type of tiny white ant that destroys yams} \tbXV{Sugo saana waare.} \tbXE{White ants destroys yams.} \tbTP{sugee} \tbVA{wasugo}
\tbLX{sukaale} \tbHM{} \tbPH{sùkààlɪ̀} \tbPS{n.~} \tbGE{a type of large dress with wide sleeves that are thrown over the shoulder} \tbTP{sukaalɛɛ} 
\tbLX{sukyiri} \tbHM{} \tbPH{sùʧírí} \tbPS{n.~} \tbSN \tbGE{heart} \tbXV{N sukyiri maŋ zaara la, a seŋ ka N gaa dɔɔta.} \tbXE{I keep experiencing heart palpitations; I need to go to the hospital.} \tbSN \tbGE{courage} \tbXV{Ka fooŋ ba taa sukyiri fo koŋ baŋ toŋ asibiti toma.} \tbXE{If you have no courage, you can't work in the hospital.} \tbOP{sukyie} \tbTP{sukyiree} \tbVA{sikyiri}
\tbLX{sukyiribaaloŋ} \tbHM{} \tbPH{sùʧìrìbààlʊ̀ŋ} \tbPS{n.~} \tbSN \tbGE{kind heart} \tbSN \tbGE{heart disease} 
\tbLX{sukyirifo᷉ɔ᷉o᷉} \tbHM{} \tbPH{súʧírífʊ́᷉ɔ́᷉ʊ́᷉} \tbPS{n.~} \tbGE{lightheartedness} \tbOP{sukyirifoɔne} \tbTP{sukyirifoɔnɛɛ} 
\tbLX{sukyiriliŋe} \tbHM{} \tbPH{sùʧìrìlíŋé} \tbPS{n.~} \tbGE{end of the ribcage} \tbOP{sukyirilinni} \tbTP{sukyirilinnee} 
\tbLX{sukyiritage} \tbHM{} \tbPH{súʧírítágɪ́} \tbPS{n.~} \tbGE{bravery} \tbOP{sukyiritagre} 
\tbLX{sulee} \tbHM{} \tbPH{sùlèè} \tbPS{n.~} \tbGE{chimney} \tbOP{sulli} \tbTP{sullee} 
\tbLX{suli} \tbHM{} \tbPH{sùlì} \tbPS{v.~} \tbGE{to hang one's head, to look down} \tbPD{sulee, sullo, sullo, sulbo, sullaa, - } 
\tbLX{sumaaroŋ} \tbHM{} \tbPH{sùmáárʊ́ŋ} \tbPS{n.~} \tbGE{coolheaded} \tbXV{A pɔge taa la sumaaroŋ ka lɛ naane o da na zɔɔ ne la a dɔɔ.} \tbXE{The woman is coolheaded otherwise she would have fought the man.} 
\tbLX{sumiime} \tbHM{} \tbPH{súmíímé} \tbPS{n.~} \tbGE{heartburn} 
\tbLX{sumosumo} \tbHM{} \tbPH{sùmòsúmó} \tbPS{n.~} \tbGE{a type of fried food made from the flour of yam} 
\tbLX{sundaa} \tbHM{} \tbPH{súndàá} \tbPS{n.~} \tbGE{a stick to which birdlime is applied to trap birds} \tbOP{sundaare} \tbTP{sundarɛɛ} 
\tbLX{sunni} \tbHM{} \tbPH{súnnì} \tbPS{n.~} \tbSN \tbGE{glue, gum, sticky substance} \tbSN \tbGE{a plant that produces gum} \tbSN \tbGE{birdlime} \tbSG{sunnuu} \tbOP{sumo} \tbTP{sunnee} \tbVA{sun (saanchi)}
\tbLX{sunni} \tbHM{} \tbPH{súnní} \tbPS{n.~} \tbGE{blue powder used in dyeing clothes during washing} 
\tbLX{sunsugiri} \tbHM{} \tbPH{súnsúgrí} \tbPS{n.~} \tbGE{white growth within a sore} \tbOP{sunsugiri} \tbTP{sunsugiree} 
\tbLX{suori} \tbHM{} \tbPH{súórì} \tbPS{v.~} \tbGE{to fall off in great numbers (leaves)} \tbPD{suoree, suoro, suoro, suoribo, suoraa, - } 
\tbLX{supaana} \tbHM{} \tbPH{sùpáánà} \tbPS{n.~} \tbGE{spanner} \tbSG{supaanoo} \tbOP{supaanare} \tbTP{supaanarɛɛ} 
\tbLX{suree} \tbHM{} \tbPH{sùréé} \tbPS{n.~} \tbGE{ladle} \tbOP{surri} 
\tbLX{suromm} \tbHM{} \tbPH{súrómm} \tbPS{adv.~} \tbGE{quiet} 
\tbLX{susu} \tbHM{} \tbPH{súsú} \tbPS{n.~} \tbGE{a cycle in which contributions of members are given to others} \tbOP{susuri} \tbTP{susuree} 
\tbLX{susu} \tbHM{} \tbPH{sùsú} \tbPS{v.~} \tbSN \tbGE{to measure} \tbSN \tbGE{to be careful, to not overdo something} \tbSN \tbGE{to think about carefully} \tbPD{susue, susuuro, susuuro, susuribo, susuuraa, - } 
\tbLX{susu} \tbHM{} \tbPH{sùsù} \tbPS{adj.~} \tbGE{fast and without stopping e.g. eating} 
\tbLX{susu} \tbHM{} \tbPH{sùsù} \tbPS{interjection.~} \tbGE{utterance to urge a dog on to pursue (animal, thief)} 
\tbLX{su-su-su} \tbHM{} \tbPH{sù-sù-sù} \tbPS{interjection.~} \tbGE{an utterance to urge a dog on to catch game} 
\tbLX{sutoloŋ} \tbHM{} \tbPH{sʊ̀tʊ́lʊ́ŋ} \tbPS{n.~} \tbGE{the state of being quick tempered} \tbXV{Daŋ ta wa deɛnɛ ne a dɔɔ na, o taa la sutoloŋ.} \tbXE{Don't you ever play with that man; he is quick tempered.} 
\tbLX{suu} \tbHM{} \tbPH{súú} \tbPS{v.~} \tbGE{to doze off for a short time} \tbPD{-, -, -, -, -, - } 
\tbLX{suuri} \tbHM{} \tbPH{súúrí} \tbPS{n.~} \tbGE{anger} \tbOP{sue} \tbTP{suuree} 
\tbLX{suuri} \tbHM{} \tbPH{súúrì} \tbPS{v.~} \tbGE{to coax, to encourage, to push to pursue e.g. dog for game} \tbPD{suuree, suuro, suuro, suuribo, suuraa, - } 
\tbLX{suuri-di-teroo} \tbHM{} \tbPH{súúr-dì-tɪ́rʊ́ʊ́} \tbPS{n.~} \tbGE{forgiveness} 
\tbLX{ta} \tbHM{} \tbPH{tá} \tbPS{adj.~} \tbGE{not (negative marker)} \tbXV{A pɔge kpa᷉a la o bie ka o ta gaa a sini.} \tbXE{The woman cautioned her son not to go to the cinema.} 
\tbLX{ta} \tbHM{} \tbPH{tá} \tbPS{v.~} \tbSN \tbGE{to reach, to arrive} \tbXV{Ka fooŋ wa ta fo yiri foŋ yeli ko ma.} \tbXE{When you arrive at your house, tell me.} \tbSN \tbGE{to approach} \tbXV{Fo maŋ gaa ka fo ta o yiri ka baare  vuoli paale ziezaa.} \tbXE{As soon as you approach the house, dogs bark all over.} \tbPD{taɛ, tara, tara, tareba, taraa, tareba } 
\tbLX{taa} \tbHM{} \tbPH{tàá} \tbPS{v.~} \tbSN \tbGE{to have} \tbXV{A bie taa la maŋgori ayi.} \tbXE{The child has two mangoes.} \tbSN \tbGE{to keep, to hold onto} \tbXV{N ba de la o zupili ka N taa.} \tbXE{My father has given his hat to me to keep.} \tbXV{A pɔge taa la poɔ.} \tbXE{The woman is pregnant.} \tbXV{o taa la yelmeŋɛ} \tbXE{S/he is honest.} \tbXV{A bibiiri taa la gɔnne.} \tbXE{The children are noisy.} \tbPD{taaɛ, tara, tara, tareba, taara, tareba } \tbSE{taa kɔŋ} \tbGE{to be hungry} 
\tbLX{taa} \tbHM{} \tbPH{táá} \tbPS{pron.~} \tbGE{each other} 
\tbLX{ta᷉a} \tbHM{} \tbPH{tà᷉à᷉} \tbPS{v.~} \tbGE{to serve very little e.g. soup} \tbXV{A zeɛre ba seŋe bone lɛnso ka ba ta᷉a᷉ a a lɛ.} \tbXE{There is very little soup left and that is why they have given out so little.} \tbPD{ta᷉aɛ, taana, taana, taanema, taanaa, - } 
\tbLX{ta᷉a} \tbHM{} \tbPH{tà᷉à᷉} \tbPS{v.~} \tbGE{to tap on the head with a finger} \tbPD{ta᷉aɛ, taana, taana, taanema, taanaa, - } 
\tbLX{taa kɔnnyuuri} \tbHM{} \tbPH{tàá kɔ̀nɲúúrì} \tbPS{v.~} \tbGE{to be thirsty} \tbPD{-, -, -, -, -, - } 
\tbLX{taala} \tbHM{} \tbPH{táálá} \tbPS{n.~} \tbGE{ingredients e.g. sauce} \tbTP{taalɛɛ} 
\tbLX{taa laare} \tbHM{} \tbPH{tàá láárɪ́} \tbPS{adj.~} \tbGE{amusing, funny} 
\tbLX{taama} \tbHM{} \tbPH{táámá} \tbPS{n.~} \tbGE{suffering} 
\tbLX{taambaa} \tbHM{} \tbPH{tàámbàá} \tbPS{n.~} \tbGE{young man} \tbOP{tambaare} \tbTP{taambaarɛɛ} 
\tbLX{taambeene} \tbHM{} \tbPH{tàámbɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{ripe shea tree fruit} \tbOP{taambeemɛ} \tbTP{taambeenɛɛ} 
\tbLX{taane} \tbHM{} \tbPH{tàànɪ̀} \tbPS{v.~} \tbGE{to stretch a great deal} \tbPD{taanɛɛ, taana, taana, taanema, taanaa, - } 
\tbLX{taa nimibaaloŋ} \tbHM{} \tbPH{} \tbPS{adj.~} \tbGE{kind} 
\tbLX{taa nimiri} \tbHM{} \tbPH{tàá nímírì} \tbPS{adj.~} \tbGE{covetous} 
\tbLX{taannuulee} \tbHM{} \tbPH{tàánnúúléé} \tbPS{n.~} \tbGE{canary} \tbOP{taannuuli} 
\tbLX{taa noɔre} \tbHM{} \tbPH{tàá nʊ́ɔ́rɪ̀} \tbPS{adj.~} \tbGE{talkative, noisy} \tbXV{A balee taa la noɔre te zuo, o maŋ vuolo la tegitegi lɛ.} \tbXE{The puppy is noisy; it always barks.} 
\tbLX{taantumma} \tbHM{} \tbPH{tá᷉â᷉ntú᷉mmá᷉} \tbPS{n.~} \tbGE{a type of caterpillar in hibernation} \tbOP{taantummamine.} 
\tbLX{taantunnuu} \tbHM{} \tbPH{tàântúnnì} \tbPS{n.~} \tbGE{shea tree caterpillar} \tbSG{taantunnuu} \tbOP{taantunni} \tbTP{taantunnee} 
\tbLX{taa nyonono} \tbHM{} \tbPH{tàá ɲʋ́nʋ́nʋ́} \tbPS{n.~} \tbSN \tbGE{selfishness} \tbSN \tbGE{sympathy} 
\tbLX{taaŋkaraa} \tbHM{} \tbPH{tàáŋkàráá} \tbPS{n.~} \tbGE{unripe shea fruit} \tbOP{taaŋkare} \tbTP{taaŋkarɛɛ} 
\tbLX{taaŋkpara} \tbHM{} \tbPH{tááŋkpárá} \tbPS{n.~} \tbGE{shea fruit with two nuts} \tbOP{taaŋkparre} \tbTP{taaŋkparɛɛ} 
\tbLX{taa poɔ} \tbHM{} \tbPH{tàá pʊ́ɔ́} \tbPS{v.~} \tbGE{to be pregnant} \tbPD{-, -, -, -, -, - } 
\tbLX{taare} \tbHM{} \tbPH{tààrɪ̀} \tbPS{v.~} \tbGE{to press hard e.g. plaster, to crush e.g. an ant} \tbPD{taarɛɛ, taara, taara, taareba, taaraa, - } 
\tbLX{taare} \tbHM{} \tbPH{táárɪ̀} \tbPS{v.~} \tbGE{to tighten, to screw} \tbPD{taarɛɛ, taara, taara, taareba, taaraa, - } 
\tbLX{ta᷉ate᷉e} \tbHM{} \tbPH{tà᷉á᷉tɪ̀᷉ɪ́᷉} \tbPS{n.~} \tbGE{a type of small bird} \tbOP{ta᷉ateene} \tbTP{ta᷉ateenɛɛ} 
\tbLX{taa toori} \tbHM{} \tbPH{táá tòòrí} \tbPS{adj.~} \tbGE{stubborn (lit. be hard-eared)} 
\tbLX{taa yelmeŋɛ} \tbHM{} \tbPH{taa yelmeŋɛ} \tbPS{adj.~} \tbGE{is honest} 
\tbLX{taayɛ} \tbHM{} \tbPH{tááyɛ̀} \tbPS{n.~} \tbSN \tbGE{tyre (rubber)} \tbSN \tbGE{catapult} \tbOP{ taayɛre} \tbTP{taayɛrɛɛ} 
\tbLX{tabole} \tbHM{} \tbPH{tábʊ̀l} \tbPS{n.~} \tbGE{table} \tbOP{tabolɔ} \tbTP{tabolɛɛ/tabollɛɛ} 
\tbLX{tagaremiri} \tbHM{} \tbPH{táŋgàrɪ̀mírì} \tbPS{n.~} \tbGE{telegram cable} \tbOP{tagaremie} \tbTP{tagaremiiree} 
\tbLX{tage} \tbHM{} \tbPH{tàgɪ̀} \tbPS{v.~} \tbSN \tbGE{to pull, to draw in or out} \tbSN \tbGE{to make thick (liquid)} \tbPD{tagɛɛ, tagra, tagra, tagreba, tagraa, tagreba } 
\tbLX{tage} \tbHM{} \tbPH{tágɪ́} \tbPS{adj.~} \tbGE{thick} \tbOP{tagre} 
\tbLX{tagekori} \tbHM{} \tbPH{tàgɪ̀kòrì} \tbPS{n.~} \tbGE{a type of thorn tree with sharp, curved thorns} \tbTP{tagekoree} 
\tbLX{tagekyoore} \tbHM{} \tbPH{tágɪ̀ʧʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{measles} \tbTP{tagekyoorɛɛ} 
\tbLX{tagetaa} \tbHM{} \tbPH{tàgɪ̀táá} \tbPS{n.~} \tbSN \tbGE{dispute} \tbSN \tbGE{conflict (literature)} 
\tbLX{tagetage} \tbHM{} \tbPH{tàgtàg} \tbPS{n.~} \tbGE{ability} \tbXV{A baala gbɛɛ kpie la, o ba taa tagetage zaa.} \tbXE{The patient's legs are paralysed; s/he cannot do anything.} 
\tbLX{tagevama} \tbHM{} \tbPH{tàgɪ̀vàmà} \tbPS{n.~} \tbSN \tbGE{smallpox} \tbSN \tbGE{chicken pox} \tbVA{teɛgara}
\tbLX{tahɛlɛ} \tbHM{} \tbPH{táhɛ́lɛ́} \tbPS{n.~} \tbSN \tbGE{extreme suffering} \tbXV{Kɔmbuli saŋa pɔgeba maŋ nyɛrɛ la tahɛlɛ ko᷉ɔ᷉ yɛlɛ.} \tbXE{During drought, women suffer severely for lack of water.} \tbSN \tbGE{carefree speech, behaviour; nonchalence} \tbXV{A bie ŋa eŋ ba naŋ ba feɛ o o yelee tahɛlɛ.} \tbXE{Since this boy was not caned he speaks with airs.} \tbOP{tahɛlle} \tbTP{tahɛlɛɛ/tahɛllɛɛ} 
\tbLX{takaare} \tbHM{} \tbPH{tákáárɪ̀} \tbPS{adj.~} \tbGE{vain, proud of oneself} 
\tbLX{takaraa} \tbHM{} \tbPH{tákáråá} \tbPS{n.~} \tbGE{small lizard} \tbOP{takarrre} \tbTP{takarrɛɛ} 
\tbLX{takoŋi} \tbHM{} \tbPH{tákòŋ} \tbPS{n.~} \tbGE{ban on performance of funeral rites} 
\tbLX{takoriko} \tbHM{} \tbPH{tàkòríkò} \tbPS{n.~} \tbGE{bridge} \tbOP{takorikori} \tbTP{takorikoree} \tbVA{tokoriko}
\tbLX{takoroo} \tbHM{} \tbPH{tákóróó} \tbPS{n.~} \tbGE{window} \tbOP{takorri} \tbTP{takoree/takorree} \tbVA{tokoroo}
\tbLX{takɔdaa} \tbHM{} \tbPH{tákɔ́dàá} \tbPS{n.~} \tbGE{Sabbath} \tbOP{takɔdaare} \tbTP{takɔdarɛɛ} \tbVA{takɔraa}
\tbLX{takɔraa} \tbHM{} \tbPH{tákɔ́ráá} \tbPS{n.~} \tbGE{Sabbath} \tbVA{takɔdaa}
\tbLX{takɔraa} \tbHM{} \tbPH{tàkɔ̀ràà} \tbPS{n.~} \tbSN \tbGE{fin (fish)} \tbSN \tbGE{a hook or spike on an object e.g. arrow} \tbOP{takɔrre} \tbTP{takɔrrɛɛ} \tbVA{kɔraa}
\tbLX{takpala} \tbHM{} \tbPH{tákpálá} \tbPS{n.~} \tbGE{great suffering} \tbTP{takpallɛɛ} 
\tbLX{takpala} \tbHM{n} \tbPH{tàkpálà} \tbPS{n.~} \tbGE{area between the throat and the chin} \tbOP{takpalle} \tbTP{takpallɛɛ} \tbVA{takpaleɛne}
\tbLX{takpaleɛne} \tbHM{} \tbPH{tàkpàlɪ̀ɛ̀nɪ́} \tbPS{n.~} \tbGE{area between the throat and chin} \tbOP{takpaleɛmɛ} \tbTP{takpaleɛnɛɛ/takpaleɛmɛɛ} \tbVA{takpalan}
\tbLX{ta kpɛ} \tbHM{} \tbPH{tá kpɛ́} \tbPS{v.~} \tbGE{to run away, to flee, to take to one's heels} \tbXV{A bie gɛrɛ la te nyɛ waabo, a leɛ takpɛ.} \tbXE{As the girl walked along, she came across a snake and took to her heels.} \tbPD{ta kpɛɛ, tara kpeɛrɛ, takpeɛrɛ, takpeɛrebɛ, takpeɛraa, - } 
\tbLX{takyeŋi} \tbHM{} \tbPH{táʧêŋ} \tbPS{n.~} \tbGE{the absence of argument} \tbOP{takyenni} 
\tbLX{takyeraa} \tbHM{} \tbPH{táʧɪ́ràá} \tbPS{n.~} \tbGE{gecko} \tbOP{takyɪrrɪ} \tbTP{takyɪrrɛɛ} 
\tbLX{takyere} \tbHM{} \tbPH{táʧɪ̀rɪ̀} \tbPS{n.~} \tbGE{temple} \tbOP{takyeɛ} \tbTP{takyerɛɛ} 
\tbLX{talaa} \tbHM{} \tbPH{tàláà} \tbPS{n.~} \tbGE{metal bowl or plate} \tbOP{talaare} \tbTP{talaarɛɛ} 
\tbLX{talane} \tbHM{} \tbPH{tàlánɪ̀} \tbPS{n.~} \tbGE{palate} \tbOP{talama} \tbTP{talannɛɛ} 
\tbLX{talantalaŋ} \tbHM{} \tbPH{tàlàntáláŋ} \tbPS{adv.~} \tbGE{haphazard} 
\tbLX{tale} \tbHM{} \tbPH{tálɪ́} \tbPS{n.~} \tbGE{lobe (of a kola nut)} \tbOP{tala} \tbTP{talɛɛ/tallɛɛ} 
\tbLX{tale} \tbHM{} \tbPH{tálɪ́} \tbPS{adj.~} \tbGE{flat open e.g. palm of the hand} \tbOP{tala} \tbTP{talɛɛ/tallɛɛ} 
\tbLX{taleri} \tbHM{} \tbPH{tálérí} \tbPS{n.~} \tbGE{small bush lizard} \tbOP{talerri} \tbTP{taleree} 
\tbLX{talle} \tbHM{} \tbPH{tàllɪ̀} \tbPS{v.~} \tbGE{to make long, quick strides; to walk fast} \tbPD{tallɛɛ, talla, talla, taleba, tallaa, - } 
\tbLX{taltal} \tbHM{} \tbPH{tàtál} \tbPS{v.~} \tbGE{to move with great pain} \tbPD{taltalɛɛ, taltala, taltala, taltaleba, taltalaa, - } 
\tbLX{tam} \tbHM{} \tbPH{tàm} \tbPS{v.~} \tbSN \tbGE{to be numb} \tbSN \tbGE{shocked; not able to react} \tbPD{tamɛɛ, tamena, tamena, tamenema, tamenaa, - } 
\tbLX{tamaa} \tbHM{} \tbPH{tàmáà} \tbPS{n.~} \tbGE{hope} 
\tbLX{tamanyini} \tbHM{} \tbPH{tàmàɲíní} \tbPS{n.~} \tbGE{gum} \tbOP{tamanyime} \tbTP{tamanyimee} \tbVA{tanyini}
\tbLX{tamaraa} \tbHM{} \tbPH{támáráà} \tbPS{n.~} \tbGE{small quiver} \tbOP{tamarre} \tbTP{tamarrɛɛ} 
\tbLX{tamaregye} \tbHM{} \tbPH{támárɪ́dzé} \tbPS{n.~} \tbGE{small pantaloons; short, tight trousers} \tbOP{tamaregyeri} \tbTP{tamaregyeree} \tbVA{tamaragye}
\tbLX{tambaa} \tbHM{} \tbPH{tǎmbáá} \tbPS{n.~} \tbGE{bow bell (a hollow, carved object that is tied to the beam of the bow and makes a sound when the bow is shot)} \tbOP{tambaare} \tbTP{tambaarɛɛ} 
\tbLX{tambalaaraa} \tbHM{} \tbPH{támbálåáráá} \tbPS{n.~} \tbGE{honey made by the yoyogibie} \tbOP{tambalaare} \tbTP{tambalaarɛɛ} \tbVA{tammalaaraa}
\tbLX{tambiri} \tbHM{} \tbPH{támbírì} \tbPS{n.~} \tbGE{bamboo} \tbOP{tambie} \tbTP{tambiiree} 
\tbLX{tambogi} \tbHM{} \tbPH{tàmbògí} \tbPS{n.~} \tbGE{cave} \tbOP{tambogiri} \tbTP{tambogiree} 
\tbLX{tambɔge} \tbHM{} \tbPH{tàmbɔ́gɪ̀} \tbPS{n.~} \tbGE{ditch or pit created by digging earth to build.(usually near the house)} \tbOP{tambɔgere} \tbTP{tambɔgerɛɛ} 
\tbLX{tammalaaraa} \tbHM{} \tbPH{támmáláåráá} \tbPS{n.~} \tbGE{} \tbVA{tambalaaraa}
\tbLX{tammiri} \tbHM{} \tbPH{tàmmírì} \tbPS{n.~} \tbGE{bowstring} \tbOP{tammie} \tbTP{tammiiree} 
\tbLX{tammo} \tbHM{} \tbPH{támmʊ́} \tbPS{n.~} \tbGE{bow} \tbOP{tama} \tbTP{tamɛɛ/tannɛɛ} 
\tbLX{tamogilo} \tbHM{} \tbPH{támógíló} \tbPS{adj.~} \tbGE{toothless} \tbTP{tamogilee} \tbVA{tomogilo}
\tbLX{tamogilo} \tbHM{} \tbPH{támógíló} \tbPS{n.~} \tbGE{one who is toothless} \tbTP{tamogilee} 
\tbLX{tamoni} \tbHM{} \tbPH{tàmónì} \tbPS{n.~} \tbGE{toothless gum} \tbOP{tamonni} \tbTP{tamonnee} \tbVA{tamoŋi, tomoŋi}
\tbLX{tampɛle} \tbHM{} \tbPH{tàmpɛ́lɪ́} \tbPS{n.~} \tbGE{a hoe blade size of a slice/slab of earth (usually cut from soft ground with a hoe)} \tbOP{tampɛlɛ} \tbTP{tampɛlɛɛ} 
\tbLX{tampɛloŋ} \tbHM{} \tbPH{tàmpɛ̀lʊ̀ŋ} \tbPS{n.~} \tbGE{ashes} \tbTP{tampɛllɛɛ} 
\tbLX{tampoli} \tbHM{} \tbPH{tàmpólí} \tbPS{n.~} \tbGE{top end of a bow} \tbOP{tampolo} \tbTP{tampolee} 
\tbLX{tampuori} \tbHM{} \tbPH{tampuori} \tbPS{n.~} \tbGE{rubbish or garbage dump} \tbOP{tampoe} \tbTP{tampuoree} 
\tbLX{tandaa} \tbHM{} \tbPH{tàndàá} \tbPS{n.~} \tbGE{bow stick} \tbOP{tandaare} \tbTP{tandarɛɛ} 
\tbLX{tandagere} \tbHM{} \tbPH{tàndàgɪ́rɪ́} \tbPS{n.~} \tbGE{piece of hard dried mud} \tbOP{tandaga} \tbTP{tandagɛɛ} 
\tbLX{tandɛre} \tbHM{} \tbPH{tàndɛ́rɪ̀} \tbPS{n.~} \tbGE{bow stick} \tbOP{tandɛɛ} \tbTP{tandɛrɛɛ} 
\tbLX{tanfaraŋ} \tbHM{} \tbPH{tánfáráŋ} \tbPS{n.~} \tbGE{a group farming agreement in which the members alternate farms} 
\tbLX{tanfoɔlaa} \tbHM{} \tbPH{tânfʊ́ɔ́làà} \tbPS{n.~} \tbGE{valley between hills} \tbOP{tanfoɔle} \tbTP{tanfoɔlɛɛ} 
\tbLX{tangyugi} \tbHM{} \tbPH{tándzúgí} \tbPS{n.~} \tbGE{in the habit of being promiscuous e.g male goat} \tbOP{tangyugri} \tbTP{tangyugree} \tbVA{tangyuu}
\tbLX{tangyuu} \tbHM{} \tbPH{tágyùú} \tbPS{n.~} \tbGE{in the habit of being promiscuous e.g (male goat)} \tbVA{tangyugi}
\tbLX{tanlori} \tbHM{} \tbPH{tánlórí} \tbPS{n.~} \tbGE{miniature bow, children's bow} \tbOP{ tanloe} \tbTP{tanloree} 
\tbLX{tanne} \tbHM{} \tbPH{tànnɪ̀} \tbPS{v.~} \tbSN \tbGE{to thunder} \tbSN \tbGE{to shout at, to scold} \tbPD{tannɛɛ, tanna, tanna, tannema, tannaa, tannema } 
\tbLX{tannetaa} \tbHM{} \tbPH{tànnɪ̀táá} \tbPS{n.~} \tbGE{quarrel} 
\tbLX{tannoɔ} \tbHM{} \tbPH{tǎnnʊ́ɔ̀} \tbPS{n.~} \tbGE{a small, dark, henlike bird that lives in highlands} \tbOP{tannoore} \tbTP{tannoorɛɛ} 
\tbLX{tantammonaa} \tbHM{} \tbPH{tántámmʊ́náá} \tbPS{n.~} \tbGE{a type of red and blac canary} \tbOP{tantammonne} \tbTP{tantammonnɛɛ} 
\tbLX{tantammonaa} \tbHM{} \tbPH{tántámmʊ́náà} \tbPS{n.~} \tbGE{red weaver bird} \tbOP{tantammonne} \tbTP{tantammonɛɛ} 
\tbLX{tantaŋ} \tbHM{} \tbPH{tàntàŋ} \tbPS{adv.~} \tbGE{immediately} 
\tbLX{tanzuŋ} \tbHM{} \tbPH{tànzûŋ} \tbPS{n.~} \tbGE{carrying pad} \tbOP{tanzunni} \tbTP{tanzunnee} \tbVA{tasere}
\tbLX{tanyini} \tbHM{} \tbPH{tànyíní} \tbPS{n.~} \tbGE{} \tbOP{tanyinni} \tbTP{tanyinee} \tbVA{tamanyini}
\tbLX{taŋa} \tbHM{} \tbPH{táŋà} \tbPS{n.~} \tbGE{mountain, hill} \tbOP{tanne} \tbTP{tannɛɛ} 
\tbLX{taŋaa} \tbHM{} \tbPH{táŋàà} \tbPS{n.~} \tbSN \tbGE{shea tree} \tbSN \tbGE{shea fruit} \tbOP{taŋema} \tbTP{taŋemɛɛ} 
\tbLX{taŋaaweɛlɛ} \tbHM{} \tbPH{táŋááwɪ̀ɛ̀lɛ́} \tbPS{n.~} \tbGE{mistletoe of shea tree} \tbSG{taŋaaweɛloo} \tbTP{taŋaaweɛlɛɛ} 
\tbLX{taŋe} \tbHM{} \tbPH{tâŋ} \tbPS{v.~} \tbGE{to keep silent, to refuse to respond (to a call)} \tbPD{taŋɛɛ, tanna, tanna, tanema, tannaa, - } 
\tbLX{taŋgara} \tbHM{} \tbPH{táŋgárá} \tbPS{n.~} \tbGE{telegraph} \tbOP{taŋgarre} \tbTP{taŋgarɛɛ/taŋgarrɛɛ} 
\tbLX{taŋgɔɔtuo} \tbHM{} \tbPH{tàŋgɔ̀ɔ̀túó} \tbPS{n.~} \tbGE{a kind of tree} \tbOP{taŋgɔtuuri} \tbTP{taŋgɔɔtuoree} 
\tbLX{taŋgboŋi} \tbHM{} \tbPH{tàŋgbôŋ} \tbPS{n.~} \tbGE{a type of short tree with broad leaves and fibrous bark} \tbOP{taŋgbonni} \tbTP{taŋgbonnee} 
\tbLX{taŋkama} \tbHM{} \tbPH{tàŋkàmà} \tbPS{n.~} \tbGE{vanity} 
\tbLX{tapolae} \tbHM{} \tbPH{tápʊ́láɪ̀} \tbPS{n.~} \tbGE{touch-light} \tbOP{tapolaɛ} \tbTP{tapolɛɛrɛɛ} \tbVA{tapolɛɛ}
\tbLX{tapolii} \tbHM{} \tbPH{tápólíí} \tbPS{n.~} \tbGE{tarpaulin} \tbOP{tapolli} \tbTP{tapollee} 
\tbLX{taporo} \tbHM{} \tbPH{tàpòrò} \tbPS{n.~} \tbGE{brain (matter)} \tbOP{taporri} \tbTP{taporee} \tbVA{kyaporo}
\tbLX{tapɔgefa᷉a} \tbHM{} \tbPH{tápɔ́gɪ́fà᷉à᷉} \tbPS{v.~} \tbGE{finish very quickly} \tbPD{-, -, -, -, -, - } 
\tbLX{taraa} \tbHM{} \tbPH{tàràà} \tbPS{adj.~} \tbGE{flat and thick} 
\tbLX{taramaŋ} \tbHM{} \tbPH{tàràmàŋ} \tbPS{adv.~} \tbGE{blindingly} 
\tbLX{tare} \tbHM{} \tbPH{tàrì} \tbPS{v.~} \tbSN \tbGE{to set sitting right on top of something e.g. pot on fire} \tbSN \tbGE{to face directly e.g. problem} \tbPD{tarɛɛ, tara, tara, tareba, taraa, - } 
\tbLX{taremana} \tbHM{} \tbPH{tàrɪ̀mànà} \tbPS{adv.~} \tbGE{completely blind} 
\tbLX{tasere} \tbHM{} \tbPH{tàsɪ́rɪ́} \tbPS{n.~} \tbGE{see tanzuŋi; a pad put on the head to carry load} \tbOP{taseɛ} \tbTP{taserɛɛ} \tbVA{tanzuŋ}
\tbLX{tasɔle} \tbHM{} \tbPH{tásɔ́lɪ́} \tbPS{n.~} \tbGE{a type of lizard with smooth pink skin} \tbOP{tasɔlɔ} \tbTP{tasɔlɛɛ} 
\tbLX{tata} \tbHM{} \tbPH{tátà} \tbPS{n.~} \tbGE{boastfulness, the act of attempting the impossible} 
\tbLX{tata yɛlɛ} \tbHM{} \tbPH{tátà yɛ́lɛ́} \tbPS{n.~} \tbGE{boastful behaviour or claim} 
\tbLX{tateŋɛ} \tbHM{} \tbPH{tátɪ́ŋɛ́} \tbPS{n.~} \tbGE{distant country} \tbOP{tatenne} \tbTP{tatennɛɛ} 
\tbLX{tatoɔroŋ} \tbHM{} \tbPH{tàtʊ́ɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{carelessness} 
\tbLX{taworo} \tbHM{} \tbPH{táwòró} \tbPS{n.~} \tbGE{towel} \tbOP{taworri} \tbTP{taworee} 
\tbLX{tayeŋine} \tbHM{} \tbPH{táyéŋíné} \tbPS{n.~} \tbGE{the state of trembling e.g. from severe hunger} \tbOP{tayenni} \tbTP{tayennee} 
\tbLX{te} \tbHM{} \tbPH{tɪ̀} \tbPS{pronoun.~} \tbSN \tbGE{we (first person plural)} \tbSN \tbGE{us (first person plural)} 
\tbLX{te} \tbHM{} \tbPH{té} \tbPS{adv.~} \tbGE{really} \tbVA{to}
\tbLX{te᷉e} \tbHM{} \tbPH{tɪ́᷉ɪ́᷉} \tbPS{n.~} \tbSN \tbGE{medicine} \tbXV{O beɛrɛ la kyɛ ba sage na nyu te᷉e.} \tbXE{She is sick but refuses to take medicine.} \tbSN \tbGE{solution} \tbXV{A pɔge taa la a bie ŋa te᷉e, o maŋ wa de o la ka o ba la kono.} \tbXE{The woman has some solution for this child, because any time she picks him up, he stops crying.} \tbSN \tbGE{poison} \tbXV{Ba de la gyaŋgbonni te᷉e ko ne a yiri gyaŋgbonni zaa.} \tbXE{The used mouse poison to kill all the mice in the house.} \tbSG{teenoo} \tbTP{teenɛɛ} 
\tbLX{teebo} \tbHM{} \tbPH{tɪ́ɪ́bʊ́} \tbPS{n.~} \tbGE{fetish} \tbOP{tebɛ} \tbTP{teerɛɛ} 
\tbLX{te᷉eda} \tbHM{} \tbPH{t̀ɪ̀᷉ɪ̀᷉dà} \tbPS{n.~} \tbGE{the act of buying medicine} \tbXV{A dɔɔ ŋa ba nɔŋe asibiti gaabo, te᷉eda ka o dɛlle.} \tbXE{This man does not like going to the hospital; he relies on buying medicine.} 
\tbLX{te᷉e-daare} \tbHM{} \tbPH{tɪ́᷉ɪ́᷉-dáárɪ́} \tbPS{n.~} \tbGE{bought medicine} 
\tbLX{te᷉e-die} \tbHM{} \tbPH{tɪ́᷉ɪ́᷉-dìé} \tbPS{n.~} \tbSN \tbGE{clinic, hospital} \tbSN \tbGE{pharmacy} \tbOP{te᷉e-deri} \tbTP{te᷉e-deree} 
\tbLX{te᷉e-doge} \tbHM{} \tbPH{tɪ́᷉ɪ́᷉-dʊ́gɪ̀} \tbPS{n.~} \tbGE{medicine pot} \tbOP{te᷉e-dogere} \tbTP{te᷉e-dogerɛɛ} 
\tbLX{tee᷉-erɛ} \tbHM{} \tbPH{tɪ́ɪ́᷉-ɪ́rɛ́} \tbPS{n.~} \tbGE{medicine man, traditional healer} \tbOP{tee᷉-erebɛ} \tbTP{tee᷉-erebɛɛ} 
\tbLX{te᷉eko᷉ɔ} \tbHM{} \tbPH{tɪ́᷈ɪ́᷉kʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{syrup (medicine)} \tbTP{te᷉ekoɔnɛɛ} 
\tbLX{te᷉ekoɔre-die} \tbHM{} \tbPH{tɪ̀᷈ɪ́᷈kʊ̀ɔ́r-díé} \tbPS{n.~} \tbGE{pharmacy} \tbOP{te᷉e᷉koɔr-deri} \tbTP{te᷉e᷉koɔr-deree} 
\tbLX{teelɛ} \tbHM{} \tbPH{tɪ́ɪ́lɛ́} \tbPS{n.~} \tbGE{sewing machine} \tbOP{teelɛre} \tbTP{teelɛrɛɛ} 
\tbLX{teene} \tbHM{} \tbPH{tɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbGE{improve upon} \tbPD{teenɛɛ, teenɛ, teenɛ, teenemɛ, teenaa, - } 
\tbLX{teene} \tbHM{} \tbPH{tɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbGE{delay, waste time} \tbPD{teenɛɛ, teenɛ, teenɛ, teenemɛ, teenaa, - } 
\tbLX{teene} \tbHM{} \tbPH{tɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbGE{to be the spokesman for a suitor in courtship} \tbPD{teenɛɛ, teenɛ, teenɛ, teenemɛ, teenaa, - } 
\tbLX{teere} \tbHM{} \tbPH{tɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to rub against} \tbSN \tbGE{to iron clothes} \tbPD{teerɛɛ, teerɛ, teerɛ, teerebɛ, teeraa, - } 
\tbLX{teɛ} \tbHM{} \tbPH{tɪ̀ɛ̀} \tbPS{v.~} \tbGE{to support} \tbPD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛ} \tbHM{} \tbPH{tɪ̀ɛ̀} \tbPS{v.~} \tbGE{to shoot with a bow} \tbPD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛ} \tbHM{} \tbPH{tɪ̀ɛ̀} \tbPS{v.~} \tbGE{to gamble} \tbPD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛ} \tbHM{} \tbPH{tɪ́ɛ́} \tbPS{n.~} \tbGE{scabbard} \tbOP{teere} \tbTP{teerɛɛ} 
\tbLX{teɛ} \tbHM{} \tbPH{tɪ́ɛ̀} \tbPS{v.~} \tbGE{to kick} \tbPD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛ} \tbHM{} \tbPH{tɪ́ɛ̀} \tbPS{v.~} \tbGE{to unset, to release (snare)} \tbPD{teɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛ} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{tree} \tbOP{teere} \tbTP{teerɛɛ} 
\tbLX{teɛ} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{a disease that causes a rash (usually on the legs)} \tbOP{teere} \tbTP{teerɛɛ} 
\tbLX{tẽɛ} \tbHM{} \tbPH{tɪ́᷉ɛ́} \tbPS{v.~} \tbSN \tbGE{to stretch out e.g. legs} \tbXV{A baala zeŋe dɛlle la a gado kyɛ te᷉ɛ o gbɛɛ.} \tbXE{The patient sat leaning against the bed with his legs stretched out.} \tbSN \tbGE{to point to} \tbXV{A pɔgelee da to᷉ɔ te᷉ɛ wuli la a nannyige.} \tbXE{The girl was able to point to the thief.} \tbSN \tbGE{to hold out to give} \tbXV{Ba de la a bondirii te᷉ɛ a pɔge ka o zagere.} \tbXE{They held out  the food to the woman, but she refused.} \tbPD{teɛ᷈ɛ, teɛnɛ, teɛnɛ, teɛnemɛ, teɛnaa, teɛnemɛ } 
\tbLX{te᷈ɛkɔɔloŋ} \tbHM{} \tbPH{tɪ̀᷉ɛ́᷉kɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{beard} \tbOP{te᷈ɛkɔɔlɔ} \tbTP{te᷈ɛkɔɔlɛɛ} 
\tbLX{teɛne} \tbHM{} \tbPH{tɪ̀ɛ́nɪ́} \tbPS{n.~} \tbGE{chin} \tbOP{teɛmɛ} \tbTP{teɛnɛɛ} 
\tbLX{teɛne} \tbHM{} \tbPH{tɪ̀ɛ̀nɪ̀} \tbPS{v.~} \tbGE{to stuff e.g. pillow} \tbPD{teɛnɛɛ, teɛnɛ, teɛnɛ, teɛnemɛ, teɛnaa, - } 
\tbLX{teɛre} \tbHM{} \tbPH{tɪ́ɛ́rɪ̀} \tbPS{v.~} \tbSN \tbGE{to think} \tbSN \tbGE{to remind} \tbPD{teɛrɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛre} \tbHM{} \tbPH{tɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to feel for texture (smoothness of flour, swelling)} \tbPD{teɛrɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, teɛrebɛ } 
\tbLX{teɛre tɔ} \tbHM{} \tbPH{tɪ́ɛ́rɪ́ tɔ̀} \tbPS{} \tbGE{hope} \tbXV{Ba de la a baala gaa ne asibiti kyɛ teɛre toɔbo kyebe.} \tbXE{The patient has been taken to the hospital but there is no hope.} 
\tbLX{tege} \tbHM{} \tbPH{tɪ́gɪ́} \tbPS{v.~} \tbGE{to be satisfied} \tbPD{tegɛɛ, tegrɛ, tegrɛ, tegrebɛ, tegraa, - } 
\tbLX{tegele} \tbHM{} \tbPH{tɪ̀glɪ̀} \tbPS{v.~} \tbGE{to shorten} \tbPD{tegelɛɛ, tegelɛ, tegelɛ, tegelebɛ, tegelaa, teglebɛ } 
\tbLX{tegelɛŋ} \tbHM{} \tbPH{tɪ́glɛ́ŋ} \tbPS{adv.~} \tbSN \tbGE{little, small (amount of liquid)} \tbSN \tbGE{short (length, time)} 
\tbLX{tegeroŋ} \tbHM{} \tbPH{tɪ́grʊ́ŋ} \tbPS{adj.~} \tbGE{heavy (weight)} \tbOP{tegerɛ} 
\tbLX{tegeroŋ} \tbHM{} \tbPH{tɪ́grʊ́ŋ} \tbPS{n.~} \tbGE{a heavy thing} \tbOP{tegerɛ} 
\tbLX{tegitegi} \tbHM{} \tbPH{tégtég} \tbPS{adv.~} \tbGE{always} 
\tbLX{tegitegi lɛ} \tbHM{} \tbPH{tégtéglɛ̀} \tbPS{adj.~} \tbGE{always like that, usually like that} 
\tbLX{tegyeerɛ} \tbHM{} \tbPH{tɪ̀dzɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{woodpecker} \tbOP{tegyeere} \tbTP{tegyeerɛɛ} \tbVA{tekoliŋkuoraa}
\tbLX{tekoliŋkuoraa} \tbHM{} \tbPH{tɪ̀kólíŋkúóráá} \tbPS{n.~} \tbGE{woodpecker} \tbOP{tekoliŋkuori} \tbTP{tekoliŋkuoree} \tbVA{tegyeerɛ, dakyeɛrɛko}
\tbLX{tekonee} \tbHM{} \tbPH{tɪ́kʊ́nɪ́ɪ́} \tbPS{adj.~} \tbGE{tiny} 
\tbLX{tekoɔ᷉} \tbHM{} \tbPH{tɪ́kʋ̃́ɔ́} \tbPS{n.~} \tbGE{sap} \tbVA{tesunni}
\tbLX{tekpiri} \tbHM{} \tbPH{tɪ̀kpírì} \tbPS{n.~} \tbGE{stump} \tbSG{tekpiruu} \tbTP{tekpiree} 
\tbLX{tekyɛ} \tbHM{} \tbPH{tɪ́ʧɛ̀} \tbPS{n.~} \tbGE{teacher} \tbOP{tekyɛre} \tbTP{tekyɛrɛɛ} 
\tbLX{tekyɛ} \tbHM{} \tbPH{tɪ̀ʧɛ̀} \tbPS{n.~} \tbGE{the act of felling trees} 
\tbLX{telare} \tbHM{} \tbPH{tɪ́làrɪ̀} \tbPS{adv.~} \tbGE{too much} 
\tbLX{teli} \tbHM{} \tbPH{télí} \tbPS{v.~} \tbGE{to be in an upright position e.g. head} \tbPD{telee, tele, tele, telibe, telaa, - } 
\tbLX{teloɔraa} \tbHM{} \tbPH{tɪ̀lʊ́ɔ́ráá} \tbPS{n.~} \tbGE{animal that lives in trees e.g. squirrel} \tbOP{teloɔre} \tbTP{teloɔrɛɛ} 
\tbLX{tembaa} \tbHM{} \tbPH{tɪ̌mbåá} \tbPS{n.~} \tbGE{a type of black ant with a painful sting} \tbOP{tembaare} \tbTP{tembaarɛɛ} 
\tbLX{tembeene} \tbHM{} \tbPH{tɪ̂mbɪ́ɪ́nɪ́} \tbPS{n.~} \tbGE{late night} \tbOP{tembeemɛ} \tbTP{tembeemɛɛ} 
\tbLX{tembie} \tbHM{} \tbPH{tɪ̀mbíé} \tbPS{n.~} \tbGE{citizen} \tbOP{tembiiri} \tbTP{tembiiree} 
\tbLX{tembimeŋɛ} \tbHM{} \tbPH{tɪ̀mbìmɪ́ŋɛ́} \tbPS{n.~} \tbGE{patriot} \tbOP{tembimenne} \tbTP{tembimennɛɛ} 
\tbLX{temm} \tbHM{} \tbPH{témm} \tbPS{adv.~} \tbGE{alone; lonely} 
\tbLX{tempoɔsori} \tbHM{} \tbPH{tîmpʊ́ɔ́sórì} \tbPS{n.~} \tbGE{street} \tbOP{tempoɔsoe} \tbTP{tempoɔsoree} 
\tbLX{tendaa} \tbHM{} \tbPH{tɪ̀ndàá} \tbPS{adv.~} \tbGE{country wide} 
\tbLX{tendaana} \tbHM{} \tbPH{tìndááná} \tbPS{n.~} \tbGE{landowner} \tbOP{tendaama} \tbTP{tendaamɛɛ} 
\tbLX{tenee} \tbHM{} \tbPH{tɪ̀nɪ́ɪ̀} \tbPS{poss. pron.~} \tbGE{we} 
\tbLX{tenfaa} \tbHM{} \tbPH{tɪ̀nfáá} \tbPS{n.~} \tbSN \tbGE{bad country, bad land} \tbSN \tbGE{hell} \tbOP{tenfaare} \tbTP{tenfaarɛɛ} 
\tbLX{tensa᷉a} \tbHM{} \tbPH{tɪ̀nsàà᷉} \tbPS{n.~} \tbGE{disruption of community, country} 
\tbLX{tensaana} \tbHM{} \tbPH{tɪ̌nsáánà} \tbPS{n.~} \tbGE{alien land, strange country} \tbOP{tensaama} \tbTP{tensaanɛɛ} 
\tbLX{tensaana} \tbHM{} \tbPH{tɪ̀nsááná} \tbPS{n.~} \tbGE{bandit} \tbOP{tensaanema} \tbTP{tensaamɛɛ} 
\tbLX{tensane} \tbHM{} \tbPH{tɪ̀nsánɪ̀} \tbPS{n.~} \tbGE{state of owing the gods of the land} \tbOP{tensama} \tbTP{tensannɛɛ} 
\tbLX{tensɛre} \tbHM{} \tbPH{tɪ́sɛ́rɪ̀} \tbPS{n.~} \tbGE{tradition} \tbVA{teŋko᷉ɔ}
\tbLX{tensoba} \tbHM{} \tbPH{tɪ̀nsʊ̀bá} \tbPS{n.~} \tbGE{landowner} \tbOP{tendemee} 
\tbLX{tensogezɔnne} \tbHM{} \tbPH{tɪ́sʊ́gɪ́zɔ́nnɪ́} \tbPS{n.~} \tbGE{night blindness} 
\tbLX{tensogɔ} \tbHM{} \tbPH{tɪ̂nsʊ́gɔ́} \tbPS{n.~} \tbGE{night} \tbOP{tensogere} \tbTP{tensogerɛɛ} \tbVA{tensoga, yensogɔ}
\tbLX{tensunni} \tbHM{} \tbPH{tɪ́nsúnní} \tbPS{n.~} \tbSN \tbGE{muddy ground} \tbSN \tbGE{fertile land} \tbOP{tensumo} 
\tbLX{tentoloŋ} \tbHM{} \tbPH{tɪ̂ntʊ́lʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{hot ground} \tbSN \tbGE{hot seaon} 
\tbLX{tentɔle} \tbHM{} \tbPH{tɪ́ntɔ́lɪ́} \tbPS{n.~} \tbGE{under pants} \tbOP{tentɔlɔ} \tbTP{tentɔlɛɛ} 
\tbLX{tentɔlvaa} \tbHM{} \tbPH{tɪ́ntɔ̀lváà} \tbPS{n.~} \tbGE{triangular underwear} \tbOP{tentɔlvaare} \tbTP{tentɔlvaarɛɛ} 
\tbLX{tentɔɔre} \tbHM{} \tbPH{tɪ̀ntɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{distant country, town, or village} \tbOP{tentɔɔbɔ} 
\tbLX{tenvelaa} \tbHM{} \tbPH{tɪ̀nvɪ̀làá} \tbPS{n.~} \tbSN \tbGE{good country, good land} \tbSN \tbGE{heaven} \tbOP{tenveɛle} 
\tbLX{teŋa} \tbHM{} \tbPH{tɪ́ŋá} \tbPS{n.~} \tbSN \tbGE{ground} \tbSN \tbGE{town, village} \tbSN \tbGE{country} \tbVA{teŋɛ}
\tbLX{teŋe} \tbHM{} \tbPH{tɪ̀ŋ} \tbPS{v.~} \tbGE{} \tbPD{teŋɛɛ, teŋenɛ, teŋenɛ, teŋenemɛ, teŋenaa, - } 
\tbLX{teŋɛ} \tbHM{} \tbPH{tɪ́ŋɛ́} \tbPS{adj.~} \tbGE{down} 
\tbLX{teŋɛ} \tbHM{} \tbPH{tíŋɛ́} \tbPS{n.~} \tbSN \tbGE{ground} \tbSN \tbGE{town, village} \tbSN \tbGE{country} \tbOP{tenne} \tbTP{tennɛɛ} \tbVA{teŋa}
\tbLX{teŋɛsoga} \tbHM{} \tbPH{tɪ́ŋɛ́sʊ́gá} \tbPS{n.~} \tbGE{floor} \tbOP{teŋɛsogre} \tbVA{teŋɛsogɔ}
\tbLX{teŋɛzu} \tbHM{} \tbPH{tɪ́ŋɛ́zù} \tbPS{n.~} \tbGE{world} \tbOP{teŋɛzuri} \tbTP{teŋɛzuree} 
\tbLX{teŋgane} \tbHM{} \tbPH{tǐŋgánɪ̀} \tbPS{n.~} \tbSN \tbGE{top layer of soil} \tbSN \tbGE{god of the land} \tbOP{teŋgama} \tbTP{teŋgamɛɛ} 
\tbLX{teŋganezu waloŋ zannoo} \tbHM{} \tbPH{tɪ̀ŋgánzu wààlʊ̀ŋ zánnʊʊ} \tbPS{n.~} \tbGE{geography} 
\tbLX{teŋgbigiri} \tbHM{} \tbPH{tɪ́ŋɛ́ gbígrúú} \tbPS{n.~} \tbGE{earthquake} 
\tbLX{teŋko᷉ɔ᷉} \tbHM{} \tbPH{tɪ́ŋkʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{tradition} \tbTP{teŋkoɔnɛɛ} \tbVA{tensɛre}
\tbLX{teŋkpoŋ} \tbHM{} \tbPH{tɪ̀ŋkpôŋ} \tbPS{n.~} \tbGE{town, city} \tbOP{teŋkponni} \tbTP{teŋkponnee} 
\tbLX{teraale} \tbHM{} \tbPH{tìrààlɪ́} \tbPS{n.~} \tbGE{perfume} \tbOP{teraalere} \tbTP{terelerɛɛ} 
\tbLX{tere} \tbHM{} \tbPH{téré} \tbPS{adj.~} \tbGE{only} 
\tbLX{tere} \tbHM{} \tbPH{tɪ̀rɪ́} \tbPS{n.~} \tbGE{spoon} \tbOP{teɛ} \tbTP{terɛɛ} 
\tbLX{tere} \tbHM{} \tbPH{tɪ̀rɪ̀} \tbPS{v.~} \tbGE{to give, to send} \tbPD{terɛɛ, terɛ, terɛ, terebɛ, teraa, - } 
\tbLX{teree} \tbHM{} \tbPH{téréé} \tbPS{n.~} \tbGE{a statement made for comments} \tbXV{A koŋkono lɔɔ la teree ko a kotuo deme.} \tbXE{The dirge singer has presented a riddle to the bereaved family.} \tbOP{terri} \tbTP{terree} 
\tbLX{teree-teree} \tbHM{} \tbPH{térée-téréé} \tbPS{adj.~} \tbGE{spread out and roundish at the end, cone-shaped} 
\tbLX{terembelli} \tbHM{} \tbPH{térémbéllí} \tbPS{adv.~} \tbGE{straight and upright} 
\tbLX{tere noɔre} \tbHM{} \tbPH{tɪ̀rɪ̀ nʊ́ɔ́rɪ̀} \tbPS{v.~} \tbGE{to leave a message} \tbPD{-, -, -, -, -, - } \tbVA{bare noɔre}
\tbLX{tere sori} \tbHM{} \tbPH{tɪ́rɪ́ sórì} \tbPS{v.~} \tbGE{allow, permit} \tbPD{-, -, -, -, -, - } 
\tbLX{teretere} \tbHM{} \tbPH{tɪ̀rɪ̀tɪ́rɪ́} \tbPS{v.~} \tbGE{to crush with feet, to trample} \tbPD{tereterɛɛ, tereterɛ, tereterebɛ, tereteraa, -, - } 
\tbLX{terri} \tbHM{} \tbPH{térrì} \tbPS{v.~} \tbGE{to smooth out the end e.g. pestle} \tbPD{terree, terre, terre, teribe, terraa, - } 
\tbLX{tesogɔ} \tbHM{} \tbPH{tɪ̀sʊ́gɔ́} \tbPS{n.~} \tbGE{trunk (of tree)} \tbOP{tesogere} \tbTP{tesogerɛɛ} \tbVA{tesoga}
\tbLX{tesunni} \tbHM{} \tbPH{tèsúnnì} \tbPS{n.~} \tbGE{sap} \tbSG{tesunnuu} \tbTP{tesunnee} \tbVA{tekoɔ᷉}
\tbLX{tetebaa} \tbHM{} \tbPH{tɪ́tɪ́báá} \tbPS{n.~} \tbGE{stinger; the tip of the tail of a bee usually used to sting} \tbOP{teterre} \tbTP{teterɛɛ} 
\tbLX{tetere} \tbHM{} \tbPH{tɪ́tɪ̀rɪ̀} \tbPS{adj.~} \tbGE{small in quantity} 
\tbLX{tevaa} \tbHM{} \tbPH{tɪ̀váà} \tbPS{adj.~} \tbGE{green} \tbOP{tevaare} \tbTP{tevaarɛɛ} 
\tbLX{tewɔne} \tbHM{} \tbPH{tɪ̀wɔ́nɪ́} \tbPS{n.~} \tbSN \tbGE{berry} \tbSN \tbGE{fruit of a tree} \tbOP{tewɔmɔ} \tbTP{tewɔnnɛɛ} 
\tbLX{tewɔŋko᷈ɔ᷉} \tbHM{} \tbPH{tɪ́wɔ́ŋkʊ́ɔ́} \tbPS{n.~} \tbGE{juice} \tbTP{tewɔŋkoɔnɛɛ} 
\tbLX{tɛ} \tbHM{} \tbPH{tɛ́} \tbPS{v.~} \tbGE{to spread in an orderly manner} \tbXV{Dasɛrɛɛ ka Dagaaba maŋ tɛ ne ba die kyɛ lɔɔ.} \tbXE{The Dagaaba arrange small sticks on planks of wood and on which they put local mortar to roof a building.} \tbPD{tɛɛ, teɛrɛ, teɛrɛ, teɛrebɛ, teɛraa, - } 
\tbLX{tɛbaa} \tbHM{} \tbPH{tɛ́báá} \tbPS{n.~} \tbGE{snare} \tbOP{tɛrre} \tbTP{tɛrrɛɛ} 
\tbLX{tɛɛse} \tbHM{} \tbPH{tɛ́ɛ́sɪ́} \tbPS{n.~} \tbGE{test} \tbOP{tɛɛsɪrɪ} \tbTP{tɛɛsɪrɛɛ} 
\tbLX{tɛɛtɛɛ} \tbHM{} \tbPH{tɛ́ɛ́tɛ̀ɛ̀} \tbPS{adj.~} \tbGE{different} 
\tbLX{tɛge} \tbHM{} \tbPH{tɛ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to exchange, to barter} \tbSN \tbGE{to change (action, form)} \tbSN \tbGE{to stop doing something suddenly and for a short time} \tbPD{-, -, -, -, -, - } 
\tbLX{tɛge} \tbHM{} \tbPH{tɛ̀gɪ́} \tbPS{n.~} \tbGE{meaning} \tbTP{tɛgrɛɛ} \tbVA{tɛgɛ}
\tbLX{tɛgele} \tbHM{} \tbPH{tɛ́glɪ̀} \tbPS{v.~} \tbSN \tbGE{to place a container with the bottom on the ground} \tbSN \tbGE{to collect in a small pool e.g water} \tbPD{tɛgelɛɛ, tɛgelɛ, tɛgelɛ, tɛgelebɛ, tɛgelaa, - } 
\tbLX{tɛgere} \tbHM{} \tbPH{tɛ̀grɪ̀} \tbPS{v.~} \tbGE{to apply the sap of a leaf to the body (local method of tattooing)} \tbPD{tɛgerɛɛ, tɛgerɛ, tɛgerɛ, tɛgerebɛ, tɛgraa, - } 
\tbLX{tɛgɛ} \tbHM{} \tbPH{tɛ́gɛ́} \tbPS{n.~} \tbGE{meaning} \tbVA{tɛge}
\tbLX{tɛgɛtɛgɛ} \tbHM{} \tbPH{tɛ́gɛ́tɛ́gɛ́} \tbPS{adv.~} \tbGE{little e.g. liquid} 
\tbLX{tɛlɛntɛlɛŋ} \tbHM{} \tbPH{tɛ̀lɛ̀ntɛ́lɛ́ŋ} \tbPS{adv.~} \tbGE{sprawled out} 
\tbLX{tɛll} \tbHM{} \tbPH{tɛ̀ll} \tbPS{adj.~} \tbGE{long, straight, and rigid} 
\tbLX{tɛmbiri} \tbHM{} \tbPH{tɛ̌mbìrí} \tbPS{n.~} \tbSN \tbGE{white ants (termites)} \tbSN \tbGE{the habitat of these termites} \tbOP{tɛmbie} \tbTP{tɛmbiiree} 
\tbLX{tɛmbogi} \tbHM{} \tbPH{tɛ̀mbògí} \tbPS{n.~} \tbGE{ditch where mortar is prepared} \tbOP{tɛmbogiri} \tbTP{tɛmbogiree} 
\tbLX{tɛmburo} \tbHM{} \tbPH{tɛ̀mbùró} \tbPS{n.~} \tbGE{dry soil} \tbTP{tɛmburee} 
\tbLX{tɛmpelaa} \tbHM{} \tbPH{tɛ̌mpɪ̀làá} \tbPS{n.~} \tbGE{white soil (used to plaster walls)} \tbOP{tɛmpeɛle} \tbTP{tɛmpeɛlɛɛ} 
\tbLX{tɛmpoɔroŋ} \tbHM{} \tbPH{tɛ̀mpʊ́ɔ́rʊ́ŋ} \tbPS{n.~} \tbGE{fertile soil} \tbOP{tɛmpoɔrɔ} 
\tbLX{tɛne} \tbHM{} \tbPH{tɛ́nɪ́} \tbPS{n.~} \tbSN \tbGE{soil, earth, mud} \tbSN \tbGE{the spirit of a land} \tbTP{tɛnɛɛ} 
\tbLX{tɛnnɛ} \tbHM{} \tbPH{tɛ̌nnɛ̀} \tbPS{n.~} \tbGE{a type of grass usually found in front of houses} \tbOP{tɛnne} \tbTP{tɛnnɛɛ} 
\tbLX{tɛŋ} \tbHM{} \tbPH{tɛ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to sweeten} \tbSN \tbGE{to ferment} \tbPD{tɛŋɛɛ, tɛnɛ, tɛnɛ, tɛnɛ, tɛnemɛ, tɛnaa } 
\tbLX{tɛŋkubo} \tbHM{} \tbPH{tɛ̌ŋkúbò} \tbPS{n.~} \tbGE{gravel} 
\tbLX{tɛre} \tbHM{} \tbPH{tɛ̀rɪ̀} \tbPS{v.~} \tbGE{to strain one's legs} \tbXV{A mɔmoɔre da tɛre la gbɛɛ eŋe taa kyɛ prri wale.} \tbXE{The wrestlers strained their legs and sweated too.} \tbPD{tɛrɛɛ, tɛrɛ, tɛrɛ, tɛrebɛ, tɛraa, tɛrebɛ } 
\tbLX{tɛretɛre} \tbHM{} \tbPH{tɛ̀rɪ̀tɛ́rɪ́} \tbPS{v.~} \tbGE{to move in protest e.g. pampered child} \tbXV{A bie zagere ka kyɛŋe kyɛ maŋ tɛretɛre.} \tbXE{The child has refused to walk up and is moving under protest.} \tbPD{tɛretɛrɛɛ, tɛretɛrɛ, tɛretɛrɛ, tɛretɛrebɛ, tɛretɛraa, - } 
\tbLX{tɛrewɛrɛ} \tbHM{} \tbPH{tɛ́rɪ́wɛ́rɛ́} \tbPS{adv.~} \tbGE{standing with legs bent backwards} 
\tbLX{tɛrɛɛtɛrɛɛ} \tbHM{} \tbPH{tɛ́rɛ́ɛ́tɛ́rɛ́ɛ́} \tbPS{adv.~} \tbGE{flat and even} 
\tbLX{tɛrɛsɛ} \tbHM{} \tbPH{tɛ̀rɛ̀sɛ̀} \tbPS{adv.~} \tbGE{clumsily} \tbXV{O tegɛɛ la a deɛ leɛ e tɛrɛsɛ lɛ o naŋ zeŋ.} \tbXE{He is so full with food that he can hardly move.} 
\tbLX{tɛrre} \tbHM{} \tbPH{tɛ́rrɪ̀} \tbPS{v.~} \tbGE{to level out} \tbPD{tɛrrɛɛ, tɛrrɛ, tɛrrɛ, tɛrebɛ, tɛrraa, - } 
\tbLX{tɛrre} \tbHM{} \tbPH{tɛ́rrɪ̀} \tbPS{v.~} \tbGE{to push a container away along the ground/floor usually in anger} \tbPD{tɛrrɛɛ, tɛrrɛ, tɛrrɛ, tɛrebɛ, tɛrraa, - } 
\tbLX{tɛtɛgere} \tbHM{} \tbPH{tɛ̀tɛ̀gɪ̀rɪ́} \tbPS{n.~} \tbGE{muddy thing} \tbOP{tɛtɛgɛ} 
\tbLX{tɛtɛrewɛrɛ} \tbHM{} \tbPH{tɛ̀tɛ̀rɪ́wɛ̌rɛ́} \tbPS{adv.~} \tbGE{daftly} 
\tbLX{ti} \tbHM{} \tbPH{tì} \tbPS{v.~} \tbGE{to vomit} \tbPD{tiee, tiire, tiire, tiiribe, tiiraa, tiiribe } 
\tbLX{ti} \tbHM{} \tbPH{tí} \tbPS{v.~} \tbGE{to put hand on e.g wall} \tbXV{A baala ti la dankyini lɛnso ka o ba le.} \tbXE{The patient held the wall and that is why s/he did not fall.} \tbPD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\tbLX{ti} \tbHM{} \tbPH{tí} \tbPS{v.~} \tbGE{to support by the use of a walking stick} \tbPD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\tbLX{ti} \tbHM{} \tbPH{tí} \tbPS{v.~} \tbGE{to put up sticks with their tops meeting (making a farm hut)} \tbPD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\tbLX{ti} \tbHM{} \tbPH{tí} \tbPS{v.~} \tbGE{to begin, to start} \tbPD{tie, tiire, tiire, tiiribe, tiiraa, - } 
\tbLX{tigi} \tbHM{} \tbPH{tìgì} \tbPS{v.~} \tbGE{to gather e.g. crowd} \tbPD{tigee, tigre, tigre, tigribe, tigraa, - } 
\tbLX{tigiki} \tbHM{} \tbPH{tìgìkì́} \tbPS{adj.~} \tbGE{dark and deep set (eyes)} \tbVA{tigtig}
\tbLX{tigikpoŋi} \tbHM{} \tbPH{tìgìkpóŋ} \tbPS{n.~} \tbGE{big feast, solemnity} \tbOP{tigikponni} \tbTP{tikponnee} 
\tbLX{tigiri} \tbHM{} \tbPH{tígrí} \tbPS{n.~} \tbGE{feast, festival} \tbOP{tige} \tbTP{tigee} 
\tbLX{tigiri} \tbHM{} \tbPH{tígrì} \tbPS{v.~} \tbGE{to grow fat} \tbPD{tigiree, tigire, tigire, tigiribe, tigiraa, - } 
\tbLX{tigiti} \tbHM{} \tbPH{tígítì} \tbPS{n.~} \tbGE{ticket} \tbOP{tigitiri} \tbTP{tigitiree} 
\tbLX{tigitigi} \tbHM{} \tbPH{tìgìtígí} \tbPS{v.~} \tbGE{to shiver violently e.g. when one has a fever} \tbPD{tigitigee, tigitigre, tigitigre, tigitigribe, tigitigraa, - } 
\tbLX{ti᷉i} \tbHM{} \tbPH{tí᷉í᷉} \tbPS{adv.~} \tbGE{closely and carefully e.g. look, observe} 
\tbLX{tiili} \tbHM{} \tbPH{tíílì} \tbPS{v.~} \tbGE{to place against an upright object e.g a plank against a tree/wall} \tbPD{tiilee, tiile, tiile, tiilibe, tiilaa, - } 
\tbLX{tiiri} \tbHM{} \tbPH{tíírí} \tbPS{n.~} \tbGE{vomit} \tbTP{tiiree} 
\tbLX{tiiri} \tbHM{} \tbPH{tììrì} \tbPS{v.~} \tbGE{to come very close to} \tbPD{tiiree, tiire, tiire, tiiribe, tiiraa, - } 
\tbLX{tiiri} \tbHM{} \tbPH{tììrì} \tbPS{v.~} \tbGE{to be fed up with} \tbPD{tiiree, tiire, tiire, tiiribe, tiiraa, - } 
\tbLX{tiiruŋ} \tbHM{} \tbPH{tíírǔŋ} \tbPS{n.~} \tbSN \tbGE{a type of plant with sour sap used for cooking} \tbSN \tbGE{a type of medicinal herbal concoction} 
\tbLX{tiiruŋ} \tbHM{} \tbPH{tììrǔŋ} \tbPS{adj.~} \tbGE{sour (taste)} 
\tbLX{tilento} \tbHM{} \tbPH{tíléntó} \tbPS{n.~} \tbSN \tbGE{hornbill} \tbSN \tbGE{pelican} \tbOP{tilentori} \tbTP{tilentoree} 
\tbLX{tiŋi} \tbHM{} \tbPH{tíŋ} \tbPS{v.~} \tbGE{to make firm (object fixed in the ground)} \tbPD{tiŋee, tiŋine, tiŋine, tiŋinime, tiŋinaa, - } 
\tbLX{tiŋimeŋ} \tbHM{} \tbPH{tìŋìméŋ} \tbPS{adj.~} \tbGE{dark} 
\tbLX{tiretire} \tbHM{} \tbPH{tírétíré} \tbPS{adv.~} \tbGE{tasteless} 
\tbLX{to} \tbHM{} \tbPH{tʊ́} \tbPS{v.~} \tbGE{to insult} \tbPD{toɛ, toora, toora, tooreba, tooraa, tooreba } 
\tbLX{to} \tbHM{} \tbPH{tó} \tbPS{adv.~} \tbGE{really} \tbXV{A bie ŋa zoro to.} \tbXE{This child runs really fast.} \tbVA{te}
\tbLX{tobaa} \tbHM{} \tbPH{tóbáá} \tbPS{n.~} \tbGE{fox} \tbOP{tobaare} \tbTP{tobaarɛɛ} 
\tbLX{tobie sawɔlɔ} \tbHM{} \tbPH{tóbíè sáwɔ́lɔ́} \tbPS{n.~} \tbGE{fried food made from boabab seed} 
\tbLX{tobiri} \tbHM{} \tbPH{tóbírì} \tbPS{n.~} \tbGE{boabab seed} \tbSG{tobiruu} \tbOP{tobie} \tbTP{tobiiree} 
\tbLX{togi} \tbHM{} \tbPH{tògì} \tbPS{v.~} \tbSN \tbGE{take down e.g. from head, fire} \tbSN \tbGE{to drip} \tbPD{togee, togiro, togiro, togiribo, togiraa, - } 
\tbLX{togifa} \tbHM{} \tbPH{tògífá} \tbPS{n.~} \tbGE{sixpence} \tbOP{togifare} \tbTP{togifarɛɛ} 
\tbLX{togilogo} \tbHM{} \tbPH{tóglógó} \tbPS{adv.~} \tbSN \tbGE{small (space)} \tbXV{Ba boma koŋ kpɛ a die, o deɛ waa la togilogoo lɛ.} \tbXE{The room cannot contain their belongings; it is very small.} \tbSN \tbGE{small and shallow (hole)} \tbXV{Ba tu la a bulee fa᷉a᷉ kyɛ deɛ e boŋkaŋa togilogoo lɛ.} \tbXE{They were unable to dig the well; they just etched a tiny hole.} 
\tbLX{togilogo} \tbHM{} \tbPH{tòglògò} \tbPS{adv.~} \tbGE{deep} \tbXV{Te tu la tambɔge togilogoo kaŋa a na kuire sagere ennɛ.} \tbXE{We dug a deep ditch into which we will dump refuse.} 
\tbLX{togiroo} \tbHM{} \tbPH{tógróó} \tbPS{n.~} \tbGE{raised wooded ground (Southern Dagaare dialect)} \tbOP{togiri} \tbTP{togiree} \tbVA{tuu}
\tbLX{togitogi} \tbHM{} \tbPH{tógtóg} \tbPS{adv.~} \tbGE{never, not at all} \tbXV{N gɛrɛ ka o gaaɛ N ba nyɛ o togitogi.} \tbXE{When I got there, he had left; I never saw him at all.} 
\tbLX{togoko} \tbHM{} \tbPH{tògòkò} \tbPS{adv.~} \tbGE{refusing to listen, hear; not minding} 
\tbLX{tokele} \tbHM{} \tbPH{tòkélé} \tbPS{n.~} \tbGE{hard dawdaw spice seed (not processable)} \tbOP{tokelee} \tbVA{dɔkele, kele}
\tbLX{tokokyile} \tbHM{} \tbPH{tòkóʧìlé} \tbPS{n.~} \tbGE{a bottle-like container made by removing the seeds from the baobab fruit} \tbOP{tokokyilli} \tbTP{tokokyilee} 
\tbLX{tokoroo} \tbHM{} \tbPH{tókóróó} \tbPS{n.~} \tbGE{window} \tbOP{tokorri} \tbTP{tokoree/tokorree} \tbVA{takoroo}
\tbLX{tokparaa} \tbHM{} \tbPH{tòkpáráà} \tbPS{n.~} \tbGE{boabab leaves (vegetable for soup)} \tbTP{tokparɛɛ} 
\tbLX{tolee} \tbHM{} \tbPH{tóléé} \tbPS{n.~} \tbGE{small mortar} \tbOP{tobilii} \tbTP{tobilee} 
\tbLX{tolema} \tbHM{} \tbPH{tʊ́lɪ́má} \tbPS{adv.~} \tbGE{avariciously} 
\tbLX{tolimala} \tbHM{} \tbPH{tòlmálá} \tbPS{n.~} \tbGE{great enjoyment, merrymaking (edibles)} \tbTP{tolimalɛɛ} \tbVA{tolomala}
\tbLX{toliŋka} \tbHM{} \tbPH{tólíŋkà} \tbPS{n.~} \tbGE{a car} \tbOP{toliŋkaɛ} \tbTP{toliŋkarɛɛ} 
\tbLX{toll} \tbHM{} \tbPH{tòll} \tbPS{adv.~} \tbGE{large (heap)} 
\tbLX{toll} \tbHM{} \tbPH{tóll} \tbPS{adj.~} \tbGE{small (heap)} 
\tbLX{tolomboll} \tbHM{} \tbPH{tólómbóll} \tbPS{adv.~} \tbGE{lonely-looking} 
\tbLX{toloŋ} \tbHM{} \tbPH{tʊ́lʊ́ŋ} \tbPS{n.~} \tbGE{heat} \tbXV{Toloŋ maŋ vɛŋ la ka te puri wale.} \tbXE{Heat induces swet.} \tbOP{tola} 
\tbLX{toloŋ} \tbHM{} \tbPH{tʊ́lʊ́ŋ} \tbPS{adj.~} \tbSN \tbGE{hot} \tbXV{A zeɛre ba taa diibu, a toloŋ.} \tbXE{The soup cannot be eaten just yet; it is very hot.} \tbSN \tbGE{fast, quick} \tbXV{Ka fooŋ ba kyɛŋ gbɛ toloŋ fo koŋ pɔge o.} \tbXE{If you don't hasten your steps, you won't meet her.} \tbOP{tola} 
\tbLX{tom} \tbHM{} \tbPH{tôm} \tbPS{v.~} \tbGE{to become sweet, to become enjoyable} \tbPD{tomee, tomino, tomino, tominimo, tominaa, - } 
\tbLX{tomomo} \tbHM{} \tbPH{tómómó} \tbPS{adv.~} \tbGE{very sweet} 
\tbLX{tone} \tbHM{} \tbPH{tʊ́nɪ́} \tbPS{n.~} \tbGE{work} \tbOP{toma} 
\tbLX{tonemaa} \tbHM{} \tbPH{tʊ̊nɪ́máá} \tbPS{adj.~} \tbGE{at once, instantly} 
\tbLX{tonni} \tbHM{} \tbPH{tònnì} \tbPS{v.~} \tbGE{to move blindly; groping} \tbPD{tonnee, tonno, tonno, tonimo, tonnaa, - } 
\tbLX{tonni} \tbHM{} \tbPH{tónní} \tbPS{v.~} \tbGE{to drip} \tbPD{tonnee, tonno, tonno, tonimo, tonnaa, - } 
\tbLX{tontombone} \tbHM{} \tbPH{tʊ̀ntʊ̀mbʊ́nɪ́} \tbPS{n.~} \tbGE{tool} \tbOP{tontombomai} 
\tbLX{tontonɔ} \tbHM{} \tbPH{tʊ̌tʊ́ná} \tbPS{n.~} \tbSN \tbGE{worker} \tbSN \tbGE{industrious person} \tbOP{tontonema} \tbVA{tontona}
\tbLX{tontoŋ} \tbHM{} \tbPH{tóntóŋ} \tbPS{adv.~} \tbSN \tbGE{very sour} \tbXV{A buuluŋ waa la tontoŋ zuo a nyuubu tuo.} \tbXE{The poridge is very sour; it is difficult to eat.} \tbSN \tbGE{very painful, sore e.g. over-night bruised body} \tbXV{A dɔɔ za᷉a de la o zu za ne a gbori, zenɛ a zaaŋ mii tontoŋ lɛ.} \tbXE{The man banged his head against the door yesterday and today it really hurts.} 
\tbLX{toŋe} \tbHM{} \tbPH{tʊ̀ŋɪ̀} \tbPS{v.~} \tbSN \tbGE{to work} \tbSN \tbGE{to send (someone to do something)} \tbPD{toŋɛɛ, tona, tona, tonema, tonaa, tonema } 
\tbLX{toŋkpeɛne} \tbHM{} \tbPH{tʊ̂ŋkpɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{miracles, powerful deeds} \tbTP{toŋkpeɛnɛɛ} 
\tbLX{too} \tbHM{} \tbPH{tòò} \tbPS{use word.~} \tbGE{okay} \tbVA{tɔɔ}
\tbLX{toobini} \tbHM{} \tbPH{tòòbíní} \tbPS{n.~} \tbGE{earwax (ear shit)} \tbOP{n/a} \tbTP{toobinee} 
\tbLX{toogee} \tbHM{} \tbPH{tóógéé} \tbPS{interjection.~} \tbGE{a state indicating an incorrect utterance which is therefore withdrawn} \tbXV{Yeli ka toogee ka lɛ naane o na ŋmɛ fo la.} \tbXE{Better say it is a mistaken statement; otherwise he will beat you.} 
\tbLX{too-gbuli} \tbHM{} \tbPH{tòò-gbúlí} \tbPS{n.~} \tbGE{unpierced ear} \tbOP{too-gbulo} \tbTP{too-gbulee} 
\tbLX{toogbulo} \tbHM{} \tbPH{tòògbùló} \tbPS{n.~} \tbSN \tbGE{hardness of hearing, deafness, blocked ears} \tbXV{Maa teɛre ka toogbulo ka taa, N maŋ yeli la a yɛlɛ ka a deɛ kaara n niŋeŋ.} \tbXE{I think she is hard of hearing, because when I talk to her she just stares into my face.} \tbSN \tbGE{an animal without a mark of identification on the ear} \tbXV{N peroo na naŋ bɔre lambori ŋmaa la pelaa kyɛ taa toogbulo.} \tbXE{My lost sheep has a white spot on its side but unmarked ears.} 
\tbLX{toogyɛle} \tbHM{} \tbPH{tòòdzɛ́lɪ́} \tbPS{n.~} \tbGE{earlobe} \tbOP{toogyɛlɛ} 
\tbLX{tookuriseɛ} \tbHM{} \tbPH{tòòkúrísɪ́ɛ́} \tbPS{n.~} \tbGE{the base of the ear (back)} \tbOP{tookuriseere} \tbTP{tookuriseerɛɛ} 
\tbLX{tookpane} \tbHM{} \tbPH{tòòkpànɪ̀} \tbPS{n.~} \tbGE{earring} \tbOP{tookpama} \tbTP{tookpanɛɛ/tookpannɛɛ} 
\tbLX{toole} \tbHM{} \tbPH{tʊ̀ʊ̀lɪ̀} \tbPS{adj.~} \tbGE{used with ‘ba’ little e.g food} \tbXV{a bondirii ba toole a bie koŋ tege.} \tbXE{The food is little, the child will not eat to his fill.} 
\tbLX{toolere} \tbHM{} \tbPH{tòòléré} \tbPS{n.~} \tbGE{peak, climax, most enjoyable part (of music)} \tbOP{toolerri} \tbTP{tooleree/toolerree} 
\tbLX{toolii} \tbHM{} \tbPH{tòòlíí} \tbPS{n.~} \tbGE{joke} \tbOP{tooliiri} \tbTP{tooliiree} 
\tbLX{toopa᷉apere} \tbHM{} \tbPH{tòòpà᷉à᷉pɪ́rɪ̀} \tbPS{} \tbGE{eardrum} 
\tbLX{toopurimo} \tbHM{} \tbPH{tòópúrímó} \tbPS{n.~} \tbGE{draining ear} \tbTP{toopurimee} 
\tbLX{toore} \tbHM{} \tbPH{tʊ́ʊ́rɪ̀} \tbPS{v.~} \tbGE{to spit} \tbPD{toorɛɛ, toorɔ, toorɔ, tooreba, tooraa, tooreba } 
\tbLX{toore} \tbHM{} \tbPH{tʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to push into e.g foot into socks} \tbSN \tbGE{to push along a surface, to drag along} \tbPD{toorɛɛ, toorɔ, toorɔ, tooreba, tooraa, - } 
\tbLX{toori} \tbHM{} \tbPH{tòòrí} \tbPS{n.~} \tbGE{ear} \tbOP{tobo} \tbTP{tooree} 
\tbLX{toori} \tbHM{} \tbPH{tòòrí} \tbPS{adj.~} \tbGE{naughty e.g. child} \tbXV{A bie na sa᷉a᷉ la baaloŋ kyɛ o taa la toori yaga.} \tbXE{That child appears to be the quiet type but he is very naughty.} 
\tbLX{toori} \tbHM{} \tbPH{tòòrí} \tbPS{adj.~} \tbGE{musically exceptional} \tbXV{A bawaseɛrebɛ sereŋ taa la toori, ba gbɛɛ lanna lere la velaa.} \tbXE{The dancers have a good ear for music; their step falls on the beat.} 
\tbLX{toovaa} \tbHM{} \tbPH{tòòváà} \tbPS{n.~} \tbGE{ear flap; the visible part of the ear (pinna)} \tbOP{toovaare} \tbTP{toovaarɛɛ} 
\tbLX{toovoɔre} \tbHM{} \tbPH{tòòvʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{ear canal; ear piercing} \tbOP{toovɔ᷉ɛ} \tbTP{toovoɔrɛɛ} 
\tbLX{to᷉ɔ} \tbHM{} \tbPH{tʊ́᷉ɔ́᷉} \tbPS{v.~} \tbSN \tbGE{to subdue, to conquer} \tbXV{A bibiiri naŋ mɔ buo la to᷉ɔ o tɔ?} \tbXE{Which of boys overcame the other in the wrestling competition?.} \tbSN \tbGE{to be able, to be capable} \tbXV{N gaa la kyɛ N koŋ to᷉ɔ leɛ wa.} \tbXE{I'll go, but I'll not be able to come back.} \tbPD{to᷉ɔɛ, toɔnɔ, toɔnɔ, toɔnema, toɔnaa, - } 
\tbLX{toɔle} \tbHM{} \tbPH{tʊ́ɔ́lɪ̀} \tbPS{v.~} \tbGE{to send a message} \tbPD{toɔlɛɛ, toɔlɔ, toɔlɔ, toɔlbɔ, toɔlaa, - } 
\tbLX{toɔlɔ} \tbHM{} \tbPH{tʊ́ɔ́lɔ̀} \tbPS{n.~} \tbGE{message} \tbVA{toɔla}
\tbLX{toɔne} \tbHM{} \tbPH{tʊ̀ɔ̀nɪ̀} \tbPS{v.~} \tbGE{to tear off in strips (fibre from the stalk)} \tbPD{toɔnɛɛ, toɔnɔ, toɔnɔ, toɔnema, toɔnaa, - } 
\tbLX{toɔraa} \tbHM{} \tbPH{tʊ́ɔ́ráá} \tbPS{n.~} \tbSN \tbGE{self} \tbXV{Ka fooŋ beɛrɛ kyɛ zagere te᷉e᷉ nyuubu fo toɔraa ka fo erɛ.} \tbXE{If you are sick but refuse medication, you will be hurting yourself.} \tbSN \tbGE{portion, share} \tbXV{Ba poŋ la nɛne ka N de n toɔraa ko n ma.} \tbXE{They shared the meat and I gave my share to my mother.} \tbOP{tɔɛ} \tbVA{toɔre}
\tbLX{toɔre} \tbHM{} \tbPH{tʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{mortar, pounding pot} \tbTP{tɔɛ} \tbVA{toɔraa}
\tbLX{toɔre} \tbHM{} \tbPH{tʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{one's share of something} \tbXV{A seemaa yɛ naŋ maale, N toɔre bɛ?} \tbXE{Where is my share of the food you cooked?.} \tbTP{tɔɛ} \tbVA{toɔraa}
\tbLX{toɔre} \tbHM{} \tbPH{tʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to be careless; to not pay enough attention} \tbXV{O toɔre la a ne,ne bare ka baare de.} \tbXE{S/he left the meat carelessly and dogs ate it.} \tbPD{-, toɔrɔ, -, -, -, - } 
\tbLX{topɛge} \tbHM{} \tbPH{tòpɛ́gɪ́} \tbPS{n.~} \tbGE{boabab fruit shell} \tbOP{topɛgre} \tbTP{topɛgrɛɛ} 
\tbLX{toraza} \tbHM{} \tbPH{tʊ́rázà} \tbPS{n.~} \tbGE{trousers} \tbOP{torazare} \tbTP{torazarɛɛ} 
\tbLX{tore} \tbHM{} \tbPH{tʊ̀rɪ̀} \tbPS{v.~} \tbGE{to sprout, to break through the seed coat} \tbPD{torɛɛ, tora, tora, tora, toraa, - } 
\tbLX{tore} \tbHM{} \tbPH{tʊ̀rɪ̀} \tbPS{v.~} \tbGE{to close one's mouth tightly (usually in anger)} \tbPD{torɛɛ, tora, tora, tora, toraa, - } 
\tbLX{torebogi} \tbHM{} \tbPH{tʊ̀rɪ̀bógí} \tbPS{n.~} \tbGE{frontier, boundary} \tbOP{torebogri} \tbTP{torebogree} 
\tbLX{tori} \tbHM{} \tbPH{tórí} \tbPS{adv.~} \tbGE{continuously (doing one thing)} \tbXV{Ko᷉ɔ yoŋi ka a baala tori nyuuro.} \tbXE{The patient just goes on drinking water only.} 
\tbLX{tori} \tbHM{1} \tbPH{tórí} \tbPS{v.~} \tbSN \tbGE{to be straight} \tbSN \tbGE{to be right, to be correct} \tbPD{toree, toro, toro, toribo, toraa, toribo } 
\tbLX{tori} \tbHM{2} \tbPH{tórí} \tbPS{v.~} \tbGE{to speak with anger, to blame, to complain} \tbPD{toree, toro, toro, toribo, toraa, - } 
\tbLX{tori-buuluŋ} \tbHM{} \tbPH{tórí-búúlúŋ} \tbPS{n.~} \tbGE{porridge made from corn and boabab flour} 
\tbLX{toriko} \tbHM{} \tbPH{tòríkò} \tbPS{n.~} \tbGE{cart} \tbOP{torikori} \tbTP{torikoree} 
\tbLX{toriko} \tbHM{} \tbPH{tòríkò} \tbPS{n.~} \tbGE{large metal snare for trapping large animals} \tbOP{torikori} \tbTP{torikoree} 
\tbLX{torizeɛ} \tbHM{} \tbPH{tórízɪ́ɛ́} \tbPS{n.~} \tbGE{red beret} \tbOP{torizeere} \tbTP{torizerɛɛ} 
\tbLX{to-uri} \tbHM{} \tbPH{tò-ùrí} \tbPS{n.~} \tbGE{tiny spikes covering the fruit of boabab} \tbTP{to-uree} 
\tbLX{tozɔŋe} \tbHM{} \tbPH{tòzɔ́ŋ} \tbPS{n.~} \tbGE{boabab fruit powder} \tbTP{tozɔnnɛɛ} 
\tbLX{tɔ} \tbHM{} \tbPH{tɔ́} \tbPS{n.~} \tbSN \tbGE{colleague, an equal} \tbXV{N baŋ ka n tɔ kaŋa la toŋ a tomɔ ŋa.} \tbXE{I guess this work is done by a neighbour.} \tbSN \tbGE{same type, same kind} \tbXV{N meŋ boɔrɔ la a fo kparoo ŋa tɔ.} \tbXE{I also want the same type of shirt that you have.} \tbOP{taaba} 
\tbLX{tɔ} \tbHM{} \tbPH{tɔ̀} \tbPS{v.~} \tbSN \tbGE{to pound} \tbSN \tbGE{to touch} \tbXV{O de la o nu tɔ ne a gane na o naŋ boɔrɔ.} \tbXE{She used her hand to touch the book she wanted.} \tbSN \tbGE{to head-butt} \tbXV{Ka fooŋ deɛnɛ ne naadɔgerɔ bile a naabo na tɔ fo la.} \tbXE{If you joke with the frehly born calf, its mother will head-butt you.} \tbSN \tbGE{to get to the end of} \tbXV{Te kyɛŋe te tɔ la a sori tɛgɛ.} \tbXE{We walked until we came to the end of the road.} \tbPD{tɔe, toɔrɔ, toɔrɔ, toɔreba, toɔraa, toɔreba } 
\tbLX{tɔba} \tbHM{} \tbPH{tɔ̀bà} \tbPS{n.~} \tbGE{tobacco} \tbVA{tɔbɔ}
\tbLX{tɔbaa} \tbHM{} \tbPH{tɔ̀bàà} \tbPS{n.~} \tbGE{a type of sharp axe used for carving} \tbOP{tɔbre} \tbTP{tɔbrɛɛ} 
\tbLX{tɔbɔ} \tbHM{} \tbPH{tɔ̀bɔ̀} \tbPS{n.~} \tbGE{tobacco} \tbSG{tɔɔroo} \tbTP{tɔɔrɛɛ} \tbVA{tɔba}
\tbLX{tɔge} \tbHM{} \tbPH{tɔ́gɪ́} \tbPS{n.~} \tbSN \tbGE{an earthenware container with a hole in the bottom for sieving} \tbSN \tbGE{fish trap} \tbOP{tɔgere} \tbTP{tɔgerɛɛ} 
\tbLX{tɔge (Eng)} \tbHM{} \tbPH{tɔ́gɪ́} \tbPS{v.~} \tbGE{to stop} \tbPD{tɔgɛɛ, tɔgerɔ, tɔgerɔ, tɔgereba, tɔgeraa, - } 
\tbLX{tɔgele} \tbHM{} \tbPH{tɔ̀glɪ̀} \tbPS{v.~} \tbGE{to imitate} \tbXV{A bie tɔgele la a biwɔɔ kyɛnne a meŋ deɛ te kyɛnɛ wɔɔreŋ.} \tbXE{The boy imitated the way the lame boy walks and now he limps.} \tbPD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\tbLX{tɔgele} \tbHM{} \tbPH{tɔ̀glɪ̀} \tbPS{v.~} \tbGE{to collect from above e.g. rain water} \tbXV{Ka fooŋ tɔgele saa ko᷉ɔ᷉ a maŋ maaroŋ.} \tbXE{Gathered rain water is always cold.} \tbPD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\tbLX{tɔgele} \tbHM{} \tbPH{tɔ̀glɪ̀} \tbPS{v.~} \tbGE{to turn to face e.g. in a direction} \tbXV{O leɛ la o niŋe tɔgele sapare a na nyɛ a saa naŋ ko.} \tbXE{She has turned towards the east to see the clouds.} \tbPD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\tbLX{tɔgele} \tbHM{} \tbPH{tɔ̀glɪ̀} \tbPS{v.~} \tbGE{to help deliver a baby} \tbPD{tɔglɛɛ, tɔglɔ, tɔglɔ, -, tɔglaa, - } 
\tbLX{tɔle} \tbHM{} \tbPH{tɔ́lɪ́} \tbPS{v.~} \tbGE{to pass by} \tbPD{tɔlɛɛ, tɔlɔ, tɔlɔ, tɔlebɔ, tɔlaa, - } 
\tbLX{tɔll} \tbHM{} \tbPH{tɔ̀ll} \tbPS{adj.~} \tbGE{long and cylindrical} 
\tbLX{tɔnne} \tbHM{} \tbPH{tɔ́nnɪ̀} \tbPS{v.~} \tbGE{to join} \tbPD{tɔnnɛɛ, tɔnnɔ, tɔnnɔ, tɔnema, tɔnnaa, tɔnema } 
\tbLX{tɔnne} \tbHM{} \tbPH{tɔ̀nnɪ̀} \tbPS{v.~} \tbGE{to sell (usually at a low price)} \tbPD{tɔnnɛɛ, tɔnnɔ, tɔnnɔ, tɔnema, tɔnnaa, tɔnema } 
\tbLX{tɔnɔ} \tbHM{} \tbPH{tɔ̀nɔ́} \tbPS{n.~} \tbGE{profit, gain, benefit} \tbTP{tɔnɛɛ} \tbVA{tɔna}
\tbLX{tɔntɔbaa} \tbHM{} \tbPH{tɔ̀ntɔ́báá} \tbPS{adv.~} \tbGE{at once} 
\tbLX{tɔŋe} \tbHM{} \tbPH{tɔ̀ŋ} \tbPS{v.~} \tbGE{to respond to, to answer, to reply} \tbPD{tɔŋɛɛ, tɔŋena/tɔnnɔ, tɔnnɔ, tɔnema, tɔnnaa, tɔnema } 
\tbLX{tɔŋe} \tbHM{} \tbPH{tɔ́ŋ} \tbPS{v.~} \tbSN \tbGE{to be in a queue} \tbSN \tbGE{to light a fire} \tbSN \tbGE{to fetch fire} \tbPD{tɔŋɛɛ, tɔnɔ, tɔnɔ, tɔnema, tɔnnaa, - } \tbVA{kyɔŋ}
\tbLX{tɔŋe} \tbHM{} \tbPH{tɔ̂ŋ} \tbPS{v.~} \tbGE{to light a fire from another fire} \tbPD{tɔŋɛɛ, tɔŋena/tɔnnɔ, tɔnnɔ, tɔnemɔ, tɔnnaa, tɔnema } 
\tbLX{tɔŋemɔ} \tbHM{} \tbPH{tɔ́ŋɪ́mɔ̀} \tbPS{n.~} \tbGE{colleague} \tbOP{tɔŋemɔmine} \tbVA{tɔŋema}
\tbLX{tɔɔ} \tbHM{} \tbPH{tɔ̀ɔ̀} \tbPS{use word.~} \tbGE{all right, okay} \tbVA{too}
\tbLX{tɔɔbɔle} \tbHM{} \tbPH{tɔ̀ɔ̀bɔ́lɪ́} \tbPS{n.~} \tbGE{cylinder-shaped pounded tobacco} \tbOP{tɔɔbɔlɔ} \tbTP{tɔɔbɔlɛɛ} 
\tbLX{tɔɔlaa} \tbHM{} \tbPH{tɔ̀ɔ̀láá} \tbPS{n.~} \tbGE{tobacco pipe} \tbOP{tɔɔlaare} \tbTP{tɔɔlaarɛɛ} 
\tbLX{tɔɔlaka᷉a} \tbHM{} \tbPH{tɔ̀ɔ̀làká᷉à᷉} \tbPS{n.~} \tbGE{nicotine} \tbTP{tɔɔlaka᷉anɛɛ} 
\tbLX{tɔɔre} \tbHM{} \tbPH{tɔ̀ɔ̀rɪ̀} \tbPS{n.~} \tbGE{ball of tobacco} \tbOP{tɔbɔ} \tbTP{tɔɔrɛɛ} 
\tbLX{tɔɔre} \tbHM{} \tbPH{tɔ̀ɔ̀rɪ́} \tbPS{adj.~} \tbGE{far} \tbOP{tɔbɔ} 
\tbLX{tɔɔvaa} \tbHM{} \tbPH{tɔ̀ɔ̀váà} \tbPS{n.~} \tbGE{tobacco leaf} \tbOP{tɔɔvaare} 
\tbLX{tɔɔwuoraa} \tbHM{} \tbPH{tɔ̀ɔ̀wúóráá} \tbPS{n.~} \tbGE{a shell used in gathering and collecting powdered tobacco (snuff)} \tbOP{tɔɔwuori} 
\tbLX{tɔɔzɔŋe} \tbHM{} \tbPH{tɔ̀ɔ̀zɔ̂ŋ} \tbPS{n.~} \tbGE{powdered tobacco; snuff} \tbTP{tɔɔzɔnnɛɛ} 
\tbLX{tɔre} \tbHM{} \tbPH{tɔ̀rɪ́} \tbPS{pronoun.~} \tbGE{self} \tbXV{N na de la a bondirii tte ko o N tɔre.} \tbXE{I will take the food to him myself.} 
\tbLX{tɔre} \tbHM{} \tbPH{tɔ̀rɪ̀} \tbPS{v.~} \tbGE{to share in pieces} \tbXV{A bondirii ba seŋe bone kyɛ te na de tɔre la a bibiiri.} \tbXE{We have very little food left but we shall share it in the little bits that we can.} \tbPD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\tbLX{tɔre} \tbHM{} \tbPH{tɔ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to break easily e.g. weak twine} \tbXV{A miri poɔ̃ɛ la lɛnso ka O gaa ka o lene a tuobu ka o tɔre.} \tbXE{The rope was rotten and that was why when s/he attempted to tie up the luggage it gave way easily.} \tbSN \tbGE{to cut a small piece} \tbXV{N baŋ ka a nɛne ba seŋe bone kyɛ N na tɔre la fe᷉e ko fo.} \tbXE{I know the piece meat is small but I'll give you a little bit.} \tbSN \tbGE{die (ailing child)} \tbPD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\tbLX{tɔre} \tbHM{} \tbPH{tɔ̀rɪ̀} \tbPS{v.~} \tbGE{to cut a small piece} \tbXV{N baŋe ka a nɛne ba seŋe bone kyɛ N na tɔre la fe᷉e ko fo.} \tbXE{I know the piece meat is small but I'll give you a little bit.} \tbPD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\tbLX{tɔre} \tbHM{} \tbPH{tɔ́rɪ́} \tbPS{v.~} \tbGE{die e.g a person who has been sick for long} \tbPD{tɔrɛɛ, tɔrɔ, tɔrɔ, tɔreba, tɔraa, - } 
\tbLX{tɔre} \tbHM{} \tbPH{tɔ̀rɪ́} \tbPS{adj.~} \tbGE{valuable} \tbXV{O yeli ka ba zu la o libie kyɛ a ba e libi tɔre zaa.} \tbXE{She said they had stolen her money but it was not worth anything.} 
\tbLX{tɔ teere} \tbHM{} \tbPH{tɔ̀̀ tɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{without any hope/hit the rocks} \tbXV{A pɔge bie baaloŋ la so ka o sori na yiibu tɔ teere.} \tbXE{The woman's intended journey hit the rocks because of her child's sickness.} \tbSN \tbGE{to worsen} \tbXV{O gaa wɛre la o nimie na a deɛ tɔ teere.} \tbXE{The eye operation has worsened his sight.} \tbPD{tɔ teerɛɛ, toɔrɔ teerɛ, tɔteerɛ, tɔteerɛ, tɔteeraa, tɔteerebɛ } 
\tbLX{tɔ zɔŋɔ kyi} \tbHM{} \tbPH{tɔ̀ zɔ̀ŋɔ̀ kyî} \tbPS{v.~} \tbSN \tbGE{to beat about the bush} \tbSN \tbGE{to be ignorant} \tbPD{tɔɛ zɔŋ kyi, toɔrɔ zɔŋ kyi, -, -, -, - } 
\tbLX{tu} \tbHM{} \tbPH{tù} \tbPS{v.~} \tbGE{to follow a route, to take a route} \tbPD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\tbLX{tu} \tbHM{} \tbPH{tù} \tbPS{v.~} \tbGE{to recount, to narrate} \tbPD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\tbLX{tu} \tbHM{} \tbPH{tù} \tbPS{v.~} \tbGE{to bring back an attempted (female) divorcee} \tbPD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\tbLX{tu} \tbHM{} \tbPH{tú} \tbPS{v.~} \tbSN \tbGE{to dig} \tbXV{A sakuuri tu la bulee ba dabɔge poɔ.} \tbXE{The school has dug a well in their garden.} \tbSN \tbGE{to uproot} \tbXV{A koɔraa tu la waare paale o wasege.} \tbXE{The farmer has dug up yams and filled the yam barn.} \tbPD{tue, tuuro, tuuro, tuuribo, tuuraa, - } 
\tbLX{tu ba} \tbHM{} \tbPH{tú bà} \tbPS{v.~} \tbGE{to take off quickly e.g. in running} \tbPD{tu baɛ, tuuro baara, tu baara, tu baareba, -, - } 
\tbLX{tuli} \tbHM{} \tbPH{tùlì} \tbPS{v.~} \tbGE{to make a mistake, to err, to be wrong} \tbPD{tulee, tullo, tullo, tulbo, tulaa, - } 
\tbLX{tulli} \tbHM{} \tbPH{tùllì} \tbPS{v.~} \tbSN \tbGE{to change (state), disguise} \tbXV{A bibiiri da tulli la ba menne a tigiri daare ka fo koŋ la baŋ ba.} \tbXE{The boys diguised themselves during the festival, so that you could not recognise them.} \tbSN \tbGE{to confuse} \tbXV{N bare la gyɛle ka a pɔge tulli ma kyɛ zo ne a bie.} \tbXE{I was so careless that the woman tricked me and escaped with the child.} \tbPD{tullee, tullo, tullo, tulibo, tullaa, - } 
\tbLX{tulli} \tbHM{} \tbPH{tùllì} \tbPS{v.~} \tbGE{to incriminate spiritually} \tbXV{O tulli la omeŋɛ lɛnso ka o bageboge dagoli ba la nyɛrɛ.} \tbXE{She has wronged the spirits and that is why she cannot see things in the spiritual world.} \tbPD{tullee, tullo, tullo, tulibo, tullaa, - } 
\tbLX{tulundaa} \tbHM{} \tbPH{túlúndàá} \tbPS{n.~} \tbGE{pestle} \tbOP{tulundaare} \tbTP{tunlundaarɛɛ} \tbVA{tulindaa, tundaa}
\tbLX{tumakuli} \tbHM{} \tbPH{tùmákúlí} \tbPS{n.~} \tbGE{burri; a type of small plant with thorny fruits} \tbSG{tumakuluu} \tbOP{tumakulli} \tbTP{tumakulee} 
\tbLX{tumo} \tbHM{} \tbPH{tùmó} \tbPS{n.~} \tbGE{one born with the feet coming first instead of the head} \tbTP{tumee} 
\tbLX{tumo} \tbHM{} \tbPH{tùmó} \tbPS{n.~} \tbGE{glutton} \tbTP{tumee} 
\tbLX{tumpaane} \tbHM{} \tbPH{túmpáánɪ́} \tbPS{n.~} \tbGE{large talking drum; a drum that gives messages} \tbOP{tumpaama} \tbTP{tumpaanɛɛ} \tbVA{tumpanne}
\tbLX{tumpaane} \tbHM{} \tbPH{túmpáánɪ́} \tbPS{n.~} \tbGE{slabs of bean flour boiled in leaves} \tbOP{tumpaama} \tbTP{tumpaanɛɛ} \tbVA{tumpanne}
\tbLX{tu noɔre} \tbHM{} \tbPH{tù nʊ́ɔrɪ̀} \tbPS{v.~} \tbGE{to narrate, to give an account of} \tbXV{Ba yeli ka a bibiiri na naŋ zɔɔ taa ka kaŋa zaa maŋ tu o noɔre.} \tbXE{Each of the boys who fought has been asked to give an account of what led to the fight.} \tbPD{-, -, -, -, -, - } 
\tbLX{tuntundaana} \tbHM{} \tbPH{túntúndááná} \tbPS{n.~} \tbGE{mosquito net} \tbOP{tuntundaane} \tbTP{tuntundaanɛɛ} 
\tbLX{tunturi} \tbHM{} \tbPH{túntúrì} \tbPS{n.~} \tbGE{a type of very large hog with powerful claws and tail for digging} \tbOP{tuntue} \tbTP{tuturee} 
\tbLX{tuŋi} \tbHM{} \tbPH{tùŋ} \tbPS{v.~} \tbGE{to thread (beads)} \tbPD{tuŋee, tuno, tuno, tunimo, tunaa, tunimo } 
\tbLX{tuŋi} \tbHM{} \tbPH{tùŋ} \tbPS{v.~} \tbSN \tbGE{to dip e.g. hand in pocket} \tbSN \tbGE{to insert a finger into the vagina (defile, rape)} \tbPD{tuŋee, tuno, tuno, tunimo, tunaa, tunimo } 
\tbLX{tuŋi} \tbHM{} \tbPH{tûŋ} \tbPS{v.~} \tbSN \tbGE{to stumble about to fall, to totter} \tbSN \tbGE{to err, to make a mistake} \tbPD{tuŋee, tuŋino, tuŋino, tuŋinimo, tuŋinaa, tuŋinimo } \tbVA{tummu (of -bu form)}
\tbLX{tuŋi} \tbHM{} \tbPH{túŋ} \tbPS{v.~} \tbGE{to overgrow e.g. hair, grass of lawn} \tbPD{-, -, -, -, -, - } 
\tbLX{tuo} \tbHM{} \tbPH{tùò} \tbPS{v.~} \tbSN \tbGE{to carry e.g. on head} \tbSN \tbGE{to swell e.g. blister} \tbPD{-, tuoro, tuoro, tuoribo, -, - } 
\tbLX{tuo} \tbHM{} \tbPH{tùó} \tbPS{n.~} \tbSN \tbGE{baobab tree} \tbSN \tbGE{baobab fruit} \tbOP{tori} \tbTP{toree} 
\tbLX{tuo} \tbHM{} \tbPH{túó} \tbPS{n.~} \tbSN \tbGE{suffering} \tbXV{Tuo yoŋ ka zɔɔre taa.} \tbXE{There is only suffering in war.} \tbSN \tbGE{difficulty} \tbXV{A daŋmaa ŋa ŋmaabo tuo la so ka o ba wa.} \tbXE{The difficulty involved in cutting this log has disuaded him from coming.} \tbOP{tuuri} \tbTP{tuuree} 
\tbLX{tuo} \tbHM{} \tbPH{túó} \tbPS{adj.~} \tbSN \tbGE{bitter e.g. quinine} \tbSN \tbGE{hot e.g. pepper} \tbOP{tuuri} \tbTP{tuuree} 
\tbLX{tuobu} \tbHM{} \tbPH{tùòbú} \tbPS{n.~} \tbSN \tbGE{load} \tbSN \tbGE{burden} \tbSN \tbGE{weight} 
\tbLX{tuori} \tbHM{} \tbPH{tùòrì} \tbPS{v.~} \tbSN \tbGE{to meet face to face} \tbSN \tbGE{to welcome} \tbPD{tuoree, tuoro, tuoro, tuoribo, tuoraa, tuoribo } 
\tbLX{tuosoba} \tbHM{} \tbPH{túósʊ́bá} \tbPS{n.~} \tbGE{person who does not want to make offers to any other person} 
\tbLX{turi} \tbHM{} \tbPH{túrì} \tbPS{n.~} \tbGE{thousand} \tbSG{turree} \tbTP{turee} 
\tbLX{turi} \tbHM{} \tbPH{tùrì} \tbPS{v.~} \tbSN \tbGE{to give from above or below} \tbXV{A pɔge wuo a kyi peɛ zu de turi o serɛ.} \tbXE{The woman gathered the millet from the top of the roof and passed it down to her husband.} \tbSN \tbGE{to offer one's hand} \tbPD{turee, turo, turo, turibo, turaa, - } 
\tbLX{tu sori} \tbHM{} \tbPH{tù sórí} \tbPS{v.~} \tbGE{to be right} \tbXV{Te yɔɔ la a gama kyɛ ka dɔɔ wa wuo a zaa gaa ne, a lɛ ba tu sori.} \tbXE{We paid for the books but the man came and took them away; that is not right.} \tbPD{-, -, -, -, -, - } 
\tbLX{tutɔ} \tbHM{} \tbPH{tútɔ́} \tbPS{n.~} \tbGE{companion} \tbOP{tutɔmine} 
\tbLX{tuu} \tbHM{} \tbPH{túú} \tbPS{n.~} \tbSN \tbGE{thicket, forest, wood} \tbSN \tbGE{raised ground (usually on a farm)} \tbSN \tbGE{eagle} \tbOP{tuuri} \tbTP{tuuree} \tbVA{togroo}
\tbLX{tuu} \tbHM{} \tbPH{túú} \tbPS{n.~} \tbSN \tbGE{thicket, forest, wood} \tbSN \tbGE{raised ground (usually on a farm)} \tbOP{tuuri} \tbTP{tuuree} \tbVA{togroo}
\tbLX{tuu} \tbHM{} \tbPH{túú} \tbPS{n.~} \tbGE{eagle} \tbOP{tuuri} \tbTP{tuuree} 
\tbLX{tuuli} \tbHM{} \tbPH{tùùlì} \tbPS{v.~} \tbSN \tbGE{to trace e.g. line} \tbSN \tbGE{to use a finger to apply e.g. ointment} \tbSN \tbGE{to go about, to roam around} \tbSN \tbGE{to carried away by water} \tbPD{tuulee, tuulo, tuulo, tuulbo, tuulaa, - } 
\tbLX{tuuli} \tbHM{} \tbPH{túúlì} \tbPS{v.~} \tbSN \tbGE{to give to drink e.g. a child} \tbSN \tbGE{not deeply emersed (arrow hitting game)} \tbPD{tuulee, tuulo, tuulo, tuulbo, tuulaa, - } 
\tbLX{tuuree} \tbHM{} \tbPH{túúréé} \tbPS{n.~} \tbGE{Europe or America} 
\tbLX{tuuri} \tbHM{} \tbPH{tùùrì} \tbPS{v.~} \tbGE{to pick out, to sort out, to separate} \tbPD{tuuree, tuuro, tuuro, tuuribo, tuuraa, - } 
\tbLX{ugi} \tbHM{} \tbPH{úgì} \tbPS{v.~} \tbGE{to expand e.g. cooked rice} \tbPD{ugee, ugro, ugro, ugribo, ugraa, - } 
\tbLX{ulee} \tbHM{} \tbPH{ùléè} \tbPS{n.~} \tbGE{tree branch} \tbOP{ulli} \tbTP{ullee} 
\tbLX{uli} \tbHM{} \tbPH{úlí} \tbPS{n.~} \tbGE{fog, mist} \tbOP{ulli} \tbTP{ullee} 
\tbLX{uli} \tbHM{} \tbPH{ùlì} \tbPS{v.~} \tbSN \tbGE{to be dusty} \tbSN \tbGE{to become moldy e.g. food} \tbXV{Ka boroboroŋ wa kɔɔre o maŋ uli la.} \tbXE{When bread is kept for long it becomes mouldy.} \tbPD{ulee, ullo, ullo, ulibo, ullaa, - } 
\tbLX{ulimo} \tbHM{} \tbPH{úlmó} \tbPS{adj.~} \tbGE{grey} 
\tbLX{unni} \tbHM{} \tbPH{ùnnì} \tbPS{v.~} \tbSN \tbGE{unburying something buried} \tbXV{Polisiri yeli ka ba ba baŋ a dɔɔ ku᷉u᷉, pare a gaa te unni a ku᷉u᷉.} \tbXE{The police said they did not know the cause of the man's death, so they went and exhumed the corpse.} \tbSN \tbGE{to bring back past memories} \tbXV{Ka Dɛre ne o pɔgeŋ wa zɔɔrɔ zaa a pɔge maŋ unni la yelikorɔ.} \tbXE{Each time Dɛre and his wife quarrel, the wife will always recount past events.} \tbPD{unnee, unno, unno, unimo, unnaa, - } 
\tbLX{unni} \tbHM{} \tbPH{únnì} \tbPS{v.~} \tbGE{to remind} \tbXV{N deɛmɛ za᷉a᷉ wa unni ma la a koɔbo.} \tbXE{My in-law reminded me about the day of the group farming.} \tbPD{unnee, unno, unno, unimo, unnaa, - } 
\tbLX{uoni} \tbHM{} \tbPH{úònì} \tbPS{n.~} \tbGE{dry season} \tbOP{uomo} \tbTP{uonee} 
\tbLX{uori} \tbHM{} \tbPH{úórì} \tbPS{v.~} \tbGE{to warm oneself (by fire, sun)} \tbPD{uoree, uoro, uoro, uoribo, uoraa, - } 
\tbLX{uori} \tbHM{} \tbPH{úórì} \tbPS{v.~} \tbGE{to mourn} \tbPD{uoree, uoro, uoro, uoribo, uoraa, - } 
\tbLX{uori} \tbHM{} \tbPH{úórì} \tbPS{v.~} \tbGE{to peer (shading one's face with the hand)} \tbPD{uoree, uoro, uoro, uoribo, uoraa, - } 
\tbLX{uori} \tbHM{} \tbPH{ùòrì} \tbPS{v.~} \tbSN \tbGE{to bellow (bull)} \tbSN \tbGE{to groan; a sound uttered in pain or grief} \tbPD{uoree, uoro, uoro, uoribo, uoraa, - } 
\tbLX{uri} \tbHM{} \tbPH{ùrí} \tbPS{n.~} \tbSN \tbGE{itchy stuff} \tbSN \tbGE{chaff} \tbOP{n/a} \tbTP{urree} 
\tbLX{uri} \tbHM{} \tbPH{ùrì} \tbPS{v.~} \tbSN \tbGE{to take unaware} \tbSN \tbGE{to be shocked} \tbXV{Maa ne a baala naŋ yeli yɛlɛ ba kɔɔre, o ku᷉u᷉ uri ma la.} \tbXE{I spoke with the patient not long ago; I'm shocked at his sudden death.} \tbPD{uree, uro, uro, uribo, uraa, - } 
\tbLX{uri} \tbHM{} \tbPH{úrì} \tbPS{v.~} \tbGE{to make burn e.g fire} \tbPD{-, -, -, -, -, - } 
\tbLX{uri} \tbHM{} \tbPH{ùrí} \tbPS{n.~} \tbGE{hair-like projections on the straw of the grass family and other plants that cause itchiness of the skin} 
\tbLX{uri} \tbHM{} \tbPH{úrí} \tbPS{v.~} \tbGE{to scrub a very dirty object e.g. cooking pot} \tbPD{uree, uro, uro, uribo, uraa, - } 
\tbLX{u᷉u} \tbHM{} \tbPH{ú᷉ú᷉} \tbPS{v.~} \tbGE{to close tightly (fist, eyes)} \tbPD{u᷉ue, uuno, uuno, uuimo, uunaa, - } 
\tbLX{u᷈u} \tbHM{} \tbPH{ù᷉ù᷈} \tbPS{v.~} \tbGE{to bury} \tbPD{u᷉ue, uuno, uuno, uuimo, uunaa, - } 
\tbLX{uuni} \tbHM{} \tbPH{úúnì} \tbPS{v.~} \tbGE{to crouch, to hide} \tbXV{Ba digi la a bie ka o zo te uuni tuu poɔ.} \tbXE{As they chased the boy, he ran and hid in a thicket.} \tbPD{uunee, uuno, uuno, uunimo, uunaa, - } 
\tbLX{uuni} \tbHM{} \tbPH{úúnì} \tbPS{v.~} \tbGE{to incubate, to brood, to sit on (a fowl on its eggs)} \tbXV{Noɔ maŋ uuni la daare ata kyɛ wɛge.} \tbXE{A fowl sits on its eggs before they hatch.} \tbPD{uunee, uuno, uuno, uunimo, uunaa, - } 
\tbLX{uuni} \tbHM{} \tbPH{úúnì} \tbPS{v.~} \tbGE{to form pus e.g. a swelling} \tbXV{Goɔ daare kyɔge la n nubiri ka o mɔre eŋ kõɔ ka ba wɛre.} \tbXE{A thorn pierced my finger and it got swollen, formed pus, and was operated upon.} \tbPD{uunee, uuno, uuno, uunimo, uunaa, - } 
\tbLX{uuri} \tbHM{} \tbPH{úúrí} \tbPS{v.~} \tbGE{to hoot, to call out e.g. lost person} \tbPD{uuree, uuro, uuro, uuribo, uuraa, - } 
\tbLX{uuri} \tbHM{} \tbPH{ùùrì} \tbPS{v.~} \tbSN \tbGE{to smoke out} \tbSN \tbGE{to warm a place with smoke} \tbPD{uuree, uuro, uuro, uuribo, uuraa, - } 
\tbLX{uuruŋ} \tbHM{} \tbPH{úúrúŋ} \tbPS{n.~} \tbGE{dust} \tbTP{uuurunnee} 
\tbLX{va} \tbHM{} \tbPH{và} \tbPS{v.~} \tbSN \tbGE{to whip} \tbSN \tbGE{to thresh (millet)} \tbPD{vaɛ, -, vaara, vareba, vaaraa, - } 
\tbLX{va} \tbHM{} \tbPH{vá} \tbPS{v.~} \tbGE{used with ‘-de’(va-de); snatch; to run directly and suddenly at someone seizing him} \tbPD{-, -, -, -, -, - } 
\tbLX{va} \tbHM{} \tbPH{và} \tbPS{v.~} \tbGE{to graze a target; miss narrowly} \tbPD{vaɛ, -, vaara, vareba, vaaraa, - } 
\tbLX{va} \tbHM{} \tbPH{vá} \tbPS{v.~} \tbGE{to rush, to dash} \tbXV{A dɔbɔ naŋ ŋmeɛrɛ la ka kaŋa wa va de o tɔ gaa lɔɔ.} \tbXE{As the men argued, one suddenly rushed on the other and threw him down.} \tbPD{vaɛ, vaara, vaara, vareba, vaaraa, - } 
\tbLX{vaa} \tbHM{} \tbPH{váà} \tbPS{n.~} \tbGE{leaf} \tbSG{vabo} \tbOP{vaare} \tbTP{vaarɛɛ} 
\tbLX{vaa} \tbHM{} \tbPH{vàà} \tbPS{v.~} \tbGE{collect or take everything from a container} \tbPD{vaaɛ, vaara, vaara, vaareba, vaaraa, - } 
\tbLX{vaa} \tbHM{} \tbPH{vàà} \tbPS{v.~} \tbGE{to draw e.g. bowstring} \tbPD{vaaɛ, vaara, vaara, vaareba, vaaraa, - } 
\tbLX{vaa} \tbHM{} \tbPH{vàà} \tbPS{v.~} \tbGE{to lift an opponent's leg (wrestling)} \tbPD{vaaɛ, vaara, vaara, vaareba, vaaraa, - } 
\tbLX{vaagaviige} \tbHM{} \tbPH{vààgàvíígè} \tbPS{v.~} \tbGE{I command you (to die)} \tbPD{-, -, -, -, -, - } 
\tbLX{vaa-iri} \tbHM{} \tbPH{vàà-ìrì} \tbPS{v.~} \tbGE{to stand up quickly} \tbPD{vaa-iree, vaara-ire, vaa-ire, vaa-iribe, -, - } 
\tbLX{vaale} \tbHM{} \tbPH{vààlɪ̀} \tbPS{v.~} \tbGE{to gather e.g. rubbish, garbage} \tbPD{-, -, -, -, -, - } 
\tbLX{vaalee} \tbHM{} \tbPH{váálɪ́ɪ́} \tbPS{n.~} \tbGE{speck} \tbSG{vaaloo} \tbOP{vaala} \tbTP{vaalɛɛ} \tbVA{vaale}
\tbLX{vaali} \tbHM{} \tbPH{vạ̀ạ̀lì} \tbPS{v.~} \tbGE{move fast e.g in standing up} \tbPD{-, -, -, -, -, - } 
\tbLX{vaane} \tbHM{} \tbPH{vàànɪ̀} \tbPS{v.~} \tbSN \tbGE{to shrink, contract} \tbSN \tbGE{to wrinkle} \tbPD{vaanɛɛ, vaana, vaana, vaaneba, vaanaa, - } 
\tbLX{vaane} \tbHM{} \tbPH{váánɪ́} \tbPS{adj.~} \tbGE{shrunken} 
\tbLX{va᷉ao} \tbHM{} \tbPH{và᷉à᷈ʊ᷉} \tbPS{adv.~} \tbGE{courageously} 
\tbLX{vaare} \tbHM{} \tbPH{vààrɪ̀} \tbPS{v.~} \tbGE{to wrap oneself, usually with a cloth} \tbPD{vaarɛɛ, vaara, vaara, vaareba, vaaraa, - } 
\tbLX{vaare} \tbHM{} \tbPH{váárɪ̀} \tbPS{v.~} \tbGE{to tug, to pull strongly and suddenly towards oneself} \tbXV{A bie da nyɔge la a pedaa miri kyɛ ka a peroo vaare o lɔɔ.} \tbXE{The boy held onto the rope around the ram's neck, but it tugged him to the ground.} \tbPD{vaarɛɛ, vaara, vaara, vaareba, vaaraa, - } 
\tbLX{vaare} \tbHM{} \tbPH{váárɪ̀} \tbPS{v.~} \tbGE{to rinse} \tbXV{Fanfane fuuri naŋ be la laare poɔ a seŋ ka vaare.} \tbXE{There are still soapsuds in the bowls, so they need to be rinsed.} \tbPD{vaarɛɛ, vaara, vaara, vaareba, vaaraa, - } 
\tbLX{vabɔre} \tbHM{} \tbPH{vábɔ́rɪ̀} \tbPS{n.~} \tbGE{spring} \tbTP{vabɔrrɛɛ} 
\tbLX{vãee} \tbHM{} \tbPH{vã̀ɪ̃́ɪ̃̀} \tbPS{adv.~} \tbGE{aromatically} 
\tbLX{vage} \tbHM{} \tbPH{vágɪ́} \tbPS{v.~} \tbGE{onomatopoeia; sounding like a bird pecking at a soft object} \tbPD{vagɛɛ, vagra, vagra, vagreba, vagraa, - } 
\tbLX{vage} \tbHM{} \tbPH{vàgɪ̀} \tbPS{adv.~} \tbGE{sound of pecking} 
\tbLX{vagere} \tbHM{} \tbPH{vàgɪ̀rɪ̀} \tbPS{adv.~} \tbGE{pecking} 
\tbLX{vakogilo} \tbHM{} \tbPH{vákóglo} \tbPS{n.~} \tbGE{dried leaves (that once worn by women)} \tbSG{vakogiluu} 
\tbLX{vakuoŋaa} \tbHM{} \tbPH{vákúóŋáá} \tbPS{n.~} \tbGE{dry leaf} \tbOP{vakuoni} 
\tbLX{vakuoŋaa} \tbHM{} \tbPH{vákúóŋáá} \tbPS{n.~} \tbGE{bad luck} \tbOP{vakuoni} 
\tbLX{valenvo᷉ɔ} \tbHM{} \tbPH{válɪ́nvʊ́᷉ɔ̀᷉} \tbPS{n.~} \tbGE{mason wasp, potter wasp} \tbSG{valenvoɔnoo} \tbOP{valenvoɔne} \tbTP{valenvoɔnɛɛ} \tbVA{valenvõa}
\tbLX{valere} \tbHM{} \tbPH{vàlɪ́rɪ́} \tbPS{v.~} \tbGE{to intertwine e.g. weaving} \tbPD{-, -, -, -, -, - } 
\tbLX{valimmu} \tbHM{} \tbPH{vàlímmú} \tbPS{n.~} \tbGE{a type of small tree with soft large leaves which women used to wear to cover parts of their private parts} 
\tbLX{valimmu} \tbHM{} \tbPH{vàlǐmmú} \tbPS{n.~} \tbGE{a type of broad-leafed plant} \tbOP{valinni} \tbTP{valinnee} 
\tbLX{valinvaa} \tbHM{} \tbPH{vàlìnv́áà} \tbPS{n.~} \tbSN \tbGE{leaf of the palm tree} \tbSN \tbGE{the leaf of the valimmu tree} \tbOP{valinvaare} \tbTP{valinvaarɛɛ} 
\tbLX{valinvuuluŋ} \tbHM{} \tbPH{v̀àlìnvúúlúŋ} \tbPS{n.~} \tbGE{soot} \tbOP{valinvuulo/valinvuuli} \tbTP{valinvuulee} \tbVA{valanvuuluŋ, vuuluŋ}
\tbLX{valogema} \tbHM{} \tbPH{válʊ́gɪ́mà} \tbPS{n.~} \tbGE{green, fresh, and tender leaves} 
\tbLX{vama} \tbHM{} \tbPH{vàmà} \tbPS{adv.~} \tbGE{effortlessly} 
\tbLX{vamaaroŋ} \tbHM{} \tbPH{vámáárʊ́ŋ} \tbPS{n.~} \tbGE{green leaf} \tbOP{vamaara} \tbTP{vamaarɛɛ} 
\tbLX{vanne} \tbHM{} \tbPH{vánnɪ̀} \tbPS{v.~} \tbGE{to stretch, to grow longer} \tbPD{vannɛɛ, vanna, vanna, vannaa, -, - } 
\tbLX{vapeɛlegboro} \tbHM{} \tbPH{vápíɛ́lgbóró} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} 
\tbLX{vapɛge} \tbHM{} \tbPH{vápɛ́gɪ́} \tbPS{n.~} \tbGE{fallen leaf} \tbOP{vapɛgre} \tbTP{vapgrɛɛ} 
\tbLX{vapielee} \tbHM{} \tbPH{vápìèlèè} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{vapieli} 
\tbLX{vapɔrɔ} \tbHM{} \tbPH{vápɔ̀rɔ̀} \tbPS{n.~} \tbGE{fresh green leaves} \tbTP{vapɔrrɛɛ} 
\tbLX{vare} \tbHM{} \tbPH{vàrɪ̀} \tbPS{v.~} \tbGE{to wrinkle} \tbPD{varɛɛ, vara, vara, vareba, varaa, vareba } 
\tbLX{vare} \tbHM{} \tbPH{vàrɪ̀} \tbPS{v.~} \tbGE{to make laundry not well washed e.g. clothes} \tbPD{varɛɛ, vara, vara, vareba, varaa, vareba } 
\tbLX{vare} \tbHM{} \tbPH{vàrɪ̀} \tbPS{v.~} \tbGE{to make the layout of a mound} \tbPD{varɛɛ, vara, vara, vareba, varaa, vareba } 
\tbLX{vare} \tbHM{} \tbPH{várɪ́} \tbPS{v.~} \tbGE{to jump, to leap, to hop} \tbPD{varoo, vara, vara, varibo, varaa, variba } 
\tbLX{vare} \tbHM{} \tbPH{várɪ́} \tbPS{v.~} \tbGE{to castrate} \tbPD{varoo, vara, vara, varibo, varaa, variba } 
\tbLX{varekpara} \tbHM{} \tbPH{vàrìkpàrà} \tbPS{adv.~} \tbGE{long and loose e.g. twine} 
\tbLX{varekpenɛ} \tbHM{} \tbPH{várɪ́kpɪ́nɛ́} \tbPS{n.~} \tbGE{thick paste from beans} \tbSG{varekpenoo} \tbTP{varekpennɛɛ} 
\tbLX{varema} \tbHM{} \tbPH{várɪ́má} \tbPS{adj.~} \tbSN \tbGE{tough e.g. meat, rope} \tbSN \tbGE{difficult} \tbXV{A pɔgelee na nɔŋe la laare kyɛ o boɔbo yɛlɛ varema.} \tbXE{That girl likes smiling but she is very difficult when it comes to dating her.} 
\tbLX{vare niŋe} \tbHM{} \tbPH{vàrɪ̀ níŋé} \tbPS{v.~} \tbGE{make a face; usually in disgust or anger} \tbPD{-, -, -, -, -, - } 
\tbLX{varevare} \tbHM{} \tbPH{vàrɪ̀várɪ́} \tbPS{v.~} \tbGE{to heckle} \tbPD{varevarɛɛ, varevara, varevara, varevareba, varevaraa, - } 
\tbLX{varevare} \tbHM{} \tbPH{vàrɪ̀várɪ́} \tbPS{v.~} \tbGE{to pull along} \tbPD{varevarɛɛ, varevara, varevara, varevareba, varevaraa, - } 
\tbLX{varoo} \tbHM{} \tbPH{várʊ̀ʊ̀} \tbPS{adj.~} \tbGE{tough e.g. meat} \tbVA{varema}
\tbLX{ve} \tbHM{} \tbPH{vɪ́} \tbPS{v.~} \tbGE{to knit} \tbPD{veɛɛ, verɛ, verɛ, verebɛ, veraa, - } 
\tbLX{vebo} \tbHM{} \tbPH{vɪ̀bʊ̀} \tbPS{adj.~} \tbGE{striking directly e.g. arrow} 
\tbLX{vedeɛ} \tbHM{} \tbPH{vɪ̀dɪ̀ɛ́} \tbPS{n.~} \tbGE{sugarcane} \tbOP{vederre} \tbTP{vedeerɛɛ} 
\tbLX{veene} \tbHM{} \tbPH{vɪ̀ɪ̀nɪ̀} \tbPS{n.~} \tbGE{a type of small bird with very lean legs (short form of daveene)} \tbOP{veemɛ} \tbTP{veenɛɛ} \tbVA{daveene}
\tbLX{veene} \tbHM{} \tbPH{vɪ́ɪ́nɪ́} \tbPS{v.~} \tbGE{to tie tightly} \tbPD{veenɛɛ, veenɛ, veenɛ, veenemɛ, veenaa, - } 
\tbLX{veɛle} \tbHM{} \tbPH{vɪ̀ɛ̀lɪ̀} \tbPS{v.~} \tbGE{to be beautiful} \tbPD{veɛlɛɛ, veɛlɛ, veɛlɛ, veɛle, -, - } 
\tbLX{veɛlɛ} \tbHM{} \tbPH{vɪ̀ɛ̀lɛ́} \tbPS{adj.~} \tbGE{beautiful} 
\tbLX{veɛlɛyeŋe} \tbHM{} \tbPH{vɪ̀ɛ̀lɛ́yéŋé} \tbPS{n.~} \tbGE{one with physical beauty only} 
\tbLX{veɛloŋ} \tbHM{} \tbPH{vɪ̀ɛ́lʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{beauty} \tbSN \tbGE{goodness} \tbSN \tbGE{light} 
\tbLX{veɛre} \tbHM{} \tbPH{vɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to push aside growth e.g in searching for something in an overgrown bush} \tbPD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa, - } 
\tbLX{veɛre} \tbHM{} \tbPH{vɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to fly low and slowly} \tbPD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa, - } 
\tbLX{veɛre} \tbHM{} \tbPH{vɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to evade e.g. a verbal attack} \tbPD{veɛrɛɛ, veɛrɛ, veɛrɛ, veɛrebɛ, veɛraa, - } 
\tbLX{veɛrenyɔge} \tbHM{} \tbPH{vɪ̀ɛ̀rɪ̀ɲɔ́gɪ́} \tbPS{n.~} \tbGE{rodent} 
\tbLX{velaa} \tbHM{} \tbPH{vɪ̀làà} \tbPS{adj.~} \tbSN \tbGE{good, nice} \tbSN \tbGE{beautiful} \tbOP{veɛle} 
\tbLX{vell} \tbHM{} \tbPH{vɪ̀ll} \tbPS{adj.~} \tbGE{full to the brim} 
\tbLX{velle} \tbHM{} \tbPH{vɪ̀llɪ̀} \tbPS{v.~} \tbGE{to hold the legs tightly in such a way as to make the person lose balance} \tbXV{A pɔge velle la a dɔɔ gbɛɛ ka o le.} \tbXE{The woman held the man’s legs so tightly that he lost balance and fell.} \tbPD{vellɛɛ, vellɛ, vellɛ, velbɛ, vellaa, - } 
\tbLX{velɔɔ} \tbHM{} \tbPH{vɪ̀lɔ̀ɔ̀} \tbPS{adj.~} \tbGE{turning instantly} 
\tbLX{venenne} \tbHM{} \tbPH{vɪ̀nɪ̀nnɪ̀} \tbPS{adj.~} \tbGE{elastic} 
\tbLX{venne} \tbHM{} \tbPH{vɪ́nnɪ́} \tbPS{v.~} \tbGE{to throw off (wrestling)} \tbPD{vennɛɛ, vennɛ, vennɛ, venemɛ, vennaa, - } 
\tbLX{venoo} \tbHM{} \tbPH{vɪ́nʊ̀ʊ̀} \tbPS{n.~} \tbGE{wine} 
\tbLX{venɔɔ} \tbHM{} \tbPH{vɪ̀nɔ̀ɔ̀} \tbPS{adj.~} \tbGE{thick and sticky e.g. grease} 
\tbLX{venveŋ} \tbHM{} \tbPH{vɪ́nvɪ́ŋ} \tbPS{n.~} \tbGE{climax} 
\tbLX{venveŋ} \tbHM{} \tbPH{vɪ́nvɪ́ŋ} \tbPS{adv.~} \tbGE{at the maximum} \tbXV{A daa naŋ da w ate kpɛ venveŋ saŋa la ka ba da moɔle a zuyɔɔ yɛlɛ.} \tbXE{It was at the climax of the market activities that they announced the payment of taxes.} 
\tbLX{vere} \tbHM{} \tbPH{vɪ̀rɪ̀} \tbPS{v.~} \tbGE{to shoot with a catapult} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ } 
\tbLX{vere} \tbHM{} \tbPH{vɪ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to sprain} \tbSN \tbGE{to dislocate (joint)} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ } 
\tbLX{vere} \tbHM{} \tbPH{vɪ̀rɪ̀} \tbPS{v.~} \tbGE{to make an effort, to strain} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, verebɛ } 
\tbLX{vere} \tbHM{} \tbPH{vɪ́rɪ́} \tbPS{v.~} \tbGE{to fence e.g. garden} \tbPD{verɛɛ, verɛ, verɛ, verebɛ, veraa, - } 
\tbLX{verevere} \tbHM{} \tbPH{vɪ̀rɪ̀vɪ́rɪ́} \tbPS{v.~} \tbGE{to pull very hard on e.g. tug of war} \tbPD{vereverɛɛ, vereverɛ, vereverɛ, vereverebɛ, vereveraa, - } 
\tbLX{verɛnverɛŋ} \tbHM{} \tbPH{vɪ̀rɛ̀nvɪ́rɛ́ŋ} \tbPS{adj.~} \tbGE{found everywhere} 
\tbLX{verɛverɛ} \tbHM{} \tbPH{vɪ̀rɛ̀vɪ̀rɛ̀} \tbPS{adj.~} \tbGE{scattered all over} 
\tbLX{vɛ᷉ɛ} \tbHM{} \tbPH{vɛ̀᷉ɛ̀᷉} \tbPS{adv.~} \tbGE{less sunny, dusky} 
\tbLX{vɛle} \tbHM{} \tbPH{vɛ̀lɪ̀} \tbPS{v.~} \tbGE{to become clear e.g. day, mind} \tbPD{vɛlɛɛ, vɛllɛ, vɛllɛ, vɛlebɛ, vɛllaa, - } 
\tbLX{vɛlevɛle} \tbHM{} \tbPH{vɛ̀lvɛ̀l} \tbPS{adj.~} \tbGE{very long and and wavy} 
\tbLX{vɛmm} \tbHM{} \tbPH{vɛ́mm} \tbPS{adv.~} \tbGE{smoothly} 
\tbLX{vɛnɛvɛnɛ} \tbHM{} \tbPH{vɛ̀nɛ̀vɛ̀nɛ̀} \tbPS{adj.~} \tbSN \tbGE{sticky e.g. honey} \tbSN \tbGE{oily} 
\tbLX{vɛnlɛmm} \tbHM{} \tbPH{vɛ́nlɛ́mm} \tbPS{adj.~} \tbGE{slim and beautiful} 
\tbLX{vɛŋe} \tbHM{} \tbPH{vɛ́ŋ} \tbPS{v.~} \tbSN \tbGE{to leave} \tbSN \tbGE{to let go} \tbSN \tbGE{to stop} \tbPD{vɛŋɛɛ, vɛnɛ, vɛnɛ, vɛnemɛ, vɛnaa, - } 
\tbLX{vɛre} \tbHM{} \tbPH{vɛ̀rɪ̀} \tbPS{v.~} \tbGE{to prepare mortar} \tbPD{vɛrɛɛ, vɛrɛ, vɛrɛ, vɛrebɛ, vɛraa, - } 
\tbLX{vɛre} \tbHM{} \tbPH{vɛ́rɪ́} \tbPS{v.~} \tbGE{to scatter thick liquid e.g. mortar} \tbPD{vɛrɛɛ, vɛrɛ, vɛrɛ, vɛrebɛ, vɛraa, - } 
\tbLX{vɛre} \tbHM{} \tbPH{vɛ̀rɪ̀} \tbPS{v.~} \tbGE{mix water with soil e.g mortar} \tbPD{-, -, -, -, -, - } 
\tbLX{vɛre} \tbHM{} \tbPH{vɛ́rɪ́} \tbPS{n.~} \tbGE{stony farmland} \tbOP{vɛrɛɛ} 
\tbLX{vɛrɛɛ} \tbHM{} \tbPH{vɛ̀rɛ̀ɛ̀} \tbPS{adj.~} \tbGE{lying low e.g. dew-laden grass} 
\tbLX{vɛvɛ} \tbHM{} \tbPH{vɛ́vɛ́} \tbPS{adj.~} \tbGE{glidingly e.g. a bird} 
\tbLX{vɛvɛre} \tbHM{} \tbPH{vɛ̀vɛ̀rɪ́} \tbPS{n.~} \tbGE{mud, marsh} \tbTP{vɛvɛrɛɛ} 
\tbLX{vi} \tbHM{} \tbPH{vî} \tbPS{n.~} \tbGE{shame, disgrace} \tbOP{viri} \tbTP{viree} 
\tbLX{vigi} \tbHM{} \tbPH{vígì} \tbPS{n.~} \tbGE{owl} \tbOP{vigiri} \tbTP{vigree} \tbVA{vig/vib}
\tbLX{vigi} \tbHM{} \tbPH{vígì} \tbPS{v.~} \tbGE{to throw out in a whole (solid waste)} \tbPD{vigee, vigre, vigre, vigribe, vigraa, - } 
\tbLX{viiri} \tbHM{} \tbPH{vììrì} \tbPS{v.~} \tbSN \tbGE{to turn around, to rotate} \tbSN \tbGE{to visit a place repeatedly} \tbPD{viiree, viire, viire, viiribe, viiraa, viiribe } 
\tbLX{vilenkyimm} \tbHM{} \tbPH{vílénkyímm} \tbPS{adj.~} \tbGE{twisted} 
\tbLX{vilenva} \tbHM{} \tbPH{vílénvà} \tbPS{n.~} \tbGE{going out of sight} \tbXV{A karema yeli ka o erɛ vilenva ka a karembiiri yaarɛɛ.} \tbXE{No sooner had the teacher left than the students dispersed.} 
\tbLX{vilenvileŋ} \tbHM{} \tbPH{vílénvíléŋ} \tbPS{adv.~} \tbGE{round, circular} 
\tbLX{vili} \tbHM{} \tbPH{vílí} \tbPS{v.~} \tbGE{to entwine, to tie, to wrap around} \tbPD{vilee, vile, vile, vilbe, vilaqa, - } 
\tbLX{vilime} \tbHM{} \tbPH{vílímé} \tbPS{adj.~} \tbGE{circular} \tbVA{vilenvileŋ}
\tbLX{virenvireŋ} \tbHM{} \tbPH{vírénvíréŋ} \tbPS{adv.~} \tbGE{totally e.g. refuse} 
\tbLX{viretaŋgbelli} \tbHM{} \tbPH{vìrèntáŋgbéllí} \tbPS{expression.~} \tbGE{your first and last} 
\tbLX{vitaŋkpulee} \tbHM{} \tbPH{vítáŋkpúléé} \tbPS{n.~} \tbGE{a type of small owl} \tbOP{vitaŋkpulli} 
\tbLX{vo} \tbHM{} \tbPH{vò} \tbPS{v.~} \tbGE{to make less crowded} \tbPD{voe, voro, voro, -, -, - } 
\tbLX{vogɛɛ} \tbHM{} \tbPH{vʊ̀gɛ̀ɛ̀} \tbPS{adv.~} \tbGE{voluminous e.g. pair of trousers} 
\tbLX{vokyɔ} \tbHM{} \tbPH{vʊ̀ʧɔ̀} \tbPS{n.~} \tbSN \tbGE{a type of tree with long fruit} \tbSN \tbGE{the fruit of this tree (used in kneading)} \tbOP{vokyɔre} \tbTP{vokyɔrɛɛ} 
\tbLX{volavola} \tbHM{} \tbPH{vʋ̀làvʋ̀là} \tbPS{adv.~} \tbGE{slimly} \tbVA{volɔvolɔ}
\tbLX{volɔvolɔ} \tbHM{} \tbPH{vʋ̀lɔ̀vʋ̀lɔ̀} \tbPS{adv.~} \tbGE{slimly} \tbVA{volavola}
\tbLX{vonne} \tbHM{} \tbPH{vʊ̀nnɪ̀} \tbPS{v.~} \tbGE{to throw far away} \tbPD{vonnɛɛ, vonna, vonna, vonema, vonnaa, - } 
\tbLX{voŋe} \tbHM{} \tbPH{vʊ́ŋ} \tbPS{v.~} \tbGE{to stay put and refuse to move} \tbPD{voŋɛɛ, vonɔ, vonɔ, vonema, vonaa, - } 
\tbLX{vo᷉o} \tbHM{} \tbPH{vʊ᷉ʊ᷉} \tbPS{v.~} \tbGE{to mix, to disturb e.g. water} \tbPD{voo᷉ɛ, voonɔ, voonɔ, voonema, voonaa, - } \tbVA{voŋe}
\tbLX{voone} \tbHM{} \tbPH{vʊ̀ʊ̀nɪ̀} \tbPS{v.~} \tbGE{to sound, to vibrate} \tbPD{voonɛɛ, voona/voonɔ, voona/voonɔ, -, -, - } 
\tbLX{vooraa} \tbHM{} \tbPH{vʊ́ʊ́ráá} \tbPS{n.~} \tbGE{a type of lumpy paste made from bean flour} \tbOP{voore} 
\tbLX{voore} \tbHM{} \tbPH{vʊ̀ʊ̀rɪ̀} \tbPS{v.~} \tbGE{echo, resound (noise)} \tbPD{voorɛɛ, voora, voora, vooreba, vooraa, - } 
\tbLX{voore} \tbHM{} \tbPH{vʊ́ʊ́rɪ́} \tbPS{v.~} \tbSN \tbGE{to breathe, to take a breath} \tbSN \tbGE{to be relieved, to be free, to rest} \tbXV{A saama kulee la ka N naa voore.} \tbXE{The guets have gone and now I'm relieved.} \tbPD{voorɛɛ, voora, voora, vooreba, vooraa, vooreba } 
\tbLX{voore fõõ} \tbHM{} \tbPH{vʊ́ʊ́rɪ́ fo᷉o᷉} \tbPS{v.~} \tbGE{to take a deep breath, to sigh} \tbPD{-, -, -, -, -, - } 
\tbLX{Vooronsoŋ} \tbHM{} \tbPH{vʊ́ʊ́rʊ́nsʊ́ŋ} \tbPS{n.~} \tbGE{Holy Spirit} 
\tbLX{vooroŋ} \tbHM{} \tbPH{vʊ́ʊ́rʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{life} \tbSN \tbGE{breathed air, breath} \tbOP{voora} \tbTP{voorɛɛ} 
\tbLX{voowoɔ} \tbHM{} \tbPH{vʊ́ʊ́wʊ́ɔ́} \tbPS{n.~} \tbGE{womb} \tbOP{voowoore} \tbTP{voowoorɛɛ} 
\tbLX{voɔ} \tbHM{} \tbPH{vʊ́ɔ̀} \tbPS{v.~} \tbGE{to pull out, to uproot} \tbPD{voɔɛ, voɔrɔ, voɔrɔ, voɔreba, vɔraa, - } 
\tbLX{voɔbale} \tbHM{} \tbPH{vʊ́ɔ́bàlɪ̀} \tbPS{n.~} \tbGE{arrow with many hooks} 
\tbLX{voɔloŋ} \tbHM{} \tbPH{vʊ̀ɔ̀lʊ́ŋ} \tbPS{n.~} \tbGE{slimy sap (from a type of tree) used to settle sediments in liquids} 
\tbLX{voɔloŋ} \tbHM{} \tbPH{vʊ̀ɔ̀lʊ́ŋ} \tbPS{adj.~} \tbGE{slimy} 
\tbLX{voɔre} \tbHM{} \tbPH{vʊ́ɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{opening, hole (through an object)} \tbSN \tbGE{a small door, usually at the back of a compound house (comparable to an emergency exit)} \tbSN \tbGE{small space} \tbXV{N daare ba nyɛ voɔre gaa a kuori.} \tbXE{I did not get the chance to attend the funeral.} \tbOP{vɔ᷉ɛ} \tbTP{vɔrɛɛ} 
\tbLX{voɔre} \tbHM{} \tbPH{vʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to be worn out, to be spent, to be tired and weak} \tbPD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\tbLX{voɔre} \tbHM{} \tbPH{vʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to get scorched e.g. plants during drought} \tbPD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\tbLX{voɔre} \tbHM{} \tbPH{vʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbGE{to smear} \tbPD{voɔrɛɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\tbLX{voraa} \tbHM{} \tbPH{vʋ́ráá} \tbPS{n.~} \tbGE{a hand drill; see kyaleere} 
\tbLX{vore} \tbHM{} \tbPH{vʊ́rɪ́} \tbPS{v.~} \tbGE{to make a hole through, to pierce e.g. ears, lips} \tbPD{vorɛɛ, vorɔ, vorɔ, voreba, voraa, voreba } 
\tbLX{vore} \tbHM{} \tbPH{vʊ́rɪ́} \tbPS{v.~} \tbGE{to weave} \tbPD{vorɛɛ, vorɔ, vorɔ, voreba, voraa, voreba } 
\tbLX{vore} \tbHM{} \tbPH{vʊ́rɪ̀} \tbPS{n.~} \tbGE{foetus} \tbOP{voɛ} \tbTP{vorɛɛ} 
\tbLX{vore} \tbHM{} \tbPH{vʊ́rɪ̀} \tbPS{adj.~} \tbGE{alive} \tbOP{voɛ} \tbTP{vorɛɛ} 
\tbLX{vori} \tbHM{} \tbPH{vórì} \tbPS{v.~} \tbGE{to rot, to spoil completely} \tbPD{voree, voro, voro, voribo, voraa, - } 
\tbLX{vorikporo} \tbHM{} \tbPH{vóríkpóró} \tbPS{adv.~} \tbGE{watery manner} 
\tbLX{vorovoro} \tbHM{} \tbPH{vòròvòrò} \tbPS{adj.~} \tbGE{watery} 
\tbLX{vɔ} \tbHM{} \tbPH{vɔ́} \tbPS{v.~} \tbGE{to perform a type of ritual for a dead male} \tbPD{vɔɛ, voɔrɔ, voɔrɔ, voɔreba, voɔraa, - } 
\tbLX{vɔga} \tbHM{} \tbPH{vɔ́gà} \tbPS{n.~} \tbGE{track, trail} \tbVA{vɔgɔ}
\tbLX{vɔgaa} \tbHM{} \tbPH{vɔ̀gáá} \tbPS{n.~} \tbGE{silk cotton tree} \tbOP{vɔgre} \tbTP{vɔgrɛɛ} 
\tbLX{vɔge} \tbHM{} \tbPH{vɔ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to take off} \tbSN \tbGE{to uncover} \tbPD{vɔgɛɛ, vɔgrɔ, vɔgrɔ, vɔgreba, vɔgraa, - } 
\tbLX{vɔge} \tbHM{} \tbPH{vɔ̀gɪ̀} \tbPS{v.~} \tbSN \tbGE{to scald} \tbSN \tbGE{to get burnt from hot substance} \tbPD{vɔgɛɛ, vɔgrɔ, vɔgrɔ, vɔgreba, vɔgraa, - } 
\tbLX{vɔgele} \tbHM{} \tbPH{vɔ̀glɪ̀} \tbPS{v.~} \tbSN \tbGE{to cover} \tbSN \tbGE{to put on top e.g. hat} \tbSN \tbGE{to overturn e.g. container} \tbPD{vɔglɛɛ, vɔglɔ, vɔglɔ, vɔgleba, vɔglaa, vɔgleba } 
\tbLX{vɔgɔ} \tbHM{} \tbPH{vɔ́gɔ̀} \tbPS{n.~} \tbGE{track, trail} \tbOP{vɔgre} \tbTP{vɔgrɛɛ} \tbVA{vɔga}
\tbLX{vɔlaa} \tbHM{} \tbPH{vɔ́láà} \tbPS{n.~} \tbGE{great desire for} 
\tbLX{vɔle} \tbHM{} \tbPH{vɔ́lɪ̀} \tbPS{n.~} \tbGE{great desire for meat; carnivorousness} \tbOP{vɔlɛɛ} \tbTP{vɔllɛɛ} 
\tbLX{vɔle} \tbHM{} \tbPH{vɔ́lɪ́} \tbPS{v.~} \tbGE{to swallow} \tbPD{vɔlɛɛ, vɔlɔ, vɔlɔ, vɔlebɔ, vɔlaa, vɔlebɔ } 
\tbLX{vɔll} \tbHM{} \tbPH{vɔ̀ll} \tbPS{adj.~} \tbGE{beautifully; usually in the state of growing up} 
\tbLX{vɔrekpɔrɔ} \tbHM{} \tbPH{vɔ̀rɪ̀kpɔ̀rɔ̀} \tbPS{adv.~} \tbSN \tbGE{wet and unwieldy} \tbXV{Saseɛ fue la ka saa meŋ mi kõɔ lɔre a sagere ka a zie e lɛ vɔrekpɔrɔ.} \tbXE{The rain which followed the rainstorm has soaked the garbage in the yard making it wet and disorderly.} \tbSN \tbGE{sound (onomatopoeic); fall by a long winding thing e.g. snake, rope} 
\tbLX{vu} \tbHM{} \tbPH{vú} \tbPS{v.~} \tbSN \tbGE{to crawl e.g. child} \tbSN \tbGE{to drag along the ground} \tbPD{vue, vuuro, vuuro, vuribo, vuuraa, vuribo } 
\tbLX{vubiri} \tbHM{} \tbPH{vúbìrí} \tbPS{n.~} \tbGE{mound for yam plant} \tbOP{vubie} \tbTP{vubiiree} 
\tbLX{vuduori} \tbHM{} \tbPH{vúdúórì} \tbPS{n.~} \tbGE{the act of making yam mounds} 
\tbLX{vuduoro} \tbHM{} \tbPH{vúdúóró} \tbPS{n.~} \tbGE{one who makes yam mounds} \tbOP{vuduoribo} 
\tbLX{vulaŋkaŋkpogili} \tbHM{} \tbPH{vùlàŋkǎŋkpóglí} \tbPS{n.~} \tbGE{final thing} \tbOP{vulaŋkaŋkpolo} \tbTP{vulaŋkaŋkpolee} 
\tbLX{vuloŋkpogili} \tbHM{} \tbPH{vùlòŋkpóglí} \tbPS{n.~} \tbGE{beetle} \tbOP{vuloŋkpogilo} \tbTP{vuloŋkpogilee} \tbVA{vulaŋkpogli}
\tbLX{vulotokpolo} \tbHM{} \tbPH{vùlòtókpóló} \tbPS{n.~} \tbGE{piece of work meant to be one’s work for the day} \tbOP{vulotokpolli} \tbTP{vulotokpolee} 
\tbLX{vulotokpolo} \tbHM{} \tbPH{vùlòtókpóló} \tbPS{n.~} \tbGE{stanza (literature)} \tbOP{vulotokpolli} \tbTP{vulotokpolee} 
\tbLX{vunni} \tbHM{} \tbPH{vùnnì} \tbPS{v.~} \tbGE{to throw far away} \tbPD{vunnee, vunno, vunno, vunimo, vunnaa, - } 
\tbLX{vuo} \tbHM{} \tbPH{vùó} \tbPS{n.~} \tbSN \tbGE{space,  passage} \tbSN \tbGE{chance, time} \tbSN \tbGE{occasion} 
\tbLX{vuo} \tbHM{} \tbPH{vùó} \tbPS{adj.~} \tbGE{hollow} \tbOP{vori} \tbTP{voree} 
\tbLX{vuoli} \tbHM{} \tbPH{vúólì} \tbPS{v.~} \tbSN \tbGE{to bark (dog)} \tbSN \tbGE{to hoot (to put to shame)} \tbPD{vuolee, vuolo, vuolo, vuolibo, vuolaa, - } \tbVA{wuoli}
\tbLX{vuoli kaa} \tbHM{} \tbPH{vúólì káá} \tbPS{v.~} \tbGE{to peer at} \tbPD{-, -, -, -, -, - } 
\tbLX{vutaŋkpulee} \tbHM{} \tbPH{vútàŋkpùlèè} \tbPS{n.~} \tbGE{gate pillar} \tbOP{vutaŋkpulli} \tbTP{vutaŋkpullee} 
\tbLX{vuu} \tbHM{} \tbPH{vúù} \tbPS{v.~} \tbGE{to drag} \tbXV{Ba vuu la a kamaana boɔraa boŋyeni na.} \tbXE{They dragged that one sack of maize.} \tbPD{vuue, vuuro, vuuro, vuuribo, vuuraa, - } 
\tbLX{vu᷈u} \tbHM{} \tbPH{vù᷉ú᷉} \tbPS{n.~} \tbGE{fire} \tbSG{vuunuu} \tbTP{vuunee} 
\tbLX{vu᷉udindigili} \tbHM{} \tbPH{vù᷉ú᷉díndíglí} \tbPS{n.~} \tbSN \tbGE{electricity} \tbSN \tbGE{electric light} \tbVA{vu᷉udigili}
\tbLX{vu᷉udiraa} \tbHM{} \tbPH{vù᷉ù᷉dìráà} \tbPS{n.~} \tbGE{flame} \tbOP{vu᷉udirri} \tbTP{vu᷉udiree} 
\tbLX{vu᷉udirri} \tbHM{} \tbPH{vù᷉ùdìrrì} \tbPS{n.~} \tbGE{a stream of burning} 
\tbLX{vu᷈ugyɛle} \tbHM{} \tbPH{vù᷈ù᷈gyɛ́lɪ́} \tbPS{n.~} \tbGE{lightbulb} \tbOP{vù᷈ùgyɛlɛ} \tbTP{vù᷈ùgyɛlɛɛ} \tbVA{vu᷈u᷈nimiri}
\tbLX{vuuli} \tbHM{} \tbPH{vúúlí} \tbPS{v.~} \tbGE{to singe} \tbPD{vuulee, vuulo, vuulo, vuulibo, vuulaa, - } 
\tbLX{vũulunsaa} \tbHM{} \tbPH{vùùlúnsáà} \tbPS{n.~} \tbGE{first rain after the dry season} 
\tbLX{vuuluŋ} \tbHM{} \tbPH{vúúlúŋ} \tbPS{n.~} \tbGE{an animal believed to have supernatural power and usually attack people in the night} \tbVA{shortened form of valinvuuluŋ}
\tbLX{vuuluŋ} \tbHM{} \tbPH{vúúlúŋ} \tbPS{n.~} \tbGE{short form of valinvuuluŋ (soot)} \tbVA{shortened form of valinvuuluŋ}
\tbLX{vuunimiri} \tbHM{} \tbPH{vùùnímírì} \tbPS{n.~} \tbGE{lightbulb} \tbOP{vuunimie} \tbTP{vuunimiiree} \tbVA{vu᷈ugyɛle}
\tbLX{vuuri} \tbHM{} \tbPH{vùùrì} \tbPS{v.~} \tbGE{to feign; to pretend or attempt to do something (which doesn't happen)} \tbPD{vuuree, vuuro, vuuro, vuuribo, vuuraa, vuuribo } 
\tbLX{vuuri} \tbHM{} \tbPH{vúúrì} \tbPS{n.~} \tbGE{small oar-like stick for stirring porridge} \tbOP{vue} \tbTP{vuuree} 
\tbLX{vuuri} \tbHM{} \tbPH{vúúrì} \tbPS{n.~} \tbGE{yam farm} \tbOP{vue} \tbTP{vuuree} 
\tbLX{vu᷉usaalbiri} \tbHM{} \tbPH{vù᷉ùsáálbírí} \tbPS{n.~} \tbGE{burning charcoal} \tbVA{vu᷉usaale}
\tbLX{vu᷉usaale} \tbHM{} \tbPH{vù᷉ù᷉sáálɪ́} \tbPS{n.~} \tbGE{burning coal} \tbOP{vu᷉usaala} \tbTP{vu᷉usaaloo} \tbVA{vu᷉usaalbiri}
\tbLX{vu᷉uvaale} \tbHM{} \tbPH{vù᷉úváálɪ́} \tbPS{n.~} \tbGE{a type of bird which hovers around bushes that are being burnt to catch fleeing insects} \tbOP{vu᷉uvaala} 
\tbLX{vu᷈uyɛlɛ} \tbHM{} \tbPH{vù᷈ù᷉gyɛ́lɛ́} \tbPS{n.~} \tbGE{lightbulb} \tbOP{vu᷈u᷈gyɛlɛ} \tbTP{vu᷈u᷈gyɛlɛɛ} 
\tbLX{vu᷈uzeɛle} \tbHM{} \tbPH{vù᷉ù᷉zɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{flying burnt particles from a fire e.g. grass} \tbSG{vu᷈uzeɛloo} \tbOP{vu᷈uzeɛlɛ} \tbTP{vu᷈uzeɛlɛɛ} 
\tbLX{vu᷈uzele} \tbHM{} \tbPH{vù᷈ù᷈zɪ̀lɪ́} \tbPS{n.~} \tbGE{flame, tongue of fire} \tbOP{vu᷈uzelɛ} \tbTP{vu᷈uzelɛɛ} 
\tbLX{Wa} \tbHM{} \tbPH{wǎ} \tbPS{n.~} \tbGE{name of the regional headquarters of the Upper West Region} 
\tbLX{wa} \tbHM{} \tbPH{wà} \tbPS{v.~} \tbGE{to come} \tbPD{waɛ, waana, waana, wanema, waanaa, wanema } 
\tbLX{wa} \tbHM{} \tbPH{wá} \tbPS{part.~} \tbGE{by the time; before the end of a given time} \tbXV{Te na wa tara ka zivɛlmɛ la.} \tbXE{By the time we arrive, it will be daybreak.} 
\tbLX{waa} \tbHM{} \tbPH{wàá} \tbPS{n.~} \tbGE{yam} \tbOP{waare} \tbTP{waarɛɛ} \tbVA{nyuuri}
\tbLX{waa} \tbHM{} \tbPH{wàá} \tbPS{auxiliary verb.~} \tbSN \tbGE{to be} \tbXV{A tomɔ waa la yaga.} \tbXE{The work is plenty.} \tbSN \tbGE{to resemble, to be like} \tbXV{A biiri waa ŋa yeere.} \tbXE{The children are like twins.} 
\tbLX{waabo} \tbHM{} \tbPH{wáábʊ́} \tbPS{n.~} \tbGE{snake} \tbOP{weere/wamine} \tbTP{weerɛɛ} 
\tbLX{waale} \tbHM{} \tbPH{wáálɪ̀} \tbPS{v.~} \tbGE{to whisper} \tbPD{waalɛɛ, waala, waala, waaleba, waalaa, waaleba } 
\tbLX{waale} \tbHM{} \tbPH{wáálɪ̀} \tbPS{v.~} \tbGE{to parboil} \tbPD{waalɛɛ, waala, waala, waaleba, waalaa, waaleba } 
\tbLX{waale} \tbHM{} \tbPH{wáálɪ̀} \tbPS{v.~} \tbGE{to bruise slightly} \tbPD{waalɛɛ, waala, waala, waaleba, waalaa, waaleba } 
\tbLX{waaleŋ} \tbHM{} \tbPH{wààlɪ̂ŋ} \tbPS{n.~} \tbGE{a type of gunpowder (imported)} 
\tbLX{waaleŋ} \tbHM{} \tbPH{wààlɪ̂ŋ} \tbPS{adj.~} \tbGE{like a native of Wa} 
\tbLX{waaloŋ} \tbHM{} \tbPH{wààlʊ̂ŋ} \tbPS{n.~} \tbSN \tbGE{nature, character} \tbSN \tbGE{resemblance} 
\tbLX{Waaloo} \tbHM{} \tbPH{wáálʊ̀ʊ̀} \tbPS{n.~} \tbGE{native of Wa} \tbOP{waala} 
\tbLX{waane} \tbHM{} \tbPH{wáánɪ̀} \tbPS{v.~} \tbSN \tbGE{to dress tightly (pull trousers to navel level)} \tbSN \tbGE{to propel, to spin} \tbPD{waanɛɛ, waana, waana, waanema, waanaa, - } 
\tbLX{waare} \tbHM{} \tbPH{wààrɪ̀} \tbPS{v.~} \tbGE{be touched by (liquid, powder)} \tbXV{A bie ɔŋ la tɛne waare o ma.} \tbXE{The child has smeared the mother with soil.} \tbPD{waarɛɛ, waara, waara, waareba, waaraa, - } 
\tbLX{waare} \tbHM{} \tbPH{wáárɪ́} \tbPS{v.~} \tbSN \tbGE{to burn the cover (when you attempt to burn green grass, it is only the blades that burn leaving the straw)} \tbXV{A moɔ da e la mɔmaara lɛnso ka ba di kyɛ waare.} \tbXE{The grass was green and that is why when they attempted burning it it did not burn.} \tbXV{A naŋkpaana nyege la a mɔmaara kyɛ a ba di, a waarɛɛ la.} \tbXE{The hunter tried to burn the green grass, but it did not burn well; it only burnt halfway.} \tbSN \tbGE{to be painful} \tbXV{Ba tɔ la a bie fele ka o eŋɛ zaa waare.} \tbXE{They hit on his sore and it was quite painful.} \tbPD{waarɛɛ, waara, waara, waareba, -, - } 
\tbLX{waaroŋ} \tbHM{} \tbPH{wààrʊ́ŋ} \tbPS{n.~} \tbGE{organised group hunting} 
\tbLX{wadɛre} \tbHM{} \tbPH{wàdɛ́rɪ́} \tbPS{n.~} \tbGE{} \tbOP{wadɛrɛɛ} \tbVA{nyudɛre}
\tbLX{wadɛrɛɛ} \tbHM{} \tbPH{wàdɛ́rɛ́ɛ́} \tbPS{n.~} \tbGE{} \tbTP{wadɛrrɛɛ} \tbVA{nyudɛrɛɛ}
\tbLX{wae} \tbHM{} \tbPH{wáɪ́} \tbPS{adj.~} \tbGE{nine} 
\tbLX{waee} \tbHM{} \tbPH{wáɪ́ɪ́} \tbPS{adv.~} \tbGE{at once} 
\tbLX{wagere} \tbHM{} \tbPH{wágrɪ̀} \tbPS{n.~} \tbGE{time for something, period} \tbOP{wagerɛɛ} 
\tbLX{wagere} \tbHM{} \tbPH{wàgrɪ̀} \tbPS{v.~} \tbGE{to bite on the surface (dog)} \tbPD{wagerɛɛ, wagera, wagera, wagereba, wageraa, - } 
\tbLX{wagyɛ} \tbHM{} \tbPH{wàdzɛ̀} \tbPS{n.~} \tbGE{blanket, cloth} \tbOP{wagyɛre} \tbTP{wagyɛrɛɛ} 
\tbLX{walaa} \tbHM{} \tbPH{wáláà} \tbPS{n.~} \tbGE{antelope} \tbOP{walle} \tbTP{wallɛɛ} 
\tbLX{walanse} \tbHM{} \tbPH{wálánsɪ̀} \tbPS{n.~} \tbGE{radio, wireless set} \tbOP{walansere} \tbTP{walanserɛɛ} 
\tbLX{wale} \tbHM{} \tbPH{wálɪ́} \tbPS{n.~} \tbGE{sweat} \tbTP{walɛɛ/wallɛɛ} 
\tbLX{walpelaa} \tbHM{} \tbPH{wàlpɪ̀làà} \tbPS{n.~} \tbGE{a type of antelope} \tbOP{walpeɛle} 
\tbLX{walpieli} \tbHM{} \tbPH{wàlpíélí} \tbPS{n.~} \tbGE{} \tbSG{walpieluu} \tbOP{walpiele} \tbTP{walpielee} \tbVA{walpieluu}
\tbLX{walpieluu} \tbHM{} \tbPH{wàpíélúú} \tbPS{n.~} \tbGE{} \tbOP{walpiele} \tbTP{walpielee} \tbVA{walpieli}
\tbLX{walsugo} \tbHM{} \tbPH{wálsúgó} \tbPS{n.~} \tbGE{group of deer} \tbOP{walsugri} \tbTP{walsugree} 
\tbLX{walsunni} \tbHM{} \tbPH{wàlsúnnì} \tbPS{n.~} \tbGE{perspiration, extreme sweat} \tbTP{wasunnee} 
\tbLX{walwoɔ} \tbHM{} \tbPH{wàlwʊ́ɔ́} \tbPS{n.~} \tbSN \tbGE{newcomer, novice} \tbSN \tbGE{a bag made of antelope skin} \tbOP{walwoore} \tbTP{walwooɛɛ} 
\tbLX{wanwɛ} \tbHM{} \tbPH{wánwɛ́} \tbPS{v.~} \tbGE{to be gone forever; to fail to return as expected} \tbXV{Ba toŋ la a bie ka o deɛ gaa te wanwɛ.} \tbXE{When they sent the child, he left and never returned.} \tbPD{wanwɛɛ, wanaweɛrɛ, wanweɛrɛ, wanweɛrebɛ, wanweɛraa, - } 
\tbLX{waŋene} \tbHM{} \tbPH{wáŋɪ́nɪ́} \tbPS{n.~} \tbGE{great one} \tbOP{wanne} \tbTP{wannɛɛ} 
\tbLX{waŋmane} \tbHM{} \tbPH{wàŋmánɪ́} \tbPS{n.~} \tbGE{a hundred yams} \tbOP{waŋmama} \tbTP{waŋmamɛɛ} 
\tbLX{ware} \tbHM{} \tbPH{wárɪ̀} \tbPS{n.~} \tbGE{coins; money} \tbOP{warre} \tbTP{warɛɛ} \tbVA{warebiri}
\tbLX{ware} \tbHM{} \tbPH{wárí} \tbPS{adv.~} \tbGE{all gone (people on a journey)} \tbXV{A teŋɛ dɔbɔ zaaŋ e weɛ, ware.} \tbXE{All the men in the town have gone to the farm; not even one was left behind.} 
\tbLX{ware} \tbHM{} \tbPH{wárɪ́} \tbPS{n.~} \tbGE{long draught} \tbTP{warɛɛ} 
\tbLX{warebiri} \tbHM{} \tbPH{wárɪ́bírì} \tbPS{n.~} \tbGE{see ware} \tbOP{warebie} \tbTP{warebiiree} \tbVA{ware}
\tbLX{warema} \tbHM{} \tbPH{wárɪ́má} \tbPS{n.~} \tbGE{brick, block} \tbOP{waremare} \tbTP{waremarɛɛ} 
\tbLX{wasaaloŋ} \tbHM{} \tbPH{wàsáálʊ́ŋ} \tbPS{n.~} \tbGE{water yam} \tbTP{wasaalɛɛ.} 
\tbLX{wasege} \tbHM{} \tbPH{wàsɪ́gɪ́} \tbPS{n.~} \tbGE{yam barn} \tbOP{wasegre} \tbTP{wasegrɛɛ} \tbVA{nyusege}
\tbLX{wasɔŋɔ} \tbHM{} \tbPH{wàsɔ́ŋ} \tbPS{n.~} \tbGE{a basket of yam} \tbOP{wasɔnne} \tbTP{wasɔnnɛɛ} 
\tbLX{wasugo} \tbHM{} \tbPH{wàsúgó} \tbPS{n.~} \tbGE{see sugo (termites)} \tbOP{wasugiri} \tbTP{wasugiree} 
\tbLX{waweɛlɛ} \tbHM{} \tbPH{wàwɪ́ɛ́lɛ́} \tbPS{n.~} \tbGE{seed yam} \tbSG{waweɛloo} \tbTP{waweɛlɛɛ} \tbVA{nyuweɛlɛ}
\tbLX{wazɪɛ} \tbHM{} \tbPH{wázɪ̀ɛ́} \tbPS{n.~} \tbGE{a type of brown snake} \tbOP{wazeere} \tbTP{wazeerɛɛ} 
\tbLX{we} \tbHM{} \tbPH{wé} \tbPS{interjection.~} \tbGE{expression of despair} 
\tbLX{webe} \tbHM{} \tbPH{wébé} \tbPS{interjection.~} \tbGE{exclamation of exhaustion} 
\tbLX{we᷉derɛ} \tbHM{} \tbPH{wɪ́᷉dɪ́rɛ́} \tbPS{n.~} \tbGE{leader, forerunner} \tbOP{we᷉derebɛ} \tbTP{we᷉derebɛɛ} 
\tbLX{weene} \tbHM{} \tbPH{wɪ̀ɪ̀nɪ̀} \tbPS{v.~} \tbGE{to begin to boil} \tbPD{weenɛɛ, weenɛ, weenɛ, -, -, - } 
\tbLX{weɛ} \tbHM{} \tbPH{wɪ̀ɛ́} \tbPS{n.~} \tbSN \tbGE{farm} \tbSN \tbGE{wilderness} \tbOP{wɛre} \tbTP{wɛrɛɛ} 
\tbLX{weɛ} \tbHM{} \tbPH{wɪ̀ɛ̀} \tbPS{v.~} \tbSN \tbGE{to make a furrow/trench e.g for water to pass} \tbSN \tbGE{cut open; usually part of the body in a bid to remove a foreign body or pus} \tbSN \tbGE{to push growth aside e.g. in a search for something in a thick growth} \tbPD{weɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\tbLX{weɛ} \tbHM{} \tbPH{wɪ́ɛ́} \tbPS{v.~} \tbGE{to save} \tbXV{A kpogili la weɛ o ka lɛ naane o zu da na ŋmare la.} \tbXE{He would have broken his head but he was saved by the crash helmet.} \tbPD{weɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\tbLX{weɛ} \tbHM{} \tbPH{wɪ́ɛ́} \tbPS{adj.~} \tbGE{not related to} 
\tbLX{we᷉ɛ} \tbHM{} \tbPH{wí᷉ɛ̀᷉} \tbPS{adv.~} \tbGE{in front, ahead} \tbXV{Dabuo la na de o sakuuri maakyeebo we᷉ɛ.} \tbXE{Dabuo will lead his school at the parade ground.} 
\tbLX{we᷉ɛ} \tbHM{} \tbPH{wɪ̀᷉ɛ̀᷉} \tbPS{v.~} \tbGE{to be wild, to be untamed, to be out of control} \tbPD{we᷉ɛɛ, weɛnɛ, weɛnɛ, weɛnemɛ, weɛnaa, - } 
\tbLX{weɛle} \tbHM{} \tbPH{wɪ́ɛ́lɪ́} \tbPS{v.~} \tbGE{to branch out in lobes (yam seed)} \tbPD{weɛlɛɛ, weɛlɛ, weɛlɛ, weɛlbɛ, weɛlaa, - } 
\tbLX{weɛle} \tbHM{} \tbPH{wɪ́ɛ́lɪ́} \tbPS{v.~} \tbGE{to miss out, to lose} \tbPD{weɛlɛɛ, weɛlɛ, weɛlɛ, weɛlbɛ, weɛlaa, - } 
\tbLX{weɛle} \tbHM{} \tbPH{wɪ̀ɛ́lɪ́} \tbPS{n.~} \tbGE{twins} \tbTP{weɛlɛɛ} 
\tbLX{weɛlɛ} \tbHM{} \tbPH{wɪ̀ɛ̀lɛ́} \tbPS{n.~} \tbGE{mistletoe (a parasitic plant whose fruit is used to produce gum)} \tbTP{weɛlɛɛ} 
\tbLX{weɛlɛ} \tbHM{} \tbPH{wɪ́ɛ̀lɛ̀} \tbPS{adj.~} \tbSN \tbGE{in vain, for nothing} \tbXV{Fo gɛrɛ la be weɛlɛ, o koŋ leɛ ka fo.} \tbXE{You are going there for nothing; she will not mind you.} \tbSN \tbGE{free} \tbXV{A doroba de la a pɔgenaŋaa gaa ne a dɔɔta weɛlɛ.} \tbXE{The driver took the old woman to the hospital free of charge.} 
\tbLX{weɛmɛ} \tbHM{} \tbPH{wɪ̀ɛ̀mɛ́} \tbPS{adj.~} \tbGE{wild, untamed} 
\tbLX{weɛre} \tbHM{} \tbPH{wɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to search through e.g. brush} \tbPD{weɛrɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\tbLX{wegere} \tbHM{} \tbPH{wɪ̀grɪ̀} \tbPS{v.~} \tbGE{to squeeze through e.g. through a thick crowd} \tbPD{wegerɛɛ, wegerɛ, wegerɛ, wegerebɛ, wegeraa, - } 
\tbLX{welaa} \tbHM{} \tbPH{wɪ́láá} \tbPS{adj.~} \tbGE{harbouring evil intentions} 
\tbLX{wele} \tbHM{} \tbPH{wɪ́lɪ́} \tbPS{v.~} \tbGE{to rummage} \tbPD{welɛɛ, welɛ, welɛ, welebɛ, welaa, - } 
\tbLX{wele} \tbHM{} \tbPH{wɪ́lɪ́} \tbPS{v.~} \tbGE{to dismantle} \tbPD{welɛɛ, welɛ, welɛ, welebɛ, welaa, - } 
\tbLX{wele} \tbHM{} \tbPH{wɪ́lɪ́} \tbPS{v.~} \tbGE{to reduce in size e.g. obesity} \tbPD{welɛɛ, welɛ, welɛ, welebɛ, welaa, - } 
\tbLX{wele} \tbHM{} \tbPH{wɪ́lɪ́} \tbPS{n.~} \tbGE{a type of ornament for the arm} \tbOP{welɛ} \tbTP{welɛɛ} 
\tbLX{welɛnwelɛŋ} \tbHM{} \tbPH{wɪ̀lɛ́nwɪ̀lɛ́ŋ} \tbPS{adv.~} \tbGE{vigorous, rolling (boiling liquid)} 
\tbLX{welɛwelɛ} \tbHM{} \tbPH{wɪ̀lɛ̀wɪ̀lɛ̀} \tbPS{adv.~} \tbGE{intermittent spackle} 
\tbLX{weli} \tbHM{} \tbPH{wélí} \tbPS{n.~} \tbGE{snail} \tbOP{wele} \tbTP{welee} 
\tbLX{weliweli} \tbHM{} \tbPH{wélwél} \tbPS{adv.~} \tbGE{clearly visible} 
\tbLX{welle} \tbHM{} \tbPH{wɪ́llɪ̀} \tbPS{v.~} \tbSN \tbGE{to scrutinize, to examine closely} \tbSN \tbGE{to rummage} \tbPD{wellɛɛ, wellɛ, wellɛ, welebɛ, wellaa, - } 
\tbLX{welli} \tbHM{} \tbPH{wéllì} \tbPS{v.~} \tbGE{to make very clear, clean, or neat} \tbPD{wellee, welle, welle, welibe, wellaa, - } 
\tbLX{wenne} \tbHM{} \tbPH{wɪ̀nnɪ̀} \tbPS{v.~} \tbGE{to go into the interior e.g. of bush or a crowd} \tbPD{wennɛɛ, wennɛ, wennɛ, wenemɛ, wennaa, - } 
\tbLX{were-bare} \tbHM{} \tbPH{wìrɪ̀-bàrì} \tbPS{v.~} \tbGE{to ignore} \tbPD{were-barɛɛ, werɛ-bara, were-bara, were-bareba, were-baraa, - } 
\tbLX{werekao} \tbHM{} \tbPH{wírɪ́káʊ̀} \tbPS{n.~} \tbGE{wound, mark, defect} \tbXV{Lɔɔre la le ne o ka a o gbɛre nyɛ werekao.} \tbXE{A car crash has caused a wound on his leg.} \tbOP{werekarre} \tbTP{werekarɛɛ} 
\tbLX{werewere} \tbHM{} \tbPH{wɪ́rɪ́wɪ́rɪ́} \tbPS{adv.~} \tbGE{smartly, quickly} 
\tbLX{werɛ} \tbHM{} \tbPH{wɪ̀rɛ́} \tbPS{adj.~} \tbGE{small species, not of high breed (groundnuts)} \tbTP{werɛɛ} 
\tbLX{werɛ} \tbHM{} \tbPH{wɪ́rɛ́} \tbPS{n.~} \tbGE{mountain} \tbTP{werɛɛ/werrɛɛ} 
\tbLX{werɛ} \tbHM{} \tbPH{wɪ́rɛ́} \tbPS{adj.~} \tbGE{in vain} \tbXV{O baloo na zaa kyɛ werɛ la, o ba nyɛ a libie.} \tbXE{All his suffering has been in vain; he did not get the money.} 
\tbLX{werɛwerɛ} \tbHM{} \tbPH{wɪ́rɛ́wɪ́rɛ́} \tbPS{adj.~} \tbGE{tiny things} 
\tbLX{werɛwerɛ} \tbHM{} \tbPH{wɪ̀rɛ̀wɪ̀rɛ̀} \tbPS{adj.~} \tbGE{coarse (flour)} 
\tbLX{weri} \tbHM{} \tbPH{wérí} \tbPS{adv.~} \tbGE{valuable, good quantity} 
\tbLX{weroo} \tbHM{} \tbPH{wɪ́rʊ́ʊ́} \tbPS{adj.~} \tbGE{small in stature} \tbOP{werɛ} 
\tbLX{wɛ} \tbHM{} \tbPH{wɛ̀} \tbPS{adv.~} \tbSN \tbGE{of cause} \tbXV{Ba nyɛ la a bonzuuri o dieŋ, ona la  zu wɛ.} \tbXE{They found the stolen items in his room, of cause he stole them.} \tbSN \tbGE{then} \tbXV{Ka fo wa pagra foŋ de we᷉ɛ᷉ wɛ.} \tbXE{If you are in a haste then take the lead.} 
\tbLX{wɛ} \tbHM{} \tbPH{wɛ́} \tbPS{v.~} \tbSN \tbGE{to go bad} \tbXV{Sugo kpɛ la a dɔɔ wasege, a waare zaaŋ wɛ.} \tbXE{Termites have invaded the man's yam barn; all the yams are destroyed.} \tbSN \tbGE{to be lost forever} \tbPD{wɛɛ, weɛrɛ, weɛrɛ, weɛrebɛ, weɛraa, - } 
\tbLX{wɛ} \tbHM{} \tbPH{wɛ́} \tbPS{adj.~} \tbGE{not strong or courageous} \tbXV{Nyɛ dɔɔ naŋ wɛ, haali o koŋ to᷉ɔ᷉ zɛle a pelee.} \tbXE{See how useless the man is; he cannot even lift the small basket.} 
\tbLX{wɛbaa} \tbHM{} \tbPH{wɛ̀báá} \tbPS{n.~} \tbGE{fox} \tbOP{wɛbaare} \tbTP{wɛbaarɛɛ} 
\tbLX{wɛdoŋɔ} \tbHM{} \tbPH{wɛ̀dʊ́ŋɔ́} \tbPS{n.~} \tbGE{a non-domesticated animal} \tbOP{wɛdonne.} \tbTP{wɛdonnɛɛ.} \tbVA{mɔdoŋɔ, wɛdoŋa,}
\tbLX{wɛge} \tbHM{} \tbPH{wɛ́gɪ̀} \tbPS{v.~} \tbGE{to hatch} \tbPD{wɛgɛɛ, wɛgrɛ, wɛgrɛ, wɛgrebɛ, wɛgraa, wɛgrebɛ } 
\tbLX{wɛge} \tbHM{} \tbPH{wɛ́gɪ̀} \tbPS{v.~} \tbGE{to multiply} \tbPD{wɛgɛɛ, wɛgrɛ, wɛgrɛ, wɛgrebɛ, wɛgraa, wɛgrebɛ } 
\tbLX{wɛge} \tbHM{} \tbPH{wɛ̀gɪ̀} \tbPS{v.~} \tbGE{to remove with a pointed implement e.g. corn from the cob} \tbPD{wɛgɛɛ, wɛgrɛ, wɛgrɛ, wɛgrebɛ, wɛgraa, - } 
\tbLX{wɛge} \tbHM{} \tbPH{wɛ́gɪ̀} \tbPS{n.~} \tbGE{beam, large plank used in making mud roofing} \tbOP{wɛgere} \tbTP{wɛgerɛɛ} \tbVA{wɛgɛ, dawɛge}
\tbLX{wɛgere} \tbHM{} \tbPH{wɛ̀grɪ̀} \tbPS{v.~} \tbGE{to force out, to squeeze out e.g. seed from pod} \tbPD{wɛgerɛɛ, wɛgerɛ, wɛgerɛ, wɛgerebɛ, wɛgeraa, - } 
\tbLX{wɛgɛ} \tbHM{} \tbPH{wɛ́gɛ́} \tbPS{n.~} \tbGE{beam} \tbOP{wɛgre} \tbTP{wɛgɛɛ} \tbVA{wɛge}
\tbLX{wɛkaraa} \tbHM{} \tbPH{wɛ̀káráá} \tbPS{n.~} \tbGE{mounds of a farm made without first clearing the land} \tbOP{wɛkare} \tbTP{wɛkarɛɛ} 
\tbLX{wɛle} \tbHM{} \tbPH{wɛ́lɪ̀} \tbPS{v.~} \tbSN \tbGE{to split or break into two} \tbSN \tbGE{to separate, to disperse} \tbPD{wɛlɛɛ, wɛlɛ, wɛlɛ, wɛlebɛ, wɛlaa, wɛlebɛ } 
\tbLX{wɛlɛmbɛll} \tbHM{} \tbPH{wɛ̀lɛ̀mbɛ̀ll} \tbPS{adv.~} \tbGE{massively moving} 
\tbLX{wɛnaabo} \tbHM{} \tbPH{wɛ̀náábʊ̀} \tbPS{n.~} \tbGE{buffalo} \tbOP{wɛnii} \tbTP{wɛniinee} 
\tbLX{wɛnoɔ} \tbHM{} \tbPH{wɛ́nʊ́ɔ́} \tbPS{n.~} \tbGE{dark grey hen} \tbOP{wɛnoore} \tbTP{wɛnoorɛɛ} 
\tbLX{wɛnu} \tbHM{} \tbPH{wɛ́nú} \tbPS{n.~} \tbGE{an arm of the main body of hunters} \tbOP{wɛnuuri.} \tbTP{wɛnuuree.} 
\tbLX{wɛnwɛrɛɛ} \tbHM{} \tbPH{wɛ̌nwɛ́rɛ́ɛ́} \tbPS{n.~} \tbGE{a type of small plant with edible leaves} \tbSG{wɛnwɛroo} \tbTP{wɛnwɛrrɛɛ} 
\tbLX{wɛnyuo} \tbHM{} \tbPH{wɛ́nyúó} \tbPS{n.~} \tbGE{leopard} \tbOP{wɛnyuuri} \tbTP{wɛnyuuree} 
\tbLX{wɛre} \tbHM{} \tbPH{wɛ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to cut open} \tbSN \tbGE{to operate} \tbSN \tbGE{to split, to saw} \tbSN \tbGE{to make a furrow or gutter} \tbPD{wɛrɛɛ, wɛrɛ, wɛrɛ, wɛrebɛ, wɛraa, - } 
\tbLX{wi} \tbHM{} \tbPH{wí} \tbPS{v.~} \tbGE{to carve; to engrave} \tbPD{wie, wiire, wiire, wiiribe, wiiraa, - } 
\tbLX{wi᷉eu} \tbHM{} \tbPH{wí᷉é᷉ú᷉} \tbPS{adj.~} \tbGE{fast, quick} \tbVA{wieŋ}
\tbLX{wigiri} \tbHM{} \tbPH{wígrì} \tbPS{v.~} \tbGE{to toss to turn over (grain in a basket)} \tbPD{wiigree, wigire, wigire, wigiribe, wigiraa, - } 
\tbLX{wiiruŋ} \tbHM{} \tbPH{wíírúŋ} \tbPS{n.~} \tbGE{design} 
\tbLX{willi} \tbHM{} \tbPH{wìllì} \tbPS{v.~} \tbSN \tbGE{to overturn to capsize} \tbXV{A saseɛ da waa la kpeɛŋaa, a da da a gboe zaa willi a mane poɔ.} \tbXE{The wind was so strong that it capsized all the boats on the sea.} \tbSN \tbGE{to heave over} \tbPD{willee, wille, wille, wilibe, willaa, wilibe } 
\tbLX{wire} \tbHM{} \tbPH{wíré} \tbPS{adv.~} \tbGE{gone far} 
\tbLX{wirewire} \tbHM{} \tbPH{wíréwíré} \tbPS{adj.~} \tbGE{far away} 
\tbLX{wiri} \tbHM{} \tbPH{wìrí} \tbPS{n.~} \tbGE{horse} \tbOP{wie} \tbTP{wiree/wirree} 
\tbLX{wiri} \tbHM{} \tbPH{wìrì} \tbPS{v.~} \tbGE{to turn inside out e.g. pocket} \tbPD{wiree, wire, wire, wiribe, wiraa, - } 
\tbLX{wiri-buuli} \tbHM{} \tbPH{wìrì-búúló} \tbPS{n.~} \tbGE{bangles for a horse} \tbOP{wiri-buulo} \tbTP{wiri-buulee} 
\tbLX{wiridaa} \tbHM{} \tbPH{wìrìdáá} \tbPS{n.~} \tbGE{male horse} \tbOP{wiridaare} \tbTP{wiridaarɛɛ} 
\tbLX{wiriduoruŋ} \tbHM{} \tbPH{wìrìdúórúŋ} \tbPS{n.~} \tbSN \tbGE{jaundice} \tbSN \tbGE{yellow fever} 
\tbLX{wiri-eɛraa} \tbHM{} \tbPH{wìrì-ɪ́ɛ́ráá} \tbPS{n.~} \tbGE{metal spur worn on the heel for prodding the horse} \tbOP{wiri-eɛre} 
\tbLX{wirigoɔ} \tbHM{} \tbPH{wìrìgʊ́ɔ́} \tbPS{n.~} \tbGE{a type of thorny plant} \tbOP{wirigoore} \tbTP{wirigoorɛɛ} 
\tbLX{wirikogi} \tbHM{} \tbPH{wìrìkógí} \tbPS{n.~} \tbGE{horse saddle} \tbOP{wirikogri} \tbTP{wirikogree} 
\tbLX{wirikuoluu} \tbHM{} \tbPH{wìrìkúólúú} \tbPS{n.~} \tbGE{a type of honey-making insect} \tbOP{wirikuolo} \tbTP{wirikuolee} 
\tbLX{wirilee} \tbHM{} \tbPH{wìrìléé} \tbPS{n.~} \tbGE{colt} \tbOP{wiribilii} 
\tbLX{wirinɔbɔŋ} \tbHM{} \tbPH{wìrìnɔ́bɔ́ŋ} \tbPS{n.~} \tbGE{metal ring placed arond the muzzle of a horse} \tbOP{wirinɔbɔnne} \tbVA{wirinɔbɔŋenɔ}
\tbLX{wirinyaŋaa} \tbHM{} \tbPH{wìrìɲáŋáá} \tbPS{n.~} \tbGE{female horse} \tbOP{wirinyaŋene} \tbTP{wirinyaŋenɛɛ} 
\tbLX{wirizoore} \tbHM{} \tbPH{wìrìzʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{horse tail} \tbOP{wirizoɛ} 
\tbLX{wirizoore} \tbHM{} \tbPH{wìrìzʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{a type of grass} \tbOP{wirizoɛ} 
\tbLX{wirizoore} \tbHM{} \tbPH{wìrìzʊ́ʊ́rɪ́} \tbPS{n.~} \tbGE{a type of sickness} \tbOP{wirizoɛ} 
\tbLX{wirizɔɔnɔ} \tbHM{} \tbPH{wìrìzɔ́ɔ́nɔ́} \tbPS{n.~} \tbGE{horse rider} \tbOP{wirizɔɔnema} \tbTP{wirizɔɔmɛɛ} 
\tbLX{wofe᷉ɛ} \tbHM{} \tbPH{wʊ̀fɪ́ɛ́} \tbPS{n.~} \tbGE{a bag made from animal skin without hair} \tbOP{wofeɛne} \tbTP{wofeɛnɛɛ} \tbVA{wofeɛmaa}
\tbLX{wofeɛmaa} \tbHM{} \tbPH{wʊ́fɪ́ɛ́máá} \tbPS{n.~} \tbGE{bag made from animal skin without hair} \tbOP{wofeɛne} \tbTP{wofeɛnɛɛ/wofeɛmɛɛ} \tbVA{wofe᷉ɛ, wofuoraa, wokpele}
\tbLX{wofuoraa} \tbHM{} \tbPH{wʊ̀fúóráá} \tbPS{n.~} \tbGE{} \tbOP{wofuori} \tbTP{wofuoree} \tbVA{wofeɛmaa}
\tbLX{Wogi} \tbHM{} \tbPH{wògí} \tbPS{n.~} \tbGE{a town in the Upper West Region of Ghana} 
\tbLX{wogi} \tbHM{} \tbPH{wógì} \tbPS{adj.~} \tbGE{long, tall} 
\tbLX{wogiri} \tbHM{} \tbPH{wógrí} \tbPS{v.~} \tbGE{to cause to shake e.g. carried container of pebbles} \tbPD{wogiree, wogiro, wogiro, wogiribo, wogiraa, - } \tbVA{wɔgere}
\tbLX{wogiri} \tbHM{} \tbPH{wògrì} \tbPS{v.~} \tbGE{to steal cunningly} \tbPD{wogiree, wogiro, wogiro, wogiribo, wogiraa, - } 
\tbLX{wogiroŋ} \tbHM{} \tbPH{wògròŋ} \tbPS{adj.~} \tbGE{large, big e.g. container} \tbOP{wogiroŋ wogiroŋ} 
\tbLX{wogoko} \tbHM{} \tbPH{wógókó} \tbPS{adv.~} \tbGE{intense (painful bite)} 
\tbLX{wokpale} \tbHM{} \tbPH{wʊ̀kpálɪ́} \tbPS{n.~} \tbGE{} \tbOP{wokpala} \tbTP{wokpalɛɛ} \tbVA{wofeɛmaa}
\tbLX{wola} \tbHM{} \tbPH{wʊ́là} \tbPS{interrogative pronoun.~} \tbSN \tbGE{how} \tbXV{Wola ka te na e maale a pɔgebilii bayi bama yelbawontaa ŋa?} \tbXE{How are we going to settle the misunderstanding between the two girls?.} \tbSN \tbGE{why} \tbXV{Wola la ka fo la zeŋe zomm lɛ?} \tbXE{Why are you sitting so quietly?.} \tbSN \tbGE{what} \tbXV{O yeli ka wola?} \tbXE{What did s/he say?.} 
\tbLX{wolaa} \tbHM{} \tbPH{wʊ́láá} \tbPS{adj.~} \tbGE{dangerous} \tbOP{wolle} 
\tbLX{woli} \tbHM{} \tbPH{wólí} \tbPS{n.~} \tbGE{leather rope worn on the waist} \tbOP{wolo} \tbTP{wollee} 
\tbLX{woliwoli} \tbHM{} \tbPH{wòlwòlí} \tbPS{v.~} \tbGE{to be restless (of a person)} \tbPD{wolwolee, wolwolo, wolwolo, wolwolibo, wolwolaa, wolwolibo } 
\tbLX{wolle} \tbHM{} \tbPH{wʊ̀llɪ̀} \tbPS{v.~} \tbGE{to rattle, to speak fluently (langauge)} \tbPD{wollɛɛ, wolla, wolla, wolba, wollaa, - } 
\tbLX{wolle} \tbHM{} \tbPH{wʊ́llɪ́} \tbPS{v.~} \tbGE{to be mischievous} \tbPD{-, wolla, wolba, wollaa, -, - } 
\tbLX{wolomboll} \tbHM{} \tbPH{wòlòmbòll} \tbPS{adj.~} \tbGE{in disarray or confusion e.g. mob} 
\tbLX{wolonwoloŋ} \tbHM{} \tbPH{wólónwólóŋ} \tbPS{adv.~} \tbGE{haphazard, in a troubled state} 
\tbLX{woloŋ} \tbHM{} \tbPH{wʊ́lʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{disease} \tbSN \tbGE{problem} 
\tbLX{wolowolo} \tbHM{} \tbPH{wólówóló} \tbPS{adj.~} \tbGE{restive} 
\tbLX{wolɔ} \tbHM{} \tbPH{wʊ́lɔ̀} \tbPS{v.~} \tbGE{to deceive, to hoax, to humbug} \tbPD{wolɛɛ, wolɔ, wola, woleba, wolaa, - } 
\tbLX{wonaa} \tbHM{} \tbPH{wónáá} \tbPS{n.~} \tbGE{that which pesters} 
\tbLX{wontaa} \tbHM{} \tbPH{wǒntáá} \tbPS{n.~} \tbGE{understanding between people} 
\tbLX{woŋi} \tbHM{} \tbPH{wòŋ} \tbPS{v.~} \tbGE{to hear} \tbXV{N ba teɛre ka o woŋ a yɛlɛ te naŋ yeli.} \tbXE{I don't think she has heard what we said.} \tbPD{woŋee, wono, wono, wonimo, wonaa, wonimo } \tbSE{woŋ noɔ} \tbGE{to enjoy, to be glad} \tbSE{woŋ tuo} \tbGE{to suffer} 
\tbLX{woŋi} \tbHM{} \tbPH{wòŋ} \tbPS{v.~} \tbGE{to be deaf} \tbXV{Ka fo toboŋ beɛrɛ ka fo ba kaa a zu velaa fo na woŋ la.} \tbXE{If you have a disease of the ears and you do not take good care of them, you will be deaf.} \tbPD{woŋee, wono, wono, wonimo, wonaa, wonimo } \tbSE{woŋ noɔ} \tbGE{to enjoy, to be glad} \tbSE{woŋ tuo} \tbGE{to suffer} 
\tbLX{woŋi} \tbHM{} \tbPH{wòŋ} \tbPS{v.~} \tbGE{to feel, to suffer} \tbXV{Ba ŋmɛ la baa ka a woŋi ka o nyɛ, bini.} \tbXE{When they hit the dog it felt so much pain that it defecated.} \tbPD{woŋee, wono, wono, wonimo, wonaa, wonimo } \tbSE{woŋ noɔ} \tbGE{to enjoy, to be glad} \tbSE{woŋ tuo} \tbGE{to suffer} 
\tbLX{woŋi} \tbHM{} \tbPH{wòŋ} \tbPS{v.~} \tbGE{to be of the correct quantity} \tbXV{Zenɛ, Ayuo zeɛre ŋa nyɛnnoo ba woŋ.} \tbXE{Today, there was not the right amount of salt in Ayuo's soup.} \tbPD{woŋee, wono, wono, wonimo, wonaa, wonimo } \tbSE{woŋ noɔ} \tbGE{to enjoy, to be glad} \tbSE{woŋ tuo} \tbGE{to suffer} 
\tbLX{woŋo} \tbHM{} \tbPH{wòŋó} \tbPS{n.~} \tbGE{deaf person} \tbOP{wonni} \tbTP{wonnee} \tbVA{woŋi}
\tbLX{woo} \tbHM{} \tbPH{wʊ́ʊ̀} \tbPS{n.~} \tbGE{water iguana} \tbOP{woore} \tbTP{woorɛɛ} 
\tbLX{woo} \tbHM{} \tbPH{wóó} \tbPS{interjection.~} \tbGE{a cry of pity} 
\tbLX{woore} \tbHM{} \tbPH{wʊ̀ʊ̀rɪ́} \tbPS{n.~} \tbGE{suffering} 
\tbLX{woɔ} \tbHM{} \tbPH{wʊ́ɔ́} \tbPS{n.~} \tbGE{bag} \tbOP{woore} \tbTP{woorɛɛ} 
\tbLX{woɔ} \tbHM{} \tbPH{wʊ̀ɔ̀} \tbPS{v.~} \tbSN \tbGE{to remove from the top of e.g. basket of grain} \tbXV{A seŋka᷉a᷉ paale la a pelee zuo, yɛ woɔ ka a ta yaara.} \tbXE{The basket is too full; remove some of the groundnuts so it does not spill.} \tbSN \tbGE{to become smaller} \tbXV{Dabuo da ba poɔrɔ kyɛ beri ayi ŋa o woɔ la o zaa.} \tbXE{Dabuo used to be plump, but recently he has gotten smaller.} \tbPD{woɔɛ, woɔrɔ, woɔrɔ, woɔreba, woɔraa, - } 
\tbLX{woɔle} \tbHM{} \tbPH{wʊ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to remove top matter; to skim e.g. yeast from pito} \tbPD{woɔlɛɛ, woɔla, woɔla, woɔleba, woɔlaa, - } 
\tbLX{wore} \tbHM{} \tbPH{wʊ́rɪ́} \tbPS{adv.~} \tbGE{too much; excessive} 
\tbLX{woroto} \tbHM{} \tbPH{wʊ̀rʊ̀tʊ̀} \tbPS{adv.~} \tbGE{crashingly} 
\tbLX{worɔworɔ} \tbHM{} \tbPH{wʊ̀rɔ̀wʊ̀rɔ̀} \tbPS{adv.~} \tbGE{full of lumps} 
\tbLX{woyaara} \tbHM{} \tbPH{wʊ̀yáárá} \tbPS{n.~} \tbGE{part of an animal-skin bag} \tbTP{woyaarɛɛ} 
\tbLX{wɔge} \tbHM{} \tbPH{wɔ̀gɪ̀} \tbPS{v.~} \tbGE{to treat fresh skin in preparation for making a bag} \tbPD{wɔgɛɛ, wɔgrɔ, wɔgrɔ, wɔgreba, wɔgraa, - } 
\tbLX{wɔge} \tbHM{} \tbPH{wɔ̀gɪ̀} \tbPS{v.~} \tbGE{to cause to suffer} \tbPD{wɔgɛɛ, wɔgrɔ, wɔgrɔ, wɔgreba, wɔgraa, - } 
\tbLX{wɔgele} \tbHM{} \tbPH{wɔ̀glɪ̀} \tbPS{v.~} \tbGE{to physically suffer inwardly intensely} \tbPD{wɔglɛɛ, wɔglɔ, wɔglɔ, wɔgleba, wɔglaa, - } 
\tbLX{wɔgele} \tbHM{} \tbPH{wɔ̀glɪ̀} \tbPS{v.~} \tbGE{to be burnt by touching a hot object} \tbPD{wɔglɛɛ, wɔglɔ, wɔglɔ, wɔgleba, wɔglaa, - } 
\tbLX{wɔgele} \tbHM{} \tbPH{wɔ̀glɪ̀} \tbPS{v.~} \tbGE{to treat beans (by soaking and drying)} \tbPD{wɔglɛɛ, wɔglɔ, wɔglɔ, wɔgleba, wɔglaa, - } 
\tbLX{wɔgere} \tbHM{} \tbPH{wɔ́grɪ́} \tbPS{v.~} \tbGE{to shake e.g. objects in a tin} \tbPD{wɔgrɛɛ, wɔgrɔ, wɔgrɔ, wɔgreba, wɔgraa, - } \tbVA{wogiri}
\tbLX{wɔne} \tbHM{} \tbPH{wɔ́nɪ́} \tbPS{n.~} \tbSN \tbGE{small swelling} \tbSN \tbGE{pimple} \tbOP{wɔmɔ} \tbTP{wɔnnɛɛ} 
\tbLX{wɔne} \tbHM{} \tbPH{wɔ́nɪ́} \tbPS{n.~} \tbGE{fruit} \tbOP{wɔmɔ} \tbTP{wɔnnɛɛ} 
\tbLX{wɔŋe} \tbHM{} \tbPH{wɔ́ŋ} \tbPS{v.~} \tbSN \tbGE{to bear fruit} \tbSN \tbGE{to develop rashes, pimples} \tbPD{wɔŋɛɛ, wɔnɔ, wɔnɔ, wɔnema, wɔnaa, - } 
\tbLX{wɔŋkampɔre} \tbHM{} \tbPH{wɔ́ŋkámpɔ́rɪ́} \tbPS{n.~} \tbSN \tbGE{a yamlike plant that bears fruit which is boiled and eaten} \tbSN \tbGE{the fruit of this plant} \tbOP{wɔŋkampɔrɔ} \tbTP{wɔŋkampɔrɛɛ} \tbVA{feree}
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ̀ɔ́} \tbPS{n.~} \tbGE{lame person} \tbOP{ wɔɔre} 
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ́ɔ́} \tbPS{v.~} \tbGE{to provide (food for the kitchen)} \tbPD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ́ɔ́} \tbPS{v.~} \tbGE{to pluck all the fruits from a tree} \tbPD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ́ɔ́} \tbPS{v.~} \tbGE{to destroy} \tbPD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ́ɔ́} \tbPS{v.~} \tbGE{to whip mercilessly (figurative)} \tbPD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ̀ɔ̀} \tbPS{v.~} \tbGE{to be lame} \tbPD{wɔɔɛ, wɔɔrɔ, wɔɔrɔ, wɔɔreba, wɔɔraa, - } 
\tbLX{wɔɔ} \tbHM{} \tbPH{wɔ́ɔ̀} \tbPS{n.~} \tbGE{elephant} \tbOP{wɔɔre} \tbTP{wɔɔrɛɛ} 
\tbLX{wɔɔgbɛre} \tbHM{1} \tbPH{wɔ́ɔ̀gbɛ́rɪ̀} \tbPS{n.~} \tbGE{elephantiasis} 
\tbLX{wɔɔgbɛre} \tbHM{2} \tbPH{wɔ́ɔ̀gbɛ́rɪ̀} \tbPS{n.~} \tbSN \tbGE{leg of an elephant} \tbSN \tbGE{difficult task (figurative)} \tbOP{wɔɔgbɛɛ} \tbTP{wɔɔgbɛrɛɛ} 
\tbLX{wɔɔkye} \tbHM{} \tbPH{wɔ̀ɔ́ʧɪ̀} \tbPS{n.~} \tbGE{watch} \tbOP{wɔɔkyere} \tbTP{wɔɔkyerɛɛ} 
\tbLX{wɔɔnu} \tbHM{} \tbPH{wɔ́ɔ́nù} \tbPS{n.~} \tbGE{trunk of an elephant} \tbOP{wɔɔnuuri} \tbTP{wɔɔnuuree} 
\tbLX{wɔɔnyene} \tbHM{} \tbPH{wɔ́ɔ́ɲɪ́nɪ̀} \tbPS{n.~} \tbGE{tusk of an elephant} \tbOP{wɔɔnyemɛ} \tbTP{wɔɔnyennɛɛ/wɔɔnyemɛɛ} 
\tbLX{wɔɔreŋ} \tbHM{} \tbPH{wɔ̀ɔ́rɪ́ŋ} \tbPS{n.~} \tbSN \tbGE{lameness} \tbSN \tbGE{the act of limping} \tbVA{wɔɔleŋ}\tbSE{} \tbGE{} 
\tbLX{wulee} \tbHM{} \tbPH{wúléé} \tbPS{n.~} \tbGE{whistle ; fife, flute} \tbOP{wubilii} \tbTP{wullee} 
\tbLX{wuli} \tbHM{} \tbPH{wùlì} \tbPS{v.~} \tbSN \tbGE{to show, to point to} \tbSN \tbGE{to direct, to guide} \tbSN \tbGE{to teach, to advise} \tbPD{wulee, wullo, wullo, wulibo, wulaa, wulibo } 
\tbLX{wuliwuli} \tbHM{} \tbPH{wùlwúlí} \tbPS{v.~} \tbGE{to rub over gently but fast} \tbPD{wuliwulee, wuliwulo, wuliwulo, wuliwulibo, wuliwulaa, - } 
\tbLX{wuliwullo} \tbHM{} \tbPH{wùlwúlló} \tbPS{n.~} \tbGE{teacher; one who directs} \tbOP{wulwulbo} \tbTP{wulwulbee} 
\tbLX{wuluu} \tbHM{} \tbPH{wúlúú} \tbPS{n.~} \tbSN \tbGE{teaching, education} \tbSN \tbGE{the act of showing} 
\tbLX{wuluu bonteɛre} \tbHM{} \tbPH{wúlúú bʊ́ntɪ́ɛ́rɪ́} \tbPS{n.~} \tbGE{teaching or learning materials} \tbSG{wuluu bonteɛraa} 
\tbLX{wuluu yelsɛgreminni} \tbHM{} \tbPH{wúlúú yèlsɛ́grɪ́bínní} \tbPS{n.~} \tbGE{lesson notes, lesson plan} 
\tbLX{wuo} \tbHM{} \tbPH{wùò} \tbPS{v.~} \tbGE{to gather, to collect} \tbPD{-, wuoro, wuoro, wuoribo, wuoraa, wuori } 
\tbLX{wuo} \tbHM{} \tbPH{wúò} \tbPS{v.~} \tbGE{to expose (a secret)} \tbPD{wuoe, wuoro, wuoro, wuoribo, wuoraa, - } 
\tbLX{wuo} \tbHM{} \tbPH{wúò} \tbPS{v.~} \tbGE{to formally introduce dirge singing at the funeral} \tbPD{wuoe, wuoro, wuoro, wuoribo, wuoraa, - } 
\tbLX{wuoli} \tbHM{} \tbPH{wúólì} \tbPS{v.~} \tbSN \tbGE{to despise, to hoot at} \tbSN \tbGE{to bark} \tbPD{wuolee, wuolo, wuolo, wuolibo, wuolaa, - } \tbVA{vuoli}
\tbLX{wuoraa} \tbHM{} \tbPH{wúóráá} \tbPS{n.~} \tbSN \tbGE{an implement for gathering} \tbSN \tbGE{a round-shaped piece of calabash with a hole in the middle used like a funnel in putting powdered tobacco into small containers} \tbOP{wuori} 
\tbLX{wurimo} \tbHM{} \tbPH{wúrímó} \tbPS{n.~} \tbGE{small, round cakes made from beans} \tbSG{wurimuu} \tbTP{wurimoree} 
\tbLX{wurowuro} \tbHM{} \tbPH{wùròwùrò} \tbPS{adv.~} \tbGE{not well cooked; underdone (round beans)} 
\tbLX{wuu} \tbHM{} \tbPH{wùù} \tbPS{adv.~} \tbSN \tbGE{continuously} \tbXV{Yuoni ŋa saa miire la wuu lɛ.} \tbXE{This year it has been raining unceasingly.} \tbSN \tbGE{having many instances} \tbXV{Zenɛ, koe deɛ waa la wuu lɛ.} \tbXE{Today, funerals are commonplace.} 
\tbLX{wuuri} \tbHM{} \tbPH{wúúrí} \tbPS{v.~} \tbSN \tbGE{to hoot (with laughter)} \tbSN \tbGE{to shout in excitement} \tbPD{wuuree, wuuro, wuuro, wuuribo, wuuraa, - } 
\tbLX{ya} \tbHM{} \tbPH{yá} \tbPS{v.~} \tbGE{to be mad, to be insane} \tbPD{yae,, yaara, yaara, yareba, yaaraa, - } 
\tbLX{yaa} \tbHM{} \tbPH{yàá} \tbPS{n.~} \tbGE{tomb} \tbOP{yaare} \tbTP{yarɛɛ} 
\tbLX{yaa} \tbHM{} \tbPH{yáà} \tbPS{v.~} \tbGE{used with 'sane'; to take revenge, to retaliate} \tbPD{yaaɛ, yaara, yaara, yaareba, yaaraa, yaareba } 
\tbLX{yaa} \tbHM{} \tbPH{yàà} \tbPS{interj.~} \tbGE{expression indicating disbelief} 
\tbLX{yaa} \tbHM{} \tbPH{yàà} \tbPS{v.~} \tbGE{remove from liquid} \tbPD{-, -, -, -, -, - } 
\tbLX{yaa} \tbHM{} \tbPH{yàà} \tbPS{v.~} \tbGE{to remove from liquid} \tbPD{yaaɛ, yaara, yaara, yaareba, yaaraa, - } 
\tbLX{yaa} \tbHM{} \tbPH{yàà} \tbPS{use word.~} \tbGE{please} \tbXV{Yaa, vɛŋe ka o meŋ tu o deme.} \tbXE{Oh, do allow him to tell his part of the story too.} 
\tbLX{yaa} \tbHM{} \tbPH{yáá} \tbPS{adv.~} \tbGE{really?} \tbXV{O yeli ka boŋ yaa?} \tbXE{What did she really say?.} 
\tbLX{yaa} \tbHM{} \tbPH{yáá} \tbPS{n.~} \tbGE{a strap; usually fixed to something for carrying e.g a bag} \tbOP{yaare} \tbTP{yaarɛɛ} 
\tbLX{yaale} \tbHM{} \tbPH{yáálɪ̀} \tbPS{v.~} \tbGE{to nurse (a baby)} \tbPD{yaalɛɛ, yaala, yaala, yaaleba, yaalaa, - } 
\tbLX{yaale} \tbHM{} \tbPH{yáálɪ̀} \tbPS{v.~} \tbGE{to swing a container in circles} \tbPD{yaalɛɛ, yaala, yaala, yaaleba, yaalaa, - } 
\tbLX{yaale} \tbHM{} \tbPH{yáálɪ̀} \tbPS{v.~} \tbGE{to get to hoplessness} \tbPD{yaalɛɛ, yaala, yaala, yaaleba, yaalaa, - } 
\tbLX{yaane} \tbHM{} \tbPH{yáánɪ́} \tbPS{n.~} \tbSN \tbGE{greeting(s)} \tbXV{Dagaaba ba deɛnɛ ne ba yaane puoruu.} \tbXE{The Dagaaba do not joke with their greetings.} \tbSN \tbGE{thanks, gratitude} \tbXV{Yɛ taa la yaane puoruu ne yɛ tonsonne yɛ naŋ toŋe.} \tbXE{You deserve our gratitude for your good work.} \tbTP{yaanɛɛ} 
\tbLX{yaane} \tbHM{} \tbPH{yáánɪ́} \tbPS{n.~} \tbGE{exhaustion} \tbXV{N koŋ la to᷉ɔ᷉ gaa yaane kpɛ ma la.} \tbXE{I can't go again; I'm tired.} \tbTP{yaanɛɛ} 
\tbLX{yaantulimo} \tbHM{} \tbPH{yáántúlímó} \tbPS{n.~} \tbGE{great-grandchild} \tbTP{yantulimee} 
\tbLX{yaara} \tbHM{} \tbPH{yààrá} \tbPS{adv.~} \tbGE{easily} \tbXV{Kyaanaa ba yaara ŋmaroo.} \tbXE{A mirror breaks easily.} 
\tbLX{yaare} \tbHM{} \tbPH{yààrɪ̀} \tbPS{v.~} \tbSN \tbGE{to spill, to scatter, to spread} \tbXV{A boɔraa la puri lɛnso ka a a kyi yaare a sori zu zaa.} \tbXE{The bag is burst and that is why the millet is scattered all over the road.} \tbSN \tbGE{to enlarge, to widen} \tbXV{A pelee ŋa noɔre yaare gaŋ la n boɔbo.} \tbXE{The mouth of this basket is too large for my liking.} \tbSN \tbGE{to perform final rites of widowhood} \tbXV{Ka banaŋ ba yaare kodɔɔ bee kopɔge o ba taa sori na toŋ sɛŋgaŋ toma.} \tbXE{Until the final widowhood rites are performed neither the widow nor the widower can engage in sex.} \tbSN \tbGE{to pester (typically used with noɔre)} \tbXV{Ba yaare la n noɔre ne ba pɔge na yɛlɛ ka N te bale.} \tbXE{They pestered me about their woman's case until I became fed up.} \tbPD{yaarɛɛ, yaara, yaara, yaareba, yaaraa, yaareba } 
\tbLX{yaaroŋ} \tbHM{} \tbPH{yààrʊ̂ŋ} \tbPS{n.~} \tbGE{salt} \tbVA{nyɛnnoŋ}
\tbLX{ya᷉ayiri} \tbHM{} \tbPH{yá᷉á᷉yírì} \tbPS{n.~} \tbGE{grandparent's house} \tbOP{ya᷉ayie} \tbTP{ya᷉ayiree} 
\tbLX{yabapa} \tbHM{} \tbPH{yàbàpà} \tbPS{adv.~} \tbGE{resoundingly e.g. noise of a large crowd} \tbXV{A pɔgebeɛlebɛ kyɛ la yieluŋ yabapa pa pa tu a sori gɛrɛ ne a pɔge.} \tbXE{The bridemaids sang earnestly along the road as they took the bride along.} 
\tbLX{Yaga} \tbHM{} \tbPH{yágá} \tbPS{n.~} \tbGE{name of a town in the Dagaaba land} 
\tbLX{yaga} \tbHM{} \tbPH{yágà} \tbPS{adj.~} \tbGE{many, much, plenty} \tbOP{yagere} \tbTP{yagerɛɛ} \tbVA{yage}
\tbLX{yage} \tbHM{} \tbPH{yágɪ̀} \tbPS{v.~} \tbSN \tbGE{to dredge to enlarge or deepen} \tbXV{Uoni saŋa Dagaaba maŋ yage la baa, a na nyɛ ko᷉ɔ᷉.} \tbXE{During the dry season, the Dagaaba dredge wells in order to get water.} \tbSN \tbGE{to increase in size e.g. a sore} \tbXV{Ka onaŋ ba gaa asibiti o fele na na yage la.} \tbXE{If she does not go to the hospital, her sore will develop into an ulcer.} \tbPD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } 
\tbLX{yage} \tbHM{} \tbPH{yàgɪ̀} \tbPS{v.~} \tbSN \tbGE{to take off e.g. a shirt} \tbSN \tbGE{to take down from a hanging position e.g. a bag from a hook} \tbPD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } \tbSE{yage bonsuuraa} \tbGE{to undress} 
\tbLX{yage} \tbHM{} \tbPH{yàgɪ̀} \tbPS{v.~} \tbGE{to show signs of butter forming e.g in the last stage in the preparation of shea butter} \tbPD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } \tbSE{yage bonsuuraa} \tbGE{to undress} 
\tbLX{yage} \tbHM{} \tbPH{yágɪ́} \tbPS{v.~} \tbGE{to remove objects from a container e.g. bag, hole} \tbXV{N ba boɔrɔ ka N yage a woɔ poɔ boma, a taa la dabe᷉ɛ.} \tbXE{I don't want to empty the contents of the bag because they are frightening.} \tbPD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } 
\tbLX{yage} \tbHM{} \tbPH{yágɪ́} \tbPS{v.~} \tbGE{to open one's eyes widely (in a concentrated look)} \tbXV{A bie nyɛ la a saana a deɛ yage nimie eŋ.} \tbXE{When the child saw the stranger, he gave him a fixed look.} \tbPD{yagɛɛ, yagra, yagra, yagreba, yagraa, - } 
\tbLX{yagekɔɔlare} \tbHM{} \tbPH{yágɪ́kɔ́ɔ́lárɪ́} \tbPS{n.~} \tbGE{jawbone} \tbOP{yagekɔɔlaɛ} \tbTP{yagekɔɔlarɛɛ} \tbVA{yagekɔɔre}
\tbLX{yagekɔɔre} \tbHM{} \tbPH{yàgɪ̀kɔ́ɔ́rɪ́} \tbPS{n.~} \tbGE{cheek} \tbOP{yagekɔbɔ} \tbTP{yagekɔɔrɛɛ} \tbVA{yagekɔɔlare}
\tbLX{yagelaa} \tbHM{} \tbPH{yágɪ́láà} \tbPS{n.~} \tbSN \tbGE{a rope holder hung from the ceiling to hold objects} \tbSN \tbGE{a basket with calabashes that is hung on a yagelaa} \tbOP{yagele} \tbTP{yagelɛɛ} 
\tbLX{yagele} \tbHM{} \tbPH{yáglɪ̀} \tbPS{v.~} \tbGE{to hang (up)} \tbPD{yagelɛɛ, yagela, yagela, yageleba, yagelaa, - } 
\tbLX{yageraa} \tbHM{} \tbPH{} \tbPS{n.~} \tbGE{clay} \tbOP{yagere} \tbVA{yagere}
\tbLX{yagere} \tbHM{} \tbPH{yágrɪ́} \tbPS{v.~} \tbGE{to separate stones from grain} \tbPD{yagerɛɛ, yagera, yagera, yagereba, yageraa, - } 
\tbLX{yagere} \tbHM{} \tbPH{yágrɪ́} \tbPS{n.~} \tbGE{modelling clay} \tbTP{yagerɛɛ} \tbVA{yageraa}
\tbLX{yagere} \tbHM{} \tbPH{yàgrɪ́} \tbPS{n.~} \tbGE{cheek} \tbOP{yaga} \tbTP{yagrɛɛ} 
\tbLX{yakpolo} \tbHM{} \tbPH{yàkpólò} \tbPS{n.~} \tbGE{graveyard} \tbTP{yakpolee} 
\tbLX{yammo} \tbHM{} \tbPH{yámmʊ́} \tbPS{n.~} \tbSN \tbGE{bile} \tbSN \tbGE{gallbladder} \tbOP{yanne} \tbTP{yaanɛɛ} 
\tbLX{yan-yaa} \tbHM{} \tbPH{yán-yáà} \tbPS{n.~} \tbGE{insane person} \tbOP{yan-yaare} \tbTP{yan-yaarɛɛ} 
\tbLX{yanyaare} \tbHM{} \tbPH{yàɲáárɪ̀} \tbPS{n.~} \tbGE{madness} 
\tbLX{yaŋaa} \tbHM{} \tbPH{yáŋàá} \tbPS{n.~} \tbGE{grandchild} \tbOP{yaŋene/yanne, yaamine} \tbTP{yaŋenɛɛ/yannɛɛ} 
\tbLX{yeebaa} \tbHM{} \tbPH{yɪ́ɪ́báá} \tbPS{n.~} \tbGE{twin} \tbOP{yeere} \tbTP{yeerɛɛ} 
\tbLX{yeelaa} \tbHM{} \tbPH{yɪ́ɪ́láá} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{yeele} 
\tbLX{yeele} \tbHM{} \tbPH{yɪ̀ɪ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to make neat} \tbSN \tbGE{to use water to extract or separate wanted things from unwanted ones e.g. grain from stones} \tbPD{yeelɛɛ, yeelɛ, yeelɛ, yeelebɛ, yeelaa, - } 
\tbLX{yeele} \tbHM{} \tbPH{yɪ̀ɪ̀lɪ̀} \tbPS{adv.~} \tbGE{extremely clean or beautiful} 
\tbLX{yeere} \tbHM{} \tbPH{yɪ́ɪ́rɪ́} \tbPS{n.~} \tbGE{twins} 
\tbLX{yeere} \tbHM{} \tbPH{yɪ̀ɪ̀rɪ̀} \tbPS{v.~} \tbGE{to experience emotional distress, to suffer} \tbXV{O pɔge naŋ kpi zaa, o deɛ tori la yeeroo yeerɛ.} \tbXE{Since his wife's death he has been suffering emotional distress.} \tbPD{yeerɛɛ, yeerɛ, yeerɛ, yeerebɛ, yeeraa, - } 
\tbLX{yeɛle} \tbHM{} \tbPH{yɪɛlɪ} \tbPS{v.~} \tbGE{to winnow, to throw grain into the air to separate it from the husk} \tbXV{A dɔbɔ ŋmɛ la a kyi ka a pɔgeba te yeɛle.} \tbXE{The men threshed the millet and the women winnowed it.} \tbPD{-, yeɛlɛ, yeɛlɛ, yeɛlebɛ, -, - } 
\tbLX{yeɛle} \tbHM{} \tbPH{yɪɛlɪ} \tbPS{v.~} \tbGE{to come down in tiny bits e.g. a drizzle} \tbXV{A saa ba mi soŋ, yeɛloo ka o yeɛle.} \tbXE{It did not really rain; it just drizzled.} \tbPD{-, yeɛlɛ, yeɛlɛ, yeɛlebɛ, -, - } 
\tbLX{yeɛle} \tbHM{} \tbPH{yɪ́ɛ́lɪ̀} \tbPS{v.~} \tbGE{to follow in significant numbers} \tbXV{Ba nyɔge la nannyige ka noba deɛ yeɛle tu.} \tbXE{A thief was arrested and many people ran after the team.} \tbPD{-, yeɛlɛ, yeɛlɛ, yeɛlebɛ, -, - } 
\tbLX{yeɛre} \tbHM{} \tbPH{yɪ̀ɛ̀rɪ̀} \tbPS{v.~} \tbGE{to want attention, to act childishly} \tbPD{yeɛrɛɛ, yeɛrɛ, yeɛrɛ, yeɛrebɛ, yeɛraa, - } 
\tbLX{yeɛre} \tbHM{} \tbPH{yìɛ́rɪ́} \tbPS{n.~} \tbGE{underside of the jaw} \tbOP{yɛrɛɛ} 
\tbLX{yelbaaloŋ} \tbHM{} \tbPH{yèlbààlʊ́ŋ} \tbPS{n.~} \tbGE{an issue under control} 
\tbLX{yelbaboɔraa} \tbHM{} \tbPH{yèlbʊ́ɔ́ráà} \tbPS{n.~} \tbSN \tbGE{something that is sought, desire, wish} \tbSN \tbGE{theme (literature)} \tbOP{yelbaboɔre} 
\tbLX{yelbalaa} \tbHM{} \tbPH{yèbáláá} \tbPS{n.~} \tbGE{a particular issue} 
\tbLX{yelbalee} \tbHM{} \tbPH{yèlbàlɪ̀ɪ̀} \tbPS{n.~} \tbGE{a particular issue} \tbXV{A yelbalee ŋa eŋɛ la ka o ne o pɔge dare zɔɔ taa.} \tbXE{It was because of this particular issue that he quarreled with his wife.} \tbVA{yelbalaa}
\tbLX{yelbanna} \tbHM{} \tbPH{yèbánná} \tbPS{n.~} \tbGE{one who knows} 
\tbLX{yelbaŋ} \tbHM{} \tbPH{yèlbàŋ} \tbPS{n.~} \tbGE{the act of knowing} 
\tbLX{yelbaŋena} \tbHM{} \tbPH{yèbáŋɪ́ná} \tbPS{n.~} \tbGE{one who knows} \tbOP{yelbaŋenema} \tbTP{yelbaŋenemɛɛ} \tbVA{yelbanna}
\tbLX{yelbeɛ} \tbHM{} \tbPH{yèlbɪ́ɛ́} \tbPS{n.~} \tbGE{mysterious, miraculous thing} \tbOP{yelbeere} \tbTP{yelbeerɛɛ} 
\tbLX{yelbieri} \tbHM{} \tbPH{yèlbíérí} \tbPS{n.~} \tbSN \tbGE{evil thing} \tbSN \tbGE{sin} \tbOP{yelbebe} 
\tbLX{yelbiri} \tbHM{} \tbPH{yèlbírì} \tbPS{n.~} \tbSN \tbGE{word} \tbSN \tbGE{something that is alluded to} \tbOP{yelbie} 
\tbLX{yelbitɔnnaa} \tbHM{} \tbPH{yélbítɔ̀nnáà} \tbPS{n.~} \tbGE{conjunction} \tbOP{yelbitɔnne} \tbTP{yelbitɔnnɛɛ} 
\tbLX{yelboɔraa} \tbHM{} \tbPH{yèlbʊ́ɔ́ráá} \tbPS{n.~} \tbGE{something that is wanted} \tbOP{yelboɔre} 
\tbLX{yelboɔraa} \tbHM{} \tbPH{yèlbʋ́ráá} \tbPS{n.~} \tbSN \tbGE{that which one likes} \tbSN \tbGE{objective} \tbSN \tbGE{something negative that affects one but caused by one’s self} \tbVA{yeldaaraa}
\tbLX{yelbuli} \tbHM{} \tbPH{yèlbúlí} \tbPS{n.~} \tbGE{main point} \tbOP{yelbulo} 
\tbLX{yeldaa} \tbHM{} \tbPH{yèldáà} \tbPS{n.~} \tbGE{simple sentence} \tbOP{yedaare} 
\tbLX{yeldaaraa} \tbHM{} \tbPH{yèldàárà} \tbPS{n.~} \tbSN \tbGE{that which one likes} \tbSN \tbGE{objective} \tbSN \tbGE{something negative that affects one but caused by one’s self} \tbOP{yelboɔre} \tbTP{yeldaare} \tbVA{yelboɔraa}
\tbLX{yeldeebo} \tbHM{} \tbPH{yéldɪ́ɪ́bʊ́} \tbPS{n.~} \tbGE{acceptance} 
\tbLX{yeldegere} \tbHM{} \tbPH{yèldɪ̀grɪ́} \tbPS{n.~} \tbGE{idiom} \tbOP{yeldegɛ} \tbTP{yeldɛgɛɛ} 
\tbLX{yeldɛgere} \tbHM{} \tbPH{yèldɛ́gɪ́rɪ̀} \tbPS{n.~} \tbGE{unpleasant or dirty talk} \tbOP{yeldɛgɛ} \tbTP{yeldɛgerɛɛ} 
\tbLX{yel-eebo} \tbHM{} \tbPH{yèl-ɪ́ɪ́bʊ́} \tbPS{n.~} \tbGE{behaviour} 
\tbLX{yel-eraa} \tbHM{} \tbPH{yèl-ɪ́ráà} \tbPS{n.~} \tbGE{something to be done} \tbOP{yel-erre} 
\tbLX{yelɛnyelɛŋ} \tbHM{} \tbPH{yɪ́lɛɲɪ́lɛ́ŋ} \tbPS{adj.~} \tbGE{clear and clean (water)} 
\tbLX{yelfaa} \tbHM{} \tbPH{yèlfáà} \tbPS{n.~} \tbGE{evil} \tbOP{yelfaare} \tbTP{yelfaarɛɛ} 
\tbLX{yelfa-e} \tbHM{} \tbPH{yèlfà-ì} \tbPS{n.~} \tbGE{wrongdoing} \tbOP{yelfa-erre} 
\tbLX{yelfa-erɛ} \tbHM{} \tbPH{yèlfà-ɪ́rɛ́} \tbPS{n.~} \tbGE{evildoer} \tbOP{yelfa-erebɛ} \tbTP{yelfa-erebɛɛ} 
\tbLX{yelferaa} \tbHM{} \tbPH{yèlfɪ̀ráà} \tbPS{n.~} \tbGE{problem, something that matters} \tbOP{yelferre} \tbTP{yelferrɛɛ} 
\tbLX{yelfogilo} \tbHM{} \tbPH{yèlfógló} \tbPS{n.~} \tbGE{useless talk} \tbSG{yelfogiluu} 
\tbLX{yelfoli} \tbHM{} \tbPH{yèlfòlí} \tbPS{n.~} \tbGE{obscene talk} \tbOP{yelfolo} 
\tbLX{yelfuoni} \tbHM{} \tbPH{yèlfùònì} \tbPS{n.~} \tbGE{imagery (literature)} \tbOP{yelfuomo} \tbTP{yelfuomee} 
\tbLX{yelgalema} \tbHM{} \tbPH{yèlgálɪ́má} \tbPS{n.~} \tbGE{strange thing} \tbTP{yelgalemɛɛ} \tbVA{yelgalee}
\tbLX{yelgbɔge} \tbHM{} \tbPH{yèlgbɔ́gɪ́} \tbPS{n.~} \tbSN \tbGE{incomplete talk} \tbSN \tbGE{gr. clause} 
\tbLX{yeli} \tbHM{} \tbPH{yélí} \tbPS{n.~} \tbSN \tbGE{something of concern} \tbXV{Ta bɔ yeli ko o, vɛŋe o ka o gɛrɛ.} \tbXE{Don't put him into any trouble; allow him to go.} \tbSN \tbGE{matter} \tbXV{Yeli ne o yelmeŋɛ N zãa ba gaa be.} \tbXE{As a matter of fact I did not go there.} \tbOP{yɛlɛ} 
\tbLX{yeli} \tbHM{} \tbPH{yélí} \tbPS{n.~} \tbGE{sentence} \tbXV{Fo na baŋ de la eronwuluu yelbiri maale ne yɛlɛ?} \tbXE{Can you construct a sentence with an adverb?.} \tbOP{yɛlɛ} 
\tbLX{yeli} \tbHM{} \tbPH{yèlì} \tbPS{v.~} \tbSN \tbGE{to speak, to talk, to inform} \tbXV{N teɛre ka o yeli la a yɛlɛ ko o ma.} \tbXE{I guess she has has informed her mother about it.} \tbSN \tbGE{to rebuke} \tbXV{O daare yeli la o pɔgeyaa lɛɛ ka o te kono.} \tbXE{He rebuked his daughter until she wept.} \tbPD{yelee, yele, yele, yelibe, yelaa, yelibe } \tbSE{waale yeli} \tbGE{to whisper} 
\tbLX{yelkaama} \tbHM{} \tbPH{yèlkáámá} \tbPS{n.~} \tbSN \tbGE{beautiful talk} \tbSN \tbGE{literature} \tbTP{yelkaamɛɛ} 
\tbLX{yelka᷉asɛgere} \tbHM{} \tbPH{yèlká᷉á᷉sɛ́grɪ́} \tbPS{n.~} \tbGE{written literature} 
\tbLX{yelka᷉a᷉yelli} \tbHM{} \tbPH{yèlká᷉á᷉yéllí} \tbPS{n.~} \tbGE{oral literature} 
\tbLX{yelkabeɛ} \tbHM{} \tbPH{yélkábɪ́ɛ́} \tbPS{n.~} \tbGE{strange thing} \tbOP{yelkabeere} \tbTP{yelkabeerɛɛ} 
\tbLX{yelkpeɛŋaa} \tbHM{} \tbPH{yèlkpɪ́ɛ́ŋàá} \tbPS{n.~} \tbGE{difficult thing} \tbOP{ yelkpeɛne} 
\tbLX{yel-lige} \tbHM{} \tbPH{yélígè} \tbPS{n.~} \tbGE{a mystery} \tbXV{Lɛ a pɔge naŋ e saŋe a baaloŋ e la ye-lige.} \tbXE{The way the woman was able to cure the patient is a mystery.} 
\tbLX{yel-lige} \tbHM{} \tbPH{yèl-lígè} \tbPS{adj.~} \tbGE{hidden} \tbXV{A dɔɔ na maŋ tona la yel-lige toma.} \tbXE{That man performs mysterious deeds.} \tbOP{yel-ligri} 
\tbLX{yelmaalaa} \tbHM{} \tbPH{yèlmààláà} \tbPS{n.~} \tbGE{rite, ritual} \tbOP{yelimaale} 
\tbLX{yelmaalaa} \tbHM{} \tbPH{yèlmààláà} \tbPS{n.~} \tbGE{something that needs to rectified} \tbOP{yelimaale} 
\tbLX{yelmaaroŋ} \tbHM{} \tbPH{yèlmáárʊ́ŋ} \tbPS{n.~} \tbGE{peaceful talk} 
\tbLX{yelmannaa} \tbHM{} \tbPH{yèlmànnáà} \tbPS{n.~} \tbSN \tbGE{comparison} \tbSN \tbGE{narrative, narration} \tbOP{yelmanne} \tbTP{yelmannɛɛ} 
\tbLX{yelmanne} \tbHM{} \tbPH{yélmánnɪ́} \tbPS{n.~} \tbGE{the act of narrating} 
\tbLX{yelmeŋɛ} \tbHM{} \tbPH{yèlmɪ́ŋɛ́} \tbPS{n.~} \tbGE{truth} \tbOP{yelimenne} \tbTP{yelimennɛɛ} 
\tbLX{yelnimiri} \tbHM{} \tbPH{yènímírì} \tbPS{n.~} \tbGE{main idea} 
\tbLX{yelnoɔ} \tbHM{} \tbPH{yèlnʊ́ɔ́} \tbPS{n.~} \tbSN \tbGE{interesting thing} \tbSN \tbGE{good news} \tbOP{yelinoore} \tbTP{yelinoorɛɛ} 
\tbLX{yelnɔnaa} \tbHM{} \tbPH{yèlnɔ́náà} \tbPS{n.~} \tbGE{something that is liked} \tbOP{yelinɔnne} 
\tbLX{yelpaala} \tbHM{} \tbPH{yèlpáálà} \tbPS{n.~} \tbGE{new thing} \tbOP{yelipaale} 
\tbLX{yelpɛgere} \tbHM{} \tbPH{yèlpɛ́grɪ́} \tbPS{n.~} \tbGE{sayings that have no value} \tbOP{yelpɛgɛ} 
\tbLX{yelsaana} \tbHM{} \tbPH{yèlsáánà} \tbPS{n.~} \tbGE{strange or new thing} \tbOP{yelisaama} \tbTP{yelisaamɛɛ} 
\tbLX{yelsaana} \tbHM{} \tbPH{yèlsááná} \tbPS{n.~} \tbSN \tbGE{destroyer} \tbSN \tbGE{one who meddles in issues} \tbOP{ yelisaanema} 
\tbLX{yelsaanaa} \tbHM{} \tbPH{yèlsáánáá} \tbPS{n.~} \tbGE{something that has gone wrong} \tbOP{yelisnaane} \tbTP{yelisaanɛɛ} 
\tbLX{yelsoŋ} \tbHM{} \tbPH{yèlsʊ̀ŋ} \tbPS{n.~} \tbGE{good deed} \tbOP{yelisonne} 
\tbLX{yelsɔgelaa} \tbHM{} \tbPH{yèlsɔ́gláà} \tbPS{n.~} \tbGE{mysterious thing, secret} \tbOP{yelisɔgelɔ} 
\tbLX{yeltegeroŋ} \tbHM{} \tbPH{yèltɪ́grʊ́ŋ} \tbPS{n.~} \tbGE{difficult position} \tbOP{yelitegerɛ} 
\tbLX{yeltoloŋ} \tbHM{} \tbPH{yèltʊ́lʊ́ŋ} \tbPS{n.~} \tbGE{serious matter, something that needs urgent attention} \tbOP{yelitolɔ} 
\tbLX{yeltɔgelaa} \tbHM{} \tbPH{yèltɔ̀gláà} \tbPS{n.~} \tbGE{that which can be imitated} \tbOP{yelitɔgele} 
\tbLX{yeltɔre} \tbHM{} \tbPH{yèltɔ̀rɪ́} \tbPS{n.~} \tbGE{something of worth or value} 
\tbLX{yeltulimo} \tbHM{} \tbPH{yèltúli̊mó} \tbPS{n.~} \tbGE{opposite} \tbVA{yelitulo}
\tbLX{yeltuluu} \tbHM{} \tbPH{yèltúlúú} \tbPS{n.~} \tbGE{mistaken talk, error} 
\tbLX{yeltuo} \tbHM{} \tbPH{yèltúó} \tbPS{n.~} \tbGE{difficult thing} \tbOP{yelituuri} \tbTP{yelituuree} 
\tbLX{yeltuuraa} \tbHM{} \tbPH{yèltùùráà} \tbPS{n.~} \tbGE{rite, tradition} \tbOP{yelituuri} 
\tbLX{yeluu} \tbHM{} \tbPH{yélúú} \tbPS{n.~} \tbGE{saying} 
\tbLX{yelwiiraa} \tbHM{} \tbPH{yèlwííráá} \tbPS{n.~} \tbGE{poem} \tbOP{yeliwiiri} 
\tbLX{yelwiiri} \tbHM{} \tbPH{yèlwíírí} \tbPS{n.~} \tbGE{poetry} 
\tbLX{yelwonaa} \tbHM{} \tbPH{yèlwònáà} \tbPS{n.~} \tbSN \tbGE{that which is a challenge; problem} \tbSN \tbGE{something that is heard} \tbOP{yeliwonni} 
\tbLX{yelwono} \tbHM{} \tbPH{yèlwónó} \tbPS{n.~} \tbSN \tbGE{listener, hearer} \tbSN \tbGE{obedient person} \tbOP{yeliwonimo} 
\tbLX{yelwontaa} \tbHM{} \tbPH{yèlwòntàá} \tbPS{n.~} \tbGE{compromise, understanding} 
\tbLX{yelyelbaraa} \tbHM{} \tbPH{yèlyèlbàráà} \tbPS{n.~} \tbGE{statement (grammar)} \tbOP{yelyelbarre} 
\tbLX{yelyeluu} \tbHM{} \tbPH{yèlyélúú} \tbPS{n.~} \tbGE{issues that cause anger} 
\tbLX{yelzagera} \tbHM{} \tbPH{yèlzågrá} \tbPS{n.~} \tbGE{one who refuses, one who disobeys} \tbOP{yelzagereba} 
\tbLX{yeme} \tbHM{} \tbPH{yèmé} \tbPS{n.~} \tbGE{slave} \tbOP{yenni} \tbTP{yennee} \tbVA{yieme}
\tbLX{yempɔge} \tbHM{} \tbPH{yèmpɔ́gɪ́} \tbPS{n.~} \tbSN \tbGE{female slave} \tbSN \tbGE{name traditionally given to a female child born outside the house} \tbOP{yelmpɔgeba} \tbTP{yempɔgrɛɛ} 
\tbLX{yendɔɔ} \tbHM{} \tbPH{yèndɔ́ɔ́} \tbPS{n.~} \tbSN \tbGE{male slave} \tbSN \tbGE{name traditionally given to a male child born outside the house} 
\tbLX{yeni} \tbHM{} \tbPH{yénì} \tbPS{adj.~} \tbGE{one} \tbOP{yeme} \tbTP{yennee} 
\tbLX{yensogɔ} \tbHM{} \tbPH{yénsʊ́gɔ́} \tbPS{n.~} \tbGE{night} \tbOP{yensogere} \tbTP{yensogerɛɛ} \tbVA{yensoga, tensogɔ, tensoga}
\tbLX{yeŋ} \tbHM{} \tbPH{yêŋ} \tbPS{adv.~} \tbGE{where} 
\tbLX{yeŋe} \tbHM{} \tbPH{yèŋé} \tbPS{adv.~} \tbGE{outside} \tbVA{yeŋi}
\tbLX{yeŋi} \tbHM{} \tbPH{yèŋí} \tbPS{adv.~} \tbGE{outside} \tbVA{yeŋe}
\tbLX{yɛ} \tbHM{} \tbPH{yɛ̀} \tbPS{pronoun.~} \tbGE{you (second person plural)} 
\tbLX{yɛ} \tbHM{} \tbPH{yɛ̀} \tbPS{v.~} \tbGE{to become exhausted, to get fed up} \tbPD{yɛɛ, yeɛrɛ, yeɛrɛ, yeɛrebɛ, yeɛraa, yɛyeɛrɛ } 
\tbLX{yɛ} \tbHM{} \tbPH{yɛ̀} \tbPS{v.~} \tbGE{become  tired} \tbXV{Ka a tontonmaŋ wa yɛ ba iri.} \tbXE{When the workers are tired they can stop.} \tbPD{-, -, -, -, -, - } \tbVA{ yɛɛ}
\tbLX{yɛ} \tbHM{} \tbPH{yɛ̀} \tbPS{adj.~} \tbGE{tired} \tbXV{Maa eŋ yɛɛ la ne a zeɛre ŋa.} \tbXE{ I am tired of this sauce.} \tbVA{ yɛɛ}
\tbLX{yɛɛ} \tbHM{} \tbPH{yɛ̀ɛ́} \tbPS{adj.~} \tbGE{spent out; reminants; not of substance (of crops that sprout from a previously cultivated field)} \tbOP{yɛre} \tbTP{yɛrɛɛ} 
\tbLX{yɛle} \tbHM{} \tbPH{yɛ́lɪ̀} \tbPS{n.~} \tbGE{cockroach} \tbOP{yɛlɛ} \tbTP{yɛlɛɛ} 
\tbLX{yɛle} \tbHM{} \tbPH{yɛ̀lɪ̀} \tbPS{v.~} \tbSN \tbGE{to multiply e.g. animals reproducing} \tbSN \tbGE{to enlarge} \tbPD{yɛlɛɛ, yɛllɛ, yɛllɛ, yɛlebɛ, yɛllaa, - } 
\tbLX{yɛlɛ} \tbHM{} \tbPH{yɛ́lɛ́} \tbPS{n.~} \tbSN \tbGE{an issue} \tbSN \tbGE{discussion, conversation} \tbSN \tbGE{serious matter} 
\tbLX{yɛloŋ} \tbHM{} \tbPH{yɛ́lʊ́ŋ} \tbPS{adj.~} \tbGE{wide} \tbOP{yɛlle} 
\tbLX{yɛmbaŋ} \tbHM{} \tbPH{yɛ́mbàŋ} \tbPS{n.~} \tbGE{cleverness, wisdom} 
\tbLX{yɛmbaŋena} \tbHM{} \tbPH{yɛ́mbáŋɪ́ná} \tbPS{n.~} \tbGE{one who has good sense, wise person} \tbOP{yɛmbaŋnema} 
\tbLX{yɛmbile} \tbHM{} \tbPH{yɛ́mbílé} \tbPS{n.~} \tbGE{small intelligence} \tbOP{yɛmbilii} \tbTP{yɛmbillee} 
\tbLX{yɛnkyeerɛ} \tbHM{} \tbPH{yɛ́nʧɪ́ɪ́rɛ́} \tbPS{n.~} \tbGE{little mind} 
\tbLX{yɛnkyeeroo} \tbHM{} \tbPH{yɛ́nʧɪ́ɪ́rʊ̊ʊ́} \tbPS{n.~} \tbGE{prudence} 
\tbLX{yɛnlige} \tbHM{} \tbPH{yɛ́nlígè} \tbPS{n.~} \tbGE{unclear mind} 
\tbLX{yɛnnyaa} \tbHM{} \tbPH{yɛ́nɲáà} \tbPS{n.~} \tbGE{the act of developing intelligence} 
\tbLX{yɛnwuli} \tbHM{} \tbPH{yɛ́nwúlì} \tbPS{n.~} \tbGE{the act of teaching sense} 
\tbLX{yɛnwuluu} \tbHM{} \tbPH{yɛ́nwúlúú} \tbPS{n.~} \tbGE{the act of teaching sense or conscience} 
\tbLX{yɛnyɛŋ} \tbHM{} \tbPH{yɛ́ɲɛ́ŋ} \tbPS{adj.~} \tbGE{near} 
\tbLX{yɛŋ} \tbHM{} \tbPH{yɛ̂ŋ} \tbPS{adv.~} \tbGE{kindly e.g. in an appeal} \tbXV{Yɛŋ de a bie gaa ne sakuuri.} \tbXE{Try and take the child to school.} 
\tbLX{yɛŋe} \tbHM{} \tbPH{yɛ̂ŋ} \tbPS{adj.~} \tbGE{intelligent} \tbXV{A pɔgelee na taa la yɛŋe.} \tbXE{That girl is intelligent.} 
\tbLX{yɛŋe} \tbHM{} \tbPH{yɛ̂ŋ} \tbPS{n.~} \tbGE{intelligence, knowledge} \tbXV{A bie yɛŋ ba seŋ bone.} \tbXE{The child has very little knowledge.} \tbOP{yɛnne} \tbTP{yɛnnɛɛ} 
\tbLX{yɛre} \tbHM{} \tbPH{yɛ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to wear, to dress} \tbXV{Ba zaa yɛre la tigiri bonsuuri.} \tbXE{They all dressed their best for the feast.} \tbSN \tbGE{to hang around the neck and shoulders} \tbXV{O yɛre bɔnzɔl velaa kaŋa.} \tbXE{She wore a beautiful chain.} \tbPD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\tbLX{yɛre} \tbHM{} \tbPH{yɛ̀rɪ̀} \tbPS{v.~} \tbGE{to trade, to do business} \tbXV{Ka fooŋ na yɛre ferebo la ka fo taa libima.} \tbXE{If you want to engage in business, you must have some capital.} \tbPD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\tbLX{yɛre} \tbHM{} \tbPH{yɛ́rɪ̀} \tbPS{v.~} \tbGE{to spread out e.g. to dry} \tbPD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\tbLX{yɛre} \tbHM{} \tbPH{yɛ́rɪ̀} \tbPS{v.~} \tbGE{to hunt in a group} \tbPD{yɛrɛɛ, yɛrɛ, yɛrɛ, yɛreba, yɛraa, yɛreba } 
\tbLX{yɛre} \tbHM{} \tbPH{yɛ́rɪ́} \tbPS{adj.~} \tbGE{wild; used with crops e.g groundnuts} \tbXV{A bibiiri gaa pɛ la seŋkãa yɛɛ.} \tbXE{The children went harvesting wild groundnuts.} \tbOP{yɛɛ} \tbTP{yɛrɛɛ} 
\tbLX{yɛreyɛrɛ} \tbHM{} \tbPH{yɛ̀ryɛ́rɛ́} \tbPS{n.~} \tbGE{trader, businessperson, merchant} \tbOP{yɛreyɛrebɛ} \tbTP{yɛreyɛrebɛɛ} 
\tbLX{yɛroŋ} \tbHM{} \tbPH{yɛ́rʊ́ŋ} \tbPS{n.~} \tbGE{trade, business} 
\tbLX{yi} \tbHM{} \tbPH{yí} \tbPS{v.~} \tbSN \tbGE{to go out, to exit} \tbXV{Te naŋ yi ka te gaa a kpaaroŋ la ka saa deɛ wa miire.} \tbXE{As we went out to go to the farm, it started raining.} \tbSN \tbGE{to divorce (by a woman)} \tbXV{A dɔɔ nɔŋ la o pɔge ŋmeɛbo lɛnso ka o yi.} \tbXE{The man likes beating his wife and that is why she has divorced him.} \tbPD{yie, yire, yire, yiribe, yiraa, yiribe } 
\tbLX{yi} \tbHM{} \tbPH{yí} \tbPS{v.~} \tbGE{to be clean} \tbXV{Fo ba pɛge a laa ne fanfane a zuiŋ o ba yi.} \tbXE{You did not wash the bowl with soap and so it is not clean.} \tbPD{yie, yire, yire, yiribe, yiraa, yiribe } 
\tbLX{yi} \tbHM{} \tbPH{yí} \tbPS{v.~} \tbGE{to resemble} \tbXV{A bie ŋa yi la o ma yaga.} \tbXE{This child resembles the mother very much.} \tbPD{yie, yire, yire, yiribe, yiraa, yiribe } 
\tbLX{yi} \tbHM{} \tbPH{yí} \tbPS{v.~} \tbGE{of land;  become infertile, to be barren e.g land} \tbXV{Ba kɔ la a koɔlɔ ŋa ka o te yi, ka fooŋ kɔ o a ba eŋ kuoluŋ fo koŋ nyɛ bonezaa.} \tbXE{Several years of farming on this land has made it infertile and if you farm it without applying fertilizer to it you won’t get any good yield.} \tbPD{yie, yire, yire, yiribe, yiraa, yiribe } 
\tbLX{yi} \tbHM{} \tbPH{yí} \tbPS{v.~} \tbGE{to lose taste e.g. beer} \tbXV{Ba naŋ zaameŋ ba pɔge a da᷉a᷉ noɔre na la so ka o yi.} \tbXE{Because the beer was not covered, it has lost its taste.} \tbPD{yie, yire, yire, yiribe, yiraa, yiribe } 
\tbLX{yidaampɔge} \tbHM{} \tbPH{yídààmpɔ́gɪ́} \tbPS{n.~} \tbGE{head woman in a family; mistress of a compound} \tbOP{yidaampɔgeba} \tbTP{yidaampɔgrɛɛ} 
\tbLX{yidaana} \tbHM{} \tbPH{yídááná} \tbPS{n.~} \tbGE{house owner} \tbOP{yidaama} \tbTP{yidaanɛɛ/yidaamɛɛ} 
\tbLX{yidaandɔɔ} \tbHM{} \tbPH{yídààndɔ́ɔ́} \tbPS{n.~} \tbGE{male head of a family; master of of compound} \tbOP{yidaandɔbɔ} \tbTP{yidaandɔɔrɛɛ} 
\tbLX{yideme} \tbHM{} \tbPH{yídèmé} \tbPS{n.~} \tbGE{family group; family members} \tbSG{yidemuu} \tbTP{yidemee} 
\tbLX{yi᷉e} \tbHM{} \tbPH{yì᷉è᷉} \tbPS{v.~} \tbGE{to faint} \tbPD{yi᷉ee, yiene, yiene, yienime, yienaa, - } 
\tbLX{yieli} \tbHM{} \tbPH{yíélì} \tbPS{v.~} \tbGE{to sing} \tbPD{yielee, yiele, yiele, yielibe, -, - } 
\tbLX{yieluŋ} \tbHM{} \tbPH{yíélúŋ} \tbPS{n.~} \tbGE{song; music} \tbOP{yiele} 
\tbLX{yieme} \tbHM{} \tbPH{yìèmé} \tbPS{n.~} \tbGE{slave} 
\tbLX{yieri} \tbHM{} \tbPH{yìèrì} \tbPS{v.~} \tbGE{to extend, to spread out} \tbXV{Ka dabeɛŋ wa kpɛ karaa, o maŋ yieri la o zu ne o kɔkɔrɛɛ.} \tbXE{When a cobra is frightened, it spreads out its head and neck.} \tbPD{yieree, yiere, yiere, yieribe, yieraa, - } 
\tbLX{yi᷉eu} \tbHM{} \tbPH{yí᷉é᷉ú᷉} \tbPS{n.~} \tbGE{spectacle, wonders} \tbOP{yieme} 
\tbLX{yigileŋ} \tbHM{} \tbPH{yígléŋ} \tbPS{adv.~} \tbGE{perched in a lonely manner} \tbXV{A bie dɔŋɛɛ la ka o zu deɛ waa yigileŋ lɛ.} \tbXE{The child has grown so lean that his head is literally perching on his neck.} 
\tbLX{yi kyeɛ} \tbHM{} \tbPH{yí ʧɪ́ɛ́} \tbPS{adj.~} \tbGE{uninteresting, boring} 
\tbLX{yi kyeɛ} \tbHM{} \tbPH{yí ʧɪ́ɛ́} \tbPS{v.~} \tbGE{to fade} \tbPD{-, -, -, -, -, - } 
\tbLX{yinni} \tbHM{} \tbPH{yìnnì} \tbPS{v.~} \tbSN \tbGE{to forget} \tbXV{N yeli la a kpaaroŋ yɛlɛ ko fo ta erɛ wa yinni.} \tbXE{I told you about the meeting, so don't forget.} \tbSN \tbGE{to glean (farm field)} \tbXV{Ka banaŋ wa pɛ seŋka᷉a᷉ baare bibiiri maŋ gaa te yinni la a gbannaa poɔ.} \tbXE{After groundnuts are harvested, children would normally go to the field to glean.} \tbPD{yinnee, yinne, yinne, yinime, yinnaa, yinime } \tbVA{inni}
\tbLX{yiri} \tbHM{} \tbPH{yírì} \tbPS{n.~} \tbSN \tbGE{house, home, compound} \tbSN \tbGE{family, family members} \tbOP{yie} \tbTP{yiree} 
\tbLX{yi sori} \tbHM{} \tbPH{yi23 sórí} \tbPS{v.~} \tbGE{to travel, to go on a trip} \tbPD{-, -, -, -, -, - } 
\tbLX{yivilime} \tbHM{} \tbPH{yívílímé} \tbPS{n.~} \tbGE{round compound} 
\tbLX{yobie} \tbHM{} \tbPH{yóbíé} \tbPS{n.~} \tbGE{ill luck, bad name} 
\tbLX{yofaa} \tbHM{} \tbPH{yófáá} \tbPS{n.~} \tbGE{a name that stigmatizes} \tbOP{yofaare} \tbTP{yofaarɛɛ.} 
\tbLX{yofuu} \tbHM{} \tbPH{yófúù} \tbPS{n.~} \tbGE{abstract noun} \tbOP{yofuuri} \tbTP{yofuuree} 
\tbLX{yogi} \tbHM{} \tbPH{yógí} \tbPS{v.~} \tbGE{to eat, especially plenty food that is tasty} \tbPD{yogee, yogro, yogro, yogro, yogribo, yograa } \tbVA{mogi}
\tbLX{yogi} \tbHM{} \tbPH{yògì} \tbPS{adj.~} \tbGE{enough, of considerable quantity} \tbXV{A zɛvaare ba yogi.} \tbXE{There is not enough of the vegetable.} 
\tbLX{yogili} \tbHM{} \tbPH{yòglì} \tbPS{v.~} \tbGE{to pamper, to treat kindly or gently} \tbPD{yoglee, yoglo, yoglo, yoglibo, yoglaa, - } 
\tbLX{yogoko} \tbHM{} \tbPH{yógókó} \tbPS{adv.~} \tbGE{very sweet} 
\tbLX{yo-iri} \tbHM{} \tbPH{yó-írì} \tbPS{n.~} \tbGE{glory, praise, honour,} 
\tbLX{yokyigiraa} \tbHM{} \tbPH{yòkyìgìráà} \tbPS{n.~} \tbGE{fishing pot} 
\tbLX{yololo} \tbHM{} \tbPH{yólóló} \tbPS{adv.~} \tbGE{very cold} 
\tbLX{yoluŋ} \tbHM{} \tbPH{yólúŋ} \tbPS{n.~} \tbGE{cone-shaped cage} \tbOP{yolli} \tbTP{yollee} 
\tbLX{yoluŋ} \tbHM{} \tbPH{yólúŋ} \tbPS{adj.~} \tbGE{emptiness e.g wearing a blouse or shirt without a singlet or a vest under it} \tbOP{yolli} \tbTP{yolee} 
\tbLX{yombo} \tbHM{} \tbPH{yómbó} \tbPS{n.~} \tbGE{black hair dye} \tbOP{yombori} \tbTP{yomboree} 
\tbLX{yontaa} \tbHM{} \tbPH{yòntáå} \tbPS{n.~} \tbSN \tbGE{wife in a polygamous marriage} \tbSN \tbGE{relationship between wives in a polygamous marriage} \tbSN \tbGE{enemy (figurative)} \tbOP{yontaare} \tbTP{yontaarɛɛ} 
\tbLX{yoŋ} \tbHM{} \tbPH{yôŋ} \tbPS{adj.~} \tbSN \tbGE{alone} \tbSN \tbGE{only} \tbOP{yonni} 
\tbLX{yoo} \tbHM{} \tbPH{yóó} \tbPS{interjection.~} \tbSN \tbGE{okay} \tbXV{Yoo, bieo te na gaa la.} \tbXE{Okay tomorrow we shall go.} \tbSN \tbGE{be informed} \tbXV{A toma ŋa fo yoŋiŋ so a.} \tbXE{Be informed that that work is for you alone.} 
\tbLX{yoɔ} \tbHM{} \tbPH{yʊ̀ɔ̀} \tbPS{v.~} \tbGE{to stop raining} \tbPD{-, -, -, -, -, - } \tbVA{gyere}
\tbLX{yoɔbo} \tbHM{} \tbPH{yʊ́ɔ́bʊ̀} \tbPS{v.~} \tbSN \tbGE{the act of roaming, wandering} \tbSN \tbGE{hunting expedition} \tbPD{-, -, -, -, -, - } 
\tbLX{yoɔbo} \tbHM{} \tbPH{yʊ́ɔ́bʊ́} \tbPS{n.~} \tbGE{the stoppage of rain} 
\tbLX{yoɔbo} \tbHM{} \tbPH{yʊ̀ɔ̀bʊ́} \tbPS{adj.~} \tbGE{short form of six} 
\tbLX{yoɔle} \tbHM{} \tbPH{yʋ̀ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to escape slowly e.g. from a punctured tyre} \tbPD{-, -, -, -, -, - } 
\tbLX{yoɔle} \tbHM{} \tbPH{yʋ́ɔ́lɪ́} \tbPS{v.~} \tbGE{to melt, to disintegrate slowly e.g salt or sugar in sauce or tea} \tbPD{-, -, -, -, -, - } 
\tbLX{yoɔne} \tbHM{} \tbPH{yʊ́ɔ́nɪ́} \tbPS{n.~} \tbGE{a type of small, spotted amphibean that looks like a frog and is found in wet sand} \tbOP{yoɔmɔ/yoɔma} \tbTP{yoɔnɛɛ} 
\tbLX{Yoɔraa} \tbHM{} \tbPH{yʋ́ɔ́ráá} \tbPS{n.~} \tbGE{a traditional name given to a child believed to be reincarnated} 
\tbLX{yoɔraa} \tbHM{} \tbPH{yʊ́ɔ́ráà} \tbPS{n.~} \tbGE{tourist, one who roams} \tbOP{yoɔrebɔ/yoɔreba} 
\tbLX{yoɔre} \tbHM{} \tbPH{yʊ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{penis} \tbOP{yɔɛ} \tbTP{yɔrɛɛ} 
\tbLX{yoɔre} \tbHM{} \tbPH{yʊ̀ɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to cause to be exhausted or emotionless} \tbSN \tbGE{to bother, irritate; to pester someone} \tbPD{yoɔrɛɛ, yoɔrɔ, yoɔrɔ, yoɔreba, yoɔraa, - } 
\tbLX{yopo᷉i} \tbHM{} \tbPH{yòpó᷉i᷈} \tbPS{adj.~} \tbGE{short form of ayoɔpõi (seven)} 
\tbLX{yopɔgeraa} \tbHM{} \tbPH{yòpɔ̀gràà} \tbPS{n.~} \tbGE{fishing pot} \tbOP{yopɔgere} \tbVA{yokyigiraa}
\tbLX{yopɔgeraa} \tbHM{} \tbPH{yòpɔ́gɪ́ráá} \tbPS{n.~} \tbGE{lid, cover of a pot} 
\tbLX{yoreyoree} \tbHM{} \tbPH{yʊ̀rɪ̀yʊ́rɪ́} \tbPS{v.~} \tbGE{to break into little bits} \tbPD{-, -, -, -, -, - } 
\tbLX{yori} \tbHM{} \tbPH{yòrì} \tbPS{v.~} \tbGE{to make a loop} \tbPD{yoree, yoro, yoro, yoribo, yoraa, - } 
\tbLX{yori} \tbHM{} \tbPH{yórì} \tbPS{v.~} \tbGE{to dismantle a loop} \tbPD{yoree, yoro, yoro, yoribo, yoraa, - } 
\tbLX{yorombiri} \tbHM{} \tbPH{yórómbírí} \tbPS{n.~} \tbGE{herring} \tbOP{yorombie} \tbTP{yorombireeyorombiiree} 
\tbLX{yorɔyorɔ} \tbHM{} \tbPH{yʊ́rɔ́yʊ́rɔ́} \tbPS{adv.~} \tbGE{smallish; usually in size} \tbVA{yorayora}
\tbLX{yosa᷈a} \tbHM{} \tbPH{yósà᷈à᷈} \tbPS{n.~} \tbGE{slander} \tbOP{yosa᷉a} 
\tbLX{yoyogibiri} \tbHM{} \tbPH{yóyògìbírì} \tbPS{n.~} \tbGE{a type of small flying insect that makes tambalaaraa (a type of honey)} \tbSG{yoyogibiruu} \tbOP{yoyogibie} \tbTP{yoyogibiree/yoyogibiiree} 
\tbLX{yɔ} \tbHM{} \tbPH{yɔ́} \tbPS{v.~} \tbGE{to wander, to roam, to travel} \tbPD{yɔɛ, yoɔrɔ, yoɔrɔ, yoɔreba, -, - } 
\tbLX{yɔge} \tbHM{} \tbPH{yɔ̀gɪ̀} \tbPS{v.~} \tbGE{to loosen e.g. hard substance (soil)} \tbPD{yɔgɛɛ, yɔgrɔ, yɔgrɔ, yɔgrebɔ, yɔgraa, - } 
\tbLX{yɔgele} \tbHM{} \tbPH{yɔ́glɪ̀} \tbPS{v.~} \tbGE{to be near, to approach (time)} \tbXV{Se᷉ɛ᷉ yɔgelɔ la ka suŋkpulo kyoɔrɔ.} \tbXE{The farming season is coming round; hence, dawdaw trees are flowering.} \tbPD{yɔgelɛɛ, yɔgelɔ, yɔglɔ, yɔgeleba, yɔgelaa, - } 
\tbLX{yɔgere} \tbHM{} \tbPH{yɔ̀grɪ́} \tbPS{n.~} \tbGE{farm field around the compound} \tbOP{yɔgɔ} \tbTP{yɔgɛɛ} 
\tbLX{yɔgere} \tbHM{} \tbPH{yɔ́grɪ̀} \tbPS{v.~} \tbGE{to loosen e.g of a string tied to something} \tbPD{yɔgerɛɛ, yɔgerɔ, yɔgerɔ, yɔgereba, yɔgeraa, - } 
\tbLX{yɔgere} \tbHM{} \tbPH{yɔ́grɪ́} \tbPS{v.~} \tbGE{to be loose, to be slack} \tbXV{Ba leŋ la a tuobu kyɛ ka a leɛ yɔgere.} \tbXE{The load was tied but it is loose again.} \tbPD{-, -, -, -, -, - } 
\tbLX{yɔgevaa} \tbHM{} \tbPH{yɔ́gɪ̀váárɪ̀} \tbPS{n.~} \tbGE{pumpkin leaf} \tbOP{yɔgevaare} \tbTP{yɔgevaarɛɛ} 
\tbLX{yɔgevaare} \tbHM{} \tbPH{yɔ̀gɪ́váárɪ̀} \tbPS{n.~} \tbGE{pumpkin leaves used as vegetable} \tbTP{yɔgevaarɛɛ} 
\tbLX{yɔgevaare} \tbHM{} \tbPH{yɔ̀gváárɪ̀} \tbPS{n.~} \tbGE{leaves of pumpkin} \tbSG{yɔgevaabo} \tbOP{yɔgevaarɛɛ} 
\tbLX{yɔgevaare-waabo} \tbHM{} \tbPH{yɔ̀gɪ̀váárɪ̀-wáábʊ̀} \tbPS{n.~} \tbGE{green mamba} \tbOP{yɔgevaare-weere} \tbTP{yɔgevaare-weerɛɛ} 
\tbLX{yɔgoo} \tbHM{} \tbPH{yɔ̀gʊ́ʊ̊} \tbPS{n.~} \tbSN \tbGE{pumpkin} \tbSN \tbGE{the plant which bears the fruit} \tbOP{yɔgɔ} 
\tbLX{yɔgoo} \tbHM{} \tbPH{yɔ̀gʊ́ʊ̊} \tbPS{n.~} \tbSN \tbGE{a type of large, white ant} \tbSN \tbGE{the home of these ants} \tbOP{yɔgɔ} 
\tbLX{yɔkɔɔloŋ} \tbHM{} \tbPH{yɔ̌kɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{pubic hair of a male} 
\tbLX{yɔle} \tbHM{} \tbPH{yɔ́lɪ́} \tbPS{v.~} \tbGE{to slip off as a result of being loose e.g. shorts} \tbPD{yɔlɛɛ, yɔlɔ, yɔlɔ, yɔlɔ, yɔlaa, - } 
\tbLX{yɔlpelaa} \tbHM{} \tbPH{yɔ̀lpɪ̀láà} \tbPS{n.~} \tbGE{a type of tree with edible fruits} \tbOP{yɔlpeɛle} 
\tbLX{yɔlzeɛ} \tbHM{} \tbPH{yɔ̀lzɪ́ɛ́} \tbPS{n.~} \tbGE{a type of tree with grey bark} \tbOP{yɔlzeere} 
\tbLX{yɔmɔ} \tbHM{} \tbPH{yɔ́mɔ́} \tbPS{adj.~} \tbGE{free and plentiful} \tbOP{yɔnne} 
\tbLX{yɔɔ} \tbHM{} \tbPH{yɔ́ɔ̀} \tbPS{n.~} \tbGE{sibling (used between individuals of the same sex)} \tbOP{yɔɔmine/yɔɔre} \tbTP{yɔɔrɛɛ} 
\tbLX{yɔɔ} \tbHM{} \tbPH{yɔ́ɔ̀} \tbPS{n.~} \tbGE{strophantus (a type of tree with highly poisonous sap)} \tbOP{yɔɔre} \tbTP{yɔɔrɛɛ} 
\tbLX{yɔɔ} \tbHM{} \tbPH{yɔ́ɔ́} \tbPS{v.~} \tbGE{to pay, to redeem} \tbPD{yɔɔɛ, yɔɔrɔ, yɔɔrɔ, yɔɔreba, yɔɔraa, yɔɔre } 
\tbLX{yɔɔ} \tbHM{} \tbPH{yɔ̀ɔ̀} \tbPS{adv.~} \tbGE{freely e.g. of liquid being poured} 
\tbLX{yɔɔdɔɔ} \tbHM{} \tbPH{yɔ́ɔ́dɔ́ɔ́} \tbPS{n.~} \tbGE{brother (used by females to address male siblings)} \tbOP{yɔɔdɔɔbɔ} \tbTP{yɔɔdɔɔrɛɛ} 
\tbLX{yɔɔlee} \tbHM{} \tbPH{yɔ́ɔ́léé} \tbPS{n.~} \tbGE{younger sibling (used between individuals of the same sex)} \tbOP{yɔɔbilii} 
\tbLX{yɔɔpɔge} \tbHM{} \tbPH{yɔ́ɔ́pɔ́gɪ́} \tbPS{n.~} \tbGE{sister (used by males to address female siblings)} \tbOP{yɔɔpɔgebɔ/yɔɔpɔgeba.} \tbVA{yɔɔpuulee}
\tbLX{yɔɔpuulee} \tbHM{} \tbPH{yɔ́ɔ̀púúléé} \tbPS{n.~} \tbGE{female sibling (usually less older) see kpɪ́ɛ́mɛ́} \tbOP{yɔɔpuuli} \tbVA{yɔɔpɔge}
\tbLX{yɔɔre} \tbHM{} \tbPH{yɔ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{payment} \tbTP{yɔɔrɛɛ} 
\tbLX{yɔre} \tbHM{} \tbPH{yɔ́rɪ̀} \tbPS{v.~} \tbGE{to fall apart, to disintegrate} \tbPD{yɔrɛɛ, yɔrɔ, yɔrɔ, yɔreba, yɔraa, - } 
\tbLX{yɔre} \tbHM{} \tbPH{yɔ̀rɪ̀} \tbPS{v.~} \tbSN \tbGE{to become intense e.g. rainfall} \tbSN \tbGE{to become widespread (disease)} \tbSN \tbGE{that which is plenty} \tbPD{yɔrɛɛ, yɔrɔ, yɔrɔ, yɔreba, yɔraa, - } 
\tbLX{Yɔrebaa} \tbHM{} \tbPH{yɔ̀rɪ́báá} \tbPS{n.~} \tbGE{Yoruba (an ethnic group found in Nigeria)} \tbSG{yɔrebaaroo} \tbOP{yɔrebaare} \tbTP{yɔrebaarɛɛ} 
\tbLX{yɔrebɔnɔ} \tbHM{} \tbPH{yɔ̀rɪ̀bɔ̀nɔ́} \tbPS{adv.~} \tbGE{loosely} 
\tbLX{yɔrɔyɔrɔ} \tbHM{} \tbPH{yɔ̀rɔ̀yɔ̀rɔ̀} \tbPS{adj.~} \tbGE{not firm, loose e.g. mud} 
\tbLX{yuli} \tbHM{} \tbPH{yúlí} \tbPS{v.~} \tbGE{to leak slowly} \tbPD{yulee, yulo, yulo, yulibo, yulaa, yulibo } 
\tbLX{yuloo} \tbHM{} \tbPH{yúlóó} \tbPS{n.~} \tbGE{object on which bows are hung} \tbOP{yulli} \tbTP{yullee} 
\tbLX{yuo} \tbHM{} \tbPH{yùò} \tbPS{v.~} \tbGE{to open} \tbPD{yuoe, yuoro, yuoro, -, yuoraa, - } 
\tbLX{yuo} \tbHM{} \tbPH{yúó} \tbPS{adj.~} \tbGE{different, of another kind} \tbXV{Zenɛ kpare yuo la ka a baala su.} \tbXE{Today the patient is wearing a different shirt.} 
\tbLX{yuonaa} \tbHM{} \tbPH{yùònáá} \tbPS{adv.~} \tbGE{this year} \tbVA{shortened form of yuoni ŋa}
\tbLX{yuoni} \tbHM{} \tbPH{yúóní} \tbPS{n.~} \tbGE{year} \tbOP{yuomo} \tbTP{yuonee} 
\tbLX{Yuoni} \tbHM{} \tbPH{yúóní} \tbPS{n.~} \tbGE{name given to a child born after nine months of pregnancy} 
\tbLX{yuoni} \tbHM{} \tbPH{yúóní} \tbPS{n.~} \tbGE{the crop of a bird} \tbOP{yuomo} \tbTP{yuonee} 
\tbLX{yuoni ŋa} \tbHM{} \tbPH{yúóní ŋá} \tbPS{adv.~} \tbGE{this year} 
\tbLX{yuoni-yuoni} \tbHM{} \tbPH{yúóní-yúóní} \tbPS{adv.~} \tbGE{annually} 
\tbLX{yuori} \tbHM{} \tbPH{yúórí} \tbPS{n.~} \tbGE{jar} \tbOP{yoe} \tbTP{yoree} 
\tbLX{yuori} \tbHM{} \tbPH{yǔórì} \tbPS{n.~} \tbSN \tbGE{name} \tbSN \tbGE{reputation} \tbSN \tbGE{namesake} \tbOP{yoe} \tbTP{yoree} 
\tbLX{yuuli} \tbHM{} \tbPH{yùùlì} \tbPS{v.~} \tbGE{to spread small amount on top of e.g. dry cement} \tbPD{yuulee, yuulo, yuulo, yuulo, yuulaa, - } 
\tbLX{za} \tbHM{} \tbPH{zá} \tbPS{v.~} \tbGE{to throw} \tbPD{zaɛ, zaara, zaara, zareba, zaaraa, zareba } 
\tbLX{za} \tbHM{} \tbPH{zá} \tbPS{v.~} \tbGE{to palpitate (of heart)} \tbPD{zaɛ, zaara, zaara, zareba, zaaraa, zareba } 
\tbLX{zaa} \tbHM{} \tbPH{zàá} \tbPS{adj.~} \tbGE{all, every} \tbXV{A kõɔ zaa ka ba nyu.} \tbXE{They have drank all the water.} \tbVA{haa}
\tbLX{za᷉a} \tbHM{} \tbPH{zá᷉à} \tbPS{adv.~} \tbGE{yesterday; short form of zaameŋ} \tbVA{zaameŋ}
\tbLX{za᷉a} \tbHM{} \tbPH{zà᷉à᷉à} \tbPS{adv.~} \tbGE{hazy} 
\tbLX{zaale} \tbHM{} \tbPH{záálɪ́} \tbPS{v.~} \tbGE{to move water around in a container by swinging the container around} \tbPD{zaalɛɛ, zaala, zaala, zaaleba, zaalaa, - } 
\tbLX{zaale} \tbHM{} \tbPH{záálɪ́} \tbPS{v.~} \tbGE{to commit suicide by hanging oneself} \tbPD{zaalɛɛ, zaala, zaala, zaaleba, zaalaa, - } 
\tbLX{zaameŋ} \tbHM{} \tbPH{zààmɪ̂ŋ} \tbPS{adv.~} \tbGE{yesterday} \tbVA{za᷉a}
\tbLX{zaayee} \tbHM{} \tbPH{zàáyéé} \tbPS{adv.~} \tbGE{surprisingly} \tbXV{N da teɛre ka o be la a dieŋ zaayee a bie yoŋ la.} \tbXE{I thought she was in the room, but to my surprise it was only the child that was there.} 
\tbLX{zabage} \tbHM{} \tbPH{zábágɪ́} \tbPS{n.~} \tbGE{foe, enemy, invader} \tbVA{zabɔge}
\tbLX{zaboraa} \tbHM{} \tbPH{zábʊ́ráá} \tbPS{n.~} \tbGE{person under review} \tbOP{zaborre} \tbTP{zaborɛɛ} 
\tbLX{zabɔge} \tbHM{} \tbPH{zábàgɪ́} \tbPS{n.~} \tbGE{foe, enemy, invader} \tbOP{zabɔgre} \tbTP{zabɔgrɛɛ} \tbVA{zabage}
\tbLX{zaga} \tbHM{} \tbPH{zágà} \tbPS{n.~} \tbSN \tbGE{animal or poultry pen} \tbSN \tbGE{mortuary} \tbOP{zagere} \tbTP{zagerɛɛ} \tbVA{zage}
\tbLX{zage} \tbHM{} \tbPH{zágɪ́} \tbPS{n.~} \tbGE{head cold} \tbTP{zagrɛɛ} 
\tbLX{zage} \tbHM{} \tbPH{zágɪ́} \tbPS{v.~} \tbGE{to be worn out} \tbPD{zagɛɛ, zagra, zagra, zagreba, zagraa, - } 
\tbLX{zage} \tbHM{} \tbPH{zàgɪ́} \tbPS{n.~} \tbGE{enclosure for animals, pen, kraal} \tbOP{zagere} \tbTP{zagerɛɛ} \tbVA{zaga}
\tbLX{zagedie} \tbHM{} \tbPH{zàgɪ̀dìè} \tbPS{n.~} \tbGE{main chamber in a rodent's burrow} \tbOP{zagederi} \tbTP{zagederee} 
\tbLX{zagela} \tbHM{} \tbPH{záglá} \tbPS{n.~} \tbGE{one being referred to} \tbXV{Zagela maŋ yeli ka o puoro la kyɛ zeɛre o taaba.} \tbXE{Some people claim to be godly but hate their neigbours.} 
\tbLX{zagelaa} \tbHM{} \tbPH{zágláà} \tbPS{adj.~} \tbGE{empty} \tbOP{zagele} \tbTP{zagelɛɛ} 
\tbLX{zagepɛmpɛlɛŋ} \tbHM{} \tbPH{zàgɪ̀pɛ́mpɛ́lɛ́ŋ} \tbPS{n.~} \tbGE{spleen} \tbOP{zagepɛmpɛlle} \tbTP{zagepɛmpɛlɛɛ} \tbVA{hɔɔpɛmpɛlɛŋ}
\tbLX{zagere} \tbHM{} \tbPH{zàgrɪ̀} \tbPS{v.~} \tbGE{to refuse, to reject} \tbPD{zagerɛɛ, zagera, zagera, zagereba, zageraa, zagreba } 
\tbLX{zagezage} \tbHM{} \tbPH{zàgzág} \tbPS{v.~} \tbGE{to harass} \tbPD{zagezagɛɛ, zagezagera, -, zagezagereba, zagezageraa, - } 
\tbLX{zalball} \tbHM{} \tbPH{zàlbàll} \tbPS{adv.~} \tbGE{airborne} 
\tbLX{zalle} \tbHM{} \tbPH{zállɪ̀} \tbPS{v.~} \tbSN \tbGE{to swing} \tbSN \tbGE{to commit suicide by hanging oneself} \tbPD{zallɛɛ, zalla, zalla, zaleba, zallaa, zaleba } \tbVA{zɛlle}
\tbLX{zamaa} \tbHM{} \tbPH{zàmàà} \tbPS{adj.~} \tbGE{confused} 
\tbLX{zamasoba} \tbHM{} \tbPH{zámásúbá} \tbPS{n.~} \tbGE{one that is homeless} 
\tbLX{zambaa} \tbHM{} \tbPH{zàmbáá} \tbPS{n.~} \tbGE{a type of small striped bush mouse} \tbOP{zambaare} \tbTP{zambaarɛɛ} \tbVA{zambateraa}
\tbLX{zambadaa} \tbHM{} \tbPH{zàmbàdàá} \tbPS{n.~} \tbGE{stilt} \tbOP{zambadaare} \tbTP{zambadaarɛɛ} 
\tbLX{zambateraa} \tbHM{} \tbPH{zàmbàtɪ̀ràà} \tbPS{n.~} \tbGE{a type of striped mouse} \tbOP{zambaterre} \tbTP{zambaterɛɛ} \tbVA{zamba}
\tbLX{zambayɛlɛ} \tbHM{} \tbPH{zámbáyɛ́lɛ̀} \tbPS{adv.~} \tbGE{lying in a manner with face up} \tbVA{zamayɛlɛ}
\tbLX{zambɛre} \tbHM{} \tbPH{zàmbɛ̀rɪ̀} \tbPS{n.~} \tbGE{blacksmith’s shop} \tbOP{zambɛrre} 
\tbLX{zambɛre} \tbHM{} \tbPH{zàmbɛ̀rɪ̀} \tbPS{n.~} \tbSN \tbGE{a blacksmith’s shop} \tbSN \tbGE{god of the forge} \tbVA{zɛmbɛre}
\tbLX{Zambɔ} \tbHM{} \tbPH{zàmbɔ́} \tbPS{n.~} \tbGE{name of a village near Lawra in the Upper West Region of Ghana} 
\tbLX{zamm} \tbHM{} \tbPH{zàmm} \tbPS{adv.~} \tbGE{huge but light} 
\tbLX{zammo} \tbHM{} \tbPH{zàmmʊ́} \tbPS{n.~} \tbGE{onion} \tbOP{zama} \tbTP{zannɛɛ} \tbVA{gaabu (Hausa)}\tbVA{zamadaare}
\tbLX{zammuli} \tbHM{} \tbPH{zàmmúlí} \tbPS{n.~} \tbGE{tree} 
\tbLX{zamparemuu} \tbHM{} \tbPH{zâmpárɪ̀mùú} \tbPS{n.~} \tbGE{wrist guard for archery} \tbOP{zamparemuuri} \tbTP{zamparemuuree} 
\tbLX{zampoŋ} \tbHM{} \tbPH{zàmpóŋ} \tbPS{n.~} \tbGE{hedgehog} \tbOP{zamponni} \tbTP{zamponnee} 
\tbLX{zampoone} \tbHM{} \tbPH{zàmpʊ́ʊ́nɪ́} \tbPS{adj.~} \tbGE{young and helpless e.g. chick} \tbOP{zampooma} \tbTP{zampoomɛɛ} 
\tbLX{zane} \tbHM{} \tbPH{zánɪ́} \tbPS{n.~} \tbGE{wrist guard for archery} \tbOP{zama} \tbTP{zamɛɛ} 
\tbLX{zanne} \tbHM{} \tbPH{zánnɪ́} \tbPS{v.~} \tbGE{to thresh neatly or completely} \tbPD{zannɛɛ, zanna, zanna, zanema, -, - } 
\tbLX{zanne} \tbHM{} \tbPH{zànnɪ́} \tbPS{v.~} \tbGE{to learn} \tbPD{zannɛɛ, zanna, zanna, zanema, zannaa, zanema } 
\tbLX{zanne} \tbHM{} \tbPH{zànnɪ́} \tbPS{v.~} \tbGE{to dream} \tbPD{zannɛɛ, zanna, zanna, zanema, zannaa, zanema } 
\tbLX{zannoŋ} \tbHM{} \tbPH{zànnʊ̌ŋ} \tbPS{n.~} \tbGE{dream} \tbOP{zanonne} \tbTP{zanonnɛɛ} \tbVA{zannoo}
\tbLX{zannoo} \tbHM{} \tbPH{zánnʋ́ʋ́} \tbPS{n.~} \tbGE{dream} \tbVA{zannoŋ}
\tbLX{zannoo} \tbHM{} \tbPH{zànnʊ́ʊ́} \tbPS{n.~} \tbGE{education, the act of learning} 
\tbLX{zannzanna} \tbHM{} \tbPH{zànnzánná} \tbPS{n.~} \tbGE{learner, pupil} \tbSG{zannzanemɛɛ} \tbTP{zannzanema} 
\tbLX{zannyuu} \tbHM{} \tbPH{zànɲúú} \tbPS{n.~} \tbGE{garlic} \tbOP{zannyuuri} \tbTP{zannyuuree} 
\tbLX{zanzakpiri-duoruŋ} \tbHM{} \tbPH{zánzákpìri-dúórúŋ} \tbPS{n.~} \tbGE{ringworm} 
\tbLX{zanzakpirime} \tbHM{} \tbPH{zánzákpìrìmé} \tbPS{n.~} \tbGE{small bat} \tbOP{zanzakpirri} \tbTP{zanzakpirree} \tbVA{zinzakpirime}
\tbLX{zanzansɔglaa} \tbHM{} \tbPH{zánzánsɔ́glàá} \tbPS{n.~} \tbGE{a type of blackish-grey fruit bat} \tbOP{zanzansoglɔ} \tbTP{zanzansɔglɛɛ} 
\tbLX{zanzantokele} \tbHM{} \tbPH{zánzàntókélé} \tbPS{n.~} \tbGE{a type of bat} \tbOP{zanzantokelli} \tbTP{zanzantokelee} 
\tbLX{zanzanzeɛ} \tbHM{} \tbPH{zánzànzɪ́ɛ́} \tbPS{n.~} \tbGE{a type of reddish-brown bat} \tbOP{zanzanzeere} \tbTP{zanzanzeerɛɛ} 
\tbLX{zanzaŋ} \tbHM{} \tbPH{zánzáŋ} \tbPS{adv.~} \tbGE{intensely e.g. heat} \tbXV{A yɛlɛ pa᷉a᷉ ta la a zanzaŋ.} \tbXE{The matter has now got to the climax.} 
\tbLX{zanzaŋaa} \tbHM{} \tbPH{zánzáŋàá} \tbPS{n.~} \tbGE{bat} \tbOP{zanzanne} \tbTP{zanzannɛɛ} 
\tbLX{zanzaŋaa} \tbHM{} \tbPH{zánzáŋàá} \tbPS{n.~} \tbSN \tbGE{a type of shellfish similar to a clam} \tbSN \tbGE{the shell of  this shelfish} \tbOP{zanzanne} \tbTP{zanzannɛɛ} 
\tbLX{zanzaŋkpara} \tbHM{} \tbPH{zànzáŋkpàrà} \tbPS{n.~} \tbGE{the act of jumping up and down in hysteria} \tbOP{zanzaŋkparre} \tbTP{zanzaŋkparɛɛ} 
\tbLX{zanzunmuli} \tbHM{} \tbPH{zaǹzùmmùlì} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} 
\tbLX{zaŋaa} \tbHM{} \tbPH{záŋáá} \tbPS{n.~} \tbGE{shell} \tbOP{zanne} \tbTP{zannɛɛ} 
\tbLX{zaŋe} \tbHM{} \tbPH{záŋ} \tbPS{v.~} \tbGE{to itch} \tbPD{zaŋɛɛ, zaŋena, zaŋena, zaŋeneba, zaŋenaa, - } 
\tbLX{zaŋgalaŋ} \tbHM{} \tbPH{záŋgáláŋ} \tbPS{n.~} \tbGE{veranda} \tbOP{zaŋgalle} \tbTP{zaŋgalɛɛ} 
\tbLX{zaŋgbaale} \tbHM{} \tbPH{zàŋgbààlɪ̀} \tbPS{n.~} \tbGE{dome-shaped cage} \tbOP{zaŋgbaala} \tbTP{zaŋgbaalɛɛ} 
\tbLX{zaŋkpegiri} \tbHM{} \tbPH{zàŋkpégrí} \tbPS{n.~} \tbGE{a type of hard pimple} \tbOP{zaŋkpege} \tbTP{zaŋkpegee} \tbVA{kpaŋkpazegele}
\tbLX{zare} \tbHM{} \tbPH{zárɪ̀} \tbPS{n.~} \tbGE{large metal hammer used in the forge} \tbOP{ zaɛ} \tbTP{zarɛɛ} 
\tbLX{zarema} \tbHM{} \tbPH{zárɪ́má} \tbPS{n.~} \tbGE{rags} \tbTP{zaremɛɛ} \tbVA{zɛremɛ}
\tbLX{zazaakɔɔ} \tbHM{} \tbPH{zàzáákɔ́ɔ́} \tbPS{adj.~} \tbGE{steady} \tbVA{dedeekɔɔ}
\tbLX{ze᷈e} \tbHM{} \tbPH{zɪ́᷈ɪ́᷈} \tbPS{n.~} \tbGE{blood} \tbTP{zeenɛɛ} 
\tbLX{ze᷈ete᷉e} \tbHM{} \tbPH{zɪ́᷈ɪ̀᷈tɪ́᷈ɪ́᷈} \tbPS{n.~} \tbGE{medicine that helps improve the level of blood in the body} \tbTP{ze᷉eteenɛɛ} 
\tbLX{ze᷉ewoɔ} \tbHM{} \tbPH{zɪ́᷉ɪ́᷉wʊ́ɔ́} \tbPS{n.~} \tbGE{blood vessel} \tbOP{ze᷉ewoore} \tbTP{ze᷉ewoorɛɛ} 
\tbLX{zeɛ} \tbHM{} \tbPH{zɪ̀ɛ́} \tbPS{n.~} \tbGE{late millet} \tbSG{zeroo (for plant)} \tbOP{zeere} \tbTP{zeerɛɛ} 
\tbLX{zeɛ} \tbHM{} \tbPH{zɪ̀ɛ̀} \tbPS{v.~} \tbGE{to snatch, to seize} \tbPD{zeɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, zeɛrebɛ } 
\tbLX{zeɛ} \tbHM{} \tbPH{zɪ̀ɛ́} \tbPS{adj.~} \tbGE{reddish brown} \tbOP{zeere} \tbTP{zerɛɛ} 
\tbLX{zeɛ} \tbHM{} \tbPH{zɪ̀ɛ́} \tbPS{adj.~} \tbGE{serious} \tbXV{Yelzeɛ la ka ba yele, azuiŋ ta laara.} \tbXE{It is a serious case, so don't laugh.} \tbOP{zeere} \tbTP{zerɛɛ} 
\tbLX{ze᷉ɛ} \tbHM{} \tbPH{zɪ̀᷉ɛ́᷉} \tbPS{n.~} \tbGE{acid} \tbTP{zeɛnɛɛ} 
\tbLX{ze᷉ɛ} \tbHM{} \tbPH{zɪ́᷉ɛ́᷉} \tbPS{n.~} \tbGE{fortification for one who has commited murder or killed an animal believed to have supernatural powers} \tbTP{zeɛnɛɛ} 
\tbLX{ze᷉ɛ} \tbHM{} \tbPH{zɪ̀᷉ɛ̀} \tbPS{v.~} \tbGE{to be needed, to be necessary} \tbPD{ze᷉ɛɛ, zeɛnɛ, zeɛnɛ, zeɛnemɛ, zeɛnaa, - } 
\tbLX{ze᷉ɛda᷉a} \tbHM{} \tbPH{zí᷉ɛ́᷉dá᷉á᷉} \tbPS{n.~} \tbGE{local beer prepared solely for ritual purposes} \tbTP{ze᷉ɛdaanɛɛ} 
\tbLX{zeɛ-genegene} \tbHM{} \tbPH{zɪ̀ɛ́-gènègènè} \tbPS{adj.~} \tbGE{brown} 
\tbLX{ze᷉ɛgoɔ} \tbHM{} \tbPH{zɪ́᷉ɛ́᷉gʊ́ɔ̀} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbOP{ze᷉ɛgoore} 
\tbLX{zeɛmaa} \tbHM{} \tbPH{zɪ́ɛ́máá} \tbPS{n.~} \tbGE{open ground where millet is gathered, threshed, and winnowed out} \tbOP{zeɛne} \tbTP{zeɛnɛɛ/zeɛmɛɛ} \tbVA{gbeŋgbelaa}
\tbLX{Zeɛmɛ} \tbHM{} \tbPH{zɪ́ɛ́mɛ̀} \tbPS{n.~} \tbGE{name traditionally given to the firstborn of twins (male)} 
\tbLX{Ze᷉ɛpɔge} \tbHM{} \tbPH{zɪ̀᷉ɛ́᷉pɔ́gɪ́} \tbPS{n.~} \tbGE{name traditionally given to the firstborn of twins (female)} \tbOP{Ze᷉ɛpɔgemine} 
\tbLX{zeɛraa} \tbHM{} \tbPH{zɪ́ɛ́ráá} \tbPS{n.~} \tbGE{sieve} \tbOP{zeɛre} \tbTP{zeɛrɛɛ} 
\tbLX{zeɛraa} \tbHM{} \tbPH{zɪ́ɛ́ráá} \tbPS{n.~} \tbGE{hawk} \tbOP{zeɛre} \tbTP{zeɛrɛɛ} 
\tbLX{zeɛre} \tbHM{} \tbPH{zɪ̀ɛ̀rɪ́} \tbPS{n.~} \tbSN \tbGE{sauce} \tbSN \tbGE{soup} \tbOP{ zɛrɛɛ} 
\tbLX{zeɛre} \tbHM{} \tbPH{zɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to hate, to despise} \tbPD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\tbLX{zeɛre} \tbHM{} \tbPH{zɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to trot} \tbPD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\tbLX{zeɛre} \tbHM{} \tbPH{zɪ́ɛ́rɪ̀} \tbPS{v.~} \tbGE{to be in haste, to go after something eagerly} \tbPD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\tbLX{zege} \tbHM{} \tbPH{zɪ̀gɪ̀} \tbPS{v.~} \tbGE{to grow to maturity and full strength} \tbPD{zegɛɛ, zegerɛ, zegerɛ, zegerebɛ, zegeraa, - } 
\tbLX{zegere} \tbHM{} \tbPH{zɪ́grɪ̀} \tbPS{v.~} \tbGE{to shrug; raise one’s shoulders, indicating lack of concern} \tbPD{zegerɛɛ, zegerɛ, zegerɛ, zegerebɛ, zegeraa, - } 
\tbLX{zegere} \tbHM{} \tbPH{zɪ̀grɪ̀} \tbPS{v.~} \tbGE{to grow to maturity} \tbPD{zegerɛɛ, zegerɛ, zegerɛ, zegerebɛ, zegeraa, - } 
\tbLX{zegezege} \tbHM{} \tbPH{zɪ̀gzɪ̀g} \tbPS{v.~} \tbGE{to be impatient} \tbPD{zegezegɛɛ, zegezegrɛ, zegezegerɛ, zegezegerebɛ, zegezegeraa, - } 
\tbLX{zegɛ} \tbHM{} \tbPH{zɪ́gɛ́} \tbPS{n.~} \tbSN \tbGE{area behind and between the shoulders} \tbSN \tbGE{hump (of a cow)} \tbOP{zegere} \tbTP{zegerɛɛ} 
\tbLX{zegitege} \tbHM{} \tbPH{zègìtégé} \tbPS{adv.~} \tbGE{not moving at all, standing still in the way} 
\tbLX{zele} \tbHM{} \tbPH{zɪ̀lɪ́} \tbPS{n.~} \tbGE{tongue} \tbOP{zelɛ} \tbTP{zelɛɛ/zellɛɛ} 
\tbLX{zelegannoɔre} \tbHM{} \tbPH{zɪ̀lɪ́gànʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{a type small bird with blue feathers and a long beak that is fond of flowers} \tbOP{zelegannɔɛ} 
\tbLX{zelɛzelɛ} \tbHM{} \tbPH{zɪ̀lɛ́zɪ̀lɛ́} \tbPS{adj.~} \tbGE{slim, at the point of giving way} 
\tbLX{zelgannoɔre} \tbHM{} \tbPH{zɪ̌lgǎnnʊ́ɔ́rɪ̀} \tbPS{n.~} \tbGE{a type of small bird with a very long curved beak} \tbOP{zelgannɔɛ} \tbTP{zelgannɔrɛɛ} 
\tbLX{zell} \tbHM{} \tbPH{zɪ́ll} \tbPS{adv.~} \tbGE{filled to the brim} 
\tbLX{zelpi᷉i} \tbHM{} \tbPH{zɪ̌pí᷉ì᷉} \tbPS{n.~} \tbGE{a type of disease that affects the tongue of dogs causing them to not eat} \tbOP{zelpiime} \tbTP{zelpiimee} 
\tbLX{zemberemɛ} \tbHM{} \tbPH{zɪ́mbɪ́rɪ́mɛ́} \tbPS{n.~} \tbGE{a type of tree whose fruit gives red juice} \tbOP{zemberre} \tbTP{zemberɛɛ/zemberrɛɛ} 
\tbLX{Zempɛnne} \tbHM{} \tbPH{zɪ̀mpɛ́nnɪ́} \tbPS{n.~} \tbGE{name of a village} 
\tbLX{zenɛ} \tbHM{} \tbPH{zɪ́nɛ̀} \tbPS{n.~} \tbGE{today} 
\tbLX{zenɛbie} \tbHM{} \tbPH{zɪ́nɛ̀bìé} \tbPS{adv.~} \tbGE{nowadays, the present time} 
\tbLX{zenzie} \tbHM{} \tbPH{zɪ̀nzíé} \tbPS{n.~} \tbSN \tbGE{sitting place} \tbSN \tbGE{settlement} \tbOP{zenziiri} \tbTP{zenziiree} 
\tbLX{zeŋe} \tbHM{} \tbPH{zɪ̀ŋ} \tbPS{v.~} \tbSN \tbGE{to sit, to be seated} \tbSN \tbGE{to settle} \tbPD{zeŋɛɛ, zenɛ, zenɛ, zenemɛ, zenaa, zenemɛ } 
\tbLX{zeŋgbiri} \tbHM{} \tbPH{zɪ̀ŋgbírì} \tbPS{n.~} \tbGE{sleeping sickness} 
\tbLX{zɛ} \tbHM{} \tbPH{zɛ̀} \tbPS{v.~} \tbGE{to smear} \tbPD{zɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, zeɛrebɛ } 
\tbLX{zɛ} \tbHM{} \tbPH{zɛ́} \tbPS{v.~} \tbGE{to sift} \tbPD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\tbLX{zɛ} \tbHM{} \tbPH{zɛ́} \tbPS{v.~} \tbGE{to hate} \tbPD{zeɛrɛɛ, zeɛrɛ, zeɛrɛ, zeɛrebɛ, zeɛraa, - } 
\tbLX{zɛdambo} \tbHM{} \tbPH{zɛ̀dàmbó} \tbPS{n.~} \tbGE{unseasoned soup} \tbOP{zɛdambori} \tbTP{zɛdamboree} 
\tbLX{zɛdogelee} \tbHM{} \tbPH{zɛ̀dʊ́gléé} \tbPS{n.~} \tbGE{earthenware cooking pot for soup} \tbOP{zɛdogebilii} \tbVA{saa-doge}
\tbLX{zɛɛ} \tbHM{} \tbPH{zɛ̀ɛ̀} \tbPS{adv.~} \tbGE{sticking on to all over e.g. fruits on a tree} 
\tbLX{zɛge} \tbHM{} \tbPH{zɛ́gɪ̀} \tbPS{v.~} \tbSN \tbGE{to raise, to lift above} \tbSN \tbGE{to promote} \tbPD{zɛgɛɛ, zɛgrɛ, zɛgrɛ, zɛgrebɛ, zɛgraa, - } 
\tbLX{zɛgelɛkɛlɛ} \tbHM{} \tbPH{zɛ́glɛ́kɛ́lɛ́} \tbPS{adv.~} \tbGE{not missing out} \tbXV{A pɔgeba zaa ne ba biiri da la, zɛgelɛkɛlɛ.} \tbXE{Each of the women had a child.} 
\tbLX{zɛgere} \tbHM{} \tbPH{zɛ́grɪ̀} \tbPS{v.~} \tbSN \tbGE{to shake (hands)} \tbSN \tbGE{to shake in the hand to weigh} \tbPD{zɛgerɛɛ, zɛgerɛ, zɛgerɛ, zɛgerebɛ, zɛgeraa, - } 
\tbLX{zɛgeroŋ} \tbHM{} \tbPH{zɛ̊grʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{something that has worth} \tbSN \tbGE{something that has scent} 
\tbLX{zɛko᷉ɔ} \tbHM{} \tbPH{zɛ̀kʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{soup that is not thick; broth-like} \tbOP{zɛkoɔnɛɛ} 
\tbLX{zɛle} \tbHM{} \tbPH{zɛ́lɪ́} \tbPS{v.~} \tbSN \tbGE{to apply poison; to be filled with venom} \tbSN \tbGE{to be furious} \tbSN \tbGE{to be bitter (taste)} \tbPD{-, -, -, -, -, - } 
\tbLX{zɛle} \tbHM{} \tbPH{zɛ́lɪ́} \tbPS{n.~} \tbGE{afterbirth (nonhuman)} \tbOP{zɛlɛ} \tbTP{zɛlɛɛ} 
\tbLX{zɛle} \tbHM{} \tbPH{zɛ̀lɪ̀} \tbPS{v.~} \tbGE{to lift, to pick up} \tbPD{zɛlɛɛ, zɛllɛ, zɛllɛ, zɛlebɛ, zɛllaa, zɛlebɛ } 
\tbLX{zɛle} \tbHM{} \tbPH{zɛ̀lɪ̀} \tbPS{v.~} \tbGE{to beg (Northern Dagaare dialect)} \tbPD{zɛlɛɛ, zɛllɛ, zɛllɛ, zɛlebɛ, zɛllaa, zɛlebɛ } 
\tbLX{zɛlezɛle} \tbHM{} \tbPH{zɛ̀lɪ̀zɛ́lɪ́} \tbPS{v.~} \tbGE{to move a heavy load with difficulty} \tbPD{zɛlezɛlɛɛ, zɛlezɛlɛ, zɛlezɛlɛ, zɛlezɛlebɛ, zɛlezɛlaa, - } 
\tbLX{zɛlɛŋkpɛ} \tbHM{} \tbPH{zɛ̀lɛ̀ŋkpɛ̀} \tbPS{adj.~} \tbGE{hanging loosely} 
\tbLX{zɛlle} \tbHM{} \tbPH{zɛ́llɪ̀} \tbPS{v.~} \tbSN \tbGE{to swing} \tbSN \tbGE{to commit suicide by hanging oneself} \tbPD{-, -, -, -, -, - } \tbVA{zalle}
\tbLX{zɛmbɛre} \tbHM{} \tbPH{zɛ̀mbɛ̀rɪ̀} \tbPS{n.~} \tbSN \tbGE{a blacksmith’s shop} \tbSN \tbGE{god of the forge} \tbOP{ zɛmbɛrɛɛ} \tbVA{zambɛre}
\tbLX{zɛmɛzɛmɛ} \tbHM{} \tbPH{zɛ̀mɛ̀zɛ̀mɛ̀} \tbPS{adv.~} \tbGE{long and flexible e.g. cane} 
\tbLX{zɛmiime} \tbHM{} \tbPH{zɛ̀míímé} \tbPS{n.~} \tbGE{sour sauce} 
\tbLX{zɛmm} \tbHM{} \tbPH{zɛ̀mm} \tbPS{adv.~} \tbGE{quietly, gently, and slowly} 
\tbLX{zɛnzɛŋ} \tbHM{} \tbPH{zɛ̀nzɛ́ŋ} \tbPS{v.~} \tbSN \tbGE{to be impatient} \tbSN \tbGE{to be eager, to be zealous} \tbPD{zɛnzɛŋɛɛ, zɛnzɛnɛ, zɛnzɛnɛ, zɛnzɛnemɛ, zɛnzɛnaa, zɛnzɛnemɛ } 
\tbLX{zɛnzɛŋ} \tbHM{} \tbPH{zɛ̀nzɛ̌ŋ} \tbPS{n.~} \tbGE{impatience} \tbOP{zɛnzɛnne} \tbTP{zɛnzɛnɛɛ} 
\tbLX{zɛŋ} \tbHM{} \tbPH{zɛ̀ŋ} \tbPS{v.~} \tbGE{to lift} \tbPD{zɛŋɛɛ, zɛnnɛ, zɛnnɛ, zɛnemɛ, zɛnnaa, - } 
\tbLX{zɛŋe} \tbHM{} \tbPH{zɛ̂ŋ} \tbPS{n.~} \tbGE{to shake e.g. tree} \tbOP{ zenɛ  zenzenɛ  zenemɛ} 
\tbLX{zɛrekpɛrɛ} \tbHM{} \tbPH{zɛ̀rɪ̀kpɛ̀rɛ̀} \tbPS{adv.~} \tbGE{hanging or swelling all over e.g. fruits} 
\tbLX{zɛremɛ} \tbHM{} \tbPH{zɛ́rɪ́mɛ́} \tbPS{n.~} \tbGE{rags} \tbVA{zarema}
\tbLX{zɛroo} \tbHM{} \tbPH{zɛ̀rʊ̀ʊ̀} \tbPS{adv.~} \tbGE{fleshy and loose (hanging)} 
\tbLX{zɛvaa} \tbHM{} \tbPH{zɛ́váà} \tbPS{n.~} \tbGE{vegetable} \tbSG{zɛvaabo} \tbOP{zɛvaare} \tbTP{zɛvaarɛɛ} 
\tbLX{zibalaa} \tbHM{} \tbPH{zìbàlàà} \tbPS{n.~} \tbGE{odd place} 
\tbLX{zibalaa} \tbHM{} \tbPH{zìbáláá} \tbPS{n.~} \tbGE{detour, diversion (road)} \tbOP{ziballe} 
\tbLX{ziberaa} \tbHM{} \tbPH{zìbèráà} \tbPS{n.~} \tbGE{the place where a person or something is} \tbOP{ ziberri} 
\tbLX{ziberaa} \tbHM{} \tbPH{zìbɪ̀ráà} \tbPS{n.~} \tbGE{late night} \tbOP{ziberre} 
\tbLX{zibiri} \tbHM{} \tbPH{zìbírì} \tbPS{n.~} \tbGE{clitoris} \tbOP{zibie} \tbTP{zibiiree} \tbVA{gyambiri}
\tbLX{zibu} \tbHM{} \tbPH{zíé} \tbPS{n.~} \tbGE{python} \tbVA{zigi}
\tbLX{zie} \tbHM{} \tbPH{zìé} \tbPS{n.~} \tbGE{place} \tbOP{ziiri} \tbTP{ziiree} 
\tbLX{zie} \tbHM{} \tbPH{zìé} \tbPS{n.~} \tbGE{weather} \tbOP{ziiri} \tbTP{ziiree} 
\tbLX{zie} \tbHM{} \tbPH{zìé} \tbPS{n.~} \tbGE{time} \tbOP{ziiri} \tbTP{ziiree} 
\tbLX{ziezaa} \tbHM{} \tbPH{zíézàá} \tbPS{adv.~} \tbGE{everywhere} 
\tbLX{zigi} \tbHM{} \tbPH{zígí} \tbPS{n.~} \tbGE{python} \tbOP{zigri} \tbTP{zigree/zigee} \tbVA{zibu}
\tbLX{zigi} \tbHM{} \tbPH{zígí} \tbPS{v.~} \tbSN \tbGE{to grow in size} \tbSN \tbGE{to grow more leaves (yam vine)} \tbPD{zigee, zigire, zigire, zigiribe, zigiraa, - } 
\tbLX{zigiree} \tbHM{} \tbPH{zígréé} \tbPS{n.~} \tbGE{gravel, laterite} \tbOP{zigiri} 
\tbLX{zigizukpara} \tbHM{} \tbPH{zìgìzúkpárá} \tbPS{n.~} \tbGE{a monster; in the form of a two-headed python} \tbOP{zigizukparre} 
\tbLX{zigoo} \tbHM{} \tbPH{zìgòò} \tbPS{adj.~} \tbGE{showing no concern for others} \tbXV{A bie kono  tuuro la o ma kyɛ ka o deɛ ba taa zigoo zaa.} \tbXE{The child is crying while running after the mother but woman did not show any concern.} 
\tbLX{zii} \tbHM{} \tbPH{zìì} \tbPS{adj.~} \tbGE{thick; foliage of a tree. (Used only in this situation)} 
\tbLX{zi᷉i} \tbHM{} \tbPH{zì᷈ì᷈} \tbPS{adj.~} \tbGE{not clear, blurry} 
\tbLX{zikuoŋaa} \tbHM{} \tbPH{zìkúóŋáá} \tbPS{n.~} \tbGE{dry place} \tbOP{zikuoni} \tbTP{zikuonee} 
\tbLX{zikpeɛraa} \tbHM{} \tbPH{zìkpɪ́ɛ́ráà} \tbPS{n.~} \tbGE{inhabitance, residence} \tbOP{zikpeɛre} 
\tbLX{zikyɛnfeɛle} \tbHM{} \tbPH{zìʧɛ̀nfɪ́ɛ́lɪ́} \tbPS{n.~} \tbGE{cold early morning} \tbOP{zikyɛnfeɛlɛ} 
\tbLX{zimaale} \tbHM{} \tbPH{zìmààlɪ̀} \tbPS{n.~} \tbGE{development} 
\tbLX{zimaaliire} \tbHM{} \tbPH{zìmáálììré} \tbPS{n.~} \tbGE{dusk, twilight (after sunset)} \tbOP{zimaaliiri} \tbTP{zimaaliiree} 
\tbLX{zimaane} \tbHM{} \tbPH{zìmáánɪ́} \tbPS{adv.~} \tbGE{evening} 
\tbLX{zimaaroŋ} \tbHM{} \tbPH{zìmàárʊ́ŋ} \tbPS{n.~} \tbGE{cool or wet place} \tbOP{zimaara} 
\tbLX{zimaavɛ᷉ɛ} \tbHM{} \tbPH{zìmáávɛ̀᷉ɛ̀᷉} \tbPS{n.~} \tbGE{evening; between four-thirty and five o'clock} 
\tbLX{zimaavɛnɛvɛnɛ} \tbHM{} \tbPH{zìmáávɛ̀nɛ̀vɛ̀nɛ̀} \tbPS{n.~} \tbGE{dusk} 
\tbLX{zimezime} \tbHM{} \tbPH{zímézímé} \tbPS{adv.~} \tbGE{of considerable weight} 
\tbLX{zimɛrɛmɛrɛ} \tbHM{} \tbPH{zìmɛ̀rɛ̀mɛ̀rɛ̀} \tbPS{n.~} \tbGE{muddy place} 
\tbLX{zinziŋ} \tbHM{} \tbPH{zíŋzíŋ} \tbPS{adv.~} \tbGE{very heavy} 
\tbLX{zinyaane} \tbHM{} \tbPH{zìɲàànɪ́} \tbPS{n.~} \tbGE{dawn} \tbTP{zinyaanɛɛ} \tbVA{zivɛlemɛ}
\tbLX{ziŋi} \tbHM{} \tbPH{zîŋ} \tbPS{v.~} \tbGE{to pound vigorously} \tbPD{ziŋee, ziŋine, ziŋine, ziŋime, ziŋinaa, - } 
\tbLX{zipigi} \tbHM{} \tbPH{zìpìgí} \tbPS{n.~} \tbGE{a type of large brown antelope with white spots on its sides} \tbOP{zipigri} \tbTP{zipigree} 
\tbLX{ziponsentaa} \tbHM{} \tbPH{zìpónsɪ́ntáá} \tbPS{n.~} \tbGE{midnight} 
\tbLX{zirii} \tbHM{} \tbPH{zìríì} \tbPS{n.~} \tbGE{a lie or lies} \tbTP{zirree} \tbVA{ziriŋ}
\tbLX{ziriŋ} \tbHM{} \tbPH{zìrǐŋ} \tbPS{n.~} \tbGE{lies} \tbTP{zirree} \tbVA{zirii}
\tbLX{ziriŋma} \tbHM{} \tbPH{zìríŋmà} \tbPS{n.~} \tbGE{the act of lying} 
\tbLX{ziriŋmaara} \tbHM{} \tbPH{zìríŋmáárá} \tbPS{n.~} \tbGE{liar} \tbOP{ ziriŋmaareba} 
\tbLX{zisɔgelaa} \tbHM{} \tbPH{zìsɔ́glåá} \tbPS{n.~} \tbSN \tbGE{dark place} \tbSN \tbGE{hiding place} \tbOP{ zisɔgele} 
\tbLX{zitɔɔre} \tbHM{} \tbPH{zìtɔ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{distant place} \tbOP{zitɔbɔ} 
\tbLX{zivelaa} \tbHM{} \tbPH{zìvɪ̀làá} \tbPS{n.~} \tbGE{good place, nice place} \tbOP{ziveɛle} \tbTP{ziveɛlɛɛ} 
\tbLX{zivɛlemɛ} \tbHM{} \tbPH{zìvɛ́lɪ́mɛ́} \tbPS{n.~} \tbGE{dawn} \tbTP{zivɛlemɛɛ} \tbVA{zinyaane}
\tbLX{ziyuo} \tbHM{} \tbPH{zìyùò} \tbPS{n.~} \tbGE{modernization} 
\tbLX{zo} \tbHM{} \tbPH{zò} \tbPS{v.~} \tbSN \tbGE{to run} \tbSN \tbGE{to run away, to escape} \tbSN \tbGE{to flow (river water)} \tbPD{zoe, zoro, zoro, zoribo, zoraa, zoribo } \tbVA{zoori (of nominal form)}
\tbLX{zoge} \tbHM{} \tbPH{zʊ̀gɪ̀} \tbPS{v.~} \tbGE{to thicken, to become hard e.g. palm of a one who does rough/odd jobs} \tbPD{zogɛɛ, zogerɔ/zogera, zogerɔ/zogera, zogereba, zogeraa, - } 
\tbLX{zogele} \tbHM{} \tbPH{zʊ̀glɪ̀} \tbPS{v.~} \tbGE{to develop large rashes} \tbPD{zogelɛɛ, zogelɔ/zogela, zogelɔ/zogela, zogeleba, zogelaa, - } 
\tbLX{zogelɔŋ} \tbHM{} \tbPH{zʊ̀glɔŋ} \tbPS{adj.~} \tbGE{thick and attached to} \tbVA{zogelaŋ}
\tbLX{zogere} \tbHM{} \tbPH{zʊ́grɪ̀} \tbPS{v.~} \tbGE{to shake vigorously} \tbPD{zogrɛɛ, zogra, zogra, zoreba, zograa, - } 
\tbLX{zogere} \tbHM{} \tbPH{zʊ̀grɪ̀} \tbPS{v.~} \tbGE{to become thick or hard} \tbPD{zogerɛɛ, zogera, zogera, zogereba, zogeraa, - } 
\tbLX{zogetogo} \tbHM{} \tbPH{zʊ̀gtʊ̀gʊ̀} \tbPS{adv.~} \tbGE{thick and wool-like} 
\tbLX{zogboŋgboli} \tbHM{} \tbPH{zʊ̀gbòŋgbòlì} \tbPS{n.~} \tbGE{the thickest part of an animal’s tail that is long} \tbOP{zogboŋgbolo} \tbTP{zogboŋgbolee} 
\tbLX{zokpore} \tbHM{} \tbPH{zʊ́kpʊ́rɪ́} \tbPS{n.~} \tbGE{the bottom end of the spinal cord} \tbOP{zokpoɛ} \tbTP{zokporɛɛ} 
\tbLX{zolaa} \tbHM{} \tbPH{zʊ́láá} \tbPS{n.~} \tbGE{pot with a very small mouth} \tbOP{zolaare} \tbTP{zolaarɛɛ} 
\tbLX{zolaa} \tbHM{} \tbPH{zʊ̀làà} \tbPS{adj.~} \tbGE{big, large} 
\tbLX{zoma} \tbHM{} \tbPH{zʊ̀má} \tbPS{n.~} \tbGE{instinct} \tbVA{zomɔ}
\tbLX{zombaa} \tbHM{} \tbPH{zʊ́mbáá} \tbPS{n.~} \tbGE{catfish} \tbOP{zombaare} \tbTP{zombaarɛɛ} 
\tbLX{zomberemɛ} \tbHM{} \tbPH{zʊ́mbɪ́rɪ́mɛ́} \tbPS{n.~} \tbGE{a type of tree with medicinal properties} \tbTP{zomberemɛɛ} 
\tbLX{Zombɛnte} \tbHM{} \tbPH{zúmbɛ́ntɪ́} \tbPS{n.~} \tbGE{a religious festival of Southern Dagao} \tbOP{zombɛntere} \tbTP{zombɛnterɛɛ} 
\tbLX{zombɛraa} \tbHM{} \tbPH{zʊ́bɛ́ráá} \tbPS{n.~} \tbGE{fishhook} \tbOP{ zombɛrre} 
\tbLX{zombiri} \tbHM{} \tbPH{zʊ̀mbírì} \tbPS{n.~} \tbGE{insult, abuse} \tbSG{zombiruu} \tbOP{zombie} \tbTP{zombiiree} 
\tbLX{zomm} \tbHM{} \tbPH{zómm} \tbPS{adv.~} \tbGE{still, calm, unmoving} 
\tbLX{zommo} \tbHM{} \tbPH{zʊ́mmʊ́} \tbPS{n.~} \tbGE{fish} \tbOP{zoma} \tbTP{zonnɛɛ} 
\tbLX{zomɔ} \tbHM{} \tbPH{zʊ̀mɔ́} \tbPS{n.~} \tbGE{instinct} \tbVA{zoma}
\tbLX{zompɛge} \tbHM{} \tbPH{zʊ́mpɛ́gɪ̀} \tbPS{n.~} \tbGE{fish scale} \tbOP{zompɛgre/zompɛgɛ} \tbTP{zompɛgrɛɛ} 
\tbLX{zompi᷉i᷉} \tbHM{} \tbPH{zʊ́pí᷉ì᷉} \tbPS{n.~} \tbGE{fishhook} \tbOP{zompiime} \tbTP{zompiimee/zompiinee} 
\tbLX{zone} \tbHM{} \tbPH{zʊ́nɪ́} \tbPS{n.~} \tbGE{insult, abuse} \tbOP{zoma} \tbTP{zonnɛɛ} 
\tbLX{zonkyoo} \tbHM{} \tbPH{zʊ́nʧóò} \tbPS{n.~} \tbGE{fishing net} \tbOP{zonkyoori} \tbTP{zonkyooree} 
\tbLX{zonne} \tbHM{} \tbPH{zúnnɪ́} \tbPS{v.~} \tbGE{to take short, heavy steps} \tbPD{zonnɛɛ, zonnɔ/zonna, zonnɔ/zonna, zonemɔ/zonema, zonnaa, - } 
\tbLX{zonne} \tbHM{} \tbPH{zúnnɪ́} \tbPS{v.~} \tbGE{to shake (from sudden shock)} \tbPD{zonnɛɛ, zonnɔ/zonna, zonnɔ/zonna, zonemɔ/zonema, zonnaa, - } 
\tbLX{zonni} \tbHM{} \tbPH{zònnì} \tbPS{v.~} \tbGE{to hesitate} \tbPD{zonnee, zonno, zonno, zonimo, zonnaa, zonimo } 
\tbLX{zonnyɔge} \tbHM{} \tbPH{zʊ́nɲɔ́gɪ́} \tbPS{n.~} \tbGE{the practice of fishing} 
\tbLX{zonnyɔgerɔ} \tbHM{} \tbPH{zʊ̌nɲɔ́grɔ̀} \tbPS{n.~} \tbGE{fisherman} \tbOP{zonnyɔgerebɔ/zonnyɔgereba} \tbTP{zonnyɔgerebɛɛ} 
\tbLX{zonsaalaa} \tbHM{} \tbPH{zʊ́nsààglàá} \tbPS{n.~} \tbGE{a type of fish with a flat head and two fins and a very slippery body, found in rivers and ponds, referred to as mudfish} \tbOP{zonsaale} \tbTP{zonsaalɛɛ} 
\tbLX{zonsɔgelaa} \tbHM{} \tbPH{zʊ́nsɔ́gláá} \tbPS{n.~} \tbGE{a type of black mudfish with a flat head} \tbOP{zonsɔgelɔ} \tbTP{zonsɔgelɛɛ} 
\tbLX{zontalane} \tbHM{} \tbPH{zʊ̀ntàlánɪ̀} \tbPS{n.~} \tbGE{gill} \tbOP{zontalama} \tbTP{zontalanne} 
\tbLX{zonwaabo} \tbHM{} \tbPH{zʊ́nwàábʊ́} \tbPS{n.~} \tbGE{eel} \tbOP{zonweere} \tbTP{zonweerɛɛ} 
\tbLX{zo nyonono} \tbHM{} \tbPH{zò ɲʊ́nʊ́nʊ́} \tbPS{v.~} \tbGE{to have mercy} \tbPD{-, -, -, -, -, - } 
\tbLX{zoŋe} \tbHM{} \tbPH{zʊ́ŋ} \tbPS{v.~} \tbGE{to shake (person)} \tbPD{zoŋɛɛ, zoŋenɔ, zoŋenɔ, zoŋenema, zoŋenaa, - } 
\tbLX{zoŋkoroŋ} \tbHM{} \tbPH{zʊ̀ŋkʊ̀rʊ̀ŋ} \tbPS{adv.~} \tbGE{grotesque; hideous} 
\tbLX{zoŋkɔɔre} \tbHM{} \tbPH{zʊ́ŋkɔ́ɔ̀ré} \tbPS{n.~} \tbGE{fish bone} \tbOP{zoŋkɔbɔ} \tbTP{zoŋkɔɔrɛɛ} 
\tbLX{zo᷉o} \tbHM{} \tbPH{zʊ̀᷈ʊ̀᷉} \tbPS{adv.~} \tbGE{dull} 
\tbLX{zoore} \tbHM{} \tbPH{zʊ́ʊ́ré} \tbPS{n.~} \tbGE{tail} \tbOP{zoɛ} \tbTP{zoorɛɛ} 
\tbLX{zoore} \tbHM{} \tbPH{zʊ́ʊ́rɪ̀} \tbPS{n.~} \tbGE{smoke} \tbVA{nyoore}
\tbLX{zo᷈ɔ} \tbHM{} \tbPH{zʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{rust (typically used with le)} \tbXV{A soɔ lɔre la ko᷉ɔ poɔ ana la so ka o le zo᷉ɔ.} \tbXE{The knife got soaked in water and that is why it is rusted.} \tbTP{zoɔnɛɛ} 
\tbLX{zoɔle} \tbHM{} \tbPH{zʊ́ɔ̀lɪ̀} \tbPS{v.~} \tbGE{to wither} \tbPD{zoɔlɛɛ, zoɔlɔ, zoɔlɔ, zoɔlba, zoɔlaa, - } 
\tbLX{zori} \tbHM{} \tbPH{zòrì} \tbPS{v.~} \tbGE{to hold the hand in anticipation for food} \tbPD{zoree, zoro, zoro, zoribo, zoraa, - } 
\tbLX{zorre} \tbHM{} \tbPH{zʊ́rrɪ̀} \tbPS{v.~} \tbGE{shudder} \tbPD{zorrɛɛ, zorra, zorra, zoreba, zorraa, - } 
\tbLX{zo vi} \tbHM{} \tbPH{zò ví} \tbPS{v.~} \tbSN \tbGE{to be shy} \tbSN \tbGE{to be ashamed} \tbPD{-, -, -, -, -, - } 
\tbLX{zɔ} \tbHM{} \tbPH{zɔ̌} \tbPS{n.~} \tbSN \tbGE{father} \tbSN \tbGE{friend (between males)} \tbOP{zɔmine/zɔre} \tbTP{zɔrɛɛ} 
\tbLX{zɔba} \tbHM{} \tbPH{zɔ̀bá} \tbPS{n.~} \tbGE{the act of pestering} \tbVA{zɔbɔ}
\tbLX{zɔbɔ} \tbHM{} \tbPH{zɔ̀bɔ́} \tbPS{n.~} \tbGE{the act of pestering} \tbVA{zɔba}
\tbLX{zɔla} \tbHM{} \tbPH{zɔ̀lá} \tbPS{n.~} \tbGE{a type of guinea corn} \tbVA{zɔlɔ}
\tbLX{zɔle} \tbHM{} \tbPH{zɔ́lɪ̀} \tbPS{n.~} \tbGE{fool} \tbOP{zɔlɔ} \tbTP{zɔlɛɛ} 
\tbLX{zɔle} \tbHM{} \tbPH{zɔ́lɪ́} \tbPS{v.~} \tbGE{to be related to people} \tbPD{zɔlɛɛ, zɔlɔ, zɔlɔ, zɔleba, zɔlaa, - } 
\tbLX{zɔlɔ} \tbHM{} \tbPH{zɔ̀lɔ́} \tbPS{n.~} \tbGE{a type of guinea corn} \tbTP{zɔllɛɛ} \tbVA{zɔla}
\tbLX{zɔmpɔrɔŋ} \tbHM{} \tbPH{zɔ́mpɔ́rɔ́ŋ} \tbPS{adv.~} \tbGE{of considerable size, state} 
\tbLX{zɔnne} \tbHM{} \tbPH{zɔ̀nnɪ́} \tbPS{n.~} \tbGE{blindness} 
\tbLX{zɔnnoŋ} \tbHM{} \tbPH{zɔ̀nnʊ́ŋ} \tbPS{n.~} \tbGE{blindness} 
\tbLX{zɔnzɔrema} \tbHM{} \tbPH{zɔ̀nzɔ́rɪ́má} \tbPS{n.~} \tbGE{cobweb, dirt hanging on walls and ceiling} \tbVA{zɔnzɔremɔ}
\tbLX{zɔnzɔremɔ} \tbHM{} \tbPH{zɔ̀nzɔ̀rɪ́mɔ́} \tbPS{n.~} \tbGE{cobweb, dirt hanging on walls and ceiling} \tbTP{zɔnzɔremɛɛ} \tbVA{zɔnzɔrema}
\tbLX{zɔŋa} \tbHM{} \tbPH{zɔ̀ŋá} \tbPS{n.~} \tbGE{blind person} \tbVA{zɔŋɔ, zɔŋe}
\tbLX{zɔŋe} \tbHM{} \tbPH{zɔ́ŋ} \tbPS{n.~} \tbSN \tbGE{flour} \tbSN \tbGE{pulp} \tbTP{zɔnnɛɛ} 
\tbLX{zɔŋe} \tbHM{} \tbPH{zɔ̀ŋ} \tbPS{v.~} \tbGE{to be blind} \tbPD{zɔŋɛɛ, zɔnnɔ, zɔnnɔ, zɔnemɔ, zɔnnaa, zɔnemɔ } 
\tbLX{zɔŋe} \tbHM{} \tbPH{zɔ̌ŋ} \tbPS{n.~} \tbGE{flour} \tbOP{zɔnne} \tbTP{zɔnnɛɛ} \tbVA{zɔŋɔ}
\tbLX{zɔŋe} \tbHM{} \tbPH{zɔ̀ŋɪ́} \tbPS{n.~} \tbGE{blind person} \tbVA{zɔŋɔ, zɔŋa}
\tbLX{zɔŋko᷉ɔ᷉} \tbHM{} \tbPH{zɔ̌ŋkʊ́᷉ɔ́} \tbPS{n.~} \tbSN \tbGE{food for workers at work} \tbSN \tbGE{water mixed with flour} \tbOP{zɔŋkoɔnɛɛ} 
\tbLX{zɔŋkɔŋ-ɔnnɔ} \tbHM{} \tbPH{zɔ̌ŋkɔ́ŋ-ɔ́nnɔ́} \tbPS{n.~} \tbGE{one who cooks workers' food} \tbOP{zɔŋkɔŋ-ɔnemɔ} \tbTP{zɔŋkɔŋ-ɔnemɛɛ.} 
\tbLX{zɔŋɔ} \tbHM{} \tbPH{zɔ̀ŋɔ́} \tbPS{n.~} \tbGE{blind person} \tbOP{zɔnne} \tbTP{zɔnnɛɛ} \tbVA{zɔŋa, zɔŋe}
\tbLX{zɔɔ} \tbHM{} \tbPH{zɔ̀ɔ́} \tbPS{n.~} \tbGE{ambush} \tbOP{zɔɔre} \tbTP{zɔɔrɛɛ} 
\tbLX{zɔɔ} \tbHM{} \tbPH{zɔ̀ɔ̀} \tbPS{adj.~} \tbGE{long and flowing e.g. hair} 
\tbLX{zɔɔ} \tbHM{} \tbPH{zɔ̀ɔ̀} \tbPS{v.~} \tbGE{to fight, to quarrel} \tbPD{zɔɔɛ, zɔɔrɔ, zɔɔrɔ, zɔɔrebɔ/zɔɔreba, -, - } 
\tbLX{zɔ᷉ɔ} \tbHM{} \tbPH{zɔ̀᷉ɔ̀᷉} \tbPS{v.~} \tbGE{to ride} \tbPD{zɔ᷉ɔɛ, zɔɔnɔ, zɔɔnɔ, zɔɔnemɔ, zɔɔnaa, - } 
\tbLX{zɔɔ-bɔ} \tbHM{} \tbPH{zɔ̀ɔ̀-bɔ́} \tbPS{n.~} \tbGE{the act of irritating others} 
\tbLX{zɔɔre} \tbHM{} \tbPH{zɔ̀ɔ̀rɪ́} \tbPS{n.~} \tbGE{quarrel, fight} 
\tbLX{zɔɔzeele} \tbHM{} \tbPH{zɔ̀ɔ̀zɪ́ɪ́lɪ́} \tbPS{n.~} \tbGE{silky flower on a corncob} \tbSG{zɔɔzeeloo} \tbOP{zɔɔzeelɛ} \tbTP{zɔɔzeelɛɛ} 
\tbLX{zɔzɔ} \tbHM{} \tbPH{zɔ̀zɔ̀} \tbPS{adv.~} \tbGE{reduced  to minimal (rainfall)} 
\tbLX{zu} \tbHM{} \tbPH{zû} \tbPS{n.~} \tbGE{head} \tbOP{zuri} \tbTP{zuree} 
\tbLX{zu} \tbHM{} \tbPH{zú} \tbPS{v.~} \tbGE{to steal} \tbPD{zuee, zuuro, -, -, zuraa, - } 
\tbLX{zu} \tbHM{} \tbPH{zû} \tbPS{adv.~} \tbGE{on top of} \tbXV{A gane dɔgele la a tabol zu.} \tbXE{The book is on the table.} 
\tbLX{zubɛroŋ} \tbHM{} \tbPH{zúbɛ́rʊ́ŋ} \tbPS{n.~} \tbSN \tbGE{strong and dangerous animal} \tbSN \tbGE{leader (often political)} \tbOP{zubɛrɛ} \tbTP{zubɛrɛɛ} 
\tbLX{zubie} \tbHM{} \tbPH{zúbíé} \tbPS{n.~} \tbGE{misfortune, ill luck} \tbTP{zubiiree} 
\tbLX{zudaa} \tbHM{} \tbPH{zúdàá} \tbPS{n.~} \tbGE{backbone; that which one relies on} \tbOP{zudaare} \tbTP{zudaarɛɛ} 
\tbLX{zufaa} \tbHM{} \tbPH{zúfáá} \tbPS{n.~} \tbGE{ill fate} \tbOP{zufaare} \tbTP{zufaarɛɛ.} 
\tbLX{zugane} \tbHM{} \tbPH{zúgánɪ́} \tbPS{n.~} \tbGE{skin of the head} \tbOP{zugama} \tbTP{zugamɛɛ} 
\tbLX{zugi} \tbHM{} \tbPH{zùgì} \tbPS{v.~} \tbGE{to turn a container upside down to bring out all the contents} \tbPD{zugee, zugiro, zugiro, zugiribo, zugiraa, - } 
\tbLX{zugi} \tbHM{} \tbPH{zùgì} \tbPS{v.~} \tbGE{to press the bellows of a furnace} \tbPD{zugee, zugiro, zugiro, zugiribo, zugiraa, - } 
\tbLX{zugilaa} \tbHM{} \tbPH{zùgìláá} \tbPS{n.~} \tbGE{bellows} \tbOP{zugilaare} \tbTP{zugilaarɛɛ} 
\tbLX{zugiloŋ} \tbHM{} \tbPH{zúglóŋ} \tbPS{adj.~} \tbGE{unproportioned} 
\tbLX{zugiraa} \tbHM{} \tbPH{zúgráa} \tbPS{n.~} \tbGE{blacksmith's bellows} \tbOP{zugiri} \tbTP{zugiree} 
\tbLX{zugiri} \tbHM{} \tbPH{zúgrì} \tbPS{v.~} \tbGE{to shake in order to know whether there something inside e.g. closed tin} \tbPD{zugiree, zugiro, zugiro, zugiribo, zugiraa, - } 
\tbLX{zugizaga} \tbHM{} \tbPH{zùgzàgà} \tbPS{adj.~} \tbGE{huge and heavy e.g. mallet} 
\tbLX{zugi zugi} \tbHM{} \tbPH{zùgzùg} \tbPS{adv.~} \tbGE{moving through in numbers e.g. spectators} \tbXV{Ba ŋmɛ la a bɔle baare ka a noba ta zugi zugi zugi ba naŋ yire.} \tbXE{The football match is over and the spectators are trouping out.} 
\tbLX{zugboli} \tbHM{} \tbPH{zúgbòlí} \tbPS{n.~} \tbSN \tbGE{a shaven head} \tbSN \tbGE{a head that is carrying something without a pad} \tbVA{zukpoli}
\tbLX{zuiŋ} \tbHM{} \tbPH{zúíŋ} \tbPS{conj.~} \tbSN \tbGE{because, as a result} \tbXV{A bie baaloŋ zuiŋ o ma ba gaa toma.} \tbXE{As a result of the child's illness, the mother didn't go to work.} \tbSN \tbGE{so} \tbXV{A saama na waa la yaga a zuiŋ maale a bondirii yaga.} \tbXE{There will be many guests, so cook plenty of food.} 
\tbLX{zukoɔre} \tbHM{} \tbPH{zúkʊ́ɔ́rɪ́} \tbPS{n.~} \tbGE{skull} \tbOP{zukɔɛ} \tbTP{zukɔrɛɛ} \tbVA{zuŋmane}
\tbLX{zukɔgeraa} \tbHM{} \tbPH{zúkɔ́gráá} \tbPS{n.~} \tbGE{pillow} \tbOP{zukɔgere} \tbTP{zukɔgerɛɛ} 
\tbLX{zukɔɔloŋ} \tbHM{} \tbPH{zúkɔ́ɔ́lʊ́ŋ} \tbPS{n.~} \tbGE{hair} \tbOP{zukɔɔlɔ} \tbTP{zukɔɔlɛɛ} 
\tbLX{zukuri} \tbHM{} \tbPH{zùkúrí} \tbPS{n.~} \tbGE{back (body)} \tbOP{zukurri} \tbTP{zukuree} 
\tbLX{zukpala} \tbHM{} \tbPH{zúkpálà} \tbPS{n.~} \tbGE{bald head} \tbTP{zukpallɛɛ} 
\tbLX{zukpeɛne} \tbHM{} \tbPH{zúkpɪ́ɛ́nɪ́} \tbPS{n.~} \tbGE{hardheadedness, stubbornness} 
\tbLX{zukpeɛŋaa} \tbHM{} \tbPH{zùkpɪ́ɛ́ŋáá} \tbPS{adj.~} \tbGE{hardheaded} \tbOP{zukpeɛne} 
\tbLX{zukpoli} \tbHM{} \tbPH{zúkpólí} \tbPS{n.~} \tbSN \tbGE{a shaven head} \tbSN \tbGE{a head that is carrying something without a pad} \tbOP{zukpolo} \tbTP{zukpolee} \tbVA{zugboli}
\tbLX{zukyuu} \tbHM{} \tbPH{zúʧùú} \tbPS{n.~} \tbGE{ritual hair} \tbOP{zukyuuri} \tbTP{zukyuuree} 
\tbLX{zulee} \tbHM{} \tbPH{zùlèè} \tbPS{adv.~} \tbGE{large at one end} 
\tbLX{zuleɛrɛ} \tbHM{} \tbPH{zúlɪ́ɛ́rɛ́} \tbPS{n.~} \tbGE{successor} \tbOP{zuleɛrebɛ} \tbTP{zuleɛrebɛɛ} 
\tbLX{zuli} \tbHM{} \tbPH{zùlì} \tbPS{v.~} \tbSN \tbGE{to be unbalanced, to lean to one side} \tbSN \tbGE{to slope} \tbPD{zulee, zullo, zullo, zulbo, zullaa, - } 
\tbLX{zulibala} \tbHM{} \tbPH{zùlìbàlà} \tbPS{adj.~} \tbGE{disproportionately big} 
\tbLX{zulimo} \tbHM{} \tbPH{zúlímó} \tbPS{adj.~} \tbSN \tbGE{uneven, sloped} \tbSN \tbGE{unbalanced (weight)} 
\tbLX{zuluŋ} \tbHM{} \tbPH{zùlûŋ} \tbPS{adj.~} \tbGE{deep} \tbOP{zulli} \tbTP{zullee} 
\tbLX{zuluŋ} \tbHM{} \tbPH{zùlǔŋ} \tbPS{n.~} \tbGE{army ant, soldier ant} \tbTP{zullee} 
\tbLX{zu-maala} \tbHM{} \tbPH{zú-máálá} \tbPS{n.~} \tbGE{hairdresser} \tbOP{zu-maaleba} 
\tbLX{zumaaroŋ} \tbHM{} \tbPH{zúmáárʊ̌ŋ} \tbPS{n.~} \tbGE{headache (concentrated around the forehead)} \tbOP{zumaara} \tbTP{zumaarɛɛ} 
\tbLX{zunoɔ} \tbHM{} \tbPH{zúnʊ̀ɔ́} \tbPS{n.~} \tbGE{fortune, favour, luck} \tbOP{zunoore} \tbTP{zunoorɛɛ} 
\tbLX{zunzugiri} \tbHM{} \tbPH{zùnzùgrí} \tbPS{n.~} \tbGE{large isolated cloud} \tbSG{zunzugiruu} \tbOP{zunzugo} \tbTP{zunzugiree} 
\tbLX{zuŋ} \tbHM{} \tbPH{zǔŋ} \tbPS{adv.~} \tbGE{on top of} \tbVA{shortened form of zu poɔ}
\tbLX{zuŋmaara} \tbHM{} \tbPH{zúŋmáárà} \tbPS{n.~} \tbGE{executioner} \tbOP{zuŋmaareba} 
\tbLX{zuŋmane} \tbHM{} \tbPH{zúŋmánɪ́} \tbPS{n.~} \tbGE{skull} \tbOP{zuŋmama} \tbTP{zuŋmamɛɛ} \tbVA{zukoɔre}
\tbLX{zuŋmanekoɔ} \tbHM{} \tbPH{zúŋmánɪ́kʊ̀ɔ́} \tbPS{n.~} \tbGE{suffering} 
\tbLX{zuŋmane ko᷉ɔ} \tbHM{} \tbPH{zúŋmánɪ́ kʊ́᷉ɔ́᷉} \tbPS{n.~} \tbGE{great suffering (typically used with nyu)} \tbXV{Deyɛŋ boŋkoɔre ba maale ka nembaaleba da nyu ba zuŋmane ko᷉ɔ.} \tbXE{Last year, there was a poor crop yield and poor people suffered greatly.} 
\tbLX{zuo} \tbHM{} \tbPH{zùò} \tbPS{v.~} \tbSN \tbGE{to be more than} \tbSN \tbGE{to be too much} \tbSN \tbGE{to pass someone by} \tbPD{-, -, -, -, -, - } 
\tbLX{zu-ɔɔ} \tbHM{} \tbPH{zú-ɔ̀ɔ̀} \tbPS{n.~} \tbGE{the condition of having a headache} \tbOP{zu-ɔɔre} 
\tbLX{zupili} \tbHM{} \tbPH{zúpílí} \tbPS{n.~} \tbGE{hat, cap} \tbSG{zupiluu} \tbOP{zupile} \tbTP{zuplee} 
\tbLX{zupiltobo} \tbHM{} \tbPH{zúpìlìtòbó} \tbPS{n.~} \tbGE{a hat that covers the ears} \tbTP{zupiltobee} 
\tbLX{zupoli} \tbHM{} \tbPH{zúpólí} \tbPS{n.~} \tbGE{sinciput (the front of the skull)} \tbSG{zupollee} \tbOP{zupolo} \tbTP{zupolli} 
\tbLX{zupono} \tbHM{} \tbPH{zú-pónó} \tbPS{n.~} \tbGE{barber} \tbXV{zu-ponimo} \tbXE{} 
\tbLX{zu pozu} \tbHM{} \tbPH{zú pòzú} \tbPS{v.~} \tbGE{to gossip} \tbPD{-, -, -, -, -, - } 
\tbLX{zusa᷉a} \tbHM{} \tbPH{zúsà᷉à᷉} \tbPS{n.~} \tbGE{immorality} 
\tbLX{zusãa} \tbHM{} \tbPH{zúsã̀ã̀} \tbPS{n.~} \tbGE{the practice of vice} 
\tbLX{zusaana} \tbHM{} \tbPH{zúsááná} \tbPS{n.~} \tbGE{one who preaches vice} 
\tbLX{zusaanaa} \tbHM{} \tbPH{zúsã́ánáá} \tbPS{n.~} \tbGE{the act of engaging in evil practices} 
\tbLX{zusoga} \tbHM{} \tbPH{zúsʋ́gá} \tbPS{adv.~} \tbGE{above, high up} \tbVA{zusogɔ}
\tbLX{zusogɔ} \tbHM{} \tbPH{zúsʊ́gá} \tbPS{adv.~} \tbGE{above, high up} \tbVA{zusoga}
\tbLX{zutage} \tbHM{} \tbPH{zú-tàgɪ́} \tbPS{n.~} \tbGE{hardheaded} \tbOP{zutagere} 
\tbLX{zutalaa} \tbHM{} \tbPH{zútálàà} \tbPS{n.~} \tbGE{headdress} \tbOP{zutalle} \tbTP{zutallɛɛ} 
\tbLX{zutuo} \tbHM{} \tbPH{zútúó} \tbPS{n.~} \tbGE{misfortune} \tbOP{zutuuri} \tbTP{zutuuree} 
\tbLX{zuu} \tbHM{} \tbPH{zùù} \tbPS{adv.~} \tbGE{huge and heavy} 
\tbLX{zuuni} \tbHM{} \tbPH{zúúní} \tbPS{v.~} \tbGE{to stoop} \tbPD{zuunee, zuuno, zuuno, -, -, - } 
\tbLX{zuyɔɔ} \tbHM{} \tbPH{zúyɔ́ɔ́} \tbPS{n.~} \tbGE{head tax} \tbTP{zuyɔɔrɛɛ} 
\tbLX{zuzeɛ} \tbHM{} \tbPH{zúzɪ̀ɛ́} \tbPS{n.~} \tbGE{leader, forerunner} \tbOP{zuzeere} \tbTP{zuzeerɛɛ} 
\tbLX{zuzeɛ} \tbHM{} \tbPH{zúzɪ̀ɛ́} \tbPS{n.~} \tbGE{red-haired person} \tbOP{zuzeere} \tbTP{zuzeerɛɛ} 
\tbLX{zu zɔŋe} \tbHM{} \tbPH{zú zɔ̂ŋ} \tbPS{v.~} \tbGE{to be ashamed} \tbPD{zue zɔŋe, -, -, -, -, - } 
\tbLX{zuzuu} \tbHM{} \tbPH{zùzúú} \tbPS{adj.~} \tbGE{monstrous} 
\tbLX{zzzzzzz} \tbHM{} \tbPH{ajsdf} \tbPS{n.~} \tbGE{hopefully works} 
